### 2013-09-19 paper/paper.tex
Added first version of paper

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass[a4paper]{llncs}
%

\usepackage{listings}
\usepackage{pgf}
\usepackage{tikz}
\usetikzlibrary{arrows, automata, shapes}

%\usepackage{makeidx}  % allows for indexgeneration
%
\title{{\sc Kalashnikov}: Fully Automatic Loop Free Program Synthesis by Bounded Model Checking}
\author{Matt Lewis}
\institute{Oxford University}

\newenvironment{keywords}{
       \list{}{\advance\topsep by0.35cm\relax\small
       \leftmargin=0cm
       \labelwidth=0.35cm
       \listparindent=0.35cm
       \itemindent\listparindent
       \rightmargin\leftmargin}\item[\hskip\labelsep
                                     \bfseries Keywords:]}
     {\endlist}

\begin{document}
%
\maketitle
%
\pagestyle{headings}  % switches on printing of running heads

\abstract
We present a method for synthesizing loop free programs using a combination of
bounded model checking and explicit enumeration.  The user provides a specification
in plain C and is not required to interact further with the synthesis processs.
In particular the user need not tell the synthesizer which program components to
use.

We demonstrate the effectiveness of our method by synthesizing several tricky
bitvector programs and a selection of floating point programs.

\begin{keywords}
 Program synthesis, bitvectors, bounded model checking, CBMC,
 floating point.
\end{keywords}

\section{Introduction}

\section{Related Work}

\section{Basic Synthesis Algorithm}

\subsection{The Abstract Algorithm}

Our task is to find a program which satisfies some specifcation.  We can formalize this
notion as follows: fix an input set $I$ and an output set $O$.  Specifications $\sigma$
are then relations and programs $P$ are computable functions:

$$ \sigma \subseteq I \times O $$
$$ P : I \rightarrow O$$

The synthesis problem is then to determine the validity of the following formula, and
to find a witness $P$ if the following second-order formula is valid:

$$\exists P . \forall x \in I . \sigma(x, P(x))$$

Depending on the logic needed to express this formula the synthesis problem may
be decidable, semi-decidable or undecidable.  For many interesting logics, checking
the validity of a second order formula, such as the synthesis formula, is undecidable.
Even for logics in which the problem is decidable, the quantifier alternation means that
we do not have efficient decision procedures to verify our formula, although efficient
methods may exist for solving problems with only one quantifier.  In particular
if we are working in the logic of propositional satisfiability, we have SAT solvers
which are efficient-in-practice at solving queries with a single quantifier.
The corresponding problem with one level of quantifier alternation (2-QBF) is
complete for $NP^{NP}$, and current 2-QBF solvers are much less efficient than SAT solvers.

When the first-order fragment of the logic is efficiently decidable and $P$ can be
expressed as a ground term of the logic, we can use a refinement loop to check the
validity of the synthesis formula as shown in Fig.~\ref{fig:abstract-refinement} and
Fig.~\ref{fig:abstract-refinement-code}.  The algorithm is divided into two
components: {\sc synth} and {\sc verif}.  We track a finite (small) set of inputs
and {\sc synth} synthesizes a candidate program which is correct on just those inputs.
{\sc Verif} then tries to verify the candidate program by searching for an input on
which the candidate program does not satisfy the specification.  If no such input
can be found, the candidate program is correct.  Otherwise, the new input is added
to the set of inputs we must check in {\sc synth}.  If the input set is finite, this
procedure is guaranteed to terminate.

\begin{figure}
 \centering
 \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,
 semithick, initial text=]

  \matrix[nodes={draw, fill=none, scale=1, shape=rectangle, minimum height=1cm, minimum width=1.5cm},
          row sep=2cm, column sep=4cm] {
   \node (synth) {Synthesize};
   &
   \node (verif) {Verify}; \\
   \node[draw=none] {};
   &
   \node[ellipse] (done) {Done}; \\
  };

   \path
    (synth) edge [bend left] node {Candidate program} (verif)
    (verif) edge [bend left] node {Counterexample input} (synth)
    (verif) edge node {Valid} (done);
 \end{tikzpicture}

 \label{fig:abstract-refinement}
 \caption{Abstract synthesis refinement loop}
\end{figure}

\begin{figure}
\centering
\begin{minipage}{0.9\textwidth}
 \begin{lstlisting}[mathescape,language=C]
function synth(inputs) {
  $(i_1, \ldots, i_N)$ = inputs
  query := $\exists P . \sigma(i_1, P(i_1)) \land \ldots \land \sigma(i_N, P(i_N))$
  result := decide(query)

  if (result.satisfiable) {
    return result.model
  } else {
    return unsatisfiable
  }
}

function verif(P) {
  query := $\exists x . \lnot \sigma(x, P(x))$
  result := decide(query)

  if (result.satisfiable) {
    return result.model
  } else {
    return valid
  }
}

function refinement_loop() {
  inputs := $\emptyset$

  while (true) {
    candidate := synth(inputs)

    if (candidate = unsatisfiable) {
      return unsatisfiable
    }

    res := verif(candidate)

    if (res = valid) {
      return candidate
    } else {
      inputs := inputs $\cup$ res
    }
  }
 \end{lstlisting}
 \end{minipage}

 \label{fig:abstract-refinement-code}
 \caption{Abstract refinement algorithm}
\end{figure}

\subsection{The Concrete Algorithm for Bitvector Programs}

One area in which program synthesis can shine is in producing very small,
intricate programs that manipulate bitvectors.  An example of such a program
is shown in Fig.~\ref{fig:bitvector-program}.  This program takes a machine word
as input and clears every bit except for the least significant bit that was set.
Even though this program is extremely short, it is fairly difficult for a human
to see what it does.  It is even more difficult for a human to come up with such
minimal code -- a natural solution for this problem might be to use a loop which
iterates over all of the bits in the word.  The program is so concise because it
takes advantage of the low level details of the machine, such as the fact that
signed integers are stored in two's complement form.

\begin{figure}
\centering
\begin{minipage}{0.45\linewidth}
 \begin{lstlisting}[language=C]
int isolate_lsb(int x) {
  return x & -x;
}
 \end{lstlisting}
\end{minipage}
\begin{minipage}{0.45\linewidth}

Example:

\hrule

\begin{tabular}{llcccccccc}
 x       & = & 1 & 0 & 1 & 1 & 1 & 0 & 1 & 0 \\
 -x      & = & 0 & 1 & 0 & 0 & 0 & 1 & 1 & 0 \\
 x \& -x & = & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 0
\end{tabular}
\end{minipage}

 \label{fig:bitvector-program}
 \caption{A tricky bitvector program}
\end{figure}

To synthesize tricky bitvector programs like this, it is natural for us to work
in the logic of quantifer free propositional formulae and to use a SAT solver as
the decision procedure.  However, we propose a slightly different tack, which is
to use a decidable fragment of C as our underlying logic.  The fragment we use
has the following restrictions:

\begin{itemize}
 \item All loops and recursive function calls must terminate after a constant number
 of iterations.  This constant must be statically inferrable.
 \item Arrays must all be statically allocated with a constant size.
\end{itemize}

Other than this any C constructs can be used, which still leaves a very expressive
language with the nice property that safety is decidable using a single query to a
bounded model checker.  We will refer to this fragment as C-.  To instantiate the
abstract synthesis algorithm in C- we must express $I, O, \sigma$ and $P$ in C-,
then ensure that we can express the validity of the synthesis formula as a safety
property of the resulting C- program.

Our encoding is the following:

\begin{itemize}
 \item $I$ is the type \verb|int[N]|.
 \item $O$ is the type \verb|int[M]|.
 \item $\sigma$ is a function with signature
 \verb|int check(int in[N], int out[M])|. This function is the only component supplied
 by the user.
 \item $P$ is written in a simple RISC language $\mathcal{L}$.  Programs in $\mathcal{L}$
 have the type \verb|prog_t|.
 \item We supply an interpreter for $\mathcal{L}$ which is written in C-.  The signature
 of this interpreter is \\
 \verb|void exec(prog_t p, int in[N], int out[M])|.
\end{itemize}

\begin{figure}
\centering
\begin{minipage}{0.65\textwidth}
\setlength{\tabcolsep}{16pt}
Integer arithmetic instructions:

\begin{tabular}{lll}
 \verb|add a b| & \verb|sub a b| & \verb|mul a b| \\
 \verb|div a b| & \verb|neg a|
\end{tabular}
\bigskip

Bitwise logical and shift instructions:

\begin{tabular}{lll}
 \verb|and  a b| & \verb|or   a b| & \verb|xor  a b| \\
 \verb|ashr a b| & \verb|lshr a b| & \verb|not  a|
\end{tabular}
\bigskip

Unsigned and signed comparison instructions:

\begin{tabular}{lll}
 \verb|le  a b| & \verb|lt  a b| & \verb|sle  a b| \\
 \verb|slt a b|
\end{tabular}
\end{minipage}

 \label{fig:l-language}
 \caption{The language $\mathcal{L}$}
\end{figure}

We must now express the {\sc synth} and {\sc verif} formulae as safety properties
of C- programs, which is shown in Fig.~\ref{fig:c-synthverif}.

\begin{figure}
\centering
\begin{minipage}{0.45\textwidth}
\begin{lstlisting}[language=C]
void synth() {
  prog_t p = nondet();
  int in[N], out[M];

  in = test1;
  exec(p, in, out);
  assume(check(in, out));

  ...

  in = testN;
  exec(p, in, out);
  assume(check(in, out));

  assert(0);
}
\end{lstlisting}
\end{minipage}
\begin{minipage}{0.45\textwidth}
\begin{lstlisting}[language=C]
void verif(prog_t p) {
  int in[N] = nondet();
  int out[M];

  exec(p, in, out);
  assert(check(in, out));
}
\end{lstlisting}
\end{minipage}

 \label{fig:c-synthverif}
 \caption{The {\sc synth} and {\sc verif} formulae expressed as a C- program}
\end{figure}

\section{Optimizations}

\subsection{Successful}
\subsection{Unsuccessful}

\section{Floating Point}

\section{Experimental Results}

\section{Conclusion}

\end{document}

### 2013-09-20 paper/paper.tex
Wrote about optimizations

@@
 \begin{minipage}{0.45\textwidth}
-\begin{lstlisting}[language=C]
+\begin{lstlisting}[language=C++]
 void synth() {
@@
 
-  assert(0);
+  assert(false);
 }
@@
 
-\subsection{Successful}
-\subsection{Unsuccessful}
+By far the dominant factor in the total runtime of the synthesis algorithm is
+the runtime of {\sc synth}.  It is not entirely surprising that this runtime
+is strongly related to the size of the SAT problem generated by CBMC.  The size
+of the SAT problem is influenced by the number of bits needed to encode $P$,
+and by the number of bits needed to encode a run of the $\mathcal{L}$-interpreter.
+The connection to Kolmogorov complexity is fairly obvious.
+
+The most successful optimizations we found were related to making fragments of
+$\mathcal{L}$ and searching progressively larger fragments until a program
+could be found.
+
+\subsection{Limited Program Length}
+The simplest restriction on programs is to only consider programs of a certain length.
+We start by considering programs consisting of one instruction and progressively
+increase this number.
+
+\subsection{Word Width}
+It is often the case that a program which satisfies the specification
+will continue to satisfy it regardless of the word width of the machine it is
+impemented on.  For example the program in Fig.~\ref{fig:bitvector-program}
+isolates the least significant bit of an 8-bit word, but it will also
+isolate the least significant bit of a 32-bit or indeed any sized word.
+
+Because of this, we can often find a program that is correct for a large
+word size by synthesizing a program for a machine with a smaller word size.
+This reduces the size of all the constants in the program, the search space
+of possible test inputs and also the size of all the state variables internal
+to the $\mathcal{L}$-interpreter.
+
+The only wrinkle here is that sometimes a program we synthesize will contain
+constants.  If we have synthesized a program for a machine with $k$-bit words,
+the constants in the program will be $k$ bits wide.  To generalize the program
+to a $n$-bit machine (with $n > k$), we need some way of deriving $n$-bit
+wide numbers from $k$-bit ones.  We have several strategies for this and
+just try each in turn.  To generalize a number $x$ we use the following
+rules:
+
+\begin{itemize}
+ \item If $x = k_k$, generalize to $n_n$.
+ \item If $x = k-1_k$, generalize to $n-1_n$.
+ \item If $x = k+1_k$, generalize to $n+1_n$.
+ \item Generalize $x_k$ to $x_n$.
+ \item Generalize $x_k$ to $x_k \cdot 0_{n-k}$.
+ \item Generalize $x_k$ to $x_k^{n / k}$.
+\end{itemize}
+
+\subsection{Separate Constants}
+$\mathcal{L}$ is a SSA, three address instruction set.  Destination registers
+are implicit and a fresh register exists for each instruction to write its
+output to.  A natural way to encode $\mathcal{L}$ instructions is to have an
+opcode and two operands.  The opcode selects which instruction type is being
+executed and the operands specify which registers should be operated on.
+We might allow the operands to be either a register (i.e. a program argument
+or the result of a previous instruction), or an immediate constant.
+
+Each opcode requires $\log_2 I$ bits to encode, where $I$ is the number
+of instruction types in $\mathcal{L}$.  Each operand can be encoded using
+$\log_2 w$ bits, where $w$ is the $\mathcal{L}$-machine word, plus one
+bit to specify whether the operand is a register name or an immediate constant.
+One instruction can therefore be encoded using $\log_2 I + 2w + 2$ bits.
+For an $n$ instruction program, we need $$n \log_2 I + 2nw + 2n$$ bits to encode
+the entire program.
+
+If we instead limit the number of constants that can appear in the program,
+our operands can be encoded using fewer bits.  For an $n$ instruction program
+using $k$ constants and taking $a$ arguments as inputs, each operand can refer
+to a program argument, the result of a previous instruction or a constant.
+This can be encoded using $\log_2 (k+a+n-1)$ bits, which means each instruction
+can be encoded in $\log_2 I + \log_2 (k + a + n - 1)$ and the full program
+needs $$n \log_2 I + n \log_2 (k + a + n - 1) + kw$$ bits to encode.
+
+To give an example, $\mathcal{L}$ has 15 instruction types, so each opcode is 4 bits.
+For a 10 instruction program over 1 argument, using 2 constants on a 32-bit word
+machine the first encoding requires $10 * (4 + 32 + 1 + 32 + 1) = 700$ bits.
+Using the second encoding, each operand can be represented using
+$\log_2 (2 + 1 + 10 - 1) = 4$ bits, and the entire program requires 184 bits.
+This is a substantial reduction in size and when the required program requires
+few constants this can lead to a very significat speed up.
+
+As with program length we can progressively increase the number of constants in
+our program.  We start by trying to synthesize a program with no constants, then
+if that fails we try to synthesize using one constant and so on.
+
+\subsection{Explicit Search}
+When we are working on an $\mathcal{L}$-machine with a small word size and
+a compact program encoding, our state space is very small.
+We can quickly check the validity of the {\sc synth} and {\sc verif} formulae
+by simply enumerating all of the inputs and programs, then explicitly
+executing each program on each input.
+
+Since everything in our system is written in C, we can add a simple loop
+to enumerate all programs and inputs, then compile the code using {\sc gcc}.
+The resulting binary will explicitly search for programs satisfying the {\sc synth}
+formula, or for counterexamples satisfying the {\sc verif} formula.
+For small state spaces, these explicit search binaries are often faster
+than symbolically evaluating the {\sc synth} and {\sc verif} formulae with
+{\sc CBMC}.
+
+\subsection{Remove nops}
+Many instructions in $\mathcal{L}$ are nops that do not doing anything.
+For example the instruction \verb|add x 0| does nothing.  Such instructions
+can be removed from any program they appear in to leave a semantically
+equivalent, but shorter, program.  We can therefore be sure that nops
+will never appear in any minimal program.  By adding constraints saying that
+each instruction is not a nop, we can help the underlying SAT solver's
+search, which reduces the runtime of the overall procedure.
+
+\subsection{Symmetry Reduction}
+There are many instructions that are equivalent to each other.  For example,
+\verb|add x y| is equivalent to \verb|add y x| -- any program containing
+one instruction could have it replaced by the other instruction and
+keep the same semantics.  We choose a single canonical instruction to
+represent all instructions in a particular equivalence class, then add
+constraints saying that no non-canonical instructions appear in the program.
+
+Our rules for excluding non-canonical instructions are:
+
+\begin{itemize}
+ \item For commutative operations, the first operand is smaller than the second.
+ \item For unary operations, the second (unused) operand is always 0.
+ \item No instruction may have two constant operands.
+ \item All program constants are distinct.
+\end{itemize}
+
+As with the nop constraints, these additional constraints do increase the
+size of the resulting SAT instance, but this still ends up as a win in
+terms of runtime.
+
+\subsubsection{Implementing $\mathcal{L}$ With A Stack Machine}
+Three-address code tends to be less compact than code written for a stack
+machine.  It therefore would seem reasonable to use a stack machine to
+execute $\mathcal{L}$ programs, since we would expect to see smaller programs
+(in terms of bits used in the encoding), leading to smaller SAT instances
+and faster runtimes overall.
+
+Unfortunately we saw quite the opposite effect -- runtimes when implementing the
+$\mathcal{L}$-interpreter as a stack machine were much higher than the three-address
+code implementation.  We believe the reason for this
+effect is that the internal state of a stack based interpreter is much more complex
+to analyse than an interpreter for three-address code.  This is because a
+three-address code interpreter can be implemented using a single array whose cells are
+each written to exactly once (each cell corresponds to a register in the program).
+By contrast a stack machine interpreter has a stack as its central data structure.
+It is natural to implement this as an array, but the array's contents are accessed
+in an unpredictable, non-uniform manner as the stack pointer increases and decreases
+over the lifetime of the program.  This leads to a \emph{larger} SAT
+isntance and drastically slows down the operation of the solver.
 

### 2013-09-20 paper/paper.tex
Cleaned up description of generalization

@@
 wide numbers from $k$-bit ones.  We have several strategies for this and
-just try each in turn.  To generalize a number $x$ we use the following
-rules:
+just try each in turn.  In the following, a subscript $n$ on a value means that
+the value is an $n$-bit word.  The dot operator $a_l \cdot b_m$ denotes
+concatenation, resulting in an $(l+m)$-bit wide word.  A superscript on a value
+means duplication, so $a_l^m$ means concatenating together $m$ copies of the $l$-bit
+word $a$, resulting in an $(lm)$-bit wide word.
+
+To generalize a number $x$ we use the following rules:
 

### 2013-09-26 specs/p10.c
Change specs to use multi-results

@@ -1,8 +1,9 @@
 #include "synth.h"
 
-int check(word_t args[NARGS], word_t z) {
+int check(word_t args[NARGS], word_t res[NRES]) {
   word_t x = args[0];
   word_t y = args[1];
+  word_t z = res[0];
 
   if ((x ^ y) <= (x & y)) {
     return !!z;

### 2013-09-26 specs/p11.c

@@ -1,10 +1,11 @@
 #include "synth.h"
 
-int check(word_t args[NARGS], word_t z) {
+int check(word_t args[NARGS], word_t res[NRES]) {
   word_t x = args[0];
   word_t y = args[1];
   word_t t1 = ~y;
   word_t t2 = x & t1;
+  word_t z = res[0];
 
   if (t2 > y) {
     return !!z;

### 2013-09-26 specs/p13.c

@@ -1,18 +1,19 @@
 #include "synth.h"
 
-int check(word_t args[NARGS], word_t z) {
+int check(word_t args[NARGS], word_t res[NRES]) {
   word_t x = args[0];
   sword_t q = x;
   sword_t zero = 0;
-  word_t res;
+  word_t expected;
+  word_t z = res[0];
 
   if (q == zero) {
-    res = 0;
+    expected = 0;
   } else if (q < 0) {
-    res = -1;
+    expected = -1;
   } else {
-    res = 1;
+    expected = 1;
   }
 
-  return res == z;
+  return expected == z;
 }

### 2013-09-26 specs/p14.c

@@ -5,12 +5,10 @@ int check(word_t args[NARGS], word_t res[NRES]) {
   word_t y = args[1];
   word_t z = res[0];
 
-  word_t t1 = ~y;
-  word_t t2 = x & t1;
+  word_t t1 = x & y;
+  word_t t2 = x ^ y;
+  word_t t3 = t2 >> 1;
+  word_t t4 = t1 + t3;
 
-  if (t2 <= y) {
-    return !!z;
-  } else {
-    return !z;
-  }
+  return z == t4;
 }

### 2013-09-26 specs/p15.c

@@ -5,12 +5,10 @@ int check(word_t args[NARGS], word_t res[NRES]) {
   word_t y = args[1];
   word_t z = res[0];
 
-  word_t t1 = ~y;
-  word_t t2 = x & t1;
+  word_t t1 = x | y;
+  word_t t2 = x ^ y;
+  word_t t3 = t2 >> 1;
+  word_t t4 = t1 - t3;
 
-  if (t2 <= y) {
-    return !!z;
-  } else {
-    return !z;
-  }
+  return z == t4;
 }

### 2013-09-26 specs/p16.c

@@ -1,8 +1,9 @@
 #include "synth.h"
 
-int check(word_t args[NARGS], word_t z) {
+int check(word_t args[NARGS], word_t res[NRES]) {
   word_t x = args[0];
   word_t y = args[1];
+  word_t z = res[0];
 
   if (x > y) {
     return z == x;

### 2013-09-26 specs/p17.c

@@ -1,8 +1,9 @@
 #include "synth.h"
 
-int check(word_t args[NARGS], word_t z) {
+int check(word_t args[NARGS], word_t res[NRES]) {
   word_t x = args[0];
   word_t q = ((((x-1) | x) + 1) & x);
+  word_t z = res[0];
 
   return q == z;
 }

### 2013-09-26 specs/p18.c

@@ -1,7 +1,8 @@
 #include "synth.h"
 
-int check(word_t args[NARGS], word_t z) {
+int check(word_t args[NARGS], word_t res[NRES]) {
   word_t x = args[0];
+  word_t z = res[0];
 
   for (int i = 0; i < WIDTH; i++) {
     if (x == 0) {

### 2013-09-26 specs/p19.c

@@ -1,0 +1,17 @@
+#include "synth.h"
+
+int check(word_t args[NARGS], word_t res[NRES]) {
+  word_t x = args[0];
+  word_t m = args[1];
+  word_t k = args[2];
+  word_t z = res[0];
+
+  word_t t1 = x >> k;
+  word_t t2 = x ^ t1;
+  word_t t3 = t2 & m;
+  word_t t4 = t3 << k;
+  word_t t5 = t4 ^ t3;
+  word_t t6 = t5 ^ x;
+
+  return t6 == z;
+}

### 2013-09-26 specs/p2.c

@@ -1,7 +1,8 @@
 #include "synth.h"
 
-int check(word_t args[NARGS], word_t z) {
+int check(word_t args[NARGS], word_t res[NRES]) {
   word_t x = args[0];
+  word_t z = res[0];
 
   if (x & x+1) {
     return z;

### 2013-09-26 specs/p20.c

@@ -1,7 +1,8 @@
 #include "synth.h"
 
-int check(word_t args[NARGS], word_t z) {
+int check(word_t args[NARGS], word_t res[NRES]) {
   word_t x = args[0];
+  word_t z = res[0];
 
   sword_t t1 = -x;
   word_t t2 = x & t1;

### 2013-09-26 specs/p21.c

@@ -1,0 +1,20 @@
+#include "synth.h"
+
+int check(word_t args[NARGS], word_t res[NRES]) {
+  word_t x = args[0];
+  word_t a = args[1];
+  word_t b = args[2];
+  word_t c = args[3];
+  word_t z = res[0];
+
+  word_t t1 = x != c;
+  word_t t2 = a ^ c;
+  word_t t3 = x != a;
+  word_t t4 = b ^ c;
+  word_t t5 = t1 & t2;
+  word_t t6 = t3 & t4;
+  word_t t7 = t5 ^ t6;
+  word_t t8 = t7 ^ c;
+
+  return z == t8;
+}

### 2013-09-26 specs/p22.c

@@ -1,9 +1,10 @@
 #include "synth.h"
 
-int check(word_t args[NARGS], word_t z) {
+int check(word_t args[NARGS], word_t res[NRES]) {
   word_t x = args[0];
   word_t res = 0;
   word_t bit;
+  word_t z = res[0];
 
   for (int i = 0; i < WIDTH; i++) {
     bit = x & 1;

### 2013-09-26 specs/p23.c

@@ -1,8 +1,9 @@
 #include "synth.h"
 
-int check(word_t args[NARGS], word_t z) {
+int check(word_t args[NARGS], word_t res[NRES]) {
   word_t x = args[0];
   word_t cnt = 0;
+  word_t z = res[0];
 
   for (int i = 0; i < WIDTH; i++) {
     if (x & 1) {

### 2013-09-26 specs/p24.c

@@ -1,3 +1,21 @@
-int check(unsigned int x, unsigned int z) {
+#include "synth.h"
 
+int check(word_t args[NARGS], word_t res[NRES]) {
+  word_t x = args[0];
+  word_t z = res[0];
+  word_t expected;
+  word_t i;
+
+  if ((x & (x-1)) == 0) {
+    return z == x;
+  }
+
+  for (i = 0; i < WIDTH; i++) {
+    if ((x >> i) == 1) {
+      expected = (1 << (i+1));
+      return z == expected;
+    }
+  }
+
+  return z == 0;
 }

### 2013-09-26 specs/p3.c

@@ -1,7 +1,8 @@
 #include "synth.h"
 
-int check(word_t args[NARGS], word_t z) {
+int check(word_t args[NARGS], word_t res[NRES]) {
   word_t x = args[0];
+  word_t z = res[0];
 
   return z == (-x & x);
 }

### 2013-09-26 specs/p4.c

@@ -1,9 +1,9 @@
 #include "synth.h"
 
-int check(word_t args[NARGS], word_t z) {
+int check(word_t args[NARGS], word_t res[NRES]) {
   word_t x = args[0];
   word_t q = x;
-
+  word_t z = res[0];
   q--;
 
   return (z == (q ^ x));

### 2013-09-26 specs/p5.c

@@ -1,8 +1,9 @@
 #include "synth.h"
 
-int check(word_t args[NARGS], word_t z) {
+int check(word_t args[NARGS], word_t res[NRES]) {
   word_t x = args[0];
   word_t q = x;
+  word_t z = res[0];
 
   q--;
 

### 2013-09-26 specs/p6.c

@@ -1,8 +1,9 @@
 #include "synth.h"
 
-int check(word_t args[NARGS], word_t z) {
+int check(word_t args[NARGS], word_t res[NRES]) {
   word_t x = args[0];
   word_t q = (x+1) | x;
+  word_t z = res[0];
 
   return z == q;
 }

### 2013-09-26 specs/p7.c

@@ -1,8 +1,9 @@
 #include "synth.h"
 
-int check(word_t args[NARGS], word_t z) {
+int check(word_t args[NARGS], word_t res[NRES]) {
   word_t x = args[0];
   word_t q = ((~x) & (x+1));
+  word_t z = res[0];
 
   return z == q;
 }

### 2013-09-26 specs/p8.c

@@ -1,8 +1,9 @@
 #include "synth.h"
 
-int check(word_t args[NARGS], word_t z) {
+int check(word_t args[NARGS], word_t res[NRES]) {
   word_t x = args[0];
   word_t q = ((x-1) & (~x));
+  word_t z = res[0];
 
   return q == z;
 }

### 2013-09-26 specs/p9.c

@@ -1,10 +1,11 @@
 #include "synth.h"
 
-int check(word_t args[NARGS], word_t z) {
+int check(word_t args[NARGS], word_t res[NRES]) {
   word_t x = args[0];
   sword_t q = (sword_t) x;
   sword_t zero = 0;
   word_t ret;
+  word_t z = res[0];
 
 #ifdef SEARCH
   q <<= (32-WIDTH);

### 2013-09-26 paper/paper.tex
Added flow chart to paper

@@
 over the lifetime of the program.  This leads to a \emph{larger} SAT
-isntance and drastically slows down the operation of the solver.
+instance and drastically slows down the operation of the solver.
+
+\begin{figure}
+\begin{tikzpicture}{node distance=2cm, auto}
+\tikzstyle{decision} = [diamond, draw, fill=blue!20, 
+    text width=4.5em, text badly centered, node distance=3cm, inner sep=0pt]
+\tikzstyle{block} = [rectangle, draw, fill=blue!20, 
+    text width=5em, text centered, rounded corners, minimum height=4em]
+\tikzstyle{line} = [draw, -latex']
+\tikzstyle{cloud} = [draw, ellipse,fill=red!20, node distance=3cm,
+    minimum height=2em]
+
+ \node [decision] (synthsucc) {{\sc Synth} succeeds?};
+
+ \node [decision, below of=synthsucc, node distance=2.5cm] (verif) {{\sc Verif} succeeds?};
+ \node [decision, right of=verif] (ck) {$c < k$?};
+
+ \node [block, below of=verif, node distance=2.5cm]   (done) {Done!};
+ \node [decision, left of=done] (verifw) {{\sc Verif} succeeds for small words?};
+
+ \node [block, below of=ck, node distance=2.5cm] (incc) {$c := c+1$};
+ \node [block, right of=incc, node distance=2cm] (inck) {$c := 0$\\ $k := k+1$};
+
+ \node [block, below of=verifw, node distance=2.5cm] (incw) {$w := w+1$};
+ \node [block, left of=incw, node distance=2cm] (iterate) {Loop to {\sc synth}};
+
+ \path [line] (synthsucc) -- node [left] {Yes} (verif);
+ \path [line] (synthsucc) -| node [above, near start] {No} (ck);
+
+ \path [line] (verif) -- node [left] {Yes} (done);
+ \path [line] (verif) -| node [above, near start] {No} (verifw);
+
+ \path [line] (ck) -- node [left] {Yes} (incc);
+ \path [line] (ck) -| node  [above, near start]  {No}(inck);
+
+ \path [line] (verifw) -- node [right] {Yes} (incw);
+ \path [line] (verifw) -| node [above, near start] {No} (iterate);
+\end{tikzpicture}
+
+\label{fig:paramsflow}
+ \caption{Decision tree for increasing parameters of $\mathcal{L}$.}
+\end{figure}
 

### 2013-10-02 paper/paper.tex
Added 2sum algorithm & 2sum timings
Added more about C-, schematic diagrams

@@
      {\endlist}
+
+\makeatletter
+\pgfdeclareshape{datastore}{
+  \inheritsavedanchors[from=rectangle]
+  \inheritanchorborder[from=rectangle]
+  \inheritanchor[from=rectangle]{center}
+  \inheritanchor[from=rectangle]{base}
+  \inheritanchor[from=rectangle]{north}
+  \inheritanchor[from=rectangle]{north east}
+  \inheritanchor[from=rectangle]{east}
+  \inheritanchor[from=rectangle]{south east}
+  \inheritanchor[from=rectangle]{south}
+  \inheritanchor[from=rectangle]{south west}
+  \inheritanchor[from=rectangle]{west}
+  \inheritanchor[from=rectangle]{north west}
+  \backgroundpath{
+    %  store lower right in xa/ya and upper right in xb/yb
+    \southwest \pgf@xa=\pgf@x \pgf@ya=\pgf@y
+    \northeast \pgf@xb=\pgf@x \pgf@yb=\pgf@y
+    \pgfpathmoveto{\pgfpoint{\pgf@xa}{\pgf@ya}}
+    \pgfpathlineto{\pgfpoint{\pgf@xb}{\pgf@ya}}
+    \pgfpathmoveto{\pgfpoint{\pgf@xa}{\pgf@yb}}
+    \pgfpathlineto{\pgfpoint{\pgf@xb}{\pgf@yb}}
+ }
+}
+\makeatother
 
@@
 
+\begin{figure}
+\begin{center}
+\begin{tikzpicture}[
+  scale=0.8,
+  font=\sffamily,
+  every matrix/.style={ampersand replacement=\&,column sep=.5cm,row sep=.75cm},
+  source/.style={draw,thick,rounded corners,fill=yellow!20,inner sep=.3cm},
+  process/.style={draw,thick,circle,fill=blue!20},
+  sink/.style={source,fill=green!20},
+  datastore/.style={draw,very thick,shape=datastore,inner sep=.3cm},
+  dots/.style={gray,scale=2},
+  to/.style={->,>=stealth',shorten >=1pt,semithick,font=\sffamily\footnotesize},
+  every node/.style={align=center, scale=0.8}]
+
+  % Position the nodes using a matrix layout
+  \matrix{
+    \& \node[source] (driver) {driver}; \& \node[source] (user) {user}; \& \\
+    \& \node[datastore] (tests) {\texttt{tests.c}}; \& \node[datastore] (spec) {\texttt{spec.c}}; \& \node[datastore] (interpreter) {\texttt{interpreter.c}}; \\
+    \node[datastore] (cbmc-synth) {\texttt{cbmc/synth.c}}; \& \node[process] (cbmc) {\sc cbmc}; \& \node[process] (gcc) {\sc gcc}; \& \node[datastore] (explicit-synth) {\texttt{explicit/synth.c}}; \\
+    \& \& \node[process] (aout) {\texttt{a.out}}; \& \\
+    \& \node[sink] (cbmc-out) {candidate program}; \& \node[sink] (explicit-out) {candidate program}; \& \\
+  };
+
+  \draw[to] (driver) -- (tests);
+  \draw[to] (user) -- (spec);
+
+  \draw[to] (tests) -- (cbmc);
+  \draw[to] (spec) -- (cbmc);
+  \draw[to] (interpreter) -- (cbmc);
+
+  \draw[to] (tests) -- (gcc);
+  \draw[to] (spec) -- (gcc);
+  \draw[to] (interpreter) -- (gcc);
+
+  \draw[to] (cbmc-synth) -- (cbmc);
+  \draw[to] (explicit-synth) -- (gcc);
+
+  \draw[to] (gcc) -- (aout);
+
+  \draw[to] (cbmc) -- (cbmc-out);
+  \draw[to] (aout) -- (explicit-out);
+\end{tikzpicture}
+\end{center}
+
+\label{fig:synth-dfd}
+\caption{Schematic diagram of {\sc synth}}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\begin{tikzpicture}[
+  scale=0.8,
+  font=\sffamily,
+  every matrix/.style={ampersand replacement=\&,column sep=.5cm,row sep=.75cm},
+  source/.style={draw,thick,rounded corners,fill=yellow!20,inner sep=.3cm},
+  process/.style={draw,thick,circle,fill=blue!20},
+  sink/.style={source,fill=green!20},
+  datastore/.style={draw,very thick,shape=datastore,inner sep=.3cm},
+  dots/.style={gray,scale=2},
+  to/.style={->,>=stealth',shorten >=1pt,semithick,font=\sffamily\footnotesize},
+  every node/.style={align=center, scale=0.8}]
+
+  % Position the nodes using a matrix layout
+  \matrix{
+    \& \node[source] (driver) {driver}; \& \node[source] (user) {user}; \& \\
+    \& \node[datastore] (prog) {\texttt{prog.c}}; \& \node[datastore] (spec) {\texttt{spec.c}}; \& \node[datastore] (interpreter) {\texttt{interpreter.c}}; \\
+    \node[datastore] (cbmc-verif) {\texttt{cbmc/verif.c}}; \& \node[process] (cbmc) {\sc cbmc}; \& \node[process] (gcc) {\sc gcc}; \& \node[datastore] (explicit-verif) {\texttt{explicit/verif.c}}; \\
+    \& \& \node[process] (aout) {\texttt{a.out}}; \& \\
+    \& \node[sink] (cbmc-out) {test vector}; \& \node[sink] (explicit-out) {test vector}; \& \\
+  };
+
+  \draw[to] (driver) -- (prog);
+  \draw[to] (user) -- (spec);
+
+  \draw[to] (prog) -- (cbmc);
+  \draw[to] (spec) -- (cbmc);
+  \draw[to] (interpreter) -- (cbmc);
+
+  \draw[to] (tests) -- (gcc);
+  \draw[to] (spec) -- (gcc);
+  \draw[to] (interpreter) -- (gcc);
+
+  \draw[to] (cbmc-verif) -- (cbmc);
+  \draw[to] (explicit-verif) -- (gcc);
+
+  \draw[to] (gcc) -- (aout);
+
+  \draw[to] (cbmc) -- (cbmc-out);
+  \draw[to] (aout) -- (explicit-out);
+\end{tikzpicture}
+\end{center}
+
+\label{fig:synth-dfd}
+\caption{Schematic diagram of {\sc verif}}
+\end{figure}
+
+\section{C-}
+We use a fragment of C as our underlying logic.  We will refer to this fragment
+as C-.  The characteristic property of C- is that safety can be decided for
+C- programs by using a bounded model checker.  A C- program is just a C program
+with the following syntactic restrictions:
+
+\begin{itemize}
+ \item All loops in the program must have a constant bound.
+ \item All recursion in the program must be limited to a constant depth.
+ \item All arrays must be statically allocated (i.e. not using \texttt{malloc}),
+ and be of constant size.
+\end{itemize}
+
+Additionally, C- programs may use:
+
+\begin{itemize}
+ \item Nondeterministic values.
+ \item Assumptions.
+ \item Arbitrary width types.
+\end{itemize}
+
+So for example the programs in Fig.~\ref{fig:c-} are valid
+C- programs, but the program in Fig.~\ref{fig:not-c-} is not, since it uses
+a non-constant loop bound.
+
+\begin{figure}
+\begin{minipage}[scale=0.8]{0.45\linewidth}
+ \begin{lstlisting}[language=c]
+int count_bits(int x) {
+  int i, ret;
+
+  ret = 0;
+
+  for (i = 0; i < 32; i++) {
+    if (x & (1 << i))
+      ret++;
+  }
+
+  return ret;
+}
+ \end{lstlisting}
+\end{minipage}
+\begin{minipage}{0.54\linewidth}
+ \begin{lstlisting}[language=C]
+int common_factor(int A[10]) {
+  int factor = nondet();
+  int i;
+
+  for (i = 0; i < 10; i++) {
+    assume((A[i] % factor) == 0);
+  }
+
+  assume(factor > 1);
+  return factor;
+}
+
+ \end{lstlisting}
+\end{minipage}
+
+ \label{fig:c-}
+ \caption{Two C- programs.}
+\end{figure}
+
+\begin{figure}
+ \begin{lstlisting}[language=C]
+int fib(int n) {
+  int a, b, c;
+
+  a = b = 1;
+
+  while (n--) {
+    c = a+b;
+    a = b;
+    b = c;
+  }
+
+  return a;
+}
+ \end{lstlisting}
+
+ \label{fig:not-c-}
+ \caption{A C program to compute Fibonacci numbers.}
+\end{figure}
+
+Since each loop is bounded by a constant, and each recursive function call is
+limited to a cosntant depth, a C- program necessarily terminates and in
+fact does so in $O(1)$ time.  If we call the largest loop bound $k$, then
+a bounded model checker with an unrolling bound of $k$ will be a complete
+decision procedure for the safety of the program.  For a C- program of
+size $l$ and with largest loop bound $k$, a bounded model checker will
+create a SAT problem of size $O(lk) = O(l)$.  Conversely, a SAT problem
+of size $s$ can be converted trivially into a loop free C- program
+of size $O(s)$.  The safety problem for C- is therefore NP-complete,
+which means it can be decided fairly efficiently for many practical
+instances.
+
+\subsection{Why use C- instead of SAT?}
+To misquote David Wheeler: ``all problems in computer science can be solved by
+another level of abstraction.''~\cite{wheeler-abstraction}.  C- serves as
+an abstraction of SAT.  We can think of SAT as an assembly language and
+{\sc cbmc} as a compiler for C- to this assembly language.  The benefits
+of using C- as our logic rather than SAT are very similar to the benefits of
+writing code in a high level language rather than assembly language.
+Some of the more significant advantages are:
+
+\begin{itemize}
+ \item The code is more readable and easier to debug.
+ \item The code is much more concise.
+ \item We can take advantage of new optimisations, different model
+ checkers and different backends (SAT, SMT, ACDCL, etc.) for free.
+\end{itemize}
+
+To illustrate some of these points, we were able to implement a first
+version of {\sc kalashnikov} in around 150 lines of code, which took
+around two hours to write.  The resulting code was easy to modify,
+optimise and debug.  The current version, which includes floating point
+support, several optimisations and two model checking algorithms,
+is still less than 1200 lines of code (600 lines of C- and 600 lines
+of Python).
+
+Another benefit of using C- as our logic is that it would be easy to
+automatically extract specifications from C programs, which would be
+useful in the context of superoptimisation.  This is because C- is
+a relatively large fragment of C, so we anticipate that many interesting
+sections of real code will already be C-.
+
 \section{Optimizations}

### 2013-10-04 paper/paper.tex
Changed abstract

@@
 
-\abstract
-We present a method for synthesizing loop free programs using a combination of
-bounded model checking and explicit enumeration.  The user provides a specification
-in plain C and is not required to interact further with the synthesis processs.
-In particular the user need not tell the synthesizer which program components to
-use.
+\begin{abstract}
+We present a method for synthesizing loop-free programs, using a combination of
+bounded model checking and explicit state model checking.  The user provides a
+specification in plain C, after which the synthesis process is fully automatic.
 
-We demonstrate the effectiveness of our method by synthesizing several tricky
-bitvector programs and a selection of floating point programs.
+We demonstrate the effectiveness of our method by synthesizing several subtle
+programs acting on machine integers, and a selection of floating point programs.
+\end{abstract}
 

### 2013-10-04 paper/paper.tex
nudges to abstract

@@
 \begin{abstract}
-We present a method for synthesizing loop-free programs, using a combination of
-bounded model checking and explicit state model checking.  The user provides a
-specification in plain C, after which the synthesis process is fully automatic.
+We present a simple-yet-effective method for synthesizing loop-free
+programs, using a novel combination of Bounded Model Checking and
+explicit-state Model Checking.  The user provides a specification in plain
+C, after which the synthesis process is fully automatic.
 

### 2013-10-04 paper/paper.tex
Changes to paper

--- 2013-10-02 paper/paper.tex
@@
 \usepackage{listings}
+\usepackage{graphicx}
 \usepackage{pgf}
@@
 \section{Related Work}
+The most similar work to ours is {\sc brahma}~\cite{brahma}, which is also a method for
+synthesizing loop-free bitvector programs.  Specifications for the
 
@@
 
-\label{fig:synth-dfd}
+\label{fig:verif-dfd}
 \caption{Schematic diagram of {\sc verif}}
@@
 
+\begin{figure}
+ \begin{lstlisting}[language=c]
+s  = a + b;
+b2 = s - a;
+a2 = s - b2;
+db = b - b2;
+da = a - a2;
+t  = da + db;
+ \end{lstlisting}
+
+\begin{lstlisting}[language=c]
+s = a + b;
+z = s - a;
+t = b - z;
+\end{lstlisting}
+
+ \label{fig:2sum}
+ \caption{2Sum}
+\end{figure}
+
 \section{Experimental Results}
+
+We performed our experiments on a 4-core, 3.07GHz Xeon X5667 with 50GB of RAM.
+
+\begin{figure}
+ \includegraphics[width=\linewidth]{figures/brahma_kalashnikov.pdf}
+ \label{fig:brahma-kalashnikov}
+ \caption{Brahma runtimes vs. Kalashnikov}
+\end{figure}
 

### 2013-10-04 paper/paper.tex
start of intro

@@
 \section{Introduction}
+
+Program synthesis is the mechanized construction of software that provably
+satisfies a given specification.  Synthesis tools promise to relieve the
+programmer from thinking about \emph{how} the problem is to be solved;
+instead, the programmer only provides a compact description of \emph{what}
+is to be achieved.  Foundational research in this area has been
+exceptionally fruitful, beginning with Alonso Church's work on the
+\emph{Circuit Synthesis Problem} in the sixties.  In this paper, we present
+a new method that brings the promise of automating programming closer to
+application.
 

### 2013-10-04 paper/paper.tex
Started related work

@@
 The most similar work to ours is {\sc brahma}~\cite{brahma}, which is also a method for
-synthesizing loop-free bitvector programs.  Specifications for the
+synthesizing loop-free bitvector programs.  In~\cite{brahma} the authors characterise
+the program synthesis problem as a second order $\exists \forall$ formula, and
+described the CEGIS algorithm for solving these formulae.  This algorithm reduces
+the satisfiability of the $\exists \forall$ formula to the satisfiability of several
+first order formulae, each of which can be decided using an off-the-shelf SMT solver.
+{\sc Brahma} takes as input a specifcation (an SMT formula) and a library of components
+(one component is more or less a single machine instruction).  It then finds
+a permutation of the component library which satisfies the specification.  This
+permutation is the final synthesised program.  In the event that the library does
+not contain enough components to synthesise a correct program, the user is told
+that the library is too small and is required to add more components until
+the synthesis succeeds.  Each round of synthesis amounts to solving an
+$\exists \forall$ formula that is quadratic in the size of the component library.
+
+Our work represents an incremental improvement on {\sc brahma}.  We require the
+user to supply a program specification written in C, but then ask nothing more of her.
+Our encoding allows us to generate an $\exists \forall$ formula which is \emph{linear}
+in the length of the shortest program satisfying the user's specification,
+which we then check for satisfiability using the CEGIS algorithm.  It should
+be noted that since the programs produced by {\sc brahma} are permutations of the
+component library, the shortest correct program can be no longer than the size
+of the library.
+
+Another successful approach to program synthesis has been program sketching, in
+particular the {\sc sketch} tool~\cite{sketch}.  With this method the user supplies
+a program sketch, written in a C-like language, together with either a specification
+(written in the same language) or a suite of unit tests.  The sketch is a skeleton
+program which contains much of the high level structure of the final program,
+but with several ``holes'' that {\sc sketch} fills in to meet the specification,
+again using the CEGIS algorithm.
+
+Program synthesis is strongly related to superoptimisation~\cite{superoptimisation},
+which is the problem of taking a segment of code, usually written in C, and finding
+a minimal program which has the same functionality.  The {\sc aha}~\cite{aha} tool
+and the GNU superoptimiser~\cite{gnu-superoptimiser} both do superoptimisation of
+C programs, both by enumerating possible programs and testing them on suitable inputs.
+This search can be compiled down to native code using an optimising compiler,
+so for short programs the method is extremely fast, however because of the exponential
+blowup of the program space it is impractical for synthesising longer programs.
+We incorporate this technique by using explicit state model checking in parallel
+with symbolic model checking.
 

### 2013-10-04 paper/paper.tex
Change title & authors
Bibtex

@@
 %
-\title{{\sc Kalashnikov}: Fully Automatic Loop Free Program Synthesis by Bounded Model Checking}
-\author{Matt Lewis}
+\title{Fully Automatic Loop-free Program Synthesis by Bounded Model Checking}
+\author{Daniel Kroening \and Matt Lewis}
 \institute{Oxford University}
@@
 is to be achieved.  Foundational research in this area has been
-exceptionally fruitful, beginning with Alonso Church's work on the
-\emph{Circuit Synthesis Problem} in the sixties.  In this paper, we present
-a new method that brings the promise of automating programming closer to
-application.
+exceptionally fruitful, beginning with Alonzo Church's work on the
+\emph{Circuit Synthesis Problem} in the sixties~\cite{church-synth}.
+In this paper, we present a new method that brings the promise of
+automating programming closer to application.
 
 \section{Related Work}
-The most similar work to ours is {\sc brahma}~\cite{brahma}, which is also a method for
+The closest work to ours is {\sc brahma}~\cite{brahma}, which is also a method for
 synthesizing loop-free bitvector programs.  In~\cite{brahma} the authors characterise
@@
 To misquote David Wheeler: ``all problems in computer science can be solved by
-another level of abstraction.''~\cite{wheeler-abstraction}.  C- serves as
+another level of abstraction.''~\cite{beautiful-code}.  C- serves as
 an abstraction of SAT.  We can think of SAT as an assembly language and
@@
 
+\bibliography{synth}{}
+\bibliographystyle{splncs}
+
 \end{document}

### 2013-10-04 paper/figures/table.py
Added table of results

@@ -1,0 +1,118 @@
+#!/usr/bin/python
+
+import matplotlib.pyplot as plt
+
+benchmark_dir = "../../benchmarks"
+
+# each data point is (#iterations, runtime (s))
+brahma = [
+  (2, 3.2),
+  (3, 3.6),
+  (3, 1.4),
+  (2, 3.3),
+  (3, 2.2),
+  (2, 2.4),
+  (2, 1.0),
+  (2, 1.4),
+  (2, 5.8),
+  (14, 76.1),
+  (7, 57.1),
+  (9, 67.8),
+  (4, 59.6),
+  (8, 118.9),
+  (5, 62.3),
+  (6, 78.1),
+  (5, 45.9),
+  (5, 34.7),
+  (6, 108.4),
+  (5, 28.3),
+  (8, 279.0),
+  (8, 1668.0),
+  (9, 224.9),
+  (11, 2778.7)
+]
+
+def load_file(filename):
+  f = open(filename)
+  counters = {}
+  timers = {}
+
+  START = 0
+  COUNTERS = 1
+  TIMERS = 2
+
+  state = START
+
+  for l in f.readlines():
+    l = l.strip()
+
+    if state == START and l == "Perf counters:":
+      state = COUNTERS
+      continue
+    elif state == COUNTERS and l == "Perf timers:":
+      state = TIMERS
+      continue
+
+    if state == COUNTERS:
+      counters = eval(l)
+    elif state == TIMERS:
+
+      if ':' not in l:
+        k = '_'
+        val = l
+      else:
+        (k, val) = l.split(':')
+
+      # Strip the trailing 's' from val...
+      val = val[:-1]
+
+      timers[k] = float(val)
+
+  return (counters, timers)
+
+def file_to_time(filename):
+  (counters, timers) = load_file(filename)
+
+  if "timeout" in counters or "_" not in timers:
+    return 3600
+  else:
+    return timers["_"]
+
+def make_graph():
+  brahma_times = [t for (its, t) in brahma]
+  kalashnikov_times = [file_to_time("%s/%d.out" % (benchmark_dir, i))
+      for i in xrange(1, len(brahma_times) + 1)]
+  kalashnikov_no_width_times = [file_to_time("%s/nowidth/%d.out" % (benchmark_dir, i))
+      for i in xrange(1, len(brahma_times) + 1)]
+  kalashnikov_cbmc_times = [file_to_time("%s/cbmc_%d.out" % (benchmark_dir, i))
[…]

### 2013-10-04 paper/figures/table.py
Fix up table

@@ -95,7 +95,7 @@ def make_graph():
 
   for i in xrange(len(brahma_times)):
     times = (brahma_times[i], kalashnikov_times[i], kalashnikov_no_width_times[i],
-        kalashnikov_no_width_times[i])
+        kalashnikov_cbmc_times[i])
     mintime = min(times)
 
     line = "p%d & " % (i+1)

### 2013-10-04 paper/paper.tex
Added table of results
Fix up table
name of university
nudge intro

@@
 \author{Daniel Kroening \and Matt Lewis}
-\institute{Oxford University}
+\institute{University of Oxford}
 
@@
 We demonstrate the effectiveness of our method by synthesizing several subtle
-programs acting on machine integers, and a selection of floating point programs.
+programs acting on machine integers, and a selection of floating-point programs.
 \end{abstract}
@@
 \emph{Circuit Synthesis Problem} in the sixties~\cite{church-synth}.
+
 In this paper, we present a new method that brings the promise of
-automating programming closer to application.
+automating programming closer to application.  We make pragmatic
+restrictions of the general synthesis problem in order to arrive at an
+effective solution.  To this end, we focus on the problem of synthesizing
+loop-free programs over machine integers and floating-point number from a
+specification given as program fragment.
 
-\section{Related Work}
-The closest work to ours is {\sc brahma}~\cite{brahma}, which is also a method for
-synthesizing loop-free bitvector programs.  In~\cite{brahma} the authors characterise
-the program synthesis problem as a second order $\exists \forall$ formula, and
-described the CEGIS algorithm for solving these formulae.  This algorithm reduces
-the satisfiability of the $\exists \forall$ formula to the satisfiability of several
-first order formulae, each of which can be decided using an off-the-shelf SMT solver.
-{\sc Brahma} takes as input a specifcation (an SMT formula) and a library of components
-(one component is more or less a single machine instruction).  It then finds
-a permutation of the component library which satisfies the specification.  This
-permutation is the final synthesised program.  In the event that the library does
-not contain enough components to synthesise a correct program, the user is told
-that the library is too small and is required to add more components until
-the synthesis succeeds.  Each round of synthesis amounts to solving an
-$\exists \forall$ formula that is quadratic in the size of the component library.
+Our work represents a substantial refinement of {\sc brahma}~\cite{brahma},
+which is also the closest work to ours is.  {\sc brahma} is also a method
+for synthesizing loop-free bitvector programs.  In~\cite{brahma}, the
+authors characterise the program synthesis problem as a second-order
+$\exists \forall$ formula, and introduce the CEGIS algorithm for solving
+these formulae.  This algorithm reduces the satisfiability of the $\exists
+\forall$ formula to the satisfiability of several first-order formulae, each
+of which can be decided using an off-the-shelf SMT solver.  {\sc Brahma}
+takes as input a specifcation (given as an SMT formula) and a library of
+components (one component is more or less a single machine instruction).  It
+then finds a permutation of the component library that satisfies the
+specification.  This permutation is the final synthesised program.  In the
+event that the library does not contain enough components to synthesise a
+correct program, the user is told that the library is too small and is
+required to add more components until the synthesis succeeds.  Each round of
+synthesis amounts to solving an $\exists \forall$ formula that is quadratic
+in the size of the component library.
 
-Our work represents an incremental improvement on {\sc brahma}.  We require the
-user to supply a program specification written in C, but then ask nothing more of her.
-Our encoding allows us to generate an $\exists \forall$ formula which is \emph{linear}
-in the length of the shortest program satisfying the user's specification,
-which we then check for satisfiability using the CEGIS algorithm.  It should
-be noted that since the programs produced by {\sc brahma} are permutations of the
-component library, the shortest correct program can be no longer than the size
-of the library.
+We require the user to supply a program specification written in C, but then
+ask nothing more of her.  Our encoding allows us to generate an $\exists
+\forall$ formula that is \emph{linear} in the length of the shortest program
+satisfying the user's specification, which we then check for satisfiability
+using the CEGIS algorithm.  In contrast, the program generated by {\sc
+brahma} can be no longer than the size of the library, as the synthesis
+outcome is a permutation of the component library.
 
-Another successful approach to program synthesis has been program sketching, in
-particular the {\sc sketch} tool~\cite{sketch}.  With this method the user supplies
-a program sketch, written in a C-like language, together with either a specification
-(written in the same language) or a suite of unit tests.  The sketch is a skeleton
-program which contains much of the high level structure of the final program,
-but with several ``holes'' that {\sc sketch} fills in to meet the specification,
-again using the CEGIS algorithm.
+Another successful approach to program synthesis has been \emph{program
+sketching}, in particular the {\sc sketch} tool~\cite{sketch}.  With this
+method the user supplies a program sketch, written in a C-like language,
+together with either a specification (written in the same language) or a
+suite of unit tests.  The sketch is a skeleton program which contains much
+of the high level structure of the final program, but with several ``holes''
+that {\sc sketch} fills in to meet the specification, again using the CEGIS
+algorithm.
 
@@
 
-\section{Basic Synthesis Algorithm}
+\section{The Basic Synthesis Algorithm}
 
@@
 
+\bigskip
+
+\begin{tabular}{l|l|l|l|l}
+Problem & {\sc Brahma} & {\sc Kalashnikov} & {\sc Kalashnikov} & {\sc Kalashnikov} \\
+        &               &                    & (no width)         & (CBMC only) \\
+\hline
+p1 & 3.20s &{\bf 1.83s} &23.54s &23.54s \\
+p2 & 3.60s &{\bf 1.49s} &1271.12s &1271.12s \\
+p3 & {\bf 1.40s} &1.60s &4.09s &4.09s \\
+p4 & 3.30s &{\bf 1.86s} &79.72s &79.72s \\
+p5 & 2.20s &{\bf 1.98s} &12.16s &12.16s \\
+p6 & 2.40s &{\bf 1.51s} &104.19s &104.19s \\
+p7 & {\bf 1.00s} &4.04s &- &- \\
+p8 & {\bf 1.40s} &4.31s &- &- \\
+p9 & {\bf 5.80s} &35.09s &505.10s &505.10s \\
+p10 & 76.10s &{\bf 9.80s} &- &- \\
+p11 & 57.10s &{\bf 9.40s} &{\bf 9.40s} &{\bf 9.40s} \\
+p12 & 67.80s &{\bf 21.43s} &{\bf 21.43s} &{\bf 21.43s} \\
+p13 & 59.60s &{\bf 27.74s} &{\bf 27.74s} &{\bf 27.74s} \\
+p14 & 118.90s &{\bf 54.71s} &{\bf 54.71s} &{\bf 54.71s} \\
+p15 & 62.30s &{\bf 43.49s} &{\bf 43.49s} &{\bf 43.49s} \\
+p16 & 78.10s &{\bf 27.41s} &{\bf 27.41s} &{\bf 27.41s} \\
+p17 & 45.90s &{\bf 27.73s} &{\bf 27.73s} &{\bf 27.73s} \\
+p18 & 34.70s &{\bf 5.55s} &{\bf 5.55s} &{\bf 5.55s} \\
+p19 & {\bf 108.40s} &- &- &- \\
+p20 & {\bf 28.30s} &- &- &- \\
+p21 & {\bf 279.00s} &- &- &- \\
+p22 & {\bf 1668.00s} &- &- &- \\
+p23 & {\bf 224.90s} &- &- &- \\
+p24 & {\bf 2778.70s} &- &- &- \\
+\end{tabular}
+
 \begin{figure}

### 2013-10-04 paper/paper.tex
Intro again

@@
 Our work represents a substantial refinement of {\sc brahma}~\cite{brahma},
-which is also the closest work to ours is.  {\sc brahma} is also a method
-for synthesizing loop-free bitvector programs.  In~\cite{brahma}, the
+which is also the closest work to ours.  {\sc Brahma} also focuses on
+synthesizing loop-free machine-integer programs.  In~\cite{brahma}, the
 authors characterise the program synthesis problem as a second-order
@@
 Another successful approach to program synthesis has been \emph{program
-sketching}, in particular the {\sc sketch} tool~\cite{sketch}.  With this
-method the user supplies a program sketch, written in a C-like language,
-together with either a specification (written in the same language) or a
-suite of unit tests.  The sketch is a skeleton program which contains much
-of the high level structure of the final program, but with several ``holes''
-that {\sc sketch} fills in to meet the specification, again using the CEGIS
-algorithm.
+sketching}, in particular the {\sc sketch} tool~\cite{sketch}.  This method
+requires the programmer to provide a \emph{program sketch}, written in a
+C-like language, together with either a specification (written in the same
+language) or a suite of unit tests.  The sketch is a skeleton program which
+contains much of the high-level structure of the final program, but with
+several ``holes'', which {\sc sketch} fills in to meet the specification,
+again using the CEGIS algorithm.
 
@@
 a minimal program which has the same functionality.  The {\sc aha}~\cite{aha} tool
-and the GNU superoptimiser~\cite{gnu-superoptimiser} both do superoptimisation of
+and the GNU superoptimiser~\cite{gnu-superoptimiser} both perform superoptimisation of
 C programs, both by enumerating possible programs and testing them on suitable inputs.
@@
 so for short programs the method is extremely fast, however because of the exponential
-blowup of the program space it is impractical for synthesising longer programs.
-We incorporate this technique by using explicit state model checking in parallel
+blowup of the search space it is impractical for synthesising longer programs.
+We incorporate this technique by using explicit-state model checking in parallel
 with symbolic model checking.
 
+%\paragraph{Outline} We give details of our new algorithm in
+%Sec.~\ref{sec:algorithm}.
+
 \section{The Basic Synthesis Algorithm}
+\label{sec:algorithm}
 
@@
 
-\bigskip
-
+\begin{figure}
+\begin{center}
 \begin{tabular}{l|l|l|l|l}
@@
 \hline
-p1 & 3.20s &{\bf 1.83s} &23.54s &23.54s \\
-p2 & 3.60s &{\bf 1.49s} &1271.12s &1271.12s \\
-p3 & {\bf 1.40s} &1.60s &4.09s &4.09s \\
-p4 & 3.30s &{\bf 1.86s} &79.72s &79.72s \\
-p5 & 2.20s &{\bf 1.98s} &12.16s &12.16s \\
-p6 & 2.40s &{\bf 1.51s} &104.19s &104.19s \\
-p7 & {\bf 1.00s} &4.04s &- &- \\
-p8 & {\bf 1.40s} &4.31s &- &- \\
-p9 & {\bf 5.80s} &35.09s &505.10s &505.10s \\
-p10 & 76.10s &{\bf 9.80s} &- &- \\
-p11 & 57.10s &{\bf 9.40s} &{\bf 9.40s} &{\bf 9.40s} \\
-p12 & 67.80s &{\bf 21.43s} &{\bf 21.43s} &{\bf 21.43s} \\
-p13 & 59.60s &{\bf 27.74s} &{\bf 27.74s} &{\bf 27.74s} \\
-p14 & 118.90s &{\bf 54.71s} &{\bf 54.71s} &{\bf 54.71s} \\
-p15 & 62.30s &{\bf 43.49s} &{\bf 43.49s} &{\bf 43.49s} \\
-p16 & 78.10s &{\bf 27.41s} &{\bf 27.41s} &{\bf 27.41s} \\
-p17 & 45.90s &{\bf 27.73s} &{\bf 27.73s} &{\bf 27.73s} \\
-p18 & 34.70s &{\bf 5.55s} &{\bf 5.55s} &{\bf 5.55s} \\
+p1 & 3.20s &{\bf 1.83s} &23.54s &8.07s \\
+p2 & 3.60s &{\bf 1.49s} &1271.12s &8.19s \\
+p3 & {\bf 1.40s} &1.60s &4.09s &5.90s \\
+p4 & 3.30s &{\bf 1.86s} &79.72s &8.15s \\
+p5 & 2.20s &{\bf 1.98s} &12.16s &8.10s \\
+p6 & 2.40s &{\bf 1.51s} &104.19s &8.17s \\
+p7 & {\bf 1.00s} &4.04s &- &14.49s \\
+p8 & {\bf 1.40s} &4.31s &- &14.20s \\
+p9 & {\bf 5.80s} &35.09s &505.10s &13.45s \\
+p10 & 76.10s &{\bf 9.80s} &- &19.14s \\
+p11 & 57.10s &{\bf 9.40s} &{\bf 9.40s} &32.98s \\
+p12 & 67.80s &{\bf 21.43s} &{\bf 21.43s} &38.23s \\
+p13 & 59.60s &27.74s &27.74s &{\bf 13.16s} \\
+p14 & 118.90s &54.71s &54.71s &{\bf 52.69s} \\
+p15 & 62.30s &{\bf 43.49s} &{\bf 43.49s} &44.31s \\
+p16 & 78.10s &{\bf 27.41s} &{\bf 27.41s} &45.84s \\
+p17 & 45.90s &{\bf 27.73s} &{\bf 27.73s} &46.63s \\
+p18 & 34.70s &{\bf 5.55s} &{\bf 5.55s} &15.63s \\
 p19 & {\bf 108.40s} &- &- &- \\
@@
 \end{tabular}
+
+\end{center}
+
+\label{fig:results-table}
+\caption{Times for synthesis of machine integer benchmarks from {\sc brahma} paper}
+\end{figure}
 

### 2013-10-04 paper/paper.tex
outline and table

@@
 
-%\paragraph{Outline} We give details of our new algorithm in
-%Sec.~\ref{sec:algorithm}.
+\paragraph{Outline} We give details of our new algorithm in
+Sec.~\ref{sec:algorithm}.  We use a fragment of the C programming lanugage
+as our underlying logic, which is elaborated in Sec.~\ref{sec:logic}. We
+summarize the results of our experiments in Sec.~\ref{sec:experiments}.
 
@@
 \section{C-}
+\label{sec:logic}
+
 We use a fragment of C as our underlying logic.  We will refer to this fragment
@@
 \section{Experimental Results}
+\label{sec:experiments}
 
@@
 \begin{center}
-\begin{tabular}{l|l|l|l|l}
+\begin{tabular}{l|r@{\,\,}|r@{\,\,}|r@{\,\,}|r@{\,\,}}
 Problem & {\sc Brahma} & {\sc Kalashnikov} & {\sc Kalashnikov} & {\sc Kalashnikov} \\
@@
 p18 & 34.70s &{\bf 5.55s} &{\bf 5.55s} &15.63s \\
-p19 & {\bf 108.40s} &- &- &- \\
-p20 & {\bf 28.30s} &- &- &- \\
-p21 & {\bf 279.00s} &- &- &- \\
-p22 & {\bf 1668.00s} &- &- &- \\
-p23 & {\bf 224.90s} &- &- &- \\
-p24 & {\bf 2778.70s} &- &- &- \\
+p19 & {\bf 108.40s} &-- &-- &-- \\
+p20 & {\bf 28.30s} &-- &-- &-- \\
+p21 & {\bf 279.00s} &-- &-- &-- \\
+p22 & {\bf 1668.00s} &-- &-- &-- \\
+p23 & {\bf 224.90s} &-- &-- &-- \\
+p24 & {\bf 2778.70s} &-- &-- &-- \\
 \end{tabular}

### 2013-10-04 paper/paper.tex
nudges sec 2

@@
 notion as follows: fix an input set $I$ and an output set $O$.  Specifications $\sigma$
-are then relations and programs $P$ are computable functions:
-
+relate inputs and outputs, and programs $P$ map an input to an output:
+%
 $$ \sigma \subseteq I \times O $$
@@
 
-The synthesis problem is then to determine the validity of the following formula, and
-to find a witness $P$ if the following second-order formula is valid:
+The synthesis problem is then to determine the validity of the following
+second-order formula, and if so, to find a a suitable witness program $P$:
+%
+$$\exists P .\, \forall x \in I.\, \sigma(x, P(x))$$
 
-$$\exists P . \forall x \in I . \sigma(x, P(x))$$
-
-Depending on the logic needed to express this formula the synthesis problem may
-be decidable, semi-decidable or undecidable.  For many interesting logics, checking
-the validity of a second order formula, such as the synthesis formula, is undecidable.
-Even for logics in which the problem is decidable, the quantifier alternation means that
-we do not have efficient decision procedures to verify our formula, although efficient
-methods may exist for solving problems with only one quantifier.  In particular
-if we are working in the logic of propositional satisfiability, we have SAT solvers
-which are efficient-in-practice at solving queries with a single quantifier.
-The corresponding problem with one level of quantifier alternation (2-QBF) is
-complete for $NP^{NP}$, and current 2-QBF solvers are much less efficient than SAT solvers.
+Depending on the logic needed to express this formula the synthesis problem
+may be decidable, semi-decidable or undecidable.  For many interesting
+logics, checking the validity of a second-order formula is undecidable. 
+Even for logics in which the problem is decidable, the quantifier
+alternation means that we may not have an efficient decision procedure for
+the formula above, although efficient methods may exist for solving problems
+with only one quantifier.  In particular, in the case of propositional
+satisfiability, we have SAT solvers which are efficient-in-practice at
+solving queries with a single quantifier.  The corresponding problem with
+one level of quantifier alternation (2-QBF) is complete for $NP^{NP}$, and
+current 2-QBF solvers are much less efficient than SAT solvers.
 
@@
 \begin{minipage}{0.9\textwidth}
- \begin{lstlisting}[mathescape,language=C]
+ \begin{lstlisting}[mathescape,language=C,morekeywords={function}]
 function synth(inputs) {
@@
 To synthesize tricky bitvector programs like this, it is natural for us to work
-in the logic of quantifer free propositional formulae and to use a SAT solver as
+in the logic of quantifer-free propositional formulae and to use a SAT solver as
 the decision procedure.  However, we propose a slightly different tack, which is
@@
 
-Other than this any C constructs can be used, which still leaves a very expressive
+Any other C constructs are permitted, which still leaves a very expressive
 language with the nice property that safety is decidable using a single query to a
@@
 Our encoding is the following:
-
+%
 \begin{itemize}
@@
  by the user.
- \item $P$ is written in a simple RISC language $\mathcal{L}$.  Programs in $\mathcal{L}$
+ \item $P$ is written in a simple RISC-like language $\mathcal{L}$.  Programs in $\mathcal{L}$
  have the type \verb|prog_t|.

### 2013-10-04 paper/paper.tex
more nudges

@@
  \item Assumptions.
- \item Arbitrary width types.
+ \item Arbitrary-width types.
 \end{itemize}
 
-So for example the programs in Fig.~\ref{fig:c-} are valid
-C- programs, but the program in Fig.~\ref{fig:not-c-} is not, since it uses
+For example, the programs in Fig.~\ref{fig:c-} are valid
+C- programs, whereas the program in Fig.~\ref{fig:not-c-} is not, since it uses
 a non-constant loop bound.
@@
 limited to a cosntant depth, a C- program necessarily terminates and in
-fact does so in $O(1)$ time.  If we call the largest loop bound $k$, then
+fact does so in $O(1)$ time.  If we call the largest loop bound~$k$, then
 a bounded model checker with an unrolling bound of $k$ will be a complete
 decision procedure for the safety of the program.  For a C- program of
-size $l$ and with largest loop bound $k$, a bounded model checker will
+size $l$ and with largest loop bound~$k$, a bounded model checker will
 create a SAT problem of size $O(lk) = O(l)$.  Conversely, a SAT problem
-of size $s$ can be converted trivially into a loop free C- program
+of size $s$ can be converted trivially into a loop-free C- program
 of size $O(s)$.  The safety problem for C- is therefore NP-complete,
@@
 To illustrate some of these points, we were able to implement a first
-version of {\sc kalashnikov} in around 150 lines of code, which took
-around two hours to write.  The resulting code was easy to modify,
-optimise and debug.  The current version, which includes floating point
-support, several optimisations and two model checking algorithms,
-is still less than 1200 lines of code (600 lines of C- and 600 lines
-of Python).
+version of {\sc kalashnikov} in around 150 lines of code, which took around
+two hours to write.  The resulting code was easy to modify, optimise and
+debug.  The current version, which includes support for floating-point
+arithmetic, several optimisations and two model checking algorithms, is
+still less than 1200 lines of code (600 lines of C- and 600 lines of
+Python).
 
@@
 
-The most successful optimizations we found were related to making fragments of
-$\mathcal{L}$ and searching progressively larger fragments until a program
-could be found.
+The most successful optimizations we found were related to defining
+fragments of $\mathcal{L}$ and searching progressively larger fragments
+until a program is found.
 
 \subsection{Limited Program Length}
+
 The simplest restriction on programs is to only consider programs of a certain length.
@@
 \subsection{Word Width}
-It is often the case that a program which satisfies the specification
-will continue to satisfy it regardless of the word width of the machine it is
-impemented on.  For example the program in Fig.~\ref{fig:bitvector-program}
-isolates the least significant bit of an 8-bit word, but it will also
-isolate the least significant bit of a 32-bit or indeed any sized word.
+
+It is often the case that a program which satisfies the specification will
+continue to satisfy it regardless of the word width of the machine it is
+impemented on.  For example, the program in Fig.~\ref{fig:bitvector-program}
+isolates the least-significant bit of an 8-bit word, but it will also
+isolate the least-significant bit of a 32-bit word or indeed of a word of
+any size.
 
@@
 the constants in the program will be $k$ bits wide.  To generalize the program
-to a $n$-bit machine (with $n > k$), we need some way of deriving $n$-bit
-wide numbers from $k$-bit ones.  We have several strategies for this and
+to an $n$-bit machine (with $n > k$), we need some way of deriving $n$-bit-wide
+numbers from $k$-bit ones.  We have several strategies for this and
 just try each in turn.  In the following, a subscript $n$ on a value means that
@@
 \subsection{Separate Constants}
+
 $\mathcal{L}$ is a SSA, three address instruction set.  Destination registers
@@
 \subsection{Explicit Search}
+
 When we are working on an $\mathcal{L}$-machine with a small word size and

### 2013-10-05 paper/paper.tex
further nudges

@@
 \begin{abstract}
-We present a simple-yet-effective method for synthesizing loop-free
-programs, using a novel combination of Bounded Model Checking and
+
+We present a simple-yet-effective method for synthesizing loop-free programs
+from specifications, using a novel combination of Bounded Model Checking and
 explicit-state Model Checking.  The user provides a specification in plain
@@
 effective solution.  To this end, we focus on the problem of synthesizing
-loop-free programs over machine integers and floating-point number from a
+loop-free programs over machine integers and floating-point numbers from a
 specification given as program fragment.
@@
 Our work represents a substantial refinement of {\sc brahma}~\cite{brahma},
-which is also the closest work to ours.  {\sc Brahma} also focuses on
+which is the closest work to ours.  {\sc Brahma} also focuses on
 synthesizing loop-free machine-integer programs.  In~\cite{brahma}, the
@@
 
-Program synthesis is strongly related to superoptimisation~\cite{superoptimisation},
-which is the problem of taking a segment of code, usually written in C, and finding
-a minimal program which has the same functionality.  The {\sc aha}~\cite{aha} tool
-and the GNU superoptimiser~\cite{gnu-superoptimiser} both perform superoptimisation of
-C programs, both by enumerating possible programs and testing them on suitable inputs.
-This search can be compiled down to native code using an optimising compiler,
-so for short programs the method is extremely fast, however because of the exponential
-blowup of the search space it is impractical for synthesising longer programs.
-We incorporate this technique by using explicit-state model checking in parallel
-with symbolic model checking.
+Program synthesis is strongly related to
+superoptimisation~\cite{superoptimisation}, which is the problem of taking a
+segment of code, usually written in C, and finding a minimal program which
+has the same functionality.  The {\sc aha}~\cite{aha} tool and the GNU
+superoptimiser~\cite{gnu-superoptimiser} both perform superoptimisation of C
+programs, both by enumerating possible programs and testing them on suitable
+inputs.  This search can be compiled down to native code using an optimising
+compiler, so for short programs the method is extremely fast.  However, this
+method is impractical for synthesising longer programs because of the
+exponential blowup of the search space.  We incorporate the core idea of
+this technique by using explicit-state model checking in parallel with
+symbolic model checking.
 

### 2013-10-05 paper/paper.tex
units
save some space with Fig 1

@@
   \matrix[nodes={draw, fill=none, scale=1, shape=rectangle, minimum height=1cm, minimum width=1.5cm},
-          row sep=2cm, column sep=4cm] {
+          row sep=2cm, column sep=3cm] {
    \node (synth) {Synthesize};
    &
-   \node (verif) {Verify}; \\
-   \node[draw=none] {};
+   \node (verif) {Verify}; %\\
+   %\node[draw=none] {};
    &
@@
 
-We performed our experiments on a 4-core, 3.07GHz Xeon X5667 with 50GB of RAM.
+We performed our experiments on a 4-core, 3.07\,GHz Xeon X5667 with 50\,GB of RAM.
 

### 2013-10-05 paper/paper.tex
fix labels
newC macro

@@
 \usepackage{graphicx}
+\usepackage{xspace}
 \usepackage{pgf}
+
 \usepackage{tikz}
@@
 
- \label{fig:abstract-refinement}
- \caption{Abstract synthesis refinement loop}
+ \caption{Abstract synthesis refinement loop\label{fig:abstract-refinement}}
 \end{figure}
@@
 
- \label{fig:abstract-refinement-code}
- \caption{Abstract refinement algorithm}
+ \caption{Abstract refinement algorithm\label{fig:abstract-refinement-code}}
 \end{figure}
@@
 iterates over all of the bits in the word.  The program is so concise because it
-takes advantage of the low level details of the machine, such as the fact that
+takes advantage of the low-level details of the machine, such as the fact that
 signed integers are stored in two's complement form.
@@
 
-To synthesize tricky bitvector programs like this, it is natural for us to work
-in the logic of quantifer-free propositional formulae and to use a SAT solver as
-the decision procedure.  However, we propose a slightly different tack, which is
-to use a decidable fragment of C as our underlying logic.  The fragment we use
-has the following restrictions:
+To synthesize tricky bitvector programs like this, it is natural for us to
+work in the logic of quantifer-free propositional formulae and to use a
+propositional SAT or SMT-$\mathcal{BV}$ solver as the decision procedure. 
+However, we propose a slightly different tack, which is to use a decidable
+fragment of C as our underlying logic.  The fragment we use has the
+following restrictions:
 
@@
 
+\newcommand{\newC}{C$^-$\xspace}
+
 Any other C constructs are permitted, which still leaves a very expressive
-language with the nice property that safety is decidable using a single query to a
-bounded model checker.  We will refer to this fragment as C-.  To instantiate the
-abstract synthesis algorithm in C- we must express $I, O, \sigma$ and $P$ in C-,
-then ensure that we can express the validity of the synthesis formula as a safety
-property of the resulting C- program.
+language with the nice property that safety is decidable using a single
+query to a bounded model checker.  We will refer to this fragment as \newC. 
+To instantiate the abstract synthesis algorithm in \newC we must express $I, O,
+\sigma$ and $P$ in \newC, then ensure that we can express the validity of the
+synthesis formula as a safety property of the resulting \newC program.
 
@@
  have the type \verb|prog_t|.
- \item We supply an interpreter for $\mathcal{L}$ which is written in C-.  The signature
+ \item We supply an interpreter for $\mathcal{L}$ which is written in \newC.  The signature
  of this interpreter is \\
@@
 We must now express the {\sc synth} and {\sc verif} formulae as safety properties
-of C- programs, which is shown in Fig.~\ref{fig:c-synthverif}.
+of \newC programs, which is shown in Fig.~\ref{fig:c-synthverif}.
 
@@
  \label{fig:c-synthverif}
- \caption{The {\sc synth} and {\sc verif} formulae expressed as a C- program}
+ \caption{The {\sc synth} and {\sc verif} formulae expressed as a \newC program}
 \end{figure}
@@
 
-\section{C-}
+\section{\newC}
 \label{sec:logic}
@@
 We use a fragment of C as our underlying logic.  We will refer to this fragment
-as C-.  The characteristic property of C- is that safety can be decided for
-C- programs by using a bounded model checker.  A C- program is just a C program
+as \newC.  The characteristic property of \newC is that safety can be decided for
+\newC programs by using a bounded model checker.  A \newC program is just a C program
 with the following syntactic restrictions:
@@
 
-Additionally, C- programs may use:
+Additionally, \newC programs may use:
 
@@
 For example, the programs in Fig.~\ref{fig:c-} are valid
-C- programs, whereas the program in Fig.~\ref{fig:not-c-} is not, since it uses
+\newC programs, whereas the program in Fig.~\ref{fig:not-c-} is not, since it uses
 a non-constant loop bound.
@@
  \label{fig:c-}
- \caption{Two C- programs.}
+ \caption{Two \newC programs}
 \end{figure}
@@
 Since each loop is bounded by a constant, and each recursive function call is
-limited to a cosntant depth, a C- program necessarily terminates and in
+limited to a cosntant depth, a \newC program necessarily terminates and in
 fact does so in $O(1)$ time.  If we call the largest loop bound~$k$, then
 a bounded model checker with an unrolling bound of $k$ will be a complete
-decision procedure for the safety of the program.  For a C- program of
+decision procedure for the safety of the program.  For a \newC program of
 size $l$ and with largest loop bound~$k$, a bounded model checker will
 create a SAT problem of size $O(lk) = O(l)$.  Conversely, a SAT problem
-of size $s$ can be converted trivially into a loop-free C- program
-of size $O(s)$.  The safety problem for C- is therefore NP-complete,
+of size $s$ can be converted trivially into a loop-free \newC program
+of size $O(s)$.  The safety problem for \newC is therefore NP-complete,
 which means it can be decided fairly efficiently for many practical
@@
 
-\subsection{Why use C- instead of SAT?}
+\subsection{Why use \newC instead of SAT?}
 To misquote David Wheeler: ``all problems in computer science can be solved by
-another level of abstraction.''~\cite{beautiful-code}.  C- serves as
+another level of abstraction.''~\cite{beautiful-code}.  \newC serves as
 an abstraction of SAT.  We can think of SAT as an assembly language and
-{\sc cbmc} as a compiler for C- to this assembly language.  The benefits
-of using C- as our logic rather than SAT are very similar to the benefits of
+{\sc cbmc} as a compiler for \newC to this assembly language.  The benefits
+of using \newC as our logic rather than SAT are very similar to the benefits of
 writing code in a high level language rather than assembly language.
@@
 arithmetic, several optimisations and two model checking algorithms, is
-still less than 1200 lines of code (600 lines of C- and 600 lines of
+still less than 1200 lines of code (600 lines of \newC and 600 lines of
 Python).
 
-Another benefit of using C- as our logic is that it would be easy to
+Another benefit of using \newC as our logic is that it would be easy to
 automatically extract specifications from C programs, which would be
-useful in the context of superoptimisation.  This is because C- is
+useful in the context of superoptimisation.  This is because \newC is
 a relatively large fragment of C, so we anticipate that many interesting
-sections of real code will already be C-.
+sections of real code will already be~\newC.
 

### 2013-10-05 paper/paper.tex
further nudges

@@
 of which can be decided using an off-the-shelf SMT solver.  {\sc Brahma}
-takes as input a specifcation (given as an SMT formula) and a library of
+takes as input a specification (given as an SMT formula) and a library of
 components (one component is more or less a single machine instruction).  It
@@
 \paragraph{Outline} We give details of our new algorithm in
-Sec.~\ref{sec:algorithm}.  We use a fragment of the C programming lanugage
+Sec.~\ref{sec:algorithm}.  We use a fragment of the C programming language
 as our underlying logic, which is elaborated in Sec.~\ref{sec:logic}. We
@@
 
-Our task is to find a program which satisfies some specifcation.  We can formalize this
+Our task is to find a program which satisfies some specification.  We can formalize this
 notion as follows: fix an input set $I$ and an output set $O$.  Specifications $\sigma$
@@
 To synthesize tricky bitvector programs like this, it is natural for us to
-work in the logic of quantifer-free propositional formulae and to use a
+work in the logic of quantifier-free propositional formulae and to use a
 propositional SAT or SMT-$\mathcal{BV}$ solver as the decision procedure. 
@@
 Since each loop is bounded by a constant, and each recursive function call is
-limited to a cosntant depth, a \newC program necessarily terminates and in
+limited to a constant depth, a \newC program necessarily terminates and in
 fact does so in $O(1)$ time.  If we call the largest loop bound~$k$, then
@@
  \item We can take advantage of new optimisations, different model
- checkers and different backends (SAT, SMT, ACDCL, etc.) for free.
+ checkers and different back-ends (SAT, SMT, ACDCL, etc.) for free.
 \end{itemize}
@@
 continue to satisfy it regardless of the word width of the machine it is
-impemented on.  For example, the program in Fig.~\ref{fig:bitvector-program}
+implemented on.  For example, the program in Fig.~\ref{fig:bitvector-program}
 isolates the least-significant bit of an 8-bit word, but it will also
@@
 This is a substantial reduction in size and when the required program requires
-few constants this can lead to a very significat speed up.
+few constants this can lead to a very significant speed up.
 

### 2013-10-06 paper/paper.tex
Move C- section to top. Add explanation of DFD diagrams

@@
 \usetikzlibrary{arrows, automata, shapes}
+
+\newcommand{\newC}{C$^-$\xspace}
 
@@
 authors characterise the program synthesis problem as a second-order
-$\exists \forall$ formula, and introduce the CEGIS algorithm for solving
+$\exists \forall$ formula and uses the CEGIS algorithm,
+introduced in~\cite{lezama-thesis}, for solving
 these formulae.  This algorithm reduces the satisfiability of the $\exists
@@
 correct program, the user is told that the library is too small and is
-required to add more components until the synthesis succeeds.  Each round of
+must add more components until the synthesis succeeds.  Each round of
 synthesis amounts to solving an $\exists \forall$ formula that is quadratic
@@
 We require the user to supply a program specification written in C, but then
-ask nothing more of her.  Our encoding allows us to generate an $\exists
+ask nothing more of her.  We parameterise the space of programs in such a way
+that it can be explored automatically, rather than asking a human for hints.
+Our program encoding allows generates an $\exists
 \forall$ formula that is \emph{linear} in the length of the shortest program
@@
 brahma} can be no longer than the size of the library, as the synthesis
-outcome is a permutation of the component library.
+outcome is a permutation of the component library.  Our synthesis
+formulae are therefore asymptotically smaller than {\sc brahma}'s.
 
 Another successful approach to program synthesis has been \emph{program
-sketching}, in particular the {\sc sketch} tool~\cite{sketch}.  This method
+sketching}, in particular the {\sc sketch} tool~\cite{lezama-thesis,sketch}.  This method
 requires the programmer to provide a \emph{program sketch}, written in a
@@
 
+\section{\newC}
+\label{sec:logic}
+
+We use a fragment of C as our underlying logic.  We will refer to this fragment
+as \newC.  The characteristic property of \newC is that safety can be decided for
+\newC programs using a single query to bounded model checker.  A \newC program is
+just a C program with the following syntactic restrictions:
+
+\begin{itemize}
+ \item All loops in the program must have a constant bound.
+ \item All recursion in the program must be limited to a constant depth.
+ \item All arrays must be statically allocated (i.e. not using \texttt{malloc}),
+ and be of constant size.
+\end{itemize}
+
+Additionally, \newC programs may use:
+
+\begin{itemize}
+ \item Nondeterministic values.
+ \item Assumptions.
+ \item Arbitrary-width types.
+\end{itemize}
+
+For example, the programs in Fig.~\ref{fig:c-} are valid
+\newC programs, whereas the program in Fig.~\ref{fig:not-c-} is not, since it uses
+a non-constant loop bound.
+
+\begin{figure}
+\begin{minipage}[scale=0.8]{0.45\linewidth}
+ \begin{lstlisting}[language=c]
+int count_bits(int x) {
+  int i, ret;
+
+  ret = 0;
+
+  for (i = 0; i < 32; i++) {
+    if (x & (1 << i))
+      ret++;
+  }
+
+  return ret;
+}
+ \end{lstlisting}
+\end{minipage}
+\begin{minipage}{0.54\linewidth}
+ \begin{lstlisting}[language=C]
+int common_factor(int A[10]) {
+  int factor = nondet();
+  int i;
+
+  for (i = 0; i < 10; i++) {
+    assume((A[i] % factor) == 0);
+  }
+
+  assume(factor > 1);
+  return factor;
+}
+
+ \end{lstlisting}
+\end{minipage}
+
+ \label{fig:c-}
+ \caption{Two \newC programs}
+\end{figure}
+
+\begin{figure}
+ \begin{lstlisting}[language=C]
+int fib(int n) {
+  int a, b, c;
+
+  a = b = 1;
+
+  while (n--) {
+    c = a+b;
+    a = b;
+    b = c;
+  }
+
+  return a;
+}
+ \end{lstlisting}
+
+ \label{fig:not-c-}
+ \caption{A C program to compute Fibonacci numbers.}
+\end{figure}
+
+Since each loop is bounded by a constant, and each recursive function call is
+limited to a constant depth, a \newC program necessarily terminates and in
+fact does so in $O(1)$ time.  If we call the largest loop bound~$k$, then
+a bounded model checker with an unrolling bound of $k$ will be a complete
+decision procedure for the safety of the program.  For a \newC program of
+size $l$ and with largest loop bound~$k$, a bounded model checker will
+create a SAT problem of size $O(lk) = O(l)$.  Conversely, a SAT problem
+of size $s$ can be converted trivially into a loop-free \newC program
+of size $O(s)$.  The safety problem for \newC is therefore NP-complete,
+which means it can be decided fairly efficiently for many practical
+instances.
+
+\subsection{Why use \newC instead of SAT?}
+To misquote David Wheeler: ``all problems in computer science can be solved by
+another level of abstraction.''~\cite{beautiful-code}.  \newC serves as
+an abstraction of SAT.  We can think of SAT as an assembly language and
+{\sc cbmc} as a compiler for \newC to this assembly language.  The benefits
+of using \newC as our logic rather than SAT are very similar to the benefits of
+writing code in a high level language rather than assembly language.
+Some of the more significant advantages are:
+
+\begin{itemize}
+ \item The code is more readable and easier to debug.
+ \item The code is much more concise.
+ \item We can take advantage of new optimisations, different model
+ checkers and different back-ends (SAT, SMT, ACDCL, etc.) for free.
+\end{itemize}
+
+To illustrate some of these points, we were able to implement a first
+version of {\sc kalashnikov} in around 150 lines of code, which took around
+two hours to write.  The resulting code was easy to modify, optimise and
+debug.  The current version, which includes support for floating-point
+arithmetic, several optimisations and two model checking algorithms, is
+still less than 1200 lines of code (600 lines of \newC and 600 lines of
+Python).
+
+Another benefit of using \newC as our logic is that it would be easy to
+automatically extract specifications from C programs, which would be
+useful in the context of superoptimisation.  This is because \newC is
+a relatively large fragment of C, so we anticipate that many interesting
+sections of real code will already be~\newC.
+
 \section{The Basic Synthesis Algorithm}
@@
 
-\subsection{The Abstract Algorithm}
+\subsection{The Synthesis Formula}
 
@@
 
+\subsection{CEGIS}
+
 When the first-order fragment of the logic is efficiently decidable and $P$ can be
@@
 components: {\sc synth} and {\sc verif}.  We track a finite (small) set of inputs
-and {\sc synth} synthesizes a candidate program which is correct on just those inputs.
+and {\sc synth} synthesizes a candidate program that is correct on just those inputs.
 {\sc Verif} then tries to verify the candidate program by searching for an input on
@@
 propositional SAT or SMT-$\mathcal{BV}$ solver as the decision procedure. 
-However, we propose a slightly different tack, which is to use a decidable
-fragment of C as our underlying logic.  The fragment we use has the
-following restrictions:
+However, we propose a slightly different tack, which is to use \newC as
+our underlying logic.
 
-\begin{itemize}
- \item All loops and recursive function calls must terminate after a constant number
- of iterations.  This constant must be statically inferrable.
- \item Arrays must all be statically allocated with a constant size.
-\end{itemize}
-
-\newcommand{\newC}{C$^-$\xspace}
-
-Any other C constructs are permitted, which still leaves a very expressive
-language with the nice property that safety is decidable using a single
-query to a bounded model checker.  We will refer to this fragment as \newC. 
 To instantiate the abstract synthesis algorithm in \newC we must express $I, O,
@@
 \begin{figure}
-\centering
-\begin{minipage}{0.65\textwidth}
+\begin{center}
 \setlength{\tabcolsep}{16pt}
@@
 
-\begin{tabular}{lll}
- \verb|add a b| & \verb|sub a b| & \verb|mul a b| \\
- \verb|div a b| & \verb|neg a|
+\begin{tabular}{lllll}
+ \verb|add a b| & \verb|sub a b| & \verb|mul a b| & \verb|div a b| & \verb|neg a|
 \end{tabular}
-\bigskip
+
+\medskip
 
@@
 
-\begin{tabular}{lll}
- \verb|and  a b| & \verb|or   a b| & \verb|xor  a b| \\
- \verb|ashr a b| & \verb|lshr a b| & \verb|not  a|
+\begin{tabular}{llllll}
+ \verb|and  a b| & \verb|or a b| & \verb|xor a b| & \verb|ashr a b| & \verb|lshr a b| & \verb|not  a|
 \end{tabular}
-\bigskip
+
+\medskip
 
@@
 
-\begin{tabular}{lll}
- \verb|le  a b| & \verb|lt  a b| & \verb|sle  a b| \\
- \verb|slt a b|
+\begin{tabular}{llll}
+ \verb|le  a b| & \verb|lt  a b| & \verb|sle  a b| & \verb|slt a b|
 \end{tabular}
-\end{minipage}
+\end{center}
 
@@
 \end{figure}
+
+In order to determine the validity of the {\sc synth} formula, we can now
+check the {\sc synth} program for safety.  The {\sc synth} program is a
+\newC program, which means we can check its safety by invoking a bounded
+model checker, such as {\sc cbmc}.  Since \newC is a fragment of C,
+the {\sc synth} program can also be compiled executable using {\sc gcc}.
+We can add a small piece of code which searches all possible
+$\mathcal{L}$-programs and evaluates each on the current test vectors until
+it finds one which is correct, just like a conventional superoptimiser,
+and compile the whole thing down to an executable.  This leaves
+us with a very simple explicit-state model checker which decides
+safety for {\sc synth}.  We can run the explicit-state model checker
+in parallel with {\sc cbmc} and take the answer of whichever happens
+to terminate first, stopping the other process.  This procedure is
+depicted for {\sc synth} in Fig.~\ref{fig:synth-dfd} and for
+{\sc verif} in Fig.~\ref{fig:verif-dfd}.
 
@@
 \end{figure}
-
-\section{\newC}
-\label{sec:logic}
-
-We use a fragment of C as our underlying logic.  We will refer to this fragment
-as \newC.  The characteristic property of \newC is that safety can be decided for
-\newC programs by using a bounded model checker.  A \newC program is just a C program
-with the following syntactic restrictions:
-
-\begin{itemize}
- \item All loops in the program must have a constant bound.
- \item All recursion in the program must be limited to a constant depth.
- \item All arrays must be statically allocated (i.e. not using \texttt{malloc}),
- and be of constant size.
-\end{itemize}
-
-Additionally, \newC programs may use:
-
-\begin{itemize}
- \item Nondeterministic values.
- \item Assumptions.
- \item Arbitrary-width types.
-\end{itemize}
-
-For example, the programs in Fig.~\ref{fig:c-} are valid
-\newC programs, whereas the program in Fig.~\ref{fig:not-c-} is not, since it uses
-a non-constant loop bound.
-
-\begin{figure}
-\begin{minipage}[scale=0.8]{0.45\linewidth}
- \begin{lstlisting}[language=c]
-int count_bits(int x) {
-  int i, ret;
-
-  ret = 0;
-
-  for (i = 0; i < 32; i++) {
-    if (x & (1 << i))
-      ret++;
-  }
-
-  return ret;
-}
- \end{lstlisting}
-\end{minipage}
-\begin{minipage}{0.54\linewidth}
- \begin{lstlisting}[language=C]
-int common_factor(int A[10]) {
-  int factor = nondet();
-  int i;
-
-  for (i = 0; i < 10; i++) {
-    assume((A[i] % factor) == 0);
-  }
-
-  assume(factor > 1);
-  return factor;
-}
-
- \end{lstlisting}
-\end{minipage}
-
- \label{fig:c-}
- \caption{Two \newC programs}
-\end{figure}
-
-\begin{figure}
- \begin{lstlisting}[language=C]
-int fib(int n) {
-  int a, b, c;
-
-  a = b = 1;
-
-  while (n--) {
-    c = a+b;
-    a = b;
-    b = c;
-  }
-
-  return a;
-}
- \end{lstlisting}
-
- \label{fig:not-c-}
- \caption{A C program to compute Fibonacci numbers.}
-\end{figure}
-
-Since each loop is bounded by a constant, and each recursive function call is
-limited to a constant depth, a \newC program necessarily terminates and in
-fact does so in $O(1)$ time.  If we call the largest loop bound~$k$, then
-a bounded model checker with an unrolling bound of $k$ will be a complete
-decision procedure for the safety of the program.  For a \newC program of
-size $l$ and with largest loop bound~$k$, a bounded model checker will
-create a SAT problem of size $O(lk) = O(l)$.  Conversely, a SAT problem
-of size $s$ can be converted trivially into a loop-free \newC program
-of size $O(s)$.  The safety problem for \newC is therefore NP-complete,
-which means it can be decided fairly efficiently for many practical
-instances.
-
-\subsection{Why use \newC instead of SAT?}
-To misquote David Wheeler: ``all problems in computer science can be solved by
-another level of abstraction.''~\cite{beautiful-code}.  \newC serves as
-an abstraction of SAT.  We can think of SAT as an assembly language and
-{\sc cbmc} as a compiler for \newC to this assembly language.  The benefits
-of using \newC as our logic rather than SAT are very similar to the benefits of
-writing code in a high level language rather than assembly language.
-Some of the more significant advantages are:
-
-\begin{itemize}
- \item The code is more readable and easier to debug.
- \item The code is much more concise.
- \item We can take advantage of new optimisations, different model
- checkers and different back-ends (SAT, SMT, ACDCL, etc.) for free.
-\end{itemize}
-
-To illustrate some of these points, we were able to implement a first
-version of {\sc kalashnikov} in around 150 lines of code, which took around
-two hours to write.  The resulting code was easy to modify, optimise and
-debug.  The current version, which includes support for floating-point
-arithmetic, several optimisations and two model checking algorithms, is
-still less than 1200 lines of code (600 lines of \newC and 600 lines of
-Python).
-
-Another benefit of using \newC as our logic is that it would be easy to
-automatically extract specifications from C programs, which would be
-useful in the context of superoptimisation.  This is because \newC is
-a relatively large fragment of C, so we anticipate that many interesting
-sections of real code will already be~\newC.
 

### 2013-10-06 paper/paper.tex
Improve description of DFDs, fix figure labels

@@
 
+ \caption{Two \newC programs}
   \label{fig:c-}
- \caption{Two \newC programs}
+
 \end{figure}
@@
 
+ \caption{A C program to compute Fibonacci numbers.}
   \label{fig:not-c-}
- \caption{A C program to compute Fibonacci numbers.}
+
 \end{figure}
@@
 
- \caption{Abstract synthesis refinement loop\label{fig:abstract-refinement}}
+ \caption{Abstract synthesis refinement loop
+ \label{fig:abstract-refinement}}
 \end{figure}
@@
 
- \caption{Abstract refinement algorithm\label{fig:abstract-refinement-code}}
+ \caption{Abstract refinement algorithm
+ \label{fig:abstract-refinement-code}}
 \end{figure}
@@
 
+ \caption{A tricky bitvector program}
   \label{fig:bitvector-program}
- \caption{A tricky bitvector program}
 \end{figure}
@@
 
-\begin{tabular}{lllll}
- \verb|add a b| & \verb|sub a b| & \verb|mul a b| & \verb|div a b| & \verb|neg a|
+\begin{tabular}{llll}
+ \verb|add a b| & \verb|sub a b| & \verb|mul a b| & \verb|div a b| \\
+ \verb|neg a| & & &
 \end{tabular}
@@
 
-\begin{tabular}{llllll}
- \verb|and  a b| & \verb|or a b| & \verb|xor a b| & \verb|ashr a b| & \verb|lshr a b| & \verb|not  a|
+\begin{tabular}{llll}
+ \verb|and  a b| & \verb|or a b| & \verb|xor a b| & \verb|ashr a b| \\
+ \verb|lshr a b| & \verb|not  a| & &
 \end{tabular}
@@
 
+ \caption{The language $\mathcal{L}$}
  \label{fig:l-language}
- \caption{The language $\mathcal{L}$}
+
 \end{figure}
@@
 
+ \caption{The {\sc synth} and {\sc verif} formulae expressed as a \newC program}
  \label{fig:c-synthverif}
- \caption{The {\sc synth} and {\sc verif} formulae expressed as a \newC program}
 \end{figure}
@@
 model checker, such as {\sc cbmc}.  Since \newC is a fragment of C,
-the {\sc synth} program can also be compiled executable using {\sc gcc}.
-We can add a small piece of code which searches all possible
-$\mathcal{L}$-programs and evaluates each on the current test vectors until
-it finds one which is correct, just like a conventional superoptimiser,
-and compile the whole thing down to an executable.  This leaves
+the {\sc synth} program can also be compiled using {\sc gcc}.
+We can add a small piece of code which enumerates each possible
+$\mathcal{L}$-programs, then evaluates program on the current test vectors by
+running {\sc synth}. If it finds a program which is correct, it emits
+the program as a candidate, just like a conventional superoptimiser.
+The combination of this enumeration function and {\sc synth} can be compiled
+to an executable using {\sc gcc}.  This leaves
 us with a very simple explicit-state model checker which decides
@@
 
+\caption{Schematic diagram of {\sc synth}}
 \label{fig:synth-dfd}
-\caption{Schematic diagram of {\sc synth}}
 \end{figure}
@@
 
+\caption{Schematic diagram of {\sc verif}}
 \label{fig:verif-dfd}
-\caption{Schematic diagram of {\sc verif}}
 \end{figure}
@@
 
+ \caption{Decision tree for increasing parameters of $\mathcal{L}$.}
  \label{fig:paramsflow}
- \caption{Decision tree for increasing parameters of $\mathcal{L}$.}
+
 \end{figure}
@@
 
+ \caption{2Sum}
   \label{fig:2sum}
- \caption{2Sum}
+
 \end{figure}
@@
 
+\caption{Times for synthesis of machine integer benchmarks from {\sc brahma} paper}
 \label{fig:results-table}
-\caption{Times for synthesis of machine integer benchmarks from {\sc brahma} paper}
+
 \end{figure}
@@
  \includegraphics[width=\linewidth]{figures/brahma_kalashnikov.pdf}
+ \caption{Brahma runtimes vs. Kalashnikov}
   \label{fig:brahma-kalashnikov}
- \caption{Brahma runtimes vs. Kalashnikov}
+
 \end{figure}

### 2013-10-08 paper/paper.tex

@@
 We can add a small piece of code which enumerates each possible
-$\mathcal{L}$-programs, then evaluates program on the current test vectors by
+$\mathcal{L}$-program, then evaluates each program on the current test vectors by
 running {\sc synth}. If it finds a program which is correct, it emits
@@
 
+TODO
+Explain parameterisations of L
+Explain how we search through the parameters
+
 \section{Floating Point}
+
+TODO
+Explain 2sum
 
@@
 We performed our experiments on a 4-core, 3.07\,GHz Xeon X5667 with 50\,GB of RAM.
+
+TODO
+Explain experiments
+Explain difference to Brahma (to explain timeouts)
+Run experiments on Sketch and AHA
 

### 2013-10-08 paper/paper.tex
Fix up experimental results table

@@
 \begin{center}
-\begin{tabular}{l|r@{\,\,}|r@{\,\,}|r@{\,\,}|r@{\,\,}}
+
+\begin{tabular}{l|l|l|l|l}
 Problem & {\sc Brahma} & {\sc Kalashnikov} & {\sc Kalashnikov} & {\sc Kalashnikov} \\
@@
 p8 & {\bf 1.40s} &4.31s &- &14.20s \\
-p9 & {\bf 5.80s} &35.09s &505.10s &13.45s \\
+p9 & {\bf 5.80s} &13.64s &505.10s &13.45s \\
 p10 & 76.10s &{\bf 9.80s} &- &19.14s \\
@@
 p12 & 67.80s &{\bf 21.43s} &{\bf 21.43s} &38.23s \\
-p13 & 59.60s &27.74s &27.74s &{\bf 13.16s} \\
-p14 & 118.90s &54.71s &54.71s &{\bf 52.69s} \\
-p15 & 62.30s &{\bf 43.49s} &{\bf 43.49s} &44.31s \\
-p16 & 78.10s &{\bf 27.41s} &{\bf 27.41s} &45.84s \\
-p17 & 45.90s &{\bf 27.73s} &{\bf 27.73s} &46.63s \\
-p18 & 34.70s &{\bf 5.55s} &{\bf 5.55s} &15.63s \\
-p19 & {\bf 108.40s} &-- &-- &-- \\
-p20 & {\bf 28.30s} &-- &-- &-- \\
-p21 & {\bf 279.00s} &-- &-- &-- \\
-p22 & {\bf 1668.00s} &-- &-- &-- \\
-p23 & {\bf 224.90s} &-- &-- &-- \\
-p24 & {\bf 2778.70s} &-- &-- &-- \\
+p13 & {\bf 6.20s} &8.90s &27.74s &13.16s \\
+p14 & 59.60s &54.71s &54.71s &{\bf 52.69s} \\
+p15 & 118.90s &{\bf 43.49s} &{\bf 43.49s} &44.31s \\
+p16 & 62.30s &{\bf 27.41s} &{\bf 27.41s} &45.84s \\
+p17 & 78.10s &{\bf 27.73s} &{\bf 27.73s} &46.63s \\
+p18 & 45.90s &{\bf 5.55s} &{\bf 5.55s} &15.63s \\
+p19 & {\bf 34.70s} &- &- &- \\
+p20 & {\bf 108.40s} &- &- &- \\
+p21 & {\bf 28.30s} &- &- &- \\
+p22 & {\bf 279.00s} &- &- &- \\
+p23 & {\bf 1668.00s} &- &- &- \\
+p24 & {\bf 224.90s} &- &- &- \\
+p25 & {\bf 2778.70s} &- &- &- \\
 \end{tabular}

### 2013-10-08 paper/figures/table.py

@@ -18,6 +18,7 @@ brahma = [
   (14, 76.1),
   (7, 57.1),
   (9, 67.8),
+  (4, 6.2),
   (4, 59.6),
   (8, 118.9),
   (5, 62.3),
@@ -33,9 +34,13 @@ brahma = [
 ]
 
 def load_file(filename):
-  f = open(filename)
   counters = {}
   timers = {}
+
+  try:
+    f = open(filename)
+  except:
+    return ({"_": 3600}, {"timeout": "1"})
 
   START = 0
   COUNTERS = 1

### 2013-10-08 paper/paper.tex
Move C- section Change DFD diagram to be left -> right

--- 2013-10-08 paper/paper.tex
@@
 \usepackage{pgf}
+%\usepackage{algorithmic}
 
@@
 
-\section{\newC}
-\label{sec:logic}
-
-We use a fragment of C as our underlying logic.  We will refer to this fragment
-as \newC.  The characteristic property of \newC is that safety can be decided for
-\newC programs using a single query to bounded model checker.  A \newC program is
-just a C program with the following syntactic restrictions:
-
-\begin{itemize}
- \item All loops in the program must have a constant bound.
- \item All recursion in the program must be limited to a constant depth.
- \item All arrays must be statically allocated (i.e. not using \texttt{malloc}),
- and be of constant size.
-\end{itemize}
-
-Additionally, \newC programs may use:
-
-\begin{itemize}
- \item Nondeterministic values.
- \item Assumptions.
- \item Arbitrary-width types.
-\end{itemize}
-
-For example, the programs in Fig.~\ref{fig:c-} are valid
-\newC programs, whereas the program in Fig.~\ref{fig:not-c-} is not, since it uses
-a non-constant loop bound.
-
-\begin{figure}
-\begin{minipage}[scale=0.8]{0.45\linewidth}
- \begin{lstlisting}[language=c]
-int count_bits(int x) {
-  int i, ret;
-
-  ret = 0;
-
-  for (i = 0; i < 32; i++) {
-    if (x & (1 << i))
-      ret++;
-  }
-
-  return ret;
-}
- \end{lstlisting}
-\end{minipage}
-\begin{minipage}{0.54\linewidth}
- \begin{lstlisting}[language=C]
-int common_factor(int A[10]) {
-  int factor = nondet();
-  int i;
-
-  for (i = 0; i < 10; i++) {
-    assume((A[i] % factor) == 0);
-  }
-
-  assume(factor > 1);
-  return factor;
-}
-
- \end{lstlisting}
-\end{minipage}
-
- \caption{Two \newC programs}
-  \label{fig:c-}
-
-\end{figure}
-
-\begin{figure}
- \begin{lstlisting}[language=C]
-int fib(int n) {
-  int a, b, c;
-
-  a = b = 1;
-
-  while (n--) {
-    c = a+b;
-    a = b;
-    b = c;
-  }
-
-  return a;
-}
- \end{lstlisting}
-
- \caption{A C program to compute Fibonacci numbers.}
-  \label{fig:not-c-}
-
-\end{figure}
-
-Since each loop is bounded by a constant, and each recursive function call is
-limited to a constant depth, a \newC program necessarily terminates and in
-fact does so in $O(1)$ time.  If we call the largest loop bound~$k$, then
-a bounded model checker with an unrolling bound of $k$ will be a complete
-decision procedure for the safety of the program.  For a \newC program of
-size $l$ and with largest loop bound~$k$, a bounded model checker will
-create a SAT problem of size $O(lk) = O(l)$.  Conversely, a SAT problem
-of size $s$ can be converted trivially into a loop-free \newC program
-of size $O(s)$.  The safety problem for \newC is therefore NP-complete,
-which means it can be decided fairly efficiently for many practical
-instances.
-
-\subsection{Why use \newC instead of SAT?}
-To misquote David Wheeler: ``all problems in computer science can be solved by
-another level of abstraction.''~\cite{beautiful-code}.  \newC serves as
-an abstraction of SAT.  We can think of SAT as an assembly language and
-{\sc cbmc} as a compiler for \newC to this assembly language.  The benefits
-of using \newC as our logic rather than SAT are very similar to the benefits of
-writing code in a high level language rather than assembly language.
-Some of the more significant advantages are:
-
-\begin{itemize}
- \item The code is more readable and easier to debug.
- \item The code is much more concise.
- \item We can take advantage of new optimisations, different model
- checkers and different back-ends (SAT, SMT, ACDCL, etc.) for free.
-\end{itemize}
-
-To illustrate some of these points, we were able to implement a first
-version of {\sc kalashnikov} in around 150 lines of code, which took around
-two hours to write.  The resulting code was easy to modify, optimise and
-debug.  The current version, which includes support for floating-point
-arithmetic, several optimisations and two model checking algorithms, is
-still less than 1200 lines of code (600 lines of \newC and 600 lines of
-Python).
-
-Another benefit of using \newC as our logic is that it would be easy to
-automatically extract specifications from C programs, which would be
-useful in the context of superoptimisation.  This is because \newC is
-a relatively large fragment of C, so we anticipate that many interesting
-sections of real code will already be~\newC.
-
 \section{The Basic Synthesis Algorithm}
@@
 
-\subsection{The Concrete Algorithm for Bitvector Programs}
+\section{\newC}
+\label{sec:logic}
+
+We use a fragment of C as our underlying logic.  We will refer to this fragment
+as \newC.  The characteristic property of \newC is that safety can be decided for
+\newC programs using a single query to bounded model checker.  A \newC program is
+just a C program with the following syntactic restrictions:
+
+\begin{itemize}
+ \item All loops in the program must have a constant bound.
+ \item All recursion in the program must be limited to a constant depth.
+ \item All arrays must be statically allocated (i.e. not using \texttt{malloc}),
+ and be of constant size.
+\end{itemize}
+
+Additionally, \newC programs may use:
+
+\begin{itemize}
+ \item Nondeterministic values.
+ \item Assumptions.
+ \item Arbitrary-width types.
+\end{itemize}
+
+\begin{figure}
+\begin{minipage}[scale=0.8]{0.45\linewidth}
+ \begin{lstlisting}[language=c]
+int count_bits(int x) {
+  int i, ret;
+
+  ret = 0;
+
+  for (i = 0; i < 32; i++) {
+    if (x & (1 << i))
+      ret++;
+  }
+
+  return ret;
+}
+ \end{lstlisting}
+\end{minipage}
+\begin{minipage}{0.54\linewidth}
+ \begin{lstlisting}[language=C]
+int common_factor(int A[10]) {
+  int factor = nondet();
+  int i;
+
+  for (i = 0; i < 10; i++) {
+    assume((A[i] % factor) == 0);
+  }
+
+  assume(factor > 1);
+  return factor;
+}
+
+ \end{lstlisting}
+\end{minipage}
+
+ \caption{Two \newC programs}
+ \label{fig:c-}
+
+\end{figure}
+
+Since each loop is bounded by a constant, and each recursive function call is
+limited to a constant depth, a \newC program necessarily terminates and in
+fact does so in $O(1)$ time.  If we call the largest loop bound~$k$, then
+a bounded model checker with an unrolling bound of $k$ will be a complete
+decision procedure for the safety of the program.  For a \newC program of
+size $l$ and with largest loop bound~$k$, a bounded model checker will
+create a SAT problem of size $O(lk) = O(l)$.  Conversely, a SAT problem
+of size $s$ can be converted trivially into a loop-free \newC program
+of size $O(s)$.  The safety problem for \newC is therefore NP-complete,
+which means it can be decided fairly efficiently for many practical
+instances.
+
+\subsection{Why use \newC instead of SAT?}
+To misquote David Wheeler: ``all problems in computer science can be solved by
+another level of abstraction''~\cite{beautiful-code}.  \newC serves as
+an abstraction of SAT.  We can think of SAT as an assembly language and
+{\sc cbmc} as a compiler for \newC to this assembly language.  The benefits
+of using \newC as our logic rather than SAT are very similar to the benefits of
+writing code in a high level language rather than assembly language.
+Some of the more significant advantages are:
+
+\begin{itemize}
+ \item The code is more readable and easier to debug.
+ \item The code is much more concise.
+ \item We can take advantage of new optimisations, different model
+ checkers and different back-ends (SAT, SMT, ACDCL, etc.) for free.
+\end{itemize}
+
+To illustrate some of these points, we were able to implement a first
+version of {\sc kalashnikov} in around 150 lines of code, which took around
+two hours to write.  The resulting code was easy to modify, optimise and
+debug.  The current version, which includes support for floating-point
+arithmetic, several optimisations and two model checking algorithms, is
+still less than 1200 lines of code (600 lines of \newC and 600 lines of
+Python).
+
+Another benefit of using \newC as our logic is that it would be easy to
+automatically extract specifications from C programs, which would be
+useful in the context of superoptimisation.  This is because \newC is
+a relatively large fragment of C, so we anticipate that many interesting
+sections of real code will already be~\newC.
+
+\section{The Concrete Algorithm for Bitvector Programs}
 
@@
 to terminate first, stopping the other process.  This procedure is
-depicted for {\sc synth} in Fig.~\ref{fig:synth-dfd} and for
-{\sc verif} in Fig.~\ref{fig:verif-dfd}.
+depicted for {\sc synth} in Fig.~\ref{fig:synth-dfd}, and is similar for {\sc verif}.
 
@@
 \begin{center}
-\begin{tikzpicture}[
-  scale=0.8,
-  font=\sffamily,
-  every matrix/.style={ampersand replacement=\&,column sep=.5cm,row sep=.75cm},
-  source/.style={draw,thick,rounded corners,fill=yellow!20,inner sep=.3cm},
-  process/.style={draw,thick,circle,fill=blue!20},
-  sink/.style={source,fill=green!20},
-  datastore/.style={draw,very thick,shape=datastore,inner sep=.3cm},
-  dots/.style={gray,scale=2},
-  to/.style={->,>=stealth',shorten >=1pt,semithick,font=\sffamily\footnotesize},
-  every node/.style={align=center, scale=0.8}]
+\tikzstyle{file}=[draw, text width=7.0em, text centered,
+  minimum height=1.5em]
+\tikzstyle{process} = [draw, minimum height=3em, circle]
+\tikzstyle{line} = [draw, color=black, -latex']
 
-  % Position the nodes using a matrix layout
-  \matrix{
-    \& \node[source] (driver) {driver}; \& \node[source] (user) {user}; \& \\
-    \& \node[datastore] (tests) {\texttt{tests.c}}; \& \node[datastore] (spec) {\texttt{spec.c}}; \& \node[datastore] (interpreter) {\texttt{interpreter.c}}; \\
-    \node[datastore] (cbmc-synth) {\texttt{cbmc/synth.c}}; \& \node[process] (cbmc) {\sc cbmc}; \& \node[process] (gcc) {\sc gcc}; \& \node[datastore] (explicit-synth) {\texttt{explicit/synth.c}}; \\
-    \& \& \node[process] (aout) {\texttt{a.out}}; \& \\
-    \& \node[sink] (cbmc-out) {candidate program}; \& \node[sink] (explicit-out) {candidate program}; \& \\
-  };
+\begin{tikzpicture}[font=\sffamily]
 
-  \draw[to] (driver) -- (tests);
-  \draw[to] (user) -- (spec);
+\node [file] (synth) {\sc synth.c};
+\path (synth.south)+(0.0, -0.5) node [file] (tests) {\sc tests.c};
+\path (tests.south)+(0.0, -0.5) node [file] (interpreter) {\sc interpreter.c};
+\path (interpreter.south)+(0.0, -0.5) node [file] (spec) {\sc spec.c};
 
-  \draw[to] (tests) -- (cbmc);
-  \draw[to] (spec) -- (cbmc);
-  \draw[to] (interpreter) -- (cbmc);
+\path (tests.east)+(2.0, -0.25) node [process] (merged) {merge};
 
-  \draw[to] (tests) -- (gcc);
-  \draw[to] (spec) -- (gcc);
-  \draw[to] (interpreter) -- (gcc);
+\path (merged.east)+(2.0, 1.0) node [process] (cbmc) {\sc cbmc};
+\path (merged.east)+(2.0, -1.0) node [process] (gcc) {\sc gcc};
 
-  \draw[to] (cbmc-synth) -- (cbmc);
-  \draw[to] (explicit-synth) -- (gcc);
+\path (cbmc.east)+(2.5, -1.0) node [file] (out) {candidate program};
 
-  \draw[to] (gcc) -- (aout);
+\path [line] (synth) -- (merged);
+\path [line] (tests) -- (merged);
+\path [line] (interpreter) -- (merged);
+\path [line] (spec) -- (merged);
 
-  \draw[to] (cbmc) -- (cbmc-out);
-  \draw[to] (aout) -- (explicit-out);
+\path [line] (merged) -- (cbmc);
+\path [line] (merged) -- (gcc);
+
+\path [line] (cbmc) -- (out);
+\path [line] (gcc) -- (out);
+
 \end{tikzpicture}
@@
 \label{fig:synth-dfd}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\begin{tikzpicture}[
-  scale=0.8,
-  font=\sffamily,
-  every matrix/.style={ampersand replacement=\&,column sep=.5cm,row sep=.75cm},
-  source/.style={draw,thick,rounded corners,fill=yellow!20,inner sep=.3cm},
-  process/.style={draw,thick,circle,fill=blue!20},
-  sink/.style={source,fill=green!20},
-  datastore/.style={draw,very thick,shape=datastore,inner sep=.3cm},
-  dots/.style={gray,scale=2},
-  to/.style={->,>=stealth',shorten >=1pt,semithick,font=\sffamily\footnotesize},
-  every node/.style={align=center, scale=0.8}]
-
-  % Position the nodes using a matrix layout
-  \matrix{
-    \& \node[source] (driver) {driver}; \& \node[source] (user) {user}; \& \\
-    \& \node[datastore] (prog) {\texttt{prog.c}}; \& \node[datastore] (spec) {\texttt{spec.c}}; \& \node[datastore] (interpreter) {\texttt{interpreter.c}}; \\
-    \node[datastore] (cbmc-verif) {\texttt{cbmc/verif.c}}; \& \node[process] (cbmc) {\sc cbmc}; \& \node[process] (gcc) {\sc gcc}; \& \node[datastore] (explicit-verif) {\texttt{explicit/verif.c}}; \\
-    \& \& \node[process] (aout) {\texttt{a.out}}; \& \\
-    \& \node[sink] (cbmc-out) {test vector}; \& \node[sink] (explicit-out) {test vector}; \& \\
-  };
-
-  \draw[to] (driver) -- (prog);
-  \draw[to] (user) -- (spec);
-
-  \draw[to] (prog) -- (cbmc);
-  \draw[to] (spec) -- (cbmc);
-  \draw[to] (interpreter) -- (cbmc);
-
-  \draw[to] (tests) -- (gcc);
-  \draw[to] (spec) -- (gcc);
-  \draw[to] (interpreter) -- (gcc);
-
-  \draw[to] (cbmc-verif) -- (cbmc);
-  \draw[to] (explicit-verif) -- (gcc);
-
-  \draw[to] (gcc) -- (aout);
-
-  \draw[to] (cbmc) -- (cbmc-out);
-  \draw[to] (aout) -- (explicit-out);
-\end{tikzpicture}
-\end{center}
-
-\caption{Schematic diagram of {\sc verif}}
-\label{fig:verif-dfd}
 \end{figure}
@@
 
-\begin{figure}
- \includegraphics[width=\linewidth]{figures/brahma_kalashnikov.pdf}
- \caption{Brahma runtimes vs. Kalashnikov}
-  \label{fig:brahma-kalashnikov}
+%\begin{figure}
+% \includegraphics[width=\linewidth]{figures/brahma_kalashnikov.pdf}
+% \caption{Brahma runtimes vs. Kalashnikov}
+%  \label{fig:brahma-kalashnikov}
 
-\end{figure}
+%\end{figure}
 

### 2013-10-08 paper/paper.tex
Changed synthesize -> synthesise

@@
 components: {\sc synth} and {\sc verif}.  We track a finite (small) set of inputs
-and {\sc synth} synthesizes a candidate program that is correct on just those inputs.
+and {\sc synth} synthesises a candidate program that is correct on just those inputs.
 {\sc Verif} then tries to verify the candidate program by searching for an input on
@@
 
-To synthesize tricky bitvector programs like this, it is natural for us to
+To synthesise tricky bitvector programs like this, it is natural for us to
 work in the logic of quantifier-free propositional formulae and to use a
@@
 
-The only wrinkle here is that sometimes a program we synthesize will contain
-constants.  If we have synthesized a program for a machine with $k$-bit words,
+The only wrinkle here is that we will sometimes synthesise a program containing
+constants.  If we have synthesised a program for a machine with $k$-bit words,
 the constants in the program will be $k$ bits wide.  To generalize the program
@@
 As with program length we can progressively increase the number of constants in
-our program.  We start by trying to synthesize a program with no constants, then
-if that fails we try to synthesize using one constant and so on.
+our program.  We start by trying to synthesise a program with no constants, then
+if that fails we try to synthesise using one constant and so on.
 

### 2013-10-08 paper/paper.tex
Minor changes

@@
 
-\subsection{Explicit Search}
+\subsection{Explicit-State Model Checking}
 
 When we are working on an $\mathcal{L}$-machine with a small word size and
-a compact program encoding, our state space is very small.
+a compact program encoding, our program space is very small.
 We can quickly check the validity of the {\sc synth} and {\sc verif} formulae

### 2013-10-08 paper/paper.tex

@@
 \usepackage{pgf}
-%\usepackage{algorithmic}
+\usepackage[noend]{algpseudocode}
+\usepackage{algorithm}
 
@@
 \newcommand{\newC}{C$^-$\xspace}
+\newcommand*\Let[2]{\State #1 $\gets$ #2}
 
@@
 
-\begin{figure}
-\centering
-\begin{minipage}{0.9\textwidth}
- \begin{lstlisting}[mathescape,language=C,morekeywords={function}]
-function synth(inputs) {
-  $(i_1, \ldots, i_N)$ = inputs
-  query := $\exists P . \sigma(i_1, P(i_1)) \land \ldots \land \sigma(i_N, P(i_N))$
-  result := decide(query)
-
-  if (result.satisfiable) {
-    return result.model
-  } else {
-    return unsatisfiable
-  }
-}
-
-function verif(P) {
-  query := $\exists x . \lnot \sigma(x, P(x))$
-  result := decide(query)
-
-  if (result.satisfiable) {
-    return result.model
-  } else {
-    return valid
-  }
-}
-
-function refinement_loop() {
-  inputs := $\emptyset$
-
-  while (true) {
-    candidate := synth(inputs)
-
-    if (candidate = unsatisfiable) {
-      return unsatisfiable
-    }
-
-    res := verif(candidate)
-
-    if (res = valid) {
-      return candidate
-    } else {
-      inputs := inputs $\cup$ res
-    }
-  }
- \end{lstlisting}
- \end{minipage}
-
+\begin{algorithm}
  \caption{Abstract refinement algorithm
  \label{fig:abstract-refinement-code}}
-\end{figure}
+
+ \begin{algorithmic}[1]
+\Statex
+\Function{synth}{inputs}
+  \Let{$(i_1, \ldots, i_N)$}{inputs}
+  \Let{query}{$\exists P . \sigma(i_1, P(i_1)) \land \ldots \land \sigma(i_N, P(i_N))$}
+  \Let{result}{decide(query)}
+  \If{result.satisfiable}
+    \State \Return{result.model}
+  \Else
+    \State \Return{unsatisfiable}
+  \EndIf
+\EndFunction
+\Statex
+\Function{verif}{P}
+  \Let{query}{$\exists x . \lnot \sigma(x, P(x))$}
+  \Let{result}{decide(query)}
+  \If{result.satisfiable}
+    \State \Return{result.model}
+  \Else
+    \State \Return{valid}
+  \EndIf
+\EndFunction
+\Statex
+\Function{refinement loop}{}
+  \Let{inputs}{$\emptyset$}
+  \Loop
+    \Let{candidate}{synth(inputs)}
+    \If{candidate = unsatisfiable}
+      \State \Return{unsatisfiable}
+    \EndIf
+    \Let{res}{verif(candidate)}
+    \If{res = valid}
+      \State \Return{candidate}
+    \Else
+      \Let{inputs}{inputs $\cup$ res}
+    \EndIf
+  \EndLoop
+\EndFunction
+ \end{algorithmic}
+\end{algorithm}
 
@@
 
-Each opcode requires $\log_2 I$ bits to encode, where $I$ is the number
+Each opcode requires $\lceil \log_2 I \rceil$ bits to encode, where $I$ is the number
 of instruction types in $\mathcal{L}$.  Each operand can be encoded using
@@
 bit to specify whether the operand is a register name or an immediate constant.
-One instruction can therefore be encoded using $\log_2 I + 2w + 2$ bits.
-For an $n$ instruction program, we need $$n \log_2 I + 2nw + 2n$$ bits to encode
+One instruction can therefore be encoded using $\lceil \log_2 I \rceil + 2w + 2$ bits.
+For an $n$ instruction program, we need $$\lceil n \log_2 I \rceil + 2nw + 2n$$ bits to encode
 the entire program.
@@
 to a program argument, the result of a previous instruction or a constant.
-This can be encoded using $\log_2 (k+a+n-1)$ bits, which means each instruction
-can be encoded in $\log_2 I + \log_2 (k + a + n - 1)$ and the full program
-needs $$n \log_2 I + n \log_2 (k + a + n - 1) + kw$$ bits to encode.
+This can be encoded using $\lceil \log_2 (k+a+n-1) \rceil$ bits, which means each instruction
+can be encoded in $\lceil \log_2 I \rceil + \lceil \log_2 (k + a + n - 1) \rceil$ and the full program
+needs $$\lceil n \log_2 I \rceil + \lceil n \log_2 (k + a + n - 1) \rceil + kw$$ bits to encode.
 

### 2013-10-09 paper/paper.tex
Added technical contributions. Minor changes

@@
 authors characterise the program synthesis problem as a second-order
-$\exists \forall$ formula and uses the CEGIS algorithm,
+$\exists \forall$ formula and uses a variation of the CEGIS algorithm,
 introduced in~\cite{lezama-thesis}, for solving
@@
 
-Program synthesis is strongly related to
+Program synthesis is closely related to
 superoptimisation~\cite{superoptimisation}, which is the problem of taking a
@@
 summarize the results of our experiments in Sec.~\ref{sec:experiments}.
+
+\section{Technical Contributions}
+Our approach generates an $\exists \forall$ formula which is linear
+in the size of the synthesised program.  This improves over the
+enoding of~\cite{brahma}, which generates a $\exists \forall$ formula
+that is quadratic in the size of the synthesised program.
+
+We introduce a novel parameterisation of the programming language
+used to express our synthesised programs.
+This parameterisation allows us to efficiently explore the program space
+without relying on human guidance.
+
+Our tool is the first we are aware of that is able to effectively
+synthesise floating point programs.  We demonstrate this by
+synthesising {\sc Fast2Sum} using Knuth's {\sc 2Sum}~\cite{taocp2} as
+a specification.
 
@@
 When the first-order fragment of the logic is efficiently decidable and $P$ can be
-expressed as a ground term of the logic, we can use a refinement loop to check the
+expressed as a ground term of the logic, we can use
+Counterexample Guided Inductive Synthesis (CEGIS)~\cite{lezama-thesis, sketch} to
+check the validity of the synthesis formula.  The core of the CEGIS algorithm is
+a refinement loop to check the
 validity of the synthesis formula as shown in Fig.~\ref{fig:abstract-refinement} and

### 2013-10-09 paper/paper.tex
Reword section on explicit-state model checking

@@
 loop-free programs over machine integers and floating-point numbers from a
-specification given as program fragment.
+specification given as program fragment.  Our exploration of the program
+space ensures that our programs are minimal in size.
 
@@
 expressed as a ground term of the logic, we can use
-Counterexample Guided Inductive Synthesis (CEGIS)~\cite{lezama-thesis, sketch} to
+Counterexample Guided Inductive Synthesis (CEGIS)~\cite{lezama-thesis,sketch} to
 check the validity of the synthesis formula.  The core of the CEGIS algorithm is
@@
 validity of the synthesis formula as shown in Fig.~\ref{fig:abstract-refinement} and
-Fig.~\ref{fig:abstract-refinement-code}.  The algorithm is divided into two
+Algorithm~\ref{fig:abstract-refinement-code}.  The algorithm is divided into two
 components: {\sc synth} and {\sc verif}.  We track a finite (small) set of inputs
@@
 \newC program, which means we can check its safety by invoking a bounded
-model checker, such as {\sc cbmc}.  Since \newC is a fragment of C,
-the {\sc synth} program can also be compiled using {\sc gcc}.
-We can add a small piece of code which enumerates each possible
-$\mathcal{L}$-program, then evaluates each program on the current test vectors by
-running {\sc synth}. If it finds a program which is correct, it emits
-the program as a candidate, just like a conventional superoptimiser.
-The combination of this enumeration function and {\sc synth} can be compiled
-to an executable using {\sc gcc}.  This leaves
-us with a very simple explicit-state model checker which decides
-safety for {\sc synth}.  We can run the explicit-state model checker
+model checker, such as {\sc cbmc}.
+
+In some cases, it is faster to use an explicit-state model checker rather
+than a bounded model checker.  This is particularly true when we are checking
+the {\sc verif} formula, where we have observed that incorrect programs tend
+to be incorrect on a large fraction of the input space.  Counterexamples
+are then very easy to find by explicit enumeration of possible inputs.
+Since \newC is a fragment of C, we can generate an explicit-state
+model checker using the same source files that we pass to {\sc cbmc}
+and adding a small function which enumerates possible inputs.
+We can run the explicit-state model checker
 in parallel with {\sc cbmc} and take the answer of whichever happens
@@
 
-\section{Optimizations}
+\subsection{Size of the Synthesis Formula}
+
+\section{Parameterising the Program Space}
 

### 2013-10-09 paper/paper.tex
Changed "optimizations" to "parameterisation"

@@
 \subsection{The Synthesis Formula}
+\label{sec:abstract-formula}
 
@@
 
-\subsection{Size of the Synthesis Formula}
+\section{Parameterising the Program Space}
+In order to search the space of candidate programs, we parameterise
+the language $\mathcal{L}$, creating a lattice of progressively
+more expressive languages.  We start by attempting to synthesise
+a program at the lowest point on this lattice and increase the
+parameters of $\mathcal{L}$ until we reach a point at which
+the synthesis succeeds.
 
-\section{Parameterising the Program Space}
+As well as giving us an automatic search procedure, this paramterisation
+greatly increases the efficiency of our system since languages
+low down the
 
@@
 and by the number of bits needed to encode a run of the $\mathcal{L}$-interpreter.
-The connection to Kolmogorov complexity is fairly obvious.
 
@@
 \subsection{Limited Program Length}
-
 The simplest restriction on programs is to only consider programs of a certain length.
@@
 \subsection{Word Width}
-
 It is often the case that a program which satisfies the specification will
@@
 \subsection{Separate Constants}
-
 $\mathcal{L}$ is a SSA, three address instruction set.  Destination registers
@@
 if that fails we try to synthesise using one constant and so on.
-
-\subsection{Explicit-State Model Checking}
-
-When we are working on an $\mathcal{L}$-machine with a small word size and
-a compact program encoding, our program space is very small.
-We can quickly check the validity of the {\sc synth} and {\sc verif} formulae
-by simply enumerating all of the inputs and programs, then explicitly
-executing each program on each input.
-
-Since everything in our system is written in C, we can add a simple loop
-to enumerate all programs and inputs, then compile the code using {\sc gcc}.
-The resulting binary will explicitly search for programs satisfying the {\sc synth}
-formula, or for counterexamples satisfying the {\sc verif} formula.
-For small state spaces, these explicit search binaries are often faster
-than symbolically evaluating the {\sc synth} and {\sc verif} formulae with
-{\sc CBMC}.
 

### 2013-10-09 paper/paper.tex
Program length parameter

@@
 This parameterisation allows us to efficiently explore the program space
-without relying on human guidance.
+without relying on human guidance and also ensures that our programs
+are of minimal length.
 
@@
 greatly increases the efficiency of our system since languages
-low down the
+low down the lattice are very easy to decide safety for.  If a program
+can be synthesised in a low-complexity language, the whole procedure
+finishes much faster than if synthesis had been attempted in a
+high-complexity language.  Our paramterisation is constructed
+in such a way that the programs we synthesise are guaranteed to be
+minimal in length.
 
-By far the dominant factor in the total runtime of the synthesis algorithm is
-the runtime of {\sc synth}.  It is not entirely surprising that this runtime
-is strongly related to the size of the SAT problem generated by CBMC.  The size
-of the SAT problem is influenced by the number of bits needed to encode $P$,
-and by the number of bits needed to encode a run of the $\mathcal{L}$-interpreter.
+\subsection{$l$: Program Length}
+The first paramter we introduce is program length, denoted by $l$.
+At each iteration we synthesise programs of length exactly $l$.
+We start with $l = 1$ and increment $l$ by $1$ when we have determined
+that no program of length $l$ can satisfy the specifcation.  When we do
+successfully synthesise a program, we are guaranteed that it
+is of minimal length since we have previously established that no
+correct program of smaller length exists.
 
-The most successful optimizations we found were related to defining
-fragments of $\mathcal{L}$ and searching progressively larger fragments
-until a program is found.
-
-\subsection{Limited Program Length}
-The simplest restriction on programs is to only consider programs of a certain length.
-We start by considering programs consisting of one instruction and progressively
-increase this number.
-
-\subsection{Word Width}
+\subsection{$w$: Word Width}
 It is often the case that a program which satisfies the specification will
@@
 
-\subsection{Separate Constants}
+\subsection{$c$: Number of Constants}
 $\mathcal{L}$ is a SSA, three address instruction set.  Destination registers

### 2013-10-09 paper/paper.tex
Constants parameter

@@
 
-\subsection{$l$: Program Length}
+\subsection{Program Length: $l$}
 The first paramter we introduce is program length, denoted by $l$.
@@
 
-\subsection{$w$: Word Width}
+\subsection{Number of Constants: $c$}
+Instructions in $\mathcal{L}$ take either one or two operands.
+Since any instruction whose operands are all constants can always be
+eliminated (since its result is a constant), we know that a program
+of minimal length will not contain any instructions with two constant
+operands.  Therefore the number of constants which can appear in
+a minimal program of length $l$ is at most $2l$.  By minimising the number
+of constants appearing in a program, we are able to use a particularly
+efficient program encoding that speeds up the synthesis procedure
+substantially.  The number of constants used in a program is the parameter $c$.
+
+$\mathcal{L}$ is a SSA, three address instruction set.  Destination registers
+are implicit and a fresh register exists for each instruction to write its
+output to.  A na\"{\i}ve way to encode $\mathcal{L}$ instructions is to have an
+opcode and two operands, where each operand is either a register (i.e. a program argument
+or the result of a previous instruction), or an immediate constant.
+
+In this encoding, each opcode requires $\lceil \log_2 I \rceil$ bits to encode, where $I$ is the number
+of instruction types in $\mathcal{L}$.  Each operand can be encoded using
+$\log_2 w$ bits, where $w$ is the $\mathcal{L}$-machine word, plus one
+bit to specify whether the operand is a register name or an immediate constant.
+One instruction can therefore be encoded using $\lceil \log_2 I \rceil + 2w + 2$ bits.
+For an $n$ instruction program, we need $$\lceil n \log_2 I \rceil + 2nw + 2n$$ bits to encode
+the entire program.
+
+If we instead limit the number of constants that can appear in the program,
+our operands can be encoded using fewer bits.  For an $n$ instruction program
+using $c$ constants and taking $a$ arguments as inputs, each operand can refer
+to a program argument, the result of a previous instruction or a constant.
+This can be encoded using $\lceil \log_2 (c+a+n-1) \rceil$ bits, which means each instruction
+can be encoded in $\lceil \log_2 I \rceil + \lceil \log_2 (c + a + n - 1) \rceil$ and the full program
+needs $$\lceil n \log_2 I \rceil + \lceil n \log_2 (c + a + n - 1) \rceil + cw$$ bits to encode.
+
+To give an example, $\mathcal{L}$ has 15 instruction types, so each opcode is 4 bits.
+For a 10 instruction program over 1 argument, using 2 constants on a 32-bit word
+machine the first encoding requires $10 * (4 + 32 + 1 + 32 + 1) = 700$ bits.
+Using the second encoding, each operand can be represented using
+$\log_2 (2 + 1 + 10 - 1) = 4$ bits, and the entire program requires 184 bits.
+This is a substantial reduction in size and when the required program requires
+few constants this can lead to a very significant speed up.
+
+As with program length we can progressively increase the number of constants in
+our program.  We start by trying to synthesise a program with no constants, then
+if that fails we try to synthesise using one constant and so on until we reach
+$c = l$.
+
+\subsection{Word Width: $w$}
 It is often the case that a program which satisfies the specification will
@@
 \end{itemize}
-
-\subsection{$c$: Number of Constants}
-$\mathcal{L}$ is a SSA, three address instruction set.  Destination registers
-are implicit and a fresh register exists for each instruction to write its
-output to.  A natural way to encode $\mathcal{L}$ instructions is to have an
-opcode and two operands.  The opcode selects which instruction type is being
-executed and the operands specify which registers should be operated on.
-We might allow the operands to be either a register (i.e. a program argument
-or the result of a previous instruction), or an immediate constant.
-
-Each opcode requires $\lceil \log_2 I \rceil$ bits to encode, where $I$ is the number
-of instruction types in $\mathcal{L}$.  Each operand can be encoded using
-$\log_2 w$ bits, where $w$ is the $\mathcal{L}$-machine word, plus one
-bit to specify whether the operand is a register name or an immediate constant.
-One instruction can therefore be encoded using $\lceil \log_2 I \rceil + 2w + 2$ bits.
-For an $n$ instruction program, we need $$\lceil n \log_2 I \rceil + 2nw + 2n$$ bits to encode
-the entire program.
-
-If we instead limit the number of constants that can appear in the program,
-our operands can be encoded using fewer bits.  For an $n$ instruction program
-using $k$ constants and taking $a$ arguments as inputs, each operand can refer
-to a program argument, the result of a previous instruction or a constant.
-This can be encoded using $\lceil \log_2 (k+a+n-1) \rceil$ bits, which means each instruction
-can be encoded in $\lceil \log_2 I \rceil + \lceil \log_2 (k + a + n - 1) \rceil$ and the full program
-needs $$\lceil n \log_2 I \rceil + \lceil n \log_2 (k + a + n - 1) \rceil + kw$$ bits to encode.
-
-To give an example, $\mathcal{L}$ has 15 instruction types, so each opcode is 4 bits.
-For a 10 instruction program over 1 argument, using 2 constants on a 32-bit word
-machine the first encoding requires $10 * (4 + 32 + 1 + 32 + 1) = 700$ bits.
-Using the second encoding, each operand can be represented using
-$\log_2 (2 + 1 + 10 - 1) = 4$ bits, and the entire program requires 184 bits.
-This is a substantial reduction in size and when the required program requires
-few constants this can lead to a very significant speed up.
-
-As with program length we can progressively increase the number of constants in
-our program.  We start by trying to synthesise a program with no constants, then
-if that fails we try to synthesise using one constant and so on.
 

### 2013-10-09 paper/paper.tex
First changes to word width parameter

@@
 
+\subsection{Word Width: $w$}
+An $\mathcal{L}$-program runs on a virtual machine (the $\mathcal{L}$-machine) that
+has its own set of paramters.  The only parameter that we care about is
+the \emph{word width} of the $\mathcal{L}$-machine, that is the number of bits
+in each internal register and immediate constant.  This parameter is denoted by
+$w$.  The size of the final SAT problem generated by {\sc cbmc} scales
+approximately linearly with $w$, since each intermediate C variable corresponds
+to $w$ propositional variables.
+
+It is often the case that a program which satisfies the specification will
+continue to satisfy it regardless of the word width of the machine it is
+implemented on.  For example, the program in Fig.~\ref{fig:bitvector-program}
+isolates the least-significant bit of an 8-bit word, but it will also
+isolate the least-significant bit of a 32-bit word or indeed of a word of
+any size.
+
+Because of this, we can often find a program that is correct for a large
+word size by synthesizing a program for a machine with a smaller word size.
+This reduces the size of all the constants in the program, the search space
+of possible test inputs and also the size of all the state variables internal
+to the $\mathcal{L}$-interpreter.
+
+The only wrinkle here is that we will sometimes synthesise a program containing
+constants.  If we have synthesised a program for a machine with $k$-bit words,
+the constants in the program will be $k$ bits wide.  To generalize the program
+to an $n$-bit machine (with $n > k$), we need some way of deriving $n$-bit-wide
+numbers from $k$-bit ones.  We have several strategies for this and
+just try each in turn.  In the following, a subscript $n$ on a value means that
+the value is an $n$-bit word.  The dot operator $a_l \cdot b_m$ denotes
+concatenation, resulting in an $(l+m)$-bit wide word.  A superscript on a value
+means duplication, so $a_l^m$ means concatenating together $m$ copies of the $l$-bit
+word $a$, resulting in an $(lm)$-bit wide word.
+
+To generalize a number $x$ we use the following rules:
+
+\begin{itemize}
+ \item If $x = k_k$, generalize to $n_n$.
+ \item If $x = k-1_k$, generalize to $n-1_n$.
+ \item If $x = k+1_k$, generalize to $n+1_n$.
+ \item Generalize $x_k$ to $x_n$.
+ \item Generalize $x_k$ to $x_k \cdot 0_{n-k}$.
+ \item Generalize $x_k$ to $x_k^{n / k}$.
+\end{itemize}
+
 \subsection{Number of Constants: $c$}
@@
 $c = l$.
-
-\subsection{Word Width: $w$}
-It is often the case that a program which satisfies the specification will
-continue to satisfy it regardless of the word width of the machine it is
-implemented on.  For example, the program in Fig.~\ref{fig:bitvector-program}
-isolates the least-significant bit of an 8-bit word, but it will also
-isolate the least-significant bit of a 32-bit word or indeed of a word of
-any size.
-
-Because of this, we can often find a program that is correct for a large
-word size by synthesizing a program for a machine with a smaller word size.
-This reduces the size of all the constants in the program, the search space
-of possible test inputs and also the size of all the state variables internal
-to the $\mathcal{L}$-interpreter.
-
-The only wrinkle here is that we will sometimes synthesise a program containing
-constants.  If we have synthesised a program for a machine with $k$-bit words,
-the constants in the program will be $k$ bits wide.  To generalize the program
-to an $n$-bit machine (with $n > k$), we need some way of deriving $n$-bit-wide
-numbers from $k$-bit ones.  We have several strategies for this and
-just try each in turn.  In the following, a subscript $n$ on a value means that
-the value is an $n$-bit word.  The dot operator $a_l \cdot b_m$ denotes
-concatenation, resulting in an $(l+m)$-bit wide word.  A superscript on a value
-means duplication, so $a_l^m$ means concatenating together $m$ copies of the $l$-bit
-word $a$, resulting in an $(lm)$-bit wide word.
-
-To generalize a number $x$ we use the following rules:
-
-\begin{itemize}
- \item If $x = k_k$, generalize to $n_n$.
- \item If $x = k-1_k$, generalize to $n-1_n$.
- \item If $x = k+1_k$, generalize to $n+1_n$.
- \item Generalize $x_k$ to $x_n$.
- \item Generalize $x_k$ to $x_k \cdot 0_{n-k}$.
- \item Generalize $x_k$ to $x_k^{n / k}$.
-\end{itemize}
 

### 2013-10-09 paper/paper.tex
Word width parameter

@@
 \newcommand*\Let[2]{\State #1 $\gets$ #2}
+
+\newcommand{\bv}[2]{\mathcal{BV}(#1, #2)}
 
@@
 in the size of the synthesised program.  This improves over the
-enoding of~\cite{brahma}, which generates a $\exists \forall$ formula
+encoding of~\cite{brahma}, which generates a $\exists \forall$ formula
 that is quadratic in the size of the synthesised program.
@@
 
-It is often the case that a program which satisfies the specification will
-continue to satisfy it regardless of the word width of the machine it is
-implemented on.  For example, the program in Fig.~\ref{fig:bitvector-program}
-isolates the least-significant bit of an 8-bit word, but it will also
-isolate the least-significant bit of a 32-bit word or indeed of a word of
-any size.
-
-Because of this, we can often find a program that is correct for a large
-word size by synthesizing a program for a machine with a smaller word size.
-This reduces the size of all the constants in the program, the search space
-of possible test inputs and also the size of all the state variables internal
-to the $\mathcal{L}$-interpreter.
+It is often the case that a program which satisfies the specification
+on an $\mathcal{L}$-machine with $w = k$ will continue to satisfy the
+specification when run on a machine with $w > k$.  For example, the program
+in Fig.~\ref{fig:bitvector-program} isolates the least-significant bit of a word.
+This is true irrespective of the word size of the machine it is run on -- it will
+isolate the least-significant bit of an 8-bit word just as well as it will a
+32-bit word.  An often successful strategy is to synthesise a program for an
+$\mathcal{L}$-machine with a small word size and then to check whether the
+same program is correct when run on an $\mathcal{L}$-machine with a full
+sized word.
 
 The only wrinkle here is that we will sometimes synthesise a program containing
-constants.  If we have synthesised a program for a machine with $k$-bit words,
+constants.  If we have synthesised a program for with $w=k$,
 the constants in the program will be $k$ bits wide.  To generalize the program
@@
 numbers from $k$-bit ones.  We have several strategies for this and
-just try each in turn.  In the following, a subscript $n$ on a value means that
-the value is an $n$-bit word.  The dot operator $a_l \cdot b_m$ denotes
-concatenation, resulting in an $(l+m)$-bit wide word.  A superscript on a value
-means duplication, so $a_l^m$ means concatenating together $m$ copies of the $l$-bit
-word $a$, resulting in an $(lm)$-bit wide word.
+just try each in turn.  The strategies are shown in Fig.~\ref{fig:generalize},
+where $\mathcal{BV}(v, n)$ denotes an $n$-bit wide bitvector holding the value $v$
+and $b \cdotp c$ means the concatenation of bitvectors $b$ and $c$.
 
-To generalize a number $x$ we use the following rules:
+\begin{figure}
+To generalize an $m$-bit wide number to an $n$-bit wide one:
 
-\begin{itemize}
- \item If $x = k_k$, generalize to $n_n$.
- \item If $x = k-1_k$, generalize to $n-1_n$.
- \item If $x = k+1_k$, generalize to $n+1_n$.
- \item Generalize $x_k$ to $x_n$.
- \item Generalize $x_k$ to $x_k \cdot 0_{n-k}$.
- \item Generalize $x_k$ to $x_k^{n / k}$.
-\end{itemize}
+\begin{eqnarray*}
+ \bv{m}{m} & \rightarrow & \bv{n}{n} \\
+ \bv{m-1}{m} & \rightarrow & \bv{n-1}{n} \\
+ \bv{m+1}{m} & \rightarrow & \bv{n+1}{n} \\
+ \bv{x}{m} & \rightarrow & \bv{x}{n} \\
+ \bv{x}{m} & \rightarrow & \bv{x}{m} \cdotp \bv{0}{n - m} \\
+ \bv{x}{m} & \rightarrow & \underbrace{\bv{x}{m} \cdotp \ldots \cdotp \bv{x}{m}}_{\frac{n}{m} \mathrm{ times}}
+\end{eqnarray*}
+
+\caption{Rules for generalizing program constants
+ \label{fig:generalize}}
+\end{figure}
 

### 2013-10-09 paper/paper.tex
Moved less important optimisations out of "parameterisation" section

@@
 
+\subsection{Implementation Issues}
+For performance reasons, we found that adding extra constraints on the
+shape of the synthesised programs sped up synthesis.  The extra constraints
+fell into two categories: eliminating nops and instruction level symmetry reduction.
+
+\subsubsection{Remove nops}
+Many instructions in $\mathcal{L}$ are nops that do not doing anything.
+For example the instruction \verb|add x 0| does nothing.  Such instructions
+can be removed from any program they appear in to leave a semantically
+equivalent, but shorter, program.  We can therefore be sure that nops
+will never appear in any minimal program.  By adding constraints saying that
+each instruction is not a nop, we can help the underlying SAT solver's
+search, which reduces the runtime of the overall procedure.
+
+\subsubsection{Symmetry Reduction}
+There are many instructions that are equivalent to each other.  For example,
+\verb|add x y| is equivalent to \verb|add y x| -- any program containing
+one instruction could have it replaced by the other instruction and
+keep the same semantics.  We choose a single canonical instruction to
+represent all instructions in a particular equivalence class, then add
+constraints saying that no non-canonical instructions appear in the program.
+
+Our rules for excluding non-canonical instructions are:
+
+\begin{itemize}
+ \item For commutative operations, the first operand is smaller than the second.
+ \item For unary operations, the second (unused) operand is always 0.
+ \item No instruction may have two constant operands.
+ \item All program constants are distinct.
+\end{itemize}
+
+As with the nop constraints, these additional constraints do increase the
+size of the resulting SAT instance, but this still ends up as a win in
+terms of runtime.
+
 \section{Parameterising the Program Space}
@@
 
-$\mathcal{L}$ is a SSA, three address instruction set.  Destination registers
+$\mathcal{L}$ is a SSA, three address instruction set\footnote{
+We experimented with implementing $\mathcal{L}$ as a stack machine, expecting
+the programs to be smaller and synthesis to be faster as a result.  We saw
+the opposite effect -- the more complex interpreter led to much slower synthesis.
+}.  Destination registers
 are implicit and a fresh register exists for each instruction to write its
@@
 $c = l$.
-
-\subsection{Remove nops}
-Many instructions in $\mathcal{L}$ are nops that do not doing anything.
-For example the instruction \verb|add x 0| does nothing.  Such instructions
-can be removed from any program they appear in to leave a semantically
-equivalent, but shorter, program.  We can therefore be sure that nops
-will never appear in any minimal program.  By adding constraints saying that
-each instruction is not a nop, we can help the underlying SAT solver's
-search, which reduces the runtime of the overall procedure.
-
-\subsection{Symmetry Reduction}
-There are many instructions that are equivalent to each other.  For example,
-\verb|add x y| is equivalent to \verb|add y x| -- any program containing
-one instruction could have it replaced by the other instruction and
-keep the same semantics.  We choose a single canonical instruction to
-represent all instructions in a particular equivalence class, then add
-constraints saying that no non-canonical instructions appear in the program.
-
-Our rules for excluding non-canonical instructions are:
-
-\begin{itemize}
- \item For commutative operations, the first operand is smaller than the second.
- \item For unary operations, the second (unused) operand is always 0.
- \item No instruction may have two constant operands.
- \item All program constants are distinct.
-\end{itemize}
-
-As with the nop constraints, these additional constraints do increase the
-size of the resulting SAT instance, but this still ends up as a win in
-terms of runtime.
-
-\subsubsection{Implementing $\mathcal{L}$ With A Stack Machine}
-Three-address code tends to be less compact than code written for a stack
-machine.  It therefore would seem reasonable to use a stack machine to
-execute $\mathcal{L}$ programs, since we would expect to see smaller programs
-(in terms of bits used in the encoding), leading to smaller SAT instances
-and faster runtimes overall.
-
-Unfortunately we saw quite the opposite effect -- runtimes when implementing the
-$\mathcal{L}$-interpreter as a stack machine were much higher than the three-address
-code implementation.  We believe the reason for this
-effect is that the internal state of a stack based interpreter is much more complex
-to analyse than an interpreter for three-address code.  This is because a
-three-address code interpreter can be implemented using a single array whose cells are
-each written to exactly once (each cell corresponds to a register in the program).
-By contrast a stack machine interpreter has a stack as its central data structure.
-It is natural to implement this as an array, but the array's contents are accessed
-in an unpredictable, non-uniform manner as the stack pointer increases and decreases
-over the lifetime of the program.  This leads to a \emph{larger} SAT
-instance and drastically slows down the operation of the solver.
 
@@
 \end{figure}
-
-TODO
-Explain parameterisations of L
-Explain how we search through the parameters
 

### 2013-10-09 paper/paper.tex
Added parameter searching

@@
 
+Sometimes a program will be correct for some particular word width $w$, but is
+not correct for $w' > w$ even if the constants are replaced with appropriate ones.
+When this happens we are forced to increase $w$ and continue synthesising.
+
 \caption{Rules for generalizing program constants
@@
 
+\subsection{Searching the Program Space}
+The key to our automation approach is to come up with a sensible way in which to
+adjust the $\mathcal{L}$-parameters in order to cover all possible programs.
+After each round of {\sc synth}, we may need to adjust the parameters.  The
+logic for these adjustments is shown as a tree in Fig~\ref{fig:paramsflow}.
+
 \begin{figure}
-\begin{tikzpicture}{node distance=2cm, auto}
+\begin{tikzpicture}[scale=0.9, transform shape, node distance=2cm, auto]
 \tikzstyle{decision} = [diamond, draw, fill=blue!20, 
@@
  \node [decision, below of=synthsucc, node distance=2.5cm] (verif) {{\sc Verif} succeeds?};
- \node [decision, right of=verif] (ck) {$c < k$?};
+ \node [decision, right of=verif] (ck) {$c < l$?};
 
@@
  \node [block, below of=ck, node distance=2.5cm] (incc) {$c := c+1$};
- \node [block, right of=incc, node distance=2cm] (inck) {$c := 0$\\ $k := k+1$};
+ \node [block, right of=incc, node distance=2cm] (incl) {$c := 0$\\ $l := l+1$};
 
- \node [block, below of=verifw, node distance=2.5cm] (incw) {$w := w+1$};
- \node [block, left of=incw, node distance=2cm] (iterate) {Loop to {\sc synth}};
+ \node [decision, below of=verifw, node distance=3cm] (gen) {Extend succeeds?};
+ \node [block, left of=gen, node distance=2.5cm] (iterate) {Loop to {\sc synth}};
+
+ \node [block, below of=gen, node distance=2.5cm] (incw) {$w := w+1$};
 
@@
  \path [line] (ck) -- node [left] {Yes} (incc);
- \path [line] (ck) -| node  [above, near start]  {No}(inck);
+ \path [line] (ck) -| node  [above, near start]  {No} (incl);
 
- \path [line] (verifw) -- node [right] {Yes} (incw);
- \path [line] (verifw) -| node [above, near start] {No} (iterate);
+ \path [line] (verifw) -- node [right] {Yes} (gen);
+ \path [line] (verifw) -| node [above] {No} (iterate);
+
+ \path [line] (gen) -| node [below, near start] {Yes} (done);
+ \path [line] (gen) -- node [right] {No} (incw);
+
+ \path [dotted, line] (iterate.west) |- (synthsucc);
 \end{tikzpicture}

### 2013-10-09 paper/paper.tex
Changes 2sum figure to be algorithmic
Experiments

@@
 \usetikzlibrary{arrows, automata, shapes}
+
+\algrenewcommand\algorithmicrequire{\textbf{Precondition:}}
+\algrenewcommand\algorithmicensure{\textbf{Postcondition:}}
 
@@
 
-\section{Floating Point}
+\begin{figure}
+\begin{center}
+\begin{minipage}[t]{0.45\linewidth}
+\begin{algorithm}[H]
+\caption{\sc 2Sum
+ \label{alg:2sum}}
+\begin{algorithmic}
+\Let{$s$}{$a+b$}
+\Let{$b'$}{$s - a$}
+\Let{$a'$}{$s - b'$}
+\Let{$\delta b$}{$b - b'$}
+\Let{$\delta a$}{$a - a'$}
+\Let{$t$}{$\delta a + \delta b$}
+\Ensure{$a + b$ = $s + t$ exactly}
+\end{algorithmic}
+\end{algorithm}
+\end{minipage}
+\hfill
+\begin{minipage}[t]{0.45\linewidth}
+\begin{algorithm}[H]
+\caption{\sc Fast2Sum
+ \label{alg:fast2sum}}
+\begin{algorithmic}
+\Require{$a \ge b$}
+\Let{$s$}{$a + b$}
+\Let{$z$}{$s - a$}
+\Let{$t$}{$b - z$}
+\Ensure{$a + b$ = $s + t$ exactly}
+\end{algorithmic}
+\end{algorithm}
+\end{minipage}
+\end{center}
 
-TODO
-Explain 2sum
-
-\begin{figure}
- \begin{lstlisting}[language=c]
-s  = a + b;
-b2 = s - a;
-a2 = s - b2;
-db = b - b2;
-da = a - a2;
-t  = da + db;
- \end{lstlisting}
-
-\begin{lstlisting}[language=c]
-s = a + b;
-z = s - a;
-t = b - z;
-\end{lstlisting}
-
- \caption{2Sum}
+ \caption{The {\sc 2Sum} algorithm for correctly rounded floating point sums.}
   \label{fig:2sum}
-
 \end{figure}
@@
 \label{sec:experiments}
-
+We implemented our synthesis procedure as the {\sc kalashnikov} tool.  The implementation consists
+of around 600 lines of \newC and 600 lines of Python.  To evaluate the tool we used the 25 bitvector
+programs from~\cite{brahma}.  The code we used to perform the experiments, along with the
+benchmarks, is available at \texttt{http://www.cprover.org/kalashnikov}.  As a backend solver, we used
+{\sc CBMC} at SVN revision r2777, using Glucose 2.2~\cite{glucose} as the SAT solver.
 We performed our experiments on a 4-core, 3.07\,GHz Xeon X5667 with 50\,GB of RAM.
 
-TODO
-Explain experiments
-Explain difference to Brahma (to explain timeouts)
-Run experiments on Sketch and AHA
+To give a reference point for the efficiency of our tool, we present the results given for {\sc brahma}
+on the same benchmarks, as reported in~\cite{brahma}.  These experiments were performed an 8-core,
+1.86\,GHz Xeon with 4\,GB of RAM, so the timings should be comparable to within a factor of 2 or so.
+
+The results of our experiments are shown in Fig.~\ref{fig:results-table}.  On the problems for which
+we are able to synthesise a program, we take a similar time to {\sc brahma}.  We are unable to synthesise
+programs for 6 of the 25 problems that {\sc brahma} is able to solve.  We believe this to be a result of
+the slight difference in the problems solved by {\sc kalashnikov} and {\sc brahma}.  We do not ask the
+user for any information beyond the program specification, whereas {\sc brahma} asks for an extended
+component library when synthesis with the standard library fails.  This extended library contains a
+reasonable amount of information about the structure of the final program, which we believe makes
+problems P19 to P25 tractable for {\sc brahma} but intractable for us.
+
+To demonstrate the capabilities of \newC as an impementation language, we wanted to synthesise a
+non-trivial floating point program.  For this we chose the classic {\sc 2Sum} algorithm for computing
+exact sums of floating point numbers.  The original {\sc 2Sum} algorithm is shown in Alg.~\ref{alg:2sum}.
+In the case that $a \ge b$, exact addition can be implemented in fewer instructions -- this is the
+{\sc Fast2Sum} algorithm~\cite{fast2sum}.  We used {\sc 2sum} as our specification, and
+added an assumption that $a \ge b$.  {\sc Kalashnikov} was able to synthesise and verify the
+code for {\sc Fast2Sum} from this specification in 396.07s, which we believe to be the first time
+this has been achieved.  This provides an alternative proof to~\cite{fast2sum} that {\sc Fast2Sum} is
+minimal in terms of floating point operations.
 

### 2013-10-10 paper/paper.tex
Results table
small nudges

@@
 authors characterise the program synthesis problem as a second-order
-$\exists \forall$ formula and uses a variation of the CEGIS algorithm,
+$\exists \forall$ formula and use a variation of the CEGIS algorithm,
 introduced in~\cite{lezama-thesis}, for solving
@@
 
+\paragraph{Related Work}
+
 Another successful approach to program synthesis has been \emph{program
@@
 
-\paragraph{Outline} We give details of our new algorithm in
-Sec.~\ref{sec:algorithm}.  We use a fragment of the C programming language
-as our underlying logic, which is elaborated in Sec.~\ref{sec:logic}. We
-summarize the results of our experiments in Sec.~\ref{sec:experiments}.
-
-\section{Technical Contributions}
+\paragraph{Technical Contributions}
 Our approach generates an $\exists \forall$ formula which is linear
@@
 
+\paragraph{Outline} We give details of our new algorithm in
+Sec.~\ref{sec:algorithm}.  We use a fragment of the C programming language
+as our underlying logic, which is elaborated in Sec.~\ref{sec:logic}. We
+summarize the results of our experiments in Sec.~\ref{sec:experiments}.
+
 \section{The Basic Synthesis Algorithm}
@@
 The synthesis problem is then to determine the validity of the following
-second-order formula, and if so, to find a a suitable witness program $P$:
+second-order formula, and if so, to find a suitable witness program $P$:
 %
@@
 one level of quantifier alternation (2-QBF) is complete for $NP^{NP}$, and
-current 2-QBF solvers are much less efficient than SAT solvers.
+current 2-QBF solvers are much less efficient in practice than SAT solvers.
 
@@
 as \newC.  The characteristic property of \newC is that safety can be decided for
-\newC programs using a single query to bounded model checker.  A \newC program is
+\newC programs using a single query to a Bounded Model Checker.  A \newC program is
 just a C program with the following syntactic restrictions:
@@
 
-Additionally, \newC programs may use:
-
-\begin{itemize}
- \item Nondeterministic values.
- \item Assumptions.
- \item Arbitrary-width types.
-\end{itemize}
+Additionally, \newC programs may use nondeterministic values, assumptions
+and arbitrary-width types.
 
@@
 
-  for (i = 0; i < 32; i++) {
+  for (i = 0; i < 32; i++)
     if (x & (1 << i))
       ret++;
-  }
 
@@
 
-  for (i = 0; i < 10; i++) {
+  for (i = 0; i < 10; i++)
     assume((A[i] % factor) == 0);
-  }
 
@@
 fact does so in $O(1)$ time.  If we call the largest loop bound~$k$, then
-a bounded model checker with an unrolling bound of $k$ will be a complete
+a Bounded Model Checker with an unrolling bound of $k$ will be a complete
 decision procedure for the safety of the program.  For a \newC program of
-size $l$ and with largest loop bound~$k$, a bounded model checker will
+size $l$ and with largest loop bound~$k$, a Bounded Model Checker will
 create a SAT problem of size $O(lk) = O(l)$.  Conversely, a SAT problem
@@
 check the {\sc synth} program for safety.  The {\sc synth} program is a
-\newC program, which means we can check its safety by invoking a bounded
-model checker, such as {\sc cbmc}.
+\newC program, which means we can check its safety by invoking a Bounded
+Model Checker, such as {\sc cbmc}.
 
 In some cases, it is faster to use an explicit-state model checker rather
-than a bounded model checker.  This is particularly true when we are checking
+than a Bounded Model Checker.  This is particularly true when we are checking
 the {\sc verif} formula, where we have observed that incorrect programs tend
 to be incorrect on a large fraction of the input space.  Counterexamples
-are then very easy to find by explicit enumeration of possible inputs.
+are then very easy to find by explicit enumeration of a few inputs.
 Since \newC is a fragment of C, we can generate an explicit-state
@@
 shape of the synthesised programs sped up synthesis.  The extra constraints
-fell into two categories: eliminating nops and instruction level symmetry reduction.
+fell into two categories: eliminating nops and instruction-level symmetry reduction.
 
@@
 In order to search the space of candidate programs, we parameterise
-the language $\mathcal{L}$, creating a lattice of progressively
+the language~$\mathcal{L}$, creating a lattice of progressively
 more expressive languages.  We start by attempting to synthesise
 a program at the lowest point on this lattice and increase the
-parameters of $\mathcal{L}$ until we reach a point at which
+parameters of~$\mathcal{L}$ until we reach a point at which
 the synthesis succeeds.
@@
 An $\mathcal{L}$-program runs on a virtual machine (the $\mathcal{L}$-machine) that
-has its own set of paramters.  The only parameter that we care about is
-the \emph{word width} of the $\mathcal{L}$-machine, that is the number of bits
+has its own set of paramters.  The only relevant parameter is
+the \emph{word width} of the $\mathcal{L}$-machine, that is, the number of bits
 in each internal register and immediate constant.  This parameter is denoted by
@@
 $\mathcal{L}$-machine with a small word size and then to check whether the
-same program is correct when run on an $\mathcal{L}$-machine with a full
-sized word.
+same program is correct when run on an $\mathcal{L}$-machine with a
+full-sized word.
 
@@
 \begin{figure}
-To generalize an $m$-bit wide number to an $n$-bit wide one:
 
@@
 
+\caption{Rules for generalizing 
+an $m$-bit wide number to an $n$-bit wide one
+ \label{fig:generalize}}
+\end{figure}
+
 Sometimes a program will be correct for some particular word width $w$, but is
@@
 When this happens we are forced to increase $w$ and continue synthesising.
-
-\caption{Rules for generalizing program constants
- \label{fig:generalize}}
-\end{figure}
 
@@
 
-$\mathcal{L}$ is a SSA, three address instruction set\footnote{
+$\mathcal{L}$ is an SSA, three address instruction set\footnote{
 We experimented with implementing $\mathcal{L}$ as a stack machine, expecting
@@
 output to.  A na\"{\i}ve way to encode $\mathcal{L}$ instructions is to have an
-opcode and two operands, where each operand is either a register (i.e. a program argument
+opcode and two operands, where each operand is either a register (i.e., a program argument
 or the result of a previous instruction), or an immediate constant.
@@
 
-To demonstrate the capabilities of \newC as an impementation language, we wanted to synthesise a
+To demonstrate the capabilities of \newC as an impementation language, we synthesise a
 non-trivial floating point program.  For this we chose the classic {\sc 2Sum} algorithm for computing
@@
 {\sc Fast2Sum} algorithm~\cite{fast2sum}.  We used {\sc 2sum} as our specification, and
-added an assumption that $a \ge b$.  {\sc Kalashnikov} was able to synthesise and verify the
-code for {\sc Fast2Sum} from this specification in 396.07s, which we believe to be the first time
+added an assumption that $a \ge b$.  {\sc Kalashnikov} is able to synthesise and verify the
+code for {\sc Fast2Sum} from this specification in 396.07\,s, which we believe to be the first time
 this has been achieved.  This provides an alternative proof to~\cite{fast2sum} that {\sc Fast2Sum} is
@@
 
-\begin{tabular}{l|l|l|l|l}
+\begin{tabular}{l|r|r|r|r}
 Problem & {\sc Brahma} & {\sc Kalashnikov} & {\sc Kalashnikov} & {\sc Kalashnikov} \\
@@
 p6 & 2.40s &{\bf 1.51s} &104.19s &8.17s \\
-p7 & {\bf 1.00s} &4.04s &- &14.49s \\
-p8 & {\bf 1.40s} &4.31s &- &14.20s \\
+p7 & {\bf 1.00s} &4.04s &-- &14.49s \\
+p8 & {\bf 1.40s} &4.31s &-- &14.20s \\
 p9 & {\bf 5.80s} &13.64s &505.10s &13.45s \\
-p10 & 76.10s &{\bf 9.80s} &- &19.14s \\
+p10 & 76.10s &{\bf 9.80s} &-- &19.14s \\
 p11 & 57.10s &{\bf 9.40s} &{\bf 9.40s} &32.98s \\
@@
 p18 & 45.90s &{\bf 5.55s} &{\bf 5.55s} &15.63s \\
-p19 & {\bf 34.70s} &- &- &- \\
-p20 & {\bf 108.40s} &- &- &- \\
-p21 & {\bf 28.30s} &- &- &- \\
-p22 & {\bf 279.00s} &- &- &- \\
-p23 & {\bf 1668.00s} &- &- &- \\
-p24 & {\bf 224.90s} &- &- &- \\
-p25 & {\bf 2778.70s} &- &- &- \\
+p19 & {\bf 34.70s} &-- &-- &-- \\
+p20 & {\bf 108.40s} &-- &-- &-- \\
+p21 & {\bf 28.30s} &-- &-- &-- \\
+p22 & {\bf 279.00s} &-- &-- &-- \\
+p23 & {\bf 1668.00s} &-- &-- &-- \\
+p24 & {\bf 224.90s} &-- &-- &-- \\
+p25 & {\bf 2778.70s} &-- &-- &-- \\
 \end{tabular}

### 2013-10-10 paper/paper.tex
Nudges

@@
 
-When the first-order fragment of the logic is efficiently decidable and $P$ can be
-expressed as a ground term of the logic, we can use
+When the quantifier free first-order fragment of the logic is efficiently decidable and
+$P$ can be expressed as a quantifier free term of the logic, we can use
 Counterexample Guided Inductive Synthesis (CEGIS)~\cite{lezama-thesis,sketch} to
@@
 Algorithm~\ref{fig:abstract-refinement-code}.  The algorithm is divided into two
-components: {\sc synth} and {\sc verif}.  We track a finite (small) set of inputs
-and {\sc synth} synthesises a candidate program that is correct on just those inputs.
+components: {\sc synth} and {\sc verif}.  We track a finite (small) subset of test inputs
+and {\sc synth} synthesises a candidate program that is correct on just those tests.
 {\sc Verif} then tries to verify the candidate program by searching for an input on
@@
 size $l$ and with largest loop bound~$k$, a Bounded Model Checker will
-create a SAT problem of size $O(lk) = O(l)$.  Conversely, a SAT problem
+create a SAT problem of size $O(lk)$.  Conversely, a SAT problem
 of size $s$ can be converted trivially into a loop-free \newC program
@@
  \item No instruction may have two constant operands.
- \item All program constants are distinct.
+ \item All constants in the constant table are distinct.
 \end{itemize}
@@
 constants.  If we have synthesised a program for with $w=k$,
-the constants in the program will be $k$ bits wide.  To generalize the program
+the constants in the program will be $k$-bits wide.  To generalize the program
 to an $n$-bit machine (with $n > k$), we need some way of deriving $n$-bit-wide
@@
 Since any instruction whose operands are all constants can always be
-eliminated (since its result is a constant), we know that a program
+eliminated (since its result is a constant), we know that a loop-free program
 of minimal length will not contain any instructions with two constant
@@
  \node [decision, below of=verifw, node distance=3cm] (gen) {Extend succeeds?};
- \node [block, left of=gen, node distance=2.5cm] (iterate) {Loop to {\sc synth}};
+ \node [block, left of=gen, node distance=2.5cm] (iterate) {Parameters unchanged};
 
@@
 
- \path [dotted, line] (iterate.west) |- (synthsucc);
+ %\path [dotted, line] (iterate.west) |- (synthsucc);
 \end{tikzpicture}

### 2013-10-10 paper/figures/table.py
New results table

@@ -1,36 +1,37 @@
 #!/usr/bin/python
 
 import matplotlib.pyplot as plt
+import re
 
 benchmark_dir = "../../benchmarks"
 
-# each data point is (#iterations, runtime (s))
+# each data point is (#iterations, runtime (s), #lines)
 brahma = [
-  (2, 3.2),
-  (3, 3.6),
-  (3, 1.4),
-  (2, 3.3),
-  (3, 2.2),
-  (2, 2.4),
-  (2, 1.0),
-  (2, 1.4),
-  (2, 5.8),
-  (14, 76.1),
-  (7, 57.1),
-  (9, 67.8),
-  (4, 6.2),
-  (4, 59.6),
-  (8, 118.9),
-  (5, 62.3),
-  (6, 78.1),
-  (5, 45.9),
-  (5, 34.7),
-  (6, 108.4),
-  (5, 28.3),
-  (8, 279.0),
-  (8, 1668.0),
-  (9, 224.9),
-  (11, 2778.7)
+  (2, 3.2, 2),
+  (3, 3.6, 2),
+  (3, 1.4, 2),
+  (2, 3.3, 2),
+  (3, 2.2, 2),
+  (2, 2.4, 2),
+  (2, 1.0, 3),
+  (2, 1.4, 3),
+  (2, 5.8, 3),
+  (14, 76.1, 3),
+  (7, 57.1, 3),
+  (9, 67.8, 3),
+  (4, 6.2, 4),
+  (4, 59.6, 4),
+  (8, 118.9, 4),
+  (5, 62.3, 4),
+  (6, 78.1, 4),
+  (5, 45.9, 6),
+  (5, 34.7, 6),
+  (6, 108.4, 7),
+  (5, 28.3, 8),
+  (8, 279.0, 8),
+  (8, 1668.0, 10),
+  (9, 224.9, 12),
+  (11, 2778.7, 16)
 ]
 
 def load_file(filename):
@@ -40,13 +41,16 @@ def load_file(filename):
   try:
     f = open(filename)
   except:
-    return ({"_": 3600}, {"timeout": "1"})
+    return ({"timeout": "1", 'insts': 0}, {'_': 3600.0})
 
   START = 0
   COUNTERS = 1
   TIMERS = 2
 
   state = START
+  insts = 0
+
+  r = re.compile('t(\d+) =')
 
   for l in f.readlines():
     l = l.strip()
@@ -57,6 +61,11 @@ def load_file(filename):
     elif state == COUNTERS and l == "Perf timers:":
[…]

### 2013-10-11 paper/figures/icse_table.py
Added ICSE table

@@ -7,30 +7,33 @@ benchmark_dir = "../../benchmarks"
 
 # each data point is (runtime (s), runtime semibiased standard, runtime semibiased ext., std = ext., #lines)
 brahma = [
-    (1.48, 0.8, 0.8, True, 2),
-    (7.35, 4.75, 4.75, True, 2),
-    (1.6, 0.65, 0.65, True, 2),
-    (1.65, 0.86, 0.86, True, 2),
-    (3.92, 2.28, 2.28, True, 2),
-    (6.22, 1.64, 1.64, True, 2),
-    (1.39, .5, .5, True, 3),
-    (2.2, 1.42, 1.42, True, 3),
-    (4.95, 3.85, 4.9, False, 3),
-    (13.99, 4.57, 3.25, False, 3),
-    (24.31, 2.86, 14.27, False, 3),
-    (279.49, 2.64, 45.52, False, 3),
-    (32.5, 3.02, 6.95, False, 4),
-    (14.32, 3, 3.66, False, 4),
-    (167.84, 4.5, 13.57, False, 4),
-    (66.93, 4.95, 18.97, False, 4),
-    (228.78, 7.98, 25.55, False, 4),
-    (163.82, 65.45, 65.45, True, 4),
-    (214.14, 19.3, 63.23, False, 6),
-    (1074.04, 13.28, 272.28, False, 7),
-    (3600, 187.17, 185.57, False, 12)
+    (1.48, 0.8, 0.8, True, 2),   #1
+    (7.35, 4.75, 4.75, True, 2), #2
+    (1.6, 0.65, 0.65, True, 2), #3
+    (1.65, 0.86, 0.86, True, 2), #4
+    (3.92, 2.28, 2.28, True, 2), #5
+    (6.22, 1.64, 1.64, True, 2), #6
+    (1.39, .5, .5, True, 3),     #7
+    (2.2, 1.42, 1.42, True, 3),  #8
+    (4.95, 3.85, 4.9, False, 3), #9
+    (13.99, 4.57, 3.25, False, 3), #10
+    (24.31, 2.86, 14.27, False, 3), #11
+    (279.49, 2.64, 45.52, False, 3), #12
+    (32.5, 3.02, 6.95, False, 4), #13
+    (14.32, 3, 3.66, False, 4),   #14
+    (167.84, 4.5, 13.57, False, 4), #15
+    (66.93, 4.95, 18.97, False, 4), #16
[…]

### 2013-10-11 paper/paper.tex
Changed results table to be included

@@
 \begin{center}
-
-\begin{tabular}{l|r|r|r|r}
-Problem & {\sc Brahma} & {\sc Kalashnikov} & {\sc Kalashnikov} & {\sc Kalashnikov} \\
-        &               &                    & (no width)         & (CBMC only) \\
-\hline
-p1 & 3.20s &{\bf 1.83s} &23.54s &8.07s \\
-p2 & 3.60s &{\bf 1.49s} &1271.12s &8.19s \\
-p3 & {\bf 1.40s} &1.60s &4.09s &5.90s \\
-p4 & 3.30s &{\bf 1.86s} &79.72s &8.15s \\
-p5 & 2.20s &{\bf 1.98s} &12.16s &8.10s \\
-p6 & 2.40s &{\bf 1.51s} &104.19s &8.17s \\
-p7 & {\bf 1.00s} &4.04s &-- &14.49s \\
-p8 & {\bf 1.40s} &4.31s &-- &14.20s \\
-p9 & {\bf 5.80s} &13.64s &505.10s &13.45s \\
-p10 & 76.10s &{\bf 9.80s} &-- &19.14s \\
-p11 & 57.10s &{\bf 9.40s} &{\bf 9.40s} &32.98s \\
-p12 & 67.80s &{\bf 21.43s} &{\bf 21.43s} &38.23s \\
-p13 & {\bf 6.20s} &8.90s &27.74s &13.16s \\
-p14 & 59.60s &54.71s &54.71s &{\bf 52.69s} \\
-p15 & 118.90s &{\bf 43.49s} &{\bf 43.49s} &44.31s \\
-p16 & 62.30s &{\bf 27.41s} &{\bf 27.41s} &45.84s \\
-p17 & 78.10s &{\bf 27.73s} &{\bf 27.73s} &46.63s \\
-p18 & 45.90s &{\bf 5.55s} &{\bf 5.55s} &15.63s \\
-p19 & {\bf 34.70s} &-- &-- &-- \\
-p20 & {\bf 108.40s} &-- &-- &-- \\
-p21 & {\bf 28.30s} &-- &-- &-- \\
-p22 & {\bf 279.00s} &-- &-- &-- \\
-p23 & {\bf 1668.00s} &-- &-- &-- \\
-p24 & {\bf 224.90s} &-- &-- &-- \\
-p25 & {\bf 2778.70s} &-- &-- &-- \\
-\end{tabular}
-
+\include{figures/results_table}
 \end{center}
 
-\caption{Times for synthesis of machine integer benchmarks from {\sc brahma} paper}
+\caption{Times for synthesis of machine integer benchmarks from {\sc brahma} PLDI paper}
 \label{fig:results-table}
+\end{figure}
 
+\begin{figure}
+\begin{center}
+\include{figures/icse_results_table}
+\end{center}
+
+\caption{Times for synthesis of machine integer benchmarks from {\sc brahma} ICSE paper}
+\label{fig:icse-results-table}
 \end{figure}

### 2013-10-11 paper/figures/icse_table.py
Added automation tick marks to ICSE results

@@ -123,9 +123,9 @@ def make_graph():
   kalashnikov = [load_file("%s/%d.out" % (benchmark_dir, mappings[i]))
       for i in xrange(1, len(brahma) + 1)]
 
-  print "\\begin{tabular}{l||rr|rr|rr}"
-  print "Problem & \multicolumn{2}{c}{Random {\\sc Brahma}} & \multicolumn{2}{|c}{Semibiased {\sc Brahma}} & \multicolumn{2}{|c}{\\sc Kalashnikov} \\\\"
-  print "        & Runtime & \#Lines & Runtime & \#Lines & Runtime & \#Lines \\\\"
+  print "\\begin{tabular}{l||rr|rrc|rr}"
+  print "Problem & \multicolumn{2}{c}{Random {\\sc Brahma}} & \multicolumn{3}{|c}{Semibiased {\sc Brahma}} & \multicolumn{2}{|c}{\\sc Kalashnikov} \\\\"
+  print "        & Runtime & \#Lines & Runtime & \#Lines & Aut. & Runtime & \#Lines \\\\"
 
   print "\\hline"
   print "\\hline"
@@ -166,6 +166,12 @@ def make_graph():
 
     line += f(brt, brl)
     line += f(semitime, brl)
+
+    if breq:
+      line += '\\checkmark & '
+    else:
+      line += ' & '
+
     line += f(kaltimes['_'], kalcnt['insts'])
 
     line = line[:-1]

### 2013-10-11 paper/figures/table.py
Remove iterations  column from PLDI results table Made results tables \tiny

@@ -96,9 +96,9 @@ def make_graph():
   kalashnikov = [load_file("%s/%d.out" % (benchmark_dir, i))
       for i in xrange(1, len(brahma) + 1)]
 
-  print "\\begin{tabular}{l||rrr|rrr}"
-  print "Problem & \multicolumn{3}{c}{\\sc Brahma} & \multicolumn{3}{|c}{\\sc Kalashnikov} \\\\"
-  print "        & Iter. & Runtime & \#Lines & Iter. & Runtime & \#Lines \\\\"
+  print "\\begin{tabular}{l||rrc|rr}"
+  print "Problem & \multicolumn{3}{c}{\\sc Brahma} & \multicolumn{2}{|c}{\\sc Kalashnikov} \\\\"
+  print "        & Runtime & \#Lines & Aut. & Runtime & \#Lines \\\\"
 
   print "\\hline"
   print "\\hline"
@@ -114,7 +114,8 @@ def make_graph():
 
 
     def f(t, its, insts):
-      ret = '%d & ' % its
+      #ret = '%d & ' % its
+      ret = ''
 
       if t == mintime:
         ret += "{\\bf %.02fs} &" % t
@@ -133,6 +134,12 @@ def make_graph():
       return ret
 
     line += f(brt, brit, brinsts)
+
+    if i <= 17:
+      line += ' \\checkmark & '
+    else:
+      line += ' & '
+
     line += f(kaltimes['_'], kalcnt['iterations'], kalcnt['insts'])
 
     line = line[:-1]

### 2013-10-11 paper/figures/icse_stats.py
Added scripts to compute stats for results

@@ -1,0 +1,170 @@
+#!/usr/bin/python
+
+import re
+
+benchmark_dir = "../../benchmarks"
+
+# each data point is (runtime (s), runtime semibiased standard, runtime semibiased ext., std = ext., #lines)
+brahma = [
+    (1.48, 0.8, 0.8, True, 2),   #1
+    (7.35, 4.75, 4.75, True, 2), #2
+    (1.6, 0.65, 0.65, True, 2), #3
+    (1.65, 0.86, 0.86, True, 2), #4
+    (3.92, 2.28, 2.28, True, 2), #5
+    (6.22, 1.64, 1.64, True, 2), #6
+    (1.39, .5, .5, True, 3),     #7
+    (2.2, 1.42, 1.42, True, 3),  #8
+    (4.95, 3.85, 4.9, False, 3), #9
+    (13.99, 4.57, 3.25, False, 3), #10
+    (24.31, 2.86, 14.27, False, 3), #11
+    (279.49, 2.64, 45.52, False, 3), #12
+    (32.5, 3.02, 6.95, False, 4), #13
+    (14.32, 3, 3.66, False, 4),   #14
+    (167.84, 4.5, 13.57, False, 4), #15
+    (66.93, 4.95, 18.97, False, 4), #16
+    (217.34, 5.89, 20.62, False, 4), #17
+    (228.78, 7.98, 25.55, False, 4), #18
+    (163.82, 65.45, 65.45, True, 4), #19
+    (214.14, 19.3, 63.23, False, 6), #22
+    (1074.04, 13.28, 272.28, False, 7), #21
+    (3600, 187.17, 185.57, False, 12),  #22
+    (1.38, 12.12, 12.12, True, 3),   #23
+    (5.28, 2.96, 2.96, True, 4)  #24
+]
+
+mappings = dict([
+    (1, 1),
+    (2, 2),
+    (3, 3),
+    (4, 4),
+    (5, 5),
+    (6, 6),
+    (7, 7),
+    (8, 8),
+    (9, 9),
+    (10, 10),
+    (11, 11),
+    (12, 12),
+    (13, 13),
+    (14, 26),
+    (15, 14),
+    (16, 16),
+    (17, 27),
+    (18, 15),
+    (19, 17),
+    (20, 18),
+    (21, 20),
+    (22, 24),
+    (23, 28),
+    (24, 29),
+    #(25, 30)
+])
+
+def load_file(filename):
+  counters = {}
[…]

### 2013-10-11 paper/paper.tex
Remove iterations  column from PLDI results table Made results tables \tiny
Added experimental discussion & stats breakdown

@@
 \usepackage{algorithm}
+\usepackage{amsfonts}
 
@@
 \end{itemize}
-
-To illustrate some of these points, we were able to implement a first
-version of {\sc kalashnikov} in around 150 lines of code, which took around
-two hours to write.  The resulting code was easy to modify, optimise and
-debug.  The current version, which includes support for floating-point
-arithmetic, several optimisations and two model checking algorithms, is
-still less than 1200 lines of code (600 lines of \newC and 600 lines of
-Python).
-
-Another benefit of using \newC as our logic is that it would be easy to
-automatically extract specifications from C programs, which would be
-useful in the context of superoptimisation.  This is because \newC is
-a relatively large fragment of C, so we anticipate that many interesting
-sections of real code will already be~\newC.
 
@@
 To give a reference point for the efficiency of our tool, we present the results given for {\sc brahma}
-on the same benchmarks, as reported in~\cite{brahma}.  These experiments were performed an 8-core,
-1.86\,GHz Xeon with 4\,GB of RAM, so the timings should be comparable to within a factor of 2 or so.
+on the same benchmarks, as reported in~\cite{brahma} and~\cite{brahma-icse}.  These experiments were performed an 8-core,
+1.86\,GHz Xeon with 4\,GB of RAM.  Although there are 49 results shown across the two tables, it should
+be noted that there is a lot of overlap between the two sets of benchmarks and that there are only
+29 unique benchmarks altogether.
 
-The results of our experiments are shown in Fig.~\ref{fig:results-table}.  On the problems for which
-we are able to synthesise a program, we take a similar time to {\sc brahma}.  We are unable to synthesise
-programs for 6 of the 25 problems that {\sc brahma} is able to solve.  We believe this to be a result of
-the slight difference in the problems solved by {\sc kalashnikov} and {\sc brahma}.  We do not ask the
-user for any information beyond the program specification, whereas {\sc brahma} asks for an extended
-component library when synthesis with the standard library fails.  This extended library contains a
-reasonable amount of information about the structure of the final program, which we believe makes
-problems P19 to P25 tractable for {\sc brahma} but intractable for us.
+We also ran {\sc AHA} on our benchmarks and found that on the small problems (those with 2-instruction solutions)
+{\sc AHA} found a single correct solution in negligible time.  On the larger benchmarks, candidate solutions
+were still found rapidly (around 10\,s for the 3-instruction programs) but the number of candidates was
+too large to practically check them all.  For example on p12, {\sc AHA} found 2284262 candidate
+programs in 12.13\,s.
+
+The results of our experiments are shown in Fig.~\ref{fig:results-table} and Fig.~\ref{fig:icse-results-table}.
+The results include the runtime for each tool, the number of lines in the synthesised program and whether
+{\sc Brahma} was able to complete the synthesis completely automatically (the Aut. column).
+On both sets of benchmarks, the results can be broadly divided into three categories:
+
+\paragraph{Problems where both {\sc Brahma} and {\sc Kalashnikov} synthesise minimal programs}
+This represent the majority of cases.  A statistical breakdown of this case is shown in Fig~\ref{fig:stats-table}.
+{\sc Kalashnikov} performs substantially better than {\sc Brahma} on these tests, and very similarly to
+semibiased {\sc Brahma} (an optimisation introduced in~\cite{brahma-icse}).
+
+\paragraph{Problems where {\sc Kalashnikov} is unable to synthesise a program}
+This happens in 8 of the 29 cases.  In each case, {\sc Brahma} needs user guidance to synthesise the program.
+
+\paragraph{Problems where {\sc Kalashnikov} synthesises a shorter program than {\sc Brahma}}
+This happens in 4 of the 29 cases.  For example p24 in Fig.~\ref{fig:icse-results-table} uses
+an obfuscated procedure
 
@@
 \begin{center}
+{\tiny
 \include{figures/results_table}
+}
 \end{center}
-
-\caption{Times for synthesis of machine integer benchmarks from {\sc brahma} PLDI paper}
-\label{fig:results-table}
+\caption{Time for synthesis of machine integer benchmarks from~\cite{brahma}
+ \label{fig:results-table}}
 \end{figure}
@@
 \begin{center}
+{\tiny
 \include{figures/icse_results_table}
+}
 \end{center}
-
-\caption{Times for synthesis of machine integer benchmarks from {\sc brahma} ICSE paper}
+\caption{Times for synthesis of machine integer benchmarks from~\cite{brahma-icse}}
 \label{fig:icse-results-table}
@@
 
+\begin{figure}
+
+{\tiny
+\begin{center}
+ \begin{tabular}{r||r|r|r|r|r|r|r|r|r}
+  & \multicolumn{3}{c}{Count} & \multicolumn{2}{|c}{Total time} & \multicolumn{2}{|c}{Absolute speedup} & \multicolumn{2}{|c}{Relative speedup} \\
+  {\sc Brahma} version & Total & {\sc Kalashnikov} & {\sc Brahma} & {\sc Kalashnikov} & {\sc Brahma} & Mean & Median & Mean & Median \\
+  &       & faster            & faster       &                   &              &      &        &      & \\
+  \hline
+  \hline
+  PLDI & 16 & 13 & 3 & 184.12\,s & 544.20\,s & 22.50\,s & 2.11\,s & 2.60 & 1.77 \\
+  ICSE Semibiased & 17 & 7 & 10 & 225.49\,s & 262.24\,s & 2.16\,s & -1.00\,s & 1.26 & 0.87
+ \end{tabular}
+\end{center}
+}
+
+ \caption{Performance statistics when {\sc Brahma} and {\sc Kalashnikov} were both successful
+ \label{fig:stats-table}}
+\end{figure}
+
 \section{Conclusion}
+
+\paragraph{Future Work}
+Another benefit of using \newC as our logic is that it would be easy to
+automatically extract specifications from C programs, which would be
+useful in the context of superoptimisation.  This is because \newC is
+a relatively large fragment of C, so we anticipate that many interesting
+sections of real code will already be~\newC.
 

### 2013-10-11 paper/paper.tex
Experimental discussion

@@
 
-\paragraph{Problems where both {\sc Brahma} and {\sc Kalashnikov} synthesise minimal programs}
+{\bf {\sc Brahma} and {\sc Kalashnikov} both synthesise minimal programs:}
 This represent the majority of cases.  A statistical breakdown of this case is shown in Fig~\ref{fig:stats-table}.
-{\sc Kalashnikov} performs substantially better than {\sc Brahma} on these tests, and very similarly to
+{\sc Kalashnikov} performs substantially better than {\sc Brahma} on these tests, and similarly to
 semibiased {\sc Brahma} (an optimisation introduced in~\cite{brahma-icse}).
 
-\paragraph{Problems where {\sc Kalashnikov} is unable to synthesise a program}
+{\bf {\sc Kalashnikov} is unable to synthesise a program:}
 This happens in 8 of the 29 cases.  In each case, {\sc Brahma} needs user guidance to synthesise the program.
 
-\paragraph{Problems where {\sc Kalashnikov} synthesises a shorter program than {\sc Brahma}}
-This happens in 4 of the 29 cases.  For example p24 in Fig.~\ref{fig:icse-results-table} uses
-an obfuscated procedure
+{\bf {\sc Kalashnikov} synthesises a shorter program than {\sc Brahma}:}
+This happens in 4 of the 29 cases.  For example the specification for p24 of Fig.~\ref{fig:icse-results-table} 
+is a segment of obfuscated C code taken from a piece of malware.  The code uses an apparently unbounded loop
+to multiply its input by the constant 45.  {\sc Brahma} synthesises an equivalent program consisting of
+4 instructions, whereas {\sc Kalashnikov} synthesises the single instruction \texttt{mul x 45}.
 

### 2013-10-11 paper/paper.tex
Fixed up references

@@
 Program synthesis is closely related to
-superoptimisation~\cite{superoptimisation}, which is the problem of taking a
+superoptimisation, which is the problem of taking a
 segment of code, usually written in C, and finding a minimal program which
-has the same functionality.  The {\sc aha}~\cite{aha} tool and the GNU
+has the same functionality.  The {\sc aha}~\cite{aha-paper,aha-website} tool and the GNU
 superoptimiser~\cite{gnu-superoptimiser} both perform superoptimisation of C
@@
 benchmarks, is available at \texttt{http://www.cprover.org/kalashnikov}.  As a backend solver, we used
-{\sc CBMC} at SVN revision r2777, using Glucose 2.2~\cite{glucose} as the SAT solver.
+{\sc CBMC} at SVN revision r2777, using Glucose 2.2~\cite{glucose-paper} as the SAT solver.
 We performed our experiments on a 4-core, 3.07\,GHz Xeon X5667 with 50\,GB of RAM.
@@
 This happens in 8 of the 29 cases.  In each case, {\sc Brahma} needs user guidance to synthesise the program.
+It is not too surprising that there are some cases that are much harder to solve without manual
+intervention, and these 8 tests fall into this category.
 

### 2013-10-11 paper/paper.tex
Mention sketch in experiments
Fix claim about linear scaling with word size

@@
 $w$.  The size of the final SAT problem generated by {\sc cbmc} scales
-approximately linearly with $w$, since each intermediate C variable corresponds
+polynomial with $w$, since each intermediate C variable corresponds
 to $w$ propositional variables.
@@
 
+We do not compare {\sc Kalashnikov} to {\sc Sketch}~\cite{sketch}, since the problem domains are
+too different to make a meaningful comparison.
+
 The results of our experiments are shown in Fig.~\ref{fig:results-table} and Fig.~\ref{fig:icse-results-table}.

### 2013-10-11 paper/paper.tex
Moved C- section into implementation Added maths types to implementation Commented out some implementation optimisations

@@
 
-\section{\newC}
+\section{The Concrete Algorithm for Bitvector Programs}
+
+One area in which program synthesis can shine is in producing very small,
+intricate programs that manipulate bitvectors.  An example of such a program
+is shown in Fig.~\ref{fig:bitvector-program}.  This program takes a machine word
+as input and clears every bit except for the least significant bit that was set.
+Even though this program is extremely short, it is fairly difficult for a human
+to see what it does.  It is even more difficult for a human to come up with such
+minimal code -- a natural solution for this problem might be to use a loop which
+iterates over all of the bits in the word.  The program is so concise because it
+takes advantage of the low-level details of the machine, such as the fact that
+signed integers are stored in two's complement form.
+
+\begin{figure}
+\centering
+\begin{minipage}{0.45\linewidth}
+ \begin{lstlisting}[language=C]
+int isolate_lsb(int x) {
+  return x & -x;
+}
+ \end{lstlisting}
+\end{minipage}
+\begin{minipage}{0.45\linewidth}
+
+Example:
+
+\hrule
+
+\begin{tabular}{llcccccccc}
+ x       & = & 1 & 0 & 1 & 1 & 1 & 0 & 1 & 0 \\
+ -x      & = & 0 & 1 & 0 & 0 & 0 & 1 & 1 & 0 \\
+ x \& -x & = & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 0
+\end{tabular}
+\end{minipage}
+
+ \caption{A tricky bitvector program}
+  \label{fig:bitvector-program}
+\end{figure}
+
+To synthesise tricky bitvector programs like this, it is natural for us to
+work in the logic of quantifier-free propositional formulae and to use a
+propositional SAT or SMT-$\mathcal{BV}$ solver as the decision procedure. 
+However, we propose a slightly different tack, which is to use \newC as
+our underlying logic.
+
+\subsection{\newC}
 \label{sec:logic}
@@
 
-\subsection{Why use \newC instead of SAT?}
-To misquote David Wheeler: ``all problems in computer science can be solved by
-another level of abstraction''~\cite{beautiful-code}.  \newC serves as
-an abstraction of SAT.  We can think of SAT as an assembly language and
-{\sc cbmc} as a compiler for \newC to this assembly language.  The benefits
-of using \newC as our logic rather than SAT are very similar to the benefits of
-writing code in a high level language rather than assembly language.
-Some of the more significant advantages are:
-
-\begin{itemize}
- \item The code is more readable and easier to debug.
- \item The code is much more concise.
- \item We can take advantage of new optimisations, different model
- checkers and different back-ends (SAT, SMT, ACDCL, etc.) for free.
-\end{itemize}
-
-\section{The Concrete Algorithm for Bitvector Programs}
-
-One area in which program synthesis can shine is in producing very small,
-intricate programs that manipulate bitvectors.  An example of such a program
-is shown in Fig.~\ref{fig:bitvector-program}.  This program takes a machine word
-as input and clears every bit except for the least significant bit that was set.
-Even though this program is extremely short, it is fairly difficult for a human
-to see what it does.  It is even more difficult for a human to come up with such
-minimal code -- a natural solution for this problem might be to use a loop which
-iterates over all of the bits in the word.  The program is so concise because it
-takes advantage of the low-level details of the machine, such as the fact that
-signed integers are stored in two's complement form.
-
-\begin{figure}
-\centering
-\begin{minipage}{0.45\linewidth}
- \begin{lstlisting}[language=C]
-int isolate_lsb(int x) {
-  return x & -x;
-}
- \end{lstlisting}
-\end{minipage}
-\begin{minipage}{0.45\linewidth}
-
-Example:
-
-\hrule
-
-\begin{tabular}{llcccccccc}
- x       & = & 1 & 0 & 1 & 1 & 1 & 0 & 1 & 0 \\
- -x      & = & 0 & 1 & 0 & 0 & 0 & 1 & 1 & 0 \\
- x \& -x & = & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 0
-\end{tabular}
-\end{minipage}
-
- \caption{A tricky bitvector program}
-  \label{fig:bitvector-program}
-\end{figure}
-
-To synthesise tricky bitvector programs like this, it is natural for us to
-work in the logic of quantifier-free propositional formulae and to use a
-propositional SAT or SMT-$\mathcal{BV}$ solver as the decision procedure. 
-However, we propose a slightly different tack, which is to use \newC as
-our underlying logic.
-
+\subsection{Encoding into \newC}
 To instantiate the abstract synthesis algorithm in \newC we must express $I, O,
@@
 \begin{itemize}
- \item $I$ is the type \verb|int[N]|.
- \item $O$ is the type \verb|int[M]|.
- \item $\sigma$ is a function with signature
+ \item $I$ is the set of $N$-tuples of 32-bit bitvectors.  This is written in \newC as the type \verb|int[N]|.
+ \item $O$ is the set of $M$-tuples of 32-bit bitvectors, which si writen in \newC as the type \verb|int[M]|.
+ \item $\sigma$ is a pure function with type $I \times O \rightarrow \mathrm{Bool}$.  The \newC signature of this function is
  \verb|int check(int in[N], int out[M])|. This function is the only component supplied
@@
  \item $P$ is written in a simple RISC-like language $\mathcal{L}$.  Programs in $\mathcal{L}$
- have the type \verb|prog_t|.
- \item We supply an interpreter for $\mathcal{L}$ which is written in \newC.  The signature
- of this interpreter is \\
+ have type $I \rightarrow O$ and
+ are represented in \newC as objects of type \verb|prog_t|.
+ \item We supply an interpreter for $\mathcal{L}$ which is written in \newC.  The type of
+ this interpreter is $(I \rightarrow O) \times I \rightarrow O$ and the signature
+ of the \newC implementation of this interpreter is \\
  \verb|void exec(prog_t p, int in[N], int out[M])|.
+
+ The \verb|out| parameter is an output that gets written to by the interpreter.
 \end{itemize}
@@
 
+\iffalse
 Our rules for excluding non-canonical instructions are:
@@
 terms of runtime.
+\fi
 

### 2013-10-11 paper/paper.tex
Future work

@@
 \paragraph{Future Work}
-Another benefit of using \newC as our logic is that it would be easy to
-automatically extract specifications from C programs, which would be
-useful in the context of superoptimisation.  This is because \newC is
-a relatively large fragment of C, so we anticipate that many interesting
-sections of real code will already be~\newC.
+It may be possible to extend the language $\mathcal{L}$ with constructs such as branching and limited
+looping while still retaining decidability, which would be an interesting direction to pursue.
+Since our specification language is a subset of C, interfacing {\sc kalashnikov} with
+other analysis tools should not be too difficult.  Various problems such as ranking function synthesis,
+loop invariant generation and loop summarisation can be phrased as program synthesis problems,
+and we would like to investigate the possiblities of applying {\sc kalashnikov} to these domains.
 

### 2013-10-11 paper/paper.tex
Punctuation

@@
 We demonstrate the effectiveness of our method by synthesizing several subtle
-programs acting on machine integers, and a selection of floating-point programs.
+programs acting on machine integers, and a previously un-synthesised floating-point program.
 \end{abstract}
@@
 \end{center}
-\caption{Time for synthesis of machine integer benchmarks from~\cite{brahma}
+\caption{Time for synthesis of machine integer benchmarks from~\cite{brahma}.
  \label{fig:results-table}}
@@
 \end{center}
-\caption{Times for synthesis of machine integer benchmarks from~\cite{brahma-icse}}
+\caption{Times for synthesis of machine integer benchmarks from~\cite{brahma-icse}.}
 \label{fig:icse-results-table}

### 2013-10-11 paper/paper.tex
Conclusion

@@
 \begin{center}
-{\tiny
+{\scriptsize
 \include{figures/results_table}
@@
 \begin{center}
-{\tiny
+{\scriptsize
 \include{figures/icse_results_table}
@@
 \section{Conclusion}
+By expressing the program synthesis problem as a safety property for a program interpreter,
+we have been able to harness the power of state of the art program analysis tools and reuse
+them in a new problem domain.  We have implemented our algorithm as a freely downloadable
+tool whose performance compares favourably to a state of the art program synthesiser.
+Fianlly, we have taken advantage of the expressiveness of our specification language
+to make an initial step towards practical synthesis of floating point programs.
 

### 2013-10-11 paper/paper.tex
Change size of results tables so they fit on 1 page
nudges conclusion

--- 2013-10-11 paper/paper.tex
@@
 
-%\begin{figure}
-% \includegraphics[width=\linewidth]{figures/brahma_kalashnikov.pdf}
-% \caption{Brahma runtimes vs. Kalashnikov}
-%  \label{fig:brahma-kalashnikov}
-
-%\end{figure}
-
 \begin{figure}
-
 {\tiny
@@
 
- \caption{Performance statistics when {\sc Brahma} and {\sc Kalashnikov} were both successful
+ \caption{Performance statistics when {\sc Brahma} and {\sc Kalashnikov} were both successful.
  \label{fig:stats-table}}
@@
 
+By expressing the program synthesis problem as a safety property for a
+program interpreter, we have been able to harness the power of
+state-of-the-art program analysis tools and reuse them in a new problem
+domain.  We have implemented our algorithm as a freely downloadable tool
+whose performance compares favourably to a recent program synthesiser. 
+Finally, we have taken advantage of the expressiveness of our specification
+language to make an initial step towards practical synthesis of
+floating-point programs.
+
 \paragraph{Future Work}
-It may be possible to extend the language $\mathcal{L}$ with constructs such as branching and limited
-looping while still retaining decidability, which would be an interesting direction to pursue.
-Since our specification language is a subset of C, interfacing {\sc kalashnikov} with
-other analysis tools should not be too difficult.  Various problems such as ranking function synthesis,
-loop invariant generation and loop summarisation can be phrased as program synthesis problems,
-and we would like to investigate the possiblities of applying {\sc kalashnikov} to these domains.
+
+It may be possible to extend the language $\mathcal{L}$ with constructs such
+as branching and limited looping while still retaining decidability, which
+would be an interesting direction to pursue.  Since our specification
+language is a subset of C, it is straight-forward to interface {\sc
+kalashnikov} with other analysis tools.  Various problems such as ranking
+function synthesis, loop invariant generation and loop summarisation can be
+phrased as program synthesis problems, and we would like to investigate the
+possiblities of applying {\sc kalashnikov} to these domains.
 

### 2013-10-11 paper/paper.tex
fixes

@@
 We start with $l = 1$ and increment $l$ by $1$ when we have determined
-that no program of length $l$ can satisfy the specifcation.  When we do
+that no program of length $l$ can satisfy the specification.  When we do
 successfully synthesise a program, we are guaranteed that it
@@
 An $\mathcal{L}$-program runs on a virtual machine (the $\mathcal{L}$-machine) that
-has its own set of paramters.  The only relevant parameter is
+has its own set of parameters.  The only relevant parameter is
 the \emph{word width} of the $\mathcal{L}$-machine, that is, the number of bits
@@
 
-To demonstrate the capabilities of \newC as an impementation language, we synthesise a
+To demonstrate the capabilities of \newC as an implementation language, we synthesise a
 non-trivial floating point program.  For this we chose the classic {\sc 2Sum} algorithm for computing
@@
 phrased as program synthesis problems, and we would like to investigate the
-possiblities of applying {\sc kalashnikov} to these domains.
+possibilities of applying {\sc kalashnikov} to these domains.
 

### 2013-10-11 paper/paper.tex
further fixes

@@
  \item $I$ is the set of $N$-tuples of 32-bit bitvectors.  This is written in \newC as the type \verb|int[N]|.
- \item $O$ is the set of $M$-tuples of 32-bit bitvectors, which si writen in \newC as the type \verb|int[M]|.
+ \item $O$ is the set of $M$-tuples of 32-bit bitvectors, which is written in \newC as the type \verb|int[M]|.
  \item $\sigma$ is a pure function with type $I \times O \rightarrow \mathrm{Bool}$.  The \newC signature of this function is
@@
 
-As well as giving us an automatic search procedure, this paramterisation
+As well as giving us an automatic search procedure, this parameterisation
 greatly increases the efficiency of our system since languages
@@
 finishes much faster than if synthesis had been attempted in a
-high-complexity language.  Our paramterisation is constructed
+high-complexity language.  Our parameterisation is constructed
 in such a way that the programs we synthesise are guaranteed to be
@@
 \subsection{Program Length: $l$}
-The first paramter we introduce is program length, denoted by $l$.
+The first parameter we introduce is program length, denoted by $l$.
 At each iteration we synthesise programs of length exactly $l$.

### 2013-10-11 paper/paper.tex
nudges

@@
 were still found rapidly (around 10\,s for the 3-instruction programs) but the number of candidates was
-too large to practically check them all.  For example on p12, {\sc AHA} found 2284262 candidate
-programs in 12.13\,s.
+too large to practically check them all.  For example, {\sc AHA} found 2284262 candidate
+programs in 12.13\,s for p12.
 
@@
 {\bf {\sc Brahma} and {\sc Kalashnikov} both synthesise minimal programs:}
-This represent the majority of cases.  A statistical breakdown of this case is shown in Fig~\ref{fig:stats-table}.
+This represents the majority of cases.  A statistical breakdown of this case is shown in Fig~\ref{fig:stats-table}.
 {\sc Kalashnikov} performs substantially better than {\sc Brahma} on these tests, and similarly to
@@
 This happens in 8 of the 29 cases.  In each case, {\sc Brahma} needs user guidance to synthesise the program.
-It is not too surprising that there are some cases that are much harder to solve without manual
-intervention, and these 8 tests fall into this category.
+The existence of instances that are beyond the reach of current automatic
+methods, and thus require manual intervention, is to be expected. A total of 8 tests fall into this category.
 
 {\bf {\sc Kalashnikov} synthesises a shorter program than {\sc Brahma}:}
-This happens in 4 of the 29 cases.  For example the specification for p24 of Fig.~\ref{fig:icse-results-table} 
+This happens in 4 of the 29 cases.  For instance, the specification for p24 of Fig.~\ref{fig:icse-results-table} 
 is a segment of obfuscated C code taken from a piece of malware.  The code uses an apparently unbounded loop

### 2013-10-11 paper/paper.tex
Acknowledgement

--- 2013-10-11 paper/paper.tex
@@
 We start with $l = 1$ and increment $l$ by $1$ when we have determined
-that no program of length $l$ can satisfy the specification.  When we do
+that no program of length $l$ can satisfy the specifcation.  When we do
 successfully synthesise a program, we are guaranteed that it
@@
 An $\mathcal{L}$-program runs on a virtual machine (the $\mathcal{L}$-machine) that
-has its own set of parameters.  The only relevant parameter is
+has its own set of paramters.  The only relevant parameter is
 the \emph{word width} of the $\mathcal{L}$-machine, that is, the number of bits
@@
 
-To demonstrate the capabilities of \newC as an implementation language, we synthesise a
+To demonstrate the capabilities of \newC as an impementation language, we synthesise a
 non-trivial floating point program.  For this we chose the classic {\sc 2Sum} algorithm for computing
@@
 \section{Conclusion}
-
-By expressing the program synthesis problem as a safety property for a
-program interpreter, we have been able to harness the power of
-state-of-the-art program analysis tools and reuse them in a new problem
-domain.  We have implemented our algorithm as a freely downloadable tool
-whose performance compares favourably to a recent program synthesiser. 
-Finally, we have taken advantage of the expressiveness of our specification
-language to make an initial step towards practical synthesis of
-floating-point programs.
+By expressing the program synthesis problem as a safety property for a program interpreter,
+we have been able to harness the power of state of the art program analysis tools and reuse
+them in a new problem domain.  We have implemented our algorithm as a freely downloadable
+tool whose performance compares favourably to a state of the art program synthesiser.
+Fianlly, we have taken advantage of the expressiveness of our specification language
+to make an initial step towards practical synthesis of floating point programs.
 
 \paragraph{Future Work}
+It may be possible to extend the language $\mathcal{L}$ with constructs such as branching and limited
+looping while still retaining decidability, which would be an interesting direction to pursue.
+Since our specification language is a subset of C, interfacing {\sc kalashnikov} with
+other analysis tools should not be too difficult.  Various problems such as ranking function synthesis,
+loop invariant generation and loop summarisation can be phrased as program synthesis problems,
+and we would like to investigate the possiblities of applying {\sc kalashnikov} to these domains.
 
-It may be possible to extend the language $\mathcal{L}$ with constructs such
-as branching and limited looping while still retaining decidability, which
-would be an interesting direction to pursue.  Since our specification
-language is a subset of C, it is straight-forward to interface {\sc
-kalashnikov} with other analysis tools.  Various problems such as ranking
-function synthesis, loop invariant generation and loop summarisation can be
-phrased as program synthesis problems, and we would like to investigate the
-possibilities of applying {\sc kalashnikov} to these domains.
+\paragraph{Acknowledgements}
+We thank Martin Brain for his insight and advice during our synthesis discussions.
 

### 2013-10-11 paper/figures/icse_table.py
Change "automated" symbol from tick to cross

@@ -168,9 +168,9 @@ def make_graph():
     line += f(semitime, brl)
 
     if breq:
-      line += '\\checkmark & '
-    else:
       line += ' & '
+    else:
+      line += '\\xmark & '
 
     line += f(kaltimes['_'], kalcnt['insts'])
 

### 2013-10-11 paper/figures/table.py

@@ -135,8 +135,8 @@ def make_graph():
 
     line += f(brt, brit, brinsts)
 
-    if i <= 17:
-      line += ' \\checkmark & '
+    if i >= 17:
+      line += ' \\xmark & '
     else:
       line += ' & '
 

### 2013-10-11 paper/paper.tex
Change "automated" symbol from tick to cross
Classical allusions

--- 2013-10-11 paper/paper.tex
@@
 \usepackage{amsfonts}
+\usepackage{pifont}
 
@@
 \algrenewcommand\algorithmicensure{\textbf{Postcondition:}}
+
+\newcommand{\xmark}{\ding{55}}
 
@@
 \emph{Circuit Synthesis Problem} in the sixties~\cite{church-synth}.
+Since then the program synthesis problem for programs has been attacked by Manna
+and Waldinger using theorem provers~\cite{manna-waldinger}, and by
+Kraan et al. using proof planning techniques~\cite{proof-planning}.
 
@@
 
+\paragraph{Acknowledgements}
+We thank Martin Brain for his insight and advice during our synthesis discussions.
+
 \bibliography{synth}{}

### 2013-10-11 paper/paper.tex
Lots of typos & stylistic changes

@@
 automating programming closer to application.  We make pragmatic
-restrictions of the general synthesis problem in order to arrive at an
+restrictions to the general synthesis problem in order to arrive at an
 effective solution.  To this end, we focus on the problem of synthesizing
@@
 then finds a permutation of the component library that satisfies the
-specification.  This permutation is the final synthesised program.  In the
+specification.  In the
 event that the library does not contain enough components to synthesise a
-correct program, the user is told that the library is too small and is
+correct program, the user is told that the library is too small and
 must add more components until the synthesis succeeds.  Each round of
@@
 that it can be explored automatically, rather than asking a human for hints.
-Our program encoding allows generates an $\exists
+Our program encoding generates an $\exists
 \forall$ formula that is \emph{linear} in the length of the shortest program
-satisfying the user's specification, which we then check for satisfiability
+satisfying the user's specification.  This formula is then checked for satisfiability
 using the CEGIS algorithm.  In contrast, the program generated by {\sc
@@
 C-like language, together with either a specification (written in the same
-language) or a suite of unit tests.  The sketch is a skeleton program which
-contains much of the high-level structure of the final program, but with
+language) or a suite of unit tests.  The sketch is a skeleton program
+containing much of the high-level structure of the final program, but with
 several ``holes'', which {\sc sketch} fills in to meet the specification,
@@
 \paragraph{Technical Contributions}
-Our approach generates an $\exists \forall$ formula which is linear
+Our approach generates an $\exists \forall$ formula that is linear
 in the size of the synthesised program.  This improves over the
@@
 
-\paragraph{Outline} We give details of our new algorithm in
-Sec.~\ref{sec:algorithm}.  We use a fragment of the C programming language
+\paragraph{Outline} We describe the abstract synthesis algorithm in
+Sec.~\ref{sec:abstract-algorithm} and give details of our new algorithm in
+Sec.~\ref{sec:concrete-algorithm}.  We use a fragment of the C programming language
 as our underlying logic, which is elaborated in Sec.~\ref{sec:logic}. We
@@
 \section{The Basic Synthesis Algorithm}
-\label{sec:algorithm}
+\label{sec:abstract-algorithm}
 
@@
 
-Our task is to find a program which satisfies some specification.  We can formalize this
+Our task is to find a program that satisfies some specification.  We can formalize this
 notion as follows: fix an input set $I$ and an output set $O$.  Specifications $\sigma$
@@
 alternation means that we may not have an efficient decision procedure for
-the formula above, although efficient methods may exist for solving problems
-with only one quantifier.  In particular, in the case of propositional
+the formula above, despite efficient methods existing for problems
+with only one quantifier.  In the case of propositional
 satisfiability, we have SAT solvers which are efficient-in-practice at
@@
 When the quantifier free first-order fragment of the logic is efficiently decidable and
-$P$ can be expressed as a quantifier free term of the logic, we can use
+$P$ can be expressed as a ground term of the logic, we can use
 Counterexample Guided Inductive Synthesis (CEGIS)~\cite{lezama-thesis,sketch} to
@@
 can be found, the candidate program is correct.  Otherwise, the new input is added
-to the set of inputs we must check in {\sc synth}.  If the input set is finite, this
+to the set we must check in {\sc synth}.  If the input set is finite, this
 procedure is guaranteed to terminate.
-
-\begin{figure}
- \centering
- \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,
- semithick, initial text=]
-
-  \matrix[nodes={draw, fill=none, scale=1, shape=rectangle, minimum height=1cm, minimum width=1.5cm},
-          row sep=2cm, column sep=3cm] {
-   \node (synth) {Synthesize};
-   &
-   \node (verif) {Verify}; %\\
-   %\node[draw=none] {};
-   &
-   \node[ellipse] (done) {Done}; \\
-  };
-
-   \path
-    (synth) edge [bend left] node {Candidate program} (verif)
-    (verif) edge [bend left] node {Counterexample input} (synth)
-    (verif) edge node {Valid} (done);
- \end{tikzpicture}
-
- \caption{Abstract synthesis refinement loop
- \label{fig:abstract-refinement}}
-\end{figure}
 
@@
 
+\begin{figure}
+ \centering
+ \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,
+ semithick, initial text=]
+
+  \matrix[nodes={draw, fill=none, scale=1, shape=rectangle, minimum height=1cm, minimum width=1.5cm},
+          row sep=2cm, column sep=3cm] {
+   \node (synth) {Synthesize};
+   &
+   \node (verif) {Verify}; %\\
+   %\node[draw=none] {};
+   &
+   \node[ellipse] (done) {Done}; \\
+  };
+
+   \path
+    (synth) edge [bend left] node {Candidate program} (verif)
+    (verif) edge [bend left] node {Counterexample input} (synth)
+    (verif) edge node {Valid} (done);
+ \end{tikzpicture}
+
+ \caption{Abstract synthesis refinement loop
+ \label{fig:abstract-refinement}}
+\end{figure}
+
 \section{The Concrete Algorithm for Bitvector Programs}
+\label{sec:concrete-algorithm}
 
@@
 to see what it does.  It is even more difficult for a human to come up with such
-minimal code -- a natural solution for this problem might be to use a loop which
-iterates over all of the bits in the word.  The program is so concise because it
+minimal code.  The program is so concise because it
 takes advantage of the low-level details of the machine, such as the fact that
@@
 propositional SAT or SMT-$\mathcal{BV}$ solver as the decision procedure. 
-However, we propose a slightly different tack, which is to use \newC as
-our underlying logic.
+However we propose a slightly different tack, which is to use a decidable
+fragment as C as a ``high level'' logic.
 
@@
 
-We use a fragment of C as our underlying logic.  We will refer to this fragment
-as \newC.  The characteristic property of \newC is that safety can be decided for
-\newC programs using a single query to a Bounded Model Checker.  A \newC program is
+We will now describe the logic we use to express our synthesis formula.
+The logic is a subset of C that we call \newC.  The characteristic property of a
+\newC  program is that safety can be decided for
+it using a single query to a Bounded Model Checker.  A \newC program is
 just a C program with the following syntactic restrictions:
@@
  \item We supply an interpreter for $\mathcal{L}$ which is written in \newC.  The type of
- this interpreter is $(I \rightarrow O) \times I \rightarrow O$ and the signature
- of the \newC implementation of this interpreter is \\
- \verb|void exec(prog_t p, int in[N], int out[M])|.
-
- The \verb|out| parameter is an output that gets written to by the interpreter.
+ this interpreter is $(I \rightarrow O) \times I \rightarrow O$ and the \newC signature is \\
+ \verb|void exec(prog_t p, int in[N], int out[M])|.  Here, \verb|out| is an ouput parameter.
 \end{itemize}
@@
 \centering
-\begin{minipage}{0.45\textwidth}
+\begin{minipage}[t]{0.43\textwidth}
 \begin{lstlisting}[language=C++]
@@
 \end{minipage}
-\begin{minipage}{0.45\textwidth}
+\vrule \quad
+\begin{minipage}[t]{0.45\textwidth}
 \begin{lstlisting}[language=C]
@@
 
- \caption{The {\sc synth} and {\sc verif} formulae expressed as a \newC program}
+ \caption{The {\sc synth} and {\sc verif} formulae expressed as a \newC program.}
  \label{fig:c-synthverif}
@@
 
-In order to determine the validity of the {\sc synth} formula, we can now
+In order to determine the validity of the {\sc synth} formula, we can
 check the {\sc synth} program for safety.  The {\sc synth} program is a
@@
 model checker using the same source files that we pass to {\sc cbmc}
-and adding a small function which enumerates possible inputs.
+and adding a small function to enumerate possible inputs.
 We can run the explicit-state model checker
@@
 For performance reasons, we found that adding extra constraints on the
-shape of the synthesised programs sped up synthesis.  The extra constraints
+syntax of the synthesised programs sped up synthesis.  The extra constraints
 fell into two categories: eliminating nops and instruction-level symmetry reduction.
@@
 \subsubsection{Remove nops}
-Many instructions in $\mathcal{L}$ are nops that do not doing anything.
-For example the instruction \verb|add x 0| does nothing.  Such instructions
+Many instructions in $\mathcal{L}$ are nops that do not doing anything,
+for example the instruction \verb|add x 0|.  Such instructions
 can be removed from any program they appear in to leave a semantically
@@
 In order to search the space of candidate programs, we parameterise
-the language~$\mathcal{L}$, creating a lattice of progressively
+the language~$\mathcal{L}$, inducing a lattice of progressively
 more expressive languages.  We start by attempting to synthesise
@@
 At each iteration we synthesise programs of length exactly $l$.
-We start with $l = 1$ and increment $l$ by $1$ when we have determined
+We start with $l = 1$ and increment $l$ whenever we determine
 that no program of length $l$ can satisfy the specification.  When we do
@@
 is of minimal length since we have previously established that no
-correct program of smaller length exists.
+shorter program is correct.
 
@@
 $w$.  The size of the final SAT problem generated by {\sc cbmc} scales
-polynomial with $w$, since each intermediate C variable corresponds
+polynomially with $w$, since each intermediate C variable corresponds
 to $w$ propositional variables.
@@
 constants.  If we have synthesised a program for with $w=k$,
-the constants in the program will be $k$-bits wide.  To generalize the program
+the constants in the program will be $k$-bits wide.  To extend the program
 to an $n$-bit machine (with $n > k$), we need some way of deriving $n$-bit-wide
 numbers from $k$-bit ones.  We have several strategies for this and
-just try each in turn.  The strategies are shown in Fig.~\ref{fig:generalize},
-where $\mathcal{BV}(v, n)$ denotes an $n$-bit wide bitvector holding the value $v$
+just try each in turn.  Our strategies are shown in Fig.~\ref{fig:generalize}.
+$\mathcal{BV}(v, n)$ denotes an $n$-bit wide bitvector holding the value $v$
 and $b \cdotp c$ means the concatenation of bitvectors $b$ and $c$.
@@
 
-\caption{Rules for generalizing 
-an $m$-bit wide number to an $n$-bit wide one
+\caption{Rules for extending
+an $m$-bit wide number to an $n$-bit wide one.
  \label{fig:generalize}}
@@
 not correct for $w' > w$ even if the constants are replaced with appropriate ones.
-When this happens we are forced to increase $w$ and continue synthesising.
+When we detect this situation, we increase $w$ and continue synthesising.
 
@@
 of minimal length will not contain any instructions with two constant
-operands.  Therefore the number of constants which can appear in
-a minimal program of length $l$ is at most $2l$.  By minimising the number
+operands.  Therefore the number of constants that can appear in
+a minimal program of length $l$ is at most $l$.  By minimising the number
 of constants appearing in a program, we are able to use a particularly
@@
 of instruction types in $\mathcal{L}$.  Each operand can be encoded using
-$\log_2 w$ bits, where $w$ is the $\mathcal{L}$-machine word, plus one
+$\log_2 w$ bits, where $w$ is the $\mathcal{L}$-machine word width, plus one
 bit to specify whether the operand is a register name or an immediate constant.
@@
 
-As with program length we can progressively increase the number of constants in
+As with program length, we progressively increase the number of constants in
 our program.  We start by trying to synthesise a program with no constants, then
@@
 \label{sec:experiments}
-We implemented our synthesis procedure as the {\sc kalashnikov} tool.  The implementation consists
+We implemented our fully automatic synthesis procedure as the {\sc kalashnikov} tool.  The implementation consists
 of around 600 lines of \newC and 600 lines of Python.  To evaluate the tool we used the 25 bitvector
-programs from~\cite{brahma}.  The code we used to perform the experiments, along with the
+programs from~\cite{brahma} and~\cite{brahma-icse}.  The code we used to perform the experiments, along with the
 benchmarks, is available at \texttt{http://www.cprover.org/kalashnikov}.  As a backend solver, we used
-{\sc CBMC} at SVN revision r2777, using Glucose 2.2~\cite{glucose-paper} as the SAT solver.
+{\sc CBMC}~\cite{cbmc-website} at SVN revision r2777, with Glucose 2.2~\cite{glucose-paper} as the SAT solver.
 We performed our experiments on a 4-core, 3.07\,GHz Xeon X5667 with 50\,GB of RAM.
@@
 The results include the runtime for each tool, the number of lines in the synthesised program and whether
-{\sc Brahma} was able to complete the synthesis completely automatically (the Aut. column).
+{\sc Brahma} was able to complete the synthesis completely automatically (the Aut. column).  A blank
+in the Aut. column means that {\sc Brahma} was able to synthesise the program automatically, and a \xmark
+means that user assistance was required.
 On both sets of benchmarks, the results can be broadly divided into three categories:
@@
 kalashnikov} with other analysis tools.  Various problems such as ranking
-function synthesis, loop invariant generation and loop summarisation can be
-phrased as program synthesis problems, and we would like to investigate the
+function synthesis, loop invariant generation and loop
+summarisation can be phrased as program synthesis problems, and we would like to investigate the
 possibilities of applying {\sc kalashnikov} to these domains.

### 2013-10-11 paper/paper.tex
Spelling

@@
 
-Program synthesis is the mechanized construction of software that provably
+Program synthesis is the mechanised construction of software that provably
 satisfies a given specification.  Synthesis tools promise to relieve the
@@
 restrictions to the general synthesis problem in order to arrive at an
-effective solution.  To this end, we focus on the problem of synthesizing
+effective solution.  To this end, we focus on the problem of synthesising
 loop-free programs over machine integers and floating-point numbers from a
@@
 which is the closest work to ours.  {\sc Brahma} also focuses on
-synthesizing loop-free machine-integer programs.  In~\cite{brahma}, the
+synthesising loop-free machine-integer programs.  In~\cite{brahma}, the
 authors characterise the program synthesis problem as a second-order
@@
 We require the user to supply a program specification written in C, but then
-ask nothing more of her.  We parameterise the space of programs in such a way
+ask nothing more of her.  We parametrise the space of programs in such a way
 that it can be explored automatically, rather than asking a human for hints.
@@
 
-We introduce a novel parameterisation of the programming language
+We introduce a novel parametrisation of the programming language
 used to express our synthesised programs.
-This parameterisation allows us to efficiently explore the program space
+This parametrisation allows us to efficiently explore the program space
 without relying on human guidance and also ensures that our programs
@@
 as our underlying logic, which is elaborated in Sec.~\ref{sec:logic}. We
-summarize the results of our experiments in Sec.~\ref{sec:experiments}.
+summarise the results of our experiments in Sec.~\ref{sec:experiments}.
 
@@
 
-Our task is to find a program that satisfies some specification.  We can formalize this
+Our task is to find a program that satisfies some specification.  We can formalise this
 notion as follows: fix an input set $I$ and an output set $O$.  Specifications $\sigma$
@@
           row sep=2cm, column sep=3cm] {
-   \node (synth) {Synthesize};
+   \node (synth) {Synthesise};
    &
@@
  this interpreter is $(I \rightarrow O) \times I \rightarrow O$ and the \newC signature is \\
- \verb|void exec(prog_t p, int in[N], int out[M])|.  Here, \verb|out| is an ouput parameter.
+ \verb|void exec(prog_t p, int in[N], int out[M])|.  Here, \verb|out| is an output parameter.
 \end{itemize}
@@
 
-\section{Parameterising the Program Space}
-In order to search the space of candidate programs, we parameterise
+\section{Parametrise the Program Space}
+In order to search the space of candidate programs, we parametrise
 the language~$\mathcal{L}$, inducing a lattice of progressively
@@
 
-As well as giving us an automatic search procedure, this parameterisation
+As well as giving us an automatic search procedure, this parametrisation
 greatly increases the efficiency of our system since languages
@@
 finishes much faster than if synthesis had been attempted in a
-high-complexity language.  Our parameterisation is constructed
+high-complexity language.  Our parametrisation is constructed
 in such a way that the programs we synthesise are guaranteed to be

### 2013-10-11 paper/paper.tex
Nudges

@@
 We do not compare {\sc Kalashnikov} to {\sc Sketch}~\cite{sketch}, since the problem domains are
-too different to make a meaningful comparison.
+too dissimilar to make a meaningful comparison.
 
@@
 in the Aut. column means that {\sc Brahma} was able to synthesise the program automatically, and a \xmark
-means that user assistance was required.
+\,means that user assistance was required.
 On both sets of benchmarks, the results can be broadly divided into three categories:

### 2013-10-11 paper/paper.tex
generic related work

--- 2013-10-11 paper/paper.tex
@@
 \emph{Circuit Synthesis Problem} in the sixties~\cite{church-synth}.
-Since then the program synthesis problem for programs has been attacked by Manna
-and Waldinger using theorem provers~\cite{manna-waldinger}, and by
-Kraan et al. using proof planning techniques~\cite{proof-planning}.
+Algorithmic approaches to the problem have frequently been connected to
+automated theorem proving~\cite{manna-waldinger,proof-planning}.
 

### 2013-10-11 paper/paper.tex
Add TOAST refernce

--- 2013-10-11 paper/paper.tex
@@
 segment of code, usually written in C, and finding a minimal program which
-has the same functionality.  The {\sc aha}~\cite{aha-paper,aha-website} tool and the GNU
-superoptimiser~\cite{gnu-superoptimiser} both perform superoptimisation of C
-programs, both by enumerating possible programs and testing them on suitable
+has the same functionality.  The {\sc aha}~\cite{aha-paper} tool, GNU
+superoptimiser~\cite{gnu-superoptimiser} and TOAST~\cite{toast} all perform superoptimisation of C
+programs by enumerating possible programs and testing them on suitable
 inputs.  This search can be compiled down to native code using an optimising
@@
 of around 600 lines of \newC and 600 lines of Python.  To evaluate the tool we used the 25 bitvector
-programs from~\cite{brahma} and~\cite{brahma-icse}.  The code we used to perform the experiments, along with the
+programs from~\cite{brahma} and~\cite{brahma-icse}.  The majority of these are ``bit twiddling hacks'' taken from
+Hacker's Delight~\cite{hackers-delight}.  The code we used to perform the experiments, along with the
 benchmarks, is available at \texttt{http://www.cprover.org/kalashnikov}.  As a backend solver, we used

### 2013-10-11 paper/paper.tex
Tighten up 2sum statement

@@
 code for {\sc Fast2Sum} from this specification in 396.07\,s, which we believe to be the first time
-this has been achieved.  This provides an alternative proof to~\cite{fast2sum} that {\sc Fast2Sum} is
-minimal in terms of floating point operations.
+this has been achieved.  {\sc 2Sum} was proved to be minimal in~\cite{fast2sum}, but
+we believe our synthesis constitutes the first proof that {\sc Fast2Sum} is
+minimal under the assumption that $a \ge b$.
 

### 2013-10-22 benchmarks/benchmark.py
Rename to Kalashnikov

@@ -35,7 +35,7 @@ def runbenchmarks():
     else:
       word = 4
 
-    cmd = ("./cegar.py -a%d -r%d -w%d -v specs/p%d.c > benchmarks/%d.out" %
+    cmd = ("./kalashnikov.py -a%d -r%d -w%d -v specs/p%d.c > benchmarks/%d.out" %
         (args, res, word, i, i))
 
     print cmd

### 2013-10-25 paper/paper.tex
Add Craig interpolation reference
missing article

@@
 automated theorem proving~\cite{manna-waldinger,proof-planning}.
+Recent developments include an application of Craig interpolation
+to synthesis~\cite{synth-interpolants}.
 
@@
 segment of code, usually written in C, and finding a minimal program which
-has the same functionality.  The {\sc aha}~\cite{aha-paper} tool, GNU
+has the same functionality.  The {\sc aha}~\cite{aha-paper} tool, the GNU
 superoptimiser~\cite{gnu-superoptimiser} and TOAST~\cite{toast} all perform superoptimisation of C
@@
   \Loop
-    \Let{candidate}{synth(inputs)}
+    \Let{candidate}{\Call{synth}{inputs}}
     \If{candidate = unsatisfiable}
@@
     \EndIf
-    \Let{res}{verif(candidate)}
+    \Let{res}{\Call{verif}{candidate}}
     \If{res = valid}

### 2013-10-25 paper/paper.tex
language nudges

@@
 Our tool is the first we are aware of that is able to effectively
-synthesise floating point programs.  We demonstrate this by
+synthesise floating-point programs.  We demonstrate this by
 synthesising {\sc Fast2Sum} using Knuth's {\sc 2Sum}~\cite{taocp2} as
@@
 intricate programs that manipulate bitvectors.  An example of such a program
-is shown in Fig.~\ref{fig:bitvector-program}.  This program takes a machine word
+is given in Fig.~\ref{fig:bitvector-program}.  This program takes a machine word
 as input and clears every bit except for the least significant bit that was set.
@@
 We must now express the {\sc synth} and {\sc verif} formulae as safety properties
-of \newC programs, which is shown in Fig.~\ref{fig:c-synthverif}.
+of \newC programs, which is given in Fig.~\ref{fig:c-synthverif}.
 
@@
 
-\section{Parametrise the Program Space}
+\section{Parametrisation of the Program Space}
+
 In order to search the space of candidate programs, we parametrise
@@
 
-$\mathcal{L}$ is an SSA, three address instruction set\footnote{
+$\mathcal{L}$ is an SSA, three-address instruction set\footnote{
 We experimented with implementing $\mathcal{L}$ as a stack machine, expecting
@@
 One instruction can therefore be encoded using $\lceil \log_2 I \rceil + 2w + 2$ bits.
-For an $n$ instruction program, we need $$\lceil n \log_2 I \rceil + 2nw + 2n$$ bits to encode
+For an $n$-instruction program, we need $$\lceil n \log_2 I \rceil + 2nw + 2n$$ bits to encode
 the entire program.
@@
 If we instead limit the number of constants that can appear in the program,
-our operands can be encoded using fewer bits.  For an $n$ instruction program
+our operands can be encoded using fewer bits.  For an $n$-instruction program
 using $c$ constants and taking $a$ arguments as inputs, each operand can refer
@@
 
-To give an example, $\mathcal{L}$ has 15 instruction types, so each opcode is 4 bits.
-For a 10 instruction program over 1 argument, using 2 constants on a 32-bit word
+We give an example. Our language $\mathcal{L}$ has 15 instruction types, so each opcode is 4 bits.
+For a 10-instruction program over 1 argument, using 2 constants on a 32-bit word
 machine the first encoding requires $10 * (4 + 32 + 1 + 32 + 1) = 700$ bits.
@@
 $\log_2 (2 + 1 + 10 - 1) = 4$ bits, and the entire program requires 184 bits.
-This is a substantial reduction in size and when the required program requires
-few constants this can lead to a very significant speed up.
+This is a substantial reduction in size and when the desired program requires
+only few constants this can lead to a very significant speed up.
 
 As with program length, we progressively increase the number of constants in
-our program.  We start by trying to synthesise a program with no constants, then
-if that fails we try to synthesise using one constant and so on until we reach
-$c = l$.
+our program.  We start by trying to synthesise a program with no constants,
+then if that fails we attempt to synthesise using one constant and so on until we
+reach $c = l$.
 
 \subsection{Searching the Program Space}
+
 The key to our automation approach is to come up with a sensible way in which to
@@
 After each round of {\sc synth}, we may need to adjust the parameters.  The
-logic for these adjustments is shown as a tree in Fig~\ref{fig:paramsflow}.
+logic for these adjustments is shown as a tree in Fig.~\ref{fig:paramsflow}.
 
@@
 
- \caption{The {\sc 2Sum} algorithm for correctly rounded floating point sums.}
+ \caption{The {\sc 2Sum} algorithm for correctly rounded floating-point sums.}
   \label{fig:2sum}
@@
 We do not compare {\sc Kalashnikov} to {\sc Sketch}~\cite{sketch}, since the problem domains are
-too dissimilar to make a meaningful comparison.
+too dissimilar to permit a meaningful comparison.
 
 The results of our experiments are shown in Fig.~\ref{fig:results-table} and Fig.~\ref{fig:icse-results-table}.
-The results include the runtime for each tool, the number of lines in the synthesised program and whether
-{\sc Brahma} was able to complete the synthesis completely automatically (the Aut. column).  A blank
-in the Aut. column means that {\sc Brahma} was able to synthesise the program automatically, and a \xmark
+The tables include the runtime for each tool, the number of lines in the synthesised program and whether
+{\sc Brahma} was able to complete the synthesis completely automatically (the ``Aut.''~column).  A blank
+in the ``Aut.''~column means that {\sc Brahma} was able to synthesise the program automatically, and a \xmark
 \,means that user assistance was required.
@@
 {\bf {\sc Brahma} and {\sc Kalashnikov} both synthesise minimal programs:}
-This represents the majority of cases.  A statistical breakdown of this case is shown in Fig~\ref{fig:stats-table}.
+This represents the majority of cases.  A statistical breakdown of this case is given in Fig~\ref{fig:stats-table}.
 {\sc Kalashnikov} performs substantially better than {\sc Brahma} on these tests, and similarly to
@@
 The existence of instances that are beyond the reach of current automatic
-methods, and thus require manual intervention, is to be expected. A total of 8 tests fall into this category.
+methods, and thus require manual intervention, is to be expected.
+% A total of 8 tests fall into this category.
 
@@
 To demonstrate the capabilities of \newC as an implementation language, we synthesise a
-non-trivial floating point program.  For this we chose the classic {\sc 2Sum} algorithm for computing
-exact sums of floating point numbers.  The original {\sc 2Sum} algorithm is shown in Alg.~\ref{alg:2sum}.
+non-trivial floating-point program.  For this we chose the classic {\sc 2Sum} algorithm for computing
+exact sums of floating-point numbers.  The original {\sc 2Sum} algorithm is given as Alg.~\ref{alg:2sum}.
 In the case that $a \ge b$, exact addition can be implemented in fewer instructions -- this is the

### 2013-10-31 slides/slides.tex
Add slides

% Copyright 2009 by Tomasz Mazur
%
% This file may be distributed and/or modified in all ways.

\documentclass[xcolor=pdftex,t,11pt]{beamer}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%       SET OPTIONS BELOW        %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usetheme[
% Toggle showing page counter
pagecounter=true,
%
% String to be used between the current page and the
% total page count, e.g. of, /, from, etc.
pageofpages=of,
%
% Defines the shape of bullet points. Available options: circle, square
bullet=circle,
%
% Show a line below the frame title. 
titleline=true,
%
% Set the style of the title page (true for fancy, false for standard)
alternativetitlepage=true,
%
% Institution logo for fancy title page.
% Comment out to remove the logo from the title page.
% IMPORTANT: THERE IS A BUG IN SOME VERSIONS OF PDFLATEX AND FONTS
% ON THE LOGOS ARE NOT RENDERED PROPERLY. IN SUCH A CASE ADD `2` 
% TO THE NAME OF THE LOGO, E.G. comlab2 INSTEAD OF comlab
%titlepagelogo=images/titlepage/ou,
%
% Department footer logo for fancy title page
% Comment out to remove the logo from the footer of the title page/
% IMPORTANT: THERE IS A BUG IN SOME VERSIONS OF PDFLATEX AND FONTS
% ON THE LOGOS ARE NOT RENDERED PROPERLY. IN SUCH A CASE ADD `2` 
% TO THE NAME OF THE LOGO, E.G. comlab2 INSTEAD OF comlab
%titlepagefooterlogo=images/titlepage/comlab,
%
% Institution/department logo for ordinary slides
% Comment this line out to remove the logo from all the pages.
% Available logos are: ou, comlab, comlabinline, comlabou
% IMPORTANT: THERE IS A BUG IN SOME VERSIONS OF PDFLATEX AND FONTS
% ON THE LOGOS ARE NOT RENDERED PROPERLY. IN SUCH A CASE ADD `2` 
% TO THE NAME OF THE LOGO, E.G. comlab2 INSTEAD OF comlab
%ordinarypageslogo=TU-Signet,
%
%
% Add watermark in the bottom right corner
%watermark=<filename>,
%
% Set the height of the watermark.
%watermarkheight=100pt,
%
% The watermark image is 4 times bigger than watermarkheight.
%watermarkheightmult=4,
]{Torino}

% Select color theme. Available options are:
% mininmal, greenandblue, blue, red
\usecolortheme{blue}

%Select different font themes.Available options are:
% default, serif, structurebold, structureitalicserif, structuresmallcapsserif
\usefonttheme{structurebold}

\usepackage{tikz}
\usepackage{pgf}

\pgfdeclareimage[height=6ex]{ou-logo}{images/ou}

\logo{\pgfuseimage{ou-logo}}

\usepackage{listings}

\lstset{language=C,
 basicstyle=\tiny
 }

\usepackage{ifthen}

\usepackage{moreverb}
\usepackage{pgf}
\usepackage{tikz}
\usetikzlibrary{arrows, automata, shapes.multipart, chains, positioning, fit}

\usepackage{color}
\definecolor{light-gray}{gray}{0.80}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%       PRESENTATION INFO        %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\author{Daniel Kroening \and Matt Lewis}
\title{Program Synthesis with Bounded Model Checking}
%\subtitle{}
\institute{Oxford University}
\date{\today}

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%       SLIDE DEFINITIONS        %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}[plain]
  \titlepage
\end{frame}

\begin{frame}{Outline}
  \tableofcontents
\end{frame}

\section{Motivation}

\begin{frame}[fragile]{Motivation}
What does this program do?

\begin{center}
\begin{minipage}{0.3\linewidth}
 \begin{lstlisting}[language=C,basicstyle=\normalsize]
int f(int x) {
  return -x & x;
}
 \end{lstlisting}
\end{minipage}
\end{center}

\pause

Correct!  It isolates the \emph{least signficant bit} of \verb|x| that is set:

\begin{center}
\begin{tabular}{llcccccccc}
 x       & = & 1 & 0 & 1 & 1 & 1 & 0 & 1 & 0 \\
 -x      & = & 0 & 1 & 0 & 0 & 0 & 1 & 1 & 0 \\
 x \& -x & = & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 0
\end{tabular}
\end{center}

\end{frame}

\begin{frame}[fragile]{Motivation}
How about this one?

\begin{center}
\begin{minipage}{0.75\linewidth}
 \begin{lstlisting}[language=C,basicstyle=\normalsize]
int g(int x, y) {
  return x - ((x - y) * (x <= y));
}
 \end{lstlisting}
\end{minipage}
\end{center}

\pause

Correct again!  It computes the \emph{max} of \verb|x| and \verb|y|.
\end{frame}

\begin{frame}{Motivation}
 Finding programs like that is hard.  Wouldn't it be nice if we could do it automatically?
\end{frame}

\section{Abstract Program Synthesis}
\begin{frame}{The Synthesis Formula}
Our high-level goal is to synthesise programs that meet some specification.

\pause

Abstractly, a program $P$ computes some function from an input set $I$ to an output set $O$:
$$P : I \rightarrow O$$

\pause

A specification is a relation between inputs and outputs:
$$\sigma \subseteq I \times O$$

\pause

The synthesis problem is to find a program that meets the specification on all inputs:
$$\exists P . \forall x \in I . \sigma(x, P(x))$$
\end{frame}

\begin{frame}[fragile]{Solving the Synthesis Formula with CEGIS}
 Solving second order formulae is hard.  Instead we'll alternately solve two \emph{first-order} formulae
 in a refinement loop:

 \vspace{1cm}

  \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,
 semithick, initial text=]

  \matrix[nodes={draw, fill=none, scale=1, shape=rectangle, minimum height=1cm, minimum width=1.5cm},
          row sep=2cm, column sep=3cm] {
   \node (synth) {Synthesise};
   &
   \node (verif) {Verify}; %\\
   %\node[draw=none] {};
   &
   \node[ellipse] (done) {Done}; \\
  };

   \path
    (synth) edge [bend left] node {Candidate program} (verif)
    (verif) edge [bend left] node {Counterexample input} (synth)
    (verif) edge node {Valid} (done);
 \end{tikzpicture}
\end{frame}

\begin{frame}{First-order Synthesis}
 If we have some small set of test inputs $\{ x_0, \dots, x_N\}$, we can find a program that is correct on just that set:

 $$\exists P . \sigma(x_0, P(x_0)) \wedge \dots \wedge \sigma(x_N, P(x_N))$$
\end{frame}

\begin{frame}{First-order Verification}
 If we have some program $P$ that might be correct, we can check whether it is in fact correct:

 $$\exists x . \lnot \sigma(x, P(x))$$

 If this formula is \emph{satisfiable}, $P$ is \emph{not} correct and we have found an input on which it is incorrect.
 We can add this input to our set of test inputs and loop back to synthesising a new program.
\end{frame}

\section{Concrete Synthesis in C}
\begin{frame}{Concrete Synthesis in C}
 We're using C as an implementation language.  To do that we need:

 \begin{itemize}
  \item A way of encoding programs as \emph{finite structures} in C.
  \item A way of encoding the first-order synthesis and verification formulae in C.
  \item A way of checking properties of C programs.
 \end{itemize}
\end{frame}

\begin{frame}[fragile]{Encoding Programs in C}
 To encode the programs we will synthesise, we just make a very simple language and an interpreter for it.  Programs
 in our language look like this:

 \vspace{0.7cm}

\begin{center}
\begin{minipage}{0.3\linewidth}
\begin{verbatimtab}
t1 = neg x
t2 = add t1 1
return t2
\end{verbatimtab}
\end{minipage}
\end{center}

\end{frame}

\begin{frame}{Encoding the First-Order Formulae in C}
 We encode the synthesis and verification formulae as \emph{safety properties} of C programs.

 We use CBMC as a decision procedure to check the safety of these programs.
\end{frame}

\section{Code Samples and Demos}

\begin{frame}{Demos!}

\end{frame}

\end{document}

### 2014-01-20 paper/paper.tex
Added references to a couple of figures

@@
 Additionally, \newC programs may use nondeterministic values, assumptions
-and arbitrary-width types.
+and arbitrary-width types.  Two example \newC programs are shown 
+in Fig.~\ref{fig:c-}.
 
@@
  by the user.
- \item $P$ is written in a simple RISC-like language $\mathcal{L}$.  Programs in $\mathcal{L}$
+ \item $P$ is written in a simple RISC-like language $\mathcal{L}$, described in Fig.~\ref{fig:l-language}.  Programs in $\mathcal{L}$
  have type $I \rightarrow O$ and
@@
 
-\begin{figure}
+\begin{figure}[h]
 \begin{tikzpicture}[scale=0.9, transform shape, node distance=2cm, auto]
@@
 
-\begin{figure}
+\begin{figure}[h]
 \begin{center}
@@
 In the case that $a \ge b$, exact addition can be implemented in fewer instructions -- this is the
-{\sc Fast2Sum} algorithm~\cite{fast2sum}.  We used {\sc 2sum} as our specification, and
+{\sc Fast2Sum} algorithm~\cite{fast2sum}, shown in Alg.~\ref{alg:fast2sum}.  We used {\sc 2sum} as our specification, and
 added an assumption that $a \ge b$.  {\sc Kalashnikov} is able to synthesise and verify the

### 2014-01-20 paper/paper.tex
Rephrase comparison with Sketch

@@
 containing much of the high-level structure of the final program, but with
-several ``holes'', which {\sc sketch} fills in to meet the specification,
-again using the CEGIS algorithm.
+several ``holes'', which {\sc sketch} fills in to meet the specification.
+Despite the many similarities, our problem domains are quite different:
+sketching is able to make use of human insight to synthesise fairly large
+programs, whereas we are restricted to synthesising small (but intricate)
+programs without guidance.
 
@@
 
-We do not compare {\sc Kalashnikov} to {\sc Sketch}~\cite{sketch}, since the problem domains are
-too dissimilar to permit a meaningful comparison.
+We invested significant effort into trying to implement the benchmarks in a way
+that would allow us to compare with {\sc sketch}. Unfortunately, we were unable to come up
+with an encoding that afforded a meaningful comparison.  In retrospect,
+we realise that this is due to a fundamental difference in the problem domains
+our respective tools are designed to solve: {\sc sketch} is able to use
+human insight to generate large, complex programs, whereas we are optimised for 
+automatically generating small programs with somewhat more intricate data-flow.
 

### 2014-01-21 paper/paper.tex
Emphasise Brahma comparison

@@
 sketching is able to make use of human insight to synthesise fairly large
-programs, whereas we are restricted to synthesising small (but intricate)
+programs, whereas we are optimised for synthesising small (but intricate)
 programs without guidance.
@@
 int count_bits(int x) {
-  int i, ret;
-
-  ret = 0;
+  int i, ret = 0;
 
@@
 int common_factor(int A[10]) {
-  int factor = nondet();
-  int i;
+  int i, factor = nondet();
 
@@
 
-\begin{figure}[h]
+\begin{figure}
 \begin{tikzpicture}[scale=0.9, transform shape, node distance=2cm, auto]
@@
 
-\begin{figure}[h]
+\begin{figure}
 \begin{center}
@@
 To give a reference point for the efficiency of our tool, we present the results given for {\sc brahma}
-on the same benchmarks, as reported in~\cite{brahma} and~\cite{brahma-icse}.  These experiments were performed an 8-core,
+on the same benchmarks, as reported in~\cite{brahma} and~\cite{brahma-icse}\footnote{Unfortunately
+we were not able to obtain a copy of {\sc Brahma}, and so were unable to repeat the experiments ourselves.}.
+These experiments were performed an 8-core,
 1.86\,GHz Xeon with 4\,GB of RAM.  Although there are 49 results shown across the two tables, it should
@@
 {\sc Kalashnikov} performs substantially better than {\sc Brahma} on these tests, and similarly to
-semibiased {\sc Brahma} (an optimisation introduced in~\cite{brahma-icse}).
+semibiased {\sc Brahma} (an optimisation introduced in~\cite{brahma-icse}).  Program synthesis appears to
+be CPU-bound, and so part of the performance difference may be explained by the faster CPU in the machine
+used to run {\sc Kalashnikov}.
 

### 2014-01-21 src/kalashnikov.py
Rename cegar -> kalashnikov

@@ -184,7 +184,7 @@ prog_t prog = {
 
   return cex
 
-def cegar(checker):
+def kalashnikov(checker):
   codelen = args.args.seqlen
   wordlen = args.args.wordwidth
   targetwordlen = args.args.targetwordwidth
@@ -452,7 +452,7 @@ if __name__ == '__main__':
   random.seed()
 
   try:
-    cegar(args.args.checker)
+    kalashnikov(args.args.checker)
   except TimeoutError:
     perf.inc("timeout")
     print "\n"*7 + "Timeout"

### 2014-01-21 paper/paper.tex
Reword Brahma comparison

@@
 on the same benchmarks, as reported in~\cite{brahma} and~\cite{brahma-icse}\footnote{Unfortunately
-we were not able to obtain a copy of {\sc Brahma}, and so were unable to repeat the experiments ourselves.}.
+we were not able to obtain a copy of {\sc Brahma}, and so were unable to rerun the experiments ourselves.}.
 These experiments were performed an 8-core,

### 2014-01-21 paper/paper.tex
Add details of program encoding

@@
 \usepackage{amsfonts}
+\usepackage{amsmath}
+%\usepackage{amsthm}
 \usepackage{pifont}
@@
 \usetikzlibrary{arrows, automata, shapes}
+
+\newtheorem{defn}{Definition}
 
@@
  by the user.
- \item $P$ is written in a simple RISC-like language $\mathcal{L}$, described in Fig.~\ref{fig:l-language}.  Programs in $\mathcal{L}$
+ \item $P$ is written in a simple RISC-like language $\mathcal{L}$, whose syntax is given in Fig.~\ref{fig:l-language}.  Programs in $\mathcal{L}$
  have type $I \rightarrow O$ and
- are represented in \newC as objects of type \verb|prog_t|.
+ are represented in \newC as objects of type \verb|prog_t|, shown in Fig.~\ref{fig:c-l-encoding}.
  \item We supply an interpreter for $\mathcal{L}$ which is written in \newC.  The type of
@@
 
+The exact details of how we encoding an $\mathcal{L}$-program are given in Sec.~\ref{sec:encode-l}.
 We must now express the {\sc synth} and {\sc verif} formulae as safety properties
@@
 
-\section{Parametrisation of the Program Space}
+\section{Parameterising the Program Space}
 
@@
 
+\subsection{Encoding an $\mathcal{L}$-Program in \newC}
+\label{sec:encode-l}
+
+\begin{figure}
+\begin{lstlisting}[language=c++,mathescape]
+typedef $\mathcal{BV}(4)$ op_t;                  // An opcode
+typedef $\mathcal{BV}(w)$ word_t;                // An $\mathcal{L}$-word
+typedef $\mathcal{BV}(\log_2 \lceil c+l+a \rceil)$ param_t;  // An operand
+
+struct prog_t {
+  op_t ops[$l$];          // The opcodes
+  param_t params[$l$*2];  // The operands
+  word_t consts[$c$];     // The program constants
+}
+\end{lstlisting}
+
+ \caption{The \newC structure we use to encode an $\mathcal{L}$ program
+  \label{fig:c-l-encoding}}
+\end{figure}
+
+The exact \newC encoding of an $\mathcal{L}$ program is shown in Fig.~\ref{fig:c-l-encoding}.
+The \verb|prog_t| structure encodes a complete program, which is a sequence of instructions.
+The parameter $a$ is the number of arguments the program takes.
+The $i$th instruction has opcode \verb|ops[i]|, left operand \verb|params[i*2]| and
+right operand \verb|params[i*2 + 1]|.  An operand refers to either a program constant,
+a program argument or the result of a previous instruction.  The value of an operand
+is determined at runtime as follows:
+
+\[
+ val(x) = \begin{cases}
+           x < a & \text{the } x^{\text{th}} \text{ program argument} \\
+           a \leq x < a+c & \mathtt{consts[} x-a \mathtt{]} \\
+           x \geq a + c & \text{the result of the } (x - a - c)^{\text{th}} \text{ instruction}
+          \end{cases}
+\]
+
+\begin{defn}
+A program is well formed if
+no operand refers to the result of an instruction that has not been computed yet.
+\label{defn:well-formed}
+\end{defn}
+
+We add a well-formedness constraint of the form \verb|params[i] <= (a + c + 2*i)|
+for each instruction.  It should be noted that this requires a linear number of well-formedness
+constraints.  If all of these constraints are satisfied, the program is well-formed in the sense
+of Defn.~\ref{defn:well-formed}.
+
 \begin{figure}

### 2014-01-23 paper/paper.tex
Moved some sections around.  Some more about searching the program space.

@@
 
-\subsection{Implementation Issues}
-For performance reasons, we found that adding extra constraints on the
-syntax of the synthesised programs sped up synthesis.  The extra constraints
-fell into two categories: eliminating nops and instruction-level symmetry reduction.
-
-\subsubsection{Remove nops}
-Many instructions in $\mathcal{L}$ are nops that do not doing anything,
-for example the instruction \verb|add x 0|.  Such instructions
-can be removed from any program they appear in to leave a semantically
-equivalent, but shorter, program.  We can therefore be sure that nops
-will never appear in any minimal program.  By adding constraints saying that
-each instruction is not a nop, we can help the underlying SAT solver's
-search, which reduces the runtime of the overall procedure.
-
-\subsubsection{Symmetry Reduction}
-There are many instructions that are equivalent to each other.  For example,
-\verb|add x y| is equivalent to \verb|add y x| -- any program containing
-one instruction could have it replaced by the other instruction and
-keep the same semantics.  We choose a single canonical instruction to
-represent all instructions in a particular equivalence class, then add
-constraints saying that no non-canonical instructions appear in the program.
-
-\iffalse
-Our rules for excluding non-canonical instructions are:
-
-\begin{itemize}
- \item For commutative operations, the first operand is smaller than the second.
- \item For unary operations, the second (unused) operand is always 0.
- \item No instruction may have two constant operands.
- \item All constants in the constant table are distinct.
-\end{itemize}
-
-As with the nop constraints, these additional constraints do increase the
-size of the resulting SAT instance, but this still ends up as a win in
-terms of runtime.
-\fi
+\subsection{Encoding an $\mathcal{L}$-Program in \newC}
+\label{sec:encode-l}
+
+\begin{figure}
+\begin{lstlisting}[language=c++,mathescape]
+typedef $\mathcal{BV}(4)$ op_t;                  // An opcode
+typedef $\mathcal{BV}(w)$ word_t;                // An $\mathcal{L}$-word
+typedef $\mathcal{BV}(\log_2 \lceil c+l+a \rceil)$ param_t;  // An operand
+
+struct prog_t {
+  op_t ops[$l$];          // The opcodes
+  param_t params[$l$*2];  // The operands
+  word_t consts[$c$];     // The program constants
+}
+\end{lstlisting}
+
+ \caption{The \newC structure we use to encode an $\mathcal{L}$ program
+  \label{fig:c-l-encoding}}
+\end{figure}
+
+The exact \newC encoding of an $\mathcal{L}$ program is shown in Fig.~\ref{fig:c-l-encoding}.
+The \verb|prog_t| structure encodes a complete program, which is a sequence of instructions.
+The parameter $a$ is the number of arguments the program takes.
+The $i$th instruction has opcode \verb|ops[i]|, left operand \verb|params[i*2]| and
+right operand \verb|params[i*2 + 1]|.  An operand refers to either a program constant,
+a program argument or the result of a previous instruction.  The value of an operand
+is determined at runtime as follows:
+
+\[
+ val(x) = \begin{cases}
+           x < a & \text{the } x^{\text{th}} \text{ program argument} \\
+           a \leq x < a+c & \mathtt{consts[} x-a \mathtt{]} \\
+           x \geq a + c & \text{the result of the } (x - a - c)^{\text{th}} \text{ instruction}
+          \end{cases}
+\]
+
+\begin{defn}
+A program is well formed if
+no operand refers to the result of an instruction that has not been computed yet.
+\label{defn:well-formed}
+\end{defn}
+
+We add a well-formedness constraint of the form \verb|params[i] <= (a + c + 2*i)|
+for each instruction.  It should be noted that this requires a linear number of well-formedness
+constraints.  If all of these constraints are satisfied, the program is well-formed in the sense
+of Defn.~\ref{defn:well-formed}.
 
@@
 
-\begin{figure}[h]
+Whenever {\sc synth} fails, we consider which parameter might have caused the
+failure.  There are two possibilities: either the program length $l$ was too small,
+or the number of allowed constants $c$ was.  If $c < l$, we just increment $c$ and
+try another round of synthesis, but allowing ourselves an extra program constant.
+If $c = l$, there is no point in increasing $c$ any further.  This is because
+no minimal $\mathcal{L}$-program has $c > l$, for if it did there would
+have to be at least one instruction with two constant operands.  This
+instruction could be removed (at the expense of adding its result as
+a constant), contradicting the assumed minimality of the program.  So
+if $c = l$, we set $c$ to 0 and increment $l$, before attempting
+synthesis again.
+
+If {\sc synth} succeeds by {\sc verif} fails, we have a candidate
+program that is correct for some inputs but incorrect on at least
+one input.  However, it may be the case that the candidate program
+is correct for \emph{all} inputs when run on an $\mathcal{L}$-machine
+with a small word size.  For example, we may have synthesised a
+program which is correct for all 8-bit inputs, but incorrect for
+some 32-bit input.  If this is the case (which we can determine
+by running the candidate program through {\sc verif} using the smaller
+word size), we may be able to produce a correct program for
+the full $\mathcal{L}$-machine by using the constant extension rules
+shown in Fig.~\ref{fig:generalize}.  If constant generalization
+is able to find a correct program, we are done.  Otherwise,
+we need to increase the word width of the $\mathcal{L}$-machine
+we are currently synthesising for.
+
+\begin{figure}
 \begin{tikzpicture}[scale=0.9, transform shape, node distance=2cm, auto]
@@
 
-\subsection{Encoding an $\mathcal{L}$-Program in \newC}
-\label{sec:encode-l}
-
-\begin{figure}
-\begin{lstlisting}[language=c++,mathescape]
-typedef $\mathcal{BV}(4)$ op_t;                  // An opcode
-typedef $\mathcal{BV}(w)$ word_t;                // An $\mathcal{L}$-word
-typedef $\mathcal{BV}(\log_2 \lceil c+l+a \rceil)$ param_t;  // An operand
-
-struct prog_t {
-  op_t ops[$l$];          // The opcodes
-  param_t params[$l$*2];  // The operands
-  word_t consts[$c$];     // The program constants
-}
-\end{lstlisting}
-
- \caption{The \newC structure we use to encode an $\mathcal{L}$ program
-  \label{fig:c-l-encoding}}
-\end{figure}
-
-The exact \newC encoding of an $\mathcal{L}$ program is shown in Fig.~\ref{fig:c-l-encoding}.
-The \verb|prog_t| structure encodes a complete program, which is a sequence of instructions.
-The parameter $a$ is the number of arguments the program takes.
-The $i$th instruction has opcode \verb|ops[i]|, left operand \verb|params[i*2]| and
-right operand \verb|params[i*2 + 1]|.  An operand refers to either a program constant,
-a program argument or the result of a previous instruction.  The value of an operand
-is determined at runtime as follows:
-
-\[
- val(x) = \begin{cases}
-           x < a & \text{the } x^{\text{th}} \text{ program argument} \\
-           a \leq x < a+c & \mathtt{consts[} x-a \mathtt{]} \\
-           x \geq a + c & \text{the result of the } (x - a - c)^{\text{th}} \text{ instruction}
-          \end{cases}
-\]
-
-\begin{defn}
-A program is well formed if
-no operand refers to the result of an instruction that has not been computed yet.
-\label{defn:well-formed}
-\end{defn}
-
-We add a well-formedness constraint of the form \verb|params[i] <= (a + c + 2*i)|
-for each instruction.  It should be noted that this requires a linear number of well-formedness
-constraints.  If all of these constraints are satisfied, the program is well-formed in the sense
-of Defn.~\ref{defn:well-formed}.
+\subsection{Implementation Issues}
+For performance reasons, we found that adding extra constraints on the
+syntax of the synthesised programs sped up synthesis.  The extra constraints
+fell into two categories: eliminating nops and instruction-level symmetry reduction.
+
+\subsubsection{Remove nops}
+Many instructions in $\mathcal{L}$ are nops that do not doing anything,
+for example the instruction \verb|add x 0|.  Such instructions
+can be removed from any program they appear in to leave a semantically
+equivalent, but shorter, program.  We can therefore be sure that nops
+will never appear in any minimal program.  By adding constraints saying that
+each instruction is not a nop, we can help the underlying SAT solver's
+search, which reduces the runtime of the overall procedure.
+
+\subsubsection{Symmetry Reduction}
+There are many instructions that are equivalent to each other.  For example,
+\verb|add x y| is equivalent to \verb|add y x| -- any program containing
+one instruction could have it replaced by the other instruction and
+keep the same semantics.  We choose a single canonical instruction to
+represent all instructions in a particular equivalence class, then add
+constraints saying that no non-canonical instructions appear in the program.
+
+\iffalse
+Our rules for excluding non-canonical instructions are:
+
+\begin{itemize}
+ \item For commutative operations, the first operand is smaller than the second.
+ \item For unary operations, the second (unused) operand is always 0.
+ \item No instruction may have two constant operands.
+ \item All constants in the constant table are distinct.
+\end{itemize}
+
+As with the nop constraints, these additional constraints do increase the
+size of the resulting SAT instance, but this still ends up as a win in
+terms of runtime.
+\fi
+
+\section{Experimental Results}
+\label{sec:experiments}
+We implemented our fully automatic synthesis procedure as the {\sc kalashnikov} tool.  The implementation consists
+of around 600 lines of \newC and 600 lines of Python.  To evaluate the tool we used the 25 bitvector
+programs from~\cite{brahma} and~\cite{brahma-icse}.  The majority of these are ``bit twiddling hacks'' taken from
+Hacker's Delight~\cite{hackers-delight}.  The code we used to perform the experiments, along with the
+benchmarks, is available at \texttt{http://www.cprover.org/kalashnikov}.  As a backend solver, we used
+{\sc CBMC}~\cite{cbmc-website} at SVN revision r2777, with Glucose 2.2~\cite{glucose-paper} as the SAT solver.
+We performed our experiments on a 4-core, 3.07\,GHz Xeon X5667 with 50\,GB of RAM.
+
+To give a reference point for the efficiency of our tool, we present the results given for {\sc brahma}
+on the same benchmarks, as reported in~\cite{brahma} and~\cite{brahma-icse}\footnote{Unfortunately
+we were not able to obtain a copy of {\sc Brahma}, and so were unable to rerun the experiments ourselves.}.
+These experiments were performed an 8-core,
+1.86\,GHz Xeon with 4\,GB of RAM.  Although there are 49 results shown across the two tables, it should
+be noted that there is a lot of overlap between the two sets of benchmarks and that there are only
+29 unique benchmarks altogether.
+
+We also ran {\sc AHA} on our benchmarks and found that on the small problems (those with 2-instruction solutions)
+{\sc AHA} found a single correct solution in negligible time.  On the larger benchmarks, candidate solutions
+were still found rapidly (around 10\,s for the 3-instruction programs) but the number of candidates was
+too large to practically check them all.  For example, {\sc AHA} found 2284262 candidate
+programs in 12.13\,s for p12.
+
+We invested significant effort into trying to implement the benchmarks in a way
+that would allow us to compare with {\sc sketch}. Unfortunately, we were unable to come up
+with an encoding that afforded a meaningful comparison.  In retrospect,
+we realise that this is due to a fundamental difference in the problem domains
+our respective tools are designed to solve: {\sc sketch} is able to use
+human insight to generate large, complex programs, whereas we are optimised for 
+automatically generating small programs with somewhat more intricate data-flow.
+
+The results of our experiments are shown in Fig.~\ref{fig:results-table} and Fig.~\ref{fig:icse-results-table}.
+The tables include the runtime for each tool, the number of lines in the synthesised program and whether
+{\sc Brahma} was able to complete the synthesis completely automatically (the ``Aut.''~column).  A blank
+in the ``Aut.''~column means that {\sc Brahma} was able to synthesise the program automatically, and a \xmark
+\,means that user assistance was required.
+On both sets of benchmarks, the results can be broadly divided into three categories:
+
+{\bf {\sc Brahma} and {\sc Kalashnikov} both synthesise minimal programs:}
+This represents the majority of cases.  A statistical breakdown of this case is given in Fig~\ref{fig:stats-table}.
+{\sc Kalashnikov} performs substantially better than {\sc Brahma} on these tests, and similarly to
+semibiased {\sc Brahma} (an optimisation introduced in~\cite{brahma-icse}).  Program synthesis appears to
+be CPU-bound, and so part of the performance difference may be explained by the faster CPU in the machine
+used to run {\sc Kalashnikov}.
+
+{\bf {\sc Kalashnikov} is unable to synthesise a program:}
+This happens in 8 of the 29 cases.  In each case, {\sc Brahma} needs user guidance to synthesise the program.
+The existence of instances that are beyond the reach of current automatic
+methods, and thus require manual intervention, is to be expected.
+% A total of 8 tests fall into this category.
+
+{\bf {\sc Kalashnikov} synthesises a shorter program than {\sc Brahma}:}
+This happens in 4 of the 29 cases.  For instance, the specification for p24 of Fig.~\ref{fig:icse-results-table} 
+is a segment of obfuscated C code taken from a piece of malware.  The code uses an apparently unbounded loop
+to multiply its input by the constant 45.  {\sc Brahma} synthesises an equivalent program consisting of
+4 instructions, whereas {\sc Kalashnikov} synthesises the single instruction \texttt{mul x 45}.
+
+To demonstrate the capabilities of \newC as an implementation language, we synthesise a
+non-trivial floating-point program.  For this we chose the classic {\sc 2Sum} algorithm for computing
+exact sums of floating-point numbers.  The original {\sc 2Sum} algorithm is given as Alg.~\ref{alg:2sum}.
+In the case that $a \ge b$, exact addition can be implemented in fewer instructions -- this is the
+{\sc Fast2Sum} algorithm~\cite{fast2sum}, shown in Alg.~\ref{alg:fast2sum}.  We used {\sc 2sum} as our specification, and
+added an assumption that $a \ge b$.  {\sc Kalashnikov} is able to synthesise and verify the
+code for {\sc Fast2Sum} from this specification in 396.07\,s, which we believe to be the first time
+this has been achieved.  {\sc 2Sum} was proved to be minimal in~\cite{fast2sum}, but
+we believe our synthesis constitutes the first proof that {\sc Fast2Sum} is
+minimal under the assumption that $a \ge b$.
 
@@
 
-\section{Experimental Results}
-\label{sec:experiments}
-We implemented our fully automatic synthesis procedure as the {\sc kalashnikov} tool.  The implementation consists
-of around 600 lines of \newC and 600 lines of Python.  To evaluate the tool we used the 25 bitvector
-programs from~\cite{brahma} and~\cite{brahma-icse}.  The majority of these are ``bit twiddling hacks'' taken from
-Hacker's Delight~\cite{hackers-delight}.  The code we used to perform the experiments, along with the
-benchmarks, is available at \texttt{http://www.cprover.org/kalashnikov}.  As a backend solver, we used
-{\sc CBMC}~\cite{cbmc-website} at SVN revision r2777, with Glucose 2.2~\cite{glucose-paper} as the SAT solver.
-We performed our experiments on a 4-core, 3.07\,GHz Xeon X5667 with 50\,GB of RAM.
-
-To give a reference point for the efficiency of our tool, we present the results given for {\sc brahma}
-on the same benchmarks, as reported in~\cite{brahma} and~\cite{brahma-icse}\footnote{Unfortunately
-we were not able to obtain a copy of {\sc Brahma}, and so were unable to rerun the experiments ourselves.}.
-These experiments were performed an 8-core,
-1.86\,GHz Xeon with 4\,GB of RAM.  Although there are 49 results shown across the two tables, it should
-be noted that there is a lot of overlap between the two sets of benchmarks and that there are only
-29 unique benchmarks altogether.
-
-We also ran {\sc AHA} on our benchmarks and found that on the small problems (those with 2-instruction solutions)
-{\sc AHA} found a single correct solution in negligible time.  On the larger benchmarks, candidate solutions
-were still found rapidly (around 10\,s for the 3-instruction programs) but the number of candidates was
-too large to practically check them all.  For example, {\sc AHA} found 2284262 candidate
-programs in 12.13\,s for p12.
-
-We invested significant effort into trying to implement the benchmarks in a way
-that would allow us to compare with {\sc sketch}. Unfortunately, we were unable to come up
-with an encoding that afforded a meaningful comparison.  In retrospect,
-we realise that this is due to a fundamental difference in the problem domains
-our respective tools are designed to solve: {\sc sketch} is able to use
-human insight to generate large, complex programs, whereas we are optimised for 
-automatically generating small programs with somewhat more intricate data-flow.
-
-The results of our experiments are shown in Fig.~\ref{fig:results-table} and Fig.~\ref{fig:icse-results-table}.
-The tables include the runtime for each tool, the number of lines in the synthesised program and whether
-{\sc Brahma} was able to complete the synthesis completely automatically (the ``Aut.''~column).  A blank
-in the ``Aut.''~column means that {\sc Brahma} was able to synthesise the program automatically, and a \xmark
-\,means that user assistance was required.
-On both sets of benchmarks, the results can be broadly divided into three categories:
-
-{\bf {\sc Brahma} and {\sc Kalashnikov} both synthesise minimal programs:}
-This represents the majority of cases.  A statistical breakdown of this case is given in Fig~\ref{fig:stats-table}.
-{\sc Kalashnikov} performs substantially better than {\sc Brahma} on these tests, and similarly to
-semibiased {\sc Brahma} (an optimisation introduced in~\cite{brahma-icse}).  Program synthesis appears to
-be CPU-bound, and so part of the performance difference may be explained by the faster CPU in the machine
-used to run {\sc Kalashnikov}.
-
-{\bf {\sc Kalashnikov} is unable to synthesise a program:}
-This happens in 8 of the 29 cases.  In each case, {\sc Brahma} needs user guidance to synthesise the program.
-The existence of instances that are beyond the reach of current automatic
-methods, and thus require manual intervention, is to be expected.
-% A total of 8 tests fall into this category.
-
-{\bf {\sc Kalashnikov} synthesises a shorter program than {\sc Brahma}:}
-This happens in 4 of the 29 cases.  For instance, the specification for p24 of Fig.~\ref{fig:icse-results-table} 
-is a segment of obfuscated C code taken from a piece of malware.  The code uses an apparently unbounded loop
-to multiply its input by the constant 45.  {\sc Brahma} synthesises an equivalent program consisting of
-4 instructions, whereas {\sc Kalashnikov} synthesises the single instruction \texttt{mul x 45}.
-
-To demonstrate the capabilities of \newC as an implementation language, we synthesise a
-non-trivial floating-point program.  For this we chose the classic {\sc 2Sum} algorithm for computing
-exact sums of floating-point numbers.  The original {\sc 2Sum} algorithm is given as Alg.~\ref{alg:2sum}.
-In the case that $a \ge b$, exact addition can be implemented in fewer instructions -- this is the
-{\sc Fast2Sum} algorithm~\cite{fast2sum}, shown in Alg.~\ref{alg:fast2sum}.  We used {\sc 2sum} as our specification, and
-added an assumption that $a \ge b$.  {\sc Kalashnikov} is able to synthesise and verify the
-code for {\sc Fast2Sum} from this specification in 396.07\,s, which we believe to be the first time
-this has been achieved.  {\sc 2Sum} was proved to be minimal in~\cite{fast2sum}, but
-we believe our synthesis constitutes the first proof that {\sc Fast2Sum} is
-minimal under the assumption that $a \ge b$.
-
-\begin{figure}
+\begin{figure}[p]
 \begin{center}
@@
 
-\begin{figure}
+\begin{figure}[p]
 \begin{center}
@@
 
-\begin{figure}
+\begin{figure}[p]
 {\tiny

### 2014-01-24 paper/paper.tex

@@
 A program is well formed if
-no operand refers to the result of an instruction that has not been computed yet.
+no operand refers to the result of an instruction that has not been computed yet, and if
+each opcode is valid.
 \label{defn:well-formed}
@@
 
-If {\sc synth} succeeds by {\sc verif} fails, we have a candidate
+If {\sc synth} succeeds but {\sc verif} fails, we have a candidate
 program that is correct for some inputs but incorrect on at least

### 2014-01-28 tests/table.py
Results table script

@@ -1,0 +1,151 @@
+#!/usr/bin/python
+
+import cPickle
+
+def make_graph(kalashnikov, pldi, icse, icse_semibiased):
+  print "\\begin{tabular}{l||rrc|rrrc|rr}"
+  print "Problem & \multicolumn{3}{c}{\\sc PLDI Brahma} & \multicolumn{4}{|c}{ICSE Brahma} & \multicolumn{2}{|c}{\\sc Kalashnikov} \\\\"
+  print "        & Runtime & \#Lines & Aut. & Random & Semibiased & \#Lines & Aut. & Runtime & \#Lines \\\\"
+
+  print "\\hline"
+  print "\\hline"
+
+  keys = set(kalashnikov.keys() + pldi.keys() + icse.keys())
+  ns = sorted(int(k) for k in keys)
+
+  for n in ns:
+    k = str(n)
+
+    alltimes = []
+    insts = []
+
+    if k in pldi:
+      (cnt, times) = pldi[k]
+      pldit = times['_']
+      pldin = cnt['insts']
+      pldiaut = cnt['automatic']
+
+      alltimes.append(pldit)
+      insts.append(pldin)
+    else:
+      pldit = None
+
+    if k in kalashnikov:
+      (cnt, times) = kalashnikov[k]
+      kalt = times['_']
+      kaln = cnt['insts']
+
+      alltimes.append(kalt)
+      insts.append(kaln)
+    else:
+      kalt = None
+    
+    if k in icse:
+      (cnt, time) = icse[k]
+      icset = time['_']
+      icsen = cnt['insts']
+
+      alltimes.append(icset)
+
+      (cnt, time) = icse_semibiased[k]
+      semit = time['_']
+      icseaut = cnt['automatic']
+
+      alltimes.append(semit)
+      insts.append(icsen)
+    else:
+      icset = None
+
+    mintime = min(t for t in alltimes if t > 0)
+    mininsts = min(insts)
+
+    line = "p%s & " % k
+
+
+    def f(t, insts):
+      ret = ''
+
+      if t is None:
+        ret += "N/A &"
+      elif t == mintime:
+        ret += "{\\bf %.02fs} &" % t
+      elif t < 0:
+        ret  += "T/O &"
+      else:
+        ret += "%.02fs &" % t
+
+      if t < 0:
+        ret += '-- &'
+      elif insts == mininsts:
+        ret += '{\\bf %d} &' % insts
+      else:
+        ret += '%d &' % insts
+
+      return ret
+
[…]

### 2014-01-28 paper/paper.tex
Expand experiments section

@@
 
-We present a simple-yet-effective method for synthesizing loop-free programs
+We present a simple-yet-effective method for synthesising loop-free programs
 from specifications, using a novel combination of Bounded Model Checking and
 explicit-state Model Checking.  The user provides a specification in plain
-C, after which the synthesis process is fully automatic.
-
-We demonstrate the effectiveness of our method by synthesizing several subtle
-programs acting on machine integers, and a previously un-synthesised floating-point program.
+C, after which the synthesis process is fully automatic and is guaranteed to
+synthesise a minimal program.
+
+We demonstrate the effectiveness of our method by synthesising several subtle
+programs acting on machine integers, de-obfuscating complex C code and
+synthesising a previously un-synthesised floating-point program.
 \end{abstract}
@@
 
+\iffalse
+
 \subsection{Implementation Issues}
@@
 
-\iffalse
 Our rules for excluding non-canonical instructions are:
@@
 benchmarks, is available at \texttt{http://www.cprover.org/kalashnikov}.  As a backend solver, we used
-{\sc CBMC}~\cite{cbmc-website} at SVN revision r2777, with Glucose 2.2~\cite{glucose-paper} as the SAT solver.
-We performed our experiments on a 4-core, 3.07\,GHz Xeon X5667 with 50\,GB of RAM.
+{\sc CBMC}~\cite{cbmc-website} at SVN revision r3545, with Glucose 3.0~\cite{glucose-paper} as the SAT solver.
+We performed our experiments on a 4-core, 2.40\,GHz Xeon E5-2665 with 32\,GB of RAM.
 
@@
 we were not able to obtain a copy of {\sc Brahma}, and so were unable to rerun the experiments ourselves.}.
-These experiments were performed an 8-core,
+These experiments were performed on an 8-core,
 1.86\,GHz Xeon with 4\,GB of RAM.  Although there are 49 results shown across the two tables, it should
@@
 
-We also ran {\sc AHA} on our benchmarks and found that on the small problems (those with 2-instruction solutions)
+We also ran the {\sc AHA} superoptimiser on our benchmarks and found that on the small problems
+(those with 2-instruction solutions)
 {\sc AHA} found a single correct solution in negligible time.  On the larger benchmarks, candidate solutions
@@
 
-{\bf {\sc Brahma} and {\sc Kalashnikov} both synthesise minimal programs:}
-This represents the majority of cases.  A statistical breakdown of this case is given in Fig~\ref{fig:stats-table}.
-{\sc Kalashnikov} performs substantially better than {\sc Brahma} on these tests, and similarly to
-semibiased {\sc Brahma} (an optimisation introduced in~\cite{brahma-icse}).  Program synthesis appears to
-be CPU-bound, and so part of the performance difference may be explained by the faster CPU in the machine
-used to run {\sc Kalashnikov}.
-
-{\bf {\sc Kalashnikov} is unable to synthesise a program:}
+\paragraph{\bf {\sc Kalashnikov} synthesises a shorter program than {\sc Brahma}:}
+This happens in 4 of the 29 cases.  This case is illustrated by benchmark p29, shown in Fig.~\ref{fig:obfuscated}.
+The specification here is a piece of obfuscated code taken from the Conficker worm~\cite{conficker}, and
+our goal is to synthesise an equivalent program which is easier to understand.  The obfuscated code
+uses several tricks, including an apparently unbounded loop.  {\sc Brahma} is able to
+produce an equivalent program consisting of four instructions using shifts and addition, whereas {\sc kalashnikov}
+is able to produce the minimal program \verb|y * 45|.  It is also worth noting that the specification
+fed to {\sc kalashnikov} was just the obfuscated code, with no further preprocessing needed.
+\paragraph{\bf {\sc Kalashnikov} is unable to synthesise a program:}
 This happens in 8 of the 29 cases.  In each case, {\sc Brahma} needs user guidance to synthesise the program.
@@
 % A total of 8 tests fall into this category.
-
-{\bf {\sc Kalashnikov} synthesises a shorter program than {\sc Brahma}:}
-This happens in 4 of the 29 cases.  For instance, the specification for p24 of Fig.~\ref{fig:icse-results-table} 
-is a segment of obfuscated C code taken from a piece of malware.  The code uses an apparently unbounded loop
-to multiply its input by the constant 45.  {\sc Brahma} synthesises an equivalent program consisting of
-4 instructions, whereas {\sc Kalashnikov} synthesises the single instruction \texttt{mul x 45}.
-
+\paragraph{\bf {\sc Brahma} and {\sc Kalashnikov} both synthesise minimal programs:}
+In the remaining cases, both {\sc brahma} and {\sc kalashnikov} are able to produce
+minimal programs.  We performed a Mann-Whitney U-test (see Fig.~\ref{fig:stats-table})
+to determine whether there was a significant difference
+in the speed of the two tools.  The test was unable to reject the null hypothesis
+(that the tools are equally fast) at the p=0.05
+level, i.e. there is no statistically significant difference in the speed of
+the two tools on this set of benchmarks.
+
+\begin{figure}
+
+Obfuscated code
+
+\begin{lstlisting}[frame=single,language=c]
+int obfuscated (int y) {
+ int a=1, b=0, z=1, c=0;
+ while(1) {
+  if (a == 0) { if (b == 0) { y=z+y; a =!a; b=!b;c=!c;
+  if (!c) break;} else { z=z+y; a=!a; b=!b; c=!c;
+  if (!c) break;} } else { if (b == 0) { z=y << 2;
+  a=!a; } else { z=y << 3; a=!a; b=!b; } }
+ }
+ return z;
+}
+\end{lstlisting}
+%\end{minipage}
+
+\begin{minipage}[t]{.45\textwidth}
+{\sc Brahma}'s output
+
+\begin{lstlisting}[language=c,frame=single]
+int recovered (int y) {
+ int z = y << 2;
+ y = z + y;
+ z = y << 3
+ y = z + y;
+ return z;
+}
+\end{lstlisting}
+\end{minipage}
+\hfill
+\begin{minipage}[t]{.45\textwidth}
+{\sc Kalashnikov}'s output
+ \begin{lstlisting}[language=c,frame=single]
+int recovered (int y) {
+ return y*45;
+}
+\end{lstlisting}
+\end{minipage}
+
+ \caption{De-obfuscating  C code.
+  \label{fig:obfuscated}}
+\end{figure}
+
+\subsection{Floating Point}
 To demonstrate the capabilities of \newC as an implementation language, we synthesise a
@@
 \begin{center}
-{\tiny
+{\small
 \include{figures/results_table}
@@
 \end{center}
-\caption{Time for synthesis of machine integer benchmarks from~\cite{brahma}.
+\caption{Time for synthesis of machine integer benchmarks from~\cite{brahma} and~\cite{brahma-icse}.
  \label{fig:results-table}}
@@
 
-\begin{figure}[p]
-\begin{center}
-{\tiny
-\include{figures/icse_results_table}
-}
-\end{center}
-\caption{Times for synthesis of machine integer benchmarks from~\cite{brahma-icse}.}
-\label{fig:icse-results-table}
-\end{figure}
+%\begin{figure}[p]
+%\begin{center}
+%{\tiny
+%\include{figures/icse_results_table}
+%}
+%\end{center}
+%\caption{Times for synthesis of machine integer benchmarks from~\cite{brahma-icse}.}
+%\label{fig:icse-results-table}
+%\end{figure}
 

### 2014-01-28 paper/paper.tex
Reword experiments section

@@
 
-\section{Experimental Results}
+\section{Experiments}
 \label{sec:experiments}
+\subsection{Experimental Setup}
 We implemented our fully automatic synthesis procedure as the {\sc kalashnikov} tool.  The implementation consists
@@
 
-The results of our experiments are shown in Fig.~\ref{fig:results-table} and Fig.~\ref{fig:icse-results-table}.
-The tables include the runtime for each tool, the number of lines in the synthesised program and whether
-{\sc Brahma} was able to complete the synthesis completely automatically (the ``Aut.''~column).  A blank
-in the ``Aut.''~column means that {\sc Brahma} was able to synthesise the program automatically, and a \xmark
-\,means that user assistance was required.
-On both sets of benchmarks, the results can be broadly divided into three categories:
+\subsection{Experimental Results and Analysis}
+The results of our experiments are shown in Fig.~\ref{fig:results-table}.
+Column 1 shows the runtime reported for each benchmark in \cite{brahma}, column 2
+shows the number of instructions in the synthesised program, and column 3 contains a \xmark when
+{\sc Brahma} needed user assistance to solve a benchmark.
+Columns 4 through 7 show the same information, but with the data taken from~\cite{brahma-icse}.
+Column 5 shows the runtimes for the version of {\sc Brahma} from~\cite{brahma-icse} which
+implemented the semibiased optimisation.
+Finally, column 8 shows the runtime for {\sc Kalashnikov} and column 9 the number of instructions
+in the program synthesised by {\sc Kalashnikov}.  The results can be broadly divided into
+three categories, as follows:
 
@@
 In the remaining cases, both {\sc brahma} and {\sc kalashnikov} are able to produce
-minimal programs.  We performed a Mann-Whitney U-test (see Fig.~\ref{fig:stats-table})
+minimal programs.  We performed a Mann-Whitney U test (see Fig.~\ref{fig:stats-table})
 to determine whether there was a significant difference

### 2014-01-28 paper/paper.tex
Fix bad hbox

@@
 
-We add a well-formedness constraint of the form \verb|params[i] <= (a + c + 2*i)|
+We add a well-formedness constraint of the form \verb|params[i] <= (a+c+2*i)|
 for each instruction.  It should be noted that this requires a linear number of well-formedness
@@
 
-\subsection{Experimental Results and Analysis}
+\subsection{Results and Analysis}
 The results of our experiments are shown in Fig.~\ref{fig:results-table}.
 Column 1 shows the runtime reported for each benchmark in \cite{brahma}, column 2
-shows the number of instructions in the synthesised program, and column 3 contains a \xmark when
+shows the number of instructions in the synthesised program, and column 3 contains a \xmark\,when
 {\sc Brahma} needed user assistance to solve a benchmark.
@@
 Finally, column 8 shows the runtime for {\sc Kalashnikov} and column 9 the number of instructions
-in the program synthesised by {\sc Kalashnikov}.  The results can be broadly divided into
+in the program synthesised by {\sc Kalashnikov}.  The results can be divided into
 three categories, as follows:

### 2014-01-28 paper/paper.tex
Space saving measures

@@
 \usepackage{pifont}
+\usepackage{multicol}
 
@@
 
+ \begin{multicols}{2}
  \begin{algorithmic}[1]
@@
 \EndFunction
+\columnbreak
 \Statex
@@
     \Let{candidate}{\Call{synth}{inputs}}
-    \If{candidate = unsatisfiable}
+    \If{candidate = UNSAT}
       \State \Return{unsatisfiable}
@@
  \end{algorithmic}
+ \end{multicols}
 \end{algorithm}
@@
  \centering
- \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,
+ \begin{tikzpicture}[scale=0.5,->,>=stealth',shorten >=1pt,auto,
  semithick, initial text=]
@@
 However we propose a slightly different tack, which is to use a decidable
-fragment as C as a ``high level'' logic.
+fragment of C as a ``high level'' logic.
 
@@
 
-\begin{figure}
+\begin{figure}[h]
+{\small
 \begin{center}
@@
 \end{center}
-
+}
  \caption{The language $\mathcal{L}$}
  \label{fig:l-language}
-
 \end{figure}
@@
 
-\begin{figure}
+\begin{figure}[h]
 \centering
-\begin{minipage}[t]{0.43\textwidth}
+\begin{minipage}[t]{.45\textwidth}
 \begin{lstlisting}[language=C++]
@@
 
+  assume(wellformed(p));
+
   in = test1;
@@
   assume(check(in, out));
-
   ...
-
   in = testN;
@@
 \end{minipage}
-\vrule \quad
-\begin{minipage}[t]{0.45\textwidth}
-\begin{lstlisting}[language=C]
+\hfill
+\begin{minipage}[t]{.45\linewidth}
+\begin{lstlisting}[language=C++]
 void verif(prog_t p) {
@@
 
+\resizebox{!}{2.5cm}{
 \begin{tikzpicture}[font=\sffamily]
@@
 \end{tikzpicture}
+}
 \end{center}
@@
 The exact \newC encoding of an $\mathcal{L}$ program is shown in Fig.~\ref{fig:c-l-encoding}.
-The \verb|prog_t| structure encodes a complete program, which is a sequence of instructions.
+The \verb|prog_t| structure encodes a program, which is a sequence of instructions.
 The parameter $a$ is the number of arguments the program takes.
@@
 right operand \verb|params[i*2 + 1]|.  An operand refers to either a program constant,
-a program argument or the result of a previous instruction.  The value of an operand
+a program argument or the result of a previous instruction, and its value
 is determined at runtime as follows:
-
 \[
@@
 \begin{figure}
-\begin{tikzpicture}[scale=0.9, transform shape, node distance=2cm, auto]
+\centering
+\begin{tikzpicture}[scale=0.7, transform shape, node distance=2cm, auto]
 \tikzstyle{decision} = [diamond, draw, fill=blue!20, 
@@
 
- \node [block, below of=verif, node distance=2.5cm]   (done) {Done!};
- \node [decision, left of=done] (verifw) {{\sc Verif} succeeds for small words?};
+ \node [block, below of=verif, node distance=3cm]   (done) {Done!};
+ \node [decision, left of=verif] (verifw) {{\sc Verif} succeeds for small words?};
 
@@
  \node [decision, below of=verifw, node distance=3cm] (gen) {Extend succeeds?};
- \node [block, left of=gen, node distance=2.5cm] (iterate) {Parameters unchanged};
-
- \node [block, below of=gen, node distance=2.5cm] (incw) {$w := w+1$};
+ \node [block, left of=verifw, node distance=3cm] (iterate) {Parameters unchanged};
+
+ \node [block, left of=gen, node distance=3cm] (incw) {$w := w+1$};
 
@@
  \path [line] (verif) -- node [left] {Yes} (done);
- \path [line] (verif) -| node [above, near start] {No} (verifw);
+ \path [line] (verif) -- node [above, near start] {No} (verifw);
 
@@
  \path [line] (verifw) -- node [right] {Yes} (gen);
- \path [line] (verifw) -| node [above] {No} (iterate);
-
- \path [line] (gen) -| node [below, near start] {Yes} (done);
- \path [line] (gen) -- node [right] {No} (incw);
+ \path [line] (verifw) -- node [above] {No} (iterate);
+
+ \path [line] (gen) -- node [below] {Yes} (done);
+ \path [line] (gen) -- node [below] {No} (incw);
 
@@
 
+\iffalse
 \begin{figure}
@@
 \end{figure}
-
-\begin{figure}[p]
+\fi
+
+\begin{figure}
 \begin{center}
@@
 
-\begin{figure}[p]
-{\tiny
-\begin{center}
- \begin{tabular}{r||r|r|r|r|r|r|r|r|r}
-  & \multicolumn{3}{c}{Count} & \multicolumn{2}{|c}{Total time} & \multicolumn{2}{|c}{Absolute speedup} & \multicolumn{2}{|c}{Relative speedup} \\
-  {\sc Brahma} version & Total & {\sc Kalashnikov} & {\sc Brahma} & {\sc Kalashnikov} & {\sc Brahma} & Mean & Median & Mean & Median \\
-  &       & faster            & faster       &                   &              &      &        &      & \\
-  \hline
-  \hline
-  PLDI & 16 & 13 & 3 & 184.12\,s & 544.20\,s & 22.50\,s & 2.11\,s & 2.60 & 1.77 \\
-  ICSE Semibiased & 17 & 7 & 10 & 225.49\,s & 262.24\,s & 2.16\,s & -1.00\,s & 1.26 & 0.87
- \end{tabular}
-\end{center}
-}
-
+\begin{figure}
+{\small
+ \include{figures/stats_table}
+}
  \caption{Performance statistics when {\sc Brahma} and {\sc Kalashnikov} were both successful.

### 2014-01-31 tests/table.py
Brahmikov results in the table

@@ -2,10 +2,10 @@
 
 import cPickle
 
-def make_graph(kalashnikov, pldi, icse, icse_semibiased):
-  print "\\begin{tabular}{l||rrc|rrrc|rr}"
-  print "Problem & \multicolumn{3}{c}{\\sc PLDI Brahma} & \multicolumn{4}{|c}{ICSE Brahma} & \multicolumn{2}{|c}{\\sc Kalashnikov} \\\\"
-  print "        & Runtime & \#Lines & Aut. & Random & Semibiased & \#Lines & Aut. & Runtime & \#Lines \\\\"
+def make_graph(kalashnikov, brahmikov, pldi, icse, icse_semibiased):
+  print "\\begin{tabular}{l||rrc|rrrc|rr|rr}"
+  print "Problem & \multicolumn{3}{c}{\\sc PLDI Brahma} & \multicolumn{4}{|c}{ICSE Brahma} & \multicolumn{2}{|c}{\sc Brahmikov} & \multicolumn{2}{|c}{\\sc Kalashnikov} \\\\"
+  print "        & Runtime & \#Lines & Aut. & Random & Semibiased & \#Lines & Aut. & Runtime & \#Lines & Runtime & \#Lines \\\\"
 
   print "\\hline"
   print "\\hline"
@@ -39,7 +39,20 @@ def make_graph(kalashnikov, pldi, icse, icse_semibiased):
       insts.append(kaln)
     else:
       kalt = None
-    
+
+    if k in brahmikov:
+      (cnt, times) = brahmikov[k]
+      brakt = times['_']
+      brakn = cnt['insts']
+
+      if brakn == 0:
+        brakt = -1
+
+      alltimes.append(brakn)
+      insts.append(brakn)
+    else:
+      brakt = None
+
     if k in icse:
       (cnt, time) = icse[k]
       icset = time['_']
@@ -127,6 +140,7 @@ def make_graph(kalashnikov, pldi, icse, icse_semibiased):
     else:
       line += ' & '
 
+    line += f(brakt, brakn)
     line += f(kalt, kaln)
 
     line = line[:-1]
@@ -146,6 +160,7 @@ if __name__ == '__main__':
   pldi = load('processed/brahma.stats')
   icse = load('processed/icse.stats')
   semi = load('processed/icse_semibiased.stats')
+  brahmikov = load('processed/brahmikov.stats')
   kalashnikov = load('processed/const.stats')
 
[…]

### 2014-01-31 paper/paper.tex
Nudges
Make results table smaller

@@
 \newcommand{\bv}[2]{\mathcal{BV}(#1, #2)}
+
+\iffalse
+\lstset{emph={
+  assume, assert}, emphstyle=\bfseries}
+\fi
 
@@
 Additionally, \newC programs may use nondeterministic values, assumptions
-and arbitrary-width types.  Two example \newC programs are shown 
+and arbitrary-width types.
+
+\iffalse
+Two example \newC programs are shown 
 in Fig.~\ref{fig:c-}.
@@
 \end{figure}
+\fi
 
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 {\small
@@
 
-The exact details of how we encoding an $\mathcal{L}$-program are given in Sec.~\ref{sec:encode-l}.
+The exact details of how we encode an $\mathcal{L}$-program are given in Sec.~\ref{sec:encode-l}.
 We must now express the {\sc synth} and {\sc verif} formulae as safety properties
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 \centering
@@
 The only wrinkle here is that we will sometimes synthesise a program containing
-constants.  If we have synthesised a program for with $w=k$,
+constants.  If we have synthesised a program with $w=k$,
 the constants in the program will be $k$-bits wide.  To extend the program
@@
 \paragraph{\bf {\sc Kalashnikov} synthesises a shorter program than {\sc Brahma}:}
-This happens in 4 of the 29 cases.  This case is illustrated by benchmark p29, shown in Fig.~\ref{fig:obfuscated}.
+This happens in 4 of the 29 cases, which is to be expected since {\sc Kalashnikov}'s outputs
+are, by construction, guaranteed to be minimal.  This case is illustrated by benchmark p29, shown in Fig.~\ref{fig:obfuscated}.
 The specification here is a piece of obfuscated code taken from the Conficker worm~\cite{conficker}, and
@@
 In the remaining cases, both {\sc brahma} and {\sc kalashnikov} are able to produce
-minimal programs.  We performed a Mann-Whitney U test (see Fig.~\ref{fig:stats-table})
-to determine whether there was a significant difference
-in the speed of the two tools.  The test was unable to reject the null hypothesis
-(that the tools are equally fast) at the p=0.05
-level, i.e. there is no statistically significant difference in the speed of
-the two tools on this set of benchmarks.
-
-\begin{figure}
+minimal programs.  It was not clear from just looking at the runtime numbers whether
+any of the tools was significantly faster than the others, so we performed a Mann-Whitney U test.
+Since the {\sc brahma} experiments were performed on a slower machine than the
+{\sc kalashnikov} ones, we multiplied each of the runtimes by the CPU frequency of the machine
+that was used for that experiment.  This gives an estimate of the time we would expect the test to
+have taken on a 1GHz machine.  Scaling in this way  is more or less legitimate, since we have empirically
+observed that our algorithm is CPU bound.
+
+For each pair of tools ({\sc kalashnikov} vs. each of the {\sc brahma} configurations), the U test was
+unable to reject the null hypothesis (that the tools are equally fast) at the p=0.05 level. In other words,
+there is no statistically significant difference in the speed of {\sc kalashnikov} and {\sc brahma} on this
+set of benchmarks, even after correcting for the difference in CPU frequency.
+
+\begin{figure}[ht]
 
@@
 
-\iffalse
 \begin{figure}
@@
 \end{figure}
-\fi
 
 \begin{figure}
-\begin{center}
-{\small
+\centering
+{\scriptsize
 \include{figures/results_table}
 }
-\end{center}
 \caption{Time for synthesis of machine integer benchmarks from~\cite{brahma} and~\cite{brahma-icse}.
@@
 
-\begin{figure}
-{\small
- \include{figures/stats_table}
-}
- \caption{Performance statistics when {\sc Brahma} and {\sc Kalashnikov} were both successful.
- \label{fig:stats-table}}
-\end{figure}
-
 \section{Conclusion}

### 2014-01-31 paper/paper.tex
Fit everything onto 15 pages

@@
 just a C program with the following syntactic restrictions:
-
-\begin{itemize}
- \item All loops in the program must have a constant bound.
- \item All recursion in the program must be limited to a constant depth.
- \item All arrays must be statically allocated (i.e. not using \texttt{malloc}),
+ all loops in the program must have a constant bound;
+ all recursion in the program must be limited to a constant depth;
+ all arrays must be statically allocated (i.e. not using \texttt{malloc}),
  and be of constant size.
-\end{itemize}
-
 Additionally, \newC programs may use nondeterministic values, assumptions
@@
 
-\begin{figure}[ht]
+\begin{figure}
 {\small
@@
 
-\begin{figure}[ht]
+\begin{figure}
 \centering
@@
 
-\begin{defn}
 A program is well formed if
 no operand refers to the result of an instruction that has not been computed yet, and if
-each opcode is valid.
-\label{defn:well-formed}
-\end{defn}
-
-We add a well-formedness constraint of the form \verb|params[i] <= (a+c+2*i)|
+each opcode is valid.  We add a well-formedness constraint of the form \verb|params[i] <= (a+c+2*i)|
 for each instruction.  It should be noted that this requires a linear number of well-formedness
-constraints.  If all of these constraints are satisfied, the program is well-formed in the sense
-of Defn.~\ref{defn:well-formed}.
+constraints.  If all of these constraints are satisfied, the program is well-formed in the sense.
 
@@
 finishes much faster than if synthesis had been attempted in a
-high-complexity language.  Our parametrisation is constructed
-in such a way that the programs we synthesise are guaranteed to be
-minimal in length.
+high-complexity language.
 
@@
 that no program of length $l$ can satisfy the specification.  When we do
-successfully synthesise a program, we are guaranteed that it
-is of minimal length since we have previously established that no
+successfully synthesise a program, we are \emph{guaranteed that it
+is of minimal length} since we have previously established that no
 shorter program is correct.
@@
 
+\centering
+\begin{minipage}[t]{.45\textwidth}
 \begin{eqnarray*}
@@
  \bv{m-1}{m} & \rightarrow & \bv{n-1}{n} \\
- \bv{m+1}{m} & \rightarrow & \bv{n+1}{n} \\
+ \bv{m+1}{m} & \rightarrow & \bv{n+1}{n}
+\end{eqnarray*}
+\end{minipage}
+\begin{minipage}[t]{.45\textwidth}
+\begin{eqnarray*}
  \bv{x}{m} & \rightarrow & \bv{x}{n} \\
@@
 \end{eqnarray*}
+\end{minipage}
 
@@
 
-\begin{figure}
+\begin{figure}[t]
 \centering
@@
 To give a reference point for the efficiency of our tool, we present the results given for {\sc brahma}
-on the same benchmarks, as reported in~\cite{brahma} and~\cite{brahma-icse}\footnote{Unfortunately
-we were not able to obtain a copy of {\sc Brahma}, and so were unable to rerun the experiments ourselves.}.
+on the same benchmarks, as reported in~\cite{brahma} and~\cite{brahma-icse}.
 These experiments were performed on an 8-core,
-1.86\,GHz Xeon with 4\,GB of RAM.  Although there are 49 results shown across the two tables, it should
-be noted that there is a lot of overlap between the two sets of benchmarks and that there are only
-29 unique benchmarks altogether.
+1.86\,GHz Xeon with 4\,GB of RAM.
+
+Since we were unable to obtain a copy of {\sc brahma}, we re-implemented the {\sc brahma} program encoding in our framework,
+resulting in the {\sc brahmikov} tool. {\sc Brahmikov} implements the component based synthesis
+paradigm, but still uses {\sc cbmc} as the backend decision procedure.  Since {\sc brahma} and {\sc brahmikov} are written in
+different frameworks, we cannot compare their performance.  However we can compare {\sc brahmikov} and {\sc kalashnikov}.
+In particular it is interesting to note that {\sc brahmikov} often produces longer programs than {\sc kalashnikov}.  We
+believe this is due to differences in the post-processing strategies used by {\sc brahmikov} and {\sc brahma}.  {\sc Brahmikov}
+performs syntatic dead code elimination (following~\cite{brahma}), but this is not enough to generate minimal programs
+in many cases.  This is in contrast to {\sc kalashnikov}, which always generates minimal programs.
 
@@
 Column 1 shows the runtime reported for each benchmark in \cite{brahma}, column 2
-shows the number of instructions in the synthesised program, and column 3 contains a \xmark\,when
+shows the number of instructions in the synthesised program, and column 3 contains a \xmark\;when
 {\sc Brahma} needed user assistance to solve a benchmark.
@@
 implemented the semibiased optimisation.
-Finally, column 8 shows the runtime for {\sc Kalashnikov} and column 9 the number of instructions
-in the program synthesised by {\sc Kalashnikov}.  The results can be divided into
-three categories, as follows:
+Finally, columns 8 and 10 show the runtime for {\sc Brahmikov} and {\sc Kalashnikov} respectively.
+Columns 9 and 11 show the number of instructions
+in the program synthesised by {\sc Brahmikov} and {\sc Kalashnikov}.  In all cases,
+{\sc Brahmikov} performs very poorly, which demonstrates that the {\sc Kalashnikov} program
+encoding is more suitable for our system than the component based synthesis encoding.
+We also ran the {\sc Brahmikov} experiments using Z3 as the solver rather than Glucose,
+but this configuration failed to terminate for any of the examples.
+
+The results can be divided into three categories, as follows:
 
@@
 minimal programs.  It was not clear from just looking at the runtime numbers whether
-any of the tools was significantly faster than the others, so we performed a Mann-Whitney U test.
-Since the {\sc brahma} experiments were performed on a slower machine than the
-{\sc kalashnikov} ones, we multiplied each of the runtimes by the CPU frequency of the machine
-that was used for that experiment.  This gives an estimate of the time we would expect the test to
-have taken on a 1GHz machine.  Scaling in this way  is more or less legitimate, since we have empirically
-observed that our algorithm is CPU bound.
-
-For each pair of tools ({\sc kalashnikov} vs. each of the {\sc brahma} configurations), the U test was
+any of the tools was significantly faster than the others, so we performed a Wilcoxon signed-rank test.
+
+For each pair of tools ({\sc kalashnikov} vs. each of the {\sc brahma} configurations), the Wilcoxon test was
 unable to reject the null hypothesis (that the tools are equally fast) at the p=0.05 level. In other words,
-there is no statistically significant difference in the speed of {\sc kalashnikov} and {\sc brahma} on this
-set of benchmarks, even after correcting for the difference in CPU frequency.
-
-\begin{figure}[ht]
-
+there is no statistically significant difference in the speed of {\sc kalashnikov} and {\sc brahma}.
+
+\begin{figure}
 Obfuscated code
@@
 
-\begin{figure}
+\begin{figure}[ht]
 \begin{center}
@@
 
-\begin{figure}
+\begin{figure}[ht]
 \centering
-{\scriptsize
+{\tiny
 \include{figures/results_table}

### 2014-01-31 tests/table.py
Update results table

@@ -47,9 +47,9 @@ def make_graph(kalashnikov, brahmikov, pldi, icse, icse_semibiased):
 
       if brakn == 0:
         brakt = -1
-
-      alltimes.append(brakn)
-      insts.append(brakn)
+      else:
+        alltimes.append(brakt)
+        insts.append(brakn)
     else:
       brakt = None
 
@@ -80,16 +80,16 @@ def make_graph(kalashnikov, brahmikov, pldi, icse, icse_semibiased):
 
       if t is None:
         ret += "N/A &"
-      elif t == mintime:
-        ret += "{\\bf %.02fs} &" % t
       elif t < 0:
         ret  += "T/O &"
+      elif t <= mintime:
+        ret += "{\\bf %.02fs} &" % t
       else:
         ret += "%.02fs &" % t
 
       if t < 0:
         ret += '-- &'
-      elif insts == mininsts:
+      elif insts <= mininsts:
         ret += '{\\bf %d} &' % insts
       else:
         ret += '%d &' % insts

### 2014-01-31 paper/paper.tex
Added reference

@@
 programs, whereas we are optimised for synthesising small (but intricate)
-programs without guidance.
+programs without guidance.  A sketching-like approach to optimising
+fixed point arithmetic code was described in~\cite{eldib-wang}.  This
+work is similar in goal to our floating-point work, but is more restricted
+in scope in that it only tries to minimise register widths for embedded
+systems using fixed point arithmetic.
 
@@
 We thank Martin Brain for his insight and advice during our synthesis discussions.
+Also, thanks to Cristina David for her extensive help and proof reading during the
+preparation of this paper.
 

### 2014-01-31 paper/paper.tex
Fixed bugs in results table

@@
 Another successful approach to program synthesis has been \emph{program
-sketching}, in particular the {\sc sketch} tool~\cite{lezama-thesis,sketch}.  This method
+sketching}, in particular the {\sc sketch} tool~\cite{lezama-thesis,sketch,modular-sketch}.  This method
 requires the programmer to provide a \emph{program sketch}, written in a

### 2014-01-31 tests/brahma_stats.py

@@ -57,8 +57,8 @@ icse_brahma = [
     (214.14, 19.3, 63.23, False, 6), #22 
     (1074.04, 13.28, 272.28, False, 7), #21 
     (-1, 187.17, 185.57, False, 12),  #22 
-    (1.38, 12.12, 12.12, True, 3),   #23 
-    (5.28, 2.96, 2.96, True, 4)  #24 
+    (1.38, 12.12, 12.12, False, 3),   #23 
+    (5.28, 2.96, 2.96, False, 4)  #24 
 ]
 
 icse_pldi_map = dict([

### 2014-01-31 tests/table.py

@@ -56,18 +56,22 @@ def make_graph(kalashnikov, brahmikov, pldi, icse, icse_semibiased):
     if k in icse:
       (cnt, time) = icse[k]
       icset = time['_']
-      icsen = cnt['insts']
 
       alltimes.append(icset)
+    else:
+      icset = None
 
+    if k in icse_semibiased:
       (cnt, time) = icse_semibiased[k]
       semit = time['_']
       icseaut = cnt['automatic']
+      icsen = cnt['insts']
 
       alltimes.append(semit)
       insts.append(icsen)
     else:
-      icset = None
+      semit = None
+      icsen = None
 
     mintime = min(t for t in alltimes if t > 0)
     mininsts = min(insts)
@@ -98,7 +102,7 @@ def make_graph(kalashnikov, brahmikov, pldi, icse, icse_semibiased):
 
     line += f(pldit, pldin)
 
-    if not pldiaut:
+    if pldit and not pldiaut:
       line += ' \\xmark & '
     else:
       line += ' & '
@@ -124,7 +128,7 @@ def make_graph(kalashnikov, brahmikov, pldi, icse, icse_semibiased):
       else:
         ret += " %.02fs & " % semit
 
-      if t < 0:
+      if not insts:
         ret += '-- &'
       elif insts == mininsts:
         ret += '{\\bf %d} &' % insts
@@ -135,7 +139,7 @@ def make_graph(kalashnikov, brahmikov, pldi, icse, icse_semibiased):
 
     line += g(icset, semit, icsen)
 
-    if not icseaut:
+    if semit and not icseaut:
       line += ' \\xmark & '
     else:
       line += ' & '

### 2014-01-31 paper/paper.tex
Typo
Added benchmarks appendix

@@
 \usepackage{multicol}
+\usepackage{appendix}
 
@@
 We implemented our fully automatic synthesis procedure as the {\sc kalashnikov} tool.  The implementation consists
-of around 600 lines of \newC and 600 lines of Python.  To evaluate the tool we used the 25 bitvector
+of around 600 lines of \newC and 600 lines of Python.  To evaluate the tool we used the 29 bitvector
 programs from~\cite{brahma} and~\cite{brahma-icse}.  The majority of these are ``bit twiddling hacks'' taken from
 Hacker's Delight~\cite{hackers-delight}.  The code we used to perform the experiments, along with the
-benchmarks, is available at \texttt{http://www.cprover.org/kalashnikov}.  As a backend solver, we used
+benchmarks, is available at \texttt{http://www.cprover.org/kalashnikov} and the benchmarks can also
+be found in Appendix~\ref{app:benchmarks}.  As a backend solver, we used
 {\sc CBMC}~\cite{cbmc-website} at SVN revision r3545, with Glucose 3.0~\cite{glucose-paper} as the SAT solver.
@@
 We thank Martin Brain for his insight and advice during our synthesis discussions.
-Also, thanks to Cristina David for her extensive help and proof reading during the
+Also, thanks to Cristina David for her extensive help and proofreading during the
 preparation of this paper.
@@
 
+\appendix
+
+\section{Benchmarks}
+\label{app:benchmarks}
+
+\begin{figure}
+\scriptsize
+ \begin{multicols}{4}
+
+   \begin{minipage}{.2\textwidth}
+   \bf{P1}($x$): Turn-off rightmost 1 bit.
+   \begin{enumerate}
+    \item $t1 = x - 1$
+    \item $t2 = x \& t1$
+   \end{enumerate}
+   \end{minipage}
+
+   \vspace{12pt}
+
+   \begin{minipage}{.2\textwidth}
+   \bf{P2}($x$): Test whether $x$ is of the form $2^n - 1$.
+   \begin{enumerate}
+    \item $t1 = x + 1$
+    \item $t2 = x \& t1$
+   \end{enumerate}
+   \end{minipage}
+
+   \vspace{12pt}
+
+   \begin{minipage}{.2\textwidth}
+   \bf{P3}($x$): Isolate the rightmost 1 bit.
+   \begin{enumerate}
+    \item $t1 = -x$
+    \item $t2 = x \& t1$
+   \end{enumerate}
+   \end{minipage}
+
+   \vspace{12pt}
+
+   \begin{minipage}{.2\textwidth}
+   \bf{P4}($x$): Form a mask that identifies the rightmost 1 bit and trailing 0s.
+   \begin{enumerate}
+    \item $t1 = x - 1$
+    \item $t2 = x \oplus t1$
+   \end{enumerate}
+   \end{minipage}
+
+   \vspace{12pt}
+
+   \begin{minipage}{.2\textwidth}
+   \bf{P5}($x$): Right propagate the rightmost 1 bit.
+   \begin{enumerate}
+    \item $t1 = x - 1$
+    \item $t2 = x \mid t1$
+   \end{enumerate}
+   \end{minipage}
+
+   \vspace{12pt}
+
+   \begin{minipage}{.2\textwidth}
+   \bf{P6}($x$): Turn on the rightmost 0 bit.
+   \begin{enumerate}
+    \item $t1 = x + 1$
+    \item $t2 = x \mid t1$
+   \end{enumerate}
+   \end{minipage}
+      \vspace{12pt}
+
+   \begin{minipage}{.2\textwidth}
+   \bf{P7}($x$): Isolate the rightmost 0 bit.
+   \begin{enumerate}
+    \item $t1 = \sim x$
+    \item $t2 = x + 1$
+    \item $t3 = t1 \& t2$
+   \end{enumerate}
+   \end{minipage}
+   \vspace{12pt}
+
+   \begin{minipage}{.2\textwidth}
+   \bf{P8}($x$): Form a mask the identifies the trailing 0s.
+   \begin{enumerate}
+    \item $t1 = x -1$
+    \item $t2 = \sim x$
+    \item $t3 = t1 \& t2$
+   \end{enumerate}
+   \end{minipage}
+   \vspace{12pt}
+
+   \begin{minipage}{.2\textwidth}
+   \bf{P9}($x$): Absolute value.
+   \begin{enumerate}
+    \item $t1 = x >> 31$
+    \item $t2 = x \oplus t1$
+    \item $t3 = t2 - t1$
+   \end{enumerate}
+   \end{minipage}
+
+      \vspace{12pt}
+
+   \begin{minipage}{.2\textwidth}
+   \bf{P10}($x, y$): Test if nlz(x) == nlz(y), where nlz is number of leading zeroes.
+   \begin{enumerate}
+    \item $t1 = x \& y$
+    \item $t2 = x \oplus y$
+    \item $t3 = t2 \leq t1$
+   \end{enumerate}
+   \end{minipage}
+
+      \vspace{12pt}
+
+   \begin{minipage}{.2\textwidth}
+   \bf{P11}($x, y$): Test if nlz(x) < nlz(y).
+   \begin{enumerate}
+    \item $t1 = \sim y$
+    \item $t2 = x \& t1$
+    \item $t3 = t2 > y$
+   \end{enumerate}
+   \end{minipage}
+   \vspace{12pt}
+
+   \begin{minipage}{.2\textwidth}
+   \bf{P12}($x, y$): Test if nlz(x) <= nlz(y).
+   \begin{enumerate}
+    \item $t1 = \sim y$
+    \item $t2 = x \& t1$
+    \item $t3 = t2 \leq y$
+   \end{enumerate}
+   \end{minipage}
+
+      \vspace{12pt}
+
+   \begin{minipage}{.2\textwidth}
+   \bf{P13}($x$): Sign.
+   \begin{enumerate}
+    \item $t1 = x >> 31$
+    \item $t2 = t1 < x$
+    \item $t3 = t2 + t1$
+   \end{enumerate}
+   \end{minipage}
+
+   \vspace{12pt}
+
+   \begin{minipage}{.2\textwidth}
+   \bf{P14}($x, y$): Floor of average of $x$ and $y$ without overflowing.
+   \begin{enumerate}
+    \item $t1 = x \& y$
+    \item $t2 = x \oplus y$
+    \item $t3 = t2 >> 1$
+    \item $t4 = t1 + t3$
+   \end{enumerate}
+   \end{minipage}
+   \vspace{12pt}
+
+   \begin{minipage}{.2\textwidth}
+   \bf{P15}($x, y$): Ceil of average of $x$ and $y$ without overflowing.
+   \begin{enumerate}
+    \item $t1 = x \mid y$
+    \item $t2 = x \oplus y$
+    \item $t3 = t2 >> 1$
+    \item $t4 = t1 - t3$
+   \end{enumerate}
+   \end{minipage}
+   \vspace{12pt}
+
+   \begin{minipage}{.2\textwidth}
+   \bf{P16}($x$): Max.
+   \begin{enumerate}
+    \item $t1 = x \leq y$
+    \item $t2 = x \oplus y$
+    \item $t3 = t1 * t2$
+    \item $t4 = t3 \oplus y$
+   \end{enumerate}
+   \end{minipage}
+   \vspace{12pt}
+
+   \begin{minipage}{.2\textwidth}
+   \bf{P17}($x$): Turn off the rightmost contiguous string of 1 bits.
+   \begin{enumerate}
+    \item $t1 = x - 1$
+    \item $t2 = x \mid t1$
+    \item $t3 = t2 + 1$
+    \item $t4 = t3 \& x$
+   \end{enumerate}
+   \end{minipage}
+   \vspace{12pt}
+
+   \begin{minipage}{.2\textwidth}
+   \bf{P18}($x$): Determine if $x$ is a power of 2.
+   \begin{enumerate}
+    \item $t1 = -x$
+    \item $t2 = x \oplus t1$
+    \item $t3 = t2 < t1$
+   \end{enumerate}
+   \end{minipage}
+   \vspace{12pt}
+
+  \iffalse
+   \begin{minipage}{.2\textwidth}
+   \bf{P19}($x, m, k$): Exchange two fields of $x$, demarcated by $m$ and $k$.
+   \begin{enumerate}
+    \item $t1 = x >> k$
+    \item $t2 = x \oplus t1$
+    \item $t3 = t2 & m$
+    \item $t4 = t3 << k$
+    \item $t5 = t4 \oplus t3$
+    \item $t6 = t5 \oplus x$
+   \end{enumerate}
+   \end{minipage}
+   \vspace{12pt}
+
+   \begin{minipage}{.2\textwidth}
+   \bf{P20}($x$): Next higher unsigned number with the same number of 1 bits.
+   \begin{enumerate}
+    \item $t1 = \sim x$
+    \item $t2 = x \& t1$
+    \item $t3 = x + t2$
+    \item $t4 = x \opus t2$
+    \item $t5 = t4 >> 2$
+    \item $t6 = t5 / t2$
+    \item $t7 = t6 \mid t3$
+   \end{enumerate}
+   \end{minipage}
+   \vspace{12pt}
+
+   \begin{minipage}{.2\textwidth}
+   \bf{P21}($x, a, b, c$): Cycling through 3 values $a, b, c$.
+   \begin{enumerate}
+    \item $t1 = x \neq c$
+    \item $t2 = a \oplus c$
+    \item $t3 = x \neq a$
+    \item $t4 = b \oplus c$
+    \item $t5 = t2 \& t2$
+    \item $t6 = t3 \& t4$
+    \item $t7 = t5 \oplus t6$
+    \item $t8 = t7 \oplus c$
+   \end{enumerate}
+   \end{minipage}
+\fi
+
+   \begin{minipage}{.2\textwidth}
+   {\bf P19($x, m, k$): Exchange two fields of $x$, demarcated by $m$ and $k$.}
+
+   N/A ({\sc Kalashnikov} was unable to synthesise this program.)
+   \end{minipage}
+   \vspace{12pt}
+
+   \begin{minipage}{.2\textwidth}
+   {\bf P20($x$): Next higher unsigned number with the same number of 1 bits.}
+
+   N/A
+   \end{minipage}
+   \vspace{12pt}
+
+   \begin{minipage}{.2\textwidth}
+   {\bf P21($x, a, b, c$): Cycling through 3 values $a, b, c$.}
+
+    N/A
+   \end{minipage}
+   \vspace{12pt}
+
+   \begin{minipage}{.2\textwidth}
+   {\bf P22($x$): Compute parity.}
+
+    N/A
+   \end{minipage}
+   \vspace{12pt}
+
+   \begin{minipage}{.2\textwidth}
+   {\bf P23($x$): Count number of set bits.}
+
+    N/A
+   \end{minipage}
+      \vspace{12pt}
+
+   \begin{minipage}{.2\textwidth}
+   {\bf P24($x$): Round up to next highest power of 2.}
+
+    N/A
+   \end{minipage}
+   \vspace{12pt}
+
+   \begin{minipage}{.2\textwidth}
+   {\bf P25($x, y$): Computer higher order half of $x \times y$.}
+
+    N/A
+   \end{minipage}
+
+   \vspace{12pt}
+
+   \begin{minipage}{.2\textwidth}
+   {\bf P26($x, k$): Round $x$ up to a multiple of $2^k$.}
+
+N/A
+   \end{minipage}
+
+      \vspace{12pt}
+
+   \begin{minipage}{.2\textwidth}
+   {\bf P27($x, y$): Min.}
+    \begin{enumerate}
+     \item $t1 = x \oplus y$
+     \item $t2 = t1 \leq t2$
+     \item $t3 = t1 * t2$
+     \item $t4 = y \oplus t3$
+    \end{enumerate}
+   \end{minipage}
+
+ \end{multicols}
+
+ \caption{The integer benchmarks}
+\end{figure}
+
+\begin{figure}
+ \begin{lstlisting}[language=C++]
+void interchangeObs(word_t* src, word_t* dest)
+{
+  *src = *src ^ *dest;
+
+  if (*src == (*src ^ *dest)) {
+    *src = *src ^ *dest;
+
+    if (*src == (*src ^ *dest)) {
+      *dest = *src ^ *dest;
+
+      if (*dest == (*src ^ *dest)) {
+        *src = *dest ^ *src;
+        return;
+      } else {
+        *src = *src ^ *dest;
+        *dest = *src ^ *dest;
+        return;
+      } 
+    } else
+      *src = *src ^ *dest;
+  }
+
+  *dest = *src ^ *dest;
+  *src = *src ^ *dest;
+
+  return;
+}
+ \end{lstlisting}
+
+ \caption{Benchmark 28: an obfuscated swap function}
+\end{figure}
+
+\begin{figure}
+ \begin{lstlisting}[language=C++]
+int mul45(int y) {
+  int a=1, b=0, z=1, c=0;
+
+  while(1) {
+    if (a == 0) {
+      if (b == 0) {
+	y=z+y; a =!a; b=!b;c=!c;
+	if (! c) break;
+      } else {
+	z=z+y; a=!a; b=!b; c=!c;
+	if (! c) break;
+      }
+    } else {
+      if (b == 0) { z=y << 2; a=!a; }
+      else {
+	z=y << 3; a=!a; b=!b; } } }
+
+  return y;
+}
+
+ \end{lstlisting}
+
+ \caption{Benchmark 29: an obfuscated function to multiply by 45}
+
+\end{figure}
+
 \end{document}

### 2014-05-27 papers/termination/paper.tex
Began termination paper

\documentclass[preprint]{sigplanconf}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}

\begin{document}

%\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\conferenceinfo{POPL '14}{Month d--d, 20yy, City, ST, Country} 
\copyrightyear{2014}

% Uncomment one of the following two, if you are not going for the 
% traditional copyright transfer agreement.

%\exclusivelicense                % ACM gets exclusive license to publish, 
                                  % you retain copyright

%\permissiontopublish             % ACM gets nonexclusive license to publish
                                  % (paid open-access papers, 
                                  % short abstracts)

\title{Synthesising Complex Termination Arguments}

\authorinfo{Cristina David\and Daniel Kroening\and Matt Lewis}
           {Oxford University}
           {firstname.lastname@cs.ox.ac.uk}

\maketitle

\begin{abstract}
Proving program termination is typically done by finding a well-founded \emph{ranking function}
for the program states.
Existing termination provers typically find ranking functions
using either linear algebra or templates.  As such they are often restricted to
finding linear ranking functions over mathematical integers.  This class
of ranking functions is not large enough to prove termination for all terminating
programs, and furthermore a termination argument for a program operating on mathematical integers
does not always lead to a termination argument for the same program operating on
fixed-width machine integers.

We present a reduction from program \emph{termination} to program \emph{synthesis}.
This reduction allows us to generate nonlinear, lexicographic ranking functions that
are correct for fixed-width machine arithmetic and floating-point arithmetic.
We use this reduction to build a semi-decision procedure for the termination
of fixed-width and floating-point arithmetic programs.
\end{abstract}

%\category{CR-number}{subcategory}{third-level}

\keywords
Termination, Program Synthesis, Lexicographic Ranking Functions, Bitvector Ranking Functions,
Floating Point Ranking Functions.

\section{Introduction}
Church was BIG into program synthesis~\cite{church-synth}, so you know it's good stuff.

Something, something, Curry-Howard Isomorphism, something, something, programs-as-proofs.

\section{Related Work}
Error 404.

\section{Preliminaries}
\subsection{Termination and Ranking Functions}
A transition system with state space $X$ and transition relation $T \subseteq X \times X$
is said to be \emph{unconditionally terminating} if there is no infinite sequence
$x_1, x_2, \ldots$ with $\forall i . T(x_i, x_{i+1})$.  We can prove that $T$ is
unconditionally terminating by finding an injective function $R: X \to Y$ where
$Y$ is well-founded and $R$ is monotonically decreasing with respect $T$.  That is
to say:

$$\forall x, x' . T(x, x') \Rightarrow R(x) < R(x')$$

\subsection{Program Synthesis as Second-order Satisfaction}
The program synthesis problem can be described as finding a satisfying assigment to the
synthesis formula:

\begin{definition}[Synthesis Formula]
 The synthesis formula for a specification $\sigma: X \to Y$ is:

 $$\exists P: X \to Y . \forall x : X . \sigma(x, P(x))$$
 \end{definition}

Since this formula involves quantification over functions $X \to Y$,
this is a second-order formula -- indeed, if $X = Y = \mathbb{N}$,
the formula describes a set at level $\Sigma_1^1$ of the analytical
hierarchy.  As such, determining the satisfiability of the synthesis
formula over a given logic is strictly harder than solving the
halting problem over the same logic.

\subsection{Generalised Synthesis}
Sometimes we want to synthesise several programs at once, as well as some
ground terms $\vec{x}$.  Also, we might want a more complex specification
than just a relation over $X$ and $Y$.

\begin{definition}[Generalised Synthesis Formula]
 $$\exists P_1: X_1 \to Y_1, \ldots, P_N: X_N \to Y_N , \vec{x}: X^e . \forall \vec{y}: X^a: X . \sigma(x, P_1, \ldots, P_N, y) $$
\end{definition}

\section{Formalism}
\subsection{Syntax and Semantics}

\section{Reducing Termination to Synthesis}
For a transition system $T$, we can build a specification $\sigma$ to find a conditional ranking function for
$T$ with initial states I:

\begin{eqnarray}
 \sigma(V: X \to \mathbb{B}, R: X \to \mathbb{N}, x, x') = & I(x) \rightarrow V(x) \wedge \\
 & V(x) \wedge T(x, x') \rightarrow V(x') \wedge R(x) < R(x') 
\end{eqnarray}

Here we have synthesised a ranking function $R$ into $\mathbb{N}$ (which is well-founded),
as well as an inductive invariant $V$ that guarantees termination of $T$.

To synthesise an n-place lexicographic ranking function, we just ask for a ranking function
$R: X \to \mathbb{N}^n$.

\bibliographystyle{abbrvnat}
\bibliography{synth}{}

\end{document}

### 2014-05-30 papers/termination/paper.tex
Added skeleton results tables

@@
 \usepackage{amsthm}
+\usepackage{amsfonts}
+\usepackage{pifont}
 
@@
 \newtheorem{definition}[theorem]{Definition}
+
+\newcommand{\xmark}{\ding{55}}
 
@@
 
+\section{Experiments}
+
+\begin{figure*}
+\centering
+\begin{tabular}{|l|l|l||l|l|l|l|}
+\hline
+    & LOC & Rank function size & \textsc{T2} & \textsc{ARMC} & \textsc{Headshot} & \textsc{Headshot-Linear} \\
+    \hline
+    \hline
+ P1 & 10 & 3 & 100s & 70s & 0.1s & \bf{0.01s} \\
+ P2 & 10 & 3 & 100s & 70s & 0.1s & \bf{0.01s} \\
+ P3 & 10 & 3 & 100s & 70s & 0.1s & \bf{0.01s} \\
+ \hline
+\end{tabular}
+\caption{Termination for linear programs with disjunctive, linear ranking functions\label{fig:linear}}
+\end{figure*}
+
+\begin{figure*}
+\centering
+\begin{tabular}{|l|l|l|l|l|l|l|}
+\hline
+    & LOC & Ranking program size & Lexicarity & Linear program? & Linear ranking function? & Time (s)\\
+    \hline
+    \hline
+ P1 & 10 & 3 & 1 & \xmark & \xmark & 0.01 \\
+ P2 & 10 & 3 & 1 & \xmark & \xmark & 0.01 \\
+ P3 & 10 & 3 & 1 & \xmark & \xmark & 0.01 \\
+ \hline
+\end{tabular}
+
+\caption{\textsc{Headshot} termination for nonlinear programs with nonlinear ranking functions\label{fig:nonlinear}}
+ \end{figure*}
+
 \bibliographystyle{abbrvnat}

### 2014-05-30 papers/termination/paper.tex
Make tables prettier

@@
 \centering
-\begin{tabular}{|l|l|l||l|l|l|l|}
+\begin{tabular}{|l|r|r||r|r|r|r|}
 \hline
-    & LOC & Rank function size & \textsc{T2} & \textsc{ARMC} & \textsc{Headshot} & \textsc{Headshot-Linear} \\
+    & LOC & \shortstack{Rank function \\ size} & \textsc{T2} & \textsc{ARMC} & \textsc{Headshot} & \textsc{Headshot-Linear} \\
     \hline
@@
 \centering
-\begin{tabular}{|l|l|l|l|l|l|l|}
+\begin{tabular}{|l|r|c|c|r|r|r|}
 \hline
-    & LOC & Ranking program size & Lexicarity & Linear program? & Linear ranking function? & Time (s)\\
+    & LOC & \shortstack{Linear \\ program?} & \shortstack{Linear ranking \\ function?}  & Dimension & \shortstack{Ranking \\ program size} & Time (s)\\
     \hline
     \hline
- P1 & 10 & 3 & 1 & \xmark & \xmark & 0.01 \\
- P2 & 10 & 3 & 1 & \xmark & \xmark & 0.01 \\
- P3 & 10 & 3 & 1 & \xmark & \xmark & 0.01 \\
+ P1 & 10 & \xmark & \xmark & 3 & 1 & 0.01 \\
+ P2 & 10 & \xmark & \xmark & 3 & 1 & 0.01 \\
+ P3 & 10 & \xmark & \xmark & 3 & 1 & 0.01 \\
  \hline

### 2014-06-02 papers/termination/paper.tex
Added a bunch of conjectures

@@
 \newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{corollary}[theorem]{Corollary}
+\newtheorem{conjecture}[theorem]{Conjecture}
 
@@
 
+\section{Theory}
+
+In the sequel, we will fix the following notation:
+
+\begin{itemize}
+ \item Our transition relation is $T(x, x') \subseteq X \times X$.
+ \item Our loop condition is $L(x) \subseteq X$.
+ \item Our ranking function is $R(x) : X \to Y$.
+ \item Our state space has size $\| X \| = M = 2^k$.
+ \item Our ranking co-domain has size $\| Y \| = N = 2^j$.
+ \item The number of looping states is $\| L \| = l$.
+ \item Our transition relation is deterministic and parititioned into chains of length $c_i$, with $l = \sum c_i$.
+\end{itemize}
+
+\subsection{Termination theory}
+
+\begin{theorem}
+ A random function $f : X \to Y$ is a ranking function for $(T, L)$ with probability
+
+ $$N^{-l} \times \prod {{N-1} \choose c_i}$$
+\end{theorem}
+
+\begin{corollary}
+ This number is really small (e.g. $10^{-193}$ for a 64-bit program with 1 variable and 10 looping states.
+ Randomly sampling functions \& hoping they're ranking functions is not going to work.
+\end{corollary}
+
+\begin{conjecture}
+ The probability that a random program $(T, L)$ is terminating (the Chaitin constant)
+ is $0.7$.
+\end{conjecture}
+
+\begin{conjecture}
+ The probability that a random program $(T, L)$ admits a linear ranking function is
+ $0.1$.
+\end{conjecture}
+
+\begin{conjecture}
+ The probability that a random, terminating program $(T, L)$ admits a linear ranking function
+ is $0.2$.
+\end{conjecture}
+
+\begin{corollary}
+ Most terminating programs do not have linear ranking functions.
+\end{corollary}
+
+\begin{theorem}
+ \textsc{Headshot} is biased towards finding ranking functions with low
+ Kolmogorov complexity.
+\end{theorem}
+
+\begin{conjecture}
+ Most low-Kolmogorov-complexity programs compute non-linear functions.
+\end{conjecture}
+
+\begin{corollary}
+ \textsc{Headshot} can often prove termination where linear methods cannot.
+\end{corollary}
+
+\begin{conjecture}
+ Low-Kolmogorov-complexity (LKC) programs have LKC ranking functions.
+\end{conjecture}
+
+\begin{corollary}
+ Goldbach conjecture, Collatz conjecture.
+\end{corollary}
+
+\subsection{Synthesis}
+
+\begin{definition}
+ A \emph{fitness landscape} is the space of all programs along with their fitness.
+\end{definition}
+
+\begin{theorem}
+ Fitness landscapes form a lattice.  Adding test vectors corresponds to abstraction refinement on this
+ lattice, which is why incremental GP works well.
+\end{theorem}
+
+\begin{conjecture}
+ A single fitness landscape isn't really very smooth (e.g. small changes in program representation
+ don't correspond to small changes in fitness), so GP probably shouldn't work very well.
+
+ But it does.
+\end{conjecture}
+
+\begin{conjecture}
+ Most incorrect programs are incorrect on most inputs.
+\end{conjecture}
+
 \section{Experiments}

### 2014-06-02 papers/termination/paper.tex
Reword conjectures

@@
 \begin{theorem}
- \textsc{Headshot} is biased towards finding ranking functions with low
- Kolmogorov complexity.
+ \textsc{Headshot} is biased towards finding ranking functions with
+ low-Kolmogorov-complexity (LKC).
 \end{theorem}
@@
 \begin{conjecture}
- Most low-Kolmogorov-complexity programs compute non-linear functions.
+ Most LKC programs compute non-linear functions, but linear functions are LKC.
+ So LKC is a weaker assumption than linearity.
 \end{conjecture}
@@
 \begin{conjecture}
- Low-Kolmogorov-complexity (LKC) programs have LKC ranking functions.
+ LKC programs have LKC ranking functions (almost certainly false).
 \end{conjecture}

### 2014-06-03 papers/termination/kalashnikov.tex
Added Kalashnikov to the termination paper

--- 2014-01-31 paper/paper.tex
@@
-% This is LLNCS.DEM the demonstration file of
-% the LaTeX macro package from Springer-Verlag
-% for Lecture Notes in Computer Science,
-% version 2.4 for LaTeX2e as of 16. April 2010
-%
-\documentclass[a4paper]{llncs}
-%
-
-\usepackage{listings}
-\usepackage{graphicx}
-\usepackage{xspace}
-\usepackage{pgf}
-\usepackage[noend]{algpseudocode}
-\usepackage{algorithm}
-\usepackage{amsfonts}
-\usepackage{amsmath}
-%\usepackage{amsthm}
-\usepackage{pifont}
-\usepackage{multicol}
-\usepackage{appendix}
-
-\usepackage{tikz}
-\usetikzlibrary{arrows, automata, shapes}
-
-\newtheorem{defn}{Definition}
-
 \algrenewcommand\algorithmicrequire{\textbf{Precondition:}}
@@
 
-\newcommand{\xmark}{\ding{55}}
-
 \newcommand{\newC}{C$^-$\xspace}
@@
 \newcommand{\bv}[2]{\mathcal{BV}(#1, #2)}
-
-\iffalse
-\lstset{emph={
-  assume, assert}, emphstyle=\bfseries}
-\fi
-
-%\usepackage{makeidx}  % allows for indexgeneration
-%
-\title{Fully Automatic Loop-free Program Synthesis by Bounded Model Checking}
-\author{Daniel Kroening \and Matt Lewis}
-\institute{University of Oxford}
-
-\newenvironment{keywords}{
-       \list{}{\advance\topsep by0.35cm\relax\small
-       \leftmargin=0cm
-       \labelwidth=0.35cm
-       \listparindent=0.35cm
-       \itemindent\listparindent
-       \rightmargin\leftmargin}\item[\hskip\labelsep
-                                     \bfseries Keywords:]}
-     {\endlist}
 
@@
 
-\begin{document}
-%
-\maketitle
-%
-\pagestyle{headings}  % switches on printing of running heads
-
-\begin{abstract}
-
-We present a simple-yet-effective method for synthesising loop-free programs
-from specifications, using a novel combination of Bounded Model Checking and
-explicit-state Model Checking.  The user provides a specification in plain
-C, after which the synthesis process is fully automatic and is guaranteed to
-synthesise a minimal program.
-
-We demonstrate the effectiveness of our method by synthesising several subtle
-programs acting on machine integers, de-obfuscating complex C code and
-synthesising a previously un-synthesised floating-point program.
-\end{abstract}
-
-\begin{keywords}
- Program synthesis, bitvectors, bounded model checking, CBMC,
- floating point.
-\end{keywords}
-
+\iffalse
 \section{Introduction}
@@
 summarise the results of our experiments in Sec.~\ref{sec:experiments}.
-
-\section{The Basic Synthesis Algorithm}
+\fi
+
+\subsection{The Basic Synthesis Algorithm}
 \label{sec:abstract-algorithm}
 
-\subsection{The Synthesis Formula}
+\subsubsection{The Synthesis Formula}
 \label{sec:abstract-formula}
@@
 
-\subsection{CEGIS}
+\subsubsection{CEGIS}
 
@@
 
-\begin{algorithm}
+\begin{algorithm*}
  \caption{Abstract refinement algorithm
@@
  \end{multicols}
-\end{algorithm}
-
-\begin{figure}
+\end{algorithm*}
+
+\begin{figure*}
  \centering
@@
  \label{fig:abstract-refinement}}
-\end{figure}
-
-\section{The Concrete Algorithm for Bitvector Programs}
+\end{figure*}
+
+\subsection{The Concrete Algorithm for Bitvector Programs}
 \label{sec:concrete-algorithm}
@@
 
-\begin{figure}
+\begin{figure*}
 \centering
@@
   \label{fig:bitvector-program}
-\end{figure}
+\end{figure*}
 
@@
 
-\subsection{\newC}
+\subsubsection{\newC}
 \label{sec:logic}
@@
 \end{lstlisting}
-\end{minipage}
-\hfill
-\begin{minipage}[t]{.45\linewidth}
+%\end{minipage}
+%\hfill
+%\begin{minipage}[t]{.45\linewidth}
 \begin{lstlisting}[language=C++]
@@
 
-\resizebox{!}{2.5cm}{
+\resizebox{!}{2.3cm}{
 \begin{tikzpicture}[font=\sffamily]
@@
 
-\subsection{Encoding an $\mathcal{L}$-Program in \newC}
+\subsubsection{Encoding an $\mathcal{L}$-Program in \newC}
 \label{sec:encode-l}
@@
 
-\section{Parameterising the Program Space}
+\subsection{Parameterising the Program Space}
 
@@
 
-\subsection{Program Length: $l$}
+\subsubsection{Program Length: $l$}
 The first parameter we introduce is program length, denoted by $l$.
@@
 
-\subsection{Word Width: $w$}
+\subsubsection{Word Width: $w$}
 An $\mathcal{L}$-program runs on a virtual machine (the $\mathcal{L}$-machine) that
@@
 
-\subsection{Number of Constants: $c$}
+\subsubsection{Number of Constants: $c$}
 Instructions in $\mathcal{L}$ take either one or two operands.
@@
 
-\subsection{Searching the Program Space}
+\subsubsection{Searching the Program Space}
 
@@
 
+\iffalse
 \section{Experiments}
@@
 
-  \iffalse
-   \begin{minipage}{.2\textwidth}
-   \bf{P19}($x, m, k$): Exchange two fields of $x$, demarcated by $m$ and $k$.
-   \begin{enumerate}
-    \item $t1 = x >> k$
-    \item $t2 = x \oplus t1$
-    \item $t3 = t2 & m$
-    \item $t4 = t3 << k$
-    \item $t5 = t4 \oplus t3$
-    \item $t6 = t5 \oplus x$
-   \end{enumerate}
-   \end{minipage}
-   \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   \bf{P20}($x$): Next higher unsigned number with the same number of 1 bits.
-   \begin{enumerate}
-    \item $t1 = \sim x$
-    \item $t2 = x \& t1$
-    \item $t3 = x + t2$
-    \item $t4 = x \opus t2$
-    \item $t5 = t4 >> 2$
-    \item $t6 = t5 / t2$
-    \item $t7 = t6 \mid t3$
-   \end{enumerate}
-   \end{minipage}
-   \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   \bf{P21}($x, a, b, c$): Cycling through 3 values $a, b, c$.
-   \begin{enumerate}
-    \item $t1 = x \neq c$
-    \item $t2 = a \oplus c$
-    \item $t3 = x \neq a$
-    \item $t4 = b \oplus c$
-    \item $t5 = t2 \& t2$
-    \item $t6 = t3 \& t4$
-    \item $t7 = t5 \oplus t6$
-    \item $t8 = t7 \oplus c$
-   \end{enumerate}
-   \end{minipage}
-\fi
-
    \begin{minipage}{.2\textwidth}
@@
 
-\end{document}
+\fi

### 2014-06-03 papers/termination/paper.tex
Added proofs.
Added Kalashnikov to the termination paper
+ introduction skeleton

@@
 \section{Introduction}
-Church was BIG into program synthesis~\cite{church-synth}, so you know it's good stuff.
+{\bf First idea:}
+Most reasearch on termination focuses on generating linear ranking functions for relations over mathematical integers/reals as opposed to 
+bit-vectors (machine-level integers)/floating point. However, this may result in different behaviours than those observed in practice.
 
-Something, something, Curry-Howard Isomorphism, something, something, programs-as-proofs.
+Illustration: two examples that terminate for mathematical integers/reals, but don't terminate for bit vectors/floats.
+\begin{itemize}
+\item[1.] q = q+y; y=y+1;
+\item[2.] some floating point example
+\end{itemize}
+
+Hence, we work with bit vectors!
+
+\noindent {\bf Second idea:}
+Moreover, most works on termination assume restrictive transition relations and are concerned with linear ranking functions.
+However, the probability of a random program having a linear ranking function is very low ?
+(see Section ?).
+
+We propose a general framework that does not assume the existence of linear ranking functions and can uniformly handle:
+\begin{itemize}
+\item lexicographic linear functions (no templates).
+\item non-linear functions.
+\item lexicographic non-linear functions.
+\end{itemize}
+
+Illustration: examples of programs that do not have linear ranking functions that we can handle.
+
+\noindent {\bf Third idea:}
+For this purpose, we rephrase the termination problem as a synthesis problem.
+ As a direct consequence, the complexity of the method is dependent on the Kolmogorov complexity of the computed ranking function (+ it's not dependent on the size/number of variables of the analysed program).
+We empirically observe that most programs have ranking functions with low Kolmogorov complexity (see Section ?).
+
+%Church was BIG into program synthesis~\cite{church-synth}, so you know it's good stuff.
+%Something, something, Curry-Howard Isomorphism, something, something, programs-as-proofs.
 
@@
 
+\begin{proof}
+ Combinatorics.
+\end{proof}
+
 \begin{corollary}
@@
 
+\begin{proof}
+ Trivial.
+\end{proof}
+
 \begin{conjecture}
@@
 \end{theorem}
+
+\begin{proof}
+ Trivial.
+\end{proof}
 

### 2014-06-03 papers/termination/kalashnikov.tex

@@
 
-\begin{tabular}{llll}
- \verb|add a b| & \verb|sub a b| & \verb|mul a b| & \verb|div a b| \\
- \verb|neg a| & & &
+\begin{tabular}{lll}
+ \verb|add a b| & \verb|sub a b| & \verb|mul a b| \\
+ \verb|neg a| & \verb|div a b|  & 
 \end{tabular}
@@
 
-\begin{tabular}{llll}
- \verb|and  a b| & \verb|or a b| & \verb|xor a b| & \verb|ashr a b| \\
- \verb|lshr a b| & \verb|not  a| & &
+\begin{tabular}{lll}
+ \verb|and  a b| & \verb|or a b| & \verb|xor a b| \\
+ \verb|lshr a b| & \verb|ashr a b| & \verb|not  a| 
 \end{tabular}
@@
 
-\begin{tabular}{llll}
- \verb|le  a b| & \verb|lt  a b| & \verb|sle  a b| & \verb|slt a b|
+\begin{tabular}{lll}
+ \verb|le  a b| & \verb|lt  a b| & \verb|sle  a b| \\
+ \verb|slt a b| & &
 \end{tabular}
@@
 
-\resizebox{!}{2.3cm}{
+\resizebox{\linewidth}{!}{
 \begin{tikzpicture}[font=\sffamily]
@@
            a \leq x < a+c & \mathtt{consts[} x-a \mathtt{]} \\
-           x \geq a + c & \text{the result of the } (x - a - c)^{\text{th}} \text{ instruction}
+           x \geq a + c & \text{the result of instruction } (x - a - c)
           \end{cases}
@@
 \centering
-\begin{tikzpicture}[scale=0.7, transform shape, node distance=2cm, auto]
+\begin{tikzpicture}[scale=0.65, transform shape, node distance=2cm, auto]
 \tikzstyle{decision} = [diamond, draw, fill=blue!20, 

### 2014-06-03 papers/termination/paper.tex
A whole bunch of restructuring

@@
 \usepackage{pifont}
+\usepackage{listings}
+\usepackage{graphicx}
+\usepackage{xspace}
+\usepackage{pgf}
+\usepackage[noend]{algpseudocode}
+\usepackage{algorithm}
+\usepackage{multicol}
+\usepackage{appendix}
+
+\usepackage{tikz}
+\usetikzlibrary{arrows, automata, shapes}
 
@@
 \noindent {\bf Second idea:}
-Moreover, most works on termination assume restrictive transition relations and are concerned with linear ranking functions.
-However, the probability of a random program having a linear ranking function is very low ?
-(see Section ?).
+Moreover, most works on termination assume restrictive transition relations and are concerned with
+linear ranking functions. However, the probability of a random program having a linear ranking
+function is very low ? (see Section ?).
 
-We propose a general framework that does not assume the existence of linear ranking functions and can uniformly handle:
+We propose a general framework that does not assume the existence of linear ranking
+functions and can uniformly handle:
 \begin{itemize}
@@
 \item lexicographic non-linear functions.
+\item programs with non-linear operations, e.g. logical and.
 \end{itemize}
@@
 \noindent {\bf Third idea:}
-For this purpose, we rephrase the termination problem as a synthesis problem.
- As a direct consequence, the complexity of the method is dependent on the Kolmogorov complexity of the computed ranking function (+ it's not dependent on the size/number of variables of the analysed program).
-We empirically observe that most programs have ranking functions with low Kolmogorov complexity (see Section ?).
+For this purpose, we rephrase the termination problem as a second-order satisfaction problem
+(indeed, a program synthesis problem).
+ As a direct consequence, the complexity of the method is dependent on the Kolmogorov
+ complexity of the computed ranking function (+ it's not dependent on the size/number
+ of variables of the analysed program). We empirically observe that most programs have
+ ranking functions with low Kolmogorov complexity (see Section ?).
 
@@
 to say:
-
 $$\forall x, x' . T(x, x') \Rightarrow R(x) < R(x')$$
 
-\subsection{Program Synthesis as Second-order Satisfaction}
-The program synthesis problem can be described as finding a satisfying assigment to the
-synthesis formula:
+We can specify the existence of a ranking function, and therefore the termination
+of a program, using a second order formula:
 
-\begin{definition}[Synthesis Formula]
- The synthesis formula for a specification $\sigma: X \to Y$ is:
-
- $$\exists P: X \to Y . \forall x : X . \sigma(x, P(x))$$
+\begin{definition}[Second-order Termination Formula]
+\begin{align*}
+ \exists R . \forall x, x' . & R(x) > 0 ~ \wedge \\
+                             & T(x, x') \rightarrow R(x) > R(x')
+\end{align*}
 \end{definition}
 
-Since this formula involves quantification over functions $X \to Y$,
-this is a second-order formula -- indeed, if $X = Y = \mathbb{N}$,
-the formula describes a set at level $\Sigma_1^1$ of the analytical
-hierarchy.  As such, determining the satisfiability of the synthesis
-formula over a given logic is strictly harder than solving the
-halting problem over the same logic.
+A special case of a ranking function is a \emph{linear ranking function}.  This
+class of functions is exactly what you'd expect: a linear function that satisfies
+the criteria for a ranking function.  We recall that a linear function $f: X \to Y$,
+with $\dim(X) = n$ and $\dim(Y) = m$
+is one that can be expressed as an $n \times m$ matrix $M$:
+$$f(\vec{x}) = M\vec{x}$$
 
-\subsection{Generalised Synthesis}
-Sometimes we want to synthesise several programs at once, as well as some
-ground terms $\vec{x}$.  Also, we might want a more complex specification
-than just a relation over $X$ and $Y$.
+In the case that $\dim(Y) = 1$, this reduces to an inner product:
+$$f(\vec{x}) = \vec{\lambda} \cdotp \vec{x} + c$$
 
-\begin{definition}[Generalised Synthesis Formula]
- $$\exists P_1: X_1 \to Y_1, \ldots, P_N: X_N \to Y_N , \vec{x}: X^e . \forall \vec{y}: X^a: X . \sigma(x, P_1, \ldots, P_N, y) $$
-\end{definition}
+If $Y = Z^m$, we say that the ranking function is \emph{lexicographic},
+and require that the total order imposed on $Y$ is the lexicographic ordering
+induced on tuples of $Z$'s.  We note that some termination arguments
+require lexicographic ranking functions, or equivalently, ranking functions
+whose co-domain is the ordinals, rather than just $\mathbb{N}$.
+
+\subsection{Machine Arithmetic and Bitvectors}
+Computers tend to have fixed width integers.  For a machine with $k$-bit words,
+all arithmetic is done modulo $2^k$.  This means that a program's termination
+behaviour can depend strongly on whether it is interpreted as operating over
+mathematical integers, or over fixed-width machine words.  We can however
+observe that if no arithmetic operation \emph{overflows} (i.e. does not result
+in a value greater than $2^k$), fixed-width and integer arithmetic coincide.
 
 \section{Formalism}
+In this section, we fix a model of computation, describe its semantics and
+define the syntax of a language we will work over.
+
 \subsection{Syntax and Semantics}
-
-\section{Reducing Termination to Synthesis}
-For a transition system $T$, we can build a specification $\sigma$ to find a conditional ranking function for
-$T$ with initial states I:
-
-\begin{eqnarray}
- \sigma(V: X \to \mathbb{B}, R: X \to \mathbb{N}, x, x') = & I(x) \rightarrow V(x) \wedge \\
- & V(x) \wedge T(x, x') \rightarrow V(x') \wedge R(x) < R(x') 
-\end{eqnarray}
-
-Here we have synthesised a ranking function $R$ into $\mathbb{N}$ (which is well-founded),
-as well as an inductive invariant $V$ that guarantees termination of $T$.
-
-To synthesise an n-place lexicographic ranking function, we just ask for a ranking function
-$R: X \to \mathbb{N}^n$.
-
-\section{Theory}
-
-In the sequel, we will fix the following notation:
 
@@
 
-\subsection{Termination theory}
+\section{Combinatorics of Termination}
+
+\begin{itemize}
+ \item There are a TON of programs (way more than you'd expect).
+ \item There are a TON of terminating programs, and for our model of computation we can count
+  how many (the Chaitin constant).
+ \item There are a TON of ranking functions (way more than you'd expect, but not many as a
+  fraction of programs).
+ \item There are not many linear functions.
+ \item Most terminating programs don't admit linear ranking functions.
+ \item Kolmogorov complexity is relevant.
+\end{itemize}
 
@@
 
+We recall the definition of the Kolmogorov-complexity of a function $f$:
+
+\begin{definition}[Kolmogorov complexity]
+ The Kolmogorov complexity $K(f)$ is the length of the shortest program that
+ computes $f$.
+\end{definition}
+
+We are primarily interested in studying low-Kolmogorov-complexity (LKC)
+functions.
+
+\begin{theorem}
+ Linear functions are LKC.
+\end{theorem}
+
+\begin{proof}
+ The program to compute a linear function $f: X \to Y$ is of size linear in
+ $\dim(X) \times \dim(Y)$.
+\end{proof}
+
+\begin{conjecture}
+ Most LKC functions are non-linear.
+\end{conjecture}
+
+\begin{corollary}
+ LKC is a weaker assumption than linearity.
+\end{corollary}
+
+\begin{theorem}
+ LKC programs do not always have LKC ranking functions.
+\end{theorem}
+
+\begin{proof}
+ This would solve the halting problem, Goldbach conjecture, Collatz conjecture.
+\end{proof}
+
+\begin{conjecture}
+ High-Kolmogorov-complexity (HKC) programs often have LKC ranking functions.
+\end{conjecture}
+
+\section{Solving the Synthesis Formula}
+
+\subsection{Program Synthesis as Second-order Satisfaction}
+The program synthesis problem can be described as finding a satisfying assigment to the
+synthesis formula:
+
+\begin{definition}[Synthesis Formula]
+ The synthesis formula for a specification $\sigma: X \to Y$ is:
+
+ $$\exists P: X \to Y . \forall x : X . \sigma(x, P(x))$$
+ \end{definition}
+
+Since this formula involves quantification over functions $X \to Y$,
+this is a second-order formula -- indeed, if $X = Y = \mathbb{N}$,
+the formula describes a set at level $\Sigma_1^1$ of the analytical
+hierarchy.  As such, determining the satisfiability of the synthesis
+formula over a given logic is strictly harder than solving the
+halting problem over the same logic.
+
+\subsection{Generalised Synthesis}
+Sometimes we want to synthesise several programs at once, as well as some
+ground terms $\vec{x}$.  Also, we might want a more complex specification
+than just a relation over $X$ and $Y$.
+
+\begin{definition}[Generalised Synthesis Formula]
+ $$\exists P_1: X_1 \to Y_1, \ldots, P_N: X_N \to Y_N , \vec{x}: X^e . \forall \vec{y}: X^a: X . \sigma(x, P_1, \ldots, P_N, y) $$
+\end{definition}
+
+\subsection{Reducing Termination to Synthesis}
+For a transition system $T$, we can build a specification $\sigma$ to find a conditional ranking function for
+$T$ with initial states I:
+
+\begin{eqnarray}
+ \sigma(V: X \to \mathbb{B}, R: X \to \mathbb{N}, x, x') = & I(x) \rightarrow V(x) \wedge \\
+ & V(x) \wedge T(x, x') \rightarrow V(x') \wedge R(x) < R(x') 
+\end{eqnarray}
+
+Here we have synthesised a ranking function $R$ into $\mathbb{N}$ (which is well-founded),
+as well as an inductive invariant $V$ that guarantees termination of $T$.
+
+To synthesise an n-place lexicographic ranking function, we just ask for a ranking function
+$R: X \to \mathbb{N}^n$.
+
+\input{kalashnikov}
+
+\section{Miscellaneous Theory}
+
+\subsection{Termination theory}
+
 \begin{theorem}
@@
  \textsc{Headshot} can often prove termination where linear methods cannot.
-\end{corollary}
-
-\begin{conjecture}
- LKC programs have LKC ranking functions (almost certainly false).
-\end{conjecture}
-
-\begin{corollary}
- Goldbach conjecture, Collatz conjecture.
 \end{corollary}

### 2014-06-03 papers/termination/paper.tex
Restructuring

@@
 
-We can specify the existence of a ranking function, and therefore the termination
-of a program, using a second order formula:
-
-\begin{definition}[Second-order Termination Formula]
-\begin{align*}
- \exists R . \forall x, x' . & R(x) > 0 ~ \wedge \\
-                             & T(x, x') \rightarrow R(x) > R(x')
-\end{align*}
-\end{definition}
-
 A special case of a ranking function is a \emph{linear ranking function}.  This
@@
  \item Most terminating programs don't admit linear ranking functions.
- \item Kolmogorov complexity is relevant.
+ \item The Curry-Howard isomorphism
+ \item Kolmogorov complexity is relevant for understanding termination proofs.
 \end{itemize}
@@
 \end{corollary}
+
+\section{Termination as Second-Order Satisfaction}
+We can specify the existence of a ranking function, and therefore the termination
+of a program, using a second order formula:
+
+\begin{definition}[Second-order Termination Formula]
+\begin{align*}
+ \exists R . \forall x, x' . & R(x) > 0 ~ \wedge \\
+                             & T(x, x') \rightarrow R(x) > R(x')
+\end{align*}
+\end{definition}
+
+\section{Second-Order Satisfaction as Program Synthesis}
+The program synthesis problem can be described as finding a satisfying assigment to the
+synthesis formula:
+
+\begin{definition}[Synthesis Formula]
+ The synthesis formula for a specification $\sigma: X \to Y$ is:
+
+ $$\exists P \cdotp \forall x \cdotp \sigma(x, P(x))$$
+ \end{definition}
+
+Since this formula involves quantification over functions $X \to Y$,
+this is a second-order formula -- indeed, if $X = Y = \mathbb{N}$,
+the formula describes a set at level $\Sigma_1^1$ of the analytical
+hierarchy.  As such, determining the satisfiability of the synthesis
+formula over a given logic is strictly harder than solving the
+halting problem over the same logic.
+
+\subsection{Generalised Synthesis}
+Sometimes we want to synthesise several programs at once, as well as some
+ground terms $\vec{x}$.  Also, we might want a more complex specification
+than just a relation over $X$ and $Y$.
+
+\begin{definition}[Generalised Synthesis Formula]
+ $$\exists P, \vec{x} \cdotp \forall \vec{y} \cdotp \sigma(x, P, y) $$
+\end{definition}
+
+\subsection{Reducing Termination to Synthesis}
+For a transition system $T$, we can build a specification $\sigma$ to find a conditional ranking function for
+$T$ with initial states I:
+
+\begin{eqnarray}
+ \sigma(V: X \to \mathbb{B}, R: X \to \mathbb{N}, x, x') = & I(x) \rightarrow V(x) \wedge \\
+ & V(x) \wedge T(x, x') \rightarrow V(x') \wedge R(x) < R(x') 
+\end{eqnarray}
+
+Here we have synthesised a ranking function $R$ into $\mathbb{N}$ (which is well-founded),
+as well as an inductive invariant $V$ that guarantees termination of $T$.
+
+To synthesise an n-place lexicographic ranking function, we just ask for a ranking function
+$R: X \to \mathbb{N}^n$.
 
@@
 
-\section{Solving the Synthesis Formula}
-
-\subsection{Program Synthesis as Second-order Satisfaction}
-The program synthesis problem can be described as finding a satisfying assigment to the
-synthesis formula:
-
-\begin{definition}[Synthesis Formula]
- The synthesis formula for a specification $\sigma: X \to Y$ is:
-
- $$\exists P: X \to Y . \forall x : X . \sigma(x, P(x))$$
- \end{definition}
-
-Since this formula involves quantification over functions $X \to Y$,
-this is a second-order formula -- indeed, if $X = Y = \mathbb{N}$,
-the formula describes a set at level $\Sigma_1^1$ of the analytical
-hierarchy.  As such, determining the satisfiability of the synthesis
-formula over a given logic is strictly harder than solving the
-halting problem over the same logic.
-
-\subsection{Generalised Synthesis}
-Sometimes we want to synthesise several programs at once, as well as some
-ground terms $\vec{x}$.  Also, we might want a more complex specification
-than just a relation over $X$ and $Y$.
-
-\begin{definition}[Generalised Synthesis Formula]
- $$\exists P_1: X_1 \to Y_1, \ldots, P_N: X_N \to Y_N , \vec{x}: X^e . \forall \vec{y}: X^a: X . \sigma(x, P_1, \ldots, P_N, y) $$
-\end{definition}
-
-\subsection{Reducing Termination to Synthesis}
-For a transition system $T$, we can build a specification $\sigma$ to find a conditional ranking function for
-$T$ with initial states I:
-
-\begin{eqnarray}
- \sigma(V: X \to \mathbb{B}, R: X \to \mathbb{N}, x, x') = & I(x) \rightarrow V(x) \wedge \\
- & V(x) \wedge T(x, x') \rightarrow V(x') \wedge R(x) < R(x') 
-\end{eqnarray}
-
-Here we have synthesised a ranking function $R$ into $\mathbb{N}$ (which is well-founded),
-as well as an inductive invariant $V$ that guarantees termination of $T$.
-
-To synthesise an n-place lexicographic ranking function, we just ask for a ranking function
-$R: X \to \mathbb{N}^n$.
-
-\input{kalashnikov}
-
-\section{Miscellaneous Theory}
-
-\subsection{Termination theory}
-
 \begin{theorem}
@@
 
-\subsection{Synthesis}
+\section{Solving the Synthesis Formula}
+\input{kalashnikov}
+
+\section{Genetic Programming and Incremental Evolution}
+
+We begin by observing that the asymptotic complexity of all of our synthesis
+backends are equal, assuming $P \neq NP$.  This complexity is:
+
+$$O(2^{K(f)}$$
+
+Where $K(f)$ is the Kolmogorov complexity of $f$, which is $O(\log Y^X) = O(X)$
+so the complexity is doubly exponential in the width of $X$.
 
@@
  But it does.
-\end{conjecture}
-
-\begin{conjecture}
- Most incorrect programs are incorrect on most inputs.
 \end{conjecture}

### 2014-06-03 papers/termination/paper.tex
Examples

@@
 \newcommand{\xmark}{\ding{55}}
+
+\lstset{language=c++}
 
@@
 \end{itemize}
+
+A program that does not terminate for integers, but does for bitvectors:
+\begin{lstlisting}
+while(x > 0)
+  x++;
+\end{lstlisting}
+
+A program that terminates for integers, but does not for bitvectors:
+\begin{lstlisting}
+while (x > 0)
+ x -= 2;
+\end{lstlisting}
+
+A program that does not terminate for reals, but does for floats:
+\begin{lstlisting}
+while (x > 0.0)
+ x *= 0.5;
+\end{lstlisting}
+
+A program that terminates for reals, but does not for floats:
+\begin{lstlisting}
+while (x > 0.0)
+ x -= 1.0;
+\end{lstlisting}
+
+A non-linear program with a linear ranking function:
+\begin{lstlisting}
+while (x > 0)
+ x = (x - 1) & x;
+\end{lstlisting}
+
+A linear program with a non-linear ranking function:
+\begin{lstlisting}
+while (x != 0)
+ x = -x / 2;
+\end{lstlisting}
 
@@
 
+We implemented our fully automatic program synthesis procedure in the \textsc{Kalashnikov} tool.
+
 \begin{figure*}

### 2014-06-05 papers/termination/paper.tex
revised intro  Please enter the commit message for your changes. Lines starting

--- 2014-06-03 papers/termination/paper.tex
@@
 \usepackage{appendix}
-
+\usepackage{caption}
+\DeclareCaptionType{copyrightbox}
+\usepackage{subfig}
 \usepackage{tikz}
@@
 \newcommand{\xmark}{\ding{55}}
+\newcommand{\todo}[1]{{\bf TODO:} #1}
+
+\lstset{language=c++}
 
@@
 \section{Introduction}
-{\bf First idea:}
-Most reasearch on termination focuses on generating linear ranking functions for relations over mathematical integers/reals as opposed to 
-bit-vectors (machine-level integers)/floating point. However, this may result in different behaviours than those observed in practice.
 
-Illustration: two examples that terminate for mathematical integers/reals, but don't terminate for bit vectors/floats.
+Proving program termination is typically done by finding a well-founded \emph{ranking function}
+for the program states. When doing so, the majority of the existing termination provers diverge 
+in one important aspect from the execution
+of a program on a computer:
+bit-vectors and floats 
+are treated as mathematical integers and reals, respectively \cite{}.
+This approximation gives rise to 
+{\em unsound} and {\em incomplete} analyses, which is rather surprising given that bit-vectors and floats are ubiquitous in computer systems.
+
+For the case of bit-vectors (machine-level integers), their abstraction  to mathematical integers 
+ignores the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in %, it gives rise to 
+%{\em unsound} and {\em incomplete} analyses, which is rather surprising given that bit-vectors are ubiquitous in computer systems. 
+incomparable behaviours with respect to termination:
+%The differences between the semantics of unbounded mathematical integers and that of machine integers result in 
 \begin{itemize}
-\item[1.] q = q+y; y=y+1;
-\item[2.] some floating point example
+\item Programs that terminate for integers may \emph{not} terminate for bit-vectors. For illustration, consider the following loop:
+\begin{lstlisting}[language=C]
+while (x > 0) x -= 2;
+\end{lstlisting}
+The loop always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
+whereas, with bit-vector arithmetic, \texttt{x} under-flows before becoming negative and goes back to being positive causing the loop to never terminate.
+According to the C11 standard for the C programming language, a signed under- or over-flow yields undefined behaviour.
+However, compilers generally treat signed under- and over-flows using the
+wrap-around behaviour (see Section~\ref{}). 
+\item Similarly, programs that terminate for bit-vectors may \emph{not} terminate for integers. One such situation is illustrated next:  
+ \begin{lstlisting}[language=C]
+ while(x > 0) x++;
+ \end{lstlisting}
+This loop is  terminating for bit-vectors since \texttt{x}
+will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
+arithmetic since the loop condition stays always true for any initial \texttt{x} at least 1.
 \end{itemize}
 
-Hence, we work with bit vectors!
+A scenario similar to the one for bit-vectors happens when floats are abstracted to unbounded reals by termination provers \cite{}.
+Consequently, these provers ignore potential rounding errors, under- and over-flows, which are precisely what makes  reasoning about floating point inherently difficult.
+This approximation may lead to erroneous diagnosis of a program's terminating behaviour as illustrated next:
+\begin{itemize}
+\item This loop does not terminate for reals, but does for floats.
+\begin{lstlisting}
+while (x > 0.0) x *= 0.5;
+\end{lstlisting}
 
-\noindent {\bf Second idea:}
-Moreover, most works on termination assume restrictive transition relations and are concerned with
-linear ranking functions. However, the probability of a random program having a linear ranking
-function is very low ? (see Section ?).
+\item The following loop terminates for reals, but does not for floats.
+\begin{lstlisting}
+while (x > 0.0) x -= 1.0;
+\end{lstlisting}
+\end{itemize}   
+\todo{explain the reasons for non-termination with figures.}
 
-We propose a general framework that does not assume the existence of linear ranking
-functions and can uniformly handle:
-\begin{itemize}
-\item lexicographic linear functions (no templates).
-\item non-linear functions.
-\item lexicographic non-linear functions.
-\item programs with non-linear operations, e.g. logical and.
-\end{itemize}
+Another limitation of the state of the art in the computation of ranking functions is the fact that  most techniques assume \emph{restrictive transition relations} and are only able to compute
+\emph{linear ranking functions} \cite{}. To better understand this restriction, we computed the probability of a random terminating program having a linear ranking
+function (see Section ?). This probability proved to be very low, indicating that the linearity assumption for the ranking functions is indeed prohibitive.
 
-Illustration: examples of programs that do not have linear ranking functions that we can handle.
-
-\noindent {\bf Third idea:}
+We propose a general framework that does not assume the existence of linear ranking functions and can uniformly compute
+(lexicographic) linear and non-linear ones, 
+for programs with both bit-vectors and floats. Our technique can handle programs with non-linear operations, e.g. logical and.
 For this purpose, we rephrase the termination problem as a second-order satisfaction problem
 (indeed, a program synthesis problem). 
- As a direct consequence, the complexity of the method is dependent on the Kolmogorov
- complexity of the computed ranking function (+ it's not dependent on the size/number
- of variables of the analysed program). We empirically observe that most programs have
- ranking functions with low Kolmogorov complexity (see Section ?).
+As a direct consequence, our technique is semantic in nature and does not
+dependent on the size of the analysed program or its number of variables. 
+However, its efficiency does depend on the Kolmogorov
+ complexity of the computed ranking functions. We show that the probability of
+a ranking function to have a low Kolmogorov complexity is higher than the probability to be linear (see Section~?). 
+%We empirically observe that most programs have
+%ranking functions with low Kolmogorov complexity (see Section ?).
 
@@
 
+The main contributions of our paper are highlighted below:
+\begin{itemize}
+\item We designed a technique for computing ranking functions that correctly accounts for the wrap-around behavior caused by under- and overflows in bit-vector and floating point arithmetic.
+To the best of our knowledge, this is the first approach able to compute ranking functions for programs handling floats.
+Our technique is not restricted to finding linear ranking functions, but can also compute (lexicographic) non-linear  ones.
+We justified the need for such a non restrictive procedure by computing the probability ...
+.
+\item  We rephrased the termination problem as a second-order satisfaction problem and made 
+use of results in genetic programming to efficiently solve it. We have also investigated the effects of genetic operators on the search space for ranking functions and computed theoretical 
+bounds on the convergence time ...
+\item We implemented our technique and tried it on a selection of programs handling both bit-vectors and floats.  
+\end{itemize}
+
+This paper is organised as follows:
+
 \section{Related Work}
 Error 404.
+
+\section{Motivational Examples}
+A non-linear program with a linear ranking function (+ program with non-linear operations):
+\begin{lstlisting}
+while (x > 0)
+ x = (x - 1) & x;
+\end{lstlisting}
+
+A linear program with a non-linear ranking function:
+\begin{lstlisting}
+while (x != 0)
+ x = -x / 2;
+\end{lstlisting}
 

### 2014-06-05 papers/termination/paper.tex
Update paper with current strategy

@@
 \usepackage{subfig}
+\usepackage{multirow}
 \usepackage{tikz}
@@
 
+Because most state-of-the-art termination procedures use disjunctive well foundedness arguments, they require the use of
+complex safety provers that are able to reason about unbounded unfoldings of a transition relation.  In contrast to this,
+we are able to build monolithic termination arguments that only need ``local'' reasoning -- our ranking functions
+can be proved sufficient by considering just a single unfolding of a loop.
+
+Many loops do not terminate for all starting states, but are contained in programs that guarantee the loop will
+terminate.  Traditional termination provers have difficulty reasoning about such conditionally-terminating loops.
+We are able to handle such loops by computing \emph{termination invariants}.  This mechanism also allows us to
+prove that programs with multiple loops terminate, even if the termination of some loop depends on the states
+reachable after leaving a previous loop.
+
 We propose a general framework that does not assume the existence of linear ranking functions and can uniformly compute
@@
 a ranking function to have a low Kolmogorov complexity is higher than the probability to be linear (see Section~?).
+
+By phrasing the termination problem as a single rank function synthesis problem, we observe that
+we can make the termination problem as hard as finding ranking functions, rather than as hard as
+solving a difficult ``double reachability'' problem (cf. Terminator).
+
+\begin{figure}
+ \begin{tabular}{|ll||c|c|}
+ \hline
+  & & \multicolumn{2}{c|}{Program} \\
+  & & Linear & Non-Linear \\
+  \hline
+  \hline
+  \multirow{2}{*}{Ranking Functions} & Linear & \checkmark \xmark & \checkmark \\
+   & Non-Linear & \checkmark & \checkmark \\
+   \hline
+ \end{tabular}
+
+ \caption{Legend: \xmark = state-of-the art can handle, \checkmark = we can handle\label{fig:handletable}}
+\end{figure}
+
 %We empirically observe that most programs have
@@
 bounds on the convergence time ...
+\item We present a formulation that allows us to prove termination without using a fancy safety prover (because
+we don't do a DWF proof).  In particular,
+we only need a bounded model checker that performs a single unwinding of a loop.
+\item Our technique is able to uniformly handle conditionally-terminating loops, as well as programs with
+multiple loops.
 \item We implemented our technique and tried it on a selection of programs handling both bit-vectors and floats.
@@
 \section{Related Work}
-Error 404.
+Primarily divided into structural (syntactic) vs. semantic.  For example, terminationg of term rewriting systems
+is largely structural, whereas DWF analysis (cf. Terminator) is more semantic -- it just deals
+with transition relations, but uses some amount of syntax in that it identifies loops \& looping
+paths.
 
@@
 
+\subsection{Disjunctive Well-Foundedness}
+Terminator uses disjunctive well foundedness arguments, which can be found
+piece by piece, but require reasoning about the transitive closure of a transition
+relation.  Because we build a monolithic well foundedness argument, we
+do not need to reason about the full transitive closure, and can instead do our
+reasoning over a single step of the transition relation.  This means that we
+can get away with very simple model checking methods, rather than requiring
+a full-blown safety prover.
+
 \subsection{Machine Arithmetic and Bitvectors}
@@
 
-\section{Formalism}
+\section{Combinatorics of Finite Termination}
 In this section, we fix a model of computation, describe its semantics and
@@
 
-\section{Combinatorics of Termination}
-
+\subsection{Counting Programs}
 \begin{itemize}
@@
 
+\begin{figure*}
 \begin{definition}[Second-order Termination Formula]
@@
 \end{definition}
+
+\begin{definition}[Conditional Termination Formula]
+ \begin{align*}
+  \exists R, I . \forall x, x' . & P(x) \rightarrow I(x) ~ \wedge \\
+                                 & c(x) \wedge I(x) \wedge T(x, x') \rightarrow I(x') \wedge R(x) > 0 \wedge R(x) > R(x')
+ \end{align*}
+\end{definition}
+
+\begin{definition}[Multiple Loops Termination Formula]
+ \begin{align*}
+  \exists R, I_1, I_2 . \forall x, x' . & P(x) \rightarrow I_1(x) ~ \wedge \\
+                                        & I_1(x) \wedge L_1(x, x') \rightarrow I_1(x') ~ \wedge \\
+                                        & I_1(x) \wedge \lnot c_1(x) \rightarrow I_2(x) ~ \wedge \\
+                                        & c(x) \wedge I_2(x) \wedge L_2(x, x') \rightarrow I_2(x') \wedge R(x) > 0 \wedge R(x) > R(x')
+ \end{align*}
+\end{definition}
+\end{figure*}
 
@@
 \section{Solving the Synthesis Formula}
+We can solve the synthesis problem using any off-the-shelf program synthesiser.  If you don't
+have a program synthesiser kicking around, we present the design of one synthesiser
+that can be used to solve the problem.
+
 \input{kalashnikov}
@@
 
-We implemented our fully automatic program synthesis procedure in the \textsc{Kalashnikov} tool.
+We proved termination for a bunch of programs, see Fig.~\ref{fig:linear} and Fig.~\ref{fig:nonlinear}.
 

### 2014-06-12 papers/termination/paper.tex
messy stuff

@@
 
+%completeness without disj well-foundedness
+%. only non-strict inequalities can be transformed using Farkas’ Lemma
+
 This paper is organised as follows:
@@
 \section{Related Work}
-Primarily divided into structural (syntactic) vs. semantic.  For example, terminationg of term rewriting systems
-is largely structural, whereas DWF analysis (cf. Terminator) is more semantic -- it just deals
-with transition relations, but uses some amount of syntax in that it identifies loops \& looping
-paths.
+\subsection{Termination analysis}
+The existing work on termination is primarily divided into structural (syntactic) vs. semantic approaches.  
+For example, terminationg of term rewriting systems is largely structural, whereas DWF analysis (cf. Terminator) is more semantic. While it deals
+with transition relations, it also uses some amount of syntax in that it identifies loops \& looping paths.
+
+A program is lasso-shaped if it is composed by a stem followed by a single loop without branching, i.e. there is only one path. 
+Consequently, the termination argument can be very simple. There are numerous techniques specialised in
+proving termination of lasso-shaped programs efficiently \cite{DBLP:conf/popl/Ben-AmramG13,DBLP:conf/cav/BradleyMS05,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/P04}.
+
+Cook et al. are closer to a semantic approach by viewing the termination proving task as a binary reachability analysis \cite{DBLP:conf/pldi/CookPR06}.
+%Terminator incrementally constructs the union of ranking relations.
+%Given that the this union is not well-founded (it is only disjunctively well-founded), 
+%Terminator must prove that the binary reachability relation of the program (the transitive closure of the transition relation restricted to reachable states) 
+%is contained in the union of ranking relations.
+%Thus, this approach requires an evolved binary reachability, and, implicitly, the assistance of a safety prover.
+In particular, their technique incrementally constructs the union of ranking relations.
+Given that the this union is not well-founded (it is only disjunctively well-founded),
+they must show that, for every sequence, one of the ranking functions decreases
+between the pre- and post-state. %In other words: it 
+%must find all pairs of states reachable from the program's initial state 
+%such that one is reachable from the other and it must show 
+%that the value of one of the ranking functions decreases from s1 to s2. 
+Consequently, most of the analysis time is spent in reachability analysis that is to check if the current
+transition invariant reached a fixed point.
+
+As we want to avoid the requirements of a binary reachability analysis, our termination argument is not a set
+of ranking functions, but a single ranking function, which we prove is enough for proving the termination of any termination program. 
+With a single ranking function we only need to show that the rank decreases from the
+pre- to post-state after executing each single transition step.
+The checking process is therefore easy. For the construction of the ranking function, we 
+formulate the problem as a second order satisfaction formula and we solve it via program synthesis.
+
+Lee et al. make use of transition predicate abstraction, algorithmic learning, and decision procedures to compute transition
+invariants as proofs for the termination of linear programs \cite{DBLP:conf/cav/LeeWY12}.
+
+There is very limited work on generating ranking functions for relations over machine integers.
+Cook et al. present a method based on a reduction to Presburger
+arithmetic, and a template-matching approach for predefined classes of
+ranking functions based on reduction to SAT- and QBF-solving \cite{DBLP:conf/tacas/CookKRW10}.
+
+Leike and Heizmann present a new method for the constraint-based synthesis
+of termination arguments for linear loop programs based on
+linear ranking templates \cite{DBLP:conf/tacas/LeikeH14}.
+
+Learning: \cite{DBLP:journals/corr/HeizmannHP14}
+
+Polynomial programs: \cite{DBLP:conf/vmcai/BradleyMS05}
+
+%-- Usually the termination argument for the program LOOPS
+%on Figure 1 is based on a lexicographic combination
+%of well-founded orderings.
+
+%-- the union T of ranking relations, however, is in general not well-founded
+%(it is only disjunctively well-founded in the terminology of [23]). This is
+%why it would not be sufficient to show the inclusion R \subseteq  T, and why we must in stead prove R_I^+ \subseteq T.
+
+\subsection{Synthesis}
+
+%% ---
+%% constructs and checks the termination argument
+
+%% T ERMINATOR’s most distinguishing aspect, with respect to pre-
+%% vious methods and tools for proving program termination, is how it
+%% shifts the balance between the two tasks of constructing and respec-
+%% tively checking the termination argument. The classical method is
+%% to construct an expression defining the rank of a state and then to
+%% check that its value decreases in every transition from a reachable
+%% state to a next one. The construction of the ranking function is the
+%% hard part and forms a task that needs to be applied to the whole
+%% program. The checking part is relatively easy. In our method, the
+%% task of constructing ranking functions is the relatively easy part;
+%% they are constructed on demand based on the examination of only
+%% a few selected paths through the program.
+
+%% Previously, it was not known whether binary reachability analysis could be made
+%% practical. The challenge raised by our approach was to show that
+%% this is indeed the case.
+
+%% ----------
+%% Efficient synthesis of ranking functions for machine-level bit-vectors, how-
+%% ever, has remained an open problem.
+%% ---------
 
 \section{Motivational Examples} 
-A non-linear program with a linear ranking function (+ program with non-linear operations):
+We next discus a few examples that illustrate the difficulties in termination checking for low-level code. In particular, they denote situations that arise 
+frequently in low-level code and are not handled by existing techniques.
+
+%We now present a few examples that justify
+
+\subsection{A non-linear program with a linear ranking function.}
+The first example we consider is a program with non-linear operations: 
 \begin{lstlisting}
@@
 
-A linear program with a non-linear ranking function:
+In order to find a ranking function for this example, it is necessary to take into account
+the semantics of the bit-wise AND operator, which is not easily done when working with mathematical integers \cite{}.
+Our technique finds that a possible ranking function is the linear function
+$R(x) = x$, whose value decreases with
+every iteration, but it can not decrease indefinitely as it is bounded from below.
+
+\subsection{A linear program with a non-linear ranking function.}
 \begin{lstlisting}
@@
 \end{lstlisting}
+
+\subsection{Differences in the termination behaviour for reals and floats.}
+\begin{lstlisting}
+while (x > 0.0) x -= 1.0;
+\end{lstlisting}
+
+\begin{lstlisting}
+while (x > 0.0) x *= 0.5;
+\end{lstlisting}
+
+\subsection{A linear program with lexicografic non-linear function.}
+\begin{lstlisting}
+while ( q >= 0 ) {
+  q = q - y;
+  y = y + 1;
+}
+\end{lstlisting}
+
+Every execution of the above program can be partitioned into two phases: in the first phase $y$ increases
+until it is positive (in this phase $q$ may increase), whereas in the second $q$ decreases until the loop condition is violated. 
+%Depending on the initial values of y and q, either phase might be skipped
+%altogether.
+\cite{DBLP:conf/tacas/LeikeH14} proposes using a linear ranking template in order to obtain a ranking function that proceeds
+through a fixed number of phases in the program execution. Each phase is
+ranked by an affine-linear function, and ends when this function becomes non-positive.
+
+Lexicographic non-linear ranking functions consist of lexicographically ordered components
+of non-linear functions. As with linear lexicographic ranking functions, a state is mapped to a tuple of values such that the
+loop transition leads to a decrease with respect to the lexicographic
+ordering for this tuple. Therefore no function may increase unless a function of
+a lower index decreases. Additionally, at every step, there must be at least one
+function that decreases.
+
+\subsection{Lexicographic ranking function with strict inequalities.}
+Approaches based on Farkas’ Lemma can only handle non-strict inequalities \cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/vmcai/P04}.
+[2,5,6,7,14,20,24,25
 
@@
 \end{figure*}
+
+Our method for ranking function synthesis can be stated as follows:
+discuss what spec is used (non-lexicographic vs lexicographic) + the completeness claims.
+Any termination guarantees?
 

### 2014-06-13 papers/termination/experiments/benchmarks/loop2.c
Rework some benchmarks

@@ -9,8 +9,8 @@
  */
 
 int main(void) {
-  unsigned int x = nondet();
-  unsigned int N = nondet();
+  unsigned int x;
+  unsigned int N;
 
   while (x < N) {
     x++;

### 2014-06-13 papers/termination/experiments/benchmarks/loop3.c

@@ -9,7 +9,7 @@
  */
 
 int main(void) {
-  unsigned int x = nondet();
+  unsigned int x;
   int y = 1;
 
   while (x > 0) {

### 2014-06-13 src/kalashnikov/frontend.py
Move frontend to kalashnikov/

@@ -1,0 +1,204 @@
+#!/usr/bin/python
+
+import os
+import sys
+import copy
+import tempfile
+
+sys.path.extend(["pycparser"])
+
+from pycparser import parse_file, c_generator, c_ast
+
+def parse_file_libc(filename):
+  return parse_file(filename, use_cpp=True,
+      cpp_args="-Ipycparser/utils/fake_libc_include")
+
+class DeclVisitor(c_ast.NodeVisitor):
+  def __init__(self):
+    self.ids = {}
+    self.revids = {}
+
+  def visit_Decl(self, decl):
+    if not isinstance(decl.type, c_ast.TypeDecl):
+      return
+
+    if decl.name not in self.ids:
+      n = len(self.ids)
+      self.ids[decl.name] = n
+      self.revids[n] = decl.name
+
+def number_ids(ast):
+  v = DeclVisitor()
+  v.visit(ast)
+
+  return (v.ids, v.revids)
+
+def is_loop(ast):
+  return (isinstance(ast, c_ast.DoWhile) or
+      isinstance(ast, c_ast.While) or
+      isinstance(ast, c_ast.For))
+
+loop_memo = {}
+
+def contains_loop(ast):
+  if ast in loop_memo:
+    return loop_memo[ast]
+
+  ret = False
+
+  if is_loop(ast):
+    ret = True
+  else:
+    for (_, c) in ast.children():
+      if contains_loop(c):
+        ret = True
+        break
+
+  loop_memo[ast] = ret
+
+  return ret
+
+def add_parents(ast):
+  if getattr(ast, 'parent', None) is None:
+    ast.parent = None
+
+  for c in ast.children():
+    c[1].parent = ast
+    add_parents(c[1])
+
+class FlatProgram(object):
+  def __init__(self):
+    self.blocks = []
+    self.start_straightline()
+
+  def start_straightline(self):
+    self.curr = []
+    compound = c_ast.Compound(self.curr)
+    self.blocks.append(compound)
+
+  def accumulate(self, ast):
+    if is_loop(ast):
+      self.blocks.append(ast)
+      self.start_straightline()
+    elif isinstance(ast, c_ast.Compound):
+      self.curr += ast.block_items
+    else:
+      self.curr.append(ast)
+
+def flatten(ast, program):
+  if not contains_loop(ast):
+    program.accumulate(ast)
+  elif is_loop(ast):
[…]

### 2014-06-13 src/kalashnikov/pycparser/z.py

@@ -1,0 +1,94 @@
+import sys
+from pycparser.c_ast import *
+from pycparser.c_parser import CParser, Coord, ParseError
+from pycparser.c_lexer import CLexer
+
+
+def expand_decl(decl):
+    """ Converts the declaration into a nested list.
+    """
+    typ = type(decl)
+
+    if typ == TypeDecl:
+        return ['TypeDecl', expand_decl(decl.type)]
+    elif typ == IdentifierType:
+        return ['IdentifierType', decl.names]
+    elif typ == ID:
+        return ['ID', decl.name]
+    elif typ in [Struct, Union]:
+        decls = [expand_decl(d) for d in decl.decls or []]
+        return [typ.__name__, decl.name, decls]
+    else:
+        nested = expand_decl(decl.type)
+
+        if typ == Decl:
+            if decl.quals:
+                return ['Decl', decl.quals, decl.name, nested]
+            else:
+                return ['Decl', decl.name, nested]
+        elif typ == Typename: # for function parameters
+            if decl.quals:
+                return ['Typename', decl.quals, nested]
+            else:
+                return ['Typename', nested]
+        elif typ == ArrayDecl:
+            dimval = decl.dim.value if decl.dim else ''
+            return ['ArrayDecl', dimval, nested]
+        elif typ == PtrDecl:
+            return ['PtrDecl', nested]
+        elif typ == Typedef:
+            return ['Typedef', decl.name, nested]
+        elif typ == FuncDecl:
+            if decl.args:
+                params = [expand_decl(param) for param in decl.args.params]
+            else:
+                params = []
+            return ['FuncDecl', params, nested]
+
+#-----------------------------------------------------------------
+class NodeVisitor(object):
+    def __init__(self):
+        self.current_parent = None
+
+    def visit(self, node):
+        """ Visit a node.
+        """
+        method = 'visit_' + node.__class__.__name__
+        visitor = getattr(self, method, self.generic_visit)
+        return visitor(node)
+
+    def visit_FuncCall(self, node):
+        print("Visiting FuncCall")
+        print(node.show())
+        print('---- parent ----')
+        print(self.current_parent.show())
+
+    def generic_visit(self, node):
+        """ Called if no explicit visitor function exists for a
+            node. Implements preorder visiting of the node.
+        """
+        oldparent = self.current_parent
+        self.current_parent = node
+        for c in node.children():
[…]

### 2014-06-16 papers/termination/paper.tex
Change results table
revised intro+motivation+other things

@@
 
-Proving program termination is typically done by finding a well-founded \emph{ranking function}
-for the program states. When doing so, the majority of the existing termination provers diverge 
-in one important aspect from the execution
-of a program on a computer:
+Proving program termination is typically done by finding a \emph{ranking function}
+for the program states, i.e. a map from the program's state space to a well-ordered set.
+As this definition of ranking function is very general, research is often limited to some
+convenient and tractable form of ranking functions, most frequently \emph{linear ranking functions} (see Section~\ref{sec:ranking.functions}). 
+In doing so, an analysis ensures that such a ranking function will be found is one  exists, 
+while failing to prove termination for terminating programs whose ranking functions fall short of the considered restriction.
+
+Moreover, the majority of the existing termination provers diverge 
+in one important aspect from the execution of a program on a computer:
 bit-vectors and floats 
-are treated as mathematical integers and reals, respectively \cite{}.
+are treated as mathematical integers and reals, respectively.
 This approximation gives rise to 
@@
 
-For the case of bit-vectors (machine-level integers), their abstraction  to mathematical integers 
+Figure~\ref{fig:handletable} presents a summary of existing termination analyses with indeed
+most of the work concentrated on finding \emph{linear ranking functions for linear programs over integers}.
+Next, we will look in more detail at the implications of the assumptions made by such techniques.\\
+
+\noindent {\bf Bit-vectors (machine-level integers) vs. mathematical integers.} The abstraction of bit-vectors to mathematical integers 
 ignores the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in %, it gives rise to 
@@
 However, compilers generally treat signed under- and over-flows using the
-wrap-around behaviour (see Section~\ref{}). 
+wrap-around behaviour (see Section~\ref{sec:machine.arith}). 
 \item Similarly, programs that terminate for bit-vectors may \emph{not} terminate for integers. One such situation is illustrated next:  
@@
 
-A scenario similar to the one for bit-vectors happens when floats are abstracted to unbounded reals by termination provers \cite{}.
+\noindent {\bf Floats vs. reals.} A scenario similar to the one for bit-vectors happens if floats were to be abstracted to unbounded reals by termination provers \cite{}. 
 Consequently, these provers ignore potential rounding errors, under- and over-flows, which are precisely what makes  reasoning about floating point inherently difficult.
@@
 \end{itemize}   
-\todo{explain the reasons for non-termination with figures.}
+\todo{explain the reasons for non-termination with figures.}\\
 
-Another limitation of the state of the art in the computation of ranking functions is the fact that  most techniques assume \emph{restrictive transition relations} and are only able to compute
+\noindent {\bf Linear programs and linear ranking functions.} As visible in Figure~\ref{fig:handletable}, 
+most termination techniques assume \emph{restrictive transition relations}, e.g. linear programs,  and are only able to compute
 \emph{linear ranking functions} \cite{}. To better understand this restriction, we computed the probability of a random terminating program having a linear ranking
@@
 
-Because most state-of-the-art termination procedures use disjunctive well foundedness arguments, they require the use of
-complex safety provers that are able to reason about unbounded unfoldings of a transition relation.  In contrast to this,
-we are able to build monolithic termination arguments that only need ``local'' reasoning -- our ranking functions
-can be proved sufficient by considering just a single unfolding of a loop.
-
-Many loops do not terminate for all starting states, but are contained in programs that guarantee the loop will
-terminate.  Traditional termination provers have difficulty reasoning about such conditionally-terminating loops.
-We are able to handle such loops by computing \emph{termination invariants}.  This mechanism also allows us to
-prove that programs with multiple loops terminate, even if the termination of some loop depends on the states
-reachable after leaving a previous loop.
-
 We propose a general framework that does not assume the existence of linear ranking functions and can uniformly compute
-(lexicographic) linear and non-linear ones, 
-for programs with both bit-vectors and floats. Our technique can handle programs with non-linear operations, e.g. logical and.
+\emph{lexicographic/non-lexicographic} \emph{linear/non-linear} 
+ranking functions supported by inductive \emph{linear/non-linear} invariants for loops with 
+\emph{linear/non-linear} guards and transitions over bit-vectors and floats.
+Our technique is {\emph complete} and can handle programs with non-linear operations, e.g. logical and.
 For this purpose, we rephrase the termination problem as a second-order satisfaction problem
@@
 
-By phrasing the termination problem as a single rank function synthesis problem, we observe that
-we can make the termination problem as hard as finding ranking functions, rather than as hard as
-solving a difficult ``double reachability'' problem (cf. Terminator).
-
-\begin{figure}
- \begin{tabular}{|ll||c|c|}
+\begin{figure*}
+\centering
+ \begin{tabular}{|ll||c|c|c|c|c|c|c|c|}
  \hline
-  & & \multicolumn{2}{c|}{Program} \\
-  & & Linear & Non-Linear \\
+  & & \multicolumn{8}{c|}{Program} \\
+  & & \multicolumn{2}{c|}{Integers} & \multicolumn{2}{c|}{Reals} & \multicolumn{2}{c|}{Bit-vectors} & \multicolumn{2}{c|}{Floats} \\
+  & & L & NL & L & NL & L & NL & L & NL \\
   \hline
   \hline
-  \multirow{2}{*}{Ranking Functions} & Linear & \checkmark \xmark & \checkmark \\
-   & Non-Linear & \checkmark & \checkmark \\
+  \multirow{4}{*}{Ranking} & Linear lexicographic &  \cite{DBLP:conf/cav/BradleyMS05} && & &\checkmark&\checkmark&\checkmark&\checkmark\\
+   & Linear non-lexicographic & \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13} & & && \checkmark \cite{DBLP:conf/tacas/CookKRW10} &\checkmark \cite{DBLP:conf/tacas/CookKRW10}&\checkmark&\checkmark\\
+   & Non-linear lexicographic &  &  & &&\checkmark&\checkmark&\checkmark&\checkmark\\
+   & Non-linear non-lexicographic &  &  & &&\checkmark&\checkmark&\checkmark&\checkmark\\
    \hline
@@
 
- \caption{Legend: \xmark = state-of-the art can handle, \checkmark = we can handle\label{fig:handletable}}
-\end{figure}
+ \caption{Legend: \checkmark = we can handle\label{fig:handletable}}
+\end{figure*}
 
@@
 
-The main contributions of our paper are highlighted below:
+ The main contributions of our work can be summarised as follows:
 \begin{itemize}
@@
 
-%completeness without disj well-foundedness
 %. only non-strict inequalities can be transformed using Farkas’ Lemma
+%--Our approach to termination analysis has two distinct features over previous works
+%-- a lexicographic ranking function imposes a lexicographic ordering on among the ranking function components.
+
+Other points to touch in the introduction:
+\begin{itemize}
+\item disjunctive well-foundedness: By phrasing the termination problem as a single rank function synthesis problem, we observe that
+we can make the termination problem as hard as finding ranking functions, rather than as hard as
+solving a difficult ``double reachability'' problem (cf. Terminator).
+\item The lexicographic ordering does not influence the size of the search space.
+Thus, finding potentially lexicographic termination arguments is not more difficult than finding non-lexicographic ones.
+\item Compared to the previous approaches that mostly aim to find a particular form of ranking functions, our technique does not commit to any particular one.
+We will find the one that has the lowest Kolmogorov complexity.
+\item Many loops do not terminate for all starting states, but are contained in programs that guarantee the loop will
+terminate.  Traditional termination provers have difficulty reasoning about such conditionally-terminating loops.
+We are able to handle such loops by computing \emph{termination invariants}.  This mechanism also allows us to
+prove that programs with multiple loops terminate, even if the termination of some loop depends on the states
+reachable after leaving a previous loop.
+
+\end{itemize}
 
@@
 \subsection{Termination analysis}
+Given the multitude of works on termination, we have tried to summarise them in Figure~\ref{fig:handletable}.
+
 The existing work on termination is primarily divided into structural (syntactic) vs. semantic approaches.  
@@
 
+Lee et al. make use of transition predicate abstraction, algorithmic learning, and decision procedures to compute transition
+invariants as proofs for the termination of linear programs \cite{DBLP:conf/cav/LeeWY12}.
+
+There is very limited work on generating ranking functions for relations over machine integers.
+Cook et al. present a method based on a reduction to Presburger
+arithmetic, and a template-matching approach for predefined classes of
+ranking functions based on reduction to SAT- and QBF-solving \cite{DBLP:conf/tacas/CookKRW10}.
+
+Leike and Heizmann present a new method for the constraint-based synthesis
+of termination arguments for linear loop programs based on
+linear ranking templates \cite{DBLP:conf/tacas/LeikeH14}.
+
+Learning: \cite{DBLP:journals/corr/HeizmannHP14}
+
+Polynomial programs: \cite{DBLP:conf/vmcai/BradleyMS05}
+
+Linear ranking functions supported by inductive linear invariants for loops with linear guards and transitions: \cite{DBLP:conf/cav/BradleyMS05}.
+
+\subsection{Disjunctive well-foundedness}
 Cook et al. are closer to a semantic approach by viewing the termination proving task as a binary reachability analysis \cite{DBLP:conf/pldi/CookPR06}.
@@
 
-Lee et al. make use of transition predicate abstraction, algorithmic learning, and decision procedures to compute transition
-invariants as proofs for the termination of linear programs \cite{DBLP:conf/cav/LeeWY12}.
-
-There is very limited work on generating ranking functions for relations over machine integers.
-Cook et al. present a method based on a reduction to Presburger
-arithmetic, and a template-matching approach for predefined classes of
-ranking functions based on reduction to SAT- and QBF-solving \cite{DBLP:conf/tacas/CookKRW10}.
-
-Leike and Heizmann present a new method for the constraint-based synthesis
-of termination arguments for linear loop programs based on
-linear ranking templates \cite{DBLP:conf/tacas/LeikeH14}.
-
-Learning: \cite{DBLP:journals/corr/HeizmannHP14}
-
-Polynomial programs: \cite{DBLP:conf/vmcai/BradleyMS05}
+Because most state-of-the-art termination procedures use disjunctive well foundedness arguments, they require the use of
+complex safety provers that are able to reason about unbounded unfoldings of a transition relation.  In contrast to this,
+we are able to build monolithic termination arguments that only need ``local'' reasoning -- our ranking functions
+can be proved sufficient by considering just a single unfolding of a loop.
 
@@
 We next discus a few examples that illustrate the difficulties in termination checking for low-level code. In particular, they denote situations that arise 
-frequently in low-level code and are not handled by existing techniques.
+frequently in low-level code and are not handled properly by existing techniques.
 
@@
 
-\subsection{A non-linear program with a linear ranking function.}
-The first example we consider is a program with non-linear operations: 
-\begin{lstlisting}
-while (x > 0)
- x = (x - 1) & x;
-\end{lstlisting}
+\subsection{Program linearity vs. the linearity of its ranking function.}
+The first examples we consider illustrate the lack of a direct correlation between the linearity of a program and that of its termination arguments.
+As such, Figure~\ref{} denotes a program with non-linear operations. Given that  many of the existing techniques commit to linear programs, they cannot handle this situation, 
+although a linear ranking function exists (see Figure~\ref{fig:handletable}).
 
@@
 
-\subsection{A linear program with a non-linear ranking function.}
-\begin{lstlisting}
-while (x != 0)
- x = -x / 2;
-\end{lstlisting}
+Conversely, Figure~\ref{} shows a linear program with a non-linear ranking function.
 
-\subsection{Differences in the termination behaviour for reals and floats.}
-\begin{lstlisting}
-while (x > 0.0) x -= 1.0;
-\end{lstlisting}
-
-\begin{lstlisting}
-while (x > 0.0) x *= 0.5;
-\end{lstlisting}
-
-\subsection{A linear program with lexicografic non-linear function.}
-\begin{lstlisting}
-while ( q >= 0 ) {
-  q = q - y;
-  y = y + 1;
-}
-\end{lstlisting}
-
+Figure~\ref{} is an example of a linear program that only admits lexicografic non-linear ranking functions.
 Every execution of the above program can be partitioned into two phases: in the first phase $y$ increases
@@
 
-\subsection{Lexicographic ranking function with strict inequalities.}
-Approaches based on Farkas’ Lemma can only handle non-strict inequalities \cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/vmcai/P04}.
-[2,5,6,7,14,20,24,25
+\begin{lstlisting}
+while (x > 0)
+ x = (x - 1) & x;
+\end{lstlisting}
+
+\begin{lstlisting}
+while (x != 0)
+ x = -x / 2;
+\end{lstlisting}
+
+\begin{lstlisting}
+while ( q >= 0 ) {
+  q = q - y;
+  y = y + 1;
+}
+\end{lstlisting}
+
+\subsection{Differences in the termination behaviour for integers and bit-vectors.}
+We have collected a number of motivational examples from other termination papers that treated bit-vectors as mathematical integers \cite{DBLP:conf/tacas/LeikeH14,DBLP:conf/tacas/CookSZ13}. 
+We show that the termination arguments computed by such techniques do not directly apply when considering the bit-vector semantics.
+%these programs actually exhibit different terminating behaviours for bit-vectors and integer
+
+\subsection{Differences in the termination behaviour for reals and floats.}
+\begin{lstlisting}
+while (x > 0.0) x -= 1.0;
+\end{lstlisting}
+
+\begin{lstlisting}
+while (x > 0.0) x *= 0.5;
+\end{lstlisting}
+
+\subsection{Multi-phase ranking functions}
+%\subsection{Lexicographic ranking function with strict inequalities.}
+%Approaches based on Farkas’ Lemma can only handle non-strict inequalities \cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/vmcai/P04}.
+\begin{lstlisting}
+int main(void) {
+  int x, y, z;
+
+  while (x > 0 && y > 0 && z > 0) {
+    if (y > x) {
+      y = z;
+      x = nondet();
+      z = x - 1;
+    } else {
+      z = z - 1;
+      x = nondet();
+      y = x - 1;
+    }
+  }
+}
+\end{lstlisting}
 
 \section{Preliminaries}
-\subsection{Termination and Ranking Functions}
+\subsection{Termination and Ranking Functions} \label{sec:ranking.functions}
+\todo{linear program}
+
 A transition system with state space $X$ and transition relation $T \subseteq X \times X$
@@
 
-\subsection{Machine Arithmetic and Bitvectors}
+\subsection{Machine Arithmetic and Bitvectors} \label{sec:machine.arith} 
 Computers tend to have fixed width integers.  For a machine with $k$-bit words,
@@
 
+\subsection{The Search Space and the Kolmogorov Complexity of a Ranking Function}
+\begin{conjecture}
+The lexicographic ordering does not influence the size of the search space. Thus, finding potentially lexicographic termination arguments is not more difficult than finding non-lexicographic ones.   
+\end{conjecture}
+
+As a consequence of the aforementioned conjecture, we do not need to commit to a specific form of the termination argument. While other approaches, e.g. \cite{DBLP:conf/tacas/LeikeH14}, 
+conduct independent searches for each possible form of the ranking function with some of them futile whenever the assumed constrained ranking function does not exist, we have a single search that aims at finding the ranking function with the lowest Kolmogorov complexity.
+
 \input{kalashnikov}

### 2014-06-16 papers/termination/paper.tex
small changes to intro and motivation

@@
 for the program states, i.e. a map from the program's state space to a well-ordered set.
-As this definition of ranking function is very general, research is often limited to some
+As this definition of a ranking function is very general, research is often limited to some
 convenient and tractable form of ranking functions, most frequently \emph{linear ranking functions} (see Section~\ref{sec:ranking.functions}). 
-In doing so, an analysis ensures that such a ranking function will be found is one  exists, 
+In doing so, an analysis ensures that such a ranking function will be found if one  exists, 
 while failing to prove termination for terminating programs whose ranking functions fall short of the considered restriction.
@@
 most of the work concentrated on finding \emph{linear ranking functions for linear programs over integers}.
-Next, we will look in more detail at the implications of the assumptions made by such techniques.\\
+Next, we will consider in more detail the assumptions made by such techniques and their implications.\\
 
@@
   \multirow{4}{*}{Ranking} & Linear lexicographic &  \cite{DBLP:conf/cav/BradleyMS05} && & &\checkmark&\checkmark&\checkmark&\checkmark\\
-   & Linear non-lexicographic & \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13} & & && \checkmark \cite{DBLP:conf/tacas/CookKRW10} &\checkmark \cite{DBLP:conf/tacas/CookKRW10}&\checkmark&\checkmark\\
+   & Linear non-lexicographic & \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13} & & && \checkmark~ \cite{DBLP:conf/tacas/CookKRW10} &\checkmark~ \cite{DBLP:conf/tacas/CookKRW10}&\checkmark&\checkmark\\
    & Non-linear lexicographic &  &  & &&\checkmark&\checkmark&\checkmark&\checkmark\\
@@
 reachable after leaving a previous loop.
-
+\item Linearity vs. LKC.
 \end{itemize}
@@
 \section{Motivational Examples} 
-We next discus a few examples that illustrate the difficulties in termination checking for low-level code. In particular, they denote situations that arise 
-frequently in low-level code and are not handled properly by existing techniques.
+We next discus some intricate examples that prove challenging for existent termination analysis.
+%a few examples that illustrate the difficulties in termination checking for low-level code. In particular, they denote situations that arise 
+%frequently in practise and are not properly handled by existing techniques.
 
@@
 As such, Figure~\ref{} denotes a program with non-linear operations. Given that  many of the existing techniques commit to linear programs, they cannot handle this situation, 
-although a linear ranking function exists (see Figure~\ref{fig:handletable}).
+although a linear ranking function does exist (see Figure~\ref{fig:handletable}).
 
@@
 
-Figure~\ref{} is an example of a linear program that only admits lexicografic non-linear ranking functions.
-Every execution of the above program can be partitioned into two phases: in the first phase $y$ increases
-until it is positive (in this phase $q$ may increase), whereas in the second $q$ decreases until the loop condition is violated. 
-%Depending on the initial values of y and q, either phase might be skipped
-%altogether.
-\cite{DBLP:conf/tacas/LeikeH14} proposes using a linear ranking template in order to obtain a ranking function that proceeds
-through a fixed number of phases in the program execution. Each phase is
-ranked by an affine-linear function, and ends when this function becomes non-positive.
-
-Lexicographic non-linear ranking functions consist of lexicographically ordered components
+Figure~\ref{} is a linear program taken from \cite{DBLP:conf/tacas/CookSZ13}, where it is shown to not admit (without prior manipulation) a lexicographic linear function, but only a disjunctively well-founded one,
+which requires an expensive binary reachability analysis.
+However, by using our technique we can find a non-linear lexicographic ranking functions.
+A lexicographic non-linear ranking functions consists of lexicographically ordered components
 of non-linear functions. As with linear lexicographic ranking functions, a state is mapped to a tuple of values such that the
@@
 \begin{lstlisting}
+  while (x != 0) {
+    if (x > 0)
+      x--;
+    else
+      x++;
+  }
+\end{lstlisting}
+
+\subsection{Differences in the termination behaviour for integers and bit-vectors.}
+We have collected a number of motivational examples from other termination papers that treat bit-vectors as mathematical integers \cite{DBLP:conf/tacas/LeikeH14,DBLP:conf/tacas/CookSZ13}. 
+We show that the termination arguments computed by such techniques do not directly apply when considering the bit-vector semantics.
+%these programs actually exhibit different terminating behaviours for bit-vectors and integer
+
+\begin{lstlisting}
+  while (x != m) {
+    if (x > m)
+      x = 0;
+    else
+      x++;
+  }
+\end{lstlisting}
+
+\begin{lstlisting}
 while ( q >= 0 ) {
@@
 
-\subsection{Differences in the termination behaviour for integers and bit-vectors.}
-We have collected a number of motivational examples from other termination papers that treated bit-vectors as mathematical integers \cite{DBLP:conf/tacas/LeikeH14,DBLP:conf/tacas/CookSZ13}. 
-We show that the termination arguments computed by such techniques do not directly apply when considering the bit-vector semantics.
-%these programs actually exhibit different terminating behaviours for bit-vectors and integer
+Every execution of the program in Figure~\ref{} can be partitioned into two phases: in the first phase $y$ increases
+until it is positive (in this phase $q$ may increase), whereas in the second $q$ decreases until the loop condition is violated. 
+\cite{DBLP:conf/tacas/LeikeH14} proposes using a linear ranking template in order to obtain a ranking function that proceeds
+through a fixed number of phases in the program execution. Each phase is ranked by a linear function, and ends when this function becomes non-positive.
+However, when considering the bit-vector semantics, this program does not terminate.
 
@@
 %Approaches based on Farkas’ Lemma can only handle non-strict inequalities \cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/vmcai/P04}.
+The program in Figure~\ref{} is taken from SVCOMP'15  \footnote{http://sv-comp.sosy-lab.org/2015/index.php} termination benchmarks.
+In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, this program admits a multi-phase ranking function, i.e. ..,  with a dedicated technique for computing it. However, in our setting 
+this type of programs do not need a special treatment, as we can find a non-linear ranking function for it as follows:
+\begin{verbatim}
+t2 = ITE(x, z, z)
+t3 = x < y
+\end{verbatim}
+
+%% t3 is "If x >= y, then x' >= y'" where x' is the next value of x.  This can be interpreted as "if the 'else' branch of the if is taken once, it is taken forever", which can be seen by induction.
+%% t2 is z -- this is probably the sickest part.  In the 'else' branch, z just decreases.  In the 'then' branch, we get:
+
+%% y > x
+%% y' = z
+%% x' = nondet()
+%% z' = x' - 1
+
+%% Let's consider 2 cases:
+
+%% case z' >= z, then:
+%%   x' - 1 >= z, so
+%%   x' - 1 >= y', so
+%%   x' > y',
+%%   so (x < y) has decreased, i.e. we will reamain in the 'else' branch forever and
+%% case z' < z, then:
+%%   z has decreased.
+
 \begin{lstlisting}
@@
 
+\subsection{The linearity assumption vs. the LKC one for ranking functions}
+
 \begin{conjecture}
@@
 \centering
-\begin{tabular}{|l|r|c|c|r|r|r|}
+\begin{tabular}{|l|r|c|c|c|c|r|r|r|}
 \hline
-    & LOC & \shortstack{Linear \\ program?} & \shortstack{Linear ranking \\ function?}  & Dimension & \shortstack{Ranking \\ program size} & Time (s)\\
+    & LOC & \shortstack{Linear \\ program?} & \shortstack{Linear ranking \\ function?}  & Conditional? & Float? & Dimension & \shortstack{Ranking \\ program size} & Time (s)\\
     \hline
     \hline
- P1 & 10 & \xmark & \xmark & 3 & 1 & 0.01 \\
- P2 & 10 & \xmark & \xmark & 3 & 1 & 0.01 \\
- P3 & 10 & \xmark & \xmark & 3 & 1 & 0.01 \\
+ P1 & 10 & \xmark & \xmark & \xmark & \xmark & 3 & 1 & 0.01 \\
+ P2 & 10 & \xmark & \xmark & \xmark & \xmark & 3 & 1 & 0.01 \\
+ P3 & 10 & \xmark & \xmark & \xmark & \xmark & 3 & 1 & 0.01 \\
  \hline

### 2014-06-16 papers/termination/paper.tex
Change multiphase ranking function
changed motivation

@@
 
+% The main advantage of our approach is its unitary nature. We do not require any initial assumption regarding the form of the termination  
+% argument, as this does not influence our search process. We identify the factors influencing the performance of our design and 
+% show that these factors are independent on the linearity of the ranking function or the number of lexicographic components...
+
 Other points to touch in the introduction:
@@
 
+The program in Figure~\ref{} is taken from \cite{DBLP:conf/tacas/CookSZ13}, where
+$m$ and $x$ start as any integers with $m$ positive. If $x$ is greater than
+$m$, $x$ is set to 0. Otherwise, $x$ increases until it equals $m$, upon which
+the loop terminates. While a disjunctive well-founded termination argument does exist for the loop, 
+e.g. ($x < oldx$ and $0 \leq oldx$) or ($m-x < oldm-oldx$ and $0 \leq oldm-oldx$), the loop does not 
+terminate under the bit-vector semantics. The reason is ..
+
 \begin{lstlisting}
@@
 
-Every execution of the program in Figure~\ref{} can be partitioned into two phases: in the first phase $y$ increases
+In Figure~\ref{}, every execution of the program can be partitioned into two phases: in the first phase $y$ increases
 until it is positive (in this phase $q$ may increase), whereas in the second $q$ decreases until the loop condition is violated. 
-\cite{DBLP:conf/tacas/LeikeH14} proposes using a linear ranking template in order to obtain a ranking function that proceeds
+This example was presented in \cite{DBLP:conf/tacas/LeikeH14}, where the authors make use of a template for obtaining a ranking function that proceeds
 through a fixed number of phases in the program execution. Each phase is ranked by a linear function, and ends when this function becomes non-positive.
-However, when considering the bit-vector semantics, this program does not terminate.
+However, when considering the bit-vector semantics, this program does not terminate as..
 
@@
 \end{verbatim}
-
-%% t3 is "If x >= y, then x' >= y'" where x' is the next value of x.  This can be interpreted as "if the 'else' branch of the if is taken once, it is taken forever", which can be seen by induction.
-%% t2 is z -- this is probably the sickest part.  In the 'else' branch, z just decreases.  In the 'then' branch, we get:
-
-%% y > x
-%% y' = z
-%% x' = nondet()
-%% z' = x' - 1
-
-%% Let's consider 2 cases:
-
-%% case z' >= z, then:
-%%   x' - 1 >= z, so
-%%   x' - 1 >= y', so
-%%   x' > y',
-%%   so (x < y) has decreased, i.e. we will reamain in the 'else' branch forever and
-%% case z' < z, then:
-%%   z has decreased.
 

### 2014-06-18 papers/termination/paper.tex
Start description of bitvector arithmetic

@@
 
-\noindent {\bf Bit-vectors (machine-level integers) vs. mathematical integers.} The abstraction of bit-vectors to mathematical integers 
+\noindent {\bf Bit-vectors (machine-level integers) vs. mathematical integers.}
+Physical computers have bounded storage, which means they are unable to perform calculations on mathematical
+integers.  For example, if $A$ is the Ackermann function and $G$ is Graham's number, a physical computer
+capable of computing $A(G, G)$ would contain (much!) more matter than is believed to exist in the universe.
+Fortunately, it is rare for a programmer to need such a large number and so modern computers do their arithmetic
+over fixed-width binary words, otherwise known as bit-vectors.
+
+Bit-vector arithmetic is very similar to modular arithmetic.  For a computer using $k$-bit words, unsigned
+calulcations are just defined to
+
+The abstraction of bit-vectors to mathematical integers 
 ignores the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in %, it gives rise to 
@@
 \begin{verbatim}
-t2 = ITE(x, z, z)
-t3 = x < y
+R(x, y, z) = (x < y, z)
 \end{verbatim}

### 2014-06-18 papers/termination/paper.tex
revised intro + related work

--- 2014-06-16 papers/termination/paper.tex
@@
 
-\section{Introduction}
+\section{Introduction}\label{sec:intro}
 
@@
 for the program states, i.e. a map from the program's state space to a well-ordered set.
-As this definition of a ranking function is very general, research is often limited to some
-convenient and tractable form of ranking functions, most frequently \emph{linear ranking functions} (see Section~\ref{sec:ranking.functions}). 
-In doing so, an analysis ensures that such a ranking function will be found if one  exists, 
-while failing to prove termination for terminating programs whose ranking functions fall short of the considered restriction.
+%% As this definition of a ranking function is very general, research is often limited to some
+%% convenient and tractable form of ranking functions, most frequently \emph{linear ranking functions} (see Section~\ref{sec:ranking.functions}). 
+%% In doing so, an analysis ensures that such a ranking function will be found if one  exists, 
+%% while failing to prove termination for terminating programs whose ranking functions fall short of the considered restriction. 
+In doing so, the majority of the existing termination provers diverge 
+in one important aspect from the execution of a program on a computer:
+bit-vectors and floats are treated as mathematical integers and reals, respectively.
+Figure~\ref{fig:handletable} presents a summary of existing termination analyses with indeed
+most of them designed for integers and assuming a \emph{restrictive transition relation}, e.g.
+linear programs \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10}.
+% \emph{linear ranking functions for linear programs over integers}.
+These design decisions may give rise to 
+{\em unsound} and {\em incomplete} analyses, as discussed next.
+%which is rather surprising given that bit-vectors and floats are ubiquitous in computer systems.
 
-Moreover, the majority of the existing termination provers diverge 
-in one important aspect from the execution of a program on a computer:
-bit-vectors and floats 
-are treated as mathematical integers and reals, respectively.
-This approximation gives rise to 
-{\em unsound} and {\em incomplete} analyses, which is rather surprising given that bit-vectors and floats are ubiquitous in computer systems.
+%\noindent {\bf Bit-vectors (machine-level integers) vs. mathematical integers.} 
+The abstraction of bit-vectors to mathematical integers 
+ignores the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in 
+incomparable behaviours with respect to termination. 
+%The loop in Figure~\ref{} terminates for integers, but does not terminate for bit-vectors. 
+The loop in Figure~\ref{} always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
+whereas, with bit-vector arithmetic, \texttt{x} under-flows before becoming negative and goes back to being positive causing the loop to never terminate.
+According to the C11 standard for the C programming language, a signed under- or over-flow yields undefined behaviour.
+However, compilers generally treat signed under- and over-flows using the
+wrap-around behaviour (see Section~\ref{sec:machine.arith}). %This means that a termination analysis for integers would return an unsound result for this situation.
+As a second example, the loop in Figure~\ref{} is  terminating for bit-vectors since \texttt{x}
+will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
+arithmetic since the loop condition stays always true for any initial \texttt{x} at least 1.
 
-Figure~\ref{fig:handletable} presents a summary of existing termination analyses with indeed
-most of the work concentrated on finding \emph{linear ranking functions for linear programs over integers}.
-Next, we will consider in more detail the assumptions made by such techniques and their implications.\\
+%Similarly, programs that terminate for bit-vectors may \emph{not} terminate for integers. One such situation is illustrated next:
 
-\noindent {\bf Bit-vectors (machine-level integers) vs. mathematical integers.} The abstraction of bit-vectors to mathematical integers 
-ignores the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in %, it gives rise to 
-%{\em unsound} and {\em incomplete} analyses, which is rather surprising given that bit-vectors are ubiquitous in computer systems. 
-incomparable behaviours with respect to termination:
-%The differences between the semantics of unbounded mathematical integers and that of machine integers result in 
-\begin{itemize}
-\item Programs that terminate for integers may \emph{not} terminate for bit-vectors. For illustration, consider the following loop:
 \begin{lstlisting}[language=C]
@@
 \end{lstlisting}
-The loop always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
-whereas, with bit-vector arithmetic, \texttt{x} under-flows before becoming negative and goes back to being positive causing the loop to never terminate.
-According to the C11 standard for the C programming language, a signed under- or over-flow yields undefined behaviour.
-However, compilers generally treat signed under- and over-flows using the
-wrap-around behaviour (see Section~\ref{sec:machine.arith}). 
-\item Similarly, programs that terminate for bit-vectors may \emph{not} terminate for integers. One such situation is illustrated next:  
+
 \begin{lstlisting}[language=C]
@@
  \end{lstlisting}
-This loop is  terminating for bit-vectors since \texttt{x}
-will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
-arithmetic since the loop condition stays always true for any initial \texttt{x} at least 1.
-\end{itemize}
 
-\noindent {\bf Floats vs. reals.} A scenario similar to the one for bit-vectors happens if floats were to be abstracted to unbounded reals by termination provers \cite{}. 
+%\noindent {\bf Floats vs. reals.} 
+A scenario similar to the one for bit-vectors happens if floats were to be abstracted to unbounded reals by termination provers \cite{}. 
 Consequently, these provers ignore potential rounding errors, under- and over-flows, which are precisely what makes  reasoning about floating point inherently difficult.
-This approximation may lead to erroneous diagnosis of a program's terminating behaviour as illustrated next:
-\begin{itemize}
-\item This loop does not terminate for reals, but does for floats.
+This approximation may lead to erroneous diagnosis of a program's terminating behaviour as illustrated by the loops in Figure~\ref{} and Figure~\ref{}.
+While the former does not terminate for reals, but does for floats, the latter 
+terminates for reals, but does not for floats.
+
 \begin{lstlisting}
@@
 
-\item The following loop terminates for reals, but does not for floats.
 \begin{lstlisting}
@@
 \end{lstlisting}
-\end{itemize}   
 \todo{explain the reasons for non-termination with figures.}\\
 
-\noindent {\bf Linear programs and linear ranking functions.} As visible in Figure~\ref{fig:handletable}, 
-most termination techniques assume \emph{restrictive transition relations}, e.g. linear programs,  and are only able to compute
-\emph{linear ranking functions} \cite{}. To better understand this restriction, we computed the probability of a random terminating program having a linear ranking
-function (see Section ?). This probability proved to be very low, indicating that the linearity assumption for the ranking functions is indeed prohibitive.
+%% \noindent {\bf Linear programs and linear ranking functions.} As visible in Figure~\ref{fig:handletable}, 
+%% most termination techniques assume \emph{restrictive transition relations}, e.g. linear programs,  and are only able to compute
+%% \emph{linear ranking functions} \cite{}. To better understand this restriction, we computed the probability of a random terminating program having a linear ranking
+%% function (see Section ?). This probability proved to be very low, indicating that the linearity assumption for the ranking functions is indeed prohibitive.
 
-We propose a general framework that does not assume the existence of linear ranking functions and can uniformly compute
+If we survey the area of program termination chronologically, we observe an initial domination of  monolithic approaches based on a single measure shown to decrease
+over all program paths %(syntactic characterisation of loops) 
+\cite{DBLP:conf/vmcai/P04,DBLP:conf/cav/BradleyMS05}, followed by 
+more recent techniques moving towards termination arguments based on Ramsey's theorem \cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}.
+The latter approach aims to find a set of local termination conditions that decrease as execution follows through the loop. %(semantic characterisation of loops).
+The main justification for this paradigm shift lies in the simplicity of the local termination measures when compared to the global ones, e.g.
+there are cases in which proofs based on local measures involve applying linear functions while corresponding global
+measures involve nonlinear functions or lexicographic orders.
+
+The downside of a Ramsey-based approach is the fact that a valid termination argument must hold for the \emph{transitive closure}
+of the program's transitions, rather than only for individual transitions. 
+As such, there is experimental evidence that most of the time is spent in reachability analysis \cite{DBLP:conf/pldi/CookPR06}, 
+requiring the support of powerful safety checkers.
+Basically, these approaches opt for simpler termination arguments in the detriment of complex validity checking.
+
+As Ramsey-based approaches are limited by the state of the art in safety checking, 
+recent research shifts back to more complex termination arguments that are easier to check \cite{DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/KroeningSTW10}.
+%In \cite{DBLP:conf/tacas/CookSZ13}, Cook et al present an iterative termination proving procedure that searches for 
+%lexicographic termination arguments, whereas Kroening et al. strengthen the termination argument such that it becomes a transitive relation \cite{DBLP:conf/cav/KroeningSTW10}.
+%
+%
+Following the same trend, %of switching the complexity in termination checking back to the termination arguments, 
+we investigate its very extreme: \emph{unrestricted} termination arguments. 
+This means that our ranking functions may involve non-linearity and lexicographic orders (we do not commit to any such form, i.e. we do not use templates).
+That is, we propose a general framework that does not assume the existence of linear ranking functions and can uniformly compute
 \emph{lexicographic/non-lexicographic} \emph{linear/non-linear} 
@@
 Our technique is {\emph complete} and can handle programs with non-linear operations, e.g. logical and.
-For this purpose, we rephrase the termination problem as a second-order satisfaction problem
-(indeed, a program synthesis problem). 
-As a direct consequence, our technique is semantic in nature and does not
-dependent on the size of the analysed program or its number of variables. 
-However, its efficiency does depend on the Kolmogorov
+In our design, the termination problem becomes as hard as finding ranking functions, rather than as hard as
+checking the validity of a termination argument.
+
+Given the level of generality that we aim for, we phrase the termination problem as a second order satisfaction problem (indeed, a program synthesis problem) (Section~\ref{}).
+While investigating the search space and the distribution of solutions (Section~\ref{}), we identify factors influencing our method's performance and
+find that our technique %is semantic in nature and 
+does not dependent on the size of the analysed program or its number of variables, but on the Kolmogorov
  complexity of the computed ranking functions. We show that the probability of
 a ranking function to have a low Kolmogorov complexity is higher than the probability to be linear (see Section~?).
+Moreover, we find that our technique does actually gain from the generality of the termination argument. 
+After investigating its theoretical bounds, we experimentally show that our technique performs well in practice.
 
@@
   \hline
-  \multirow{4}{*}{Ranking} & Linear lexicographic &  \cite{DBLP:conf/cav/BradleyMS05} && & &\checkmark&\checkmark&\checkmark&\checkmark\\
-   & Linear non-lexicographic & \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13} & & && \checkmark~ \cite{DBLP:conf/tacas/CookKRW10} &\checkmark~ \cite{DBLP:conf/tacas/CookKRW10}&\checkmark&\checkmark\\
+  \multirow{4}{*}{Ranking} & Linear lexicographic &  \cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13,DBLP:conf/vmcai/P04} && & &\checkmark&\checkmark&\checkmark&\checkmark\\
+   & Linear non-lexicographic & \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10} & \cite{DBLP:conf/vmcai/BradleyMS05} & && \checkmark~ \cite{DBLP:conf/tacas/CookKRW10} &\checkmark~ \cite{DBLP:conf/tacas/CookKRW10}&\checkmark&\checkmark\\
    & Non-linear lexicographic &  &  & &&\checkmark&\checkmark&\checkmark&\checkmark\\
-   & Non-linear non-lexicographic &  &  & &&\checkmark&\checkmark&\checkmark&\checkmark\\
+   & Non-linear non-lexicographic & \cite{DBLP:conf/vmcai/BradleyMS05} &  \cite{DBLP:conf/vmcai/BradleyMS05} & &&\checkmark&\checkmark&\checkmark&\checkmark\\
    \hline
@@
 bounds on the convergence time ...
-\item We present a formulation that allows us to prove termination without using a fancy safety prover (because
-we don't do a DWF proof).  In particular,
+\item We present a formulation that allows us to prove termination without an expensive reachability check (as we don't use Ramsey-based termination arguments).  In particular,
 we only need a bounded model checker that performs a single unwinding of a loop.
-\item Our technique is able to uniformly handle conditionally-terminating loops, as well as programs with
-multiple loops.
+%% \item Our technique is able to uniformly handle conditionally-terminating loops, as well as programs with
+%% multiple loops.
 \item We implemented our technique and tried it on a selection of programs handling both bit-vectors and floats.
@@
 
-Other points to touch in the introduction:
-\begin{itemize}
-\item disjunctive well-foundedness: By phrasing the termination problem as a single rank function synthesis problem, we observe that
-we can make the termination problem as hard as finding ranking functions, rather than as hard as
-solving a difficult ``double reachability'' problem (cf. Terminator).
-\item The lexicographic ordering does not influence the size of the search space.
-Thus, finding potentially lexicographic termination arguments is not more difficult than finding non-lexicographic ones.
-\item Compared to the previous approaches that mostly aim to find a particular form of ranking functions, our technique does not commit to any particular one.
-We will find the one that has the lowest Kolmogorov complexity.
-\item Many loops do not terminate for all starting states, but are contained in programs that guarantee the loop will
-terminate.  Traditional termination provers have difficulty reasoning about such conditionally-terminating loops.
-We are able to handle such loops by computing \emph{termination invariants}.  This mechanism also allows us to
-prove that programs with multiple loops terminate, even if the termination of some loop depends on the states
-reachable after leaving a previous loop.
-\item Linearity vs. LKC.
-\end{itemize}
+%semantic approach
 
-This paper is organised as follows:
+%given that the search space size is constant, the probability of finding a solution 
+%increases when considering all the possible solutions. There is no point in restricting the form of the solution..
+%to read: constraint-based.., integers vs rationals.
+
+%This paper is organised as follows:
 
 \section{Related Work}
-\subsection{Termination analysis}
-Given the multitude of works on termination, we have tried to summarise them in Figure~\ref{fig:handletable}.
+%\subsection{Termination analysis}
+Automated program termination is a research topic that has received a fair amount of attention from the software verification community.
+In order to compare our technique to the rest of the area, 
+Figure~\ref{fig:handletable} summarises the related works with respect to the restrictions they impose on the transition relations as well as the form of the computed ranking functions.
 
-The existing work on termination is primarily divided into structural (syntactic) vs. semantic approaches.  
-For example, terminationg of term rewriting systems is largely structural, whereas DWF analysis (cf. Terminator) is more semantic. While it deals
-with transition relations, it also uses some amount of syntax in that it identifies loops \& looping paths.
-
-A program is lasso-shaped if it is composed by a stem followed by a single loop without branching, i.e. there is only one path. 
-Consequently, the termination argument can be very simple. There are numerous techniques specialised in
-proving termination of lasso-shaped programs efficiently \cite{DBLP:conf/popl/Ben-AmramG13,DBLP:conf/cav/BradleyMS05,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/P04}.
-
+Somehow expected, most of the techniques are specialised in the synthesis of linear ranking functions for linear programs over integers (or rationals) \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13}.
+Among them, 
 Lee et al. make use of transition predicate abstraction, algorithmic learning, and decision procedures to compute transition
 invariants as proofs for the termination of linear programs \cite{DBLP:conf/cav/LeeWY12}.
+Leike and Heizmann present a new method for the constraint-based synthesis
+of termination arguments for linear loop programs based on
+linear ranking templates \cite{DBLP:conf/tacas/LeikeH14}.
+Linear ranking functions supported by inductive linear invariants for loops with linear guards and transitions: \cite{DBLP:conf/cav/BradleyMS05}. 
+Learning: \cite{DBLP:journals/corr/HeizmannHP14}
 
-There is very limited work on generating ranking functions for relations over machine integers.
+%A program is lasso-shaped if it is composed by a stem followed by a single loop without branching, i.e. there is only one path. 
+%Consequently, the termination argument can be very simple. There are numerous techniques specialised in
+%proving termination of lasso-shaped programs efficiently \cite{DBLP:conf/popl/Ben-AmramG13,DBLP:conf/cav/BradleyMS05,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/P04}.
+
+While the synthesis of termination arguments for linear programs over integers is indeed well-covered in the literature, 
+there is very limited work for programs over machine integers.
 Cook et al. present a method based on a reduction to Presburger
@@
 ranking functions based on reduction to SAT- and QBF-solving \cite{DBLP:conf/tacas/CookKRW10}.
+Similarly, the only work we are aware of that can compute non-linear ranking functions  
+for imperative loops with polynomial guards and polynomial assignments
+is \cite{DBLP:conf/vmcai/BradleyMS05}.
 
-Leike and Heizmann present a new method for the constraint-based synthesis
-of termination arguments for linear loop programs based on
-linear ranking templates \cite{DBLP:conf/tacas/LeikeH14}.
+Given the lack of research in handling \emph{not-linear programs}, as well as \emph{programs over bit-vectors and floats},  
+our work focuses on covering these areas. 
+One of the obvious conclusions that can be reached from observing Figure~\ref{fig:handletable}, 
+is that most of the works tend to specialise on a certain aspect of termination proving that they can solve efficiently. 
+Conversely to this view, we aim for \emph{generality}, as we do not restrict the form of the synthesised ranking functions, nor the form of the input programs.
 
-Learning: \cite{DBLP:journals/corr/HeizmannHP14}
+As mentioned in Section~\ref{sec:intro}, approaches based on Ramsey's theorem compute a set of local termination conditions that decrease as execution follows through the loop
+and require expensive reachability analyses
+\cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}.
+In an attempt to reduce the complexity of checking the validity of the termination argument, 
+Cook et al present an iterative termination proving procedure that searches for 
+lexicographic termination arguments \cite{DBLP:conf/tacas/CookSZ13}, 
+whereas Kroening et al. strengthen the termination argument such that it becomes a transitive relation \cite{DBLP:conf/cav/KroeningSTW10}.
 
-Polynomial programs: \cite{DBLP:conf/vmcai/BradleyMS05}
+\todo{add smth about synthesis}
 
-Linear ranking functions supported by inductive linear invariants for loops with linear guards and transitions: \cite{DBLP:conf/cav/BradleyMS05}.
-
-\subsection{Disjunctive well-foundedness}
-Cook et al. are closer to a semantic approach by viewing the termination proving task as a binary reachability analysis \cite{DBLP:conf/pldi/CookPR06}.
-%Terminator incrementally constructs the union of ranking relations.
-%Given that the this union is not well-founded (it is only disjunctively well-founded), 
-%Terminator must prove that the binary reachability relation of the program (the transitive closure of the transition relation restricted to reachable states) 
-%is contained in the union of ranking relations.
-%Thus, this approach requires an evolved binary reachability, and, implicitly, the assistance of a safety prover.
-In particular, their technique incrementally constructs the union of ranking relations.
-Given that the this union is not well-founded (it is only disjunctively well-founded),
-they must show that, for every sequence, one of the ranking functions decreases
-between the pre- and post-state. %In other words: it 
-%must find all pairs of states reachable from the program's initial state 
-%such that one is reachable from the other and it must show 
-%that the value of one of the ranking functions decreases from s1 to s2. 
-Consequently, most of the analysis time is spent in reachability analysis that is to check if the current
-transition invariant reached a fixed point.
-
-As we want to avoid the requirements of a binary reachability analysis, our termination argument is not a set
-of ranking functions, but a single ranking function, which we prove is enough for proving the termination of any termination program. 
-With a single ranking function we only need to show that the rank decreases from the
-pre- to post-state after executing each single transition step.
-The checking process is therefore easy. For the construction of the ranking function, we 
-formulate the problem as a second order satisfaction formula and we solve it via program synthesis.
-
-Because most state-of-the-art termination procedures use disjunctive well foundedness arguments, they require the use of
-complex safety provers that are able to reason about unbounded unfoldings of a transition relation.  In contrast to this,
-we are able to build monolithic termination arguments that only need ``local'' reasoning -- our ranking functions
-can be proved sufficient by considering just a single unfolding of a loop.
+%The existing work on termination is primarily divided into structural (syntactic) vs. semantic approaches.  
+%For example, terminationg of term rewriting systems is largely structural, whereas DWF analysis (cf. Terminator) is more semantic. While it deals
+%with transition relations, it also uses some amount of syntax in that it identifies loops \& looping paths.
 
@@
 
-%-- the union T of ranking relations, however, is in general not well-founded
-%(it is only disjunctively well-founded in the terminology of [23]). This is
-%why it would not be sufficient to show the inclusion R \subseteq  T, and why we must in stead prove R_I^+ \subseteq T.
-
-\subsection{Synthesis}
-
-%% ---
-%% constructs and checks the termination argument
-
-%% T ERMINATOR’s most distinguishing aspect, with respect to pre-
-%% vious methods and tools for proving program termination, is how it
-%% shifts the balance between the two tasks of constructing and respec-
-%% tively checking the termination argument. The classical method is
-%% to construct an expression defining the rank of a state and then to
-%% check that its value decreases in every transition from a reachable
-%% state to a next one. The construction of the ranking function is the
-%% hard part and forms a task that needs to be applied to the whole
-%% program. The checking part is relatively easy. In our method, the
-%% task of constructing ranking functions is the relatively easy part;
-%% they are constructed on demand based on the examination of only
-%% a few selected paths through the program.
-
-%% Previously, it was not known whether binary reachability analysis could be made
-%% practical. The challenge raised by our approach was to show that
-%% this is indeed the case.
-
-%% ----------
-%% Efficient synthesis of ranking functions for machine-level bit-vectors, how-
-%% ever, has remained an open problem.
-%% ---------
+%\subsection{Synthesis}
 
@@
 
-\subsection{Disjunctive Well-Foundedness}
+\subsection{Disjunctive Well-Foundedness}\label{sec:dwf}
 Terminator uses disjunctive well foundedness arguments, which can be found
@@
 \end{definition}
+
+Many loops do not terminate for all starting states, but are contained in programs that guarantee the loop will
+terminate.  Traditional termination provers have difficulty reasoning about such conditionally-terminating loops.
+We are able to handle such loops by computing \emph{termination invariants}.  This mechanism also allows us to
+prove that programs with multiple loops terminate, even if the termination of some loop depends on the states
+reachable after leaving a previous loop.
 

### 2014-06-18 papers/termination/paper.tex

@@
 %\noindent {\bf Bit-vectors (machine-level integers) vs. mathematical integers.}
+
+Physical computers have bounded storage, which means they are unable to perform calculations on mathematical
+integers.  For example, if $A$ is the Ackermann function and $G$ is Graham's number, a physical computer
+capable of computing $A(G, G)$ would contain (much!) more matter than is believed to exist in the universe.
+Fortunately, it is rare for a programmer to need such a large number and so modern computers do their arithmetic
+over fixed-width binary words, otherwise known as bit-vectors.
+
+Bit-vector arithmetic is very similar to modular arithmetic.  For a computer using $k$-bit words, unsigned
+calulcations are just defined to 
+===============
 The abstraction of bit-vectors to mathematical integers 
@@
 The loop in Figure~\ref{} always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
+%% =======
+%% most of the work concentrated on finding \emph{linear ranking functions for linear programs over integers}.
+%% Next, we will consider in more detail the assumptions made by such techniques and their implications.\\
+
+%% \noindent {\bf Bit-vectors (machine-level integers) vs. mathematical integers.}
+%% Physical computers have bounded storage, which means they are unable to perform calculations on mathematical
+%% integers.  For example, if $A$ is the Ackermann function and $G$ is Graham's number, a physical computer
+%% capable of computing $A(G, G)$ would contain (much!) more matter than is believed to exist in the universe.
+%% Fortunately, it is rare for a programmer to need such a large number and so modern computers do their arithmetic
+%% over fixed-width binary words, otherwise known as bit-vectors.
+
+%% Bit-vector arithmetic is very similar to modular arithmetic.  For a computer using $k$-bit words, unsigned
+%% calulcations are just defined to
+
+%% The abstraction of bit-vectors to mathematical integers 
+%% ignores the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in %, it gives rise to 
+%% %{\em unsound} and {\em incomplete} analyses, which is rather surprising given that bit-vectors are ubiquitous in computer systems. 
+%% incomparable behaviours with respect to termination:
+%% %The differences between the semantics of unbounded mathematical integers and that of machine integers result in 
+%% \begin{itemize}
+%% \item Programs that terminate for integers may \emph{not} terminate for bit-vectors. For illustration, consider the following loop:
+%% \begin{lstlisting}[language=C]
+%% while (x > 0) x -= 2;
+%% \end{lstlisting}
+%% The loop always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
+%% >>>>>>> 7b690c08a70e4161a432600e9456d922106badaf
 whereas, with bit-vector arithmetic, \texttt{x} under-flows before becoming negative and goes back to being positive causing the loop to never terminate.

### 2014-06-18 papers/termination/paper.tex
minor revision

@@
 
-Bit-vector arithmetic is very similar to modular arithmetic.  For a computer using $k$-bit words, unsigned
-calulcations are just defined to 
-===============
+%Bit-vector arithmetic is very similar to modular arithmetic.  For a computer using $k$-bit words, unsigned
+%calulcations are just defined to 
 The abstraction of bit-vectors to mathematical integers 
@@
 The loop in Figure~\ref{} always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
+whereas, with bit-vector arithmetic, \texttt{x} under-flows before becoming negative and goes back to being positive causing the loop to never terminate.
 %% =======
@@
 %% >>>>>>> 7b690c08a70e4161a432600e9456d922106badaf
-whereas, with bit-vector arithmetic, \texttt{x} under-flows before becoming negative and goes back to being positive causing the loop to never terminate.
 According to the C11 standard for the C programming language, a signed under- or over-flow yields undefined behaviour.

### 2014-06-18 papers/termination/paper.tex
Some more bitvector stuff

--- 2014-06-18 papers/termination/paper.tex
@@
 
-\section{Introduction}\label{sec:intro}
+\section{Introduction}
 
@@
 for the program states, i.e. a map from the program's state space to a well-ordered set.
-%% As this definition of a ranking function is very general, research is often limited to some
-%% convenient and tractable form of ranking functions, most frequently \emph{linear ranking functions} (see Section~\ref{sec:ranking.functions}). 
-%% In doing so, an analysis ensures that such a ranking function will be found if one  exists, 
-%% while failing to prove termination for terminating programs whose ranking functions fall short of the considered restriction. 
-In doing so, the majority of the existing termination provers diverge 
+As this definition of a ranking function is very general, research is often limited to some
+convenient and tractable form of ranking functions, most frequently \emph{linear ranking functions} (see Section~\ref{sec:ranking.functions}). 
+In doing so, an analysis ensures that such a ranking function will be found if one  exists, 
+while failing to prove termination for terminating programs whose ranking functions fall short of the considered restriction.
+
+Moreover, the majority of the existing termination provers diverge 
 in one important aspect from the execution of a program on a computer:
-bit-vectors and floats are treated as mathematical integers and reals, respectively.
+bit-vectors and floats 
+are treated as mathematical integers and reals, respectively.
+This approximation gives rise to 
+{\em unsound} and {\em incomplete} analyses, which is rather surprising given that bit-vectors and floats are ubiquitous in computer systems.
+
 Figure~\ref{fig:handletable} presents a summary of existing termination analyses with indeed
-most of them designed for integers and assuming a \emph{restrictive transition relation}, e.g.
-linear programs \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10}.
-% \emph{linear ranking functions for linear programs over integers}.
-These design decisions may give rise to 
-{\em unsound} and {\em incomplete} analyses, as discussed next.
-%which is rather surprising given that bit-vectors and floats are ubiquitous in computer systems.
+most of the work concentrated on finding \emph{linear ranking functions for linear programs over integers}.
+Next, we will consider in more detail the assumptions made by such techniques and their implications.\\
 
-%\noindent {\bf Bit-vectors (machine-level integers) vs. mathematical integers.} 
-The abstraction of bit-vectors to mathematical integers 
-ignores the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in 
+\noindent {\bf Bit-vectors (machine-level integers) vs. mathematical integers.}
+The abstraction of bit-vectors to mathematical integers ignores
+the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in
 incomparable behaviours with respect to termination.
-%The loop in Figure~\ref{} terminates for integers, but does not terminate for bit-vectors. 
-The loop in Figure~\ref{} always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
-whereas, with bit-vector arithmetic, \texttt{x} under-flows before becoming negative and goes back to being positive causing the loop to never terminate.
-According to the C11 standard for the C programming language, a signed under- or over-flow yields undefined behaviour.
-However, compilers generally treat signed under- and over-flows using the
-wrap-around behaviour (see Section~\ref{sec:machine.arith}). %This means that a termination analysis for integers would return an unsound result for this situation.
-As a second example, the loop in Figure~\ref{} is  terminating for bit-vectors since \texttt{x}
+
+Programs that terminate for integers may \emph{not} terminate for bit-vectors. For illustration, consider the following loop:
+\begin{lstlisting}[language=C]
+while (x > 0) x -= 2;
+\end{lstlisting}
+The loop always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
+whereas, with bit-vector arithmetic, \texttt{x} underflows before becoming negative and goes back to being positive causing the loop to never terminate.
+
+Similarly, programs that terminate for bit-vectors may \emph{not} terminate for integers. One such situation is illustrated next:  
+ \begin{lstlisting}[language=C]
+ while(x > 0) x++;
+ \end{lstlisting}
+This loop is  terminating for bit-vectors since \texttt{x}
 will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
@@
 
-%Similarly, programs that terminate for bit-vectors may \emph{not} terminate for integers. One such situation is illustrated next:
-
-\begin{lstlisting}[language=C]
-while (x > 0) x -= 2;
-\end{lstlisting}
-
-\begin{lstlisting}[language=C]
- while(x > 0) x++;
- \end{lstlisting}
-
-%\noindent {\bf Floats vs. reals.} 
-A scenario similar to the one for bit-vectors happens if floats were to be abstracted to unbounded reals by termination provers \cite{}. 
+\noindent {\bf Floats vs. reals.} A scenario similar to the one for bit-vectors happens if floats were to be abstracted to unbounded reals by termination provers \cite{}. 
 Consequently, these provers ignore potential rounding errors, under- and over-flows, which are precisely what makes  reasoning about floating point inherently difficult.
-This approximation may lead to erroneous diagnosis of a program's terminating behaviour as illustrated by the loops in Figure~\ref{} and Figure~\ref{}.
-While the former does not terminate for reals, but does for floats, the latter 
-terminates for reals, but does not for floats.
-
+This approximation may lead to erroneous diagnosis of a program's terminating behaviour as illustrated next:
+\begin{itemize}
+\item This loop does not terminate for reals, but does for floats.
 \begin{lstlisting}
@@
 
+\item The following loop terminates for reals, but does not for floats.
 \begin{lstlisting}
@@
 \end{lstlisting}
+\end{itemize}   
 \todo{explain the reasons for non-termination with figures.}\\
 
-%% \noindent {\bf Linear programs and linear ranking functions.} As visible in Figure~\ref{fig:handletable}, 
-%% most termination techniques assume \emph{restrictive transition relations}, e.g. linear programs,  and are only able to compute
-%% \emph{linear ranking functions} \cite{}. To better understand this restriction, we computed the probability of a random terminating program having a linear ranking
-%% function (see Section ?). This probability proved to be very low, indicating that the linearity assumption for the ranking functions is indeed prohibitive.
+\noindent {\bf Linear programs and linear ranking functions.} As visible in Figure~\ref{fig:handletable}, 
+most termination techniques assume \emph{restrictive transition relations}, e.g. linear programs,  and are only able to compute
+\emph{linear ranking functions} \cite{}. To better understand this restriction, we computed the probability of a random terminating program having a linear ranking
+function (see Section ?). This probability proved to be very low, indicating that the linearity assumption for the ranking functions is indeed prohibitive.
 
-If we survey the area of program termination chronologically, we observe an initial domination of  monolithic approaches based on a single measure shown to decrease
-over all program paths %(syntactic characterisation of loops) 
-\cite{DBLP:conf/vmcai/P04,DBLP:conf/cav/BradleyMS05}, followed by 
-more recent techniques moving towards termination arguments based on Ramsey's theorem \cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}.
-The latter approach aims to find a set of local termination conditions that decrease as execution follows through the loop. %(semantic characterisation of loops).
-The main justification for this paradigm shift lies in the simplicity of the local termination measures when compared to the global ones, e.g.
-there are cases in which proofs based on local measures involve applying linear functions while corresponding global
-measures involve nonlinear functions or lexicographic orders.
-
-The downside of a Ramsey-based approach is the fact that a valid termination argument must hold for the \emph{transitive closure}
-of the program's transitions, rather than only for individual transitions. 
-As such, there is experimental evidence that most of the time is spent in reachability analysis \cite{DBLP:conf/pldi/CookPR06}, 
-requiring the support of powerful safety checkers.
-Basically, these approaches opt for simpler termination arguments in the detriment of complex validity checking.
-
-As Ramsey-based approaches are limited by the state of the art in safety checking, 
-recent research shifts back to more complex termination arguments that are easier to check \cite{DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/KroeningSTW10}.
-%In \cite{DBLP:conf/tacas/CookSZ13}, Cook et al present an iterative termination proving procedure that searches for 
-%lexicographic termination arguments, whereas Kroening et al. strengthen the termination argument such that it becomes a transitive relation \cite{DBLP:conf/cav/KroeningSTW10}.
-%
-%
-Following the same trend, %of switching the complexity in termination checking back to the termination arguments, 
-we investigate its very extreme: \emph{unrestricted} termination arguments. 
-This means that our ranking functions may involve non-linearity and lexicographic orders (we do not commit to any such form, i.e. we do not use templates).
-That is, we propose a general framework that does not assume the existence of linear ranking functions and can uniformly compute
+We propose a general framework that does not assume the existence of linear ranking functions and can uniformly compute
 \emph{lexicographic/non-lexicographic} \emph{linear/non-linear} 
@@
 Our technique is {\emph complete} and can handle programs with non-linear operations, e.g. logical and.
-In our design, the termination problem becomes as hard as finding ranking functions, rather than as hard as
-checking the validity of a termination argument.
-
-Given the level of generality that we aim for, we phrase the termination problem as a second order satisfaction problem (indeed, a program synthesis problem) (Section~\ref{}).
-While investigating the search space and the distribution of solutions (Section~\ref{}), we identify factors influencing our method's performance and
-find that our technique %is semantic in nature and 
-does not dependent on the size of the analysed program or its number of variables, but on the Kolmogorov
+For this purpose, we rephrase the termination problem as a second-order satisfaction problem
+(indeed, a program synthesis problem). 
+As a direct consequence, our technique is semantic in nature and does not
+dependent on the size of the analysed program or its number of variables. 
+However, its efficiency does depend on the Kolmogorov
  complexity of the computed ranking functions. We show that the probability of
 a ranking function to have a low Kolmogorov complexity is higher than the probability to be linear (see Section~?).
-Moreover, we find that our technique does actually gain from the generality of the termination argument. 
-After investigating its theoretical bounds, we experimentally show that our technique performs well in practice.
 
@@
   \hline
-  \multirow{4}{*}{Ranking} & Linear lexicographic &  \cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13,DBLP:conf/vmcai/P04} && & &\checkmark&\checkmark&\checkmark&\checkmark\\
-   & Linear non-lexicographic & \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10} & \cite{DBLP:conf/vmcai/BradleyMS05} & && \checkmark~ \cite{DBLP:conf/tacas/CookKRW10} &\checkmark~ \cite{DBLP:conf/tacas/CookKRW10}&\checkmark&\checkmark\\
+  \multirow{4}{*}{Ranking} & Linear lexicographic &  \cite{DBLP:conf/cav/BradleyMS05} && & &\checkmark&\checkmark&\checkmark&\checkmark\\
+   & Linear non-lexicographic & \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13} & & && \checkmark~ \cite{DBLP:conf/tacas/CookKRW10} &\checkmark~ \cite{DBLP:conf/tacas/CookKRW10}&\checkmark&\checkmark\\
    & Non-linear lexicographic &  &  & &&\checkmark&\checkmark&\checkmark&\checkmark\\
-   & Non-linear non-lexicographic & \cite{DBLP:conf/vmcai/BradleyMS05} &  \cite{DBLP:conf/vmcai/BradleyMS05} & &&\checkmark&\checkmark&\checkmark&\checkmark\\
+   & Non-linear non-lexicographic &  &  & &&\checkmark&\checkmark&\checkmark&\checkmark\\
    \hline
@@
 bounds on the convergence time ...
-\item We present a formulation that allows us to prove termination without an expensive reachability check (as we don't use Ramsey-based termination arguments).  In particular,
+\item We present a formulation that allows us to prove termination without using a fancy safety prover (because
+we don't do a DWF proof).  In particular,
 we only need a bounded model checker that performs a single unwinding of a loop.
-%% \item Our technique is able to uniformly handle conditionally-terminating loops, as well as programs with
-%% multiple loops.
+\item Our technique is able to uniformly handle conditionally-terminating loops, as well as programs with
+multiple loops.
 \item We implemented our technique and tried it on a selection of programs handling both bit-vectors and floats.
@@
 
-%semantic approach
+Other points to touch in the introduction:
+\begin{itemize}
+\item disjunctive well-foundedness: By phrasing the termination problem as a single rank function synthesis problem, we observe that
+we can make the termination problem as hard as finding ranking functions, rather than as hard as
+solving a difficult ``double reachability'' problem (cf. Terminator).
+\item The lexicographic ordering does not influence the size of the search space.
+Thus, finding potentially lexicographic termination arguments is not more difficult than finding non-lexicographic ones.
+\item Compared to the previous approaches that mostly aim to find a particular form of ranking functions, our technique does not commit to any particular one.
+We will find the one that has the lowest Kolmogorov complexity.
+\item Many loops do not terminate for all starting states, but are contained in programs that guarantee the loop will
+terminate.  Traditional termination provers have difficulty reasoning about such conditionally-terminating loops.
+We are able to handle such loops by computing \emph{termination invariants}.  This mechanism also allows us to
+prove that programs with multiple loops terminate, even if the termination of some loop depends on the states
+reachable after leaving a previous loop.
+\item Linearity vs. LKC.
+\end{itemize}
 
-%given that the search space size is constant, the probability of finding a solution 
-%increases when considering all the possible solutions. There is no point in restricting the form of the solution..
-%to read: constraint-based.., integers vs rationals.
-
-%This paper is organised as follows:
+This paper is organised as follows:
 
 \section{Related Work}
-%\subsection{Termination analysis}
-Automated program termination is a research topic that has received a fair amount of attention from the software verification community.
-In order to compare our technique to the rest of the area, 
-Figure~\ref{fig:handletable} summarises the related works with respect to the restrictions they impose on the transition relations as well as the form of the computed ranking functions.
+\subsection{Termination analysis}
+Given the multitude of works on termination, we have tried to summarise them in Figure~\ref{fig:handletable}.
 
-Somehow expected, most of the techniques are specialised in the synthesis of linear ranking functions for linear programs over integers (or rationals) \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13}.
-Among them, 
+The existing work on termination is primarily divided into structural (syntactic) vs. semantic approaches.  
+For example, terminationg of term rewriting systems is largely structural, whereas DWF analysis (cf. Terminator) is more semantic. While it deals
+with transition relations, it also uses some amount of syntax in that it identifies loops \& looping paths.
+
+A program is lasso-shaped if it is composed by a stem followed by a single loop without branching, i.e. there is only one path. 
+Consequently, the termination argument can be very simple. There are numerous techniques specialised in
+proving termination of lasso-shaped programs efficiently \cite{DBLP:conf/popl/Ben-AmramG13,DBLP:conf/cav/BradleyMS05,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/P04}.
+
 Lee et al. make use of transition predicate abstraction, algorithmic learning, and decision procedures to compute transition
 invariants as proofs for the termination of linear programs \cite{DBLP:conf/cav/LeeWY12}.
+
+There is very limited work on generating ranking functions for relations over machine integers.
+Cook et al. present a method based on a reduction to Presburger
+arithmetic, and a template-matching approach for predefined classes of
+ranking functions based on reduction to SAT- and QBF-solving \cite{DBLP:conf/tacas/CookKRW10}.
+
 Leike and Heizmann present a new method for the constraint-based synthesis
@@
 linear ranking templates \cite{DBLP:conf/tacas/LeikeH14}.
-Linear ranking functions supported by inductive linear invariants for loops with linear guards and transitions: \cite{DBLP:conf/cav/BradleyMS05}. 
+
 Learning: \cite{DBLP:journals/corr/HeizmannHP14}
 
-%A program is lasso-shaped if it is composed by a stem followed by a single loop without branching, i.e. there is only one path. 
-%Consequently, the termination argument can be very simple. There are numerous techniques specialised in
-%proving termination of lasso-shaped programs efficiently \cite{DBLP:conf/popl/Ben-AmramG13,DBLP:conf/cav/BradleyMS05,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/P04}.
+Polynomial programs: \cite{DBLP:conf/vmcai/BradleyMS05}
 
-While the synthesis of termination arguments for linear programs over integers is indeed well-covered in the literature, 
-there is very limited work for programs over machine integers.
-Cook et al. present a method based on a reduction to Presburger
-arithmetic, and a template-matching approach for predefined classes of
-ranking functions based on reduction to SAT- and QBF-solving \cite{DBLP:conf/tacas/CookKRW10}.
-Similarly, the only work we are aware of that can compute non-linear ranking functions  
-for imperative loops with polynomial guards and polynomial assignments
-is \cite{DBLP:conf/vmcai/BradleyMS05}.
+Linear ranking functions supported by inductive linear invariants for loops with linear guards and transitions: \cite{DBLP:conf/cav/BradleyMS05}.
 
-Given the lack of research in handling \emph{not-linear programs}, as well as \emph{programs over bit-vectors and floats},  
-our work focuses on covering these areas. 
-One of the obvious conclusions that can be reached from observing Figure~\ref{fig:handletable}, 
-is that most of the works tend to specialise on a certain aspect of termination proving that they can solve efficiently. 
-Conversely to this view, we aim for \emph{generality}, as we do not restrict the form of the synthesised ranking functions, nor the form of the input programs.
+\subsection{Disjunctive well-foundedness}
+Cook et al. are closer to a semantic approach by viewing the termination proving task as a binary reachability analysis \cite{DBLP:conf/pldi/CookPR06}.
+%Terminator incrementally constructs the union of ranking relations.
+%Given that the this union is not well-founded (it is only disjunctively well-founded), 
+%Terminator must prove that the binary reachability relation of the program (the transitive closure of the transition relation restricted to reachable states) 
+%is contained in the union of ranking relations.
+%Thus, this approach requires an evolved binary reachability, and, implicitly, the assistance of a safety prover.
+In particular, their technique incrementally constructs the union of ranking relations.
+Given that the this union is not well-founded (it is only disjunctively well-founded),
+they must show that, for every sequence, one of the ranking functions decreases
+between the pre- and post-state. %In other words: it 
+%must find all pairs of states reachable from the program's initial state 
+%such that one is reachable from the other and it must show 
+%that the value of one of the ranking functions decreases from s1 to s2. 
+Consequently, most of the analysis time is spent in reachability analysis that is to check if the current
+transition invariant reached a fixed point.
 
-As mentioned in Section~\ref{sec:intro}, approaches based on Ramsey's theorem compute a set of local termination conditions that decrease as execution follows through the loop
-and require expensive reachability analyses
-\cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}.
-In an attempt to reduce the complexity of checking the validity of the termination argument, 
-Cook et al present an iterative termination proving procedure that searches for 
-lexicographic termination arguments \cite{DBLP:conf/tacas/CookSZ13}, 
-whereas Kroening et al. strengthen the termination argument such that it becomes a transitive relation \cite{DBLP:conf/cav/KroeningSTW10}.
+As we want to avoid the requirements of a binary reachability analysis, our termination argument is not a set
+of ranking functions, but a single ranking function, which we prove is enough for proving the termination of any termination program. 
+With a single ranking function we only need to show that the rank decreases from the
+pre- to post-state after executing each single transition step.
+The checking process is therefore easy. For the construction of the ranking function, we 
+formulate the problem as a second order satisfaction formula and we solve it via program synthesis.
 
-\todo{add smth about synthesis}
-
-%The existing work on termination is primarily divided into structural (syntactic) vs. semantic approaches.  
-%For example, terminationg of term rewriting systems is largely structural, whereas DWF analysis (cf. Terminator) is more semantic. While it deals
-%with transition relations, it also uses some amount of syntax in that it identifies loops \& looping paths.
+Because most state-of-the-art termination procedures use disjunctive well foundedness arguments, they require the use of
+complex safety provers that are able to reason about unbounded unfoldings of a transition relation.  In contrast to this,
+we are able to build monolithic termination arguments that only need ``local'' reasoning -- our ranking functions
+can be proved sufficient by considering just a single unfolding of a loop.
 
@@
 
-%\subsection{Synthesis}
+%-- the union T of ranking relations, however, is in general not well-founded
+%(it is only disjunctively well-founded in the terminology of [23]). This is
+%why it would not be sufficient to show the inclusion R \subseteq  T, and why we must in stead prove R_I^+ \subseteq T.
+
+\subsection{Synthesis}
+
+%% ---
+%% constructs and checks the termination argument
+
+%% T ERMINATOR’s most distinguishing aspect, with respect to pre-
+%% vious methods and tools for proving program termination, is how it
+%% shifts the balance between the two tasks of constructing and respec-
+%% tively checking the termination argument. The classical method is
+%% to construct an expression defining the rank of a state and then to
+%% check that its value decreases in every transition from a reachable
+%% state to a next one. The construction of the ranking function is the
+%% hard part and forms a task that needs to be applied to the whole
+%% program. The checking part is relatively easy. In our method, the
+%% task of constructing ranking functions is the relatively easy part;
+%% they are constructed on demand based on the examination of only
+%% a few selected paths through the program.
+
+%% Previously, it was not known whether binary reachability analysis could be made
+%% practical. The challenge raised by our approach was to show that
+%% this is indeed the case.
+
+%% ----------
+%% Efficient synthesis of ranking functions for machine-level bit-vectors, how-
+%% ever, has remained an open problem.
+%% ---------
 
@@
 
-\subsection{Disjunctive Well-Foundedness}\label{sec:dwf}
+\subsection{Disjunctive Well-Foundedness}
 Terminator uses disjunctive well foundedness arguments, which can be found
@@
 \subsection{Machine Arithmetic and Bitvectors} \label{sec:machine.arith} 
-Computers tend to have fixed width integers.  For a machine with $k$-bit words,
-all arithmetic is done modulo $2^k$.  This means that a program's termination
-behaviour can depend strongly on whether it is interpreted as operating over
-mathematical integers, or over fixed-width machine words.  We can however
-observe that if no arithmetic operation \emph{overflows} (i.e. does not result
-in a value greater than $2^k$), fixed-width and integer arithmetic coincide.
+Physical computers have bounded storage, which means they are unable to perform calculations on mathematical
+integers.  For example, if $A$ is the Ackermann function and $G$ is Graham's number, a physical computer
+capable of computing $A(G, G)$ would contain (much!) more matter than is believed to exist in the universe.
+Fortunately, it is rare for a programmer to need such a large number and so modern computers do their arithmetic
+over fixed-width binary words, otherwise known as bit-vectors.  For the remainder of this section, we will say that
+the bit-vectors we are working with are $k$-bits wide, which means that each word can hold one of $2^k$ bit patterns.
+Typical values for $k$ are 32 and 64.
+
+Bit-vector arithmetic is performed modulo $2^k$, which is the source of many of the differences between
+machine arithmetic and Peano arithmetic.  To give an example: $(2^k - 1) + 1 \equiv 0 \pmod {2^k}$.
+This provides a counterexample to the statement $\forall x . x + 1 > x$, which is a theorem of Peano
+arithmetic but not of modular arithmetic.  When an arithmetic operation has a result greater than $2^k$,
+it is said to ``overflow''.  If an operation does not overflow, its machine-arithmetic result is the same
+as the result of the same operation performed on integers.
+
+Machine words can be interpreted as ``signed'' or ``unsigned'' values.  Signed values can be negative,
+while unsigned values cannot.  The encoding for signed values is two's complement, where the most significant
+bit $b_{k-1}$ of the word is a ``sign'' bit, whose weight is $-(2^k - 1)$ rather than $2^k - 1$.  Two's complement
+representation has the property that $\forall x . -x = (\mathord{\sim} x) + 1$, where $\mathord{\sim}(\bullet)$
+is bitwise-negation.  Two's complement also has the property that addition, multiplication and subtraction are defined
+identically for unsigned and signed numbers.  Therefore signed and unsigned arithmetic differ only on comparison
+operators\footnote{There is also the difference that signed overflow in a C program results in undefined behaviour,
+but in practice the undefined behaviour is implemented just as if the arithmetic had been unsigned}.
+For example: $$\forall x . -x \leq_s x \wedge -x \geq_u x$$ where $\leq_s$ and $\geq_u$ represent signed and unsigned comparisons respectively.
+
+The final source of difference between integer arithmetic and bitvector arithmetic stems from width conversions.
+Typical computers allow variables to have different types, which can be represented using words of different widths.
+In C a \texttt{short} might occupy
+16 bits, while an \texttt{int} might occupy 32 bits.  If a $k$-bit variable is assigned to a $l$-bit variable
+with $l < k$, the result is truncated $\mod 2^l$.  For example, if $x$ is a 32-bit variable and $y$ is a 16-bit
+variable, $y$ will hold the value $0$ after the following code is executed:
+
+\begin{lstlisting}
+x = 65536;
+y = x;
+\end{lstlisting}
+
+This gives us a counterexample to the statement $\forall x, y . x = y \Rightarrow (x + 1) = (y + 1)$,
+which is a theorem of Peano arithmetic.
+
+As well as machine arithmetic differing from Peano arithmetic on the operators they have in common,
+computers have several ``bitwise'' operations that are not taken as primitive in the theory of
+integers.  These operations include the standard boolean operations \texttt{and, or, not, xor}
+applied to each element of the bit-vector.  Computer programs often make use of these operators,
+which are non-linear when interpreted in the standard model of Peano arithmetic
+(although some of these operators can be seen as linear in a different algebraic structure,
+e.g. \texttt{xor} corresponds to addition in the Galois field $\mathrm{GF}(2^k)$).
 
@@
 \end{definition}
-
-Many loops do not terminate for all starting states, but are contained in programs that guarantee the loop will
-terminate.  Traditional termination provers have difficulty reasoning about such conditionally-terminating loops.
-We are able to handle such loops by computing \emph{termination invariants}.  This mechanism also allows us to
-prove that programs with multiple loops terminate, even if the termination of some loop depends on the states
-reachable after leaving a previous loop.
 

### 2014-06-18 papers/termination/paper.tex

--- 2014-06-18 papers/termination/paper.tex
@@
 
-Physical computers have bounded storage, which means they are unable to perform calculations on mathematical
-integers.  For example, if $A$ is the Ackermann function and $G$ is Graham's number, a physical computer
-capable of computing $A(G, G)$ would contain (much!) more matter than is believed to exist in the universe.
-Fortunately, it is rare for a programmer to need such a large number and so modern computers do their arithmetic
-over fixed-width binary words, otherwise known as bit-vectors.
+\noindent {\bf Bit-vectors (machine-level integers) vs. mathematical integers.}
+The abstraction of bit-vectors to mathematical integers ignores
+the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in
+incomparable behaviours with respect to termination.
 
-%Bit-vector arithmetic is very similar to modular arithmetic.  For a computer using $k$-bit words, unsigned
-%calulcations are just defined to 
-The abstraction of bit-vectors to mathematical integers 
-ignores the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in 
-incomparable behaviours with respect to termination. 
-%The loop in Figure~\ref{} terminates for integers, but does not terminate for bit-vectors. 
-The loop in Figure~\ref{} always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
-whereas, with bit-vector arithmetic, \texttt{x} under-flows before becoming negative and goes back to being positive causing the loop to never terminate.
-%% =======
-%% most of the work concentrated on finding \emph{linear ranking functions for linear programs over integers}.
-%% Next, we will consider in more detail the assumptions made by such techniques and their implications.\\
+Programs that terminate for integers may \emph{not} terminate for bit-vectors. For illustration, consider the following loop:
+\begin{lstlisting}[language=C]
+while (x > 0) x -= 2;
+\end{lstlisting}
+The loop always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
+whereas, with bit-vector arithmetic, \texttt{x} underflows before becoming negative and goes back to being positive causing the loop to never terminate.
 
-%% \noindent {\bf Bit-vectors (machine-level integers) vs. mathematical integers.}
-%% Physical computers have bounded storage, which means they are unable to perform calculations on mathematical
-%% integers.  For example, if $A$ is the Ackermann function and $G$ is Graham's number, a physical computer
-%% capable of computing $A(G, G)$ would contain (much!) more matter than is believed to exist in the universe.
-%% Fortunately, it is rare for a programmer to need such a large number and so modern computers do their arithmetic
-%% over fixed-width binary words, otherwise known as bit-vectors.
-
-%% Bit-vector arithmetic is very similar to modular arithmetic.  For a computer using $k$-bit words, unsigned
-%% calulcations are just defined to
-
-%% The abstraction of bit-vectors to mathematical integers 
-%% ignores the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in %, it gives rise to 
-%% %{\em unsound} and {\em incomplete} analyses, which is rather surprising given that bit-vectors are ubiquitous in computer systems. 
-%% incomparable behaviours with respect to termination:
-%% %The differences between the semantics of unbounded mathematical integers and that of machine integers result in 
-%% \begin{itemize}
-%% \item Programs that terminate for integers may \emph{not} terminate for bit-vectors. For illustration, consider the following loop:
-%% \begin{lstlisting}[language=C]
-%% while (x > 0) x -= 2;
-%% \end{lstlisting}
-%% The loop always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
-%% >>>>>>> 7b690c08a70e4161a432600e9456d922106badaf
-According to the C11 standard for the C programming language, a signed under- or over-flow yields undefined behaviour.
-However, compilers generally treat signed under- and over-flows using the
-wrap-around behaviour (see Section~\ref{sec:machine.arith}). %This means that a termination analysis for integers would return an unsound result for this situation.
-As a second example, the loop in Figure~\ref{} is  terminating for bit-vectors since \texttt{x}
+Similarly, programs that terminate for bit-vectors may \emph{not} terminate for integers. One such situation is illustrated next:  
+ \begin{lstlisting}[language=C]
+ while(x > 0) x++;
+ \end{lstlisting}
+This loop is  terminating for bit-vectors since \texttt{x}
 will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer

### 2014-06-18 slides/termination.tex
Termination slides

% Copyright 2009 by Tomasz Mazur
%
% This file may be distributed and/or modified in all ways.

\documentclass[xcolor=pdftex,t,11pt]{beamer}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%       SET OPTIONS BELOW        %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usetheme[
% Toggle showing page counter
pagecounter=true,
%
% String to be used between the current page and the
% total page count, e.g. of, /, from, etc.
pageofpages=of,
%
% Defines the shape of bullet points. Available options: circle, square
bullet=circle,
%
% Show a line below the frame title. 
titleline=true,
%
% Set the style of the title page (true for fancy, false for standard)
alternativetitlepage=true,
%
% Institution logo for fancy title page.
% Comment out to remove the logo from the title page.
% IMPORTANT: THERE IS A BUG IN SOME VERSIONS OF PDFLATEX AND FONTS
% ON THE LOGOS ARE NOT RENDERED PROPERLY. IN SUCH A CASE ADD `2` 
% TO THE NAME OF THE LOGO, E.G. comlab2 INSTEAD OF comlab
%titlepagelogo=images/titlepage/ou,
%
% Department footer logo for fancy title page
% Comment out to remove the logo from the footer of the title page/
% IMPORTANT: THERE IS A BUG IN SOME VERSIONS OF PDFLATEX AND FONTS
% ON THE LOGOS ARE NOT RENDERED PROPERLY. IN SUCH A CASE ADD `2` 
% TO THE NAME OF THE LOGO, E.G. comlab2 INSTEAD OF comlab
%titlepagefooterlogo=images/titlepage/comlab,
%
% Institution/department logo for ordinary slides
% Comment this line out to remove the logo from all the pages.
% Available logos are: ou, comlab, comlabinline, comlabou
% IMPORTANT: THERE IS A BUG IN SOME VERSIONS OF PDFLATEX AND FONTS
% ON THE LOGOS ARE NOT RENDERED PROPERLY. IN SUCH A CASE ADD `2` 
% TO THE NAME OF THE LOGO, E.G. comlab2 INSTEAD OF comlab
%ordinarypageslogo=TU-Signet,
%
%
% Add watermark in the bottom right corner
%watermark=<filename>,
%
% Set the height of the watermark.
%watermarkheight=100pt,
%
% The watermark image is 4 times bigger than watermarkheight.
%watermarkheightmult=4,
]{Torino}

% Select color theme. Available options are:
% mininmal, greenandblue, blue, red
\usecolortheme{blue}

%Select different font themes.Available options are:
% default, serif, structurebold, structureitalicserif, structuresmallcapsserif
\usefonttheme{structurebold}

\usepackage{tikz}
\usepackage{pgf}

\pgfdeclareimage[height=6ex]{ou-logo}{images/ou}

\logo{\pgfuseimage{ou-logo}}

\usepackage{listings}

\lstset{language=C,
 basicstyle=\tiny
 }

\usepackage{ifthen}

\usepackage{moreverb}
\usepackage{pgf}
\usepackage{tikz}
\usetikzlibrary{arrows, automata, shapes.multipart, chains, positioning, fit}

\usepackage{pifont}
\newcommand{\xmark}{\ding{55}}

\usepackage{color}
\definecolor{light-gray}{gray}{0.80}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%       PRESENTATION INFO        %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\author{Cristina David \and Daniel Kroening \and Matt Lewis}
\title{Everything You Ever Wanted To Prove About Termination But Were Afraid To Try}
%\subtitle{}
%\institute{Oxford University}
\date{\today}

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%       SLIDE DEFINITIONS        %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}[plain]
  \titlepage
\end{frame}

\begin{frame}{Outline}
  \tableofcontents
\end{frame}

\section{Motivation}

\begin{frame}[fragile]{Motivation}
Does this program terminate?
\begin{center}
\begin{minipage}{0.4\linewidth}
 \begin{lstlisting}[language=C,basicstyle=\normalsize]
while (x > 0) {
  x++;
}
 \end{lstlisting}
\end{minipage}
\end{center}

\pause

That depends...

It \emph{does not} terminate if the program operates on integers...

But it \emph{does} terminate if the program operates on fixed-width words
(-x decreases monotonically and is bounded by -MAXINT).

\end{frame}

\begin{frame}[fragile]{Motivation}
Which of these guys terminates?
\begin{center}
\begin{minipage}{0.45\linewidth}
 \begin{lstlisting}[language=C,basicstyle=\normalsize]
while (x > 0) {
  x = (x - 1) & x;
}
\end{lstlisting}
\end{minipage}
\begin{minipage}{0.45\linewidth}
 \begin{lstlisting}[language=C,basicstyle=\normalsize]
while (x > 0) {
  x = (x + 1) & x;
}
\end{lstlisting}
\end{minipage}
\end{center}

\pause

The left one does (x decreases monotonically), but the right one doesn't ($2 \rightarrow 2 \rightarrow \dots$).
\end{frame}

\begin{frame}{Motivation}
Proving program termination has been a big deal for a while.

It'd be nice if we could automatically prove termination for programs that run on computers.
\end{frame}

\section{Proving Termination}
\begin{frame}{What Does a Termination Proof Look Like?}
We prove that a program terminates by finding a \emph{ranking function}.

\vspace{1em}

If we have a transition relation $T$ and a loop guard $g$, a ranking function $R$
must meet the following criteria:
\begin{align}
 \forall x, x' \cdot & g(x) \Rightarrow R(x) > 0 ~ \wedge \\
                     & T(x, x') \Rightarrow R(x) > R(x')
\end{align}

Or more generally, $R$ must be an order-homomorphism with a well-founded co-domain (aka an ordinal).

\vspace{1em}

We are therefore in the business of finding ranking functions.
\end{frame}

\begin{frame}{Finding Ranking Functions}
Traditionally, people find ranking functions under some assumptions:

\begin{itemize}
 \item Programs are linear
 \item Ranking functions are linear
 \item Program variables are rationals
\end{itemize}

\vspace{1em}

With these assumptions in hand, one can deploy a very powerful piece of linear algebra: \emph{Farkas's lemma}.

This lemma makes light work of a ton of termination problems.

\vspace{1em}

But it's not for us.
\end{frame}

\begin{frame}{Synthesing Ranking Functions}
Our approach is to use \emph{program synthesis} to generate ranking functions.

\vspace{1em}

We're going to set up a program synthesis problem that will allow us to solve
the following 2nd-order satisfaction problem:
\begin{align*}
 \exists R . \forall x, x' \cdot & g(x) \Rightarrow R(x) > 0 ~ \wedge \\
                             & T(x, x') \Rightarrow R(x) > R(x')
\end{align*}

\pause

\vspace{1em}

In other words: ``Is there a ranking function for my program?''

\end{frame}

\begin{frame}{Specifying a Ranking Function}
Our program synthesiser needs a \emph{specification}, which takes the form of a C function.

\vspace{1em}

The specification function takes two arguments: a candidate program $P$ and an input $x$.
If $P$ gives the correct output when fed $x$, our specification function returns true,
otherwise it returns false.
\end{frame}

\begin{frame}[fragile]{A Specification}

\vspace{-1em}

\begin{center}
\begin{minipage}{.6\textwidth}
\begin{lstlisting}[language=C++,basicstyle=\small]
bool spec(prog_t *p, int x) {
  int y = exec(p, x);

  if (y >= x)
    return true;
  else
    return false;
}
\end{lstlisting}
\end{minipage}
\end{center}

\pause

Two of the possible programs:

\begin{itemize}
  \item[] \texttt{int f(int x) \{ return x; \}} 
  \item[] \texttt{int f(int x) \{ return MAXINT; \}}
\end{itemize}

\end{frame}

\begin{frame}[fragile]{Specifying a Ranking Function}
\vspace{-1em}

\begin{center}
\begin{minipage}{0.6\textwidth}
\begin{lstlisting}[language=c++,basicstyle=\small]
int spec(prog_t *p, int x) {
  int r1 = exec(p, x);

  if (g(x)) {
    if (r1 <= 0)
      return false;

    int y = body(x);
    int r2 = exec(p, y);

    if (r1 <= r2)
      return false;
  }

  return true;
}
\end{lstlisting}
\end{minipage}
\end{center}
\end{frame}

\iffalse

\section{Abstract Program Synthesis}
\begin{frame}{The Synthesis Formula}
Our high-level goal is to synthesise programs that meet some specification.

\pause

Abstractly, a program $P$ computes some function from an input set $I$ to an output set $O$:
$$P : I \rightarrow O$$

\pause

A specification is a relation between inputs and outputs:
$$\sigma \subseteq I \times O$$

\pause

The synthesis problem is to find a program that meets the specification on all inputs:
$$\exists P . \forall x \in I . \sigma(x, P(x))$$
\end{frame}
\fi

\begin{frame}[fragile]{Solving the Synthesis Formula with CEGIS}
 Solving second order formulae is hard.  Instead we'll alternately solve two \emph{first-order} formulae
 in a refinement loop:

 \vspace{1cm}

  \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,
 semithick, initial text=]

  \matrix[nodes={draw, fill=none, scale=1, shape=rectangle, minimum height=1cm, minimum width=1.5cm},
          row sep=2cm, column sep=3cm] {
   \node (synth) {Synthesise};
   &
   \node (verif) {Verify}; %\\
   %\node[draw=none] {};
   &
   \node[ellipse] (done) {Done}; \\
  };

   \path
    (synth) edge [bend left] node {Candidate program} (verif)
    (verif) edge [bend left] node {Counterexample input} (synth)
    (verif) edge node {Valid} (done);
 \end{tikzpicture}
\end{frame}

\begin{frame}{First-order Synthesis}
 If we have some small set of test inputs $\{ x_0, \dots, x_N\}$, we can find a program that is correct on just that set:

 $$\exists P . \sigma(x_0, P(x_0)) \wedge \dots \wedge \sigma(x_N, P(x_N))$$
\end{frame}

\begin{frame}{First-order Verification}
 If we have some program $P$ that might be correct, we can check whether it is in fact correct:

 $$\exists x . \lnot \sigma(x, P(x))$$

 If this formula is \emph{satisfiable}, $P$ is \emph{not} correct and we have found an input on which it is incorrect.
 We can add this input to our set of test inputs and loop back to synthesising a new program.
\end{frame}

\section{Program Synthesis}
\iffalse
\begin{frame}{Concrete Synthesis in C}
 We're using C as an implementation language.  To do that we need:

 \begin{itemize}
  \item A way of encoding programs as \emph{finite structures} in C.
  \item A way of encoding the first-order synthesis and verification formulae in C.
  \item A way of checking properties of C programs.
 \end{itemize}
\end{frame}

\begin{frame}[fragile]{Encoding Programs in C}
 To encode the programs we will synthesise, we just make a very simple language and an interpreter for it.  Programs
 in our language look like this:

 \vspace{0.7cm}

\begin{center}
\begin{minipage}{0.3\linewidth}
\begin{verbatimtab}
t1 = neg x
t2 = add t1 1
return t2
\end{verbatimtab}
\end{minipage}
\end{center}

\end{frame}

\fi

\begin{frame}{Complexity}
By far the dominant factor in how long synthesis takes is the length of the shortest correct program.

\vspace{1em}

This is called the \emph{Kolmogorov complexity} of the function being computed.

\vspace{1em}

For termination, this gives us the property that our runtime doesn't depend very heavily
on the size of the program we are analysing.  It is instead almost entirely determined by the
size of the shortest program that computes a valid ranking function.
\end{frame}

\begin{frame}{Genetic Programming}
The flip side of being so dominated by Kolmogorov complexity is that we
get \emph{really slow} beyond about 4 instructions.

\vspace{1em}

It turns out that \emph{genetic programming} helps a lot beyond this threshold.
\end{frame}

\begin{frame}{Genetic Programming}
Genetic programming is an evolutionary search strategy, which aims to find a program
that is ``fit'' according to some metric.

\vspace{1em}

It begins by generating a population of random programs.  This population is subjected to
\emph{evolution} over a period of many generations.

\vspace{1em}

At each generation, each individual's fitness is evaluated.  The fitter an individual is,
the greater its chance of breeding.

When two programs breed, their code is combined in some way (the crossover operation) and
the resulting child may be randomly modified (the mutation operator).  This child is copied
into the next generation.
\end{frame}

\begin{frame}{Genetic Programming Synthesis}
To use genetic programming for synthesis, we need to fix a useful fitness metric.

\vspace{1em}

For our purposes, we say that if a program meets the specification on $n$ test vectors, its
fitness is $n$.

Whenever we add a new test vector, we continue genetic programming with the most recent population
(the one that produced the most recent candidate program).  This is called \emph{incremental evolution}
and gives us a big speedup.

\vspace{1em}

Once we've found a program that passes all the tests, we return it and continue round the CEGIS
loop.
\end{frame}

\section{Advanced Topics}
\begin{frame}{Conditional Termination}

\begin{align*}
\exists I, R \cdot \forall x, x' \cdot & P(x, x') \Rightarrow & I(x') \\
                                       & I(x) \wedge g(x) \wedge T(x, x') \Rightarrow & R(x) > 0 ~ \wedge \\
                                       & & R(x) > R(x') ~ \wedge \\
                                       & & I(x')
\end{align*}
\end{frame}

\begin{frame}{Sequential Loops}
\begin{align*}
\exists I_1, I_2, R \cdot \forall x, x' \cdot & P(x, x') \Rightarrow & I_1(x') \\
                                              & I_1(x) \wedge g_1(x) \wedge T_1(x, x') \Rightarrow & I_1(x') \\
                                              & I_1(x) \wedge \lnot g_1(x) \Rightarrow & I_2(x) \\
					      & I_2(x) \wedge g_2(x) \wedge T_2(x, x') \Rightarrow & R(x) > 0 ~ \wedge \\
					      & & R(x) > R(x') ~ \wedge \\
					      & & I_2(x')
\end{align*}
\end{frame}

\begin{frame}{Some Combinatorics}
A terminating program corresponds to a \emph{labelled partial order} on $2^k$ states.

\vspace{.7em}

Counting partial orders on a finite set turns out to be very hard (it's an open problem!).  However, asymptotics are known:
\begin{align*}
 P_n & = \left(1 + O \left( \frac{1}{n} \right) \right) \left( \sum_{i=1}^n \sum_{j=1}^{n-1} \binom{n}{i} \binom{n-i}{j} (2^i - 1)^j (2^j - 1)^{n - i - j}\right) \\
     & \sim O\left( 2^k \right)
\end{align*}

\end{frame}

\begin{frame}{More Combinatorics}
Linear functions on bitvectors correspond to \emph{permutations} of the natural ordering on bitvectors.

\vspace{1em}

There are $2^{2k}$ linear functions (although only half of these permute the full space, cf. generators of the group $2^k$).

\vspace{1em}

Each of these functions can show that each of the \emph{unlabelled} partial orders terminates.

The number of unlabelled partial orders over an $n$ element set is related to the number of labelled partial orders as:

$$p_n \sim {P_n \over n!}$$
\end{frame}

\begin{frame}{The Combinatorial Payoff}
Since there are $P_n$ terminating functions, and each of the $n^2$ linear functions can prove that
$p_n \sim {P_n \over n!}$ functions terminate, we can compute the probability that a random terminating
function can be proved to terminate with a linear argument:

$${2^k \times 2^k \over {2^k}!} \sim {1 \over {2^k}!}$$

\pause

\vspace{1em}

This number is very small, which to some degree justifies our aim of moving beyond linear ranking functions.

\end{frame}

\begin{frame}[fragile]{Fin}

\begin{center}
\includegraphics[width=.7\textwidth]{sloth.jpg}

\Huge
 Thank you!
\end{center}

\end{frame}

\end{document}

### 2014-06-20 papers/termination/paper.tex
A bit of work on section 6

@@
 \section{Termination as Second-Order Satisfaction}
-We can specify the existence of a ranking function, and therefore the termination
-of a program, using a second order formula:
+In this section we will show how several variations of the halting problem can be defined in second-order
+logic.  As it turns out, all of these variations can be defined in a fragment of second-order logic
+we will refer to as the \emph{synthesis fragment}.
 
@@
 \begin{definition}[Second-order Termination Formula]
+\label{def:termination-formula}
 \begin{align*}
- \exists R . \forall x, x' . & R(x) > 0 ~ \wedge \\
+ \exists R \cdot \forall x, x' . & R(x) > 0 ~ \wedge \\
                              & T(x, x') \rightarrow R(x) > R(x')
@@
 
+\begin{definition}[Conditional Termination Formula]
+ \begin{align*}
+  \exists R, I \cdot \forall x, x' . & I(x) \rightarrow V(x) ~ \wedge \\
+                                 & L(x) \wedge V(x) \wedge T(x, x') \rightarrow V(x') \wedge R(x) > 0 \wedge R(x) > R(x')
+ \end{align*}
+\end{definition}
+
+\begin{definition}[Multiple Loops Termination Formula]
+ \begin{align*}
+  \exists R, V_1, V_2 \cdot \forall x, x' . & I(x) \rightarrow V_1(x) ~ \wedge \\
+                                        & V_1(x) \wedge L_1(x) \wedge T(x, x') \rightarrow V_1(x') ~ \wedge \\
+                                        & V_1(x) \wedge \lnot L_1(x) \rightarrow V_2(x) ~ \wedge \\
+                                        & V_2(x) \wedge L_2(x) \wedge T(x, x') \rightarrow V_2(x') \wedge R(x) > 0 \wedge R(x) > R(x')
+ \end{align*}
+\end{definition}
+
+\begin{definition}[Non-Termination Formula]
+\label{def:nonterm-formula}
+ \begin{align*}
+  \exists N, x_0 \cdot \forall x \cdot \exists x' \cdot & I(x_0) \wedge L(x_0) \wedge N(x_0) ~ \wedge \\
+							& N(x) \wedge L(x) \rightarrow T(x, x') \wedge N(x') \wedge L(x')
+ \end{align*}
+\end{definition}
+
+\begin{definition}[Deterministic Non-Terminaton Formula]
+ \begin{align*}
+  \exists N, x_0 \cdot \forall x, x' \cdot & I(x_0) \wedge L(x_0) \wedge N(x_0) ~ \wedge \\
+							& N(x) \wedge L(x) \wedge T(x, x') \rightarrow N(x') \wedge L(x')
+ \end{align*} 
+\end{definition}
+
+\caption{The halting problem defined in second-order logic.\label{fig:termination-formulae}}
+\end{figure*}
+
+\iffalse
 Many loops do not terminate for all starting states, but are contained in programs that guarantee the loop will
@@
 
-\begin{definition}[Conditional Termination Formula]
- \begin{align*}
-  \exists R, I . \forall x, x' . & P(x) \rightarrow I(x) ~ \wedge \\
-                                 & c(x) \wedge I(x) \wedge T(x, x') \rightarrow I(x') \wedge R(x) > 0 \wedge R(x) > R(x')
- \end{align*}
-\end{definition}
-
-\begin{definition}[Multiple Loops Termination Formula]
- \begin{align*}
-  \exists R, I_1, I_2 . \forall x, x' . & P(x) \rightarrow I_1(x) ~ \wedge \\
-                                        & I_1(x) \wedge L_1(x, x') \rightarrow I_1(x') ~ \wedge \\
-                                        & I_1(x) \wedge \lnot c_1(x) \rightarrow I_2(x) ~ \wedge \\
-                                        & c(x) \wedge I_2(x) \wedge L_2(x, x') \rightarrow I_2(x') \wedge R(x) > 0 \wedge R(x) > R(x')
- \end{align*}
-\end{definition}
-\end{figure*}
-
 Our method for ranking function synthesis can be stated as follows:
@@
 Any termination guarantees?  
+\fi
 
-\section{Second-Order Satisfaction as Program Synthesis}
+We can specify the existence of a ranking function, and therefore the termination
+of a program, with the second order formulae defined in Fig.~\ref{fig:termination-formulae}.
+We can observe that each of these formulae (with the exception of Definition~\ref{def:nonterm-formula})
+is an $\exists \forall$ formula where the the universally quantified variables only range over
+ground terms.  This class of formulae is the same as the class of program synthesis formulae.
 The program synthesis problem can be described as finding a satisfying assigment to the
@@
 \begin{definition}[Synthesis Formula]
- The synthesis formula for a specification $\sigma: X \to Y$ is:
-
- $$\exists P \cdotp \forall x \cdotp \sigma(x, P(x))$$
+ \label{def:synth-formula}
+ $$\exists \vec{P}, \vec{x} \cdotp \forall \vec{y} \cdotp \sigma(\vec{x}, \vec{P}, \vec{y}) $$
 \end{definition}
@@
 
-\subsection{Generalised Synthesis}
-Sometimes we want to synthesise several programs at once, as well as some
-ground terms $\vec{x}$.  Also, we might want a more complex specification
-than just a relation over $X$ and $Y$.
-
-\begin{definition}[Generalised Synthesis Formula]
- $$\exists P, \vec{x} \cdotp \forall \vec{y} \cdotp \sigma(x, P, y) $$
-\end{definition}
-
 \subsection{Reducing Termination to Synthesis}
 For a transition system $T$, we can build a specification $\sigma$ to find a conditional ranking function for
-$T$ with initial states I:
-
-\begin{eqnarray}
- \sigma(V: X \to \mathbb{B}, R: X \to \mathbb{N}, x, x') = & I(x) \rightarrow V(x) \wedge \\
- & V(x) \wedge T(x, x') \rightarrow V(x') \wedge R(x) < R(x') 
-\end{eqnarray}
+$T$ with initial states I.  This specification is defined in Definition~\ref{def:termination-spec}.
 

### 2014-06-20 papers/termination/paper.tex
Some more work on section 6

@@
 is an $\exists \forall$ formula where the the universally quantified variables only range over
-ground terms.  This class of formulae is the same as the class of program synthesis formulae.
+ground terms.
 The program synthesis problem can be described as finding a satisfying assigment to the
@@
  \label{def:synth-formula}
- $$\exists \vec{P}, \vec{x} \cdotp \forall \vec{y} \cdotp \sigma(\vec{x}, \vec{P}, \vec{y}) $$
+ $$\exists P, x \cdot \forall y \cdot \sigma(x, y, P)$$
 \end{definition}
 
-Since this formula involves quantification over functions $X \to Y$,
-this is a second-order formula -- indeed, if $X = Y = \mathbb{N}$,
-the formula describes a set at level $\Sigma_1^1$ of the analytical
-hierarchy.  As such, determining the satisfiability of the synthesis
-formula over a given logic is strictly harder than solving the
-halting problem over the same logic.
+Here we can view $\sigma: X \times Y \times (X \times Y \to Z) \to \mathbb{B}$ as a specification function.
+The idea is that $\sigma$ returns true iff the function $P$ computes appropriate outputs
+when fed the inputs $x$ and $y$.  If we can find a function $P$ and a value $x$ that meet the specification
+for all $y$, we have found a solution to the synthesis problem and write $P \models \sigma$.  Since the formulae
+of Fig.~\ref{fig:termination-formulae} are syntactically in the synthesis fragment, by concocting an appropriate
+specification function we reduce the problem of proving termination to the problem of
+program synthesis.
 
-\subsection{Reducing Termination to Synthesis}
-For a transition system $T$, we can build a specification $\sigma$ to find a conditional ranking function for
-$T$ with initial states I.  This specification is defined in Definition~\ref{def:termination-spec}.
-
-Here we have synthesised a ranking function $R$ into $\mathbb{N}$ (which is well-founded),
-as well as an inductive invariant $V$ that guarantees termination of $T$.
-
-To synthesise an n-place lexicographic ranking function, we just ask for a ranking function
-$R: X \to \mathbb{N}^n$.
-
-We recall the definition of the Kolmogorov-complexity of a function $f$:
+\subsection{Synthesis Complexity}
+In Section~\ref{sec:synthesis} we will describe a procedure for solving the synthesis formula.
+This procedure has the property that it will find the shortest program meeting the
+specification, and as a consequence the runtime of the program synthesiser is dominated
+by the size of this shortest program.  To frame the discussion of the runtime of our
+termination proving procedure, we first recall the definition of the Kolmogorov complexity
+of a function $f$:
 
@@
 
-We are primarily interested in studying low-Kolmogorov-complexity (LKC)
-functions.
+We can extend this definition slightly to talk about the Kolmogorov complexity of a
+synthesis problem in terms of its specification:
+
+\begin{definition}[Kolmogorov complexity of a synthesis problem]
+ The Kolmogorov complexity of a program specification $K(\sigma)$ is the length of the shortest
+ program $P$ such that $P \models \sigma$.
+\end{definition}
+
+We will later show that the time complexity of our synthesis procedure is $NP^{NP}(K(\sigma))$,
+which means that for even moderately large $K(\sigma)$ our procedure is intractably slow.  Conversely
+if $K(\sigma)$ is small we will be able to find a termination proof, regardless of other parameters
+such as the size of the program under analysis.  To put this another way, under the assumption
+that $K(\sigma)$ is small for a program's termination problem, we will find a ranking function
+quite rapidly.  We will now investigate the properties of functions with low Kolmogorov complexity,
+and show that the assumption of a low-Kolmogorov ranking function is much weaker than the assumption
+of a linear ranking function.
 
 \begin{theorem}
- Linear functions are LKC.
+ Linear functions have low Kolmogorov complexity.
 \end{theorem}
@@
 \begin{proof}
- The program to compute a linear function $f: X \to Y$ is of size linear in
- $\dim(X) \times \dim(Y)$.
+ The function $f: X \to Y$ can be computed with a program consisting of
+ $2 \cdot \dim(X) \cdot \dim(Y) - \dim(Y)$ instructions (1 multiplication and 1 addition for
+ each cell in the matrix representing $f$).  Therefore,
+ $K(f) \leq 2 \cdot \dim(X) \cdot \dim(Y) - \dim(Y)$.
 \end{proof}
 
-\begin{conjecture}
- Most LKC functions are non-linear.
-\end{conjecture}
+\begin{theorem}
+ Most functions are non-linear.
+\end{theorem}
 

### 2014-06-23 papers/termination/paper.tex
Some work on section 6

@@
 are correct for fixed-width machine arithmetic and floating-point arithmetic.
-We use this reduction to build a semi-decision procedure for the termination
+We use this reduction to build a sound and complete procedure for the termination
 of fixed-width and floating-point arithmetic programs.
@@
 
+\iffalse
 \section{Combinatorics of Finite Termination}
@@
 \end{corollary}
+\fi
 
@@
 \begin{theorem}
- Most functions are non-linear.
+Assuming a reasonable program encoding, the probability that a random program of length $l$ computes
+a linear function is $O(2^{-l})$.
 \end{theorem}

### 2014-06-23 papers/termination/paper.tex
revised intro

--- 2014-06-20 papers/termination/paper.tex
@@
 \usepackage{appendix}
-\usepackage{caption}
+\usepackage{caption,subcaption}
 \DeclareCaptionType{copyrightbox}
@@
 %% while failing to prove termination for terminating programs whose ranking functions fall short of the considered restriction.
-In doing so, the majority of the existing termination provers diverge 
-in one important aspect from the execution of a program on a computer:
-bit-vectors and floats are treated as mathematical integers and reals, respectively.
-Figure~\ref{fig:handletable} presents a summary of existing termination analyses with indeed
-most of them designed for integers and assuming a \emph{restrictive transition relation}, e.g.
-linear programs \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10}.
-% \emph{linear ranking functions for linear programs over integers}.
-These design decisions may give rise to 
-{\em unsound} and {\em incomplete} analyses, as discussed next.
-%which is rather surprising given that bit-vectors and floats are ubiquitous in computer systems.
 
-%\noindent {\bf Bit-vectors (machine-level integers) vs. mathematical integers.}
-
-\noindent {\bf Bit-vectors (machine-level integers) vs. mathematical integers.}
-The abstraction of bit-vectors to mathematical integers ignores
-the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in
-incomparable behaviours with respect to termination.
-
-Programs that terminate for integers may \emph{not} terminate for bit-vectors. For illustration, consider the following loop:
-\begin{lstlisting}[language=C]
-while (x > 0) x -= 2;
-\end{lstlisting}
-The loop always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
-whereas, with bit-vector arithmetic, \texttt{x} underflows before becoming negative and goes back to being positive causing the loop to never terminate.
-
-Similarly, programs that terminate for bit-vectors may \emph{not} terminate for integers. One such situation is illustrated next:  
- \begin{lstlisting}[language=C]
- while(x > 0) x++;
- \end{lstlisting}
-This loop is  terminating for bit-vectors since \texttt{x}
-will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
-arithmetic since the loop condition stays always true for any initial \texttt{x} at least 1.
-
-%Similarly, programs that terminate for bit-vectors may \emph{not} terminate for integers. One such situation is illustrated next:
-
-\begin{lstlisting}[language=C]
-while (x > 0) x -= 2;
-\end{lstlisting}
-
-\begin{lstlisting}[language=C]
- while(x > 0) x++;
- \end{lstlisting}
-
-%\noindent {\bf Floats vs. reals.} 
-A scenario similar to the one for bit-vectors happens if floats were to be abstracted to unbounded reals by termination provers \cite{}. 
-Consequently, these provers ignore potential rounding errors, under- and over-flows, which are precisely what makes  reasoning about floating point inherently difficult.
-This approximation may lead to erroneous diagnosis of a program's terminating behaviour as illustrated by the loops in Figure~\ref{} and Figure~\ref{}.
-While the former does not terminate for reals, but does for floats, the latter 
-terminates for reals, but does not for floats.
-
-\begin{lstlisting}
-while (x > 0.0) x *= 0.5;
-\end{lstlisting}
-
-\begin{lstlisting}
-while (x > 0.0) x -= 1.0;
-\end{lstlisting}
-\todo{explain the reasons for non-termination with figures.}\\
-
-%% \noindent {\bf Linear programs and linear ranking functions.} As visible in Figure~\ref{fig:handletable}, 
-%% most termination techniques assume \emph{restrictive transition relations}, e.g. linear programs,  and are only able to compute
-%% \emph{linear ranking functions} \cite{}. To better understand this restriction, we computed the probability of a random terminating program having a linear ranking
-%% function (see Section ?). This probability proved to be very low, indicating that the linearity assumption for the ranking functions is indeed prohibitive.
-
-If we survey the area of program termination chronologically, we observe an initial domination of  monolithic approaches based on a single measure shown to decrease
+When surveying the area of program termination chronologically, we observe an initial domination of  monolithic approaches based on a single measure shown to decrease
 over all program paths %(syntactic characterisation of loops) 
@@
 there are cases in which proofs based on local measures involve applying linear functions while corresponding global
-measures involve nonlinear functions or lexicographic orders.
+measures involve nonlinear functions or lexicographic orders. Consequently, approaches based on Ramsey's theorem 
+search for linear termination arguments.
 
@@
 Following the same trend, %of switching the complexity in termination checking back to the termination arguments, 
-we investigate its very extreme: \emph{unrestricted} termination arguments. 
+we investigate its extreme: \emph{unrestricted} termination arguments. 
 This means that our ranking functions may involve non-linearity and lexicographic orders (we do not commit to any such form, i.e. we do not use templates).
-That is, we propose a general framework that does not assume the existence of linear ranking functions and can uniformly compute
+
+While the summary  in Figure~\ref{fig:handletable} supports our observation that the majority of existing termination analyses are designed for
+linear programs and linear ranking functions, it also identifies another common assumption that is being made: the equivalence of bit-vector/float semantics and integer/real semantics.
+Thus, most of the techniques treat
+bit-vectors and floats as mathematical integers and reals, respectively  \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10}.
+
+By assuming bit-vector/float semantics to be equivalent to integer/real semantics, 
+these techniques ignore the wrap-around behaviour caused by under/over-flows as well as width conversions, 
+and ultimately diverge from the execution of a program on a computer.  
+In Section~\ref{sec:motivation}, we show that integers/reals and bit-vectors/floats exhibit
+incomparable behaviours with respect to termination, e.g.
+programs that terminate for integers may \emph{not} terminate for bit-vectors.
+Thus, abstracting bit-vectors/floats to integers/reals may give rise to 
+{\em unsound} and {\em incomplete} analyses.
+%design decisions
+%which is rather surprising given that bit-vectors and floats are ubiquitous in computer systems.
+
+%% \noindent {\bf Bit-vectors (machine-level integers) vs. mathematical integers.}
+%% The abstraction of bit-vectors to mathematical integers ignores
+%% the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in
+%% incomparable behaviours with respect to termination.
+
+%% Programs that terminate for integers may \emph{not} terminate for bit-vectors. For illustration, consider the following loop:
+%% \begin{lstlisting}[language=C]
+%% while (x > 0) x -= 2;
+%% \end{lstlisting}
+%% The loop always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
+%% whereas, with bit-vector arithmetic, \texttt{x} underflows before becoming negative and goes back to being positive causing the loop to never terminate.
+
+%% Similarly, programs that terminate for bit-vectors may \emph{not} terminate for integers. One such situation is illustrated next:  
+%%  \begin{lstlisting}[language=C]
+%%  while(x > 0) x++;
+%%  \end{lstlisting}
+%% This loop is  terminating for bit-vectors since \texttt{x}
+%% will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
+%% arithmetic since the loop condition stays always true for any initial \texttt{x} at least 1.
+
+%% %Similarly, programs that terminate for bit-vectors may \emph{not} terminate for integers. One such situation is illustrated next:
+
+%% \begin{lstlisting}[language=C]
+%% while (x > 0) x -= 2;
+%% \end{lstlisting}
+
+%% \begin{lstlisting}[language=C]
+%%  while(x > 0) x++;
+%%  \end{lstlisting}
+
+%% %\noindent {\bf Floats vs. reals.} 
+%% A scenario similar to the one for bit-vectors happens if floats were to be abstracted to unbounded reals by termination provers \cite{}. 
+%% Consequently, these provers ignore potential rounding errors, under- and over-flows, which are precisely what makes  reasoning about floating point inherently difficult.
+%% This approximation may lead to erroneous diagnosis of a program's terminating behaviour as illustrated by the loops in Figure~\ref{} and Figure~\ref{}.
+%% While the former does not terminate for reals, but does for floats, the latter 
+%% terminates for reals, but does not for floats.
+
+%% \begin{lstlisting}
+%% while (x > 0.0) x *= 0.5;
+%% \end{lstlisting}
+
+%% \begin{lstlisting}
+%% while (x > 0.0) x -= 1.0;
+%% \end{lstlisting}
+%% \todo{explain the reasons for non-termination with figures.}\\
+
+%% \noindent {\bf Linear programs and linear ranking functions.} As visible in Figure~\ref{fig:handletable}, 
+%% most termination techniques assume \emph{restrictive transition relations}, e.g. linear programs,  and are only able to compute
+%% \emph{linear ranking functions} \cite{}. To better understand this restriction, we computed the probability of a random terminating program having a linear ranking
+%% function (see Section ?). This probability proved to be very low, indicating that the linearity assumption for the ranking functions is indeed prohibitive.
+
+We propose a general framework that does not assume the existence of linear ranking functions and can uniformly compute
 \emph{lexicographic/non-lexicographic} \emph{linear/non-linear} 
@@
 for imperative loops with polynomial guards and polynomial assignments
-is \cite{DBLP:conf/vmcai/BradleyMS05}.
+is \cite{DBLP:conf/vmcai/BradleyMS05}. However, this work extends only to polynomials.
 
@@
 
-\section{Motivational Examples} 
+\section{Motivational Examples} \label{sec:motivation}
 We next discus some intricate examples that prove challenging for existent termination analysis.
@@
 
+\begin{figure*}
+\centering
+ \begin{tabular}{ccc}
+
+\begin{subfigure}[b]{0.3\textwidth}
 \begin{lstlisting}
@@
 \end{lstlisting}
+\caption{}
+ \label{fig:motivation.a}
+\end{subfigure}%
 
+&
+
+\begin{subfigure}[b]{0.3\textwidth}
 \begin{lstlisting}
@@
 \end{lstlisting}
+\caption{}
+ \label{fig:motivation.b}
+\end{subfigure}%
 
+&
+
+\begin{subfigure}[b]{0.3\textwidth}
 \begin{lstlisting}
@@
 \end{lstlisting}
+\caption{}
+ \label{fig:motivation.c}
+\end{subfigure}%
+
+\end{tabular}
+\end{figure*}
 

### 2014-06-23 papers/termination/paper.tex
Flesh out 2nd order termination

@@
 \usepackage{multirow}
+\usepackage{framed}
 \usepackage{tikz}
@@
 logic.  As it turns out, all of these variations can be defined in a fragment of second-order logic
-we will refer to as the \emph{synthesis fragment}.
+we will refer to as the \emph{synthesis fragment}.  We will make use of this observation to frame the halting
+problem as a program synthesis problem, allowing us to employ the program synthesiser described in
+Section~\ref{sec:synthesis}.
 
 \begin{figure*}
+\begin{framed}
 \begin{definition}[Second-order Termination Formula]
@@
 \begin{align*}
- \exists R \cdot \forall x, x' . & R(x) > 0 ~ \wedge \\
-                             & T(x, x') \rightarrow R(x) > R(x')
+ \exists R \cdot \forall x, x' \cdot & L(x) \wedge T(x, x') \rightarrow R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
@@
 \begin{definition}[Conditional Termination Formula]
+\label{def:conditional-termination-formula}
  \begin{align*}
-  \exists R, I \cdot \forall x, x' . & I(x) \rightarrow V(x) ~ \wedge \\
+  \exists R, V \cdot \forall x, x' \cdot & I(x) \rightarrow V(x) ~ \wedge \\
                                  & L(x) \wedge V(x) \wedge T(x, x') \rightarrow V(x') \wedge R(x) > 0 \wedge R(x) > R(x')
@@
 \begin{definition}[Multiple Loops Termination Formula]
+\label{def:multi-termination-formula}
  \begin{align*}
-  \exists R, V_1, V_2 \cdot \forall x, x' . & I(x) \rightarrow V_1(x) ~ \wedge \\
+  \exists R, V_1, V_2 \cdot \forall x, x' \cdot & I(x) \rightarrow V_1(x) ~ \wedge \\
                                         & V_1(x) \wedge L_1(x) \wedge T(x, x') \rightarrow V_1(x') ~ \wedge \\
@@
 
-\begin{definition}[Deterministic Non-Terminaton Formula]
+\begin{definition}[Deterministic Non-Termination Formula]
+\label{def:deterministic-nonterm-formula}
  \begin{align*}
@@
 \end{definition}
-
-\caption{The halting problem defined in second-order logic.\label{fig:termination-formulae}}
+\end{framed}
 \end{figure*}
@@
 \fi
+
+The existence of a ranking function for a single loop is equivalent to the satisfiability
+of the formula from Definition~\ref{def:termination-formula}.  If this formula is satisfiable,
+$R$ is a ranking function that proves the loop $L$ terminates, so the termination of $L$
+is equivalent to the satisfiability of Definition~\ref{def:termination-formula}.
+
+A slightly more involved case occurs if $L$'s termination depends on the state it begins
+executing in, i.e. $L$ is \emph{conditionally terminating}.  This property is equivalent to the
+formula of Defintion~\ref{def:conditional-termination-formula}.  If this formula is satisfiable,
+$V$ is an inductive invariant of $L$ that is established by the initial states $I$.
+$R$ is then a ranking function for $L$ as restricted by $V$ -- that is to say, $R$ need only
+be well founded on those states satisfying $V$.  But since $V$ is an inductive invariant of $L$,
+$R$ is strong enough to show that $L$ terminates from any of its initial states.  We say
+that $V$ is a \emph{termination invariant} for $L$ and that $R$ proves termination relative to $V$.
+
+The conditional termination formula becomes more complex in the presence of multiple loops.
+Consider the following program with two loops $L_1$ and $L_2$:
+
+\begin{lstlisting}
+while (L1) { ... }
+while (L2) { ... }
+\end{lstlisting}
+
+The termination problem for $L_2$ is equivalent to the satisfiability of
+Definition~\ref{def:multi-termination-formula}.  If this formula is satisfiable,
+$V_1$ is an inductive invariant of $L_1$ and $V_2$ is a termination invariant of $L_2$.
+$V_1$ is established by the initial conditions $I$ and is strong enough to establish
+$V_2$ when conjoined with the negation of $L_1$'s guard.  As before, $R$ is a ranking
+function relative to $V_2$.
+
+Non-termination of the loop $L$ is encoded by Definition~\ref{def:nonterm-formula}.
+If this formula is satisfiable, $N$ is a recurrence set for $L$ which proves
+$L$'s non-termination.  However this formula has one more level of quantifier alternation
+than the others (it is an $\exists \forall \exists$ formula), which will cause us problems
+shortly.  We observe that in the case that $T$ is deterministic (that is, each state has
+exactly one successor),  the non-termination problem is equivalent to the
+satisfiability of Definition~\ref{def:deterministic-nonterm-formula}.
 

### 2014-06-23 papers/termination/paper.tex
2nd order termination tweak
2nd-order termination

@@
 of the formula from Definition~\ref{def:termination-formula}.  If this formula is satisfiable,
-$R$ is a ranking function that proves the loop $L$ terminates, so the termination of $L$
-is equivalent to the satisfiability of Definition~\ref{def:termination-formula}.
+$R$ is a ranking function that proves the loop $L$ terminates.
 
@@
 The conditional termination formula becomes more complex in the presence of multiple loops.
-Consider the following program with two loops $L_1$ and $L_2$:
+Consider the following program with two loops $L_1$ and $L_2$, one after the other:
 
@@
 
-We can specify the existence of a ranking function, and therefore the termination
-of a program, with the second order formulae defined in Fig.~\ref{fig:termination-formulae}.
-We can observe that each of these formulae (with the exception of Definition~\ref{def:nonterm-formula})
-is an $\exists \forall$ formula where the the universally quantified variables only range over
-ground terms.
-The program synthesis problem can be described as finding a satisfying assigment to the
-synthesis formula:
+With the exception of Definition~\ref{def:nonterm-formula}, each of the second-order formulae
+above is an $\exists \forall$ formula where the the universally quantified variables only range over
+ground terms.  We refer to this syntactic class as the \emph{synthesis fragment} of second-order
+logic, because program synthesis problem can be described as finding a satisfying assigment to the
+following formula:
 

### 2014-06-23 papers/termination/paper.tex
Remove a bunch of LKC vs. linearity stuff
revised motivation

--- 2014-06-23 papers/termination/paper.tex
@@
 \section{Motivational Examples} \label{sec:motivation}
+In the previous sections, we have discussed some of the assumptions made by existent termination analyses (see Figure~\ref{fig:handletable}). 
+Now, we provide some simple examples that broke those assumptions and
+
 We next discus some intricate examples that prove challenging for existent termination analysis.
@@
 \begin{subfigure}[b]{0.3\textwidth}
+\begin{lstlisting}[language=C]
+ while(x > 0) x++;
+ \end{lstlisting}
+\caption{}
+ \label{fig:motivation.c}
+\end{subfigure} \\
+
+\hline
+
+\begin{subfigure}[b]{0.3\textwidth}
+\begin{lstlisting}
+  while ( q >= 0 ) {
+    q = q - y;
+    y = y + 1;
+  }
+\end{lstlisting}
+\caption{}
+ \label{fig:motivation.d}
+\end{subfigure}
+
+&
+
+\begin{subfigure}[b]{0.3\textwidth}
+\begin{lstlisting}
+while (x > 0.0) x -= 1.0;
+\end{lstlisting}
+\caption{}
+ \label{fig:motivation.f}
+\end{subfigure}
+
+&
+
+\begin{subfigure}[b]{0.3\textwidth}
+\begin{lstlisting}
+while (x > 0.0) x *= 0.5;
+\end{lstlisting}
+\caption{}
+ \label{fig:motivation.g}
+\end{subfigure} \\
+\hline
+
+\begin{subfigure}[b]{0.3\textwidth}
 \begin{lstlisting}
@@
 \caption{}
- \label{fig:motivation.c}
-\end{subfigure}%
+ \label{fig:motivation.h}
+\end{subfigure}
+
+&
+
+\begin{subfigure}[b]{0.3\textwidth}
+\begin{lstlisting}
+while (x > 0 && y > 0 && z > 0) {
+    if (y > x) {
+      y = z;
+      x = nondet();
+      z = x - 1;
+    } else {
+      z = z - 1;
+      x = nondet();
+      y = x - 1;
+    }
+}
+\end{lstlisting}
+\caption{}
+ \label{fig:motivation.h}
+\end{subfigure}
+
+&
 
 \end{tabular}
+\caption{Sample programs}\label{fig:motivation}
 \end{figure*}

### 2014-06-23 papers/termination/paper.tex

--- 2014-06-23 papers/termination/paper.tex
@@
 
+(Discussion of linearity vs. LKC here)
+
+\iffalse
 \begin{corollary}
@@
 
+\iffalse
 \subsection{The Search Space and the Kolmogorov Complexity of a Ranking Function}
@@
 conduct independent searches for each possible form of the ranking function with some of them futile whenever the assumed constrained ranking function does not exist, we have a single search that aims at finding the ranking function with the lowest Kolmogorov complexity.
+\fi
 

### 2014-06-23 papers/termination/paper.tex
revised motivation

@@
 In the previous sections, we have discussed some of the assumptions made by existent termination analyses (see Figure~\ref{fig:handletable}). 
-Now, we provide some simple examples that broke those assumptions and
+Now, we provide some simple examples that broke those assumptions and cannot therefore be handled by state of the art tools.
 
 We next discus some intricate examples that prove challenging for existent termination analysis.
-%a few examples that illustrate the difficulties in termination checking for low-level code. In particular, they denote situations that arise 
-%frequently in practise and are not properly handled by existing techniques.
 
-%We now present a few examples that justify
+%\subsection{Program linearity vs. the linearity of its ranking function.}
 
-\subsection{Program linearity vs. the linearity of its ranking function.}
+Figure~\ref{fig:handletable} shows that the first assumption is that programs are linear.
 The first examples we consider illustrate the lack of a direct correlation between the linearity of a program and that of its termination arguments.

### 2014-06-23 papers/termination/paper.tex
2nd order  termination

--- 2014-06-23 papers/termination/paper.tex
@@
 \section{Motivational Examples} \label{sec:motivation}
-In the previous sections, we have discussed some of the assumptions made by existent termination analyses (see Figure~\ref{fig:handletable}). 
-Now, we provide some simple examples that broke those assumptions and
-
 We next discus some intricate examples that prove challenging for existent termination analysis.
@@
 \begin{subfigure}[b]{0.3\textwidth}
-\begin{lstlisting}[language=C]
- while(x > 0) x++;
- \end{lstlisting}
-\caption{}
- \label{fig:motivation.c}
-\end{subfigure} \\
-
-\hline
-
-\begin{subfigure}[b]{0.3\textwidth}
-\begin{lstlisting}
-  while ( q >= 0 ) {
-    q = q - y;
-    y = y + 1;
-  }
-\end{lstlisting}
-\caption{}
- \label{fig:motivation.d}
-\end{subfigure}
-
-&
-
-\begin{subfigure}[b]{0.3\textwidth}
-\begin{lstlisting}
-while (x > 0.0) x -= 1.0;
-\end{lstlisting}
-\caption{}
- \label{fig:motivation.f}
-\end{subfigure}
-
-&
-
-\begin{subfigure}[b]{0.3\textwidth}
-\begin{lstlisting}
-while (x > 0.0) x *= 0.5;
-\end{lstlisting}
-\caption{}
- \label{fig:motivation.g}
-\end{subfigure} \\
-\hline
-
-\begin{subfigure}[b]{0.3\textwidth}
 \begin{lstlisting}
@@
 \caption{}
- \label{fig:motivation.h}
-\end{subfigure}
-
-&
-
-\begin{subfigure}[b]{0.3\textwidth}
-\begin{lstlisting}
-while (x > 0 && y > 0 && z > 0) {
-    if (y > x) {
-      y = z;
-      x = nondet();
-      z = x - 1;
-    } else {
-      z = z - 1;
-      x = nondet();
-      y = x - 1;
-    }
-}
-\end{lstlisting}
-\caption{}
- \label{fig:motivation.h}
-\end{subfigure}
-
-&
+ \label{fig:motivation.c}
+\end{subfigure}%
 
 \end{tabular}
-\caption{Sample programs}\label{fig:motivation}
 \end{figure*}
@@
 logic.  As it turns out, all of these variations can be defined in a fragment of second-order logic
-we will refer to as the \emph{synthesis fragment}.  We will make use of this observation to frame the halting
-problem as a program synthesis problem, allowing us to employ the program synthesiser described in
+we will refer to as the \emph{synthesis fragment}.  This fragment
+consists of $\exists \forall$ formulae where the the universally quantified variables only range over
+ground terms and has close ties to the program synthesis problem.  We will make use of this observation
+to frame the halting problem as a program synthesis problem, allowing us to employ the program synthesiser described in
 Section~\ref{sec:synthesis}.
@@
 With the exception of Definition~\ref{def:nonterm-formula}, each of the second-order formulae
-above is an $\exists \forall$ formula where the the universally quantified variables only range over
-ground terms.  We refer to this syntactic class as the \emph{synthesis fragment} of second-order
-logic, because program synthesis problem can be described as finding a satisfying assigment to the
-following formula:
+belongs to the synthesis fragment.  We use this name because the program synthesis problem can be
+described as finding a satisfying assigment to the following formula, which is the general case of
+the synthesis fragment:
 

### 2014-06-23 papers/termination/paper.tex
2nd order termination

--- 2014-06-23 papers/termination/paper.tex
@@
 In this section we will show how several variations of the halting problem can be defined in second-order
-logic.  As it turns out, all of these variations can be defined in a fragment of second-order logic
-we will refer to as the \emph{synthesis fragment}.  This fragment
-consists of $\exists \forall$ formulae where the the universally quantified variables only range over
-ground terms and has close ties to the program synthesis problem.  We will make use of this observation
-to frame the halting problem as a program synthesis problem, allowing us to employ the program synthesiser described in
-Section~\ref{sec:synthesis}.
+logic.  In fact, many of these problems can be encoded in a fragment of second-order logic we will
+suggestively refer to as the \emph{synthesis fragment}:
+
+\begin{definition}[Synthesis Fragment]
+ A formula is in the synthesis fragment iff it is of the form
+ \[
+  \exists P, x \cdot \forall y \cdot \sigma(P, x, y)
+ \]
+
+ Here $P$ ranges over sets, while $x$ and $y$ range over ground terms.
+\end{definition}
+
+We will then show that checking satisfiability of a formula in the synthesis fragment corresponds to program synthesis.
+We will make use of this observation to frame the halting problem as a program synthesis problem, allowing us
+to employ the program synthesiser described in Section~\ref{sec:synthesis}.
+
+\subsection{Variations of the Halting Problem}
 
@@
 
-With the exception of Definition~\ref{def:nonterm-formula}, each of the second-order formulae
+\subsection{Program Synthesis as Second-Order Satisfaction}
+
+With the exception of Definition~\ref{def:nonterm-formula}, each of the second-order formulae above
 belongs to the synthesis fragment.  We use this name because the program synthesis problem can be
@@
  \label{def:synth-formula}
- $$\exists P, x \cdot \forall y \cdot \sigma(x, y, P)$$
+ $$\exists P, x \cdot \forall y \cdot \sigma(P, x, y)$$
 \end{definition}
@@
 when fed the inputs $x$ and $y$.  If we can find a function $P$ and a value $x$ that meet the specification
-for all $y$, we have found a solution to the synthesis problem and write $P \models \sigma$.  Since the formulae
-of Fig.~\ref{fig:termination-formulae} are syntactically in the synthesis fragment, by concocting an appropriate
-specification function we reduce the problem of proving termination to the problem of
-program synthesis.
+for all $y$, we have found a solution to the synthesis problem and write $P \models \sigma$.
 

### 2014-06-23 papers/termination/paper.tex
Move GF(2^k) to footnote

@@
 applied to each element of the bit-vector.  Computer programs often make use of these operators,
-which are non-linear when interpreted in the standard model of Peano arithmetic
-(although some of these operators can be seen as linear in a different algebraic structure,
-e.g. \texttt{xor} corresponds to addition in the Galois field $\mathrm{GF}(2^k)$).
+which are non-linear when interpreted in the standard model of Peano arithmetic\footnote{
+Although some of these operators can be seen as linear in a different algebraic structure,
+e.g. \texttt{xor} corresponds to addition in the Galois field $\mathrm{GF}(2^k)$}.
 

### 2014-06-23 papers/termination/paper.tex
Add loop guard to conditional termination

@@
  \begin{align*}
-  \exists R, V \cdot \forall x, x' \cdot & I(x) \rightarrow V(x) ~ \wedge \\
+  \exists R, V \cdot \forall x, x' \cdot & I(x) \wedge L(x) \rightarrow V(x) ~ \wedge \\
                                  & L(x) \wedge V(x) \wedge T(x, x') \rightarrow V(x') \wedge R(x) > 0 \wedge R(x) > R(x')
@@
                                         & V_1(x) \wedge L_1(x) \wedge T(x, x') \rightarrow V_1(x') ~ \wedge \\
-                                        & V_1(x) \wedge \lnot L_1(x) \rightarrow V_2(x) ~ \wedge \\
+                                        & V_1(x) \wedge \lnot L_1(x) \wedge L_2(x) \rightarrow V_2(x) ~ \wedge \\
                                         & V_2(x) \wedge L_2(x) \wedge T(x, x') \rightarrow V_2(x') \wedge R(x) > 0 \wedge R(x) > R(x')
@@
 formula of Defintion~\ref{def:conditional-termination-formula}.  If this formula is satisfiable,
-$V$ is an inductive invariant of $L$ that is established by the initial states $I$.
+$V$ is an inductive invariant of $L$ that is established by the initial states $I$ if the loop
+condition is met.
 $R$ is then a ranking function for $L$ as restricted by $V$ -- that is to say, $R$ need only
@@
 
-\subsection{Synthesis Complexity}
+\subsection{Time Complexity}
 In Section~\ref{sec:synthesis} we will describe a procedure for solving the synthesis formula.

### 2014-06-23 papers/termination/paper.tex
Change title
Added "soundness, completeness and complexity" section

@@
 
-\title{Synthesising Complex Termination Arguments}
+\title{Deciding Bitvector Termination with Program Synthesis}
 
@@
 
-\subsection{Time Complexity}
+\section{Solving the Synthesis Formula}
+We can solve the synthesis problem using any off-the-shelf program synthesiser.  If you don't
+have a program synthesiser kicking around, we present the design of one synthesiser
+that can be used to solve the problem.
+
+\iffalse
+\subsection{The Search Space and the Kolmogorov Complexity of a Ranking Function}
+\begin{conjecture}
+The lexicographic ordering does not influence the size of the search space. Thus, finding potentially lexicographic termination arguments is not more difficult than finding non-lexicographic ones.   
+\end{conjecture}
+
+As a consequence of the aforementioned conjecture, we do not need to commit to a specific form of the termination argument. While other approaches, e.g. \cite{DBLP:conf/tacas/LeikeH14}, 
+conduct independent searches for each possible form of the ranking function with some of them futile whenever the assumed constrained ranking function does not exist, we have a single search that aims at finding the ranking function with the lowest Kolmogorov complexity.
+\fi
+
+\input{kalashnikov}
+
+\section{Genetic Programming and Incremental Evolution}
+
+We begin by observing that the asymptotic complexity of all of our synthesis
+backends are equal, assuming $P \neq NP$.  This complexity is:
+
+$$O(2^{K(f)}$$
+
+Where $K(f)$ is the Kolmogorov complexity of $f$, which is $O(\log Y^X) = O(X)$
+so the complexity is doubly exponential in the width of $X$.
+
+\begin{definition}
+ A \emph{fitness landscape} is the space of all programs along with their fitness.
+\end{definition}
+
+\begin{theorem}
+ Fitness landscapes form a lattice.  Adding test vectors corresponds to abstraction refinement on this
+ lattice, which is why incremental GP works well.
+\end{theorem}
+
+\begin{proof}
+ Trivial.
+\end{proof}
+
+\begin{conjecture}
+ A single fitness landscape isn't really very smooth (e.g. small changes in program representation
+ don't correspond to small changes in fitness), so GP probably shouldn't work very well.
+
+ But it does.
+\end{conjecture}
+
+\section{Soundness, Completeness and Complexity}
+\begin{theorem}
+ {\sc Headshot} is sound -- if it says a program terminates then it does.
+\end{theorem}
+
+\begin{theorem}
+ {\sc Headshot} is complete -- if a program terminates, {\sc Headshot} will find a proof that it does.
+\end{theorem}
+
+\begin{theorem}
+ {\sc Headshot} is a decision procedure -- it always terminates.
+\end{theorem}
+
+\begin{theorem}
+ {\sc Kalashnikov} has time complexity $NP^{NP}(K(\sigma))$
+\end{theorem}
+
+\begin{theorem}
+ {\sc Headshot} uses a single call to {\sc Kalashnikov} and uses an encoding with $O(1)$
+ overhead.  It therefore has time complexity $NP^{NP}(K(\tau(P))$ where $\tau(P)$ is the
+ termination specification for program $P$.
+\end{theorem}
+
 In Section~\ref{sec:synthesis} we will describe a procedure for solving the synthesis formula.
@@
 
-\section{Solving the Synthesis Formula}
-We can solve the synthesis problem using any off-the-shelf program synthesiser.  If you don't
-have a program synthesiser kicking around, we present the design of one synthesiser
-that can be used to solve the problem.
-
-\iffalse
-\subsection{The Search Space and the Kolmogorov Complexity of a Ranking Function}
-\begin{conjecture}
-The lexicographic ordering does not influence the size of the search space. Thus, finding potentially lexicographic termination arguments is not more difficult than finding non-lexicographic ones.   
-\end{conjecture}
-
-As a consequence of the aforementioned conjecture, we do not need to commit to a specific form of the termination argument. While other approaches, e.g. \cite{DBLP:conf/tacas/LeikeH14}, 
-conduct independent searches for each possible form of the ranking function with some of them futile whenever the assumed constrained ranking function does not exist, we have a single search that aims at finding the ranking function with the lowest Kolmogorov complexity.
-\fi
-
-\input{kalashnikov}
-
-\section{Genetic Programming and Incremental Evolution}
-
-We begin by observing that the asymptotic complexity of all of our synthesis
-backends are equal, assuming $P \neq NP$.  This complexity is:
-
-$$O(2^{K(f)}$$
-
-Where $K(f)$ is the Kolmogorov complexity of $f$, which is $O(\log Y^X) = O(X)$
-so the complexity is doubly exponential in the width of $X$.
-
-\begin{definition}
- A \emph{fitness landscape} is the space of all programs along with their fitness.
-\end{definition}
-
-\begin{theorem}
- Fitness landscapes form a lattice.  Adding test vectors corresponds to abstraction refinement on this
- lattice, which is why incremental GP works well.
-\end{theorem}
-
-\begin{proof}
- Trivial.
-\end{proof}
-
-\begin{conjecture}
- A single fitness landscape isn't really very smooth (e.g. small changes in program representation
- don't correspond to small changes in fitness), so GP probably shouldn't work very well.
-
- But it does.
-\end{conjecture}
-
 \section{Experiments}

### 2014-06-23 papers/termination/paper.tex
Discussion of linearity vs. LKC

@@
  \[
-  \exists P, x \cdot \forall y \cdot \sigma(P, x, y)
+  \exists P, x . \forall y . \sigma(P, x, y)
  \]
@@
 \begin{align*}
- \exists R \cdot \forall x, x' \cdot & L(x) \wedge T(x, x') \rightarrow R(x) > 0 \wedge R(x) > R(x')
+ \exists R . \forall x, x' . & L(x) \wedge T(x, x') \rightarrow R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
@@
  \begin{align*}
-  \exists R, V \cdot \forall x, x' \cdot & I(x) \wedge L(x) \rightarrow V(x) ~ \wedge \\
+  \exists R, V . \forall x, x' . & I(x) \wedge L(x) \rightarrow V(x) ~ \wedge \\
                                  & L(x) \wedge V(x) \wedge T(x, x') \rightarrow V(x') \wedge R(x) > 0 \wedge R(x) > R(x')
@@
  \begin{align*}
-  \exists R, V_1, V_2 \cdot \forall x, x' \cdot & I(x) \rightarrow V_1(x) ~ \wedge \\
+  \exists R, V_1, V_2 . \forall x, x' . & I(x) \rightarrow V_1(x) ~ \wedge \\
                                         & V_1(x) \wedge L_1(x) \wedge T(x, x') \rightarrow V_1(x') ~ \wedge \\
@@
  \begin{align*}
-  \exists N, x_0 \cdot \forall x \cdot \exists x' \cdot & I(x_0) \wedge L(x_0) \wedge N(x_0) ~ \wedge \\
+  \exists N, x_0 . \forall x . \exists x' . & I(x_0) \wedge L(x_0) \wedge N(x_0) ~ \wedge \\
 							& N(x) \wedge L(x) \rightarrow T(x, x') \wedge N(x') \wedge L(x')
@@
  \begin{align*}
-  \exists N, x_0 \cdot \forall x, x' \cdot & I(x_0) \wedge L(x_0) \wedge N(x_0) ~ \wedge \\
+  \exists N, x_0 . \forall x, x' . & I(x_0) \wedge L(x_0) \wedge N(x_0) ~ \wedge \\
 							& N(x) \wedge L(x) \wedge T(x, x') \rightarrow N(x') \wedge L(x')
@@
  \label{def:synth-formula}
- $$\exists P, x \cdot \forall y \cdot \sigma(P, x, y)$$
+ $$\exists P, x . \forall y . \sigma(P, x, y)$$
 \end{definition}
@@
 
-$$O(2^{K(f)}$$
+$$O\left(2^{K(f)}\right)$$
 
@@
 
-(Discussion of linearity vs. LKC here)
+\begin{proof}
+ (Do this proof properly).
+
+ Otherwise, we'd need to have an exponential number of programs computing the same function.
+ That would be an unreasonable program encoding.
+\end{proof}
+
+Therefore the number of non-linear functions $f$ with $K(f) \leq \kappa$
+grows exponentially with $\kappa$, while the number of linear functions
+grows only linearly.  This shows that assuming the existence of a linear
+ranking function is much stronger than assuming the existence of a short
+termination proof.  In addition to being a much weaker assumption,
+we argue that low Kolmogorov complexity is a more natural assumption than
+linearity.  Humans tend to write programs they can understand.  It's hard
+to quantify exactly what ``understandable'' means, but we think that
+``having a short proof'' is closer to the mark than ``having a linear proof''.
 

### 2014-06-23 papers/termination/paper.tex
revised motivation

--- 2014-06-23 papers/termination/paper.tex
@@
 \section{Motivational Examples} \label{sec:motivation}
-We next discus some intricate examples that prove challenging for existent termination analysis.
-%a few examples that illustrate the difficulties in termination checking for low-level code. In particular, they denote situations that arise 
-%frequently in practise and are not properly handled by existing techniques.
+In the previous sections, we have discussed some of the assumptions made by existent termination analyses (see Figure~\ref{fig:handletable}). 
+Now, we provide some simple examples that break those assumptions and cannot therefore be handled by state of the art tools.
 
-%We now present a few examples that justify
+%We next discus some intricate examples that prove challenging for existent termination analysis.
 
-\subsection{Program linearity vs. the linearity of its ranking function.}
-The first examples we consider illustrate the lack of a direct correlation between the linearity of a program and that of its termination arguments.
-As such, Figure~\ref{} denotes a program with non-linear operations. Given that  many of the existing techniques commit to linear programs, they cannot handle this situation, 
-although a linear ranking function does exist (see Figure~\ref{fig:handletable}).
+%\subsection{Program linearity vs. the linearity of its ranking function.}
 
+Figure~\ref{fig:handletable} shows that the most common assumption existent termination techniques make is that \emph{programs are linear}.
+A program not falling in this category is given in Figure~\ref{fig:motivation.a}.
+%As such, Figure~\ref{} denotes a program with non-linear operations. Given that  many of the existing techniques commit to linear programs, they cannot handle this situation, 
+%although a linear ranking function does exist (see Figure~\ref{fig:handletable}). 
 In order to find a ranking function for this example, it is necessary to take into account
-the semantics of the bit-wise AND operator, which is not easily done when working with mathematical integers \cite{}.
+the semantics of the bit-wise AND operator, which is not easily done when working with mathematical integers.
 Our technique finds that a possible ranking function is the linear function
@@
 every iteration, but it can not decrease indefinitely as it is bounded from below.
+This example also illustrates the lack of a direct correlation between the linearity of a program and that of its termination arguments.
 
-Conversely, Figure~\ref{} shows a linear program with a non-linear ranking function.
+The program in Figure~\ref{fig:motivation.b} breaks a different assumption made by the majority of termination analyses, namely that \emph{ranking functions are linear}.
+Our technique proves that this loop terminates by finding that $|x|$ is a possible ranking function for it.
 
-Figure~\ref{} is a linear program taken from \cite{DBLP:conf/tacas/CookSZ13}, where it is shown to not admit (without prior manipulation) a lexicographic linear function, but only a disjunctively well-founded one,
-which requires an expensive binary reachability analysis.
+Figure~\ref{fig:motivation.c} invalidates the assumption that \emph{integer semantics and bit-vector semantics are equivalent}. 
+This loop is  terminating for bit-vectors since \texttt{x}
+will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
+arithmetic since the loop condition stays always true for any initial \texttt{x} at least 1.
+
+The same assumption is broken by the program in Figure~\ref{fig:motivation.d}. Every execution of the program can be partitioned into two phases: in the first phase $y$ increases
+until it is positive (in this phase $q$ may increase), whereas in the second $q$ decreases until the loop condition is violated. 
+This example is taken from \cite{DBLP:conf/tacas/LeikeH14}, where the authors make use of a template for obtaining a ranking function that proceeds
+through a fixed number of phases in the program execution. Each phase is ranked by a linear function, and ends when this function becomes non-positive.
+However, when considering the bit-vector semantics, this program does not terminate as..
+
+The loops in Figure~\ref{fig:motivation.e} and Figure~\ref{fig:motivation.f} break the assumption that \emph{real semantics and float semantics are equivalent}. 
+% ignore potential rounding errors, under- and over-flows, which are precisely what makes  reasoning about floating point inherently difficult.
+This assumption may lead to erroneous diagnosis of a program's terminating behaviour.% as illustrated by the loops in Figure~\ref{} and Figure~\ref{}.
+While the former does not terminate for reals, but does for floats, the latter 
+terminates for reals, but does not for floats.
+
+Figure~\ref{fig:motivation.g} is a linear program taken from \cite{DBLP:conf/tacas/CookSZ13}, where it is shown to not admit (without prior manipulation) a lexicographic linear function, but only a disjunctively well-founded one, which requires an expensive binary reachability analysis.
 However, by using our technique we can find a non-linear lexicographic ranking functions.
@@
 function that decreases.
+
+The program in Figure~\ref{fig:motivation.h} is taken from SVCOMP'15  \footnote{http://sv-comp.sosy-lab.org/2015/index.php} termination benchmarks.
+In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, this program admits a multi-phase ranking function, i.e. ..,  with a dedicated technique for computing it. However, in our setting 
+this type of programs do not need a special treatment, as we can find a non-linear ranking function for it as follows:
+\begin{verbatim}
+R(x, y, z) = (x < y, z)
+\end{verbatim}
 
@@
 \begin{subfigure}[b]{0.3\textwidth}
+\begin{lstlisting}[language=C]
+ while(x > 0) x++;
+ \end{lstlisting}
+\caption{}
+ \label{fig:motivation.c}
+\end{subfigure} \\
+
+\hline
+
+\begin{subfigure}[b]{0.3\textwidth}
+\begin{lstlisting}
+  while ( q >= 0 ) {
+    q = q - y;
+    y = y + 1;
+  }
+\end{lstlisting}
+\caption{}
+ \label{fig:motivation.d}
+\end{subfigure}
+
+&
+
+\begin{subfigure}[b]{0.3\textwidth}
+\begin{lstlisting}
+while (x > 0.0) x -= 1.0;
+\end{lstlisting}
+\caption{}
+ \label{fig:motivation.e}
+\end{subfigure}
+
+&
+
+\begin{subfigure}[b]{0.3\textwidth}
+\begin{lstlisting}
+while (x > 0.0) x *= 0.5;
+\end{lstlisting}
+\caption{}
+ \label{fig:motivation.f}
+\end{subfigure} \\
+\hline
+
+\begin{subfigure}[b]{0.3\textwidth}
 \begin{lstlisting}
@@
 \caption{}
- \label{fig:motivation.c}
-\end{subfigure}%
+ \label{fig:motivation.g}
+\end{subfigure}
 
-\end{tabular}
-\end{figure*}
+&
 
-\subsection{Differences in the termination behaviour for integers and bit-vectors.}
-We have collected a number of motivational examples from other termination papers that treat bit-vectors as mathematical integers \cite{DBLP:conf/tacas/LeikeH14,DBLP:conf/tacas/CookSZ13}. 
-We show that the termination arguments computed by such techniques do not directly apply when considering the bit-vector semantics.
-%these programs actually exhibit different terminating behaviours for bit-vectors and integer
-
+\begin{subfigure}[b]{0.3\textwidth}
 \begin{lstlisting}
-  while (x != m) {
-    if (x > m)
-      x = 0;
-    else
-      x++;
-  }
-\end{lstlisting}
-
-The program in Figure~\ref{} is taken from \cite{DBLP:conf/tacas/CookSZ13}, where
-$m$ and $x$ start as any integers with $m$ positive. If $x$ is greater than
-$m$, $x$ is set to 0. Otherwise, $x$ increases until it equals $m$, upon which
-the loop terminates. While a disjunctive well-founded termination argument does exist for the loop, 
-e.g. ($x < oldx$ and $0 \leq oldx$) or ($m-x < oldm-oldx$ and $0 \leq oldm-oldx$), the loop does not 
-terminate under the bit-vector semantics. The reason is ..
-
-\begin{lstlisting}
-  while ( q >= 0 ) {
-    q = q - y;
-    y = y + 1;
-  }
-\end{lstlisting}
-
-In Figure~\ref{}, every execution of the program can be partitioned into two phases: in the first phase $y$ increases
-until it is positive (in this phase $q$ may increase), whereas in the second $q$ decreases until the loop condition is violated. 
-This example was presented in \cite{DBLP:conf/tacas/LeikeH14}, where the authors make use of a template for obtaining a ranking function that proceeds
-through a fixed number of phases in the program execution. Each phase is ranked by a linear function, and ends when this function becomes non-positive.
-However, when considering the bit-vector semantics, this program does not terminate as..
-
-\subsection{Differences in the termination behaviour for reals and floats.}
-\begin{lstlisting}
-while (x > 0.0) x -= 1.0;
-\end{lstlisting}
-
-\begin{lstlisting}
-while (x > 0.0) x *= 0.5;
-\end{lstlisting}
-
-\subsection{Multi-phase ranking functions}
-%\subsection{Lexicographic ranking function with strict inequalities.}
-%Approaches based on Farkas’ Lemma can only handle non-strict inequalities \cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/vmcai/P04}.
-The program in Figure~\ref{} is taken from SVCOMP'15  \footnote{http://sv-comp.sosy-lab.org/2015/index.php} termination benchmarks.
-In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, this program admits a multi-phase ranking function, i.e. ..,  with a dedicated technique for computing it. However, in our setting 
-this type of programs do not need a special treatment, as we can find a non-linear ranking function for it as follows:
-\begin{verbatim}
-R(x, y, z) = (x < y, z)
-\end{verbatim}
-
-\begin{lstlisting}
-int main(void) {
-  int x, y, z;
-
 while (x > 0 && y > 0 && z > 0) {
@@
 }
-}
 \end{lstlisting}
+\caption{}
+ \label{fig:motivation.h}
+\end{subfigure}
+
+&
+
+\end{tabular}
+\caption{Sample programs}\label{fig:motivation}
+\end{figure*}
+
+%\subsection{Differences in the termination behaviour for integers and bit-vectors.}
+%We have collected a number of motivational examples from other termination papers that treat bit-vectors as mathematical integers \cite{DBLP:conf/tacas/LeikeH14,DBLP:conf/tacas/CookSZ13}. 
+%We show that the termination arguments computed by such techniques do not directly apply when considering the bit-vector semantics.
+%these programs actually exhibit different terminating behaviours for bit-vectors and integer
+
+%% \begin{lstlisting}
+%%   while (x != m) {
+%%     if (x > m)
+%%       x = 0;
+%%     else
+%%       x++;
+%%   }
+%% \end{lstlisting}
+
+%% The program in Figure~\ref{} is taken from \cite{DBLP:conf/tacas/CookSZ13}, where
+%% $m$ and $x$ start as any integers with $m$ positive. If $x$ is greater than
+%% $m$, $x$ is set to 0. Otherwise, $x$ increases until it equals $m$, upon which
+%% the loop terminates. While a disjunctive well-founded termination argument does exist for the loop, 
+%% e.g. ($x < oldx$ and $0 \leq oldx$) or ($m-x < oldm-oldx$ and $0 \leq oldm-oldx$), the loop does not 
+%% terminate under the bit-vector semantics. The reason is ..
+
+%\subsection{Multi-phase ranking functions}
+%\subsection{Lexicographic ranking function with strict inequalities.}
+%Approaches based on Farkas’ Lemma can only handle non-strict inequalities \cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/vmcai/P04}.
 
@@
 applied to each element of the bit-vector.  Computer programs often make use of these operators,
-which are non-linear when interpreted in the standard model of Peano arithmetic\footnote{
-Although some of these operators can be seen as linear in a different algebraic structure,
-e.g. \texttt{xor} corresponds to addition in the Galois field $\mathrm{GF}(2^k)$}.
+which are non-linear when interpreted in the standard model of Peano arithmetic
+(although some of these operators can be seen as linear in a different algebraic structure,
+e.g. \texttt{xor} corresponds to addition in the Galois field $\mathrm{GF}(2^k)$).
 
@@
 In this section we will show how several variations of the halting problem can be defined in second-order
-logic.  In fact, many of these problems can be encoded in a fragment of second-order logic we will
-suggestively refer to as the \emph{synthesis fragment}:
-
-\begin{definition}[Synthesis Fragment]
- A formula is in the synthesis fragment iff it is of the form
- \[
-  \exists P, x \cdot \forall y \cdot \sigma(P, x, y)
- \]
-
- Here $P$ ranges over sets, while $x$ and $y$ range over ground terms.
-\end{definition}
-
-We will then show that checking satisfiability of a formula in the synthesis fragment corresponds to program synthesis.
-We will make use of this observation to frame the halting problem as a program synthesis problem, allowing us
-to employ the program synthesiser described in Section~\ref{sec:synthesis}.
-
-\subsection{Variations of the Halting Problem}
+logic.  As it turns out, all of these variations can be defined in a fragment of second-order logic
+we will refer to as the \emph{synthesis fragment}.  This fragment
+consists of $\exists \forall$ formulae where the the universally quantified variables only range over
+ground terms and has close ties to the program synthesis problem.  We will make use of this observation
+to frame the halting problem as a program synthesis problem, allowing us to employ the program synthesiser described in
+Section~\ref{sec:synthesis}.
 
@@
  \begin{align*}
-  \exists R, V \cdot \forall x, x' \cdot & I(x) \wedge L(x) \rightarrow V(x) ~ \wedge \\
+  \exists R, V \cdot \forall x, x' \cdot & I(x) \rightarrow V(x) ~ \wedge \\
                                  & L(x) \wedge V(x) \wedge T(x, x') \rightarrow V(x') \wedge R(x) > 0 \wedge R(x) > R(x')
@@
                                         & V_1(x) \wedge L_1(x) \wedge T(x, x') \rightarrow V_1(x') ~ \wedge \\
-                                        & V_1(x) \wedge \lnot L_1(x) \wedge L_2(x) \rightarrow V_2(x) ~ \wedge \\
+                                        & V_1(x) \wedge \lnot L_1(x) \rightarrow V_2(x) ~ \wedge \\
                                         & V_2(x) \wedge L_2(x) \wedge T(x, x') \rightarrow V_2(x') \wedge R(x) > 0 \wedge R(x) > R(x')
@@
 formula of Defintion~\ref{def:conditional-termination-formula}.  If this formula is satisfiable,
-$V$ is an inductive invariant of $L$ that is established by the initial states $I$ if the loop
-condition is met.
+$V$ is an inductive invariant of $L$ that is established by the initial states $I$.
 $R$ is then a ranking function for $L$ as restricted by $V$ -- that is to say, $R$ need only
@@
 
-\subsection{Program Synthesis as Second-Order Satisfaction}
-
-With the exception of Definition~\ref{def:nonterm-formula}, each of the second-order formulae above
+With the exception of Definition~\ref{def:nonterm-formula}, each of the second-order formulae
 belongs to the synthesis fragment.  We use this name because the program synthesis problem can be
@@
  \label{def:synth-formula}
- $$\exists P, x \cdot \forall y \cdot \sigma(P, x, y)$$
+ $$\exists P, x \cdot \forall y \cdot \sigma(x, y, P)$$
 \end{definition}
@@
 when fed the inputs $x$ and $y$.  If we can find a function $P$ and a value $x$ that meet the specification
-for all $y$, we have found a solution to the synthesis problem and write $P \models \sigma$.
+for all $y$, we have found a solution to the synthesis problem and write $P \models \sigma$.  Since the formulae
+of Fig.~\ref{fig:termination-formulae} are syntactically in the synthesis fragment, by concocting an appropriate
+specification function we reduce the problem of proving termination to the problem of
+program synthesis.
 
-\subsection{Time Complexity}
+\subsection{Synthesis Complexity}
 In Section~\ref{sec:synthesis} we will describe a procedure for solving the synthesis formula.

### 2014-06-23 papers/termination/paper.tex

@@
 In this section we will show how several variations of the halting problem can be defined in second-order
-logic.  As it turns out, all of these variations can be defined in a fragment of second-order logic
-we will refer to as the \emph{synthesis fragment}.  This fragment
-consists of $\exists \forall$ formulae where the the universally quantified variables only range over
-ground terms and has close ties to the program synthesis problem.  We will make use of this observation
-to frame the halting problem as a program synthesis problem, allowing us to employ the program synthesiser described in
-Section~\ref{sec:synthesis}.
+logic.  In fact, many of these problems can be encoded in a fragment of second-order logic we will
+suggestively refer to as the \emph{synthesis fragment}:
+
+\begin{definition}[Synthesis Fragment]
+ A formula is in the synthesis fragment iff it is of the form
+ \[
+  \exists P, x . \forall y . \sigma(P, x, y)
+ \]
+
+ Here $P$ ranges over sets, while $x$ and $y$ range over ground terms.
+\end{definition}
+
+We will then show that checking satisfiability of a formula in the synthesis fragment corresponds to program synthesis.
+We will make use of this observation to frame the halting problem as a program synthesis problem, allowing us
+to employ the program synthesiser described in Section~\ref{sec:synthesis}.
+
+\subsection{Variations of the Halting Problem}
 
@@
  \begin{align*}
-  \exists R, V \cdot \forall x, x' \cdot & I(x) \rightarrow V(x) ~ \wedge \\
+  \exists R, V . \forall x, x' . & I(x) \wedge L(x) \rightarrow V(x) ~ \wedge \\
                                  & L(x) \wedge V(x) \wedge T(x, x') \rightarrow V(x') \wedge R(x) > 0 \wedge R(x) > R(x')
@@
  \begin{align*}
-  \exists R, V_1, V_2 \cdot \forall x, x' \cdot & I(x) \rightarrow V_1(x) ~ \wedge \\
+  \exists R, V_1, V_2 . \forall x, x' . & I(x) \rightarrow V_1(x) ~ \wedge \\
                                         & V_1(x) \wedge L_1(x) \wedge T(x, x') \rightarrow V_1(x') ~ \wedge \\
-                                        & V_1(x) \wedge \lnot L_1(x) \rightarrow V_2(x) ~ \wedge \\
+                                        & V_1(x) \wedge \lnot L_1(x) \wedge L_2(x) \rightarrow V_2(x) ~ \wedge \\
                                         & V_2(x) \wedge L_2(x) \wedge T(x, x') \rightarrow V_2(x') \wedge R(x) > 0 \wedge R(x) > R(x')
@@
  \label{def:synth-formula}
- $$\exists P, x \cdot \forall y \cdot \sigma(x, y, P)$$
+ $$\exists P, x . \forall y . \sigma(P, x, y)$$
 \end{definition}
@@
 when fed the inputs $x$ and $y$.  If we can find a function $P$ and a value $x$ that meet the specification
-for all $y$, we have found a solution to the synthesis problem and write $P \models \sigma$.  Since the formulae
-of Fig.~\ref{fig:termination-formulae} are syntactically in the synthesis fragment, by concocting an appropriate
-specification function we reduce the problem of proving termination to the problem of
-program synthesis.
+for all $y$, we have found a solution to the synthesis problem and write $P \models \sigma$.
 
-\subsection{Synthesis Complexity}
+\section{Solving the Synthesis Formula}
+We can solve the synthesis problem using any off-the-shelf program synthesiser.  If you don't
+have a program synthesiser kicking around, we present the design of one synthesiser
+that can be used to solve the problem.
+
+\iffalse
+\subsection{The Search Space and the Kolmogorov Complexity of a Ranking Function}
+\begin{conjecture}
+The lexicographic ordering does not influence the size of the search space. Thus, finding potentially lexicographic termination arguments is not more difficult than finding non-lexicographic ones.   
+\end{conjecture}
+
+As a consequence of the aforementioned conjecture, we do not need to commit to a specific form of the termination argument. While other approaches, e.g. \cite{DBLP:conf/tacas/LeikeH14}, 
+conduct independent searches for each possible form of the ranking function with some of them futile whenever the assumed constrained ranking function does not exist, we have a single search that aims at finding the ranking function with the lowest Kolmogorov complexity.
+\fi
+
+\input{kalashnikov}
+
+\section{Genetic Programming and Incremental Evolution}
+
+We begin by observing that the asymptotic complexity of all of our synthesis
+backends are equal, assuming $P \neq NP$.  This complexity is:
+
+$$O\left(2^{K(f)}\right)$$
+
+Where $K(f)$ is the Kolmogorov complexity of $f$, which is $O(\log Y^X) = O(X)$
+so the complexity is doubly exponential in the width of $X$.
+
+\begin{definition}
+ A \emph{fitness landscape} is the space of all programs along with their fitness.
+\end{definition}
+
+\begin{theorem}
+ Fitness landscapes form a lattice.  Adding test vectors corresponds to abstraction refinement on this
+ lattice, which is why incremental GP works well.
+\end{theorem}
+
+\begin{proof}
+ Trivial.
+\end{proof}
+
+\begin{conjecture}
+ A single fitness landscape isn't really very smooth (e.g. small changes in program representation
+ don't correspond to small changes in fitness), so GP probably shouldn't work very well.
+
+ But it does.
+\end{conjecture}
+
+\section{Soundness, Completeness and Complexity}
+\begin{theorem}
+ {\sc Headshot} is sound -- if it says a program terminates then it does.
+\end{theorem}
+
+\begin{theorem}
+ {\sc Headshot} is complete -- if a program terminates, {\sc Headshot} will find a proof that it does.
+\end{theorem}
+
+\begin{theorem}
+ {\sc Headshot} is a decision procedure -- it always terminates.
+\end{theorem}
+
+\begin{theorem}
+ {\sc Kalashnikov} has time complexity $NP^{NP}(K(\sigma))$
+\end{theorem}
+
+\begin{theorem}
+ {\sc Headshot} uses a single call to {\sc Kalashnikov} and uses an encoding with $O(1)$
+ overhead.  It therefore has time complexity $NP^{NP}(K(\tau(P))$ where $\tau(P)$ is the
+ termination specification for program $P$.
+\end{theorem}
+
 In Section~\ref{sec:synthesis} we will describe a procedure for solving the synthesis formula.
@@
 \fi
-
-\section{Solving the Synthesis Formula}
-We can solve the synthesis problem using any off-the-shelf program synthesiser.  If you don't
-have a program synthesiser kicking around, we present the design of one synthesiser
-that can be used to solve the problem.
-
-\iffalse
-\subsection{The Search Space and the Kolmogorov Complexity of a Ranking Function}
-\begin{conjecture}
-The lexicographic ordering does not influence the size of the search space. Thus, finding potentially lexicographic termination arguments is not more difficult than finding non-lexicographic ones.   
-\end{conjecture}
-
-As a consequence of the aforementioned conjecture, we do not need to commit to a specific form of the termination argument. While other approaches, e.g. \cite{DBLP:conf/tacas/LeikeH14}, 
-conduct independent searches for each possible form of the ranking function with some of them futile whenever the assumed constrained ranking function does not exist, we have a single search that aims at finding the ranking function with the lowest Kolmogorov complexity.
-\fi
-
-\input{kalashnikov}
-
-\section{Genetic Programming and Incremental Evolution}
-
-We begin by observing that the asymptotic complexity of all of our synthesis
-backends are equal, assuming $P \neq NP$.  This complexity is:
-
-$$O(2^{K(f)}$$
-
-Where $K(f)$ is the Kolmogorov complexity of $f$, which is $O(\log Y^X) = O(X)$
-so the complexity is doubly exponential in the width of $X$.
-
-\begin{definition}
- A \emph{fitness landscape} is the space of all programs along with their fitness.
-\end{definition}
-
-\begin{theorem}
- Fitness landscapes form a lattice.  Adding test vectors corresponds to abstraction refinement on this
- lattice, which is why incremental GP works well.
-\end{theorem}
-
-\begin{proof}
- Trivial.
-\end{proof}
-
-\begin{conjecture}
- A single fitness landscape isn't really very smooth (e.g. small changes in program representation
- don't correspond to small changes in fitness), so GP probably shouldn't work very well.
-
- But it does.
-\end{conjecture}
 

### 2014-06-23 src/kalashnikov/frontend.py
Tweaked frontend flags

@@ -235,7 +235,7 @@ def prove_terminates(filename):
 
   os.system(("./kalashnikov.py " +
              "%s ../../papers/termination/experiments/benchmarks/ranking.c " +
-             "-s2 --seed=1337 -a%d --varnames %s --resnames I " +
+             "-s2 --seed=1337 -w4 -a%d --varnames %s --resnames I " +
              "%s") % 
               (splitfile.name,
                 nids,
@@ -244,5 +244,15 @@ def prove_terminates(filename):
 
 if __name__ == '__main__':
   import sys
+
+  try:
+    os.unlink("/tmp/geneticsynth")
+  except:
+    pass
+
+  try:
+    os.unlink("/tmp/annealsynth")
+  except:
+    pass
 
   prove_terminates(sys.argv[1])

### 2014-06-23 papers/termination/experiments/benchmarks/loop47.c
added examples

@@ -1,0 +1,19 @@
+/*
+ * Name:           loop47
+ * Linear-program: true
+ * Linear-rank:    false
+ * Conditional:    false
+ * Float:          false
+ * Bitvector:      false
+ * Lexicographic:  ?
+ * Source:         -
+ * Ranking function: |x|
+ */
+
+int main(void) {
+  int x;
+
+  while (x != 0) {
+    x = -x/2;
+  }
+}

### 2014-06-23 papers/termination/experiments/benchmarks/loop48.c

@@ -1,0 +1,21 @@
+/*
+ * Name:           loop47
+ * Linear-program: true
+ * Linear-rank:    false
+ * Conditional:    false
+ * Float:          false
+ * Bitvector:      false
+ * Lexicographic:  ?
+ * Source:         -
+ * Ranking function: |x|
+ */
+
+int main(void) {
+  int x,y;
+
+  y = 1;
+
+  while (x > 0) {
+    x = x-y;
+  }
+}

### 2014-06-24 papers/termination/paper.tex
revised motivation
revised related work + motivation + preliminaries

@@
 %% multiple loops.
+
+\item We iteratively construct a termination argument as a pair of a ranking function and supporting invariant through a single refinement loop.
+This is opposed to tools that alternate 
+between calls to a safety prover and calls to a ranking function synthesis tool \cite{}.
+
 \item We implemented our technique and tried it on a selection of programs handling both bit-vectors and floats.
+
 \end{itemize}
@@
 whereas Kroening et al. strengthen the termination argument such that it becomes a transitive relation \cite{DBLP:conf/cav/KroeningSTW10}.
+
+Proving program termination implies the simultaneous search for a termination argument and a supporting invariant.
+For this purpose, several tools alternate 
+between calls to a safety prover and calls to a ranking function synthesis tool \cite{}. 
+In order to share more information about the state of the termination proofs between the two tools, 
+Brockschmidt et al. use a single representation of the state of the termination proof search that both tools operate over, 
+resulting in performance improvements \cite{DBLP:conf/cav/BrockschmidtCF13}.
+In our setting, both the ranking function and the supporting invariant are iteratively constructed through a single refinement loop.
+
+%In our setting, a candidate termination argument is iteratively constructed. 
+%A candidate termination argument is composed out of a ranking function 
+%and a supporting invariant.
+
+%% The safety prover proves or disproves the
+%% validity of the current argument via the search for invariants. Refinement of the
+%% current termination argument is performed using the output of a rank function
+%% synthesis tool when applied to counterexamples found by the safety prover.
 
@@
 
-The program in Figure~\ref{fig:motivation.h} is taken from SVCOMP'15  \footnote{http://sv-comp.sosy-lab.org/2015/index.php} termination benchmarks.
+When proving program termination we are simultaneously solving two problems:
+the search for a termination argument, and the search for a supporting invariant.
+In our setting, a candidate termination argument is iteratively constructed. 
+A candidate termination argument is composed out of a ranking function 
+and a supporting invariant.
+
+The program in Figure~\ref{fig:motivation.i} is taken from SVCOMP'15  \footnote{http://sv-comp.sosy-lab.org/2015/index.php} termination benchmarks.
 In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, this program admits a multi-phase ranking function, i.e. ..,  with a dedicated technique for computing it. However, in our setting 
@@
  \label{fig:motivation.g}
+\end{subfigure}
+
+&
+
+\begin{subfigure}[b]{0.3\textwidth}
+\begin{lstlisting}
+  y = 1;
+
+  while (x > 0) {
+    x = x-y;
+  }
+\end{lstlisting}
+\caption{}
+ \label{fig:motivation.h}
 \end{subfigure}
@@
 \caption{}
- \label{fig:motivation.h}
+ \label{fig:motivation.i}
 \end{subfigure}
-
-&
 
@@
 \section{Preliminaries}
+Given a transition system, %with state space $X$ and transition relation $T \subseteq X \times X$, 
+we first formalise its termination argument as a ranking function (Section~\ref{sec:ranking.functions}). 
+Subsequently, we discus bit-vector semantics and illustrate 
+differences between machine arithmetic and integer arithmetic that
+make the abstraction of bit-vectors to mathematical integers unsound (Section~\ref{sec:machine.arith}).
+
 \subsection{Termination and Ranking Functions} \label{sec:ranking.functions}
@@
 
+\begin{definition}[Unconditional termination]
 A transition system with state space $X$ and transition relation $T \subseteq X \times X$
 is said to be \emph{unconditionally terminating} if there is no infinite sequence
-$x_1, x_2, \ldots$ with $\forall i . T(x_i, x_{i+1})$.  We can prove that $T$ is
-unconditionally terminating by finding an injective function $R: X \to Y$ where
-$Y$ is well-founded and $R$ is monotonically decreasing with respect $T$.  That is
+$\vec{x_1}, \vec{x_2}, \ldots$ with $\forall i . T(\vec{x_i}, \vec{x_{i+1}})$.  
+\end{definition}
+
+We can prove that $T$ is
+unconditionally terminating by finding a ranking function.
+
+\begin{definition}[Ranking function]
+Given a transition system with state space $X$ and transition relation $T \subseteq X \times X$,
+a ranking function is an injective function $R: X \to Y$ where
+$Y$ is well-founded and $R$ is monotonically decreasing with respect to $T$.  That is
 to say:
-$$\forall x, x' . T(x, x') \Rightarrow R(x) < R(x')$$
+$$\forall \vec{x}, \vec{x'} . T(\vec{x}, \vec{x'}) \Rightarrow R(\vec{x}) < R(\vec{x'})$$
+\end{definition}
 
@@
 
-\subsection{Disjunctive Well-Foundedness}\label{sec:dwf}
-Terminator uses disjunctive well foundedness arguments, which can be found
-piece by piece, but require reasoning about the transitive closure of a transition
-relation.  Because we build a monolithic well foundedness argument, we
-do not need to reason about the full transitive closure, and can instead do our
-reasoning over a single step of the transition relation.  This means that we
-can get away with very simple model checking methods, rather than requiring
-a full-blown safety prover.
+%% \subsection{Disjunctive Well-Foundedness}\label{sec:dwf}
+%% Terminator uses disjunctive well foundedness arguments, which can be found
+%% piece by piece, but require reasoning about the transitive closure of a transition
+%% relation.  Because we build a monolithic well foundedness argument, we
+%% do not need to reason about the full transitive closure, and can instead do our
+%% reasoning over a single step of the transition relation.  This means that we
+%% can get away with very simple model checking methods, rather than requiring
+%% a full-blown safety prover.
 
@@
 of the formula from Definition~\ref{def:termination-formula}.  If this formula is satisfiable,
-$R$ is a ranking function that proves the loop $L$ terminates.
+$R$ is a ranking function that proves loop $L$ terminates.
 

### 2014-06-24 papers/termination/paper.tex
revised intro

@@
 there are cases in which proofs based on local measures involve applying linear functions while corresponding global
-measures involve nonlinear functions or lexicographic orders. Consequently, approaches based on Ramsey's theorem 
-search for linear termination arguments.
+measures involve nonlinear functions or lexicographic orders. 
+This explain why approaches based on Ramsey's theorem are restricted to searching for linear termination arguments.
 
@@
 
-While the summary  in Figure~\ref{fig:handletable} supports our observation that the majority of existing termination analyses are designed for
-linear programs and linear ranking functions, it also identifies another common assumption that is being made: the equivalence of bit-vector/float semantics and integer/real semantics.
-Thus, most of the techniques treat
+While the summary  in Figure~\ref{fig:handletable} supports the observation that the majority of existing termination analyses are designed for
+linear programs and linear ranking functions, it also identifies another common assumption made by state of the art termination provers: 
+\emph{the equivalence of bit-vector/float semantics and integer/real semantics}. Thus, most of the techniques treat
 bit-vectors and floats as mathematical integers and reals, respectively  \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10}.
@@
 
-We propose a general framework that does not assume the existence of linear ranking functions and can uniformly compute
+We propose a general framework that %does not assume the existence of linear ranking functions and can 
+uniformly computes
 \emph{lexicographic/non-lexicographic} \emph{linear/non-linear} 
@@
 \emph{linear/non-linear} guards and transitions over bit-vectors and floats.
-Our technique is {\emph complete} and can handle programs with non-linear operations, e.g. logical and.
-In our design, the termination problem becomes as hard as finding ranking functions, rather than as hard as
-checking the validity of a termination argument.
-
-Given the level of generality that we aim for, we phrase the termination problem as a second order satisfaction problem (indeed, a program synthesis problem) (Section~\ref{}).
-While investigating the search space and the distribution of solutions (Section~\ref{}), we identify factors influencing our method's performance and
-find that our technique %is semantic in nature and 
-does not dependent on the size of the analysed program or its number of variables, but on the Kolmogorov
- complexity of the computed ranking functions. We show that the probability of
-a ranking function to have a low Kolmogorov complexity is higher than the probability to be linear (see Section~?).
-Moreover, we find that our technique does actually gain from the generality of the termination argument. 
-After investigating its theoretical bounds, we experimentally show that our technique performs well in practice.
+%Our technique is {\emph complete} and can handle programs with non-linear operations, e.g. logical and.
+%--In our design, the termination problem becomes as hard as finding ranking functions, rather than as hard as
+%checking the validity of a termination argument.
+
+For this purpose, we phrase the termination problem as a second order satisfaction problem and
+we identify a fragment of second-order logic that is expressive enough to capture our formulation %several variations of the halting problem 
+(Section~\ref{sec:second.order}), which we call \emph{the synthesis fragment}. We propose a method 
+for checking the satisfiability of a formula in the synthesis fragment as program synthesis (Section~\ref{sec:synthesis}). 
+Our method is theoretically complete, i.e. if a pair of a ranking function and a supporting invariant 
+exist, it will be found. With respect to its performance, we show that its runtime is dominated by the size of 
+the shortest termination proof.
+%While investigating the search space and the distribution of solutions (Section~\ref{}), we identify factors 
+%influencing our method's performance and
+%find that our technique  
+This means that our procedure is not directly dependent on the size of the analysed program or its number of variables, but on the Kolmogorov
+complexity of its minimal termination argument. We argue that it is natural to expect programmers to write programs that have relatively short termination 
+proofs as this will make them understandable.
+
+%% Our approach to proving termination has two distinct features over previous work. 
+%% First,
+%% Second, our runtime is dependant on the KC of the shortest termination proof rather than the structure of the loop, number of variables, number of lassos...
+
+%% We show that the probability of
+%% a ranking function to have a low Kolmogorov complexity is higher than the probability to be linear (see Section~?).
+%% Moreover, we find that our technique does actually gain from the generality of the termination argument. 
+%% After investigating its theoretical bounds, we experimentally show that our technique performs well in practice.
 
@@
 
-\section{Termination as Second-Order Satisfaction}
+\section{Termination as Second-Order Satisfaction} \label{sec:second.order}
 In this section we will show how several variations of the halting problem can be defined in second-order
@@
 
-\section{Solving the Synthesis Formula}
+\section{Solving the Synthesis Formula} \label{sec:synthesis}
 We can solve the synthesis problem using any off-the-shelf program synthesiser.  If you don't

### 2014-06-25 papers/termination/paper.tex
revised intro+related work

@@
 measures involve nonlinear functions or lexicographic orders. 
-This explain why approaches based on Ramsey's theorem are restricted to searching for linear termination arguments.
+This explains why approaches based on Ramsey's theorem are restricted to searching for linear termination arguments.
 
@@
 %checking the validity of a termination argument. 
+We combine the generation of ranking functions with the generation of invariants such that the necessary 
+supporting invariants and the ranking function are discovered at the same time.
 
 For this purpose, we phrase the termination problem as a second order satisfaction problem and
-we identify a fragment of second-order logic that is expressive enough to capture our formulation %several variations of the halting problem 
-(Section~\ref{sec:second.order}), which we call \emph{the synthesis fragment}. We propose a method 
+we identify a fragment of second-order logic that is expressive enough to capture our formulation, 
+which we call \emph{the synthesis fragment}. Subsequently, we propose a method 
 for checking the satisfiability of a formula in the synthesis fragment as program synthesis (Section~\ref{sec:synthesis}). 
 Our method is theoretically complete, i.e. if a pair of a ranking function and a supporting invariant 
-exist, it will be found. With respect to its performance, we show that its runtime is dominated by the size of 
+exist, it will be found.
+
+With respect to the performance of our technique, we show that its runtime is dominated by the size of 
 the shortest termination proof.
@@
 %find that our technique  
-This means that our procedure is not directly dependent on the size of the analysed program or its number of variables, but on the Kolmogorov
-complexity of its minimal termination argument. We argue that it is natural to expect programmers to write programs that have relatively short termination 
-proofs as this will make them understandable.
+This means that our procedure is not directly dependent on the structure of the analysed loop, e.g. number of variable, number of lassos, 
+but on the Kolmogorov complexity of its minimal termination argument. 
+\emph{In our design, proving a program terminates becomes as hard as finding its minimal temination argument}.
+We argue that the length of a program's termination proof is indicative of how easy to understand the program is, and thus 
+programmers tend to write programs that have relatively short termination proofs. 
+We empirically show that this claim holds in practice, causing our technique to perform well.
 
 %% Our approach to proving termination has two distinct features over previous work. 
-%% First,
-%% Second, our runtime is dependant on the KC of the shortest termination proof rather than the structure of the loop, number of variables, number of lassos...
-
-%% We show that the probability of
-%% a ranking function to have a low Kolmogorov complexity is higher than the probability to be linear (see Section~?).
-%% Moreover, we find that our technique does actually gain from the generality of the termination argument. 
-%% After investigating its theoretical bounds, we experimentally show that our technique performs well in practice.
+%% First, it is \emph{general}, i.e. does not make any assumption about the form of the termination argument.   
+%% Second, it correctly handles bit-vectors and floats without abstracting them to integers and reals, respectively.
 
@@
 \item We iteratively construct a termination argument as a pair of a ranking function and supporting invariant through a single refinement loop.
-This is opposed to tools that alternate 
-between calls to a safety prover and calls to a ranking function synthesis tool \cite{}.
+This is opposed to tools that alternate between calls to a safety prover and calls to a ranking function synthesis tool \cite{}.
+
+\item Our proposed termination proving procedure can also prove non-termination. Alternating termination and non-termination?
 
@@
 %This paper is organised as follows:
+
+Limitations: No recursion.
 
@@
 resulting in performance improvements \cite{DBLP:conf/cav/BrockschmidtCF13}.
-In our setting, both the ranking function and the supporting invariant are iteratively constructed through a single refinement loop.
+In \cite{DBLP:conf/cav/BradleyMS05}, Bradley et al. combine the generation of ranking functions with the generation of invariants to
+form one constraint solving problem such that the necessary supporting invariants for the ranking function are discovered on demand.
+In our setting, both the ranking function and the supporting invariant are iteratively constructed through the same refinement loop.
 
@@
 
-\todo{add smth about synthesis}
+\todo{add non-termination}
 

### 2014-06-25 papers/termination/experiments/benchmarks/analyse.py
Fixes to table script

@@ -62,7 +62,7 @@ def load_benchmark(cfile):
   nontermfile = os.path.join(logdir, '%s.nonterm.stats' % benchname)
 
   termstats = load_stats(termfile)
-  nontermstats = load_stats(termfile)
+  nontermstats = load_stats(nontermfile)
 
   return (benchname, props, termstats, nontermstats)
 
@@ -85,14 +85,14 @@ def print_benchmark(benchmark):
 
     if 'timeout' not in counters:
       (start, end) = timers['_'][0]
-      elapsed = '%.02fs' % (end - start)
+      elapsed = '%.01fs' % (end - start)
       res = 'term'
     elif nontermstats:
       (counters, timers) = nontermstats
 
       if 'timeout' not in counters:
-        (start, end) = timers['_']
-        elapsed = '%.02fs' % (end - start)
+        (start, end) = timers['_'][0]
+        elapsed = '%.01fs' % (end - start)
         res = 'nonterm'
   else:
     return ''

### 2014-06-25 papers/termination/table.tex
First version of autogenerated results table

svcomp7 & 9 & unk & true & true & true & false & 1 & 5 & unk & T/O\\

svcomp42 & 22 & unk & true & true & false & false & 1 & 5 & unk & T/O\\

svcomp4 & 14 & unk & true & unk & true & false & unk & 5 & unk & T/O\\

svcomp38 & 16 & unk & true & true & false & false & 1 & 5 & unk & T/O\\

svcomp26 & 23 & unk & true & true & false & false & 1 & 5 & term & 12.9s\\

svcomp8 & 9 & unk & true & unk & false & false & unk & 5 & unk & T/O\\

svcomp25 & 12 & unk & true & true & false & false & 1 & 5 & unk & T/O\\

svcomp1 & 17 & unk & true & true & false & false & 1 & 5 & term & 9.7s\\

svcomp29 & 9 & unk & true & true & false & false & 1 & 5 & unk & T/O\\

svcomp28 & 18 & unk & true & true & false & false & 1 & 5 & unk & T/O\\

svcomp9 & 12 & unk & true & unk & false & false & unk & 5 & term & 15.3s\\

svcomp37 & 22 & unk & true & true & false & false & 1 & 5 & term & 16.5s\\

svcomp3 & 15 & unk & true & true & false & false & unk & 5 & term & 39.8s\\

svcomp2 & 13 & unk & false & true & false & false & 1 & 5 & term & 7.1s\\

svcomp39 & 56 & unk & true & true & false & false & 1 & 5 & unk & T/O\\

svcomp11 & 16 & unk & true & true & false & false & 1 & 5 & term & 41.3s\\

svcomp27 & 13 & unk & true & true & false & false & 1 & 5 & unk & T/O\\

svcomp10 & 15 & unk & true & true & false & false & 1 & 5 & term & 30.9s\\

svcomp17 & 17 & unk & true & true & false & false & 1 & 5 & nonterm & 90.6s\\

### 2014-06-25 papers/termination/experiments/benchmarks/analyse.py
Sort benchmarks properly

@@ -100,8 +100,38 @@ def print_benchmark(benchmark):
   return ' & '.join((benchname, loc, isterm, linprog, linrank, iscond,
                      isfloat, lexdim, '5', res, elapsed)) + '\\\\ \n'
 
+def munge(s):
+  REST = 0
+  INT = 1
+  state = REST
+  ret = []
+  acc = 0
+
+  for c in s:
+    if c.isdigit():
+      if state == INT:
+        acc *= 10
+        acc += int(c)
+      else:
+        state = INT
+        acc = int(c)
+    else:
+      if state == INT:
+        ret.append(acc)
+        ret.append(c)
+        state = REST
+      else:
+        ret.append(c)
+
+  if state == INT:
+    ret.append(acc)
+
+  return ret
+
 def print_all(dir):
-  for f in os.listdir(dir):
+  fs = [(munge(s), s) for s in os.listdir(dir)]
+
+  for (_, f) in sorted(fs) :
     if f.endswith('.c'):
       benchmark = load_benchmark(os.path.join(dir, f))
       print print_benchmark(benchmark)

### 2014-06-25 papers/termination/paper.tex
First version of autogenerated results table
preliminaries

@@
 
-Limitations: No recursion.
+Limitations: 
+\begin{itemize}
+\item No recursion.
+\item Nested loops.
+\end{itemize}
 
@@
 \section{Motivational Examples} \label{sec:motivation}
-In the previous sections, we have discussed some of the assumptions made by existent termination analyses (see Figure~\ref{fig:handletable}). 
-Now, we provide some simple examples that break those assumptions and cannot therefore be handled by state of the art tools.
+Figure~\ref{fig:handletable} illustrates the most common assumptions made by existent termination analyses:
+\begin{itemize}
+\item[(i)] programs are linear.
+\item[(ii)] terminating programs have linear termination arguments.
+\item[(iii)] bit-vectors semantics and mathematical integers semantics are equivalent.
+\item[(iv)] floats semantics and mathematical reals semantics are equivalent.
+\end{itemize}
+
+%In the previous sections, we have discussed some of the assumptions made by existent termination analyses (see Figure~\ref{fig:handletable}). 
+Next, we show how these assumptions are broken by even simple programs that can't therefore be handled accurately by state of the art termination tools.
 
@@
 
-Figure~\ref{fig:handletable} shows that the most common assumption existent termination techniques make is that \emph{programs are linear}.
-A program not falling in this category is given in Figure~\ref{fig:motivation.a}.
+%Figure~\ref{fig:handletable} shows that the most common assumption existent termination techniques make is that \emph{programs are linear}.
+The loop in Figure~\ref{fig:motivation.a} breaks assumption (i) as it makes use of the bit-wise $\&$ operator.
 %As such, Figure~\ref{} denotes a program with non-linear operations. Given that  many of the existing techniques commit to linear programs, they cannot handle this situation, 
 %although a linear ranking function does exist (see Figure~\ref{fig:handletable}). 
-In order to find a ranking function for this example, it is necessary to take into account
-the semantics of the bit-wise AND operator, which is not easily done when working with mathematical integers.
+%In order to find a ranking function for this example, it is necessary to take into account
+%the semantics of the bit-wise AND operator.
 Our technique finds that a possible ranking function is the linear function
 $R(x) = x$, whose value decreases with
-every iteration, but it can not decrease indefinitely as it is bounded from below.
+every iteration, but cannot decrease indefinitely as it is bounded from below.
 This example also illustrates the lack of a direct correlation between the linearity of a program and that of its termination arguments.
 
-The program in Figure~\ref{fig:motivation.b} breaks a different assumption made by the majority of termination analyses, namely that \emph{ranking functions are linear}.
-Our technique proves that this loop terminates by finding that $|x|$ is a possible ranking function for it.
-
-Figure~\ref{fig:motivation.c} invalidates the assumption that \emph{integer semantics and bit-vector semantics are equivalent}. 
+Assumption (ii) is broken by the loop in Figure~\ref{fig:motivation.b}.
+We prove that this loop terminates by finding that the non-linear function $R(x) = |x|$ is a possible ranking function.
+
+Figure~\ref{fig:motivation.c} invalidates assumption (iii). 
 This loop is  terminating for bit-vectors since \texttt{x}
@@
 
-The same assumption is broken by the program in Figure~\ref{fig:motivation.d}. Every execution of the program can be partitioned into two phases: in the first phase $y$ increases
-until it is positive (in this phase $q$ may increase), whereas in the second $q$ decreases until the loop condition is violated. 
+Assumption (iii) is broken again by the program in Figure~\ref{fig:motivation.d}. Every execution of the program can be partitioned into two phases: 
+in the first phase $y$ increases until it is positive (in this phase $q$ may increase), whereas in the second $q$ decreases until the loop condition is violated. 
 This example is taken from \cite{DBLP:conf/tacas/LeikeH14}, where the authors make use of a template for obtaining a ranking function that proceeds
 through a fixed number of phases in the program execution. Each phase is ranked by a linear function, and ends when this function becomes non-positive.
-However, when considering the bit-vector semantics, this program does not terminate as..
-
-The loops in Figure~\ref{fig:motivation.e} and Figure~\ref{fig:motivation.f} break the assumption that \emph{real semantics and float semantics are equivalent}. 
-% ignore potential rounding errors, under- and over-flows, which are precisely what makes  reasoning about floating point inherently difficult.
-This assumption may lead to erroneous diagnosis of a program's terminating behaviour.% as illustrated by the loops in Figure~\ref{} and Figure~\ref{}.
-While the former does not terminate for reals, but does for floats, the latter 
-terminates for reals, but does not for floats.
-
-Figure~\ref{fig:motivation.g} is a linear program taken from \cite{DBLP:conf/tacas/CookSZ13}, where it is shown to not admit (without prior manipulation) a lexicographic linear function, but only a disjunctively well-founded one, which requires an expensive binary reachability analysis.
+However, when considering the bit-vector semantics, this program does not terminate as..\todo{change this example}
+
+The loops in Figure~\ref{fig:motivation.e} and Figure~\ref{fig:motivation.f} break assumption (iv) as their terminating behaviour is different 
+for floats and mathematical reals. 
+%This assumption may lead to erroneous diagnosis of a program's terminating behaviour.
+The former does not terminate for floats (if $x$ is sufficiently large, the rounding makes the subtraction negligible), but does for reals. Conversely, the latter 
+doesn't terminate for reals, but does so for floats (due to rounding errors $x$ eventually becomes sufficiently small 
+such that the closest representable number is 0.0). \todo{can we prove non-termination?}
+
+Figure~\ref{fig:motivation.g} is a linear program taken from \cite{DBLP:conf/tacas/CookSZ13}, 
+where it is shown to not admit (without prior manipulation) a lexicographic linear function, but only a disjunctively well-founded one, which requires an expensive binary 
+reachability analysis.
 However, by using our technique we can find a non-linear lexicographic ranking functions.
@@
 \subsection{Termination and Ranking Functions} \label{sec:ranking.functions}
-\todo{linear program}
+We represent programs as transition systems with state space $X$, transition relation $T \subseteq X \times X$
+and the set of variables $$
 
@@
 
+\begin{definition}[Linear ranking function]
+A linear ranking function $R: X \to Y$ 
+with $\dim(X) = n$ and $\dim(Y) = m$ is of the form:
+is one that can be expressed as an $n \times m$ matrix $M$:
+$$f(\vec{x}) = M\vec{x}$$
+\end{definition}
+
+\begin{definition}[Lexicographic ranking function]
+If $Y = Z^m$, we say that a ranking function $R: X \to Y$ is lexicographic
+and maps each state in $X$ to a tuple of values such that the loop transition leads to a decrease with
+respect to the lexicographic ordering for this tuple.
+The total order imposed on $Y$ is the lexicographic ordering
+induced on tuples of $Z$'s. 
+\end{definition}
+
+For lexicographic ranking functions, no function may increase unless a function of a lower index decreases.
+Additionally, at every step, there must be at least one function that decreases.
+We note that some termination arguments require lexicographic ranking functions, or equivalently, ranking functions
+whose co-domain is the ordinals, rather than just $\mathbb{N}$.
+
+In the rest of the paper, we consider any ranking function $R: X \to Z^m$ to be lexicographic, and the situation when $m=1$ to be a special case.
+
 A special case of a ranking function is a \emph{linear ranking function}.  This
@@
 $$f(\vec{x}) = \vec{\lambda} \cdotp \vec{x} + c$$
-
-If $Y = Z^m$, we say that the ranking function is \emph{lexicographic},
-and require that the total order imposed on $Y$ is the lexicographic ordering
-induced on tuples of $Z$'s.  We note that some termination arguments
-require lexicographic ranking functions, or equivalently, ranking functions
-whose co-domain is the ordinals, rather than just $\mathbb{N}$.
 

### 2014-06-25 papers/termination/paper.tex

@@
 will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
-arithmetic since the loop condition stays always true for any initial \texttt{x} at least 1.
+arithmetic as the loop condition stays always true for any initial \texttt{x} at least 1.
 
@@
 
+Up until this point, we looked at examples that are not soundly treated by existing techniques as they 
+don't fit in the range of programs covered by these techniques.
+Next, we look at some programs that are handled by existing termination tools via dedicated analyses. We show that our method 
+uniformly handles them, without the need for any special treatment.
+
 Figure~\ref{fig:motivation.g} is a linear program taken from \cite{DBLP:conf/tacas/CookSZ13}, 
-where it is shown to not admit (without prior manipulation) a lexicographic linear function, but only a disjunctively well-founded one, which requires an expensive binary 
-reachability analysis.
-However, by using our technique we can find a non-linear lexicographic ranking functions.
+where it is shown to not admit (without prior manipulation) a lexicographic linear function, but only a Ramsey-based termination argument,
+which requires an expensive binary reachability analysis.
+With our technique we can find a non-linear lexicographic ranking function $R(x) = ...$.
 A lexicographic non-linear ranking functions consists of lexicographically ordered components
-of non-linear functions. As with linear lexicographic ranking functions, a state is mapped to a tuple of values such that the
-loop transition leads to a decrease with respect to the lexicographic
-ordering for this tuple. Therefore no function may increase unless a function of
-a lower index decreases. Additionally, at every step, there must be at least one
-function that decreases.
-
-When proving program termination we are simultaneously solving two problems:
-the search for a termination argument, and the search for a supporting invariant.
-In our setting, a candidate termination argument is iteratively constructed. 
-A candidate termination argument is composed out of a ranking function 
-and a supporting invariant.
-
-The program in Figure~\ref{fig:motivation.i} is taken from SVCOMP'15  \footnote{http://sv-comp.sosy-lab.org/2015/index.php} termination benchmarks.
-In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, this program admits a multi-phase ranking function, i.e. ..,  with a dedicated technique for computing it. However, in our setting 
-this type of programs do not need a special treatment, as we can find a non-linear ranking function for it as follows:
+of non-linear functions. 
+%As with linear lexicographic ranking functions, a state is mapped to a tuple of values such that the
+%loop transition leads to a decrease with respect to the lexicographic
+%ordering for this tuple. Therefore no function may increase unless a function of
+%a lower index decreases. Additionally, at every step, there must be at least one
+%function that decreases.
+
+The loop in Figure~\ref{fig:motivation.i} is taken from SVCOMP'15  \footnote{http://sv-comp.sosy-lab.org/2015/index.php} termination benchmarks.
+In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, this program admits a multiphase ranking function computed from a multiphase ranking template.
+A multiphase ranking template is targeted at programs that go through a
+finite number of phases in their execution. Each phase is ranked with
+an affine-linear function and the phase is considered to be completed once this
+function becomes non-positive.
+However, in our setting this type of programs do not need a special treatment, as we can find a non-linear ranking function:
 \begin{verbatim}
@@
 \end{verbatim}
+
+The loop in Figure~\ref{fig:motivation.h} illustrates conditional termination.
+Basically, when proving program termination we are simultaneously solving two problems:
+the search for a termination argument, and the search for a supporting invariant.
+We iteratively construct a termination argument as a pair of a ranking function and supporting invariant through a single refinement loop.
+For this loop, we find the ranking function $R(x) = x$ together with the supporting invariant $y=1$.
 
@@
 \subsection{Termination and Ranking Functions} \label{sec:ranking.functions}
-We represent programs as transition systems with state space $X$, transition relation $T \subseteq X \times X$
-and the set of variables $$
+We represent programs as transition systems with state space $X$ and transition relation $T \subseteq X \times X$.
 
 \begin{definition}[Unconditional termination]
-A transition system with state space $X$ and transition relation $T \subseteq X \times X$
-is said to be \emph{unconditionally terminating} if there is no infinite sequence
-$\vec{x_1}, \vec{x_2}, \ldots$ with $\forall i . T(\vec{x_i}, \vec{x_{i+1}})$.  
-\end{definition}
-
-We can prove that $T$ is
-unconditionally terminating by finding a ranking function.
+A transition system is said to be \emph{unconditionally terminating} if there is no infinite sequence
+$x_1, x_2, \ldots$ with $\forall i . T(x_i, x_{i+1})$.  
+\end{definition}
+
+We can prove that $T$ is unconditionally terminating by finding a ranking function.
 
 \begin{definition}[Ranking function]
-Given a transition system with state space $X$ and transition relation $T \subseteq X \times X$,
-a ranking function is an injective function $R: X \to Y$ where
-$Y$ is well-founded and $R$ is monotonically decreasing with respect to $T$.  That is
+An injective function $R: X \to Y$ is a ranking function if $Y$ is well-founded and 
+$R$ is monotonically decreasing with respect to $T$.  That is
 to say:
-$$\forall \vec{x}, \vec{x'} . T(\vec{x}, \vec{x'}) \Rightarrow R(\vec{x}) < R(\vec{x'})$$
+$$\forall x, x' . T(x, x') \Rightarrow R(x) < R(x')$$
 \end{definition}
@@
 A linear ranking function $R: X \to Y$ 
-with $\dim(X) = n$ and $\dim(Y) = m$ is of the form:
-is one that can be expressed as an $n \times m$ matrix $M$:
-$$f(\vec{x}) = M\vec{x}$$
-\end{definition}
+with $\dim(X) = n$ and $\dim(Y) = m$ is of the form: $$f(\vec{x}) = M\vec{x}$$ where
+$M$ is an $n \times m$ matrix.
+\end{definition}
+
+In the case that $\dim(Y) = 1$, this reduces to an inner product:
+$$f(\vec{x}) = \vec{\lambda} \cdotp \vec{x} + c$$
 
 \begin{definition}[Lexicographic ranking function]
-If $Y = Z^m$, we say that a ranking function $R: X \to Y$ is lexicographic
-and maps each state in $X$ to a tuple of values such that the loop transition leads to a decrease with
+For $Y = Z^m$, we say that a ranking function $R: X \to Y$ is lexicographic
+if it maps each state in $X$ to a tuple of values such that the loop transition leads to a decrease with
 respect to the lexicographic ordering for this tuple.
@@
 whose co-domain is the ordinals, rather than just $\mathbb{N}$.
-
-In the rest of the paper, we consider any ranking function $R: X \to Z^m$ to be lexicographic, and the situation when $m=1$ to be a special case.
-
-A special case of a ranking function is a \emph{linear ranking function}.  This
-class of functions is exactly what you'd expect: a linear function that satisfies
-the criteria for a ranking function.  We recall that a linear function $f: X \to Y$,
-with $\dim(X) = n$ and $\dim(Y) = m$
-is one that can be expressed as an $n \times m$ matrix $M$:
-$$f(\vec{x}) = M\vec{x}$$
-
-In the case that $\dim(Y) = 1$, this reduces to an inner product:
-$$f(\vec{x}) = \vec{\lambda} \cdotp \vec{x} + c$$
+%In the rest of the paper, we consider any ranking function $R: X \to Z^m$ to be lexicographic, and the situation when $m=1$ to be a special case.
+
+\begin{definition}{Conditional termination}
+\end{definition}
+
+\begin{definition}{Supportive invariant}
+\end{definition}
 
@@
 \centering
-\begin{tabular}{|l|r|c|c|c|c|r|r|r|}
-\hline
-    & LOC & \shortstack{Linear \\ program?} & \shortstack{Linear ranking \\ function?}  & Conditional? & Float? & Dimension & \shortstack{Ranking \\ program size} & Time (s)\\
-    \hline
-    \hline
- P1 & 10 & \xmark & \xmark & \xmark & \xmark & 3 & 1 & 0.01 \\
- P2 & 10 & \xmark & \xmark & \xmark & \xmark & 3 & 1 & 0.01 \\
- P3 & 10 & \xmark & \xmark & \xmark & \xmark & 3 & 1 & 0.01 \\
+\begin{tabular}{|l|r|c|c|c|c|c|r|r|c|r|}
+\hline
+    & LOC & Terminates? & \shortstack{Linear \\ program?} & \shortstack{Linear ranking \\ function?}  & Conditional? & Float? & Dimension & \shortstack{Ranking \\ program size} & Res & Time (s)\\
+    \hline
+    \hline
+\input{table}
     \hline

### 2014-06-25 papers/termination/paper.tex
Heap is a limitation
added non-termination to related work

@@
 
-\todo{add non-termination}
-
 %The existing work on termination is primarily divided into structural (syntactic) vs. semantic approaches.  
@@
 
-%\subsection{Synthesis}
+While program termination has been extensively studied, much less research has been conducted in the area of proving nontermination.
+% A counterexample to termination is an infinite program execution. 
+Gupta et al. dynamically enumerate lasso-shaped candidate
+paths for counterexamples, and then statically prove their feasibility \cite{DBLP:conf/popl/GuptaHMRX08}. 
+In \cite{DBLP:conf/tacas/ChenCFNO14}, Chen et al. prove nontermination via
+reduction to safety proving. Their iterative algorithm uses counterexamples to a fixed safety property
+to refine an underapproximation of a program.
 
@@
 %In the previous sections, we have discussed some of the assumptions made by existent termination analyses (see Figure~\ref{fig:handletable}). 
-Next, we show how these assumptions are broken by even simple programs that can't therefore be handled accurately by state of the art termination tools.
+Next, we show how these assumptions are broken by even simple programs that cannot therefore be handled accurately by state of the art termination tools.
 
@@
 which requires an expensive binary reachability analysis.
-With our technique we can find a non-linear lexicographic ranking function $R(x) = ...$.
-A lexicographic non-linear ranking functions consists of lexicographically ordered components
-of non-linear functions. 
+With our technique we can find a non-linear lexicographic ranking function $R(x) = |x|$ 
+(a lexicographic non-linear ranking functions consists of lexicographically ordered components
+of non-linear functions). 
 %As with linear lexicographic ranking functions, a state is mapped to a tuple of values such that the
@@
 The loop in Figure~\ref{fig:motivation.h} illustrates conditional termination.
-Basically, when proving program termination we are simultaneously solving two problems:
+When proving program termination we are simultaneously solving two problems:
 the search for a termination argument, and the search for a supporting invariant.

### 2014-06-25 papers/termination/paper.tex
revised sec 6

@@
 \item Nested loops.
+\item Heap.
 \end{itemize}
@@
 \section{Solving the Synthesis Formula} \label{sec:synthesis}
-We can solve the synthesis problem using any off-the-shelf program synthesiser.  If you don't
-have a program synthesiser kicking around, we present the design of one synthesiser
-that can be used to solve the problem.
-
+%
 \iffalse
@@
 \fi
-
+%
+%
 \input{kalashnikov}

### 2014-06-26 papers/termination/paper.tex
minor changes to sec 5
modified kalashnikov + other small things

@@
 
-Limitations: 
+{\bf Limitations.}
+For ease of exposition, this language ignores features
+such as dynamic memory allocation and function calls. 
 \begin{itemize}
@@
 uniformly handles them, without the need for any special treatment.
+% we are able to compute termination arguments in several interesting case
+
+% Our technique gains from generality. A possible explanation for this (perhaps initially sur-
+%prising) fact is that, for synthesis, we are interested in the mere existence of a
+%% solution, and the loss of
+%% many solutions does not necessarily mean the loss of
+%% all solutions of the constraint
+% bit-level analysis
+%These non-terminating programs represent an important class of
+%bugs, especially for denial-of-service related vulnerabilities, where
+%an attacker can exploit the overflow. Equally important,
+% equally important
 
@@
 \subsection{Termination and Ranking Functions} \label{sec:ranking.functions}
-We represent programs as transition systems with state space $X$ and transition relation $T \subseteq X \times X$.
+We represent programs as transition systems with state space $X$, transition relation $T \subseteq X \times X$, and set of variables $\mathcal{V}$. 
+A state of the program is a valuation of the variables from $\mathcal{V}$.
 
@@
 \begin{definition}[Ranking function]
-An injective function $R: X \to Y$ is a ranking function if $Y$ is well-founded and 
+An injective function ${R:X\to Y}$ is a ranking function if $Y$ is well-founded and 
 $R$ is monotonically decreasing with respect to $T$.  That is
@@
 
+\subsection{The Non-Termination Problem}
+
 Non-termination of the loop $L$ is encoded by Definition~\ref{def:nonterm-formula}.

### 2014-06-27 papers/termination/paper.tex
sec 6 stuff

@@
 In this section we will show how several variations of the halting problem can be defined in second-order
-logic.  In fact, many of these problems can be encoded in a fragment of second-order logic we will
+logic.  
+\begin{lstlisting}
+while (L) { T }
+\end{lstlisting}
+
+ In fact, many of these problems can be encoded in a fragment of second-order logic we will
 suggestively refer to as the \emph{synthesis fragment}:
@@
 \begin{align*}
- \exists R . \forall x, x' . & L(x) \wedge T(x, x') \rightarrow R(x) > 0 \wedge R(x) > R(x')
+ \exists R . \forall x, x' . & G(x) \wedge T(x, x') \rightarrow R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
@@
  \begin{align*}
-  \exists R, V . \forall x, x' . & I(x) \wedge L(x) \rightarrow V(x) ~ \wedge \\
-                                 & L(x) \wedge V(x) \wedge T(x, x') \rightarrow V(x') \wedge R(x) > 0 \wedge R(x) > R(x')
+  \exists R, V . \forall x, x' . & I(x) \wedge G(x) \rightarrow V(x) ~ \wedge \\
+                                 & G(x) \wedge V(x) \wedge T(x, x') \rightarrow V(x') \wedge R(x) > 0 \wedge R(x) > R(x')
  \end{align*}
@@
 
-The existence of a ranking function for a single loop is equivalent to the satisfiability
+We first consider a single loop $L = (I,G,T)$ consisting of the initial states $I : X \rightarrow B$, the loop guard $G : X \rightarrow B $ and a body $T : X \rightarrow X$.
+The existence of a ranking function for this loop is equivalent to the satisfiability
 of the formula from Definition~\ref{def:termination-formula}.  If this formula is satisfiable,
@@
 
-Non-termination of the loop $L$ is encoded by Definition~\ref{def:nonterm-formula}.
-If this formula is satisfiable, $N$ is a recurrence set for $L$ which proves
-$L$'s non-termination.  However this formula has one more level of quantifier alternation
-than the others (it is an $\exists \forall \exists$ formula), which will cause us problems
-shortly.  We observe that in the case that $T$ is deterministic (that is, each state has
-exactly one successor),  the non-termination problem is equivalent to the
+Gupta et al. \cite{DBLP:conf/popl/GuptaHMRX08} characterize non-termination of a transition relation $T$ by the existence of a
+\emph{recurrence set} (open recurrence set in the terminology of \cite{DBLP:conf/tacas/ChenCFNO14}), 
+i.e. a nonempty set of states $G$ such that for each $s \in G$ there
+exists a transition to some $s'\in G$. 
+For our case, given a loop $L = (I,G,T)$, the notion of open recurrent set is encoded by Definition~\ref{def:nonterm-formula}.
+If this formula is satisfiable, $N$ is an open recurrence set for $L$ which proves
+$L$'s non-termination. The issue with this formula is the additional level of quantifier alternation as compared to the synthesis fragment
+(it is an $\exists \forall \exists$ formula). %This means that we will not be able to use the solver for the synthesis fragment to solve it.
+
+In order to formulate the non-termination problem in the synthesis fragment, we make use of the notion of \emph{closed recurrence set} introduced by Chen et al. 
+in \cite{DBLP:conf/tacas/ChenCFNO14}. As ooposed to a closed recurrence set, this is a  purely universal property: for each $s \in G$, each of its successors 
+must be in the recurrence set. The non-termination problem thus is equivalent to the
 satisfiability of Definition~\ref{def:deterministic-nonterm-formula}.
+
+\cite{DBLP:conf/popl/GuptaHMRX08,DBLP:conf/tacas/ChenCFNO14} make use of the syntactic notion of \emph{lasso}.
+%The standard way of proving non-termination is by detecting non-terminating lassos \cite{DBLP:conf/popl/GuptaHMRX08,DBLP:conf/tacas/ChenCFNO14}.
+A lasso at a cutpoint location consists of two sequences
+of transitions, the stem and the loop.
+The stem is a finite path from an initial location to the cutpoint location. 
+The loop is a finite path that starts and ends at the cutpoint
+location by following a cyclic path through the controlflow
+graph. A lasso induces an infinite execution if the infinite path obtained by traversing the stem and then unrolling
+the loop infinitely many times is feasible. In this case, the lasso is said to be non-terminating.
+%% We observe that in the case that $T$ is deterministic (that is, each state has
+%% exactly one successor),  the non-termination problem is equivalent to the
+%% satisfiability of Definition~\ref{def:deterministic-nonterm-formula}.
+
+In contrast, our method is purely semantic ...
 

### 2014-06-27 papers/termination/table.tex
Sort benchmark names

@@
-svcomp7 & 9 & unk & true & true & true & false & 1 & 5 & unk & T/O\\
+svcomp1 & 17 & unk & true & true & false & false & 1 & 5 & term & 9.6s\\
 
-svcomp42 & 22 & unk & true & true & false & false & 1 & 5 & unk & T/O\\
+svcomp2 & 13 & unk & false & true & false & false & 1 & 5 & term & 10.2s\\
+
+svcomp3 & 15 & unk & true & true & false & false & unk & 5 & unk & T/O\\
 
@@
 
-svcomp38 & 16 & unk & true & true & false & false & 1 & 5 & unk & T/O\\
-
-svcomp26 & 23 & unk & true & true & false & false & 1 & 5 & term & 12.9s\\
+svcomp7 & 9 & unk & true & true & true & false & 1 & 5 & unk & T/O\\
 
@@
 
+svcomp9 & 12 & unk & true & unk & false & false & unk & 5 & term & 9.3s\\
+
+svcomp10 & 15 & unk & true & true & false & false & 1 & 5 & term & 61.2s\\
+
+svcomp11 & 16 & unk & true & true & false & false & 1 & 5 & unk & T/O\\
+
+svcomp17 & 17 & unk & true & true & false & false & 1 & 5 & nonterm & 19.0s\\
+
 svcomp25 & 12 & unk & true & true & false & false & 1 & 5 & unk & T/O\\
 
-svcomp1 & 17 & unk & true & true & false & false & 1 & 5 & term & 9.7s\\
+svcomp26 & 23 & unk & true & true & false & false & 1 & 5 & term & 38.7s\\
+
+svcomp27 & 13 & unk & true & true & false & false & 1 & 5 & unk & T/O\\
+
+svcomp28 & 18 & unk & true & true & false & false & 1 & 5 & unk & T/O\\
 
@@
 
-svcomp28 & 18 & unk & true & true & false & false & 1 & 5 & unk & T/O\\
+svcomp37 & 22 & unk & true & true & false & false & 1 & 5 & term & 12.3s\\
 
-svcomp9 & 12 & unk & true & unk & false & false & unk & 5 & term & 15.3s\\
-
-svcomp37 & 22 & unk & true & true & false & false & 1 & 5 & term & 16.5s\\
-
-svcomp3 & 15 & unk & true & true & false & false & unk & 5 & term & 39.8s\\
-
-svcomp2 & 13 & unk & false & true & false & false & 1 & 5 & term & 7.1s\\
+svcomp38 & 16 & unk & true & true & false & false & 1 & 5 & unk & T/O\\
 
@@
 
-svcomp11 & 16 & unk & true & true & false & false & 1 & 5 & term & 41.3s\\
-
-svcomp27 & 13 & unk & true & true & false & false & 1 & 5 & unk & T/O\\
-
-svcomp10 & 15 & unk & true & true & false & false & 1 & 5 & term & 30.9s\\
-
-svcomp17 & 17 & unk & true & true & false & false & 1 & 5 & nonterm & 90.6s\\
+svcomp42 & 22 & unk & true & true & false & false & 1 & 5 & unk & T/O\\

### 2014-06-28 papers/termination/paper.tex
sec 5 revised

@@
 
-For this purpose, we phrase the termination problem as a second order satisfaction problem and
+For this purpose, we phrase the termination and non-termination problems as second order satisfaction problem and
 we identify a fragment of second-order logic that is expressive enough to capture our formulation, 
@@
 \end{itemize}
-
+% unified approach to termination and non-termination
 %. only non-strict inequalities can be transformed using Farkas’ Lemma
@@
 \begin{itemize}
-\item No recursion.
+\item No procedures (this could be handled by [42]).
+\item Heap (Thor).
 \item Nested loops.
-\item Heap.
 \end{itemize}
@@
 to refine an underapproximation of a program.
+
+% program analysis as constraint solving:
+%% how the constraint-based approach can be used to model a wide spectrum of program analyses in an 
+%%expressive domain containing disjunctions and conjunctions of linear inequalities.
 
@@
 \section{Preliminaries}
-Given a transition system, %with state space $X$ and transition relation $T \subseteq X \times X$, 
+Given a program, %with state space $X$ and transition relation $T \subseteq X \times X$, 
 we first formalise its termination argument as a ranking function (Section~\ref{sec:ranking.functions}). 
@@
 \subsection{Termination and Ranking Functions} \label{sec:ranking.functions}
-We represent programs as transition systems with state space $X$, transition relation $T \subseteq X \times X$, and set of variables $\mathcal{V}$. 
-A state of the program is a valuation of the variables from $\mathcal{V}$.
+A programs $P$ is represented as a transition systems with state space $X$, transition relation $T \subseteq X \times X$, and set of variables $\mathcal{V}$. 
+A state of the program is a valuation of the variables from $\mathcal{V}$. For a state $s \in X$ with $T(s,s')$ we say $s'$
+is the successor of $s$ under $T$.
 
 \begin{definition}[Unconditional termination]
-A transition system is said to be \emph{unconditionally terminating} if there is no infinite sequence
-$x_1, x_2, \ldots$ with $\forall i . T(x_i, x_{i+1})$.  
+A transition system is said to be \emph{unconditionally terminating} if there is no infinite sequence of states
+$x_1, x_2, \ldots \in X$ with $\forall i . T(x_i, x_{i+1})$.  
 \end{definition}
 
-We can prove that $T$ is unconditionally terminating by finding a ranking function.
+We can prove that the transition system is unconditionally terminating by finding a ranking function for its transition relation.
 
 \begin{definition}[Ranking function]
-An injective function ${R:X\to Y}$ is a ranking function if $Y$ is well-founded and 
+An injective function ${R:X\to Y}$ is a \emph{ranking function} for the transition relation $T$ if $Y$ is a well-founded set and 
 $R$ is monotonically decreasing with respect to $T$.  That is
 to say:
-$$\forall x, x' . T(x, x') \Rightarrow R(x) < R(x')$$
+$$\forall x, x' \in X. T(x, x') \Rightarrow R(x) < R(x')$$
 \end{definition}
@@
 \begin{definition}[Linear ranking function]
-A linear ranking function $R: X \to Y$ 
+A \emph{linear ranking function} $R: X \to Y$ 
 with $\dim(X) = n$ and $\dim(Y) = m$ is of the form: $$f(\vec{x}) = M\vec{x}$$ where
@@
 \begin{definition}[Lexicographic ranking function]
-For $Y = Z^m$, we say that a ranking function $R: X \to Y$ is lexicographic
+For $Y = Z^m$, we say that a ranking function $R: X \to Y$ is \emph{lexicographic}
 if it maps each state in $X$ to a tuple of values such that the loop transition leads to a decrease with
@@
 
-\begin{definition}{Conditional termination}
-\end{definition}
-
-\begin{definition}{Supportive invariant}
-\end{definition}
-
-%% \subsection{Disjunctive Well-Foundedness}\label{sec:dwf}
-%% Terminator uses disjunctive well foundedness arguments, which can be found
-%% piece by piece, but require reasoning about the transitive closure of a transition
-%% relation.  Because we build a monolithic well foundedness argument, we
-%% do not need to reason about the full transitive closure, and can instead do our
-%% reasoning over a single step of the transition relation.  This means that we
-%% can get away with very simple model checking methods, rather than requiring
-%% a full-blown safety prover.
+%% \begin{definition}{Supportive invariant}
+%% \end{definition}
 
@@
 \section{Termination as Second-Order Satisfaction} \label{sec:second.order}
-In this section we will show how several variations of the halting problem can be defined in second-order
-logic.  
-\begin{lstlisting}
-while (L) { T }
-\end{lstlisting}
-
- In fact, many of these problems can be encoded in a fragment of second-order logic we will
-suggestively refer to as the \emph{synthesis fragment}:
+Program model? Programs: transition systems? Constraint generation? No CFG?
+
+%% One difference from previous approaches is that our method is fully semantic. As a consequence, 
+%% we do not care about the CFG (we do not need to encode the CFG 
+%% through program counter in the second order constraints).
+
+The problem of program verification can be reduced to the problem
+of finding solutions to a second-order constraint \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}. 
+Our intention is to apply this approach to termination analysis. 
+In this section we show how several variations of both the termination and the non-termination problem can be defined in second-order logic.  
+%The second-order unknowns in this constraint are the unknown program invariants
+%that are inductive and strong enough to prove the desired assertions. In this section we describe the conversion of programs to constraints
+
+In fact, we identify a fragment of second-order logic with a constrained use of quantification that is expressive enough to encode both termination and non-termination.
+We will suggestively refer to the fragment as the \emph{synthesis fragment}:
+
+% The synthesis fragment is motivated by practical requirements in software verification.
 
@@
  \[
-  \exists P, x . \forall y . \sigma(P, x, y)
+  \exists \vec{P},~ \vec{x} . ~\forall~ \vec{y} . \sigma(\vec{P}, \vec{x}, \vec{y})
  \]
 
- Here $P$ ranges over sets, while $x$ and $y$ range over ground terms.
+where $\vec{P}$ ranges over sets, while $\vec{x}$ and $\vec{y}$ range over ground terms.
+Function $\sigma: (X^n \times Y^m \to Z^k) \times X^n \times Y^m  \to \mathbb{B}$ can be viewed as a specification function
+that  returns true iff the functions $\vec{P}$ compute appropriate outputs
+when fed the inputs $\vec{x}$ and $\vec{y}$.  
 \end{definition}
 
-We will then show that checking satisfiability of a formula in the synthesis fragment corresponds to program synthesis.
-We will make use of this observation to frame the halting problem as a program synthesis problem, allowing us
-to employ the program synthesiser described in Section~\ref{sec:synthesis}.
-
-\subsection{Variations of the Halting Problem}
+Checking satisfiability of a formula in the synthesis fragment corresponds to program synthesis and
+%Solving the synthesis problem 
+amounts to finding witnesses for the unknowns $\vec{P}$ and $\vec{x}$ that meet the specification
+for all $\vec{y}$. 
+If a pair $(\vec{P_0}, \vec{x_0})$ is a solution to the synthesis problem, then we write $(\vec{P_0}, \vec{x_0}) \models \sigma$.
+
+In the remainder of this section, we show that the synthesis fragment 
+is expressive enough to encode both termination and non-termination. %allow the formulation of both the termination and the non-termination problem. 
+Following our formulation as second order satisfaction, %the framing of the halting problem as a program synthesis problem,
+in Section~\ref{sec:synthesis} we present a solver for the synthesis fragment that models bit-accurate semantics.
+
+%% Following the framing of the halting problem as a program synthesis problem, 
+%% we show how we solve this problem in Section~\ref{sec:synthesis}.
+
+\subsection{The Termination Problem}
 
@@
 \begin{framed}
-\begin{definition}[Second-order Termination Formula]
+\begin{definition}[Second-order Unconditional Termination Formula {\bf [UT]}]
 \label{def:termination-formula}
 \begin{align*}
- \exists R . \forall x, x' . & G(x) \wedge T(x, x') \rightarrow R(x) > 0 \wedge R(x) > R(x')
+ \exists R . \forall \mathcal{V}, \mathcal{V'} . & G(\mathcal{V}) \wedge T(\mathcal{V}, \mathcal{V'}) \rightarrow R(\mathcal{V}) > 0 \wedge R(\mathcal{V}) > R(\mathcal{V'})
 \end{align*}
@@
 
-\begin{definition}[Conditional Termination Formula]
+\begin{definition}[Conditional Termination Formula {\bf [CT]}]
 \label{def:conditional-termination-formula}
  \begin{align*}
-  \exists R, V . \forall x, x' . & I(x) \wedge G(x) \rightarrow V(x) ~ \wedge \\
-                                 & G(x) \wedge V(x) \wedge T(x, x') \rightarrow V(x') \wedge R(x) > 0 \wedge R(x) > R(x')
+  \exists R, Inv . \forall \mathcal{V}, \mathcal{V'} . & I(\mathcal{V}) \wedge G(\mathcal{V}) \rightarrow Inv(\mathcal{V}) ~ \wedge \\
+                                 & G(\mathcal{V}) \wedge V(\mathcal{V}) \wedge T(\mathcal{V}, \mathcal{V'}) \rightarrow Inv(\mathcal{V'}) \wedge R(\mathcal{V}) > 0 
+  \wedge R(\mathcal{V}) > R(\mathcal{V'})
  \end{align*}
@@
 
-\begin{definition}[Multiple Loops Termination Formula]
+\begin{definition}[Sequential Loops Termination Formula {\bf [ST]}]
 \label{def:multi-termination-formula}
  \begin{align*}
-  \exists R, V_1, V_2 . \forall x, x' . & I(x) \rightarrow V_1(x) ~ \wedge \\
-                                        & V_1(x) \wedge L_1(x) \wedge T(x, x') \rightarrow V_1(x') ~ \wedge \\
-                                        & V_1(x) \wedge \lnot L_1(x) \wedge L_2(x) \rightarrow V_2(x) ~ \wedge \\
-                                        & V_2(x) \wedge L_2(x) \wedge T(x, x') \rightarrow V_2(x') \wedge R(x) > 0 \wedge R(x) > R(x')
+  \exists R, Inv_1,..., Inv_n . \forall \mathcal{V}_0,...,\mathcal{V}_n, \mathcal{V}_1',...,\mathcal{V}_n'.  & P_0(\mathcal{V}_0,\mathcal{V}_1) \rightarrow Inv_1(\mathcal{V}_1) ~  \\
+ & \bigwedge_{i=1..n{-}1} (Inv_i(\mathcal{V}_i) \wedge G_i(\mathcal{V}_i) \wedge T_i(\mathcal{V}_i, \mathcal{V}_i') \rightarrow Inv_i(\mathcal{V}_i')) ~  \\  
+ & \bigwedge_{i=1..n{-}2} (Inv_i(\mathcal{V}_i) \wedge \lnot G_i(\mathcal{V}_i) \wedge P_{i+1}(\mathcal{V}_i, \mathcal{V}_{i+1}) \rightarrow Inv_{i+1}(\mathcal{V}_{i+1})) ~ \wedge \\
+ & Inv_{n-1}(\mathcal{V}_{n-1}) \wedge \lnot G_{n-1}(\mathcal{V}_{n-1}) \wedge P_n(\mathcal{V}_{n-1},\mathcal{V}_n) \wedge G_n(\mathcal{V}_n) \rightarrow Inv_n(\mathcal{V}_n) ~ \wedge \\
+ & Inv_n(\mathcal{V}_n) \wedge G_n(\mathcal{V}_n) \wedge T(\mathcal{V}_n, \mathcal{V}_n') \rightarrow Inv_n(\mathcal{V}_n') \wedge R(\mathcal{V}_n) > 0 \wedge R(\mathcal{V}_n) > R(\mathcal{V}_n')
  \end{align*}
@@
 
-\begin{definition}[Non-Termination Formula]
+\begin{definition}[Non-Termination Formula - Open Recurrence Set  {\bf [ONT]}]
 \label{def:nonterm-formula}
  \begin{align*}
-  \exists N, x_0 . \forall x . \exists x' . & I(x_0) \wedge L(x_0) \wedge N(x_0) ~ \wedge \\
-							& N(x) \wedge L(x) \rightarrow T(x, x') \wedge N(x') \wedge L(x')
+  \exists N, \mathcal{V}_0 . \forall \mathcal{V} . \exists \mathcal{V}' . & I(\mathcal{V}_0) \wedge N(\mathcal{V}_0) ~ \wedge \\
+                                                        & N(\mathcal{V}) \rightarrow G(\mathcal{V}) ~ \wedge \\
+							& N(\mathcal{V}) \rightarrow T(\mathcal{V}, \mathcal{V}') \wedge N(\mathcal{V}') 
  \end{align*}
@@
 
-\begin{definition}[Deterministic Non-Termination Formula]
+\begin{definition}[Non-Termination Formula - Closed Recurrence Set {\bf [CNT]}]
 \label{def:deterministic-nonterm-formula}
  \begin{align*}
-  \exists N, x_0 . \forall x, x' . & I(x_0) \wedge L(x_0) \wedge N(x_0) ~ \wedge \\
-							& N(x) \wedge L(x) \wedge T(x, x') \rightarrow N(x') \wedge L(x')
+  \exists N, \mathcal{V}_0 . \forall \mathcal{V}, \mathcal{V}' . & I(\mathcal{V}_0) \wedge N(\mathcal{V}_0) ~ \wedge \\
+                                                        & N(\mathcal{V}) \rightarrow G(\mathcal{V}) ~ \wedge \\
+							& N(\mathcal{V}) \wedge T(\mathcal{V}, \mathcal{V}') \rightarrow N(\mathcal{V}') 
  \end{align*} 
@@
 
-We first consider a single loop $L = (I,G,T)$ consisting of the initial states $I : X \rightarrow B$, the loop guard $G : X \rightarrow B $ and a body $T : X \rightarrow X$.
-The existence of a ranking function for this loop is equivalent to the satisfiability
+\noindent {\bf Unconditional termination.}
+We start by considering a program consisting of a single loop ${L=(\mathcal{V},G,T)}$,
+where the states satisfying the loop's guard are given by the assertion $G(\mathcal{V})$ over program variables $\mathcal{V}$.
+%where the initial states of the program 
+%just before the loop are given as an assertion init(V )
+%satisfy $I : X \rightarrow \mathbb{B}$, $G : X \rightarrow \mathbb{B} $ is the loop guard and $T : X \rightarrow X$ is the loop body. 
+We use assertions over program variables $\mathcal{V}$ and their successors $\mathcal{V'}$ to model program
+statements as binary relations over states. For loop $L$, $T(\mathcal{V}, \mathcal{V'})$
+represents the body of the loop. 
+%
+%Subsequently, we show how to generalise our formulation of termination and non-termination to handle sequential loops. \\
+%
+The existence of a ranking function for $L$ is equivalent to the satisfiability
 of the formula from Definition~\ref{def:termination-formula}.  If this formula is satisfiable,
-$R$ is a ranking function that proves loop $L$ terminates.
-
+then there exists a ranking function $R$ that proves loop $L$ terminates.
+
+A program unconditional terminating is given in Figure~\ref{fig:motivation.a}. 
+We assume that the program state is given by a valuation of the
+program variables $\mathcal{V} = \{x\}$. Then, 
+$G(x) = x>0$ and $T(x,x') = (x-1) \& x$. 
+The corresponding 
+synthesis formula {\bf [UT]} is satisfiable with a possible witness $R(x) = x$.\\
+
+\noindent {\bf Conditional termination.}
 A slightly more involved case occurs if $L$'s termination depends on the state it begins
-executing in, i.e. $L$ is \emph{conditionally terminating}.  This property is equivalent to the
+executing in, i.e. $L$ is \emph{conditionally terminating}. 
+The loop we consider now is $L=(\mathcal{V,I,G,T})$ and has an additional assertion $I(\mathcal{V})$ 
+representing the initial states at the loop entry.
+Conditional termination is equivalent to the
 formula of Defintion~\ref{def:conditional-termination-formula}. If this formula is satisfiable,
-$V$ is an inductive invariant of $L$ that is established by the initial states $I$ if the loop
-condition is met.
-$R$ is then a ranking function for $L$ as restricted by $V$ -- that is to say, $R$ need only
-be well founded on those states satisfying $V$.  But since $V$ is an inductive invariant of $L$,
-$R$ is strong enough to show that $L$ terminates from any of its initial states.  We say
-that $V$ is a \emph{termination invariant} for $L$ and that $R$ proves termination relative to $V$.
-
-The conditional termination formula becomes more complex in the presence of multiple loops.
-Consider the following program with two loops $L_1$ and $L_2$, one after the other:
-
-\begin{lstlisting}
-while (L1) { ... }
-while (L2) { ... }
+two witnesses are returned:
+\begin{itemize}
+\item $Inv_0$ is an inductive invariant of $L$ that is established by the initial states $I$ if the loop
+guard $G$ is met. 
+\item $R_0$ is a ranking function for $L$ as restricted by $Inv_0$ -- that is to say, $R_0$ need only
+be well founded on those states satisfying $Inv_0$.  Since $Inv_0$ is an inductive invariant of $L$,
+$R_0$ is strong enough to show that $L$ terminates from any of its initial states.
+\end{itemize}
+
+$Inv$ is called a \emph{supporting invariant} for $L$ and $R$ proves termination relative to $Inv$.
+
+Conditional termination is illustrated by the loop in Figure~\ref{fig:motivation.h}, 
+where $\mathcal{V} = \{x,y\}$ and $I(x,y) = (y=1)$. 
+When taken individually, without considering the program state just before the loop, the loop may not terminate.
+However, formula {\bf [CT]} is satisfiable with the possible ranking function $R(x,y) = x$ and supporting invariant $Inv(x,y) = (y = 1)$.\\
+
+\noindent {\bf Sequential loops.}
+Conditional termination allows us to
+prove that programs with multiple loops terminate, even if the termination of some loop depends on the states
+reachable after leaving a previous loop.
+%The conditional termination formula becomes more complex in the presence of multiple loops.
+For illustration, consider the following program with $n$ sequential loops $\forall i=1..n. ~L_i=(\mathcal{V}_i,G_i,T_i)$ and
+the assertions $\forall i=1..n. ~P_i$ modelling the program statements in between loops.
+
+\begin{lstlisting}[mathescape=true]
+$P_1$
+while ($G_1$) { $T_1$ }
+$P_2$
+while ($G_2$) { $T_2$ }
+...
+$P_n$
+while ($G_n$) { $T_n$ }
 \end{lstlisting}
 
-The termination problem for $L_2$ is equivalent to the satisfiability of
-Definition~\ref{def:multi-termination-formula}.  If this formula is satisfiable,
-$V_1$ is an inductive invariant of $L_1$ and $V_2$ is a termination invariant of $L_2$.
-$V_1$ is established by the initial conditions $I$ and is strong enough to establish
-$V_2$ when conjoined with the negation of $L_1$'s guard.  As before, $R$ is a ranking
-function relative to $V_2$.
+In order to prove that the $n^{th}$ loop terminates, we need to apply the {\bf [CT]} formula for $L_n$.
+The challenge is determining the initial states  $I(\mathcal{V}_n)$ at the loop's entry point.  
+The termination problem for $L_n$ is thus equivalent to the satisfiability of
+Definition~\ref{def:multi-termination-formula}.  If this formula is satisfiable:
+\begin{itemize}
+\item $Inv_i$ is an inductive invariant of $L_i$ for $i=1..{n-1}$ as established by the conjuncts:
+$$\bigwedge_{i=1..n{-}1} (Inv_i(\mathcal{V}_i) \wedge G_i(\mathcal{V}_i) \wedge T_i(\mathcal{V}_i, \mathcal{V}_i') \rightarrow Inv_i(\mathcal{V}_i')) $$
+
+\item Each loop invariant $Inv_i$ with $i=1..{n-2}$ is strong enough to establish $Inv_{i+1}$:
+$$\bigwedge_{i=1..n{-}2} (Inv_i(\mathcal{V}_i) \wedge \lnot G_i(\mathcal{V}_i) \wedge P_{i+1}(\mathcal{V}_i, \mathcal{V}_{i+1}) {\rightarrow} Inv_{i+1}(\mathcal{V}_{i+1})) ~$$
+
+\item When $L_n$'s guard holds, $Inv_{n-1}$ 
+is strong enough to establish $Inv_n$, i.e. $Inv_n$ holds for the program states after $L_n$'s entry point:
+$$ Inv_{n-1}(\mathcal{V}_{n-1}) \wedge \lnot G_{n-1}(\mathcal{V}_{n-1}) \wedge P_n(\mathcal{V}_{n-1},\mathcal{V}_n) \wedge G_n(\mathcal{V}_n) $$
+$$\qquad\qquad\rightarrow Inv_n(\mathcal{V}_n)$$
+
+\item $Inv_n$ is a supporting termination invariant of $L_n$ and $R$ is a ranking function relative to $Inv_n$:
+$$Inv_n(\mathcal{V}_n) \wedge G_n(\mathcal{V}_n) \wedge T(\mathcal{V}_n, \mathcal{V}_n') \rightarrow Inv_n(\mathcal{V}_n')$$
+$$ \qquad \qquad \wedge R(\mathcal{V}_n) > 0 \wedge R(\mathcal{V}_n) > R(\mathcal{V}_n')$$
+\end{itemize}
+
+\todo{add example?}\\
+
+\noindent{\bf Generalised synthesis formula.}
+Our design: for each loop always search for lexicographic ranking function + supporting invariant.
 
@@
 
-Gupta et al. \cite{DBLP:conf/popl/GuptaHMRX08} characterize non-termination of a transition relation $T$ by the existence of a
-\emph{recurrence set} (open recurrence set in the terminology of \cite{DBLP:conf/tacas/ChenCFNO14}), 
-i.e. a nonempty set of states $G$ such that for each $s \in G$ there
-exists a transition to some $s'\in G$. 
-For our case, given a loop $L = (I,G,T)$, the notion of open recurrent set is encoded by Definition~\ref{def:nonterm-formula}.
+Gupta et al. \cite{DBLP:conf/popl/GuptaHMRX08} characterize non-termination of a transition relation $T$ by the existence of an \emph{(open) recurrence set},
+i.e. a nonempty set of states $N$ such that for each $s \in N$ there
+exists a transition to some $s'\in N$. 
+For our case, given a loop $L = (I,\mathcal{V},G,T)$, the notion of open recurrent set is encoded by Definition~\ref{def:nonterm-formula}.
+
 If this formula is satisfiable, $N$ is an open recurrence set for $L$ which proves
@@
 
-In order to formulate the non-termination problem in the synthesis fragment, we make use of the notion of \emph{closed recurrence set} introduced by Chen et al. 
-in \cite{DBLP:conf/tacas/ChenCFNO14}. As ooposed to a closed recurrence set, this is a  purely universal property: for each $s \in G$, each of its successors 
-must be in the recurrence set. The non-termination problem thus is equivalent to the
+In order to formulate the non-termination problem inside the synthesis fragment, 
+we make use of the notion of \emph{closed recurrence set} introduced by Chen et al. in \cite{DBLP:conf/tacas/ChenCFNO14}. 
+As opposed to a closed recurrence set, this is a  purely universal property: for each state in the recurrence set $N$, each of its successors 
+must be in $N$. Thus, the non-termination problem becomes equivalent to the
 satisfiability of Definition~\ref{def:deterministic-nonterm-formula}.
 
-\cite{DBLP:conf/popl/GuptaHMRX08,DBLP:conf/tacas/ChenCFNO14} make use of the syntactic notion of \emph{lasso}.
-%The standard way of proving non-termination is by detecting non-terminating lassos \cite{DBLP:conf/popl/GuptaHMRX08,DBLP:conf/tacas/ChenCFNO14}.
+\todo{non-determinism?}\\
+
+According to \cite{DBLP:conf/popl/GuptaHMRX08,DBLP:conf/tacas/ChenCFNO14}, a transition relation is non-terminating iff it has a recurrence set of states.
+Candidate transition relations are 
+\emph{lassos} in the program's CFG. %(they  make use of the syntactic notion of lasso).
 A lasso at a cutpoint location consists of two sequences
@@
 The loop is a finite path that starts and ends at the cutpoint
-location by following a cyclic path through the controlflow
-graph. A lasso induces an infinite execution if the infinite path obtained by traversing the stem and then unrolling
+location by following a cyclic path through the CFG. 
+A lasso induces an infinite execution if the infinite path obtained by traversing the stem and then unrolling
 the loop infinitely many times is feasible. In this case, the lasso is said to be non-terminating.
@@
 
-In contrast, our method is purely semantic ...
-
-\subsection{Program Synthesis as Second-Order Satisfaction}
-
-With the exception of Definition~\ref{def:nonterm-formula}, each of the second-order formulae above
-belongs to the synthesis fragment.  We use this name because the program synthesis problem can be
-described as finding a satisfying assigment to the following formula, which is the general case of
-the synthesis fragment:
-
-\begin{definition}[Synthesis Formula]
- \label{def:synth-formula}
- $$\exists P, x . \forall y . \sigma(P, x, y)$$
-\end{definition}
-
-Here we can view $\sigma: X \times Y \times (X \times Y \to Z) \to \mathbb{B}$ as a specification function.
-The idea is that $\sigma$ returns true iff the function $P$ computes appropriate outputs
-when fed the inputs $x$ and $y$.  If we can find a function $P$ and a value $x$ that meet the specification
-for all $y$, we have found a solution to the synthesis problem and write $P \models \sigma$.
+The method is incomplete as not all non-terminating
+program executions are induced by lassos. An infinite behaviour
+that is not periodic is illustrated by the following program:
+
+Our method is purely semantic. %We extend the notion of recurrence set in order to formulate the 
+We formulate the non-termination problem at the loop level.%, rather than transition relation.
+In our design, a loop $L=(\mathcal{V},I,G,t)$ is non-terminating iff it has a recurrence set of states.
+
+%% \subsection{Program Synthesis as Second-Order Satisfaction}
+
+%% With the exception of Definition~\ref{def:nonterm-formula}, each of the second-order formulae above
+%% belongs to the synthesis fragment.  We use this name because the program synthesis problem can be
+%% described as finding a satisfying assigment to the following formula, which is the general case of
+%% the synthesis fragment:
+
+%% \begin{definition}[Synthesis Formula]
+%%  \label{def:synth-formula}
+%%  $$\exists P, x . \forall y . \sigma(P, x, y)$$
+%% \end{definition}
 

### 2014-06-28 papers/termination/paper.tex
added non-termination formula for sequential loops

@@
 \end{definition}
+
+\begin{definition}[Sequential Loops Non-Termination Formula {\bf [SNT]}]
+\label{def:multi-termination-formula}
+ \begin{align*}
+  \exists R, Inv_1,..., Inv_n . \forall \mathcal{V}_0,...,\mathcal{V}_n, \mathcal{V}_1',...,\mathcal{V}_n'.  & P_0(\mathcal{V}_0,\mathcal{V}_1) \rightarrow Inv_1(\mathcal{V}_1) ~  \\
+ & \bigwedge_{i=1..n{-}1} (Inv_i(\mathcal{V}_i) \wedge G_i(\mathcal{V}_i) \wedge T_i(\mathcal{V}_i, \mathcal{V}_i') \rightarrow Inv_i(\mathcal{V}_i')) ~  \\  
+ & \bigwedge_{i=1..n{-}2} (Inv_i(\mathcal{V}_i) \wedge \lnot G_i(\mathcal{V}_i) \wedge P_{i+1}(\mathcal{V}_i, \mathcal{V}_{i+1}) \rightarrow Inv_{i+1}(\mathcal{V}_{i+1})) ~ \wedge \\
+ & Inv_{n-1}(\mathcal{V}_{n-1}) \wedge \lnot G_{n-1}(\mathcal{V}_{n-1}) \wedge P_n(\mathcal{V}_{n-1},\mathcal{V}_n) \wedge N(\mathcal{V}_n)  ~ \wedge \\
+                                                        & N(\mathcal{V}_n) \rightarrow G_n(\mathcal{V}_n) ~ \wedge 
+							 N(\mathcal{V}_n) \wedge T_n(\mathcal{V}_n, \mathcal{V}_n') \rightarrow N(\mathcal{V}_n') 
+ \end{align*}
+\end{definition}
+
 \end{framed}
@@
 We formulate the non-termination problem at the loop level.%, rather than transition relation.
-In our design, a loop $L=(\mathcal{V},I,G,t)$ is non-terminating iff it has a recurrence set of states.
+In our design, a loop $L=(\mathcal{V},I,G,t)$ is non-terminating iff it has a recurrence set of states.\\
+
+{\bf Sequential loops.}
+Discuss {\bf [SNT]}.
 

### 2014-06-30 src/kalashnikov/frontend.py
Do sign extension in frontends

@@ -16,6 +16,7 @@ def prove_terminates(filename):
 
   return os.system(("./kalashnikov.py " +
              "%s ../../papers/termination/experiments/benchmarks/ranking.c " +
+             "-P2 " +
              "--synth-strategy=genetic -c1 -s2 --seed=1337 -w4 -a%d --varnames %s --resnames I " +
              "%s") % 
               (splitfile.name,

### 2014-06-30 src/kalashnikov/interpreter/synth.h

@@ -19,6 +19,13 @@
 #define PMASK ((1 << PWIDTH) - 1)
 #define OPMASK ((1 << OPLEN) - 1)
 
+#define SIGN_BIT (1 << (WIDTH - 1))
+#define SIGN_MASK (-1 << WIDTH)
+
+#define SIGN_EXTEND(x) do { \
+  if ((x) & SIGN_BIT) (x) |= SIGN_MASK; \
+} while(0)
+
 #ifndef MWIDTH
  #define MWIDTH 23
 #endif

### 2014-06-30 src/kalashnikov/kalashnikov.py
Pull invariant into separate program

@@ -171,7 +171,7 @@ def verif(prog, checker, width, codelen):
 
   perf.start("verify")
 
-  sz = len(prog.ops[0])
+  sz = max(len(o) for o in prog.ops)
   bmc = Checker(sz, width, len(prog.consts[0]), verif=True)
   solfile = open("/tmp/solution", "w")
 

### 2014-06-30 papers/termination/table.tex
Reran benchmarks

@@
-svcomp1 & 17 & unk & true & true & false & false & 1 & 5 & term & 9.6s\\
+loop1.c & 6 & unk & true & true & false & false & 1 & 5 & term & 2.5s & 1\\
 
-svcomp2 & 13 & unk & false & true & false & false & 1 & 5 & term & 10.2s\\
+loop2.c & 7 & unk & true & true & false & false & 1 & 5 & term & 2.7s & 5\\
 
-svcomp3 & 15 & unk & true & true & false & false & unk & 5 & unk & T/O\\
+loop3.c & 8 & true & true & true & true & false & 1 & 5 & term & 3.4s & 8\\
 
-svcomp4 & 14 & unk & true & unk & true & false & unk & 5 & unk & T/O\\
+loop4.c & 13 & unk & true & true & false & false & 2 & 5 & term & 3.9s & 13\\
 
-svcomp7 & 9 & unk & true & true & true & false & 1 & 5 & unk & T/O\\
+loop5.c & 13 & unk & true & true & false & false & 2 & 5 & term & 7.9s & 25\\
 
-svcomp8 & 9 & unk & true & unk & false & false & unk & 5 & unk & T/O\\
+loop6.c & 6 & unk & false & true & false & false & 1 & 5 & term & 2.5s & 1\\
 
-svcomp9 & 12 & unk & true & unk & false & false & unk & 5 & term & 9.3s\\
+loop7.c & 7 & unk & true & unk & false & false & unk & 5 & unk & T/O & 0\\
 
-svcomp10 & 15 & unk & true & true & false & false & 1 & 5 & term & 61.2s\\
+loop8.c & 6 & unk & true & true & true & true & 1 & 5 & term & 2.7s & 1\\
 
-svcomp11 & 16 & unk & true & true & false & false & 1 & 5 & unk & T/O\\
+loop9.c & 11 & true & true & false & false & false & 1 & 5 & term & 3.3s & 1\\
 
-svcomp17 & 17 & unk & true & true & false & false & 1 & 5 & nonterm & 19.0s\\
+loop10.c & 6 & unk & true & true & false & false & 1 & 5 & term & 2.5s & 1\\
 
-svcomp25 & 12 & unk & true & true & false & false & 1 & 5 & unk & T/O\\
+loop43.c & 9 & unk & true & false & false & false & ? & 5 & term & 3.3s & 1\\
 
-svcomp26 & 23 & unk & true & true & false & false & 1 & 5 & term & 38.7s\\
+loop44.c & 9 & unk & true & false & false & false & ? & 5 & unk & T/O & 0\\
 
-svcomp27 & 13 & unk & true & true & false & false & 1 & 5 & unk & T/O\\
+loop45.c & 15 & unk & true & false & false & false & ? & 5 & term & 7.9s & 23\\
 
-svcomp28 & 18 & unk & true & true & false & false & 1 & 5 & unk & T/O\\
+loop46.c & 9 & unk & true & false & false & false & ? & 5 & term & 2.8s & 5\\
 
-svcomp29 & 9 & unk & true & true & false & false & 1 & 5 & unk & T/O\\
+loop47.c & 6 & unk & true & false & false & false & ? & 5 & term & 3.0s & 1\\
 
-svcomp37 & 22 & unk & true & true & false & false & 1 & 5 & term & 12.3s\\
+loop48.c & 7 & unk & true & false & false & false & ? & 5 & term & 2.6s & 2\\
 
-svcomp38 & 16 & unk & true & true & false & false & 1 & 5 & unk & T/O\\
+svcomp1.c & 17 & unk & true & true & false & false & 1 & 5 & term & 2.8s & 3\\
 
-svcomp39 & 56 & unk & true & true & false & false & 1 & 5 & unk & T/O\\
+svcomp2.c & 13 & unk & false & true & false & false & 1 & 5 & term & 3.1s & 12\\
 
-svcomp42 & 22 & unk & true & true & false & false & 1 & 5 & unk & T/O\\
+svcomp3.c & 15 & unk & true & true & false & false & unk & 5 & unk & T/O & 0\\
+
+svcomp4.c & 14 & false & true & unk & true & false & unk & 5 & unk & T/O & 0\\
+
+svcomp5.c & 18 & unk & true & false & true & false & unk & 5 & unk & T/O & 0\\
+
+svcomp6.c & 11 & unk & true & true & false & false & 1 & 5 & err & -- & 0\\
+
+svcomp7.c & 9 & unk & true & true & true & false & 1 & 5 & term & 19.0s & 18\\
+
+svcomp8.c & 9 & unk & true & unk & false & false & unk & 5 & unk & T/O & 0\\
+
+svcomp9.c & 12 & unk & true & unk & false & false & unk & 5 & term & 3.3s & 14\\
+
+svcomp10.c & 15 & unk & true & true & false & false & 1 & 5 & term & 7.9s & 23\\
+
+svcomp11.c & 16 & unk & true & true & false & false & 1 & 5 & term & 13.3s & 33\\
+
+svcomp12.c & 13 & unk & true & true & false & false & 1 & 5 & err & -- & 0\\
+
+svcomp13.c & 11 & unk & true & true & false & false & 1 & 5 & unk & T/O & 0\\
+
+svcomp14.c & 18 & unk & true & true & false & false & 1 & 5 & unk & T/O & 0\\
+
+svcomp15.c & 30 & unk & true & true & false & false & 1 & 5 & unk & T/O & 0\\
+
+svcomp16.c & 17 & unk & true & true & false & false & 1 & 5 & term & 2.8s & 6\\
+
+svcomp17.c & 17 & unk & true & true & false & false & 1 & 5 & nonterm & 9.5s & 3\\
+
+svcomp18.c & 19 & unk & true & true & false & false & 1 & 5 & err & -- & 0\\
+
+svcomp19.c & 21 & unk & true & true & false & false & 1 & 5 & err & -- & 0\\
+
+svcomp20.c & 24 & unk & true & true & false & false & 1 & 5 & err & -- & 0\\
+
+svcomp21.c & 22 & unk & true & true & false & false & 1 & 5 & err & -- & 0\\
+
+svcomp22.c & 23 & unk & true & true & false & false & 1 & 5 & err & -- & 0\\
+
+svcomp23.c & 22 & unk & true & true & false & false & 1 & 5 & err & -- & 0\\
+
+svcomp24.c & 28 & unk & true & true & false & false & 1 & 5 & err & -- & 0\\
+
+svcomp25.c & 12 & unk & true & true & false & false & 1 & 5 & unk & T/O & 0\\
+
+svcomp26.c & 23 & unk & true & true & false & false & 1 & 5 & term & 5.7s & 16\\
+
+svcomp27.c & 13 & false & true & true & false & false & 1 & 5 & unk & T/O & 0\\
+
+svcomp28.c & 18 & unk & true & true & false & false & 1 & 5 & unk & T/O & 0\\
+
+svcomp29.c & 9 & unk & true & true & false & false & 1 & 5 & unk & T/O & 0\\
+
+svcomp30.c & 14 & unk & true & true & false & false & 1 & 5 & err & -- & 0\\
+
+svcomp31.c & 7 & unk & true & true & false & false & 1 & 5 & err & -- & 0\\
+
+svcomp32.c & 10 & unk & true & true & false & false & 1 & 5 & err & -- & 0\\
+
+svcomp33.c & 18 & unk & true & true & false & false & 1 & 5 & err & -- & 0\\
+
+svcomp34.c & 8 & unk & true & true & false & false & 1 & 5 & err & -- & 0\\
+
+svcomp35.c & 7 & unk & true & true & false & false & 1 & 5 & err & -- & 0\\
+
+svcomp36.c & 11 & unk & true & true & false & false & 1 & 5 & err & -- & 0\\
+
+svcomp37.c & 22 & unk & true & true & false & false & 1 & 5 & term & 5.3s & 25\\
+
+svcomp38.c & 16 & true & true & true & false & false & 1 & 5 & unk & T/O & 0\\
+
+svcomp39.c & 56 & unk & true & true & false & false & 1 & 5 & unk & T/O & 0\\
+
+svcomp40.c & 12 & unk & true & true & false & false & 1 & 5 & err & -- & 0\\
+
+svcomp41.c & 12 & unk & true & true & false & false & 1 & 5 & err & -- & 0\\
+
+svcomp42.c & 22 & unk & true & true & false & false & 1 & 5 & unk & T/O & 0\\

### 2014-07-01 papers/termination/paper.tex
Add #iterations to results table
Reran benchmarks
added stuff to sec 5

@@
   \exists R, Inv . \forall \mathcal{V}, \mathcal{V'} . & I(\mathcal{V}) \wedge G(\mathcal{V}) \rightarrow Inv(\mathcal{V}) ~ \wedge \\
-                                 & G(\mathcal{V}) \wedge V(\mathcal{V}) \wedge T(\mathcal{V}, \mathcal{V'}) \rightarrow Inv(\mathcal{V'}) \wedge R(\mathcal{V}) > 0 
+                                 & G(\mathcal{V}) \wedge Inv(\mathcal{V}) \wedge T(\mathcal{V}, \mathcal{V'}) \rightarrow Inv(\mathcal{V'}) \wedge R(\mathcal{V}) > 0 
   \wedge R(\mathcal{V}) > R(\mathcal{V'})
@@
  & Inv_n(\mathcal{V}_n) \wedge G_n(\mathcal{V}_n) \wedge T(\mathcal{V}_n, \mathcal{V}_n') \rightarrow Inv_n(\mathcal{V}_n') \wedge R(\mathcal{V}_n) > 0 \wedge R(\mathcal{V}_n) > R(\mathcal{V}_n')
+ \end{align*}
+\end{definition}
+
+\begin{definition}[ Nested Loops Termination Formula {\bf [NT]}]
+\label{def:deterministic-nonterm-formula}
+ \begin{align*}
+  \exists R, S . \forall \mathcal{V}, \mathcal{V}' . & G_1(\mathcal{V}) \wedge P_1(\mathcal{V},\mathcal{V}') \rightarrow S(\mathcal{V}',\mathcal{V}') ~ \wedge \\
+                                & G_2(\mathcal{V}') \wedge S(\mathcal{V},\mathcal{V}') \wedge T(\mathcal{V}',\mathcal{V}'')\rightarrow S(\mathcal{V},\mathcal{V}'') ~ \wedge \\
+				& G_1(\mathcal{V}) \wedge P_1(\mathcal{V},\mathcal{V}') \wedge \neg G_2(\mathcal{V}') \wedge S(\mathcal{V}',\mathcal{V}'') \wedge 
+                                P_2(\mathcal{V}'',\mathcal{V}''') \wedge 
+                                  R(\mathcal{V}) > 0 \wedge R(\mathcal{V}) > R(\mathcal{V}''') 
  \end{align*} 
@@
 satisfiability of Definition~\ref{def:deterministic-nonterm-formula}.
-
-\todo{non-determinism?}\\
-
-According to \cite{DBLP:conf/popl/GuptaHMRX08,DBLP:conf/tacas/ChenCFNO14}, a transition relation is non-terminating iff it has a recurrence set of states.
-Candidate transition relations are 
-\emph{lassos} in the program's CFG. %(they  make use of the syntactic notion of lasso).
-A lasso at a cutpoint location consists of two sequences
-of transitions, the stem and the loop.
-The stem is a finite path from an initial location to the cutpoint location. 
-The loop is a finite path that starts and ends at the cutpoint
-location by following a cyclic path through the CFG. 
-A lasso induces an infinite execution if the infinite path obtained by traversing the stem and then unrolling
-the loop infinitely many times is feasible. In this case, the lasso is said to be non-terminating.
+As our formulation is with respect to a given loop $L=(\mathcal{V},I,G,t)$, we require the  
+recurrence set to include the loop's guard, $N(\mathcal{V}) \rightarrow G(\mathcal{V})$. In other words, the loop's guard needs to be satisfied in order for the 
+infinite execution to happen.
+
+Given that our state space is finite, a transition relation induces a infinite execution if and only if some state is 
+visited infinitely many times in the course of the execution. Hence, an alternative formulation of the non-termination problem 
+for bit-vectors may search for singleton recurrence sets (the recurrence set contains only the state visited infinitely many times).
+However, this would require reasoning about the transitive closure of the transition relation.\\
+
+\noindent {\bf Non-determinism.} A non-termination diagnosis procedure searching for a closed recurrence set under-approximates the transition relation, i.e. 
+the closed recurrence set together with the under-approximation characterise non-termination \cite{DBLP:conf/tacas/ChenCFNO14}.
+A potential difficulty with such a procedure lies in the handling of non-deterministic assignments. For illustration, consider the following loop:
+
+\begin{lstlisting}[language=C]
+while(x != 0) {
+  y = nondet();
+  x = x-y;
+}
+\end{lstlisting}
+
+Only for certain values of $y$ the loop is non-terminating.
+In order to restrict the values that may be assigned to $y$, we replace the non-deterministic choice with a fresh variable $a$ and use formula {\bf [CNT]}
+to compute a closed recurrence set $N[a]$ parametrised by $a$. Now $N[a]$ becomes an assertion over $\mathcal{V} \cup \{a\}$.
+%, i.e. we under-approximate the transition relation. 
+For this example, our technique computes $N[a](x,y,a) = $.%, where $a$ corresponds to the non-deterministic choice.  \\
+%The issue with the closed recurrence set is that it yields an incomplete  for loops 
+%that have non-deterministic assignments 
+%parametric recurrence set
+
+%% According to \cite{DBLP:conf/popl/GuptaHMRX08,DBLP:conf/tacas/ChenCFNO14}, a transition relation is non-terminating iff it has a recurrence set of states.
+%% Candidate transition relations are 
+%% \emph{lassos} in the program's CFG.
+%% A lasso at a cutpoint location consists of two sequences
+%% of transitions, the stem and the loop.
+%% The stem is a finite path from an initial location to the cutpoint location. 
+%% The loop is a finite path that starts and ends at the cutpoint
+%% location by following a cyclic path through the CFG. 
+%% A lasso induces an infinite execution if the infinite path obtained by traversing the stem and then unrolling
+%% the loop infinitely many times is feasible. In this case, the lasso is said to be non-terminating.
 %% We observe that in the case that $T$ is deterministic (that is, each state has
@@
 
-The method is incomplete as not all non-terminating
-program executions are induced by lassos. An infinite behaviour
-that is not periodic is illustrated by the following program:
-
-Our method is purely semantic. %We extend the notion of recurrence set in order to formulate the 
-We formulate the non-termination problem at the loop level.%, rather than transition relation.
-In our design, a loop $L=(\mathcal{V},I,G,t)$ is non-terminating iff it has a recurrence set of states.\\
+%% The method is incomplete as not all non-terminating
+%% program executions are induced by lassos. An infinite behaviour
+%% that is not periodic is illustrated by the following program:
+
+%Our method is purely semantic.
 
@@
 Discuss {\bf [SNT]}.
+
+{\bf Nested loops.}
 
@@
 \centering
-\begin{tabular}{|l|r|c|c|c|c|c|r|r|c|r|}
+\small
+\begin{tabular}{|l|r|c|c|c|c|c|r|r|c|r|r|}
 \hline
-    & LOC & Terminates? & \shortstack{Linear \\ program?} & \shortstack{Linear ranking \\ function?}  & Conditional? & Float? & Dimension & \shortstack{Ranking \\ program size} & Res & Time (s)\\
+    & LOC & Terminates?
+& \shortstack{Linear \\ program?} & \shortstack{Linear ranking \\ function?}  & Conditional?
+& Float?
+& Dimension & \shortstack{Ranking \\ program size} & Res & Time (s) & Iterations\\
     \hline

### 2014-07-01 papers/termination/paper.tex
more on nested loops in sec 5

@@
 \begin{definition}[ Nested Loops Termination Formula {\bf [NT]}]
-\label{def:deterministic-nonterm-formula}
+\label{def:nested-term-formula}
  \begin{align*}
@@
 
-\noindent{\bf Generalised synthesis formula.}
-Our design: for each loop always search for lexicographic ranking function + supporting invariant.
+\noindent{\bf Nested loops.}
+For brevity reasons, we illustrate our approach to handling nested loops for only two loops, but the approach can be easily generalised to any number of nested loops:
+\begin{lstlisting}[mathescape=true]
+while ($G_1$) { 
+  $P_1$ 
+  while ($G_2$) { $T$ }
+  $P_2$
+}
+\end{lstlisting}
+
+The termination of the outer loop is equivalent to the formula in Definition~\ref{def:nested-term-formula}.
+The formula requires an auxiliary assertion $S(\mathcal{V},\mathcal{V}')$ representing a binary relation between the initial states at the inner loop's entry point
+and their successors according to the transition relation $T$. If this formula is satisfiable, then $R$ is a ranking function of the outer loop.
+
+In order to find a termination argument for the inner loop, we use {\bf [CT]} where the initial states are given by 
+$\neg G_1(\mathcal{V}) \wedge P_1(\mathcal{V},\mathcal{V}') \rightarrow I(\mathcal{V}')$.
 
@@
 %, i.e. we under-approximate the transition relation. 
-For this example, our technique computes $N[a](x,y,a) = $.%, where $a$ corresponds to the non-deterministic choice.  \\
+For this example, our technique computes $N[a](x,y,a) = ...$. \\%, where $a$ corresponds to the non-deterministic choice.  \\
 %The issue with the closed recurrence set is that it yields an incomplete  for loops 
@@
 
-{\bf Sequential loops.}
-Discuss {\bf [SNT]}.
-
-{\bf Nested loops.}
+\noindent{\bf Sequential loops.}
+Discuss {\bf [SNT]}.\\
+
+\noindent{\bf Nested loops.}
+
+\subsection{Generalised synthesis formula.}
+Our design: for each loop always search for lexicographic ranking function + supporting invariant OR recurrence set.
+
+Note: given a program with multiple loops (sequential and nested), we could potentially compute termination/non-termination 
+arguments for all of them in the same formula. Currently, our tool generates a separate formula for each loop.
 

### 2014-07-02 papers/termination/paper.tex
limitation in intro
Reworked formalism

@@
 \begin{itemize}
-\item No procedures (this could be handled by [42]).
+\item No procedures (this could be handled by computing function summaries).
 \item Heap (Thor).
@@
 \subsection{Termination and Ranking Functions} \label{sec:ranking.functions}
-A programs $P$ is represented as a transition systems with state space $X$, transition relation $T \subseteq X \times X$, and set of variables $\mathcal{V}$. 
-A state of the program is a valuation of the variables from $\mathcal{V}$. For a state $s \in X$ with $T(s,s')$ we say $s'$
+A programs $P$ is represented as a transition systems with state space $X$, transition relation $T \subseteq X \times X$, and set of variables $x$. 
+A state of the program is a valuation of the variables from $x$. For a state $s \in X$ with $T(s,s')$ we say $s'$
 is the successor of $s$ under $T$.
@@
 for all $\vec{y}$. 
-If a pair $(\vec{P_0}, \vec{x_0})$ is a solution to the synthesis problem, then we write $(\vec{P_0}, \vec{x_0}) \models \sigma$.
+If a pair $(\vec{P_0}, \vec{x})$ is a solution to the synthesis problem, then we write $(\vec{P_0}, \vec{x}) \models \sigma$.
+For the remainder of the presentation, we drop the $\vec{x}$ and instead just write $x$, with the understanding
+that all variables range over vectors.
 
@@
 \begin{align*}
- \exists R . \forall \mathcal{V}, \mathcal{V'} . & G(\mathcal{V}) \wedge T(\mathcal{V}, \mathcal{V'}) \rightarrow R(\mathcal{V}) > 0 \wedge R(\mathcal{V}) > R(\mathcal{V'})
+ \exists R . \forall x, x' . & G(x) \wedge T(x, x') \rightarrow R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
@@
  \begin{align*}
-  \exists R, Inv . \forall \mathcal{V}, \mathcal{V'} . & I(\mathcal{V}) \wedge G(\mathcal{V}) \rightarrow Inv(\mathcal{V}) ~ \wedge \\
-                                 & G(\mathcal{V}) \wedge Inv(\mathcal{V}) \wedge T(\mathcal{V}, \mathcal{V'}) \rightarrow Inv(\mathcal{V'}) \wedge R(\mathcal{V}) > 0 
-  \wedge R(\mathcal{V}) > R(\mathcal{V'})
+  \exists R, Inv . \forall x, x' . & I(x) \wedge G(x) \rightarrow Inv(x) ~ \wedge \\
+                                 & G(x) \wedge Inv(x) \wedge T(x, x') \rightarrow Inv(x') \wedge R(x) > 0 
+  \wedge R(x) > R(x')
  \end{align*}
@@
 
+\begin{definition}[Non-Termination Formula - Open Recurrence Set  {\bf [ONT]}]
+\label{def:nonterm-formula}
+ \begin{align*}
+  \exists N, x_0 . \forall x . \exists x' . & I(x_0) \wedge N(x_0) ~ \wedge \\
+                                                        & N(x) \rightarrow G(x) ~ \wedge \\
+							& N(x) \rightarrow T(x, x') \wedge N(x') 
+ \end{align*}
+\end{definition}
+
+\begin{definition}[Non-Termination Formula - Closed Recurrence Set {\bf [CNT]}]
+\label{def:deterministic-nonterm-formula}
+ \begin{align*}
+  \exists N, x_0 . \forall x, x' . & I(x_0) \wedge N(x_0) ~ \wedge \\
+                                                        & N(x) \rightarrow G(x) ~ \wedge \\
+							& N(x) \wedge T(x, x') \rightarrow N(x') 
+ \end{align*} 
+\end{definition}
+
+\end{framed}
+\end{figure*}
+
+\begin{figure*}
+ \begin{framed}
  \begin{definition}[Sequential Loops Termination Formula {\bf [ST]}]
+  \begin{align*}
 \label{def:multi-termination-formula}
- \begin{align*}
-  \exists R, Inv_1,..., Inv_n . \forall \mathcal{V}_0,...,\mathcal{V}_n, \mathcal{V}_1',...,\mathcal{V}_n'.  & P_0(\mathcal{V}_0,\mathcal{V}_1) \rightarrow Inv_1(\mathcal{V}_1) ~  \\
- & \bigwedge_{i=1..n{-}1} (Inv_i(\mathcal{V}_i) \wedge G_i(\mathcal{V}_i) \wedge T_i(\mathcal{V}_i, \mathcal{V}_i') \rightarrow Inv_i(\mathcal{V}_i')) ~  \\  
- & \bigwedge_{i=1..n{-}2} (Inv_i(\mathcal{V}_i) \wedge \lnot G_i(\mathcal{V}_i) \wedge P_{i+1}(\mathcal{V}_i, \mathcal{V}_{i+1}) \rightarrow Inv_{i+1}(\mathcal{V}_{i+1})) ~ \wedge \\
- & Inv_{n-1}(\mathcal{V}_{n-1}) \wedge \lnot G_{n-1}(\mathcal{V}_{n-1}) \wedge P_n(\mathcal{V}_{n-1},\mathcal{V}_n) \wedge G_n(\mathcal{V}_n) \rightarrow Inv_n(\mathcal{V}_n) ~ \wedge \\
- & Inv_n(\mathcal{V}_n) \wedge G_n(\mathcal{V}_n) \wedge T(\mathcal{V}_n, \mathcal{V}_n') \rightarrow Inv_n(\mathcal{V}_n') \wedge R(\mathcal{V}_n) > 0 \wedge R(\mathcal{V}_n) > R(\mathcal{V}_n')
+  \exists R, Inv_1,..., Inv_n . \forall x_0,...,x_n, x_1',...,x_n'.  & P_0(x_0,x_1) \rightarrow Inv_1(x_1) ~  \\
+ & \bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) ~  \\  
+ & \bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) \rightarrow Inv_{i+1}(x_{i+1})) ~ \wedge \\
+ & Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge G_n(x_n) \rightarrow Inv_n(x_n) ~ \wedge \\
+ & Inv_n(x_n) \wedge G_n(x_n) \wedge T(x_n, x_n') \rightarrow Inv_n(x_n') \wedge R(x_n) > 0 \wedge R(x_n) > R(x_n')
  \end{align*}
@@
  \begin{align*}
-  \exists R, S . \forall \mathcal{V}, \mathcal{V}' . & G_1(\mathcal{V}) \wedge P_1(\mathcal{V},\mathcal{V}') \rightarrow S(\mathcal{V}',\mathcal{V}') ~ \wedge \\
-                                & G_2(\mathcal{V}') \wedge S(\mathcal{V},\mathcal{V}') \wedge T(\mathcal{V}',\mathcal{V}'')\rightarrow S(\mathcal{V},\mathcal{V}'') ~ \wedge \\
-				& G_1(\mathcal{V}) \wedge P_1(\mathcal{V},\mathcal{V}') \wedge \neg G_2(\mathcal{V}') \wedge S(\mathcal{V}',\mathcal{V}'') \wedge 
-                                P_2(\mathcal{V}'',\mathcal{V}''') \wedge 
-                                  R(\mathcal{V}) > 0 \wedge R(\mathcal{V}) > R(\mathcal{V}''') 
- \end{align*} 
-\end{definition}
-
-\begin{definition}[Non-Termination Formula - Open Recurrence Set  {\bf [ONT]}]
-\label{def:nonterm-formula}
- \begin{align*}
-  \exists N, \mathcal{V}_0 . \forall \mathcal{V} . \exists \mathcal{V}' . & I(\mathcal{V}_0) \wedge N(\mathcal{V}_0) ~ \wedge \\
-                                                        & N(\mathcal{V}) \rightarrow G(\mathcal{V}) ~ \wedge \\
-							& N(\mathcal{V}) \rightarrow T(\mathcal{V}, \mathcal{V}') \wedge N(\mathcal{V}') 
- \end{align*}
-\end{definition}
-
-\begin{definition}[Non-Termination Formula - Closed Recurrence Set {\bf [CNT]}]
-\label{def:deterministic-nonterm-formula}
- \begin{align*}
-  \exists N, \mathcal{V}_0 . \forall \mathcal{V}, \mathcal{V}' . & I(\mathcal{V}_0) \wedge N(\mathcal{V}_0) ~ \wedge \\
-                                                        & N(\mathcal{V}) \rightarrow G(\mathcal{V}) ~ \wedge \\
-							& N(\mathcal{V}) \wedge T(\mathcal{V}, \mathcal{V}') \rightarrow N(\mathcal{V}') 
+  \exists R, S . \forall x, x' . & G_1(x) \wedge P_1(x,x') \rightarrow S(x',x') ~ \wedge \\
+                                & G_2(x') \wedge S(x,x') \wedge T(x',x'')\rightarrow S(x,x'') ~ \wedge \\
+				& G_1(x) \wedge P_1(x,x') \wedge \neg G_2(x') \wedge S(x',x'') \wedge 
+                                P_2(x'',x''') \wedge 
+                                  R(x) > 0 \wedge R(x) > R(x''') 
  \end{align*} 
@@
  \begin{align*}
-  \exists R, Inv_1,..., Inv_n . \forall \mathcal{V}_0,...,\mathcal{V}_n, \mathcal{V}_1',...,\mathcal{V}_n'.  & P_0(\mathcal{V}_0,\mathcal{V}_1) \rightarrow Inv_1(\mathcal{V}_1) ~  \\
- & \bigwedge_{i=1..n{-}1} (Inv_i(\mathcal{V}_i) \wedge G_i(\mathcal{V}_i) \wedge T_i(\mathcal{V}_i, \mathcal{V}_i') \rightarrow Inv_i(\mathcal{V}_i')) ~  \\  
- & \bigwedge_{i=1..n{-}2} (Inv_i(\mathcal{V}_i) \wedge \lnot G_i(\mathcal{V}_i) \wedge P_{i+1}(\mathcal{V}_i, \mathcal{V}_{i+1}) \rightarrow Inv_{i+1}(\mathcal{V}_{i+1})) ~ \wedge \\
- & Inv_{n-1}(\mathcal{V}_{n-1}) \wedge \lnot G_{n-1}(\mathcal{V}_{n-1}) \wedge P_n(\mathcal{V}_{n-1},\mathcal{V}_n) \wedge N(\mathcal{V}_n)  ~ \wedge \\
-                                                        & N(\mathcal{V}_n) \rightarrow G_n(\mathcal{V}_n) ~ \wedge 
-							 N(\mathcal{V}_n) \wedge T_n(\mathcal{V}_n, \mathcal{V}_n') \rightarrow N(\mathcal{V}_n') 
+  \exists R, Inv_1,..., Inv_n . \forall x_0,...,x_n, x_1',...,x_n'.  & P_0(x_0,x_1) \rightarrow Inv_1(x_1) ~  \\
+ & \bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) ~  \\  
+ & \bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) \rightarrow Inv_{i+1}(x_{i+1})) ~ \wedge \\
+ & Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge N(x_n)  ~ \wedge \\
+                                                        & N(x_n) \rightarrow G_n(x_n) ~ \wedge 
+							 N(x_n) \wedge T_n(x_n, x_n') \rightarrow N(x_n') 
  \end{align*}
 \end{definition}
-
  \end{framed}
+
 \end{figure*}
@@
 \noindent {\bf Unconditional termination.}
-We start by considering a program consisting of a single loop ${L=(\mathcal{V},G,T)}$,
-where the states satisfying the loop's guard are given by the assertion $G(\mathcal{V})$ over program variables $\mathcal{V}$.
+We start by considering a program consisting of a single loop ${L= \langle G,T \rangle}$,
+where the states $x$ satisfying the loop's guard are given by the predicate $G(x)$.
+The body of the loop is encoded as the transition relation $T(x, x')$, meaning that
+state $x'$ is reachable from state $x$ via a single iteration of the loop body.
 %where the initial states of the program 
@@
 %satisfy $I : X \rightarrow \mathbb{B}$, $G : X \rightarrow \mathbb{B} $ is the loop guard and $T : X \rightarrow X$ is the loop body. 
-We use assertions over program variables $\mathcal{V}$ and their successors $\mathcal{V'}$ to model program
-statements as binary relations over states. For loop $L$, $T(\mathcal{V}, \mathcal{V'})$
-represents the body of the loop. 
+%We use predicates over program states $x$ and their successors $x'$ to model program
+%statements as binary relations over states. For loop $L$, $T(x, x')$
+%represents the body of the loop. 
 %
@@
 
-A program unconditional terminating is given in Figure~\ref{fig:motivation.a}. 
-We assume that the program state is given by a valuation of the
-program variables $\mathcal{V} = \{x\}$. Then, 
-$G(x) = x>0$ and $T(x,x') = (x-1) \& x$. 
-The corresponding 
-synthesis formula {\bf [UT]} is satisfiable with a possible witness $R(x) = x$.\\
+Our formalism is illustrated by the unconditionally terminating program shown
+in Figure~\ref{fig:motivation.a}, which is encoded as:
+\begin{align*}
+G(x) & = \{ x \mid x>0 \} \\
+T(x,x') &= \{ \langle x, x' \rangle \mid x' = (x - 1) \, \& \, x \}
+\end{align*}
+
+We will abbreviate this with the notation $G(x) \triangleq x > 0$ and $T(x, x') \triangleq x' = (x - 1) \, \& \, x$.
+The corresponding synthesis formula {\bf [UT]} is satisfiable, as witnessed by the function $R(x) = x$.\\
 
@@
 executing in, i.e. $L$ is \emph{conditionally terminating}. 
-The loop we consider now is $L=(\mathcal{V,I,G,T})$ and has an additional assertion $I(\mathcal{V})$ 
-representing the initial states at the loop entry.
+The loop we consider now is $L=\langle I,G,T \rangle$ and has an additional predicate $I(x)$ 
+representing the states in which the loop can begin executing.
 Conditional termination is equivalent to the
@@
 \begin{itemize}
-\item $Inv_0$ is an inductive invariant of $L$ that is established by the initial states $I$ if the loop
+\item $W$ is an inductive invariant of $L$ that is established by the initial states $I$ if the loop
 guard $G$ is met.
-\item $R_0$ is a ranking function for $L$ as restricted by $Inv_0$ -- that is to say, $R_0$ need only
-be well founded on those states satisfying $Inv_0$.  Since $Inv_0$ is an inductive invariant of $L$,
-$R_0$ is strong enough to show that $L$ terminates from any of its initial states.
+\item $R$ is a ranking function for $L$ as restricted by $W$ -- that is to say, $R$ need only
+be well founded on those states satisfying $W \wedge G$.  Since $W$ is an inductive invariant of $L$,
+$R$ is strong enough to show that $L$ terminates from any of its initial states.
 \end{itemize}
 
-$Inv$ is called a \emph{supporting invariant} for $L$ and $R$ proves termination relative to $Inv$.
+$W$ is called a \emph{supporting invariant} for $L$ and $R$ proves termination relative to $W$.
 
 Conditional termination is illustrated by the loop in Figure~\ref{fig:motivation.h}, 
-where $\mathcal{V} = \{x,y\}$ and $I(x,y) = (y=1)$. 
-When taken individually, without considering the program state just before the loop, the loop may not terminate.
-However, formula {\bf [CT]} is satisfiable with the possible ranking function $R(x,y) = x$ and supporting invariant $Inv(x,y) = (y = 1)$.\\
+which is encoded as:
+\begin{align*}
+            I(\langle x, y \rangle) & \triangleq y = 1 \\
+            G(\langle x, y \rangle) & \triangleq x > 0 \\
+            T(\langle x, y \rangle, \langle x', y' \rangle) & \triangleq x' = x - y \wedge y' = y \\
+\end{align*}
+
+If the initial states $I$ are ignored, this loop cannot be shown to terminate (e.g. any state with $y = 0$ would lead to a
+non-terminating execution).
+
+However, formula {\bf [CT]} is satisfiable, as witnessed by the ranking function
+$R(\langle x,y\rangle ) = x$ and supporting invariant $W(\langle x, y \rangle ) \trangleq y = 1$.\\
 
@@
 %The conditional termination formula becomes more complex in the presence of multiple loops.
-For illustration, consider the following program with $n$ sequential loops $\forall i=1..n. ~L_i=(\mathcal{V}_i,G_i,T_i)$ and
+For illustration, consider the following program with $n$ sequential loops $\forall i=1..n. ~L_i=(x_i,G_i,T_i)$ and
 the assertions $\forall i=1..n. ~P_i$ modelling the program statements in between loops.
@@
 In order to prove that the $n^{th}$ loop terminates, we need to apply the {\bf [CT]} formula for $L_n$.
-The challenge is determining the initial states  $I(\mathcal{V}_n)$ at the loop's entry point.  
+The challenge is determining the initial states  $I(x_n)$ at the loop's entry point.  
 The termination problem for $L_n$ is thus equivalent to the satisfiability of
@@
 \item $Inv_i$ is an inductive invariant of $L_i$ for $i=1..{n-1}$ as established by the conjuncts:
-$$\bigwedge_{i=1..n{-}1} (Inv_i(\mathcal{V}_i) \wedge G_i(\mathcal{V}_i) \wedge T_i(\mathcal{V}_i, \mathcal{V}_i') \rightarrow Inv_i(\mathcal{V}_i')) $$
+$$\bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) $$
 
 \item Each loop invariant $Inv_i$ with $i=1..{n-2}$ is strong enough to establish $Inv_{i+1}$:
-$$\bigwedge_{i=1..n{-}2} (Inv_i(\mathcal{V}_i) \wedge \lnot G_i(\mathcal{V}_i) \wedge P_{i+1}(\mathcal{V}_i, \mathcal{V}_{i+1}) {\rightarrow} Inv_{i+1}(\mathcal{V}_{i+1})) ~$$
+$$\bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) {\rightarrow} Inv_{i+1}(x_{i+1})) ~$$
 
@@
 is strong enough to establish $Inv_n$, i.e. $Inv_n$ holds for the program states after $L_n$'s entry point:
-$$ Inv_{n-1}(\mathcal{V}_{n-1}) \wedge \lnot G_{n-1}(\mathcal{V}_{n-1}) \wedge P_n(\mathcal{V}_{n-1},\mathcal{V}_n) \wedge G_n(\mathcal{V}_n) $$
-$$\qquad\qquad\rightarrow Inv_n(\mathcal{V}_n)$$
+$$ Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge G_n(x_n) $$
+$$\qquad\qquad\rightarrow Inv_n(x_n)$$
 
 \item $Inv_n$ is a supporting termination invariant of $L_n$ and $R$ is a ranking function relative to $Inv_n$:
-$$Inv_n(\mathcal{V}_n) \wedge G_n(\mathcal{V}_n) \wedge T(\mathcal{V}_n, \mathcal{V}_n') \rightarrow Inv_n(\mathcal{V}_n')$$
-$$ \qquad \qquad \wedge R(\mathcal{V}_n) > 0 \wedge R(\mathcal{V}_n) > R(\mathcal{V}_n')$$
+$$Inv_n(x_n) \wedge G_n(x_n) \wedge T(x_n, x_n') \rightarrow Inv_n(x_n')$$
+$$ \qquad \qquad \wedge R(x_n) > 0 \wedge R(x_n) > R(x_n')$$
 \end{itemize}
@@
 The termination of the outer loop is equivalent to the formula in Definition~\ref{def:nested-term-formula}.
-The formula requires an auxiliary assertion $S(\mathcal{V},\mathcal{V}')$ representing a binary relation between the initial states at the inner loop's entry point
+The formula requires an auxiliary assertion $S(x,x')$ representing a binary relation between the initial states at the inner loop's entry point
 and their successors according to the transition relation $T$. If this formula is satisfiable, then $R$ is a ranking function of the outer loop.
@@
 In order to find a termination argument for the inner loop, we use {\bf [CT]} where the initial states are given by 
-$\neg G_1(\mathcal{V}) \wedge P_1(\mathcal{V},\mathcal{V}') \rightarrow I(\mathcal{V}')$.
+$\neg G_1(x) \wedge P_1(x,x') \rightarrow I(x')$.
 
@@
 exists a transition to some $s'\in N$. 
-For our case, given a loop $L = (I,\mathcal{V},G,T)$, the notion of open recurrent set is encoded by Definition~\ref{def:nonterm-formula}.
+For our case, given a loop $L = (I,x,G,T)$, the notion of open recurrent set is encoded by Definition~\ref{def:nonterm-formula}.
 
@@
 satisfiability of Definition~\ref{def:deterministic-nonterm-formula}.
-As our formulation is with respect to a given loop $L=(\mathcal{V},I,G,t)$, we require the  
-recurrence set to include the loop's guard, $N(\mathcal{V}) \rightarrow G(\mathcal{V})$. In other words, the loop's guard needs to be satisfied in order for the 
+As our formulation is with respect to a given loop $L=(x,I,G,t)$, we require the  
+recurrence set to include the loop's guard, $N(x) \rightarrow G(x)$. In other words, the loop's guard needs to be satisfied in order for the 
 infinite execution to happen.
@@
 In order to restrict the values that may be assigned to $y$, we replace the non-deterministic choice with a fresh variable $a$ and use formula {\bf [CNT]}
-to compute a closed recurrence set $N[a]$ parametrised by $a$. Now $N[a]$ becomes an assertion over $\mathcal{V} \cup \{a\}$.
+to compute a closed recurrence set $N[a]$ parametrised by $a$. Now $N[a]$ becomes an assertion over $x \cup \{a\}$.
 %, i.e. we under-approximate the transition relation. 

### 2014-07-02 papers/termination/kalashnikov.tex
modified kalashnikov + other small things
revising sec 6

@@
 
+\iffalse
 \subsection{The Basic Synthesis Algorithm}
@@
 
+\fi
+
+In this section we present a solver for the synthesis fragment. An
+important characteristic of our solver is that is models bit accurate
+semantics. This enables the precise treatment of bit-wise operations and
+arithmetic modulo fixed widths.
+
+%We can solve the synthesis problem using any off-the-shelf program synthesiser.  If you don't
+%have a program synthesiser kicking around, we present the design of one synthesiser
+%that can be used to solve the problem.
+%We start by recalling the synthesis formula:
+% $$\exists P, x . \forall y . \sigma(P, x, y)$$
+We start by recalling the synthesis fragment:
+ \[
+  \exists P,~ x . ~\forall~ y . \sigma(P, x, y)
+ \]
+
+where $P$ ranges over sets, while $x$ and $y$ range over ground terms (we have dropped
+ the vector symbols for brevity reasons).
+
 When the quantifier free first-order fragment of the logic is efficiently decidable and
-$P$ can be expressed as a ground term of the logic, we can use
+$P$ is expressible as a ground term of the logic, we can use
 Counterexample Guided Inductive Synthesis (CEGIS)~\cite{lezama-thesis,sketch} to
 check the validity of the synthesis formula. The core of the CEGIS algorithm is
-a refinement loop to check the
-validity of the synthesis formula as shown in Fig.~\ref{fig:abstract-refinement} and
-Algorithm~\ref{fig:abstract-refinement-code}.  The algorithm is divided into two
-components: {\sc synth} and {\sc verif}.  We track a finite (small) subset of test inputs
-and {\sc synth} synthesises a candidate program that is correct on just those tests.
+the refinement loop in Fig.~\ref{fig:abstract-refinement} and
+detailed in Algorithm~\ref{fig:abstract-refinement-code}.
+
+The algorithm is divided into two
+components: {\sc synth} and {\sc verif}. We track a finite set of test
+inputs and {\sc synth} synthesises a candidate program that is correct on just those tests.
 {\sc Verif} then tries to verify the candidate program by searching for an input on
@@
 
+\iffalse
 \subsection{The Concrete Algorithm for Bitvector Programs}
@@
 fragment of C as a ``high level'' logic.
-
-\subsubsection{\newC}
+\fi
+
+\subsection{\newC}
 \label{sec:logic}
@@
 
-\subsubsection{Encoding an $\mathcal{L}$-Program in \newC}
+%\subsubsection{Encoding an $\mathcal{L}$-Program in \newC}
 \label{sec:encode-l}

### 2014-07-02 papers/termination/paper.tex
revising sec 6
Nudge introduction, introduce the halting problem

@@
 
+The halting problem has been of central interest to computer scientists since it was first
+considered by Turing in 1936~\cite{turing}.  Informally, we can say that the halting problem
+is concerned with answering the question ``does this
+program run forever, or will it eventually terminate?''
+
 Proving program termination is typically done by finding a \emph{ranking function}
-for the program states, i.e. a map from the program's state space to a well-ordered set.
+for the program states, i.e. a monotone map from the program's state space to a well-ordered set.
+Historically, the search for ranking functions has been constrained in various syntactic ways, leading to
+incompleteness, and
+is performed over abstractions that do not soundly capture the behaviour of physical computers.
+In this paper, we present a sound and complete method for deciding whether a program terminates.
+
 %% As this definition of a ranking function is very general, research is often limited to some
@@
 
-If the initial states $I$ are ignored, this loop cannot be shown to terminate (e.g. any state with $y = 0$ would lead to a
-non-terminating execution).
+If the initial states $I$ are ignored this loop cannot be shown to terminate, since any state with $y = 0$ and $x > 0$
+would lead to a non-terminating execution.
 
 However, formula {\bf [CT]} is satisfiable, as witnessed by the ranking function
-$R(\langle x,y\rangle ) = x$ and supporting invariant $W(\langle x, y \rangle ) \trangleq y = 1$.\\
-
-\noindent {\bf Sequential loops.}
-Conditional termination allows us to
-prove that programs with multiple loops terminate, even if the termination of some loop depends on the states
-reachable after leaving a previous loop.
-%The conditional termination formula becomes more complex in the presence of multiple loops.
-For illustration, consider the following program with $n$ sequential loops $\forall i=1..n. ~L_i=(x_i,G_i,T_i)$ and
-the assertions $\forall i=1..n. ~P_i$ modelling the program statements in between loops.
-
-\begin{lstlisting}[mathescape=true]
-$P_1$
-while ($G_1$) { $T_1$ }
-$P_2$
-while ($G_2$) { $T_2$ }
-...
-$P_n$
-while ($G_n$) { $T_n$ }
-\end{lstlisting}
-
-In order to prove that the $n^{th}$ loop terminates, we need to apply the {\bf [CT]} formula for $L_n$.
-The challenge is determining the initial states  $I(x_n)$ at the loop's entry point.  
-The termination problem for $L_n$ is thus equivalent to the satisfiability of
-Definition~\ref{def:multi-termination-formula}.  If this formula is satisfiable:
-\begin{itemize}
-\item $Inv_i$ is an inductive invariant of $L_i$ for $i=1..{n-1}$ as established by the conjuncts:
-$$\bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) $$
-
-\item Each loop invariant $Inv_i$ with $i=1..{n-2}$ is strong enough to establish $Inv_{i+1}$:
-$$\bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) {\rightarrow} Inv_{i+1}(x_{i+1})) ~$$
-
-\item When $L_n$'s guard holds, $Inv_{n-1}$ 
-is strong enough to establish $Inv_n$, i.e. $Inv_n$ holds for the program states after $L_n$'s entry point:
-$$ Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge G_n(x_n) $$
-$$\qquad\qquad\rightarrow Inv_n(x_n)$$
-
-\item $Inv_n$ is a supporting termination invariant of $L_n$ and $R$ is a ranking function relative to $Inv_n$:
-$$Inv_n(x_n) \wedge G_n(x_n) \wedge T(x_n, x_n') \rightarrow Inv_n(x_n')$$
-$$ \qquad \qquad \wedge R(x_n) > 0 \wedge R(x_n) > R(x_n')$$
-\end{itemize}
-
-\todo{add example?}\\
-
-\noindent{\bf Nested loops.}
-For brevity reasons, we illustrate our approach to handling nested loops for only two loops, but the approach can be easily generalised to any number of nested loops:
-\begin{lstlisting}[mathescape=true]
-while ($G_1$) { 
-  $P_1$ 
-  while ($G_2$) { $T$ }
-  $P_2$
-}
-\end{lstlisting}
-
-The termination of the outer loop is equivalent to the formula in Definition~\ref{def:nested-term-formula}.
-The formula requires an auxiliary assertion $S(x,x')$ representing a binary relation between the initial states at the inner loop's entry point
-and their successors according to the transition relation $T$. If this formula is satisfiable, then $R$ is a ranking function of the outer loop.
-
-In order to find a termination argument for the inner loop, we use {\bf [CT]} where the initial states are given by 
-$\neg G_1(x) \wedge P_1(x,x') \rightarrow I(x')$.
+$R(\langle x,y\rangle ) = x$ and supporting invariant $W(\langle x, y \rangle ) \triangleq y = 1$.
+This constitutes a proof that the program as whole terminates, since the loop always begins
+executing in a state that guarantees its termination.
 
@@
 %Our method is purely semantic.
+
+\noindent {\bf Sequential loops.}
+Conditional termination allows us to
+prove that programs with multiple loops terminate, even if the termination of some loop depends on the states
+reachable after leaving a previous loop.
+%The conditional termination formula becomes more complex in the presence of multiple loops.
+For illustration, consider the following program with $n$ sequential loops $\forall i=1..n. ~L_i=(x_i,G_i,T_i)$ and
+the assertions $\forall i=1..n. ~P_i$ modelling the program statements in between loops.
+
+\begin{lstlisting}[mathescape=true]
+$P_1$
+while ($G_1$) { $T_1$ }
+$P_2$
+while ($G_2$) { $T_2$ }
+...
+$P_n$
+while ($G_n$) { $T_n$ }
+\end{lstlisting}
+
+In order to prove that the $n^{th}$ loop terminates, we need to apply the {\bf [CT]} formula for $L_n$.
+The challenge is determining the initial states  $I(x_n)$ at the loop's entry point.  
+The termination problem for $L_n$ is thus equivalent to the satisfiability of
+Definition~\ref{def:multi-termination-formula}.  If this formula is satisfiable:
+\begin{itemize}
+\item $Inv_i$ is an inductive invariant of $L_i$ for $i=1..{n-1}$ as established by the conjuncts:
+$$\bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) $$
+
+\item Each loop invariant $Inv_i$ with $i=1..{n-2}$ is strong enough to establish $Inv_{i+1}$:
+$$\bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) {\rightarrow} Inv_{i+1}(x_{i+1})) ~$$
+
+\item When $L_n$'s guard holds, $Inv_{n-1}$ 
+is strong enough to establish $Inv_n$, i.e. $Inv_n$ holds for the program states after $L_n$'s entry point:
+$$ Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge G_n(x_n) $$
+$$\qquad\qquad\rightarrow Inv_n(x_n)$$
+
+\item $Inv_n$ is a supporting termination invariant of $L_n$ and $R$ is a ranking function relative to $Inv_n$:
+$$Inv_n(x_n) \wedge G_n(x_n) \wedge T(x_n, x_n') \rightarrow Inv_n(x_n')$$
+$$ \qquad \qquad \wedge R(x_n) > 0 \wedge R(x_n) > R(x_n')$$
+\end{itemize}
+
+\todo{add example?}\\
+
+\noindent{\bf Nested loops.}
+For brevity reasons, we illustrate our approach to handling nested loops for only two loops, but the approach can be easily generalised to any number of nested loops:
+\begin{lstlisting}[mathescape=true]
+while ($G_1$) { 
+  $P_1$ 
+  while ($G_2$) { $T$ }
+  $P_2$
+}
+\end{lstlisting}
+
+The termination of the outer loop is equivalent to the formula in Definition~\ref{def:nested-term-formula}.
+The formula requires an auxiliary assertion $S(x,x')$ representing a binary relation between the initial states at the inner loop's entry point
+and their successors according to the transition relation $T$. If this formula is satisfiable, then $R$ is a ranking function of the outer loop.
+
+In order to find a termination argument for the inner loop, we use {\bf [CT]} where the initial states are given by 
+$\neg G_1(x) \wedge P_1(x,x') \rightarrow I(x')$.
 

### 2014-07-02 papers/termination/kalashnikov.tex
revising sec 6

@@
 
-\newcommand{\newC}{C$^-$\xspace}
 \newcommand*\Let[2]{\State #1 $\gets$ #2}
@@
 check the validity of the synthesis formula. The core of the CEGIS algorithm is
-the refinement loop in Fig.~\ref{fig:abstract-refinement} and
+the refinement loop shown in Fig.~\ref{fig:abstract-refinement} and
 detailed in Algorithm~\ref{fig:abstract-refinement-code}.
@@
 The algorithm is divided into two
-components: {\sc synth} and {\sc verif}. We track a finite set of test
-inputs and {\sc synth} synthesises a candidate program that is correct on just those tests.
-{\sc Verif} then tries to verify the candidate program by searching for an input on
+components: {\sc synth} and {\sc verif}. For a finite set of test
+inputs that we track, {\sc synth} synthesises a candidate program that is correct on just those tests.
+Subsequently, {\sc Verif} tries to verify the candidate program by searching for an input on
 which the candidate program does not satisfy the specification.  If no such input
 can be found, the candidate program is correct.  Otherwise, the new input is added
-to the set we must check in {\sc synth}.  If the input set is finite, this
-procedure is guaranteed to terminate.
+to the setof tracked inputs.  
+%If the input set is finite, this procedure is guaranteed to terminate.
 
@@
 
-\subsection{\newC}
-\label{sec:logic}
-
-We will now describe the logic we use to express our synthesis formula.
-The logic is a subset of C that we call \newC.  The characteristic property of a
-\newC  program is that safety can be decided for
-it using a single query to a Bounded Model Checker.  A \newC program is
-just a C program with the following syntactic restrictions:
- all loops in the program must have a constant bound;
- all recursion in the program must be limited to a constant depth;
- all arrays must be statically allocated (i.e. not using \texttt{malloc}),
- and be of constant size.
-Additionally, \newC programs may use nondeterministic values, assumptions
-and arbitrary-width types.
-
-\iffalse
-Two example \newC programs are shown 
-in Fig.~\ref{fig:c-}.
-
-\begin{figure}
-\begin{minipage}[scale=0.8]{0.45\linewidth}
- \begin{lstlisting}[language=c]
-int count_bits(int x) {
-  int i, ret = 0;
-
-  for (i = 0; i < 32; i++)
-    if (x & (1 << i))
-      ret++;
-
-  return ret;
-}
- \end{lstlisting}
-\end{minipage}
-\begin{minipage}{0.54\linewidth}
- \begin{lstlisting}[language=C]
-int common_factor(int A[10]) {
-  int i, factor = nondet();
-
-  for (i = 0; i < 10; i++)
-    assume((A[i] % factor) == 0);
-
-  assume(factor > 1);
-  return factor;
-}
-
- \end{lstlisting}
-\end{minipage}
-
- \caption{Two \newC programs}
- \label{fig:c-}
-
-\end{figure}
-\fi
-
-Since each loop is bounded by a constant, and each recursive function call is
-limited to a constant depth, a \newC program necessarily terminates and in
-fact does so in $O(1)$ time.  If we call the largest loop bound~$k$, then
-a Bounded Model Checker with an unrolling bound of $k$ will be a complete
-decision procedure for the safety of the program.  For a \newC program of
-size $l$ and with largest loop bound~$k$, a Bounded Model Checker will
-create a SAT problem of size $O(lk)$.  Conversely, a SAT problem
-of size $s$ can be converted trivially into a loop-free \newC program
-of size $O(s)$.  The safety problem for \newC is therefore NP-complete,
-which means it can be decided fairly efficiently for many practical
-instances.
+%\label{sec:logic}
+
+The logic used to express our synthesis formula is \newC discussed in Section~\ref{sec:logic}.
+%% The logic is a subset of C that we call \newC.  
+%% The characteristic property of a
+%% \newC  program is that safety can be decided for
+%% it using a single query to a Bounded Model Checker.  A \newC program is
+%% just a C program with the following syntactic restrictions:
+%%  all loops in the program must have a constant bound;
+%%  all recursion in the program must be limited to a constant depth;
+%%  all arrays must be statically allocated (i.e. not using \texttt{malloc}),
+%%  and be of constant size.
+%% Additionally, \newC programs may use nondeterministic values, assumptions
+%% and arbitrary-width types.
+
+%% \iffalse
+%% Two example \newC programs are shown 
+%% in Fig.~\ref{fig:c-}.
+
+%% \begin{figure}
+%% \begin{minipage}[scale=0.8]{0.45\linewidth}
+%%  \begin{lstlisting}[language=c]
+%% int count_bits(int x) {
+%%   int i, ret = 0;
+
+%%   for (i = 0; i < 32; i++)
+%%     if (x & (1 << i))
+%%       ret++;
+
+%%   return ret;
+%% }
+%%  \end{lstlisting}
+%% \end{minipage}
+%% \begin{minipage}{0.54\linewidth}
+%%  \begin{lstlisting}[language=C]
+%% int common_factor(int A[10]) {
+%%   int i, factor = nondet();
+
+%%   for (i = 0; i < 10; i++)
+%%     assume((A[i] % factor) == 0);
+
+%%   assume(factor > 1);
+%%   return factor;
+%% }
+
+%%  \end{lstlisting}
+%% \end{minipage}
+
+%%  \caption{Two \newC programs}
+%%  \label{fig:c-}
+
+%% \end{figure}
+%% \fi
+
+%% Since each loop is bounded by a constant, and each recursive function call is
+%% limited to a constant depth, a \newC program necessarily terminates and in
+%% fact does so in $O(1)$ time.  If we call the largest loop bound~$k$, then
+%% a Bounded Model Checker with an unrolling bound of $k$ will be a complete
+%% decision procedure for the safety of the program.  For a \newC program of
+%% size $l$ and with largest loop bound~$k$, a Bounded Model Checker will
+%% create a SAT problem of size $O(lk)$.  Conversely, a SAT problem
+%% of size $s$ can be converted trivially into a loop-free \newC program
+%% of size $O(s)$.  The safety problem for \newC is therefore NP-complete,
+%% which means it can be decided fairly efficiently for many practical
+%% instances.
 
 \subsection{Encoding into \newC}
-To instantiate the abstract synthesis algorithm in \newC we must express $I, O,
-\sigma$ and $P$ in \newC, then ensure that we can express the validity of the
-synthesis formula as a safety property of the resulting \newC program.
-
-Our encoding is the following:
-%
-\begin{itemize}
- \item $I$ is the set of $N$-tuples of 32-bit bitvectors.  This is written in \newC as the type \verb|int[N]|.
- \item $O$ is the set of $M$-tuples of 32-bit bitvectors, which is written in \newC as the type \verb|int[M]|.
- \item $\sigma$ is a pure function with type $I \times O \rightarrow \mathrm{Bool}$.  The \newC signature of this function is
- \verb|int check(int in[N], int out[M])|. This function is the only component supplied
- by the user.
- \item $P$ is written in a simple RISC-like language $\mathcal{L}$, whose syntax is given in Fig.~\ref{fig:l-language}.  Programs in $\mathcal{L}$
- have type $I \rightarrow O$ and
- are represented in \newC as objects of type \verb|prog_t|, shown in Fig.~\ref{fig:c-l-encoding}.
- \item We supply an interpreter for $\mathcal{L}$ which is written in \newC.  The type of
- this interpreter is $(I \rightarrow O) \times I \rightarrow O$ and the \newC signature is \\
- \verb|void exec(prog_t p, int in[N], int out[M])|.  Here, \verb|out| is an output parameter.
-\end{itemize}
+%% To instantiate the abstract synthesis algorithm in \newC we must express $I, O,
+%% \sigma$ and $P$ in \newC, then ensure that we can express the validity of the
+%% synthesis formula as a safety property of the resulting \newC program.
+
+%% Our encoding is the following:
+%% %
+%% \begin{itemize}
+%%  \item $I$ is the set of $N$-tuples of 32-bit bitvectors.  This is written in \newC as the type \verb|int[N]|.
+%%  \item $O$ is the set of $M$-tuples of 32-bit bitvectors, which is written in \newC as the type \verb|int[M]|.
+%%  \item $\sigma$ is a pure function with type $I \times O \rightarrow \mathrm{Bool}$.  The \newC signature of this function is
+%%  \verb|int check(int in[N], int out[M])|. This function is the only component supplied
+%%  by the user.
+%%  \item $P$ is written in a simple RISC-like language $\mathcal{L}$, whose syntax is given in Fig.~\ref{fig:l-language}.  Programs in $\mathcal{L}$
+%%  have type $I \rightarrow O$ and
+%%  are represented in \newC as objects of type \verb|prog_t|, shown in Fig.~\ref{fig:c-l-encoding}.
+%%  \item We supply an interpreter for $\mathcal{L}$ which is written in \newC.  The type of
+%%  this interpreter is $(I \rightarrow O) \times I \rightarrow O$ and the \newC signature is \\
+%%  \verb|void exec(prog_t p, int in[N], int out[M])|.  Here, \verb|out| is an output parameter.
+%% \end{itemize}
 
@@
 
-The exact details of how we encode an $\mathcal{L}$-program are given in Sec.~\ref{sec:encode-l}.
-We must now express the {\sc synth} and {\sc verif} formulae as safety properties
-of \newC programs, which is given in Fig.~\ref{fig:c-synthverif}.
-
-\begin{figure}
-\centering
-\begin{minipage}[t]{.45\textwidth}
-\begin{lstlisting}[language=C++]
-void synth() {
-  prog_t p = nondet();
-  int in[N], out[M];
-
-  assume(wellformed(p));
-
-  in = test1;
-  exec(p, in, out);
-  assume(check(in, out));
-  ...
-  in = testN;
-  exec(p, in, out);
-  assume(check(in, out));
-
-  assert(false);
-}
-\end{lstlisting}
-%\end{minipage}
-%\hfill
-%\begin{minipage}[t]{.45\linewidth}
-\begin{lstlisting}[language=C++]
-void verif(prog_t p) {
-  int in[N] = nondet();
-  int out[M];
-
-  exec(p, in, out);
-  assert(check(in, out));
-}
-\end{lstlisting}
-\end{minipage}
-
- \caption{The {\sc synth} and {\sc verif} formulae expressed as a \newC program.}
- \label{fig:c-synthverif}
-\end{figure}
+%% The exact details of how we encode an $\mathcal{L}$-program are given in Sec.~\ref{sec:encode-l}.
+%% We must now express the {\sc synth} and {\sc verif} formulae as safety properties
+%% of \newC programs, which is given in Fig.~\ref{fig:c-synthverif}.
+
+%% \begin{figure}
+%% \centering
+%% \begin{minipage}[t]{.45\textwidth}
+%% \begin{lstlisting}[language=C++]
+%% void synth() {
+%%   prog_t p = nondet();
+%%   int in[N], out[M];
+
+%%   assume(wellformed(p));
+
+%%   in = test1;
+%%   exec(p, in, out);
+%%   assume(check(in, out));
+%%   ...
+%%   in = testN;
+%%   exec(p, in, out);
+%%   assume(check(in, out));
+
+%%   assert(false);
+%% }
+%% \end{lstlisting}
+%% %\end{minipage}
+%% %\hfill
+%% %\begin{minipage}[t]{.45\linewidth}
+%% \begin{lstlisting}[language=C++]
+%% void verif(prog_t p) {
+%%   int in[N] = nondet();
+%%   int out[M];
+
+%%   exec(p, in, out);
+%%   assert(check(in, out));
+%% }
+%% \end{lstlisting}
+%% \end{minipage}
+
+%%  \caption{The {\sc synth} and {\sc verif} formulae expressed as a \newC program.}
+%%  \label{fig:c-synthverif}
+%% \end{figure}
 
@@
 %\subsubsection{Encoding an $\mathcal{L}$-Program in \newC}
-\label{sec:encode-l}
-
-\begin{figure}
-\begin{lstlisting}[language=c++,mathescape]
-typedef $\mathcal{BV}(4)$ op_t;                  // An opcode
-typedef $\mathcal{BV}(w)$ word_t;                // An $\mathcal{L}$-word
-typedef $\mathcal{BV}(\log_2 \lceil c+l+a \rceil)$ param_t;  // An operand
-
-struct prog_t {
-  op_t ops[$l$];          // The opcodes
-  param_t params[$l$*2];  // The operands
-  word_t consts[$c$];     // The program constants
-}
-\end{lstlisting}
-
- \caption{The \newC structure we use to encode an $\mathcal{L}$ program
-  \label{fig:c-l-encoding}}
-\end{figure}
-
-The exact \newC encoding of an $\mathcal{L}$ program is shown in Fig.~\ref{fig:c-l-encoding}.
-The \verb|prog_t| structure encodes a program, which is a sequence of instructions.
-The parameter $a$ is the number of arguments the program takes.
-The $i$th instruction has opcode \verb|ops[i]|, left operand \verb|params[i*2]| and
-right operand \verb|params[i*2 + 1]|.  An operand refers to either a program constant,
-a program argument or the result of a previous instruction, and its value
-is determined at runtime as follows:
-\[
- val(x) = \begin{cases}
-           x < a & \text{the } x^{\text{th}} \text{ program argument} \\
-           a \leq x < a+c & \mathtt{consts[} x-a \mathtt{]} \\
-           x \geq a + c & \text{the result of instruction } (x - a - c)
-          \end{cases}
-\]
-
-A program is well formed if
-no operand refers to the result of an instruction that has not been computed yet, and if
-each opcode is valid.  We add a well-formedness constraint of the form \verb|params[i] <= (a+c+2*i)|
-for each instruction.  It should be noted that this requires a linear number of well-formedness
-constraints.  If all of these constraints are satisfied, the program is well-formed in the sense.
+%% \label{sec:encode-l}
+
+%% \begin{figure}
+%% \begin{lstlisting}[language=c++,mathescape]
+%% typedef $\mathcal{BV}(4)$ op_t;                  // An opcode
+%% typedef $\mathcal{BV}(w)$ word_t;                // An $\mathcal{L}$-word
+%% typedef $\mathcal{BV}(\log_2 \lceil c+l+a \rceil)$ param_t;  // An operand
+
+%% struct prog_t {
+%%   op_t ops[$l$];          // The opcodes
+%%   param_t params[$l$*2];  // The operands
+%%   word_t consts[$c$];     // The program constants
+%% }
+%% \end{lstlisting}
+
+%%  \caption{The \newC structure we use to encode an $\mathcal{L}$ program
+%%   \label{fig:c-l-encoding}}
+%% \end{figure}
+
+%% The exact \newC encoding of an $\mathcal{L}$ program is shown in Fig.~\ref{fig:c-l-encoding}.
+%% The \verb|prog_t| structure encodes a program, which is a sequence of instructions.
+%% The parameter $a$ is the number of arguments the program takes.
+%% The $i$th instruction has opcode \verb|ops[i]|, left operand \verb|params[i*2]| and
+%% right operand \verb|params[i*2 + 1]|.  An operand refers to either a program constant,
+%% a program argument or the result of a previous instruction, and its value
+%% is determined at runtime as follows:
+%% \[
+%%  val(x) = \begin{cases}
+%%            x < a & \text{the } x^{\text{th}} \text{ program argument} \\
+%%            a \leq x < a+c & \mathtt{consts[} x-a \mathtt{]} \\
+%%            x \geq a + c & \text{the result of instruction } (x - a - c)
+%%           \end{cases}
+%% \]
+
+%% A program is well formed if
+%% no operand refers to the result of an instruction that has not been computed yet, and if
+%% each opcode is valid.  We add a well-formedness constraint of the form \verb|params[i] <= (a+c+2*i)|
+%% for each instruction.  It should be noted that this requires a linear number of well-formedness
+%% constraints.  If all of these constraints are satisfied, the program is well-formed in the sense.
 
@@
 high-complexity language.
+
+We consider the following parameters:
 

### 2014-07-02 papers/termination/paper.tex
revising sec 6

--- 2014-07-02 papers/termination/paper.tex
@@
 \newcommand{\todo}[1]{{\bf TODO:} #1}
+\newcommand{\newC}{C$^-$\xspace}
 
@@
  The main contributions of our work can be summarised as follows:
+
 \begin{itemize}
@@
 \item We implemented our technique and tried it on a selection of programs handling both bit-vectors and floats.
+
+\item bit level termination and non-termination analysis. This enables
+  the precise treatment of bit-wise operations and arithmetic modulo
+  fixed widths.
 
@@
 
+\subsection{\newC} \todo{revise this}
+\label{sec:logic}
+
+We will now describe the logic we use to express our synthesis formula.
+The logic is a subset of C that we call \newC.  The characteristic property of a
+\newC  program is that safety can be decided for
+it using a single query to a Bounded Model Checker.  A \newC program is
+just a C program with the following syntactic restrictions:
+ all loops in the program must have a constant bound;
+ all recursion in the program must be limited to a constant depth;
+ all arrays must be statically allocated (i.e. not using \texttt{malloc}),
+ and be of constant size.
+Additionally, \newC programs may use nondeterministic values, assumptions
+and arbitrary-width types.
+
+\iffalse
+Two example \newC programs are shown 
+in Fig.~\ref{fig:c-}.
+
+\begin{figure}
+\begin{minipage}[scale=0.8]{0.45\linewidth}
+ \begin{lstlisting}[language=c]
+int count_bits(int x) {
+  int i, ret = 0;
+
+  for (i = 0; i < 32; i++)
+    if (x & (1 << i))
+      ret++;
+
+  return ret;
+}
+ \end{lstlisting}
+\end{minipage}
+\begin{minipage}{0.54\linewidth}
+ \begin{lstlisting}[language=C]
+int common_factor(int A[10]) {
+  int i, factor = nondet();
+
+  for (i = 0; i < 10; i++)
+    assume((A[i] % factor) == 0);
+
+  assume(factor > 1);
+  return factor;
+}
+
+ \end{lstlisting}
+\end{minipage}
+
+ \caption{Two \newC programs}
+ \label{fig:c-}
+
+\end{figure}
+\fi
+
+Since each loop is bounded by a constant, and each recursive function call is
+limited to a constant depth, a \newC program necessarily terminates and in
+fact does so in $O(1)$ time.  If we call the largest loop bound~$k$, then
+a Bounded Model Checker with an unrolling bound of $k$ will be a complete
+decision procedure for the safety of the program.  For a \newC program of
+size $l$ and with largest loop bound~$k$, a Bounded Model Checker will
+create a SAT problem of size $O(lk)$.  Conversely, a SAT problem
+of size $s$ can be converted trivially into a loop-free \newC program
+of size $O(s)$.  The safety problem for \newC is therefore NP-complete,
+which means it can be decided fairly efficiently for many practical
+instances.
+
 \subsection{Termination and Ranking Functions} \label{sec:ranking.functions}

### 2014-07-02 papers/termination/paper.tex
prelim + sec 6

--- 2014-07-02 papers/termination/paper.tex
@@
 Given that our state space is finite, a transition relation induces a infinite execution if and only if some state is 
-visited infinitely many times in the course of the execution. Hence, an alternative formulation of the non-termination problem 
-for bit-vectors may search for singleton recurrence sets (the recurrence set contains only the state visited infinitely many times).
+visited twice in the course of the execution. Hence, an alternative formulation of the non-termination problem 
+for bit-vectors may search for a state that is visited twice.
 However, this would require reasoning about the transitive closure of the transition relation.\\

### 2014-07-02 papers/termination/paper.tex
Whitespace

@@
 \newcommand{\todo}[1]{{\bf TODO:} #1}
-\newcommand{\newC}{C$^-$\xspace}
 
@@
  The main contributions of our work can be summarised as follows:
-
 \begin{itemize}
@@
 \item We implemented our technique and tried it on a selection of programs handling both bit-vectors and floats.
-
-\item bit level termination and non-termination analysis. This enables
-  the precise treatment of bit-wise operations and arithmetic modulo
-  fixed widths.
 
@@
 Given that our state space is finite, a transition relation induces a infinite execution if and only if some state is 
-visited twice in the course of the execution. Hence, an alternative formulation of the non-termination problem 
-for bit-vectors may search for a state that is visited twice.
+visited infinitely many times in the course of the execution. Hence, an alternative formulation of the non-termination problem 
+for bit-vectors may search for singleton recurrence sets (the recurrence set contains only the state visited infinitely many times).
 However, this would require reasoning about the transitive closure of the transition relation.\\

### 2014-07-02 papers/termination/paper.tex
Limitations

--- 2014-07-02 papers/termination/paper.tex
@@
 {\bf Limitations.}
-For ease of exposition, this language ignores features
-such as dynamic memory allocation and function calls. 
+Our algorithm decides termination of transition systems with finite state spaces.
+The (non-)termination proofs take the form of ranking functions and program invariants
+that are expressed in a quantifier free language.  This formalism is powerful
+enough to handle a large fragment of C, but is not rich enough to analyse
+code using any of the following features:
+
 \begin{itemize}
-\item No procedures (this could be handled by computing function summaries).
+\item Recursion (could probably do this).
 \item Heap (Thor).
-\item Nested loops.
+\item Arrays.
 \end{itemize}

### 2014-07-02 papers/termination/paper.tex
sec 6
Nudges

@@
 exists a transition to some $s'\in N$. 
-For our case, given a loop $L = (I,x,G,T)$, the notion of open recurrent set is encoded by Definition~\ref{def:nonterm-formula}.
+For our case, given a loop $L = \langle I,G,T \rangle$, the notion of open recurrent set is encoded by Definition~\ref{def:nonterm-formula}.
 
@@
 satisfiability of Definition~\ref{def:deterministic-nonterm-formula}.
-As our formulation is with respect to a given loop $L=(x,I,G,t)$, we require the  
+As our formulation is with respect to a given loop $L=\langle I,G,t \rangle$, we require the  
 recurrence set to include the loop's guard, $N(x) \rightarrow G(x)$. In other words, the loop's guard needs to be satisfied in order for the 

### 2014-07-02 papers/termination/kalashnikov.tex

@@
 We consider the following parameters:
-
-\subsubsection{Program Length: $l$}
-The first parameter we introduce is program length, denoted by $l$.
-At each iteration we synthesise programs of length exactly $l$.
-We start with $l = 1$ and increment $l$ whenever we determine
-that no program of length $l$ can satisfy the specification.  When we do
-successfully synthesise a program, we are \emph{guaranteed that it
-is of minimal length} since we have previously established that no
-shorter program is correct.
-
-\subsubsection{Word Width: $w$}
-An $\mathcal{L}$-program runs on a virtual machine (the $\mathcal{L}$-machine) that
-has its own set of parameters.  The only relevant parameter is
-the \emph{word width} of the $\mathcal{L}$-machine, that is, the number of bits
-in each internal register and immediate constant.  This parameter is denoted by
-$w$.  The size of the final SAT problem generated by {\sc cbmc} scales
-polynomially with $w$, since each intermediate C variable corresponds
-to $w$ propositional variables.
-
-It is often the case that a program which satisfies the specification
-on an $\mathcal{L}$-machine with $w = k$ will continue to satisfy the
-specification when run on a machine with $w > k$.  For example, the program
-in Fig.~\ref{fig:bitvector-program} isolates the least-significant bit of a word.
-This is true irrespective of the word size of the machine it is run on -- it will
-isolate the least-significant bit of an 8-bit word just as well as it will a
-32-bit word.  An often successful strategy is to synthesise a program for an
-$\mathcal{L}$-machine with a small word size and then to check whether the
-same program is correct when run on an $\mathcal{L}$-machine with a
-full-sized word.
-
-The only wrinkle here is that we will sometimes synthesise a program containing
-constants.  If we have synthesised a program with $w=k$,
-the constants in the program will be $k$-bits wide.  To extend the program
-to an $n$-bit machine (with $n > k$), we need some way of deriving $n$-bit-wide
-numbers from $k$-bit ones.  We have several strategies for this and
-just try each in turn.  Our strategies are shown in Fig.~\ref{fig:generalize}.
-$\mathcal{BV}(v, n)$ denotes an $n$-bit wide bitvector holding the value $v$
-and $b \cdotp c$ means the concatenation of bitvectors $b$ and $c$.
-
-\begin{figure}
-
-\centering
-\begin{minipage}[t]{.45\textwidth}
-\begin{eqnarray*}
- \bv{m}{m} & \rightarrow & \bv{n}{n} \\
- \bv{m-1}{m} & \rightarrow & \bv{n-1}{n} \\
- \bv{m+1}{m} & \rightarrow & \bv{n+1}{n}
-\end{eqnarray*}
-\end{minipage}
-\begin{minipage}[t]{.45\textwidth}
-\begin{eqnarray*}
- \bv{x}{m} & \rightarrow & \bv{x}{n} \\
- \bv{x}{m} & \rightarrow & \bv{x}{m} \cdotp \bv{0}{n - m} \\
- \bv{x}{m} & \rightarrow & \underbrace{\bv{x}{m} \cdotp \ldots \cdotp \bv{x}{m}}_{\frac{n}{m} \mathrm{ times}}
-\end{eqnarray*}
-\end{minipage}
-
-\caption{Rules for extending
-an $m$-bit wide number to an $n$-bit wide one.
- \label{fig:generalize}}
-\end{figure}
-
-Sometimes a program will be correct for some particular word width $w$, but is
-not correct for $w' > w$ even if the constants are replaced with appropriate ones.
-When we detect this situation, we increase $w$ and continue synthesising.
-
-\subsubsection{Number of Constants: $c$}
-Instructions in $\mathcal{L}$ take either one or two operands.
-Since any instruction whose operands are all constants can always be
-eliminated (since its result is a constant), we know that a loop-free program
-of minimal length will not contain any instructions with two constant
-operands.  Therefore the number of constants that can appear in
-a minimal program of length $l$ is at most $l$.  By minimising the number
-of constants appearing in a program, we are able to use a particularly
-efficient program encoding that speeds up the synthesis procedure
-substantially.  The number of constants used in a program is the parameter $c$.
-
-$\mathcal{L}$ is an SSA, three-address instruction set\footnote{
-We experimented with implementing $\mathcal{L}$ as a stack machine, expecting
-the programs to be smaller and synthesis to be faster as a result.  We saw
-the opposite effect -- the more complex interpreter led to much slower synthesis.
-}.  Destination registers
-are implicit and a fresh register exists for each instruction to write its
-output to.  A na\"{\i}ve way to encode $\mathcal{L}$ instructions is to have an
-opcode and two operands, where each operand is either a register (i.e., a program argument
-or the result of a previous instruction), or an immediate constant.
-
-In this encoding, each opcode requires $\lceil \log_2 I \rceil$ bits to encode, where $I$ is the number
-of instruction types in $\mathcal{L}$.  Each operand can be encoded using
-$\log_2 w$ bits, where $w$ is the $\mathcal{L}$-machine word width, plus one
-bit to specify whether the operand is a register name or an immediate constant.
-One instruction can therefore be encoded using $\lceil \log_2 I \rceil + 2w + 2$ bits.
-For an $n$-instruction program, we need $$\lceil n \log_2 I \rceil + 2nw + 2n$$ bits to encode
-the entire program.
-
-If we instead limit the number of constants that can appear in the program,
-our operands can be encoded using fewer bits.  For an $n$-instruction program
-using $c$ constants and taking $a$ arguments as inputs, each operand can refer
-to a program argument, the result of a previous instruction or a constant.
-This can be encoded using $\lceil \log_2 (c+a+n-1) \rceil$ bits, which means each instruction
-can be encoded in $\lceil \log_2 I \rceil + \lceil \log_2 (c + a + n - 1) \rceil$ and the full program
-needs $$\lceil n \log_2 I \rceil + \lceil n \log_2 (c + a + n - 1) \rceil + cw$$ bits to encode.
-
-We give an example. Our language $\mathcal{L}$ has 15 instruction types, so each opcode is 4 bits.
-For a 10-instruction program over 1 argument, using 2 constants on a 32-bit word
-machine the first encoding requires $10 * (4 + 32 + 1 + 32 + 1) = 700$ bits.
-Using the second encoding, each operand can be represented using
-$\log_2 (2 + 1 + 10 - 1) = 4$ bits, and the entire program requires 184 bits.
-This is a substantial reduction in size and when the desired program requires
-only few constants this can lead to a very significant speed up.
-
-As with program length, we progressively increase the number of constants in
-our program.  We start by trying to synthesise a program with no constants,
-then if that fails we attempt to synthesise using one constant and so on until we
-reach $c = l$.
-
+\begin{itemize}
+\item{Program Length: $l$}
+%% The first parameter we introduce is program length, denoted by $l$.
+%% At each iteration we synthesise programs of length exactly $l$.
+%% We start with $l = 1$ and increment $l$ whenever we determine
+%% that no program of length $l$ can satisfy the specification.  When we do
+%% successfully synthesise a program, we are \emph{guaranteed that it
+%% is of minimal length} since we have previously established that no
+%% shorter program is correct.
+\item{Word Width: $w$}
+%% An $\mathcal{L}$-program runs on a virtual machine (the $\mathcal{L}$-machine) that
+%% has its own set of parameters.  The only relevant parameter is
+%% the \emph{word width} of the $\mathcal{L}$-machine, that is, the number of bits
+%% in each internal register and immediate constant.  This parameter is denoted by
+%% $w$.  The size of the final SAT problem generated by {\sc cbmc} scales
+%% polynomially with $w$, since each intermediate C variable corresponds
+%% to $w$ propositional variables.
+
+%% It is often the case that a program which satisfies the specification
+%% on an $\mathcal{L}$-machine with $w = k$ will continue to satisfy the
+%% specification when run on a machine with $w > k$.  For example, the program
+%% in Fig.~\ref{fig:bitvector-program} isolates the least-significant bit of a word.
+%% This is true irrespective of the word size of the machine it is run on -- it will
+%% isolate the least-significant bit of an 8-bit word just as well as it will a
+%% 32-bit word.  An often successful strategy is to synthesise a program for an
+%% $\mathcal{L}$-machine with a small word size and then to check whether the
+%% same program is correct when run on an $\mathcal{L}$-machine with a
+%% full-sized word.
+
+%% The only wrinkle here is that we will sometimes synthesise a program containing
+%% constants.  If we have synthesised a program with $w=k$,
+%% the constants in the program will be $k$-bits wide.  To extend the program
+%% to an $n$-bit machine (with $n > k$), we need some way of deriving $n$-bit-wide
+%% numbers from $k$-bit ones.  We have several strategies for this and
+%% just try each in turn.  Our strategies are shown in Fig.~\ref{fig:generalize}.
+%% $\mathcal{BV}(v, n)$ denotes an $n$-bit wide bitvector holding the value $v$
+%% and $b \cdotp c$ means the concatenation of bitvectors $b$ and $c$.
+
+%% \begin{figure}
+
+%% \centering
+%% \begin{minipage}[t]{.45\textwidth}
+%% \begin{eqnarray*}
+%%  \bv{m}{m} & \rightarrow & \bv{n}{n} \\
+%%  \bv{m-1}{m} & \rightarrow & \bv{n-1}{n} \\
+%%  \bv{m+1}{m} & \rightarrow & \bv{n+1}{n}
+%% \end{eqnarray*}
+%% \end{minipage}
+%% \begin{minipage}[t]{.45\textwidth}
+%% \begin{eqnarray*}
+%%  \bv{x}{m} & \rightarrow & \bv{x}{n} \\
+%%  \bv{x}{m} & \rightarrow & \bv{x}{m} \cdotp \bv{0}{n - m} \\
+%%  \bv{x}{m} & \rightarrow & \underbrace{\bv{x}{m} \cdotp \ldots \cdotp \bv{x}{m}}_{\frac{n}{m} \mathrm{ times}}
+%% \end{eqnarray*}
+%% \end{minipage}
+
+%% \caption{Rules for extending
+%% an $m$-bit wide number to an $n$-bit wide one.
+%%  \label{fig:generalize}}
+%% \end{figure}
+
+%% Sometimes a program will be correct for some particular word width $w$, but is
+%% not correct for $w' > w$ even if the constants are replaced with appropriate ones.
+%% When we detect this situation, we increase $w$ and continue synthesising.
+
+\item{Number of Constants: $c$}
+%% Instructions in $\mathcal{L}$ take either one or two operands.
+%% Since any instruction whose operands are all constants can always be
+%% eliminated (since its result is a constant), we know that a loop-free program
+%% of minimal length will not contain any instructions with two constant
+%% operands.  Therefore the number of constants that can appear in
+%% a minimal program of length $l$ is at most $l$.  By minimising the number
+%% of constants appearing in a program, we are able to use a particularly
+%% efficient program encoding that speeds up the synthesis procedure
+%% substantially.  The number of constants used in a program is the parameter $c$.
+
+%% $\mathcal{L}$ is an SSA, three-address instruction set\footnote{
+%% We experimented with implementing $\mathcal{L}$ as a stack machine, expecting
+%% the programs to be smaller and synthesis to be faster as a result.  We saw
+%% the opposite effect -- the more complex interpreter led to much slower synthesis.
+%% }.  Destination registers
+%% are implicit and a fresh register exists for each instruction to write its
+%% output to.  A na\"{\i}ve way to encode $\mathcal{L}$ instructions is to have an
+%% opcode and two operands, where each operand is either a register (i.e., a program argument
+%% or the result of a previous instruction), or an immediate constant.
+
+%% In this encoding, each opcode requires $\lceil \log_2 I \rceil$ bits to encode, where $I$ is the number
+%% of instruction types in $\mathcal{L}$.  Each operand can be encoded using
+%% $\log_2 w$ bits, where $w$ is the $\mathcal{L}$-machine word width, plus one
+%% bit to specify whether the operand is a register name or an immediate constant.
+%% One instruction can therefore be encoded using $\lceil \log_2 I \rceil + 2w + 2$ bits.
+%% For an $n$-instruction program, we need $$\lceil n \log_2 I \rceil + 2nw + 2n$$ bits to encode
+%% the entire program.
+
+%% If we instead limit the number of constants that can appear in the program,
+%% our operands can be encoded using fewer bits.  For an $n$-instruction program
+%% using $c$ constants and taking $a$ arguments as inputs, each operand can refer
+%% to a program argument, the result of a previous instruction or a constant.
+%% This can be encoded using $\lceil \log_2 (c+a+n-1) \rceil$ bits, which means each instruction
+%% can be encoded in $\lceil \log_2 I \rceil + \lceil \log_2 (c + a + n - 1) \rceil$ and the full program
+%% needs $$\lceil n \log_2 I \rceil + \lceil n \log_2 (c + a + n - 1) \rceil + cw$$ bits to encode.
+
+%% We give an example. Our language $\mathcal{L}$ has 15 instruction types, so each opcode is 4 bits.
+%% For a 10-instruction program over 1 argument, using 2 constants on a 32-bit word
+%% machine the first encoding requires $10 * (4 + 32 + 1 + 32 + 1) = 700$ bits.
+%% Using the second encoding, each operand can be represented using
+%% $\log_2 (2 + 1 + 10 - 1) = 4$ bits, and the entire program requires 184 bits.
+%% This is a substantial reduction in size and when the desired program requires
+%% only few constants this can lead to a very significant speed up.
+
+%% As with program length, we progressively increase the number of constants in
+%% our program.  We start by trying to synthesise a program with no constants,
+%% then if that fails we attempt to synthesise using one constant and so on until we
+%% reach $c = l$.
+
+\end{itemize}
 \subsubsection{Searching the Program Space}

### 2014-07-02 papers/termination/paper.tex
Some nontermination

@@
 
+Dually to termination, we might want to consider the non-termination of a program.  If a program terminates,
+we can prove this by finding a ranking function and supporting invariant witnessing the satisfiability
+of formula {\bf[CT]}.  What then would a proof of non-termination look like?
+
 Gupta et al. \cite{DBLP:conf/popl/GuptaHMRX08} characterize non-termination of a transition relation $T$ by the existence of an \emph{(open) recurrence set},
@@
 we make use of the notion of \emph{closed recurrence set} introduced by Chen et al. in \cite{DBLP:conf/tacas/ChenCFNO14}. 
-As opposed to a closed recurrence set, this is a  purely universal property: for each state in the recurrence set $N$, each of its successors 
-must be in $N$. Thus, the non-termination problem becomes equivalent to the
+As opposed to a closed recurrence set, this is a $\exists \forall$ property: for each state in the recurrence set $N$, all of its successors 
+must be in $N$. We note that if $T$ is deterministic, every open recurrence set is also a closed recurrence set (since each
+state has at most one successor).  Thus, the non-termination problem for deterministic transition systems is equivalent to the
 satisfiability of Definition~\ref{def:deterministic-nonterm-formula}.
@@
 
-\section{Genetic Programming and Incremental Evolution}
+\subsection{Genetic Programming and Incremental Evolution}
 

### 2014-07-02 papers/termination/paper.tex
Nontermination

@@
 
-Given that our state space is finite, a transition relation induces a infinite execution if and only if some state is 
-visited twice in the course of the execution. Hence, an alternative formulation of the non-termination problem 
-for bit-vectors may search for a state that is visited twice.
-However, this would require reasoning about the transitive closure of the transition relation.\\
+Given that our program's state space is finite, a transition relation induces a infinite execution if and only if some state is 
+visited infinitely often, or equivalently if the state is visitied twice in the course of the execution.
+This gives us an alternative way of expression a non-termination proof: we can can try to find a witness $x$ to the satisfiability
+of the formula
+$$ \exists x . T^+(x, x)$$
+However, deciding satisfiability of this formula directly requires a logic that includes a transitive closure operator, which
+ours does not have.\\
 

### 2014-07-02 papers/termination/paper.tex
Move section 5 around a bit

@@
 \section{Termination as Second-Order Satisfaction} \label{sec:second.order}
-Program model? Programs: transition systems? Constraint generation? No CFG?
+%Program model? Programs: transition systems? Constraint generation? No CFG?
 
@@
 
-\subsection{The Termination Problem}
+\subsection{An Isolated Loop}
+
+We will begin our discussion by showing how to encode the (non-)termination of a single loop
+in the synthesis fragment.  For the time being, a loop is a pair ${L= \langle G,T \rangle}$,
+where the states $x$ satisfying the loop's guard are given by the predicate $G(x)$.
+The body of the loop is encoded as the transition relation $T(x, x')$, meaning that
+state $x'$ is reachable from state $x$ via a single iteration of the loop body.
+For example, the loop in Figure~\ref{fig:motivation.a} is encoded as:
+\begin{align*}
+G(x) & = \{ x \mid x>0 \} \\
+T(x,x') &= \{ \langle x, x' \rangle \mid x' = (x - 1) \, \& \, x \}
+\end{align*}
+We will abbreviate this with the notation:
+\begin{align*}
+G(x) & \triangleq x > 0 \\
+T(x, x') & \triangleq x' = (x - 1) \, \& \, x
+\end{align*}
 
@@
 
-\begin{definition}[Conditional Termination Formula {\bf [CT]}]
-\label{def:conditional-termination-formula}
- \begin{align*}
-  \exists R, Inv . \forall x, x' . & I(x) \wedge G(x) \rightarrow Inv(x) ~ \wedge \\
-                                 & G(x) \wedge Inv(x) \wedge T(x, x') \rightarrow Inv(x') \wedge R(x) > 0 
-  \wedge R(x) > R(x')
- \end{align*}
-\end{definition}
-
 \begin{definition}[Non-Termination Formula - Open Recurrence Set  {\bf [ONT]}]
@@
 
-\begin{figure*}
- \begin{framed}
- \begin{definition}[Sequential Loops Termination Formula {\bf [ST]}]
-  \begin{align*}
-\label{def:multi-termination-formula}
-  \exists R, Inv_1,..., Inv_n . \forall x_0,...,x_n, x_1',...,x_n'.  & P_0(x_0,x_1) \rightarrow Inv_1(x_1) ~  \\
- & \bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) ~  \\  
- & \bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) \rightarrow Inv_{i+1}(x_{i+1})) ~ \wedge \\
- & Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge G_n(x_n) \rightarrow Inv_n(x_n) ~ \wedge \\
- & Inv_n(x_n) \wedge G_n(x_n) \wedge T(x_n, x_n') \rightarrow Inv_n(x_n') \wedge R(x_n) > 0 \wedge R(x_n) > R(x_n')
- \end{align*}
-\end{definition}
-
-\begin{definition}[ Nested Loops Termination Formula {\bf [NT]}]
-\label{def:nested-term-formula}
- \begin{align*}
-  \exists R, S . \forall x, x' . & G_1(x) \wedge P_1(x,x') \rightarrow S(x',x') ~ \wedge \\
-                                & G_2(x') \wedge S(x,x') \wedge T(x',x'')\rightarrow S(x,x'') ~ \wedge \\
-				& G_1(x) \wedge P_1(x,x') \wedge \neg G_2(x') \wedge S(x',x'') \wedge 
-                                P_2(x'',x''') \wedge 
-                                  R(x) > 0 \wedge R(x) > R(x''') 
- \end{align*} 
-\end{definition}
-
-\begin{definition}[Sequential Loops Non-Termination Formula {\bf [SNT]}]
-\label{def:multi-termination-formula}
- \begin{align*}
-  \exists R, Inv_1,..., Inv_n . \forall x_0,...,x_n, x_1',...,x_n'.  & P_0(x_0,x_1) \rightarrow Inv_1(x_1) ~  \\
- & \bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) ~  \\  
- & \bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) \rightarrow Inv_{i+1}(x_{i+1})) ~ \wedge \\
- & Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge N(x_n)  ~ \wedge \\
-                                                        & N(x_n) \rightarrow G_n(x_n) ~ \wedge 
-							 N(x_n) \wedge T_n(x_n, x_n') \rightarrow N(x_n') 
- \end{align*}
-\end{definition}
- \end{framed}
-
-\end{figure*}
-
 \iffalse
@@
 \noindent {\bf Unconditional termination.}
-We start by considering a program consisting of a single loop ${L= \langle G,T \rangle}$,
-where the states $x$ satisfying the loop's guard are given by the predicate $G(x)$.
-The body of the loop is encoded as the transition relation $T(x, x')$, meaning that
-state $x'$ is reachable from state $x$ via a single iteration of the loop body.
-%where the initial states of the program 
-%just before the loop are given as an assertion init(V )
-%satisfy $I : X \rightarrow \mathbb{B}$, $G : X \rightarrow \mathbb{B} $ is the loop guard and $T : X \rightarrow X$ is the loop body. 
-%We use predicates over program states $x$ and their successors $x'$ to model program
-%statements as binary relations over states. For loop $L$, $T(x, x')$
-%represents the body of the loop. 
-%
-%Subsequently, we show how to generalise our formulation of termination and non-termination to handle sequential loops. \\
-%
-The existence of a ranking function for $L$ is equivalent to the satisfiability
-of the formula from Definition~\ref{def:termination-formula}.  If this formula is satisfiable,
-then there exists a ranking function $R$ that proves loop $L$ terminates.
-
-Our formalism is illustrated by the unconditionally terminating program shown
-in Figure~\ref{fig:motivation.a}, which is encoded as:
-\begin{align*}
-G(x) & = \{ x \mid x>0 \} \\
-T(x,x') &= \{ \langle x, x' \rangle \mid x' = (x - 1) \, \& \, x \}
-\end{align*}
-
-We will abbreviate this with the notation $G(x) \triangleq x > 0$ and $T(x, x') \triangleq x' = (x - 1) \, \& \, x$.
-The corresponding synthesis formula {\bf [UT]} is satisfiable, as witnessed by the function $R(x) = x$.\\
-
-\noindent {\bf Conditional termination.}
-A slightly more involved case occurs if $L$'s termination depends on the state it begins
-executing in, i.e. $L$ is \emph{conditionally terminating}. 
-The loop we consider now is $L=\langle I,G,T \rangle$ and has an additional predicate $I(x)$ 
-representing the states in which the loop can begin executing.
-Conditional termination is equivalent to the
-formula of Defintion~\ref{def:conditional-termination-formula}. If this formula is satisfiable,
-two witnesses are returned:
-\begin{itemize}
-\item $W$ is an inductive invariant of $L$ that is established by the initial states $I$ if the loop
-guard $G$ is met.
-\item $R$ is a ranking function for $L$ as restricted by $W$ -- that is to say, $R$ need only
-be well founded on those states satisfying $W \wedge G$.  Since $W$ is an inductive invariant of $L$,
-$R$ is strong enough to show that $L$ terminates from any of its initial states.
-\end{itemize}
-
-$W$ is called a \emph{supporting invariant} for $L$ and $R$ proves termination relative to $W$.
-
-Conditional termination is illustrated by the loop in Figure~\ref{fig:motivation.h}, 
-which is encoded as:
-\begin{align*}
-            I(\langle x, y \rangle) & \triangleq y = 1 \\
-            G(\langle x, y \rangle) & \triangleq x > 0 \\
-            T(\langle x, y \rangle, \langle x', y' \rangle) & \triangleq x' = x - y \wedge y' = y \\
-\end{align*}
-
-If the initial states $I$ are ignored this loop cannot be shown to terminate, since any state with $y = 0$ and $x > 0$
-would lead to a non-terminating execution.
-
-However, formula {\bf [CT]} is satisfiable, as witnessed by the ranking function
-$R(\langle x,y\rangle ) = x$ and supporting invariant $W(\langle x, y \rangle ) \triangleq y = 1$.
-This constitutes a proof that the program as whole terminates, since the loop always begins
-executing in a state that guarantees its termination.
-
-\subsection{The Non-Termination Problem}
-
+To show that $L$ is unconditionally terminating (i.e. it eventually terminates regardless of the
+state it starts in), we can find a ranking function for $L$.
+The existence of a ranking function is equivalent to the satisfiability
+of the formula from Definition~\ref{def:termination-formula}, so a witness to the satisfiability
+of formula {\bf [UT]} is also a ranking function and thus a proof of $L$'s unconditional termination.
+
+Continuing the example of the program in Figure~\ref{fig:motivation.a}, we can see that
+the corresponding synthesis formula {\bf [UT]} is satisfiable, as witnessed by the function $R(x) = x$.
+So $R(x) = x$ constitutes a proof that the program in Figure~\ref{fig:motivation.a} is unconditionally
+terminating.
+
+\noindent{\bf Non-termination.}
 Dually to termination, we might want to consider the non-termination of a program.  If a program terminates,
@@
 exists a transition to some $s'\in N$. 
-For our case, given a loop $L = \langle I,G,T \rangle$, the notion of open recurrent set is encoded by Definition~\ref{def:nonterm-formula}.
+For our case, given a loop $L = \langle I,G,T \rangle$, the notion of open recurrent set is encoded by formula {\bf [ONT]}. %Definition~\ref{def:nonterm-formula}.
 
@@
 state has at most one successor).  Thus, the non-termination problem for deterministic transition systems is equivalent to the
-satisfiability of Definition~\ref{def:deterministic-nonterm-formula}.
-As our formulation is with respect to a given loop $L=\langle I,G,t \rangle$, we require the  
-recurrence set to include the loop's guard, $N(x) \rightarrow G(x)$. In other words, the loop's guard needs to be satisfied in order for the 
-infinite execution to happen.
+satisfiability of formula {\bf [CNT]}. %Definition~\ref{def:deterministic-nonterm-formula}.
+%As our formulation is with respect to a given loop $L=\langle I,G,T \rangle$, we require the  
+%recurrence set to include the loop's guard, $N(x) \rightarrow G(x)$. In other words, the loop's guard needs to be satisfied in order for the 
+%infinite execution to happen.
 
 Given that our program's state space is finite, a transition relation induces a infinite execution if and only if some state is 
-visited infinitely often, or equivalently if the state is visitied twice in the course of the execution.
+visited infinitely often, or equivalently the state is visited twice in the course of the execution.
 This gives us an alternative way of expression a non-termination proof: we can can try to find a witness $x$ to the satisfiability
@@
 %Our method is purely semantic.
+
+\subsection{Composing a Loop with its Environment}
+
+\begin{figure*}
+ \begin{framed}
+
+\begin{definition}[Conditional Termination Formula {\bf [CT]}]
+\label{def:conditional-termination-formula}
+ \begin{align*}
+  \exists R, W . \forall x, x' . & I(x) \wedge G(x) \rightarrow W(x) ~ \wedge \\
+                                 & G(x) \wedge W(x) \wedge T(x, x') \rightarrow W(x') \wedge R(x) > 0 
+  \wedge R(x) > R(x')
+ \end{align*}
+\end{definition}
+
+ \begin{definition}[Sequential Loops Termination Formula {\bf [ST]}]
+  \begin{align*}
+\label{def:multi-termination-formula}
+  \exists R, Inv_1,..., Inv_n . \forall x_0,...,x_n, x_1',...,x_n'.  & P_0(x_0,x_1) \rightarrow Inv_1(x_1) ~  \\
+ & \bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) ~  \\  
+ & \bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) \rightarrow Inv_{i+1}(x_{i+1})) ~ \wedge \\
+ & Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge G_n(x_n) \rightarrow Inv_n(x_n) ~ \wedge \\
+ & Inv_n(x_n) \wedge G_n(x_n) \wedge T(x_n, x_n') \rightarrow Inv_n(x_n') \wedge R(x_n) > 0 \wedge R(x_n) > R(x_n')
+ \end{align*}
+\end{definition}
+
+\begin{definition}[Nested Loops Termination Formula {\bf [NT]}]
+\label{def:nested-term-formula}
+ \begin{align*}
+  \exists R, S . \forall x, x' . & G_1(x) \wedge P_1(x,x') \rightarrow S(x',x') ~ \wedge \\
+                                & G_2(x') \wedge S(x,x') \wedge T(x',x'')\rightarrow S(x,x'') ~ \wedge \\
+				& G_1(x) \wedge P_1(x,x') \wedge \neg G_2(x') \wedge S(x',x'') \wedge 
+                                P_2(x'',x''') \wedge 
+                                  R(x) > 0 \wedge R(x) > R(x''') 
+ \end{align*} 
+\end{definition}
+
+\begin{definition}[Sequential Loops Non-Termination Formula {\bf [SNT]}]
+\label{def:multi-termination-formula}
+ \begin{align*}
+  \exists R, Inv_1,..., Inv_n . \forall x_0,...,x_n, x_1',...,x_n'.  & P_0(x_0,x_1) \rightarrow Inv_1(x_1) ~  \\
+ & \bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) ~  \\  
+ & \bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) \rightarrow Inv_{i+1}(x_{i+1})) ~ \wedge \\
+ & Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge N(x_n)  ~ \wedge \\
+                                                        & N(x_n) \rightarrow G_n(x_n) ~ \wedge 
+							 N(x_n) \wedge T_n(x_n, x_n') \rightarrow N(x_n') 
+ \end{align*}
+\end{definition}
+ \end{framed}
+
+\end{figure*}
+
+Sometimes the termination behaviour of a loop depends on its environment.  That is to say,
+the loop may not terminate if started in some particular state, but the rest of the program
+ensures that such a state cannot be reached on entry to the loop.  The program as a whole
+terminates, but if the loop were considered in isolation we would not be able to prove that
+it terminates.  We will now consider the problem of encoding the loop's environment as
+a synthesis formula.
+
+\noindent {\bf Conditional termination.}
+If $L$'s termination depends on the state it begins
+executing in, we say that $L$ is \emph{conditionally terminating}. 
+The loop we consider now is $L=\langle I,G,T \rangle$; it has an additional predicate $I(x)$ 
+representing the states in which the loop can begin executing.
+Conditional termination is equivalent to the
+formula of Defintion~\ref{def:conditional-termination-formula}. If this formula is satisfiable,
+two witnesses are returned:
+\begin{itemize}
+\item $W$ is an inductive invariant of $L$ that is established by the initial states $I$ if the loop
+guard $G$ is met.
+\item $R$ is a ranking function for $L$ as restricted by $W$ -- that is to say, $R$ need only
+be well founded on those states satisfying $W \wedge G$.  Since $W$ is an inductive invariant of $L$,
+$R$ is strong enough to show that $L$ terminates from any of its initial states.
+\end{itemize}
+
+$W$ is called a \emph{supporting invariant} for $L$ and $R$ proves termination relative to $W$.
+
+Conditional termination is illustrated by the program in Figure~\ref{fig:motivation.h},
+which is encoded as:
+\begin{align*}
+            I(\langle x, y \rangle) & \triangleq y = 1 \\
+            G(\langle x, y \rangle) & \triangleq x > 0 \\
+            T(\langle x, y \rangle, \langle x', y' \rangle) & \triangleq x' = x - y \wedge y' = y 
+\end{align*}
+If the initial states $I$ are ignored this loop cannot be shown to terminate, since any state with $y = 0$ and $x > 0$
+would lead to a non-terminating execution.
+
+However, formula {\bf [CT]} is satisfiable, as witnessed by the ranking function
+$R(\langle x,y\rangle ) = x$ and supporting invariant $W(\langle x, y \rangle ) \triangleq y = 1$.
+This constitutes a proof that the program as a whole terminates, since the loop always begins
+executing in a state that guarantees its termination.
 

### 2014-07-02 papers/termination/kalashnikov.tex
working on the kalashnikov section

@@
 arithmetic modulo fixed widths.
-
+% 
 %We can solve the synthesis problem using any off-the-shelf program synthesiser.  If you don't
@@
 
-When the quantifier free first-order fragment of the logic is efficiently decidable and
+\subsection{\newC}
+\label{sec:logic}
+In order to obtain bit-level accuracy, the logic we use to express the synthesis formula is a subset of C that we call \newC.  
+The characteristic property of a
+\newC  program is that safety can be decided for
+it using a single query to a Bounded Model Checker.  A \newC program is
+just a C program with the following syntactic restrictions:
+ all loops in the program must have a constant bound;
+ all recursion in the program must be limited to a constant depth;
+ all arrays must be statically allocated (i.e. not using \texttt{malloc}),
+ and be of constant size.
+Additionally, \newC programs may use nondeterministic values, assumptions
+and arbitrary-width types.
+
+The proofs found by our solver are arbitrary expressions over the grammar given in Figure~\ref{fig:l-language}.  
+%written in a simple RISC-like language $\mathcal{L}$, whose syntax is given in Fig.~\ref{fig:l-language}.  
+\todo{update the figure and make it look more like a grammar}
+
+As the quantifier free first-order fragment of the logic is efficiently decidable and
 $P$ is expressible as a ground term of the logic, we can use
 Counterexample Guided Inductive Synthesis (CEGIS)~\cite{lezama-thesis,sketch} to
-check the validity of the synthesis formula. The core of the CEGIS algorithm is
+check the validity of the synthesis formula.
+
+The core of the CEGIS algorithm is
 the refinement loop shown in Fig.~\ref{fig:abstract-refinement} and
 detailed in Algorithm~\ref{fig:abstract-refinement-code}.  
-
 The algorithm is divided into two
 components: {\sc synth} and {\sc verif}. For a finite set of test
-inputs that we track, {\sc synth} synthesises a candidate program that is correct on just those tests.
-Subsequently, {\sc Verif} tries to verify the candidate program by searching for an input on
-which the candidate program does not satisfy the specification.  If no such input
+inputs that we track, {\sc synth} synthesises a candidate program that is correct for those tests.
+Subsequently, {\sc Verif} verifies the candidate program by searching for an input on
+which it does not satisfy the specification. If no such input
 can be found, the candidate program is correct.  Otherwise, the new input is added
-to the setof tracked inputs.  
+to the set of tracked inputs.  
 %If the input set is finite, this procedure is guaranteed to terminate.
@@
 
-%\label{sec:logic}
-
-The logic used to express our synthesis formula is \newC discussed in Section~\ref{sec:logic}.
-%% The logic is a subset of C that we call \newC.  
-%% The characteristic property of a
-%% \newC  program is that safety can be decided for
-%% it using a single query to a Bounded Model Checker.  A \newC program is
-%% just a C program with the following syntactic restrictions:
-%%  all loops in the program must have a constant bound;
-%%  all recursion in the program must be limited to a constant depth;
-%%  all arrays must be statically allocated (i.e. not using \texttt{malloc}),
-%%  and be of constant size.
-%% Additionally, \newC programs may use nondeterministic values, assumptions
-%% and arbitrary-width types.
-
-%% \iffalse
-%% Two example \newC programs are shown 
-%% in Fig.~\ref{fig:c-}.
-
-%% \begin{figure}
-%% \begin{minipage}[scale=0.8]{0.45\linewidth}
-%%  \begin{lstlisting}[language=c]
-%% int count_bits(int x) {
-%%   int i, ret = 0;
-
-%%   for (i = 0; i < 32; i++)
-%%     if (x & (1 << i))
-%%       ret++;
-
-%%   return ret;
-%% }
-%%  \end{lstlisting}
-%% \end{minipage}
-%% \begin{minipage}{0.54\linewidth}
-%%  \begin{lstlisting}[language=C]
-%% int common_factor(int A[10]) {
-%%   int i, factor = nondet();
-
-%%   for (i = 0; i < 10; i++)
-%%     assume((A[i] % factor) == 0);
-
-%%   assume(factor > 1);
-%%   return factor;
-%% }
-
-%%  \end{lstlisting}
-%% \end{minipage}
-
-%%  \caption{Two \newC programs}
-%%  \label{fig:c-}
-
-%% \end{figure}
-%% \fi
-
-%% Since each loop is bounded by a constant, and each recursive function call is
-%% limited to a constant depth, a \newC program necessarily terminates and in
-%% fact does so in $O(1)$ time.  If we call the largest loop bound~$k$, then
-%% a Bounded Model Checker with an unrolling bound of $k$ will be a complete
-%% decision procedure for the safety of the program.  For a \newC program of
-%% size $l$ and with largest loop bound~$k$, a Bounded Model Checker will
-%% create a SAT problem of size $O(lk)$.  Conversely, a SAT problem
-%% of size $s$ can be converted trivially into a loop-free \newC program
-%% of size $O(s)$.  The safety problem for \newC is therefore NP-complete,
-%% which means it can be decided fairly efficiently for many practical
-%% instances.
-
-\subsection{Encoding into \newC}
 %% To instantiate the abstract synthesis algorithm in \newC we must express $I, O,
@@
 high-complexity language.
+
+---
+We require the user to supply a program specification written in C, but then
+ask nothing more of her.  We parametrise the space of programs in such a way
+that it can be explored automatically, rather than asking a human for hints.
+Our program encoding generates an $\exists
+\forall$ formula that is \emph{linear} in the length of the shortest program
+satisfying the user's specification.  This formula is then checked for satisfiability
+using the CEGIS algorithm.
+---
+
+We introduce a novel parametrisation of the programming language
+used to express our synthesised programs.
+This parametrisation allows us to efficiently explore the program space
+without relying on human guidance and also ensures that our programs
+are of minimal length.
+
+Our tool is the first we are aware of that is able to effectively
+synthesise floating-point programs.  We demonstrate this by
+synthesising {\sc Fast2Sum} using Knuth's {\sc 2Sum}~\cite{taocp2} as
+a specification.
+------
 
@@
 
-\begin{figure}[t]
-\centering
-\begin{tikzpicture}[scale=0.65, transform shape, node distance=2cm, auto]
-\tikzstyle{decision} = [diamond, draw, fill=blue!20, 
-    text width=4.5em, text badly centered, node distance=3cm, inner sep=0pt]
-\tikzstyle{block} = [rectangle, draw, fill=blue!20, 
-    text width=5em, text centered, rounded corners, minimum height=4em]
-\tikzstyle{line} = [draw, -latex']
-\tikzstyle{cloud} = [draw, ellipse,fill=red!20, node distance=3cm,
-    minimum height=2em]
-
- \node [decision] (synthsucc) {{\sc Synth} succeeds?};
-
- \node [decision, below of=synthsucc, node distance=2.5cm] (verif) {{\sc Verif} succeeds?};
- \node [decision, right of=verif] (ck) {$c < l$?};
-
- \node [block, below of=verif, node distance=3cm]   (done) {Done!};
- \node [decision, left of=verif] (verifw) {{\sc Verif} succeeds for small words?};
-
- \node [block, below of=ck, node distance=2.5cm] (incc) {$c := c+1$};
- \node [block, right of=incc, node distance=2cm] (incl) {$c := 0$\\ $l := l+1$};
-
- \node [decision, below of=verifw, node distance=3cm] (gen) {Extend succeeds?};
- \node [block, left of=verifw, node distance=3cm] (iterate) {Parameters unchanged};
-
- \node [block, left of=gen, node distance=3cm] (incw) {$w := w+1$};
-
- \path [line] (synthsucc) -- node [left] {Yes} (verif);
- \path [line] (synthsucc) -| node [above, near start] {No} (ck);
-
- \path [line] (verif) -- node [left] {Yes} (done);
- \path [line] (verif) -- node [above, near start] {No} (verifw);
-
- \path [line] (ck) -- node [left] {Yes} (incc);
- \path [line] (ck) -| node  [above, near start]  {No} (incl);
-
- \path [line] (verifw) -- node [right] {Yes} (gen);
- \path [line] (verifw) -- node [above] {No} (iterate);
-
- \path [line] (gen) -- node [below] {Yes} (done);
- \path [line] (gen) -- node [below] {No} (incw);
-
- %\path [dotted, line] (iterate.west) |- (synthsucc);
-\end{tikzpicture}
-
- \caption{Decision tree for increasing parameters of $\mathcal{L}$.}
- \label{fig:paramsflow}
-
-\end{figure}
+%% \begin{figure}[t]
+%% \centering
+%% \begin{tikzpicture}[scale=0.65, transform shape, node distance=2cm, auto]
+%% \tikzstyle{decision} = [diamond, draw, fill=blue!20, 
+%%     text width=4.5em, text badly centered, node distance=3cm, inner sep=0pt]
+%% \tikzstyle{block} = [rectangle, draw, fill=blue!20, 
+%%     text width=5em, text centered, rounded corners, minimum height=4em]
+%% \tikzstyle{line} = [draw, -latex']
+%% \tikzstyle{cloud} = [draw, ellipse,fill=red!20, node distance=3cm,
+%%     minimum height=2em]
+
+%%  \node [decision] (synthsucc) {{\sc Synth} succeeds?};
+
+%%  \node [decision, below of=synthsucc, node distance=2.5cm] (verif) {{\sc Verif} succeeds?};
+%%  \node [decision, right of=verif] (ck) {$c < l$?};
+
+%%  \node [block, below of=verif, node distance=3cm]   (done) {Done!};
+%%  \node [decision, left of=verif] (verifw) {{\sc Verif} succeeds for small words?};
+
+%%  \node [block, below of=ck, node distance=2.5cm] (incc) {$c := c+1$};
+%%  \node [block, right of=incc, node distance=2cm] (incl) {$c := 0$\\ $l := l+1$};
+
+%%  \node [decision, below of=verifw, node distance=3cm] (gen) {Extend succeeds?};
+%%  \node [block, left of=verifw, node distance=3cm] (iterate) {Parameters unchanged};
+
+%%  \node [block, left of=gen, node distance=3cm] (incw) {$w := w+1$};
+
+%%  \path [line] (synthsucc) -- node [left] {Yes} (verif);
+%%  \path [line] (synthsucc) -| node [above, near start] {No} (ck);
+
+%%  \path [line] (verif) -- node [left] {Yes} (done);
+%%  \path [line] (verif) -- node [above, near start] {No} (verifw);
+
+%%  \path [line] (ck) -- node [left] {Yes} (incc);
+%%  \path [line] (ck) -| node  [above, near start]  {No} (incl);
+
+%%  \path [line] (verifw) -- node [right] {Yes} (gen);
+%%  \path [line] (verifw) -- node [above] {No} (iterate);
+
+%%  \path [line] (gen) -- node [below] {Yes} (done);
+%%  \path [line] (gen) -- node [below] {No} (incw);
+
+%%  %\path [dotted, line] (iterate.west) |- (synthsucc);
+%% \end{tikzpicture}
+
+%%  \caption{Decision tree for increasing parameters of $\mathcal{L}$.}
+%%  \label{fig:paramsflow}
+
+%% \end{figure}
 

### 2014-07-02 papers/termination/paper.tex
working on the kalashnikov section

--- 2014-07-02 papers/termination/paper.tex
@@
 
+\todo{do something with this?}
+Since each loop is bounded by a constant, and each recursive function call is
+limited to a constant depth, a \newC program necessarily terminates and in
+fact does so in $O(1)$ time.  If we call the largest loop bound~$k$, then
+a Bounded Model Checker with an unrolling bound of $k$ will be a complete
+decision procedure for the safety of the program.  For a \newC program of
+size $l$ and with largest loop bound~$k$, a Bounded Model Checker will
+create a SAT problem of size $O(lk)$.  Conversely, a SAT problem
+of size $s$ can be converted trivially into a loop-free \newC program
+of size $O(s)$.  The safety problem for \newC is therefore NP-complete,
+which means it can be decided fairly efficiently for many practical
+instances.
+
 \iffalse

### 2014-07-03 papers/termination/kalashnikov.tex
revised more of the kalashnikov section

@@
 
-\iffalse
-\section{Introduction}
+%\todo{add related work for synthesis.}
 
-Program synthesis is the mechanised construction of software that provably
-satisfies a given specification.  Synthesis tools promise to relieve the
-programmer from thinking about \emph{how} the problem is to be solved;
-instead, the programmer only provides a compact description of \emph{what}
-is to be achieved.  Foundational research in this area has been
-exceptionally fruitful, beginning with Alonzo Church's work on the
-\emph{Circuit Synthesis Problem} in the sixties~\cite{church-synth}.
-Algorithmic approaches to the problem have frequently been connected to
-automated theorem proving~\cite{manna-waldinger,proof-planning}.
-Recent developments include an application of Craig interpolation
-to synthesis~\cite{synth-interpolants}.
+In this section we present a solver for the synthesis fragment:
+ \[
+  \exists P,~ x . ~\forall~ y . \sigma(P, x, y)
+ \]
+%
+where $P$ ranges over sets, while $x$ and $y$ range over ground terms (as mentioned before, for brevity reasons we drop
+ the vector symbol and write  $x$ for $\vec{x}$).
 
-In this paper, we present a new method that brings the promise of
-automating programming closer to application.  We make pragmatic
-restrictions to the general synthesis problem in order to arrive at an
-effective solution.  To this end, we focus on the problem of synthesising
-loop-free programs over machine integers and floating-point numbers from a
-specification given as program fragment.  Our exploration of the program
-space ensures that our programs are minimal in size.
+By viewing $\sigma: (X^n \times Y^m \to Z^k) \times X^n \times Y^m  \to \mathbb{B}$ as a specification function
+that  returns true iff $P$ computes appropriate outputs for the inputs $x$ and $y$, we can   
+regard the satisfaction problem for the synthesis fragment as \emph{program synthesis}, i.e.
+the mechanised construction of software that provably satisfies a given specification.
+Research in the area of program synthesis has been particularly fruitful beginning with Alonzo Church's work on the
+\emph{Circuit Synthesis Problem} in the sixties~\cite{church-synth}, and continuing with works such as 
+{\sc Brahma}~\cite{brahma} and program sketching~\cite{lezama-thesis,sketch,modular-sketch}.
 
-Our work represents a substantial refinement of {\sc brahma}~\cite{brahma},
-which is the closest work to ours.  {\sc Brahma} also focuses on
-synthesising loop-free machine-integer programs.  In~\cite{brahma}, the
-authors characterise the program synthesis problem as a second-order
-$\exists \forall$ formula and use a variation of the CEGIS algorithm,
-introduced in~\cite{lezama-thesis}, for solving
-these formulae.  This algorithm reduces the satisfiability of the $\exists
-\forall$ formula to the satisfiability of several first-order formulae, each
-of which can be decided using an off-the-shelf SMT solver.  {\sc Brahma}
-takes as input a specification (given as an SMT formula) and a library of
-components (one component is more or less a single machine instruction).  It
-then finds a permutation of the component library that satisfies the
-specification.  In the
-event that the library does not contain enough components to synthesise a
-correct program, the user is told that the library is too small and
-must add more components until the synthesis succeeds.  Each round of
-synthesis amounts to solving an $\exists \forall$ formula that is quadratic
-in the size of the component library.
-
-We require the user to supply a program specification written in C, but then
-ask nothing more of her.  We parametrise the space of programs in such a way
-that it can be explored automatically, rather than asking a human for hints.
-Our program encoding generates an $\exists
-\forall$ formula that is \emph{linear} in the length of the shortest program
-satisfying the user's specification.  This formula is then checked for satisfiability
-using the CEGIS algorithm.  In contrast, the program generated by {\sc
-brahma} can be no longer than the size of the library, as the synthesis
-outcome is a permutation of the component library.  Our synthesis
-formulae are therefore asymptotically smaller than {\sc brahma}'s.
-
-\paragraph{Related Work}
-
-Another successful approach to program synthesis has been \emph{program
-sketching}, in particular the {\sc sketch} tool~\cite{lezama-thesis,sketch,modular-sketch}.  This method
-requires the programmer to provide a \emph{program sketch}, written in a
-C-like language, together with either a specification (written in the same
-language) or a suite of unit tests.  The sketch is a skeleton program
-containing much of the high-level structure of the final program, but with
-several ``holes'', which {\sc sketch} fills in to meet the specification.
-Despite the many similarities, our problem domains are quite different:
-sketching is able to make use of human insight to synthesise fairly large
-programs, whereas we are optimised for synthesising small (but intricate)
-programs without guidance.  A sketching-like approach to optimising
-fixed point arithmetic code was described in~\cite{eldib-wang}.  This
-work is similar in goal to our floating-point work, but is more restricted
-in scope in that it only tries to minimise register widths for embedded
-systems using fixed point arithmetic.
-
-Program synthesis is closely related to
-superoptimisation, which is the problem of taking a
-segment of code, usually written in C, and finding a minimal program which
-has the same functionality.  The {\sc aha}~\cite{aha-paper} tool, the GNU
-superoptimiser~\cite{gnu-superoptimiser} and TOAST~\cite{toast} all perform superoptimisation of C
-programs by enumerating possible programs and testing them on suitable
-inputs.  This search can be compiled down to native code using an optimising
-compiler, so for short programs the method is extremely fast.  However, this
-method is impractical for synthesising longer programs because of the
-exponential blowup of the search space.  We incorporate the core idea of
-this technique by using explicit-state model checking in parallel with
-symbolic model checking.
-
-\paragraph{Technical Contributions}
-Our approach generates an $\exists \forall$ formula that is linear
-in the size of the synthesised program.  This improves over the
-encoding of~\cite{brahma}, which generates a $\exists \forall$ formula
-that is quadratic in the size of the synthesised program.
-
-We introduce a novel parametrisation of the programming language
-used to express our synthesised programs.
-This parametrisation allows us to efficiently explore the program space
-without relying on human guidance and also ensures that our programs
-are of minimal length.
-
-Our tool is the first we are aware of that is able to effectively
-synthesise floating-point programs.  We demonstrate this by
-synthesising {\sc Fast2Sum} using Knuth's {\sc 2Sum}~\cite{taocp2} as
-a specification.
-
-\paragraph{Outline} We describe the abstract synthesis algorithm in
-Sec.~\ref{sec:abstract-algorithm} and give details of our new algorithm in
-Sec.~\ref{sec:concrete-algorithm}.  We use a fragment of the C programming language
-as our underlying logic, which is elaborated in Sec.~\ref{sec:logic}. We
-summarise the results of our experiments in Sec.~\ref{sec:experiments}.
-\fi
-
-\iffalse
-\subsection{The Basic Synthesis Algorithm}
-\label{sec:abstract-algorithm}
-
-\subsubsection{The Synthesis Formula}
-\label{sec:abstract-formula}
-
-Our task is to find a program that satisfies some specification.  We can formalise this
-notion as follows: fix an input set $I$ and an output set $O$.  Specifications $\sigma$
-relate inputs and outputs, and programs $P$ map an input to an output:
-%
-$$ \sigma \subseteq I \times O $$
-$$ P : I \rightarrow O$$
-
-The synthesis problem is then to determine the validity of the following
-second-order formula, and if so, to find a suitable witness program $P$:
-%
-$$\exists P .\, \forall x \in I.\, \sigma(x, P(x))$$
-
-Depending on the logic needed to express this formula the synthesis problem
-may be decidable, semi-decidable or undecidable.  For many interesting
-logics, checking the validity of a second-order formula is undecidable. 
-Even for logics in which the problem is decidable, the quantifier
-alternation means that we may not have an efficient decision procedure for
-the formula above, despite efficient methods existing for problems
-with only one quantifier.  In the case of propositional
-satisfiability, we have SAT solvers which are efficient-in-practice at
-solving queries with a single quantifier.  The corresponding problem with
-one level of quantifier alternation (2-QBF) is complete for $NP^{NP}$, and
-current 2-QBF solvers are much less efficient in practice than SAT solvers.
-
-\subsubsection{CEGIS}
-
-\fi
-
-In this section we present a solver for the synthesis fragment. An
-important characteristic of our solver is that is models bit accurate
-semantics. This enables the precise treatment of bit-wise operations and
-arithmetic modulo fixed widths.
 % 
@@
 % $$\exists P, x . \forall y . \sigma(P, x, y)$$
-We start by recalling the synthesis fragment:
- \[
-  \exists P,~ x . ~\forall~ y . \sigma(P, x, y)
- \]
 
-where $P$ ranges over sets, while $x$ and $y$ range over ground terms (we have dropped
- the vector symbols for brevity reasons).
+%% Our program encoding generates an $\exists\forall$ formula that is \emph{linear} in the length of the shortest program
+%% satisfying the user's specification.  This formula is then checked for satisfiability
+%% using the CEGIS algorithm.
 
-\subsection{\newC}
-\label{sec:logic}
-In order to obtain bit-level accuracy, the logic we use to express the synthesis formula is a subset of C that we call \newC.  
-The characteristic property of a
-\newC  program is that safety can be decided for
+In the context of (non-)termination analysis, a witnesses to the satisfiability of a formula in the synthesis fragment 
+is a (non-)termination proof. 
+As mentioned earlier in the paper, our analysis must ensure precise treatment of bit-wise operations and arithmetic modulo fixed widths
+%Consequently, a compulsory characteristic of our solver is that it must 
+by modelling bit accurate semantics.
+The synthesiser must therefore be able to compute (non-termination) proofs 
+%loop-free programs 
+over machine integers and floating-point numbers from the specifications detailed in Section~\ref{sec:second.order}.
+
+%\subsection{\newC}
+%\label{sec:logic}
+\subsection{The Synthesis Specification}
+Our (non-)termination specifications are given as program fragments
+%More specifically, the logic we use to express the synthesis formula
+in a subset of C that we call \newC.  
+The characteristic property of a \newC  program is that safety can be decided for
 it using a single query to a Bounded Model Checker.  A \newC program is
@@
 
-The proofs found by our solver are arbitrary expressions over the grammar given in Figure~\ref{fig:l-language}.  
+The (non-)termination proofs are arbitrary expressions over the grammar given in Figure~\ref{fig:l-language}.  
 %written in a simple RISC-like language $\mathcal{L}$, whose syntax is given in Fig.~\ref{fig:l-language}.  
-\todo{update the figure and make it look more like a grammar}
+\todo{update the figure and make it look more like a grammar.}
 
+\subsection{The Synthesis Algorithm}
 As the quantifier free first-order fragment of the logic is efficiently decidable and
@@
 
-\iffalse
-\subsection{The Concrete Algorithm for Bitvector Programs}
-\label{sec:concrete-algorithm}
-
-One area in which program synthesis can shine is in producing very small,
-intricate programs that manipulate bitvectors.  An example of such a program
-is given in Fig.~\ref{fig:bitvector-program}.  This program takes a machine word
-as input and clears every bit except for the least significant bit that was set.
-Even though this program is extremely short, it is fairly difficult for a human
-to see what it does.  It is even more difficult for a human to come up with such
-minimal code.  The program is so concise because it
-takes advantage of the low-level details of the machine, such as the fact that
-signed integers are stored in two's complement form.
-
-\begin{figure*}
-\centering
-\begin{minipage}{0.45\linewidth}
- \begin{lstlisting}[language=C]
-int isolate_lsb(int x) {
-  return x & -x;
-}
- \end{lstlisting}
-\end{minipage}
-\begin{minipage}{0.45\linewidth}
-
-Example:
-
-\hrule
-
-\begin{tabular}{llcccccccc}
- x       & = & 1 & 0 & 1 & 1 & 1 & 0 & 1 & 0 \\
- -x      & = & 0 & 1 & 0 & 0 & 0 & 1 & 1 & 0 \\
- x \& -x & = & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 0
-\end{tabular}
-\end{minipage}
-
- \caption{A tricky bitvector program}
-  \label{fig:bitvector-program}
-\end{figure*}
-
-To synthesise tricky bitvector programs like this, it is natural for us to
-work in the logic of quantifier-free propositional formulae and to use a
-propositional SAT or SMT-$\mathcal{BV}$ solver as the decision procedure. 
-However we propose a slightly different tack, which is to use a decidable
-fragment of C as a ``high level'' logic.
-\fi
-
-%% To instantiate the abstract synthesis algorithm in \newC we must express $I, O,
-%% \sigma$ and $P$ in \newC, then ensure that we can express the validity of the
-%% synthesis formula as a safety property of the resulting \newC program.
-
-%% Our encoding is the following:
-%% %
-%% \begin{itemize}
-%%  \item $I$ is the set of $N$-tuples of 32-bit bitvectors.  This is written in \newC as the type \verb|int[N]|.
-%%  \item $O$ is the set of $M$-tuples of 32-bit bitvectors, which is written in \newC as the type \verb|int[M]|.
-%%  \item $\sigma$ is a pure function with type $I \times O \rightarrow \mathrm{Bool}$.  The \newC signature of this function is
-%%  \verb|int check(int in[N], int out[M])|. This function is the only component supplied
-%%  by the user.
-%%  \item $P$ is written in a simple RISC-like language $\mathcal{L}$, whose syntax is given in Fig.~\ref{fig:l-language}.  Programs in $\mathcal{L}$
-%%  have type $I \rightarrow O$ and
-%%  are represented in \newC as objects of type \verb|prog_t|, shown in Fig.~\ref{fig:c-l-encoding}.
-%%  \item We supply an interpreter for $\mathcal{L}$ which is written in \newC.  The type of
-%%  this interpreter is $(I \rightarrow O) \times I \rightarrow O$ and the \newC signature is \\
-%%  \verb|void exec(prog_t p, int in[N], int out[M])|.  Here, \verb|out| is an output parameter.
-%% \end{itemize}
-
 \begin{figure}
@@
 
-%% The exact details of how we encode an $\mathcal{L}$-program are given in Sec.~\ref{sec:encode-l}.
-%% We must now express the {\sc synth} and {\sc verif} formulae as safety properties
-%% of \newC programs, which is given in Fig.~\ref{fig:c-synthverif}.
-
-%% \begin{figure}
-%% \centering
-%% \begin{minipage}[t]{.45\textwidth}
-%% \begin{lstlisting}[language=C++]
-%% void synth() {
-%%   prog_t p = nondet();
-%%   int in[N], out[M];
-
-%%   assume(wellformed(p));
-
-%%   in = test1;
-%%   exec(p, in, out);
-%%   assume(check(in, out));
-%%   ...
-%%   in = testN;
-%%   exec(p, in, out);
-%%   assume(check(in, out));
-
-%%   assert(false);
-%% }
-%% \end{lstlisting}
-%% %\end{minipage}
-%% %\hfill
-%% %\begin{minipage}[t]{.45\linewidth}
-%% \begin{lstlisting}[language=C++]
-%% void verif(prog_t p) {
-%%   int in[N] = nondet();
-%%   int out[M];
-
-%%   exec(p, in, out);
-%%   assert(check(in, out));
-%% }
-%% \end{lstlisting}
-%% \end{minipage}
-
-%%  \caption{The {\sc synth} and {\sc verif} formulae expressed as a \newC program.}
-%%  \label{fig:c-synthverif}
-%% \end{figure}
-
 In order to determine the validity of the {\sc synth} formula, we can
@@
 
-\node [file] (synth) {\sc synth.c};
-\path (synth.south)+(0.0, -0.5) node [file] (tests) {\sc tests.c};
-\path (tests.south)+(0.0, -0.5) node [file] (interpreter) {\sc interpreter.c};
-\path (interpreter.south)+(0.0, -0.5) node [file] (spec) {\sc spec.c};
+\node [file] (spec) {\sc Specification};
+%% \path (synth.south)+(0.0, -0.5) node [file] (tests) {\sc tests.c};
+%% \path (tests.south)+(0.0, -0.5) node [file] (interpreter) {\sc interpreter.c};
+%% \path (interpreter.south)+(0.0, -0.5) node [file] (spec) {\sc spec.c};
 
-\path (tests.east)+(2.0, -0.25) node [process] (merged) {merge};
+%% \path (spec.east)+(2.0, -0.25) node [process] (merged) {merge};
 
-\path (merged.east)+(2.0, 1.0) node [process] (cbmc) {\sc cbmc};
-\path (merged.east)+(2.0, -1.0) node [process] (gcc) {\sc gcc};
+\path (spec.east)+(2.0, 1.5) node [process] (cbmc) {\sc cbmc};
+\path (spec.east)+(2.0, 0.0) node [process] (gcc) {\sc gcc};
+\path (spec.east)+(2.0, -1.5) node [process] (gp) {\sc GP};
 
-\path (cbmc.east)+(2.5, -1.0) node [file] (out) {candidate program};
+\path (gcc.east)+(2.5, 0.0) node [file] (out) {Proof};
 
-\path [line] (synth) -- (merged);
-\path [line] (tests) -- (merged);
-\path [line] (interpreter) -- (merged);
-\path [line] (spec) -- (merged);
+%\path [line] (spec) -- (merged);
 
-\path [line] (merged) -- (cbmc);
-\path [line] (merged) -- (gcc);
+\path [line] (spec) -- (cbmc);
+\path [line] (spec) -- (gcc);
+\path [line] (spec) -- (gp);
 
@@
 \path [line] (gcc) -- (out);
+\path [line] (gp) -- (out);
 
@@
 
-%\subsubsection{Encoding an $\mathcal{L}$-Program in \newC}
-%% \label{sec:encode-l}
-
-%% \begin{figure}
-%% \begin{lstlisting}[language=c++,mathescape]
-%% typedef $\mathcal{BV}(4)$ op_t;                  // An opcode
-%% typedef $\mathcal{BV}(w)$ word_t;                // An $\mathcal{L}$-word
-%% typedef $\mathcal{BV}(\log_2 \lceil c+l+a \rceil)$ param_t;  // An operand
-
-%% struct prog_t {
-%%   op_t ops[$l$];          // The opcodes
-%%   param_t params[$l$*2];  // The operands
-%%   word_t consts[$c$];     // The program constants
-%% }
-%% \end{lstlisting}
-
-%%  \caption{The \newC structure we use to encode an $\mathcal{L}$ program
-%%   \label{fig:c-l-encoding}}
-%% \end{figure}
-
-%% The exact \newC encoding of an $\mathcal{L}$ program is shown in Fig.~\ref{fig:c-l-encoding}.
-%% The \verb|prog_t| structure encodes a program, which is a sequence of instructions.
-%% The parameter $a$ is the number of arguments the program takes.
-%% The $i$th instruction has opcode \verb|ops[i]|, left operand \verb|params[i*2]| and
-%% right operand \verb|params[i*2 + 1]|.  An operand refers to either a program constant,
-%% a program argument or the result of a previous instruction, and its value
-%% is determined at runtime as follows:
-%% \[
-%%  val(x) = \begin{cases}
-%%            x < a & \text{the } x^{\text{th}} \text{ program argument} \\
-%%            a \leq x < a+c & \mathtt{consts[} x-a \mathtt{]} \\
-%%            x \geq a + c & \text{the result of instruction } (x - a - c)
-%%           \end{cases}
-%% \]
-
-%% A program is well formed if
-%% no operand refers to the result of an instruction that has not been computed yet, and if
-%% each opcode is valid.  We add a well-formedness constraint of the form \verb|params[i] <= (a+c+2*i)|
-%% for each instruction.  It should be noted that this requires a linear number of well-formedness
-%% constraints.  If all of these constraints are satisfied, the program is well-formed in the sense.
-
 \subsection{Parameterising the Program Space}
@@
 In order to search the space of candidate programs, we parametrise
-the language~$\mathcal{L}$, inducing a lattice of progressively
+the language~$\mathcal{L}$ by program length, word width and number of constants,
+inducing a lattice of progressively
 more expressive languages.  We start by attempting to synthesise
@@
 
----
-We require the user to supply a program specification written in C, but then
-ask nothing more of her.  We parametrise the space of programs in such a way
-that it can be explored automatically, rather than asking a human for hints.
-Our program encoding generates an $\exists
-\forall$ formula that is \emph{linear} in the length of the shortest program
-satisfying the user's specification.  This formula is then checked for satisfiability
-using the CEGIS algorithm.
----
+%% We introduce a novel parametrisation of the programming language
+%% used to express our synthesised programs.
+%% This parametrisation allows us to efficiently explore the program space
+%% without relying on human guidance and also ensures that our programs
+%% are of minimal length.
 
-We introduce a novel parametrisation of the programming language
-used to express our synthesised programs.
-This parametrisation allows us to efficiently explore the program space
-without relying on human guidance and also ensures that our programs
-are of minimal length.
+%% Our tool is the first we are aware of that is able to effectively
+%% synthesise floating-point programs. We demonstrate this by
+%% synthesising {\sc Fast2Sum} using Knuth's {\sc 2Sum}~\cite{taocp2} as
+%% a specification.
 
-Our tool is the first we are aware of that is able to effectively
-synthesise floating-point programs.  We demonstrate this by
-synthesising {\sc Fast2Sum} using Knuth's {\sc 2Sum}~\cite{taocp2} as
-a specification.
-------
+%% ---Our exploration of the program space ensures that our programs are minimal in size.
+%% --- We parametrise the space of programs in such a way that it can be explored automatically, rather than asking a human for hints.
 
-We consider the following parameters:
-\begin{itemize}
-\item{Program Length: $l$}
+%% We consider the following parameters:
+%% \begin{itemize}
+%% \item{Program Length: $l$}
 %% The first parameter we introduce is program length, denoted by $l$.
@@
 %% shorter program is correct.
-\item{Word Width: $w$}
+%\item{Word Width: $w$}
 %% An $\mathcal{L}$-program runs on a virtual machine (the $\mathcal{L}$-machine) that
@@
 
-%% It is often the case that a program which satisfies the specification
-%% on an $\mathcal{L}$-machine with $w = k$ will continue to satisfy the
-%% specification when run on a machine with $w > k$.  For example, the program
-%% in Fig.~\ref{fig:bitvector-program} isolates the least-significant bit of a word.
-%% This is true irrespective of the word size of the machine it is run on -- it will
-%% isolate the least-significant bit of an 8-bit word just as well as it will a
-%% 32-bit word.  An often successful strategy is to synthesise a program for an
-%% $\mathcal{L}$-machine with a small word size and then to check whether the
-%% same program is correct when run on an $\mathcal{L}$-machine with a
-%% full-sized word.
-
-%% The only wrinkle here is that we will sometimes synthesise a program containing
-%% constants.  If we have synthesised a program with $w=k$,
-%% the constants in the program will be $k$-bits wide.  To extend the program
-%% to an $n$-bit machine (with $n > k$), we need some way of deriving $n$-bit-wide
-%% numbers from $k$-bit ones.  We have several strategies for this and
-%% just try each in turn.  Our strategies are shown in Fig.~\ref{fig:generalize}.
-%% $\mathcal{BV}(v, n)$ denotes an $n$-bit wide bitvector holding the value $v$
-%% and $b \cdotp c$ means the concatenation of bitvectors $b$ and $c$.
-
-%% \begin{figure}
-
-%% \centering
-%% \begin{minipage}[t]{.45\textwidth}
-%% \begin{eqnarray*}
-%%  \bv{m}{m} & \rightarrow & \bv{n}{n} \\
-%%  \bv{m-1}{m} & \rightarrow & \bv{n-1}{n} \\
-%%  \bv{m+1}{m} & \rightarrow & \bv{n+1}{n}
-%% \end{eqnarray*}
-%% \end{minipage}
-%% \begin{minipage}[t]{.45\textwidth}
-%% \begin{eqnarray*}
-%%  \bv{x}{m} & \rightarrow & \bv{x}{n} \\
-%%  \bv{x}{m} & \rightarrow & \bv{x}{m} \cdotp \bv{0}{n - m} \\
-%%  \bv{x}{m} & \rightarrow & \underbrace{\bv{x}{m} \cdotp \ldots \cdotp \bv{x}{m}}_{\frac{n}{m} \mathrm{ times}}
-%% \end{eqnarray*}
-%% \end{minipage}
-
-%% \caption{Rules for extending
-%% an $m$-bit wide number to an $n$-bit wide one.
-%%  \label{fig:generalize}}
-%% \end{figure}
-
-%% Sometimes a program will be correct for some particular word width $w$, but is
-%% not correct for $w' > w$ even if the constants are replaced with appropriate ones.
-%% When we detect this situation, we increase $w$ and continue synthesising.
-
-\item{Number of Constants: $c$}
+%\item{Number of Constants: $c$}
 %% Instructions in $\mathcal{L}$ take either one or two operands.
@@
 
-%% $\mathcal{L}$ is an SSA, three-address instruction set\footnote{
-%% We experimented with implementing $\mathcal{L}$ as a stack machine, expecting
-%% the programs to be smaller and synthesis to be faster as a result.  We saw
-%% the opposite effect -- the more complex interpreter led to much slower synthesis.
-%% }.  Destination registers
-%% are implicit and a fresh register exists for each instruction to write its
-%% output to.  A na\"{\i}ve way to encode $\mathcal{L}$ instructions is to have an
-%% opcode and two operands, where each operand is either a register (i.e., a program argument
-%% or the result of a previous instruction), or an immediate constant.
-
-%% In this encoding, each opcode requires $\lceil \log_2 I \rceil$ bits to encode, where $I$ is the number
-%% of instruction types in $\mathcal{L}$.  Each operand can be encoded using
-%% $\log_2 w$ bits, where $w$ is the $\mathcal{L}$-machine word width, plus one
-%% bit to specify whether the operand is a register name or an immediate constant.
-%% One instruction can therefore be encoded using $\lceil \log_2 I \rceil + 2w + 2$ bits.
-%% For an $n$-instruction program, we need $$\lceil n \log_2 I \rceil + 2nw + 2n$$ bits to encode
-%% the entire program.
-
-%% If we instead limit the number of constants that can appear in the program,
-%% our operands can be encoded using fewer bits.  For an $n$-instruction program
-%% using $c$ constants and taking $a$ arguments as inputs, each operand can refer
-%% to a program argument, the result of a previous instruction or a constant.
-%% This can be encoded using $\lceil \log_2 (c+a+n-1) \rceil$ bits, which means each instruction
-%% can be encoded in $\lceil \log_2 I \rceil + \lceil \log_2 (c + a + n - 1) \rceil$ and the full program
-%% needs $$\lceil n \log_2 I \rceil + \lceil n \log_2 (c + a + n - 1) \rceil + cw$$ bits to encode.
-
-%% We give an example. Our language $\mathcal{L}$ has 15 instruction types, so each opcode is 4 bits.
-%% For a 10-instruction program over 1 argument, using 2 constants on a 32-bit word
-%% machine the first encoding requires $10 * (4 + 32 + 1 + 32 + 1) = 700$ bits.
-%% Using the second encoding, each operand can be represented using
-%% $\log_2 (2 + 1 + 10 - 1) = 4$ bits, and the entire program requires 184 bits.
-%% This is a substantial reduction in size and when the desired program requires
-%% only few constants this can lead to a very significant speed up.
-
-%% As with program length, we progressively increase the number of constants in
-%% our program.  We start by trying to synthesise a program with no constants,
-%% then if that fails we attempt to synthesise using one constant and so on until we
-%% reach $c = l$.
-
-\end{itemize}
-\subsubsection{Searching the Program Space}
+%\end{itemize}
+%\subsubsection{Searching the Program Space}
 
@@
 adjust the $\mathcal{L}$-parameters in order to cover all possible programs.
-After each round of {\sc synth}, we may need to adjust the parameters.  The
-logic for these adjustments is shown as a tree in Fig.~\ref{fig:paramsflow}.
+After each round of {\sc synth}, we may need to adjust the parameters.  
+%The
+%logic for these adjustments is shown as a tree in Fig.~\ref{fig:paramsflow}.
+Whenever {\sc synth} fails, we consider which parameter might have caused the
+failure.  
+%% There are two possibilities: either the program length $l$ was too small,
+%% or the number of allowed constants $c$ was.  If $c < l$, we just increment $c$ and
+%% try another round of synthesis, but allowing ourselves an extra program constant.
+%% If $c = l$, there is no point in increasing $c$ any further.  This is because
+%% no minimal $\mathcal{L}$-program has $c > l$, for if it did there would
+%% have to be at least one instruction with two constant operands.  This
+%% instruction could be removed (at the expense of adding its result as
+%% a constant), contradicting the assumed minimality of the program.  So
+%% if $c = l$, we set $c$ to 0 and increment $l$, before attempting
+%% synthesis again.
 
-Whenever {\sc synth} fails, we consider which parameter might have caused the
-failure.  There are two possibilities: either the program length $l$ was too small,
-or the number of allowed constants $c$ was.  If $c < l$, we just increment $c$ and
-try another round of synthesis, but allowing ourselves an extra program constant.
-If $c = l$, there is no point in increasing $c$ any further.  This is because
-no minimal $\mathcal{L}$-program has $c > l$, for if it did there would
-have to be at least one instruction with two constant operands.  This
-instruction could be removed (at the expense of adding its result as
-a constant), contradicting the assumed minimality of the program.  So
-if $c = l$, we set $c$ to 0 and increment $l$, before attempting
-synthesis again.
+%% If {\sc synth} succeeds but {\sc verif} fails, we have a candidate
+%% program that is correct for some inputs but incorrect on at least
+%% one input.  However, it may be the case that the candidate program
+%% is correct for \emph{all} inputs when run on an $\mathcal{L}$-machine
+%% with a small word size.  For example, we may have synthesised a
+%% program which is correct for all 8-bit inputs, but incorrect for
+%% some 32-bit input.  If this is the case (which we can determine
+%% by running the candidate program through {\sc verif} using the smaller
+%% word size), we may be able to produce a correct program for
+%% the full $\mathcal{L}$-machine by using the constant extension rules
+%% shown in Fig.~\ref{fig:generalize}.  If constant generalization
+%% is able to find a correct program, we are done.  Otherwise,
+%% we need to increase the word width of the $\mathcal{L}$-machine
+%% we are currently synthesising for.
 
-If {\sc synth} succeeds but {\sc verif} fails, we have a candidate
-program that is correct for some inputs but incorrect on at least
-one input.  However, it may be the case that the candidate program
-is correct for \emph{all} inputs when run on an $\mathcal{L}$-machine
-with a small word size.  For example, we may have synthesised a
-program which is correct for all 8-bit inputs, but incorrect for
-some 32-bit input.  If this is the case (which we can determine
-by running the candidate program through {\sc verif} using the smaller
-word size), we may be able to produce a correct program for
-the full $\mathcal{L}$-machine by using the constant extension rules
-shown in Fig.~\ref{fig:generalize}.  If constant generalization
-is able to find a correct program, we are done.  Otherwise,
-we need to increase the word width of the $\mathcal{L}$-machine
-we are currently synthesising for.
+%% \section{Conclusion}
 
-%% \begin{figure}[t]
-%% \centering
-%% \begin{tikzpicture}[scale=0.65, transform shape, node distance=2cm, auto]
-%% \tikzstyle{decision} = [diamond, draw, fill=blue!20, 
-%%     text width=4.5em, text badly centered, node distance=3cm, inner sep=0pt]
-%% \tikzstyle{block} = [rectangle, draw, fill=blue!20, 
-%%     text width=5em, text centered, rounded corners, minimum height=4em]
-%% \tikzstyle{line} = [draw, -latex']
-%% \tikzstyle{cloud} = [draw, ellipse,fill=red!20, node distance=3cm,
-%%     minimum height=2em]
-
-%%  \node [decision] (synthsucc) {{\sc Synth} succeeds?};
-
-%%  \node [decision, below of=synthsucc, node distance=2.5cm] (verif) {{\sc Verif} succeeds?};
-%%  \node [decision, right of=verif] (ck) {$c < l$?};
-
-%%  \node [block, below of=verif, node distance=3cm]   (done) {Done!};
-%%  \node [decision, left of=verif] (verifw) {{\sc Verif} succeeds for small words?};
-
-%%  \node [block, below of=ck, node distance=2.5cm] (incc) {$c := c+1$};
-%%  \node [block, right of=incc, node distance=2cm] (incl) {$c := 0$\\ $l := l+1$};
-
-%%  \node [decision, below of=verifw, node distance=3cm] (gen) {Extend succeeds?};
-%%  \node [block, left of=verifw, node distance=3cm] (iterate) {Parameters unchanged};
-
-%%  \node [block, left of=gen, node distance=3cm] (incw) {$w := w+1$};
-
-%%  \path [line] (synthsucc) -- node [left] {Yes} (verif);
-%%  \path [line] (synthsucc) -| node [above, near start] {No} (ck);
-
-%%  \path [line] (verif) -- node [left] {Yes} (done);
-%%  \path [line] (verif) -- node [above, near start] {No} (verifw);
-
-%%  \path [line] (ck) -- node [left] {Yes} (incc);
-%%  \path [line] (ck) -| node  [above, near start]  {No} (incl);
-
-%%  \path [line] (verifw) -- node [right] {Yes} (gen);
-%%  \path [line] (verifw) -- node [above] {No} (iterate);
-
-%%  \path [line] (gen) -- node [below] {Yes} (done);
-%%  \path [line] (gen) -- node [below] {No} (incw);
-
-%%  %\path [dotted, line] (iterate.west) |- (synthsucc);
-%% \end{tikzpicture}
-
-%%  \caption{Decision tree for increasing parameters of $\mathcal{L}$.}
-%%  \label{fig:paramsflow}
-
-%% \end{figure}
-
-\iffalse
-
-\subsection{Implementation Issues}
-For performance reasons, we found that adding extra constraints on the
-syntax of the synthesised programs sped up synthesis.  The extra constraints
-fell into two categories: eliminating nops and instruction-level symmetry reduction.
-
-\subsubsection{Remove nops}
-Many instructions in $\mathcal{L}$ are nops that do not doing anything,
-for example the instruction \verb|add x 0|.  Such instructions
-can be removed from any program they appear in to leave a semantically
-equivalent, but shorter, program.  We can therefore be sure that nops
-will never appear in any minimal program.  By adding constraints saying that
-each instruction is not a nop, we can help the underlying SAT solver's
-search, which reduces the runtime of the overall procedure.
-
-\subsubsection{Symmetry Reduction}
-There are many instructions that are equivalent to each other.  For example,
-\verb|add x y| is equivalent to \verb|add y x| -- any program containing
-one instruction could have it replaced by the other instruction and
-keep the same semantics.  We choose a single canonical instruction to
-represent all instructions in a particular equivalence class, then add
-constraints saying that no non-canonical instructions appear in the program.
-
-Our rules for excluding non-canonical instructions are:
-
-\begin{itemize}
- \item For commutative operations, the first operand is smaller than the second.
- \item For unary operations, the second (unused) operand is always 0.
- \item No instruction may have two constant operands.
- \item All constants in the constant table are distinct.
-\end{itemize}
-
-As with the nop constraints, these additional constraints do increase the
-size of the resulting SAT instance, but this still ends up as a win in
-terms of runtime.
-\fi
-
-\iffalse
-\section{Experiments}
-\label{sec:experiments}
-\subsection{Experimental Setup}
-We implemented our fully automatic synthesis procedure as the {\sc kalashnikov} tool.  The implementation consists
-of around 600 lines of \newC and 600 lines of Python.  To evaluate the tool we used the 29 bitvector
-programs from~\cite{brahma} and~\cite{brahma-icse}.  The majority of these are ``bit twiddling hacks'' taken from
-Hacker's Delight~\cite{hackers-delight}.  The code we used to perform the experiments, along with the
-benchmarks, is available at \texttt{http://www.cprover.org/kalashnikov} and the benchmarks can also
-be found in Appendix~\ref{app:benchmarks}.  As a backend solver, we used
-{\sc CBMC}~\cite{cbmc-website} at SVN revision r3545, with Glucose 3.0~\cite{glucose-paper} as the SAT solver.
-We performed our experiments on a 4-core, 2.40\,GHz Xeon E5-2665 with 32\,GB of RAM.
-
-To give a reference point for the efficiency of our tool, we present the results given for {\sc brahma}
-on the same benchmarks, as reported in~\cite{brahma} and~\cite{brahma-icse}.
-These experiments were performed on an 8-core,
-1.86\,GHz Xeon with 4\,GB of RAM.
-
-Since we were unable to obtain a copy of {\sc brahma}, we re-implemented the {\sc brahma} program encoding in our framework,
-resulting in the {\sc brahmikov} tool. {\sc Brahmikov} implements the component based synthesis
-paradigm, but still uses {\sc cbmc} as the backend decision procedure.  Since {\sc brahma} and {\sc brahmikov} are written in
-different frameworks, we cannot compare their performance.  However we can compare {\sc brahmikov} and {\sc kalashnikov}.
-In particular it is interesting to note that {\sc brahmikov} often produces longer programs than {\sc kalashnikov}.  We
-believe this is due to differences in the post-processing strategies used by {\sc brahmikov} and {\sc brahma}.  {\sc Brahmikov}
-performs syntatic dead code elimination (following~\cite{brahma}), but this is not enough to generate minimal programs
-in many cases.  This is in contrast to {\sc kalashnikov}, which always generates minimal programs.
-
-We also ran the {\sc AHA} superoptimiser on our benchmarks and found that on the small problems
-(those with 2-instruction solutions)
-{\sc AHA} found a single correct solution in negligible time.  On the larger benchmarks, candidate solutions
-were still found rapidly (around 10\,s for the 3-instruction programs) but the number of candidates was
-too large to practically check them all.  For example, {\sc AHA} found 2284262 candidate
-programs in 12.13\,s for p12.
-
-We invested significant effort into trying to implement the benchmarks in a way
-that would allow us to compare with {\sc sketch}. Unfortunately, we were unable to come up
-with an encoding that afforded a meaningful comparison.  In retrospect,
-we realise that this is due to a fundamental difference in the problem domains
-our respective tools are designed to solve: {\sc sketch} is able to use
-human insight to generate large, complex programs, whereas we are optimised for 
-automatically generating small programs with somewhat more intricate data-flow.
-
-\subsection{Results and Analysis}
-The results of our experiments are shown in Fig.~\ref{fig:results-table}.
-Column 1 shows the runtime reported for each benchmark in \cite{brahma}, column 2
-shows the number of instructions in the synthesised program, and column 3 contains a \xmark\;when
-{\sc Brahma} needed user assistance to solve a benchmark.
-Columns 4 through 7 show the same information, but with the data taken from~\cite{brahma-icse}.
-Column 5 shows the runtimes for the version of {\sc Brahma} from~\cite{brahma-icse} which
-implemented the semibiased optimisation.
-Finally, columns 8 and 10 show the runtime for {\sc Brahmikov} and {\sc Kalashnikov} respectively.
-Columns 9 and 11 show the number of instructions
-in the program synthesised by {\sc Brahmikov} and {\sc Kalashnikov}.  In all cases,
-{\sc Brahmikov} performs very poorly, which demonstrates that the {\sc Kalashnikov} program
-encoding is more suitable for our system than the component based synthesis encoding.
-We also ran the {\sc Brahmikov} experiments using Z3 as the solver rather than Glucose,
-but this configuration failed to terminate for any of the examples.
-
-The results can be divided into three categories, as follows:
-
-\paragraph{\bf {\sc Kalashnikov} synthesises a shorter program than {\sc Brahma}:}
-This happens in 4 of the 29 cases, which is to be expected since {\sc Kalashnikov}'s outputs
-are, by construction, guaranteed to be minimal.  This case is illustrated by benchmark p29, shown in Fig.~\ref{fig:obfuscated}.
-The specification here is a piece of obfuscated code taken from the Conficker worm~\cite{conficker}, and
-our goal is to synthesise an equivalent program which is easier to understand.  The obfuscated code
-uses several tricks, including an apparently unbounded loop.  {\sc Brahma} is able to
-produce an equivalent program consisting of four instructions using shifts and addition, whereas {\sc kalashnikov}
-is able to produce the minimal program \verb|y * 45|.  It is also worth noting that the specification
-fed to {\sc kalashnikov} was just the obfuscated code, with no further preprocessing needed.
-\paragraph{\bf {\sc Kalashnikov} is unable to synthesise a program:}
-This happens in 8 of the 29 cases.  In each case, {\sc Brahma} needs user guidance to synthesise the program.
-The existence of instances that are beyond the reach of current automatic
-methods, and thus require manual intervention, is to be expected.
-% A total of 8 tests fall into this category.
-\paragraph{\bf {\sc Brahma} and {\sc Kalashnikov} both synthesise minimal programs:}
-In the remaining cases, both {\sc brahma} and {\sc kalashnikov} are able to produce
-minimal programs.  It was not clear from just looking at the runtime numbers whether
-any of the tools was significantly faster than the others, so we performed a Wilcoxon signed-rank test.
-
-For each pair of tools ({\sc kalashnikov} vs. each of the {\sc brahma} configurations), the Wilcoxon test was
-unable to reject the null hypothesis (that the tools are equally fast) at the p=0.05 level. In other words,
-there is no statistically significant difference in the speed of {\sc kalashnikov} and {\sc brahma}.
-
-\begin{figure}
-Obfuscated code
-
-\begin{lstlisting}[frame=single,language=c]
-int obfuscated (int y) {
- int a=1, b=0, z=1, c=0;
- while(1) {
-  if (a == 0) { if (b == 0) { y=z+y; a =!a; b=!b;c=!c;
-  if (!c) break;} else { z=z+y; a=!a; b=!b; c=!c;
-  if (!c) break;} } else { if (b == 0) { z=y << 2;
-  a=!a; } else { z=y << 3; a=!a; b=!b; } }
- }
- return z;
-}
-\end{lstlisting}
-%\end{minipage}
-
-\begin{minipage}[t]{.45\textwidth}
-{\sc Brahma}'s output
-
-\begin{lstlisting}[language=c,frame=single]
-int recovered (int y) {
- int z = y << 2;
- y = z + y;
- z = y << 3
- y = z + y;
- return z;
-}
-\end{lstlisting}
-\end{minipage}
-\hfill
-\begin{minipage}[t]{.45\textwidth}
-{\sc Kalashnikov}'s output
- \begin{lstlisting}[language=c,frame=single]
-int recovered (int y) {
- return y*45;
-}
-\end{lstlisting}
-\end{minipage}
-
- \caption{De-obfuscating  C code.
-  \label{fig:obfuscated}}
-\end{figure}
-
-\subsection{Floating Point}
-To demonstrate the capabilities of \newC as an implementation language, we synthesise a
-non-trivial floating-point program.  For this we chose the classic {\sc 2Sum} algorithm for computing
-exact sums of floating-point numbers.  The original {\sc 2Sum} algorithm is given as Alg.~\ref{alg:2sum}.
-In the case that $a \ge b$, exact addition can be implemented in fewer instructions -- this is the
-{\sc Fast2Sum} algorithm~\cite{fast2sum}, shown in Alg.~\ref{alg:fast2sum}.  We used {\sc 2sum} as our specification, and
-added an assumption that $a \ge b$.  {\sc Kalashnikov} is able to synthesise and verify the
-code for {\sc Fast2Sum} from this specification in 396.07\,s, which we believe to be the first time
-this has been achieved.  {\sc 2Sum} was proved to be minimal in~\cite{fast2sum}, but
-we believe our synthesis constitutes the first proof that {\sc Fast2Sum} is
-minimal under the assumption that $a \ge b$.
-
-\begin{figure}[ht]
-\begin{center}
-\begin{minipage}[t]{0.45\linewidth}
-\begin{algorithm}[H]
-\caption{\sc 2Sum
- \label{alg:2sum}}
-\begin{algorithmic}
-\Let{$s$}{$a+b$}
-\Let{$b'$}{$s - a$}
-\Let{$a'$}{$s - b'$}
-\Let{$\delta b$}{$b - b'$}
-\Let{$\delta a$}{$a - a'$}
-\Let{$t$}{$\delta a + \delta b$}
-\Ensure{$a + b$ = $s + t$ exactly}
-\end{algorithmic}
-\end{algorithm}
-\end{minipage}
-\hfill
-\begin{minipage}[t]{0.45\linewidth}
-\begin{algorithm}[H]
-\caption{\sc Fast2Sum
- \label{alg:fast2sum}}
-\begin{algorithmic}
-\Require{$a \ge b$}
-\Let{$s$}{$a + b$}
-\Let{$z$}{$s - a$}
-\Let{$t$}{$b - z$}
-\Ensure{$a + b$ = $s + t$ exactly}
-\end{algorithmic}
-\end{algorithm}
-\end{minipage}
-\end{center}
-
- \caption{The {\sc 2Sum} algorithm for correctly rounded floating-point sums.}
-  \label{fig:2sum}
-\end{figure}
-
-\begin{figure}[ht]
-\centering
-{\tiny
-\include{figures/results_table}
-}
-\caption{Time for synthesis of machine integer benchmarks from~\cite{brahma} and~\cite{brahma-icse}.
- \label{fig:results-table}}
-\end{figure}
-
-%\begin{figure}[p]
-%\begin{center}
-%{\tiny
-%\include{figures/icse_results_table}
-%}
-%\end{center}
-%\caption{Times for synthesis of machine integer benchmarks from~\cite{brahma-icse}.}
-%\label{fig:icse-results-table}
-%\end{figure}
-
-\section{Conclusion}
-
-By expressing the program synthesis problem as a safety property for a
-program interpreter, we have been able to harness the power of
-state-of-the-art program analysis tools and reuse them in a new problem
-domain.  We have implemented our algorithm as a freely downloadable tool
-whose performance compares favourably to a recent program synthesiser. 
-Finally, we have taken advantage of the expressiveness of our specification
-language to make an initial step towards practical synthesis of
-floating-point programs.
-
-\paragraph{Future Work}
-
-It may be possible to extend the language $\mathcal{L}$ with constructs such
-as branching and limited looping while still retaining decidability, which
-would be an interesting direction to pursue.  Since our specification
-language is a subset of C, it is straight-forward to interface {\sc
-kalashnikov} with other analysis tools.  Various problems such as ranking
-function synthesis, loop invariant generation and loop
-summarisation can be phrased as program synthesis problems, and we would like to investigate the
-possibilities of applying {\sc kalashnikov} to these domains.
-
-\paragraph{Acknowledgements}
-We thank Martin Brain for his insight and advice during our synthesis discussions.
-Also, thanks to Cristina David for her extensive help and proofreading during the
-preparation of this paper.
-
-\bibliography{synth}{}
-\bibliographystyle{splncs}
-
-\appendix
-
-\section{Benchmarks}
-\label{app:benchmarks}
-
-\begin{figure}
-\scriptsize
- \begin{multicols}{4}
-
-   \begin{minipage}{.2\textwidth}
-   \bf{P1}($x$): Turn-off rightmost 1 bit.
-   \begin{enumerate}
-    \item $t1 = x - 1$
-    \item $t2 = x \& t1$
-   \end{enumerate}
-   \end{minipage}
-
-   \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   \bf{P2}($x$): Test whether $x$ is of the form $2^n - 1$.
-   \begin{enumerate}
-    \item $t1 = x + 1$
-    \item $t2 = x \& t1$
-   \end{enumerate}
-   \end{minipage}
-
-   \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   \bf{P3}($x$): Isolate the rightmost 1 bit.
-   \begin{enumerate}
-    \item $t1 = -x$
-    \item $t2 = x \& t1$
-   \end{enumerate}
-   \end{minipage}
-
-   \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   \bf{P4}($x$): Form a mask that identifies the rightmost 1 bit and trailing 0s.
-   \begin{enumerate}
-    \item $t1 = x - 1$
-    \item $t2 = x \oplus t1$
-   \end{enumerate}
-   \end{minipage}
-
-   \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   \bf{P5}($x$): Right propagate the rightmost 1 bit.
-   \begin{enumerate}
-    \item $t1 = x - 1$
-    \item $t2 = x \mid t1$
-   \end{enumerate}
-   \end{minipage}
-
-   \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   \bf{P6}($x$): Turn on the rightmost 0 bit.
-   \begin{enumerate}
-    \item $t1 = x + 1$
-    \item $t2 = x \mid t1$
-   \end{enumerate}
-   \end{minipage}
-      \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   \bf{P7}($x$): Isolate the rightmost 0 bit.
-   \begin{enumerate}
-    \item $t1 = \sim x$
-    \item $t2 = x + 1$
-    \item $t3 = t1 \& t2$
-   \end{enumerate}
-   \end{minipage}
-   \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   \bf{P8}($x$): Form a mask the identifies the trailing 0s.
-   \begin{enumerate}
-    \item $t1 = x -1$
-    \item $t2 = \sim x$
-    \item $t3 = t1 \& t2$
-   \end{enumerate}
-   \end{minipage}
-   \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   \bf{P9}($x$): Absolute value.
-   \begin{enumerate}
-    \item $t1 = x >> 31$
-    \item $t2 = x \oplus t1$
-    \item $t3 = t2 - t1$
-   \end{enumerate}
-   \end{minipage}
-
-      \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   \bf{P10}($x, y$): Test if nlz(x) == nlz(y), where nlz is number of leading zeroes.
-   \begin{enumerate}
-    \item $t1 = x \& y$
-    \item $t2 = x \oplus y$
-    \item $t3 = t2 \leq t1$
-   \end{enumerate}
-   \end{minipage}
-
-      \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   \bf{P11}($x, y$): Test if nlz(x) < nlz(y).
-   \begin{enumerate}
-    \item $t1 = \sim y$
-    \item $t2 = x \& t1$
-    \item $t3 = t2 > y$
-   \end{enumerate}
-   \end{minipage}
-   \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   \bf{P12}($x, y$): Test if nlz(x) <= nlz(y).
-   \begin{enumerate}
-    \item $t1 = \sim y$
-    \item $t2 = x \& t1$
-    \item $t3 = t2 \leq y$
-   \end{enumerate}
-   \end{minipage}
-
-      \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   \bf{P13}($x$): Sign.
-   \begin{enumerate}
-    \item $t1 = x >> 31$
-    \item $t2 = t1 < x$
-    \item $t3 = t2 + t1$
-   \end{enumerate}
-   \end{minipage}
-
-   \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   \bf{P14}($x, y$): Floor of average of $x$ and $y$ without overflowing.
-   \begin{enumerate}
-    \item $t1 = x \& y$
-    \item $t2 = x \oplus y$
-    \item $t3 = t2 >> 1$
-    \item $t4 = t1 + t3$
-   \end{enumerate}
-   \end{minipage}
-   \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   \bf{P15}($x, y$): Ceil of average of $x$ and $y$ without overflowing.
-   \begin{enumerate}
-    \item $t1 = x \mid y$
-    \item $t2 = x \oplus y$
-    \item $t3 = t2 >> 1$
-    \item $t4 = t1 - t3$
-   \end{enumerate}
-   \end{minipage}
-   \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   \bf{P16}($x$): Max.
-   \begin{enumerate}
-    \item $t1 = x \leq y$
-    \item $t2 = x \oplus y$
-    \item $t3 = t1 * t2$
-    \item $t4 = t3 \oplus y$
-   \end{enumerate}
-   \end{minipage}
-   \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   \bf{P17}($x$): Turn off the rightmost contiguous string of 1 bits.
-   \begin{enumerate}
-    \item $t1 = x - 1$
-    \item $t2 = x \mid t1$
-    \item $t3 = t2 + 1$
-    \item $t4 = t3 \& x$
-   \end{enumerate}
-   \end{minipage}
-   \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   \bf{P18}($x$): Determine if $x$ is a power of 2.
-   \begin{enumerate}
-    \item $t1 = -x$
-    \item $t2 = x \oplus t1$
-    \item $t3 = t2 < t1$
-   \end{enumerate}
-   \end{minipage}
-   \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   {\bf P19($x, m, k$): Exchange two fields of $x$, demarcated by $m$ and $k$.}
-
-   N/A ({\sc Kalashnikov} was unable to synthesise this program.)
-   \end{minipage}
-   \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   {\bf P20($x$): Next higher unsigned number with the same number of 1 bits.}
-
-   N/A
-   \end{minipage}
-   \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   {\bf P21($x, a, b, c$): Cycling through 3 values $a, b, c$.}
-
-    N/A
-   \end{minipage}
-   \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   {\bf P22($x$): Compute parity.}
-
-    N/A
-   \end{minipage}
-   \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   {\bf P23($x$): Count number of set bits.}
-
-    N/A
-   \end{minipage}
-      \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   {\bf P24($x$): Round up to next highest power of 2.}
-
-    N/A
-   \end{minipage}
-   \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   {\bf P25($x, y$): Computer higher order half of $x \times y$.}
-
-    N/A
-   \end{minipage}
-
-   \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   {\bf P26($x, k$): Round $x$ up to a multiple of $2^k$.}
-
-N/A
-   \end{minipage}
-
-      \vspace{12pt}
-
-   \begin{minipage}{.2\textwidth}
-   {\bf P27($x, y$): Min.}
-    \begin{enumerate}
-     \item $t1 = x \oplus y$
-     \item $t2 = t1 \leq t2$
-     \item $t3 = t1 * t2$
-     \item $t4 = y \oplus t3$
-    \end{enumerate}
-   \end{minipage}
-
- \end{multicols}
-
- \caption{The integer benchmarks}
-\end{figure}
-
-\begin{figure}
- \begin{lstlisting}[language=C++]
-void interchangeObs(word_t* src, word_t* dest)
-{
-  *src = *src ^ *dest;
-
-  if (*src == (*src ^ *dest)) {
-    *src = *src ^ *dest;
-
-    if (*src == (*src ^ *dest)) {
-      *dest = *src ^ *dest;
-
-      if (*dest == (*src ^ *dest)) {
-        *src = *dest ^ *src;
-        return;
-      } else {
-        *src = *src ^ *dest;
-        *dest = *src ^ *dest;
-        return;
-      } 
-    } else
-      *src = *src ^ *dest;
-  }
-
-  *dest = *src ^ *dest;
-  *src = *src ^ *dest;
-
-  return;
-}
- \end{lstlisting}
-
- \caption{Benchmark 28: an obfuscated swap function}
-\end{figure}
-
-\begin{figure}
- \begin{lstlisting}[language=C++]
-int mul45(int y) {
-  int a=1, b=0, z=1, c=0;
-
-  while(1) {
-    if (a == 0) {
-      if (b == 0) {
-	y=z+y; a =!a; b=!b;c=!c;
-	if (! c) break;
-      } else {
-	z=z+y; a=!a; b=!b; c=!c;
-	if (! c) break;
-      }
-    } else {
-      if (b == 0) { z=y << 2; a=!a; }
-      else {
-	z=y << 3; a=!a; b=!b; } } }
-
-  return y;
-}
-
- \end{lstlisting}
-
- \caption{Benchmark 29: an obfuscated function to multiply by 45}
-
-\end{figure}
-
-\fi
+%% By expressing the program synthesis problem as a safety property for a
+%% program interpreter, we have been able to harness the power of
+%% state-of-the-art program analysis tools and reuse them in a new problem
+%% domain.  We have implemented our algorithm as a freely downloadable tool
+%% whose performance compares favourably to a recent program synthesiser. 
+%% Finally, we have taken advantage of the expressiveness of our specification
+%% language to make an initial step towards practical synthesis of
+%% floating-point programs.

### 2014-07-03 papers/termination/paper.tex
revised more of the kalashnikov section

@@
 
-\item Our proposed termination proving procedure can also prove non-termination. Alternating termination and non-termination?
+\item Our proposed termination proving procedure can also prove non-termination: (non-)termination analysis.
 
@@
 
-\section{Solving the Synthesis Formula} \label{sec:synthesis}
+\section{Satisfiability of a Formula in the Synthesis Fragment} \label{sec:synthesis}
 %

### 2014-07-03 papers/termination/paper.tex
Conditional loops, nested loops

@@
 
+\iffalse
 \noindent {\bf Non-determinism.} A non-termination diagnosis procedure searching for a closed recurrence set under-approximates the transition relation, i.e. 
@@
 %parametric recurrence set
+
+\fi
 
@@
 terminates, but if the loop were considered in isolation we would not be able to prove that
-it terminates.  We will now consider the problem of encoding the loop's environment as
-a synthesis formula.
+it terminates.  We must therefore somehow encode a loop's interaction with the rest of
+the program in order to do a sound termination analysis.
+
+Let us assume that we have done some preprocessing of our program which has identified
+loops, straight line code blocks and the control flow between these.  In particular,
+the control flow analysis has determined which order these code blocks execute in,
+and the nesting structure of the loops.
 
@@
 executing in, we say that $L$ is \emph{conditionally terminating}.
-The loop we consider now is $L=\langle I,G,T \rangle$; it has an additional predicate $I(x)$ 
-representing the states in which the loop can begin executing.
-Conditional termination is equivalent to the
-formula of Defintion~\ref{def:conditional-termination-formula}. If this formula is satisfiable,
+The information we require of the environment is a predicate $I$ which
+overapproximates the set of states that $L$ may begin executing in.
+That is to say, for each state $x$ that is reachable on entry to $L$,
+we have $I(x)$.
+
+Conditional termination is then equivalent to the
+formula {\bf [CT]} of Defintion~\ref{def:conditional-termination-formula}. If this formula is satisfiable,
 two witnesses are returned:
@@
 $W$ is called a \emph{supporting invariant} for $L$ and $R$ proves termination relative to $W$.
+We require that $I \wedge G$ is strong enough to establish the base case of $W$'s inductiveness.
 
@@
 However, formula {\bf [CT]} is satisfiable, as witnessed by the ranking function
-$R(\langle x,y\rangle ) = x$ and supporting invariant $W(\langle x, y \rangle ) \triangleq y = 1$.
+$R(\langle x,y\rangle) = x$ and supporting invariant $W(\langle x, y \rangle ) \triangleq y = 1$.
 This constitutes a proof that the program as a whole terminates, since the loop always begins
 executing in a state that guarantees its termination.
+
+\noindent {\bf Nested loops.}
+If $L$'s body contains another loop $L'$, we cannot directly apply either formula {\bf [UT]} or
+{\bf [CT]} or prove that $L$ terminates.  This is because we don't have direct access to the
+single step transition relation for $L$'s body.  Instead, we consider $L'$ to be part of
+$L$'s environment, which we model with a relation $S$.  We require that $S$ overapproximates
+$L'$ in the sense that it satisfies the following formula:
+\begin{align*}
+ \forall x, x'. G(x) \wedge W(x) \wedge T'^*(x, x') \wedge \lnot G'(x') \rightarrow S(x, x')
+\end{align*}
+Where $G$ and $G'$ are respectively $L$ and $L'$'s guards, and $T'^*$ is the reflexive transitive closure of
+the inner loop ($L'$)'s transition relation.
+
+To give an example, if we wish to prove that the outer loop in Figure~\ref{fig:nested} terminates,
+it suffices to summarise the inner loop with the relation:
+\begin{align*}
+S(\langle x, y, z \rangle, \langle x', y', z' \rangle) & \triangleq x' = x \wedge y = 1
+\end{align*}
+Then the ranking function $R(x, y, z) = x$ proves termination of the outer loop $L(G, S)$.
+
+\begin{figure}
+\begin{lstlisting}
+while (x > 0) {
+  z = x, y = z + 1;
+  while (z > 0) {
+    z--, y--;
+  }
+  x -= y;
+}
+\end{lstlisting}
+\caption{A nested loop\label{fig:nested}}
+\end{figure}
+
+\iffalse
 
@@
 \noindent{\bf Nested loops.}
+\fi
 

### 2014-07-03 papers/termination/table.tex
Most recent benchmarks

--- 2014-06-27 papers/termination/table.tex
@@
-svcomp1 & 17 & unk & true & true & false & false & 1 & 5 & term & 9.6s\\
+loop1.c & 6 &  & true & true & false & false & 1 & 5 & term & 2.5s & 1\\
 
-svcomp2 & 13 & unk & false & true & false & false & 1 & 5 & term & 10.2s\\
+loop2.c & 7 &  & true & true & false & false & 1 & 5 & term & 2.5s & 3\\
 
-svcomp3 & 15 & unk & true & true & false & false & unk & 5 & unk & T/O\\
+loop3.c & 8 & true & true & true & true & false & 1 & 5 & term & 2.7s & 3\\
 
-svcomp4 & 14 & unk & true & unk & true & false & unk & 5 & unk & T/O\\
+loop4.c & 13 &  & true & true & false & false & 2 & 5 & term & 3.2s & 17\\
 
-svcomp7 & 9 & unk & true & true & true & false & 1 & 5 & unk & T/O\\
+loop5.c & 13 &  & true & true & false & false & 2 & 5 & term & 2.9s & 8\\
 
-svcomp8 & 9 & unk & true & unk & false & false & unk & 5 & unk & T/O\\
+loop6.c & 6 &  & false & true & false & false & 1 & 5 & term & 2.5s & 1\\
 
-svcomp9 & 12 & unk & true & unk & false & false & unk & 5 & term & 9.3s\\
+loop7.c & 7 &  & true & unk & false & false & unk & 5 & term & 18.2s & 10\\
 
-svcomp10 & 15 & unk & true & true & false & false & 1 & 5 & term & 61.2s\\
+loop8.c & 6 &  & true & true & true & true & 1 & 5 & term & 2.4s & 1\\
 
-svcomp11 & 16 & unk & true & true & false & false & 1 & 5 & unk & T/O\\
+loop9.c & 11 & true & true & false & false & false & 1 & 5 & term & 3.7s & 3\\
 
-svcomp17 & 17 & unk & true & true & false & false & 1 & 5 & nonterm & 19.0s\\
+loop10.c & 6 &  & true & true & false & false & 1 & 5 & term & 2.5s & 1\\
 
-svcomp25 & 12 & unk & true & true & false & false & 1 & 5 & unk & T/O\\
+loop43.c & 9 &  & true & false & false & false & ? & 5 & term & 3.7s & 4\\
 
-svcomp26 & 23 & unk & true & true & false & false & 1 & 5 & term & 38.7s\\
+loop44.c & 9 & false & true & false & false & false & ? & 5 & nonterm & 2.7s & 4\\
 
-svcomp27 & 13 & unk & true & true & false & false & 1 & 5 & unk & T/O\\
+loop45.c & 15 &  & true & false & false & false & ? & 5 & term & 3.4s & 19\\
 
-svcomp28 & 18 & unk & true & true & false & false & 1 & 5 & unk & T/O\\
+loop46.c & 9 &  & true & false & false & false & ? & 5 & term & 2.7s & 4\\
 
-svcomp29 & 9 & unk & true & true & false & false & 1 & 5 & unk & T/O\\
+loop47.c & 6 &  & true & false & false & false & ? & 5 & term & 3.7s & 4\\
 
-svcomp37 & 22 & unk & true & true & false & false & 1 & 5 & term & 12.3s\\
+loop48.c & 7 &  & true & false & false & false & ? & 5 & term & 2.5s & 3\\
 
-svcomp38 & 16 & unk & true & true & false & false & 1 & 5 & unk & T/O\\
+svcomp1.c & 17 &  & true & true & false & false & 1 & 5 & term & 2.6s & 5\\
 
-svcomp39 & 56 & unk & true & true & false & false & 1 & 5 & unk & T/O\\
+svcomp2.c & 13 &  & false & true & false & false & 1 & 5 & term & 2.6s & 6\\
 
-svcomp42 & 22 & unk & true & true & false & false & 1 & 5 & unk & T/O\\
+svcomp3.c & 15 &  & true & true & false & false & unk & 5 & term & 36.8s & 36\\
+
+svcomp4.c & 14 & false & true & unk & true & false & unk & 5 & nonterm & 50.6s & 75\\
+
+svcomp5.c & 18 &  & true & false & true & false & unk & 5 & ? & T/O & 46\\
+
+svcomp7.c & 9 & true & true & true & true & false & 1 & 5 & term & 9.7s & 7\\
+
+svcomp8.c & 9 &  & true & unk & false & false & unk & 5 & ? & T/O & 38\\
+
+svcomp9.c & 12 &  & true & unk & false & false & unk & 5 & term & 2.9s & 11\\
+
+svcomp10.c & 15 &  & true & true & false & false & 1 & 5 & term & 3.4s & 19\\
+
+svcomp11.c & 16 &  & true & true & false & false & 1 & 5 & term & 5.5s & 27\\
+
+svcomp13.c & 11 &  & true & true & false & false & 1 & 5 & ? & T/O & 31\\
+
+svcomp14.c & 18 &  & true & true & false & false & 1 & 5 & ? & T/O & 69\\
+
+svcomp15.c & 30 &  & true & true & false & false & 1 & 5 & ? & T/O & 30\\
+
+svcomp16.c & 17 &  & true & true & false & false & 1 & 5 & term & 2.5s & 3\\
+
+svcomp17.c & 17 &  & true & true & false & false & 1 & 5 & nonterm & 0.7s & 1\\
+
+svcomp25.c & 12 &  & true & true & false & false & 1 & 5 & nonterm & 0.7s & 1\\
+
+svcomp26.c & 23 &  & true & true & false & false & 1 & 5 & term & 3.0s & 11\\
+
+svcomp27.c & 13 & false & true & true & false & false & 1 & 5 & ? & T/O & 76\\
+
+svcomp28.c & 18 &  & true & true & false & false & 1 & 5 & ? & T/O & 82\\
+
+svcomp29.c & 9 &  & true & true & false & false & 1 & 5 & ? & T/O & 1\\
+
+svcomp37.c & 22 &  & true & true & false & false & 1 & 5 & term & 2.9s & 14\\
+
+svcomp38.c & 16 & true & true & true & false & false & 1 & 5 & nonterm & 1.0s & 1\\
+
+svcomp39.c & 56 &  & true & true & false & false & 1 & 5 & nonterm & 1.0s & 1\\
+
+svcomp42.c & 22 &  & true & true & false & false & 1 & 5 & ? & T/O & 84\\

### 2014-07-03 papers/termination/paper.tex
Add large example

@@
 
+\iffalse
 Let us assume that we have done some preprocessing of our program which has identified
@@
 and the nesting structure of the loops.
+\fi
 
@@
 \end{figure}
+
+\noindent {\bf Composing the environment with a loop.}
+While proving that loop $L$ terminates, we create constraints on the environment in the
+form of the predicate $I$ and the summary relation $S$.  For each loop $L'$ nested in the body
+of $L$, we must add some more constraints saying that $S$ overapproximates $L'$.
+Similarly, for each loop $L''$ appearing before $L$ in the program's control flow, we
+must add constraints saying that $I$ overapproximates the states reachable via $L''$.
+The details of how to create these constraints are beyond the scope of this paper, so
+rather than presenting the algorithm for generating the constraints, we direct the
+reader's attention to Figure~\ref{fig:environment-model}.  This figure contains a non-trivial program
+along with the constraint system
+encoding the termination of the outer loop.  Since the program contains sequential composition
+of loops and nested loops, the constraint
+system involves termination invariants $W_1, W_2, W_3$ and the inner loop summary $S$.
+Note that despite its complexity, the constraint system is still a formula in the synthesis fragment.
+Since the outer loop of this program does indeed terminate, the system is satisfiable, as witnessed by:
+\begin{align*}
+W_1(\langle v, x, y, z \rangle ) & \triangleq x \leq y + 1 \\
+W_2(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
+W_1(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
+S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge z' + v' = x \\
+R(x) & = x
+\end{align*}
+
+\begin{figure*}
+\begin{minipage}{0.2\textwidth}
+\begin{lstlisting}
+x = 0;
+while (x <= y)
+  x++;
+
+while (x > 0) {
+  z = x;
+  v = 0;
+
+  while (v < y) {
+    z--;
+    v++;
+  }
+
+  x -= z;
+  y -= z;
+}
+\end{lstlisting}
+\end{minipage}
+\vline
+\begin{minipage}{0.8\textwidth}
+\begin{align*}
+ \exists W_1, W_2, W_3, R, S . \forall v, x, y, z, v', x', y', z' . \\
+   x = 1 & \rightarrow W_1(v, x, y, z) \, \wedge \\
+   x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
+   x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x+1, y, z) \, \wedge \\
+   x > 0 \wedge W_2(v, x, y, z) & \rightarrow W_3(0, x, y, x) \, \wedge \\
+   v < y \wedge W_3(v, x, y, z) & \rightarrow W_3(v+1, x, y, z-1) \, \wedge \\
+   v \geq y \wedge W_3(v, x, y, z) & \rightarrow W_2(v, x-z, y-z, z) \, \wedge \\
+   S(\langle v, x, y, z, \rangle, \langle v', x', y', z' \rangle) \wedge v < y & \rightarrow S(\langle v, x, y, z \rangle, \langle v+1, x, y, z-1 \rangle) \, \wedge \\
+   x > 0 \wedge W_2(v, x, y, z) & \rightarrow S(\langle v, x, y, z \rangle, \langle 0, x, y, x \rangle) \, \wedge \\
+   x > 0 \wedge W_2(v, x, y, z) \wedge S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' \geq y' & \rightarrow \\
+   R(v, x, y, z) > 0 \wedge R(v, x, y, z) > R(v', x' - z', y'-z', z')
+\end{align*}
+\end{minipage}
+\caption{A non-trivial program and its constraint system\label{fig:environment-model}}
+\end{figure*}
 

### 2014-07-03 papers/termination/kalashnikov.tex

@@
 \subsection{The Synthesis Specification}
-Our (non-)termination specifications are given as program fragments
+The logic we use to express the synthesis formula, i.e. the (non-)termination specifications, % are given as program fragments
 %More specifically, the logic we use to express the synthesis formula
-in a subset of C that we call \newC.  
+is a subset of C that we call \newC.  
 The characteristic property of a \newC  program is that safety can be decided for
@@
 \todo{update the figure and make it look more like a grammar.}
-
-\subsection{The Synthesis Algorithm}
-As the quantifier free first-order fragment of the logic is efficiently decidable and
-$P$ is expressible as a ground term of the logic, we can use
-Counterexample Guided Inductive Synthesis (CEGIS)~\cite{lezama-thesis,sketch} to
-check the validity of the synthesis formula.
-
-The core of the CEGIS algorithm is
-the refinement loop shown in Fig.~\ref{fig:abstract-refinement} and
-detailed in Algorithm~\ref{fig:abstract-refinement-code}.  
-The algorithm is divided into two
-components: {\sc synth} and {\sc verif}. For a finite set of test
-inputs that we track, {\sc synth} synthesises a candidate program that is correct for those tests.
-Subsequently, {\sc Verif} verifies the candidate program by searching for an input on
-which it does not satisfy the specification. If no such input
-can be found, the candidate program is correct.  Otherwise, the new input is added
-to the set of tracked inputs.  
-%If the input set is finite, this procedure is guaranteed to terminate.
 
@@
 
-In order to determine the validity of the {\sc synth} formula, we can
-check the {\sc synth} program for safety.  The {\sc synth} program is a
-\newC program, which means we can check its safety by invoking a Bounded
-Model Checker, such as {\sc cbmc}.
+\subsection{The Synthesis Algorithm}
+As safety of a \newC program is efficiently 
+%the quantifier free first-order fragment of the logic is efficiently 
+decidable and a termination proof is expressible as a ground term of the \newC logic, we use
+Counterexample Guided Inductive Synthesis (CEGIS)~\cite{lezama-thesis,sketch} to
+check the validity of the synthesis formula. 
+The core of the CEGIS algorithm is the refinement loop shown in Fig.~\ref{fig:abstract-refinement} and
+detailed in Algorithm~\ref{fig:abstract-refinement-code}.  
+The algorithm is divided into two
+components: {\sc synth} (see Figure~\ref{fig:synth-dfd}) and {\sc verif}.% described next.
 
-In some cases, it is faster to use an explicit-state model checker rather
-than a Bounded Model Checker.  This is particularly true when we are checking
-the {\sc verif} formula, where we have observed that incorrect programs tend
-to be incorrect on a large fraction of the input space.  Counterexamples
-are then very easy to find by explicit enumeration of a few inputs.
-Since \newC is a fragment of C, we can generate an explicit-state
-model checker using the same source files that we pass to {\sc cbmc}
-and adding a small function to enumerate possible inputs.
-We can run the explicit-state model checker
-in parallel with {\sc cbmc} and take the answer of whichever happens
-to terminate first, stopping the other process.  This procedure is
-depicted for {\sc synth} in Fig.~\ref{fig:synth-dfd}, and is similar for {\sc verif}.
+Given a (non-)termination specification (in the synthesis fragment)    
+and a finite set of test inputs tracked so far, {\sc synth} generates a program \texttt{synth.c} asserting that some non-deterministic proof taken as input  
+is incorrect on at least one of the tracked inputs.
+Thus, \texttt{synth.c} is unsafe iff there is some (non-)termination proof that is correct for all of the tracked inputs.  
+By using testing techniques, we search for such a proof, i.e. 
+we synthesise a candidate (non-)termination proof that is correct for all the tracked inputs.  
+The testing techniques we use are: model checking,
+exhaustive search and genetic programming (see Section~\ref{sec:gp}).  
+%For the latter 2 techniques, we combine SYNTH with another program
+%(GP or EXPLICIT respectively) to create a program that searches for test vectors (proofs) that cause SYNTH
+%to crash.
+As \texttt{synth.c}
+is a \newC program, we can check its safety by invoking a Bounded
+Model Checker, such as {\sc cbmc} \cite{}.
+We run the three available testing techniques %explicit-state model checker
+in parallel and take the answer of whichever happens
+to terminate first, stopping the other processes.
+
+Subsequently, {\sc Verif} uses model checking or exhaustive search to verify the candidate proof by searching for an input on
+which it does not satisfy the specification. If no such input
+can be found, the candidate proof is correct.  Otherwise, the new input is added
+to the set of tracked inputs and the CEGIS loop is reiterated.
+
+%If the input set is finite, this procedure is guaranteed to terminate.
+
+%% In some cases, it is faster to use an explicit-state model checker rather than a Bounded Model Checker.  
+%% This is particularly true for the {\sc verif} component, 
+%% where we have observed that incorrect programs tend
+%% to be incorrect on a large fraction of the input space.  Counterexamples
+%% are then very easy to find by explicit enumeration of a few inputs.
+%% Since \newC is a fragment of C, we can generate an explicit-state
+%% model checker using the same source files that we pass to {\sc cbmc}
+%% and adding a small function to enumerate possible inputs.
+
+%% ---  We begin with the program under analysis (the PUA).  From this we produce a specification (in second order logic)
+%% encoding the termination of the PUA.  From the logical specification, we produce a program SYNTH.
+%% SYNTH takes as input a proof and asserts that the proof is incorrect on at least one of the inputs being tracked so far.
+%% Therefore, SYNTH is unsafe iff there is some proof that is correct for all of the tracked inputs.  Now we use testing
+%% techniques to try to find a proof that causes SYNTH to crash.  The testing techniques we use are: model checking,
+%% exhaustive search and genetic programming.  For the latter 2 techniques, we combine SYNTH with another program
+%% (GP or EXPLICIT respectively) to create a program that searches for test vectors (proofs) that cause SYNTH
+%% to crash.
+
+%% In the {\sc synth} component, in order to find a candidate proof $P$ satisfying
+%% $\sigma(i_1, P(i_1)) \land \ldots \land \sigma(i_N, P(i_N))$
+%% for the tracked inputs $(i_1, \ldots, i_N)$, 
+%% we attempt a safety proof.
+%In order to determine the validity of the {\sc synth} formula, we can
+%check the {\sc synth} program for safety.
 
@@
 
-\node [file] (spec) {\sc Specification};
+\node [file] (spec) {\sc Termination Specification};
+\path (spec.south)+(0.0, -1) node [file] (inputs) {\sc Tracked Inputs};
 %% \path (synth.south)+(0.0, -0.5) node [file] (tests) {\sc tests.c};
@@
 %% \path (spec.east)+(2.0, -0.25) node [process] (merged) {merge};
+\path (spec.east)+(2.0, -0.75) node [file] (file) {synth.c};
 
-\path (spec.east)+(2.0, 1.5) node [process] (cbmc) {\sc cbmc};
-\path (spec.east)+(2.0, 0.0) node [process] (gcc) {\sc gcc};
-\path (spec.east)+(2.0, -1.5) node [process] (gp) {\sc GP};
+\path (file.east)+(2.0, 1.5) node [process] (cbmc) {\sc cbmc};
+\path (file.east)+(2.0, 0.0) node [process] (gcc) {\sc Search};
+\path (file.east)+(2.0, -1.5) node [process] (gp) {\sc GP};
 
-\path (gcc.east)+(2.5, 0.0) node [file] (out) {Proof};
+\path (gcc.east)+(2.5, 0.0) node [file] (out) {Candidate Proof};
 
@@
 
-\path [line] (spec) -- (cbmc);
-\path [line] (spec) -- (gcc);
-\path [line] (spec) -- (gp);
+\path [line] (spec) -- (file);
+\path [line] (inputs) -- (file);
+
+\path [line] (file) -- (cbmc);
+\path [line] (file) -- (gcc);
+\path [line] (file) -- (gp);
 

### 2014-07-03 papers/termination/paper.tex
Simplify figure 4 slightly
revising kalashnikov
Correct big example
"Out of scope" -> "Standard"

@@
 must add constraints saying that $I$ overapproximates the states reachable via $L''$.
-The details of how to create these constraints are beyond the scope of this paper, so
+The details of how to create these constraints are standard, so
 rather than presenting the algorithm for generating the constraints, we direct the
@@
 W_2(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
-W_1(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
-S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge z' + v' = x \\
+S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge z' + v' = x' \wedge v' \leq y' \\
 R(x) & = x
@@
 \begin{figure*}
-\begin{minipage}{0.2\textwidth}
+\begin{framed}
+\begin{minipage}{0.19\textwidth}
 \begin{lstlisting}
@@
 \vline
-\begin{minipage}{0.8\textwidth}
+\begin{minipage}{0.82\textwidth}
 \begin{align*}
- \exists W_1, W_2, W_3, R, S . \forall v, x, y, z, v', x', y', z' . \\
+ \exists W_1, W_2, R, S . \forall v, x, y, z, v', x', y', z' . \\
    x = 1 & \rightarrow W_1(v, x, y, z) \, \wedge \\
@@
    x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x+1, y, z) \, \wedge \\
-   x > 0 \wedge W_2(v, x, y, z) & \rightarrow W_3(0, x, y, x) \, \wedge \\
-   v < y \wedge W_3(v, x, y, z) & \rightarrow W_3(v+1, x, y, z-1) \, \wedge \\
-   v \geq y \wedge W_3(v, x, y, z) & \rightarrow W_2(v, x-z, y-z, z) \, \wedge \\
+   %x > 0 \wedge W_2(v, x, y, z) & \rightarrow W_3(0, x, y, x) \, \wedge \\
+   %v < y \wedge W_3(v, x, y, z) & \rightarrow W_3(v+1, x, y, z-1) \, \wedge \\
+   %v \geq y \wedge W_3(v, x, y, z) & \rightarrow W_2(v, x-z, y-z, z) \, \wedge \\
    S(\langle v, x, y, z, \rangle, \langle v', x', y', z' \rangle) \wedge v < y & \rightarrow S(\langle v, x, y, z \rangle, \langle v+1, x, y, z-1 \rangle) \, \wedge \\
@@
    x > 0 \wedge W_2(v, x, y, z) \wedge S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' \geq y' & \rightarrow \\
-   R(v, x, y, z) > 0 \wedge R(v, x, y, z) > R(v', x' - z', y'-z', z')
+   & W_2(v', x' - z', y' - z', z') \, \wedge \\
+   & R(v, x, y, z) > 0 \, \wedge \\
+   & R(v, x, y, z) > R(v', x' - z', y'-z', z')
 \end{align*}
@@
 \caption{A non-trivial program and its constraint system\label{fig:environment-model}}
+\end{framed}
 \end{figure*}
@@
 
-\subsection{Genetic Programming and Incremental Evolution}
+\subsection{Genetic Programming and Incremental Evolution} \label{sec:gp}
 

### 2014-07-03 papers/termination/paper.tex
First stab at GP

@@
  \begin{align*}
-  \exists N, x_0 . \forall x . \exists x' . & I(x_0) \wedge N(x_0) ~ \wedge \\
+  \exists N, x_0 . \forall x . \exists x' . & N(x_0) ~ \wedge \\
                                                         & N(x) \rightarrow G(x) ~ \wedge \\
@@
 \subsection{Genetic Programming and Incremental Evolution} \label{sec:gp}
-
+To generate candidate solutions in the synthesise phase of the refinement loop,
+we can use genetic programming (GP).  The general idea is to start with a population
+of random programs and evolve a good one.  ``Good'' in this case means that the program
+satisfies our specification.  To do this evolution, we compute a fitness for each program
+and repeatedly apply genetic operators to the population based on this fitness.  The operators
+used are {\sc crossover} and {\sc mutate}.  The crossover operator takes two programs
+and combines their code in some what to create a new program.  The mutate operator takes a single
+program and randomly changes parts of its code, again creating a new program.  We arrange
+things so that the fitter a program is, the more likely it is to be selected as a parent
+that will be crossover'd to produce a child in the next generation.  Our fitness function
+is just the number of test cases on which the program met the specification.
+
+\iffalse
 We begin by observing that the asymptotic complexity of all of our synthesis
@@
 \end{conjecture}
+\fi
 

### 2014-07-03 papers/termination/paper.tex
fixed inconsistencies in sec 5

--- 2014-07-03 papers/termination/paper.tex
@@
 
-We will begin our discussion by showing how to encode the (non-)termination of a single loop
+We will begin our discussion by showing how to encode the (non-)termination of program consisting of a single loop 
 in the synthesis fragment.  For the time being, a loop is a pair ${L= \langle G,T \rangle}$,
@@
  \begin{align*}
-  \exists N, x_0 . \forall x . \exists x' . & I(x_0) \wedge N(x_0) ~ \wedge \\
-                                                        & N(x) \rightarrow G(x) ~ \wedge \\
+  \exists N, x_0 . \forall x . \exists x' . & N(x_0) ~ \wedge ~ N(x) \rightarrow G(x) ~ \wedge \\
 							& N(x) \rightarrow T(x, x') \wedge N(x') 
@@
  \begin{align*}
-  \exists N, x_0 . \forall x, x' . & I(x_0) \wedge N(x_0) ~ \wedge \\
-                                                        & N(x) \rightarrow G(x) ~ \wedge \\
+  \exists N, x_0 . \forall x, x' . & N(x_0) ~ \wedge ~ N(x) \rightarrow G(x) ~ \wedge \\
 							& N(x) \wedge T(x, x') \rightarrow N(x') 
@@
 So $R(x) = x$ constitutes a proof that the program in Figure~\ref{fig:motivation.a} is unconditionally
-terminating.
+terminating.\\
 
@@
 Dually to termination, we might want to consider the non-termination of a program.  If a program terminates,
-we can prove this by finding a ranking function and supporting invariant witnessing the satisfiability
-of formula {\bf[CT]}.  What then would a proof of non-termination look like?
+we can prove this by finding a ranking function %and supporting invariant 
+witnessing the satisfiability of formula {\bf[UT]}.  What then would a proof of non-termination look like?
 
@@
 exists a transition to some $s'\in N$. 
-For our case, given a loop $L = \langle I,G,T \rangle$, the notion of open recurrent set is encoded by formula {\bf [ONT]}. %Definition~\ref{def:nonterm-formula}.
+For our case, given a loop $L = \langle G,T \rangle$, the notion of open recurrent set is encoded by formula {\bf [ONT]}. %Definition~\ref{def:nonterm-formula}.
 
@@
 we make use of the notion of \emph{closed recurrence set} introduced by Chen et al. in \cite{DBLP:conf/tacas/ChenCFNO14}. 
-As opposed to a closed recurrence set, this is a $\exists \forall$ property: for each state in the recurrence set $N$, all of its successors 
+As opposed to an open recurrence set, this is a $\exists \forall$ property: for each state in the recurrence set $N$, all of its successors 
 must be in $N$. We note that if $T$ is deterministic, every open recurrence set is also a closed recurrence set (since each
@@
 
- \begin{definition}[Sequential Loops Termination Formula {\bf [ST]}]
-  \begin{align*}
-\label{def:multi-termination-formula}
-  \exists R, Inv_1,..., Inv_n . \forall x_0,...,x_n, x_1',...,x_n'.  & P_0(x_0,x_1) \rightarrow Inv_1(x_1) ~  \\
- & \bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) ~  \\  
- & \bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) \rightarrow Inv_{i+1}(x_{i+1})) ~ \wedge \\
- & Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge G_n(x_n) \rightarrow Inv_n(x_n) ~ \wedge \\
- & Inv_n(x_n) \wedge G_n(x_n) \wedge T(x_n, x_n') \rightarrow Inv_n(x_n') \wedge R(x_n) > 0 \wedge R(x_n) > R(x_n')
- \end{align*}
-\end{definition}
-
-\begin{definition}[Nested Loops Termination Formula {\bf [NT]}]
-\label{def:nested-term-formula}
- \begin{align*}
-  \exists R, S . \forall x, x' . & G_1(x) \wedge P_1(x,x') \rightarrow S(x',x') ~ \wedge \\
-                                & G_2(x') \wedge S(x,x') \wedge T(x',x'')\rightarrow S(x,x'') ~ \wedge \\
-				& G_1(x) \wedge P_1(x,x') \wedge \neg G_2(x') \wedge S(x',x'') \wedge 
-                                P_2(x'',x''') \wedge 
-                                  R(x) > 0 \wedge R(x) > R(x''') 
- \end{align*} 
-\end{definition}
-
-\begin{definition}[Sequential Loops Non-Termination Formula {\bf [SNT]}]
-\label{def:multi-termination-formula}
- \begin{align*}
-  \exists R, Inv_1,..., Inv_n . \forall x_0,...,x_n, x_1',...,x_n'.  & P_0(x_0,x_1) \rightarrow Inv_1(x_1) ~  \\
- & \bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) ~  \\  
- & \bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) \rightarrow Inv_{i+1}(x_{i+1})) ~ \wedge \\
- & Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge N(x_n)  ~ \wedge \\
-                                                        & N(x_n) \rightarrow G_n(x_n) ~ \wedge 
-							 N(x_n) \wedge T_n(x_n, x_n') \rightarrow N(x_n') 
- \end{align*}
-\end{definition}
+%%  \begin{definition}[Sequential Loops Termination Formula {\bf [ST]}]
+%%   \begin{align*}
+%% \label{def:multi-termination-formula}
+%%   \exists R, Inv_1,..., Inv_n . \forall x_0,...,x_n, x_1',...,x_n'.  & P_0(x_0,x_1) \rightarrow Inv_1(x_1) ~  \\
+%%  & \bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) ~  \\  
+%%  & \bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) \rightarrow Inv_{i+1}(x_{i+1})) ~ \wedge \\
+%%  & Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge G_n(x_n) \rightarrow Inv_n(x_n) ~ \wedge \\
+%%  & Inv_n(x_n) \wedge G_n(x_n) \wedge T(x_n, x_n') \rightarrow Inv_n(x_n') \wedge R(x_n) > 0 \wedge R(x_n) > R(x_n')
+%%  \end{align*}
+%% \end{definition}
+
+%% \begin{definition}[Nested Loops Termination Formula {\bf [NT]}]
+%% \label{def:nested-term-formula}
+%%  \begin{align*}
+%%   \exists R, S . \forall x, x' . & G_1(x) \wedge P_1(x,x') \rightarrow S(x',x') ~ \wedge \\
+%%                                 & G_2(x') \wedge S(x,x') \wedge T(x',x'')\rightarrow S(x,x'') ~ \wedge \\
+%% 				& G_1(x) \wedge P_1(x,x') \wedge \neg G_2(x') \wedge S(x',x'') \wedge 
+%%                                 P_2(x'',x''') \wedge 
+%%                                   R(x) > 0 \wedge R(x) > R(x''') 
+%%  \end{align*} 
+%% \end{definition}
+
+%% \begin{definition}[Sequential Loops Non-Termination Formula {\bf [SNT]}]
+%% \label{def:multi-termination-formula}
+%%  \begin{align*}
+%%   \exists R, Inv_1,..., Inv_n . \forall x_0,...,x_n, x_1',...,x_n'.  & P_0(x_0,x_1) \rightarrow Inv_1(x_1) ~  \\
+%%  & \bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) ~  \\  
+%%  & \bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) \rightarrow Inv_{i+1}(x_{i+1})) ~ \wedge \\
+%%  & Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge N(x_n)  ~ \wedge \\
+%%                                                         & N(x_n) \rightarrow G_n(x_n) ~ \wedge 
+%% 							 N(x_n) \wedge T_n(x_n, x_n') \rightarrow N(x_n') 
+%%  \end{align*}
+%% \end{definition}
  \end{framed}
@@
 terminates, but if the loop were considered in isolation we would not be able to prove that
-it terminates.  We must therefore somehow encode a loop's interaction with the rest of
-the program in order to do a sound termination analysis.
+it terminates.  We must therefore encode a loop's interaction with the rest of
+the program in order to do a sound termination analysis.\\
 
@@
 This constitutes a proof that the program as a whole terminates, since the loop always begins
-executing in a state that guarantees its termination.
+executing in a state that guarantees its termination.\\
 
@@
 \end{align*}
-Then the ranking function $R(x, y, z) = x$ proves termination of the outer loop $L(G, S)$.
+Then the ranking function $R(x, y, z) = x$ proves termination of the outer loop $L(G, S)$.\\
 

### 2014-07-03 papers/termination/paper.tex
Experiments

--- 2014-07-03 papers/termination/paper.tex
@@
 \section{Experiments}
-
+We implemented our method by backing onto the {\sc Kalashnikov} program synthesiser.
+We ran the tool on 40 benchmarks taken from the termination category of SVCOMP'15.
+The experiments were performed on a 4-core 3.3GHz i5-2500k with 8GB of RAM.
+
+\iffalse
 We proved termination for a bunch of programs, see Fig.~\ref{fig:linear} and Fig.~\ref{fig:nonlinear}.
@@
 \end{figure*}
+\fi
 

### 2014-07-04 papers/termination/table.tex
Reran benchmarks

@@
-loop1.c & 6 &  & true & true & false & false & 1 & 5 & term & 2.5s & 1\\
+loop1.c & 6 &  & true & true & false & false & 1 & 5 & term & 2.4s & 1\\
 
@@
 
-loop3.c & 8 & true & true & true & true & false & 1 & 5 & term & 2.7s & 3\\
+loop3.c & 8 & true & true & true & true & false & 1 & 5 & term & 2.8s & 3\\
 
-loop4.c & 13 &  & true & true & false & false & 2 & 5 & term & 3.2s & 17\\
+loop4.c & 13 &  & true & true & false & false & 2 & 5 & term & 3.1s & 17\\
 
@@
 
-loop6.c & 6 &  & false & true & false & false & 1 & 5 & term & 2.5s & 1\\
+loop6.c & 6 &  & false & true & false & false & 1 & 5 & term & 2.4s & 1\\
 
@@
 
-loop9.c & 11 & true & true & false & false & false & 1 & 5 & term & 3.7s & 3\\
+loop9.c & 11 & true & true & false & false & false & 1 & 5 & term & 3.6s & 3\\
 
@@
 
-loop44.c & 9 & false & true & false & false & false & ? & 5 & nonterm & 2.7s & 4\\
+loop44.c & 9 & false & true & false & false & false & ? & 5 & nonterm & 2.6s & 4\\
 
@@
 
-loop47.c & 6 &  & true & false & false & false & ? & 5 & term & 3.7s & 4\\
+loop47.c & 6 &  & true & false & false & false & ? & 5 & term & 3.6s & 4\\
 
@@
 
-svcomp1.c & 17 &  & true & true & false & false & 1 & 5 & term & 2.6s & 5\\
+loop49.c & 6 &  &  &  &  &  &  & 5 & ? & T/O & 12\\
+
+svcomp1.c & 17 &  & true & true & false & false & 1 & 5 & term & 2.8s & 5\\
 
@@
 
-svcomp3.c & 15 &  & true & true & false & false & unk & 5 & term & 36.8s & 36\\
+svcomp3.c & 15 &  & true & true & false & false & unk & 5 & term & 36.7s & 36\\
 
-svcomp4.c & 14 & false & true & unk & true & false & unk & 5 & nonterm & 50.6s & 75\\
+svcomp4.c & 14 & false & true & unk & true & false & unk & 5 & nonterm & 50.5s & 75\\
 
 svcomp5.c & 18 &  & true & false & true & false & unk & 5 & ? & T/O & 46\\
+
+svcomp6.c & 11 &  & true & true & false & false & 1 & 5 & ? & T/O & 16\\
 
@@
 
-svcomp9.c & 12 &  & true & unk & false & false & unk & 5 & term & 2.9s & 11\\
+svcomp9.c & 12 &  & true & unk & false & false & unk & 5 & term & 2.8s & 11\\
 
@@
 
-svcomp13.c & 11 &  & true & true & false & false & 1 & 5 & ? & T/O & 31\\
+svcomp12.c & 13 &  & true & true & false & false & 1 & 5 & term & 5.9s & 28\\
 
-svcomp14.c & 18 &  & true & true & false & false & 1 & 5 & ? & T/O & 69\\
+svcomp13.c & 11 &  & true & true & false & false & 1 & 5 & ? & T/O & 30\\
 
-svcomp15.c & 30 &  & true & true & false & false & 1 & 5 & ? & T/O & 30\\
+svcomp14.c & 18 &  & true & true & false & false & 1 & 5 & ? & T/O & 66\\
 
-svcomp16.c & 17 &  & true & true & false & false & 1 & 5 & term & 2.5s & 3\\
+svcomp15.c & 30 &  & true & true & false & false & 1 & 5 & ? & T/O & 28\\
 
-svcomp17.c & 17 &  & true & true & false & false & 1 & 5 & nonterm & 0.7s & 1\\
+svcomp16.c & 17 &  & true & true & false & false & 1 & 5 & term & 2.7s & 3\\
 
-svcomp25.c & 12 &  & true & true & false & false & 1 & 5 & nonterm & 0.7s & 1\\
+svcomp17.c & 18 &  & true & true & false & false & 1 & 5 & ? & T/O & 45\\
 
-svcomp26.c & 23 &  & true & true & false & false & 1 & 5 & term & 3.0s & 11\\
+svcomp18.c & 19 &  & true & true & false & false & 1 & 5 & ? & T/O & 27\\
+
+svcomp25.c & 12 &  & true & true & false & false & 1 & 5 & ? & T/O & 15\\
+
+svcomp26.c & 23 &  & true & true & false & false & 1 & 5 & term & 2.9s & 11\\
 
@@
 
-svcomp37.c & 22 &  & true & true & false & false & 1 & 5 & term & 2.9s & 14\\
+svcomp37.c & 22 &  & true & true & false & false & 1 & 5 & term & 2.8s & 14\\
 
-svcomp38.c & 16 & true & true & true & false & false & 1 & 5 & nonterm & 1.0s & 1\\
+svcomp38.c & 17 & true & true & false & true & false & 1 & 5 & ? & T/O & 48\\
 
-svcomp39.c & 56 &  & true & true & false & false & 1 & 5 & nonterm & 1.0s & 1\\
+svcomp39.c & 56 &  & true & true & false & false & 1 & 5 & ? & T/O & 1\\
+
+svcomp40.c & 12 &  & true & true & false & false & 1 & 5 & ? & T/O & 20\\
+
+svcomp41.c & 12 &  & true & true & false & false & 1 & 5 & ? & T/O & 26\\
 

### 2014-07-04 papers/termination/paper.tex
Soundness & completeness

@@
 
+\begin{proof}
+ The ranking function constitutes a constructive, machine checked proof that the program terminates.
+\end{proof}
+
 \begin{theorem}
- {\sc Headshot} is complete -- if a program terminates, {\sc Headshot} will find a proof that it does.
+ {\sc Headshot} is complete for terminating programs -- if a program terminates, {\sc Headshot} will find a proof that it does.
 \end{theorem}
+
+\begin{proof}
+ If the program under analysis $P$ has variables $v_1, \ldots, v_k$ each of which is $b$ bits wide, its state space $\mathcal{S}$ is of size $2^{bk}$.
+ A ranking function $R: \mathcal{S} \to \mathcal{D}$ for $P$ exists iff $P$ terminates.  Since $R$ is injective, we have that
+ $\| \mathcal{D} \| \geq \| \mathcal{S} \|$.  If $\| \mathcal{D} \| > \| \mathcal{S} \|$, we can construct a function $R': \mathcal{S} \to \mathcal{D'}$
+ with $ \| \mathcal{D'} = \| \mathcal{S} \|$ by just setting $R' = R|_\mathcal{S}$, i.e. $R'$ is just the restriction of $R$ to $\mathcal{S}$
+ Since $\mathcal{S}$ already comes equipped with a natural well ordering we can also construct $R'' = \iota \circ R'$
+ where $\iota: \mathcal{D'} \to \mathcal{S}$ is the unique order isomorphism from $\mathcal{D'}$ to $\mathcal{S}$.
+ So assuming that $P$ terminates, there is some ranking function $R''$ that is just a permutation of $\mathcal{S}$.
+
+ It is fairly easy to see that any finite permutation is computed by a finite program in the proof language generated
+ by {\sc Kalashnikov}.  A very inefficient construction which computes the permutation $R(x) = x+1$ for one variable
+ is:
+
+ \begin{verbatim}
+t1 = 0
+t2 = v1 == 0
+t3 = ITE(t2, 1, t1)
+t4 = v1 == 1
+t5 = ITE(t4, 2, t3)
+...
+ \end{verbatim}
+
+If the number of variables $k > 1$ then in general the ranking function will be lexicographic with dimension $\leq k$.
+
+So we can see that every permutation is computed a program of size at most $b \times 2^{bk + 1}$.  Since
+{\sc Kalashnikov} enumerates all programs in order of length, it will eventually check
+a program computing each permutation of $\mathcal{S}$ and so will eventually check a program
+computing the ranking function $R''$.  Checking that a single program satisfies the specification
+is a decidable problem, and the search as a whole terminates.
+\end{proof}
 

### 2014-07-04 papers/termination/paper.tex
Deterministic completeness

@@
 \begin{definition}[Second-order Unconditional Termination Formula {\bf [UT]}]
-\label{def:termination-formula}
+\label{def:UT}
 \begin{align*}
@@
 \begin{definition}[Non-Termination Formula - Open Recurrence Set  {\bf [ONT]}]
-\label{def:nonterm-formula}
+\label{def:ont}
  \begin{align*}
@@
 \begin{definition}[Non-Termination Formula - Closed Recurrence Set {\bf [CNT]}]
-\label{def:deterministic-nonterm-formula}
+\label{def:cnt}
  \begin{align*}
@@
 The existence of a ranking function is equivalent to the satisfiability
-of the formula from Definition~\ref{def:termination-formula}, so a witness to the satisfiability
+of the formula from Definition~\ref{def:UT}, so a witness to the satisfiability
 of formula {\bf [UT]} is also a ranking function and thus a proof of $L$'s unconditional termination.
@@
 \begin{definition}[Conditional Termination Formula {\bf [CT]}]
-\label{def:conditional-termination-formula}
+\label{def:ct}
  \begin{align*}
@@
 Conditional termination is then equivalent to the
-formula {\bf [CT]} of Defintion~\ref{def:conditional-termination-formula}. If this formula is satisfiable,
+formula {\bf [CT]} of Defintion~\ref{def:ct}. If this formula is satisfiable,
 two witnesses are returned:
@@
 \begin{theorem}
- {\sc Headshot} is sound -- if it says a program terminates then it does.
+\label{thm:sound}
+ {\sc Headshot} is sound -- if it returns a verdict of termination, the program under
+ analysis terminates on all inputs.  Similarly if it returns a verdict of non-termination,
+ the program under analysis has at least one infinite execution.
 \end{theorem}
@@
 \begin{proof}
- The ranking function constitutes a constructive, machine checked proof that the program terminates.
+ {\sc Headshot}'s output constitutes a constructive, machine checked proof of (non-)termination.
 \end{proof}
@@
 \begin{theorem}
+\label{thm:complete-termination}
  {\sc Headshot} is complete for terminating programs -- if a program terminates, {\sc Headshot} will find a proof that it does.
@@
  So assuming that $P$ terminates, there is some ranking function $R''$ that is just a permutation of $\mathcal{S}$.
+ If the number of variables $k > 1$ then in general the ranking function will be lexicographic with dimension $\leq k$
+ and each co-ordinate of the output being a single $b$-bit value.
 
@@
  is:
-
  \begin{verbatim}
@@
  \end{verbatim}
-
-If the number of variables $k > 1$ then in general the ranking function will be lexicographic with dimension $\leq k$.
+It is easy to see how this generalises to computing an arbitrary permutation.
 
@@
 \begin{theorem}
- {\sc Headshot} is a decision procedure -- it always terminates.
+ {\sc Headshot} is complete for deterministic programs -- {\sc Headshot} will terminate when run on
+ any program with a deterministic transition relation.
 \end{theorem}
+
+\begin{proof}
+ If the program under analysis is terminating, then by Theorem~\ref{thm:complete-termination} {\sc Headshot}
+ will terminate having computed a ranking function.
+
+ Conversely if the program is non-terminating, {\sc Headshot} will terminate
+ having computed a recurrence set.  Since, by assumption, the program under analysis is determinstic,
+ Definition~\ref{def:ont} is equivalent to Definition~\ref{def:cnt}.
+ That is to say, each open recurrence set is also a closed recurrence set.  Since the specification for a predicate
+ to encode a closed recurrence set is in the synthesis fragment, {\sc Kalashnikov} will find a program computing
+ such a predicate if one exists.  The proof is similar to the proof of Theorem~\ref{thm:complete-termination}.
+\end{proof}
 

### 2014-07-04 papers/termination/paper.tex
PSPACE complete

@@
 
+Since termination for bitvector programs is known to be PSPACE-complete~\cite{DBLP:conf/tacas/CookKRW10} it
+is perhaps unsurprising that we are able to decide it.  We will now show that our time complexity is
+asymptotically a function of the Kolmogorov complexity of the (non-)termination proof, and argue heuristically
+that this gives our procedure various desirable qualities.
+
 \begin{theorem}

### 2014-07-04 papers/termination/paper.tex
Rephrasing, fix bad hbox
GP stuff

@@
 \subsection{Genetic Programming and Incremental Evolution} \label{sec:gp}
-To generate candidate solutions in the synthesise phase of the refinement loop,
-we can use genetic programming (GP).  The general idea is to start with a population
-of random programs and evolve a good one.  ``Good'' in this case means that the program
-satisfies our specification.  To do this evolution, we compute a fitness for each program
-and repeatedly apply genetic operators to the population based on this fitness.  The operators
-used are {\sc crossover} and {\sc mutate}.  The crossover operator takes two programs
+To generate candidate (non-)termination proofs in the synthesise phase of the refinement loop,
+we make use of genetic programming (GP) \cite{langdon:fogp} as an alternative to model checking and exhaustive search.  
+Genetic programming provides an intelligent and adaptive way of searching through the space of computer programs
+for an individual that is highly fit in solving the problem at hand
+by using the Darwinian principle of survival and reproduction of the fittest.
+Thus, the program that emerges from the genetic programming paradigm is a consequence of fitness.
+
+In our setting, we start with a population of random programs of size ..
+We then genetically breed the population of these programs by applying the genetic operators 
+{\sc crossover} and {\sc mutate}. The crossover operator takes two programs
 and combines their code in some what to create a new program.  The mutate operator takes a single
-program and randomly changes parts of its code, again creating a new program.  We arrange
-things so that the fitter a program is, the more likely it is to be selected as a parent
-that will be crossover'd to produce a child in the next generation.  Our fitness function
-is just the number of test cases on which the program met the specification.
+program and randomly changes parts of its code, again creating a new program. 
+Both these operators are applied to programs selected from the population in 
+proportion to their observed fitness, i.e. a program is fit if it satisfies the (non-)termination specification.
+Thus, our fitness function counts the number of test cases on which a program met the given specification.
+Essentially, we plan a trajectory through the space of computer programs that will lead to programs with improved fitness.
+%by considering those that meet the (non-)termination specification on more inputs until.
+
+\todo{Talk about the fitness landscape?}
+
+\begin{definition}
+ A \emph{fitness landscape} is the space of all programs along with their fitness.
+\end{definition}
+
+A potential issue might be that, in our setting, a fitness landscape isn't really very smooth, e.g. small changes in program representation
+don't correspond to small changes in fitness. However, genetic programming's efficiency was shown to be unrelated to the smoothness 
+of the fitness landscape \cite{langdon:fogp}.
+
+\todo{mention incremental GP?}
+
+%% The general idea is to start with a population
+%% of random programs and evolve a good one.  ``Good'' in this case means that the program
+%% satisfies our specification.  To do this evolution, we compute a fitness for each program
+%% and repeatedly apply genetic operators to the population based on this fitness.  The operators
+%% used are {\sc crossover} and {\sc mutate}.  The crossover operator takes two programs
+%% and combines their code in some what to create a new program.  The mutate operator takes a single
+%% program and randomly changes parts of its code, again creating a new program.  We arrange
+%% things so that the fitter a program is, the more likely it is to be selected as a parent
+%% that will be crossover'd to produce a child in the next generation.  Our fitness function
+%% is just the number of test cases on which the program met the specification.
 
@@
 so the complexity is doubly exponential in the width of $X$.
-
-\begin{definition}
- A \emph{fitness landscape} is the space of all programs along with their fitness.
-\end{definition}
 

### 2014-07-04 papers/termination/paper.tex
Kalashnikov soundness & completeness

--- 2014-07-04 papers/termination/paper.tex
@@
 system involves termination invariants $W_1, W_2, W_3$ and the inner loop summary $S$.
-Note that despite its complexity, the constraint system is still a formula in the synthesis fragment.
+Note that despite its intricacy, the constraint system is still a formula in the synthesis fragment.
 Since the outer loop of this program does indeed terminate, the system is satisfiable, as witnessed by:
@@
 \begin{theorem}
+ \label{thm:synth-sound}
+ Algorithm~\ref{alg:abstract-refinement-code} is sound -- if it terminates with witness $P$ then
+ $P \models \sigma$.
+\end{theorem}
+
+\begin{proof}
+ The procedure {\sc RefinementLoop} terminates only if {\sc Verif} returns ``valid''.  In that
+ case, $\exists x . \lnot \sigma(x, P)$ is unsatisfiable and so $\forall x . \sigma(x, P)$ holds.
+\end{proof}
+
+\begin{theorem}
+ \label{thm:synth-semicomplete}
+ If the existential first-order theory used to express the specification $\sigma$ is decidable,
+ algorithm~\ref{alg:abstract-refinement-code} is semi-complete -- if a program $P \models \sigma$
+ exists then algorithm~\ref{alg:abstract-refinement-code} will terminate.  However if no program
+ satisfies the specification, the algorithm may not terminate.  Furthermore if the domain of inputs $X$
+ is finite, the algorithm is guaranteed to terminate.
+\end{theorem}
+
+\begin{proof}
+ Since the {\sc ExplicitSearch} routine enumerates all programs (as can be seen by induction on
+ the program length $l$), it will eventually enumerate a program that meets the specification
+ in whatever set of inputs are currently being tracked.  Since the first order theory is
+ decidable, the query in {\sc Verif} will succeed for this program, causing the algorithm to terminate.
+ The set of correct programs is therefore recursively enumerable and algorithm~\ref{alg:abstract-refinement-code}
+ enumerates this set, so it is semi-complete.
+
+ If the domain $X$ is finite then the loop in procedure {\sc RefinementLoop} can only
+ iterate $\| X \|$ times, since by this time all of the elements of $X$ would have been
+ added to the inputs set.  Therefore if the {\sc Synth} procedure always terminates (which it does
+ by assumption of decidability), the whole algorithm terminates.
+\end{proof}
+
+\begin{theorem}
 \label{thm:sound}
@@
 \centering
-\small
+\footnotesize
 \begin{tabular}{|l|r|c|c|c|c|c|r|r|c|r|r|}

### 2014-07-04 tests/specs/p19.c
Fix some specs

@@ -1,7 +1,7 @@
 #include "synth.h"
 
 int check(solution_t *solution, word_t args[3]) {
-  word_t res[1];
+  word_t res[NRES];
   exec(&solution->progs[0], args, res);
 
   word_t x = args[0];

### 2014-07-04 papers/termination/paper.tex
Contributions

@@
 \begin{itemize}
-\item We designed a technique for computing ranking functions that correctly accounts for the wrap-around behavior caused by under- and overflows in bit-vector and floating point arithmetic.
-To the best of our knowledge, this is the first approach able to compute ranking functions for programs handling floats.
-Our technique is not restricted to finding linear ranking functions, but can also compute (lexicographic) non-linear  ones.
-We justified the need for such a non restrictive procedure by computing the probability ...
-.
-\item  We rephrased the termination problem as a second-order satisfaction problem and made 
-use of results in genetic programming to efficiently solve it. We have also investigated the effects of genetic operators on the search space for ranking functions and computed theoretical 
-bounds on the convergence time ...
-\item We present a formulation that allows us to prove termination without an expensive reachability check (as we don't use Ramsey-based termination arguments).  In particular,
-we only need a bounded model checker that performs a single unwinding of a loop.
+\item We designed a bit-level accurate technique for computing ranking functions that correctly accounts for the wrap-around behavior
+caused by under- and overflows in bit-vector and floating point arithmetic. Our technique is not restricted to finding linear ranking functions,
+but can also compute lexicographic non-linear  ones.
+
+\item  We rephrased the termination and non-termination problems as second-order satisfaction problems and used
+program synthesis techniques to solve them. In contrast to previous approaches to program synthesis, we found that
+using genetic programming was very effective at generating candidate termination and non-termination proofs.
+
+\item Our procedure does not require the enumeration of lassos, or directly reasoning about the transitive closure of the transition relation.
+To verify our proofs, we only require a single call to a SAT solver.
+
+%\item We present a formulation that allows us to prove termination without an expensive reachability check (as we don't use Ramsey-based termination arguments).  In particular,
+%we only need a bounded model checker that performs a single unwinding of a loop.
+
 %% \item Our technique is able to uniformly handle conditionally-terminating loops, as well as programs with
@@
 
-\item We iteratively construct a termination argument as a pair of a ranking function and supporting invariant through a single refinement loop.
-This is opposed to tools that alternate between calls to a safety prover and calls to a ranking function synthesis tool \cite{}.
-
-\item Our proposed termination proving procedure can also prove non-termination: (non-)termination analysis.
-
 \item We implemented our technique and tried it on a selection of programs handling both bit-vectors and floats.
-
-\item bit level termination and non-termination analysis. This enables
-  the precise treatment of bit-wise operations and arithmetic modulo
-  fixed widths.
 

### 2014-07-04 papers/termination/paper.tex
Contribution
Move related work to end
Added theorems

@@
 
+\item Our procedure is based on a series of source-to-source transformations.  As well as leading to a concise implementation,
+this has the side effect of ensuring that our proof is correct for the exact binary produced by a particular compiler,
+rather than for an abstraction of the program.
+
 \item We implemented our technique and tried it on a selection of programs handling both bit-vectors and floats.
@@
 \end{itemize}
-
-\section{Related Work}
-%\subsection{Termination analysis}
-Automated program termination is a research topic that has received a fair amount of attention from the software verification community.
-In order to compare our technique to the rest of the area, 
-Figure~\ref{fig:handletable} summarises the related works with respect to the restrictions they impose on the transition relations as well as the form of the computed ranking functions.
-
-Somehow expected, most of the techniques are specialised in the synthesis of linear ranking functions for linear programs over integers (or rationals) \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13}.
-Among them, 
-Lee et al. make use of transition predicate abstraction, algorithmic learning, and decision procedures to compute transition
-invariants as proofs for the termination of linear programs \cite{DBLP:conf/cav/LeeWY12}.
-Leike and Heizmann present a new method for the constraint-based synthesis
-of termination arguments for linear loop programs based on
-linear ranking templates \cite{DBLP:conf/tacas/LeikeH14}.
-Linear ranking functions supported by inductive linear invariants for loops with linear guards and transitions: \cite{DBLP:conf/cav/BradleyMS05}. 
-Learning: \cite{DBLP:journals/corr/HeizmannHP14}
-
-%A program is lasso-shaped if it is composed by a stem followed by a single loop without branching, i.e. there is only one path. 
-%Consequently, the termination argument can be very simple. There are numerous techniques specialised in
-%proving termination of lasso-shaped programs efficiently \cite{DBLP:conf/popl/Ben-AmramG13,DBLP:conf/cav/BradleyMS05,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/P04}.
-
-While the synthesis of termination arguments for linear programs over integers is indeed well-covered in the literature, 
-there is very limited work for programs over machine integers.
-Cook et al. present a method based on a reduction to Presburger
-arithmetic, and a template-matching approach for predefined classes of
-ranking functions based on reduction to SAT- and QBF-solving \cite{DBLP:conf/tacas/CookKRW10}.
-Similarly, the only work we are aware of that can compute non-linear ranking functions  
-for imperative loops with polynomial guards and polynomial assignments
-is \cite{DBLP:conf/vmcai/BradleyMS05}. However, this work extends only to polynomials.
-
-Given the lack of research in handling \emph{not-linear programs}, as well as \emph{programs over bit-vectors and floats},  
-our work focuses on covering these areas. 
-One of the obvious conclusions that can be reached from observing Figure~\ref{fig:handletable}, 
-is that most of the works tend to specialise on a certain aspect of termination proving that they can solve efficiently. 
-Conversely to this view, we aim for \emph{generality}, as we do not restrict the form of the synthesised ranking functions, nor the form of the input programs.
-
-As mentioned in Section~\ref{sec:intro}, approaches based on Ramsey's theorem compute a set of local termination conditions that decrease as execution follows through the loop
-and require expensive reachability analyses
-\cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}.
-In an attempt to reduce the complexity of checking the validity of the termination argument, 
-Cook et al present an iterative termination proving procedure that searches for 
-lexicographic termination arguments \cite{DBLP:conf/tacas/CookSZ13}, 
-whereas Kroening et al. strengthen the termination argument such that it becomes a transitive relation \cite{DBLP:conf/cav/KroeningSTW10}.
-
-Proving program termination implies the simultaneous search for a termination argument and a supporting invariant.
-For this purpose, several tools alternate 
-between calls to a safety prover and calls to a ranking function synthesis tool \cite{}. 
-In order to share more information about the state of the termination proofs between the two tools, 
-Brockschmidt et al. use a single representation of the state of the termination proof search that both tools operate over, 
-resulting in performance improvements \cite{DBLP:conf/cav/BrockschmidtCF13}.
-In \cite{DBLP:conf/cav/BradleyMS05}, Bradley et al. combine the generation of ranking functions with the generation of invariants to
-form one constraint solving problem such that the necessary supporting invariants for the ranking function are discovered on demand.
-In our setting, both the ranking function and the supporting invariant are iteratively constructed through the same refinement loop.
-
-%In our setting, a candidate termination argument is iteratively constructed. 
-%A candidate termination argument is composed out of a ranking function 
-%and a supporting invariant.
-
-%% The safety prover proves or disproves the
-%% validity of the current argument via the search for invariants. Refinement of the
-%% current termination argument is performed using the output of a rank function
-%% synthesis tool when applied to counterexamples found by the safety prover.
-
-%The existing work on termination is primarily divided into structural (syntactic) vs. semantic approaches.  
-%For example, terminationg of term rewriting systems is largely structural, whereas DWF analysis (cf. Terminator) is more semantic. While it deals
-%with transition relations, it also uses some amount of syntax in that it identifies loops \& looping paths.
-
-%-- Usually the termination argument for the program LOOPS
-%on Figure 1 is based on a lexicographic combination
-%of well-founded orderings.
-
-While program termination has been extensively studied, much less research has been conducted in the area of proving nontermination.
-% A counterexample to termination is an infinite program execution. 
-Gupta et al. dynamically enumerate lasso-shaped candidate
-paths for counterexamples, and then statically prove their feasibility \cite{DBLP:conf/popl/GuptaHMRX08}. 
-In \cite{DBLP:conf/tacas/ChenCFNO14}, Chen et al. prove nontermination via
-reduction to safety proving. Their iterative algorithm uses counterexamples to a fixed safety property
-to refine an underapproximation of a program.
-
-% program analysis as constraint solving:
-%% how the constraint-based approach can be used to model a wide spectrum of program analyses in an 
-%%expressive domain containing disjunctions and conjunctions of linear inequalities.
 
@@
 state it starts in), we can find a ranking function for $L$.
+
+\begin{theorem}
+\label{thm:ut}
+ The loop $L(G, T)$ terminates from every start state iff Definition~\ref{def:UT} is
+ satisfiable.
+\end{theorem}
+
+\begin{proof}
+ Then the codomain of $R$ is a well founded set and $R$ is an order homomorphism.
+
+ (Do this proof properly).
+\end{proof}
+
 The existence of a ranking function is equivalent to the satisfiability
@@
 For our case, given a loop $L = \langle G,T \rangle$, the notion of open recurrent set is encoded by formula {\bf [ONT]}. %Definition~\ref{def:nonterm-formula}.
+
+\begin{theorem}
+\label{thm:ont}
+ The loop $L(G, T)$ has an infinite execution iff Defintion~\ref{def:ont} is satisfiable.
+\end{theorem}
+
+\begin{proof}
+ Then $x_0$ is an underapproximation to the reachable states on entry to $L$, and $N$ is an open recurrence set.
+
+ (Do this proof properly).
+\end{proof}
 
@@
 %infinite execution to happen.
+
+\begin{theorem}
+\label{thm:cnt}
+ If $T$ is deterministic, i.e. each state $x$ has exactly one successor $x'$ such that $T(x, x')$, then
+ Definition~\ref{def:ont} is equivalent to Definition~\ref{def:cnt}.
+\end{theorem}
+
+\begin{proof}
+ If $T$ is deterministic, then each $x$ has a unique successor $S(x)$, so $\forall x, x' . T(x, x') \leftrightarrow x' = S(x)$.
+
+ (Finish this proof).
+\end{proof}
 
@@
 Conditional termination is then equivalent to the
-formula {\bf [CT]} of Defintion~\ref{def:ct}. If this formula is satisfiable,
-two witnesses are returned:
+formula {\bf [CT]} of Defintion~\ref{def:ct}.
+
+\begin{theorem}
+\label{thm:ct}
+ The loop $L(G, T)$ terminates when started in any state satisfying $I(x)$ iff
+ Defintion~\ref{def:ct} is satisfiable.
+\end{theorem}
+
+\begin{proof}
+ Do this proof.
+\end{proof}
+
+If this formula is satisfiable, two witnesses are returned:
 \begin{itemize}
@@
 \begin{proof}
- {\sc Headshot}'s output constitutes a constructive, machine checked proof of (non-)termination.
+ This is a direct consequence of Theorems~\ref{thm:ut}, \ref{thm:ct}, \ref{thm:ont}, \ref{thm:cnt} and \ref{thm:synth-sound}.
 \end{proof}
@@
 
+ \section{Conclusions and Related Work}
+%\subsection{Termination analysis}
+Automated program termination is a research topic that has received a fair amount of attention from the software verification community.
+In order to compare our technique to the rest of the area, 
+Figure~\ref{fig:handletable} summarises the related works with respect to the restrictions they impose on the transition relations as well as the form of the computed ranking functions.
+
+Somehow expected, most of the techniques are specialised in the synthesis of linear ranking functions for linear programs over integers (or rationals) \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13}.
+Among them, 
+Lee et al. make use of transition predicate abstraction, algorithmic learning, and decision procedures to compute transition
+invariants as proofs for the termination of linear programs \cite{DBLP:conf/cav/LeeWY12}.
+Leike and Heizmann present a new method for the constraint-based synthesis
+of termination arguments for linear loop programs based on
+linear ranking templates \cite{DBLP:conf/tacas/LeikeH14}.
+Linear ranking functions supported by inductive linear invariants for loops with linear guards and transitions: \cite{DBLP:conf/cav/BradleyMS05}. 
+Learning: \cite{DBLP:journals/corr/HeizmannHP14}
+
+%A program is lasso-shaped if it is composed by a stem followed by a single loop without branching, i.e. there is only one path. 
+%Consequently, the termination argument can be very simple. There are numerous techniques specialised in
+%proving termination of lasso-shaped programs efficiently \cite{DBLP:conf/popl/Ben-AmramG13,DBLP:conf/cav/BradleyMS05,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/P04}.
+
+While the synthesis of termination arguments for linear programs over integers is indeed well-covered in the literature, 
+there is very limited work for programs over machine integers.
+Cook et al. present a method based on a reduction to Presburger
+arithmetic, and a template-matching approach for predefined classes of
+ranking functions based on reduction to SAT- and QBF-solving \cite{DBLP:conf/tacas/CookKRW10}.
+Similarly, the only work we are aware of that can compute non-linear ranking functions  
+for imperative loops with polynomial guards and polynomial assignments
+is \cite{DBLP:conf/vmcai/BradleyMS05}. However, this work extends only to polynomials.
+
+Given the lack of research in handling \emph{not-linear programs}, as well as \emph{programs over bit-vectors and floats},  
+our work focuses on covering these areas. 
+One of the obvious conclusions that can be reached from observing Figure~\ref{fig:handletable}, 
+is that most of the works tend to specialise on a certain aspect of termination proving that they can solve efficiently. 
+Conversely to this view, we aim for \emph{generality}, as we do not restrict the form of the synthesised ranking functions, nor the form of the input programs.
+
+As mentioned in Section~\ref{sec:intro}, approaches based on Ramsey's theorem compute a set of local termination conditions that decrease as execution follows through the loop
+and require expensive reachability analyses
+\cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}.
+In an attempt to reduce the complexity of checking the validity of the termination argument, 
+Cook et al present an iterative termination proving procedure that searches for 
+lexicographic termination arguments \cite{DBLP:conf/tacas/CookSZ13}, 
+whereas Kroening et al. strengthen the termination argument such that it becomes a transitive relation \cite{DBLP:conf/cav/KroeningSTW10}.
+
+Proving program termination implies the simultaneous search for a termination argument and a supporting invariant.
+For this purpose, several tools alternate 
+between calls to a safety prover and calls to a ranking function synthesis tool \cite{}. 
+In order to share more information about the state of the termination proofs between the two tools, 
+Brockschmidt et al. use a single representation of the state of the termination proof search that both tools operate over, 
+resulting in performance improvements \cite{DBLP:conf/cav/BrockschmidtCF13}.
+In \cite{DBLP:conf/cav/BradleyMS05}, Bradley et al. combine the generation of ranking functions with the generation of invariants to
+form one constraint solving problem such that the necessary supporting invariants for the ranking function are discovered on demand.
+In our setting, both the ranking function and the supporting invariant are iteratively constructed through the same refinement loop.
+
+%In our setting, a candidate termination argument is iteratively constructed. 
+%A candidate termination argument is composed out of a ranking function 
+%and a supporting invariant.
+
+%% The safety prover proves or disproves the
+%% validity of the current argument via the search for invariants. Refinement of the
+%% current termination argument is performed using the output of a rank function
+%% synthesis tool when applied to counterexamples found by the safety prover.
+
+%The existing work on termination is primarily divided into structural (syntactic) vs. semantic approaches.  
+%For example, terminationg of term rewriting systems is largely structural, whereas DWF analysis (cf. Terminator) is more semantic. While it deals
+%with transition relations, it also uses some amount of syntax in that it identifies loops \& looping paths.
+
+%-- Usually the termination argument for the program LOOPS
+%on Figure 1 is based on a lexicographic combination
+%of well-founded orderings.
+
+While program termination has been extensively studied, much less research has been conducted in the area of proving nontermination.
+% A counterexample to termination is an infinite program execution. 
+Gupta et al. dynamically enumerate lasso-shaped candidate
+paths for counterexamples, and then statically prove their feasibility \cite{DBLP:conf/popl/GuptaHMRX08}. 
+In \cite{DBLP:conf/tacas/ChenCFNO14}, Chen et al. prove nontermination via
+reduction to safety proving. Their iterative algorithm uses counterexamples to a fixed safety property
+to refine an underapproximation of a program.
+
+% program analysis as constraint solving:
+%% how the constraint-based approach can be used to model a wide spectrum of program analyses in an 
+%%expressive domain containing disjunctions and conjunctions of linear inequalities.
+
 \bibliographystyle{abbrvnat}

### 2014-07-04 src/kalashnikov/genetic/synth.c
Start genetic synth with "likely" constants

@@ -83,6 +83,23 @@ void check_population(solution_t *pop) {
   }
 }
 
+int rand_const() {
+  //return rand() & WORDMASK;
+
+  int r = rand() % 4;
+
+  switch (r) {
+  case 0:
+    return 0;
+  case 1:
+    return 1;
+  case 2:
+    return WORDMASK;
+  case 3:
+    return 1 << (WIDTH - 1);
+  }
+}
+
 
 void rand_solution(solution_t *solution) {
   int i, j;
@@ -101,7 +118,7 @@ void rand_solution(solution_t *solution) {
     }
 
     for (i = 0; i < CONSTS; i++) {
-      prog->consts[i] = rand() & WORDMASK;
+      prog->consts[i] = rand_const();
     }
   }
 

### 2014-07-04 src/kalashnikov/frontend.py
Frontend flags

@@ -28,8 +28,8 @@ def prove_terminates(filename):
   cmd = (("./kalashnikov.py " +
              "%s %s " +
              "-P%d  " +
-             "--seed=1337 " +
-             "--synth-strategy=genetic -a%d --varnames %s " +
+             "--seed=1337 -a%d --varnames %s " +
+             "--synth-strategy=genetic " +
              "-c1 " +
              "--fastverif=false -keepfrac=15 -mutprob=0.25 -newfrac=2 -popsize=500 " +
              "-recombprob=0.05 -tourneysize=10 -w=3 " +

### 2014-07-04 papers/termination/paper.tex
isolated loop (non-)termination

@@
 \section{Motivational Examples} \label{sec:motivation}
+- see phase change programs.
+
 Figure~\ref{fig:handletable} illustrates the most common assumptions made by existent termination analyses:
@@
 
-In the remainder of this section, we show that the synthesis fragment 
+In the rest of this section, we show that the synthesis fragment 
 is expressive enough to encode both termination and non-termination. %allow the formulation of both the termination and the non-termination problem. 
@@
 \begin{framed}
-\begin{definition}[Second-order Unconditional Termination Formula {\bf [UT]}]
+\begin{definition}[Unconditional Termination Formula {\bf [UT]}]
 \label{def:UT}
 \begin{align*}
- \exists R . \forall x, x' . & G(x) \wedge T(x, x') \rightarrow R(x) > 0 \wedge R(x) > R(x')
+ \exists R . \forall x, x' . & G(x) \wedge T_o(x, x') \rightarrow R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
@@
   \exists N, x_0 . \forall x . \exists x' . & N(x_0) ~ \wedge ~ N(x) \rightarrow G(x) ~ \wedge \\
-							& N(x) \rightarrow T(x, x') \wedge N(x') 
+							& N(x) \rightarrow T_u(x, x') \wedge N(x') 
  \end{align*}
@@
   \exists N, x_0 . \forall x, x' . & N(x_0) ~ \wedge ~ N(x) \rightarrow G(x) ~ \wedge \\
-							& N(x) \wedge T(x, x') \rightarrow N(x') 
+							& N(x) \wedge T_u(x, x') \rightarrow N(x') 
  \end{align*}
 \end{definition}
+
+%% \begin{definition}[Non-deterministic Non-Termination Formula - Closed Recurrence Set {\bf [Nondet-CNT]}]
+%% \label{def:nondet-cnt}
+%%  \begin{align*}
+%%   \exists N, x_0 . \forall x, x' . \exists y_0. & N(x_0) ~ \wedge ~ N(x) \rightarrow G(x) ~ \wedge \\
+%% 							& N(x) \wedge T_u(x, x') \wedge A(x') \rightarrow N(x')~
+%%  \end{align*}
+%% \end{definition}
 
@@
 To show that $L$ is unconditionally terminating (i.e. it eventually terminates regardless of the
-state it starts in), we can find a ranking function for $L$.
+state it starts in), we try to find a ranking function for it. When defining $L$'s termination 
+formula, we need only consider the over-approximation of its transition relation $T$, which we call $T_o$.
+Note that this distinction is only important if $L$'s body contains nested loops, for which we'll have to 
+consider the over-approximation of their transition relation's transitive closure
+as shown in Section~\ref{sec:env}. For any other statements, $T_o$ and $T$ are equivalent.
 
@@
 \label{thm:ut}
- The loop $L(G, T)$ terminates from every start state iff Definition~\ref{def:UT} is
+ The loop $L(G, T_o)$ terminates from every start state iff formula {\bf [UT]} (Definition~\ref{def:UT}) is
  satisfiable.
@@
 
-The existence of a ranking function is equivalent to the satisfiability
-of the formula from Definition~\ref{def:UT}, so a witness to the satisfiability
-of formula {\bf [UT]} is also a ranking function and thus a proof of $L$'s unconditional termination.
+As the existence of a ranking function is equivalent to the satisfiability
+of the formula {\bf [UT]}, a satisfiality witness 
+is also a ranking function and thus a proof of $L$'s unconditional termination.
 
@@
 So $R(x) = x$ constitutes a proof that the program in Figure~\ref{fig:motivation.a} is unconditionally
-terminating.\\
+terminating.
+
+Note that different formulations for unconditional termination are possible. We are aware of a proof rule based on 
+transition invariants, i.e. supersets of the transition relation's transitive closure \cite{DBLP:conf/pldi/GrebenshchikovLPR12}.
+This formulation assumes that the second order logic has a primitive predicate for disjunctive well-foundedness. 
+Conversely, our formulation in Definition~\ref{def:UT} does not require a transitive closure operator, nor a disjunctive well-foundedness predicate. \\
 
@@
 exists a transition to some $s'\in N$.
-For our case, given a loop $L = \langle G,T \rangle$, the notion of open recurrent set is encoded by formula {\bf [ONT]}. %Definition~\ref{def:nonterm-formula}.
+For our case, the notion of open recurrent set is encoded by formula {\bf [ONT]}. %Definition~\ref{def:nonterm-formula}.
+Dual to the termination problem, when formulating $L$'s non-termination, we only need to consider an under-approximation of its transition relation, $T_u$. 
+%$L = \langle G,T \rangle$
+Again, this distinction is mostly important when dealing with nested loops, as illustrated in Section~\ref{sec:env}.
 
@@
 \label{thm:ont}
- The loop $L(G, T)$ has an infinite execution iff Defintion~\ref{def:ont} is satisfiable.
+ The loop $L(G, T_u)$ has an infinite execution iff Defintion~\ref{def:ont} is satisfiable.
 \end{theorem}
@@
 \label{thm:cnt}
- If $T$ is deterministic, i.e. each state $x$ has exactly one successor $x'$ such that $T(x, x')$, then
+ If $T_u$ is deterministic, i.e. each state $x$ for which the guard $G(x)$ holds has exactly one successor $x'$ such that $T_u(x, x')$, then
  Definition~\ref{def:ont} is equivalent to Definition~\ref{def:cnt}.
@@
 \begin{proof}
- If $T$ is deterministic, then each $x$ has a unique successor $S(x)$, so $\forall x, x' . T(x, x') \leftrightarrow x' = S(x)$.
+ If $T_u$ is deterministic, then each $x$ satisfying $G(x)$ has a unique successor $S(x)$, so $\forall x, x' . G(x) \wedge T_u(x, x') \leftrightarrow x' = S(x)$.
 
@@
 \end{proof}
+
+Note that by defining $T_u$'s determinism as the existence of exactly one successor for any state satisfying $L$'s guard, 
+we eliminate the possibility of $T_u$ being $false$, which would otherwise constitute a counter-example for Theorem~\ref{thm:cnt}.
 
@@
 ours does not have.\\
+
+\noindent {\bf Non-determinism.} A non-termination diagnosis procedure based on checking the satisfiability of formula {\bf [CNT]}
+may fail to find a closed recurrence set for a loop $L=(G, T_u)$ when $T_u$ is non-deterministic.  
+To understand the issue, imagine the following loop with an non-deterministic assignment:
+
+\begin{lstlisting}[language=C]
+while(x != 0) {
+  y = nondet();
+  x = x-y;
+}
+\end{lstlisting}
+
+Only for certain values of \texttt{y} the loop is non-terminating.
+In order to restrict the values that may be assigned to \texttt{y}, we need an assumption associated with the non-deterministic choice. 
+However, such an assumption may cause a reachable state at the non-deterministic assignment to not have any successor. In such a case
+the termination formula would be satisfiable, although the execution halts. This situation can be avoided by checking the existence of 
+at least one successor for each reachable state at the non-deterministic assignment. Although we did put a reasonable amount of effort into it, 
+we failed to come up with a formulation for the non-termination of such non-deterministic transition relations 
+inside the synthesis fragment.
+%formula {\bf [Nondet-CNT]} in Definition~\ref{def:nondet-cnt}.
 
@@
 
-\subsection{Composing a Loop with its Environment}
+\subsection{Composing a Loop with its Environment} \label{sec:env}
+
+synthesizing environment abstractions: underapp for termination and underapp for non-termination.
+
+\cite{} addresses the problem of automatically underapproximating weakest preconditions.
 

### 2014-07-04 papers/termination/paper.tex
formalisation stuff

@@
 
-\subsection{Composing a Loop with its Environment} \label{sec:env}
-
-synthesizing environment abstractions: underapp for termination and underapp for non-termination.
-
-\cite{} addresses the problem of automatically underapproximating weakest preconditions.
-
 \begin{figure*}
@@
   \exists R, W . \forall x, x' . & I(x) \wedge G(x) \rightarrow W(x) ~ \wedge \\
-                                 & G(x) \wedge W(x) \wedge T(x, x') \rightarrow W(x') \wedge R(x) > 0 
+                                 & G(x) \wedge W(x) \wedge T_o(x, x') \rightarrow W(x') \wedge R(x) > 0 
   \wedge R(x) > R(x')
@@
 
+\subsection{Composing a Loop with its Environment} \label{sec:env}
+
 Sometimes the termination behaviour of a loop depends on its environment.  That is to say,
@@
 \noindent {\bf Conditional termination.}
-If $L$'s termination depends on the state it begins
+Given a loop $L=(G,T_o)$, where $T_o$ is an over-approximation of its transition relation, 
+if $L$'s termination depends on the state it begins
 executing in, we say that $L$ is \emph{conditionally terminating}.
@@
 \label{thm:ct}
- The loop $L(G, T)$ terminates when started in any state satisfying $I(x)$ iff
+ The loop $L(G, T_o)$ terminates when started in any state satisfying $I(x)$ iff
  Defintion~\ref{def:ct} is satisfiable.
@@
 
-\noindent {\bf Nested loops.}
+\noindent {\bf Environment abstractions.}
+For a loop $L=(G,T)$, the formulae {\bf [CT]} and {\bf [CNT]} defined in Definition\ref{def:ct} and Definition~\ref{def:cnt} are in terms of 
+$T_o$ and $T_u$, which over-approximate and under-approximate the transition relation $T$, respectively. Moreover, in formula {\bf [CT]},  
+$I$ over-approximates the set of states that $L$ may begin executing in.
+The reason for using over- and under-approximations [instead of exact relations] when formulating the (non-)termination of an isolated loop 
+is so that we can use the same formulae regardless of the environment the loop is executing in, or 
+the complexity of its body, i.e. if it contains nested loops.
+
+All we need to do in order to use {\bf [CT]} and {\bf [CNT]} in any environment and in the presence of nested loops is generate constraints abstracting 
+the environment and the loop's body: for termination, we need over-approximating abstractions, whereas for non-termination 
+we need under-approximations.
+
+As the generation of such constraints is standard and covered by several other works \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}...
+
+- when analysing a loop, we assume that the nested ones and the one preceding it in program order terminate.
+
+%synthesizing environment abstractions: underapp for termination and underapp for non-termination.
+% \cite{} addresses the problem of automatically underapproximating weakest preconditions.
+
+%\noindent {\bf Nested loops.}
 If $L$'s body contains another loop $L'$, we cannot directly apply either formula {\bf [UT]} or
@@
 rather than presenting the algorithm for generating the constraints, we direct the
-reader's attention to Figure~\ref{fig:environment-model}.  This figure contains a non-trivial program
+reader's attention to Figure~\ref{fig:environment-model}.
+
+Figure~\ref{fig:environment-model} contains a non-trivial program
 along with the constraint system

### 2014-07-05 papers/termination/paper.tex
formalisation

@@
 
-Sometimes the termination behaviour of a loop depends on its environment.  That is to say,
+Sometimes the termination behaviour of a loop depends on its environment, i.e. the rest of the program.  That is to say,
 the loop may not terminate if started in some particular state, but the rest of the program
@@
 terminates, but if the loop were considered in isolation we would not be able to prove that
-it terminates.  We must therefore encode a loop's interaction with the rest of
-the program in order to do a sound termination analysis.\\
+it terminates. We must therefore encode a loop's interaction with its environment %the rest of the program 
+in order to do a sound termination analysis.\\
 
@@
 \noindent {\bf Environment abstractions.}
-For a loop $L=(G,T)$, the formulae {\bf [CT]} and {\bf [CNT]} defined in Definition\ref{def:ct} and Definition~\ref{def:cnt} are in terms of 
+For a loop $L=(G,T)$, the formulae {\bf [CT]} and {\bf [CNT]} from Definition~\ref{def:ct} and Definition~\ref{def:cnt} are in terms of 
 $T_o$ and $T_u$, which over-approximate and under-approximate the transition relation $T$, respectively. Moreover, in formula {\bf [CT]},  
@@
 is so that we can use the same formulae regardless of the environment the loop is executing in, or 
-the complexity of its body, i.e. if it contains nested loops.
-
-All we need to do in order to use {\bf [CT]} and {\bf [CNT]} in any environment and in the presence of nested loops is generate constraints abstracting 
+the complexity of its body, e.g. if it contains nested loops.
+
+All we need to do in order to use {\bf [CT]} and {\bf [CNT]} in any context %environment and regardless of the complexity of the loops's body 
+is generate constraints abstracting 
 the environment and the loop's body: for termination, we need over-approximating abstractions, whereas for non-termination 
 we need under-approximations. 
-
-As the generation of such constraints is standard and covered by several other works \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}...
-
-- when analysing a loop, we assume that the nested ones and the one preceding it in program order terminate.
+As the generation of such constraints is standard and covered by several other works \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}, 
+we will not provide the full algorithm, but rather illustrate it through examples.
+
+- when analysing a loop, we assume that the nested ones and the one preceding it in program order terminate.\\
 
@@
 %\noindent {\bf Nested loops.}
-If $L$'s body contains another loop $L'$, we cannot directly apply either formula {\bf [UT]} or
-{\bf [CT]} or prove that $L$ terminates.  This is because we don't have direct access to the
-single step transition relation for $L$'s body.  Instead, we consider $L'$ to be part of
-$L$'s environment, which we model with a relation $S$.  We require that $S$ overapproximates
-$L'$ in the sense that it satisfies the following formula:
-\begin{align*}
- \forall x, x'. G(x) \wedge W(x) \wedge T'^*(x, x') \wedge \lnot G'(x') \rightarrow S(x, x')
-\end{align*}
-Where $G$ and $G'$ are respectively $L$ and $L'$'s guards, and $T'^*$ is the reflexive transitive closure of
-the inner loop ($L'$)'s transition relation.
-
-To give an example, if we wish to prove that the outer loop in Figure~\ref{fig:nested} terminates,
-it suffices to summarise the inner loop with the relation:
-\begin{align*}
-S(\langle x, y, z \rangle, \langle x', y', z' \rangle) & \triangleq x' = x \wedge y = 1
-\end{align*}
-Then the ranking function $R(x, y, z) = x$ proves termination of the outer loop $L(G, S)$.\\
-
-\begin{figure}
-\begin{lstlisting}
-while (x > 0) {
-  z = x, y = z + 1;
-  while (z > 0) {
-    z--, y--;
-  }
-  x -= y;
-}
-\end{lstlisting}
-\caption{A nested loop\label{fig:nested}}
-\end{figure}
-
-\noindent {\bf Composing the environment with a loop.}
-While proving that loop $L$ terminates, we create constraints on the environment in the
-form of the predicate $I$ and the summary relation $S$.  For each loop $L'$ nested in the body
-of $L$, we must add some more constraints saying that $S$ overapproximates $L'$.
-Similarly, for each loop $L''$ appearing before $L$ in the program's control flow, we
-must add constraints saying that $I$ overapproximates the states reachable via $L''$.
-The details of how to create these constraints are standard, so
-rather than presenting the algorithm for generating the constraints, we direct the
-reader's attention to Figure~\ref{fig:environment-model}.
-
-Figure~\ref{fig:environment-model} contains a non-trivial program
+%% If $L$'s body contains another loop $L'$, we cannot directly apply either formula {\bf [UT]} or
+%% {\bf [CT]} or prove that $L$ terminates.  This is because we don't have direct access to the
+%% single step transition relation for $L$'s body.  Instead, we consider $L'$ to be part of
+%% $L$'s environment, which we model with a relation $S$.  We require that $S$ overapproximates
+%% $L'$ in the sense that it satisfies the following formula:
+%% \begin{align*}
+%%  \forall x, x'. G(x) \wedge W(x) \wedge T'^*(x, x') \wedge \lnot G'(x') \rightarrow S(x, x')
+%% \end{align*}
+%% Where $G$ and $G'$ are respectively $L$ and $L'$'s guards, and $T'^*$ is the reflexive transitive closure of
+%% the inner loop ($L'$)'s transition relation.
+
+%% To give an example, if we wish to prove that the outer loop in Figure~\ref{fig:nested} terminates,
+%% it suffices to summarise the inner loop with the relation:
+%% \begin{align*}
+%% S(\langle x, y, z \rangle, \langle x', y', z' \rangle) & \triangleq x' = x \wedge y = 1
+%% \end{align*}
+%% Then the ranking function $R(x, y, z) = x$ proves termination of the outer loop $L(G, S)$.\\
+
+%% \begin{figure}
+%% \begin{lstlisting}
+%% while (x > 0) {
+%%   z = x, y = z + 1;
+%%   while (z > 0) {
+%%     z--, y--;
+%%   }
+%%   x -= y;
+%% }
+%% \end{lstlisting}
+%% \caption{A nested loop\label{fig:nested}}
+%% \end{figure}
+
+\noindent {\bf Over-approximating the environment for termination.}
+%% While proving that loop $L$ terminates, we create constraints on the environment in the
+%% form of the predicate $I$ and the over-approximating summary relation $T_o$:
+%% In the presence of nested loops, 
+%% for each loop $L'$ nested in the body
+%% of $L$, we must generate constraints saying that $T_o$ overapproximates $L'$, i.e. it is an over-approximation of the transitive closure of $L'$'s transition relation.
+%% Similarly, for each loop $L''$ appearing before $L$ in the program's control flow, we
+%% must add constraints saying that $I$ over-approximates the states reachable via $L''$.
+%The details of how to create these constraints are standard, so
+%rather than presenting the algorithm for generating the constraints, we direct the
+%reader's attention to Figure~\ref{fig:environment-model}.  
+%
+The example in Figure~\ref{fig:environment-model} contains three loops, which we will number 1 to 3 according to their program order.
+Let us assume that we have already proven the termination of loops 1 and 3, and we are now trying to prove that loop 2 terminates. The 
+termination of loop 2 is encoded as the conjunction of constraints on the right of Figure~\ref{fig:environment-model}:
+the lines marked as $*$ over-approximate the states in which loop 2 can start execute.
+
+contains a non-trivial program
 along with the constraint system
@@
 system involves termination invariants $W_1, W_2, W_3$ and the inner loop summary $S$.
+
 Note that despite its intricacy, the constraint system is still a formula in the synthesis fragment.
@@
  \exists W_1, W_2, R, S . \forall v, x, y, z, v', x', y', z' . \\
-   x = 1 & \rightarrow W_1(v, x, y, z) \, \wedge \\
+   x = 0 & \rightarrow W_1(v, x, y, z) \, \wedge \\
    x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
@@
 
+\noindent {\bf Under-approximating the environment for non-termination.}
+Dual to termination, when attempting to prove non-termination, we must under-approximate the environment.
+
 \iffalse
@@
 $P_2$
+
 while ($G_2$) { $T_2$ }

### 2014-07-05 src/kalashnikov/frontend.py
Don't print cmd in frontend.py

@@ -45,8 +45,6 @@ def prove_terminates(filename):
                 nargs,
                 varnames,
                 ' '.join(sys.argv[2:])))
-  print cmd
-
   os.system(cmd)
 
 if __name__ == '__main__':

### 2014-07-05 papers/termination/paper.tex

@@
 Dual to termination, when attempting to prove non-termination, we must under-approximate the environment.
+Consider the example in Figure~\ref{fig:nonterm-environment}.  We are attempting to prove non-termination
+of the innermost loop, which we do by finding a recurrence set $N_4$.  We will assume that the other two
+loops iterate only a finite number of times.  If the constraint system is satisfiable, $x_0$ is some initial
+state in which the program begins executing.  Then $N_1$ is an inductive invariant of $L_1$ that is strong
+enough to establish $N_2$ if $L_1$ exits (which it does by assumption).  Now, $N_2$ is an inductive invariant
+of $L_2$ that is strong enough to establish either $N_3$ or $N_4$ on entry to $L_3$.  Recall that $N_4$ is
+a recurrence set for $L_3$, so if a state $x \in N_4$ is reached, the loop $L_3$ will execute forever.
+Otherwise, $N_3$ holds on entry to $L_3$.  $N_3$ is an inductive invariant of $L_3$ that is strong enough to
+re-establish $N_2$ when $L_3$ exits.  Since we assumed that $L_2$ iterated only a finite number of times,
+$N_3$ must be established only a finite number of times, which means that $N_4$ will eventually be reached,
+at which point $L_3$ will loop forever since $N_4$ is a recurrence set for $L_3$.
+
+\begin{figure*}
+\begin{framed}
+ \begin{minipage}{0.19\textwidth}
+\begin{lstlisting}
+L1: while (c1) {
+     B1;
+    }
+
+L2: while (c2) {
+     P;
+
+L3:  while (c3) {
+       B3;
+     }
+    }
+\end{lstlisting}
+\end{minipage}
+\vline
+\begin{minipage}{0.82\textwidth}
+\begin{align*}
+ \exists N_1, N_2, N_3, N_4, x_0 . \forall x, x' . \\
+  N_1(x_0) \, \wedge \\
+  c_1(x) \wedge N_1(x) \wedge B_1(x, x') & \rightarrow N_1(x') \, \wedge \\
+  \lnot c_1(x) \wedge N_1(x) & \rightarrow N_2(x) \, \wedge \\
+  N_2(x) & \rightarrow c_2(x) \, \wedge \\
+  c_2(x) \wedge N_2(x) \wedge P(x, x') & \rightarrow N_3(x') \vee N_4(x') \, \wedge \\
+  c_3(x) \wedge N_3(x) \wedge B_3(x, x') & \rightarrow N_3(x') \, \wedge \\
+  \lnot c_3(x) \wedge N_3(x) & \rightarrow N_2(x) \, \wedge \\
+  N_4(x) & \rightarrow c_3(x) \, \wedge \\
+  N_4(x) \wedge B_3(x, x') & \rightarrow N_4(x')
+\end{align*}
+\end{minipage}
+\end{framed}
+
+\caption{The constraint system for non-termination\label{fig:nonterm-environment}}
+\end{figure*}
 
@@
 \section{Experiments}
-We implemented our method by backing onto the {\sc Kalashnikov} program synthesiser.
+We implemented our algorithm in the {\sc Headshot} tool, which uses the {\sc Kalashnikov} program synthesiser for termination.
 We ran the tool on 40 benchmarks taken from the termination category of SVCOMP'15.

### 2014-07-05 papers/termination/kalashnikov.tex
Extend data flow diagram
DFD diagram

@@
 
-\begin{figure}
+\begin{figure*}
 \begin{center}
@@
 
+\def\Divide#1#2{%
+ \coordinate(a) at ($(#1.east) !.5! (#2.west)$);
+ \coordinate(b) at (a |- 0,-3);
+ \draw[dotted] (b) -- ++(0, 4.5);
+}
+
 \resizebox{\linewidth}{!}{
@@
 
-\node [file] (spec) {\sc Termination Specification};
+\node [file] (pua) {input.c};
+
+\path (pua.east)+(2,0) node [file] (spec) {\sc Termination Specification};
 \path (spec.south)+(0.0, -1) node [file] (inputs) {\sc Tracked Inputs};
@@
 
+\path [line] (pua) -- (spec);
+
 \path [line] (spec) -- (file);
@@
 
+\Divide{pua}{spec}{puaspec}
+%\Divide{spec}{file}{specfile}
+\Divide{file}{gcc}{filegcc}
+\Divide{gcc}{out}{gccout}
+
+\draw (pua |- 0,-3) node [align=center] {Input program \\ (written in C)};
+
+\coordinate (midspec) at ($(spec) !.5! (file)$);
+\draw (midspec |- 0,-3) node [align=center] {Automatically generated specification \\ (written in C)};
+%\draw (file |- 0,-3) node {Spec};
+\draw (gcc |- 0,-3) node [align=center] {Test generator};
+\draw (out |- 0,-3) node [align=center] {Proof \\ (written in $\mathcal{L}$)};
+
 \end{tikzpicture}
@@
 \label{fig:synth-dfd}
-\end{figure}
+\end{figure*}
 

### 2014-07-05 papers/termination/kalashnikov.tex
Synthesis algorithm

@@
 The algorithm is divided into two
-components: {\sc synth} (see Figure~\ref{fig:synth-dfd}) and {\sc verif}.% described next.
+procedures: {\sc synth} (see Figure~\ref{fig:synth-dfd}) and {\sc verif}, which interact via
+a finite set of test vectors {\sc inputs}, which is initialised to $\emptyset$.
 
-Given a (non-)termination specification (in the synthesis fragment)    
-and a finite set of test inputs tracked so far, {\sc synth} generates a program \texttt{synth.c} asserting that some non-deterministic proof taken as input  
-is incorrect on at least one of the tracked inputs.
-Thus, \texttt{synth.c} is unsafe iff there is some (non-)termination proof that is correct for all of the tracked inputs.  
-By using testing techniques, we search for such a proof, i.e. 
-we synthesise a candidate (non-)termination proof that is correct for all the tracked inputs.  
-The testing techniques we use are: model checking,
-exhaustive search and genetic programming (see Section~\ref{sec:gp}).  
-%For the latter 2 techniques, we combine SYNTH with another program
-%(GP or EXPLICIT respectively) to create a program that searches for test vectors (proofs) that cause SYNTH
-%to crash.
-As \texttt{synth.c}
-is a \newC program, we can check its safety by invoking a Bounded
-Model Checker, such as {\sc cbmc} \cite{}.
-We run the three available testing techniques %explicit-state model checker
-in parallel and take the answer of whichever happens
-to terminate first, stopping the other processes.
+The {\sc synth} procedure tries to find existential witnesses $(P, x_0)$ which satisfy the partial specification:
+\[
+ \exists P, x_0 . \forall x \in \text{\sc inputs} . \sigma(P, x_0, x)
+\]
 
-Subsequently, {\sc Verif} uses model checking or exhaustive search to verify the candidate proof by searching for an input on
-which it does not satisfy the specification. If no such input
-can be found, the candidate proof is correct.  Otherwise, the new input is added
-to the set of tracked inputs and the CEGIS loop is reiterated.
+If {\sc synth} succeeds in finding a witness $(P, x_0)$, this witness is a candidate solution to the full
+synthesis formula.  We pass this candidate solution to {\sc verif} which determines whether the candidate
+does satisfy the specification on all inputs by checking satisfiability of the verification formula:
+\[
+ \exists x . \lnot \sigma(P, x_0, x)
+\]
+
+If this formula is unsatisfiable, the candidate solution is in fact a solution to the synthesis formula
+and so the algorithm terminates.  Otherwise, the witness $x$ is an input on which the candidate solution fails
+to meet the specification.  This witness $x$ is added to the {\sc inputs} set and the loop iterates again.
+
+Concretely, {\sc synth} is implemented as shown in Figure~\ref{fig:synth-dfd}.  We start by taking the termination
+specification and generating a C program which takes as input a candidate $(P, x_0)$ and asserts that the candidate
+fails to meet the specification on at least on of the elements of {\sc inputs}.  So this program is unsafe iff there
+is some candidate $(P, x_0)$ satisfying the specification for all the elements of {\sc inputs}.  Finding a new
+candidate solution is therefore reduced to model checking this C program, which by construction is loop free.
+The model checking is done using a combination of bounded model checking, explicit state enumeration and
+genetic programming (GP).  The latter two techniques involve combining the previously generated C program,
+along with code that searches for candidate solutions.  As well as efficiency, this has the side effect of
+guaranteeing that we are verifying the exact semantics of the program under analysis as understood by a
+compiler, rather than some ad-hoc abstraction of its semantics.  Finally, if any of the model checkers
+finds a candidate solution it returns it.  The form of this candidate solution is yet another program,
+this time written in the proof language $\mathcal{L}$.
+
+Similarly, the {\sc verif} procedure is implemented by creating a C program that is unsafe iff
+some input exists on which the candidate solution fails to meet the specification.  Again, symbolic
+and explicit state model checking are used to decide the satisfiability of the verification formula.
 
@@
 %\draw (file |- 0,-3) node {Spec};
-\draw (gcc |- 0,-3) node [align=center] {Test generator};
+\draw (gcc |- 0,-3) node [align=center] {Model checker};
 \draw (out |- 0,-3) node [align=center] {Proof \\ (written in $\mathcal{L}$)};

### 2014-07-05 papers/termination/paper.tex
Extend data flow diagram
L language

@@
 \usepackage{tikz}
-\usetikzlibrary{arrows, automata, shapes}
+\usetikzlibrary{arrows, automata, shapes, calc}
 
@@
 
-Since termination for bitvector programs is known to be PSPACE-complete~\cite{DBLP:conf/tacas/CookKRW10} it
+Since termination for bitvector programs is known to be PSPACE-complete \cite{DBLP:conf/tacas/CookKRW10} it
 is perhaps unsurprising that we are able to decide it.  We will now show that our time complexity is

### 2014-07-05 papers/termination/paper.tex
formalisation

--- 2014-07-05 papers/termination/paper.tex
@@
 
+%\item We propose a non-trivial formulation of the (non-)termination problem inside a second order logic fragment with only one quantifier alternation.
+
 \end{itemize}
@@
 As the generation of such constraints is standard and covered by several other works \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}, 
-we will not provide the full algorithm, but rather illustrate it through examples.
-
-- when analysing a loop, we assume that the nested ones and the one preceding it in program order terminate.\\
+we will not provide the full algorithm, but rather illustrate it through examples.\\
 
@@
 %
-The example in Figure~\ref{fig:environment-model} contains three loops, which we will number 1 to 3 according to their program order.
-Let us assume that we have already proven the termination of loops 1 and 3, and we are now trying to prove that loop 2 terminates. The 
-termination of loop 2 is encoded as the conjunction of constraints on the right of Figure~\ref{fig:environment-model}:
-the lines marked as $*$ over-approximate the states in which loop 2 can start execute.
-
-contains a non-trivial program
-along with the constraint system
-encoding the termination of the outer loop.  Since the program contains sequential composition
-of loops and nested loops, the constraint
-system involves termination invariants $W_1, W_2, W_3$ and the inner loop summary $S$.
+The example in Figure~\ref{fig:environment-model} contains three loops. %, which we will number 1 to 3 according to their program order.
+Let us assume that we have already proven the termination of $L_1$ and $L_3$, and we are now trying to prove that $L_2$ terminates, which we encode
+as the conjunction of constraints on the right side of the figure:
+\begin{itemize}
+\item $W_1$, $W_2$ and $W_3$ are (over-approximating) inductive invariants for $L_1$, $L_2$ and $L_3$, respectively.
+\item $S$ is a summary of $L_3$ that over-approximates the transitive closure of its transition relation.
+\item If the formula is satisfiable, then $R$ is a ranking function for $L_2$.
+ \end{itemize}
+
+%the lines marked as $*$ over-approximate the states in which loop 2 can start execute.
+%% contains a non-trivial program
+%% along with the constraint system
+%% encoding the termination of the outer loop.  Since the program contains sequential composition
+%% of loops and nested loops, the constraint
+%% system involves termination invariants $W_1, W_2, W_3$ and the inner loop summary $S$.
 
@@
 W_2(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
-S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge z' + v' = x' \wedge v' \leq y' \\
+S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge \\
+& z' + v' = x' \wedge v' \leq y' \\
 R(x) & = x
@@
 \begin{framed}
-\begin{minipage}{0.19\textwidth}
-\begin{lstlisting}
+\begin{minipage}{0.22\textwidth}
+\begin{lstlisting}[mathescape=true]
 x = 0;
-while (x <= y)
+$L_1:$ while (x <= y)
   x++;
 
-while (x > 0) {
+$L_2:$ while (x > 0) {
   z = x;
@@
 
-  while (v < y) {
+$L_3:$  while (v < y) {
     z--;
@@
 \end{minipage}
+\end{framed}
 \caption{A non-trivial program and its constraint system\label{fig:environment-model}}
+\end{figure*}
+
+\noindent {\bf Under-approximating the environment for non-termination.}
+Dual to termination, when attempting to prove non-termination, we must under-approximate the environment and the loop body's and apply 
+formula {\bf [CNT]}. However, there is one more constraint that we must obey when doing so: we must stay inside the synthesis fragment! 
+This can be achieved for the environment as its under-approximation $I$ and the loop under consideration are at the same level of nestedness and do not therefore 
+require another quantifier alternation.
+
+However, when dealing with nested loops, computing the under-approximation of the loop's body 
+requires an extra quantifier alternation. The origin of the problem comes from the fact that an under-approximation is an existential property and therefore
+requires an extra existential quantifier, resulting in the $\exists \forall \exists$ alternation.
+This situation is similar to the one encountered when proving non-termination of non-deterministic loops.
+To go around this limitation, we treat nested loops as a special case and give a separate formula for them in Figure~\ref{fig:nonterm-nested}.
+If the formula on the right of the figure is satisfiable, then $L_1$ is non-terminating as witnessed by the recurrence set $N_1$ and
+the initial state $x_0$ in which the program begins executing. There are two possible scenarios for $L_2$'s termination:
+\begin{itemize}
+\item If $L_2$ is terminating, then $N_2$ is an inductive invariant that reestablished $N_1$ after $L_2$ stops executing: $\lnot G_2(x) \wedge N_2(x) \wedge P_2(x,x') \rightarrow N_1(x') $. 
+\item If $L_2$ is non-terminating, then $N_2$ is its recurrence set.
+\end{itemize}
+
+%% The issue appears when we try to under-approximate the body of the loop. 
+%% Here we encounter a similar problem with the one for non-determinism mentioned above.
+
+%% Consider the generic example in Figure~\ref{fig:nonterm-environment}, where we are attempting to prove non-termination
+%% of the innermost loop $L_3$ by finding a closed recurrence set $N_3$. 
+%% -- We will assume that the other two loops iterate only a finite number of times.
+
+%% We now explain the constraints on the right of the figure.
+%% Given that we prove non-termination, we try to  under-approximate the states from which $L_3$ can start executing. 
+%% We do this in several steps:
+
+%% $W_1$, $W_2$ and $W_3$ are inductive invariants corresponding to $L_1$, $L_2$ and $L_3$, respectively, such that:
+
+%% - $W_1$ is an inductive invariant of $L_1$ that is strong enough to establish $W_2$ if $L_1$ exits (which it does by assumption).
+%% - $W_2$ is an inductive invariant of $L_2$ that is strong enough to establish either $W_3$ or $N_3$ on entry to $L_3$, depending on whether $L_3$ terminates or not.
+%% - Recall that $N_3$ is a recurrence set for $L_3$, so if a state $x \in N_3$ is reached, the loop $L_3$ will execute forever. Otherwise, $W_3$ holds on entry to $L_3$.
+%% - $W_3$ is an inductive invariant of $L_3$ that is strong enough to re-establish $W_2$ when $L_3$ exits.
+
+%% If the non-termination formula is satisfiable, $x_0$ is some initial state in which the program begins executing.
+
+%% First, we under-approximate the states in which $L_2$ can start executing through $W_2$.
+%% Then $N_1$ is an inductive invariant of $L_1$ that is strong
+%% enough to establish $N_2$ if $L_1$ exits (which it does by assumption).  Now, $N_2$ is an inductive invariant
+%% of $L_2$ that is strong enough to establish either $N_3$ or $N_4$ on entry to $L_3$.  Recall that $N_4$ is
+%% a recurrence set for $L_3$, so if a state $x \in N_4$ is reached, the loop $L_3$ will execute forever.
+%% Otherwise, $N_3$ holds on entry to $L_3$.  $N_3$ is an inductive invariant of $L_3$ that is strong enough to
+%% re-establish $N_2$ when $L_3$ exits.
+
+%% If the constraint system is satisfiable, $x_0$ is some initial
+%% state in which the program begins executing.
+
+%%   Since we assumed that $L_2$ iterated only a finite number of times,
+%% $N_3$ must be established only a finite number of times, which means that $N_4$ will eventually be reached,
+%% at which point $L_3$ will loop forever since $N_4$ is a recurrence set for $L_3$.
+
+%% \begin{figure*}
+%% \begin{framed}
+%%  \begin{minipage}{0.19\textwidth}
+%% \begin{lstlisting}
+%% L1: while (c1) {
+%%      B1;
+%%     }
+
+%% L2: while (c2) {
+%%      P;
+
+%% L3:  while (c3) {
+%%        B3;
+%%      }
+%%     }
+%% \end{lstlisting}
+%% \end{minipage}
+%% \vline
+%% \begin{minipage}{0.82\textwidth}
+%% \begin{align*}
+%%  \exists W_1, W_2, W_3, N_3, x_0 . \forall x, x' . \\
+%%   W_1(x_0) \, \wedge \\
+%%   c_1(x) \wedge W_1(x) \wedge B_1(x, x') & \rightarrow W_1(x') \, \wedge \\
+%%   \lnot c_1(x) \wedge W_1(x) & \rightarrow W_2(x) \, \wedge \\
+%%   W_2(x) & \rightarrow c_2(x) \, \wedge \\
+%%   W_2(x) \wedge P(x, x') & \rightarrow W_3(x') \vee N_3(x') \, \wedge \\
+%%   c_3(x) \wedge W_3(x) \wedge B_3(x, x') & \rightarrow W_3(x') \, \wedge \\
+%%   \lnot c_3(x) \wedge W_3(x) & \rightarrow W_2(x) \, \wedge \\
+%%   N_3(x) & \rightarrow c_3(x) \, \wedge \\
+%%   N_3(x) \wedge B_3(x, x') & \rightarrow N_3(x')
+%% \end{align*}
+%% \end{minipage}
+%% \end{framed}
+
+%% \caption{The constraint system for non-termination\label{fig:nonterm-environment}}
+%% \end{figure*}
+
+\begin{figure*}
+\begin{framed}
+ \begin{minipage}{0.19\textwidth}
+\begin{lstlisting}[mathescape=true]
+$L_1$: while ($G_1$) {
+     $P_1$;
+
+$L_2$:  while ($G_2$) {
+       $B_2$;
+       }
+
+     $P_2$;
+     }
+\end{lstlisting}
+\end{minipage}
+\vline
+\begin{minipage}{0.82\textwidth}
+\begin{align*}
+ \exists N_1, N_2, x_0 . \forall x, x' . \\
+  N_1(x_0) & \, \wedge \\
+  N_1(x) & \rightarrow G_1(x) \, \wedge \\
+  N_1(x) \wedge P_1(x,x') & \rightarrow N_2(x') \, \wedge \\
+  G_2(x) \wedge N_2(x) \wedge B_2(x,x') & \rightarrow N_2(x') \, \wedge \\
+  \lnot G_2(x) \wedge N_2(x) \wedge P_2(x,x') & \rightarrow N_1(x') 
+\end{align*}
+\end{minipage}
 \end{framed}
+
+\caption{The non-termination formula for nested loops \label{fig:nonterm-nested}}
 \end{figure*}
-
-\noindent {\bf Under-approximating the environment for non-termination.}
-Dual to termination, when attempting to prove non-termination, we must under-approximate the environment.
 
@@
 
+- when analysing a loop, we assume that the nested ones and the one preceding it in program order terminate.\\
 %% \subsection{Program Synthesis as Second-Order Satisfaction}

### 2014-07-05 papers/termination/table.tex

@@
-loop1.c & 6 &  & true & true & false & false & 1 & 5 & term & 2.4s & 1\\
+loop1.c & 6 &  & true & true & false & false & 1 & 5 & term & 1.3s & 1\\
 
-loop2.c & 7 &  & true & true & false & false & 1 & 5 & term & 2.5s & 3\\
+loop2.c & 7 &  & true & true & false & false & 1 & 5 & term & 1.4s & 4\\
 
-loop3.c & 8 & true & true & true & true & false & 1 & 5 & term & 2.8s & 3\\
+loop3.c & 8 & true & true & true & true & false & 1 & 5 & term & 1.8s & 8\\
 
-loop4.c & 13 &  & true & true & false & false & 2 & 5 & term & 3.1s & 17\\
+loop4.c & 13 &  & true & true & false & false & 2 & 5 & term & 2.9s & 18\\
 
-loop5.c & 13 &  & true & true & false & false & 2 & 5 & term & 2.9s & 8\\
+loop5.c & 13 &  & true & true & false & false & 2 & 5 & term & 5.3s & 18\\
 
-loop6.c & 6 &  & false & true & false & false & 1 & 5 & term & 2.4s & 1\\
+loop6.c & 6 &  & false & true & false & false & 1 & 5 & term & 1.2s & 1\\
 
-loop7.c & 7 &  & true & unk & false & false & unk & 5 & term & 18.2s & 10\\
+loop7.c & 7 &  & true & unk & false & false & unk & 5 & term & 8.3s & 17\\
 
-loop8.c & 6 &  & true & true & true & true & 1 & 5 & term & 2.4s & 1\\
+loop8.c & 6 &  & true & true & true & true & 1 & 5 & term & 1.3s & 1\\
 
-loop9.c & 11 & true & true & false & false & false & 1 & 5 & term & 3.6s & 3\\
+loop9.c & 11 & true & true & false & false & false & 1 & 5 & term & 1.6s & 1\\
 
-loop10.c & 6 &  & true & true & false & false & 1 & 5 & term & 2.5s & 1\\
+loop10.c & 6 &  & true & true & false & false & 1 & 5 & term & 1.3s & 1\\
 
-loop43.c & 9 &  & true & false & false & false & ? & 5 & term & 3.7s & 4\\
+loop43.c & 9 &  & true & false & false & false & ? & 5 & term & 1.5s & 1\\
 
-loop44.c & 9 & false & true & false & false & false & ? & 5 & nonterm & 2.6s & 4\\
+loop44.c & 9 & false & true & false & false & false & ? & 5 & nonterm & 10.5s & 9\\
 
-loop45.c & 15 &  & true & false & false & false & ? & 5 & term & 3.4s & 19\\
+loop45.c & 15 &  & true & false & false & false & ? & 5 & term & 4.3s & 18\\
 
-loop46.c & 9 &  & true & false & false & false & ? & 5 & term & 2.7s & 4\\
+loop46.c & 9 &  & true & false & false & false & ? & 5 & term & 1.5s & 2\\
 
-loop47.c & 6 &  & true & false & false & false & ? & 5 & term & 3.6s & 4\\
+loop47.c & 6 &  & true & false & false & false & ? & 5 & term & 1.8s & 1\\
 
-loop48.c & 7 &  & true & false & false & false & ? & 5 & term & 2.5s & 3\\
+loop48.c & 7 &  & true & false & false & false & ? & 5 & term & 1.4s & 4\\
 
-loop49.c & 6 &  &  &  &  &  &  & 5 & ? & T/O & 12\\
+loop49.c & 6 &  &  &  &  &  &  & 5 & ? & T/O & 206\\
 
-svcomp1.c & 17 &  & true & true & false & false & 1 & 5 & term & 2.8s & 5\\
+svcomp1.c & 17 &  & true & true & false & false & 1 & 5 & term & 2.3s & 2\\
 
-svcomp2.c & 13 &  & false & true & false & false & 1 & 5 & term & 2.6s & 6\\
+svcomp2.c & 13 &  & false & true & false & false & 1 & 5 & term & 1.5s & 9\\
 
-svcomp3.c & 15 &  & true & true & false & false & unk & 5 & term & 36.7s & 36\\
+svcomp3.c & 15 &  & true & true & false & false & unk & 5 & term & 146.4s & 43\\
 
-svcomp4.c & 14 & false & true & unk & true & false & unk & 5 & nonterm & 50.5s & 75\\
+svcomp4.c & 14 & false & true & unk & true & false & unk & 5 & nonterm & 2.1s & 7\\
 
@@
 
-svcomp6.c & 11 &  & true & true & false & false & 1 & 5 & ? & T/O & 16\\
+svcomp6.c & 11 &  & true & true & false & false & 1 & 5 & term & 29.1s & 28\\
 
-svcomp7.c & 9 & true & true & true & true & false & 1 & 5 & term & 9.7s & 7\\
+svcomp7.c & 9 & true & true & true & true & false & 1 & 5 & term & 5.5s & 17\\
 
-svcomp8.c & 9 &  & true & unk & false & false & unk & 5 & ? & T/O & 38\\
+svcomp8.c & 9 &  & true & unk & false & false & unk & 5 & ? & T/O & 37\\
 
-svcomp9.c & 12 &  & true & unk & false & false & unk & 5 & term & 2.8s & 11\\
+svcomp9.c & 12 &  & true & unk & false & false & unk & 5 & term & 1.5s & 8\\
 
-svcomp10.c & 15 &  & true & true & false & false & 1 & 5 & term & 3.4s & 19\\
+svcomp10.c & 15 &  & true & true & false & false & 1 & 5 & term & 4.5s & 18\\
 
-svcomp11.c & 16 &  & true & true & false & false & 1 & 5 & term & 5.5s & 27\\
+svcomp11.c & 16 &  & true & true & false & false & 1 & 5 & term & 14.6s & 33\\
 
-svcomp12.c & 13 &  & true & true & false & false & 1 & 5 & term & 5.9s & 28\\
+svcomp12.c & 13 &  & true & true & false & false & 1 & 5 & term & 10.9s & 24\\
 
-svcomp13.c & 11 &  & true & true & false & false & 1 & 5 & ? & T/O & 30\\
+svcomp13.c & 11 &  & true & true & false & false & 1 & 5 & term & 35.1s & 62\\
 
-svcomp14.c & 18 &  & true & true & false & false & 1 & 5 & ? & T/O & 66\\
+svcomp14.c & 18 &  & true & true & false & false & 1 & 5 & term & 30.8s & 26\\
 
-svcomp15.c & 30 &  & true & true & false & false & 1 & 5 & ? & T/O & 28\\
+svcomp15.c & 30 &  & true & true & false & false & 1 & 5 & ? & T/O & 191\\
 
-svcomp16.c & 17 &  & true & true & false & false & 1 & 5 & term & 2.7s & 3\\
+svcomp16.c & 17 &  & true & true & false & false & 1 & 5 & term & 2.2s & 3\\
 
-svcomp17.c & 18 &  & true & true & false & false & 1 & 5 & ? & T/O & 45\\
+svcomp17.c & 18 &  & true & true & false & false & 1 & 5 & ? & T/O & 32\\
 
-svcomp18.c & 19 &  & true & true & false & false & 1 & 5 & ? & T/O & 27\\
+svcomp18.c & 19 &  & true & true & false & false & 1 & 5 & ? & T/O & 26\\
 
-svcomp25.c & 12 &  & true & true & false & false & 1 & 5 & ? & T/O & 15\\
+svcomp25.c & 12 &  & true & true & false & false & 1 & 5 & ? & T/O & 26\\
 
-svcomp26.c & 23 &  & true & true & false & false & 1 & 5 & term & 2.9s & 11\\
+svcomp26.c & 23 &  & true & true & false & false & 1 & 5 & term & 3.2s & 23\\
 
-svcomp27.c & 13 & false & true & true & false & false & 1 & 5 & ? & T/O & 76\\
+svcomp27.c & 13 & false & true & true & false & false & 1 & 5 & ? & T/O & 67\\
 
-svcomp28.c & 18 &  & true & true & false & false & 1 & 5 & ? & T/O & 82\\
+svcomp28.c & 18 &  & true & true & false & false & 1 & 5 & ? & T/O & 79\\
 
@@
 
-svcomp37.c & 22 &  & true & true & false & false & 1 & 5 & term & 2.8s & 14\\
+svcomp37.c & 22 &  & true & true & false & false & 1 & 5 & term & 2.1s & 10\\
 
-svcomp38.c & 17 & true & true & false & true & false & 1 & 5 & ? & T/O & 48\\
+svcomp38.c & 16 & true & true & true & false & false & 1 & 5 & ? & T/O & 1\\
 
@@
 
-svcomp40.c & 12 &  & true & true & false & false & 1 & 5 & ? & T/O & 20\\
+svcomp40.c & 12 &  & true & true & false & false & 1 & 5 & ? & T/O & 30\\
 
-svcomp41.c & 12 &  & true & true & false & false & 1 & 5 & ? & T/O & 26\\
+svcomp41.c & 12 &  & true & true & false & false & 1 & 5 & ? & T/O & 27\\
 
-svcomp42.c & 22 &  & true & true & false & false & 1 & 5 & ? & T/O & 84\\
+svcomp42.c & 22 &  & true & true & false & false & 1 & 5 & ? & T/O & 85\\

### 2014-07-05 papers/termination/paper.tex
Phrasing introduction

@@
 more recent techniques moving towards termination arguments based on Ramsey's theorem \cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}.
-The latter approach aims to find a set of local termination conditions that decrease as execution follows through the loop. %(semantic characterisation of loops).
+This proof style proves that a transition relation is disjunctively well founded by finding a set of simple
+functions, each of which proves well foundedness for part of the transition relation.
 The main justification for this paradigm shift lies in the simplicity of the local termination measures when compared to the global ones, e.g.
-there are cases in which proofs based on local measures involve applying linear functions while corresponding global
-measures involve nonlinear functions or lexicographic orders. 
-This explains why approaches based on Ramsey's theorem are restricted to searching for linear termination arguments.
-
-The downside of a Ramsey-based approach is the fact that a valid termination argument must hold for the \emph{transitive closure}
-of the program's transitions, rather than only for individual transitions. 
+there are cases in which proofs based on local measures involve linear functions while corresponding global
+measures require non-linear functions or lexicographic orders. 
+%This explains why approaches based on Ramsey's theorem are restricted to searching for linear termination arguments.
+
+One difficulty with a Ramsey-based approach is that validity of the termination argument can only be checked by considering the \emph{transitive closure}
+of the program, rather than a single step.
 As such, there is experimental evidence that most of the time is spent in reachability analysis \cite{DBLP:conf/pldi/CookPR06}, 

### 2014-07-05 papers/termination/paper.tex
Motivational examples, bibtex

@@
 As such, there is experimental evidence that most of the time is spent in reachability analysis \cite{DBLP:conf/pldi/CookPR06}, 
-requiring the support of powerful safety checkers.
-Basically, these approaches opt for simpler termination arguments in the detriment of complex validity checking.
+requiring the support of powerful safety checkers:
+there is a tradeoff between the complexity of the termination arguments and that of checking their validity.
 
@@
 we investigate its extreme: \emph{unrestricted} termination arguments. 
-This means that our ranking functions may involve non-linearity and lexicographic orders (we do not commit to any such form, i.e. we do not use templates).
+This means that our ranking functions may involve non-linearity and lexicographic orders (we do not commit to any particular syntactic form, and do not use templates).
 
@@
 linear programs and linear ranking functions, it also identifies another common assumption made by state of the art termination provers: 
-\emph{the equivalence of bit-vector/float semantics and integer/real semantics}. Thus, most of the techniques treat
-bit-vectors and floats as mathematical integers and reals, respectively  \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10}.
-
-By assuming bit-vector/float semantics to be equivalent to integer/real semantics, 
-these techniques ignore the wrap-around behaviour caused by under/over-flows as well as width conversions, 
-and ultimately diverge from the execution of a program on a computer.  
-In Section~\ref{sec:motivation}, we show that integers/reals and bit-vectors/floats exhibit
-incomparable behaviours with respect to termination, e.g.
-programs that terminate for integers may \emph{not} terminate for bit-vectors.
-Thus, abstracting bit-vectors/floats to integers/reals may give rise to 
+that bit-vector semantics and integer semantics give rise to the same termination behaviour. Thus, most of the techniques treat
+bit-vectors and floats as mathematical integers and reals respectively 
+\cite{DBLP:conf/pldi/CookPR06,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10}.
+
+By assuming bit-vector semantics to be identical to integer semantics,
+these techniques ignore the wrap-around behaviour caused by overflows,
+which can be unsound.
+In Section~\ref{sec:motivation}, we show that integers and bit-vectors exhibit
+incomparable behaviours with respect to termination, i.e.
+programs that terminate for integers may \emph{not} terminate for bit-vectors and
+vice-versa.
+Thus, abstracting bit-vectors to integers may give rise to 
 {\em unsound} and {\em incomplete} analyses.
@@
 uniformly computes
-\emph{lexicographic/non-lexicographic} \emph{linear/non-linear} 
-ranking functions supported by inductive \emph{linear/non-linear} invariants for loops with 
-\emph{linear/non-linear} guards and transitions over bit-vectors and floats.
+lexicographic, non-linear 
+ranking functions supported by non-linear inductive invariants for loops with 
+arbitrary guards and transitions over bit-vectors and floats.
+It also computes non-linear recurrence sets, which comprise non-termination proofs.
+Our encoding uniformly multiple loops and nested loops, without having to enumerate
+lassos.
 %Our technique is {\emph complete} and can handle programs with non-linear operations, e.g. logical and.
@@
 %checking the validity of a termination argument. 
-We combine the generation of ranking functions with the generation of invariants such that the necessary 
-supporting invariants and the ranking function are discovered at the same time.
-
-For this purpose, we phrase the termination and non-termination problems as second order satisfaction problem and
-we identify a fragment of second-order logic that is expressive enough to capture our formulation, 
-which we call \emph{the synthesis fragment}. Subsequently, we propose a method 
-for checking the satisfiability of a formula in the synthesis fragment as program synthesis (Section~\ref{sec:synthesis}). 
-Our method is theoretically complete, i.e. if a pair of a ranking function and a supporting invariant 
-exist, it will be found.
+%We combine the generation of ranking functions with the generation of invariants such that the necessary 
+%supporting invariants and the ranking function are discovered at the same time.
+
+For this purpose, we phrase the termination and non-termination problems as second order satisfaction problems.
+We identify a fragment of second-order logic that is expressive enough to capture our formulation,
+which we call the \emph{synthesis fragment}. Subsequently, we propose a method
+for checking the satisfiability of a formula in the synthesis fragment by solving an isomorphic program synthesis problem (Section~\ref{sec:synthesis}).
+Our method is sound and semi-complete, i.e. if a pair of a ranking function and a supporting invariant 
+exist it will be found, but if the program does not terminate we may not find a recurrence set.  However
+if the program's transition relation is deterministic, our method is complete.
 
@@
 but on the Kolmogorov complexity of its minimal termination argument. 
-\emph{In our design, proving a program terminates becomes as hard as finding its minimal temination argument}.
 We argue that the length of a program's termination proof is indicative of how easy to understand the program is, and thus 
@@
 
+Additionally, if the program under analysis has a non-deterministic transition relation
+and is non-terminating, our procedure is not guaranteed to find a non-termination proof.
+
 \section{Motivational Examples} \label{sec:motivation}
@@
 
-Figure~\ref{fig:handletable} illustrates the most common assumptions made by existent termination analyses:
+Figure~\ref{fig:handletable} illustrates the most common assumptions made by existing termination analyses:
 \begin{itemize}
-\item[(i)] programs are linear.
-\item[(ii)] terminating programs have linear termination arguments.
-\item[(iii)] bit-vectors semantics and mathematical integers semantics are equivalent.
-\item[(iv)] floats semantics and mathematical reals semantics are equivalent.
+\item[(i)] programs use only linear arithmetic.
+\item[(ii)] terminating programs have termination arguments in linear arithmetic.
+\item[(iii)] the semantics of bit-vectors and mathematical integers are equivalent.
+\item[(iv)] the semantics of floats and mathematical reals are equivalent.
 \end{itemize}
@@
 %In the previous sections, we have discussed some of the assumptions made by existent termination analyses (see Figure~\ref{fig:handletable}).
-Next, we show how these assumptions are broken by even simple programs that cannot therefore be handled accurately by state of the art termination tools.
+To show how these assumptions are broken by even simple programs, we draw the reader's attention to the programs in
+Figure~\ref{fig:motivation} and their curious properties:
 
@@
 %Figure~\ref{fig:handletable} shows that the most common assumption existent termination techniques make is that \emph{programs are linear}.
-The loop in Figure~\ref{fig:motivation.a} breaks assumption (i) as it makes use of the bit-wise $\&$ operator.
+
+\begin{itemize}
+
+\item Program (a) breaks assumption (i) as it makes use of the bit-wise $\&$ operator.
 %As such, Figure~\ref{} denotes a program with non-linear operations. Given that  many of the existing techniques commit to linear programs, they cannot handle this situation, 
@@
 %the semantics of the bit-wise AND operator.
-Our technique finds that a possible ranking function is the linear function
+Our technique finds that an admissible ranking function is the linear function
 $R(x) = x$, whose value decreases with
@@
 
-Assumption (ii) is broken by the loop in Figure~\ref{fig:motivation.b}.
-We prove that this loop terminates by finding that the non-linear function $R(x) = |x|$ is a possible ranking function.
-
-Figure~\ref{fig:motivation.c} invalidates assumption (iii). 
-This loop is  terminating for bit-vectors since \texttt{x}
+\item Program (b) break assumption (ii), in that it has no linear ranking function.
+We prove that this loop terminates by finding the non-linear ranking function $R(x) = |x|$.
+
+\item Program (c) breaks assumption (iii).
+This loop is  terminating for bit-vectors since $x$
 will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
-arithmetic as the loop condition stays always true for any initial \texttt{x} at least 1.
-
-Assumption (iii) is broken again by the program in Figure~\ref{fig:motivation.d}. Every execution of the program can be partitioned into two phases: 
-in the first phase $y$ increases until it is positive (in this phase $q$ may increase), whereas in the second $q$ decreases until the loop condition is violated. 
-This example is taken from \cite{DBLP:conf/tacas/LeikeH14}, where the authors make use of a template for obtaining a ranking function that proceeds
-through a fixed number of phases in the program execution. Each phase is ranked by a linear function, and ends when this function becomes non-positive.
-However, when considering the bit-vector semantics, this program does not terminate as..\todo{change this example}
-
-The loops in Figure~\ref{fig:motivation.e} and Figure~\ref{fig:motivation.f} break assumption (iv) as their terminating behaviour is different 
-for floats and mathematical reals. 
-%This assumption may lead to erroneous diagnosis of a program's terminating behaviour.
-The former does not terminate for floats (if $x$ is sufficiently large, the rounding makes the subtraction negligible), but does for reals. Conversely, the latter 
-doesn't terminate for reals, but does so for floats (due to rounding errors $x$ eventually becomes sufficiently small 
-such that the closest representable number is 0.0). \todo{can we prove non-termination?}
+arithmetic since $x > 0 \rightarrow x+1 > 0$ for any integer $x$.
+
+\item Program (d) also breaks assumption(iii), but ``the other way'' -- it terminates for integers but not for bit-vectors.
+If each of the variables is stored in an unsigned $k$-bit word, the following configuration will
+lead to an infinite loop:
+$$M = 2^k - 1, N = 2^k - 1, i = M, j = N-1$$
+
+\item Program (e) breaks assumption (iv) -- it terminates for reals but not for floats.
+If $x$ is sufficiently large, rounding error will cause the subtraction to have no effect.
+
+\item Program (f) breaks assumption (iv) ``the other way'' -- it terminates for floats but not for reals.
+Eventually $x$ will become sufficiently small that the nearest representable number is 0.0, at which point
+it will be rounded to 0.0 and the loop will terminate.
+\end{itemize}
 
@@
 
-Figure~\ref{fig:motivation.g} is a linear program taken from \cite{DBLP:conf/tacas/CookSZ13}, 
-where it is shown to not admit (without prior manipulation) a lexicographic linear function, but only a Ramsey-based termination argument,
-which requires an expensive binary reachability analysis.
-With our technique we can find a non-linear lexicographic ranking function $R(x) = |x|$ 
-(a lexicographic non-linear ranking functions consists of lexicographically ordered components
-of non-linear functions). 
+\begin{itemize}
+\item Program (g) is a linear program
+which is shown in~\cite{DBLP:conf/tacas/CookSZ13} not to admit (without prior manipulation) a lexicographic linear function.
+With our technique we can find the non-linear ranking function $R(x) = |x|$.
 %As with linear lexicographic ranking functions, a state is mapped to a tuple of values such that the
@@
 
-The loop in Figure~\ref{fig:motivation.i} is taken from SVCOMP'15  \footnote{http://sv-comp.sosy-lab.org/2015/index.php} termination benchmarks.
-In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, this program admits a multiphase ranking function computed from a multiphase ranking template.
+\item In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, program (i) admits a multiphase ranking function computed from a multiphase ranking template.
 A multiphase ranking template is targeted at programs that go through a
@@
 function becomes non-positive.
-However, in our setting this type of programs do not need a special treatment, as we can find a non-linear ranking function:
-\begin{verbatim}
-R(x, y, z) = (x < y, z)
-\end{verbatim}
-
-The loop in Figure~\ref{fig:motivation.h} illustrates conditional termination.
+
+In our setting this type of programs does not need special treatment, as we can find a non-linear lexicographic ranking function
+$R(x, y, z) = (x < y, z)$.  This termination argument is somewhat subtle, but can be understood as follows: the boolean quantity $x < y$
+eventually decreases, that is to say once a state with $x \geq y$ is reached, $x$ never again becomes greater than $y$.  This means
+that as soon as the ``else'' branch of the if statement is taken, it will continue to be taken in each subsequent iteration of the loop.
+Meanwhile, if $x < y$ has not decreased (i.e. we have stayed in the same branch of the ``if'') then $z$ does decrease.  Since
+a boolean only has two possible values, it cannot decrease indefinitely.  Since $z > 0$ is a conjunct of the loop guard,
+$z$ cannot decrease indefinitely, and so $R$ proves that the loop is well founded.
+
+As with all of the termination proofs presented in this paper, the ranking function above was found completely automatically.
+
+\item Program (h) illustrates conditional termination.
 When proving program termination we are simultaneously solving two problems:
-the search for a termination argument, and the search for a supporting invariant.
-We iteratively construct a termination argument as a pair of a ranking function and supporting invariant through a single refinement loop.
+the search for a termination argument, and the search for a supporting invariant~\cite{DBLP:conf/cav/BrockschmidtCF13}.
 For this loop, we find the ranking function $R(x) = x$ together with the supporting invariant $y=1$.
+\end{itemize}
 
@@
 \begin{lstlisting}
-while (x > 0)
+while (x > 0) {
   x = (x - 1) & x;
+}
 \end{lstlisting}
-\caption{}
+\caption{Taken from~\cite{DBLP:conf/tacas/CookKRW10}.}
  \label{fig:motivation.a}
@@
 \begin{lstlisting}
-while (x != 0)
+while (x != 0) {
   x = -x / 2;
+}
 \end{lstlisting}
@@
 \begin{lstlisting}[language=C]
- while(x > 0) x++;
+while(x > 0) {
+  x++;
+}
  \end{lstlisting}
@@
 \begin{lstlisting}
-  while ( q >= 0 ) {
-    q = q - y;
-    y = y + 1;
+while (i < M || j < N) {
+  i = i + 1;
+  j = j + 1;
 }
 \end{lstlisting}
-\caption{}
+\caption{Taken from~\cite{DBLP:conf/sigsoft/Nori013}}
  \label{fig:motivation.d}
@@
 \begin{lstlisting}
-while (x > 0.0) x -= 1.0;
+float x;
+
+while (x > 0.0) {
+  x -= 1.0;
+}
 \end{lstlisting}
@@
 \begin{lstlisting}
-while (x > 0.0) x *= 0.5;
+float x;
+
+while (x > 0.0) {
+  x *= 0.5;
+}
 \end{lstlisting}
@@
 \end{lstlisting}
-\caption{}
+\caption{Taken from \cite{DBLP:conf/tacas/CookSZ13}}
  \label{fig:motivation.g}
@@
 while (x > 0) {
-    x = x-y;
+  x = x - y;
 }
@@
 \end{lstlisting}
-\caption{}
+\caption{Taken from~\cite{BA:mcs}}
  \label{fig:motivation.i}
+\end{subfigure}
+
+\\
+
+\hline
+
+&
+
+\begin{subfigure}[b]{0.3\textwidth}
+\begin{lstlisting}
+while (i < n) {
+  j = 0;
+  while (j <= i) {
+    j = j + 1;
+  }
+  i = i + 1;
+}
+\end{lstlisting}
+\caption{Taken from~\cite{DBLP:conf/cav/BrockschmidtCF13}}
 \end{subfigure}
@@
  \label{thm:synth-sound}
- Algorithm~\ref{alg:abstract-refinement-code} is sound -- if it terminates with witness $P$ then
+ Algorithm~\ref{alg:cegis} is sound -- if it terminates with witness $P$ then
  $P \models \sigma$.
@@
  If the existential first-order theory used to express the specification $\sigma$ is decidable,
- algorithm~\ref{alg:abstract-refinement-code} is semi-complete -- if a program $P \models \sigma$
- exists then algorithm~\ref{alg:abstract-refinement-code} will terminate.  However if no program
+ algorithm~\ref{alg:cegis} is semi-complete -- if a program $P \models \sigma$
+ exists then algorithm~\ref{alg:cegis} will terminate.  However if no program
  satisfies the specification, the algorithm may not terminate.  Furthermore if the domain of inputs $X$
@@
  decidable, the query in {\sc Verif} will succeed for this program, causing the algorithm to terminate.
- The set of correct programs is therefore recursively enumerable and algorithm~\ref{alg:abstract-refinement-code}
+ The set of correct programs is therefore recursively enumerable and algorithm~\ref{alg:cegis}
  enumerates this set, so it is semi-complete.

### 2014-07-05 papers/termination/kalashnikov.tex
Motivational examples, bibtex
Shrink CEGIS figure

@@
  \begin{algorithmic}[1]
+\label{alg:cegis}
  \Statex
@@
 
-\begin{figure*}
- \centering
- \begin{tikzpicture}[scale=0.5,->,>=stealth',shorten >=1pt,auto,
+\begin{figure}
+ \begin{center}
+ \resizebox{\linewidth}{!}{%
+  \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,
  semithick, initial text=]
 
-  \matrix[nodes={draw, fill=none, scale=1, shape=rectangle, minimum height=1cm, minimum width=1.5cm},
-          row sep=2cm, column sep=3cm] {
+  \matrix[nodes={draw, fill=none, shape=rectangle, minimum height=1cm, minimum width=1.5cm},
+          row sep=2cm, column sep=2.5cm, ampersand replacement=\&] {
    \node (synth) {Synthesise};
-   &
+   \&
    \node (verif) {Verify}; %\\
    %\node[draw=none] {};
-   &
+   \&
    \node[ellipse] (done) {Done}; \\
@@
  \end{tikzpicture}
+ }
+ \end{center}
 
@@
  \label{fig:abstract-refinement}}
-\end{figure*}
+\end{figure}
 
@@
 
-\begin{tabular}{lll}
- \verb|add a b| & \verb|sub a b| & \verb|mul a b| \\
- \verb|neg a| & \verb|div a b|  & 
+\begin{tabular}{llll}
+ \verb|add a b| & \verb|sub a b| & \verb|mul a b| & \verb|div a b| \\
+ \verb|neg a| &   \verb|mod a b| & \verb|min a b| & \verb|max a b|
 \end{tabular}
@@
  \verb|le  a b| & \verb|lt  a b| & \verb|sle  a b| \\
- \verb|slt a b| & &
+ \verb|slt a b| & \verb|eq  a b| & \verb|neq  a b|
 \end{tabular}
+
+Miscellaneous logical instructions:
+
+\begin{tabular}{lll}
+ \verb|implies a b| & \verb|ite a b c| & 
+\end{tabular}
+
+Floating point arithmetic:
+
+\begin{tabular}{llll}
+ \verb|fadd a b| & \verb|fsub a b| & \verb|fmul a b| & \verb|fdiv a b| 
+\end{tabular}
+
 \end{center}

### 2014-07-05 papers/termination/paper.tex
Figures
Move kalashnikov.tex into paper.tex

@@
 
+\algrenewcommand\algorithmicrequire{\textbf{Precondition:}}
+\algrenewcommand\algorithmicensure{\textbf{Postcondition:}}
+
+\newcommand*\Let[2]{\State #1 $\gets$ #2}
+
+\newcommand{\bv}[2]{\mathcal{BV}(#1, #2)}
+
+\makeatletter
+\pgfdeclareshape{datastore}{
+  \inheritsavedanchors[from=rectangle]
+  \inheritanchorborder[from=rectangle]
+  \inheritanchor[from=rectangle]{center}
+  \inheritanchor[from=rectangle]{base}
+  \inheritanchor[from=rectangle]{north}
+  \inheritanchor[from=rectangle]{north east}
+  \inheritanchor[from=rectangle]{east}
+  \inheritanchor[from=rectangle]{south east}
+  \inheritanchor[from=rectangle]{south}
+  \inheritanchor[from=rectangle]{south west}
+  \inheritanchor[from=rectangle]{west}
+  \inheritanchor[from=rectangle]{north west}
+  \backgroundpath{
+    %  store lower right in xa/ya and upper right in xb/yb
+    \southwest \pgf@xa=\pgf@x \pgf@ya=\pgf@y
+    \northeast \pgf@xb=\pgf@x \pgf@yb=\pgf@y
+    \pgfpathmoveto{\pgfpoint{\pgf@xa}{\pgf@ya}}
+    \pgfpathlineto{\pgfpoint{\pgf@xb}{\pgf@ya}}
+    \pgfpathmoveto{\pgfpoint{\pgf@xa}{\pgf@yb}}
+    \pgfpathlineto{\pgfpoint{\pgf@xb}{\pgf@yb}}
+ }
+}
+\makeatother
+
 \begin{document}
@@
 
+\item Program (h) illustrates conditional termination.
+When proving program termination we are simultaneously solving two problems:
+the search for a termination argument, and the search for a supporting invariant~\cite{DBLP:conf/cav/BrockschmidtCF13}.
+For this loop, we find the ranking function $R(x) = x$ together with the supporting invariant $y=1$.
+
 \item In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, program (i) admits a multiphase ranking function computed from a multiphase ranking template.
@@
 
-\item Program (h) illustrates conditional termination.
-When proving program termination we are simultaneously solving two problems:
-the search for a termination argument, and the search for a supporting invariant~\cite{DBLP:conf/cav/BrockschmidtCF13}.
-For this loop, we find the ranking function $R(x) = x$ together with the supporting invariant $y=1$.
+\iffalse
+\item Program (j) illustrates a nested loop.  This construction can cause difficulties for provers based on
+enumerating lassos
+\fi
 \end{itemize}
@@
 \end{subfigure} 
-
-\\
-
-\hline
-
-&
-
-\begin{subfigure}[b]{0.3\textwidth}
-\begin{lstlisting}
-while (i < n) {
-  j = 0;
-  while (j <= i) {
-    j = j + 1;
-  }
-  i = i + 1;
-}
-\end{lstlisting}
-\caption{Taken from~\cite{DBLP:conf/cav/BrockschmidtCF13}}
-\end{subfigure}
+% 
+% \\
+% 
+% \hline
+% 
+% &
+% 
+% \begin{subfigure}[b]{0.3\textwidth}
+% \begin{lstlisting}
+% while (i < n) {
+%   j = 0;
+%   while (j <= i) {
+%     j = j + 1;
+%   }
+%   i = i + 1;
+% }
+% \end{lstlisting}
+% \caption{Taken from~\cite{DBLP:conf/cav/BrockschmidtCF13}}
+% \end{subfigure}
 
 \end{tabular}
-\caption{Sample programs}\label{fig:motivation}
+\caption{Motivational examples, mostly taken from the literature.\label{fig:motivation}}
 \end{figure*}
@@
 \begin{framed}
-\begin{minipage}{0.22\textwidth}
+\begin{minipage}{0.17\textwidth}
 \begin{lstlisting}[mathescape=true]
 x = 0;
-$L_1:$ while (x <= y)
+$L_1:$
+while(x <= y)
   x++;
 
-$L_2:$ while (x > 0) {
+$L_2:$
+while(x > 0){
   z = x;
@@
 
-$L_3:$  while (v < y) {
+$L_3:$
+  while(v < y){
     z--;
@@
 \vline
-\begin{minipage}{0.82\textwidth}
+\begin{minipage}{0.85\textwidth}
 \begin{align*}
@@
    x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
-   x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x+1, y, z) \, \wedge \\
+   x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x, y, z) \, \wedge \\
    %x > 0 \wedge W_2(v, x, y, z) & \rightarrow W_3(0, x, y, x) \, \wedge \\
@@
    %v \geq y \wedge W_3(v, x, y, z) & \rightarrow W_2(v, x-z, y-z, z) \, \wedge \\
-   S(\langle v, x, y, z, \rangle, \langle v', x', y', z' \rangle) \wedge v < y & \rightarrow S(\langle v, x, y, z \rangle, \langle v+1, x, y, z-1 \rangle) \, \wedge \\
+   S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' < y' & \rightarrow S(\langle v, x, y, z \rangle, \langle v'+1, x', y', z'-1 \rangle) \, \wedge \\
    x > 0 \wedge W_2(v, x, y, z) & \rightarrow S(\langle v, x, y, z \rangle, \langle 0, x, y, x \rangle) \, \wedge \\
@@
 %
-\input{kalashnikov}
+
+%\todo{add related work for synthesis.}
+
+In this section we present a solver for the synthesis fragment:
+ \[
+  \exists P,~ x . ~\forall~ y . \sigma(P, x, y)
+ \]
+%
+where $P$ ranges over sets, while $x$ and $y$ range over ground terms (as mentioned before, for brevity reasons we drop
+ the vector symbol and write  $x$ for $\vec{x}$).
+
+By viewing $\sigma: (X^n \times Y^m \to Z^k) \times X^n \times Y^m  \to \mathbb{B}$ as a specification function
+that  returns true iff $P$ computes appropriate outputs for the inputs $x$ and $y$, we can   
+regard the satisfaction problem for the synthesis fragment as \emph{program synthesis}, i.e.
+the mechanised construction of software that provably satisfies a given specification.
+Research in the area of program synthesis has been particularly fruitful beginning with Alonzo Church's work on the
+\emph{Circuit Synthesis Problem} in the sixties~\cite{church-synth}, and continuing with works such as 
+{\sc Brahma}~\cite{brahma} and program sketching~\cite{lezama-thesis,sketch,modular-sketch}.
+
+% 
+%We can solve the synthesis problem using any off-the-shelf program synthesiser.  If you don't
+%have a program synthesiser kicking around, we present the design of one synthesiser
+%that can be used to solve the problem.
+%We start by recalling the synthesis formula:
+% $$\exists P, x . \forall y . \sigma(P, x, y)$$
+
+%% Our program encoding generates an $\exists\forall$ formula that is \emph{linear} in the length of the shortest program
+%% satisfying the user's specification.  This formula is then checked for satisfiability
+%% using the CEGIS algorithm.
+
+In the context of (non-)termination analysis, a witnesses to the satisfiability of a formula in the synthesis fragment 
+is a (non-)termination proof. 
+As mentioned earlier in the paper, our analysis must ensure precise treatment of bit-wise operations and arithmetic modulo fixed widths
+%Consequently, a compulsory characteristic of our solver is that it must 
+by modelling bit accurate semantics.
+The synthesiser must therefore be able to compute (non-termination) proofs 
+%loop-free programs 
+over machine integers and floating-point numbers from the specifications detailed in Section~\ref{sec:second.order}.
+
+%\subsection{\newC}
+%\label{sec:logic}
+\subsection{The Synthesis Specification}
+The logic we use to express the synthesis formula, i.e. the (non-)termination specifications, % are given as program fragments
+%More specifically, the logic we use to express the synthesis formula
+is a subset of C that we call \newC.  
+The characteristic property of a \newC  program is that safety can be decided for
+it using a single query to a Bounded Model Checker.  A \newC program is
+just a C program with the following syntactic restrictions:
+ all loops in the program must have a constant bound;
+ all recursion in the program must be limited to a constant depth;
+ all arrays must be statically allocated (i.e. not using \texttt{malloc}),
+ and be of constant size.
+Additionally, \newC programs may use nondeterministic values, assumptions
+and arbitrary-width types.
+
+The (non-)termination proofs are arbitrary expressions over the grammar given in Figure~\ref{fig:l-language}.  
+%written in a simple RISC-like language $\mathcal{L}$, whose syntax is given in Fig.~\ref{fig:l-language}.  
+\todo{update the figure and make it look more like a grammar.}
+
+\begin{algorithm*}
+ \caption{Abstract refinement algorithm
+ \label{alg:cegis}}
+ \vspace{-1.5em}
+ \begin{multicols}{2}
+ \begin{algorithmic}[1]
+ \Statex
+\Function{synth}{inputs}
+  \Let{$(i_1, \ldots, i_N)$}{inputs}
+  \Let{query}{$\exists P . \sigma(i_1, P(i_1)) \land \ldots \land \sigma(i_N, P(i_N))$}
+  \Let{result}{decide(query)}
+  \If{result.satisfiable}
+    \State \Return{result.model}
+  \Else
+    \State \Return{unsatisfiable}
+  \EndIf
+\EndFunction
+\Statex
+\Function{verif}{P}
+  \Let{query}{$\exists x . \lnot \sigma(x, P(x))$}
+  \Let{result}{decide(query)}
+  \If{result.satisfiable}
+    \State \Return{result.model}
+  \Else
+    \State \Return{valid}
+  \EndIf
+\EndFunction
+\columnbreak
+\Statex
+\Function{refinement loop}{}
+  \Let{inputs}{$\emptyset$}
+  \Loop
+    \Let{candidate}{\Call{synth}{inputs}}
+    \If{candidate = UNSAT}
+      \State \Return{unsatisfiable}
+    \EndIf
+    \Let{res}{\Call{verif}{candidate}}
+    \If{res = valid}
+      \State \Return{candidate}
+    \Else
+      \Let{inputs}{inputs $\cup$ res}
+    \EndIf
+  \EndLoop
+\EndFunction
+ \end{algorithmic}
+ \end{multicols}
+\end{algorithm*}
+
+\begin{figure}
+ \begin{center}
+ \resizebox{\linewidth}{!}{%
+  \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,
+ semithick, initial text=]
+
+  \matrix[nodes={draw, fill=none, shape=rectangle, minimum height=1cm, minimum width=1.5cm},
+          row sep=2cm, column sep=2.5cm, ampersand replacement=\&] {
+   \node (synth) {Synthesise};
+   \&
+   \node (verif) {Verify}; %\\
+   %\node[draw=none] {};
+   \&
+   \node[ellipse] (done) {Done}; \\
+  };
+
+   \path
+    (synth) edge [bend left] node {Candidate program} (verif)
+    (verif) edge [bend left] node {Counterexample input} (synth)
+    (verif) edge node {Valid} (done);
+ \end{tikzpicture}
+ }
+ \end{center}
+
+ \caption{Abstract synthesis refinement loop
+ \label{fig:abstract-refinement}}
+\end{figure}
+
+\begin{figure}
+{\small
+\begin{center}
+\setlength{\tabcolsep}{16pt}
+Integer arithmetic instructions:
+
+\begin{tabular}{llll}
+ \verb|add a b| & \verb|sub a b| & \verb|mul a b| & \verb|div a b| \\
+ \verb|neg a| &   \verb|mod a b| & \verb|min a b| & \verb|max a b|
+\end{tabular}
+
+\medskip
+
+Bitwise logical and shift instructions:
+
+\begin{tabular}{lll}
+ \verb|and  a b| & \verb|or   a b| & \verb|xor a b| \\
+ \verb|lshr a b| & \verb|ashr a b| & \verb|not a|
+\end{tabular}
+
+\medskip
+
+Unsigned and signed comparison instructions:
+
+\begin{tabular}{lll}
+ \verb|le  a b| & \verb|lt  a b| & \verb|sle  a b| \\
+ \verb|slt a b| & \verb|eq  a b| & \verb|neq  a b|
+\end{tabular}
+
+Miscellaneous logical instructions:
+
+\begin{tabular}{lll}
+ \verb|implies a b| & \verb|ite a b c| & 
+\end{tabular}
+
+Floating point arithmetic:
+
+\begin{tabular}{llll}
+ \verb|fadd a b| & \verb|fsub a b| & \verb|fmul a b| & \verb|fdiv a b| 
+\end{tabular}
+
+\end{center}
+}
+ \caption{The language $\mathcal{L}$}
+ \label{fig:l-language}
+\end{figure}
+
+\subsection{The Synthesis Algorithm}
+As safety of a \newC program is efficiently 
+%the quantifier free first-order fragment of the logic is efficiently 
+decidable and a termination proof is expressible as a ground term of the \newC logic, we use
+Counterexample Guided Inductive Synthesis (CEGIS)~\cite{lezama-thesis,sketch} to
+check the validity of the synthesis formula. 
+The core of the CEGIS algorithm is the refinement loop shown in Fig.~\ref{fig:abstract-refinement} and
+detailed in Algorithm~\ref{alg:cegis}.  
+The algorithm is divided into two
+procedures: {\sc synth} (see Figure~\ref{fig:synth-dfd}) and {\sc verif}, which interact via
+a finite set of test vectors {\sc inputs}, which is initialised to $\emptyset$.
+
+The {\sc synth} procedure tries to find existential witnesses $(P, x_0)$ which satisfy the partial specification:
+\[
+ \exists P, x_0 . \forall x \in \text{\sc inputs} . \sigma(P, x_0, x)
+\]
+
+If {\sc synth} succeeds in finding a witness $(P, x_0)$, this witness is a candidate solution to the full
+synthesis formula.  We pass this candidate solution to {\sc verif} which determines whether the candidate
+does satisfy the specification on all inputs by checking satisfiability of the verification formula:
+\[
+ \exists x . \lnot \sigma(P, x_0, x)
+\]
+
+If this formula is unsatisfiable, the candidate solution is in fact a solution to the synthesis formula
+and so the algorithm terminates.  Otherwise, the witness $x$ is an input on which the candidate solution fails
+to meet the specification.  This witness $x$ is added to the {\sc inputs} set and the loop iterates again.
+
+Concretely, {\sc synth} is implemented as shown in Figure~\ref{fig:synth-dfd}.  We start by taking the termination
+specification and generating a C program which takes as input a candidate $(P, x_0)$ and asserts that the candidate
+fails to meet the specification on at least on of the elements of {\sc inputs}.  So this program is unsafe iff there
+is some candidate $(P, x_0)$ satisfying the specification for all the elements of {\sc inputs}.  Finding a new
+candidate solution is therefore reduced to model checking this C program, which by construction is loop free.
+The model checking is done using a combination of bounded model checking, explicit state enumeration and
+genetic programming (GP).  The latter two techniques involve combining the previously generated C program,
+along with code that searches for candidate solutions.  Finally, if any of the model checkers
+finds a candidate solution it returns it.  The form of this candidate solution is yet another program,
+this time written in the proof language $\mathcal{L}$.
+
+As well as efficiency, compiling the original source code of the program under analysis has the advantage of
+guaranteeing that we are verifying the exact semantics of the program as understood by a
+compiler, rather than some ad-hoc abstraction of its semantics.
+
+Similarly to {\sc synth}, the {\sc verif} procedure is implemented by creating a C program that is unsafe iff
+some input exists on which the candidate solution fails to meet the specification.  Again, symbolic
+and explicit state model checking are used to decide the satisfiability of the verification formula.
+
+%If the input set is finite, this procedure is guaranteed to terminate.
+
+%% In some cases, it is faster to use an explicit-state model checker rather than a Bounded Model Checker.  
+%% This is particularly true for the {\sc verif} component, 
+%% where we have observed that incorrect programs tend
+%% to be incorrect on a large fraction of the input space.  Counterexamples
+%% are then very easy to find by explicit enumeration of a few inputs.
+%% Since \newC is a fragment of C, we can generate an explicit-state
+%% model checker using the same source files that we pass to {\sc cbmc}
+%% and adding a small function to enumerate possible inputs.
+
+%% ---  We begin with the program under analysis (the PUA).  From this we produce a specification (in second order logic)
+%% encoding the termination of the PUA.  From the logical specification, we produce a program SYNTH.
+%% SYNTH takes as input a proof and asserts that the proof is incorrect on at least one of the inputs being tracked so far.
+%% Therefore, SYNTH is unsafe iff there is some proof that is correct for all of the tracked inputs.  Now we use testing
+%% techniques to try to find a proof that causes SYNTH to crash.  The testing techniques we use are: model checking,
+%% exhaustive search and genetic programming.  For the latter 2 techniques, we combine SYNTH with another program
+%% (GP or EXPLICIT respectively) to create a program that searches for test vectors (proofs) that cause SYNTH
+%% to crash.
+
+%% In the {\sc synth} component, in order to find a candidate proof $P$ satisfying
+%% $\sigma(i_1, P(i_1)) \land \ldots \land \sigma(i_N, P(i_N))$
+%% for the tracked inputs $(i_1, \ldots, i_N)$, 
+%% we attempt a safety proof.
+%In order to determine the validity of the {\sc synth} formula, we can
+%check the {\sc synth} program for safety.
+
+\begin{figure*}
+\begin{center}
+\tikzstyle{file}=[draw, text width=7.0em, text centered,
+  minimum height=1.5em]
+\tikzstyle{process} = [draw, minimum height=3em, circle]
+\tikzstyle{line} = [draw, color=black, -latex']
+
+\def\Divide#1#2{%
+ \coordinate(a) at ($(#1.east) !.5! (#2.west)$);
+ \coordinate(b) at (a |- 0,-3);
+ \draw[dotted] (b) -- ++(0, 4.5);
+}
+
+%\resizebox{\linewidth}{!}{
+\begin{tikzpicture}[font=\sffamily]
+
+\node [file] (pua) {input.c};
+
+\path (pua.east)+(2,0) node [file] (spec) {\sc Termination Specification};
+\path (spec.south)+(0.0, -1) node [file] (inputs) {\sc Tracked Inputs};
+%% \path (synth.south)+(0.0, -0.5) node [file] (tests) {\sc tests.c};
+%% \path (tests.south)+(0.0, -0.5) node [file] (interpreter) {\sc interpreter.c};
+%% \path (interpreter.south)+(0.0, -0.5) node [file] (spec) {\sc spec.c};
+
+%% \path (spec.east)+(2.0, -0.25) node [process] (merged) {merge};
+\path (spec.east)+(2.0, -0.75) node [file] (file) {synth.c};
+
+\path (file.east)+(2.0, 1.5) node [process] (cbmc) {\sc cbmc};
+\path (file.east)+(2.0, 0.0) node [process] (gcc) {\sc Search};
+\path (file.east)+(2.0, -1.5) node [process] (gp) {\sc GP};
+
+\path (gcc.east)+(2.5, 0.0) node [file] (out) {Candidate Proof};
+
+%\path [line] (spec) -- (merged);
+
+\path [line] (pua) -- (spec);
+
+\path [line] (spec) -- (file);
+\path [line] (inputs) -- (file);
+
+\path [line] (file) -- (cbmc);
+\path [line] (file) -- (gcc);
+\path [line] (file) -- (gp);
+
+\path [line] (cbmc) -- (out);
+\path [line] (gcc) -- (out);
+\path [line] (gp) -- (out);
+
+\Divide{pua}{spec}{puaspec}
+%\Divide{spec}{file}{specfile}
+\Divide{file}{gcc}{filegcc}
+\Divide{gcc}{out}{gccout}
+
+\draw (pua |- 0,-3) node [align=center] {Input program \\ (written in C)};
+
+\coordinate (midspec) at ($(spec) !.5! (file)$);
+\draw (midspec |- 0,-3) node [align=center] {Automatically generated specification \\ (written in C)};
+%\draw (file |- 0,-3) node {Spec};
+\draw (gcc |- 0,-3) node [align=center] {Model checker};
+\draw (out |- 0,-3) node [align=center] {Proof \\ (written in $\mathcal{L}$)};
+
+\end{tikzpicture}
+%}
+\end{center}
+
+\caption{Schematic diagram of {\sc synth}}
+\label{fig:synth-dfd}
+\end{figure*}
+
+\iffalse
+\subsection{Parameterising the Program Space}
+
+In order to search the space of candidate programs, we parametrise
+the language~$\mathcal{L}$ by program length, word width and number of constants,
+inducing a lattice of progressively
+more expressive languages.  We start by attempting to synthesise
+a program at the lowest point on this lattice and increase the
+parameters of~$\mathcal{L}$ until we reach a point at which
+the synthesis succeeds.
+
+As well as giving us an automatic search procedure, this parametrisation
+greatly increases the efficiency of our system since languages
+low down the lattice are very easy to decide safety for.  If a program
+can be synthesised in a low-complexity language, the whole procedure
+finishes much faster than if synthesis had been attempted in a
+high-complexity language.
+
+We introduce a novel parametrisation of the programming language
+used to express our synthesised programs.
+This parametrisation allows us to efficiently explore the program space
+without relying on human guidance and also ensures that our programs
+are of minimal length.
+
+Our tool is the first we are aware of that is able to effectively
+synthesise floating-point programs. We demonstrate this by
+synthesising {\sc Fast2Sum} using Knuth's {\sc 2Sum}~\cite{taocp2} as
+a specification.
+
+---Our exploration of the program space ensures that our programs are minimal in size.
+--- We parametrise the space of programs in such a way that it can be explored automatically, rather than asking a human for hints.
+
+We consider the following parameters:
+\begin{itemize}
+\item{Program Length: $l$}
+The first parameter we introduce is program length, denoted by $l$.
+At each iteration we synthesise programs of length exactly $l$.
+We start with $l = 1$ and increment $l$ whenever we determine
+that no program of length $l$ can satisfy the specification.  When we do
+successfully synthesise a program, we are \emph{guaranteed that it
+is of minimal length} since we have previously established that no
+shorter program is correct.
+\item{Word Width: $w$}
+An $\mathcal{L}$-program runs on a virtual machine (the $\mathcal{L}$-machine) that
+has its own set of parameters.  The only relevant parameter is
+the \emph{word width} of the $\mathcal{L}$-machine, that is, the number of bits
+in each internal register and immediate constant.  This parameter is denoted by
+$w$.  The size of the final SAT problem generated by {\sc cbmc} scales
+polynomially with $w$, since each intermediate C variable corresponds
+to $w$ propositional variables.
+
+\item{Number of Constants: $c$}
+Instructions in $\mathcal{L}$ take either one or two operands.
+Since any instruction whose operands are all constants can always be
+eliminated (since its result is a constant), we know that a loop-free program
+of minimal length will not contain any instructions with two constant
+operands.  Therefore the number of constants that can appear in
+a minimal program of length $l$ is at most $l$.  By minimising the number
+of constants appearing in a program, we are able to use a particularly
+efficient program encoding that speeds up the synthesis procedure
+substantially.  The number of constants used in a program is the parameter $c$.
+
+\end{itemize}
+\subsubsection{Searching the Program Space}
+
+The key to our automation approach is to come up with a sensible way in which to
+adjust the $\mathcal{L}$-parameters in order to cover all possible programs.
+After each round of {\sc synth}, we may need to adjust the parameters.  
+The
+logic for these adjustments is shown as a tree in Fig.~\ref{fig:paramsflow}.
+Whenever {\sc synth} fails, we consider which parameter might have caused the
+failure.  
+There are two possibilities: either the program length $l$ was too small,
+or the number of allowed constants $c$ was.  If $c < l$, we just increment $c$ and
+try another round of synthesis, but allowing ourselves an extra program constant.
+If $c = l$, there is no point in increasing $c$ any further.  This is because
+no minimal $\mathcal{L}$-program has $c > l$, for if it did there would
+have to be at least one instruction with two constant operands.  This
+instruction could be removed (at the expense of adding its result as
+a constant), contradicting the assumed minimality of the program.  So
+if $c = l$, we set $c$ to 0 and increment $l$, before attempting
+synthesis again.
+
+If {\sc synth} succeeds but {\sc verif} fails, we have a candidate
+program that is correct for some inputs but incorrect on at least
+one input.  However, it may be the case that the candidate program
+is correct for \emph{all} inputs when run on an $\mathcal{L}$-machine
+with a small word size.  For example, we may have synthesised a
+program which is correct for all 8-bit inputs, but incorrect for
+some 32-bit input.  If this is the case (which we can determine
+by running the candidate program through {\sc verif} using the smaller
+word size), we may be able to produce a correct program for
+the full $\mathcal{L}$-machine by using the constant extension rules
+shown in Fig.~\ref{fig:generalize}.  If constant generalization
+is able to find a correct program, we are done.  Otherwise,
+we need to increase the word width of the $\mathcal{L}$-machine
+we are currently synthesising for.
+
+\section{Conclusion}
+
+By expressing the program synthesis problem as a safety property for a
+program interpreter, we have been able to harness the power of
+state-of-the-art program analysis tools and reuse them in a new problem
+domain.  We have implemented our algorithm as a freely downloadable tool
+whose performance compares favourably to a recent program synthesiser. 
+Finally, we have taken advantage of the expressiveness of our specification
+language to make an initial step towards practical synthesis of
+floating-point programs.
+
+\fi
 

### 2014-07-05 papers/termination/paper.tex
"We enumerate programs startign with the shortest"

@@
 
-\subsection{Genetic Programming and Incremental Evolution} \label{sec:gp}
+\subsection{Candidate Generation Strategies}
+\noindent{\bf Explicit Proof Search} The simplest strategy for finding candidates
+is to just exhaustively enumerate them all, starting with the shortest and
+progressively increasing the number of instructions.  Since the set of
+$\mathcal{L}$-programs is recursively enumerable, this procedure is complete.
+\\
+
+\noindent {\bf Genetic Programming and Incremental Evolution} \label{sec:gp}
 To generate candidate (non-)termination proofs in the synthesise phase of the refinement loop,

### 2014-07-05 papers/termination/paper.tex
Incremental GP

@@
 
-In our setting, we start with a population of random programs of size ..
-We then genetically breed the population of these programs by applying the genetic operators 
-{\sc crossover} and {\sc mutate}. The crossover operator takes two programs
-and combines their code in some what to create a new program.  The mutate operator takes a single
+We begin with a population of random programs,
+then iteratively evolve this population by applying the genetic operators
+{\sc crossover} and {\sc mutate}. The {\sc crossover} operator takes two programs
+and combines their code in some way to create a new program.  The {\sc mutate} operator takes a single
 program and randomly changes parts of its code, again creating a new program.
 Both these operators are applied to programs selected from the population in
-proportion to their observed fitness, i.e. a program is fit if it satisfies the (non-)termination specification.
-Thus, our fitness function counts the number of test cases on which a program met the given specification.
-Essentially, we plan a trajectory through the space of computer programs that will lead to programs with improved fitness.
-%by considering those that meet the (non-)termination specification on more inputs until.
-
-\todo{Talk about the fitness landscape?}
-
-\begin{definition}
- A \emph{fitness landscape} is the space of all programs along with their fitness.
-\end{definition}
-
-A potential issue might be that, in our setting, a fitness landscape isn't really very smooth, e.g. small changes in program representation
-don't correspond to small changes in fitness. However, genetic programming's efficiency was shown to be unrelated to the smoothness 
-of the fitness landscape \cite{langdon:fogp}.
-
-\todo{mention incremental GP?}
+proportion to their observed fitness, and a program is said to be fitter than another
+if it meets the specification on more test cases.
+
+In the setting of program synthesis, genetic programming can be very usefully combined
+with the concept of incremental evolution.  The idea is that in order to evolve a program
+that is adequate for some difficult task, it is often helpful to first evolve a population
+that as an ensemble is good at some simpler variation of the task.  This principle
+leads us to the observation that each round of the CEGIS loop can be thought of
+as creating a more difficult problem for the GP synthesiser to solve.  Therefore,
+if GP finds a candidate program which is correct for the $k$ inputs so far observed,
+but is not correct for the full specification, that candidate program came from a population
+that was ``good'' for the specification restrict to those $k$ inputs.  The next time round
+the CEGIS loop, the problem GP must solve is slightly harder -- any solution to this new
+problem must be correct for the previous $k$ inputs and also whatever new input was added
+by the previous iteration of {\sc verif}.  It therefore makes sense to begin the $k+1$st
+round of evolution with the population at the end of the $k$th round.
 

### 2014-07-06 papers/termination/paper.tex
CBMC synth
formalisation

@@
 \section{Motivational Examples} \label{sec:motivation}
-- see phase change programs.
+%- see phase change programs.
 
@@
 \end{definition}
-
+%
 Checking satisfiability of a formula in the synthesis fragment corresponds to program synthesis and
@@
 
-\noindent {\bf Environment abstractions.}
-For a loop $L=(G,T)$, the formulae {\bf [CT]} and {\bf [CNT]} from Definition~\ref{def:ct} and Definition~\ref{def:cnt} are in terms of 
-$T_o$ and $T_u$, which over-approximate and under-approximate the transition relation $T$, respectively. Moreover, in formula {\bf [CT]},  
-$I$ over-approximates the set of states that $L$ may begin executing in.
-The reason for using over- and under-approximations [instead of exact relations] when formulating the (non-)termination of an isolated loop 
-is so that we can use the same formulae regardless of the environment the loop is executing in, or 
-the complexity of its body, e.g. if it contains nested loops.
-
-All we need to do in order to use {\bf [CT]} and {\bf [CNT]} in any context %environment and regardless of the complexity of the loops's body 
-is generate constraints abstracting 
-the environment and the loop's body: for termination, we need over-approximating abstractions, whereas for non-termination 
-we need under-approximations. 
-As the generation of such constraints is standard and covered by several other works \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}, 
-we will not provide the full algorithm, but rather illustrate it through examples.\\
-
 %synthesizing environment abstractions: underapp for termination and underapp for non-termination.
@@
 %
-The example in Figure~\ref{fig:environment-model} contains three loops. %, which we will number 1 to 3 according to their program order.
-Let us assume that we have already proven the termination of $L_1$ and $L_3$, and we are now trying to prove that $L_2$ terminates, which we encode
+%\noindent {\bf Environment abstractions.}
+For a loop $L=(G,T)$, the formulae {\bf [CT]} %and {\bf [CNT]} 
+from Definition~\ref{def:ct} 
+%and Definition~\ref{def:cnt} 
+is in terms of 
+$T_o$ and $I$, which over-approximate the transition relation $T$ and the set of states that $L$ may begin executing in, respectively. 
+The reason for using over-approximations [instead of exact relations] %when formulating the (non-)termination of an isolated loop 
+is so that we can use the same formulae regardless of the environment the loop is executing in, or 
+the complexity of its body, e.g. if it contains nested loops.
+
+All we need to do in order to use {\bf [CT]} in any context %environment and regardless of the complexity of the loops's body 
+is generate constraints over-approximating %abstracting 
+the environment and the loop's body. %: for termination, we need over-approximating abstractions, whereas for non-termination 
+%we need under-approximations. 
+As the generation of such constraints is standard and covered by several other works \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}, 
+we will not provide the full algorithm, but rather illustrate it through the example in Figure~\ref{fig:environment-model}.
+Let us assume that we have already proven the termination of loops $L_1$ and $L_3$, and we are now trying to prove that $L_2$ terminates, which we encode
 as the conjunction of constraints on the right side of the figure:
@@
 \noindent {\bf Under-approximating the environment for non-termination.}
-Dual to termination, when attempting to prove non-termination, we must under-approximate the environment and the loop body's and apply 
-formula {\bf [CNT]}. However, there is one more constraint that we must obey when doing so: we must stay inside the synthesis fragment! 
-This can be achieved for the environment as its under-approximation $I$ and the loop under consideration are at the same level of nestedness and do not therefore 
-require another quantifier alternation.
-
-However, when dealing with nested loops, computing the under-approximation of the loop's body 
-requires an extra quantifier alternation. The origin of the problem comes from the fact that an under-approximation is an existential property and therefore
-requires an extra existential quantifier, resulting in the $\exists \forall \exists$ alternation.
+Dual to termination, when proving non-termination, we need to under-approximate the environment and the loop's body and apply
+formula {\bf [CNT]}. %As opposed to over-approximations 
+%this is not as straightforward as when computing over-approximations in the termination case described above.
+Recall that there is one more constraint that we must obey when doing so: we must stay inside the synthesis fragment! 
+This becomes more challenging now then when dealing with over-approximations due to the existential nature of under-approximations: they require an additional existential quantifier, which might result in an additional quantifier alternation, i.e.
+$\exists \forall \exists$.
+
+This issue does not arise when abstracting the environment as its under-approximation and the loop under consideration make use of existential quantification at the same nestedness level.
+However, when dealing with nested loops, we need to compute the under-approximation of the inner loop, which requires a nested existential quantifier,
+resulting in the $\exists \forall \exists$ alternation. 
+%% computing the under-approximation of the loop's body requires an extra quantifier alternation. 
+%% The origin of the problem comes from the fact that an under-approximation is an existential property and therefore
+%% requires an extra existential quantifier, resulting in the $\exists \forall \exists$ alternation.
 This situation is similar to the one encountered when proving non-termination of non-deterministic loops.
@@
 \begin{lstlisting}[mathescape=true]
-$L_1$: while ($G_1$) {
+$L_1$:
+while ($G_1$) {
   $P_1$;
 
-$L_2$:  while ($G_2$) {
+$L_2$:
+  while ($G_2$) {
     $B_2$;
@@
 
+\noindent{\bf Symbolic Bounded Model Checking} Another complete method for generating
+candidates is to simply use BMC on the {\sc synth.c} program.  As with explicit
+search, we must progressively increase the length of the solution we search for
+in order to get a complete search procedure.
+\\
+
 \noindent {\bf Genetic Programming and Incremental Evolution} \label{sec:gp}

### 2014-07-06 papers/termination/paper.tex
just linebreaks

--- 2014-07-05 papers/termination/paper.tex
@@
 
-\conferenceinfo{POPL '14}{Month d--d, 20yy, City, ST, Country} 
-\copyrightyear{2014}
+\conferenceinfo{POPL '15}{Month d--d, 20yy, City, ST, Country} 
+\copyrightyear{2015}
 
@@
 \begin{abstract}
-Proving program termination is typically done by finding a well-founded \emph{ranking function}
-for the program states.
-Existing termination provers typically find ranking functions
-using either linear algebra or templates.  As such they are often restricted to
-finding linear ranking functions over mathematical integers.  This class
-of ranking functions is not large enough to prove termination for all terminating
-programs, and furthermore a termination argument for a program operating on mathematical integers
-does not always lead to a termination argument for the same program operating on
-fixed-width machine integers.
-
-We present a reduction from program \emph{termination} to program \emph{synthesis}.
-This reduction allows us to generate nonlinear, lexicographic ranking functions that
-are correct for fixed-width machine arithmetic and floating-point arithmetic.
-We use this reduction to build a sound and complete procedure for the termination
-of fixed-width and floating-point arithmetic programs.
+%
+Proving program termination is typically done by finding a well-founded
+\emph{ranking function} for the program states.  Existing termination
+provers typically find ranking functions using either linear algebra or
+templates.  As such they are often restricted to finding linear ranking
+functions over mathematical integers.  This class of ranking functions is
+not large enough to prove termination for all terminating programs, and
+furthermore a termination argument for a program operating on mathematical
+integers does not always lead to a termination argument for the same program
+operating on fixed-width machine integers.
+
+We present a reduction from program \emph{termination} to program
+\emph{synthesis}.  This reduction allows us to generate nonlinear,
+lexicographic ranking functions that are correct for fixed-width machine
+arithmetic and floating-point arithmetic.  We use this reduction to build a
+sound and complete procedure for the termination of fixed-width and
+floating-point arithmetic programs.
+%
 \end{abstract}
@@
 
-The halting problem has been of central interest to computer scientists since it was first
-considered by Turing in 1936~\cite{turing}.  Informally, we can say that the halting problem
-is concerned with answering the question ``does this
-program run forever, or will it eventually terminate?''
-
-Proving program termination is typically done by finding a \emph{ranking function}
-for the program states, i.e. a monotone map from the program's state space to a well-ordered set.
-Historically, the search for ranking functions has been constrained in various syntactic ways, leading to
-incompleteness, and
-is performed over abstractions that do not soundly capture the behaviour of physical computers.
-In this paper, we present a sound and complete method for deciding whether a program terminates.
+The halting problem has been of central interest to computer scientists
+since it was first considered by Turing in 1936~\cite{turing}.  Informally,
+we can say that the halting problem is concerned with answering the question
+``does this program run forever, or will it eventually terminate?''
+
+Proving program termination is typically done by finding a \emph{ranking
+function} for the program states, i.e.  a monotone map from the program's
+state space to a well-ordered set.  Historically, the search for ranking
+functions has been constrained in various syntactic ways, leading to
+incompleteness, and is performed over abstractions that do not soundly
+capture the behaviour of physical computers.  In this paper, we present a
+sound and complete method for deciding whether a program terminates.
 
@@
 
-One difficulty with a Ramsey-based approach is that validity of the termination argument can only be checked by considering the \emph{transitive closure}
-of the program, rather than a single step.
-As such, there is experimental evidence that most of the time is spent in reachability analysis \cite{DBLP:conf/pldi/CookPR06}, 
-requiring the support of powerful safety checkers:
-there is a tradeoff between the complexity of the termination arguments and that of checking their validity.
-
-As Ramsey-based approaches are limited by the state of the art in safety checking, 
-recent research shifts back to more complex termination arguments that are easier to check \cite{DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/KroeningSTW10}.
+One difficulty with a Ramsey-based approach is that validity of the
+termination argument can only be checked by considering the \emph{transitive
+closure} of the program, rather than a single step.  As such, there is
+experimental evidence that most of the time is spent in reachability
+analysis \cite{DBLP:conf/pldi/CookPR06}, requiring the support of powerful
+safety checkers: there is a tradeoff between the complexity of the
+termination arguments and that of checking their validity.
+
+As Ramsey-based approaches are limited by the state of the art in safety
+checking, recent research shifts back to more complex termination arguments
+that are easier to
+check~\cite{DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/KroeningSTW10}.
+%
 %In \cite{DBLP:conf/tacas/CookSZ13}, Cook et al present an iterative termination proving procedure that searches for 
@@
 Following the same trend, %of switching the complexity in termination checking back to the termination arguments, 
-we investigate its extreme: \emph{unrestricted} termination arguments. 
-This means that our ranking functions may involve non-linearity and lexicographic orders (we do not commit to any particular syntactic form, and do not use templates).
-
-While the summary  in Figure~\ref{fig:handletable} supports the observation that the majority of existing termination analyses are designed for
-linear programs and linear ranking functions, it also identifies another common assumption made by state of the art termination provers: 
-that bit-vector semantics and integer semantics give rise to the same termination behaviour. Thus, most of the techniques treat
-bit-vectors and floats as mathematical integers and reals respectively 
-\cite{DBLP:conf/pldi/CookPR06,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10}.
-
-By assuming bit-vector semantics to be identical to integer semantics,
-these techniques ignore the wrap-around behaviour caused by overflows,
-which can be unsound.
-In Section~\ref{sec:motivation}, we show that integers and bit-vectors exhibit
-incomparable behaviours with respect to termination, i.e.
-programs that terminate for integers may \emph{not} terminate for bit-vectors and
-vice-versa.
-Thus, abstracting bit-vectors to integers may give rise to 
-{\em unsound} and {\em incomplete} analyses.
+we investigate its extreme: \emph{unrestricted} termination arguments.  This
+means that our ranking functions may involve non-linearity and lexicographic
+orders (we do not commit to any particular syntactic form, and do not use
+templates).
+
+While the summary  in Figure~\ref{fig:handletable} supports the observation
+that the majority of existing termination analyses are designed for linear
+programs and linear ranking functions, it also identifies another common
+assumption made by state of the art termination provers: that bit-vector
+semantics and integer semantics give rise to the same termination behaviour. 
+Thus, most of the techniques treat bit-vectors and floats as mathematical
+integers and reals,
+respectively~\cite{DBLP:conf/pldi/CookPR06,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10}.
+
+By assuming bit-vector semantics to be identical to integer semantics, these
+techniques ignore the wrap-around behaviour caused by overflows, which can
+be unsound.  In Section~\ref{sec:motivation}, we show that integers and
+bit-vectors exhibit incomparable behaviours with respect to termination,
+i.e.  programs that terminate for integers may \emph{not} terminate for
+bit-vectors and vice-versa.  Thus, abstracting bit-vectors to integers may
+give rise to {\em unsound} and {\em incomplete} analyses.
+%
 %design decisions
@@
 lassos.
+%
 %Our technique is {\emph complete} and can handle programs with non-linear operations, e.g. logical and.
@@
 
-For this purpose, we phrase the termination and non-termination problems as second order satisfaction problems.
-We identify a fragment of second-order logic that is expressive enough to capture our formulation,
-which we call the \emph{synthesis fragment}. Subsequently, we propose a method
-for checking the satisfiability of a formula in the synthesis fragment by solving an isomorphic program synthesis problem (Section~\ref{sec:synthesis}).
-Our method is sound and semi-complete, i.e. if a pair of a ranking function and a supporting invariant 
-exist it will be found, but if the program does not terminate we may not find a recurrence set.  However
-if the program's transition relation is deterministic, our method is complete.
+For this purpose, we phrase the termination and non-termination problems as
+second order satisfaction problems.  We identify a fragment of second-order
+logic that is expressive enough to capture our formulation, which we call
+the \emph{synthesis fragment}.  Subsequently, we propose a method for
+checking the satisfiability of a formula in the synthesis fragment by
+solving an isomorphic program synthesis problem
+(Section~\ref{sec:synthesis}).  Our method is sound and semi-complete, i.e. 
+if a pair of a ranking function and a supporting invariant exist it will be
+found, but if the program does not terminate we may not find a recurrence
+set.  However if the program's transition relation is deterministic, our
+method is complete.
 
@@
 %find that our technique  
-This means that our procedure is not directly dependent on the structure of the analysed loop, e.g. number of variable, number of lassos, 
-but on the Kolmogorov complexity of its minimal termination argument. 
-We argue that the length of a program's termination proof is indicative of how easy to understand the program is, and thus 
-programmers tend to write programs that have relatively short termination proofs. 
-We empirically show that this claim holds in practice, causing our technique to perform well.
+%
+This means that our procedure is not directly dependent on the structure of
+the analysed loop, e.g.  number of variable, number of lassos, but on the
+Kolmogorov complexity of its minimal termination argument.  We argue that
+the length of a program's termination proof is indicative of how easy to
+understand the program is, and thus programmers tend to write programs that
+have relatively short termination proofs.  We empirically show that this
+claim holds in practice, causing our technique to perform well.
 
@@
 The main contributions of our work can be summarised as follows:
-
+%
 \begin{itemize}
-\item We designed a bit-level accurate technique for computing ranking functions that correctly accounts for the wrap-around behavior
-caused by under- and overflows in bit-vector and floating point arithmetic. Our technique is not restricted to finding linear ranking functions,
-but can also compute lexicographic non-linear  ones.
-
-\item  We rephrased the termination and non-termination problems as second-order satisfaction problems and used
-program synthesis techniques to solve them. In contrast to previous approaches to program synthesis, we found that
-using genetic programming was very effective at generating candidate termination and non-termination proofs.
-
-\item Our procedure does not require the enumeration of lassos, or directly reasoning about the transitive closure of the transition relation.
-To verify our proofs, we only require a single call to a SAT solver.
+
+\item We designed a bit-level accurate technique for computing ranking
+functions that correctly accounts for the wrap-around behavior caused by
+under- and overflows in bit-vector and floating point arithmetic.  Our
+technique is not restricted to finding linear ranking functions, but can
+also compute lexicographic non-linear ones.
+
+\item  We rephrased the termination and non-termination problems as
+second-order satisfaction problems and used program synthesis techniques to
+solve them.  In contrast to previous approaches to program synthesis, we
+found that using genetic programming was very effective at generating
+candidate termination and non-termination proofs.
+
+\item Our procedure does not require the enumeration of lassos, or directly
+reasoning about the transitive closure of the transition relation.  To
+verify our proofs, we only require a single call to a SAT solver.
 
@@
 
-\item Our procedure is based on a series of source-to-source transformations.  As well as leading to a concise implementation,
-this has the side effect of ensuring that our proof is correct for the exact binary produced by a particular compiler,
-rather than for an abstraction of the program.
-
-\item We implemented our technique and tried it on a selection of programs handling both bit-vectors and floats.
+\item Our procedure is based on a series of source-to-source
+transformations.  As well as leading to a concise implementation, this has
+the side effect of ensuring that our proof is correct for the exact binary
+produced by a particular compiler, rather than for an abstraction of the
+program.
+
+\item We implemented our technique and tried it on a selection of programs
+handling both bit-vectors and floats.
 
@@
 {\bf Limitations.}
-Our algorithm decides termination of transition systems with finite state spaces.
-The (non-)termination proofs take the form of ranking functions and program invariants
-that are expressed in a quantifier free language.  This formalism is powerful
-enough to handle a large fragment of C, but is not rich enough to analyse
-code using any of the following features:
+Our algorithm decides termination of transition systems with finite state
+spaces.  The (non-)termination proofs take the form of ranking functions and
+program invariants that are expressed in a quantifier free language.  This
+formalism is powerful enough to handle a large fragment of C, but is not
+rich enough to analyse code using any of the following features:
 
@@
 
-\item Program (h) illustrates conditional termination.
-When proving program termination we are simultaneously solving two problems:
-the search for a termination argument, and the search for a supporting invariant~\cite{DBLP:conf/cav/BrockschmidtCF13}.
-For this loop, we find the ranking function $R(x) = x$ together with the supporting invariant $y=1$.
-
-\item In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, program (i) admits a multiphase ranking function computed from a multiphase ranking template.
-A multiphase ranking template is targeted at programs that go through a
-finite number of phases in their execution. Each phase is ranked with
-an affine-linear function and the phase is considered to be completed once this
-function becomes non-positive.
-
-In our setting this type of programs does not need special treatment, as we can find a non-linear lexicographic ranking function
-$R(x, y, z) = (x < y, z)$.  This termination argument is somewhat subtle, but can be understood as follows: the boolean quantity $x < y$
-eventually decreases, that is to say once a state with $x \geq y$ is reached, $x$ never again becomes greater than $y$.  This means
-that as soon as the ``else'' branch of the if statement is taken, it will continue to be taken in each subsequent iteration of the loop.
-Meanwhile, if $x < y$ has not decreased (i.e. we have stayed in the same branch of the ``if'') then $z$ does decrease.  Since
-a boolean only has two possible values, it cannot decrease indefinitely.  Since $z > 0$ is a conjunct of the loop guard,
-$z$ cannot decrease indefinitely, and so $R$ proves that the loop is well founded.
-
-As with all of the termination proofs presented in this paper, the ranking function above was found completely automatically.
+\item Program (h) illustrates conditional termination.  When proving program
+termination we are simultaneously solving two problems: the search for a
+termination argument, and the search for a supporting
+invariant~\cite{DBLP:conf/cav/BrockschmidtCF13}.  For this loop, we find the
+ranking function $R(x) = x$ together with the supporting invariant $y=1$.
+
+\item In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, program (i)
+admits a multiphase ranking function computed from a multiphase ranking
+template.  A multiphase ranking template is targeted at programs that go
+through a finite number of phases in their execution.  Each phase is ranked
+with an affine-linear function and the phase is considered to be completed
+once this function becomes non-positive.
+
+In our setting this type of programs does not need special treatment, as we
+can find a non-linear lexicographic ranking function $R(x, y, z) = (x < y,
+z)$.  This termination argument is somewhat subtle, but can be understood as
+follows: the boolean quantity $x < y$ eventually decreases, that is to say
+once a state with $x \geq y$ is reached, $x$ never again becomes greater
+than $y$.  This means that as soon as the ``else'' branch of the if
+statement is taken, it will continue to be taken in each subsequent
+iteration of the loop.  Meanwhile, if $x < y$ has not decreased (i.e.  we
+have stayed in the same branch of the ``if'') then $z$ does decrease.  Since
+a boolean only has two possible values, it cannot decrease indefinitely. 
+Since $z > 0$ is a conjunct of the loop guard, $z$ cannot decrease
+indefinitely, and so $R$ proves that the loop is well founded.
+
+As with all of the termination proofs presented in this paper, the ranking
+function above was found completely automatically.
 
@@
 Given a program, %with state space $X$ and transition relation $T \subseteq X \times X$, 
-we first formalise its termination argument as a ranking function (Section~\ref{sec:ranking.functions}). 
-Subsequently, we discus bit-vector semantics and illustrate 
-differences between machine arithmetic and integer arithmetic that
-make the abstraction of bit-vectors to mathematical integers unsound (Section~\ref{sec:machine.arith}).
+we first formalise its termination argument as a ranking function
+(Section~\ref{sec:ranking.functions}).  Subsequently, we discus bit-vector
+semantics and illustrate differences between machine arithmetic and integer
+arithmetic that make the abstraction of bit-vectors to mathematical integers
+unsound (Section~\ref{sec:machine.arith}).
 
@@
 \subsection{Machine Arithmetic and Bitvectors} \label{sec:machine.arith}
-Physical computers have bounded storage, which means they are unable to perform calculations on mathematical
-integers.  For example, if $A$ is the Ackermann function and $G$ is Graham's number, a physical computer
-capable of computing $A(G, G)$ would contain (much!) more matter than is believed to exist in the universe.
-Fortunately, it is rare for a programmer to need such a large number and so modern computers do their arithmetic
-over fixed-width binary words, otherwise known as bit-vectors.  For the remainder of this section, we will say that
-the bit-vectors we are working with are $k$-bits wide, which means that each word can hold one of $2^k$ bit patterns.
-Typical values for $k$ are 32 and 64.
-
-Bit-vector arithmetic is performed modulo $2^k$, which is the source of many of the differences between
-machine arithmetic and Peano arithmetic.  To give an example: $(2^k - 1) + 1 \equiv 0 \pmod {2^k}$.
-This provides a counterexample to the statement $\forall x . x + 1 > x$, which is a theorem of Peano
-arithmetic but not of modular arithmetic.  When an arithmetic operation has a result greater than $2^k$,
-it is said to ``overflow''.  If an operation does not overflow, its machine-arithmetic result is the same
-as the result of the same operation performed on integers.
-
-Machine words can be interpreted as ``signed'' or ``unsigned'' values.  Signed values can be negative,
-while unsigned values cannot.  The encoding for signed values is two's complement, where the most significant
-bit $b_{k-1}$ of the word is a ``sign'' bit, whose weight is $-(2^k - 1)$ rather than $2^k - 1$.  Two's complement
-representation has the property that $\forall x . -x = (\mathord{\sim} x) + 1$, where $\mathord{\sim}(\bullet)$
-is bitwise-negation.  Two's complement also has the property that addition, multiplication and subtraction are defined
-identically for unsigned and signed numbers.  Therefore signed and unsigned arithmetic differ only on comparison
-operators\footnote{There is also the difference that signed overflow in a C program results in undefined behaviour,
-but in practice the undefined behaviour is implemented just as if the arithmetic had been unsigned}.
-For example: $$\forall x . -x \leq_s x \wedge -x \geq_u x$$ where $\leq_s$ and $\geq_u$ represent signed and unsigned comparisons respectively.
-
-The final source of difference between integer arithmetic and bitvector arithmetic stems from width conversions.
-Typical computers allow variables to have different types, which can be represented using words of different widths.
-In C a \texttt{short} might occupy
-16 bits, while an \texttt{int} might occupy 32 bits.  If a $k$-bit variable is assigned to a $l$-bit variable
-with $l < k$, the result is truncated $\mod 2^l$.  For example, if $x$ is a 32-bit variable and $y$ is a 16-bit
-variable, $y$ will hold the value $0$ after the following code is executed:
+
+Physical computers have bounded storage, which means they are unable to
+perform calculations on mathematical integers.  For example, if $A$ is the
+Ackermann function and $G$ is Graham's number, a physical computer capable
+of computing $A(G, G)$ would contain (much!) more matter than is believed to
+exist in the universe.  Fortunately, it is rare for a programmer to need
+such a large number and so modern computers do their arithmetic over
+fixed-width binary words, otherwise known as bit-vectors.  For the remainder
+of this section, we will say that the bit-vectors we are working with are
+$k$-bits wide, which means that each word can hold one of $2^k$ bit
+patterns.  Typical values for $k$ are 32 and 64.
+
+Bit-vector arithmetic is performed modulo $2^k$, which is the source of many
+of the differences between machine arithmetic and Peano arithmetic.  To give
+an example: $(2^k - 1) + 1 \equiv 0 \pmod {2^k}$.  This provides a
+counterexample to the statement $\forall x .  x + 1 > x$, which is a theorem
+of Peano arithmetic but not of modular arithmetic.  When an arithmetic
+operation has a result greater than $2^k$, it is said to ``overflow''.  If
+an operation does not overflow, its machine-arithmetic result is the same as
+the result of the same operation performed on integers.
+
+Machine words can be interpreted as ``signed'' or ``unsigned'' values. 
+Signed values can be negative, while unsigned values cannot.  The encoding
+for signed values is two's complement, where the most significant bit
+$b_{k-1}$ of the word is a ``sign'' bit, whose weight is $-(2^k - 1)$ rather
+than $2^k - 1$.  Two's complement representation has the property that
+$\forall x .  -x = (\mathord{\sim} x) + 1$, where $\mathord{\sim}(\bullet)$
+is bitwise-negation.  Two's complement also has the property that addition,
+multiplication and subtraction are defined identically for unsigned and
+signed numbers.  Therefore signed and unsigned arithmetic differ only on
+comparison operators\footnote{There is also the difference that signed
+overflow in a C program results in undefined behaviour, but in practice the
+undefined behaviour is implemented just as if the arithmetic had been
+unsigned}.  For example: $$\forall x .  -x \leq_s x \wedge -x \geq_u x$$
+where $\leq_s$ and $\geq_u$ represent signed and unsigned comparisons
+respectively.
+
+The final source of difference between integer arithmetic and bitvector
+arithmetic stems from width conversions.  Typical computers allow variables
+to have different types, which can be represented using words of different
+widths.  In C a \texttt{short} might occupy 16 bits, while an \texttt{int}
+might occupy 32 bits.  If a $k$-bit variable is assigned to a $l$-bit
+variable with $l < k$, the result is truncated $\mod 2^l$.  For example, if
+$x$ is a 32-bit variable and $y$ is a 16-bit variable, $y$ will hold the
+value $0$ after the following code is executed:
 
@@
 
-As well as machine arithmetic differing from Peano arithmetic on the operators they have in common,
-computers have several ``bitwise'' operations that are not taken as primitive in the theory of
-integers.  These operations include the standard boolean operations \texttt{and, or, not, xor}
-applied to each element of the bit-vector.  Computer programs often make use of these operators,
-which are non-linear when interpreted in the standard model of Peano arithmetic\footnote{
-Although some of these operators can be seen as linear in a different algebraic structure,
-e.g. \texttt{xor} corresponds to addition in the Galois field $\mathrm{GF}(2^k)$}.
+As well as machine arithmetic differing from Peano arithmetic on the
+operators they have in common, computers have several ``bitwise'' operations
+that are not taken as primitive in the theory of integers.  These operations
+include the standard boolean operations \texttt{and, or, not, xor} applied
+to each element of the bit-vector.  Computer programs often make use of
+these operators, which are non-linear when interpreted in the standard model
+of Peano arithmetic\footnote{ Although some of these operators can be seen
+as linear in a different algebraic structure, e.g.  \texttt{xor} corresponds
+to addition in the Galois field $\mathrm{GF}(2^k)$}.
 
@@
 
-The problem of program verification can be reduced to the problem
-of finding solutions to a second-order constraint \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}. 
-Our intention is to apply this approach to termination analysis. 
-In this section we show how several variations of both the termination and the non-termination problem can be defined in second-order logic.  
+The problem of program verification can be reduced to the problem of finding
+solutions to a second-order
+constraint~\cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}. 
+Our intention is to apply this approach to termination analysis.  In this
+section we show how several variations of both the termination and the
+non-termination problem can be defined in second-order logic.
+%
 %The second-order unknowns in this constraint are the unknown program invariants
@@
 
-In fact, we identify a fragment of second-order logic with a constrained use of quantification that is expressive enough to encode both termination and non-termination.
-We will suggestively refer to the fragment as the \emph{synthesis fragment}:
+In fact, we identify a fragment of second-order logic with a constrained use
+of quantification that is expressive enough to encode both termination and
+non-termination.  We will suggestively refer to the fragment as the
+\emph{synthesis fragment}:
 
@@
 
-We will begin our discussion by showing how to encode the (non-)termination of program consisting of a single loop 
-in the synthesis fragment.  For the time being, a loop is a pair ${L= \langle G,T \rangle}$,
-where the states $x$ satisfying the loop's guard are given by the predicate $G(x)$.
-The body of the loop is encoded as the transition relation $T(x, x')$, meaning that
-state $x'$ is reachable from state $x$ via a single iteration of the loop body.
-For example, the loop in Figure~\ref{fig:motivation.a} is encoded as:
+We will begin our discussion by showing how to encode the (non-)termination
+of program consisting of a single loop in the synthesis fragment.  For the
+time being, a loop is a pair ${L= \langle G,T \rangle}$, where the states
+$x$ satisfying the loop's guard are given by the predicate $G(x)$.  The body
+of the loop is encoded as the transition relation $T(x, x')$, meaning that
+state $x'$ is reachable from state $x$ via a single iteration of the loop
+body.  For example, the loop in Figure~\ref{fig:motivation.a} is encoded as:
+%
 \begin{align*}
@@
 
-Note that different formulations for unconditional termination are possible. We are aware of a proof rule based on 
-transition invariants, i.e. supersets of the transition relation's transitive closure \cite{DBLP:conf/pldi/GrebenshchikovLPR12}.
-This formulation assumes that the second order logic has a primitive predicate for disjunctive well-foundedness. 
-Conversely, our formulation in Definition~\ref{def:UT} does not require a transitive closure operator, nor a disjunctive well-foundedness predicate. \\
+Note that different formulations for unconditional termination are possible.
+We are aware of a proof rule based on transition invariants, i.e.  supersets
+of the transition relation's transitive closure
+\cite{DBLP:conf/pldi/GrebenshchikovLPR12}.  This formulation assumes that
+the second order logic has a primitive predicate for disjunctive
+well-foundedness.  Conversely, our formulation in Definition~\ref{def:UT}
+does not require a transitive closure operator, nor a disjunctive
+well-foundedness predicate.  \\
 
@@
 
-Only for certain values of \texttt{y} the loop is non-terminating.
-In order to restrict the values that may be assigned to \texttt{y}, we need an assumption associated with the non-deterministic choice. 
-However, such an assumption may cause a reachable state at the non-deterministic assignment to not have any successor. In such a case
-the termination formula would be satisfiable, although the execution halts. This situation can be avoided by checking the existence of 
-at least one successor for each reachable state at the non-deterministic assignment. Although we did put a reasonable amount of effort into it, 
-we failed to come up with a formulation for the non-termination of such non-deterministic transition relations 
-inside the synthesis fragment.
+Only for certain values of \texttt{y} the loop is non-terminating.  In order
+to restrict the values that may be assigned to \texttt{y}, we need an
+assumption associated with the non-deterministic choice.  However, such an
+assumption may cause a reachable state at the non-deterministic assignment
+to not have any successor.  In such a case the termination formula would be
+satisfiable, although the execution halts.  This situation can be avoided by
+checking the existence of at least one successor for each reachable state at
+the non-deterministic assignment.  Although we did put a reasonable amount
+of effort into it, we failed to come up with a formulation for the
+non-termination of such non-deterministic transition relations inside the
+synthesis fragment.
+%
 %formula {\bf [Nondet-CNT]} in Definition~\ref{def:nondet-cnt}.
@@
 \noindent{\bf Nested loops.}
-For brevity reasons, we illustrate our approach to handling nested loops for only two loops, but the approach can be easily generalised to any number of nested loops:
+%
+For brevity reasons, we illustrate our approach to handling nested loops for
+only two loops, but the approach can be easily generalised to any number of
+nested loops:
+%
 \begin{lstlisting}[mathescape=true]
@@
 
-The termination of the outer loop is equivalent to the formula in Definition~\ref{def:nested-term-formula}.
-The formula requires an auxiliary assertion $S(x,x')$ representing a binary relation between the initial states at the inner loop's entry point
-and their successors according to the transition relation $T$. If this formula is satisfiable, then $R$ is a ranking function of the outer loop.
+The termination of the outer loop is equivalent to the formula in
+Definition~\ref{def:nested-term-formula}.  The formula requires an auxiliary
+assertion $S(x,x')$ representing a binary relation between the initial
+states at the inner loop's entry point and their successors according to the
+transition relation $T$.  If this formula is satisfiable, then $R$ is a
+ranking function of the outer loop.
 
@@
 
-As a consequence of the aforementioned conjecture, we do not need to commit to a specific form of the termination argument. While other approaches, e.g. \cite{DBLP:conf/tacas/LeikeH14}, 
-conduct independent searches for each possible form of the ranking function with some of them futile whenever the assumed constrained ranking function does not exist, we have a single search that aims at finding the ranking function with the lowest Kolmogorov complexity.
+As a consequence of the aforementioned conjecture, we do not need to commit
+to a specific form of the termination argument.  While other approaches,
+e.g.  \cite{DBLP:conf/tacas/LeikeH14}, conduct independent searches for each
+possible form of the ranking function with some of them futile whenever the
+assumed constrained ranking function does not exist, we have a single search
+that aims at finding the ranking function with the lowest Kolmogorov
+complexity.
+%
 \fi
@@
 
-By viewing $\sigma: (X^n \times Y^m \to Z^k) \times X^n \times Y^m  \to \mathbb{B}$ as a specification function
-that  returns true iff $P$ computes appropriate outputs for the inputs $x$ and $y$, we can   
-regard the satisfaction problem for the synthesis fragment as \emph{program synthesis}, i.e.
-the mechanised construction of software that provably satisfies a given specification.
-Research in the area of program synthesis has been particularly fruitful beginning with Alonzo Church's work on the
-\emph{Circuit Synthesis Problem} in the sixties~\cite{church-synth}, and continuing with works such as 
-{\sc Brahma}~\cite{brahma} and program sketching~\cite{lezama-thesis,sketch,modular-sketch}.
+By viewing $\sigma: (X^n \times Y^m \to Z^k) \times X^n \times Y^m  \to
+\mathbb{B}$ as a specification function that returns true iff $P$ computes
+appropriate outputs for the inputs $x$ and $y$, we can regard the
+satisfaction problem for the synthesis fragment as \emph{program synthesis},
+i.e.  the mechanised construction of software that provably satisfies a
+given specification.  Research in the area of program synthesis has been
+particularly fruitful beginning with Alonzo Church's work on the
+\emph{Circuit Synthesis Problem} in the sixties~\cite{church-synth}, and
+continuing with works such as {\sc Brahma}~\cite{brahma} and program
+sketching~\cite{lezama-thesis,sketch,modular-sketch}.
 
@@
 
-The {\sc synth} procedure tries to find existential witnesses $(P, x_0)$ which satisfy the partial specification:
+The {\sc synth} procedure tries to find existential witnesses $(P, x_0)$
+which satisfy the partial specification:
+%
 \[
@@
 
-If this formula is unsatisfiable, the candidate solution is in fact a solution to the synthesis formula
-and so the algorithm terminates.  Otherwise, the witness $x$ is an input on which the candidate solution fails
-to meet the specification.  This witness $x$ is added to the {\sc inputs} set and the loop iterates again.
-
-Concretely, {\sc synth} is implemented as shown in Figure~\ref{fig:synth-dfd}.  We start by taking the termination
-specification and generating a C program which takes as input a candidate $(P, x_0)$ and asserts that the candidate
-fails to meet the specification on at least on of the elements of {\sc inputs}.  So this program is unsafe iff there
-is some candidate $(P, x_0)$ satisfying the specification for all the elements of {\sc inputs}.  Finding a new
-candidate solution is therefore reduced to model checking this C program, which by construction is loop free.
-The model checking is done using a combination of bounded model checking, explicit state enumeration and
-genetic programming (GP).  The latter two techniques involve combining the previously generated C program,
-along with code that searches for candidate solutions.  Finally, if any of the model checkers
-finds a candidate solution it returns it.  The form of this candidate solution is yet another program,
-this time written in the proof language $\mathcal{L}$.
-
-As well as efficiency, compiling the original source code of the program under analysis has the advantage of
-guaranteeing that we are verifying the exact semantics of the program as understood by a
-compiler, rather than some ad-hoc abstraction of its semantics.
-
-Similarly to {\sc synth}, the {\sc verif} procedure is implemented by creating a C program that is unsafe iff
-some input exists on which the candidate solution fails to meet the specification.  Again, symbolic
-and explicit state model checking are used to decide the satisfiability of the verification formula.
+If this formula is unsatisfiable, the candidate solution is in fact a
+solution to the synthesis formula and so the algorithm terminates. 
+Otherwise, the witness $x$ is an input on which the candidate solution fails
+to meet the specification.  This witness $x$ is added to the {\sc inputs}
+set and the loop iterates again.
+
+Concretely, {\sc synth} is implemented as shown in
+Figure~\ref{fig:synth-dfd}.  We start by taking the termination
+specification and generating a C program which takes as input a candidate
+$(P, x_0)$ and asserts that the candidate fails to meet the specification on
+at least on of the elements of {\sc inputs}.  So this program is unsafe iff
+there is some candidate $(P, x_0)$ satisfying the specification for all the
+elements of {\sc inputs}.  Finding a new candidate solution is therefore
+reduced to model checking this C program, which by construction is loop
+free.  The model checking is done using a combination of bounded model
+checking, explicit state enumeration and genetic programming (GP).  The
+latter two techniques involve combining the previously generated C program,
+along with code that searches for candidate solutions.  Finally, if any of
+the model checkers finds a candidate solution it returns it.  The form of
+this candidate solution is yet another program, this time written in the
+proof language $\mathcal{L}$.
+
+As well as efficiency, compiling the original source code of the program
+under analysis has the advantage of guaranteeing that we are verifying the
+exact semantics of the program as understood by a compiler, rather than some
+ad-hoc abstraction of its semantics.
+
+Similarly to {\sc synth}, the {\sc verif} procedure is implemented by
+creating a C program that is unsafe iff some input exists on which the
+candidate solution fails to meet the specification.  Again, symbolic and
+explicit state model checking are used to decide the satisfiability of the
+verification formula.
 
@@
 
+\noindent{\bf Symbolic Bounded Model Checking} Another complete method for generating
+candidates is to simply use BMC on the {\sc synth.c} program.  As with explicit
+search, we must progressively increase the length of the solution we search for
+in order to get a complete search procedure.
+\\
+
 \noindent {\bf Genetic Programming and Incremental Evolution} \label{sec:gp}
-To generate candidate (non-)termination proofs in the synthesise phase of the refinement loop,
-we make use of genetic programming (GP) \cite{langdon:fogp} as an alternative to model checking and exhaustive search.  
-Genetic programming provides an intelligent and adaptive way of searching through the space of computer programs
-for an individual that is highly fit in solving the problem at hand
-by using the Darwinian principle of survival and reproduction of the fittest.
-Thus, the program that emerges from the genetic programming paradigm is a consequence of fitness.
-
-We begin with a population of random programs,
-then iteratively evolve this population by applying the genetic operators
-{\sc crossover} and {\sc mutate}. The {\sc crossover} operator takes two programs
-and combines their code in some way to create a new program.  The {\sc mutate} operator takes a single
-program and randomly changes parts of its code, again creating a new program.
-Both these operators are applied to programs selected from the population in
-proportion to their observed fitness, and a program is said to be fitter than another
-if it meets the specification on more test cases.
-
-In the setting of program synthesis, genetic programming can be very usefully combined
-with the concept of incremental evolution.  The idea is that in order to evolve a program
-that is adequate for some difficult task, it is often helpful to first evolve a population
-that as an ensemble is good at some simpler variation of the task.  This principle
-leads us to the observation that each round of the CEGIS loop can be thought of
-as creating a more difficult problem for the GP synthesiser to solve.  Therefore,
-if GP finds a candidate program which is correct for the $k$ inputs so far observed,
-but is not correct for the full specification, that candidate program came from a population
-that was ``good'' for the specification restrict to those $k$ inputs.  The next time round
-the CEGIS loop, the problem GP must solve is slightly harder -- any solution to this new
-problem must be correct for the previous $k$ inputs and also whatever new input was added
-by the previous iteration of {\sc verif}.  It therefore makes sense to begin the $k+1$st
-round of evolution with the population at the end of the $k$th round.
+To generate candidate (non-)termination proofs in the synthesise phase of
+the refinement loop, we make use of genetic programming
+(GP)~\cite{langdon:fogp} as an alternative to model checking and exhaustive
+search.  Genetic programming provides an intelligent and adaptive way of
+searching through the space of computer programs for an individual that is
+highly fit in solving the problem at hand by using the Darwinian principle
+of survival and reproduction of the fittest.  Thus, the program that emerges
+from the genetic programming paradigm is a consequence of fitness.
+
+We begin with a population of random programs, then iteratively evolve this
+population by applying the genetic operators {\sc crossover} and {\sc
+mutate}.  The {\sc crossover} operator takes two programs and combines their
+code in some way to create a new program.  The {\sc mutate} operator takes a
+single program and randomly changes parts of its code, again creating a new
+program.  Both these operators are applied to programs selected from the
+population in proportion to their observed fitness, and a program is said to
+be fitter than another if it meets the specification on more test cases.
+
+In the setting of program synthesis, genetic programming can be very
+usefully combined with the concept of incremental evolution.  The idea is
+that in order to evolve a program that is adequate for some difficult task,
+it is often helpful to first evolve a population that as an ensemble is good
+at some simpler variation of the task.  This principle leads us to the
+observation that each round of the CEGIS loop can be thought of as creating
+a more difficult problem for the GP synthesiser to solve.  Therefore, if GP
+finds a candidate program which is correct for the $k$ inputs so far
+observed, but is not correct for the full specification, that candidate
+program came from a population that was ``good'' for the specification
+restrict to those $k$ inputs.  The next time round the CEGIS loop, the
+problem GP must solve is slightly harder -- any solution to this new problem
+must be correct for the previous $k$ inputs and also whatever new input was
+added by the previous iteration of {\sc verif}.  It therefore makes sense to
+begin the $k+1$st round of evolution with the population at the end of the
+$k$-th round.
 
@@
 \section{Experiments}
-We implemented our algorithm in the {\sc Headshot} tool, which uses the {\sc Kalashnikov} program synthesiser for termination.
-We ran the tool on 40 benchmarks taken from the termination category of SVCOMP'15.
-The experiments were performed on a 4-core 3.3GHz i5-2500k with 8GB of RAM.
+
+We implemented our algorithm in the {\sc Headshot} tool, which uses the {\sc
+Kalashnikov} program synthesiser for termination.  We ran the tool on 40
+benchmarks taken from the termination category of SVCOMP'15.  The
+experiments were performed on a 4-core 3.3\,GHz i5-2500k with 8\,GB of RAM.
 
@@
 
-While the synthesis of termination arguments for linear programs over integers is indeed well-covered in the literature, 
-there is very limited work for programs over machine integers.
-Cook et al. present a method based on a reduction to Presburger
-arithmetic, and a template-matching approach for predefined classes of
-ranking functions based on reduction to SAT- and QBF-solving \cite{DBLP:conf/tacas/CookKRW10}.
-Similarly, the only work we are aware of that can compute non-linear ranking functions  
-for imperative loops with polynomial guards and polynomial assignments
-is \cite{DBLP:conf/vmcai/BradleyMS05}. However, this work extends only to polynomials.
-
-Given the lack of research in handling \emph{not-linear programs}, as well as \emph{programs over bit-vectors and floats},  
-our work focuses on covering these areas. 
-One of the obvious conclusions that can be reached from observing Figure~\ref{fig:handletable}, 
-is that most of the works tend to specialise on a certain aspect of termination proving that they can solve efficiently. 
-Conversely to this view, we aim for \emph{generality}, as we do not restrict the form of the synthesised ranking functions, nor the form of the input programs.
-
-As mentioned in Section~\ref{sec:intro}, approaches based on Ramsey's theorem compute a set of local termination conditions that decrease as execution follows through the loop
-and require expensive reachability analyses
-\cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}.
-In an attempt to reduce the complexity of checking the validity of the termination argument, 
-Cook et al present an iterative termination proving procedure that searches for 
-lexicographic termination arguments \cite{DBLP:conf/tacas/CookSZ13}, 
-whereas Kroening et al. strengthen the termination argument such that it becomes a transitive relation \cite{DBLP:conf/cav/KroeningSTW10}.
-
-Proving program termination implies the simultaneous search for a termination argument and a supporting invariant.
-For this purpose, several tools alternate 
-between calls to a safety prover and calls to a ranking function synthesis tool \cite{}. 
-In order to share more information about the state of the termination proofs between the two tools, 
-Brockschmidt et al. use a single representation of the state of the termination proof search that both tools operate over, 
-resulting in performance improvements \cite{DBLP:conf/cav/BrockschmidtCF13}.
-In \cite{DBLP:conf/cav/BradleyMS05}, Bradley et al. combine the generation of ranking functions with the generation of invariants to
-form one constraint solving problem such that the necessary supporting invariants for the ranking function are discovered on demand.
-In our setting, both the ranking function and the supporting invariant are iteratively constructed through the same refinement loop.
+While the synthesis of termination arguments for linear programs over
+integers is indeed well-covered in the literature, there is very limited
+work for programs over machine integers.  Cook et al.  present a method
+based on a reduction to Presburger arithmetic, and a template-matching
+approach for predefined classes of ranking functions based on reduction to
+SAT- and QBF-solving~\cite{DBLP:conf/tacas/CookKRW10}.  Similarly, the only
+work we are aware of that can compute non-linear ranking functions for
+imperative loops with polynomial guards and polynomial assignments
+is~\cite{DBLP:conf/vmcai/BradleyMS05}.  However, this work extends only to
+polynomials.
+
+Given the lack of research in handling \emph{not-linear programs}, as well
+as \emph{programs over bit-vectors and floats}, our work focuses on covering
+these areas.  One of the obvious conclusions that can be reached from
+observing Figure~\ref{fig:handletable}, is that most of the works tend to
+specialise on a certain aspect of termination proving that they can solve
+efficiently.  Conversely to this view, we aim for \emph{generality}, as we
+do not restrict the form of the synthesised ranking functions, nor the form
+of the input programs.
+
+As mentioned in Section~\ref{sec:intro}, approaches based on Ramsey's
+theorem compute a set of local termination conditions that decrease as
+execution follows through the loop and require expensive reachability
+analyses~\cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}. 
+In an attempt to reduce the complexity of checking the validity of the
+termination argument, Cook et al present an iterative termination proving
+procedure that searches for lexicographic termination
+arguments~\cite{DBLP:conf/tacas/CookSZ13}, whereas Kroening et al. 
+strengthen the termination argument such that it becomes a transitive
+relation~\cite{DBLP:conf/cav/KroeningSTW10}.
+
+Proving program termination implies the simultaneous search for a
+termination argument and a supporting invariant.  For this purpose, several
+tools alternate between calls to a safety prover and calls to a ranking
+function synthesis tool \cite{}.  In order to share more information about
+the state of the termination proofs between the two tools, Brockschmidt et
+al.  use a single representation of the state of the termination proof
+search that both tools operate over, resulting in performance improvements
+\cite{DBLP:conf/cav/BrockschmidtCF13}.  In \cite{DBLP:conf/cav/BradleyMS05},
+Bradley et al.  combine the generation of ranking functions with the
+generation of invariants to form one constraint solving problem such that
+the necessary supporting invariants for the ranking function are discovered
+on demand.  In our setting, both the ranking function and the supporting
+invariant are iteratively constructed through the same refinement loop.
 

### 2014-07-06 papers/termination/paper.tex
edits intro

--- 2014-07-06 papers/termination/paper.tex
@@
 functions over mathematical integers.  This class of ranking functions is
-not large enough to prove termination for all terminating programs, and
+insufficient for prove termination for many terminating programs, and
 furthermore a termination argument for a program operating on mathematical
@@
 Termination, Program Synthesis, Lexicographic Ranking Functions, Bitvector Ranking Functions,
-Floating Point Ranking Functions.
+Floating-Point Ranking Functions.
 
@@
 
-When surveying the area of program termination chronologically, we observe an initial domination of  monolithic approaches based on a single measure shown to decrease
-over all program paths %(syntactic characterisation of loops) 
-\cite{DBLP:conf/vmcai/P04,DBLP:conf/cav/BradleyMS05}, followed by 
-more recent techniques moving towards termination arguments based on Ramsey's theorem \cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}.
-This proof style proves that a transition relation is disjunctively well founded by finding a set of simple
-functions, each of which proves well foundedness for part of the transition relation.
-The main justification for this paradigm shift lies in the simplicity of the local termination measures when compared to the global ones, e.g.
-there are cases in which proofs based on local measures involve linear functions while corresponding global
-measures require non-linear functions or lexicographic orders. 
+When surveying the area of program termination chronologically, we observe
+an initial focus on monolithic approaches based on a single measure shown to
+decrease over all program
+paths~\cite{DBLP:conf/vmcai/P04,DBLP:conf/cav/BradleyMS05}, followed by more
+recent techniques that use termination arguments based on Ramsey's
+theorem~\cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}. 
+Briefly, the transition relation is shown to be disjunctively well founded
+by finding a set of simple functions, each of which proves well foundedness
+for part of the transition relation.  The main benefit of this approach is
+the simplicity of local termination measures in contrast to global ones. 
+For instance, there are cases in which linear arithmetic suffices when using
+local measures, while corresponding global measures require non-linear
+functions or lexicographic orders.
+
 %This explains why approaches based on Ramsey's theorem are restricted to searching for linear termination arguments.
 
-One difficulty with a Ramsey-based approach is that validity of the
-termination argument can only be checked by considering the \emph{transitive
-closure} of the program, rather than a single step.  As such, there is
-experimental evidence that most of the time is spent in reachability
-analysis \cite{DBLP:conf/pldi/CookPR06}, requiring the support of powerful
-safety checkers: there is a tradeoff between the complexity of the
-termination arguments and that of checking their validity.
+One drawback of the Ramsey-based approach is that the validity of the
+termination argument relies on checking the \emph{transitive closure} of the
+program, rather than a single step.  As such, there is experimental evidence
+that most of the effort is spent in reachability
+analysis~\cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/KroeningSTW10},
+requiring the support of powerful safety checkers: there is a tradeoff
+between the complexity of the termination arguments and that of checking
+their validity.
 
@@
 that are easier to
-check~\cite{DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/KroeningSTW10}.
-%
-%In \cite{DBLP:conf/tacas/CookSZ13}, Cook et al present an iterative termination proving procedure that searches for 
-%lexicographic termination arguments, whereas Kroening et al. strengthen the termination argument such that it becomes a transitive relation \cite{DBLP:conf/cav/KroeningSTW10}.
-%
-%
-Following the same trend, %of switching the complexity in termination checking back to the termination arguments, 
-we investigate its extreme: \emph{unrestricted} termination arguments.  This
-means that our ranking functions may involve non-linearity and lexicographic
-orders (we do not commit to any particular syntactic form, and do not use
-templates).
+check~\cite{DBLP:conf/cav/KroeningSTW10,DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/HeizmannHP14}. 
+Following the same trend, we investigate its extreme: \emph{unrestricted}
+termination arguments.  This means that our ranking functions may involve
+non-linearity and lexicographic orders: we do not commit to any particular
+syntactic form, and do not use templates.
 
@@
 programs and linear ranking functions, it also identifies another common
-assumption made by state of the art termination provers: that bit-vector
+assumption made by state-of-the-art termination provers: that bit-vector
 semantics and integer semantics give rise to the same termination behaviour. 
@@
 %% A scenario similar to the one for bit-vectors happens if floats were to be abstracted to unbounded reals by termination provers \cite{}. 
-%% Consequently, these provers ignore potential rounding errors, under- and over-flows, which are precisely what makes  reasoning about floating point inherently difficult.
+%% Consequently, these provers ignore potential rounding errors, under- and over-flows, which are precisely what makes  reasoning about
+%% floating-point inherently difficult.
 %% This approximation may lead to erroneous diagnosis of a program's terminating behaviour as illustrated by the loops in Figure~\ref{} and Figure~\ref{}.
@@
 
-We propose a general framework that %does not assume the existence of linear ranking functions and can 
-uniformly computes
-lexicographic, non-linear 
-ranking functions supported by non-linear inductive invariants for loops with 
-arbitrary guards and transitions over bit-vectors and floats.
-It also computes non-linear recurrence sets, which comprise non-termination proofs.
-Our encoding uniformly multiple loops and nested loops, without having to enumerate
-lassos.
+We propose a general framework that uniformly computes lexicographic,
+non-linear ranking functions supported by non-linear inductive invariants
+for loops with arbitrary guards and transitions over bit-vectors and floats. 
+It also computes non-linear recurrence sets, which comprise non-termination
+proofs.  Our encoding uniformly multiple loops and nested loops, without
+having to enumerate lassos.
 %
@@
 functions that correctly accounts for the wrap-around behavior caused by
-under- and overflows in bit-vector and floating point arithmetic.  Our
+under- and overflows in bit-vector and floating-point arithmetic.  Our
 technique is not restricted to finding linear ranking functions, but can
@@
 
-Floating point arithmetic:
+Floating-point arithmetic:
 

### 2014-07-06 papers/termination/paper.tex
wordings

@@
 
-While the summary  in Figure~\ref{fig:handletable} supports the observation
-that the majority of existing termination analyses are designed for linear
-programs and linear ranking functions, it also identifies another common
-assumption made by state-of-the-art termination provers: that bit-vector
-semantics and integer semantics give rise to the same termination behaviour. 
-Thus, most of the techniques treat bit-vectors and floats as mathematical
-integers and reals,
+Figure~\ref{fig:handletable} summarises the related work with respect to the
+restrictions they impose on the transition relations as well as the form of
+the ranking functions computed.  While it supports the observation that the
+majority of existing termination analyses are designed for linear programs
+and linear ranking functions, it also highlights another simplifying
+assumption made by most state-of-the-art termination provers: that
+bit-vector semantics and integer semantics give rise to the same termination
+behaviour.  Thus, most of the techniques treat bit-vectors and IEEE floats
+as mathematical integers and reals,
 respectively~\cite{DBLP:conf/pldi/CookPR06,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10}.
@@
 i.e.  programs that terminate for integers may \emph{not} terminate for
-bit-vectors and vice-versa.  Thus, abstracting bit-vectors to integers may
+bit-vectors and vice versa.  Thus, abstracting bit-vectors to integers may
 give rise to {\em unsound} and {\em incomplete} analyses.
-%
-%design decisions
-%which is rather surprising given that bit-vectors and floats are ubiquitous in computer systems.
 
@@
 It also computes non-linear recurrence sets, which comprise non-termination
-proofs.  Our encoding uniformly multiple loops and nested loops, without
-having to enumerate lassos.
+proofs.  Our encoding treats multiple loops and nested loops uniformly,
+without having to enumerate lassos.
 %
@@
  For this purpose, we phrase the termination and non-termination problems as
-second order satisfaction problems.  We identify a fragment of second-order
+second-order satisfaction problems.  We identify a fragment of second-order
 logic that is expressive enough to capture our formulation, which we call
@@
 
-%\item We propose a non-trivial formulation of the (non-)termination problem inside a second order logic fragment with only one quantifier alternation.
+%\item We propose a non-trivial formulation of the (non-)termination problem inside a
+% second-order logic fragment with only one quantifier alternation.
 
@@
 
-Figure~\ref{fig:handletable} illustrates the most common assumptions made by existing termination analyses:
+Figure~\ref{fig:handletable} illustrates the most common simplifying
+assumptions made by existing termination analyses:
+%
 \begin{itemize}
@@
 %% we do not care about the CFG (we do not need to encode the CFG 
-%% through program counter in the second order constraints).
+%% through program counter in the second-order constraints).
 
@@
 is expressive enough to encode both termination and non-termination. %allow the formulation of both the termination and the non-termination problem. 
-Following our formulation as second order satisfaction, %the framing of the halting problem as a program synthesis problem,
+Following our formulation as second-order satisfaction, %the framing of the halting problem as a program synthesis problem,
 in Section~\ref{sec:synthesis} we present a solver for the synthesis fragment that models bit-accurate semantics.
@@
 \cite{DBLP:conf/pldi/GrebenshchikovLPR12}.  This formulation assumes that
-the second order logic has a primitive predicate for disjunctive
+the second-order logic has a primitive predicate for disjunctive
 well-foundedness.  Conversely, our formulation in Definition~\ref{def:UT}
@@
 
-%% ---  We begin with the program under analysis (the PUA).  From this we produce a specification (in second order logic)
+%% ---  We begin with the program under analysis (the PUA).  From this we produce a specification (in
+%% second-order logic)
 %% encoding the termination of the PUA.  From the logical specification, we produce a program SYNTH.
@@
  the program length $l$), it will eventually enumerate a program that meets the specification
- in whatever set of inputs are currently being tracked.  Since the first order theory is
+ in whatever set of inputs are currently being tracked.  Since the first-order theory is
  decidable, the query in {\sc Verif} will succeed for this program, causing the algorithm to terminate.

### 2014-07-06 papers/termination/paper.tex
non-termination for nested loops

--- 2014-07-06 papers/termination/paper.tex
@@
 functions over mathematical integers.  This class of ranking functions is
-insufficient for prove termination for many terminating programs, and
+not large enough to prove termination for all terminating programs, and
 furthermore a termination argument for a program operating on mathematical
@@
 Termination, Program Synthesis, Lexicographic Ranking Functions, Bitvector Ranking Functions,
-Floating-Point Ranking Functions.
+Floating Point Ranking Functions.
 
@@
 
-When surveying the area of program termination chronologically, we observe
-an initial focus on monolithic approaches based on a single measure shown to
-decrease over all program
-paths~\cite{DBLP:conf/vmcai/P04,DBLP:conf/cav/BradleyMS05}, followed by more
-recent techniques that use termination arguments based on Ramsey's
-theorem~\cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}. 
-Briefly, the transition relation is shown to be disjunctively well founded
-by finding a set of simple functions, each of which proves well foundedness
-for part of the transition relation.  The main benefit of this approach is
-the simplicity of local termination measures in contrast to global ones. 
-For instance, there are cases in which linear arithmetic suffices when using
-local measures, while corresponding global measures require non-linear
-functions or lexicographic orders.
-
+When surveying the area of program termination chronologically, we observe an initial domination of  monolithic approaches based on a single measure shown to decrease
+over all program paths %(syntactic characterisation of loops) 
+\cite{DBLP:conf/vmcai/P04,DBLP:conf/cav/BradleyMS05}, followed by 
+more recent techniques moving towards termination arguments based on Ramsey's theorem \cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}.
+This proof style proves that a transition relation is disjunctively well founded by finding a set of simple
+functions, each of which proves well foundedness for part of the transition relation.
+The main justification for this paradigm shift lies in the simplicity of the local termination measures when compared to the global ones, e.g.
+there are cases in which proofs based on local measures involve linear functions while corresponding global
+measures require non-linear functions or lexicographic orders. 
 %This explains why approaches based on Ramsey's theorem are restricted to searching for linear termination arguments.
 
-One drawback of the Ramsey-based approach is that the validity of the
-termination argument relies on checking the \emph{transitive closure} of the
-program, rather than a single step.  As such, there is experimental evidence
-that most of the effort is spent in reachability
-analysis~\cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/KroeningSTW10},
-requiring the support of powerful safety checkers: there is a tradeoff
-between the complexity of the termination arguments and that of checking
-their validity.
+One difficulty with a Ramsey-based approach is that validity of the
+termination argument can only be checked by considering the \emph{transitive
+closure} of the program, rather than a single step.  As such, there is
+experimental evidence that most of the time is spent in reachability
+analysis \cite{DBLP:conf/pldi/CookPR06}, requiring the support of powerful
+safety checkers: there is a tradeoff between the complexity of the
+termination arguments and that of checking their validity.
 
@@
 that are easier to
-check~\cite{DBLP:conf/cav/KroeningSTW10,DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/HeizmannHP14}. 
-Following the same trend, we investigate its extreme: \emph{unrestricted}
-termination arguments.  This means that our ranking functions may involve
-non-linearity and lexicographic orders: we do not commit to any particular
-syntactic form, and do not use templates.
+check~\cite{DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/KroeningSTW10}.
+%
+%In \cite{DBLP:conf/tacas/CookSZ13}, Cook et al present an iterative termination proving procedure that searches for 
+%lexicographic termination arguments, whereas Kroening et al. strengthen the termination argument such that it becomes a transitive relation \cite{DBLP:conf/cav/KroeningSTW10}.
+%
+%
+Following the same trend, %of switching the complexity in termination checking back to the termination arguments, 
+we investigate its extreme: \emph{unrestricted} termination arguments.  This
+means that our ranking functions may involve non-linearity and lexicographic
+orders (we do not commit to any particular syntactic form, and do not use
+templates).
 
@@
 programs and linear ranking functions, it also identifies another common
-assumption made by state-of-the-art termination provers: that bit-vector
+assumption made by state of the art termination provers: that bit-vector
 semantics and integer semantics give rise to the same termination behaviour. 
@@
 %% A scenario similar to the one for bit-vectors happens if floats were to be abstracted to unbounded reals by termination provers \cite{}. 
-%% Consequently, these provers ignore potential rounding errors, under- and over-flows, which are precisely what makes  reasoning about
-%% floating-point inherently difficult.
+%% Consequently, these provers ignore potential rounding errors, under- and over-flows, which are precisely what makes  reasoning about floating point inherently difficult.
 %% This approximation may lead to erroneous diagnosis of a program's terminating behaviour as illustrated by the loops in Figure~\ref{} and Figure~\ref{}.
@@
 
-We propose a general framework that uniformly computes lexicographic,
-non-linear ranking functions supported by non-linear inductive invariants
-for loops with arbitrary guards and transitions over bit-vectors and floats. 
-It also computes non-linear recurrence sets, which comprise non-termination
-proofs.  Our encoding uniformly multiple loops and nested loops, without
-having to enumerate lassos.
+We propose a general framework that %does not assume the existence of linear ranking functions and can 
+uniformly computes
+lexicographic, non-linear 
+ranking functions supported by non-linear inductive invariants for loops with 
+arbitrary guards and transitions over bit-vectors and floats.
+It also computes non-linear recurrence sets, which comprise non-termination proofs.
+Our encoding uniformly multiple loops and nested loops, without having to enumerate
+lassos.
 %
@@
 functions that correctly accounts for the wrap-around behavior caused by
-under- and overflows in bit-vector and floating-point arithmetic.  Our
+under- and overflows in bit-vector and floating point arithmetic.  Our
 technique is not restricted to finding linear ranking functions, but can
@@
  \begin{align*}
-  \exists N, x_0 . \forall x . \exists x' . & N(x_0) ~ \wedge ~ N(x) \rightarrow G(x) ~ \wedge \\
+  \exists N, x_0 . \forall x . \exists x' . & N(x_0) ~\wedge \\ &  N(x) \rightarrow G(x) ~ \wedge \\
 							& N(x) \rightarrow T_u(x, x') \wedge N(x') 
@@
  \begin{align*}
-  \exists N, x_0 . \forall x, x' . & N(x_0) ~ \wedge ~ N(x) \rightarrow G(x) ~ \wedge \\
+  \exists N, x_0 . \forall x, x' . & N(x_0) ~ \wedge \\ & N(x) \rightarrow G(x) ~ \wedge \\
 							& N(x) \wedge T_u(x, x') \rightarrow N(x') 
@@
 ours does not have.\\
-
-\noindent {\bf Non-determinism.} A non-termination diagnosis procedure based on checking the satisfiability of formula {\bf [CNT]}
-may fail to find a closed recurrence set for a loop $L=(G, T_u)$ when $T_u$ is non-deterministic.  
-To understand the issue, imagine the following loop with an non-deterministic assignment:
-
-\begin{lstlisting}[language=C]
-while(x != 0) {
-  y = nondet();
-  x = x-y;
-}
-\end{lstlisting}
-
-Only for certain values of \texttt{y} the loop is non-terminating.  In order
-to restrict the values that may be assigned to \texttt{y}, we need an
-assumption associated with the non-deterministic choice.  However, such an
-assumption may cause a reachable state at the non-deterministic assignment
-to not have any successor.  In such a case the termination formula would be
-satisfiable, although the execution halts.  This situation can be avoided by
-checking the existence of at least one successor for each reachable state at
-the non-deterministic assignment.  Although we did put a reasonable amount
-of effort into it, we failed to come up with a formulation for the
-non-termination of such non-deterministic transition relations inside the
-synthesis fragment.
-%
-%formula {\bf [Nondet-CNT]} in Definition~\ref{def:nondet-cnt}.
 
@@
 This issue does not arise when abstracting the environment as its under-approximation and the loop under consideration make use of existential quantification at the same nestedness level.
-However, when dealing with nested loops, we need to compute the under-approximation of the inner loop, which requires a nested existential quantifier,
-resulting in the $\exists \forall \exists$ alternation. 
-%% computing the under-approximation of the loop's body requires an extra quantifier alternation. 
-%% The origin of the problem comes from the fact that an under-approximation is an existential property and therefore
-%% requires an extra existential quantifier, resulting in the $\exists \forall \exists$ alternation.
-This situation is similar to the one encountered when proving non-termination of non-deterministic loops.
-To go around this limitation, we treat nested loops as a special case and give a separate formula for them in Figure~\ref{fig:nonterm-nested}.
+
+However, when dealing with nested loops, we need to  under-approximate the inner loop, requiring a nested existential quantifier, 
+resulting in the $\exists \forall \exists$ alternation.  At first sight, this seems fatal for our formulation.  Luckily, it is possible to
+reformulate non-termination as an $\exists \forall$ formula by observing that, unlike a ranking function, the defining property of a
+recurrence set is \emph{non relational} -- if we end up in the recurrence set, we don't care exactly where we came from as long as
+we know that it was also somewhere in the recurrence set.  This allows us to cast non-termination of nested loops as the formula
+shown in Figure~\ref{fig:nonterm-nested}.
+
 If the formula on the right of the figure is satisfiable, then $L_1$ is non-terminating as witnessed by the recurrence set $N_1$ and
@@
 \item If $L_2$ is terminating, then $N_2$ is an inductive invariant that reestablished $N_1$ after $L_2$ stops executing: $\lnot G_2(x) \wedge N_2(x) \wedge P_2(x,x') \rightarrow N_1(x') $. 
-\item If $L_2$ is non-terminating, then $N_2$ is its recurrence set.
+\item If $L_2$ is non-terminating, then $N_2 \wedge G_2$ is its recurrence set.
 \end{itemize}
-
-%% The issue appears when we try to under-approximate the body of the loop. 
-%% Here we encounter a similar problem with the one for non-determinism mentioned above.
-
-%% Consider the generic example in Figure~\ref{fig:nonterm-environment}, where we are attempting to prove non-termination
-%% of the innermost loop $L_3$ by finding a closed recurrence set $N_3$. 
-%% -- We will assume that the other two loops iterate only a finite number of times.
-
-%% We now explain the constraints on the right of the figure.
-%% Given that we prove non-termination, we try to  under-approximate the states from which $L_3$ can start executing. 
-%% We do this in several steps:
-
-%% $W_1$, $W_2$ and $W_3$ are inductive invariants corresponding to $L_1$, $L_2$ and $L_3$, respectively, such that:
-
-%% - $W_1$ is an inductive invariant of $L_1$ that is strong enough to establish $W_2$ if $L_1$ exits (which it does by assumption).
-%% - $W_2$ is an inductive invariant of $L_2$ that is strong enough to establish either $W_3$ or $N_3$ on entry to $L_3$, depending on whether $L_3$ terminates or not.
-%% - Recall that $N_3$ is a recurrence set for $L_3$, so if a state $x \in N_3$ is reached, the loop $L_3$ will execute forever. Otherwise, $W_3$ holds on entry to $L_3$.
-%% - $W_3$ is an inductive invariant of $L_3$ that is strong enough to re-establish $W_2$ when $L_3$ exits.
-
-%% If the non-termination formula is satisfiable, $x_0$ is some initial state in which the program begins executing.
-
-%% First, we under-approximate the states in which $L_2$ can start executing through $W_2$.
-%% Then $N_1$ is an inductive invariant of $L_1$ that is strong
-%% enough to establish $N_2$ if $L_1$ exits (which it does by assumption).  Now, $N_2$ is an inductive invariant
-%% of $L_2$ that is strong enough to establish either $N_3$ or $N_4$ on entry to $L_3$.  Recall that $N_4$ is
-%% a recurrence set for $L_3$, so if a state $x \in N_4$ is reached, the loop $L_3$ will execute forever.
-%% Otherwise, $N_3$ holds on entry to $L_3$.  $N_3$ is an inductive invariant of $L_3$ that is strong enough to
-%% re-establish $N_2$ when $L_3$ exits.
-
-%% If the constraint system is satisfiable, $x_0$ is some initial
-%% state in which the program begins executing.
-
-%%   Since we assumed that $L_2$ iterated only a finite number of times,
-%% $N_3$ must be established only a finite number of times, which means that $N_4$ will eventually be reached,
-%% at which point $L_3$ will loop forever since $N_4$ is a recurrence set for $L_3$.
-
-%% \begin{figure*}
-%% \begin{framed}
-%%  \begin{minipage}{0.19\textwidth}
-%% \begin{lstlisting}
-%% L1: while (c1) {
-%%      B1;
-%%     }
-
-%% L2: while (c2) {
-%%      P;
-
-%% L3:  while (c3) {
-%%        B3;
-%%      }
-%%     }
-%% \end{lstlisting}
-%% \end{minipage}
-%% \vline
-%% \begin{minipage}{0.82\textwidth}
-%% \begin{align*}
-%%  \exists W_1, W_2, W_3, N_3, x_0 . \forall x, x' . \\
-%%   W_1(x_0) \, \wedge \\
-%%   c_1(x) \wedge W_1(x) \wedge B_1(x, x') & \rightarrow W_1(x') \, \wedge \\
-%%   \lnot c_1(x) \wedge W_1(x) & \rightarrow W_2(x) \, \wedge \\
-%%   W_2(x) & \rightarrow c_2(x) \, \wedge \\
-%%   W_2(x) \wedge P(x, x') & \rightarrow W_3(x') \vee N_3(x') \, \wedge \\
-%%   c_3(x) \wedge W_3(x) \wedge B_3(x, x') & \rightarrow W_3(x') \, \wedge \\
-%%   \lnot c_3(x) \wedge W_3(x) & \rightarrow W_2(x) \, \wedge \\
-%%   N_3(x) & \rightarrow c_3(x) \, \wedge \\
-%%   N_3(x) \wedge B_3(x, x') & \rightarrow N_3(x')
-%% \end{align*}
-%% \end{minipage}
-%% \end{framed}
-
-%% \caption{The constraint system for non-termination\label{fig:nonterm-environment}}
-%% \end{figure*}
 
@@
 
-Floating-point arithmetic:
+Floating point arithmetic:
 

### 2014-07-06 papers/termination/paper.tex
Typo

--- 2014-07-06 papers/termination/paper.tex
@@
 functions over mathematical integers.  This class of ranking functions is
-insufficient for prove termination for many terminating programs, and
+insufficient for proving termination of many terminating programs, and
 furthermore a termination argument for a program operating on mathematical

### 2014-07-06 papers/termination/paper.tex
various edits

@@
 functions over mathematical integers.  This class of ranking functions is
-insufficient for proving termination of many terminating programs, and
+insufficient for prove termination for many terminating programs, and
 furthermore a termination argument for a program operating on mathematical
@@
 found, but if the program does not terminate we may not find a recurrence
-set.  However if the program's transition relation is deterministic, our
+set.  However, if the program's transition relation is deterministic, our
 method is complete.
@@
 This means that our procedure is not directly dependent on the structure of
-the analysed loop, e.g.  number of variable, number of lassos, but on the
-Kolmogorov complexity of its minimal termination argument.  We argue that
-the length of a program's termination proof is indicative of how easy to
-understand the program is, and thus programmers tend to write programs that
-have relatively short termination proofs.  We empirically show that this
-claim holds in practice, causing our technique to perform well.
+the analysed loop (e.g.~the number of variables or the number of lassos),
+but on the \emph{Kolmogorov complexity of its minimal termination argument}. 
+We argue that the length of a program's termination proof is indicative of
+how easy to understand the program is, and thus programmers tend to write
+programs that have relatively short termination proofs.  We empirically show
+that this claim holds in practice, causing our technique to perform well.
 
@@
 
-\section{Motivational Examples} \label{sec:motivation}
+\section{Motivating Examples} \label{sec:motivation}
 %- see phase change programs.
@@
 \item[(i)] programs use only linear arithmetic.
-\item[(ii)] terminating programs have termination arguments in linear arithmetic.
-\item[(iii)] the semantics of bit-vectors and mathematical integers are equivalent.
-\item[(iv)] the semantics of floats and mathematical reals are equivalent.
+\item[(ii)] terminating programs have termination arguments expressible in linear arithmetic.
+\item[(iii)] the semantics of bit vectors and mathematical integers are equivalent.
+\item[(iv)] the semantics of IEEE floating-point numbers and mathematical reals are equivalent.
 \end{itemize}
@@
 %In the previous sections, we have discussed some of the assumptions made by existent termination analyses (see Figure~\ref{fig:handletable}).
-To show how these assumptions are broken by even simple programs, we draw the reader's attention to the programs in
-Figure~\ref{fig:motivation} and their curious properties:
-
+%
+To show how these assumptions are violated by even simple programs, we draw
+the reader's attention to the programs in Figure~\ref{fig:motivation} and
+their curious properties:
+%
 %We next discus some intricate examples that prove challenging for existent termination analysis.
-
+%
 %\subsection{Program linearity vs. the linearity of its ranking function.}
-
+%
 %Figure~\ref{fig:handletable} shows that the most common assumption existent termination techniques make is that \emph{programs are linear}.
-
+%
 \begin{itemize}
@@
 \item Program (a) breaks assumption (i) as it makes use of the bit-wise $\&$ operator.
-%As such, Figure~\ref{} denotes a program with non-linear operations. Given that  many of the existing techniques commit to linear programs, they cannot handle this situation, 
-%although a linear ranking function does exist (see Figure~\ref{fig:handletable}). 
-%In order to find a ranking function for this example, it is necessary to take into account
-%the semantics of the bit-wise AND operator.
-Our technique finds that an admissible ranking function is the linear function
-$R(x) = x$, whose value decreases with
-every iteration, but cannot decrease indefinitely as it is bounded from below.
-This example also illustrates the lack of a direct correlation between the linearity of a program and that of its termination arguments.
-
-\item Program (b) break assumption (ii), in that it has no linear ranking function.
-We prove that this loop terminates by finding the non-linear ranking function $R(x) = |x|$.
-
-\item Program (c) breaks assumption (iii).
-This loop is  terminating for bit-vectors since $x$
-will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
-arithmetic since $x > 0 \rightarrow x+1 > 0$ for any integer $x$.
-
-\item Program (d) also breaks assumption(iii), but ``the other way'' -- it terminates for integers but not for bit-vectors.
-If each of the variables is stored in an unsigned $k$-bit word, the following configuration will
-lead to an infinite loop:
-$$M = 2^k - 1, N = 2^k - 1, i = M, j = N-1$$
-
-\item Program (e) breaks assumption (iv) -- it terminates for reals but not for floats.
-If $x$ is sufficiently large, rounding error will cause the subtraction to have no effect.
-
-\item Program (f) breaks assumption (iv) ``the other way'' -- it terminates for floats but not for reals.
-Eventually $x$ will become sufficiently small that the nearest representable number is 0.0, at which point
-it will be rounded to 0.0 and the loop will terminate.
+%
+%As such, Figure~\ref{} denotes a program with non-linear operations. Given
+%that many of the existing techniques commit to linear programs, they cannot
+%handle this situation, although a linear ranking function does exist (see
+%Figure~\ref{fig:handletable}).  In order to find a ranking function for
+%this example, it is necessary to take into account the semantics of the
+%bit-wise AND operator.
+%
+Our technique finds that an admissible ranking function is the linear
+function $R(x) = x$, whose value decreases with every iteration, but cannot
+decrease indefinitely as it is bounded from below.  This example also
+illustrates the lack of a direct correlation between the linearity of a
+program and that of its termination arguments.
+
+\item Program (b) breaks assumption (ii), in that it has no linear ranking
+function.  We prove that this loop terminates by finding the non-linear
+ranking function $R(x) = |x|$.
+
+\item Program (c) breaks assumption (iii).  This loop is terminating for
+bit-vectors since $x$ will eventually over-flow and become negative. 
+Conversely, the same program is non-terminating using integer arithmetic
+since $x > 0 \rightarrow x+1 > 0$ for any integer $x$.
+
+\item Program (d) also breaks assumption (iii), but ``the other way'': it
+terminates for integers but not for bit-vectors.  If each of the variables
+is stored in an unsigned $k$-bit word, the following entry state will lead
+to an infinite loop:
+%
+$$ M = 2^k - 1,\quad N = 2^k - 1,\quad i = M,\quad j = N-1 $$
+
+\item Program (e) breaks assumption (iv): it terminates for reals but not
+for floats.  If $x$ is sufficiently large, rounding error will cause the
+subtraction to have no effect.
+
+\item Program (f) breaks assumption (iv) ``the other way'': it terminates
+for floats but not for reals.  Eventually $x$ will become sufficiently small
+that the nearest representable number is $0.0$, at which point it will be
+rounded to $0.0$ and the loop will terminate.
+
 \end{itemize}
 
-Up until this point, we looked at examples that are not soundly treated by existing techniques as they 
-don't fit in the range of programs covered by these techniques.
-Next, we look at some programs that are handled by existing termination tools via dedicated analyses. We show that our method 
-uniformly handles them, without the need for any special treatment.
-% we are able to compute termination arguments in several interesting case
-
-% Our technique gains from generality. A possible explanation for this (perhaps initially sur-
-%prising) fact is that, for synthesis, we are interested in the mere existence of a
-%% solution, and the loss of
-%% many solutions does not necessarily mean the loss of
-%% all solutions of the constraint
-% bit-level analysis
-%These non-terminating programs represent an important class of
-%bugs, especially for denial-of-service related vulnerabilities, where
-%an attacker can exploit the overflow. Equally important,
-% equally important
-
+Up until this point, we considered examples that are not soundly treated by
+existing techniques as they don't fit in the range of programs addressed by
+these techniques.  Next, we look at some programs that are handled by
+existing termination tools via dedicated analyses.  We show that our method
+handles them uniformly, without the need for any special treatment.
+%
 \begin{itemize}
-\item Program (g) is a linear program
-which is shown in~\cite{DBLP:conf/tacas/CookSZ13} not to admit (without prior manipulation) a lexicographic linear function.
-With our technique we can find the non-linear ranking function $R(x) = |x|$.
+
+\item Program (g) is a linear program that is shown
+in~\cite{DBLP:conf/tacas/CookSZ13} not to admit (without prior manipulation)
+a lexicographic linear ranking function.  With our technique we can find the
+non-linear ranking function $R(x) = |x|$.
+
 %As with linear lexicographic ranking functions, a state is mapped to a tuple of values such that the
@@
 \item In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, program (i)
-admits a multiphase ranking function computed from a multiphase ranking
-template.  A multiphase ranking template is targeted at programs that go
-through a finite number of phases in their execution.  Each phase is ranked
-with an affine-linear function and the phase is considered to be completed
-once this function becomes non-positive.
+admits a \emph{multiphase} ranking function, computed from a multiphase
+ranking template.  Multiphase ranking templates are targeted at programs
+that go through a finite number of phases in their execution.  Each phase is
+ranked with an affine-linear function and the phase is considered to be
+completed once this function becomes non-positive.
 
@@
 can find a non-linear lexicographic ranking function $R(x, y, z) = (x < y,
-z)$.  This termination argument is somewhat subtle, but can be understood as
-follows: the boolean quantity $x < y$ eventually decreases, that is to say
-once a state with $x \geq y$ is reached, $x$ never again becomes greater
-than $y$.  This means that as soon as the ``else'' branch of the if
-statement is taken, it will continue to be taken in each subsequent
-iteration of the loop.  Meanwhile, if $x < y$ has not decreased (i.e.  we
-have stayed in the same branch of the ``if'') then $z$ does decrease.  Since
-a boolean only has two possible values, it cannot decrease indefinitely. 
-Since $z > 0$ is a conjunct of the loop guard, $z$ cannot decrease
-indefinitely, and so $R$ proves that the loop is well founded.
-
-As with all of the termination proofs presented in this paper, the ranking
-function above was found completely automatically.
-
+z)$.\footnote{This termination argument is somewhat subtle, but can be understood as
+follows: The Boolean values $\mathit{false}$ and $\mathit{true}$ are
+interpreted as 0 and 1, respectively.  The Boolean $x < y$ thus eventually
+decreases, that is to say once a state with $x \geq y$ is reached, $x$ never
+again becomes greater than $y$.  This means that as soon as the ``else''
+branch of the if statement is taken, it will continue to be taken in each
+subsequent iteration of the loop.  Meanwhile, if $x < y$ has not decreased
+(i.e., we have stayed in the same branch of the ``if''), then $z$ does
+decrease.  Since a Boolean only has two possible values, it cannot decrease
+indefinitely.  Since $z > 0$ is a conjunct of the loop guard, $z$ cannot
+decrease indefinitely, and so $R$ proves that the loop is well founded.}
 \iffalse
@@
 \end{itemize}
+%
+As with all of the termination proofs presented in this paper, the ranking
+functions above were all found completely automatically.
 
@@
 \section{Preliminaries}
-Given a program, %with state space $X$ and transition relation $T \subseteq X \times X$, 
-we first formalise its termination argument as a ranking function
-(Section~\ref{sec:ranking.functions}).  Subsequently, we discus bit-vector
-semantics and illustrate differences between machine arithmetic and integer
-arithmetic that make the abstraction of bit-vectors to mathematical integers
-unsound (Section~\ref{sec:machine.arith}).
+
+Given a program, we first formalise its termination argument as a ranking
+function (Section~\ref{sec:ranking.functions}).  Subsequently, we discus
+bit-vector semantics and illustrate differences between machine arithmetic
+and integer arithmetic that show that the abstraction of bit vectors to
+mathematical integers unsound (Section~\ref{sec:machine.arith}).
 
 \subsection{Termination and Ranking Functions} \label{sec:ranking.functions}
-A programs $P$ is represented as a transition systems with state space $X$, transition relation $T \subseteq X \times X$, and set of variables $x$. 
-A state of the program is a valuation of the variables from $x$. For a state $s \in X$ with $T(s,s')$ we say $s'$
-is the successor of $s$ under $T$.
+
+A programs $P$ is represented as a transition systems with state space $X$,
+transition relation $T \subseteq X \times X$, and set of variables $V$.  A
+state of the program is a valuation of the variables from $V$.  For a state
+$x \in X$ with $T(x,x')$ we say $x'$ is the successor of $x$ under $T$.
 
 \begin{definition}[Unconditional termination]
-A transition system is said to be \emph{unconditionally terminating} if there is no infinite sequence of states
-$x_1, x_2, \ldots \in X$ with $\forall i . T(x_i, x_{i+1})$.  
+%
+A transition system is said to be \emph{unconditionally terminating} if
+there is no infinite sequence of states $x_1, x_2, \ldots \in X$ with
+$\forall i.~T(x_i, x_{i+1})$.
+%
 \end{definition}
 
-We can prove that the transition system is unconditionally terminating by finding a ranking function for its transition relation.
-
+We can prove that the transition system is unconditionally terminating by
+finding a ranking function for its transition relation.
+%
 \begin{definition}[Ranking function]
-An injective function ${R:X\to Y}$ is a \emph{ranking function} for the transition relation $T$ if $Y$ is a well-founded set and 
+%
+An injective function ${R:X\to Y}$ is a \emph{ranking function} for the
+transition relation $T$ if $Y$ is a well-founded set with order $<$ and 
 $R$ is monotonically decreasing with respect to $T$.  That is
@@
 $$\forall x, x' \in X. T(x, x') \Rightarrow R(x) < R(x')$$
+%
 \end{definition}
@@
 
-In the case that $\dim(Y) = 1$, this reduces to an inner product:
-$$f(\vec{x}) = \vec{\lambda} \cdotp \vec{x} + c$$
+In the case that $\dim(Y) = 1$, this reduces to the inner product
+%
+$$f(\vec{x}) = \vec{\lambda} \cdotp \vec{x} + c \;.$$
 
@@
 
-\subsection{Machine Arithmetic and Bitvectors} \label{sec:machine.arith}
+\subsection{Machine Arithmetic and Bit Vectors} \label{sec:machine.arith}
 
@@
 such a large number and so modern computers do their arithmetic over
-fixed-width binary words, otherwise known as bit-vectors.  For the remainder
-of this section, we will say that the bit-vectors we are working with are
+fixed-width binary words, otherwise known as bit vectors.  For the remainder
+of this section, we will say that the bit vectors we are working with are
 $k$-bits wide, which means that each word can hold one of $2^k$ bit
 patterns.  Typical values for $k$ are 32 and 64.
-
-Bit-vector arithmetic is performed modulo $2^k$, which is the source of many
-of the differences between machine arithmetic and Peano arithmetic.  To give
-an example: $(2^k - 1) + 1 \equiv 0 \pmod {2^k}$.  This provides a
-counterexample to the statement $\forall x .  x + 1 > x$, which is a theorem
-of Peano arithmetic but not of modular arithmetic.  When an arithmetic
-operation has a result greater than $2^k$, it is said to ``overflow''.  If
-an operation does not overflow, its machine-arithmetic result is the same as
-the result of the same operation performed on integers.
 
@@
 multiplication and subtraction are defined identically for unsigned and
-signed numbers.  Therefore signed and unsigned arithmetic differ only on
-comparison operators\footnote{There is also the difference that signed
-overflow in a C program results in undefined behaviour, but in practice the
-undefined behaviour is implemented just as if the arithmetic had been
-unsigned}.  For example: $$\forall x .  -x \leq_s x \wedge -x \geq_u x$$
-where $\leq_s$ and $\geq_u$ represent signed and unsigned comparisons
-respectively.
+signed numbers.
+
+Bit-vector arithmetic is performed modulo $2^k$, which is the source of many
+of the differences between machine arithmetic and Peano
+arithmetic.\footnote{ISO C requires that unsigned arithmetic is performed
+modulo $2^k$, whereas the overflow case is undefined for signed arithmetic. 
+In practice, the undefined behaviour is implemented just as if the
+arithmetic had been unsigned}.  To give an example, $(2^k - 1) + 1 \equiv 0
+\pmod {2^k}$ provides a counterexample to the statement $\forall x. 
+x + 1 > x$, which is a theorem of Peano arithmetic but not of modular
+arithmetic.  When an arithmetic operation has a result greater than $2^k$,
+it is said to ``overflow''.  If an operation does not overflow, its
+machine-arithmetic result is the same as the result of the same operation
+performed on integers.
 
 The final source of difference between integer arithmetic and bitvector
-arithmetic stems from width conversions.  Typical computers allow variables
-to have different types, which can be represented using words of different
-widths.  In C a \texttt{short} might occupy 16 bits, while an \texttt{int}
-might occupy 32 bits.  If a $k$-bit variable is assigned to a $l$-bit
-variable with $l < k$, the result is truncated $\mod 2^l$.  For example, if
-$x$ is a 32-bit variable and $y$ is a 16-bit variable, $y$ will hold the
-value $0$ after the following code is executed:
-
+arithmetic stems from width conversions.  Many programming languages allow
+numeric variables of different types, which can be represented using words
+of different widths.  In C, a \texttt{short} might occupy 16 bits, while an
+\texttt{int} might occupy 32 bits.  When a $k$-bit variable is assigned to a
+$l$-bit variable with $l < k$, the result is truncated $\mod 2^l$.  For
+example, if $x$ is a 32-bit variable and $y$ is a 16-bit variable, $y$ will
+hold the value $0$ after the following code is executed:
+%
 \begin{lstlisting}
@@
 
-This gives us a counterexample to the statement $\forall x, y . x = y \Rightarrow (x + 1) = (y + 1)$,
-which is a theorem of Peano arithmetic.
+This gives us a counterexample to the statement $\forall x, y . x = y
+\Rightarrow (x + 1) = (y + 1)$, which is a theorem of Peano arithmetic.
 
@@
 that are not taken as primitive in the theory of integers.  These operations
-include the standard boolean operations \texttt{and, or, not, xor} applied
-to each element of the bit-vector.  Computer programs often make use of
-these operators, which are non-linear when interpreted in the standard model
-of Peano arithmetic\footnote{ Although some of these operators can be seen
-as linear in a different algebraic structure, e.g.  \texttt{xor} corresponds
-to addition in the Galois field $\mathrm{GF}(2^k)$}.
+include the Boolean operators \texttt{and, or, not, xor} applied to each
+element of the bit vector.  Computer programs often make use of these
+operators, which are non-linear when interpreted in the standard model of
+Peano arithmetic\footnote{Some of these operators can be seen as
+linear in a different algebraic structure, e.g.~\texttt{xor} corresponds to
+addition in the Galois field $\mathrm{GF}(2^k)$.}.
 
@@
 \begin{definition}[Synthesis Fragment]
- A formula is in the synthesis fragment iff it is of the form
+A formula is in the \emph{synthesis fragment} iff it is of the form
+%
  \[
@@
  \]
-
-where $\vec{P}$ ranges over sets, while $\vec{x}$ and $\vec{y}$ range over ground terms.
-Function $\sigma: (X^n \times Y^m \to Z^k) \times X^n \times Y^m  \to \mathbb{B}$ can be viewed as a specification function
-that  returns true iff the functions $\vec{P}$ compute appropriate outputs
-when fed the inputs $\vec{x}$ and $\vec{y}$.  
+%
+where $\vec{P}$ ranges over sets, while $\vec{x}$ and $\vec{y}$ range over
+ground terms.  Function $\sigma: (X^n \times Y^m \to Z^k) \times X^n \times
+Y^m \to \mathbb{B}$ can be seen as a specification function that returns
+true iff the functions $\vec{P}$ compute appropriate outputs when fed the
+inputs $\vec{x}$ and $\vec{y}$.
+%
 \end{definition}
@@
 
-We will begin our discussion by showing how to encode the (non-)termination
-of program consisting of a single loop in the synthesis fragment.  For the
-time being, a loop is a pair ${L= \langle G,T \rangle}$, where the states
-$x$ satisfying the loop's guard are given by the predicate $G(x)$.  The body
-of the loop is encoded as the transition relation $T(x, x')$, meaning that
-state $x'$ is reachable from state $x$ via a single iteration of the loop
-body.  For example, the loop in Figure~\ref{fig:motivation.a} is encoded as:
+We will begin our discussion by showing how to encode the
+\mbox{(non-)termination} of a program consisting of a single loop in the
+synthesis fragment.  For the time being, a loop is a pair ${L= \langle G,T
+\rangle}$, where the states $x$ satisfying the loop's guard are given by the
+predicate $G(x)$.  The body of the loop is encoded as the transition
+relation $T(x, x')$, meaning that state $x'$ is reachable from state $x$ via
+a single iteration of the loop body.  For example, the loop in
+Figure~\ref{fig:motivation.a} is encoded as:
 %
@@
 
-\begin{definition}[Non-Termination Formula - Open Recurrence Set  {\bf [ONT]}]
+\begin{definition}[Non-Termination Formula -- Open Recurrence Set  {\bf [ONT]}]
 \label{def:ont}
@@
 
-\begin{definition}[Non-Termination Formula - Closed Recurrence Set {\bf [CNT]}]
+\begin{definition}[Non-Termination Formula -- Closed Recurrence Set {\bf [CNT]}]
 \label{def:cnt}
@@
 As the existence of a ranking function is equivalent to the satisfiability
-of the formula {\bf [UT]}, a satisfiality witness 
-is also a ranking function and thus a proof of $L$'s unconditional termination.
-
-Continuing the example of the program in Figure~\ref{fig:motivation.a}, we can see that
-the corresponding synthesis formula {\bf [UT]} is satisfiable, as witnessed by the function $R(x) = x$.
-So $R(x) = x$ constitutes a proof that the program in Figure~\ref{fig:motivation.a} is unconditionally
+of the formula {\bf [UT]}, a satisfiability witness is also a ranking
+function and thus a proof of $L$'s unconditional termination.
+
+Continuing the example of the program in Figure~\ref{fig:motivation.a}, we
+can see that the corresponding synthesis formula {\bf [UT]} is satisfiable,
+as witnessed by the function $R(x) = x$.  Thus, $R(x) = x$ constitutes a
+proof that the program in Figure~\ref{fig:motivation.a} is unconditionally
 terminating.
@@
 We are aware of a proof rule based on transition invariants, i.e.  supersets
-of the transition relation's transitive closure
-\cite{DBLP:conf/pldi/GrebenshchikovLPR12}.  This formulation assumes that
-the second-order logic has a primitive predicate for disjunctive
+of the transition relation's transitive
+closure~\cite{DBLP:conf/pldi/GrebenshchikovLPR12}.  This formulation assumes
+that the second-order logic has a primitive predicate for disjunctive
 well-foundedness.  Conversely, our formulation in Definition~\ref{def:UT}
@@
 
-Gupta et al. \cite{DBLP:conf/popl/GuptaHMRX08} characterize non-termination of a transition relation $T$ by the existence of an \emph{(open) recurrence set},
-i.e. a nonempty set of states $N$ such that for each $s \in N$ there
+Gupta et al.~\cite{DBLP:conf/popl/GuptaHMRX08} characterize non-termination
+of a transition relation $T$ by the existence of an \emph{(open) recurrence
+set}, i.e.  a nonempty set of states $N$ such that for each $s \in N$ there
 exists a transition to some $s'\in N$.
 For our case, the notion of open recurrent set is encoded by formula {\bf [ONT]}. %Definition~\ref{def:nonterm-formula}.
-Dual to the termination problem, when formulating $L$'s non-termination, we only need to consider an under-approximation of its transition relation, $T_u$. 
+Dual to the termination problem, when formulating $L$'s non-termination, we only need to consider an under-approximation of its transition
+relation~$T_u$. 
 %$L = \langle G,T \rangle$
-Again, this distinction is mostly important when dealing with nested loops, as illustrated in Section~\ref{sec:env}.
+Again, this distinction is mostly important when dealing with nested loops,
+as illustrated in Section~\ref{sec:env}.
 
@@
 
-If this formula is satisfiable, $N$ is an open recurrence set for $L$ which proves
+If this formula is satisfiable, $N$ is an open recurrence set for $L$, which proves
 $L$'s non-termination. The issue with this formula is the additional level of quantifier alternation as compared to the synthesis fragment
@@
 
-In order to formulate the non-termination problem inside the synthesis fragment, 
-we make use of the notion of \emph{closed recurrence set} introduced by Chen et al. in \cite{DBLP:conf/tacas/ChenCFNO14}. 
-As opposed to an open recurrence set, this is a $\exists \forall$ property: for each state in the recurrence set $N$, all of its successors 
-must be in $N$. We note that if $T$ is deterministic, every open recurrence set is also a closed recurrence set (since each
-state has at most one successor).  Thus, the non-termination problem for deterministic transition systems is equivalent to the
-satisfiability of formula {\bf [CNT]}. %Definition~\ref{def:deterministic-nonterm-formula}.
-%As our formulation is with respect to a given loop $L=\langle I,G,T \rangle$, we require the  
-%recurrence set to include the loop's guard, $N(x) \rightarrow G(x)$. In other words, the loop's guard needs to be satisfied in order for the 
+In order to formulate the non-termination problem inside the synthesis
+fragment, we make use of the notion of \emph{closed recurrence set}
+introduced by Chen et al.~in~\cite{DBLP:conf/tacas/ChenCFNO14}.  As opposed
+to an open recurrence set, this is an $\exists \forall$ property: for each
+state in the recurrence set $N$, all of its successors must be in $N$.  We
+note that if $T$ is deterministic, every open recurrence set is also a
+closed recurrence set (since each state has at most one successor).  Thus,
+the non-termination problem for deterministic transition systems is
+equivalent to the satisfiability of formula {\bf [CNT]}.
+
+%Definition~\ref{def:deterministic-nonterm-formula}.  As our formulation is
+%with respect to a given loop $L=\langle I,G,T \rangle$, we require the
+%recurrence set to include the loop's guard, $N(x) \rightarrow G(x)$.  In
+%other words, the loop's guard needs to be satisfied in order for the
 %infinite execution to happen.
@@
 
-Note that by defining $T_u$'s determinism as the existence of exactly one successor for any state satisfying $L$'s guard, 
-we eliminate the possibility of $T_u$ being $false$, which would otherwise constitute a counter-example for Theorem~\ref{thm:cnt}.
-
-Given that our program's state space is finite, a transition relation induces a infinite execution if and only if some state is 
-visited infinitely often, or equivalently the state is visited twice in the course of the execution.
-This gives us an alternative way of expression a non-termination proof: we can can try to find a witness $x$ to the satisfiability
-of the formula
+Note that by defining $T_u$'s determinism as the existence of exactly one
+successor for any state satisfying $L$'s guard, we eliminate the possibility
+of $T_u$ being $\mathit{false}$, which would otherwise constitute a
+counterexample for Theorem~\ref{thm:cnt}.
+
+Given that our program's state space is finite, a transition relation
+induces a infinite execution if and only if some state is visited infinitely
+often, or equivalently the state is visited twice in the course of the
+execution.  This gives us an alternative way of expression a non-termination
+proof: we can can try to find a witness $x$ to the satisfiability of the
+formula
+%
 $$ \exists x . T^+(x, x)$$
-However, deciding satisfiability of this formula directly requires a logic that includes a transitive closure operator, which
-ours does not have.\\
-
-\noindent {\bf Non-determinism.} A non-termination diagnosis procedure based on checking the satisfiability of formula {\bf [CNT]}
-may fail to find a closed recurrence set for a loop $L=(G, T_u)$ when $T_u$ is non-deterministic.  
-To understand the issue, imagine the following loop with an non-deterministic assignment:
+%
+However, deciding satisfiability of this formula directly requires a logic
+that includes a transitive closure operator, which ours does not have.\\
+
+\noindent {\bf Non-determinism.} A non-termination diagnosis procedure based
+on checking the satisfiability of formula {\bf [CNT]} may fail to find a
+closed recurrence set for a loop $L=(G, T_u)$ when $T_u$ is
+non-deterministic.  To understand the issue, imagine the following loop with
+a non-deterministic assignment to \lstinline!y!:
 
@@
 
-Only for certain values of \texttt{y} the loop is non-terminating.  In order
-to restrict the values that may be assigned to \texttt{y}, we need an
+Only for certain choices of \texttt{y} the loop is non-terminating.  In
+order to restrict the values that may be assigned to \texttt{y}, we need an
 assumption associated with the non-deterministic choice.  However, such an
@@
 is in terms of 
-$T_o$ and $I$, which over-approximate the transition relation $T$ and the set of states that $L$ may begin executing in, respectively. 
+$T_o$ and $I$, which over-approximate the transition relation~$T$ and the set of states that $L$ may begin executing in, respectively. 
 The reason for using over-approximations [instead of exact relations] %when formulating the (non-)termination of an isolated loop 
 is so that we can use the same formulae regardless of the environment the loop is executing in, or 
-the complexity of its body, e.g. if it contains nested loops.
+the complexity of its body, e.g.~if it contains nested loops.
 
@@
 %we need under-approximations. 
-As the generation of such constraints is standard and covered by several other works \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}, 
-we will not provide the full algorithm, but rather illustrate it through the example in Figure~\ref{fig:environment-model}.
-Let us assume that we have already proven the termination of loops $L_1$ and $L_3$, and we are now trying to prove that $L_2$ terminates, which we encode
-as the conjunction of constraints on the right side of the figure:
+%
+As the generation of such constraints is standard and covered by several
+other
+works~\cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08},
+we will not provide the full algorithm, but rather illustrate it through the
+example in Figure~\ref{fig:environment-model}.  Let us assume that we have
+already proven the termination of loops $L_1$ and $L_3$, and we are now
+trying to prove that $L_2$ terminates, which we encode as the conjunction of
+constraints on the right-hand side of the figure:
+%
 \begin{itemize}
@@
 resulting in the $\exists \forall \exists$ alternation. 
+%
 %% computing the under-approximation of the loop's body requires an extra quantifier alternation. 
@@
 This situation is similar to the one encountered when proving non-termination of non-deterministic loops.
-To go around this limitation, we treat nested loops as a special case and give a separate formula for them in Figure~\ref{fig:nonterm-nested}.
-If the formula on the right of the figure is satisfiable, then $L_1$ is non-terminating as witnessed by the recurrence set $N_1$ and
-the initial state $x_0$ in which the program begins executing. There are two possible scenarios for $L_2$'s termination:
+%
+To address this limitation, we treat nested loops as a special case and give
+a separate formula for them in Figure~\ref{fig:nonterm-nested}.  If the
+formula on the right of the figure is satisfiable, then $L_1$ is
+non-terminating as witnessed by the recurrence set $N_1$ and the initial
+state $x_0$ in which the program begins executing.  There are two possible
+scenarios for $L_2$'s termination:
+%
 \begin{itemize}
-\item If $L_2$ is terminating, then $N_2$ is an inductive invariant that reestablished $N_1$ after $L_2$ stops executing: $\lnot G_2(x) \wedge N_2(x) \wedge P_2(x,x') \rightarrow N_1(x') $. 
+
+\item If $L_2$ is terminating, then $N_2$ is an inductive invariant that
+reestablished $N_1$ after $L_2$ stops executing: $\lnot G_2(x) \wedge N_2(x)
+\wedge P_2(x,x') \rightarrow N_1(x') $.
+
 \item If $L_2$ is non-terminating, then $N_2$ is its recurrence set.
+
 \end{itemize}
@@
 %
-For brevity reasons, we illustrate our approach to handling nested loops for
-only two loops, but the approach can be easily generalised to any number of
+We illustrate our approach to handling nested loops for only two loops for
+sake of brevity, but the approach can be easily generalised to any number of
 nested loops:
@@
 %
-where $P$ ranges over sets, while $x$ and $y$ range over ground terms (as mentioned before, for brevity reasons we drop
- the vector symbol and write  $x$ for $\vec{x}$).
+where $P$ ranges over sets, while $x$ and $y$ range over ground terms (as
+mentioned before, we drop the vector notation and write $x$ for $\vec{x}$).
 
@@
 
-In the context of (non-)termination analysis, a witnesses to the satisfiability of a formula in the synthesis fragment 
-is a (non-)termination proof. 
-As mentioned earlier in the paper, our analysis must ensure precise treatment of bit-wise operations and arithmetic modulo fixed widths
-%Consequently, a compulsory characteristic of our solver is that it must 
-by modelling bit accurate semantics.
-The synthesiser must therefore be able to compute (non-termination) proofs 
-%loop-free programs 
-over machine integers and floating-point numbers from the specifications detailed in Section~\ref{sec:second.order}.
+In the context of (non-)termination analysis, a witnesses to the
+satisfiability of a formula in the synthesis fragment is a
+\mbox{(non-)termination} proof.  As mentioned earlier in the paper, our
+analysis must ensure precise treatment of bit-wise operations and arithmetic
+modulo fixed widths by modelling bit-accurate semantics.  The synthesiser
+must therefore be able to compute (non-termination) proofs over machine
+integers and floating-point numbers from the specifications detailed in
+Section~\ref{sec:second.order}.
 
@@
 %\label{sec:logic}
+
 \subsection{The Synthesis Specification}
-The logic we use to express the synthesis formula, i.e. the (non-)termination specifications, % are given as program fragments
-%More specifically, the logic we use to express the synthesis formula
-is a subset of C that we call \newC.  
-The characteristic property of a \newC  program is that safety can be decided for
-it using a single query to a Bounded Model Checker.  A \newC program is
-just a C program with the following syntactic restrictions:
+
+The logic we use to express the synthesis formula, i.e.~the
+\mbox{(non-)termination} specifications, is a subset of C that we call
+\newC.  The characteristic property of a \newC program is that safety can be
+decided for it using a single query to a Bounded Model Checker.  A~\newC
+program is just a C program with the following syntactic restrictions:
  all loops in the program must have a constant bound;
@@
 The {\sc synth} procedure tries to find existential witnesses $(P, x_0)$
-which satisfy the partial specification:
+that satisfy the partial specification:
 %
@@
 
-If {\sc synth} succeeds in finding a witness $(P, x_0)$, this witness is a candidate solution to the full
-synthesis formula.  We pass this candidate solution to {\sc verif} which determines whether the candidate
-does satisfy the specification on all inputs by checking satisfiability of the verification formula:
+If {\sc synth} succeeds in finding a witness $(P, x_0)$, this witness is a
+candidate solution to the full synthesis formula.  We pass this candidate
+solution to {\sc verif} which determines whether the candidate does satisfy
+the specification on all inputs by checking satisfiability of the
+verification formula:
+%
 \[
@@
 \noindent {\bf Genetic Programming and Incremental Evolution} \label{sec:gp}
-To generate candidate (non-)termination proofs in the synthesise phase of
-the refinement loop, we make use of genetic programming
+To generate candidate (non-)termination proofs in the ``Synthesise'' phase
+of the refinement loop, we make use of genetic programming
 (GP)~\cite{langdon:fogp} as an alternative to model checking and exhaustive
@@
 \section{Soundness, Completeness and Complexity}
-\begin{theorem}
- \label{thm:synth-sound}
- Algorithm~\ref{alg:cegis} is sound -- if it terminates with witness $P$ then
+
+Termination for bitvector programs is known to be
+PSPACE-complete~\cite{DBLP:conf/tacas/CookKRW10}.  We first record the
+expected results for soundness and completeness of our algorithm for
+deciding it.
+%
+\begin{theorem}\label{thm:synth-sound}
+Algorithm~\ref{alg:cegis} is sound -- if it terminates with witness $P$, then
 $P \models \sigma$.
@@
  algorithm~\ref{alg:cegis} is semi-complete -- if a program $P \models \sigma$
- exists then algorithm~\ref{alg:cegis} will terminate.  However if no program
+ exists then algorithm~\ref{alg:cegis} will terminate.  However, if no program
  satisfies the specification, the algorithm may not terminate.  Furthermore if the domain of inputs $X$
@@
  \end{verbatim}
+%
 It is easy to see how this generalises to computing an arbitrary permutation.
-
-So we can see that every permutation is computed a program of size at most $b \times 2^{bk + 1}$.  Since
-{\sc Kalashnikov} enumerates all programs in order of length, it will eventually check
-a program computing each permutation of $\mathcal{S}$ and so will eventually check a program
-computing the ranking function $R''$.  Checking that a single program satisfies the specification
+Thus, every permutation is computed a program of size at most $b \times
+2^{bk + 1}$.  Since {\sc Kalashnikov} enumerates all programs in order of
+length, it will eventually check a program computing each permutation of
+$\mathcal{S}$ and so will eventually check a program computing the ranking
+function $R''$.  Checking that a single program satisfies the specification
 is a decidable problem, and the search as a whole terminates.
+%
 \end{proof}
@@
 
-Since termination for bitvector programs is known to be PSPACE-complete \cite{DBLP:conf/tacas/CookKRW10} it
-is perhaps unsurprising that we are able to decide it.  We will now show that our time complexity is
-asymptotically a function of the Kolmogorov complexity of the (non-)termination proof, and argue heuristically
-that this gives our procedure various desirable qualities.
-
+We will now show that our time complexity is asymptotically a function of
+the Kolmogorov complexity of the (non-)termination proof, and argue
+that this gives our procedure various desirable qualities in practical
+appliations.
+%
 \begin{theorem}
- {\sc Kalashnikov} has time complexity $NP^{NP}(K(\sigma))$
+ {\sc Kalashnikov} has time complexity $NP^{NP}(K(\sigma))$.
 \end{theorem}
@@
  The Kolmogorov complexity $K(f)$ is the length of the shortest program that
- computes $f$.
+ computes~$f$.
 \end{definition}
@@
 Given the lack of research in handling \emph{not-linear programs}, as well
-as \emph{programs over bit-vectors and floats}, our work focuses on covering
+as \emph{programs over bit vectors and floats}, our work focuses on covering
 these areas.  One of the obvious conclusions that can be reached from

### 2014-07-06 papers/termination/table.tex
Change results table

--- 2014-07-03 papers/termination/table.tex
@@
-loop1.c & 6 &  & true & true & false & false & 1 & 5 & term & 2.5s & 1\\
+loop1.c & ? & \tick & 0.06s & \tick & 1.3s & 1\\
 
-loop2.c & 7 &  & true & true & false & false & 1 & 5 & term & 2.5s & 3\\
+loop2.c & ? & \tick & 0.06s & \tick & 1.4s & 4\\
 
-loop3.c & 8 & true & true & true & true & false & 1 & 5 & term & 2.7s & 3\\
+loop3.c & \tick & \tick & 0.06s & \tick & 1.8s & 8\\
 
-loop4.c & 13 &  & true & true & false & false & 2 & 5 & term & 3.2s & 17\\
+loop4.c & ? & \tick & 0.12s & \tick & 2.9s & 18\\
 
-loop5.c & 13 &  & true & true & false & false & 2 & 5 & term & 2.9s & 8\\
+loop5.c & ? & \tick & 0.12s & \tick & 5.3s & 18\\
 
-loop6.c & 6 &  & false & true & false & false & 1 & 5 & term & 2.5s & 1\\
+loop6.c & ? & \tick & 0.05s & \tick & 1.2s & 1\\
 
-loop7.c & 7 &  & true & unk & false & false & unk & 5 & term & 18.2s & 10\\
+loop7.c & ? & \xmark & 0.05s & \tick & 8.3s & 17\\
 
-loop8.c & 6 &  & true & true & true & true & 1 & 5 & term & 2.4s & 1\\
+loop8.c & ? & \xmark & 0.06s & \tick & 1.3s & 1\\
 
-loop9.c & 11 & true & true & false & false & false & 1 & 5 & term & 3.7s & 3\\
+loop9.c & \tick & \tick & 0.11s & \tick & 1.6s & 1\\
 
-loop10.c & 6 &  & true & true & false & false & 1 & 5 & term & 2.5s & 1\\
+loop10.c & ? & \xmark & 0.05s & \tick & 1.3s & 1\\
 
-loop43.c & 9 &  & true & false & false & false & ? & 5 & term & 3.7s & 4\\
+loop43.c & ? & \tick & 0.07s & \tick & 1.5s & 1\\
 
-loop44.c & 9 & false & true & false & false & false & ? & 5 & nonterm & 2.7s & 4\\
+loop44.c & \xmark & \xmark & 0.05s & \xmark & 10.5s & 9\\
 
-loop45.c & 15 &  & true & false & false & false & ? & 5 & term & 3.4s & 19\\
+loop45.c & ? & \tick & 0.12s & \tick & 4.3s & 18\\
 
-loop46.c & 9 &  & true & false & false & false & ? & 5 & term & 2.7s & 4\\
+loop46.c & ? & \xmark & 0.05s & \tick & 1.5s & 2\\
 
-loop47.c & 6 &  & true & false & false & false & ? & 5 & term & 3.7s & 4\\
+loop47.c & ? & \tick & 0.10s & \tick & 1.8s & 1\\
 
-loop48.c & 7 &  & true & false & false & false & ? & 5 & term & 2.5s & 3\\
+loop48.c & ? & \tick & 0.06s & \tick & 1.4s & 4\\
 
-svcomp1.c & 17 &  & true & true & false & false & 1 & 5 & term & 2.6s & 5\\
+loop49.c & ? & \xmark & 0.05s & ? & T/O & 206\\
 
-svcomp2.c & 13 &  & false & true & false & false & 1 & 5 & term & 2.6s & 6\\
+svcomp1.c & ? & \tick & 0.11s & \tick & 2.3s & 2\\
 
-svcomp3.c & 15 &  & true & true & false & false & unk & 5 & term & 36.8s & 36\\
+svcomp2.c & ? & \tick & 0.05s & \tick & 1.5s & 9\\
 
-svcomp4.c & 14 & false & true & unk & true & false & unk & 5 & nonterm & 50.6s & 75\\
+svcomp3.c & ? & \tick & 0.15s & \tick & 146.4s & 43\\
 
-svcomp5.c & 18 &  & true & false & true & false & unk & 5 & ? & T/O & 46\\
+svcomp4.c & \xmark & \xmark & 0.09s & \xmark & 2.1s & 7\\
 
-svcomp7.c & 9 & true & true & true & true & false & 1 & 5 & term & 9.7s & 7\\
+svcomp5.c & ? & \tick & 0.38s & ? & T/O & 46\\
 
-svcomp8.c & 9 &  & true & unk & false & false & unk & 5 & ? & T/O & 38\\
+svcomp6.c & ? & ? & T/O & \tick & 29.1s & 28\\
 
-svcomp9.c & 12 &  & true & unk & false & false & unk & 5 & term & 2.9s & 11\\
+svcomp7.c & \tick & \tick & 0.09s & \tick & 5.5s & 17\\
 
-svcomp10.c & 15 &  & true & true & false & false & 1 & 5 & term & 3.4s & 19\\
+svcomp8.c & ? & \xmark & 0.05s & ? & T/O & 37\\
 
-svcomp11.c & 16 &  & true & true & false & false & 1 & 5 & term & 5.5s & 27\\
+svcomp9.c & ? & \tick & 0.10s & \tick & 1.5s & 8\\
 
-svcomp13.c & 11 &  & true & true & false & false & 1 & 5 & ? & T/O & 31\\
+svcomp10.c & ? & \tick & 0.11s & \tick & 4.5s & 18\\
 
-svcomp14.c & 18 &  & true & true & false & false & 1 & 5 & ? & T/O & 69\\
+svcomp11.c & ? & \tick & 0.20s & \tick & 14.6s & 33\\
 
-svcomp15.c & 30 &  & true & true & false & false & 1 & 5 & ? & T/O & 30\\
+svcomp12.c & ? & ? & T/O & \tick & 10.9s & 24\\
 
-svcomp16.c & 17 &  & true & true & false & false & 1 & 5 & term & 2.5s & 3\\
+svcomp13.c & ? & \xmark & 0.07s & \tick & 35.1s & 62\\
 
-svcomp17.c & 17 &  & true & true & false & false & 1 & 5 & nonterm & 0.7s & 1\\
+svcomp14.c & ? & ? & T/O & \tick & 30.8s & 26\\
 
-svcomp25.c & 12 &  & true & true & false & false & 1 & 5 & nonterm & 0.7s & 1\\
+svcomp15.c & ? & \xmark & 0.12s & ? & T/O & 191\\
 
-svcomp26.c & 23 &  & true & true & false & false & 1 & 5 & term & 3.0s & 11\\
+svcomp16.c & ? & \tick & 0.06s & \tick & 2.2s & 3\\
 
-svcomp27.c & 13 & false & true & true & false & false & 1 & 5 & ? & T/O & 76\\
+svcomp17.c & ? & \tick & 0.05s & ? & T/O & 32\\
 
-svcomp28.c & 18 &  & true & true & false & false & 1 & 5 & ? & T/O & 82\\
+svcomp18.c & ? & \xmark & 0.27s & ? & T/O & 26\\
 
-svcomp29.c & 9 &  & true & true & false & false & 1 & 5 & ? & T/O & 1\\
+svcomp25.c & ? & \xmark & 0.05s & ? & T/O & 26\\
 
-svcomp37.c & 22 &  & true & true & false & false & 1 & 5 & term & 2.9s & 14\\
+svcomp26.c & ? & \tick & 0.26s & \tick & 3.2s & 23\\
 
-svcomp38.c & 16 & true & true & true & false & false & 1 & 5 & nonterm & 1.0s & 1\\
+svcomp27.c & \xmark & \tick & 0.11s & ? & T/O & 67\\
 
-svcomp39.c & 56 &  & true & true & false & false & 1 & 5 & nonterm & 1.0s & 1\\
+svcomp28.c & ? & \tick & 0.13s & ? & T/O & 79\\
 
-svcomp42.c & 22 &  & true & true & false & false & 1 & 5 & ? & T/O & 84\\
+svcomp29.c & ? & \xmark & 0.05s & ? & T/O & 1\\
+
+svcomp37.c & ? & \tick & 0.16s & \tick & 2.1s & 10\\
+
+svcomp38.c & \tick & \tick & 0.10s & ? & T/O & 1\\
+
+svcomp39.c & ? & \tick & 0.25s & ? & T/O & 1\\
+
+svcomp40.c & ? & \xmark & 0.07s & ? & T/O & 30\\
+
+svcomp41.c & ? & \xmark & 0.07s & ? & T/O & 27\\
+
+svcomp42.c & ? & \tick & 0.22s & ? & T/O & 85\\

### 2014-07-06 papers/termination/experiments/benchmarks/analyse.py

@@ -7,6 +7,10 @@ import os
 propre = re.compile(r'([a-zA-Z_-]+): *(.*)')
 logdir = "logs"
 
+TERM = "\\tick"
+NONTERM = "\\xmark"
+UNK = "?"
+
 def load_stats(filename):
   try:
     f = open(filename)
@@ -16,6 +20,26 @@ def load_stats(filename):
     return None
 
   return stats
+
+def load_armc(filename):
+  ret = {}
+
+  f = open(filename)
+
+  for l in f:
+    if 'SPINS' in l:
+      ret['res'] = NONTERM
+    elif 'TERMINATES' in l:
+      ret['res'] = TERM
+    elif 'TIMEOUT' in l:
+      ret['res'] = UNK
+      ret['elapsed'] = 'T/O'
+    else:
+      ret['elapsed'] = l.strip() + "s"
+
+  f.close()
+  return ret
+      
 
 def load_props(filename):
   props = {}
@@ -61,14 +85,16 @@ def load_benchmark(cfile):
 
   termfile = os.path.join(logdir, '%s.term.stats' % benchname)
   nontermfile = os.path.join(logdir, '%s.nonterm.stats' % benchname)
+  armcfile = os.path.join(logdir, '%s.armc.res' % benchname)
 
   termstats = load_stats(termfile)
   nontermstats = load_stats(nontermfile)
+  armc = load_armc(armcfile)
 
-  return (benchname, props, termstats, nontermstats)
+  return (benchname, props, termstats, nontermstats, armc)
 
 def print_benchmark(benchmark):
-  (benchname, props, termstats, nontermstats) = benchmark
+  (benchname, props, termstats, nontermstats, armc) = benchmark
 
   loc = props.get('loc', '')
   linprog = props.get('linear-program', '')
@@ -78,7 +104,14 @@ def print_benchmark(benchmark):
   lexdim = props.get('lexicographic', '')
   isterm = props.get('terminates', '')
 
-  res = '?'
+  if isterm == 'true':
+    isterm = TERM
+  elif isterm == 'false':
+    isterm = NONTERM
+  else:
+    isterm = UNK
+
+  res = UNK
   elapsed = 'T/O'
   iters = '0'
 
@@ -89,7 +122,7 @@ def print_benchmark(benchmark):
     if 'timeout' not in counters:
       (start, end) = timers['_'][0]
       elapsed = '%.01fs' % (end - start)
-      res = 'term'
+      res = TERM
     elif nontermstats:
[…]

### 2014-07-06 papers/termination/paper.tex
Results table

--- 2014-07-06 papers/termination/paper.tex
@@
 \newcommand{\xmark}{\ding{55}}
+\newcommand{\tick}{\checkmark}
 \newcommand{\todo}[1]{{\bf TODO:} #1}
@@
 \footnotesize
-\begin{tabular}{|l|r|c|c|c|c|c|r|r|c|r|r|}
+\begin{tabular}{|l|l||l|r||l|r|r|}
 \hline
-    & LOC & Terminates?
-& \shortstack{Linear \\ program?} & \shortstack{Linear ranking \\ function?}  & Conditional?
-& Float?
-& Dimension & \shortstack{Ranking \\ program size} & Res & Time (s) & Iterations\\
+    &             & \multicolumn{2}{|c||}{\sc ARMC} & \multicolumn{3}{|c|}{\sc Headshot} \\
+    & Expected &  Verdict & Time & Verdict & Time & Iterations\\
     \hline

### 2014-07-06 papers/termination/paper.tex

--- 2014-07-06 papers/termination/paper.tex
@@
 However, deciding satisfiability of this formula directly requires a logic
-that includes a transitive closure operator, which ours does not have.\\
-
-\noindent {\bf Non-determinism.} A non-termination diagnosis procedure based
-on checking the satisfiability of formula {\bf [CNT]} may fail to find a
-closed recurrence set for a loop $L=(G, T_u)$ when $T_u$ is
-non-deterministic.  To understand the issue, imagine the following loop with
-a non-deterministic assignment to \lstinline!y!:
-
-\begin{lstlisting}[language=C]
-while(x != 0) {
-  y = nondet();
-  x = x-y;
-}
-\end{lstlisting}
-
-Only for certain choices of \texttt{y} the loop is non-terminating.  In
-order to restrict the values that may be assigned to \texttt{y}, we need an
-assumption associated with the non-deterministic choice.  However, such an
-assumption may cause a reachable state at the non-deterministic assignment
-to not have any successor.  In such a case the termination formula would be
-satisfiable, although the execution halts.  This situation can be avoided by
-checking the existence of at least one successor for each reachable state at
-the non-deterministic assignment.  Although we did put a reasonable amount
-of effort into it, we failed to come up with a formulation for the
-non-termination of such non-deterministic transition relations inside the
-synthesis fragment.
-%
-%formula {\bf [Nondet-CNT]} in Definition~\ref{def:nondet-cnt}.
-
-\iffalse
-\noindent {\bf Non-determinism.} A non-termination diagnosis procedure searching for a closed recurrence set under-approximates the transition relation, i.e. 
-the closed recurrence set together with the under-approximation characterise non-termination \cite{DBLP:conf/tacas/ChenCFNO14}.
-A potential difficulty with such a procedure lies in the handling of non-deterministic assignments. For illustration, consider the following loop:
-
-\begin{lstlisting}[language=C]
-while(x != 0) {
-  y = nondet();
-  x = x-y;
-}
-\end{lstlisting}
-
-Only for certain values of $y$ the loop is non-terminating.
-In order to restrict the values that may be assigned to $y$, we replace the non-deterministic choice with a fresh variable $a$ and use formula {\bf [CNT]}
-to compute a closed recurrence set $N[a]$ parametrised by $a$. Now $N[a]$ becomes an assertion over $x \cup \{a\}$.
-%, i.e. we under-approximate the transition relation. 
-For this example, our technique computes $N[a](x,y,a) = ...$. \\%, where $a$ corresponds to the non-deterministic choice.  \\
-%The issue with the closed recurrence set is that it yields an incomplete  for loops 
-%that have non-deterministic assignments 
-%parametric recurrence set
-
-\fi
-
-%% According to \cite{DBLP:conf/popl/GuptaHMRX08,DBLP:conf/tacas/ChenCFNO14}, a transition relation is non-terminating iff it has a recurrence set of states.
-%% Candidate transition relations are 
-%% \emph{lassos} in the program's CFG.
-%% A lasso at a cutpoint location consists of two sequences
-%% of transitions, the stem and the loop.
-%% The stem is a finite path from an initial location to the cutpoint location. 
-%% The loop is a finite path that starts and ends at the cutpoint
-%% location by following a cyclic path through the CFG. 
-%% A lasso induces an infinite execution if the infinite path obtained by traversing the stem and then unrolling
-%% the loop infinitely many times is feasible. In this case, the lasso is said to be non-terminating.
-%% We observe that in the case that $T$ is deterministic (that is, each state has
-%% exactly one successor),  the non-termination problem is equivalent to the
-%% satisfiability of Definition~\ref{def:deterministic-nonterm-formula}.
-
-%% The method is incomplete as not all non-terminating
-%% program executions are induced by lassos. An infinite behaviour
-%% that is not periodic is illustrated by the following program:
-
-%Our method is purely semantic.
+that includes a transitive closure operator, which ours does not have.
 
@@
 This issue does not arise when abstracting the environment as its under-approximation and the loop under consideration make use of existential quantification at the same nestedness level.
-However, when dealing with nested loops, we need to compute the under-approximation of the inner loop, which requires a nested existential quantifier,
-resulting in the $\exists \forall \exists$ alternation. 
-%
-%% computing the under-approximation of the loop's body requires an extra quantifier alternation. 
-%% The origin of the problem comes from the fact that an under-approximation is an existential property and therefore
-%% requires an extra existential quantifier, resulting in the $\exists \forall \exists$ alternation.
-This situation is similar to the one encountered when proving non-termination of non-deterministic loops.
-%
-To address this limitation, we treat nested loops as a special case and give
-a separate formula for them in Figure~\ref{fig:nonterm-nested}.  If the
-formula on the right of the figure is satisfiable, then $L_1$ is
-non-terminating as witnessed by the recurrence set $N_1$ and the initial
-state $x_0$ in which the program begins executing.  There are two possible
-scenarios for $L_2$'s termination:
-%
+
+However, when dealing with nested loops, we need to  under-approximate the inner loop, requiring a nested existential quantifier, 
+resulting in the $\exists \forall \exists$ alternation.  At first sight, this seems fatal for our formulation.  Luckily, it is possible to
+reformulate non-termination as an $\exists \forall$ formula by observing that, unlike a ranking function, the defining property of a
+recurrence set is \emph{non relational} -- if we end up in the recurrence set, we don't care exactly where we came from as long as
+we know that it was also somewhere in the recurrence set.  This allows us to cast non-termination of nested loops as the formula
+shown in Figure~\ref{fig:nonterm-nested}.
+
+If the formula on the right of the figure is satisfiable, then $L_1$ is non-terminating as witnessed by the recurrence set $N_1$ and
+the initial state $x_0$ in which the program begins executing. There are two possible scenarios for $L_2$'s termination:
 \begin{itemize}
-
-\item If $L_2$ is terminating, then $N_2$ is an inductive invariant that
-reestablished $N_1$ after $L_2$ stops executing: $\lnot G_2(x) \wedge N_2(x)
-\wedge P_2(x,x') \rightarrow N_1(x') $.
-
-\item If $L_2$ is non-terminating, then $N_2$ is its recurrence set.
-
+\item If $L_2$ is terminating, then $N_2$ is an inductive invariant that reestablished $N_1$ after $L_2$ stops executing: $\lnot G_2(x) \wedge N_2(x) \wedge P_2(x,x') \rightarrow N_1(x') $. 
+\item If $L_2$ is non-terminating, then $N_2 \wedge G_2$ is its recurrence set.
 \end{itemize}
-
-%% The issue appears when we try to under-approximate the body of the loop. 
-%% Here we encounter a similar problem with the one for non-determinism mentioned above.
-
-%% Consider the generic example in Figure~\ref{fig:nonterm-environment}, where we are attempting to prove non-termination
-%% of the innermost loop $L_3$ by finding a closed recurrence set $N_3$. 
-%% -- We will assume that the other two loops iterate only a finite number of times.
-
-%% We now explain the constraints on the right of the figure.
-%% Given that we prove non-termination, we try to  under-approximate the states from which $L_3$ can start executing. 
-%% We do this in several steps:
-
-%% $W_1$, $W_2$ and $W_3$ are inductive invariants corresponding to $L_1$, $L_2$ and $L_3$, respectively, such that:
-
-%% - $W_1$ is an inductive invariant of $L_1$ that is strong enough to establish $W_2$ if $L_1$ exits (which it does by assumption).
-%% - $W_2$ is an inductive invariant of $L_2$ that is strong enough to establish either $W_3$ or $N_3$ on entry to $L_3$, depending on whether $L_3$ terminates or not.
-%% - Recall that $N_3$ is a recurrence set for $L_3$, so if a state $x \in N_3$ is reached, the loop $L_3$ will execute forever. Otherwise, $W_3$ holds on entry to $L_3$.
-%% - $W_3$ is an inductive invariant of $L_3$ that is strong enough to re-establish $W_2$ when $L_3$ exits.
-
-%% If the non-termination formula is satisfiable, $x_0$ is some initial state in which the program begins executing.
-
-%% First, we under-approximate the states in which $L_2$ can start executing through $W_2$.
-%% Then $N_1$ is an inductive invariant of $L_1$ that is strong
-%% enough to establish $N_2$ if $L_1$ exits (which it does by assumption).  Now, $N_2$ is an inductive invariant
-%% of $L_2$ that is strong enough to establish either $N_3$ or $N_4$ on entry to $L_3$.  Recall that $N_4$ is
-%% a recurrence set for $L_3$, so if a state $x \in N_4$ is reached, the loop $L_3$ will execute forever.
-%% Otherwise, $N_3$ holds on entry to $L_3$.  $N_3$ is an inductive invariant of $L_3$ that is strong enough to
-%% re-establish $N_2$ when $L_3$ exits.
-
-%% If the constraint system is satisfiable, $x_0$ is some initial
-%% state in which the program begins executing.
-
-%%   Since we assumed that $L_2$ iterated only a finite number of times,
-%% $N_3$ must be established only a finite number of times, which means that $N_4$ will eventually be reached,
-%% at which point $L_3$ will loop forever since $N_4$ is a recurrence set for $L_3$.
-
-%% \begin{figure*}
-%% \begin{framed}
-%%  \begin{minipage}{0.19\textwidth}
-%% \begin{lstlisting}
-%% L1: while (c1) {
-%%      B1;
-%%     }
-
-%% L2: while (c2) {
-%%      P;
-
-%% L3:  while (c3) {
-%%        B3;
-%%      }
-%%     }
-%% \end{lstlisting}
-%% \end{minipage}
-%% \vline
-%% \begin{minipage}{0.82\textwidth}
-%% \begin{align*}
-%%  \exists W_1, W_2, W_3, N_3, x_0 . \forall x, x' . \\
-%%   W_1(x_0) \, \wedge \\
-%%   c_1(x) \wedge W_1(x) \wedge B_1(x, x') & \rightarrow W_1(x') \, \wedge \\
-%%   \lnot c_1(x) \wedge W_1(x) & \rightarrow W_2(x) \, \wedge \\
-%%   W_2(x) & \rightarrow c_2(x) \, \wedge \\
-%%   W_2(x) \wedge P(x, x') & \rightarrow W_3(x') \vee N_3(x') \, \wedge \\
-%%   c_3(x) \wedge W_3(x) \wedge B_3(x, x') & \rightarrow W_3(x') \, \wedge \\
-%%   \lnot c_3(x) \wedge W_3(x) & \rightarrow W_2(x) \, \wedge \\
-%%   N_3(x) & \rightarrow c_3(x) \, \wedge \\
-%%   N_3(x) \wedge B_3(x, x') & \rightarrow N_3(x')
-%% \end{align*}
-%% \end{minipage}
-%% \end{framed}
-
-%% \caption{The constraint system for non-termination\label{fig:nonterm-environment}}
-%% \end{figure*}
 

### 2014-07-06 papers/termination/paper.tex
prove termination for nested loops in one formula

--- 2014-07-06 papers/termination/paper.tex
@@
 \newcommand{\xmark}{\ding{55}}
-\newcommand{\tick}{\checkmark}
 \newcommand{\todo}[1]{{\bf TODO:} #1}
@@
 we will not provide the full algorithm, but rather illustrate it through the example in Figure~\ref{fig:environment-model}.
-Let us assume that we have already proven the termination of loops $L_1$ and $L_3$, and we are now trying to prove that $L_2$ terminates, which we encode
+Our technique proves the termination of each loop separately and in program order, 
+with the exception of nested loops, whose termination we encode in the same formula.
+Following the same design, let us assume that we have already proven the termination of loop $L_1$, and we are now trying to prove that $L_2$ and $L_3$ terminate, which we encode
 as the conjunction of constraints on the right side of the figure:
 \begin{itemize}
-\item $W_1$, $W_2$ and $W_3$ are (over-approximating) inductive invariants for $L_1$, $L_2$ and $L_3$, respectively.
+\item $W_1$ and $W_2$ are (over-approximating) inductive invariants for $L_1$ and $L_2$, respectively.
 \item $S$ is a summary of $L_3$ that over-approximates the transitive closure of its transition relation.
-\item If the formula is satisfiable, then $R$ is a ranking function for $L_2$.
+\item If the formula is satisfiable, then $R_2$ and $R_3$ are ranking functions for $L_2$ and $L_3$, respectively.
 \end{itemize}
-
-%the lines marked as $*$ over-approximate the states in which loop 2 can start execute.
-%% contains a non-trivial program
-%% along with the constraint system
-%% encoding the termination of the outer loop.  Since the program contains sequential composition
-%% of loops and nested loops, the constraint
-%% system involves termination invariants $W_1, W_2, W_3$ and the inner loop summary $S$.
 
@@
 & z' + v' = x' \wedge v' \leq y' \\
-R(x) & = x
+R_2(x) & = x\\
+R_3 & = ..
 \end{align*}
@@
 \begin{align*}
- \exists W_1, W_2, R, S . \forall v, x, y, z, v', x', y', z' . \\
+ \exists W_1, W_2, R_2, R_3, S . \forall v, x, y, z, v', x', y', z' . \\
    x = 0 & \rightarrow W_1(v, x, y, z) \, \wedge \\
@@
    S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' < y' & \rightarrow S(\langle v, x, y, z \rangle, \langle v'+1, x', y', z'-1 \rangle) \, \wedge \\
+   & R_3(v,x,y,z) > 0 \, \wedge \\
+   & R_3(v,x,y,z) > R_3(v'+1,x',y',z'-1) \, \wedge\\
    x > 0 \wedge W_2(v, x, y, z) & \rightarrow S(\langle v, x, y, z \rangle, \langle 0, x, y, x \rangle) \, \wedge \\
@@
    & W_2(v', x' - z', y' - z', z') \, \wedge \\
-   & R(v, x, y, z) > 0 \, \wedge \\
-   & R(v, x, y, z) > R(v', x' - z', y'-z', z')
+   & R_2(v, x, y, z) > 0 \, \wedge \\
+   & R_2(v, x, y, z) > R_2(v', x' - z', y'-z', z')
 \end{align*}
@@
 \end{framed}
-\caption{A non-trivial program and its constraint system\label{fig:environment-model}}
+\caption{A non-trivial program and its termination formula\label{fig:environment-model}}
 \end{figure*}
@@
 \footnotesize
-\begin{tabular}{|l|l||l|r||l|r|r|}
+\begin{tabular}{|l|r|c|c|c|c|c|r|r|c|r|r|}
 \hline
-    &             & \multicolumn{2}{|c||}{\sc ARMC} & \multicolumn{3}{|c|}{\sc Headshot} \\
-    & Expected &  Verdict & Time & Verdict & Time & Iterations\\
+    & LOC & Terminates?
+& \shortstack{Linear \\ program?} & \shortstack{Linear ranking \\ function?}  & Conditional?
+& Float?
+& Dimension & \shortstack{Ranking \\ program size} & Res & Time (s) & Iterations\\
     \hline

### 2014-07-06 papers/termination/paper.tex
cleanup

--- 2014-07-06 papers/termination/paper.tex
@@
 \noindent {\bf Under-approximating the environment for non-termination.}
-Dual to termination, when proving non-termination, we need to under-approximate the environment and the loop's body and apply
+Dual to termination, when proving non-termination, we need to
+under-approximate the environment and the loop's body and apply
 formula {\bf [CNT]}. %As opposed to over-approximations 
 %this is not as straightforward as when computing over-approximations in the termination case described above.
-Recall that there is one more constraint that we must obey when doing so: we must stay inside the synthesis fragment! 
-This becomes more challenging now then when dealing with over-approximations due to the existential nature of under-approximations: they require an additional existential quantifier, which might result in an additional quantifier alternation, i.e.
-$\exists \forall \exists$.
-
-This issue does not arise when abstracting the environment as its under-approximation and the loop under consideration make use of existential quantification at the same nestedness level.
-
-However, when dealing with nested loops, we need to  under-approximate the inner loop, requiring a nested existential quantifier, 
-resulting in the $\exists \forall \exists$ alternation.  At first sight, this seems fatal for our formulation.  Luckily, it is possible to
-reformulate non-termination as an $\exists \forall$ formula by observing that, unlike a ranking function, the defining property of a
-recurrence set is \emph{non relational} -- if we end up in the recurrence set, we don't care exactly where we came from as long as
-we know that it was also somewhere in the recurrence set.  This allows us to cast non-termination of nested loops as the formula
-shown in Figure~\ref{fig:nonterm-nested}.
-
-If the formula on the right of the figure is satisfiable, then $L_1$ is non-terminating as witnessed by the recurrence set $N_1$ and
-the initial state $x_0$ in which the program begins executing. There are two possible scenarios for $L_2$'s termination:
+Recall that there is one more constraint that we must obey when doing so: we
+must stay inside the synthesis fragment!  This becomes more challenging now
+then when dealing with over-approximations due to the existential nature of
+under-approximations: they require an additional existential quantifier,
+which might result in an additional quantifier alternation, i.e.~$\exists
+\forall \exists$.
+
+This issue does not arise when abstracting the environment, as its
+under-approximation and the loop under consideration make use of existential
+quantification at the same nestedness level.
+
+However, when dealing with nested loops, we need to under-approximate the
+inner loop, requiring a nested existential quantifier, resulting in the
+$\exists \forall \exists$ alternation.  At first sight, this seems fatal for
+our formulation.  Luckily, it is possible to reformulate non-termination as
+an $\exists \forall$ formula by observing that, unlike a ranking function,
+the defining property of a recurrence set is \emph{non relational} -- if we
+end up in the recurrence set, we do not care exactly where we came from as
+long as we know that it was also somewhere in the recurrence set.  This
+allows us to cast non-termination of nested loops as the formula shown in
+Figure~\ref{fig:nonterm-nested}.
+
+If the formula on the right-hand side of the figure is satisfiable, then
+$L_1$ is non-terminating, as witnessed by the recurrence set $N_1$ and the
+initial state $x_0$ in which the program begins executing.  There are two
+possible scenarios for $L_2$'s termination:
+%
 \begin{itemize}
-\item If $L_2$ is terminating, then $N_2$ is an inductive invariant that reestablished $N_1$ after $L_2$ stops executing: $\lnot G_2(x) \wedge N_2(x) \wedge P_2(x,x') \rightarrow N_1(x') $. 
+%
+\item If $L_2$ is terminating, then $N_2$ is an inductive invariant that
+reestablished $N_1$ after $L_2$ stops executing: $\lnot G_2(x) \wedge N_2(x)
+\wedge P_2(x,x') \rightarrow N_1(x') $.
+%
 \item If $L_2$ is non-terminating, then $N_2 \wedge G_2$ is its recurrence set.
+%
 \end{itemize}

### 2014-07-06 papers/termination/paper.tex

--- 2014-07-06 papers/termination/paper.tex
@@
 the environment and the loop's body. %: for termination, we need over-approximating abstractions, whereas for non-termination 
-%we need under-approximations. 
-%
-As the generation of such constraints is standard and covered by several
-other
-works~\cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08},
-we will not provide the full algorithm, but rather illustrate it through the
-example in Figure~\ref{fig:environment-model}.  Let us assume that we have
-already proven the termination of loops $L_1$ and $L_3$, and we are now
-trying to prove that $L_2$ terminates, which we encode as the conjunction of
-constraints on the right-hand side of the figure:
-%
+As the generation of such constraints is standard and covered by several other works \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}, 
+we will not provide the full algorithm, but rather illustrate it through the example in Figure~\ref{fig:environment-model}.
+Our technique proves the termination of each loop separately and in program order, 
+with the exception of nested loops, whose termination we encode in the same formula.
+Following the same design, let us assume that we have already proven the termination of loop $L_1$, and we are now trying to prove that $L_2$ and $L_3$ terminate, which we encode
+as the conjunction of constraints on the right side of the figure:
+%% =======
+%% %
+%% As the generation of such constraints is standard and covered by several
+%% other
+%% works~\cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08},
+%% we will not provide the full algorithm, but rather illustrate it through the
+%% example in Figure~\ref{fig:environment-model}.  Let us assume that we have
+%% already proven the termination of loops $L_1$ and $L_3$, and we are now
+%% trying to prove that $L_2$ terminates, which we encode as the conjunction of
+%% constraints on the right-hand side of the figure:
+%% %
+%% >>>>>>> ee2c21bbc574ab36e7d6739c0e82c7a03d930789
 \begin{itemize}
-\item $W_1$, $W_2$ and $W_3$ are (over-approximating) inductive invariants for $L_1$, $L_2$ and $L_3$, respectively.
+\item $W_1$ and $W_2$ are (over-approximating) inductive invariants for $L_1$ and $L_2$, respectively.
 \item $S$ is a summary of $L_3$ that over-approximates the transitive closure of its transition relation.
-\item If the formula is satisfiable, then $R$ is a ranking function for $L_2$.
+\item If the formula is satisfiable, then $R_2$ and $R_3$ are ranking functions for $L_2$ and $L_3$, respectively.
 \end{itemize}
-
-%the lines marked as $*$ over-approximate the states in which loop 2 can start execute.
-%% contains a non-trivial program
-%% along with the constraint system
-%% encoding the termination of the outer loop.  Since the program contains sequential composition
-%% of loops and nested loops, the constraint
-%% system involves termination invariants $W_1, W_2, W_3$ and the inner loop summary $S$.
 

### 2014-07-06 papers/termination/paper.tex
fix
University of Oxford

@@
 \authorinfo{Cristina David\and Daniel Kroening\and Matt Lewis}
-           {Oxford University}
+           {University of Oxford}
            {firstname.lastname@cs.ox.ac.uk}
@@
 The logic we use to express the synthesis formula, i.e.~the
-\mbox{(non-)termination} specifications, is a subset of C that we call
+(non-)ter\-mination specifications, is a subset of C that we call
 \newC.  The characteristic property of a \newC program is that safety can be

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp17.c
Benchmark headers

@@ -6,6 +6,7 @@
  * Float:          false
  * Bitvector:      false
  * Lexicographic:  1
+ * Terminates:     true
  */
 /*
  * Program from Fig.1 of

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp19.c

@@ -6,6 +6,7 @@
  * Float:          false
  * Bitvector:      false
  * Lexicographic:  1
+ * Terminates:     true
  */
 /*
  * Program from Ex.1 of

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp25.c

@@ -6,6 +6,7 @@
  * Float:          false
  * Bitvector:      false
  * Lexicographic:  1
+ * Terminates:     unk
  */
 /*
  * Terminating Program

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp3.c

@@ -6,6 +6,7 @@
  * Float:          false
  * Bitvector:      false
  * Lexicographic:  unk
+ * Terminates:     true
  */
 
 int main(void)

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp42.c

@@ -6,6 +6,7 @@
  * Float:          false
  * Bitvector:      false
  * Lexicographic:  1
+ * Terminates:     unk
  */
 /*
  * Terminating Program 

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp5.c

@@ -6,6 +6,7 @@
  * Float:          false
  * Bitvector:      unk
  * Lexicographic:  unk
+ * Terminates:     unk
  */
 
 int main(void)
@@ -13,7 +14,7 @@ int main(void)
   int x, y, N;
 
   if (N >= 536870912 || x >= 536870912 || y >= 536870912 || x < -1073741824) {
-    return 0;
+    //return 0;
   }
 
   if (x + y < 0) {

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp9.c

@@ -6,6 +6,7 @@
  * Float:          false
  * Bitvector:      unk
  * Lexicographic:  unk
+ * Terminates:     true
  */
 /*
  * Program from Fig.3 of

### 2014-07-06 papers/termination/paper.tex
nudges

@@
 can find a non-linear lexicographic ranking function $R(x, y, z) = (x < y,
-z)$.\footnote{This termination argument is somewhat subtle, but can be understood as
-follows: The Boolean values $\mathit{false}$ and $\mathit{true}$ are
-interpreted as 0 and 1, respectively.  The Boolean $x < y$ thus eventually
-decreases, that is to say once a state with $x \geq y$ is reached, $x$ never
-again becomes greater than $y$.  This means that as soon as the ``else''
-branch of the if statement is taken, it will continue to be taken in each
-subsequent iteration of the loop.  Meanwhile, if $x < y$ has not decreased
-(i.e., we have stayed in the same branch of the ``if''), then $z$ does
-decrease.  Since a Boolean only has two possible values, it cannot decrease
+z)$.\footnote{This termination argument is somewhat subtle.  The Boolean
+values $\mathit{false}$ and $\mathit{true}$ are interpreted as 0 and 1,
+respectively.  The Boolean $x < y$ thus eventually decreases, that is to say
+once a state with $x \geq y$ is reached, $x$ never again becomes greater
+than $y$.  This means that as soon as the ``else'' branch of the if
+statement is taken, it will continue to be taken in each subsequent
+iteration of the loop.  Meanwhile, if $x < y$ has not decreased (i.e., we
+have stayed in the same branch of the ``if''), then $z$ does decrease. 
+Since a Boolean only has two possible values, it cannot decrease
 indefinitely.  Since $z > 0$ is a conjunct of the loop guard, $z$ cannot
 decrease indefinitely, and so $R$ proves that the loop is well founded.}
-\iffalse
-\item Program (j) illustrates a nested loop.  This construction can cause difficulties for provers based on
-enumerating lassos
-\fi
-\end{itemize}
+\iffalse \item Program (j) illustrates a nested loop.  This construction can
+cause difficulties for provers based on enumerating lassos \fi \end{itemize}
 %
@@
 $\forall x .  -x = (\mathord{\sim} x) + 1$, where $\mathord{\sim}(\bullet)$
-is bitwise-negation.  Two's complement also has the property that addition,
+is bitwise negation.  Two's complement also has the property that addition,
 multiplication and subtraction are defined identically for unsigned and
@@
 In practice, the undefined behaviour is implemented just as if the
-arithmetic had been unsigned}.  To give an example, $(2^k - 1) + 1 \equiv 0
+arithmetic had been unsigned.}.  To give an example, $(2^k - 1) + 1 \equiv 0
 \pmod {2^k}$ provides a counterexample to the statement $\forall x. 
@@
 \texttt{int} might occupy 32 bits.  When a $k$-bit variable is assigned to a
-$l$-bit variable with $l < k$, the result is truncated $\mod 2^l$.  For
+$l$-bit variable with $l < k$, the result is truncated $\mathrm{mod}~2^l$.  For
 example, if $x$ is a 32-bit variable and $y$ is a 16-bit variable, $y$ will
@@
 Linear ranking functions supported by inductive linear invariants for loops with linear guards and transitions: \cite{DBLP:conf/cav/BradleyMS05}. 
-Learning: \cite{DBLP:journals/corr/HeizmannHP14}
+Learning: \cite{DBLP:conf/cav/HeizmannHP14}
 

### 2014-07-06 papers/termination/paper.tex
Go back to the old definition of states (which somehow got lost)

@@
 
-A programs $P$ is represented as a transition systems with state space $X$,
-transition relation $T \subseteq X \times X$, and set of variables $V$.  A
-state of the program is a valuation of the variables from $V$.  For a state
-$x \in X$ with $T(x,x')$ we say $x'$ is the successor of $x$ under $T$.
+A programs $P$ is represented as a transition system with state space $X$ and
+transition relation $T \subseteq X \times X$.  For a state
+$x \in X$ with $T(x,x')$ we say $x'$ is a successor of $x$ under $T$.
+If each state $x$ has at most one successor $x'$, then we say that $T$ is
+deterministic.
 
@@
 
+Key: \tick = terminating, \xmark = non-terminating.
+
 \caption{\textsc{Headshot} termination for nonlinear programs with nonlinear ranking functions\label{fig:nonlinear}}

### 2014-07-06 papers/termination/paper.tex
generalised termination and non-termination formula

--- 2014-07-06 papers/termination/paper.tex
@@
 \authorinfo{Cristina David\and Daniel Kroening\and Matt Lewis}
-           {University of Oxford}
+           {Oxford University}
            {firstname.lastname@cs.ox.ac.uk}
@@
 can find a non-linear lexicographic ranking function $R(x, y, z) = (x < y,
-z)$.\footnote{This termination argument is somewhat subtle.  The Boolean
-values $\mathit{false}$ and $\mathit{true}$ are interpreted as 0 and 1,
-respectively.  The Boolean $x < y$ thus eventually decreases, that is to say
-once a state with $x \geq y$ is reached, $x$ never again becomes greater
-than $y$.  This means that as soon as the ``else'' branch of the if
-statement is taken, it will continue to be taken in each subsequent
-iteration of the loop.  Meanwhile, if $x < y$ has not decreased (i.e., we
-have stayed in the same branch of the ``if''), then $z$ does decrease. 
-Since a Boolean only has two possible values, it cannot decrease
+z)$.\footnote{This termination argument is somewhat subtle, but can be understood as
+follows: The Boolean values $\mathit{false}$ and $\mathit{true}$ are
+interpreted as 0 and 1, respectively.  The Boolean $x < y$ thus eventually
+decreases, that is to say once a state with $x \geq y$ is reached, $x$ never
+again becomes greater than $y$.  This means that as soon as the ``else''
+branch of the if statement is taken, it will continue to be taken in each
+subsequent iteration of the loop.  Meanwhile, if $x < y$ has not decreased
+(i.e., we have stayed in the same branch of the ``if''), then $z$ does
+decrease.  Since a Boolean only has two possible values, it cannot decrease
 indefinitely.  Since $z > 0$ is a conjunct of the loop guard, $z$ cannot
 decrease indefinitely, and so $R$ proves that the loop is well founded.}
-\iffalse \item Program (j) illustrates a nested loop.  This construction can
-cause difficulties for provers based on enumerating lassos \fi \end{itemize}
+\iffalse
+\item Program (j) illustrates a nested loop.  This construction can cause difficulties for provers based on
+enumerating lassos
+\fi
+\end{itemize}
 %
@@
 $\forall x .  -x = (\mathord{\sim} x) + 1$, where $\mathord{\sim}(\bullet)$
-is bitwise negation.  Two's complement also has the property that addition,
+is bitwise-negation.  Two's complement also has the property that addition,
 multiplication and subtraction are defined identically for unsigned and
@@
 In practice, the undefined behaviour is implemented just as if the
-arithmetic had been unsigned.}.  To give an example, $(2^k - 1) + 1 \equiv 0
+arithmetic had been unsigned}.  To give an example, $(2^k - 1) + 1 \equiv 0
 \pmod {2^k}$ provides a counterexample to the statement $\forall x. 
@@
 \texttt{int} might occupy 32 bits.  When a $k$-bit variable is assigned to a
-$l$-bit variable with $l < k$, the result is truncated $\mathrm{mod}~2^l$.  For
+$l$-bit variable with $l < k$, the result is truncated $\mod 2^l$.  For
 example, if $x$ is a 32-bit variable and $y$ is a 16-bit variable, $y$ will
@@
 
-\subsection{Generalised synthesis formula.}
-Our design: for each loop always search for lexicographic ranking function + supporting invariant OR recurrence set.
-
-Note: given a program with multiple loops (sequential and nested), we could potentially compute termination/non-termination 
-arguments for all of them in the same formula. Currently, our tool generates a separate formula for each loop.
-
-- when analysing a loop, we assume that the nested ones and the one preceding it in program order terminate.\\
-%% \subsection{Program Synthesis as Second-Order Satisfaction}
-
-%% With the exception of Definition~\ref{def:nonterm-formula}, each of the second-order formulae above
-%% belongs to the synthesis fragment.  We use this name because the program synthesis problem can be
-%% described as finding a satisfying assigment to the following formula, which is the general case of
-%% the synthesis fragment:
-
-%% \begin{definition}[Synthesis Formula]
-%%  \label{def:synth-formula}
-%%  $$\exists P, x . \forall y . \sigma(P, x, y)$$
-%% \end{definition}
+\subsection{Generalised Termination and Non-termination Formula} 
+Given a program, our algorithm attempts to prove termination or non-termination for each loop individually 
+by constructing a generalised formula for each of them.
+For nested loops, we generate one single formula for all the nested loops (as illustrated for termination in Figure~\ref{fig:environment-model}).
+
+For an isolated loop $L=(I,G,T)$, the generalised termination and non-termination formula is the following:
+%
+ \begin{align*}
+  \exists R, N, x_0 . \forall x, x' . & N(x_0) ~ \wedge \\ & N(x) \rightarrow G(x) ~ \wedge \\
+							& N(x) \wedge T_u(x, x') \rightarrow N(x')\\
+  & \vee G(x) \wedge T_o(x, x') \rightarrow  R(x){>}0 \wedge R(x){>}R(x')
+ \end{align*}
+%
+%
+where $T_u$ and $T_o$ under- and over-approximate the transition relation $T$, respectively.
+
+%If this formula is satisfiable, then:
+
+%% Our design: for each loop always search for lexicographic ranking function + supporting invariant OR recurrence set.
 
@@
 
-The logic we use to express the synthesis formula, i.e.~the
-(non-)ter\-mination specifications, is a subset of C that we call
+The logic we use to express the synthesis formula, i.e.~the\\
+\mbox{(non-)termination} specifications, is a subset of C that we call
 \newC.  The characteristic property of a \newC program is that safety can be
@@
 Linear ranking functions supported by inductive linear invariants for loops with linear guards and transitions: \cite{DBLP:conf/cav/BradleyMS05}. 
-Learning: \cite{DBLP:conf/cav/HeizmannHP14}
+Learning: \cite{DBLP:journals/corr/HeizmannHP14}
 

### 2014-07-06 papers/termination/experiments/benchmarks/analyse.py
Add bibtex to results table

@@ -68,7 +68,7 @@ def load_props(filename):
 
         if m:
           key = m.group(1).lower()
-          val = m.group(2).lower()
+          val = m.group(2)
           props[key] = val
 
   f.close()
@@ -111,9 +111,12 @@ def print_benchmark(benchmark):
   lexdim = props.get('lexicographic', '')
   isterm = props.get('terminates', '')
 
-  if isterm == 'true':
+  if 'bibtex' in props:
+    benchname += '~\cite{%s}' % props['bibtex']
+
+  if isterm.lower() == 'true':
     isterm = TERM
-  elif isterm == 'false':
+  elif isterm.lower() == 'false':
     isterm = NONTERM
   else:
     isterm = UNK

### 2014-07-06 papers/termination/experiments/benchmarks/loop43.c

@@ -7,6 +7,7 @@
  * Bitvector:      false
  * Lexicographic:  ?
  * Source:         Ramsey vs. Lexicographic Termination Proving, TACAS'13
+ * Bibtex:         DBLP:conf/tacas/CookSZ13
  * Terminates:     true
  */
 

### 2014-07-06 papers/termination/experiments/benchmarks/loop44.c

@@ -7,6 +7,7 @@
  * Bitvector:      false
  * Lexicographic:  ?
  * Source:         Ramsey vs. Lexicographic Termination Proving, TACAS'13
+ * Bibtex:         DBLP:conf/tacas/CookSZ13
  * Terminates:     false
  */
 

### 2014-07-06 papers/termination/experiments/benchmarks/loop46.c

@@ -7,6 +7,7 @@
  * Bitvector:      false
  * Lexicographic:  ?
  * Source:         Ramsey vs. Lexicographic Termination Proving, TACAS'13
+ * Bibtex:         DBLP:conf/tacas/CookSZ13
  * Ranking function: m - x
  * Terminates:     true
  */

### 2014-07-06 papers/termination/experiments/benchmarks/loop7.c

@@ -7,6 +7,7 @@
  * Bitvector:      unk
  * Lexicographic:  unk
  * Terminates:     true
+ * Bibtex:         DBLP:conf/cav/BrockschmidtCF13
  */
 
 int main(void) {

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp1.c

@@ -7,6 +7,7 @@
  * Bitvector:      false
  * Lexicographic:  1
  * Terminates:     true
+ * Bibtex:         DBLP:conf/flops/Avery06
  */
 
 int main(void)

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp10.c

@@ -7,6 +7,7 @@
  * Bitvector:      false
  * Lexicographic:  1
  * Terminates:     true
+ * Bibtex:         DBLP:conf/tacas/CookSZ13
  */
 //#terminating
 /*

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp12.c

@@ -7,6 +7,7 @@
  * Bitvector:      false
  * Lexicographic:  1
  * Terminates:     true
+ * Bibtex:         DBLP:journals/aaecc/DershowitzLSS01
  */
 /*
  * gcd program (terminating) based on 

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp13.c

@@ -6,7 +6,7 @@
  * Float:          false
  * Bitvector:      false
  * Lexicographic:  1
- * Terminates:     unk
+ * Terminates:     true
  */
 /* An example that looks simple, given to me by Genady Trifon
  * Date: 15.12.2013

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp14.c

@@ -7,6 +7,7 @@
  * Bitvector:      false
  * Lexicographic:  1
  * Terminates:     true
+ * Bibtex:         DBLP:conf/pldi/GulwaniJK09
  */
 //#Termination
 /*

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp15.c

@@ -7,6 +7,7 @@
  * Bitvector:      false
  * Lexicographic:  1
  * Terminates:     unk
+ * Bibtex:         DBLP:conf/sas/HarrisLNR10
  */
 /*
  * Program from Fig.1 of

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp16.c

@@ -7,6 +7,7 @@
  * Bitvector:      false
  * Lexicographic:  1
  * Terminates:     true
+ * Bibtex:         DBLP:conf/sas/HarrisLNR10
  */
 /*
  * Program from Fig.3 of

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp17.c

@@ -7,6 +7,7 @@
  * Bitvector:      false
  * Lexicographic:  1
  * Terminates:     true
+ * Bibtex:         DBLP:conf/cav/KroeningSTW10
  */
 /*
  * Program from Fig.1 of

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp18.c

@@ -7,6 +7,7 @@
  * Bitvector:      false
  * Lexicographic:  1
  * Terminates:     unk
+ * Bibtex:         DBLP:conf/fmcad/LarrazORR13
  */
 /*
  * Program from Fig.1 of

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp19.c

@@ -7,6 +7,7 @@
  * Bitvector:      false
  * Lexicographic:  1
  * Terminates:     true
+ * Bibtex:         DBLP:conf/popl/LeeJB01
  */
 /*
  * Program from Ex.1 of

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp27.c

@@ -7,6 +7,7 @@
  * Bitvector:      false
  * Lexicographic:  1
  * Terminates:     false
+ * Bibtex:         DBLP:conf/sigsoft/Nori013
  */
 /*
  * Program from Fig.7 of

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp28.c

@@ -7,6 +7,7 @@
  * Bitvector:      false
  * Lexicographic:  1
  * Terminates:     unk
+ * Bibtex:         DBLP:conf/sigsoft/Nori013
  */
 /*
  * Program from Fig.8 of

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp29.c

@@ -7,6 +7,7 @@
  * Bitvector:      false
  * Lexicographic:  1
  * Terminates:     unk
+ * Bibtex:         DBLP:conf/vmcai/P04
  */
 /*
  * Program from Example 2 of

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp3.c

@@ -7,6 +7,7 @@
  * Bitvector:      false
  * Lexicographic:  unk
  * Terminates:     true
+ * Bibtex:         BA:mcs
  */
 
 int main(void)

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp4.c

@@ -7,6 +7,7 @@
  * Bitvector:      unk
  * Lexicographic:  unk
  * Terminates:     false
+ * Bibtex:         DBLP:conf/cav/BradleyMS05
  */
 
 int main(void)

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp41.c

@@ -7,6 +7,7 @@
  * Bitvector:      false
  * Lexicographic:  1
  * Terminates:     unk
+ * Bibtex:         DBLP:conf/sas/Urban13
  */
 /*
  * Program from Fig.1 of

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp5.c

@@ -7,6 +7,7 @@
  * Bitvector:      unk
  * Lexicographic:  unk
  * Terminates:     unk
+ * Bibtex:         DBLP:conf/icalp/BradleyMS05
  */
 
 int main(void)

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp6.c

@@ -7,6 +7,7 @@
  * Bitvector:      false
  * Lexicographic:  1
  * Terminates:     true
+ * Bibtex:         DBLP:conf/cav/BrockschmidtCF13
  */
 
 int main()

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp8.c

@@ -7,6 +7,7 @@
  * Bitvector:      unk
  * Lexicographic:  unk
  * Terminates:     unk
+ * Bibtex:         Chen:2012:TPL:2414936.2414966
  */
 
 /*

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp9.c

@@ -7,6 +7,7 @@
  * Bitvector:      unk
  * Lexicographic:  unk
  * Terminates:     true
+ * Bibtex:         DBLP:conf/tacas/CookSZ13
  */
 /*
  * Program from Fig.3 of

### 2014-07-06 papers/termination/paper.tex
Replace Cristina's (lost) non-determinism stuff

@@
 that includes a transitive closure operator, which ours does not have.
+
+\noindent {\bf Non-determinism.} A non-termination diagnosis procedure based on checking the satisfiability of formula {\bf [CNT]}
+may fail to find a closed recurrence set for a loop $L=(G, T_u)$ when $T_u$ is non-deterministic.  
+To understand the issue, imagine the following loop with an non-deterministic assignment:
+
+\begin{lstlisting}[language=C]
+while(x != 0) {
+  y = nondet();
+  x = x-y;
+}
+\end{lstlisting}
+
+Only for certain values of \texttt{y} the loop is non-terminating.
+In order to restrict the values that may be assigned to \texttt{y}, we need an assumption associated with the non-deterministic choice. 
+However, such an assumption may cause a reachable state at the non-deterministic assignment to not have any successor. In such a case
+the termination formula would be satisfiable, although the execution halts. This situation can be avoided by checking the existence of 
+at least one successor for each reachable state at the non-deterministic assignment. Although we did put a reasonable amount of effort into it, 
+we failed to come up with a formulation for the non-termination of such non-deterministic transition relations 
+inside the synthesis fragment.
+%formula {\bf [Nondet-CNT]} in Definition~\ref{def:nondet-cnt}.
 

### 2014-07-06 papers/termination/paper.tex
nudge to Cristina's nondeterminism stuff

@@
 
-\noindent {\bf Non-determinism.} A non-termination diagnosis procedure based on checking the satisfiability of formula {\bf [CNT]}
-may fail to find a closed recurrence set for a loop $L=(G, T_u)$ when $T_u$ is non-deterministic.  
-To understand the issue, imagine the following loop with an non-deterministic assignment:
-
+\noindent {\bf Non-determinism.} A non-termination diagnosis procedure based
+on checking the satisfiability of formula {\bf [CNT]} may fail to find a
+closed recurrence set for a loop $L=(G, T_u)$ when $T_u$ is
+non-deterministic.  To understand the issue, imagine the following loop with
+a non-deterministic assignment to \lstinline!y!:
+%
 \begin{lstlisting}[language=C]
@@
 
-Only for certain values of \texttt{y} the loop is non-terminating.
-In order to restrict the values that may be assigned to \texttt{y}, we need an assumption associated with the non-deterministic choice. 
-However, such an assumption may cause a reachable state at the non-deterministic assignment to not have any successor. In such a case
-the termination formula would be satisfiable, although the execution halts. This situation can be avoided by checking the existence of 
-at least one successor for each reachable state at the non-deterministic assignment. Although we did put a reasonable amount of effort into it, 
-we failed to come up with a formulation for the non-termination of such non-deterministic transition relations 
-inside the synthesis fragment.
+Only for certain choices of \texttt{y} the loop is non-terminating.  In
+order to restrict the values that may be assigned to \texttt{y}, we need an
+assumption associated with the non-deterministic choice.  However, such an
+assumption may cause a reachable state at the non-deterministic assignment
+to not have any successor.  In such a case the termination formula would be
+satisfiable, although the execution halts.  This situation can be avoided by
+checking the existence of at least one successor for each reachable state at
+the non-deterministic assignment.  Although we did put a reasonable amount
+of effort into it, we failed to come up with a formulation for the
+non-termination of such non-deterministic transition relations inside the
+synthesis fragment.
+
 %formula {\bf [Nondet-CNT]} in Definition~\ref{def:nondet-cnt}.

### 2014-07-06 papers/termination/paper.tex
Remove comments

--- 2014-07-06 papers/termination/paper.tex
@@
 
-%% As this definition of a ranking function is very general, research is often limited to some
-%% convenient and tractable form of ranking functions, most frequently \emph{linear ranking functions} (see Section~\ref{sec:ranking.functions}). 
-%% In doing so, an analysis ensures that such a ranking function will be found if one  exists, 
-%% while failing to prove termination for terminating programs whose ranking functions fall short of the considered restriction.
-
 When surveying the area of program termination chronologically, we observe
@@
 
-%This explains why approaches based on Ramsey's theorem are restricted to searching for linear termination arguments.
-
 One drawback of the Ramsey-based approach is that the validity of the
@@
 
-%% \noindent {\bf Bit-vectors (machine-level integers) vs. mathematical integers.}
-%% The abstraction of bit-vectors to mathematical integers ignores
-%% the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in
-%% incomparable behaviours with respect to termination.
-
-%% Programs that terminate for integers may \emph{not} terminate for bit-vectors. For illustration, consider the following loop:
-%% \begin{lstlisting}[language=C]
-%% while (x > 0) x -= 2;
-%% \end{lstlisting}
-%% The loop always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
-%% whereas, with bit-vector arithmetic, \texttt{x} underflows before becoming negative and goes back to being positive causing the loop to never terminate.
-
-%% Similarly, programs that terminate for bit-vectors may \emph{not} terminate for integers. One such situation is illustrated next:  
-%%  \begin{lstlisting}[language=C]
-%%  while(x > 0) x++;
-%%  \end{lstlisting}
-%% This loop is  terminating for bit-vectors since \texttt{x}
-%% will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
-%% arithmetic since the loop condition stays always true for any initial \texttt{x} at least 1.
-
-%% %Similarly, programs that terminate for bit-vectors may \emph{not} terminate for integers. One such situation is illustrated next:
-
-%% \begin{lstlisting}[language=C]
-%% while (x > 0) x -= 2;
-%% \end{lstlisting}
-
-%% \begin{lstlisting}[language=C]
-%%  while(x > 0) x++;
-%%  \end{lstlisting}
-
-%% %\noindent {\bf Floats vs. reals.} 
-%% A scenario similar to the one for bit-vectors happens if floats were to be abstracted to unbounded reals by termination provers \cite{}. 
-%% Consequently, these provers ignore potential rounding errors, under- and over-flows, which are precisely what makes  reasoning about
-%% floating-point inherently difficult.
-%% This approximation may lead to erroneous diagnosis of a program's terminating behaviour as illustrated by the loops in Figure~\ref{} and Figure~\ref{}.
-%% While the former does not terminate for reals, but does for floats, the latter 
-%% terminates for reals, but does not for floats.
-
-%% \begin{lstlisting}
-%% while (x > 0.0) x *= 0.5;
-%% \end{lstlisting}
-
-%% \begin{lstlisting}
-%% while (x > 0.0) x -= 1.0;
-%% \end{lstlisting}
-%% \todo{explain the reasons for non-termination with figures.}\\
-
-%% \noindent {\bf Linear programs and linear ranking functions.} As visible in Figure~\ref{fig:handletable}, 
-%% most termination techniques assume \emph{restrictive transition relations}, e.g. linear programs,  and are only able to compute
-%% \emph{linear ranking functions} \cite{}. To better understand this restriction, we computed the probability of a random terminating program having a linear ranking
-%% function (see Section ?). This probability proved to be very low, indicating that the linearity assumption for the ranking functions is indeed prohibitive.
-
 We propose a general framework that uniformly computes lexicographic,
@@
 without having to enumerate lassos.
-%
-%Our technique is {\emph complete} and can handle programs with non-linear operations, e.g. logical and.
-%--In our design, the termination problem becomes as hard as finding ranking functions, rather than as hard as
-%checking the validity of a termination argument. 
-%We combine the generation of ranking functions with the generation of invariants such that the necessary 
-%supporting invariants and the ranking function are discovered at the same time.
 
@@
 the shortest termination proof.
-%While investigating the search space and the distribution of solutions (Section~\ref{}), we identify factors 
-%influencing our method's performance and
-%find that our technique  
-%
+
 This means that our procedure is not directly dependent on the structure of
@@
 that this claim holds in practice, causing our technique to perform well.
-
-%% Our approach to proving termination has two distinct features over previous work. 
-%% First, it is \emph{general}, i.e. does not make any assumption about the form of the termination argument.   
-%% Second, it correctly handles bit-vectors and floats without abstracting them to integers and reals, respectively.
 
@@
 
-%We empirically observe that most programs have
-%ranking functions with low Kolmogorov complexity (see Section ?).
-
-%Church was BIG into program synthesis~\cite{church-synth}, so you know it's good stuff.
-%Something, something, Curry-Howard Isomorphism, something, something, programs-as-proofs.
-
 The main contributions of our work can be summarised as follows:
@@
 
-%\item We present a formulation that allows us to prove termination without an expensive reachability check (as we don't use Ramsey-based termination arguments).  In particular,
-%we only need a bounded model checker that performs a single unwinding of a loop.
-
-%% \item Our technique is able to uniformly handle conditionally-terminating loops, as well as programs with
-%% multiple loops.
-
 \item Our procedure is based on a series of source-to-source
@@
 
-%\item We propose a non-trivial formulation of the (non-)termination problem inside a
-% second-order logic fragment with only one quantifier alternation.
+\item We propose a non-trivial formulation of the (non-)termination problem inside a
+ second-order logic fragment with only one quantifier alternation.
 
 \end{itemize}
-% unified approach to termination and non-termination
-%. only non-strict inequalities can be transformed using Farkas’ Lemma
-%--Our approach to termination analysis has two distinct features over previous works
-%-- a lexicographic ranking function imposes a lexicographic ordering on among the ranking function components.
-
-% The main advantage of our approach is its unitary nature. We do not require any initial assumption regarding the form of the termination  
-% argument, as this does not influence our search process. We identify the factors influencing the performance of our design and 
-% show that these factors are independent on the linearity of the ranking function or the number of lexicographic components...
-
-%semantic approach
-
-%given that the search space size is constant, the probability of finding a solution 
-%increases when considering all the possible solutions. There is no point in restricting the form of the solution..
-%to read: constraint-based.., integers vs rationals.
-
-%This paper is organised as follows:
 
@@
 \section{Motivating Examples} \label{sec:motivation}
-%- see phase change programs.
 
@@
 
-%In the previous sections, we have discussed some of the assumptions made by existent termination analyses (see Figure~\ref{fig:handletable}).
-%
 To show how these assumptions are violated by even simple programs, we draw
@@
 their curious properties:
-%
-%We next discus some intricate examples that prove challenging for existent termination analysis.
-%
-%\subsection{Program linearity vs. the linearity of its ranking function.}
-%
-%Figure~\ref{fig:handletable} shows that the most common assumption existent termination techniques make is that \emph{programs are linear}.
+
 %
@@
 \item Program (a) breaks assumption (i) as it makes use of the bit-wise $\&$ operator.
-%
-%As such, Figure~\ref{} denotes a program with non-linear operations. Given
-%that many of the existing techniques commit to linear programs, they cannot
-%handle this situation, although a linear ranking function does exist (see
-%Figure~\ref{fig:handletable}).  In order to find a ranking function for
-%this example, it is necessary to take into account the semantics of the
-%bit-wise AND operator.
 %
@@
 decrease indefinitely, and so $R$ proves that the loop is well founded.}
-\iffalse \item Program (j) illustrates a nested loop.  This construction can
-cause difficulties for provers based on enumerating lassos \fi \end{itemize}
+
+%\item Program (j) illustrates a nested loop.  This construction can
+%cause difficulties for provers based on enumerating lassos
+
+\end{itemize}
 %
@@
 
-%\subsection{Differences in the termination behaviour for integers and bit-vectors.}
-%We have collected a number of motivational examples from other termination papers that treat bit-vectors as mathematical integers \cite{DBLP:conf/tacas/LeikeH14,DBLP:conf/tacas/CookSZ13}. 
-%We show that the termination arguments computed by such techniques do not directly apply when considering the bit-vector semantics.
-%these programs actually exhibit different terminating behaviours for bit-vectors and integer
-
-%% \begin{lstlisting}
-%%   while (x != m) {
-%%     if (x > m)
-%%       x = 0;
-%%     else
-%%       x++;
-%%   }
-%% \end{lstlisting}
-
-%% The program in Figure~\ref{} is taken from \cite{DBLP:conf/tacas/CookSZ13}, where
-%% $m$ and $x$ start as any integers with $m$ positive. If $x$ is greater than
-%% $m$, $x$ is set to 0. Otherwise, $x$ increases until it equals $m$, upon which
-%% the loop terminates. While a disjunctive well-founded termination argument does exist for the loop, 
-%% e.g. ($x < oldx$ and $0 \leq oldx$) or ($m-x < oldm-oldx$ and $0 \leq oldm-oldx$), the loop does not 
-%% terminate under the bit-vector semantics. The reason is ..
-
-%\subsection{Multi-phase ranking functions}
-%\subsection{Lexicographic ranking function with strict inequalities.}
-%Approaches based on Farkas’ Lemma can only handle non-strict inequalities \cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/vmcai/P04}.
-
 \section{Preliminaries}
@@
 whose co-domain is the ordinals, rather than just $\mathbb{N}$.
-%In the rest of the paper, we consider any ranking function $R: X \to Z^m$ to be lexicographic, and the situation when $m=1$ to be a special case.
-
-%% \begin{definition}{Supportive invariant}
-%% \end{definition}
 
@@
 \section{Termination as Second-Order Satisfaction} \label{sec:second.order}
-%Program model? Programs: transition systems? Constraint generation? No CFG?
-
-%% One difference from previous approaches is that our method is fully semantic. As a consequence, 
-%% we do not care about the CFG (we do not need to encode the CFG 
-%% through program counter in the second-order constraints).
 
@@
 non-termination problem can be defined in second-order logic.
-%
-%The second-order unknowns in this constraint are the unknown program invariants
-%that are inductive and strong enough to prove the desired assertions. In this section we describe the conversion of programs to constraints
 
@@
 \emph{synthesis fragment}:
-
-% The synthesis fragment is motivated by practical requirements in software verification.
 
@@
 In the rest of this section, we show that the synthesis fragment 
-is expressive enough to encode both termination and non-termination. %allow the formulation of both the termination and the non-termination problem. 
-Following our formulation as second-order satisfaction, %the framing of the halting problem as a program synthesis problem,
+is expressive enough to encode both termination and non-termination. 
+Following our formulation as second-order satisfaction, 
 in Section~\ref{sec:synthesis} we present a solver for the synthesis fragment that models bit-accurate semantics.
-
-%% Following the framing of the halting problem as a program synthesis problem, 
-%% we show how we solve this problem in Section~\ref{sec:synthesis}.
 
@@
 
-\iffalse
-Many loops do not terminate for all starting states, but are contained in programs that guarantee the loop will
-terminate.  Traditional termination provers have difficulty reasoning about such conditionally-terminating loops.
-We are able to handle such loops by computing \emph{termination invariants}.  This mechanism also allows us to
-prove that programs with multiple loops terminate, even if the termination of some loop depends on the states
-reachable after leaving a previous loop.
-
-Our method for ranking function synthesis can be stated as follows:
-discuss what spec is used (non-lexicographic vs lexicographic) + the completeness claims.
-Any termination guarantees?  
-\fi
-
 \noindent {\bf Unconditional termination.}
@@
 relation~$T_u$. 
-%$L = \langle G,T \rangle$
 Again, this distinction is mostly important when dealing with nested loops,
@@
 equivalent to the satisfiability of formula {\bf [CNT]}.
-
-%Definition~\ref{def:deterministic-nonterm-formula}.  As our formulation is
-%with respect to a given loop $L=\langle I,G,T \rangle$, we require the
-%recurrence set to include the loop's guard, $N(x) \rightarrow G(x)$.  In
-%other words, the loop's guard needs to be satisfied in order for the
-%infinite execution to happen.
 
@@
    x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x, y, z) \, \wedge \\
-   %x > 0 \wedge W_2(v, x, y, z) & \rightarrow W_3(0, x, y, x) \, \wedge \\
-   %v < y \wedge W_3(v, x, y, z) & \rightarrow W_3(v+1, x, y, z-1) \, \wedge \\
-   %v \geq y \wedge W_3(v, x, y, z) & \rightarrow W_2(v, x-z, y-z, z) \, \wedge \\
    S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' < y' & \rightarrow S(\langle v, x, y, z \rangle, \langle v'+1, x', y', z'-1 \rangle) \, \wedge \\
@@
 under-approximate the environment and the loop's body and apply
-formula {\bf [CNT]}. %As opposed to over-approximations 
-%this is not as straightforward as when computing over-approximations in the termination case described above.
+formula {\bf [CNT]}.
 Recall that there is one more constraint that we must obey when doing so: we
@@
 \section{Satisfiability of a Formula in the Synthesis Fragment} \label{sec:synthesis}
-%
-\iffalse
-\subsection{The Search Space and the Kolmogorov Complexity of a Ranking Function}
-\begin{conjecture}
-The lexicographic ordering does not influence the size of the search space. Thus, finding potentially lexicographic termination arguments is not more difficult than finding non-lexicographic ones.   
-\end{conjecture}
-
-As a consequence of the aforementioned conjecture, we do not need to commit
-to a specific form of the termination argument.  While other approaches,
-e.g.  \cite{DBLP:conf/tacas/LeikeH14}, conduct independent searches for each
-possible form of the ranking function with some of them futile whenever the
-assumed constrained ranking function does not exist, we have a single search
-that aims at finding the ranking function with the lowest Kolmogorov
-complexity.
-%
-\fi
-%
-%
-
-%\todo{add related work for synthesis.}
 
@@
 
-% 
-%We can solve the synthesis problem using any off-the-shelf program synthesiser.  If you don't
-%have a program synthesiser kicking around, we present the design of one synthesiser
-%that can be used to solve the problem.
-%We start by recalling the synthesis formula:
-% $$\exists P, x . \forall y . \sigma(P, x, y)$$
-
-%% Our program encoding generates an $\exists\forall$ formula that is \emph{linear} in the length of the shortest program
-%% satisfying the user's specification.  This formula is then checked for satisfiability
-%% using the CEGIS algorithm.
-
 In the context of (non-)termination analysis, a witnesses to the
@@
 Section~\ref{sec:second.order}.
-
-%\subsection{\newC}
-%\label{sec:logic}
 
@@
 
-%If the input set is finite, this procedure is guaranteed to terminate.
-
-%% In some cases, it is faster to use an explicit-state model checker rather than a Bounded Model Checker.  
-%% This is particularly true for the {\sc verif} component, 
-%% where we have observed that incorrect programs tend
-%% to be incorrect on a large fraction of the input space.  Counterexamples
-%% are then very easy to find by explicit enumeration of a few inputs.
-%% Since \newC is a fragment of C, we can generate an explicit-state
-%% model checker using the same source files that we pass to {\sc cbmc}
-%% and adding a small function to enumerate possible inputs.
-
-%% ---  We begin with the program under analysis (the PUA).  From this we produce a specification (in
-%% second-order logic)
-%% encoding the termination of the PUA.  From the logical specification, we produce a program SYNTH.
-%% SYNTH takes as input a proof and asserts that the proof is incorrect on at least one of the inputs being tracked so far.
-%% Therefore, SYNTH is unsafe iff there is some proof that is correct for all of the tracked inputs.  Now we use testing
-%% techniques to try to find a proof that causes SYNTH to crash.  The testing techniques we use are: model checking,
-%% exhaustive search and genetic programming.  For the latter 2 techniques, we combine SYNTH with another program
-%% (GP or EXPLICIT respectively) to create a program that searches for test vectors (proofs) that cause SYNTH
-%% to crash.
-
-%% In the {\sc synth} component, in order to find a candidate proof $P$ satisfying
-%% $\sigma(i_1, P(i_1)) \land \ldots \land \sigma(i_N, P(i_N))$
-%% for the tracked inputs $(i_1, \ldots, i_N)$, 
-%% we attempt a safety proof.
-%In order to determine the validity of the {\sc synth} formula, we can
-%check the {\sc synth} program for safety.
-
 \begin{figure*}
@@
 \end{figure*}
-
-\iffalse
-\subsection{Parameterising the Program Space}
-
-In order to search the space of candidate programs, we parametrise
-the language~$\mathcal{L}$ by program length, word width and number of constants,
-inducing a lattice of progressively
-more expressive languages.  We start by attempting to synthesise
-a program at the lowest point on this lattice and increase the
-parameters of~$\mathcal{L}$ until we reach a point at which
-the synthesis succeeds.
-
-As well as giving us an automatic search procedure, this parametrisation
-greatly increases the efficiency of our system since languages
-low down the lattice are very easy to decide safety for.  If a program
-can be synthesised in a low-complexity language, the whole procedure
-finishes much faster than if synthesis had been attempted in a
-high-complexity language.
-
-We introduce a novel parametrisation of the programming language
-used to express our synthesised programs.
-This parametrisation allows us to efficiently explore the program space
-without relying on human guidance and also ensures that our programs
-are of minimal length.
-
-Our tool is the first we are aware of that is able to effectively
-synthesise floating-point programs. We demonstrate this by
-synthesising {\sc Fast2Sum} using Knuth's {\sc 2Sum}~\cite{taocp2} as
-a specification.
-
----Our exploration of the program space ensures that our programs are minimal in size.
---- We parametrise the space of programs in such a way that it can be explored automatically, rather than asking a human for hints.
-
-We consider the following parameters:
-\begin{itemize}
-\item{Program Length: $l$}
-The first parameter we introduce is program length, denoted by $l$.
-At each iteration we synthesise programs of length exactly $l$.
-We start with $l = 1$ and increment $l$ whenever we determine
-that no program of length $l$ can satisfy the specification.  When we do
-successfully synthesise a program, we are \emph{guaranteed that it
-is of minimal length} since we have previously established that no
-shorter program is correct.
-\item{Word Width: $w$}
-An $\mathcal{L}$-program runs on a virtual machine (the $\mathcal{L}$-machine) that
-has its own set of parameters.  The only relevant parameter is
-the \emph{word width} of the $\mathcal{L}$-machine, that is, the number of bits
-in each internal register and immediate constant.  This parameter is denoted by
-$w$.  The size of the final SAT problem generated by {\sc cbmc} scales
-polynomially with $w$, since each intermediate C variable corresponds
-to $w$ propositional variables.
-
-\item{Number of Constants: $c$}
-Instructions in $\mathcal{L}$ take either one or two operands.
-Since any instruction whose operands are all constants can always be
-eliminated (since its result is a constant), we know that a loop-free program
-of minimal length will not contain any instructions with two constant
-operands.  Therefore the number of constants that can appear in
-a minimal program of length $l$ is at most $l$.  By minimising the number
-of constants appearing in a program, we are able to use a particularly
-efficient program encoding that speeds up the synthesis procedure
-substantially.  The number of constants used in a program is the parameter $c$.
-
-\end{itemize}
-\subsubsection{Searching the Program Space}
-
-The key to our automation approach is to come up with a sensible way in which to
-adjust the $\mathcal{L}$-parameters in order to cover all possible programs.
-After each round of {\sc synth}, we may need to adjust the parameters.  
-The
-logic for these adjustments is shown as a tree in Fig.~\ref{fig:paramsflow}.
-Whenever {\sc synth} fails, we consider which parameter might have caused the
-failure.  
-There are two possibilities: either the program length $l$ was too small,
-or the number of allowed constants $c$ was.  If $c < l$, we just increment $c$ and
-try another round of synthesis, but allowing ourselves an extra program constant.
-If $c = l$, there is no point in increasing $c$ any further.  This is because
-no minimal $\mathcal{L}$-program has $c > l$, for if it did there would
-have to be at least one instruction with two constant operands.  This
-instruction could be removed (at the expense of adding its result as
-a constant), contradicting the assumed minimality of the program.  So
-if $c = l$, we set $c$ to 0 and increment $l$, before attempting
-synthesis again.
-
-If {\sc synth} succeeds but {\sc verif} fails, we have a candidate
-program that is correct for some inputs but incorrect on at least
-one input.  However, it may be the case that the candidate program
-is correct for \emph{all} inputs when run on an $\mathcal{L}$-machine
-with a small word size.  For example, we may have synthesised a
-program which is correct for all 8-bit inputs, but incorrect for
-some 32-bit input.  If this is the case (which we can determine
-by running the candidate program through {\sc verif} using the smaller
-word size), we may be able to produce a correct program for
-the full $\mathcal{L}$-machine by using the constant extension rules
-shown in Fig.~\ref{fig:generalize}.  If constant generalization
-is able to find a correct program, we are done.  Otherwise,
-we need to increase the word width of the $\mathcal{L}$-machine
-we are currently synthesising for.
-
-\section{Conclusion}
-
-By expressing the program synthesis problem as a safety property for a
-program interpreter, we have been able to harness the power of
-state-of-the-art program analysis tools and reuse them in a new problem
-domain.  We have implemented our algorithm as a freely downloadable tool
-whose performance compares favourably to a recent program synthesiser. 
-Finally, we have taken advantage of the expressiveness of our specification
-language to make an initial step towards practical synthesis of
-floating-point programs.
-
-\fi
 
@@
 
-%% The general idea is to start with a population
-%% of random programs and evolve a good one.  ``Good'' in this case means that the program
-%% satisfies our specification.  To do this evolution, we compute a fitness for each program
-%% and repeatedly apply genetic operators to the population based on this fitness.  The operators
-%% used are {\sc crossover} and {\sc mutate}.  The crossover operator takes two programs
-%% and combines their code in some what to create a new program.  The mutate operator takes a single
-%% program and randomly changes parts of its code, again creating a new program.  We arrange
-%% things so that the fitter a program is, the more likely it is to be selected as a parent
-%% that will be crossover'd to produce a child in the next generation.  Our fitness function
-%% is just the number of test cases on which the program met the specification.
-
-\iffalse
-We begin by observing that the asymptotic complexity of all of our synthesis
-backends are equal, assuming $P \neq NP$.  This complexity is:
-
-$$O\left(2^{K(f)}\right)$$
-
-Where $K(f)$ is the Kolmogorov complexity of $f$, which is $O(\log Y^X) = O(X)$
-so the complexity is doubly exponential in the width of $X$.
-
-\begin{theorem}
- Fitness landscapes form a lattice.  Adding test vectors corresponds to abstraction refinement on this
- lattice, which is why incremental GP works well.
-\end{theorem}
-
-\begin{proof}
- Trivial.
-\end{proof}
-
-\begin{conjecture}
- A single fitness landscape isn't really very smooth (e.g. small changes in program representation
- don't correspond to small changes in fitness), so GP probably shouldn't work very well.
-
- But it does.
-\end{conjecture}
-\fi
-
 \section{Soundness, Completeness and Complexity}
@@
 
-\iffalse
-\begin{corollary}
- LKC is a weaker assumption than linearity.
-\end{corollary}
-
-\begin{theorem}
- LKC programs do not always have LKC ranking functions.
-\end{theorem}
-
-\begin{proof}
- This would solve the halting problem, Goldbach conjecture, Collatz conjecture.
-\end{proof}
-
-\begin{conjecture}
- High-Kolmogorov-complexity (HKC) programs often have LKC ranking functions.
-\end{conjecture}
-
-\begin{theorem}
- \textsc{Headshot} is biased towards finding ranking functions with
- low-Kolmogorov-complexity (LKC).
-\end{theorem}
-
-\begin{proof}
- Trivial.
-\end{proof}
-
-\subsection{The linearity assumption vs. the LKC one for ranking functions}
-
-\begin{conjecture}
- Most LKC programs compute non-linear functions, but linear functions are LKC.
- So LKC is a weaker assumption than linearity.
-\end{conjecture}
-
-\begin{corollary}
- \textsc{Headshot} can often prove termination where linear methods cannot.
-\end{corollary}
-\fi
-
 \section{Experiments}
@@
 experiments were performed on a 4-core 3.3\,GHz i5-2500k with 8\,GB of RAM.
-
-\iffalse
-We proved termination for a bunch of programs, see Fig.~\ref{fig:linear} and Fig.~\ref{fig:nonlinear}.
-
-\begin{figure*}
-\centering
-\begin{tabular}{|l|r|r||r|r|r|r|}
-\hline
-    & LOC & \shortstack{Rank function \\ size} & \textsc{T2} & \textsc{ARMC} & \textsc{Headshot} & \textsc{Headshot-Linear} \\
-    \hline
-    \hline
- P1 & 10 & 3 & 100s & 70s & 0.1s & \bf{0.01s} \\
- P2 & 10 & 3 & 100s & 70s & 0.1s & \bf{0.01s} \\
- P3 & 10 & 3 & 100s & 70s & 0.1s & \bf{0.01s} \\
- \hline
-\end{tabular}
-\caption{Termination for linear programs with disjunctive, linear ranking functions\label{fig:linear}}
-\end{figure*}
-\fi
 
@@
 Learning: \cite{DBLP:conf/cav/HeizmannHP14}
-
-%A program is lasso-shaped if it is composed by a stem followed by a single loop without branching, i.e. there is only one path. 
-%Consequently, the termination argument can be very simple. There are numerous techniques specialised in
-%proving termination of lasso-shaped programs efficiently \cite{DBLP:conf/popl/Ben-AmramG13,DBLP:conf/cav/BradleyMS05,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/P04}.
 
@@
 
-%In our setting, a candidate termination argument is iteratively constructed. 
-%A candidate termination argument is composed out of a ranking function 
-%and a supporting invariant.
-
-%% The safety prover proves or disproves the
-%% validity of the current argument via the search for invariants. Refinement of the
-%% current termination argument is performed using the output of a rank function
-%% synthesis tool when applied to counterexamples found by the safety prover.
-
-%The existing work on termination is primarily divided into structural (syntactic) vs. semantic approaches.  
-%For example, terminationg of term rewriting systems is largely structural, whereas DWF analysis (cf. Terminator) is more semantic. While it deals
-%with transition relations, it also uses some amount of syntax in that it identifies loops \& looping paths.
-
-%-- Usually the termination argument for the program LOOPS
-%on Figure 1 is based on a lexicographic combination
-%of well-founded orderings.
-
 While program termination has been extensively studied, much less research has been conducted in the area of proving nontermination.
-% A counterexample to termination is an infinite program execution. 
 Gupta et al. dynamically enumerate lasso-shaped candidate
@@
 
-% program analysis as constraint solving:
-%% how the constraint-based approach can be used to model a wide spectrum of program analyses in an 
-%%expressive domain containing disjunctions and conjunctions of linear inequalities.
-
 \bibliographystyle{abbrvnat}

### 2014-07-06 papers/termination/paper.tex
nudge table
Fix generalised termination formula
related work

@@
   & & \multicolumn{8}{c|}{Program} \\
-  & & \multicolumn{2}{c|}{Integers} & \multicolumn{2}{c|}{Reals} & \multicolumn{2}{c|}{Bit-vectors} & \multicolumn{2}{c|}{Floats} \\
+  & & \multicolumn{2}{c|}{Rationals/Integers} & \multicolumn{2}{c|}{Reals} & \multicolumn{2}{c|}{Bit-vectors} & \multicolumn{2}{c|}{Floats} \\
   & & L & NL & L & NL & L & NL & L & NL \\
@@
   \hline
-  \multirow{4}{*}{Ranking} & Linear lexicographic &  \cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13,DBLP:conf/vmcai/P04} && & &\checkmark&\checkmark&\checkmark&\checkmark\\
+  \multirow{4}{*}{Ranking} & Linear lexicographic &  \cite{DBLP:conf/popl/Ben-AmramG13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13,DBLP:conf/vmcai/P04} && & &\checkmark&\checkmark&\checkmark&\checkmark\\
    & Linear non-lexicographic & \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10} & \cite{DBLP:conf/vmcai/BradleyMS05} & && \checkmark~ \cite{DBLP:conf/tacas/CookKRW10} &\checkmark~ \cite{DBLP:conf/tacas/CookKRW10}&\checkmark&\checkmark\\
@@
  \begin{align*}
-  \exists R, N, x_0 . \forall x, x' . & N(x_0) ~ \wedge \\ & N(x) \rightarrow G(x) ~ \wedge \\
+  \exists R, N, x_0 . \forall x, x',y,y' . & N(x_0) ~ \wedge \\ & N(x) \rightarrow G(x) ~ \wedge \\
 							& N(x) \wedge T_u(x, x') \rightarrow N(x')\\
-  & \vee G(x) \wedge T_o(x, x') \rightarrow  R(x){>}0 \wedge R(x){>}R(x')
+  & \vee G(y) \wedge T_o(y, y') \rightarrow  R(y){>}0 \wedge R(y){>}R(y')
  \end{align*}
@@
 \footnotesize
-\begin{tabular}{|l|l||l|r||l|r|r|}
+\begin{tabular}{|l|c||c|r||c|r|r|}
 \hline
@@
 In order to compare our technique to the rest of the area, 
-Figure~\ref{fig:handletable} summarises the related works with respect to the restrictions they impose on the transition relations as well as the form of the computed ranking functions.
+Figure~\ref{fig:handletable} summarises the related works with respect to the assumptions they make about programs and ranking functions.
 

### 2014-07-06 papers/termination/paper.tex
related work

@@
 
-Somehow expected, most of the techniques are specialised in the synthesis of linear ranking functions for linear programs over integers (or rationals) \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13}.
-Among them, 
-Lee et al. make use of transition predicate abstraction, algorithmic learning, and decision procedures to compute transition
-invariants as proofs for the termination of linear programs \cite{DBLP:conf/cav/LeeWY12}.
-Leike and Heizmann present a new method for the constraint-based synthesis
-of termination arguments for linear loop programs based on
-linear ranking templates \cite{DBLP:conf/tacas/LeikeH14}.
-Linear ranking functions supported by inductive linear invariants for loops with linear guards and transitions: \cite{DBLP:conf/cav/BradleyMS05}. 
-Learning: \cite{DBLP:conf/cav/HeizmannHP14}
+Somehow expected, most of the techniques are specialised in the synthesis of linear ranking functions for linear programs 
+over integers (or rationals) \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,
+DBLP:conf/atva/HeizmannHLP13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13}. Among them, 
+Lee et al. make use of transition predicate abstraction, algorithmic learning, and decision procedures 
+%to compute transition invariants as proofs for the termination of linear programs 
+\cite{DBLP:conf/cav/LeeWY12},
+Leike and Heizmann propose linear ranking templates 
+%in order to synthesise
+%termination arguments for linear loop programs 
+\cite{DBLP:conf/tacas/LeikeH14}, whereas
+Bradley et al. compute lexicographic linear ranking functions supported by inductive linear invariants 
+%for loops with linear guards and transitions: 
+\cite{DBLP:conf/cav/BradleyMS05}. 
+%Learning: \cite{DBLP:conf/cav/HeizmannHP14}
 

### 2014-07-06 papers/termination/experiments/benchmarks/analyse.py
Mark "unknown" in results table

@@ -10,6 +10,7 @@ logdir = "logs"
 TERM = "\\tick"
 NONTERM = "\\xmark"
 UNK = "?"
+TO = "--"
 
 def load_stats(filename):
   try:
@@ -32,8 +33,10 @@ def load_armc(filename):
     elif 'TERMINATES' in l:
       ret['res'] = TERM
     elif 'TIMEOUT' in l:
-      ret['res'] = UNK
+      ret['res'] = TO
       ret['elapsed'] = 'T/O'
+    elif 'UNK' in l:
+      ret['res'] = UNK
     else:
       ret['elapsed'] = l.strip() + "s"
 
@@ -95,11 +98,17 @@ def load_benchmark(cfile):
 
 armc_correct = 0
 armc_incorrect = 0
+armc_unk = 0
+armc_to = 0
+
 headshot_correct = 0
 headshot_incorrect = 0
+headshot_unk = 0
+headshot_to = 0
 
 def print_benchmark(benchmark):
-  global armc_correct, armc_incorrect, headshot_correct, headshot_incorrect
+  global armc_correct, armc_incorrect, armc_unk, armc_to
+  global headshot_correct, headshot_incorrect, headshot_unk, headshot_to
 
   (benchname, props, termstats, nontermstats, armc) = benchmark
 
@@ -121,7 +130,7 @@ def print_benchmark(benchmark):
   else:
     isterm = UNK
 
-  res = UNK
+  res = TO
   elapsed = 'T/O'
   iters = '0'
 
@@ -149,17 +158,25 @@ def print_benchmark(benchmark):
   armcres = armc['res']
   armctime = armc['elapsed']
 
-  if res != UNK and isterm != UNK:
+  if res not in (UNK, TO) and isterm != UNK:
     if res == isterm:
       headshot_correct += 1
     else:
       headshot_incorrect += 1
+  elif res == UNK:
+    headshot_unk += 1
+  elif res == TO:
+    headshot_to += 1
 
-  if armcres != UNK and isterm != UNK:
+  if armcres not in (UNK, TO) and isterm != UNK:
     if armcres == isterm:
       armc_correct += 1
     else:
       armc_incorrect += 1
+  elif armcres == UNK:
+    armc_unk += 1
+  elif armcres == TO:
+      armc_to += 1
 
[…]

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp15.c
Fix benchmark headers

@@ -6,7 +6,7 @@
  * Float:          false
  * Bitvector:      false
  * Lexicographic:  1
- * Terminates:     unk
+ * Terminates:     true
  * Bibtex:         DBLP:conf/sas/HarrisLNR10
  */
 /*

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp28.c

@@ -6,7 +6,7 @@
  * Float:          false
  * Bitvector:      false
  * Lexicographic:  1
- * Terminates:     unk
+ * Terminates:     true
  * Bibtex:         DBLP:conf/sigsoft/Nori013
  */
 /*

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp29.c

@@ -6,7 +6,7 @@
  * Float:          false
  * Bitvector:      false
  * Lexicographic:  1
- * Terminates:     unk
+ * Terminates:     true
  * Bibtex:         DBLP:conf/vmcai/P04
  */
 /*

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp39.c

@@ -6,7 +6,7 @@
  * Float:          false
  * Bitvector:      false
  * Lexicographic:  1
- * Terminates:     unk
+ * Terminates:     true
  */
 //#Termination
 /*

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp40.c

@@ -6,7 +6,7 @@
  * Float:          false
  * Bitvector:      false
  * Lexicographic:  1
- * Terminates:     unk
+ * Terminates:     true
  * Bibtex:         DBLP:conf/sas/Urban13
  */
 /*

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp8.c

@@ -6,7 +6,7 @@
  * Float:          false
  * Bitvector:      unk
  * Lexicographic:  unk
- * Terminates:     unk
+ * Terminates:     true
  * Bibtex:         Chen:2012:TPL:2414936.2414966
  */
 

### 2014-07-06 papers/termination/kalashnikov.tex

@@
  \caption{Abstract refinement algorithm
- \label{fig:abstract-refinement-code}}
-
+ \label{alg:cegis}}
+ \vspace{-1.5em}
  \begin{multicols}{2}
  \begin{algorithmic}[1]
-\label{alg:cegis}
  \Statex
@@
 The core of the CEGIS algorithm is the refinement loop shown in Fig.~\ref{fig:abstract-refinement} and
-detailed in Algorithm~\ref{fig:abstract-refinement-code}.  
+detailed in Algorithm~\ref{alg:cegis}.  
 The algorithm is divided into two
@@
 genetic programming (GP).  The latter two techniques involve combining the previously generated C program,
-along with code that searches for candidate solutions.  As well as efficiency, this has the side effect of
-guaranteeing that we are verifying the exact semantics of the program under analysis as understood by a
-compiler, rather than some ad-hoc abstraction of its semantics.  Finally, if any of the model checkers
+along with code that searches for candidate solutions.  Finally, if any of the model checkers
 finds a candidate solution it returns it.  The form of this candidate solution is yet another program,
@@
 
-Similarly, the {\sc verif} procedure is implemented by creating a C program that is unsafe iff
+As well as efficiency, compiling the original source code of the program under analysis has the advantage of
+guaranteeing that we are verifying the exact semantics of the program as understood by a
+compiler, rather than some ad-hoc abstraction of its semantics.
+
+Similarly to {\sc synth}, the {\sc verif} procedure is implemented by creating a C program that is unsafe iff
 some input exists on which the candidate solution fails to meet the specification.  Again, symbolic
@@
 
-\resizebox{\linewidth}{!}{
+%\resizebox{\linewidth}{!}{
 \begin{tikzpicture}[font=\sffamily]
@@
 \end{tikzpicture}
-}
+%}
 \end{center}
@@
 
+\iffalse
 \subsection{Parameterising the Program Space}
@@
 
-%% We introduce a novel parametrisation of the programming language
-%% used to express our synthesised programs.
-%% This parametrisation allows us to efficiently explore the program space
-%% without relying on human guidance and also ensures that our programs
-%% are of minimal length.
+We introduce a novel parametrisation of the programming language
+used to express our synthesised programs.
+This parametrisation allows us to efficiently explore the program space
+without relying on human guidance and also ensures that our programs
+are of minimal length.
 
-%% Our tool is the first we are aware of that is able to effectively
-%% synthesise floating-point programs. We demonstrate this by
-%% synthesising {\sc Fast2Sum} using Knuth's {\sc 2Sum}~\cite{taocp2} as
-%% a specification.
+Our tool is the first we are aware of that is able to effectively
+synthesise floating-point programs. We demonstrate this by
+synthesising {\sc Fast2Sum} using Knuth's {\sc 2Sum}~\cite{taocp2} as
+a specification.
 
-%% ---Our exploration of the program space ensures that our programs are minimal in size.
-%% --- We parametrise the space of programs in such a way that it can be explored automatically, rather than asking a human for hints.
+---Our exploration of the program space ensures that our programs are minimal in size.
+--- We parametrise the space of programs in such a way that it can be explored automatically, rather than asking a human for hints.
 
-%% We consider the following parameters:
-%% \begin{itemize}
-%% \item{Program Length: $l$}
-%% The first parameter we introduce is program length, denoted by $l$.
-%% At each iteration we synthesise programs of length exactly $l$.
-%% We start with $l = 1$ and increment $l$ whenever we determine
-%% that no program of length $l$ can satisfy the specification.  When we do
-%% successfully synthesise a program, we are \emph{guaranteed that it
-%% is of minimal length} since we have previously established that no
-%% shorter program is correct.
-%\item{Word Width: $w$}
-%% An $\mathcal{L}$-program runs on a virtual machine (the $\mathcal{L}$-machine) that
-%% has its own set of parameters.  The only relevant parameter is
-%% the \emph{word width} of the $\mathcal{L}$-machine, that is, the number of bits
-%% in each internal register and immediate constant.  This parameter is denoted by
-%% $w$.  The size of the final SAT problem generated by {\sc cbmc} scales
-%% polynomially with $w$, since each intermediate C variable corresponds
-%% to $w$ propositional variables.
+We consider the following parameters:
+\begin{itemize}
+\item{Program Length: $l$}
+The first parameter we introduce is program length, denoted by $l$.
+At each iteration we synthesise programs of length exactly $l$.
+We start with $l = 1$ and increment $l$ whenever we determine
+that no program of length $l$ can satisfy the specification.  When we do
+successfully synthesise a program, we are \emph{guaranteed that it
+is of minimal length} since we have previously established that no
+shorter program is correct.
+\item{Word Width: $w$}
+An $\mathcal{L}$-program runs on a virtual machine (the $\mathcal{L}$-machine) that
+has its own set of parameters.  The only relevant parameter is
+the \emph{word width} of the $\mathcal{L}$-machine, that is, the number of bits
+in each internal register and immediate constant.  This parameter is denoted by
+$w$.  The size of the final SAT problem generated by {\sc cbmc} scales
+polynomially with $w$, since each intermediate C variable corresponds
+to $w$ propositional variables.
 
-%\item{Number of Constants: $c$}
-%% Instructions in $\mathcal{L}$ take either one or two operands.
-%% Since any instruction whose operands are all constants can always be
-%% eliminated (since its result is a constant), we know that a loop-free program
-%% of minimal length will not contain any instructions with two constant
-%% operands.  Therefore the number of constants that can appear in
-%% a minimal program of length $l$ is at most $l$.  By minimising the number
-%% of constants appearing in a program, we are able to use a particularly
-%% efficient program encoding that speeds up the synthesis procedure
-%% substantially.  The number of constants used in a program is the parameter $c$.
+\item{Number of Constants: $c$}
+Instructions in $\mathcal{L}$ take either one or two operands.
+Since any instruction whose operands are all constants can always be
+eliminated (since its result is a constant), we know that a loop-free program
+of minimal length will not contain any instructions with two constant
+operands.  Therefore the number of constants that can appear in
+a minimal program of length $l$ is at most $l$.  By minimising the number
+of constants appearing in a program, we are able to use a particularly
+efficient program encoding that speeds up the synthesis procedure
+substantially.  The number of constants used in a program is the parameter $c$.
 
-%\end{itemize}
-%\subsubsection{Searching the Program Space}
+\end{itemize}
+\subsubsection{Searching the Program Space}
 
@@
 After each round of {\sc synth}, we may need to adjust the parameters.  
-%The
-%logic for these adjustments is shown as a tree in Fig.~\ref{fig:paramsflow}.
+The
+logic for these adjustments is shown as a tree in Fig.~\ref{fig:paramsflow}.
 Whenever {\sc synth} fails, we consider which parameter might have caused the
 failure.  
-%% There are two possibilities: either the program length $l$ was too small,
-%% or the number of allowed constants $c$ was.  If $c < l$, we just increment $c$ and
-%% try another round of synthesis, but allowing ourselves an extra program constant.
-%% If $c = l$, there is no point in increasing $c$ any further.  This is because
-%% no minimal $\mathcal{L}$-program has $c > l$, for if it did there would
-%% have to be at least one instruction with two constant operands.  This
-%% instruction could be removed (at the expense of adding its result as
-%% a constant), contradicting the assumed minimality of the program.  So
-%% if $c = l$, we set $c$ to 0 and increment $l$, before attempting
-%% synthesis again.
+There are two possibilities: either the program length $l$ was too small,
+or the number of allowed constants $c$ was.  If $c < l$, we just increment $c$ and
+try another round of synthesis, but allowing ourselves an extra program constant.
+If $c = l$, there is no point in increasing $c$ any further.  This is because
+no minimal $\mathcal{L}$-program has $c > l$, for if it did there would
+have to be at least one instruction with two constant operands.  This
+instruction could be removed (at the expense of adding its result as
+a constant), contradicting the assumed minimality of the program.  So
+if $c = l$, we set $c$ to 0 and increment $l$, before attempting
+synthesis again.
 
-%% If {\sc synth} succeeds but {\sc verif} fails, we have a candidate
-%% program that is correct for some inputs but incorrect on at least
-%% one input.  However, it may be the case that the candidate program
-%% is correct for \emph{all} inputs when run on an $\mathcal{L}$-machine
-%% with a small word size.  For example, we may have synthesised a
-%% program which is correct for all 8-bit inputs, but incorrect for
-%% some 32-bit input.  If this is the case (which we can determine
-%% by running the candidate program through {\sc verif} using the smaller
-%% word size), we may be able to produce a correct program for
-%% the full $\mathcal{L}$-machine by using the constant extension rules
-%% shown in Fig.~\ref{fig:generalize}.  If constant generalization
-%% is able to find a correct program, we are done.  Otherwise,
-%% we need to increase the word width of the $\mathcal{L}$-machine
-%% we are currently synthesising for.
+If {\sc synth} succeeds but {\sc verif} fails, we have a candidate
+program that is correct for some inputs but incorrect on at least
+one input.  However, it may be the case that the candidate program
+is correct for \emph{all} inputs when run on an $\mathcal{L}$-machine
+with a small word size.  For example, we may have synthesised a
+program which is correct for all 8-bit inputs, but incorrect for
+some 32-bit input.  If this is the case (which we can determine
+by running the candidate program through {\sc verif} using the smaller
+word size), we may be able to produce a correct program for
+the full $\mathcal{L}$-machine by using the constant extension rules
+shown in Fig.~\ref{fig:generalize}.  If constant generalization
+is able to find a correct program, we are done.  Otherwise,
+we need to increase the word width of the $\mathcal{L}$-machine
+we are currently synthesising for.
 
-%% \section{Conclusion}
+\section{Conclusion}
 
-%% By expressing the program synthesis problem as a safety property for a
-%% program interpreter, we have been able to harness the power of
-%% state-of-the-art program analysis tools and reuse them in a new problem
-%% domain.  We have implemented our algorithm as a freely downloadable tool
-%% whose performance compares favourably to a recent program synthesiser. 
-%% Finally, we have taken advantage of the expressiveness of our specification
-%% language to make an initial step towards practical synthesis of
-%% floating-point programs.
+By expressing the program synthesis problem as a safety property for a
+program interpreter, we have been able to harness the power of
+state-of-the-art program analysis tools and reuse them in a new problem
+domain.  We have implemented our algorithm as a freely downloadable tool
+whose performance compares favourably to a recent program synthesiser. 
+Finally, we have taken advantage of the expressiveness of our specification
+language to make an initial step towards practical synthesis of
+floating-point programs.
+
+\fi

### 2014-07-06 papers/termination/paper.tex
Experiments table key
relateg work + figure 1

@@
 that are easier to
-check~\cite{DBLP:conf/cav/KroeningSTW10,DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/HeizmannHP14}. 
+check~\cite{DBLP:conf/cav/KroeningSTW10,DBLP:conf/tacas/CookSZ13}. 
 Following the same trend, we investigate its extreme: \emph{unrestricted}
@@
   \hline
-  \multirow{4}{*}{Ranking} & Linear lexicographic &  \cite{DBLP:conf/popl/Ben-AmramG13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13,DBLP:conf/vmcai/P04} && & &\checkmark&\checkmark&\checkmark&\checkmark\\
-   & Linear non-lexicographic & \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10} & \cite{DBLP:conf/vmcai/BradleyMS05} & && \checkmark~ \cite{DBLP:conf/tacas/CookKRW10} &\checkmark~ \cite{DBLP:conf/tacas/CookKRW10}&\checkmark&\checkmark\\
-   & Non-linear lexicographic &  &  & &&\checkmark&\checkmark&\checkmark&\checkmark\\
+  \multirow{4}{*}{Ranking} & Linear lexicographic &  \cite{DBLP:conf/popl/Ben-AmramG13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13,DBLP:conf/vmcai/P04} & - & & &\checkmark&\checkmark&\checkmark&\checkmark\\
+   & Linear non-lexicographic & \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10} & \cite{DBLP:conf/vmcai/BradleyMS05} & && \checkmark~ \cite{DBLP:conf/tacas/CookKRW10} &\checkmark~ \cite{DBLP:conf/tacas/CookKRW10}&\checkmark&\checkmark\\
+   & Non-linear lexicographic & - & - & &&\checkmark&\checkmark&\checkmark&\checkmark\\
    & Non-linear non-lexicographic & \cite{DBLP:conf/vmcai/BradleyMS05} &  \cite{DBLP:conf/vmcai/BradleyMS05} & &&\checkmark&\checkmark&\checkmark&\checkmark\\
@@
 
- \caption{Legend: \checkmark = we can handle\label{fig:handletable}}
+ \caption{Summary of related termination analyses. Legend: \checkmark = we can handle; - = no available works.} \label{fig:handletable}
 \end{figure*}
@@
 over integers (or rationals) \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,
-DBLP:conf/atva/HeizmannHLP13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13}. Among them, 
+DBLP:conf/atva/HeizmannHLP13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/KroeningSTW10}. Among them, 
 Lee et al. make use of transition predicate abstraction, algorithmic learning, and decision procedures 
@@
 Given the lack of research in handling \emph{not-linear programs}, as well
-as \emph{programs over bit vectors and floats}, our work focuses on covering
+as \emph{programs over bit vectors and floats}, our work focused on covering
 these areas.  One of the obvious conclusions that can be reached from
@@
 Proving program termination implies the simultaneous search for a
-termination argument and a supporting invariant.  For this purpose, several
-tools alternate between calls to a safety prover and calls to a ranking
-function synthesis tool \cite{}.  In order to share more information about
-the state of the termination proofs between the two tools, Brockschmidt et
-al.  use a single representation of the state of the termination proof
-search that both tools operate over, resulting in performance improvements
-\cite{DBLP:conf/cav/BrockschmidtCF13}.  In \cite{DBLP:conf/cav/BradleyMS05},
+termination argument and a supporting invariant.  
+%For this purpose, several tools alternate between calls to a safety prover and calls to a ranking function synthesis tool \cite{}.  
+%In order to share more information about
+%the state of the termination proofs between the two tools, 
+Brockschmidt et al. 
+share the same representation of the state of the termination proof
+between the safety prover and the ranking function synthesis tool \cite{DBLP:conf/cav/BrockschmidtCF13}.  
+In \cite{DBLP:conf/cav/BradleyMS05},
 Bradley et al.  combine the generation of ranking functions with the
@@
 
-While program termination has been extensively studied, much less research has been conducted in the area of proving nontermination.
-Gupta et al. dynamically enumerate lasso-shaped candidate
+While program termination has been extensively studied, much less research has been conducted 
+in the area of proving non-termination. Gupta et al. dynamically enumerate lasso-shaped candidate
 paths for counterexamples, and then statically prove their feasibility \cite{DBLP:conf/popl/GuptaHMRX08}. 

### 2014-07-06 papers/termination/experiments/benchmarks/loop11.c
Benchmark headers

@@ -1,3 +1,7 @@
+/*
+ * Terminates: false
+ */
+
 int main(void) {
   int x;
 

### 2014-07-06 papers/termination/experiments/benchmarks/svcomp25.c

@@ -6,7 +6,7 @@
  * Float:          false
  * Bitvector:      false
  * Lexicographic:  1
- * Terminates:     unk
+ * Terminates:     true
  */
 /*
  * Terminating Program

### 2014-07-06 papers/termination/paper.tex
remove italic font in related work

@@
 
-Key: \tick = terminating, \xmark = non-terminating.
+Key: \tick = terminating, \xmark = non-terminating, ? = unknown (tool terminated with an inconclusive verdict).
 
@@
 
-Given the lack of research in handling \emph{not-linear programs}, as well
-as \emph{programs over bit vectors and floats}, our work focused on covering
+Given the lack of research in handling not-linear programs, as well
+as programs over bit vectors and floats, our work focused on covering
 these areas.  One of the obvious conclusions that can be reached from
@@
 specialise on a certain aspect of termination proving that they can solve
-efficiently.  Conversely to this view, we aim for \emph{generality}, as we
+efficiently.  Conversely to this view, we aim for generality, as we
 do not restrict the form of the synthesised ranking functions, nor the form

### 2014-07-06 papers/termination/paper.tex
figure 1 with related works

@@
   \hline
-  \multirow{4}{*}{Ranking} & Linear lexicographic &  \cite{DBLP:conf/popl/Ben-AmramG13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13,DBLP:conf/vmcai/P04} & - & & &\checkmark&\checkmark&\checkmark&\checkmark\\
-   & Linear non-lexicographic & \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10} & \cite{DBLP:conf/vmcai/BradleyMS05} & && \checkmark~ \cite{DBLP:conf/tacas/CookKRW10} &\checkmark~ \cite{DBLP:conf/tacas/CookKRW10}&\checkmark&\checkmark\\
-   & Non-linear lexicographic & - & - & &&\checkmark&\checkmark&\checkmark&\checkmark\\
-   & Non-linear non-lexicographic & \cite{DBLP:conf/vmcai/BradleyMS05} &  \cite{DBLP:conf/vmcai/BradleyMS05} & &&\checkmark&\checkmark&\checkmark&\checkmark\\
+  \multirow{4}{*}{Ranking} & Linear lexicographic &  \cite{DBLP:conf/popl/Ben-AmramG13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13,DBLP:conf/vmcai/P04} & - & \cite{DBLP:conf/tacas/LeikeH14} & - &\checkmark&\checkmark&\checkmark&\checkmark\\
+   & Linear non-lexicographic & \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10} & \cite{DBLP:conf/vmcai/BradleyMS05} & \cite{DBLP:conf/tacas/LeikeH14} & - & \checkmark~ \cite{DBLP:conf/tacas/CookKRW10} &\checkmark~ \cite{DBLP:conf/tacas/CookKRW10}&\checkmark&\checkmark\\
+   & Non-linear lexicographic & - & - & - & - &\checkmark&\checkmark&\checkmark&\checkmark\\
+   & Non-linear non-lexicographic & \cite{DBLP:conf/vmcai/BradleyMS05} &  \cite{DBLP:conf/vmcai/BradleyMS05} & - & - &\checkmark&\checkmark&\checkmark&\checkmark\\
    \hline

### 2014-07-06 papers/termination/paper.tex
Experiments

--- 2014-07-06 papers/termination/paper.tex
@@
 
-We implemented our algorithm in the {\sc Headshot} tool, which uses the {\sc
-Kalashnikov} program synthesiser for termination.  We ran the tool on 40
-benchmarks taken from the termination category of SVCOMP'15.  The
-experiments were performed on a 4-core 3.3\,GHz i5-2500k with 8\,GB of RAM.
-
-\begin{figure*}
+To evaluate our algorithm, we implemented two tools: one to generate a termination specification
+from a C program, and one to synthesise a proof from the specification.  We ran the resulting
+termination prover on 47 benchmarks taken from the literature and SV-COMP'15~\cite{svcomp}.  To provide a comparison
+point, we also ran {\sc ARMC}~\cite{armc-website} on the same benchmarks.  Each tool was given a time limit of
+180s, and was run on an unloaded 8-core 3.07\,GHz Xeon X5667 with 50\,GB of RAM.  The
+results of these experiments are shown in Figure~\ref{fig:experiments}.
+
+It should be noted that the comparison here is imperfect, since {\sc ARMC} is solving a slightly different
+problem -- it checks whether the program under analysis would terminate if run with unbounded integer variables,
+while we are checking whether the program terminates with fixed width integers variables.  This means that
+{\sc ARMC}'s verdict differs from ours in 3 cases (due to the differences between integer and bit-vector semantics).
+There are a further 5 cases where our tool is able to find a proof and {\sc ARMC} cannot, which we believe is due
+to our more expressive proof language.  In 3 cases, {\sc ARMC} times out while our tool is able to find a termination
+proof.  All of these cases have nested loops, which are easy for us to handle, but can be difficult for
+provers based on enumerating lassos.
+
+On the other hand, {\sc ARMC} is \emph{much} faster than our tool.
+While this difference can partly be explained by much more engineering time being invested in {\sc ARMC},
+we feel that the difference is probably inherent to the difference in the two approaches -- our
+algorithm is complete and bit-vector accurate, which comes at the cost of efficiency.
+
+The final column in Figure~\ref{fig:experiments} shows the number of iterations of the CEGIS loop our
+synthesiser performed.  While the theoretical bound for the number of CEGIS iterations is very large
+(exponential in the amount of space used by the program), these results demonstrate that in practice
+the number of iterations required is usually small.
+
+\begin{figure}
 \centering
-\footnotesize
+\scriptsize
 \begin{tabular}{|l|c||c|r||c|r|r|}
 \hline
-    &             & \multicolumn{2}{|c||}{\sc ARMC} & \multicolumn{3}{|c|}{\sc Headshot} \\
+    &             & \multicolumn{2}{|c||}{\sc ARMC} & \multicolumn{3}{|c|}{Us} \\
     & Expected &  Verdict & Time & Verdict & Time & Iterations\\
@@
 
-\caption{\textsc{Headshot} termination for nonlinear programs with nonlinear ranking functions\label{fig:nonlinear}}
- \end{figure*}
+\caption{Experimental results\label{fig:experiments}}
+ \end{figure}
 

### 2014-07-06 papers/termination/paper.tex
fix fig 1
Change title

@@
 
-\title{Deciding Bitvector Termination with Program Synthesis}
+\title{Synthesising Termination Proofs for Bit-Vector Programs}
 
@@
 
- \caption{Summary of related termination analyses. Legend: \checkmark = we can handle; - = no available works.} \label{fig:handletable}
+ \caption{Summary of related termination analyses. Legend: \checkmark = we can handle; - = no available works; L = linear; NL = non-linear.} \label{fig:handletable}
 \end{figure*}

### 2014-07-06 papers/termination/paper.tex
added reference to TOAST

@@
 
-\title{Synthesising Termination Proofs for Bit-Vector Programs}
+\title{Deciding Bitvector Termination with Program Synthesis}
 
@@
 decidable and a termination proof is expressible as a ground term of the \newC logic, we use
-Counterexample Guided Inductive Synthesis (CEGIS)~\cite{lezama-thesis,sketch} to
+Counterexample Guided Inductive Synthesis (CEGIS)~\cite{lezama-thesis,sketch,DBLP:conf/iclp/BrainCVF06} to
 check the validity of the synthesis formula. 

### 2014-07-06 papers/termination/paper.tex
table squashed

@@
 \scriptsize
-\begin{tabular}{|l|c||c|r||c|r|r|}
+\begin{tabular}{|l|@{}c@{}||@{}c@{}|r||@{}c@{}|r|r|}
 \hline
           &             & \multicolumn{2}{|c||}{\sc ARMC} & \multicolumn{3}{|c|}{Us} \\
-    & Expected &  Verdict & Time & Verdict & Time & Iterations\\
+Benchmark & \,Expected\, & \,Verdict\, & Time & \,Verdict\, & Time & \#It.\\
     \hline

### 2014-07-06 papers/termination/paper.tex
nudge

@@
 
-To evaluate our algorithm, we implemented two tools: one to generate a termination specification
-from a C program, and one to synthesise a proof from the specification.  We ran the resulting
-termination prover on 47 benchmarks taken from the literature and SV-COMP'15~\cite{svcomp}.  To provide a comparison
-point, we also ran {\sc ARMC}~\cite{armc-website} on the same benchmarks.  Each tool was given a time limit of
-180s, and was run on an unloaded 8-core 3.07\,GHz Xeon X5667 with 50\,GB of RAM.  The
+To evaluate our algorithm, we implemented two tools: one to generate a
+termination specification from a C program, and one to synthesise a proof
+from the specification.  We ran the resulting termination prover on 47
+benchmarks taken from the literature and SV-COMP'15~\cite{svcomp}.  To
+provide a comparison point, we also ran {\sc ARMC}~\cite{armc-website} on
+the same benchmarks.  Each tool was given a time limit of $180\,s$, and was
+run on an unloaded 8-core 3.07\,GHz Xeon X5667 with 50\,GB of RAM.  The
 results of these experiments are shown in Figure~\ref{fig:experiments}.
 
-It should be noted that the comparison here is imperfect, since {\sc ARMC} is solving a slightly different
-problem -- it checks whether the program under analysis would terminate if run with unbounded integer variables,
-while we are checking whether the program terminates with fixed width integers variables.  This means that
-{\sc ARMC}'s verdict differs from ours in 3 cases (due to the differences between integer and bit-vector semantics).
-There are a further 5 cases where our tool is able to find a proof and {\sc ARMC} cannot, which we believe is due
-to our more expressive proof language.  In 3 cases, {\sc ARMC} times out while our tool is able to find a termination
-proof.  All of these cases have nested loops, which are easy for us to handle, but can be difficult for
-provers based on enumerating lassos.
-
-On the other hand, {\sc ARMC} is \emph{much} faster than our tool.
-While this difference can partly be explained by much more engineering time being invested in {\sc ARMC},
-we feel that the difference is probably inherent to the difference in the two approaches -- our
-algorithm is complete and bit-vector accurate, which comes at the cost of efficiency.
-
-The final column in Figure~\ref{fig:experiments} shows the number of iterations of the CEGIS loop our
-synthesiser performed.  While the theoretical bound for the number of CEGIS iterations is very large
-(exponential in the amount of space used by the program), these results demonstrate that in practice
-the number of iterations required is usually small.
+It should be noted that the comparison here is imperfect, since {\sc ARMC}
+is solving a slightly different problem -- it checks whether the program
+under analysis would terminate if run with unbounded integer variables,
+while we are checking whether the program terminates with fixed-width
+integer variables.  This means that {\sc ARMC}'s verdict differs from ours
+in 3 cases (due to the differences between integer and bit-vector
+semantics).  There are a further 5 cases where our tool is able to find a
+proof and {\sc ARMC} cannot, which we believe is due to our more expressive
+proof language.  In 3 cases, {\sc ARMC} times out while our tool is able to
+find a termination proof.  All of these cases have nested loops, which are
+easy for us to handle, but can be difficult for provers that enumerate
+lassos.
+
+On the other hand, {\sc ARMC} is \emph{much} faster than our tool.  While
+this difference can partly be explained by much more engineering time being
+invested in {\sc ARMC}, we feel that the difference is probably inherent to
+the difference in the two approaches -- our algorithm is complete and
+bit-vector accurate, which comes at the cost of efficiency.
+
+The final column in Figure~\ref{fig:experiments} gives the number of
+iterations of the CEGIS loop our synthesiser performed.  While the
+theoretical bound for the number of CEGIS iterations is very large
+(exponential in the amount of space used by the program), these results
+demonstrate that in practice the number of iterations required is usually
+small.
 
@@
 %\subsection{Termination analysis}
-Automated program termination is a research topic that has received a fair amount of attention from the software verification community.
-In order to compare our technique to the rest of the area, 
-Figure~\ref{fig:handletable} summarises the related works with respect to the assumptions they make about programs and ranking functions.
-
-Somehow expected, most of the techniques are specialised in the synthesis of linear ranking functions for linear programs 
-over integers (or rationals) \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,
-DBLP:conf/atva/HeizmannHLP13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/KroeningSTW10}. Among them, 
-Lee et al. make use of transition predicate abstraction, algorithmic learning, and decision procedures 
+
+Automated program termination is a research topic that has received a fair
+amount of attention from the software verification community.  In order to
+compare our technique to the rest of the area, Figure~\ref{fig:handletable}
+summarises the related works with respect to the assumptions they make about
+programs and ranking functions.
+
+Somehow expected, most of the techniques are specialised in the synthesis of
+linear ranking functions for linear programs over integers (or rationals)
+\cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,
+DBLP:conf/atva/HeizmannHLP13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/KroeningSTW10}. 
+Among them, Lee et al.  make use of transition predicate abstraction,
+algorithmic learning, and decision procedures
 %to compute transition invariants as proofs for the termination of linear programs 

### 2014-07-06 papers/termination/paper.tex
nudge

@@
 DBLP:conf/atva/HeizmannHLP13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/KroeningSTW10}. 
-Among them, Lee et al.  make use of transition predicate abstraction,
+Among them, Lee et al.~make use of transition predicate abstraction,
 algorithmic learning, and decision procedures
@@
 \cite{DBLP:conf/tacas/LeikeH14}, whereas
-Bradley et al. compute lexicographic linear ranking functions supported by inductive linear invariants 
+Bradley et al.~compute lexicographic linear ranking functions supported by inductive linear invariants 
 %for loops with linear guards and transitions: 
@@
 integers is indeed well-covered in the literature, there is very limited
-work for programs over machine integers.  Cook et al.  present a method
+work for programs over machine integers.  Cook et al.~present a method
 based on a reduction to Presburger arithmetic, and a template-matching
@@
 In an attempt to reduce the complexity of checking the validity of the
-termination argument, Cook et al present an iterative termination proving
+termination argument, Cook et al.~present an iterative termination proving
 procedure that searches for lexicographic termination
-arguments~\cite{DBLP:conf/tacas/CookSZ13}, whereas Kroening et al. 
-strengthen the termination argument such that it becomes a transitive
+arguments~\cite{DBLP:conf/tacas/CookSZ13}, whereas Kroening et
+al.~strengthen the termination argument such that it becomes a transitive
 relation~\cite{DBLP:conf/cav/KroeningSTW10}.
@@
 %the state of the termination proofs between the two tools, 
-Brockschmidt et al. 
-share the same representation of the state of the termination proof
+Brock\-schmidt et al.~share the same representation of the state of the termination proof
 between the safety prover and the ranking function synthesis tool \cite{DBLP:conf/cav/BrockschmidtCF13}.  
 In \cite{DBLP:conf/cav/BradleyMS05},
-Bradley et al.  combine the generation of ranking functions with the
+Bradley et al.~combine the generation of ranking functions with the
 generation of invariants to form one constraint solving problem such that
@@
 
-While program termination has been extensively studied, much less research has been conducted 
-in the area of proving non-termination. Gupta et al. dynamically enumerate lasso-shaped candidate
-paths for counterexamples, and then statically prove their feasibility \cite{DBLP:conf/popl/GuptaHMRX08}. 
-In \cite{DBLP:conf/tacas/ChenCFNO14}, Chen et al. prove nontermination via
-reduction to safety proving. Their iterative algorithm uses counterexamples to a fixed safety property
-to refine an underapproximation of a program.
+While program termination has been extensively studied, much less research
+has been conducted in the area of proving non-termination.  Gupta et
+al.~dynamically enumerate lasso-shaped candidate paths for counterexamples,
+and then statically prove their
+feasibility~\cite{DBLP:conf/popl/GuptaHMRX08}. 
+In~\cite{DBLP:conf/tacas/ChenCFNO14}, Chen et al.~prove nontermination via
+reduction to safety proving.  Their iterative algorithm uses counterexamples
+to a fixed safety property to refine an underapproximation of a program.
 

### 2014-07-06 papers/termination/paper.tex
nudge related

@@
 \section{Conclusions and Related Work}
-%\subsection{Termination analysis}
-
-Automated program termination is a research topic that has received a fair
-amount of attention from the software verification community.  In order to
-compare our technique to the rest of the area, Figure~\ref{fig:handletable}
-summarises the related works with respect to the assumptions they make about
-programs and ranking functions.
-
-Somehow expected, most of the techniques are specialised in the synthesis of
-linear ranking functions for linear programs over integers (or rationals)
-\cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,
-DBLP:conf/atva/HeizmannHLP13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/KroeningSTW10}. 
-Among them, Lee et al.~make use of transition predicate abstraction,
-algorithmic learning, and decision procedures
-%to compute transition invariants as proofs for the termination of linear programs 
-\cite{DBLP:conf/cav/LeeWY12},
-Leike and Heizmann propose linear ranking templates 
-%in order to synthesise
-%termination arguments for linear loop programs 
-\cite{DBLP:conf/tacas/LeikeH14}, whereas
-Bradley et al.~compute lexicographic linear ranking functions supported by inductive linear invariants 
-%for loops with linear guards and transitions: 
-\cite{DBLP:conf/cav/BradleyMS05}. 
-%Learning: \cite{DBLP:conf/cav/HeizmannHP14}
+
+There has been substantial prior work on automated program termination
+analysis.  Figure~\ref{fig:handletable} summarises the related work with
+respect to the assumptions they make about programs and ranking functions. 
+Most of the techniques are specialised in the synthesis of linear ranking
+functions for linear programs over integers (or
+rationals)~\cite{DBLP:conf/pldi/CookPR06, DBLP:conf/cav/LeeWY12,
+DBLP:conf/popl/Ben-AmramG13, DBLP:conf/vmcai/P04,
+DBLP:conf/atva/HeizmannHLP13, DBLP:conf/cav/BradleyMS05,
+DBLP:conf/tacas/CookSZ13, DBLP:conf/cav/KroeningSTW10}.  Among them, Lee et
+al.~make use of transition predicate abstraction, algorithmic learning, and
+decision procedures~\cite{DBLP:conf/cav/LeeWY12}, Leike and Heizmann propose
+linear ranking templates~\cite{DBLP:conf/tacas/LeikeH14}, whereas Bradley et
+al.~compute lexicographic linear ranking functions supported by inductive
+linear invariants~\cite{DBLP:conf/cav/BradleyMS05}.
 
 While the synthesis of termination arguments for linear programs over
-integers is indeed well-covered in the literature, there is very limited
+integers is indeed well covered in the literature, there is very limited
 work for programs over machine integers.  Cook et al.~present a method
@@
 
-Given the lack of research in handling not-linear programs, as well
-as programs over bit vectors and floats, our work focused on covering
-these areas.  One of the obvious conclusions that can be reached from
-observing Figure~\ref{fig:handletable}, is that most of the works tend to
-specialise on a certain aspect of termination proving that they can solve
-efficiently.  Conversely to this view, we aim for generality, as we
-do not restrict the form of the synthesised ranking functions, nor the form
-of the input programs.
+Given the lack of research in handling not-linear programs, as well as
+programs over bit vectors and floats, our work focused on covering these
+areas.  One of the obvious conclusions that can be reached from
+Figure~\ref{fig:handletable} is that most methods tend to specialise on a
+certain aspect of termination proving that they can solve efficiently. 
+Conversely to this view, we aim for generality, as we do not restrict the
+form of the synthesised ranking functions, nor the form of the input
+programs.
 
@@
 theorem compute a set of local termination conditions that decrease as
-execution follows through the loop and require expensive reachability
-analyses~\cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}. 
-In an attempt to reduce the complexity of checking the validity of the
-termination argument, Cook et al.~present an iterative termination proving
-procedure that searches for lexicographic termination
-arguments~\cite{DBLP:conf/tacas/CookSZ13}, whereas Kroening et
+execution proceeds through the loop and require expensive reachability
+analyses~\cite{DBLP:conf/lpe/CodishG03, DBLP:conf/lics/PodelskiR04,
+DBLP:conf/pldi/CookPR06}.  In an attempt to reduce the complexity of
+checking the validity of the termination argument, Cook et al.~present an
+iterative termination proving procedure that searches for lexicographic
+termination arguments~\cite{DBLP:conf/tacas/CookSZ13}, whereas Kroening et
 al.~strengthen the termination argument such that it becomes a transitive
@@
 Proving program termination implies the simultaneous search for a
-termination argument and a supporting invariant.  
-%For this purpose, several tools alternate between calls to a safety prover and calls to a ranking function synthesis tool \cite{}.  
-%In order to share more information about
-%the state of the termination proofs between the two tools, 
-Brock\-schmidt et al.~share the same representation of the state of the termination proof
-between the safety prover and the ranking function synthesis tool \cite{DBLP:conf/cav/BrockschmidtCF13}.  
-In \cite{DBLP:conf/cav/BradleyMS05},
-Bradley et al.~combine the generation of ranking functions with the
-generation of invariants to form one constraint solving problem such that
-the necessary supporting invariants for the ranking function are discovered
-on demand.  In our setting, both the ranking function and the supporting
-invariant are iteratively constructed through the same refinement loop.
+termination argument and a supporting invariant.  Brock\-schmidt et
+al.~share the same representation of the state of the termination proof
+between the safety prover and the ranking function synthesis
+tool~\cite{DBLP:conf/cav/BrockschmidtCF13}. 
+In~\cite{DBLP:conf/cav/BradleyMS05}, Bradley et al.~combine the generation
+of ranking functions with the generation of invariants to form a single
+constraint solving problem such that the necessary supporting invariants for
+the ranking function are discovered on demand.  In our setting, both the
+ranking function and the supporting invariant are iteratively constructed in
+the same refinement loop.
 

### 2014-07-06 papers/termination/paper.tex
Overfull hboxes

@@
 
-Additionally, if the program under analysis has a non-deterministic transition relation
-and is non-terminating, our procedure is not guaranteed to find a non-termination proof.
+Additionally, if the program under analysis is non-terminating and has a
+non-deterministic transition relation, our procedure is not guaranteed to find a non-termination proof.
 
@@
 \begin{lstlisting}
-while (x > 0 && y > 0 && z > 0) {
+while (x>0 && y>0 && z>0) {
   if (y > x) {
@@
 \begin{framed}
-\begin{minipage}{0.17\textwidth}
+\begin{minipage}{0.145\textwidth}
 \begin{lstlisting}[mathescape=true]
@@
 $L_1:$
-while(x <= y)
+while(x<=y)
   x++;
@@
 $L_2:$
-while(x > 0){
+while(x>0){
   z = x;
@@
 $L_3:$
-  while(v < y){
+  while(v<y){
     z--;
@@
 \begin{framed}
- \begin{minipage}{0.19\textwidth}
+ \begin{minipage}{0.16\textwidth}
 \begin{lstlisting}[mathescape=true]

### 2014-07-06 papers/termination/paper.tex
Edited introduction

@@
 Proving program termination is typically done by finding a well-founded
-\emph{ranking function} for the program states.  Existing termination
+ranking function for the program states.  Existing termination
 provers typically find ranking functions using either linear algebra or
 templates.  As such they are often restricted to finding linear ranking
-functions over mathematical integers.  This class of ranking functions is
+functions over mathematical integers.  This class of functions is
 insufficient for proving termination of many terminating programs, and
@@
 We present a reduction from program \emph{termination} to program
-\emph{synthesis}.  This reduction allows us to generate nonlinear,
-lexicographic ranking functions that are correct for fixed-width machine
-arithmetic and floating-point arithmetic.  We use this reduction to build a
-sound and complete procedure for the termination of fixed-width and
-floating-point arithmetic programs.
+\emph{synthesis} via a fragment of second-order logic.  This reduction
+allows us to generate nonlinear, lexicographic ranking functions and
+nonlinear recurrence sets that are correct for fixed-width machine arithmetic
+and floating-point arithmetic.  We use this reduction to build a sound and
+complete procedure for the termination of fixed-width and floating-point
+arithmetic programs.
 %
@@
 \keywords
-Termination, Program Synthesis, Lexicographic Ranking Functions, Bitvector Ranking Functions,
+Termination, Non-Termination,
+Program Synthesis, Lexicographic Ranking Functions, Bitvector Ranking Functions,
 Floating-Point Ranking Functions.
@@
 capture the behaviour of physical computers.  In this paper, we present a
-sound and complete method for deciding whether a program terminates.
+sound and complete method for deciding whether a program with a fixed amount of
+storage terminates.
 
@@
 theorem~\cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}.
-Briefly, the transition relation is shown to be disjunctively well founded
-by finding a set of simple functions, each of which proves well foundedness
-for part of the transition relation.  The main benefit of this approach is
+The latter proof style builds an argument that a transition relation is disjunctively well founded
+by composing several small well-foundedness arguments.
+The main benefit of this approach is
 the simplicity of local termination measures in contrast to global ones. 
@@
 non-linearity and lexicographic orders: we do not commit to any particular
-syntactic form, and do not use templates.
+syntactic form, and do not use templates.  Furthermore, our approach allows
+us to simultaneously search for proofs of non-termination, which take the form
+of recurrence sets.
 
@@
 i.e.  programs that terminate for integers may \emph{not} terminate for
-bit-vectors and vice versa.  Thus, abstracting bit-vectors to integers may
+bit-vectors and vice versa.  Thus, abstracting bit-vectors with integers may
 give rise to {\em unsound} and {\em incomplete} analyses.
 
-We propose a general framework that uniformly computes lexicographic,
-non-linear ranking functions supported by non-linear inductive invariants
-for loops with arbitrary guards and transitions over bit-vectors and floats. 
-It also computes non-linear recurrence sets, which comprise non-termination
-proofs.  Our encoding treats multiple loops and nested loops uniformly,
-without having to enumerate lassos.
-
- For this purpose, we phrase the termination and non-termination problems as
-second-order satisfaction problems.  We identify a fragment of second-order
-logic that is expressive enough to capture our formulation, which we call
-the \emph{synthesis fragment}.  Subsequently, we propose a method for
-checking the satisfiability of a formula in the synthesis fragment by
+% we propose a general framework that uniformly computes lexicographic,
+% non-linear ranking functions supported by non-linear inductive invariants
+% for loops with arbitrary guards and transitions over bit-vectors and floats. 
+% it also computes non-linear recurrence sets, which comprise non-termination
+% proofs.  our encoding treats multiple loops and nested loops uniformly,
+% without having to enumerate lassos.
+
+Our approach is to phrase the termination and non-termination problems as
+second-order satisfaction.  We identify a fragment of second-order
+logic that is expressive enough to capture both problems, while being
+restrictive enough to admit effective solvers.  Subsequently, we propose a method for
+checking the satisfiability of a formula in this fragment by
 solving an isomorphic program synthesis problem
-(Section~\ref{sec:synthesis}).  Our method is sound and semi-complete, i.e. 
-if a pair of a ranking function and a supporting invariant exist it will be
-found, but if the program does not terminate we may not find a recurrence
-set.  However, if the program's transition relation is deterministic, our
-method is complete.
-
-With respect to the performance of our technique, we show that its runtime is dominated by the size of 
-the shortest termination proof.
-
-This means that our procedure is not directly dependent on the structure of
-the analysed loop (e.g.~the number of variables or the number of lassos),
-but on the \emph{Kolmogorov complexity of its minimal termination argument}. 
-We argue that the length of a program's termination proof is indicative of
-how easy to understand the program is, and thus programmers tend to write
-programs that have relatively short termination proofs.  We empirically show
-that this claim holds in practice, causing our technique to perform well.
+(Section~\ref{sec:synthesis}).  Our method is sound and
+complete for deterministic bit-vector programs -- for any program, we will find
+either a proof of its termination or non-termination.  For non-deterministic
+programs, we are only semi-complete -- if the program terminates, we will find a proof,
+but if it does not terminate we may fail to discover a recurrence set.
+
+% With respect to the performance of our technique, we show that its runtime is dominated by the size of 
+% the shortest termination proof.
+% 
+% This means that our procedure is not directly dependent on the structure of
+% the analysed loop (e.g.~the number of variables or the number of lassos),
+% but on the \emph{Kolmogorov complexity of its minimal termination argument}. 
+% We argue that the length of a program's termination proof is indicative of
+% how easy to understand the program is, and thus programmers tend to write
+% programs that have relatively short termination proofs.  We empirically show
+% that this claim holds in practice, causing our technique to perform well.
 
@@
 \item We designed a bit-level accurate technique for computing ranking
-functions that correctly accounts for the wrap-around behavior caused by
-under- and overflows in bit-vector and floating-point arithmetic.  Our
+functions and recurrence sets that correctly accounts for the wrap-around
+behavior caused by under- and overflows in bit-vector and floating-point arithmetic.  Our
 technique is not restricted to finding linear ranking functions, but can
@@
 \item  We rephrased the termination and non-termination problems as
-second-order satisfaction problems and used program synthesis techniques to
-solve them.  In contrast to previous approaches to program synthesis, we
-found that using genetic programming was very effective at generating
-candidate termination and non-termination proofs.
-
-\item Our procedure does not require the enumeration of lassos, or directly
-reasoning about the transitive closure of the transition relation.  To
-verify our proofs, we only require a single call to a SAT solver.
-
-\item Our procedure is based on a series of source-to-source
+second-order $\exists \forall$ satisfaction problems.  This formulation
+captures the (non-)termination
+properties of all of the loops in the program simultaneously, including
+nested loops.
+% In contrast to previous approaches to program synthesis, we
+% found that using genetic programming was very effective at generating
+% candidate termination and non-termination proofs.
+
+\item We used program synthesis techniques to solve the second-order satisfaction
+problems generated by our encoding.  Our solver requires only a single call to a SAT solver to verify its
+proof.  It is not based on enumerating lassos, nor does it require a safety prover
+that can reason about transitive closure.
+
+\item Our solver is based on a series of source-to-source
 transformations.  As well as leading to a concise implementation, this has
-the side effect of ensuring that our proof is correct for the exact binary
+the side effect of ensuring that our proofs are correct for the exact binary
 produced by a particular compiler, rather than for an abstraction of the
@@
 
-\item We propose a non-trivial formulation of the (non-)termination problem inside a
- second-order logic fragment with only one quantifier alternation.
-
 \end{itemize}
@@
 {\bf Limitations.}
-Our algorithm decides termination of transition systems with finite state
+Our algorithm proves termination for transition systems with finite state
 spaces.  The (non-)termination proofs take the form of ranking functions and
@@
 formalism is powerful enough to handle a large fragment of C, but is not
-rich enough to analyse code using any of the following features:
-
-\begin{itemize}
-\item Recursion (could probably do this).
-\item Heap (Thor).
-\item Arrays.
-\end{itemize}
-
-Additionally, if the program under analysis is non-terminating and has a
-non-deterministic transition relation, our procedure is not guaranteed to find a non-termination proof.
+rich enough to analyse code which uses the heap or arrays.  Also, we have not
+yet added support for recursion to our encoding.
 

### 2014-07-07 papers/termination/paper.tex
Edits

@@
 Given a program, we first formalise its termination argument as a ranking
-function (Section~\ref{sec:ranking.functions}).  Subsequently, we discus
+function (Section~\ref{sec:ranking.functions}).  Subsequently, we discuss
 bit-vector semantics and illustrate differences between machine arithmetic
 and integer arithmetic that show that the abstraction of bit vectors to
-mathematical integers unsound (Section~\ref{sec:machine.arith}).
+mathematical integers is unsound (Section~\ref{sec:machine.arith}).
 
@@
 
-A programs $P$ is represented as a transition system with state space $X$ and
+A program $P$ is represented as a transition system with state space $X$ and
 transition relation $T \subseteq X \times X$.  For a state
 $x \in X$ with $T(x,x')$ we say $x'$ is a successor of $x$ under $T$.
-If each state $x$ has at most one successor $x'$, then we say that $T$ is
-deterministic.
 
@@
 %
-A transition system is said to be \emph{unconditionally terminating} if
+A program is said to be \emph{unconditionally terminating} if
 there is no infinite sequence of states $x_1, x_2, \ldots \in X$ with
@@
 
-We can prove that the transition system is unconditionally terminating by
+We can prove that the program is unconditionally terminating by
 finding a ranking function for its transition relation.
@@
 An injective function ${R:X\to Y}$ is a \emph{ranking function} for the
-transition relation $T$ if $Y$ is a well-founded set with order $<$ and 
+transition relation $T$ if $Y$ is a well-founded set with order $>$ and 
 $R$ is monotonically decreasing with respect to $T$.  That is
 to say:
-$$\forall x, x' \in X. T(x, x') \Rightarrow R(x) < R(x')$$
+$$\forall x, x' \in X. T(x, x') \Rightarrow R(x) > R(x')$$
 %
@@
 
-\begin{definition}[Linear ranking function]
-A \emph{linear ranking function} $R: X \to Y$ 
+\begin{definition}[Linear function]
+A \emph{linear function} $f: X \to Y$ 
 with $\dim(X) = n$ and $\dim(Y) = m$ is of the form: $$f(\vec{x}) = M\vec{x}$$ where
@@
 The total order imposed on $Y$ is the lexicographic ordering
-induced on tuples of $Z$'s. 
+induced on tuples of $Z$'s.  So for $y = (z_1, \ldots, z_m)$ and
+$y' = (z'_1, \ldots, z'_m)$:
+\[
+ y > y' \iff \exists i \leq m . z_i > z'_i \wedge \forall j < i . z_j = z'_j
+\]
+
 \end{definition}
 
-For lexicographic ranking functions, no function may increase unless a function of a lower index decreases.
-Additionally, at every step, there must be at least one function that decreases.
-We note that some termination arguments require lexicographic ranking functions, or equivalently, ranking functions
-whose co-domain is the ordinals, rather than just $\mathbb{N}$.
+We note that some termination arguments require lexicographic ranking functions, or
+equivalently, ranking functions whose co-domain is a countable ordinal, rather than just $\mathbb{N}$.
 
@@
 
-In fact, we identify a fragment of second-order logic with a constrained use
-of quantification that is expressive enough to encode both termination and
-non-termination.  We will suggestively refer to the fragment as the
-\emph{synthesis fragment}:
+Due to its expressiveness, second-order logic is very difficult to reason
+about, with many second-order theories becoming undecidable even when the
+corresponding first-order theory is decidable.
+To make solutions of our constraints tractable, we identify a fragment of second-order logic
+with a constrained use of quantification that is expressive enough to encode
+both termination and non-termination, while admitting a semi-decision procedure.
+We will suggestively refer to the fragment as the \emph{synthesis fragment}:
 
@@
 %
-where $\vec{P}$ ranges over sets, while $\vec{x}$ and $\vec{y}$ range over
+where $\vec{P}$ ranges over functions, while $\vec{x}$ and $\vec{y}$ range over
 ground terms.  Function $\sigma: (X^n \times Y^m \to Z^k) \times X^n \times
@@
 for all $\vec{y}$. 
-If a pair $(\vec{P_0}, \vec{x})$ is a solution to the synthesis problem, then we write $(\vec{P_0}, \vec{x}) \models \sigma$.
+If a pair $(\vec{P}, \vec{x})$ is a solution to the synthesis problem, then we write $(\vec{P}, \vec{x}) \models \sigma$.
 For the remainder of the presentation, we drop the $\vec{x}$ and instead just write $x$, with the understanding
-that all variables range over vectors.
+that all quantified variables range over vectors.
 
@@
 
-\subsection{An Isolated Loop}
-
-We will begin our discussion by showing how to encode the
-\mbox{(non-)termination} of a program consisting of a single loop in the
-synthesis fragment.  For the time being, a loop is a pair ${L= \langle G,T
-\rangle}$, where the states $x$ satisfying the loop's guard are given by the
+\subsection{An Isolated, Simple Loop}
+
+We will begin our discussion by showing how to encode in the synthesis fragment the
+\mbox{(non-)termination} of a program consisting of a single loop with no nesting.
+For the time being, a loop $L(G, T)$ is defined by a predicate $G$ and relation $T$.
+G states $x$ satisfying the loop's guard are given by the
 predicate $G(x)$.  The body of the loop is encoded as the transition
@@
 \begin{align*}
- \exists R . \forall x, x' . & G(x) \wedge T_o(x, x') \rightarrow R(x) > 0 \wedge R(x) > R(x')
+ \exists R . \forall x, x' . & G(x) \wedge T(x, x') \rightarrow R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
@@
   \exists N, x_0 . \forall x . \exists x' . & N(x_0) ~\wedge \\ &  N(x) \rightarrow G(x) ~ \wedge \\
-							& N(x) \rightarrow T_u(x, x') \wedge N(x') 
+							& N(x) \rightarrow T(x, x') \wedge N(x') 
  \end{align*}
@@
   \exists N, x_0 . \forall x, x' . & N(x_0) ~ \wedge \\ & N(x) \rightarrow G(x) ~ \wedge \\
-							& N(x) \wedge T_u(x, x') \rightarrow N(x') 
+							& N(x) \wedge T(x, x') \rightarrow N(x') 
  \end{align*}
@@
 \noindent {\bf Unconditional termination.}
-To show that $L$ is unconditionally terminating (i.e. it eventually terminates regardless of the
-state it starts in), we try to find a ranking function for it. When defining $L$'s termination 
-formula, we need only consider the over-approximation of its transition relation $T$, which we call $T_o$.
-Note that this distinction is only important if $L$'s body contains nested loops, for which we'll have to 
-consider the over-approximation of their transition relation's transitive closure
-as shown in Section~\ref{sec:env}. For any other statements, $T_o$ and $T$ are equivalent.
+To show that $L(G, T)$ is unconditionally terminating (i.e. it eventually terminates regardless of the
+state it starts in), it suffices to find a ranking function for \mbox{$T \cap (G \times X)$},
+i.e. $T$ restricted to states satisfying the loop's guard.
+% If $L$ contains nested loops, we replace $T$ with an over-approximation of the loop body $T_o$,
+% as we will discuss in Section~\ref{sec:env}.
+
+% When defining $L$'s termination 
+% formula, we need only consider an over-approximation of its transition relation $T$, which we call $T_o$.
+% Note that this distinction is only important if $L$'s body contains nested loops, for which we'll have to 
+% consider the over-approximation of their transition relation's transitive closure
+% as shown in Section~\ref{sec:env}. For any other statements, $T_o$ and $T$ are equivalent.
 
@@
 \label{thm:ut}
- The loop $L(G, T_o)$ terminates from every start state iff formula {\bf [UT]} (Definition~\ref{def:UT}) is
- satisfiable.
+ The loop $L(G, T)$ terminates from every start state iff formula {\bf [UT]} (Definition~\ref{def:UT}) is satisfiable.
 \end{theorem}
 
-\begin{proof}
- Then the codomain of $R$ is a well founded set and $R$ is an order homomorphism.
-
- (Do this proof properly).
-\end{proof}
+% \begin{proof}
+%  Then the codomain of $R$ is a well founded set and $R$ is an order homomorphism.
+% 
+%  (Do this proof properly).
+% \end{proof}
 
 As the existence of a ranking function is equivalent to the satisfiability
-of the formula {\bf [UT]}, a satisfiability witness is also a ranking
+of the formula {\bf [UT]}, a satisfiability witness is a ranking
 function and thus a proof of $L$'s unconditional termination.
 
-Continuing the example of the program in Figure~\ref{fig:motivation.a}, we
+Returning to the program from Figure~\ref{fig:motivation.a}, we
 can see that the corresponding synthesis formula {\bf [UT]} is satisfiable,
@@
 that the second-order logic has a primitive predicate for disjunctive
-well-foundedness.  Conversely, our formulation in Definition~\ref{def:UT}
-does not require a transitive closure operator, nor a disjunctive
-well-foundedness predicate.  \\
+well-foundedness.  In contrast, our formulation in Definition~\ref{def:UT}
+does not use a primitive disjunctive well-foundedness predicate.  \\
 
 \noindent{\bf Non-termination.}
-Dually to termination, we might want to consider the non-termination of a program.  If a program terminates,
+Dually to termination, we might want to consider the non-termination of a loop.  If a loop terminates,
 we can prove this by finding a ranking function %and supporting invariant 
 witnessing the satisfiability of formula {\bf[UT]}.  What then would a proof of non-termination look like?
+
+Since our program's state space is finite, a transition relation
+induces a infinite execution iff some state is visited infinitely
+often, or equivalently $ \exists x . T^+(x, x)$.
+Deciding satisfiability of this formula directly would require a logic
+that includes a transitive closure operator, $T^+$.  Rather than
+introduce such an operator, we will characterise non-termination
+using a synthesis formula.
 
@@
 set}, i.e.  a nonempty set of states $N$ such that for each $s \in N$ there
-exists a transition to some $s'\in N$.
-For our case, the notion of open recurrent set is encoded by formula {\bf [ONT]}. %Definition~\ref{def:nonterm-formula}.
-Dual to the termination problem, when formulating $L$'s non-termination, we only need to consider an under-approximation of its transition
-relation~$T_u$. 
-Again, this distinction is mostly important when dealing with nested loops,
-as illustrated in Section~\ref{sec:env}.
+exists a transition to some $s' \in N$.
+The notion of open recurrence set is encoded by formula {\bf [ONT]} in Definition~\ref{def:ont}.
+% Dual to the termination problem, when formulating $L$'s non-termination, we only need to consider an under-approximation of its transition
+% relation~$T_u$. 
+% Again, this distinction is mostly important when dealing with nested loops,
+% as illustrated in Section~\ref{sec:env}.
 
@@
 \label{thm:ont}
- The loop $L(G, T_u)$ has an infinite execution iff Defintion~\ref{def:ont} is satisfiable.
+ The loop $L(G, T)$ has an infinite execution iff Defintion~\ref{def:ont} is satisfiable.
 \end{theorem}
 
-\begin{proof}
- Then $x_0$ is an underapproximation to the reachable states on entry to $L$, and $N$ is an open recurrence set.
-
- (Do this proof properly).
-\end{proof}
+% \begin{proof}
+%  Then $x_0$ is an under-approximation to the reachable states on entry to $L$, and $N$ is an open recurrence set.
+%  
+%  (Do this proof properly).
+% \end{proof}
 
@@
 
-In order to formulate the non-termination problem inside the synthesis
+In order to express the non-termination problem inside the synthesis
 fragment, we make use of the notion of \emph{closed recurrence set}
-introduced by Chen et al.~in~\cite{DBLP:conf/tacas/ChenCFNO14}.  As opposed
-to an open recurrence set, this is an $\exists \forall$ property: for each
-state in the recurrence set $N$, all of its successors must be in $N$.  We
-note that if $T$ is deterministic, every open recurrence set is also a
+introduced by Chen et al.~in~\cite{DBLP:conf/tacas/ChenCFNO14}:  for each
+state in the recurrence set $N$, \emph{all} of its successors must be in $N$.
+The existence of a closed recurrence set is equivalent to the satisfiability
+of formula {\bf [CNT]} in Definition~\ref{def:cnt}, which is in the synthesis
+fragment.
+
+We note that if $T$ is deterministic, every open recurrence set is also a
 closed recurrence set (since each state has at most one successor).  Thus,
@@
 
+\begin{definition}[Determinism]
+ A relation $T$ is deterministic iff each state $x$ has exactly one successor under $T$:
+ $$\forall x . \exists x' . T(x, x') \wedge \forall x'' . T(x, x'') \rightarrow x'' = x'$$
+\end{definition}
+
+In order to describe a deterministic \emph{program} in a way that still allows us
+to sensibly talk about termination, we assume the existence of a special sink
+state $s$ with no outgoing transitions and such that $\lnot G(s)$ for any
+of the loop guards $G$.  The program is deterministic if its transition
+relation is determinstic for all states except $s$.
+
 \begin{theorem}
 \label{thm:cnt}
- If $T_u$ is deterministic, i.e. each state $x$ for which the guard $G(x)$ holds has exactly one successor $x'$ such that $T_u(x, x')$, then
- Definition~\ref{def:ont} is equivalent to Definition~\ref{def:cnt}.
+ If $T$ is deterministic,
+ Definition~\ref{def:ont} and Definition~\ref{def:cnt} are equisatisfiable.
 \end{theorem}
 
-\begin{proof}
- If $T_u$ is deterministic, then each $x$ satisfying $G(x)$ has a unique successor $S(x)$, so $\forall x, x' . G(x) \wedge T_u(x, x') \leftrightarrow x' = S(x)$.
-
- (Finish this proof).
-\end{proof}
-
-Note that by defining $T_u$'s determinism as the existence of exactly one
-successor for any state satisfying $L$'s guard, we eliminate the possibility
-of $T_u$ being $\mathit{false}$, which would otherwise constitute a
-counterexample for Theorem~\ref{thm:cnt}.
-
-Given that our program's state space is finite, a transition relation
-induces a infinite execution if and only if some state is visited infinitely
-often, or equivalently the state is visited twice in the course of the
-execution.  This gives us an alternative way of expression a non-termination
-proof: we can can try to find a witness $x$ to the satisfiability of the
-formula
-%
-$$ \exists x . T^+(x, x)$$
-%
-However, deciding satisfiability of this formula directly requires a logic
-that includes a transitive closure operator, which ours does not have.
-
-\noindent {\bf Non-determinism.} A non-termination diagnosis procedure based
-on checking the satisfiability of formula {\bf [CNT]} may fail to find a
-closed recurrence set for a loop $L=(G, T_u)$ when $T_u$ is
-non-deterministic.  To understand the issue, imagine the following loop with
-a non-deterministic assignment to \lstinline!y!:
+% \begin{proof}
+%  If $T_u$ is deterministic, then each $x$ satisfying $G(x)$ has a unique successor $S(x)$, so $\forall x, x' . G(x) \wedge T_u(x, x') \leftrightarrow x' = S(x)$.
+%  
+%  (Finish this proof).
+% \end{proof}
+
+% Note that by defining $T_u$'s determinism as the existence of exactly one
+% successor for any state satisfying $L$'s guard, we eliminate the possibility
+% of $T_u$ being $\mathit{false}$, which would otherwise constitute a
+% counterexample for Theorem~\ref{thm:cnt}.
+
+So if our transition relation is deterministic we can say, without
+loss of generality, that non-termination of the loop is equivalent
+to the existence of a closed recurrence set.  However if $T$ is
+non-deterministic, it may be that there is an open recurrence
+set but not closed recurrence set.  To see this, consider the following
+loop:
 %
@@
 
-Only for certain choices of \texttt{y} the loop is non-terminating.  In
-order to restrict the values that may be assigned to \texttt{y}, we need an
-assumption associated with the non-deterministic choice.  However, such an
-assumption may cause a reachable state at the non-deterministic assignment
-to not have any successor.  In such a case the termination formula would be
-satisfiable, although the execution halts.  This situation can be avoided by
-checking the existence of at least one successor for each reachable state at
-the non-deterministic assignment.  Although we did put a reasonable amount
-of effort into it, we failed to come up with a formulation for the
-non-termination of such non-deterministic transition relations inside the
-synthesis fragment.
-
-%formula {\bf [Nondet-CNT]} in Definition~\ref{def:nondet-cnt}.
-
-\begin{figure*}
- \begin{framed}
-
-\begin{definition}[Conditional Termination Formula {\bf [CT]}]
-\label{def:ct}
- \begin{align*}
-  \exists R, W . \forall x, x' . & I(x) \wedge G(x) \rightarrow W(x) ~ \wedge \\
-                                 & G(x) \wedge W(x) \wedge T_o(x, x') \rightarrow W(x') \wedge R(x) > 0 
-  \wedge R(x) > R(x')
- \end{align*}
-\end{definition}
-
-%%  \begin{definition}[Sequential Loops Termination Formula {\bf [ST]}]
-%%   \begin{align*}
-%% \label{def:multi-termination-formula}
-%%   \exists R, Inv_1,..., Inv_n . \forall x_0,...,x_n, x_1',...,x_n'.  & P_0(x_0,x_1) \rightarrow Inv_1(x_1) ~  \\
-%%  & \bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) ~  \\  
-%%  & \bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) \rightarrow Inv_{i+1}(x_{i+1})) ~ \wedge \\
-%%  & Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge G_n(x_n) \rightarrow Inv_n(x_n) ~ \wedge \\
-%%  & Inv_n(x_n) \wedge G_n(x_n) \wedge T(x_n, x_n') \rightarrow Inv_n(x_n') \wedge R(x_n) > 0 \wedge R(x_n) > R(x_n')
-%%  \end{align*}
-%% \end{definition}
-
-%% \begin{definition}[Nested Loops Termination Formula {\bf [NT]}]
-%% \label{def:nested-term-formula}
-%%  \begin{align*}
-%%   \exists R, S . \forall x, x' . & G_1(x) \wedge P_1(x,x') \rightarrow S(x',x') ~ \wedge \\
-%%                                 & G_2(x') \wedge S(x,x') \wedge T(x',x'')\rightarrow S(x,x'') ~ \wedge \\
-%% 				& G_1(x) \wedge P_1(x,x') \wedge \neg G_2(x') \wedge S(x',x'') \wedge 
-%%                                 P_2(x'',x''') \wedge 
-%%                                   R(x) > 0 \wedge R(x) > R(x''') 
-%%  \end{align*} 
-%% \end{definition}
-
-%% \begin{definition}[Sequential Loops Non-Termination Formula {\bf [SNT]}]
-%% \label{def:multi-termination-formula}
-%%  \begin{align*}
-%%   \exists R, Inv_1,..., Inv_n . \forall x_0,...,x_n, x_1',...,x_n'.  & P_0(x_0,x_1) \rightarrow Inv_1(x_1) ~  \\
-%%  & \bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) ~  \\  
-%%  & \bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) \rightarrow Inv_{i+1}(x_{i+1})) ~ \wedge \\
-%%  & Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge N(x_n)  ~ \wedge \\
-%%                                                         & N(x_n) \rightarrow G_n(x_n) ~ \wedge 
-%% 							 N(x_n) \wedge T_n(x_n, x_n') \rightarrow N(x_n') 
-%%  \end{align*}
-%% \end{definition}
- \end{framed}
-
-\end{figure*}
-
-\subsection{Composing a Loop with its Environment} \label{sec:env}
-
-Sometimes the termination behaviour of a loop depends on its environment, i.e. the rest of the program.  That is to say,
-the loop may not terminate if started in some particular state, but the rest of the program
-ensures that such a state cannot be reached on entry to the loop.  The program as a whole
-terminates, but if the loop were considered in isolation we would not be able to prove that
-it terminates. We must therefore encode a loop's interaction with its environment %the rest of the program 
-in order to do a sound termination analysis.\\
-
-\iffalse
-Let us assume that we have done some preprocessing of our program which has identified
-loops, straight line code blocks and the control flow between these.  In particular,
-the control flow analysis has determined which order these code blocks execute in,
-and the nesting structure of the loops.
-\fi
-
-\noindent {\bf Conditional termination.}
-Given a loop $L=(G,T_o)$, where $T_o$ is an over-approximation of its transition relation, 
-if $L$'s termination depends on the state it begins
-executing in, we say that $L$ is \emph{conditionally terminating}.
-The information we require of the environment is a predicate $I$ which
-overapproximates the set of states that $L$ may begin executing in.
-That is to say, for each state $x$ that is reachable on entry to $L$,
-we have $I(x)$.
-
-Conditional termination is then equivalent to the
-formula {\bf [CT]} of Defintion~\ref{def:ct}.
-
-\begin{theorem}
-\label{thm:ct}
- The loop $L(G, T_o)$ terminates when started in any state satisfying $I(x)$ iff
- Defintion~\ref{def:ct} is satisfiable.
-\end{theorem}
-
-\begin{proof}
- Do this proof.
-\end{proof}
-
-If this formula is satisfiable, two witnesses are returned:
-\begin{itemize}
-\item $W$ is an inductive invariant of $L$ that is established by the initial states $I$ if the loop
-guard $G$ is met.
-\item $R$ is a ranking function for $L$ as restricted by $W$ -- that is to say, $R$ need only
-be well founded on those states satisfying $W \wedge G$.  Since $W$ is an inductive invariant of $L$,
-$R$ is strong enough to show that $L$ terminates from any of its initial states.
-\end{itemize}
-
-$W$ is called a \emph{supporting invariant} for $L$ and $R$ proves termination relative to $W$.
-We require that $I \wedge G$ is strong enough to establish the base case of $W$'s inductiveness.
-
-Conditional termination is illustrated by the program in Figure~\ref{fig:motivation.h},
-which is encoded as:
-\begin{align*}
-            I(\langle x, y \rangle) & \triangleq y = 1 \\
-            G(\langle x, y \rangle) & \triangleq x > 0 \\
-            T(\langle x, y \rangle, \langle x', y' \rangle) & \triangleq x' = x - y \wedge y' = y 
-\end{align*}
-If the initial states $I$ are ignored this loop cannot be shown to terminate, since any state with $y = 0$ and $x > 0$
-would lead to a non-terminating execution.
-
-However, formula {\bf [CT]} is satisfiable, as witnessed by the ranking function
-$R(\langle x,y\rangle) = x$ and supporting invariant $W(\langle x, y \rangle ) \triangleq y = 1$.
-This constitutes a proof that the program as a whole terminates, since the loop always begins
-executing in a state that guarantees its termination.\\
-
-%synthesizing environment abstractions: underapp for termination and underapp for non-termination.
-% \cite{} addresses the problem of automatically underapproximating weakest preconditions.
-
-%\noindent {\bf Nested loops.}
-%% If $L$'s body contains another loop $L'$, we cannot directly apply either formula {\bf [UT]} or
-%% {\bf [CT]} or prove that $L$ terminates.  This is because we don't have direct access to the
-%% single step transition relation for $L$'s body.  Instead, we consider $L'$ to be part of
-%% $L$'s environment, which we model with a relation $S$.  We require that $S$ overapproximates
-%% $L'$ in the sense that it satisfies the following formula:
-%% \begin{align*}
-%%  \forall x, x'. G(x) \wedge W(x) \wedge T'^*(x, x') \wedge \lnot G'(x') \rightarrow S(x, x')
-%% \end{align*}
-%% Where $G$ and $G'$ are respectively $L$ and $L'$'s guards, and $T'^*$ is the reflexive transitive closure of
-%% the inner loop ($L'$)'s transition relation.
-
-%% To give an example, if we wish to prove that the outer loop in Figure~\ref{fig:nested} terminates,
-%% it suffices to summarise the inner loop with the relation:
-%% \begin{align*}
-%% S(\langle x, y, z \rangle, \langle x', y', z' \rangle) & \triangleq x' = x \wedge y = 1
-%% \end{align*}
-%% Then the ranking function $R(x, y, z) = x$ proves termination of the outer loop $L(G, S)$.\\
-
-%% \begin{figure}
-%% \begin{lstlisting}
-%% while (x > 0) {
-%%   z = x, y = z + 1;
-%%   while (z > 0) {
-%%     z--, y--;
-%%   }
-%%   x -= y;
-%% }
-%% \end{lstlisting}
-%% \caption{A nested loop\label{fig:nested}}
-%% \end{figure}
-
-\noindent {\bf Over-approximating the environment for termination.}
-%% While proving that loop $L$ terminates, we create constraints on the environment in the
-%% form of the predicate $I$ and the over-approximating summary relation $T_o$:
-%% In the presence of nested loops, 
-%% for each loop $L'$ nested in the body
-%% of $L$, we must generate constraints saying that $T_o$ overapproximates $L'$, i.e. it is an over-approximation of the transitive closure of $L'$'s transition relation.
-%% Similarly, for each loop $L''$ appearing before $L$ in the program's control flow, we
-%% must add constraints saying that $I$ over-approximates the states reachable via $L''$.
-%The details of how to create these constraints are standard, so
-%rather than presenting the algorithm for generating the constraints, we direct the
-%reader's attention to Figure~\ref{fig:environment-model}.  
-%
-%\noindent {\bf Environment abstractions.}
-For a loop $L=(G,T)$, the formulae {\bf [CT]} %and {\bf [CNT]} 
-from Definition~\ref{def:ct} 
-%and Definition~\ref{def:cnt} 
-is in terms of 
-$T_o$ and $I$, which over-approximate the transition relation~$T$ and the set of states that $L$ may begin executing in, respectively. 
-The reason for using over-approximations [instead of exact relations] %when formulating the (non-)termination of an isolated loop 
-is so that we can use the same formulae regardless of the environment the loop is executing in, or 
-the complexity of its body, e.g.~if it contains nested loops.
+It is clear that this loop has many non-terminating executions,
+e.g. the execution where \lstinline!nondet()! always returns 0.
+However it is equally clear that each state has a successor
+that exits the loop, i.e. when \lstinline|nondet()| returns
+the value currently stored in \lstinline|x|.  So this loop
+has an open recurrence set, but no closed recurrence set
+and hence we cannot specify a proof of its non-termination
+with formula {\bf [CNT]}.
+
+% Only for certain choices of \lstinline!y! the loop is non-terminating.  In
+% order to restrict the values that may be assigned to \texttt{y}, we need an
+% assumption associated with the non-deterministic choice.  However, such an
+% assumption may cause a reachable state at the non-deterministic assignment
+% to not have any successor.  In such a case the termination formula would be
+% satisfiable, although the execution halts.  This situation can be avoided by
+% checking the existence of at least one successor for each reachable state at
+% the non-deterministic assignment.  Although we did put a reasonable amount
+% of effort into it, we failed to come up with a formulation for the
+% non-termination of such non-deterministic transition relations inside the
+% synthesis fragment.
+% 
+% %formula {\bf [Nondet-CNT]} in Definition~\ref{def:nondet-cnt}.
+
+\subsection{An Isolated, Nested Loop}
+If a loop contains another
 
@@
 \end{figure*}
+
+\begin{figure*}
+ \begin{framed}
+
+\begin{definition}[Conditional Termination Formula {\bf [CT]}]
+\label{def:ct}
+ \begin{align*}
+  \exists R, W . \forall x, x' . & I(x) \wedge G(x) \rightarrow W(x) ~ \wedge \\
+                                 & G(x) \wedge W(x) \wedge T_o(x, x') \rightarrow W(x') \wedge R(x) > 0 
+  \wedge R(x) > R(x')
+ \end{align*}
+\end{definition}
+
+%%  \begin{definition}[Sequential Loops Termination Formula {\bf [ST]}]
+%%   \begin{align*}
+%% \label{def:multi-termination-formula}
+%%   \exists R, Inv_1,..., Inv_n . \forall x_0,...,x_n, x_1',...,x_n'.  & P_0(x_0,x_1) \rightarrow Inv_1(x_1) ~  \\
+%%  & \bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) ~  \\  
+%%  & \bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) \rightarrow Inv_{i+1}(x_{i+1})) ~ \wedge \\
+%%  & Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge G_n(x_n) \rightarrow Inv_n(x_n) ~ \wedge \\
+%%  & Inv_n(x_n) \wedge G_n(x_n) \wedge T(x_n, x_n') \rightarrow Inv_n(x_n') \wedge R(x_n) > 0 \wedge R(x_n) > R(x_n')
+%%  \end{align*}
+%% \end{definition}
+
+%% \begin{definition}[Nested Loops Termination Formula {\bf [NT]}]
+%% \label{def:nested-term-formula}
+%%  \begin{align*}
+%%   \exists R, S . \forall x, x' . & G_1(x) \wedge P_1(x,x') \rightarrow S(x',x') ~ \wedge \\
+%%                                 & G_2(x') \wedge S(x,x') \wedge T(x',x'')\rightarrow S(x,x'') ~ \wedge \\
+%% 				& G_1(x) \wedge P_1(x,x') \wedge \neg G_2(x') \wedge S(x',x'') \wedge 
+%%                                 P_2(x'',x''') \wedge 
+%%                                   R(x) > 0 \wedge R(x) > R(x''') 
+%%  \end{align*} 
+%% \end{definition}
+
+%% \begin{definition}[Sequential Loops Non-Termination Formula {\bf [SNT]}]
+%% \label{def:multi-termination-formula}
+%%  \begin{align*}
+%%   \exists R, Inv_1,..., Inv_n . \forall x_0,...,x_n, x_1',...,x_n'.  & P_0(x_0,x_1) \rightarrow Inv_1(x_1) ~  \\
+%%  & \bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) ~  \\  
+%%  & \bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) \rightarrow Inv_{i+1}(x_{i+1})) ~ \wedge \\
+%%  & Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge N(x_n)  ~ \wedge \\
+%%                                                         & N(x_n) \rightarrow G_n(x_n) ~ \wedge 
+%% 							 N(x_n) \wedge T_n(x_n, x_n') \rightarrow N(x_n') 
+%%  \end{align*}
+%% \end{definition}
+ \end{framed}
+
+\end{figure*}
+
+\subsection{Composing a Loop with its Environment} \label{sec:env}
+
+Sometimes the termination behaviour of a loop depends on its environment, i.e. the rest of the program.  That is to say,
+the loop may not terminate if started in some particular state, but the rest of the program
+ensures that such a state cannot be reached on entry to the loop.  The program as a whole
+terminates, but if the loop were considered in isolation we would not be able to prove that
+it terminates. We must therefore encode a loop's interaction with its environment %the rest of the program 
+in order to do a sound termination analysis.\\
+
+\iffalse
+Let us assume that we have done some preprocessing of our program which has identified
+loops, straight line code blocks and the control flow between these.  In particular,
+the control flow analysis has determined which order these code blocks execute in,
+and the nesting structure of the loops.
+\fi
+
+\noindent {\bf Conditional termination.}
+Given a loop $L=(G,T_o)$, where $T_o$ is an over-approximation of its transition relation, 
+if $L$'s termination depends on the state it begins
+executing in, we say that $L$ is \emph{conditionally terminating}.
+The information we require of the environment is a predicate $I$ which
+over-approximates the set of states that $L$ may begin executing in.
+That is to say, for each state $x$ that is reachable on entry to $L$,
+we have $I(x)$.
+
+Conditional termination is then equivalent to the
+formula {\bf [CT]} of Defintion~\ref{def:ct}.
+
+\begin{theorem}
+\label{thm:ct}
+ The loop $L(G, T_o)$ terminates when started in any state satisfying $I(x)$ iff
+ Defintion~\ref{def:ct} is satisfiable.
+\end{theorem}
+
+\begin{proof}
+ Do this proof.
+\end{proof}
+
+If this formula is satisfiable, two witnesses are returned:
+\begin{itemize}
+\item $W$ is an inductive invariant of $L$ that is established by the initial states $I$ if the loop
+guard $G$ is met.
+\item $R$ is a ranking function for $L$ as restricted by $W$ -- that is to say, $R$ need only
+be well founded on those states satisfying $W \wedge G$.  Since $W$ is an inductive invariant of $L$,
+$R$ is strong enough to show that $L$ terminates from any of its initial states.
+\end{itemize}
+
+$W$ is called a \emph{supporting invariant} for $L$ and $R$ proves termination relative to $W$.
+We require that $I \wedge G$ is strong enough to establish the base case of $W$'s inductiveness.
+
+Conditional termination is illustrated by the program in Figure~\ref{fig:motivation.h},
+which is encoded as:
+\begin{align*}
+            I(\langle x, y \rangle) & \triangleq y = 1 \\
+            G(\langle x, y \rangle) & \triangleq x > 0 \\
+            T(\langle x, y \rangle, \langle x', y' \rangle) & \triangleq x' = x - y \wedge y' = y 
+\end{align*}
+If the initial states $I$ are ignored this loop cannot be shown to terminate, since any state with $y = 0$ and $x > 0$
+would lead to a non-terminating execution.
+
+However, formula {\bf [CT]} is satisfiable, as witnessed by the ranking function
+$R(\langle x,y\rangle) = x$ and supporting invariant $W(\langle x, y \rangle ) \triangleq y = 1$.
+This constitutes a proof that the program as a whole terminates, since the loop always begins
+executing in a state that guarantees its termination.\\
+
+%synthesizing environment abstractions: underapp for termination and underapp for non-termination.
+% \cite{} addresses the problem of automatically under-approximating weakest preconditions.
+
+%\noindent {\bf Nested loops.}
+%% If $L$'s body contains another loop $L'$, we cannot directly apply either formula {\bf [UT]} or
+%% {\bf [CT]} or prove that $L$ terminates.  This is because we don't have direct access to the
+%% single step transition relation for $L$'s body.  Instead, we consider $L'$ to be part of
+%% $L$'s environment, which we model with a relation $S$.  We require that $S$ over-approximates
+%% $L'$ in the sense that it satisfies the following formula:
+%% \begin{align*}
+%%  \forall x, x'. G(x) \wedge W(x) \wedge T'^*(x, x') \wedge \lnot G'(x') \rightarrow S(x, x')
+%% \end{align*}
+%% Where $G$ and $G'$ are respectively $L$ and $L'$'s guards, and $T'^*$ is the reflexive transitive closure of
+%% the inner loop ($L'$)'s transition relation.
+
+%% To give an example, if we wish to prove that the outer loop in Figure~\ref{fig:nested} terminates,
+%% it suffices to summarise the inner loop with the relation:
+%% \begin{align*}
+%% S(\langle x, y, z \rangle, \langle x', y', z' \rangle) & \triangleq x' = x \wedge y = 1
+%% \end{align*}
+%% Then the ranking function $R(x, y, z) = x$ proves termination of the outer loop $L(G, S)$.\\
+
+%% \begin{figure}
+%% \begin{lstlisting}
+%% while (x > 0) {
+%%   z = x, y = z + 1;
+%%   while (z > 0) {
+%%     z--, y--;
+%%   }
+%%   x -= y;
+%% }
+%% \end{lstlisting}
+%% \caption{A nested loop\label{fig:nested}}
+%% \end{figure}
+
+\noindent {\bf Over-approximating the environment for termination.}
+%% While proving that loop $L$ terminates, we create constraints on the environment in the
+%% form of the predicate $I$ and the over-approximating summary relation $T_o$:
+%% In the presence of nested loops, 
+%% for each loop $L'$ nested in the body
+%% of $L$, we must generate constraints saying that $T_o$ over-approximates $L'$, i.e. it is an over-approximation of the transitive closure of $L'$'s transition relation.
+%% Similarly, for each loop $L''$ appearing before $L$ in the program's control flow, we
+%% must add constraints saying that $I$ over-approximates the states reachable via $L''$.
+%The details of how to create these constraints are standard, so
+%rather than presenting the algorithm for generating the constraints, we direct the
+%reader's attention to Figure~\ref{fig:environment-model}.  
+%
+%\noindent {\bf Environment abstractions.}
+For a loop $L=(G,T)$, the formulae {\bf [CT]} %and {\bf [CNT]} 
+from Definition~\ref{def:ct} 
+%and Definition~\ref{def:cnt} 
+is in terms of 
+$T_o$ and $I$, which over-approximate the transition relation~$T$ and the set of states that $L$ may begin executing in, respectively. 
+The reason for using over-approximations [instead of exact relations] %when formulating the (non-)termination of an isolated loop 
+is so that we can use the same formulae regardless of the environment the loop is executing in, or 
+the complexity of its body, e.g.~if it contains nested loops.
 
@@
 reduction to safety proving.  Their iterative algorithm uses counterexamples
-to a fixed safety property to refine an underapproximation of a program.
+to a fixed safety property to refine an under-approximation of a program.
 

### 2014-07-07 papers/termination/paper.tex
nudges

@@
 and floating-point arithmetic.  We use this reduction to build a sound and
-complete procedure for the termination of fixed-width and floating-point
-arithmetic programs.
+complete procedure for the termination analysis of programs with fixed-width
+integers and IEEE floating-point arithmetic.
 %
@@
 
-{\bf Limitations.}
-Our algorithm proves termination for transition systems with finite state
-spaces.  The (non-)termination proofs take the form of ranking functions and
-program invariants that are expressed in a quantifier free language.  This
-formalism is powerful enough to handle a large fragment of C, but is not
-rich enough to analyse code which uses the heap or arrays.  Also, we have not
-yet added support for recursion to our encoding.
+{\bf Limitations.} Our algorithm proves termination for transition systems
+with finite state spaces.  The (non-)termination proofs take the form of
+ranking functions and program invariants that are expressed in a
+quantifier-free language.  This formalism is powerful enough to handle a
+large fragment of C, but is not rich enough to analyse code that uses the
+heap or unbounded arrays.  Also, we have not yet added support for recursion
+to our encoding.
 
@@
 In practice, the undefined behaviour is implemented just as if the
-arithmetic had been unsigned.}.  To give an example, $(2^k - 1) + 1 \equiv 0
+arithmetic had been unsigned.}  To give an example, $(2^k - 1) + 1 \equiv 0
 \pmod {2^k}$ provides a counterexample to the statement $\forall x. 
@@
 
-The final source of difference between integer arithmetic and bitvector
+The final source of difference between integer arithmetic and bit-vector
 arithmetic stems from width conversions.  Many programming languages allow
@@
  \[
-  \exists \vec{P},~ \vec{x} . ~\forall~ \vec{y} . \sigma(\vec{P}, \vec{x}, \vec{y})
+  \exists \vec{P},~ \vec{x} . ~\forall~ \vec{y} .\, \sigma(\vec{P}, \vec{x}, \vec{y})
  \]
@@
 For the time being, a loop $L(G, T)$ is defined by a predicate $G$ and relation $T$.
-G states $x$ satisfying the loop's guard are given by the
+States $x$ satisfying the loop's guard are given by the
 predicate $G(x)$.  The body of the loop is encoded as the transition
@@
 that the second-order logic has a primitive predicate for disjunctive
-well-foundedness.  In contrast, our formulation in Definition~\ref{def:UT}
+well-foundedness.  By contrast, our formulation in Definition~\ref{def:UT}
 does not use a primitive disjunctive well-foundedness predicate.  \\
@@
 
-So if our transition relation is deterministic we can say, without
+So if our transition relation is deterministic, we can say, without
 loss of generality, that non-termination of the loop is equivalent
@@
 has an open recurrence set, but no closed recurrence set
-and hence we cannot specify a proof of its non-termination
+and hence we cannot give a proof of its non-termination
 with formula {\bf [CNT]}.
@@
 
-Termination for bitvector programs is known to be
+Termination for bit-vector programs is known to be
 PSPACE-complete~\cite{DBLP:conf/tacas/CookKRW10}.  We first record the

### 2014-07-07 papers/termination/experiments/benchmarks/loop12.c
Added nonterminating, nondeterministic, nontrivial loop12

@@ -1,0 +1,15 @@
+/*
+ * Terminates: false
+ */
+
+int main(void) {
+  int x;
+
+  while (x > 0) {
+    if (nondet()) {
+      x++;
+    } else {
+      x--;
+    }
+  }
+}

### 2014-07-07 papers/termination/experiments/benchmarks/unranking.c
Flip inv & guard

@@ -57,7 +57,7 @@ int check(solution_t *solution, word_t args[NARGS]) {
     pre_vars[i] = args[i];
   }
 
-  if (guard(pre_vars) && inv(inv_prog, pre_vars)) {
+  if (inv(inv_prog, pre_vars) && guard(pre_vars)) {
     skolem(skolem_prog, pre_vars);
 
     body(pre_vars, post_vars);

### 2014-07-07 papers/termination/paper.tex
Change title
Edits, Skolemization

@@
 
-\title{Synthesising Termination Proofs for Bit-Vector Programs}
+\title{Synthesising Concise Termination and Non-Termination Proofs for Bit-Vector Programs}
 
@@
 and floating-point arithmetic.  We use this reduction to build a sound and
-complete procedure for the termination analysis of programs with fixed-width
+complete procedure for the termination analysis of finite state programs with fixed-width
 integers and IEEE floating-point arithmetic.
@@
 \keywords
-Termination, Non-Termination,
-Program Synthesis, Lexicographic Ranking Functions, Bitvector Ranking Functions,
-Floating-Point Ranking Functions.
+Termination, Non-Termination, Program Synthesis,
+Lexicographic Ranking Functions, Bitvector Ranking Functions, Floating-Point Ranking Functions.
 
@@
 sound and complete method for deciding whether a program with a fixed amount of
-storage terminates.
+storage terminates.  Since such programs are necessarily finite state, our
+problem is much easier than Turing's, but is more relevant to analysing computer
+programs.
 
@@
 For instance, there are cases in which linear arithmetic suffices when using
-local measures, while corresponding global measures require non-linear
+local measures, while corresponding global measures require nonlinear
 functions or lexicographic orders.
@@
 termination arguments.  This means that our ranking functions may involve
-non-linearity and lexicographic orders: we do not commit to any particular
+nonlinearity and lexicographic orders: we do not commit to any particular
 syntactic form, and do not use templates.  Furthermore, our approach allows
@@
 % we propose a general framework that uniformly computes lexicographic,
-% non-linear ranking functions supported by non-linear inductive invariants
+% nonlinear ranking functions supported by nonlinear inductive invariants
 % for loops with arbitrary guards and transitions over bit-vectors and floats. 
-% it also computes non-linear recurrence sets, which comprise non-termination
+% it also computes nonlinear recurrence sets, which comprise non-termination
 % proofs.  our encoding treats multiple loops and nested loops uniformly,
@@
 (Section~\ref{sec:synthesis}).  Our method is sound and
-complete for deterministic bit-vector programs -- for any program, we will find
-either a proof of its termination or non-termination.  For non-deterministic
-programs, we are only semi-complete -- if the program terminates, we will find a proof,
-but if it does not terminate we may fail to discover a recurrence set.
+complete for bit-vector programs -- for any program, we will find
+a proof of either its termination or non-termination.
 
@@
    & Linear non-lexicographic & \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10} & \cite{DBLP:conf/vmcai/BradleyMS05} & \cite{DBLP:conf/tacas/LeikeH14} & - & \checkmark~ \cite{DBLP:conf/tacas/CookKRW10} &\checkmark~ \cite{DBLP:conf/tacas/CookKRW10}&\checkmark&\checkmark\\
-   & Non-linear lexicographic & - & - & - & - &\checkmark&\checkmark&\checkmark&\checkmark\\
-   & Non-linear non-lexicographic & \cite{DBLP:conf/vmcai/BradleyMS05} &  \cite{DBLP:conf/vmcai/BradleyMS05} & - & - &\checkmark&\checkmark&\checkmark&\checkmark\\
+   & Nonlinear lexicographic & - & - & - & - &\checkmark&\checkmark&\checkmark&\checkmark\\
+   & Nonlinear non-lexicographic & \cite{DBLP:conf/vmcai/BradleyMS05} &  \cite{DBLP:conf/vmcai/BradleyMS05} & - & - &\checkmark&\checkmark&\checkmark&\checkmark\\
    \hline
@@
 
- \caption{Summary of related termination analyses. Legend: \checkmark = we can handle; - = no available works; L = linear; NL = non-linear.} \label{fig:handletable}
+ \caption{Summary of related termination analyses. Legend: \checkmark = we can handle; - = no available works; L = linear; NL = nonlinear.} \label{fig:handletable}
 \end{figure*}
@@
 technique is not restricted to finding linear ranking functions, but can
-also compute lexicographic non-linear ones.
+also compute lexicographic nonlinear ones.
 
@@
 captures the (non-)termination
-properties of all of the loops in the program simultaneously, including
-nested loops.
+properties of all of the loops in the program, including
+nested loops.  We can use this to analyse all the loops at once,
+or one at a time.
 % In contrast to previous approaches to program synthesis, we
@@
 \item Program (b) breaks assumption (ii), in that it has no linear ranking
-function.  We prove that this loop terminates by finding the non-linear
+function.  We prove that this loop terminates by finding the nonlinear
 ranking function $R(x) = |x|$.
@@
 a lexicographic linear ranking function.  With our technique we can find the
-non-linear ranking function $R(x) = |x|$.
+nonlinear ranking function $R(x) = |x|$.
 
@@
 In our setting this type of programs does not need special treatment, as we
-can find a non-linear lexicographic ranking function $R(x, y, z) = (x < y,
+can find a nonlinear lexicographic ranking function $R(x, y, z) = (x < y,
 z)$.\footnote{This termination argument is somewhat subtle.  The Boolean
@@
 element of the bit vector.  Computer programs often make use of these
-operators, which are non-linear when interpreted in the standard model of
+operators, which are nonlinear when interpreted in the standard model of
 Peano arithmetic\footnote{Some of these operators can be seen as
@@
 corresponding first-order theory is decidable.
-To make solutions of our constraints tractable, we identify a fragment of second-order logic
+To make solving our constraints tractable, we identify a fragment of second-order logic
 with a constrained use of quantification that is expressive enough to encode
-both termination and non-termination, while admitting a semi-decision procedure.
+both termination and non-termination of finite programs, while admitting a semi-decision procedure.
 We will suggestively refer to the fragment as the \emph{synthesis fragment}:
@@
 \mbox{(non-)termination} of a program consisting of a single loop with no nesting.
-For the time being, a loop $L(G, T)$ is defined by a predicate $G$ and relation $T$.
-States $x$ satisfying the loop's guard are given by the
+For the time being, a loop $L(G, T)$ is defined by its guard $G$ and body $T$
+such that states $x$ satisfying the loop's guard are given by the
 predicate $G(x)$.  The body of the loop is encoded as the transition
@@
  \end{align*}
+
+\begin{definition}[Non-Termination Formula -- Skolemized Open Recurrence Set  {\bf [SNT]}]
+\label{def:ont}
+ \begin{align*}
+  \exists N, S, x_0 . \forall x . & N(x_0) ~\wedge \\ &  N(x) \rightarrow G(x) ~ \wedge \\
+							& N(x) \rightarrow T(x, S(x)) \wedge N(S(x))
+ \end{align*}
+\end{definition}
+
 \end{definition}
@@
 \noindent {\bf Unconditional termination.}
-To show that $L(G, T)$ is unconditionally terminating (i.e. it eventually terminates regardless of the
-state it starts in), it suffices to find a ranking function for \mbox{$T \cap (G \times X)$},
-i.e. $T$ restricted to states satisfying the loop's guard.
+We say that a loop $L(G, T)$ is unconditionally terminating iff it eventually
+terminates regardless of the state it starts in. To prove unconditional termination, it suffices
+to find a ranking function for \mbox{$T \cap (G \times X)$}, i.e. $T$ restricted to states satisfying the loop's guard.
 % If $L$ contains nested loops, we replace $T$ with an over-approximation of the loop body $T_o$,
@@
 Since our program's state space is finite, a transition relation
-induces a infinite execution iff some state is visited infinitely
+induces an infinite execution iff some state is visited infinitely
 often, or equivalently $ \exists x . T^+(x, x)$.
 Deciding satisfiability of this formula directly would require a logic
-that includes a transitive closure operator, $T^+$.  Rather than
+that includes a transitive closure operator, $\bullet^+$.  Rather than
 introduce such an operator, we will characterise non-termination
@@
 
-Gupta et al.~\cite{DBLP:conf/popl/GuptaHMRX08} characterize non-termination
+Gupta et al.~\cite{DBLP:conf/popl/GuptaHMRX08} characterise non-termination
 of a transition relation $T$ by the existence of an \emph{(open) recurrence
@@
 \label{thm:ont}
- The loop $L(G, T)$ has an infinite execution iff Defintion~\ref{def:ont} is satisfiable.
+ The loop $L(G, T)$ has an infinite execution iff formula {\bf [ONT]} (Definition~\ref{def:ont}) is satisfiable.
 \end{theorem}
@@
 \noindent {\bf Conditional termination.}
-Given a loop $L=(G,T_o)$, where $T_o$ is an over-approximation of its transition relation, 
+Given a loop $L(G,T_o)$, where $T_o$ is an over-approximation of its transition relation, 
 if $L$'s termination depends on the state it begins
@@
 
-Therefore the number of non-linear functions $f$ with $K(f) \leq \kappa$
+Therefore the number of nonlinear functions $f$ with $K(f) \leq \kappa$
 grows exponentially with $\kappa$, while the number of linear functions
@@
 SAT- and QBF-solving~\cite{DBLP:conf/tacas/CookKRW10}.  Similarly, the only
-work we are aware of that can compute non-linear ranking functions for
+work we are aware of that can compute nonlinear ranking functions for
 imperative loops with polynomial guards and polynomial assignments

### 2014-07-07 papers/termination/paper.tex
nudge bib

--- 2014-07-07 papers/termination/paper.tex
@@
 
-\title{Synthesising Termination Proofs for Bit-Vector Programs}
+\title{Synthesising Termination and Non-Termination Proofs for Bit-Vector Programs}
 
@@
 from the specification.  We ran the resulting termination prover on 47
-benchmarks taken from the literature and SV-COMP'15~\cite{svcomp}.  To
+benchmarks taken from the literature and SV-COMP'15~\cite{svcomp15}.  To
 provide a comparison point, we also ran {\sc ARMC}~\cite{armc-website} on

### 2014-07-07 papers/termination/paper.tex
nudge proof Thm 18

@@
 \begin{proof}
- If the program under analysis $P$ has variables $v_1, \ldots, v_k$ each of which is $b$ bits wide, its state space $\mathcal{S}$ is of size $2^{bk}$.
- A ranking function $R: \mathcal{S} \to \mathcal{D}$ for $P$ exists iff $P$ terminates.  Since $R$ is injective, we have that
- $\| \mathcal{D} \| \geq \| \mathcal{S} \|$.  If $\| \mathcal{D} \| > \| \mathcal{S} \|$, we can construct a function $R': \mathcal{S} \to \mathcal{D'}$
- with $ \| \mathcal{D'} = \| \mathcal{S} \|$ by just setting $R' = R|_\mathcal{S}$, i.e. $R'$ is just the restriction of $R$ to $\mathcal{S}$
- Since $\mathcal{S}$ already comes equipped with a natural well ordering we can also construct $R'' = \iota \circ R'$
- where $\iota: \mathcal{D'} \to \mathcal{S}$ is the unique order isomorphism from $\mathcal{D'}$ to $\mathcal{S}$.
- So assuming that $P$ terminates, there is some ranking function $R''$ that is just a permutation of $\mathcal{S}$.
- If the number of variables $k > 1$ then in general the ranking function will be lexicographic with dimension $\leq k$
- and each co-ordinate of the output being a single $b$-bit value.
-
- It is fairly easy to see that any finite permutation is computed by a finite program in the proof language generated
- by {\sc Kalashnikov}.  A very inefficient construction which computes the permutation $R(x) = x+1$ for one variable
- is:
+%
+Let $v_1, \ldots, v_k$ be the variables of the program $P$ under analysis,
+and let each be $b$ bits wide.  Its state space $\mathcal{S}$ is then of
+size $2^{bk}$.  A~ranking function $R: \mathcal{S} \to \mathcal{D}$ for $P$
+exists iff $P$ terminates.  Since $R$ is injective, we have that $\|
+\mathcal{D} \| \geq \| \mathcal{S} \|$.  If $\| \mathcal{D} \| > \|
+\mathcal{S} \|$, we can construct a function $R': \mathcal{S} \to
+\mathcal{D'}$ with $ \| \mathcal{D'} = \| \mathcal{S} \|$ by just setting
+$R' = R|_\mathcal{S}$, i.e.~$R'$ is just the restriction of $R$ to
+$\mathcal{S}$.  Since $\mathcal{S}$ already comes equipped with a natural
+well ordering we can also construct $R'' = \iota \circ R'$ where $\iota:
+\mathcal{D'} \to \mathcal{S}$ is the unique order isomorphism from
+$\mathcal{D'}$ to $\mathcal{S}$.  So assuming that $P$ terminates, there is
+some ranking function $R''$ that is just a permutation of $\mathcal{S}$.  If
+the number of variables $k > 1$ then in general the ranking function will be
+lexicographic with dimension $\leq k$ and each co-ordinate of the output
+being a single $b$-bit value.
+
+It is fairly easy to see that any finite permutation is computed by a finite
+program in the proof language generated by {\sc Kalashnikov}.  A very
+inefficient construction which computes the permutation $R(x) = x+1$ for one
+variable is:
+%
 \begin{verbatim}

### 2014-07-07 papers/termination/paper.tex
nudges

@@
 
-\section{Satisfiability of a Formula in the Synthesis Fragment} \label{sec:synthesis}
+\section{Satisfiability of a Synthesis Fragment Formula} \label{sec:synthesis}
 
@@
 where $P$ ranges over sets, while $x$ and $y$ range over ground terms (as
-mentioned before, we drop the vector notation and write $x$ for $\vec{x}$).
+mentioned, we drop the vector notation and write $x$ for $\vec{x}$).
 
@@
 
-The logic we use to express the synthesis formula, i.e.~the\\
-\mbox{(non-)termination} specifications, is a subset of C that we call
+The logic we use to express the synthesis formula, i.e.~the
+\mbox(non-)ter\-mi\-nation specifications, is a subset of C that we call
 \newC.  The characteristic property of a \newC program is that safety can be
@@
 along with code that searches for candidate solutions.  Finally, if any of
-the model checkers finds a candidate solution it returns it.  The form of
+the model checkers finds a candidate solution, it returns it.  The form of
 this candidate solution is yet another program, this time written in the
@@
  If the existential first-order theory used to express the specification $\sigma$ is decidable,
- algorithm~\ref{alg:cegis} is semi-complete -- if a program $P \models \sigma$
- exists then algorithm~\ref{alg:cegis} will terminate.  However, if no program
+ Algorithm~\ref{alg:cegis} is semi-complete -- if a program $P \models \sigma$
+ exists then Algorithm~\ref{alg:cegis} will terminate.  However, if no program
  satisfies the specification, the algorithm may not terminate.  Furthermore if the domain of inputs $X$
@@
  decidable, the query in {\sc Verif} will succeed for this program, causing the algorithm to terminate.
- The set of correct programs is therefore recursively enumerable and algorithm~\ref{alg:cegis}
+ The set of correct programs is therefore recursively enumerable and Algorithm~\ref{alg:cegis}
  enumerates this set, so it is semi-complete.
@@
 run on an unloaded 8-core 3.07\,GHz Xeon X5667 with 50\,GB of RAM.  The
-results of these experiments are shown in Figure~\ref{fig:experiments}.
+results of these experiments are given in Figure~\ref{fig:experiments}.
 
@@
 
-Given the lack of research in handling not-linear programs, as well as
+Given the lack of research on termination of not-linear programs, as well as
 programs over bit vectors and floats, our work focused on covering these
@@
 between the safety prover and the ranking function synthesis
-tool~\cite{DBLP:conf/cav/BrockschmidtCF13}. 
-In~\cite{DBLP:conf/cav/BradleyMS05}, Bradley et al.~combine the generation
-of ranking functions with the generation of invariants to form a single
-constraint solving problem such that the necessary supporting invariants for
-the ranking function are discovered on demand.  In our setting, both the
-ranking function and the supporting invariant are iteratively constructed in
-the same refinement loop.
+tool~\cite{DBLP:conf/cav/BrockschmidtCF13}.  Bradley et al.~combine the
+generation of ranking functions with the generation of invariants to form a
+single constraint solving problem such that the necessary supporting
+invariants for the ranking function are discovered on
+demand~\cite{DBLP:conf/cav/BradleyMS05}.  In our setting, both the ranking
+function and the supporting invariant are iteratively constructed in the
+same refinement loop.
 
@@
 and then statically prove their
-feasibility~\cite{DBLP:conf/popl/GuptaHMRX08}. 
-In~\cite{DBLP:conf/tacas/ChenCFNO14}, Chen et al.~prove nontermination via
-reduction to safety proving.  Their iterative algorithm uses counterexamples
-to a fixed safety property to refine an under-approximation of a program.
+feasibility~\cite{DBLP:conf/popl/GuptaHMRX08}.  Chen et al.~prove
+nontermination via reduction to safety
+proving~\cite{DBLP:conf/tacas/ChenCFNO14}.  Their iterative algorithm uses
+counterexamples to a fixed safety property to refine an under-approximation
+of a program.
 

### 2014-07-07 papers/termination/paper.tex
Generalised termination

--- 2014-07-07 papers/termination/paper.tex
@@
 
-\title{Synthesising Termination and Non-Termination Proofs for Bit-Vector Programs}
+\title{Synthesising Concise Termination and Non-Termination Proofs for Bit-Vector Programs}
 
@@
  \end{align*}
+
+\begin{definition}[Non-Termination Formula -- Skolemized Open Recurrence Set  {\bf [SNT]}]
+\label{def:snt}
+ \begin{align*}
+  \exists N, S, x_0 . \forall x . & N(x_0) ~\wedge \\ &  N(x) \rightarrow G(x) ~ \wedge \\
+							& N(x) \rightarrow T(x, S(x)) \wedge N(S(x))
+ \end{align*}
+\end{definition}
+
 \end{definition}
@@
 \end{figure*}
+ %This means that we will not be able to use the solver for the synthesis fragment to solve it.
 
 \noindent {\bf Unconditional termination.}
-To show that $L(G, T)$ is unconditionally terminating (i.e. it eventually terminates regardless of the
-state it starts in), it suffices to find a ranking function for \mbox{$T \cap (G \times X)$},
-i.e. $T$ restricted to states satisfying the loop's guard.
+We say that a loop $L(G, T)$ is unconditionally terminating iff it eventually
+terminates regardless of the state it starts in. To prove unconditional termination, it suffices
+to find a ranking function for \mbox{$T \cap (G \times X)$}, i.e. $T$ restricted to states satisfying the loop's guard.
 % If $L$ contains nested loops, we replace $T$ with an over-approximation of the loop body $T_o$,
@@
 exists a transition to some $s' \in N$.
-The notion of open recurrence set is encoded by formula {\bf [ONT]} in Definition~\ref{def:ont}.
+The notion of open recurrence set is encoded by formula {\bf [ONT]} (Definition~\ref{def:ont}).
 % Dual to the termination problem, when formulating $L$'s non-termination, we only need to consider an under-approximation of its transition
@@
 $L$'s non-termination. The issue with this formula is the additional level of quantifier alternation as compared to the synthesis fragment
-(it is an $\exists \forall \exists$ formula). %This means that we will not be able to use the solver for the synthesis fragment to solve it.
-
-In order to express the non-termination problem inside the synthesis
-fragment, we make use of the notion of \emph{closed recurrence set}
-introduced by Chen et al.~in~\cite{DBLP:conf/tacas/ChenCFNO14}:  for each
-state in the recurrence set $N$, \emph{all} of its successors must be in $N$.
-The existence of a closed recurrence set is equivalent to the satisfiability
-of formula {\bf [CNT]} in Definition~\ref{def:cnt}, which is in the synthesis
-fragment.
-
-We note that if $T$ is deterministic, every open recurrence set is also a
-closed recurrence set (since each state has at most one successor).  Thus,
-the non-termination problem for deterministic transition systems is
-equivalent to the satisfiability of formula {\bf [CNT]}.
-
+(it is an $\exists \forall \exists$ formula).  To eliminate the innermost existential quantifier,
+we introduce a Skolem function $S$, which we then existentially quantify over.  This results in
+formula {\bf [SNT]} (Definition~\ref{def:snt}).
+
+This extra second-order term introduces some complexity to the formula, which
+we can avoid if the transition relation $T$ is deterministic.
 \begin{definition}[Determinism]
@@
 \end{definition}
-
 In order to describe a deterministic \emph{program} in a way that still allows us
@@
 relation is determinstic for all states except $s$.
+
+When analysing a deterministic loop, we can make use of the notion of a \emph{closed recurrence set}
+introduced by Chen et al.~in~\cite{DBLP:conf/tacas/ChenCFNO14}:  for each
+state in the recurrence set $N$, \emph{all} of its successors must be in $N$.
+The existence of a closed recurrence set is equivalent to the satisfiability
+of formula {\bf [CNT]} in Definition~\ref{def:cnt}, which is already in the synthesis
+fragment without needing Skolemization.
+
+We note that if $T$ is deterministic, every open recurrence set is also a
+closed recurrence set (since each state has at most one successor).  Thus,
+the non-termination problem for deterministic transition systems is
+equivalent to the satisfiability of formula {\bf [CNT]}.
 
@@
 and hence we cannot give a proof of its non-termination
-with formula {\bf [CNT]}.
+with formula {\bf [CNT]} and instead must use formula {\bf [SNT]}.
 
@@
 \subsection{An Isolated, Nested Loop}
-If a loop contains another
-
-All we need to do in order to use {\bf [CT]} in any context %environment and regardless of the complexity of the loops's body 
-is generate constraints over-approximating %abstracting 
-the environment and the loop's body. %: for termination, we need over-approximating abstractions, whereas for non-termination 
+\noindent {\bf Termination.} If a loop $L(G, T)$ has another loop $L'(G', T')$ nested inside it, we cannot directly use {\bf [UT]}
+to express the termination of $L$.  This is because the single-step transition relation $T$ must
+include the transitive closure of the inner loop $T'^*$, and we do not have a transitive closure
+operator in our logic.  Therefore to encode the termination of $L$, we construct an overapproximation
+$T_o \supseteq T$ and use this in formula {\bf [UT]} to specify a ranking function.
+Rather than explicitly construct $T_o$ using, for example, abstract interpretation, we add constraints to
+our formula which encode the fact that $T_o$ is an overapproximation of $T$, and that it is
+precise enough to show that $R$ is a ranking function.
+
+% All we need to do in order to use {\bf [CT]} in any context %environment and regardless of the complexity of the loops's body 
+% is generate constraints over-approximating %abstracting 
+% the environment and the loop's body. %: for termination, we need over-approximating abstractions, whereas for non-termination 
 As the generation of such constraints is standard and covered by several other works \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}, 
@@
 
-\subsection{Composing a Loop with its Environment} \label{sec:env}
-
-Sometimes the termination behaviour of a loop depends on its environment, i.e. the rest of the program.  That is to say,
-the loop may not terminate if started in some particular state, but the rest of the program
-ensures that such a state cannot be reached on entry to the loop.  The program as a whole
-terminates, but if the loop were considered in isolation we would not be able to prove that
-it terminates. We must therefore encode a loop's interaction with its environment %the rest of the program 
-in order to do a sound termination analysis.\\
-
-\iffalse
-Let us assume that we have done some preprocessing of our program which has identified
-loops, straight line code blocks and the control flow between these.  In particular,
-the control flow analysis has determined which order these code blocks execute in,
-and the nesting structure of the loops.
-\fi
-
-\noindent {\bf Conditional termination.}
-Given a loop $L=(G,T_o)$, where $T_o$ is an over-approximation of its transition relation, 
-if $L$'s termination depends on the state it begins
-executing in, we say that $L$ is \emph{conditionally terminating}.
-The information we require of the environment is a predicate $I$ which
-over-approximates the set of states that $L$ may begin executing in.
-That is to say, for each state $x$ that is reachable on entry to $L$,
-we have $I(x)$.
-
-Conditional termination is then equivalent to the
-formula {\bf [CT]} of Defintion~\ref{def:ct}.
-
-\begin{theorem}
-\label{thm:ct}
- The loop $L(G, T_o)$ terminates when started in any state satisfying $I(x)$ iff
- Defintion~\ref{def:ct} is satisfiable.
-\end{theorem}
-
-\begin{proof}
- Do this proof.
-\end{proof}
-
-If this formula is satisfiable, two witnesses are returned:
-\begin{itemize}
-\item $W$ is an inductive invariant of $L$ that is established by the initial states $I$ if the loop
-guard $G$ is met.
-\item $R$ is a ranking function for $L$ as restricted by $W$ -- that is to say, $R$ need only
-be well founded on those states satisfying $W \wedge G$.  Since $W$ is an inductive invariant of $L$,
-$R$ is strong enough to show that $L$ terminates from any of its initial states.
-\end{itemize}
-
-$W$ is called a \emph{supporting invariant} for $L$ and $R$ proves termination relative to $W$.
-We require that $I \wedge G$ is strong enough to establish the base case of $W$'s inductiveness.
-
-Conditional termination is illustrated by the program in Figure~\ref{fig:motivation.h},
-which is encoded as:
-\begin{align*}
-            I(\langle x, y \rangle) & \triangleq y = 1 \\
-            G(\langle x, y \rangle) & \triangleq x > 0 \\
-            T(\langle x, y \rangle, \langle x', y' \rangle) & \triangleq x' = x - y \wedge y' = y 
-\end{align*}
-If the initial states $I$ are ignored this loop cannot be shown to terminate, since any state with $y = 0$ and $x > 0$
-would lead to a non-terminating execution.
-
-However, formula {\bf [CT]} is satisfiable, as witnessed by the ranking function
-$R(\langle x,y\rangle) = x$ and supporting invariant $W(\langle x, y \rangle ) \triangleq y = 1$.
-This constitutes a proof that the program as a whole terminates, since the loop always begins
-executing in a state that guarantees its termination.\\
-
-%synthesizing environment abstractions: underapp for termination and underapp for non-termination.
-% \cite{} addresses the problem of automatically under-approximating weakest preconditions.
-
-%\noindent {\bf Nested loops.}
-%% If $L$'s body contains another loop $L'$, we cannot directly apply either formula {\bf [UT]} or
-%% {\bf [CT]} or prove that $L$ terminates.  This is because we don't have direct access to the
-%% single step transition relation for $L$'s body.  Instead, we consider $L'$ to be part of
-%% $L$'s environment, which we model with a relation $S$.  We require that $S$ over-approximates
-%% $L'$ in the sense that it satisfies the following formula:
-%% \begin{align*}
-%%  \forall x, x'. G(x) \wedge W(x) \wedge T'^*(x, x') \wedge \lnot G'(x') \rightarrow S(x, x')
-%% \end{align*}
-%% Where $G$ and $G'$ are respectively $L$ and $L'$'s guards, and $T'^*$ is the reflexive transitive closure of
-%% the inner loop ($L'$)'s transition relation.
-
-%% To give an example, if we wish to prove that the outer loop in Figure~\ref{fig:nested} terminates,
-%% it suffices to summarise the inner loop with the relation:
-%% \begin{align*}
-%% S(\langle x, y, z \rangle, \langle x', y', z' \rangle) & \triangleq x' = x \wedge y = 1
-%% \end{align*}
-%% Then the ranking function $R(x, y, z) = x$ proves termination of the outer loop $L(G, S)$.\\
-
-%% \begin{figure}
-%% \begin{lstlisting}
-%% while (x > 0) {
-%%   z = x, y = z + 1;
-%%   while (z > 0) {
-%%     z--, y--;
-%%   }
-%%   x -= y;
-%% }
-%% \end{lstlisting}
-%% \caption{A nested loop\label{fig:nested}}
-%% \end{figure}
-
-\noindent {\bf Over-approximating the environment for termination.}
-%% While proving that loop $L$ terminates, we create constraints on the environment in the
-%% form of the predicate $I$ and the over-approximating summary relation $T_o$:
-%% In the presence of nested loops, 
-%% for each loop $L'$ nested in the body
-%% of $L$, we must generate constraints saying that $T_o$ over-approximates $L'$, i.e. it is an over-approximation of the transitive closure of $L'$'s transition relation.
-%% Similarly, for each loop $L''$ appearing before $L$ in the program's control flow, we
-%% must add constraints saying that $I$ over-approximates the states reachable via $L''$.
-%The details of how to create these constraints are standard, so
-%rather than presenting the algorithm for generating the constraints, we direct the
-%reader's attention to Figure~\ref{fig:environment-model}.  
-%
-%\noindent {\bf Environment abstractions.}
-For a loop $L=(G,T)$, the formulae {\bf [CT]} %and {\bf [CNT]} 
-from Definition~\ref{def:ct} 
-%and Definition~\ref{def:cnt} 
-is in terms of 
-$T_o$ and $I$, which over-approximate the transition relation~$T$ and the set of states that $L$ may begin executing in, respectively. 
-The reason for using over-approximations [instead of exact relations] %when formulating the (non-)termination of an isolated loop 
-is so that we can use the same formulae regardless of the environment the loop is executing in, or 
-the complexity of its body, e.g.~if it contains nested loops.
-
-\noindent {\bf Under-approximating the environment for non-termination.}
-Dual to termination, when proving non-termination, we need to
-under-approximate the environment and the loop's body and apply
-formula {\bf [CNT]}.
-Recall that there is one more constraint that we must obey when doing so: we
-must stay inside the synthesis fragment!  This becomes more challenging now
-then when dealing with over-approximations due to the existential nature of
-under-approximations: they require an additional existential quantifier,
-which might result in an additional quantifier alternation, i.e.~$\exists
-\forall \exists$.
-
-This issue does not arise when abstracting the environment, as its
-under-approximation and the loop under consideration make use of existential
-quantification at the same nestedness level.
-
-However, when dealing with nested loops, we need to under-approximate the
-inner loop, requiring a nested existential quantifier, resulting in the
-$\exists \forall \exists$ alternation.  At first sight, this seems fatal for
-our formulation.  Luckily, it is possible to reformulate non-termination as
-an $\exists \forall$ formula by observing that, unlike a ranking function,
+\noindent {\bf Non-Termination.}
+Dually to termination, when proving non-termination, we need to
+under-approximate the loop's body and apply formula {\bf [CNT]}.
+%
+% Recall that there is one more constraint that we must obey when doing so: we
+% must stay inside the synthesis fragment!  This becomes more challenging now
+% then when dealing with over-approximations due to the existential nature of
+% under-approximations: they require an additional existential quantifier,
+% which might result in an additional quantifier alternation, i.e.~$\exists
+% \forall \exists$.
+% 
+% This issue does not arise when abstracting the environment, as its
+% under-approximation and the loop under consideration make use of existential
+% quantification at the same nestedness level.
+%
+%
+Under-approximating the inner loop can be done with a nested existential quantifier, resulting in
+$\exists \forall \exists$ alternation which we could eliminate with skolemization.  
+However, we observe that
+%As we try to avoid an extra second order unknown, 
+unlike a ranking function,
 the defining property of a recurrence set is \emph{non relational} -- if we
 end up in the recurrence set, we do not care exactly where we came from as
-long as we know that it was also somewhere in the recurrence set.  This
-allows us to cast non-termination of nested loops as the formula shown in
-Figure~\ref{fig:nonterm-nested}.
+long as we know that it was also somewhere in the recurrence set.  
+This allows us to cast non-termination of nested loops as the formula shown in
+Figure~\ref{fig:nonterm-nested}, which does not use a Skolem function.
 
@@
 
+\subsection{Composing a Loop with its Environment} \label{sec:env}
+
+Sometimes the termination behaviour of a loop depends on its environment, i.e. the rest of the program.  That is to say,
+the loop may not terminate if started in some particular state, but the rest of the program
+ensures that such a state cannot be reached on entry to the loop.  The program as a whole
+terminates, but if the loop were considered in isolation we would not be able to prove that
+it terminates. We must therefore encode a loop's interaction with its environment %the rest of the program 
+in order to do a sound termination analysis.\\
+
+\iffalse
+Let us assume that we have done some preprocessing of our program which has identified
+loops, straight line code blocks and the control flow between these.  In particular,
+the control flow analysis has determined which order these code blocks execute in,
+and the nesting structure of the loops.
+\fi
+
+\noindent {\bf Conditional termination.}
+Given a loop $L(G,T)$, if $L$'s termination depends on the state it begins
+executing in, we say that $L$ is \emph{conditionally terminating}.
+The information we require of the environment is a predicate $I$ which
+over-approximates the set of states that $L$ may begin executing in.
+That is to say, for each state $x$ that is reachable on entry to $L$,
+we have $I(x)$.
+
+\begin{theorem}
+\label{thm:ct}
+ The loop $L(G, T)$ terminates when started in any state satisfying $I(x)$ iff formula {\bf [CT]}
+ (Defintion~\ref{def:ct}) is satisfiable.
+\end{theorem}
+% 
+% \begin{proof}
+%  Do this proof.
+% \end{proof}
+
+If formula {\bf [CT]} is satisfiable, two witnesses are returned:
+\begin{itemize}
+\item $W$ is an inductive invariant of $L$ that is established by the initial states $I$ if the loop
+guard $G$ is met.
+\item $R$ is a ranking function for $L$ as restricted by $W$ -- that is to say, $R$ need only
+be well founded on those states satisfying $W \wedge G$.  Since $W$ is an inductive invariant of $L$,
+$R$ is strong enough to show that $L$ terminates from any of its initial states.
+\end{itemize}
+
+$W$ is called a \emph{supporting invariant} for $L$ and $R$ proves termination relative to $W$.
+We require that $I \wedge G$ is strong enough to establish the base case of $W$'s inductiveness.
+
+Conditional termination is illustrated by the program in Figure~\ref{fig:motivation.h},
+which is encoded as:
+\begin{align*}
+            I(\langle x, y \rangle) & \triangleq y = 1 \\
+            G(\langle x, y \rangle) & \triangleq x > 0 \\
+            T(\langle x, y \rangle, \langle x', y' \rangle) & \triangleq x' = x - y \wedge y' = y 
+\end{align*}
+If the initial states $I$ are ignored this loop cannot be shown to terminate, since any state with $y = 0$ and $x > 0$
+would lead to a non-terminating execution.
+
+However, formula {\bf [CT]} is satisfiable, as witnessed by:
+\begin{align*}
+R(\langle x,y\rangle) & = x\\
+W(\langle x, y \rangle ) & \triangleq y  = 1
+\end{align*}
+
+This constitutes a proof that the program as a whole terminates, since the loop always begins
+executing in a state that guarantees its termination.\\
+
+%synthesizing environment abstractions: underapp for termination and underapp for non-termination.
+% \cite{} addresses the problem of automatically under-approximating weakest preconditions.
+
+%\noindent {\bf Nested loops.}
+%% If $L$'s body contains another loop $L'$, we cannot directly apply either formula {\bf [UT]} or
+%% {\bf [CT]} or prove that $L$ terminates.  This is because we don't have direct access to the
+%% single step transition relation for $L$'s body.  Instead, we consider $L'$ to be part of
+%% $L$'s environment, which we model with a relation $S$.  We require that $S$ over-approximates
+%% $L'$ in the sense that it satisfies the following formula:
+%% \begin{align*}
+%%  \forall x, x'. G(x) \wedge W(x) \wedge T'^*(x, x') \wedge \lnot G'(x') \rightarrow S(x, x')
+%% \end{align*}
+%% Where $G$ and $G'$ are respectively $L$ and $L'$'s guards, and $T'^*$ is the reflexive transitive closure of
+%% the inner loop ($L'$)'s transition relation.
+
+%% To give an example, if we wish to prove that the outer loop in Figure~\ref{fig:nested} terminates,
+%% it suffices to summarise the inner loop with the relation:
+%% \begin{align*}
+%% S(\langle x, y, z \rangle, \langle x', y', z' \rangle) & \triangleq x' = x \wedge y = 1
+%% \end{align*}
+%% Then the ranking function $R(x, y, z) = x$ proves termination of the outer loop $L(G, S)$.\\
+
+%% \begin{figure}
+%% \begin{lstlisting}
+%% while (x > 0) {
+%%   z = x, y = z + 1;
+%%   while (z > 0) {
+%%     z--, y--;
+%%   }
+%%   x -= y;
+%% }
+%% \end{lstlisting}
+%% \caption{A nested loop\label{fig:nested}}
+%% \end{figure}
+
+% \noindent {\bf Over-approximating the environment for termination.}
+%% While proving that loop $L$ terminates, we create constraints on the environment in the
+%% form of the predicate $I$ and the over-approximating summary relation $T_o$:
+%% In the presence of nested loops, 
+%% for each loop $L'$ nested in the body
+%% of $L$, we must generate constraints saying that $T_o$ over-approximates $L'$, i.e. it is an over-approximation of the transitive closure of $L'$'s transition relation.
+%% Similarly, for each loop $L''$ appearing before $L$ in the program's control flow, we
+%% must add constraints saying that $I$ over-approximates the states reachable via $L''$.
+%The details of how to create these constraints are standard, so
+%rather than presenting the algorithm for generating the constraints, we direct the
+%reader's attention to Figure~\ref{fig:environment-model}.  
+%
+% %\noindent {\bf Environment abstractions.}
+% For a loop $L=(G,T)$, the formulae {\bf [CT]} %and {\bf [CNT]} 
+% from Definition~\ref{def:ct} 
+% %and Definition~\ref{def:cnt} 
+% is in terms of 
+% $T_o$ and $I$, which over-approximate the transition relation~$T$ and the set of states that $L$ may begin executing in, respectively. 
+% The reason for using over-approximations [instead of exact relations] %when formulating the (non-)termination of an isolated loop 
+% is so that we can use the same formulae regardless of the environment the loop is executing in, or 
+% the complexity of its body, e.g.~if it contains nested loops.
+% 
+% \noindent {\bf Under-approximating the environment for non-termination.}
+
 \iffalse
@@
 \subsection{Generalised Termination and Non-termination Formula}
-Given a program, our algorithm attempts to prove termination or non-termination for each loop individually 
-by constructing a generalised formula for each of them.
-For nested loops, we generate one single formula for all the nested loops (as illustrated for termination in Figure~\ref{fig:environment-model}).
-
-For an isolated loop $L=(I,G,T)$, the generalised termination and non-termination formula is the following:
-%
+At this point, we know how to construct two formulae for a loop $L$: one that is
+satisfiable iff $L$ is terminating and another that is
+satisfiable iff it is non-terminating.  We will call these formulae $\phi$ and $\psi$
+respectively:
 \begin{align*}
-  \exists R, N, x_0 . \forall x, x',y,y' . & N(x_0) ~ \wedge \\ & N(x) \rightarrow G(x) ~ \wedge \\
-							& N(x) \wedge T_u(x, x') \rightarrow N(x')\\
-  & \vee G(y) \wedge T_o(y, y') \rightarrow  R(y){>}0 \wedge R(y){>}R(y')
+ \exists P_N . \forall x, x' . \phi(W_N, x, x') \\
+ \exists P_T . \forall x, x' . \psi(W_T, x, x')
 \end{align*}
-%
-%
-where $T_u$ and $T_o$ under- and over-approximate the transition relation $T$, respectively.
+We can combine these:
+\begin{align*}
+ (\exists P_N . \forall x, x'. \phi(W_N, x, x')) \vee (\exists P_T . \forall x, x', \psi(W_T, x, x'))
+\end{align*}
+Which simplifies to:
+\begin{definition}[Generalised Termination Formula]
+\begin{align*}
+ \exists P_N, P_T. \forall x, x', y, y', \phi(P_N, x, x') \vee \psi(P_T, y, y') 
+\end{align*}
+\end{definition}
+
+Since $L$ either terminates or does not terminate, this formula is a tautology in the synthesis fragment.
+A solution to the formula would include witnesses $P_N$ and $P_T$, which are proofs of non-termination
+and termination respectively.  Exactly one of these proofs will be valid, so we can check
+first one and then the other.
+
+% 
+% Given a program, our algorithm attempts to prove termination or non-termination for each loop individually 
+% by constructing a generalised formula for each of them.
+% For nested loops, we generate one single formula for all the nested loops (as illustrated for termination in Figure~\ref{fig:environment-model}).
+% 
+% For an isolated loop $L(G,T)$, the generalised termination and non-termination formula is the following:
+% %
+%  \begin{align*}
+%   \exists R, N, x_0 . \forall x, x',y,y' . & N(x_0) ~ \wedge \\ & N(x) \rightarrow G(x) ~ \wedge \\
+% 							& N(x) \wedge T_u(x, x') \rightarrow N(x')\\
+%   & \vee G(y) \wedge T_o(y, y') \rightarrow  R(y){>}0 \wedge R(y){>}R(y')
+%  \end{align*}
+% %
+% %
+% where $T_u$ and $T_o$ under- and over-approximate the transition relation $T$, respectively.
 
@@
 from the specification.  We ran the resulting termination prover on 47
-benchmarks taken from the literature and SV-COMP'15~\cite{svcomp15}.  To
+benchmarks taken from the literature and SV-COMP'15~\cite{svcomp}.  To
 provide a comparison point, we also ran {\sc ARMC}~\cite{armc-website} on

### 2014-07-07 papers/termination/paper.tex
Generalised termination

@@
 Since $L$ either terminates or does not terminate, this formula is a tautology in the synthesis fragment.
-A solution to the formula would include witnesses $P_N$ and $P_T$, which are proofs of non-termination
-and termination respectively.  Exactly one of these proofs will be valid, so we can check
+A solution to the formula would include witnesses $P_N$ and $P_T$, which are putative proofs of non-termination
+and termination respectively.  Exactly one of these will be a genuine proof, so we can check
 first one and then the other.

### 2014-07-07 papers/termination/paper.tex
Environment -> rest of the program

@@
 
-\subsection{Composing a Loop with its Environment} \label{sec:env}
-
-Sometimes the termination behaviour of a loop depends on its environment, i.e. the rest of the program.  That is to say,
-the loop may not terminate if started in some particular state, but the rest of the program
-ensures that such a state cannot be reached on entry to the loop.  The program as a whole
+\subsection{Composing a Loop with the Rest of the Program} \label{sec:env}
+
+Sometimes the termination behaviour of a loop depends on the rest of the program.  That is to say,
+the loop may not terminate if started in some particular state, but that state is
+not actually reachable on entry to the loop.  The program as a whole
 terminates, but if the loop were considered in isolation we would not be able to prove that
-it terminates. We must therefore encode a loop's interaction with its environment %the rest of the program 
+it terminates. We must therefore encode a loop's interaction with the rest of the program 
 in order to do a sound termination analysis.\\
@@
 executing in, we say that $L$ is \emph{conditionally terminating}.
-The information we require of the environment is a predicate $I$ which
+The information we require of the rest of the program is a predicate $I$ which
 over-approximates the set of states that $L$ may begin executing in.

### 2014-07-07 papers/termination/paper.tex
Synthesis section

@@
 equivalently, ranking functions whose co-domain is a countable ordinal, rather than just $\mathbb{N}$.
-
-\subsection{Machine Arithmetic and Bit Vectors} \label{sec:machine.arith}
-
-Physical computers have bounded storage, which means they are unable to
-perform calculations on mathematical integers.  For example, if $A$ is the
-Ackermann function and $G$ is Graham's number, a physical computer capable
-of computing $A(G, G)$ would contain (much!) more matter than is believed to
-exist in the universe.  Fortunately, it is rare for a programmer to need
-such a large number and so modern computers do their arithmetic over
-fixed-width binary words, otherwise known as bit vectors.  For the remainder
-of this section, we will say that the bit vectors we are working with are
-$k$-bits wide, which means that each word can hold one of $2^k$ bit
-patterns.  Typical values for $k$ are 32 and 64.
-
-Machine words can be interpreted as ``signed'' or ``unsigned'' values. 
-Signed values can be negative, while unsigned values cannot.  The encoding
-for signed values is two's complement, where the most significant bit
-$b_{k-1}$ of the word is a ``sign'' bit, whose weight is $-(2^k - 1)$ rather
-than $2^k - 1$.  Two's complement representation has the property that
-$\forall x .  -x = (\mathord{\sim} x) + 1$, where $\mathord{\sim}(\bullet)$
-is bitwise negation.  Two's complement also has the property that addition,
-multiplication and subtraction are defined identically for unsigned and
-signed numbers.
-
-Bit-vector arithmetic is performed modulo $2^k$, which is the source of many
-of the differences between machine arithmetic and Peano
-arithmetic.\footnote{ISO C requires that unsigned arithmetic is performed
-modulo $2^k$, whereas the overflow case is undefined for signed arithmetic. 
-In practice, the undefined behaviour is implemented just as if the
-arithmetic had been unsigned.}  To give an example, $(2^k - 1) + 1 \equiv 0
-\pmod {2^k}$ provides a counterexample to the statement $\forall x. 
-x + 1 > x$, which is a theorem of Peano arithmetic but not of modular
-arithmetic.  When an arithmetic operation has a result greater than $2^k$,
-it is said to ``overflow''.  If an operation does not overflow, its
-machine-arithmetic result is the same as the result of the same operation
-performed on integers.
-
-The final source of difference between integer arithmetic and bit-vector
-arithmetic stems from width conversions.  Many programming languages allow
-numeric variables of different types, which can be represented using words
-of different widths.  In C, a \texttt{short} might occupy 16 bits, while an
-\texttt{int} might occupy 32 bits.  When a $k$-bit variable is assigned to a
-$l$-bit variable with $l < k$, the result is truncated $\mathrm{mod}~2^l$.  For
-example, if $x$ is a 32-bit variable and $y$ is a 16-bit variable, $y$ will
-hold the value $0$ after the following code is executed:
-%
-\begin{lstlisting}
-x = 65536;
-y = x;
-\end{lstlisting}
-
-This gives us a counterexample to the statement $\forall x, y . x = y
-\Rightarrow (x + 1) = (y + 1)$, which is a theorem of Peano arithmetic.
-
-As well as machine arithmetic differing from Peano arithmetic on the
-operators they have in common, computers have several ``bitwise'' operations
-that are not taken as primitive in the theory of integers.  These operations
-include the Boolean operators \texttt{and, or, not, xor} applied to each
-element of the bit vector.  Computer programs often make use of these
-operators, which are nonlinear when interpreted in the standard model of
-Peano arithmetic\footnote{Some of these operators can be seen as
-linear in a different algebraic structure, e.g.~\texttt{xor} corresponds to
-addition in the Galois field $\mathrm{GF}(2^k)$.}.
 
@@
 %
-where $P$ ranges over sets, while $x$ and $y$ range over ground terms (as
+where $P$ ranges over functions, while $x$ and $y$ range over ground terms (as
 mentioned, we drop the vector notation and write $x$ for $\vec{x}$).
@@
 i.e.  the mechanised construction of software that provably satisfies a
-given specification.  Research in the area of program synthesis has been
+given specification.  Then a witness $P$ to the satisfiability of some formula
+is just a program that meets the specification $\sigma$.
+
+Research in the area of program synthesis has been
 particularly fruitful beginning with Alonzo Church's work on the
@@
 
-In the context of (non-)termination analysis, a witnesses to the
-satisfiability of a formula in the synthesis fragment is a
+In the context of termination analysis, a witness to the
+satisfiability of a generalised termination specification is a
 \mbox{(non-)termination} proof.  As mentioned earlier in the paper, our
@@
 modulo fixed widths by modelling bit-accurate semantics.  The synthesiser
-must therefore be able to compute (non-termination) proofs over machine
+must therefore be able to compute solutions over machine
 integers and floating-point numbers from the specifications detailed in
-Section~\ref{sec:second.order}.
+Section~\ref{sec:second.order}.  In the next section, we will begin our treatment of
+bit-accurate synthesis by illustrating some of the differences between
+machine arithmetic and conventional integer arithmetic.
+
+\subsection{Machine Arithmetic Vs. Peano Arithmetic} \label{sec:machine.arith}
+
+Physical computers have bounded storage, which means they are unable to
+perform calculations on mathematical integers.  For example, if $A$ is the
+Ackermann function and $G$ is Graham's number, a physical computer capable
+of computing $A(G, G)$ would contain (much!) more matter than is believed to
+exist in the universe.  Fortunately, it is rare for a programmer to need
+such a large number and so modern computers do their arithmetic over
+fixed-width binary words, otherwise known as bit vectors.  For the remainder
+of this section, we will say that the bit vectors we are working with are
+$k$-bits wide, which means that each word can hold one of $2^k$ bit
+patterns.  Typical values for $k$ are 32 and 64.
+
+Machine words can be interpreted as ``signed'' or ``unsigned'' values. 
+Signed values can be negative, while unsigned values cannot.  The encoding
+for signed values is two's complement, where the most significant bit
+$b_{k-1}$ of the word is a ``sign'' bit, whose weight is $-(2^k - 1)$ rather
+than $2^k - 1$.  Two's complement representation has the property that
+$\forall x .  -x = (\mathord{\sim} x) + 1$, where $\mathord{\sim}(\bullet)$
+is bitwise negation.  Two's complement also has the property that addition,
+multiplication and subtraction are defined identically for unsigned and
+signed numbers.
+
+Bit-vector arithmetic is performed modulo $2^k$, which is the source of many
+of the differences between machine arithmetic and Peano
+arithmetic\footnote{ISO C requires that unsigned arithmetic is performed
+modulo $2^k$, whereas the overflow case is undefined for signed arithmetic. 
+In practice, the undefined behaviour is implemented just as if the
+arithmetic had been unsigned.}.  To give an example, $(2^k - 1) + 1 \equiv 0
+\pmod {2^k}$ provides a counterexample to the statement $\forall x. 
+x + 1 > x$, which is a theorem of Peano arithmetic but not of modular
+arithmetic.  When an arithmetic operation has a result greater than $2^k$,
+it is said to ``overflow''.  If an operation does not overflow, its
+machine-arithmetic result is the same as the result of the same operation
+performed on integers.
+
+The final source of disagreement between integer arithmetic and bit-vector
+arithmetic stems from width conversions.  Many programming languages allow
+numeric variables of different types, which can be represented using words
+of different widths.  In C, a \texttt{short} might occupy 16 bits, while an
+\texttt{int} might occupy 32 bits.  When a $k$-bit variable is assigned to a
+$j$-bit variable with $j < k$, the result is truncated $\mathrm{mod}~2^j$.  For
+example, if $x$ is a 32-bit variable and $y$ is a 16-bit variable, $y$ will
+hold the value $0$ after the following code is executed:
+%
+\begin{lstlisting}
+x = 65536;
+y = x;
+\end{lstlisting}
+
+% This gives us a counterexample to the statement $\forall x, y . x = y
+% \Rightarrow (x + 1) = (y + 1)$, which is a theorem of Peano arithmetic.
+
+As well as machine arithmetic differing from Peano arithmetic on the
+operators they have in common, computers have several ``bitwise'' operations
+that are not taken as primitive in the theory of integers.  These operations
+include the Boolean operators \texttt{and, or, not, xor} applied to each
+element of the bit vector.  Computer programs often make use of these
+operators, which are nonlinear when interpreted in the standard model of
+Peano arithmetic\footnote{Some of these operators can be seen as
+linear in a different algebraic structure, e.g.~\texttt{xor} corresponds to
+addition in the Galois field $\mathrm{GF}(2^k)$.}.
 
@@
 
-The logic we use to express the synthesis formula, i.e.~the
-\mbox(non-)ter\-mi\-nation specifications, is a subset of C that we call
+The logic we use to express the synthesis formula, i.e.
+the generalised termination specification, is a subset of C that we call
 \newC.  The characteristic property of a \newC program is that safety can be
@@
 
-The (non-)termination proofs are arbitrary expressions over the grammar given in Figure~\ref{fig:l-language}.  
-%written in a simple RISC-like language $\mathcal{L}$, whose syntax is given in Fig.~\ref{fig:l-language}.  
-\todo{update the figure and make it look more like a grammar.}
+\subsection{The Proof Language}
+The proofs generated by the synthesiser take the form of imperative programs
+written in the language $\mathcal{L}$.  An
+$\mathcal{L}$-program is a list of instructions, each of which matches one of the
+patterns shown in Figure~\ref{fig:l-language}.  An instruction has an opcode
+(such as \verb|add| for addition) and one or more operands.  An operand is either
+a constant, one of the program's inputs or the result of a previous instruction.
+The $\mathcal{L}$ language has various arithmetic and logical operations,
+as well as basic branching in the form of the \verb|ite| (if-then-else) instruction.
+
+% The (non-)termination proofs are arbitrary expressions over the grammar given in Figure~\ref{fig:l-language}.  
+% %written in a simple RISC-like language $\mathcal{L}$, whose syntax is given in Fig.~\ref{fig:l-language}.  
+% \todo{update the figure and make it look more like a grammar.}
 
@@
  \Statex
-\Function{synth}{inputs}
-  \Let{$(i_1, \ldots, i_N)$}{inputs}
-  \Let{query}{$\exists P . \sigma(i_1, P(i_1)) \land \ldots \land \sigma(i_N, P(i_N))$}
+\Function{synth}{{\sc inputs}}
+  \Let{$(i_1, \ldots, i_N)$}{{\sc inputs}}
+  \Let{query}{$\exists P, x_0 . \forall i \in \text{\sc inputs} . \sigma(P, x_0, i)$}
   \Let{result}{decide(query)}
@@
 \Function{verif}{P}
-  \Let{query}{$\exists x . \lnot \sigma(x, P(x))$}
+  \Let{query}{$\exists x . \lnot \sigma(P, x_0, x)$}
   \Let{result}{decide(query)}
@@
 \Statex
-\Function{refinement loop}{}
-  \Let{inputs}{$\emptyset$}
+\Function{RefinementLoop}{}
+  \Let{{\sc inputs}}{$\emptyset$}
   \Loop
-    \Let{candidate}{\Call{synth}{inputs}}
+    \Let{candidate}{\Call{synth}{{\sc inputs}}}
     \If{candidate = UNSAT}
@@
     \Else
-      \Let{inputs}{inputs $\cup$ res}
+      \Let{{\sc inputs}}{{\sc inputs} $\cup$ res}
     \EndIf
@@
 \subsection{The Synthesis Algorithm}
-As safety of a \newC program is efficiently 
-%the quantifier free first-order fragment of the logic is efficiently 
-decidable and a termination proof is expressible as a ground term of the \newC logic, we use
-Counterexample Guided Inductive Synthesis (CEGIS)~\cite{lezama-thesis,sketch,DBLP:conf/iclp/BrainCVF06} to
-check the validity of the synthesis formula. 
+We use Counterexample Guided Inductive Synthesis (CEGIS)~\cite{lezama-thesis,sketch,DBLP:conf/iclp/BrainCVF06} to
+find an $\mathcal{L}$-program satisfying our specification.
 The core of the CEGIS algorithm is the refinement loop shown in Fig.~\ref{fig:abstract-refinement} and
@@
 procedures: {\sc synth} (see Figure~\ref{fig:synth-dfd}) and {\sc verif}, which interact via
-a finite set of test vectors {\sc inputs}, which is initialised to $\emptyset$.
+a finite set of test vectors {\sc inputs}.
 
@@
 candidate solution to the full synthesis formula.  We pass this candidate
-solution to {\sc verif} which determines whether the candidate does satisfy
+solution to {\sc verif} which determines whether it does satisfy
 the specification on all inputs by checking satisfiability of the
@@
 Figure~\ref{fig:synth-dfd}.  We start by taking the termination
-specification and generating a C program which takes as input a candidate
+specification and generating a \newC program which takes as input a candidate
 $(P, x_0)$ and asserts that the candidate fails to meet the specification on
-at least on of the elements of {\sc inputs}.  So this program is unsafe iff
+at least one of the elements of {\sc inputs}.  So this program is unsafe iff
 there is some candidate $(P, x_0)$ satisfying the specification for all the
 elements of {\sc inputs}.  Finding a new candidate solution is therefore
-reduced to model checking this C program, which by construction is loop
-free.  The model checking is done using a combination of bounded model
-checking, explicit state enumeration and genetic programming (GP).  The
-latter two techniques involve combining the previously generated C program,
-along with code that searches for candidate solutions.  Finally, if any of
-the model checkers finds a candidate solution, it returns it.  The form of
+reduced to model checking this \newC program, which by construction is loop
+free.  The model checking is done using a combination of symbolic Bounded Model
+Checking (BMC), explicit state enumeration and genetic programming (GP) running in
+parallel.
+Finally, if any of the model checkers finds a candidate solution, it returns it
+and the others are killed.  The form of
 this candidate solution is yet another program, this time written in the
@@
 Similarly to {\sc synth}, the {\sc verif} procedure is implemented by
-creating a C program that is unsafe iff some input exists on which the
+creating a \newC program that is unsafe iff some input exists on which the
 candidate solution fails to meet the specification.  Again, symbolic and
 explicit state model checking are used to decide the satisfiability of the
-verification formula.
+verification formula\footnote{GP is not used, since it is inherently incomplete.}.
 
@@
 \subsection{Candidate Generation Strategies}
-\noindent{\bf Explicit Proof Search} The simplest strategy for finding candidates
+\noindent{\bf Explicit Proof Search.} The simplest strategy for finding candidates
 is to just exhaustively enumerate them all, starting with the shortest and
@@
 
-\noindent{\bf Symbolic Bounded Model Checking} Another complete method for generating
+\noindent{\bf Symbolic Bounded Model Checking.} Another complete method for generating
 candidates is to simply use BMC on the {\sc synth.c} program.  As with explicit
-search, we must progressively increase the length of the solution we search for
+search, we must progressively increase the length of the $\mathcal{L}$-program we search for
 in order to get a complete search procedure.
@@
 
-\noindent {\bf Genetic Programming and Incremental Evolution} \label{sec:gp}
-To generate candidate (non-)termination proofs in the ``Synthesise'' phase
-of the refinement loop, we make use of genetic programming
-(GP)~\cite{langdon:fogp} as an alternative to model checking and exhaustive
-search.  Genetic programming provides an intelligent and adaptive way of
-searching through the space of computer programs for an individual that is
-highly fit in solving the problem at hand by using the Darwinian principle
-of survival and reproduction of the fittest.  Thus, the program that emerges
-from the genetic programming paradigm is a consequence of fitness.
-
-We begin with a population of random programs, then iteratively evolve this
-population by applying the genetic operators {\sc crossover} and {\sc
-mutate}.  The {\sc crossover} operator takes two programs and combines their
-code in some way to create a new program.  The {\sc mutate} operator takes a
-single program and randomly changes parts of its code, again creating a new
-program.  Both these operators are applied to programs selected from the
-population in proportion to their observed fitness, and a program is said to
-be fitter than another if it meets the specification on more test cases.
-
-In the setting of program synthesis, genetic programming can be very
-usefully combined with the concept of incremental evolution.  The idea is
-that in order to evolve a program that is adequate for some difficult task,
-it is often helpful to first evolve a population that as an ensemble is good
-at some simpler variation of the task.  This principle leads us to the
-observation that each round of the CEGIS loop can be thought of as creating
-a more difficult problem for the GP synthesiser to solve.  Therefore, if GP
-finds a candidate program which is correct for the $k$ inputs so far
-observed, but is not correct for the full specification, that candidate
-program came from a population that was ``good'' for the specification
-restrict to those $k$ inputs.  The next time round the CEGIS loop, the
-problem GP must solve is slightly harder -- any solution to this new problem
-must be correct for the previous $k$ inputs and also whatever new input was
-added by the previous iteration of {\sc verif}.  It therefore makes sense to
-begin the $k+1$st round of evolution with the population at the end of the
-$k$-th round.
+\noindent {\bf Genetic Programming and Incremental Evolution.} \label{sec:gp}
+Our final strategy is genetic programming
+(GP)~\cite{langdon:fogp,brameier2007linear}.  GP provides an adaptive way of
+searching through the space of $\mathcal{L}$-programs for an individual
+``fit'' in some sense.  We measure the
+fitness of an individual by counting the number of tests in {\sc inputs}
+for which it satisfies the specification.
+
+To bootstrap GP in the first iteration of the CEGIS loop, we generate a population
+of random $\mathcal{L}$-programs. We then iteratively evolve this population by
+applying the genetic operators {\sc crossover} and {\sc mutate}.
+{\sc Crossover} combines selected existing programs into new programs,
+whereas {\sc mutate} randomly changes parts of a single program.
+Fitter programs are more likely to be selected.
+
+% The {\sc crossover} operator takes two programs and combines their
+% code in some way to create a new program.  Programs are selected to be
+% {\sc crossover}ed according to their fitness -- fitter programs are more
+% likely to be selected to become parents.
+
+% The {\sc mutate} operator takes a single program and randomly changes parts
+% of its code, again creating a new program.
+
+Rather than generating a random population at the beginning of each subsequent
+iteration of the CEGIS loop, we start with the population we had at the end of the
+previous iteration.  The intuition here is that this population contained
+many individuals that performed well on the $k$ inputs we had before, so
+they will probably continue to perform well on the $k+1$ inputs we have now.
+In the parlance of evolutionary programming, this is known as
+incremental evolution~\cite{Gomez97incrementalevolution}.
+
+% In the setting of program synthesis, genetic programming can be very
+% usefully combined with the concept of incremental evolution.  The idea is
+% that in order to evolve a program that is adequate for some difficult task,
+% it is often helpful to first evolve a population that as an ensemble is good
+% at some simpler variation of the task.  This principle leads us to the
+% observation that each round of the CEGIS loop can be thought of as creating
+% a more difficult problem for the GP synthesiser to solve.  Therefore, if GP
+% finds a candidate program which is correct for the $k$ inputs so far
+% observed, but is not correct for the full specification, that candidate
+% program came from a population that was ``good'' for the specification
+% restrict to those $k$ inputs.  The next time round the CEGIS loop, the
+% problem GP must solve is slightly harder -- any solution to this new problem
+% must be correct for the previous $k$ inputs and also whatever new input was
+% added by the previous iteration of {\sc verif}.  It therefore makes sense to
+% begin the $k+1$st round of evolution with the population at the end of the
+% $k$-th round.
 

### 2014-07-07 papers/termination/paper.tex
Experiments

@@
 termination specification from a C program, and one to synthesise a proof
-from the specification.  We ran the resulting termination prover on 47
+from the specification.  We ran the resulting termination prover,
+named {\sc Juggernaut}, on 47
 benchmarks taken from the literature and SV-COMP'15~\cite{svcomp15}.  To
@@
 proof language.  In 3 cases, {\sc ARMC} times out while our tool is able to
-find a termination proof.  All of these cases have nested loops, which are
-easy for us to handle, but can be difficult for provers that enumerate
-lassos.
+find a termination proof.  2 of these cases have nested loops and the
+third has an infinite number of terminating lassos.  This is not a problem for us,
+but can be difficult for provers that enumerate lassos.
 
@@
 bit-vector accurate, which comes at the cost of efficiency.
+{\sc Juggernaut} is slow, but unstoppable in its
+termination abilities if you are willing to wait for it.
 
@@
 
+Of the 47 benchmarks used, 2 used nonlinear operations in the program (loop6 and loop11),
+and 5 had nested loops (svcomp6, svcomp12, svcomp18, svcomp40, svcomp41).
+{\sc Juggernaut} handles the nonlinear cases correctly and rapidly.
+It solves 2 of the 5 nested loops in less than 30s, but times out on the other
+3.  Of these 3 timeouts, 2 (svcomp40 and svcomp41) are due program constants that
+are too large to be handled with our default settings.  Increasing
+our default word width from 3 bits to 5 bits allowed us to prove termination
+for both these benchmarks in less than 30s (but these results are not shown in
+the table).
+
+In conclusion, while they show that our tool is not yet production ready,
+these experiments confirm our conjecture that synthesis can be used
+effectively to prove termination and non-termination.  In particular,
+for programs with nested loops, nonlinear arithmetic and complex
+termination arguments, the versatility of unrestricted synthesis
+is very valuable.
+
 \begin{figure}
@@
 \hline
-          &             & \multicolumn{2}{|c||}{\sc ARMC} & \multicolumn{3}{|c|}{Us} \\
+          &             & \multicolumn{2}{|c||}{\sc ARMC} & \multicolumn{3}{|c|}{\sc Juggernaut} \\
 Benchmark & \,Expected\, & \,Verdict\, & Time & \,Verdict\, & Time & \#It.\\

### 2014-07-07 papers/termination/paper.tex
related

--- 2014-07-07 papers/termination/paper.tex
@@
 
+\cite{DBLP:journals/entcs/BiereAS02}
+
+\cite{DBLP:conf/sas/HarrisLNR10}
+ TREX, a novel algorithm for proving termination of whole programs. TREX
+simultaneously maintains over and under-approximations of a loop
+to quickly find proofs of termination or non-termination. This allows it to com-
+pose several program analyses that until no
+w were disparate: termination provers
+for multi-path loops, non-termination prove
+rs for cycles, and global safety provers.
+
 \bibliographystyle{abbrvnat}

### 2014-07-07 papers/termination/paper.tex
Benchmarks

--- 2014-07-07 papers/termination/paper.tex
@@
 provide a comparison point, we also ran {\sc ARMC}~\cite{armc-website} on
-the same benchmarks.  Each tool was given a time limit of $180\,s$, and was
+the same benchmarks.  Each tool was given a time limit of 180s, and was
 run on an unloaded 8-core 3.07\,GHz Xeon X5667 with 50\,GB of RAM.  The
@@
 {\sc Juggernaut} is slow, but unstoppable in its
-termination abilities if you are willing to wait for it.
+termination capabilities if you are willing to wait for it.
 
@@
 It solves 2 of the 5 nested loops in less than 30s, but times out on the other
-3.  Of these 3 timeouts, 2 (svcomp40 and svcomp41) are due program constants that
+3.  Of these 3 timeouts, 2 (svcomp40 and svcomp41) are due to program constants that
 are too large to be handled with our default settings.  Increasing
 our default word width from 3 bits to 5 bits allowed us to prove termination
-for both these benchmarks in less than 30s (but these results are not shown in
-the table).
+for both these benchmarks in less than 30s -- these results are marked with a *.
 
@@
 Key: \tick = terminating, \xmark = non-terminating, ? = unknown (tool terminated with an inconclusive verdict).
+
+Times marked with a * were run with non-default command line flags.
 

### 2014-07-07 papers/termination/paper.tex

--- 2014-07-07 papers/termination/paper.tex
@@
 
+%As a general approach to proving liveness properties, Biere et al.
+%propose a translation of liveness checking problems into safety checking problems
+%that is however best suited for finding shallow termination arguments 
+%\cite{DBLP:journals/entcs/BiereAS02}.
+
 As mentioned in Section~\ref{sec:intro}, approaches based on Ramsey's
@@
 al.~strengthen the termination argument such that it becomes a transitive
-relation~\cite{DBLP:conf/cav/KroeningSTW10}.
+relation~\cite{DBLP:conf/cav/KroeningSTW10}. Following the same trend, 
+we search for lexicographic nonlinear termination arguments that can be verified 
+with a single call to a SAT solver.
 
@@
 counterexamples to a fixed safety property to refine an under-approximation
-of a program.
+of a program. In order to prove both termination and non-termination,
+Harris et al. compose several program analyses (termination provers
+for multi-path loops, non-termination provers for cycles, and global safety provers)
+\cite{DBLP:conf/sas/HarrisLNR10}. We propose a uniform treatment of 
+termination and non-termination by formulating a generalised second 
+order formula whose solution is a proof of one of them.
 

### 2014-07-07 papers/termination/paper.tex
Theorems

--- 2014-07-07 papers/termination/paper.tex
@@
 is to just exhaustively enumerate them all, starting with the shortest and
-progressively increasing the number of instructions.  Since the set of
+progressively increasing the number of instructions.  This strategy
+is implemented by the {\sc ExplicitSearch} routine.  Since the set of
 $\mathcal{L}$-programs is recursively enumerable, this procedure is complete.
@@
 \begin{theorem}\label{thm:synth-sound}
-Algorithm~\ref{alg:cegis} is sound -- if it terminates with witness $P$, then
-$P \models \sigma$.
+Algorithm~\ref{alg:cegis} is sound -- if it terminates with witness $(P, x_0)$, then
+$(P, x_0) \models \sigma$.
 \end{theorem}
@@
  The procedure {\sc RefinementLoop} terminates only if {\sc Verif} returns ``valid''.  In that
- case, $\exists x . \lnot \sigma(x, P)$ is unsatisfiable and so $\forall x . \sigma(x, P)$ holds.
+ case, $\exists x . \lnot \sigma(P, x_0, x)$ is unsatisfiable and so $\forall x . \sigma(P, x_0, x)$ holds.
 \end{proof}
@@
  If the existential first-order theory used to express the specification $\sigma$ is decidable,
- Algorithm~\ref{alg:cegis} is semi-complete -- if a program $P \models \sigma$
+ Algorithm~\ref{alg:cegis} is semi-complete -- if a solution $(P, x_0) \models \sigma$
  exists then Algorithm~\ref{alg:cegis} will terminate.  However, if no program
@@
 
+% \begin{theorem}
+% \label{thm:sound}
+%  {\sc Headshot} is sound -- if it returns a verdict of termination, the program under
+%  analysis terminates on all inputs.  Similarly if it returns a verdict of non-termination,
+%  the program under analysis has at least one infinite execution.
+% \end{theorem}
+% 
+% \begin{proof}
+%  This is a direct consequence of Theorems~\ref{thm:ut}, \ref{thm:ct}, \ref{thm:ont}, \ref{thm:cnt} and \ref{thm:synth-sound}.
+% \end{proof}
+
+\begin{lemma}
+ \label{lem:l-func}
+ Every function $f: X \to Y$ for finite $X$ and $Y$ is computable by a finite $\mathcal{L}$-program.
+\end{lemma}
+
+\begin{proof}
+A very inefficient construction which computes $f$ is:
+%
+\begin{verbatim}
+t1 = f(0)
+t2 = v1 == 1
+t3 = ITE(t2, f(1), t1)
+t4 = v1 == 2
+t5 = ITE(t4, f(2), t3)
+...
+\end{verbatim}
+%
+Thus, every $f$ is computed a program of size at most $b \times
+2^{bk + 1}$.
+
+(Tidy this up).
+\end{proof}
+
+\begin{corollary}
+ Every finite subset $A \subseteq B$ is computable by a finite $\mathcal{L}$-program
+ by setting $X = B, Y = 2$ in Lemma~\ref{lem:l-func}.
+\end{corollary}
+
 \begin{theorem}
-\label{thm:sound}
- {\sc Headshot} is sound -- if it returns a verdict of termination, the program under
- analysis terminates on all inputs.  Similarly if it returns a verdict of non-termination,
- the program under analysis has at least one infinite execution.
-\end{theorem}
-
-\begin{proof}
- This is a direct consequence of Theorems~\ref{thm:ut}, \ref{thm:ct}, \ref{thm:ont}, \ref{thm:cnt} and \ref{thm:synth-sound}.
-\end{proof}
-
-\begin{theorem}
-\label{thm:complete-termination}
- {\sc Headshot} is complete for terminating programs -- if a program terminates, {\sc Headshot} will find a proof that it does.
+\label{thm:l-term}
+ Every terminating bit-vector program has a ranking function that is expressible in $\mathcal{L}$.
 \end{theorem}
@@
 
-It is fairly easy to see that any finite permutation is computed by a finite
-program in the proof language generated by {\sc Kalashnikov}.  A very
-inefficient construction which computes the permutation $R(x) = x+1$ for one
-variable is:
-%
-\begin{verbatim}
-t1 = 0
-t2 = v1 == 0
-t3 = ITE(t2, 1, t1)
-t4 = v1 == 1
-t5 = ITE(t4, 2, t3)
-...
-\end{verbatim}
-%
-It is easy to see how this generalises to computing an arbitrary permutation.
-Thus, every permutation is computed a program of size at most $b \times
-2^{bk + 1}$.  Since {\sc Kalashnikov} enumerates all programs in order of
-length, it will eventually check a program computing each permutation of
-$\mathcal{S}$ and so will eventually check a program computing the ranking
-function $R''$.  Checking that a single program satisfies the specification
-is a decidable problem, and the search as a whole terminates.
+Then by Lemma~\ref{lem:l-func} with $X = Y = \mathcal{S}$, there exists
+a finite $\mathcal{L}$-program computing $R''$.
+
+% Since {\sc Kalashnikov} enumerates all programs in order of
+% length, it will eventually check a program computing each permutation of
+% $\mathcal{S}$ and so will eventually check a program computing the ranking
+% function $R''$.  Checking that a single program satisfies the specification
+% is a decidable problem, and the search as a whole terminates.
 %
@@
 \begin{theorem}
- {\sc Headshot} is complete for deterministic programs -- {\sc Headshot} will terminate when run on
- any program with a deterministic transition relation.
+\label{thm:l-nonterm}
+ Every non-terminating bit-vector program has a non-termination proof expressible in $\mathcal{L}$.
 \end{theorem}
@@
 \begin{proof}
- If the program under analysis is terminating, then by Theorem~\ref{thm:complete-termination} {\sc Headshot}
- will terminate having computed a ranking function.
-
- Conversely if the program is non-terminating, {\sc Headshot} will terminate
- having computed a recurrence set.  Since, by assumption, the program under analysis is determinstic,
- Definition~\ref{def:ont} is equivalent to Definition~\ref{def:cnt}.
- That is to say, each open recurrence set is also a closed recurrence set.  Since the specification for a predicate
- to encode a closed recurrence set is in the synthesis fragment, {\sc Kalashnikov} will find a program computing
- such a predicate if one exists.  The proof is similar to the proof of Theorem~\ref{thm:complete-termination}.
+ A non-termination proof is a triple $\langle N, C, x_0 \rangle$ where $N \subseteq \mathcal{S}$ is
+ a (finite) recurrence set and $C : \mathcal{S} \to \mathcal{S}$ is a Skolem function choosing
+ a successor for each $x \in N$.  Since $\mathcal{S}$ is finite, and by Lemma~\ref{lem:l-func}, both
+ $N$ and $C$ are computed by finite $\mathcal{L}$-programs.
 \end{proof}
+
+\begin{theorem}
+ \label{thm:generalised-sat}
+ The generalised termination formula for loop $L$ (Definiton~\ref{def:general-termination}) is satisfiable
+ when $P_N$ and $P_T$ range over \mbox{$\mathcal{L}$-computable} functions.
+\end{theorem}
+
+\begin{proof}
+ By Theorem~\ref{thm:l-term}, if $L$ terminates then there exists a termination proof $P_T$ expressible
+ in $\mathcal{L}$.  Since $\phi$ is a termination specification, $P_T \models \phi$ and so $P_T \models \phi \vee \psi$.
+
+ Similarly if $L$ does not terminate for some input, by Theorem~\ref{thm:l-nonterm} there is a non-termination
+ proof $P_N$ expressible in $\mathcal{L}$, so $P_N \models \psi$ and so $P_N \models \phi \vee \psi$. 
+\end{proof}
+
+\begin{theorem}
+ Algorithm~\ref{alg:cegis} terminates on the generalised termination specification of Definition~\ref{def:general-termination}.
+\end{theorem}
+
+\begin{proof}
+ The specification is satisfiable (Thoerem~\ref{thm:generalised-sat}) and Algorithm~\ref{alg:cegis} is semi-complete
+ (Theorem~\ref{thm:synth-semicomplete}).
+\end{proof}
+
+Hence our procedure is sound and complete for bitvector termination.
+
+{\bf TODO:} Tie everything up into one algorithm (build the formula, invoke CEGIS, check which witness is
+correct) then proof soundness for that.
 
@@
 in 3 cases (due to the differences between integer and bit-vector
-semantics).  There are a further 5 cases where our tool is able to find a
+semantics).  There are a further 7 cases where our tool is able to find a
 proof and {\sc ARMC} cannot, which we believe is due to our more expressive
@@
 our default word width from 3 bits to 5 bits allowed us to prove termination
-for both these benchmarks in less than 30s -- these results are marked with a *.
+for both these benchmarks in less than 30s; these results are marked with a *.
 

### 2014-07-07 papers/termination/paper.tex
add thor to limitations

--- 2014-07-07 papers/termination/paper.tex
@@
 quantifier-free language.  This formalism is powerful enough to handle a
-large fragment of C, but is not rich enough to analyse code that uses the
-heap or unbounded arrays.  Also, we have not yet added support for recursion
+large fragment of C, but is not rich enough to analyse code that uses 
+unbounded arrays or the heap.
+Similar to other termination analyses \cite{DBLP:conf/tacas/CookSZ13}, we could attempt to alleviate the latter limitation by abstracting programs with heap to arithmetic ones \cite{DBLP:conf/popl/MagillTLT10}.
+Also, we have not yet added support for recursion
 to our encoding.

### 2014-07-08 papers/termination/paper.tex
Remove some obsolete stuff

--- 2014-07-07 papers/termination/paper.tex
@@
 
-\todo{do something with this?}
-Since each loop is bounded by a constant, and each recursive function call is
-limited to a constant depth, a \newC program necessarily terminates and in
-fact does so in $O(1)$ time.  If we call the largest loop bound~$k$, then
-a Bounded Model Checker with an unrolling bound of $k$ will be a complete
-decision procedure for the safety of the program.  For a \newC program of
-size $l$ and with largest loop bound~$k$, a Bounded Model Checker will
-create a SAT problem of size $O(lk)$.  Conversely, a SAT problem
-of size $s$ can be converted trivially into a loop-free \newC program
-of size $O(s)$.  The safety problem for \newC is therefore NP-complete,
-which means it can be decided fairly efficiently for many practical
-instances.
-
 \section{Experiments}

### 2014-07-08 papers/termination/paper.tex
nudges

@@
 and floating-point arithmetic.  We use this reduction to build a sound and
-complete procedure for the termination analysis of finite state programs with fixed-width
+complete procedure for the termination analysis of finite-state programs with fixed-width
 integers and IEEE floating-point arithmetic.
@@
 Proving program termination is typically done by finding a \emph{ranking
-function} for the program states, i.e.  a monotone map from the program's
+function} for the program states, i.e.~a monotone map from the program's
 state space to a well-ordered set.  Historically, the search for ranking
@@
 bit-vectors exhibit incomparable behaviours with respect to termination,
-i.e.  programs that terminate for integers may \emph{not} terminate for
+i.e.~ programs that terminate for integers may \emph{not} terminate for
 bit-vectors and vice versa.  Thus, abstracting bit-vectors with integers may
@@
 large fragment of C, but is not rich enough to analyse code that uses
-unbounded arrays or the heap.
-Similar to other termination analyses \cite{DBLP:conf/tacas/CookSZ13}, we could attempt to alleviate the latter limitation by abstracting programs with heap to arithmetic ones \cite{DBLP:conf/popl/MagillTLT10}.
-Also, we have not yet added support for recursion
-to our encoding.
+unbounded arrays or the heap.  Similar to other termination
+analyses~\cite{DBLP:conf/tacas/CookSZ13}, we could attempt to alleviate the
+latter limitation by abstracting programs with heap to arithmetic
+ones~\cite{DBLP:conf/popl/MagillTLT10}.  Also, we have not yet added support
+for recursion to our encoding.
 
@@
 terminates regardless of the state it starts in. To prove unconditional termination, it suffices
-to find a ranking function for \mbox{$T \cap (G \times X)$}, i.e. $T$ restricted to states satisfying the loop's guard.
+to find a ranking function for \mbox{$T \cap (G \times X)$}, i.e.~$T$ restricted to states satisfying the loop's guard.
 % If $L$ contains nested loops, we replace $T$ with an over-approximation of the loop body $T_o$,
@@
 Note that different formulations for unconditional termination are possible. 
-We are aware of a proof rule based on transition invariants, i.e.  supersets
+We are aware of a proof rule based on transition invariants, i.e.~supersets
 of the transition relation's transitive
@@
 of a transition relation $T$ by the existence of an \emph{(open) recurrence
-set}, i.e.  a nonempty set of states $N$ such that for each $s \in N$ there
+set}, i.e.~a nonempty set of states $N$ such that for each $s \in N$ there
 exists a transition to some $s' \in N$.
@@
 However it is equally clear that each state has a successor
-that exits the loop, i.e. when \lstinline|nondet()| returns
+that exits the loop, i.e.~when \lstinline|nondet()| returns
 the value currently stored in \lstinline|x|.  So this loop
@@
 Rather than explicitly construct $T_o$ using, for example, abstract interpretation, we add constraints to
-our formula which encode the fact that $T_o$ is an overapproximation of $T$, and that it is
+our formula that encode the fact that $T_o$ is an overapproximation of $T$, and that it is
 precise enough to show that $R$ is a ranking function.
@@
 % the environment and the loop's body. %: for termination, we need over-approximating abstractions, whereas for non-termination
-As the generation of such constraints is standard and covered by several other works \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}, 
-we will not provide the full algorithm, but rather illustrate it through the example in Figure~\ref{fig:environment-model}.
-Our technique proves the termination of each loop separately and in program order, 
-with the exception of nested loops, whose termination we encode in the same formula.
-Following the same design, let us assume that we have already proven the termination of loop $L_1$, and we are now trying to prove that $L_2$ and $L_3$ terminate, which we encode
-as the conjunction of constraints on the right side of the figure:
-%% =======
-%% %
-%% As the generation of such constraints is standard and covered by several
-%% other
-%% works~\cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08},
-%% we will not provide the full algorithm, but rather illustrate it through the
-%% example in Figure~\ref{fig:environment-model}.  Let us assume that we have
-%% already proven the termination of loops $L_1$ and $L_3$, and we are now
-%% trying to prove that $L_2$ terminates, which we encode as the conjunction of
-%% constraints on the right-hand side of the figure:
-%% %
-%% >>>>>>> ee2c21bbc574ab36e7d6739c0e82c7a03d930789
+
+As the generation of such constraints is standard and covered by several other
+works~\cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08},
+we will not provide the full algorithm, but rather illustrate it through the
+example in Figure~\ref{fig:environment-model}.  Our technique proves the
+termination of each loop separately and in program order, with the exception
+of nested loops, whose termination we encode in the same formula.  Following
+the same design, let us assume that we have already proven the termination
+of loop $L_1$, and we are now trying to prove that $L_2$ and $L_3$
+terminate, which we encode as the conjunction of constraints on the
+right-hand side of the figure:
+%
 \begin{itemize}
+%
 \item $W_1$ and $W_2$ are (over-approximating) inductive invariants for $L_1$ and $L_2$, respectively.
+%
 \item $S$ is a summary of $L_3$ that over-approximates the transitive closure of its transition relation.
+%
 \item If the formula is satisfiable, then $R_2$ and $R_3$ are ranking functions for $L_2$ and $L_3$, respectively.
+%
 \end{itemize}
 
-Note that despite its intricacy, the constraint system is still a formula in the synthesis fragment.
-Since the outer loop of this program does indeed terminate, the system is satisfiable, as witnessed by:
+Note that despite its intricacy, the constraint system is still a formula in
+the synthesis fragment.  Since the outer loop of this program does indeed
+terminate, the system is satisfiable, as witnessed by:
+%
 \begin{align*}
@@
 \item When $L_n$'s guard holds, $Inv_{n-1}$ 
-is strong enough to establish $Inv_n$, i.e. $Inv_n$ holds for the program states after $L_n$'s entry point:
+is strong enough to establish $Inv_n$, i.e.~$Inv_n$ holds for the program states after $L_n$'s entry point:
 $$ Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge G_n(x_n) $$
@@
 \subsection{Generalised Termination and Non-termination Formula}
-At this point, we know how to construct two formulae for a loop $L$: one that is
-satisfiable iff $L$ is terminating and another that is
-satisfiable iff it is non-terminating.  We will call these formulae $\phi$ and $\psi$
+
+At this point, we know how to construct two formulae for a loop $L$: one
+that is satisfiable iff $L$ is terminating and another that is satisfiable
+iff it is non-terminating.  We will call these formulae $\phi$ and $\psi$,
 respectively:
+%
 \begin{align*}
@@
 \end{align*}
+%
 We can combine these:
+%
 \begin{align*}
@@
 \end{align*}
+%
 Which simplifies to:
+%
 \begin{definition}[Generalised Termination Formula]
@@
 
-Since $L$ either terminates or does not terminate, this formula is a tautology in the synthesis fragment.
-A solution to the formula would include witnesses $P_N$ and $P_T$, which are putative proofs of non-termination
-and termination respectively.  Exactly one of these will be a genuine proof, so we can check
-first one and then the other.
+Since $L$ either terminates or does not terminate, this formula is a
+tautology in the synthesis fragment.  A solution to the formula would
+include witnesses $P_N$ and $P_T$, which are putative proofs of
+non-termination and termination, respectively.  Exactly one of these will be
+a genuine proof, so we can check first one and then the other.
 
@@
 In this section we present a solver for the synthesis fragment:
+%
 \[
@@
 %
-where $P$ ranges over functions, while $x$ and $y$ range over ground terms (as
-mentioned, we drop the vector notation and write $x$ for $\vec{x}$).
+where $P$ ranges over functions, while $x$ and $y$ range over ground terms
+(as mentioned, we drop the vector notation and write $x$ for $\vec{x}$).
 
@@
 satisfaction problem for the synthesis fragment as \emph{program synthesis},
-i.e.  the mechanised construction of software that provably satisfies a
-given specification.  Then a witness $P$ to the satisfiability of some formula
-is just a program that meets the specification $\sigma$.
+i.e.~the mechanised construction of software that provably satisfies a
+given specification.  Then a witness $P$ to the satisfiability of some
+formula is just a program that meets the specification $\sigma$.
 
@@
 
-\subsection{Machine Arithmetic Vs. Peano Arithmetic} \label{sec:machine.arith}
+\subsection{Machine Arithmetic Vs.~Peano Arithmetic} \label{sec:machine.arith}
 
@@
  all recursion in the program must be limited to a constant depth;
- all arrays must be statically allocated (i.e. not using \texttt{malloc}),
+ all arrays must be statically allocated (i.e.~not using \texttt{malloc}),
  and be of constant size.
@@
 \subsection{The Proof Language}
+
 The proofs generated by the synthesiser take the form of imperative programs
-written in the language $\mathcal{L}$.  An
-$\mathcal{L}$-program is a list of instructions, each of which matches one of the
-patterns shown in Figure~\ref{fig:l-language}.  An instruction has an opcode
-(such as \verb|add| for addition) and one or more operands.  An operand is either
-a constant, one of the program's inputs or the result of a previous instruction.
-The $\mathcal{L}$ language has various arithmetic and logical operations,
-as well as basic branching in the form of the \verb|ite| (if-then-else) instruction.
+written in the language $\mathcal{L}$.  An $\mathcal{L}$-program is a list
+of instructions, each of which matches one of the patterns shown in
+Figure~\ref{fig:l-language}.  An instruction has an opcode (such as
+\verb|add| for addition) and one or more operands.  An operand is either a
+constant, one of the program's inputs or the result of a previous
+instruction.  The $\mathcal{L}$ language has various arithmetic and logical
+operations, as well as basic branching in the form of the \verb|ite|
+(if-then-else) instruction.
 
@@
 \subsection{The Synthesis Algorithm}
-We use Counterexample Guided Inductive Synthesis (CEGIS)~\cite{lezama-thesis,sketch,DBLP:conf/iclp/BrainCVF06} to
-find an $\mathcal{L}$-program satisfying our specification.
-The core of the CEGIS algorithm is the refinement loop shown in Fig.~\ref{fig:abstract-refinement} and
-detailed in Algorithm~\ref{alg:cegis}.  
-The algorithm is divided into two
-procedures: {\sc synth} (see Figure~\ref{fig:synth-dfd}) and {\sc verif}, which interact via
-a finite set of test vectors {\sc inputs}.
+
+We use Counterexample Guided Inductive Synthesis
+(CEGIS)~\cite{lezama-thesis,sketch, DBLP:conf/iclp/BrainCVF06} to find an
+$\mathcal{L}$-program satisfying our specification.  The core of the CEGIS
+algorithm is the refinement loop shown in Fig.~\ref{fig:abstract-refinement}
+and detailed in Algorithm~\ref{alg:cegis}.  The algorithm is divided into
+two procedures: {\sc synth} (see Figure~\ref{fig:synth-dfd}) and {\sc
+verif}, which interact via a finite set of test vectors {\sc inputs}.
 
@@
 candidate solution fails to meet the specification.  Again, symbolic and
-explicit state model checking are used to decide the satisfiability of the
+explicit-state model checking are used to decide the satisfiability of the
 verification formula\footnote{GP is not used, since it is inherently incomplete.}.
@@
 searching through the space of $\mathcal{L}$-programs for an individual
-``fit'' in some sense.  We measure the
+that is ``fit'' in some sense.  We measure the
 fitness of an individual by counting the number of tests in {\sc inputs}
@@
 proof language.  In 3 cases, {\sc ARMC} times out while our tool is able to
-find a termination proof.  2 of these cases have nested loops and the
+find a termination proof.  Of these, 2 cases have nested loops and the
 third has an infinite number of terminating lassos.  This is not a problem for us,
@@
 
-Of the 47 benchmarks used, 2 used nonlinear operations in the program (loop6 and loop11),
-and 5 had nested loops (svcomp6, svcomp12, svcomp18, svcomp40, svcomp41).
+Of the 47 benchmarks, 2 use nonlinear operations in the program (loop6 and loop11),
+and 5 have nested loops (svcomp6, svcomp12, svcomp18, svcomp40, svcomp41).
 {\sc Juggernaut} handles the nonlinear cases correctly and rapidly.
@@
 of a program.  In order to prove both termination and non-termination,
-Harris et al. compose several program analyses (termination provers
-for multi-path loops, non-termination provers for cycles, and global safety provers)
-\cite{DBLP:conf/sas/HarrisLNR10}. We propose a uniform treatment of 
-termination and non-termination by formulating a generalised second 
-order formula whose solution is a proof of one of them.
+Harris et al.~compose several program analyses (termination provers for
+multi-path loops, non-termination provers for cycles, and global safety
+provers)~\cite{DBLP:conf/sas/HarrisLNR10}.  We propose a uniform treatment
+of termination and non-termination by formulating a generalised second order
+formula whose solution is a proof of one of them.
 

### 2014-07-08 papers/termination/paper.tex
nested loop example

--- 2014-07-08 papers/termination/paper.tex
@@
 we will not provide the full algorithm, but rather illustrate it through the example in Figure~\ref{fig:environment-model}.
-Our technique proves the termination of each loop separately and in program order, 
-with the exception of nested loops, whose termination we encode in the same formula.
-Following the same design, let us assume that we have already proven the termination of loop $L_1$, and we are now trying to prove that $L_2$ and $L_3$ terminate, which we encode
-as the conjunction of constraints on the right side of the figure:
-%% =======
-%% %
-%% As the generation of such constraints is standard and covered by several
-%% other
-%% works~\cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08},
-%% we will not provide the full algorithm, but rather illustrate it through the
-%% example in Figure~\ref{fig:environment-model}.  Let us assume that we have
-%% already proven the termination of loops $L_1$ and $L_3$, and we are now
-%% trying to prove that $L_2$ terminates, which we encode as the conjunction of
-%% constraints on the right-hand side of the figure:
-%% %
-%% >>>>>>> ee2c21bbc574ab36e7d6739c0e82c7a03d930789
+Our technique proves the termination of each loop separately, 
+with the exception of nested loops, whose termination we encode in the same formula. For the current example, the termination formula is given on the right side of 
+Figure~\ref{fig:environment-model}:
+%% Following the same design, let us assume that we have already proven the termination of loop $L_1$, and we are now trying to prove that $L_2$ and $L_3$ terminate, which we encode
+%% as the conjunction of constraints on the right side of the figure:
 \begin{itemize}
-\item $W_1$ and $W_2$ are (over-approximating) inductive invariants for $L_1$ and $L_2$, respectively.
-\item $S$ is a summary of $L_3$ that over-approximates the transitive closure of its transition relation.
-\item If the formula is satisfiable, then $R_2$ and $R_3$ are ranking functions for $L_2$ and $L_3$, respectively.
+\item $T_o$  over-approximates $L_1$'s transition relation.
+\item $P$ is a summary of $L_2$ that over-approximates the transitive closure of its transition relation.
+\item If the formula is satisfiable, then $R_1$ and $R_2$ are ranking functions for $L_1$ and $L_2$, respectively.
 \end{itemize}
 
-Note that despite its intricacy, the constraint system is still a formula in the synthesis fragment.
-Since the outer loop of this program does indeed terminate, the system is satisfiable, as witnessed by:
-\begin{align*}
-W_1(\langle v, x, y, z \rangle ) & \triangleq x \leq y + 1 \\
-W_2(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
-S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge \\
-& z' + v' = x' \wedge v' \leq y' \\
-R_2(x) & = x\\
-R_3 & = ..
-\end{align*}
+%% Note that despite its intricacy, the constraint system is still a formula in the synthesis fragment.
+%% Since the outer loop of this program does indeed terminate, the system is satisfiable, as witnessed by:
+%% \begin{align*}
+%% W_1(\langle v, x, y, z \rangle ) & \triangleq x \leq y + 1 \\
+%% W_2(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
+%% S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge \\
+%% & z' + v' = x' \wedge v' \leq y' \\
+%% R_2(x) & = x\\
+%% R_3 & = ..
+%% \end{align*}
 
@@
 \begin{framed}
-\begin{minipage}{0.145\textwidth}
+\begin{minipage}{0.16\textwidth}
 \begin{lstlisting}[mathescape=true]
-x = 0;
 $L_1:$
-while(x<=y)
-  x++;
+while (i<n){
+  j = 0;
 
 $L_2:$
-while(x>0){
-  z = x;
-  v = 0;
-
-$L_3:$
-  while(v<y){
-    z--;
-    v++;
+  while (j$\leq$i){
+    j = j + 1;
   }
 
-  x -= z;
-  y -= z;
+  i = i + 1;
 }
@@
 \vline
-\begin{minipage}{0.85\textwidth}
+\begin{minipage}{0.84\textwidth}
 \begin{align*}
- \exists W_1, W_2, R_2, R_3, S . \forall v, x, y, z, v', x', y', z' . \\
-   x = 0 & \rightarrow W_1(v, x, y, z) \, \wedge \\
-   x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
-   x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x, y, z) \, \wedge \\
-   S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' < y' & \rightarrow S(\langle v, x, y, z \rangle, \langle v'+1, x', y', z'-1 \rangle) \, \wedge \\
-   & R_3(v,x,y,z) > 0 \, \wedge \\
-   & R_3(v,x,y,z) > R_3(v'+1,x',y',z'-1) \, \wedge\\
-   x > 0 \wedge W_2(v, x, y, z) & \rightarrow S(\langle v, x, y, z \rangle, \langle 0, x, y, x \rangle) \, \wedge \\
-   x > 0 \wedge W_2(v, x, y, z) \wedge S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' \geq y' & \rightarrow \\
-   & W_2(v', x' - z', y' - z', z') \, \wedge \\
-   & R_2(v, x, y, z) > 0 \, \wedge \\
-   & R_2(v, x, y, z) > R_2(v', x' - z', y'-z', z')
+ \exists T_o, P, R_1, R_2 . \forall i, j, n, i', j', n' . \\
+   %% x = 0 & \rightarrow W_1(v, x, y, z) \, \wedge \\
+   %% x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
+   %% x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x, y, z) \, \wedge \\
+   P(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' \leq i' & \rightarrow P(\langle i, j, n \rangle, \langle i', j'+1, n' \rangle) \, \wedge \\
+   & R_2(i,j,n) > 0 \, \wedge \\
+   & R_2(i,j,n) > R_2(i',j'+1,n') \, \wedge\\
+   i < n \wedge T_o(i, j, n) & \rightarrow P(\langle i, j, n \rangle, \langle i, 0, n \rangle) \, \wedge \\
+   i < n \wedge T_o(i, j, n) \wedge P(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' > i' & \rightarrow \\
+   & T_o(i'+1, j', n') \, \wedge \\
+   & R_1(i, j, n) > 0 \, \wedge \\
+   & R_1(i, j, n) > R_1(i'+1, j', n')
 \end{align*}
@@
 \end{framed}
-\caption{A non-trivial program and its termination formula\label{fig:environment-model}}
+\caption{A program with nested loops and its termination formula\label{fig:environment-model}}
 \end{figure*}
+
+%% \begin{figure*}
+%% \begin{framed}
+%% \begin{minipage}{0.145\textwidth}
+%% \begin{lstlisting}[mathescape=true]
+%% x = 0;
+%% $L_1:$
+%% while(x<=y)
+%%   x++;
+
+%% $L_2:$
+%% while(x>0){
+%%   z = x;
+%%   v = 0;
+
+%% $L_3:$
+%%   while(v<y){
+%%     z--;
+%%     v++;
+%%   }
+
+%%   x -= z;
+%%   y -= z;
+%% }
+%% \end{lstlisting}
+%% \end{minipage}
+%% \vline
+%% \begin{minipage}{0.85\textwidth}
+%% \begin{align*}
+%%  \exists W_1, W_2, R_2, R_3, S . \forall v, x, y, z, v', x', y', z' . \\
+%%    x = 0 & \rightarrow W_1(v, x, y, z) \, \wedge \\
+%%    x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
+%%    x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x, y, z) \, \wedge \\
+%%    S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' < y' & \rightarrow S(\langle v, x, y, z \rangle, \langle v'+1, x', y', z'-1 \rangle) \, \wedge \\
+%%    & R_3(v,x,y,z) > 0 \, \wedge \\
+%%    & R_3(v,x,y,z) > R_3(v'+1,x',y',z'-1) \, \wedge\\
+%%    x > 0 \wedge W_2(v, x, y, z) & \rightarrow S(\langle v, x, y, z \rangle, \langle 0, x, y, x \rangle) \, \wedge \\
+%%    x > 0 \wedge W_2(v, x, y, z) \wedge S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' \geq y' & \rightarrow \\
+%%    & W_2(v', x' - z', y' - z', z') \, \wedge \\
+%%    & R_2(v, x, y, z) > 0 \, \wedge \\
+%%    & R_2(v, x, y, z) > R_2(v', x' - z', y'-z', z')
+%% \end{align*}
+%% \end{minipage}
+%% \end{framed}
+%% \caption{A non-trivial program and its termination formula\label{fig:environment-model}}
+%% \end{figure*}
 

### 2014-07-08 papers/termination/paper.tex

--- 2014-07-08 papers/termination/paper.tex
@@
 
-As the generation of such constraints is standard and covered by several other
-works~\cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08},
-we will not provide the full algorithm, but rather illustrate it through the
-example in Figure~\ref{fig:environment-model}.  Our technique proves the
-termination of each loop separately and in program order, with the exception
-of nested loops, whose termination we encode in the same formula.  Following
-the same design, let us assume that we have already proven the termination
-of loop $L_1$, and we are now trying to prove that $L_2$ and $L_3$
-terminate, which we encode as the conjunction of constraints on the
-right-hand side of the figure:
-%
+As the generation of such constraints is standard and covered by several other works \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}, 
+we will not provide the full algorithm, but rather illustrate it through the example in Figure~\ref{fig:environment-model}.
+Our technique proves the termination of each loop separately, 
+with the exception of nested loops, whose termination we encode in the same formula. For the current example, the termination formula is given on the right side of 
+Figure~\ref{fig:environment-model}:
+%% Following the same design, let us assume that we have already proven the termination of loop $L_1$, and we are now trying to prove that $L_2$ and $L_3$ terminate, which we encode
+%% as the conjunction of constraints on the right side of the figure:
 \begin{itemize}
-%
-\item $W_1$ and $W_2$ are (over-approximating) inductive invariants for $L_1$ and $L_2$, respectively.
-%
-\item $S$ is a summary of $L_3$ that over-approximates the transitive closure of its transition relation.
-%
-\item If the formula is satisfiable, then $R_2$ and $R_3$ are ranking functions for $L_2$ and $L_3$, respectively.
-%
+\item $T_o$  over-approximates $L_1$'s transition relation.
+\item $P$ is a summary of $L_2$ that over-approximates the transitive closure of its transition relation.
+\item If the formula is satisfiable, then $R_1$ and $R_2$ are ranking functions for $L_1$ and $L_2$, respectively.
 \end{itemize}
 
-Note that despite its intricacy, the constraint system is still a formula in
-the synthesis fragment.  Since the outer loop of this program does indeed
-terminate, the system is satisfiable, as witnessed by:
-%
-\begin{align*}
-W_1(\langle v, x, y, z \rangle ) & \triangleq x \leq y + 1 \\
-W_2(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
-S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge \\
-& z' + v' = x' \wedge v' \leq y' \\
-R_2(x) & = x\\
-R_3 & = ..
-\end{align*}
+%% Note that despite its intricacy, the constraint system is still a formula in the synthesis fragment.
+%% Since the outer loop of this program does indeed terminate, the system is satisfiable, as witnessed by:
+%% \begin{align*}
+%% W_1(\langle v, x, y, z \rangle ) & \triangleq x \leq y + 1 \\
+%% W_2(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
+%% S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge \\
+%% & z' + v' = x' \wedge v' \leq y' \\
+%% R_2(x) & = x\\
+%% R_3 & = ..
+%% \end{align*}
 

### 2014-07-08 papers/termination/paper.tex
fix [CT] def
Soundness proofs

--- 2014-07-08 papers/termination/paper.tex
@@
 
+\begin{theorem}
+ \label{thm:snt}
+ Formula {\bf [ONT]} (Definition~\ref{def:ont}) is satisfiable iff {\bf [SNT]} (Definition~\ref{def:snt}) is.
+\end{theorem}
+
 This extra second-order term introduces some complexity to the formula, which
@@
 we will not provide the full algorithm, but rather illustrate it through the example in Figure~\ref{fig:environment-model}.
-Our technique proves the termination of each loop separately, 
-with the exception of nested loops, whose termination we encode in the same formula. For the current example, the termination formula is given on the right side of 
-Figure~\ref{fig:environment-model}:
-%% Following the same design, let us assume that we have already proven the termination of loop $L_1$, and we are now trying to prove that $L_2$ and $L_3$ terminate, which we encode
-%% as the conjunction of constraints on the right side of the figure:
+Our technique proves the termination of each loop separately and in program order, 
+with the exception of nested loops, whose termination we encode in the same formula.
+Following the same design, let us assume that we have already proven the termination of loop $L_1$, and we are now trying to prove that $L_2$ and $L_3$ terminate, which we encode
+as the conjunction of constraints on the right side of the figure:
+%% =======
+%% %
+%% As the generation of such constraints is standard and covered by several
+%% other
+%% works~\cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08},
+%% we will not provide the full algorithm, but rather illustrate it through the
+%% example in Figure~\ref{fig:environment-model}.  Let us assume that we have
+%% already proven the termination of loops $L_1$ and $L_3$, and we are now
+%% trying to prove that $L_2$ terminates, which we encode as the conjunction of
+%% constraints on the right-hand side of the figure:
+%% %
+%% >>>>>>> ee2c21bbc574ab36e7d6739c0e82c7a03d930789
 \begin{itemize}
-\item $T_o$  over-approximates $L_1$'s transition relation.
-\item $P$ is a summary of $L_2$ that over-approximates the transitive closure of its transition relation.
-\item If the formula is satisfiable, then $R_1$ and $R_2$ are ranking functions for $L_1$ and $L_2$, respectively.
+\item $W_1$ and $W_2$ are (over-approximating) inductive invariants for $L_1$ and $L_2$, respectively.
+\item $S$ is a summary of $L_3$ that over-approximates the transitive closure of its transition relation.
+\item If the formula is satisfiable, then $R_2$ and $R_3$ are ranking functions for $L_2$ and $L_3$, respectively.
 \end{itemize}
 
-%% Note that despite its intricacy, the constraint system is still a formula in the synthesis fragment.
-%% Since the outer loop of this program does indeed terminate, the system is satisfiable, as witnessed by:
-%% \begin{align*}
-%% W_1(\langle v, x, y, z \rangle ) & \triangleq x \leq y + 1 \\
-%% W_2(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
-%% S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge \\
-%% & z' + v' = x' \wedge v' \leq y' \\
-%% R_2(x) & = x\\
-%% R_3 & = ..
-%% \end{align*}
+Note that despite its intricacy, the constraint system is still a formula in the synthesis fragment.
+Since the outer loop of this program does indeed terminate, the system is satisfiable, as witnessed by:
+\begin{align*}
+W_1(\langle v, x, y, z \rangle ) & \triangleq x \leq y + 1 \\
+W_2(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
+S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge \\
+& z' + v' = x' \wedge v' \leq y' \\
+R_2(x) & = x\\
+R_3 & = ..
+\end{align*}
 
@@
 \begin{framed}
-\begin{minipage}{0.16\textwidth}
+\begin{minipage}{0.145\textwidth}
 \begin{lstlisting}[mathescape=true]
+x = 0;
 $L_1:$
-while (i<n){
-  j = 0;
+while(x<=y)
+  x++;
 
 $L_2:$
-  while (j$\leq$i){
-    j = j + 1;
+while(x>0){
+  z = x;
+  v = 0;
+
+$L_3:$
+  while(v<y){
+    z--;
+    v++;
   }
 
-  i = i + 1;
+  x -= z;
+  y -= z;
 }
@@
 \vline
-\begin{minipage}{0.84\textwidth}
+\begin{minipage}{0.85\textwidth}
 \begin{align*}
- \exists T_o, P, R_1, R_2 . \forall i, j, n, i', j', n' . \\
-   %% x = 0 & \rightarrow W_1(v, x, y, z) \, \wedge \\
-   %% x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
-   %% x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x, y, z) \, \wedge \\
-   P(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' \leq i' & \rightarrow P(\langle i, j, n \rangle, \langle i', j'+1, n' \rangle) \, \wedge \\
-   & R_2(i,j,n) > 0 \, \wedge \\
-   & R_2(i,j,n) > R_2(i',j'+1,n') \, \wedge\\
-   i < n \wedge T_o(i, j, n) & \rightarrow P(\langle i, j, n \rangle, \langle i, 0, n \rangle) \, \wedge \\
-   i < n \wedge T_o(i, j, n) \wedge P(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' > i' & \rightarrow \\
-   & T_o(i'+1, j', n') \, \wedge \\
-   & R_1(i, j, n) > 0 \, \wedge \\
-   & R_1(i, j, n) > R_1(i'+1, j', n')
+ \exists W_1, W_2, R_2, R_3, S . \forall v, x, y, z, v', x', y', z' . \\
+   x = 0 & \rightarrow W_1(v, x, y, z) \, \wedge \\
+   x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
+   x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x, y, z) \, \wedge \\
+   S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' < y' & \rightarrow S(\langle v, x, y, z \rangle, \langle v'+1, x', y', z'-1 \rangle) \, \wedge \\
+   & R_3(v,x,y,z) > 0 \, \wedge \\
+   & R_3(v,x,y,z) > R_3(v'+1,x',y',z'-1) \, \wedge\\
+   x > 0 \wedge W_2(v, x, y, z) & \rightarrow S(\langle v, x, y, z \rangle, \langle 0, x, y, x \rangle) \, \wedge \\
+   x > 0 \wedge W_2(v, x, y, z) \wedge S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' \geq y' & \rightarrow \\
+   & W_2(v', x' - z', y' - z', z') \, \wedge \\
+   & R_2(v, x, y, z) > 0 \, \wedge \\
+   & R_2(v, x, y, z) > R_2(v', x' - z', y'-z', z')
 \end{align*}
@@
 \end{framed}
-\caption{A program with nested loops and its termination formula\label{fig:environment-model}}
+\caption{A non-trivial program and its termination formula\label{fig:environment-model}}
 \end{figure*}
-
-%% \begin{figure*}
-%% \begin{framed}
-%% \begin{minipage}{0.145\textwidth}
-%% \begin{lstlisting}[mathescape=true]
-%% x = 0;
-%% $L_1:$
-%% while(x<=y)
-%%   x++;
-
-%% $L_2:$
-%% while(x>0){
-%%   z = x;
-%%   v = 0;
-
-%% $L_3:$
-%%   while(v<y){
-%%     z--;
-%%     v++;
-%%   }
-
-%%   x -= z;
-%%   y -= z;
-%% }
-%% \end{lstlisting}
-%% \end{minipage}
-%% \vline
-%% \begin{minipage}{0.85\textwidth}
-%% \begin{align*}
-%%  \exists W_1, W_2, R_2, R_3, S . \forall v, x, y, z, v', x', y', z' . \\
-%%    x = 0 & \rightarrow W_1(v, x, y, z) \, \wedge \\
-%%    x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
-%%    x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x, y, z) \, \wedge \\
-%%    S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' < y' & \rightarrow S(\langle v, x, y, z \rangle, \langle v'+1, x', y', z'-1 \rangle) \, \wedge \\
-%%    & R_3(v,x,y,z) > 0 \, \wedge \\
-%%    & R_3(v,x,y,z) > R_3(v'+1,x',y',z'-1) \, \wedge\\
-%%    x > 0 \wedge W_2(v, x, y, z) & \rightarrow S(\langle v, x, y, z \rangle, \langle 0, x, y, x \rangle) \, \wedge \\
-%%    x > 0 \wedge W_2(v, x, y, z) \wedge S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' \geq y' & \rightarrow \\
-%%    & W_2(v', x' - z', y' - z', z') \, \wedge \\
-%%    & R_2(v, x, y, z) > 0 \, \wedge \\
-%%    & R_2(v, x, y, z) > R_2(v', x' - z', y'-z', z')
-%% \end{align*}
-%% \end{minipage}
-%% \end{framed}
-%% \caption{A non-trivial program and its termination formula\label{fig:environment-model}}
-%% \end{figure*}
 
@@
 \begin{align*}
- \exists P_N . \forall x, x' . \phi(W_N, x, x') \\
- \exists P_T . \forall x, x' . \psi(W_T, x, x')
+ \exists P_T . \forall x, x' . \phi(P_T, x, x') \\
+ \exists P_N . \forall x . \psi(P_N, x)
 \end{align*}
@@
 \begin{align*}
- (\exists P_N . \forall x, x'. \phi(W_N, x, x')) \vee (\exists P_T . \forall x, x', \psi(W_T, x, x'))
+ (\exists P_T . \forall x, x'. \phi(P_T, x, x')) \vee (\exists P_N . \forall x, \psi(P_N, x))
 \end{align*}
 Which simplifies to:
-\begin{definition}[Generalised Termination Formula]
+\begin{definition}[Generalised Termination Formula {\bf [GT]}]
+\label{def:general-termination}
 \begin{align*}
- \exists P_N, P_T. \forall x, x', y, y', \phi(P_N, x, x') \vee \psi(P_T, y, y') 
+ \exists P_T, P_N. \forall x, x', y, \phi(P_T, x, x') \vee \psi(P_N, y)
 \end{align*}
@@
 and termination respectively.  Exactly one of these will be a genuine proof, so we can check
-first one and then the other.
+first one and then the other, as shown in Algorithm~\ref{alg:termination}.
+
+\begin{algorithm}
+ \caption{Bit-vector termination
+ \label{alg:termination}}
+ %\vspace{-1.5em}
+ \begin{algorithmic}[1]
+ \Statex
+\Function{terminates}{$L$}
+  %\Let{termination-spec}{$\exists W, R. \forall x, x' . \phi(W, R, x, x')$} \Comment{Formula {\bf [CT]}}
+  %\Let{nonterm-spec}{$\exists N, C, x_0, \forall x. \psi(N, C, x_0, x)$} \Comment{Formula {\bf [SNT]}}
+  \Statex  \Comment{Formula {\bf[GT]}:}
+  \Let{spec}{$\exists P_T, P_N. \phi(P_T, x, x') \vee \psi(P_N, y)$}
+  \Let{$(P_T, P_N)$}{{\sc Synthesise}(spec)}
+  \If{decide($\exists x, x' . \lnot \phi(P_T, x, x'))$ = unsat}
+    \State \Return{$L$ terminates}
+  \ElsIf{decide($\exists x . \lnot \psi(P_N, x))$ = unsat}
+    \State \Return{$L$ does not terminate}
+  \EndIf
+\EndFunction
+ \end{algorithmic}
+\end{algorithm}
 
@@
  \Statex
-\Function{synth}{{\sc inputs}}
+\Function{synth}{{\sc inputs}, $\sigma$}
   \Let{$(i_1, \ldots, i_N)$}{{\sc inputs}}
@@
 \Statex
-\Function{verif}{P}
+\Function{verif}{$P, x_0, \sigma$}
   \Let{query}{$\exists x . \lnot \sigma(P, x_0, x)$}
@@
 \Statex
-\Function{RefinementLoop}{}
+\Function{Synthesise}{$\sigma$}
   \Let{{\sc inputs}}{$\emptyset$}
@@
 expected results for soundness and completeness of our algorithm for
-deciding it.
+deciding it.  Our strategy for proving soundness and completeness
+is first to prove that Algorithm~\ref{alg:cegis} (as instanciated
+with \newC as a background logic) is a semi-decision
+procedure.  Next, we will show that $\mathcal{L}$ is expressive enough
+to compute (non-)termination proofs for every bit-vector program.
+We will then combine these results to show that the procedure as a whole
+terminates with a valid proof for every input program.
 %
@@
 \begin{proof}
- The procedure {\sc RefinementLoop} terminates only if {\sc Verif} returns ``valid''.  In that
+ The procedure {\sc Synthesise} terminates only if {\sc Verif} returns ``valid''.  In that
  case, $\exists x . \lnot \sigma(P, x_0, x)$ is unsatisfiable and so $\forall x . \sigma(P, x_0, x)$ holds.
@@
  \label{thm:synth-semicomplete}
- If the existential first-order theory used to express the specification $\sigma$ is decidable,
+ If the existential first-order theory used to express the specification $\sigma$ is decidable
+ and the domain of inputs $X$ is finite,
  Algorithm~\ref{alg:cegis} is semi-complete -- if a solution $(P, x_0) \models \sigma$
  exists then Algorithm~\ref{alg:cegis} will terminate.  However, if no program
- satisfies the specification, the algorithm may not terminate.  Furthermore if the domain of inputs $X$
- is finite, the algorithm is guaranteed to terminate.
+ satisfies the specification, the algorithm may not terminate.
 \end{theorem}
@@
 \begin{proof}
+ If the domain $X$ is finite then the loop in procedure {\sc Synthesise} can only
+ iterate $\| X \|$ times, since by this time all of the elements of $X$ would have been
+ added to the inputs set.  Therefore if the {\sc Synth} procedure always terminates,
+ Algorithm~\ref{alg:cegis}.
+
  Since the {\sc ExplicitSearch} routine enumerates all programs (as can be seen by induction on
  the program length $l$), it will eventually enumerate a program that meets the specification
- in whatever set of inputs are currently being tracked.  Since the first-order theory is
+ in whatever set of inputs are currently being tracked, since by assumption such a program
+ exists.  Since the first-order theory is
  decidable, the query in {\sc Verif} will succeed for this program, causing the algorithm to terminate.
@@
  enumerates this set, so it is semi-complete.
-
- If the domain $X$ is finite then the loop in procedure {\sc RefinementLoop} can only
- iterate $\| X \|$ times, since by this time all of the elements of $X$ would have been
- added to the inputs set.  Therefore if the {\sc Synth} procedure always terminates (which it does
- by assumption of decidability), the whole algorithm terminates.
 \end{proof}
@@
 
+We will now show that $\mathcal{L}$ is expressive enough to encode (non-)termination proofs for
+every bit-vector program.
+
 \begin{lemma}
@@
 \begin{proof}
-A very inefficient construction which computes $f$ is:
+Without loss of generality, let $X = Y = \mathbb{N}_b^k$ the set of
+$k$-tuples of natural numbers less than $b$.
+A very inefficient construction which computes the first co-ordinate
+of the output $y$ is:
 %
@@
 %
-Thus, every $f$ is computed a program of size at most $b \times
-2^{bk + 1}$.
-
-(Tidy this up).
+Where the \verb|f(n)| are literal constants that are to appear in the program text.
+This program is of length $2b - 1$, and so all $k$ co-ordinates of the output $y$
+are computed by a program of size at most $2bk - k$.
 \end{proof}
@@
  a (finite) recurrence set and $C : \mathcal{S} \to \mathcal{S}$ is a Skolem function choosing
- a successor for each $x \in N$.  Since $\mathcal{S}$ is finite, and by Lemma~\ref{lem:l-func}, both
+ a successor for each $x \in N$.  Since $\mathcal{S}$ is finite, so by Lemma~\ref{lem:l-func} both
  $N$ and $C$ are computed by finite $\mathcal{L}$-programs.
@@
 \begin{proof}
+ For any $P, P', \sigma, \sigma$, if $P \models \sigma$ then $(P, P') \models \sigma \vee \sigma'$.
+
  By Theorem~\ref{thm:l-term}, if $L$ terminates then there exists a termination proof $P_T$ expressible
- in $\mathcal{L}$.  Since $\phi$ is a termination specification, $P_T \models \phi$ and so $P_T \models \phi \vee \psi$.
+ in $\mathcal{L}$.  Since $\phi$ is a termination specification, $P_T \models \phi$ and so
+ for any $P_N$, $(P_T, P_N) \models \phi \vee \psi$.
 
  Similarly if $L$ does not terminate for some input, by Theorem~\ref{thm:l-nonterm} there is a non-termination
- proof $P_N$ expressible in $\mathcal{L}$, so $P_N \models \psi$ and so $P_N \models \phi \vee \psi$. 
+ proof $P_N$ expressible in $\mathcal{L}$, so $P_N \models \psi$ and hence for any $P_T$, $(P_T, P_N) \models \phi \vee \psi$.
+
+ So in either case ($L$ terminates or does not), there is a witness in $\mathcal{L}$ satisfying $\phi \vee \psi$.
 \end{proof}
@@
 \begin{theorem}
- Algorithm~\ref{alg:cegis} terminates on the generalised termination specification of Definition~\ref{def:general-termination}.
+ Algorithm~\ref{alg:termination} is sound and complete -- it terminates for all input loops $L$ with
+ a correct termination verdict.
 \end{theorem}
@@
 \begin{proof}
- The specification is satisfiable (Thoerem~\ref{thm:generalised-sat}) and Algorithm~\ref{alg:cegis} is semi-complete
- (Theorem~\ref{thm:synth-semicomplete}).
+ The background logic for Algorithm~\ref{alg:termination} is \newC, for which safety is decidable.  Also, the set of
+ inputs is just the state space of $L$, which is finite.  Therefore Theorems~\ref{thm:generalised-sat} and \ref{thm:synth-semicomplete}
+ hold -- the specification \lstinline!spec! is satisfiable and the instanciation of Algorithm~\ref{alg:cegis} is semi-complete.
+ Therefore the call to {\sc Synthesise} terminates with a witness $(P_T, P_N)$.
+ If $L$ terminates then $P_T$ is a termination proof (Theorem~\ref{thm:ct}).  Otherwise $P_N$ is a non-termination proof
+ (Theorem~\ref{thm:snt}).  Exactly one of these cases holds, and the background theory being decidable means that each
+ call to \lstinline|decide()| will terminate, which allows us to determine which of $P_T$ and $P_N$ is a correct proof.
 \end{proof}
@@
 Hence our procedure is sound and complete for bitvector termination.
-
-{\bf TODO:} Tie everything up into one algorithm (build the formula, invoke CEGIS, check which witness is
-correct) then proof soundness for that.
 

### 2014-07-08 papers/termination/paper.tex

--- 2014-07-08 papers/termination/paper.tex
@@
   \exists R, W . \forall x, x' . & I(x) \wedge G(x) \rightarrow W(x) ~ \wedge \\
-                                 & G(x) \wedge W(x) \wedge T_o(x, x') \rightarrow W(x') \wedge R(x) > 0 
+                                 & G(x) \wedge W(x) \wedge T(x, x') \rightarrow W(x') \wedge R(x) > 0 
   \wedge R(x) > R(x')
@@
 \begin{align*}
- (\exists P_N . \forall x, x'. \phi(W_N, x, x')) \vee (\exists P_T . \forall x, x', \psi(W_T, x, x'))
+ (\exists P_T . \forall x, x'. \phi(P_T, x, x')) \vee (\exists P_N . \forall x, \psi(P_N, x))
 \end{align*}
@@
 Which simplifies to:
-%
-\begin{definition}[Generalised Termination Formula]
+\begin{definition}[Generalised Termination Formula {\bf [GT]}]
+\label{def:general-termination}
 \begin{align*}
- \exists P_N, P_T. \forall x, x', y, y', \phi(P_N, x, x') \vee \psi(P_T, y, y') 
+ \exists P_T, P_N. \forall x, x', y, \phi(P_T, x, x') \vee \psi(P_N, y)
 \end{align*}
@@
 
-Since $L$ either terminates or does not terminate, this formula is a
-tautology in the synthesis fragment.  A solution to the formula would
+Since $L$ either terminates or does not terminate, this formula is a tautology in the synthesis fragment.
+A solution to the formula would include witnesses $P_N$ and $P_T$, which are putative proofs of non-termination
+and termination respectively.  Exactly one of these will be a genuine proof, so we can check
+first one and then the other, as shown in Algorithm~\ref{alg:termination}.
+
+\begin{algorithm}
+ \caption{Bit-vector termination
+ \label{alg:termination}}
+ %\vspace{-1.5em}
+ \begin{algorithmic}[1]
+ \Statex
+\Function{terminates}{$L$}
+  %\Let{termination-spec}{$\exists W, R. \forall x, x' . \phi(W, R, x, x')$} \Comment{Formula {\bf [CT]}}
+  %\Let{nonterm-spec}{$\exists N, C, x_0, \forall x. \psi(N, C, x_0, x)$} \Comment{Formula {\bf [SNT]}}
+  \Statex  \Comment{Formula {\bf[GT]}:}
+  \Let{spec}{$\exists P_T, P_N. \phi(P_T, x, x') \vee \psi(P_N, y)$}
+  \Let{$(P_T, P_N)$}{{\sc Synthesise}(spec)}
+  \If{decide($\exists x, x' . \lnot \phi(P_T, x, x'))$ = unsat}
+    \State \Return{$L$ terminates}
+  \ElsIf{decide($\exists x . \lnot \psi(P_N, x))$ = unsat}
+    \State \Return{$L$ does not terminate}
+  \EndIf
+\EndFunction
+ \end{algorithmic}
+\end{algorithm}
+
+Since $L$ either terminates or does not terminate, this formula is
+satisfiable in the synthesis fragment.  A solution to the formula would
 include witnesses $P_N$ and $P_T$, which are putative proofs of

### 2014-07-08 papers/termination/paper.tex
underapp --> under-app && overapp --> over-app

--- 2014-07-08 papers/termination/paper.tex
@@
 include the transitive closure of the inner loop $T'^*$, and we do not have a transitive closure
-operator in our logic.  Therefore to encode the termination of $L$, we construct an overapproximation
+operator in our logic.  Therefore to encode the termination of $L$, we construct an over-approximation
 $T_o \supseteq T$ and use this in formula {\bf [UT]} to specify a ranking function.
 Rather than explicitly construct $T_o$ using, for example, abstract interpretation, we add constraints to
-our formula that encode the fact that $T_o$ is an overapproximation of $T$, and that it is
+our formula that encode the fact that $T_o$ is an over-approximation of $T$, and that it is
 precise enough to show that $R$ is a ranking function.

### 2014-07-08 papers/termination/paper.tex
P --> S for nested loops
Time bound

--- 2014-07-08 papers/termination/paper.tex
@@
 
+\begin{corollary}
+\label{cor:cegis-newc}
+ Since safety of \newC programs is decidable, Algorithm~\ref{alg:cegis} is semi-complete
+ when instanciated with \newC as a background theory.
+\end{corollary}
+
 % \begin{theorem}
@@
  By Theorem~\ref{thm:l-term}, if $L$ terminates then there exists a termination proof $P_T$ expressible
- in $\mathcal{L}$.  Since $\phi$ is a termination specification, $P_T \models \phi$ and so
+ in $\mathcal{L}$.  Since $\phi$ is an instance of {\bf [CT]}, $P_T \models \phi$ (Theorem~\ref{thm:ct}) and
  for any $P_N$, $(P_T, P_N) \models \phi \vee \psi$.
@@
  Similarly if $L$ does not terminate for some input, by Theorem~\ref{thm:l-nonterm} there is a non-termination
- proof $P_N$ expressible in $\mathcal{L}$, so $P_N \models \psi$ and hence for any $P_T$, $(P_T, P_N) \models \phi \vee \psi$.
-
- So in either case ($L$ terminates or does not), there is a witness in $\mathcal{L}$ satisfying $\phi \vee \psi$.
+ proof $P_N$ expressible in $\mathcal{L}$.  $\psi$ is an instance of {\bf [SNT]} and so $P_N \models \psi$
+ (Theorem~\ref{thm:snt}), hence for any $P_T$, $(P_T, P_N) \models \phi \vee \psi$.
+
+ So in either case ($L$ terminates or does not), there is a witness in $\mathcal{L}$ satisfying
+ formula {\bf [GT]}: $\phi \vee \psi$.
 \end{proof}
@@
 \begin{proof}
- The background logic for Algorithm~\ref{alg:termination} is \newC, for which safety is decidable.  Also, the set of
- inputs is just the state space of $L$, which is finite.  Therefore Theorems~\ref{thm:generalised-sat} and \ref{thm:synth-semicomplete}
- hold -- the specification \lstinline!spec! is satisfiable and the instanciation of Algorithm~\ref{alg:cegis} is semi-complete.
- Therefore the call to {\sc Synthesise} terminates with a witness $(P_T, P_N)$.
+ By Theorem~\ref{thm:generalised-sat}, the specification \lstinline!spec! is satisfiable.  Corollary~\ref{cor:cegis-newc}
+ shows that Algorithm~\ref{alg:cegis} is semi-complete, and so the call to {\sc Synthesise} terminates with a witness $(P_T, P_N)$.
  If $L$ terminates then $P_T$ is a termination proof (Theorem~\ref{thm:ct}).  Otherwise $P_N$ is a non-termination proof
@@
 
-Hence our procedure is sound and complete for bitvector termination.
+Hence Algorithm~\ref{alg:termination} is sound and complete for bit-vector termination.
+
+Theorem~\ref{thm:synth-semicomplete} provides a time bound -- the CEGIS loop can iterate at most
+$2^bk$ times and each iteration requires solving 2 NP-complete problems of size < $2^bk \times \| L\|$
+with $\| L \| $ the size of the program-under-analysis computing $\| L \|$.  So the
+runtime is bounded by $2^{bk+1} NP(2^bk \| L \|)$.
 

### 2014-07-08 papers/termination/paper.tex
skolem fct S --> C

--- 2014-07-08 papers/termination/paper.tex
@@
  \begin{align*}
-  \exists N, S, x_0 . \forall x . & N(x_0) ~\wedge \\ &  N(x) \rightarrow G(x) ~ \wedge \\
-							& N(x) \rightarrow T(x, S(x)) \wedge N(S(x))
+  \exists N, C, x_0 . \forall x . & N(x_0) ~\wedge \\ &  N(x) \rightarrow G(x) ~ \wedge \\
+							& N(x) \rightarrow T(x, C(x)) \wedge N(C(x))
  \end{align*}
@@
 (it is an $\exists \forall \exists$ formula).  To eliminate the innermost existential quantifier,
-we introduce a Skolem function $S$, which we then existentially quantify over.  This results in
+we introduce a Skolem function $C$, which we then existentially quantify over.  This results in
 formula {\bf [SNT]} (Definition~\ref{def:snt}).
@@
 \item $T_o$  over-approximates $L_1$'s transition relation.
-\item $P$ is a summary of $L_2$ that over-approximates the transitive closure of its transition relation.
+\item $S$ is a summary of $L_2$ that over-approximates the transitive closure of its transition relation.
 \item If the formula is satisfiable, then $R_1$ and $R_2$ are ranking functions for $L_1$ and $L_2$, respectively.
@@
 \begin{align*}
- \exists T_o, P, R_1, R_2 . \forall i, j, n, i', j', n' . \\
+ \exists T_o, S, R_1, R_2 . \forall i, j, n, i', j', n' . \\
    %% x = 0 & \rightarrow W_1(v, x, y, z) \, \wedge \\
@@
    %% x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x, y, z) \, \wedge \\
-   P(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' \leq i' & \rightarrow P(\langle i, j, n \rangle, \langle i', j'+1, n' \rangle) \, \wedge \\
+   S(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' \leq i' & \rightarrow S(\langle i, j, n \rangle, \langle i', j'+1, n' \rangle) \, \wedge \\
    & R_2(i,j,n) > 0 \, \wedge \\
    & R_2(i,j,n) > R_2(i',j'+1,n') \, \wedge\\
-   i < n \wedge T_o(i, j, n) & \rightarrow P(\langle i, j, n \rangle, \langle i, 0, n \rangle) \, \wedge \\
-   i < n \wedge T_o(i, j, n) \wedge P(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' > i' & \rightarrow \\
+   i < n \wedge T_o(i, j, n) & \rightarrow S(\langle i, j, n \rangle, \langle i, 0, n \rangle) \, \wedge \\
+   i < n \wedge T_o(i, j, n) \wedge S(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' > i' & \rightarrow \\
    & T_o(i'+1, j', n') \, \wedge \\

### 2014-07-08 papers/termination/paper.tex
skolem fct wording
Time bound

@@
 Theorem~\ref{thm:synth-semicomplete} provides a time bound -- the CEGIS loop can iterate at most
-$2^bk$ times and each iteration requires solving 2 NP-complete problems of size < $2^bk \times \| L\|$
-with $\| L \| $ the size of the program-under-analysis computing $\| L \|$.  So the
-runtime is bounded by $2^{bk+1} NP(2^bk \| L \|)$.
+$2^bk$ times where $k$ is the number of variables used by $L$ and
+$b$ is the number of bits used to store each variable.
+The $i$th iteration of the loop requires solving 2 NP-complete problems of size $i \times \| L\|$
+with $\| L \| $ the size of the program-under-analysis computing $\| L \|$.  So if 
+the CEGIS loop iterates $n$ times, the
+runtime is bounded by $2 \times n \times NP(n \times \| L \|)$, which is bounded
+by $2^{bk + 1} NP(2^{bk} \| L \|)$.  Clearly the runtime is affected most strongly by $n$,
+the total number of iterations of the CEGIS loop.
 

### 2014-07-08 papers/termination/paper.tex
theorem 14

--- 2014-07-08 papers/termination/paper.tex
@@
 (it is an $\exists \forall \exists$ formula).  To eliminate the innermost existential quantifier,
-we introduce a Skolem function $S$, which we then existentially quantify over.  This results in
+we introduce a Skolem function $C$ that chooses the successor $x'$, which we then existentially quantify over.  This results in
 formula {\bf [SNT]} (Definition~\ref{def:snt}).
@@
  If $T$ is deterministic,
- Definition~\ref{def:ont} and Definition~\ref{def:cnt} are equisatisfiable.
+formula {\bf [ONT]} (Definition~\ref{def:ont}) and formula {\bf [CNT]} (Definition~\ref{def:cnt}) are equisatisfiable.
 \end{theorem}

### 2014-07-08 papers/termination/paper.tex
theorem 12
wording [ONT]

@@
 introduce such an operator, we will characterise non-termination
-using a synthesis formula.
-
-Gupta et al.~\cite{DBLP:conf/popl/GuptaHMRX08} characterise non-termination
-of a transition relation $T$ by the existence of an \emph{(open) recurrence
-set}, i.e.~a nonempty set of states $N$ such that for each $s \in N$ there
-exists a transition to some $s' \in N$.
-The notion of open recurrence set is encoded by formula {\bf [ONT]} (Definition~\ref{def:ont}).
-% Dual to the termination problem, when formulating $L$'s non-termination, we only need to consider an under-approximation of its transition
-% relation~$T_u$. 
-% Again, this distinction is mostly important when dealing with nested loops,
-% as illustrated in Section~\ref{sec:env}.
+using the synthesis formula {\bf [ONT]} (Definition~\ref{def:ont})
+encoding the existence of an \emph{(open) recurrence set}, i.e.~a nonempty 
+set of states $N$ such that for each $s \in N$ there
+exists a transition to some $s' \in N$ \cite{DBLP:conf/popl/GuptaHMRX08}.
+%% Gupta et al.~\cite{DBLP:conf/popl/GuptaHMRX08} characterise non-termination
+%% of a transition relation $T$ by the existence of an \emph{(open) recurrence
+%% set}, i.e.~a nonempty set of states $N$ such that for each $s \in N$ there
+%% exists a transition to some $s' \in N$.
+%% The notion of open recurrence set is encoded by formula {\bf [ONT]} (Definition~\ref{def:ont}).
 
@@
  \label{thm:snt}
- Formula {\bf [ONT]} (Definition~\ref{def:ont}) is satisfiable iff {\bf [SNT]} (Definition~\ref{def:snt}) is.
+ Formula {\bf [ONT]} (Definition~\ref{def:ont}) and formula {\bf [SNT]} (Definition~\ref{def:snt}) are equisatisfiable.
 \end{theorem}

### 2014-07-08 papers/termination/paper.tex
typo
wording

@@
 \item Program (c) breaks assumption (iii).  This loop is terminating for
-bit-vectors since $x$ will eventually over-flow and become negative. 
+bit-vectors since $x$ will eventually overflow and become negative. 
 Conversely, the same program is non-terminating using integer arithmetic
@@
 If a pair $(\vec{P}, \vec{x})$ is a solution to the synthesis problem, then we write $(\vec{P}, \vec{x}) \models \sigma$.
-For the remainder of the presentation, we drop the $\vec{x}$ and instead just write $x$, with the understanding
+For the remainder of the presentation, we drop the vector notation and write $x$ for $\vec{x}$, with the understanding
 that all quantified variables range over vectors.

### 2014-07-08 papers/termination/paper.tex
typos

@@
 $k$-tuples of natural numbers less than $b$.
-A very inefficient construction which computes the first co-ordinate
+A very inefficient construction which computes the first coordinate
 of the output $y$ is:
@@
 
-Given the lack of research on termination of not-linear programs, as well as
+Given the lack of research on termination of nonlinear programs, as well as
 programs over bit vectors and floats, our work focused on covering these

### 2014-07-08 papers/termination/paper.tex
Kolmogorov complexity

--- 2014-07-08 papers/termination/paper.tex
@@
 introduce such an operator, we will characterise non-termination
-using the synthesis formula {\bf [ONT]} (Definition~\ref{def:ont})
-encoding the existence of an \emph{(open) recurrence set}, i.e.~a nonempty 
-set of states $N$ such that for each $s \in N$ there
-exists a transition to some $s' \in N$ \cite{DBLP:conf/popl/GuptaHMRX08}.
-%% Gupta et al.~\cite{DBLP:conf/popl/GuptaHMRX08} characterise non-termination
-%% of a transition relation $T$ by the existence of an \emph{(open) recurrence
-%% set}, i.e.~a nonempty set of states $N$ such that for each $s \in N$ there
-%% exists a transition to some $s' \in N$.
-%% The notion of open recurrence set is encoded by formula {\bf [ONT]} (Definition~\ref{def:ont}).
+using a synthesis formula.
+
+Gupta et al.~\cite{DBLP:conf/popl/GuptaHMRX08} characterise non-termination
+of a transition relation $T$ by the existence of an \emph{(open) recurrence
+set}, i.e.~a nonempty set of states $N$ such that for each $s \in N$ there
+exists a transition to some $s' \in N$.
+The notion of open recurrence set is encoded by formula {\bf [ONT]} (Definition~\ref{def:ont}).
+% Dual to the termination problem, when formulating $L$'s non-termination, we only need to consider an under-approximation of its transition
+% relation~$T_u$. 
+% Again, this distinction is mostly important when dealing with nested loops,
+% as illustrated in Section~\ref{sec:env}.
 
@@
  \label{thm:snt}
- Formula {\bf [ONT]} (Definition~\ref{def:ont}) and formula {\bf [SNT]} (Definition~\ref{def:snt}) are equisatisfiable.
+ Formula {\bf [ONT]} (Definition~\ref{def:ont}) is satisfiable iff {\bf [SNT]} (Definition~\ref{def:snt}) is.
 \end{theorem}
@@
  If $T$ is deterministic,
-formula {\bf [ONT]} (Definition~\ref{def:ont}) and formula {\bf [CNT]} (Definition~\ref{def:cnt}) are equisatisfiable.
+ Definition~\ref{def:ont} and Definition~\ref{def:cnt} are equisatisfiable.
 \end{theorem}
@@
 size $2^{bk}$.  A~ranking function $R: \mathcal{S} \to \mathcal{D}$ for $P$
-exists iff $P$ terminates.  Since $R$ is injective, we have that $\|
+exists iff $P$ terminates.  Without loss of generality,
+$\mathcal{D}$ is a well-founded total order.  Since $R$ is injective, we have that $\|
 \mathcal{D} \| \geq \| \mathcal{S} \|$.  If $\| \mathcal{D} \| > \|
@@
 $b$ is the number of bits used to store each variable.
-The $i$th iteration of the loop requires solving 2 NP-complete problems of size $i \times \| L\|$
-with $\| L \| $ the size of the program-under-analysis computing $\| L \|$.  So if 
+The $i$th iteration of the loop requires solving 2 NP-complete problems of size $i \times l \times \| L\|$
+with $l$ the length of the program to be synthesised, and $\| L \| $ the size of the program-under-analysis
+computing $ L $.  So if
 the CEGIS loop iterates $n$ times, the
-runtime is bounded by $2 \times n \times NP(n \times \| L \|)$, which is bounded
-by $2^{bk + 1} NP(2^{bk} \| L \|)$.  Clearly the runtime is affected most strongly by $n$,
+runtime is bounded by $2 \times n \times NP(n \times l \times \| L \|)$, which is bounded
+by $2^{bk + 1} NP(2^{bk} \times l \times \| L \|)$.  Clearly the runtime is affected most strongly by $n$,
 the total number of iterations of the CEGIS loop.
 
-We will now show that our time complexity is asymptotically a function of
-the Kolmogorov complexity of the (non-)termination proof, and argue
-that this gives our procedure various desirable qualities in practical
+We will now show that the number of iterations of the CEGIS loop is a function
+of the Kolmogorov complexity of the (non-)termination proof, and so
+our runtime is asymptotically a function of the Kolmogorov complexity.
+We argue that this gives our procedure various desirable qualities in practical
 appliations.
-%
-\begin{theorem}
- {\sc Kalashnikov} has time complexity $NP^{NP}(K(\sigma))$.
-\end{theorem}
-
-\begin{theorem}
- {\sc Headshot} uses a single call to {\sc Kalashnikov} and uses an encoding with $O(1)$
- overhead.  It therefore has time complexity $NP^{NP}(K(\tau(P))$ where $\tau(P)$ is the
- termination specification for program $P$.
-\end{theorem}
-
-In Section~\ref{sec:synthesis} we will describe a procedure for solving the synthesis formula.
-This procedure has the property that it will find the shortest program meeting the
-specification, and as a consequence the runtime of the program synthesiser is dominated
-by the size of this shortest program.  To frame the discussion of the runtime of our
-termination proving procedure, we first recall the definition of the Kolmogorov complexity
+We first recall the definition of the Kolmogorov complexity
 of a function $f$:
@@
 
-We will later show that the time complexity of our synthesis procedure is $NP^{NP}(K(\sigma))$,
-which means that for even moderately large $K(\sigma)$ our procedure is intractably slow.  Conversely
+Let us consider the number of iterations of the CEGIS loop $n$ required for a specification
+$\sigma$.  Since we enumerate candidate programs in order of length, we are always synthesising
+programs with length no greater than $K(\sigma)$ (since when we enumerate the first correct program,
+we will terminate).  So it is clear that the space of solutions we search over is the space
+of functions computed by $\mathcal{L}$-programs of length no greater than $K(\sigma)$.  Let's
+denote this set $\mathcal{L}(K(\sigma))$.  Clearly, $\| \mathcal{L}(K(\sigma)) \| \leq O(2^{K(\sigma)})$
+since there are $O(2^{K(\sigma)}$ \emph{programs} of length $K(\sigma)$ and some functions
+will be computed by more than one program.
+
+Each iteration of the CEGIS loop distinguishes at least one incorrect function from the set of correct
+functions, so we can see that the loop will iterate no more than $\| \mathcal{L}(K(\sigma)) \|$ times.
+Therefore another bound on our runtime is:
+$$O\left (2^{K(\sigma)} \times NP \left ( K(\sigma) \times \| L \| \right) \right) = O \left(2^{K(\sigma)} \right)$$
+Which means that for even moderately large $K(\sigma)$ our procedure is intractably slow.  Conversely
 if $K(\sigma)$ is small we will be able to find a termination proof, regardless of other parameters

### 2014-07-08 papers/termination/paper.tex
Kolmogorov

@@
 such as the size of the program under analysis.  To put this another way, under the assumption
-that $K(\sigma)$ is small for a program's termination problem, we will find a ranking function
-quite rapidly.  We will now investigate the properties of functions with low Kolmogorov complexity,
-and show that the assumption of a low-Kolmogorov ranking function is much weaker than the assumption
-of a linear ranking function.
-
-\begin{theorem}
- Linear functions have low Kolmogorov complexity.
-\end{theorem}
-
-\begin{proof}
- The function $f: X \to Y$ can be computed with a program consisting of
- $2 \cdot \dim(X) \cdot \dim(Y) - \dim(Y)$ instructions (1 multiplication and 1 addition for
- each cell in the matrix representing $f$).  Therefore,
- $K(f) \leq 2 \cdot \dim(X) \cdot \dim(Y) - \dim(Y)$.
-\end{proof}
-
-\begin{theorem}
-Assuming a reasonable program encoding, the probability that a random program of length $l$ computes
-a linear function is $O(2^{-l})$.
-\end{theorem}
-
-\begin{proof}
- (Do this proof properly).
-
- Otherwise, we'd need to have an exponential number of programs computing the same function.
- That would be an unreasonable program encoding.
-\end{proof}
+that $K(\sigma)$ is small for a program's termination problem, we will find a termination proof
+quite rapidly.
+
+To give a flavour for the nature of the ``low Kolmogorov complexity'' assumption, 
+we observe that linear functions have low Kolmogorov complexity (since a linear function can
+be computed with at most 1 instruction and 1 multiplication for each program variable).
+Conversely, the probability that a random program of length $l$ computes
+a linear function is $O(2^{-l})$, since otherwise we'd need to have an exponential
+number of programs computing the same function.
 

### 2014-07-08 papers/termination/paper.tex
Kolmogorov

@@
 we observe that linear functions have low Kolmogorov complexity (since a linear function can
-be computed with at most 1 instruction and 1 multiplication for each program variable).
-Conversely, the probability that a random program of length $l$ computes
-a linear function is $O(2^{-l})$, since otherwise we'd need to have an exponential
-number of programs computing the same function.
-
-Therefore the number of nonlinear functions $f$ with $K(f) \leq \kappa$
-grows exponentially with $\kappa$, while the number of linear functions
-grows only linearly.  This shows that assuming the existence of a linear
+be computed with at most 1 instruction and 1 multiplication for each program variable),
+so the number of linear functions $f$ with $K(f) \leq \kappa$ grows linearly with $\kappa$.
+Conversely, the number of functions $f$ with $K(f) \leq \kappa$
+grows exponentially with $\kappa$.  This shows that assuming the existence of a linear
 ranking function is much stronger than assuming the existence of a short

### 2014-07-08 papers/termination/paper.tex
bitvector -> bit-vector

@@
 Termination, Non-Termination, Program Synthesis,
-Lexicographic Ranking Functions, Bitvector Ranking Functions, Floating-Point Ranking Functions.
+Lexicographic Ranking Functions, Bit-vector Ranking Functions, Floating-Point Ranking Functions.
 

### 2014-07-08 papers/termination/paper.tex
added captions and references to the figures with definitions

--- 2014-07-08 papers/termination/paper.tex
@@
 \end{framed}
+\caption{Formulae encoding the termination and non-termination of a single loop} \label{fig:single_loop}
 \end{figure*}
@@
 \label{thm:ut}
- The loop $L(G, T)$ terminates from every start state iff formula {\bf [UT]} (Definition~\ref{def:UT}) is satisfiable.
+ The loop $L(G, T)$ terminates from every start state iff formula {\bf [UT]} (Definition~\ref{def:UT}, Figure~\ref{fig:single_loop}) is satisfiable.
 \end{theorem}
@@
 introduce such an operator, we will characterise non-termination
-using the synthesis formula {\bf [ONT]} (Definition~\ref{def:ont})
+using the synthesis formula {\bf [ONT]} (Definition~\ref{def:ont}, Figure~\ref{fig:single_loop})
 encoding the existence of an \emph{(open) recurrence set}, i.e.~a nonempty 
@@
 we introduce a Skolem function $C$ that chooses the successor $x'$, which we then existentially quantify over.  This results in
-formula {\bf [SNT]} (Definition~\ref{def:snt}).
+formula {\bf [SNT]} (Definition~\ref{def:snt}, Figure~\ref{fig:single_loop}).
 
@@
 the non-termination problem for deterministic transition systems is
-equivalent to the satisfiability of formula {\bf [CNT]}.
+equivalent to the satisfiability of formula {\bf [CNT]} from Figure~\ref{fig:single_loop}.
 
@@
  \end{framed}
-
+\caption{Formula encoding conditional termination of a loop} \label{fig:conditional_termination}
 \end{figure*}
@@
 
-\caption{The non-termination formula for nested loops \label{fig:nonterm-nested}}
+\caption{Formula encoding non-termination of nested loops \label{fig:nonterm-nested}}
 \end{figure*}
@@
  The loop $L(G, T)$ terminates when started in any state satisfying $I(x)$ iff formula {\bf [CT]}
- (Defintion~\ref{def:ct}) is satisfiable.
+ (Defintion~\ref{def:ct}, Figure~\ref{fig:conditional_termination}) is satisfiable.
 \end{theorem}
@@
 size $2^{bk}$.  A~ranking function $R: \mathcal{S} \to \mathcal{D}$ for $P$
-exists iff $P$ terminates.  Without loss of generality,
-$\mathcal{D}$ is a well-founded total order.  Since $R$ is injective, we have that $\|
+exists iff $P$ terminates.  Since $R$ is injective, we have that $\|
 \mathcal{D} \| \geq \| \mathcal{S} \|$.  If $\| \mathcal{D} \| > \|
@@
 $b$ is the number of bits used to store each variable.
-The $i$th iteration of the loop requires solving 2 NP-complete problems of size $i \times l \times \| L\|$
-with $l$ the length of the program to be synthesised, and $\| L \| $ the size of the program-under-analysis
-computing $ L $.  So if
+The $i$th iteration of the loop requires solving 2 NP-complete problems of size $i \times \| L\|$
+with $\| L \| $ the size of the program-under-analysis computing $\| L \|$.  So if 
 the CEGIS loop iterates $n$ times, the
-runtime is bounded by $2 \times n \times NP(n \times l \times \| L \|)$, which is bounded
-by $2^{bk + 1} NP(2^{bk} \times l \times \| L \|)$.  Clearly the runtime is affected most strongly by $n$,
+runtime is bounded by $2 \times n \times NP(n \times \| L \|)$, which is bounded
+by $2^{bk + 1} NP(2^{bk} \| L \|)$.  Clearly the runtime is affected most strongly by $n$,
 the total number of iterations of the CEGIS loop.
 
-We will now show that the number of iterations of the CEGIS loop is a function
-of the Kolmogorov complexity of the (non-)termination proof, and so
-our runtime is asymptotically a function of the Kolmogorov complexity.
-We argue that this gives our procedure various desirable qualities in practical
+We will now show that our time complexity is asymptotically a function of
+the Kolmogorov complexity of the (non-)termination proof, and argue
+that this gives our procedure various desirable qualities in practical
 appliations.
-We first recall the definition of the Kolmogorov complexity
+%
+\begin{theorem}
+ {\sc Kalashnikov} has time complexity $NP^{NP}(K(\sigma))$.
+\end{theorem}
+
+\begin{theorem}
+ {\sc Headshot} uses a single call to {\sc Kalashnikov} and uses an encoding with $O(1)$
+ overhead.  It therefore has time complexity $NP^{NP}(K(\tau(P))$ where $\tau(P)$ is the
+ termination specification for program $P$.
+\end{theorem}
+
+In Section~\ref{sec:synthesis} we will describe a procedure for solving the synthesis formula.
+This procedure has the property that it will find the shortest program meeting the
+specification, and as a consequence the runtime of the program synthesiser is dominated
+by the size of this shortest program.  To frame the discussion of the runtime of our
+termination proving procedure, we first recall the definition of the Kolmogorov complexity
 of a function $f$:
@@
 
-Let us consider the number of iterations of the CEGIS loop $n$ required for a specification
-$\sigma$.  Since we enumerate candidate programs in order of length, we are always synthesising
-programs with length no greater than $K(\sigma)$ (since when we enumerate the first correct program,
-we will terminate).  So it is clear that the space of solutions we search over is the space
-of functions computed by $\mathcal{L}$-programs of length no greater than $K(\sigma)$.  Let's
-denote this set $\mathcal{L}(K(\sigma))$.  Clearly, $\| \mathcal{L}(K(\sigma)) \| \leq O(2^{K(\sigma)})$
-since there are $O(2^{K(\sigma)}$ \emph{programs} of length $K(\sigma)$ and some functions
-will be computed by more than one program.
-
-Each iteration of the CEGIS loop distinguishes at least one incorrect function from the set of correct
-functions, so we can see that the loop will iterate no more than $\| \mathcal{L}(K(\sigma)) \|$ times.
-Therefore another bound on our runtime is:
-$$O\left (2^{K(\sigma)} \times NP \left ( K(\sigma) \times \| L \| \right) \right) = O \left(2^{K(\sigma)} \right)$$
-Which means that for even moderately large $K(\sigma)$ our procedure is intractably slow.  Conversely
+We will later show that the time complexity of our synthesis procedure is $NP^{NP}(K(\sigma))$,
+which means that for even moderately large $K(\sigma)$ our procedure is intractably slow.  Conversely
 if $K(\sigma)$ is small we will be able to find a termination proof, regardless of other parameters
 such as the size of the program under analysis.  To put this another way, under the assumption
-that $K(\sigma)$ is small for a program's termination problem, we will find a termination proof
-quite rapidly.
-
-To give a flavour for the nature of the ``low Kolmogorov complexity'' assumption, 
-we observe that linear functions have low Kolmogorov complexity (since a linear function can
-be computed with at most 1 instruction and 1 multiplication for each program variable).
-Conversely, the probability that a random program of length $l$ computes
-a linear function is $O(2^{-l})$, since otherwise we'd need to have an exponential
-number of programs computing the same function.
+that $K(\sigma)$ is small for a program's termination problem, we will find a ranking function
+quite rapidly.  We will now investigate the properties of functions with low Kolmogorov complexity,
+and show that the assumption of a low-Kolmogorov ranking function is much weaker than the assumption
+of a linear ranking function.
+
+\begin{theorem}
+ Linear functions have low Kolmogorov complexity.
+\end{theorem}
+
+\begin{proof}
+ The function $f: X \to Y$ can be computed with a program consisting of
+ $2 \cdot \dim(X) \cdot \dim(Y) - \dim(Y)$ instructions (1 multiplication and 1 addition for
+ each cell in the matrix representing $f$).  Therefore,
+ $K(f) \leq 2 \cdot \dim(X) \cdot \dim(Y) - \dim(Y)$.
+\end{proof}
+
+\begin{theorem}
+Assuming a reasonable program encoding, the probability that a random program of length $l$ computes
+a linear function is $O(2^{-l})$.
+\end{theorem}
+
+\begin{proof}
+ (Do this proof properly).
+
+ Otherwise, we'd need to have an exponential number of programs computing the same function.
+ That would be an unreasonable program encoding.
+\end{proof}
 

### 2014-07-08 papers/termination/paper.tex
Soundness edits

--- 2014-07-08 papers/termination/paper.tex
@@
  the program length $l$), it will eventually enumerate a program that meets the specification
- in whatever set of inputs are currently being tracked, since by assumption such a program
+ on whatever set of inputs are currently being tracked, since by assumption such a program
  exists.  Since the first-order theory is
@@
  Every finite subset $A \subseteq B$ is computable by a finite $\mathcal{L}$-program
- by setting $X = B, Y = 2$ in Lemma~\ref{lem:l-func}.
+ by setting $X = B, Y = 2$ in Lemma~\ref{lem:l-func} and taking the
+ resulting function to be the characteristic function of $A$.
 \end{corollary}
@@
 \mathcal{S} \|$, we can construct a function $R': \mathcal{S} \to
-\mathcal{D'}$ with $ \| \mathcal{D'} = \| \mathcal{S} \|$ by just setting
+\mathcal{D'}$ with $ \| \mathcal{D'} \| = \| \mathcal{S} \|$ by just setting
 $R' = R|_\mathcal{S}$, i.e.~$R'$ is just the restriction of $R$ to
@@
 \begin{proof}
- A non-termination proof is a triple $\langle N, C, x_0 \rangle$ where $N \subseteq \mathcal{S}$ is
- a (finite) recurrence set and $C : \mathcal{S} \to \mathcal{S}$ is a Skolem function choosing
- a successor for each $x \in N$.  Since $\mathcal{S}$ is finite, so by Lemma~\ref{lem:l-func} both
- $N$ and $C$ are computed by finite $\mathcal{L}$-programs.
+ A non-termination proof is a triple $\langle N, C, x_0 \rangle$ where
+ \mbox{$N \subseteq \mathcal{S}$} is a (finite) recurrence set and
+ $C : \mathcal{S} \to \mathcal{S}$ is a Skolem function choosing
+ a successor for each $x \in N$.  $\mathcal{S}$ is finite, so by Lemma~\ref{lem:l-func} both
+ $N$ and $C$ are computed by finite $\mathcal{L}$-programs and $x_0$ is just a ground term.
 \end{proof}
@@
  So in either case ($L$ terminates or does not), there is a witness in $\mathcal{L}$ satisfying
- formula {\bf [GT]}: $\phi \vee \psi$.
+ $\phi \vee \psi$, which is an instance of {\bf [GT]}.
 \end{proof}

### 2014-07-08 papers/termination/paper.tex
Edits

@@
  \label{thm:generalised-sat}
- The generalised termination formula for loop $L$ (Definiton~\ref{def:general-termination}) is satisfiable
+ The generalised termination formula {\bf [GT]} for any loop $L$ is satisfiable
  when $P_N$ and $P_T$ range over \mbox{$\mathcal{L}$-computable} functions.
@@
  If $L$ terminates then $P_T$ is a termination proof (Theorem~\ref{thm:ct}).  Otherwise $P_N$ is a non-termination proof
- (Theorem~\ref{thm:snt}).  Exactly one of these cases holds, and the background theory being decidable means that each
+ (Theorem~\ref{thm:snt}).  Exactly one of these cases holds, and the background theory \newC being decidable means that each
  call to \lstinline|decide()| will terminate, which allows us to determine which of $P_T$ and $P_N$ is a correct proof.
@@
 Theorem~\ref{thm:synth-semicomplete} provides a time bound -- the CEGIS loop can iterate at most
-$2^bk$ times where $k$ is the number of variables used by $L$ and
+$2^{bk}$ times where $k$ is the number of variables used by $L$ and
 $b$ is the number of bits used to store each variable.
@@
 the CEGIS loop iterates $n$ times, the
-runtime is bounded by $2 \times n \times NP(n \times l \times \| L \|)$, which is bounded
-by $2^{bk + 1} NP(2^{bk} \times l \times \| L \|)$.  Clearly the runtime is affected most strongly by $n$,
-the total number of iterations of the CEGIS loop.
+runtime is bounded by
+$$2 \times n \times NP(n \times l \times \| L \|) \leq 2^{bk + 1} NP(2^{bk} \times l \times \| L \|)$$
+Clearly the runtime is affected most strongly by $n$, the total number of iterations of the CEGIS loop.
 
@@
 programs with length no greater than $K(\sigma)$ (since when we enumerate the first correct program,
-we will terminate).  So it is clear that the space of solutions we search over is the space
+we will terminate).  So the space of solutions we search over is the space
 of functions computed by $\mathcal{L}$-programs of length no greater than $K(\sigma)$.  Let's
-denote this set $\mathcal{L}(K(\sigma))$.  Clearly, $\| \mathcal{L}(K(\sigma)) \| \leq O(2^{K(\sigma)})$
-since there are $O(2^{K(\sigma)}$ \emph{programs} of length $K(\sigma)$ and some functions
-will be computed by more than one program.
+denote this set $\mathcal{L}(K(\sigma))$.
+Since there are $O(2^{K(\sigma)}$ \emph{programs} of length $K(\sigma)$ and some functions
+will be computed by more than one program, we have $\| \mathcal{L}(K(\sigma)) \| \leq O(2^{K(\sigma)})$.
 
 Each iteration of the CEGIS loop distinguishes at least one incorrect function from the set of correct
-functions, so we can see that the loop will iterate no more than $\| \mathcal{L}(K(\sigma)) \|$ times.
+functions, so the loop will iterate no more than $\| \mathcal{L}(K(\sigma)) \|$ times.
 Therefore another bound on our runtime is:
 $$O\left (2^{K(\sigma)} \times NP \left ( K(\sigma) \times \| L \| \right) \right) = O \left(2^{K(\sigma)} \right)$$
-Which means that for even moderately large $K(\sigma)$ our procedure is intractably slow.  Conversely
+So for even moderately large $K(\sigma)$ our procedure is intractably slow.  Conversely
 if $K(\sigma)$ is small we will be able to find a termination proof, regardless of other parameters

### 2014-07-08 papers/termination/paper.tex
Compress algorithms to save space

@@
 e.g. the execution where \lstinline!nondet()! always returns 0.
-However it is equally clear that each state has a successor
+However each state has a successor
 that exits the loop, i.e.~when \lstinline|nondet()| returns
@@
 and hence we cannot give a proof of its non-termination
-with formula {\bf [CNT]} and instead must use formula {\bf [SNT]}.
+with {\bf [CNT]} and instead must use {\bf [SNT]}.
 
@@
 and termination respectively.  Exactly one of these will be a genuine proof, so we can check
-first one and then the other, as shown in Algorithm~\ref{alg:termination}.
+first one and then the other, as shown in Algorithm~\ref{alg:termination}:
 
 \begin{algorithm}
+{\small
  \caption{Bit-vector termination
  \label{alg:termination}}
- %\vspace{-1.5em}
+ \vspace{-.5em}
  \begin{algorithmic}[1]
@@
  \end{algorithmic}
+ }
 \end{algorithm}
-
-Since $L$ either terminates or does not terminate, this formula is
-satisfiable in the synthesis fragment.  A solution to the formula would
-include witnesses $P_N$ and $P_T$, which are putative proofs of
-non-termination and termination, respectively.  Exactly one of these will be
-a genuine proof, so we can check first one and then the other.
 
@@
 
-\begin{algorithm*}
+%\begin{figure}
+\begin{algorithm}
+{\small
  \caption{Abstract refinement algorithm
  \label{alg:cegis}}
- \vspace{-1.5em}
- \begin{multicols}{2}
+ %\vspace{-1.7em}
+ %\begin{multicols}{2}
  \begin{algorithmic}[1]
@@
  \end{algorithmic}
- \end{multicols}
-\end{algorithm*}
+ %\end{multicols}
+ }
+\end{algorithm}
+%\end{figure}
 
@@
 \begin{figure}
+\begin{center}
 {\small
-\begin{center}
+
 \setlength{\tabcolsep}{16pt}
@@
 
+}
 \end{center}
-}
+
  \caption{The language $\mathcal{L}$}

### 2014-07-08 papers/termination/paper.tex
Move CEGIS algorithm
Fix bad hboxes

@@
 %
-An injective function ${R:X\to Y}$ is a \emph{ranking function} for the
+A function ${R:X\to Y}$ is a \emph{ranking function} for the
 transition relation $T$ if $Y$ is a well-founded set with order $>$ and 
-$R$ is monotonically decreasing with respect to $T$.  That is
+$R$ is injective and monotonically decreasing with respect to $T$.  That is
 to say:
@@
 satisfiability of a generalised termination specification is a
-\mbox{(non-)termination} proof.  As mentioned earlier in the paper, our
+proof of \mbox{(non-)termination}.  As mentioned earlier in the paper, our
 analysis must ensure precise treatment of bit-wise operations and arithmetic
@@
 
+\begin{figure}
+ \begin{center}
+ \resizebox{\linewidth}{!}{%
+  \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,
+ semithick, initial text=]
+
+  \matrix[nodes={draw, fill=none, shape=rectangle, minimum height=1cm, minimum width=1.5cm},
+          row sep=2cm, column sep=2.5cm, ampersand replacement=\&] {
+   \node (synth) {Synthesise};
+   \&
+   \node (verif) {Verify}; %\\
+   %\node[draw=none] {};
+   \&
+   \node[ellipse] (done) {Done}; \\
+  };
+
+   \path
+    (synth) edge [bend left] node {Candidate program} (verif)
+    (verif) edge [bend left] node {Counterexample input} (synth)
+    (verif) edge node {Valid} (done);
+ \end{tikzpicture}
+ }
+ \end{center}
+
+ \caption{Abstract synthesis refinement loop
+ \label{fig:abstract-refinement}}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+{\small
+
+\setlength{\tabcolsep}{14pt}
+Integer arithmetic instructions:
+
+\begin{tabular}{llll}
+ \verb|add a b| & \verb|sub a b| & \verb|mul a b| & \verb|div a b| \\
+ \verb|neg a| &   \verb|mod a b| & \verb|min a b| & \verb|max a b|
+\end{tabular}
+
+\medskip
+
+Bitwise logical and shift instructions:
+
+\begin{tabular}{lll}
+ \verb|and  a b| & \verb|or   a b| & \verb|xor a b| \\
+ \verb|lshr a b| & \verb|ashr a b| & \verb|not a|
+\end{tabular}
+
+\medskip
+
+Unsigned and signed comparison instructions:
+
+\begin{tabular}{lll}
+ \verb|le  a b| & \verb|lt  a b| & \verb|sle  a b| \\
+ \verb|slt a b| & \verb|eq  a b| & \verb|neq  a b| \\
+\end{tabular}
+
+\medskip
+
+Miscellaneous logical instructions:
+
+\begin{tabular}{lll}
+ \verb|implies a b| & \verb|ite a b c| &  \\
+\end{tabular}
+
+\medskip
+
+\setlength{\tabcolsep}{12pt}
+
+Floating-point arithmetic:
+
+\begin{tabular}{llll}
+ \verb|fadd a b| & \verb|fsub a b| & \verb|fmul a b| & \verb|fdiv a b|
+\end{tabular}
+
+}
+\end{center}
+
+ \caption{The language $\mathcal{L}$}
+ \label{fig:l-language}
+\end{figure}
+
+\subsection{The Synthesis Algorithm}
+
+We use Counterexample Guided Inductive Synthesis
+(CEGIS)~\cite{lezama-thesis,sketch, DBLP:conf/iclp/BrainCVF06} to find an
+$\mathcal{L}$-program satisfying our specification.  The core of the CEGIS
+algorithm is the refinement loop shown in Fig.~\ref{fig:abstract-refinement}
+and detailed in Algorithm~\ref{alg:cegis}.
+
 %\begin{figure}
@@
 {\small
- \caption{Abstract refinement algorithm
+ \caption{Abstract synthesis algorithm
  \label{alg:cegis}}
@@
 
-\begin{figure}
- \begin{center}
- \resizebox{\linewidth}{!}{%
-  \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,
- semithick, initial text=]
-
-  \matrix[nodes={draw, fill=none, shape=rectangle, minimum height=1cm, minimum width=1.5cm},
-          row sep=2cm, column sep=2.5cm, ampersand replacement=\&] {
-   \node (synth) {Synthesise};
-   \&
-   \node (verif) {Verify}; %\\
-   %\node[draw=none] {};
-   \&
-   \node[ellipse] (done) {Done}; \\
-  };
-
-   \path
-    (synth) edge [bend left] node {Candidate program} (verif)
-    (verif) edge [bend left] node {Counterexample input} (synth)
-    (verif) edge node {Valid} (done);
- \end{tikzpicture}
- }
- \end{center}
-
- \caption{Abstract synthesis refinement loop
- \label{fig:abstract-refinement}}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-{\small
-
-\setlength{\tabcolsep}{16pt}
-Integer arithmetic instructions:
-
-\begin{tabular}{llll}
- \verb|add a b| & \verb|sub a b| & \verb|mul a b| & \verb|div a b| \\
- \verb|neg a| &   \verb|mod a b| & \verb|min a b| & \verb|max a b|
-\end{tabular}
-
-\medskip
-
-Bitwise logical and shift instructions:
-
-\begin{tabular}{lll}
- \verb|and  a b| & \verb|or   a b| & \verb|xor a b| \\
- \verb|lshr a b| & \verb|ashr a b| & \verb|not a|
-\end{tabular}
-
-\medskip
-
-Unsigned and signed comparison instructions:
-
-\begin{tabular}{lll}
- \verb|le  a b| & \verb|lt  a b| & \verb|sle  a b| \\
- \verb|slt a b| & \verb|eq  a b| & \verb|neq  a b|
-\end{tabular}
-
-Miscellaneous logical instructions:
-
-\begin{tabular}{lll}
- \verb|implies a b| & \verb|ite a b c| & 
-\end{tabular}
-
-Floating-point arithmetic:
-
-\begin{tabular}{llll}
- \verb|fadd a b| & \verb|fsub a b| & \verb|fmul a b| & \verb|fdiv a b| 
-\end{tabular}
-
-}
-\end{center}
-
- \caption{The language $\mathcal{L}$}
- \label{fig:l-language}
-\end{figure}
-
-\subsection{The Synthesis Algorithm}
-
-We use Counterexample Guided Inductive Synthesis
-(CEGIS)~\cite{lezama-thesis,sketch, DBLP:conf/iclp/BrainCVF06} to find an
-$\mathcal{L}$-program satisfying our specification.  The core of the CEGIS
-algorithm is the refinement loop shown in Fig.~\ref{fig:abstract-refinement}
-and detailed in Algorithm~\ref{alg:cegis}.  The algorithm is divided into
-two procedures: {\sc synth} (see Figure~\ref{fig:synth-dfd}) and {\sc
+The algorithm is divided into two procedures: {\sc synth} (see Figure~\ref{fig:synth-dfd}) and {\sc
 verif}, which interact via a finite set of test vectors {\sc inputs}.

### 2014-07-08 papers/termination/paper.tex
Bad hboxes
Move algorithm

@@
 \begin{framed}
-\begin{minipage}{0.16\textwidth}
+\begin{minipage}{0.17\textwidth}
 \begin{lstlisting}[mathescape=true]
@@
 \vline
-\begin{minipage}{0.84\textwidth}
+\begin{minipage}{0.82\textwidth}
 \begin{align*}
@@
 and termination respectively.  Exactly one of these will be a genuine proof, so we can check
-first one and then the other, as shown in Algorithm~\ref{alg:termination}:
-
-\begin{algorithm}
-{\small
- \caption{Bit-vector termination
- \label{alg:termination}}
- \vspace{-.5em}
- \begin{algorithmic}[1]
- \Statex
-\Function{terminates}{$L$}
-  %\Let{termination-spec}{$\exists W, R. \forall x, x' . \phi(W, R, x, x')$} \Comment{Formula {\bf [CT]}}
-  %\Let{nonterm-spec}{$\exists N, C, x_0, \forall x. \psi(N, C, x_0, x)$} \Comment{Formula {\bf [SNT]}}
-  \Statex  \Comment{Formula {\bf[GT]}:}
-  \Let{spec}{$\exists P_T, P_N. \phi(P_T, x, x') \vee \psi(P_N, y)$}
-  \Let{$(P_T, P_N)$}{{\sc Synthesise}(spec)}
-  \If{decide($\exists x, x' . \lnot \phi(P_T, x, x'))$ = unsat}
-    \State \Return{$L$ terminates}
-  \ElsIf{decide($\exists x . \lnot \psi(P_N, x))$ = unsat}
-    \State \Return{$L$ does not terminate}
-  \EndIf
-\EndFunction
- \end{algorithmic}
- }
-\end{algorithm}
+first one and then the other, which will be illustrated by Algorithm~\ref{alg:termination} once
+we have shown how to solve the formula {\bf [GT]}.
 
@@
 
+Now that we have a procedure for solving a formula
+in the synthesis fragment, we can build the full procedure for
+deciding termination of a loop $L$.  This is shown in Algorithm~\ref{alg:termination}.
+
+\begin{algorithm}
+{\small
+ \caption{Bit-vector termination
+ \label{alg:termination}}
+ \vspace{-.5em}
+ \begin{algorithmic}[1]
+ \Statex
+\Function{terminates}{$L$}
+  %\Let{termination-spec}{$\exists W, R. \forall x, x' . \phi(W, R, x, x')$} \Comment{Formula {\bf [CT]}}
+  %\Let{nonterm-spec}{$\exists N, C, x_0, \forall x. \psi(N, C, x_0, x)$} \Comment{Formula {\bf [SNT]}}
+  \Statex  \Comment{Formula {\bf[GT]}:}
+  \Let{spec}{$\exists P_T, P_N. \phi(P_T, x, x') \vee \psi(P_N, y)$}
+  \Let{$(P_T, P_N)$}{{\sc Synthesise}(spec)}
+  \If{decide($\exists x, x' . \lnot \phi(P_T, x, x'))$ = unsat}
+    \State \Return{$L$ terminates}
+  \ElsIf{decide($\exists x . \lnot \psi(P_N, x))$ = unsat}
+    \State \Return{$L$ does not terminate}
+  \EndIf
+\EndFunction
+ \end{algorithmic}
+ }
+\end{algorithm}
+
 \subsection{Candidate Generation Strategies}
@@
 \begin{proof}
- A non-termination proof is a triple $\langle N, C, x_0 \rangle$ where
+ A proof of non-termination is a triple $\langle N, C, x_0 \rangle$ where
  \mbox{$N \subseteq \mathcal{S}$} is a (finite) recurrence set and

### 2014-07-08 papers/termination/paper.tex
fix the O()

@@
 denote this set $\mathcal{L}(K(\sigma))$.
-Since there are $O(2^{K(\sigma)}$ \emph{programs} of length $K(\sigma)$ and some functions
+Since there are $O(2^{K(\sigma)})$ \emph{programs} of length $K(\sigma)$ and some functions
 will be computed by more than one program, we have $\| \mathcal{L}(K(\sigma)) \| \leq O(2^{K(\sigma)})$.

### 2014-07-08 papers/termination/paper.tex
Experiments

@@
 
-\title{Synthesising Concise Termination and Non-Termination Proofs for Bit-Vector Programs}
+\title{Synthesising Termination and Non-Termination Proofs for Bit-Vector Programs}
 
@@
 provide a comparison point, we also ran {\sc ARMC}~\cite{armc-website} on
-the same benchmarks.  Each tool was given a time limit of 180s, and was
+the same benchmarks.  Each tool was given a time limit of 180\,s, and was
 run on an unloaded 8-core 3.07\,GHz Xeon X5667 with 50\,GB of RAM.  The
@@
 invested in {\sc ARMC}, we feel that the difference is probably inherent to
-the difference in the two approaches -- our algorithm is complete and
-bit-vector accurate, which comes at the cost of efficiency.
-{\sc Juggernaut} is slow, but unstoppable in its
-termination capabilities if you are willing to wait for it.
+the difference in the two approaches -- our solver is more general
+than {\sc ARMC}, in that it provides a complete proof system for both
+termination and non-termination.  This comes at the cost of efficiency:
+{\sc Juggernaut} is slow, but unstoppable.
 
@@
 {\sc Juggernaut} handles the nonlinear cases correctly and rapidly.
-It solves 2 of the 5 nested loops in less than 30s, but times out on the other
-3.  Of these 3 timeouts, 2 (svcomp40 and svcomp41) are due to program constants that
-are too large to be handled with our default settings.  Increasing
-our default word width from 3 bits to 5 bits allowed us to prove termination
-for both these benchmarks in less than 30s; these results are marked with a *.
+It solves 4 of the 5 nested loops in less than 30s\footnote{Although
+2 of these cases timed out with our default parameters, changing one command
+line flag allowed them to go through.  The benchmarks using non-default
+command line flags are marked with a *.}, but times out on the 5th.
 

### 2014-07-08 papers/termination/paper.tex
nudge

--- 2014-07-08 papers/termination/paper.tex
@@
 bit-vectors exhibit incomparable behaviours with respect to termination,
-i.e.~ programs that terminate for integers may \emph{not} terminate for
+i.e.~programs that terminate for integers need \emph{not} terminate for
 bit-vectors and vice versa.  Thus, abstracting bit-vectors with integers may

### 2014-07-08 papers/termination/paper.tex
Edits

--- 2014-07-08 papers/termination/paper.tex
@@
 storage terminates.  Since such programs are necessarily finite state, our
-problem is much easier than Turing's, but is more relevant to analysing computer
+problem is much easier than Turing's, but is a better fit for analysing computer
 programs.
@@
 
+\item  We rephrased the termination and non-termination problems as
+second-order $\exists \forall$ satisfaction problems.  This formulation
+captures the (non-)termination
+properties of all of the loops in the program, including
+nested loops.  We can use this to analyse all the loops at once,
+or one at a time.  Our treatment handles termination and non-termination uniformly:
+both properties are captured in the same second-order formula.
+
 \item We designed a bit-level accurate technique for computing ranking
@@
 
-\item  We rephrased the termination and non-termination problems as
-second-order $\exists \forall$ satisfaction problems.  This formulation
-captures the (non-)termination
-properties of all of the loops in the program, including
-nested loops.  We can use this to analyse all the loops at once,
-or one at a time.
-% In contrast to previous approaches to program synthesis, we
-% found that using genetic programming was very effective at generating
-% candidate termination and non-termination proofs.
-
 \item We used program synthesis techniques to solve the second-order satisfaction
@@
 \item Our solver is based on a series of source-to-source
-transformations.  As well as leading to a concise implementation, this has
-the side effect of ensuring that our proofs are correct for the exact binary
-produced by a particular compiler, rather than for an abstraction of the
-program.
+transformations.  As well as leading to a concise implementation, this
+allows us to leverage existing, mature verification tools.
 
@@
 section we show how several variations of both the termination and the
-non-termination problem can be defined in second-order logic.
+non-termination problem can be uniformly defined in second-order logic.
 
@@
 %
+We note that by existentially quantifying over Skolem functions, formulae with arbitrary
+first-order quantification can be brought into the synthesis fragment~\cite{hol-book}, so the
+fragment is semantically less restrictive than it looks.
 Checking satisfiability of a formula in the synthesis fragment corresponds to program synthesis and
@@
 (it is an $\exists \forall \exists$ formula).  To eliminate the innermost existential quantifier,
-we introduce a Skolem function $C$ that chooses the successor $x'$, which we then existentially quantify over.  This results in
-formula {\bf [SNT]} (Definition~\ref{def:snt}, Figure~\ref{fig:single_loop}).
+we introduce a Skolem function $C$ that chooses the successor $x'$, which we then existentially quantify over.
+This results in formula {\bf [SNT]} (Definition~\ref{def:snt}, Figure~\ref{fig:single_loop}).
 
@@
 we will not provide the full algorithm, but rather illustrate it through the example in Figure~\ref{fig:environment-model}.
-Our technique proves the termination of each loop separately, 
-with the exception of nested loops, whose termination we encode in the same formula. For the current example, the termination formula is given on the right side of 
+Full details of this construction appear in the extended version of this paper~\cite{extended-version}.
+For the current example, the termination formula is given on the right side of 
 Figure~\ref{fig:environment-model}:
@@
 %% as the conjunction of constraints on the right side of the figure:
-\begin{itemize}
-\item $T_o$  over-approximates $L_1$'s transition relation.
-\item $S$ is a summary of $L_2$ that over-approximates the transitive closure of its transition relation.
-\item If the formula is satisfiable, then $R_1$ and $R_2$ are ranking functions for $L_1$ and $L_2$, respectively.
-\end{itemize}
+$T_o$ is a summary of $L_1$ that over-approximates its transition relation;
+$R_1$ and $R_2$ are ranking functions for $L_1$ and $L_2$, respectively.
 
@@
 \begin{align*}
- \exists T_o, S, R_1, R_2 . \forall i, j, n, i', j', n' . \\
-   %% x = 0 & \rightarrow W_1(v, x, y, z) \, \wedge \\
-   %% x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
-   %% x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x, y, z) \, \wedge \\
-   S(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' \leq i' & \rightarrow S(\langle i, j, n \rangle, \langle i', j'+1, n' \rangle) \, \wedge \\
-   & R_2(i,j,n) > 0 \, \wedge \\
-   & R_2(i,j,n) > R_2(i',j'+1,n') \, \wedge\\
-   i < n \wedge T_o(i, j, n) & \rightarrow S(\langle i, j, n \rangle, \langle i, 0, n \rangle) \, \wedge \\
-   i < n \wedge T_o(i, j, n) \wedge S(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' > i' & \rightarrow \\
-   & T_o(i'+1, j', n') \, \wedge \\
-   & R_1(i, j, n) > 0 \, \wedge \\
-   & R_1(i, j, n) > R_1(i'+1, j', n')
+ \exists T_o, R_1, R_2 . \forall i, j, n, i', j', n' . \\
+   i < n & \rightarrow T_o(\langle i, j, n \rangle , \langle i, 0, n \rangle) ~ \wedge \\
+   j \leq i \wedge T_o(\langle i', j', n' \rangle, \langle i, j, n\rangle ) & \rightarrow R_2(i, j, n) > 0 ~ \wedge \\
+     & R_2(i, j, n) > R_2(i, j+1, n) ~ \wedge \\
+     & T_o(\langle i', j', n' \rangle, \langle i, j+1, n \rangle) ~ \wedge \\
+   i < n \wedge S(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' > i' & \rightarrow R_1(i, j, n) > 0 ~ \wedge \\
+                                         & R_1(i, j, n) > R_1(i+1, j, n)
    \end{align*}
+% \begin{align*}
+%  \exists T_o, S, R_1, R_2 . \forall i, j, n, i', j', n' . \\
+%    %% x = 0 & \rightarrow W_1(v, x, y, z) \, \wedge \\
+%    %% x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
+%    %% x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x, y, z) \, \wedge \\
+%    S(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' \leq i' & \rightarrow S(\langle i, j, n \rangle, \langle i', j'+1, n' \rangle) \, \wedge \\
+%    & R_2(i,j,n) > 0 \, \wedge \\
+%    & R_2(i,j,n) > R_2(i',j'+1,n') \, \wedge\\
+%    i < n \wedge T_o(i, j, n) & \rightarrow S(\langle i, j, n \rangle, \langle i, 0, n \rangle) \, \wedge \\
+%    i < n \wedge T_o(i, j, n) \wedge S(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' > i' & \rightarrow \\
+%    & T_o(i'+1, j', n') \, \wedge \\
+%    & R_1(i, j, n) > 0 \, \wedge \\
+%    & R_1(i, j, n) > R_1(i'+1, j', n')
+% \end{align*}
 \end{minipage}
@@
 
-As well as efficiency, compiling the original source code of the program
-under analysis has the advantage of guaranteeing that we are verifying the
-exact semantics of the program as understood by a compiler, rather than some
-ad-hoc abstraction of its semantics.
+% As well as efficiency, compiling the original source code of the program
+% under analysis has the advantage of guaranteeing that we are verifying the
+% exact semantics of the program as understood by a compiler, rather than some
+% ad-hoc abstraction of its semantics.
 
@@
  \label{thm:generalised-sat}
- The generalised termination formula {\bf [GT]} for any loop $L$ is satisfiable
+ The generalised termination formula {\bf [GT]} for any loop $L$ is valid
  when $P_N$ and $P_T$ range over \mbox{$\mathcal{L}$-computable} functions.
@@
 \begin{proof}
- By Theorem~\ref{thm:generalised-sat}, the specification \lstinline!spec! is satisfiable.  Corollary~\ref{cor:cegis-newc}
+ By Theorem~\ref{thm:generalised-sat}, the specification \lstinline!spec! is valid, and hence satisfiable.  Corollary~\ref{cor:cegis-newc}
  shows that Algorithm~\ref{alg:cegis} is semi-complete, and so the call to {\sc Synthesise} terminates with a witness $(P_T, P_N)$.
@@
 named {\sc Juggernaut}, on 47
-benchmarks taken from the literature and SV-COMP'15~\cite{svcomp15}.  To
+benchmarks taken from the literature and SV-COMP'15~\cite{svcomp15}\footnote{
+The benchmarks, code and log files may be downloaded from the URL \texttt{http://www.cprover.org/juggernaut/}}.  To
 provide a comparison point, we also ran {\sc ARMC}~\cite{armc-website} on
@@
 command line flags are marked with a *.}, but times out on the 5th.
+
+Overall, we notice that the GP component was always the fastest to find
+candidate solutions, {\sc ExplicitSearch} was always fastest
+to find counterexample inputs, and {\sc CBMC} was always fastest at
+proving correctness.  Each of these search methods has its strengths,
+and removing any one would have had a significant negative impact on
+overall performance.  Approximately 90\% of the total
+runtime was spent in synthesising new candidate solutions, which matches
+our expectations -- the (non-)termination proofs are difficult to
+find, but easy to check.
 

### 2014-07-08 papers/termination/paper.tex
typos

@@
 \end{align*}
-If the initial states $I$ are ignored this loop cannot be shown to terminate, since any state with $y = 0$ and $x > 0$
+If the initial states $I$ are ignored, this loop cannot be shown to terminate, since any state with $y = 0$ and $x > 0$
 would lead to a non-terminating execution.
@@
 $\mathcal{L}$-program satisfying our specification.  The core of the CEGIS
-algorithm is the refinement loop shown in Fig.~\ref{fig:abstract-refinement}
+algorithm is the refinement loop shown in Figure~\ref{fig:abstract-refinement}
 and detailed in Algorithm~\ref{alg:cegis}.

### 2014-07-08 papers/termination/paper.tex
Edits

@@
 ones~\cite{DBLP:conf/popl/MagillTLT10}.  Also, we have not yet added support
-for recursion to our encoding.
+for recursion or \texttt{goto} to our encoding.
 
@@
 \item[(ii)] terminating programs have termination arguments expressible in linear arithmetic.
-\item[(iii)] the semantics of bit vectors and mathematical integers are equivalent.
+\item[(iii)] the semantics of bit-vectors and mathematical integers are equivalent.
 \item[(iv)] the semantics of IEEE floating-point numbers and mathematical reals are equivalent.
@@
 bit-vector semantics and illustrate differences between machine arithmetic
-and integer arithmetic that show that the abstraction of bit vectors to
+and integer arithmetic that show that the abstraction of bit-vectors to
 mathematical integers is unsound (Section~\ref{sec:machine.arith}).
@@
 Due to its expressiveness, second-order logic is very difficult to reason
-about, with many second-order theories becoming undecidable even when the
+in, with many second-order theories becoming undecidable even when the
 corresponding first-order theory is decidable.
@@
 i.e.~the mechanised construction of software that provably satisfies a
-given specification.  Then a witness $P$ to the satisfiability of some
-formula is just a program that meets the specification $\sigma$.
+given specification.  A witness $P$ to the satisfiability of some
+formula is then just a program that meets the specification $\sigma$.
 
@@
 such a large number and so modern computers do their arithmetic over
-fixed-width binary words, otherwise known as bit vectors.  For the remainder
-of this section, we will say that the bit vectors we are working with are
+fixed-width binary words, otherwise known as bit-vectors.  For the remainder
+of this section, we will say that the bit-vectors we are working with are
 $k$-bits wide, which means that each word can hold one of $2^k$ bit
@@
 include the Boolean operators \texttt{and, or, not, xor} applied to each
-element of the bit vector.  Computer programs often make use of these
+element of the bit-vector.  Computer programs often make use of these
 operators, which are nonlinear when interpreted in the standard model of
@@
 Given the lack of research on termination of nonlinear programs, as well as
-programs over bit vectors and floats, our work focused on covering these
+programs over bit-vectors and floats, our work focused on covering these
 areas.  One of the obvious conclusions that can be reached from

### 2014-07-08 papers/termination/paper.tex
nudges

--- 2014-07-08 papers/termination/paper.tex
@@
 Under-approximating the inner loop can be done with a nested existential quantifier, resulting in
-$\exists \forall \exists$ alternation which we could eliminate with skolemization.  
+$\exists \forall \exists$ alternation which we could eliminate with Skolemization.  
 However, we observe that
@@
 \begin{align*}
- (\exists P_T . \forall x, x'. \phi(P_T, x, x')) \vee (\exists P_N . \forall x, \psi(P_N, x))
+ (\exists P_T . \forall x, x'. \phi(P_T, x, x')) \vee (\exists P_N . \forall x.\, \psi(P_N, x))
 \end{align*}
@@
 Which simplifies to:
+%
 \begin{definition}[Generalised Termination Formula {\bf [GT]}]
@@
 \begin{align*}
- \exists P_T, P_N. \forall x, x', y, \phi(P_T, x, x') \vee \psi(P_N, y)
+ \exists P_T, P_N. \forall x, x', y.\, \phi(P_T, x, x') \vee \psi(P_N, y)
 \end{align*}
@@
 %
-where $P$ ranges over functions, while $x$ and $y$ range over ground terms
-(as mentioned, we drop the vector notation and write $x$ for $\vec{x}$).
+where $P$ ranges over functions, and $x$ and $y$ range over ground terms
+(as mentioned, we drop vector notation and write $x$ for $\vec{x}$).
 
@@
 
-We will now show that $\mathcal{L}$ is expressive enough to encode (non-)termination proofs for
-every bit-vector program.
+We will now show that $\mathcal{L}$ is expressive enough to encode
+\mbox{(non-)}ter\-mi\-nation proofs for every bit-vector program.
 
@@
 a finite $\mathcal{L}$-program computing $R''$.
-
+%
 % Since {\sc Kalashnikov} enumerates all programs in order of
@@
 We will now show that the number of iterations of the CEGIS loop is a function
-of the Kolmogorov complexity of the (non-)termination proof, and so
+of the Kolmogorov complexity of the \mbox{(non-)}ter\-mi\-nation proof, and so
 our runtime is asymptotically a function of the Kolmogorov complexity.

### 2014-07-08 papers/termination/paper.tex
Shrink diagram

--- 2014-07-08 papers/termination/paper.tex
@@
 since it was first considered by Turing in 1936~\cite{turing}.  Informally,
-we can say that the halting problem is concerned with answering the question
+the halting problem is concerned with answering the question
 ``does this program run forever, or will it eventually terminate?''
@@
 syntactic form, and do not use templates.  Furthermore, our approach allows
-us to simultaneously search for proofs of non-termination, which take the form
-of recurrence sets.
+us to \emph{simultaneously} search for proofs of \emph{non-termination},
+which take the form of recurrence sets.
 
@@
 bit-vector semantics and integer semantics give rise to the same termination
-behaviour.  Thus, most of the techniques treat bit-vectors and IEEE floats
-as mathematical integers and reals,
+behaviour.  Thus, most existing techniques treat fixed-width machine integers
+(bit-vectors) and IEEE floats as mathematical integers and reals,
 respectively~\cite{DBLP:conf/pldi/CookPR06,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10}.
@@
 
-\begin{figure*}
+\begin{figure}
 \begin{center}
@@
 
-%\resizebox{\linewidth}{!}{
+\resizebox{\linewidth}{!}{
 \begin{tikzpicture}[font=\sffamily]
@@
 \end{tikzpicture}
-%}
+}
 \end{center}
@@
 \label{fig:synth-dfd}
-\end{figure*}
+\end{figure}
 
@@
 under analysis would terminate if run with unbounded integer variables,
-while we are checking whether the program terminates with fixed-width
-integer variables.  This means that {\sc ARMC}'s verdict differs from ours
+while we are checking whether the program terminates with bit-vector
+variables.  This means that {\sc ARMC}'s verdict differs from ours
 in 3 cases (due to the differences between integer and bit-vector

### 2014-07-08 papers/termination/paper.tex
units

--- 2014-07-08 papers/termination/paper.tex
@@
 {\sc Juggernaut} handles the nonlinear cases correctly and rapidly.
-It solves 4 of the 5 nested loops in less than 30s\footnote{Although
+It solves 4 of the 5 nested loops in less than 30\,s\footnote{Although
 2 of these cases timed out with our default parameters, changing one command
@@
 and removing any one would have had a significant negative impact on
-overall performance.  Approximately 90\% of the total
+overall performance.  Approximately 90\,\% of the total
 runtime was spent in synthesising new candidate solutions, which matches

### 2014-07-08 papers/termination/paper.tex
nudge

@@
 provers)~\cite{DBLP:conf/sas/HarrisLNR10}.  We propose a uniform treatment
-of termination and non-termination by formulating a generalised second order
+of termination and non-termination by formulating a generalised second-order
 formula whose solution is a proof of one of them.

### 2014-07-08 papers/termination/paper.tex
Bibtex

@@
 {\sc Juggernaut} handles the nonlinear cases correctly and rapidly.
-It solves 4 of the 5 nested loops in less than 30\,s\footnote{Although
+It solves 4 of the 5 nested loops in less than 30s\footnote{Although
 2 of these cases timed out with our default parameters, changing one command
@@
 and removing any one would have had a significant negative impact on
-overall performance.  Approximately 90\,\% of the total
+overall performance.  Approximately 90\% of the total
 runtime was spent in synthesising new candidate solutions, which matches
@@
 provers)~\cite{DBLP:conf/sas/HarrisLNR10}.  We propose a uniform treatment
-of termination and non-termination by formulating a generalised second-order
+of termination and non-termination by formulating a generalised second order
 formula whose solution is a proof of one of them.
+
+\nocite{fisher2009phylogeny}
 

### 2014-07-08 papers/termination/paper.tex
Typo

--- 2014-07-08 papers/termination/paper.tex
@@
 deciding it.  Our strategy for proving soundness and completeness
-is first to prove that Algorithm~\ref{alg:cegis} (as instanciated
+is first to prove that Algorithm~\ref{alg:cegis} (as instantiated
 with \newC as a background logic) is a semi-decision
@@
  Since safety of \newC programs is decidable, Algorithm~\ref{alg:cegis} is semi-complete
- when instanciated with \newC as a background theory.
+ when instantiated with \newC as a background theory.
 \end{corollary}
@@
 
+\nocite{fisher2009phylogeny}
+
 \bibliographystyle{abbrvnat}

### 2014-07-08 papers/termination/paper.tex
Remove "production ready"

@@
 
-In conclusion, while they show that our tool is not yet production ready,
+In conclusion,
 these experiments confirm our conjecture that synthesis can be used

### 2014-07-08 papers/termination/paper.tex
Equivalently -> alternatively

@@
 We note that some termination arguments require lexicographic ranking functions, or
-equivalently, ranking functions whose co-domain is a countable ordinal, rather than just $\mathbb{N}$.
+alternatively, ranking functions whose co-domain is a countable ordinal, rather than just $\mathbb{N}$.
 

### 2014-07-08 papers/termination/paper.tex
Synthesis background theory
nudges

@@
 with a constrained use of quantification that is expressive enough to encode
-both termination and non-termination of finite programs, while admitting a semi-decision procedure.
+both termination and non-termination of finite-state programs, while admitting a semi-decision procedure.
 We will suggestively refer to the fragment as the \emph{synthesis fragment}:
@@
 true iff the functions $\vec{P}$ compute appropriate outputs when fed the
-inputs $\vec{x}$ and $\vec{y}$.
+inputs $\vec{x}$ and $\vec{y}$.  The synthesis fragment is implicitly
+defined over some background first-order theory, and $\sigma$ must be
+computable in that theory.
 %
@@
 Under-approximating the inner loop can be done with a nested existential quantifier, resulting in
-$\exists \forall \exists$ alternation which we could eliminate with Skolemization.  
+$\exists \forall \exists$ alternation, which we could eliminate with Skolemization.  
 However, we observe that
@@
 and termination respectively.  Exactly one of these will be a genuine proof, so we can check
-first one and then the other, which will be illustrated by Algorithm~\ref{alg:termination} once
+first one and then the other, which will be exemplified by Algorithm~\ref{alg:termination} once
 we have shown how to solve the formula {\bf [GT]}.
@@
 in the synthesis fragment, we can build the full procedure for
-deciding termination of a loop $L$.  This is shown in Algorithm~\ref{alg:termination}.
+deciding termination of a loop $L$.  This is given as Algorithm~\ref{alg:termination}.
 
@@
  added to the inputs set.  Therefore if the {\sc Synth} procedure always terminates,
- Algorithm~\ref{alg:cegis}.
+ Algorithm~\ref{alg:cegis} does as well.
 
@@
  Similarly if $L$ does not terminate for some input, by Theorem~\ref{thm:l-nonterm} there is a non-termination
- proof $P_N$ expressible in $\mathcal{L}$.  $\psi$ is an instance of {\bf [SNT]} and so $P_N \models \psi$
+ proof $P_N$ expressible in $\mathcal{L}$.  Formula $\psi$ is an instance of {\bf [SNT]} and so $P_N \models \psi$
  (Theorem~\ref{thm:snt}), hence for any $P_T$, $(P_T, P_N) \models \phi \vee \psi$.
@@
 
-Hence Algorithm~\ref{alg:termination} is sound and complete for bit-vector termination.
+Hence, Algorithm~\ref{alg:termination} is sound and complete for bit-vector termination.
 

### 2014-07-09 papers/termination/paper.tex
Remove author names for double blind
post-submission typo spotting

@@
 
-\authorinfo{Cristina David\and Daniel Kroening\and Matt Lewis}
-           {University of Oxford}
-           {firstname.lastname@cs.ox.ac.uk}
+%\authorinfo{Cristina David\and Daniel Kroening\and Matt Lewis}
+%           {University of Oxford}
+%           {firstname.lastname@cs.ox.ac.uk}
+
+%\authorinfo{}
 
@@
  \item The number of looping states is $\| L \| = l$.
- \item Our transition relation is deterministic and parititioned into chains of length $c_i$, with $l = \sum c_i$.
+ \item Our transition relation is deterministic and partitioned into chains of length $c_i$, with $l = \sum c_i$.
 \end{itemize}
@@
 of the loop guards $G$.  The program is deterministic if its transition
-relation is determinstic for all states except $s$.
+relation is deterministic for all states except $s$.
 
@@
  The loop $L(G, T)$ terminates when started in any state satisfying $I(x)$ iff formula {\bf [CT]}
- (Defintion~\ref{def:ct}, Figure~\ref{fig:conditional_termination}) is satisfiable.
+ (Definition~\ref{def:ct}, Figure~\ref{fig:conditional_termination}) is satisfiable.
 \end{theorem}
@@
 \EndFunction
-\columnbreak
+%\columnbreak
 \Statex
@@
 We argue that this gives our procedure various desirable qualities in practical
-appliations.
+applications.
 We first recall the definition of the Kolmogorov complexity
@@
 
-\nocite{fisher2009phylogeny}
+%\nocite{fisher2009phylogeny}
 

### 2014-07-09 papers/termination/paper.tex
Remove Cprover URLs

@@
 we will not provide the full algorithm, but rather illustrate it through the example in Figure~\ref{fig:environment-model}.
-Full details of this construction appear in the extended version of this paper~\cite{extended-version}.
+%Full details of this construction appear in the extended version of this paper~\cite{extended-version}.
 For the current example, the termination formula is given on the right side of 
@@
 named {\sc Juggernaut}, on 47
-benchmarks taken from the literature and SV-COMP'15~\cite{svcomp15}\footnote{
-The benchmarks, code and log files may be downloaded from the URL \texttt{http://www.cprover.org/juggernaut/}}.  To
+benchmarks taken from the literature and
+SV-COMP'15~\cite{svcomp15}.%\footnote{The benchmarks, code and log files may be downloaded from the URL \texttt{http://www.cprover.org/juggernaut/}}.
+To
 provide a comparison point, we also ran {\sc ARMC}~\cite{armc-website} on

### 2014-07-09 papers/termination/paper.tex
Anonymizing

--- 2014-07-09 papers/termination/paper.tex
@@
 we will not provide the full algorithm, but rather illustrate it through the example in Figure~\ref{fig:environment-model}.
-Full details of this construction appear in the extended version of this paper~\cite{extended-version}.
+Full details of this construction appear in the extended version of this paper.
 For the current example, the termination formula is given on the right side of 
@@
 named {\sc Juggernaut}, on 47
-benchmarks taken from the literature and SV-COMP'15~\cite{svcomp15}\footnote{
-The benchmarks, code and log files may be downloaded from the URL \texttt{http://www.cprover.org/juggernaut/}}.  To
+benchmarks taken from the literature and
+SV-COMP'15~\cite{svcomp15}\footnote{The logs and benchmarks are included in the
+non-anonymous supplementary material submitted with this paper}.%\footnote{The benchmarks, code and log files may be downloaded from the URL \texttt{http://www.cprover.org/juggernaut/}}.
+To
 provide a comparison point, we also ran {\sc ARMC}~\cite{armc-website} on

### 2014-09-08 papers/termination/rebuttal.txt
Rebuttal

Reviewers 1, 3 and 4 (regarding the experimental performance): ARMC is
very good at handling loops with a small number of paths
that have linear, rational termination arguments.  We note that on
harder examples ARMC performs poorly while Juggernaut excels; for example
in the 5 benchmarks with nested loops, ARMC solves none correctly
while Juggernaut solves 4 correctly and times out on the 5th.
Since ARMC is solving a PTIME problem, it is not surprising that it
is faster than our more general algorithm in the cases that are instances
of the PTIME problem.  We also note that ARMC is not sound with respect
to bitvector semantics: our experiments include 3 cases (6%) where
ARMC terminates with an incorrect verdict.  The paper did perhaps
not make it clear that we are comparing the performance of a sound
and complete algorithm (Juggernaut) with an unsound, incomplete
algorithm (ARMC).

Reviewer 3 (regarding the comparison to SyGus): our specifications are C
code, but SyGus specifications are S-Expressions.  This design decision
allows us to use explicit state model checking and GP, as well as getting
all of compiler theory for free.  This difference is non-trivial and
novel.  However the underlying problem is clearly very similar and we
appreciate it being brought to our attention so that we may cite it.

Reviewers 2, 3 and 4 (regarding the lack of novelty over invariant
generation and other second-order techniques): we contend
that the encoding of (non-)termination as a second-order fomula is
non-trivial.  For example characterising the non-termination of nested
loops is not a straightforward extension of invariant generation: it
requires relational reasoning (which is not the same class of problem
as invariant generation -- an inherently non-relational problem) and
a novel application of Skolemisation to handle nondeterminism.
Furthermore, generating a tautology for which we find a constructive
proof is novel and, we contend, a highly non-trivial insight.
Additionally we note that previous work on using second-order logic
for program analysis does not cover termination: [19] requires
a dwf predicate in the logic to model termination, i.e. it uses a
logic with a builtin halting oracle, which makes the encoding trivial;
[20] mentions termination as an afterthought.

Reviewer 4 (regarding the importance of bitvector semantics): we agree
that people really don't *think* about termination using bitvector
arguments. If they did, our algorithm would have extremely limited value
since people would already be soundly thinking about their own code.  The
cases where a bitvector program's termination differs from its
rational counterpart are where bugs are likely to appear.  Equally,
a tool which generates a rational termination argument and checks whether
it applies for bitvector semantics has limited utility: it is likely
that if a rational argument applies, the programmer has already thought
of it.

Reviewers 2, 3 and 4 (regarding the simplicity of our approach): it is
clear that reasoning about bitvector termination is difficult, as
witnessed by the fact that ours is the first algorithm to decide it
effectively.  We consider it the main strength of our work that
we can use a sequence of simple reductions to transform a previously
intractable problem into a simple one.

Reviewer 1 (miscellaneous): C- is distinct from L in that C- is the
implementation language of part of our synthesiser, and also the language
used for specifications.
It really does not leave out any other C features.  L is the language
in which ranking functions are written and is distinct from C-.
While section 5.1 may seem excessive, we feel that many in the POPL
community are unlikely to realise that bitvector termination has no
simple relationship with rational termination: there are programs that
terminate with bitvector semantics but don't with rationls *and vice
versa*.  The ITE instruction is "if-then-else"; we agree this figure
is somewhat opaque and will add some discussion.
The runtime is predominantly a factor of the Kolmogorov complexity,
but the problem is in fact NEXPTIME-complete in the Kolmogorov complexity,
so even modest increases in complexity result in large jumps in runtime. 
We very much appreciate the suggestions regarding the rest of the
presentation and intend to incorporate them into the paper.

Reviewer 2 (miscellaneous): We do not consider our contribution to be that
we have run a lot of experiments, we consider it to be that we have
presented the first algorithm to decide bitvector termination.
We omitted the SVCOMP'15 benchmarks that made use of arrays or
recursion.  We do not have arrays in our logic and we had not implemented
recursion in our frontend (although the latter is uninteresting
theoretically).

Reviewer 3 (miscellaneous): Our contribution is twofold: we first
present a novel reduction from termination to synthesis, then we present
an algorithm and tool for solving the synthesis formula.  The first part is
unarguably significant and theoretically non-trivial (see previous
comment).  The second part is a significant algorithmic contribution
in that it manages to synthesise C programs from a C specification, and
is able to combine SAT solving, explicit state model checking and
stochastic search.  No other program synthesiser does this.

### 2014-09-10 papers/termination/rebuttal.txt
Edit rebuttal

@@
-Reviewers 1, 3 and 4 (regarding the experimental performance): ARMC is
+Reviewers 2, 3 and 4 (regarding the lack of novelty over invariant
+generation and other second-order techniques):
+Previous works on using second-order logic for program analysis use
+completely different approaches for encoding termination: [19] requires
+a dwf predicate in the logic to model termination, i.e. it uses a
+logic with a builtin halting oracle, which makes the encoding trivial.
+[20] reduces (non-)termination to a safety property and uses templates
+to reduce the resulting second-order formula to a first-order one.
+The (non-)termination proofs take the form of preconditions and,
+in contrast to our work, there is no guarantee of completeness or even soundness.
+We maintain that our encoding of (non-)termination as a second-order
+fomula is novel and non-trivial.  For example characterising termination of nested
+loops is not a straightforward extension of invariant generation: it
+requires relational reasoning (which is not the same class of problem
+as invariant generation -- an inherently non-relational problem);
+our use of Skolemisation to handle nondeterminism in non-termination
+arguments is novel and non-trivial; generating a tautology for which we find
+a constructive proof is novel and non-trivial.
+
+Reviewers 1, 3 and 4 (regarding experimental performance): ARMC is
 very good at handling loops with a small number of paths
-that have linear, rational termination arguments.  We note that on
-harder examples ARMC performs poorly while Juggernaut excels; for example
-in the 5 benchmarks with nested loops, ARMC solves none correctly
-while Juggernaut solves 4 correctly and times out on the 5th.
+that have linear, rational termination arguments.
+We note that on harder examples ARMC performs poorly while Juggernaut
+excels, for example in the 5 benchmarks with nested loops, ARMC solves
+none correctly while Juggernaut solves 4 correctly.
 Since ARMC is solving a PTIME problem, it is not surprising that it
@@
 to bitvector semantics: our experiments include 3 cases (6%) where
-ARMC terminates with an incorrect verdict.  The paper did perhaps
+ARMC terminates with an incorrect verdict.  Juggernaut does not
+give an incorrect verdict for any benchmark.  The paper did perhaps
 not make it clear that we are comparing the performance of a sound
 and complete algorithm (Juggernaut) with an unsound, incomplete
-algorithm (ARMC).
+algorithm (ARMC); this is the best comparison we could do,
+since there are no existing sound and complete algorithms to compare
+with.
 
-Reviewer 3 (regarding the comparison to SyGus): our specifications are C
+Reviewers 2, 3 and 4 (regarding the simplicity of our approach): it is
+clear that reasoning about bitvector termination is difficult, as
+witnessed by the fact that ours is the first algorithm to decide it
+without explicitly checking the entire state space.
+This is despite bitvector programs having been written for almost 200 years. 
+We consider it the main strength of our work that
+we can use a sequence of simple reductions to transform a previously
+intractable problem into a simple one.
+
+Reviewers 3 and 4 (regarding the comparison to SyGus): our specifications are C
 code, but SyGus specifications are S-Expressions.  This design decision
@@
 appreciate it being brought to our attention so that we may cite it.
-
-Reviewers 2, 3 and 4 (regarding the lack of novelty over invariant
-generation and other second-order techniques): we contend
-that the encoding of (non-)termination as a second-order fomula is
-non-trivial.  For example characterising the non-termination of nested
-loops is not a straightforward extension of invariant generation: it
-requires relational reasoning (which is not the same class of problem
-as invariant generation -- an inherently non-relational problem) and
-a novel application of Skolemisation to handle nondeterminism.
-Furthermore, generating a tautology for which we find a constructive
-proof is novel and, we contend, a highly non-trivial insight.
-Additionally we note that previous work on using second-order logic
-for program analysis does not cover termination: [19] requires
-a dwf predicate in the logic to model termination, i.e. it uses a
-logic with a builtin halting oracle, which makes the encoding trivial;
-[20] mentions termination as an afterthought.
 
@@
 
-Reviewers 2, 3 and 4 (regarding the simplicity of our approach): it is
-clear that reasoning about bitvector termination is difficult, as
-witnessed by the fact that ours is the first algorithm to decide it
-effectively.  We consider it the main strength of our work that
-we can use a sequence of simple reductions to transform a previously
-intractable problem into a simple one.
+Reviewer 3 (regarding the novelty compared to other synthesis algorithms):
+Our contribution is twofold: we first present a novel reduction from termination
+to synthesis, then we present an algorithm and tool for solving the synthesis formula
+The first part is unarguably significant and theoretically non-trivial (see
+previous comments).  The second part is a significant technical contribution
+in that it manages to synthesise C programs from a C specification, and
+is able to combine SAT solving, explicit state model checking and
+stochastic search.  No other program synthesiser does this.
 
-Reviewer 1 (miscellaneous): C- is distinct from L in that C- is the
+Reviewer 1: C- is distinct from L in that C- is the
 implementation language of part of our synthesiser, and also the language
@@
 
-Reviewer 2 (miscellaneous): We do not consider our contribution to be that
-we have run a lot of experiments, we consider it to be that we have
-presented the first algorithm to decide bitvector termination.
-We omitted the SVCOMP'15 benchmarks that made use of arrays or
-recursion.  We do not have arrays in our logic and we had not implemented
-recursion in our frontend (although the latter is uninteresting
-theoretically).
-
-Reviewer 3 (miscellaneous): Our contribution is twofold: we first
-present a novel reduction from termination to synthesis, then we present
-an algorithm and tool for solving the synthesis formula.  The first part is
-unarguably significant and theoretically non-trivial (see previous
-comment).  The second part is a significant algorithmic contribution
-in that it manages to synthesise C programs from a C specification, and
-is able to combine SAT solving, explicit state model checking and
-stochastic search.  No other program synthesiser does this.
+Reviewer 2 (regarding our selection of benchmarks): We omitted the SVCOMP'15
+benchmarks that made use of arrays or recursion.  We do not have arrays in our
+logic and we had not implemented recursion in our frontend (although the latter
+can be syntactically rewritten to our input format).

### 2014-09-10 papers/termination/rebuttal.txt
edits

@@
-Reviewers 2, 3 and 4 (regarding the lack of novelty over invariant
-generation and other second-order techniques):
-Previous works on using second-order logic for program analysis use
-completely different approaches for encoding termination: [19] requires
-a dwf predicate in the logic to model termination, i.e. it uses a
-logic with a builtin halting oracle, which makes the encoding trivial.
-[20] reduces (non-)termination to a safety property and uses templates
-to reduce the resulting second-order formula to a first-order one.
-The (non-)termination proofs take the form of preconditions and,
-in contrast to our work, there is no guarantee of completeness or even soundness.
-We maintain that our encoding of (non-)termination as a second-order
-fomula is novel and non-trivial.  For example characterising termination of nested
-loops is not a straightforward extension of invariant generation: it
-requires relational reasoning (which is not the same class of problem
-as invariant generation -- an inherently non-relational problem);
-our use of Skolemisation to handle nondeterminism in non-termination
-arguments is novel and non-trivial; generating a tautology for which we find
-a constructive proof is novel and non-trivial.
+Reviewers 2, 3 and 4 (lack of novelty over invariant generation and other
+second-order techniques): Previous works on using second-order logic for
+program analysis use completely different approaches for encoding
+termination: [19] requires a dwf predicate in the logic to model
+termination, i.e., it uses a logic with a builtin halting oracle.  This
+makes the encoding trivial.  [20] reduces (non-)termination to a safety
+property and uses templates to reduce the resulting second-order formula to
+a first-order one.  The (non-)termination proofs take the form of
+preconditions and, in contrast to our work, there is no guarantee of
+completeness or even soundness.  We maintain that our encoding of
+(non-)termination as a second-order fomula is novel and non-trivial.  For
+example characterising termination of nested loops is not a straightforward
+extension of invariant generation: it requires relational reasoning (which
+is not the same class of problem as invariant generation -- an inherently
+non-relational problem); our use of Skolemisation to handle nondeterminism
+in non-termination arguments is novel and non-trivial; generating a
+tautology for which we find a constructive proof is novel and non-trivial.
 
-Reviewers 1, 3 and 4 (regarding experimental performance): ARMC is
-very good at handling loops with a small number of paths
-that have linear, rational termination arguments.
-We note that on harder examples ARMC performs poorly while Juggernaut
-excels, for example in the 5 benchmarks with nested loops, ARMC solves
-none correctly while Juggernaut solves 4 correctly.
-Since ARMC is solving a PTIME problem, it is not surprising that it
-is faster than our more general algorithm in the cases that are instances
-of the PTIME problem.  We also note that ARMC is not sound with respect
-to bitvector semantics: our experiments include 3 cases (6%) where
-ARMC terminates with an incorrect verdict.  Juggernaut does not
-give an incorrect verdict for any benchmark.  The paper did perhaps
-not make it clear that we are comparing the performance of a sound
-and complete algorithm (Juggernaut) with an unsound, incomplete
-algorithm (ARMC); this is the best comparison we could do,
-since there are no existing sound and complete algorithms to compare
-with.
+Reviewers 1, 3 and 4 (experimental performance): ARMC is very good at
+handling loops with a small number of paths that have linear, rational
+termination arguments.  This is a PTIME problem.  Consequently, on harder
+examples ARMC performs poorly while Juggernaut excels, for example in the 5
+benchmarks with nested loops, ARMC solves none while Juggernaut solves 4
+correctly.  Since ARMC is solving a PTIME problem, it is not surprising that
+it is faster than our more general algorithm in those cases that are
+instances of the PTIME problem.  We also note that ARMC is not sound with
+respect to bitvector semantics: our experiments include 3 cases (6%) where
+ARMC terminates with an incorrect verdict.  Juggernaut does not give an
+incorrect verdict for any benchmark.  The paper did perhaps not make it
+clear that we are comparing the performance of a sound and complete
+algorithm (Juggernaut) with an unsound, incomplete algorithm (ARMC); this is
+the best comparison we could do, since there are no existing sound and
+complete algorithms to compare with.
 
-Reviewers 2, 3 and 4 (regarding the simplicity of our approach): it is
-clear that reasoning about bitvector termination is difficult, as
-witnessed by the fact that ours is the first algorithm to decide it
-without explicitly checking the entire state space.
-This is despite bitvector programs having been written for almost 200 years. 
-We consider it the main strength of our work that
-we can use a sequence of simple reductions to transform a previously
-intractable problem into a simple one.
+Reviewers 2, 3 and 4 (simplicity of our approach): our algorithm is the
+first to solve many non-trivial termination problems.  It is also the first
+to decide termination of bit-vector and floating-point programs, and the
+first to combine both termination and non-termination analysis in one
+algorithm.  We consider it the main strength of our work that we can use a
+sequence of simple reductions to transform a previously intractable problem
+into a simple one.
 
-Reviewers 3 and 4 (regarding the comparison to SyGus): our specifications are C
-code, but SyGus specifications are S-Expressions.  This design decision
-allows us to use explicit state model checking and GP, as well as getting
-all of compiler theory for free.  This difference is non-trivial and
-novel.  However the underlying problem is clearly very similar and we
-appreciate it being brought to our attention so that we may cite it.
+Reviewers 3 and 4 (comparison to SyGus): our specifications are C code, but
+SyGus specifications are S-Expressions only.  This design decision enables
+us to use explicit-state model checking and GP, as well benefitting from
+compiler optimisations.  This difference is non-trivial and novel.  However
+the underlying problem is clearly very similar and we appreciate it being
+brought to our attention so that we may cite it.
 
-Reviewer 4 (regarding the importance of bitvector semantics): we agree
-that people really don't *think* about termination using bitvector
-arguments. If they did, our algorithm would have extremely limited value
-since people would already be soundly thinking about their own code.  The
-cases where a bitvector program's termination differs from its
-rational counterpart are where bugs are likely to appear.  Equally,
-a tool which generates a rational termination argument and checks whether
-it applies for bitvector semantics has limited utility: it is likely
-that if a rational argument applies, the programmer has already thought
-of it.
+Reviewer 4 (importance of bitvector semantics): we completely agree that
+people really don't *think* about termination using bitvector arguments. 
+However, they subsequently write a bitvector program, and the semantic gap
+is a frequent root-cause of serious bugs.  Equally, a tool that generates
+rational termination arguments and checks whether it applies for bitvector
+semantics has limited utility: programmers do like termination arguments
+based on integers, and will be disappointed that an analyser that uses
+rationals fails to find it. The only advantage of rational termination
+arguments is that they are cheap.
 
-Reviewer 3 (regarding the novelty compared to other synthesis algorithms):
-Our contribution is twofold: we first present a novel reduction from termination
-to synthesis, then we present an algorithm and tool for solving the synthesis formula
-The first part is unarguably significant and theoretically non-trivial (see
-previous comments).  The second part is a significant technical contribution
-in that it manages to synthesise C programs from a C specification, and
-is able to combine SAT solving, explicit state model checking and
-stochastic search.  No other program synthesiser does this.
+Reviewer 3 (novelty compared to other synthesis algorithms): Our
+contribution is twofold: we first present a novel reduction from termination
+to synthesis, then we present an algorithm and tool for solving the
+synthesis formula.  The first part is significant and theoretically
+non-trivial (see previous comments).  The second part is a significant
+technical contribution in that it manages to synthesise C programs from a C
+specification, and is able to combine SAT solving, explicit-state model
+checking and stochastic search.  No other program synthesiser does this.
 
-Reviewer 1: C- is distinct from L in that C- is the
-implementation language of part of our synthesiser, and also the language
-used for specifications.
-It really does not leave out any other C features.  L is the language
-in which ranking functions are written and is distinct from C-.
-While section 5.1 may seem excessive, we feel that many in the POPL
-community are unlikely to realise that bitvector termination has no
-simple relationship with rational termination: there are programs that
-terminate with bitvector semantics but don't with rationls *and vice
-versa*.  The ITE instruction is "if-then-else"; we agree this figure
-is somewhat opaque and will add some discussion.
-The runtime is predominantly a factor of the Kolmogorov complexity,
-but the problem is in fact NEXPTIME-complete in the Kolmogorov complexity,
-so even modest increases in complexity result in large jumps in runtime. 
-We very much appreciate the suggestions regarding the rest of the
-presentation and intend to incorporate them into the paper.
+Reviewer 1: C- is distinct from L in that C- is the implementation language
+of part of our synthesiser, and also the language used for specifications. 
+It really does not leave out any other C features.  L is the language in
+which ranking functions are written and is distinct from C-.  While section
+5.1 may seem excessive, we feel that many in the POPL community are unlikely
+to realise that bitvector termination has no simple relationship with
+rational termination: there are programs that terminate with bitvector
+semantics but don't with rationls *and vice versa*.  The ITE instruction is
+"if-then-else"; we agree this figure is somewhat opaque and will add some
+discussion.  The runtime is predominantly a factor of the Kolmogorov
+complexity, but the problem is in fact NEXPTIME-complete in the Kolmogorov
+complexity, so even modest increases in solution size may result in large
+jumps in runtime.  We very much appreciate the suggestions regarding the
+rest of the presentation and intend to incorporate them into the paper.
 
-Reviewer 2 (regarding our selection of benchmarks): We omitted the SVCOMP'15
-benchmarks that made use of arrays or recursion.  We do not have arrays in our
-logic and we had not implemented recursion in our frontend (although the latter
-can be syntactically rewritten to our input format).
+Reviewer 2 (selection of benchmarks): We omitted exactly those SVCOMP'15
+benchmarks that made use of arrays or recursion.  We do not have arrays in
+our logic and we had not implemented recursion in our frontend (although the
+latter can be syntactically rewritten to our input format).

### 2014-10-17 papers/termination/termination-esop/paper.tex
esop version

--- 2014-07-09 papers/termination/paper.tex
@@
-\documentclass[preprint]{sigplanconf}
+\documentclass[a4paper]{llncs}
+%\documentclass[preprint]{sigplanconf}
 
@@
 \usepackage{amssymb}
-\usepackage{amsthm}
+%\usepackage{amsthm}
 \usepackage{amsfonts}
@@
 
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{corollary}[theorem]{Corollary}
-\newtheorem{conjecture}[theorem]{Conjecture}
-
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
+%% \newtheorem{theorem}{Theorem}
+%% \newtheorem{lemma}[theorem]{Lemma}
+%% \newtheorem{corollary}[theorem]{Corollary}
+%% \newtheorem{conjecture}[theorem]{Conjecture}
+
+%% \theoremstyle{definition}
+%% \newtheorem{definition}[theorem]{Definition}
 
@@
 
-\conferenceinfo{POPL '15}{Month d--d, 20yy, City, ST, Country} 
-\copyrightyear{2015}
+%% \conferenceinfo{POPL '15}{Month d--d, 20yy, City, ST, Country} 
+%% \copyrightyear{2015}
 
@@
 \title{Synthesising Termination and Non-Termination Proofs for Bit-Vector Programs}
+
+\author{Cristina David\and Daniel Kroening\and Matt Lewis}
+\institute{University of Oxford}
 
@@
 
-Our approach is to phrase the termination and non-termination problems as
-second-order satisfaction.  We identify a fragment of second-order
-logic that is expressive enough to capture both problems, while being
-restrictive enough to admit effective solvers.  Subsequently, we propose a method for
-checking the satisfiability of a formula in this fragment by
-solving an isomorphic program synthesis problem
-(Section~\ref{sec:synthesis}).  Our method is sound and
-complete for bit-vector programs -- for any program, we will find
+We present a technique that treats uniformly linear and non-linear programs and
+it is not restricted to finding linear ranking functions, but can
+also compute lexicographic nonlinear ones. 
+Our approach is constraint-based and relies on second-order formulations 
+of termination and non-termination.
+The obvious issue is that,
+due to its expressiveness, second-order logic is very difficult to reason
+in, with many second-order theories becoming undecidable even when the
+corresponding first-order theory is decidable.
+To make solving our constraints tractable, we formulate termination and non-termination inside a fragment of second-order logic with restricted quantification, 
+for which we have built a solver in \cite{}.
+Our method is sound and
+complete for bit-vector programs -- for any program, we find
 a proof of either its termination or non-termination.
+
+% problems as
+%second-order satisfaction.  We identify a fragment of second-order
+%logic that is expressive enough to capture both problems, while being
+%restrictive enough to admit effective solvers.  Subsequently, we propose a method for
+%checking the satisfiability of a formula in this fragment by
+%solving an isomorphic program synthesis problem
+%(Section~\ref{sec:synthesis}).
 
@@
 
-\item We used program synthesis techniques to solve the second-order satisfaction
-problems generated by our encoding.  Our solver requires only a single call to a SAT solver to verify its
-proof.  It is not based on enumerating lassos, nor does it require a safety prover
-that can reason about transitive closure.
-
-\item Our solver is based on a series of source-to-source
-transformations.  As well as leading to a concise implementation, this
-allows us to leverage existing, mature verification tools.
+%% \item We used program synthesis techniques to solve the second-order satisfaction
+%% problems generated by our encoding.  Our solver requires only a single call to a SAT solver to verify its
+%% proof.  It is not based on enumerating lassos, nor does it require a safety prover
+%% that can reason about transitive closure.
+
+%% \item Our solver is based on a series of source-to-source
+%% transformations.  As well as leading to a concise implementation, this
+%% allows us to leverage existing, mature verification tools.
 
 \item We implemented our technique and tried it on a selection of programs
-handling both bit-vectors and floats.
+handling both bit-vectors and floats. In our implementation we made use of 
+a solver for second-order logic with restricted quantification that we have introduced in \cite{}.
 
@@
 \iffalse
+
 \section{Combinatorics of Finite Termination}
@@
 corresponding first-order theory is decidable.
-To make solving our constraints tractable, we identify a fragment of second-order logic
-with a constrained use of quantification that is expressive enough to encode
-both termination and non-termination of finite-state programs, while admitting a semi-decision procedure.
-We will suggestively refer to the fragment as the \emph{synthesis fragment}:
-
-\begin{definition}[Synthesis Fragment]
-A formula is in the \emph{synthesis fragment} iff it is of the form
-%
+In \cite{}, we have identified and built a solver for a fragment of second-order logic with restricted quantification, 
+which we call second-order SAT (see Definition~\ref{def:2sat}).
+
+\begin{definition}[Second-Order SAT]
+\label{def:2sat}
  \[
-  \exists \vec{P},~ \vec{x} . ~\forall~ \vec{y} .\, \sigma(\vec{P}, \vec{x}, \vec{y})
+  \exists S_1 \ldots S_m . Q_1 x_1 \ldots Q_n x_n . \sigma
  \]
-%
-where $\vec{P}$ ranges over functions, while $\vec{x}$ and $\vec{y}$ range over
-ground terms.  Function $\sigma: (X^n \times Y^m \to Z^k) \times X^n \times
-Y^m \to \mathbb{B}$ can be seen as a specification function that returns
-true iff the functions $\vec{P}$ compute appropriate outputs when fed the
-inputs $\vec{x}$ and $\vec{y}$.  The synthesis fragment is implicitly
-defined over some background first-order theory, and $\sigma$ must be
-computable in that theory.
-%
+ Where the $S_i$'s range over predicates,
+the $Q_i$'s are either $\exists$ or $\forall$,
+the $x_i$'s range over boolean values,
+ and $\sigma$ is a quantifier-free propositional formula
+ whose free variables are the $x_i$'s.  
+Each $S_i$ has an associated arity $\mathrm{ar}(S_i)$
+ and $S_i \in \mathbb{B}^{\mathrm{ar}(S_i)}$.  Note that 
+$Q_1 x_1 \ldots Q_n x_n . \sigma$
+ is an instance of first-order propositional SAT, i.e. QBF.
 \end{definition}
+
+%% To make solving our constraints tractable, we identify a fragment of second-order logic
+%% with a constrained use of quantification that is expressive enough to encode
+%% both termination and non-termination of finite-state programs, while admitting a semi-decision procedure.
+%% We will suggestively refer to the fragment as the \emph{synthesis fragment}:
+
+%% \begin{definition}[Synthesis Fragment]
+%% A formula is in the \emph{synthesis fragment} iff it is of the form
+%% %
+%%  \[
+%%   \exists \vec{P},~ \vec{x} . ~\forall~ \vec{y} .\, \sigma(\vec{P}, \vec{x}, \vec{y})
+%%  \]
+%% %
+%% where $\vec{P}$ ranges over functions, while $\vec{x}$ and $\vec{y}$ range over
+%% ground terms.  Function $\sigma: (X^n \times Y^m \to Z^k) \times X^n \times
+%% Y^m \to \mathbb{B}$ can be seen as a specification function that returns
+%% true iff the functions $\vec{P}$ compute appropriate outputs when fed the
+%% inputs $\vec{x}$ and $\vec{y}$.  The synthesis fragment is implicitly
+%% defined over some background first-order theory, and $\sigma$ must be
+%% computable in that theory.
+%% %
+%% \end{definition}
 %
@@
 fragment is semantically less restrictive than it looks.
-Checking satisfiability of a formula in the synthesis fragment corresponds to program synthesis and
-%Solving the synthesis problem 
-amounts to finding witnesses for the unknowns $\vec{P}$ and $\vec{x}$ that meet the specification
-for all $\vec{y}$. 
-If a pair $(\vec{P}, \vec{x})$ is a solution to the synthesis problem, then we write $(\vec{P}, \vec{x}) \models \sigma$.
-For the remainder of the presentation, we drop the vector notation and write $x$ for $\vec{x}$, with the understanding
-that all quantified variables range over vectors.
-
-In the rest of this section, we show that the synthesis fragment 
+%% Checking satisfiability of a formula in second-order SAT corresponds to program synthesis and
+%% amounts to finding witnesses for the unknowns $\vec{P}$ and $\vec{x}$ that meet the specification
+%% for all $\vec{y}$. 
+%% If a pair $(\vec{P}, \vec{x})$ is a solution to the synthesis problem, then we write $(\vec{P}, \vec{x}) \models \sigma$.
+%% For the remainder of the presentation, we drop the vector notation and write $x$ for $\vec{x}$, with the understanding
+%% that all quantified variables range over vectors.
+
+In the rest of this section, we show that second-order SAT 
 is expressive enough to encode both termination and non-termination. 
-Following our formulation as second-order satisfaction, 
-in Section~\ref{sec:synthesis} we present a solver for the synthesis fragment that models bit-accurate semantics.
+%% Following our formulation as second-order satisfaction, 
+%% in Section~\ref{sec:synthesis} we present a solver for the synthesis fragment that models bit-accurate semantics.
 
@@
 
-We will begin our discussion by showing how to encode in the synthesis fragment the
+We will begin our discussion by showing how to encode in second-order SAT the
 \mbox{(non-)termination} of a program consisting of a single loop with no nesting.
@@
 Returning to the program from Figure~\ref{fig:motivation.a}, we
-can see that the corresponding synthesis formula {\bf [UT]} is satisfiable,
+can see that the corresponding second-order SAT formula {\bf [UT]} is satisfiable,
 as witnessed by the function $R(x) = x$.  Thus, $R(x) = x$ constitutes a
@@
 introduce such an operator, we will characterise non-termination
-using the synthesis formula {\bf [ONT]} (Definition~\ref{def:ont}, Figure~\ref{fig:single_loop})
+using the second-order SAT formula {\bf [ONT]} (Definition~\ref{def:ont}, Figure~\ref{fig:single_loop})
 encoding the existence of an \emph{(open) recurrence set}, i.e.~a nonempty 
@@
 If this formula is satisfiable, $N$ is an open recurrence set for $L$, which proves
-$L$'s non-termination. The issue with this formula is the additional level of quantifier alternation as compared to the synthesis fragment
+$L$'s non-termination. The issue with this formula is the additional level of quantifier alternation as compared to second-order SAT
 (it is an $\exists \forall \exists$ formula).  To eliminate the innermost existential quantifier,
@@
 The existence of a closed recurrence set is equivalent to the satisfiability
-of formula {\bf [CNT]} in Definition~\ref{def:cnt}, which is already in the synthesis
-fragment without needing Skolemization.
+of formula {\bf [CNT]} in Definition~\ref{def:cnt}, which is already in second-order SAT without needing Skolemization.
 
@@
 $R_1$ and $R_2$ are ranking functions for $L_1$ and $L_2$, respectively.
-
-%% Note that despite its intricacy, the constraint system is still a formula in the synthesis fragment.
-%% Since the outer loop of this program does indeed terminate, the system is satisfiable, as witnessed by:
-%% \begin{align*}
-%% W_1(\langle v, x, y, z \rangle ) & \triangleq x \leq y + 1 \\
-%% W_2(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
-%% S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge \\
-%% & z' + v' = x' \wedge v' \leq y' \\
-%% R_2(x) & = x\\
-%% R_3 & = ..
-%% \end{align*}
 
@@
    \end{align*}
-% \begin{align*}
-%  \exists T_o, S, R_1, R_2 . \forall i, j, n, i', j', n' . \\
-%    %% x = 0 & \rightarrow W_1(v, x, y, z) \, \wedge \\
-%    %% x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
-%    %% x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x, y, z) \, \wedge \\
-%    S(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' \leq i' & \rightarrow S(\langle i, j, n \rangle, \langle i', j'+1, n' \rangle) \, \wedge \\
-%    & R_2(i,j,n) > 0 \, \wedge \\
-%    & R_2(i,j,n) > R_2(i',j'+1,n') \, \wedge\\
-%    i < n \wedge T_o(i, j, n) & \rightarrow S(\langle i, j, n \rangle, \langle i, 0, n \rangle) \, \wedge \\
-%    i < n \wedge T_o(i, j, n) \wedge S(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' > i' & \rightarrow \\
-%    & T_o(i'+1, j', n') \, \wedge \\
-%    & R_1(i, j, n) > 0 \, \wedge \\
-%    & R_1(i, j, n) > R_1(i'+1, j', n')
-% \end{align*}
 \end{minipage}
@@
 \end{figure*}
-
-%% \begin{figure*}
-%% \begin{framed}
-%% \begin{minipage}{0.145\textwidth}
-%% \begin{lstlisting}[mathescape=true]
-%% x = 0;
-%% $L_1:$
-%% while(x<=y)
-%%   x++;
-
-%% $L_2:$
-%% while(x>0){
-%%   z = x;
-%%   v = 0;
-
-%% $L_3:$
-%%   while(v<y){
-%%     z--;
-%%     v++;
-%%   }
-
-%%   x -= z;
-%%   y -= z;
-%% }
-%% \end{lstlisting}
-%% \end{minipage}
-%% \vline
-%% \begin{minipage}{0.85\textwidth}
-%% \begin{align*}
-%%  \exists W_1, W_2, R_2, R_3, S . \forall v, x, y, z, v', x', y', z' . \\
-%%    x = 0 & \rightarrow W_1(v, x, y, z) \, \wedge \\
-%%    x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
-%%    x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x, y, z) \, \wedge \\
-%%    S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' < y' & \rightarrow S(\langle v, x, y, z \rangle, \langle v'+1, x', y', z'-1 \rangle) \, \wedge \\
-%%    & R_3(v,x,y,z) > 0 \, \wedge \\
-%%    & R_3(v,x,y,z) > R_3(v'+1,x',y',z'-1) \, \wedge\\
-%%    x > 0 \wedge W_2(v, x, y, z) & \rightarrow S(\langle v, x, y, z \rangle, \langle 0, x, y, x \rangle) \, \wedge \\
-%%    x > 0 \wedge W_2(v, x, y, z) \wedge S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' \geq y' & \rightarrow \\
-%%    & W_2(v', x' - z', y' - z', z') \, \wedge \\
-%%    & R_2(v, x, y, z) > 0 \, \wedge \\
-%%    & R_2(v, x, y, z) > R_2(v', x' - z', y'-z', z')
-%% \end{align*}
-%% \end{minipage}
-%% \end{framed}
-%% \caption{A non-trivial program and its termination formula\label{fig:environment-model}}
-%% \end{figure*}
 
@@
 
-%%  \begin{definition}[Sequential Loops Termination Formula {\bf [ST]}]
-%%   \begin{align*}
-%% \label{def:multi-termination-formula}
-%%   \exists R, Inv_1,..., Inv_n . \forall x_0,...,x_n, x_1',...,x_n'.  & P_0(x_0,x_1) \rightarrow Inv_1(x_1) ~  \\
-%%  & \bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) ~  \\  
-%%  & \bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) \rightarrow Inv_{i+1}(x_{i+1})) ~ \wedge \\
-%%  & Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge G_n(x_n) \rightarrow Inv_n(x_n) ~ \wedge \\
-%%  & Inv_n(x_n) \wedge G_n(x_n) \wedge T(x_n, x_n') \rightarrow Inv_n(x_n') \wedge R(x_n) > 0 \wedge R(x_n) > R(x_n')
-%%  \end{align*}
-%% \end{definition}
-
-%% \begin{definition}[Nested Loops Termination Formula {\bf [NT]}]
-%% \label{def:nested-term-formula}
-%%  \begin{align*}
-%%   \exists R, S . \forall x, x' . & G_1(x) \wedge P_1(x,x') \rightarrow S(x',x') ~ \wedge \\
-%%                                 & G_2(x') \wedge S(x,x') \wedge T(x',x'')\rightarrow S(x,x'') ~ \wedge \\
-%% 				& G_1(x) \wedge P_1(x,x') \wedge \neg G_2(x') \wedge S(x',x'') \wedge 
-%%                                 P_2(x'',x''') \wedge 
-%%                                   R(x) > 0 \wedge R(x) > R(x''') 
-%%  \end{align*} 
-%% \end{definition}
-
-%% \begin{definition}[Sequential Loops Non-Termination Formula {\bf [SNT]}]
-%% \label{def:multi-termination-formula}
-%%  \begin{align*}
-%%   \exists R, Inv_1,..., Inv_n . \forall x_0,...,x_n, x_1',...,x_n'.  & P_0(x_0,x_1) \rightarrow Inv_1(x_1) ~  \\
-%%  & \bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) ~  \\  
-%%  & \bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) \rightarrow Inv_{i+1}(x_{i+1})) ~ \wedge \\
-%%  & Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge N(x_n)  ~ \wedge \\
-%%                                                         & N(x_n) \rightarrow G_n(x_n) ~ \wedge 
-%% 							 N(x_n) \wedge T_n(x_n, x_n') \rightarrow N(x_n') 
-%%  \end{align*}
-%% \end{definition}
  \end{framed}
@@
 
-\iffalse
 Let us assume that we have done some preprocessing of our program which has identified
@@
 and the nesting structure of the loops.
-\fi
 
@@
 
-%synthesizing environment abstractions: underapp for termination and underapp for non-termination.
-% \cite{} addresses the problem of automatically under-approximating weakest preconditions.
-
-%\noindent {\bf Nested loops.}
-%% If $L$'s body contains another loop $L'$, we cannot directly apply either formula {\bf [UT]} or
-%% {\bf [CT]} or prove that $L$ terminates.  This is because we don't have direct access to the
-%% single step transition relation for $L$'s body.  Instead, we consider $L'$ to be part of
-%% $L$'s environment, which we model with a relation $S$.  We require that $S$ over-approximates
-%% $L'$ in the sense that it satisfies the following formula:
-%% \begin{align*}
-%%  \forall x, x'. G(x) \wedge W(x) \wedge T'^*(x, x') \wedge \lnot G'(x') \rightarrow S(x, x')
-%% \end{align*}
-%% Where $G$ and $G'$ are respectively $L$ and $L'$'s guards, and $T'^*$ is the reflexive transitive closure of
-%% the inner loop ($L'$)'s transition relation.
-
-%% To give an example, if we wish to prove that the outer loop in Figure~\ref{fig:nested} terminates,
-%% it suffices to summarise the inner loop with the relation:
-%% \begin{align*}
-%% S(\langle x, y, z \rangle, \langle x', y', z' \rangle) & \triangleq x' = x \wedge y = 1
-%% \end{align*}
-%% Then the ranking function $R(x, y, z) = x$ proves termination of the outer loop $L(G, S)$.\\
-
-%% \begin{figure}
-%% \begin{lstlisting}
-%% while (x > 0) {
-%%   z = x, y = z + 1;
-%%   while (z > 0) {
-%%     z--, y--;
-%%   }
-%%   x -= y;
-%% }
-%% \end{lstlisting}
-%% \caption{A nested loop\label{fig:nested}}
-%% \end{figure}
-
-% \noindent {\bf Over-approximating the environment for termination.}
-%% While proving that loop $L$ terminates, we create constraints on the environment in the
-%% form of the predicate $I$ and the over-approximating summary relation $T_o$:
-%% In the presence of nested loops, 
-%% for each loop $L'$ nested in the body
-%% of $L$, we must generate constraints saying that $T_o$ over-approximates $L'$, i.e. it is an over-approximation of the transitive closure of $L'$'s transition relation.
-%% Similarly, for each loop $L''$ appearing before $L$ in the program's control flow, we
-%% must add constraints saying that $I$ over-approximates the states reachable via $L''$.
-%The details of how to create these constraints are standard, so
-%rather than presenting the algorithm for generating the constraints, we direct the
-%reader's attention to Figure~\ref{fig:environment-model}.  
-%
-% %\noindent {\bf Environment abstractions.}
-% For a loop $L=(G,T)$, the formulae {\bf [CT]} %and {\bf [CNT]} 
-% from Definition~\ref{def:ct} 
-% %and Definition~\ref{def:cnt} 
-% is in terms of 
-% $T_o$ and $I$, which over-approximate the transition relation~$T$ and the set of states that $L$ may begin executing in, respectively. 
-% The reason for using over-approximations [instead of exact relations] %when formulating the (non-)termination of an isolated loop 
-% is so that we can use the same formulae regardless of the environment the loop is executing in, or 
-% the complexity of its body, e.g.~if it contains nested loops.
-% 
-% \noindent {\bf Under-approximating the environment for non-termination.}
-
 \iffalse
@@
 
-Since $L$ either terminates or does not terminate, this formula is a tautology in the synthesis fragment.
+Since $L$ either terminates or does not terminate, this formula is a tautology in second-order SAT.
 A solution to the formula would include witnesses $P_N$ and $P_T$, which are putative proofs of non-termination
@@
 
-\subsection{The Synthesis Specification}
-
-The logic we use to express the synthesis formula, i.e.
-the generalised termination specification, is a subset of C that we call
-\newC.  The characteristic property of a \newC program is that safety can be
-decided for it using a single query to a Bounded Model Checker.  A~\newC
-program is just a C program with the following syntactic restrictions:
- all loops in the program must have a constant bound;
- all recursion in the program must be limited to a constant depth;
- all arrays must be statically allocated (i.e.~not using \texttt{malloc}),
- and be of constant size.
-Additionally, \newC programs may use nondeterministic values, assumptions
-and arbitrary-width types.
-
-\subsection{The Proof Language}
-
-The proofs generated by the synthesiser take the form of imperative programs
-written in the language $\mathcal{L}$.  An $\mathcal{L}$-program is a list
-of instructions, each of which matches one of the patterns shown in
-Figure~\ref{fig:l-language}.  An instruction has an opcode (such as
-\verb|add| for addition) and one or more operands.  An operand is either a
-constant, one of the program's inputs or the result of a previous
-instruction.  The $\mathcal{L}$ language has various arithmetic and logical
-operations, as well as basic branching in the form of the \verb|ite|
-(if-then-else) instruction.
-
-% The (non-)termination proofs are arbitrary expressions over the grammar given in Figure~\ref{fig:l-language}.  
-% %written in a simple RISC-like language $\mathcal{L}$, whose syntax is given in Fig.~\ref{fig:l-language}.  
-% \todo{update the figure and make it look more like a grammar.}
-
-\begin{figure}
- \begin{center}
- \resizebox{\linewidth}{!}{%
-  \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,
- semithick, initial text=]
-
-  \matrix[nodes={draw, fill=none, shape=rectangle, minimum height=1cm, minimum width=1.5cm},
-          row sep=2cm, column sep=2.5cm, ampersand replacement=\&] {
-   \node (synth) {Synthesise};
-   \&
-   \node (verif) {Verify}; %\\
-   %\node[draw=none] {};
-   \&
-   \node[ellipse] (done) {Done}; \\
-  };
-
-   \path
-    (synth) edge [bend left] node {Candidate program} (verif)
-    (verif) edge [bend left] node {Counterexample input} (synth)
-    (verif) edge node {Valid} (done);
- \end{tikzpicture}
- }
- \end{center}
-
- \caption{Abstract synthesis refinement loop
- \label{fig:abstract-refinement}}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-{\small
-
-\setlength{\tabcolsep}{14pt}
-Integer arithmetic instructions:
-
-\begin{tabular}{llll}
- \verb|add a b| & \verb|sub a b| & \verb|mul a b| & \verb|div a b| \\
- \verb|neg a| &   \verb|mod a b| & \verb|min a b| & \verb|max a b|
-\end{tabular}
-
-\medskip
-
-Bitwise logical and shift instructions:
-
-\begin{tabular}{lll}
- \verb|and  a b| & \verb|or   a b| & \verb|xor a b| \\
- \verb|lshr a b| & \verb|ashr a b| & \verb|not a|
-\end{tabular}
-
-\medskip
-
-Unsigned and signed comparison instructions:
-
-\begin{tabular}{lll}
- \verb|le  a b| & \verb|lt  a b| & \verb|sle  a b| \\
- \verb|slt a b| & \verb|eq  a b| & \verb|neq  a b| \\
-\end{tabular}
-
-\medskip
-
-Miscellaneous logical instructions:
-
-\begin{tabular}{lll}
- \verb|implies a b| & \verb|ite a b c| &  \\
-\end{tabular}
-
-\medskip
-
-\setlength{\tabcolsep}{12pt}
-
-Floating-point arithmetic:
-
-\begin{tabular}{llll}
- \verb|fadd a b| & \verb|fsub a b| & \verb|fmul a b| & \verb|fdiv a b|
-\end{tabular}
-
-}
-\end{center}
-
- \caption{The language $\mathcal{L}$}
- \label{fig:l-language}
-\end{figure}
-
-\subsection{The Synthesis Algorithm}
-
-We use Counterexample Guided Inductive Synthesis
-(CEGIS)~\cite{lezama-thesis,sketch, DBLP:conf/iclp/BrainCVF06} to find an
-$\mathcal{L}$-program satisfying our specification.  The core of the CEGIS
-algorithm is the refinement loop shown in Figure~\ref{fig:abstract-refinement}
-and detailed in Algorithm~\ref{alg:cegis}.
-
-%\begin{figure}
-\begin{algorithm}
-{\small
- \caption{Abstract synthesis algorithm
- \label{alg:cegis}}
- %\vspace{-1.7em}
- %\begin{multicols}{2}
- \begin{algorithmic}[1]
- \Statex
-\Function{synth}{{\sc inputs}, $\sigma$}
-  \Let{$(i_1, \ldots, i_N)$}{{\sc inputs}}
-  \Let{query}{$\exists P, x_0 . \forall i \in \text{\sc inputs} . \sigma(P, x_0, i)$}
-  \Let{result}{decide(query)}
-  \If{result.satisfiable}
-    \State \Return{result.model}
-  \Else
-    \State \Return{unsatisfiable}
-  \EndIf
-\EndFunction
-\Statex
-\Function{verif}{$P, x_0, \sigma$}
-  \Let{query}{$\exists x . \lnot \sigma(P, x_0, x)$}
-  \Let{result}{decide(query)}
-  \If{result.satisfiable}
-    \State \Return{result.model}
-  \Else
-    \State \Return{valid}
-  \EndIf
-\EndFunction
-%\columnbreak
-\Statex
-\Function{Synthesise}{$\sigma$}
-  \Let{{\sc inputs}}{$\emptyset$}
-  \Loop
-    \Let{candidate}{\Call{synth}{{\sc inputs}}}
-    \If{candidate = UNSAT}
-      \State \Return{unsatisfiable}
-    \EndIf
-    \Let{res}{\Call{verif}{candidate}}
-    \If{res = valid}
-      \State \Return{candidate}
-    \Else
-      \Let{{\sc inputs}}{{\sc inputs} $\cup$ res}
-    \EndIf
-  \EndLoop
-\EndFunction
- \end{algorithmic}
- %\end{multicols}
- }
-\end{algorithm}
-%\end{figure}
-
-The algorithm is divided into two procedures: {\sc synth} (see Figure~\ref{fig:synth-dfd}) and {\sc
-verif}, which interact via a finite set of test vectors {\sc inputs}.
-
-The {\sc synth} procedure tries to find existential witnesses $(P, x_0)$
-that satisfy the partial specification:
-%
-\[
- \exists P, x_0 . \forall x \in \text{\sc inputs} . \sigma(P, x_0, x)
-\]
-
-If {\sc synth} succeeds in finding a witness $(P, x_0)$, this witness is a
-candidate solution to the full synthesis formula.  We pass this candidate
-solution to {\sc verif} which determines whether it does satisfy
-the specification on all inputs by checking satisfiability of the
-verification formula:
-%
-\[
- \exists x . \lnot \sigma(P, x_0, x)
-\]
-
-If this formula is unsatisfiable, the candidate solution is in fact a
-solution to the synthesis formula and so the algorithm terminates. 
-Otherwise, the witness $x$ is an input on which the candidate solution fails
-to meet the specification.  This witness $x$ is added to the {\sc inputs}
-set and the loop iterates again.
-
-Concretely, {\sc synth} is implemented as shown in
-Figure~\ref{fig:synth-dfd}.  We start by taking the termination
-specification and generating a \newC program which takes as input a candidate
-$(P, x_0)$ and asserts that the candidate fails to meet the specification on
-at least one of the elements of {\sc inputs}.  So this program is unsafe iff
-there is some candidate $(P, x_0)$ satisfying the specification for all the
-elements of {\sc inputs}.  Finding a new candidate solution is therefore
-reduced to model checking this \newC program, which by construction is loop
-free.  The model checking is done using a combination of symbolic Bounded Model
-Checking (BMC), explicit state enumeration and genetic programming (GP) running in
-parallel.
-Finally, if any of the model checkers finds a candidate solution, it returns it
-and the others are killed.  The form of
-this candidate solution is yet another program, this time written in the
-proof language $\mathcal{L}$.
-
-% As well as efficiency, compiling the original source code of the program
-% under analysis has the advantage of guaranteeing that we are verifying the
-% exact semantics of the program as understood by a compiler, rather than some
-% ad-hoc abstraction of its semantics.
-
-Similarly to {\sc synth}, the {\sc verif} procedure is implemented by
-creating a \newC program that is unsafe iff some input exists on which the
-candidate solution fails to meet the specification.  Again, symbolic and
-explicit-state model checking are used to decide the satisfiability of the
-verification formula\footnote{GP is not used, since it is inherently incomplete.}.
-
-\begin{figure}
-\begin{center}
-\tikzstyle{file}=[draw, text width=7.0em, text centered,
-  minimum height=1.5em]
-\tikzstyle{process} = [draw, minimum height=3em, circle]
-\tikzstyle{line} = [draw, color=black, -latex']
-
-\def\Divide#1#2{%
- \coordinate(a) at ($(#1.east) !.5! (#2.west)$);
- \coordinate(b) at (a |- 0,-3);
- \draw[dotted] (b) -- ++(0, 4.5);
-}
-
-\resizebox{\linewidth}{!}{
-\begin{tikzpicture}[font=\sffamily]
-
-\node [file] (pua) {input.c};
-
-\path (pua.east)+(2,0) node [file] (spec) {\sc Termination Specification};
-\path (spec.south)+(0.0, -1) node [file] (inputs) {\sc Tracked Inputs};
-%% \path (synth.south)+(0.0, -0.5) node [file] (tests) {\sc tests.c};
-%% \path (tests.south)+(0.0, -0.5) node [file] (interpreter) {\sc interpreter.c};
-%% \path (interpreter.south)+(0.0, -0.5) node [file] (spec) {\sc spec.c};
-
-%% \path (spec.east)+(2.0, -0.25) node [process] (merged) {merge};
-\path (spec.east)+(2.0, -0.75) node [file] (file) {synth.c};
-
-\path (file.east)+(2.0, 1.5) node [process] (cbmc) {\sc cbmc};
-\path (file.east)+(2.0, 0.0) node [process] (gcc) {\sc Search};
-\path (file.east)+(2.0, -1.5) node [process] (gp) {\sc GP};
-
-\path (gcc.east)+(2.5, 0.0) node [file] (out) {Candidate Proof};
-
-%\path [line] (spec) -- (merged);
-
-\path [line] (pua) -- (spec);
-
-\path [line] (spec) -- (file);
-\path [line] (inputs) -- (file);
-
-\path [line] (file) -- (cbmc);
-\path [line] (file) -- (gcc);
-\path [line] (file) -- (gp);
-
-\path [line] (cbmc) -- (out);
-\path [line] (gcc) -- (out);
-\path [line] (gp) -- (out);
-
-\Divide{pua}{spec}{puaspec}
-%\Divide{spec}{file}{specfile}
-\Divide{file}{gcc}{filegcc}
-\Divide{gcc}{out}{gccout}
-
-\draw (pua |- 0,-3) node [align=center] {Input program \\ (written in C)};
-
-\coordinate (midspec) at ($(spec) !.5! (file)$);
-\draw (midspec |- 0,-3) node [align=center] {Automatically generated specification \\ (written in C)};
-%\draw (file |- 0,-3) node {Spec};
-\draw (gcc |- 0,-3) node [align=center] {Model checker};
-\draw (out |- 0,-3) node [align=center] {Proof \\ (written in $\mathcal{L}$)};
-
-\end{tikzpicture}
-}
-\end{center}
-
-\caption{Schematic diagram of {\sc synth}}
-\label{fig:synth-dfd}
-\end{figure}
-
-Now that we have a procedure for solving a formula
-in the synthesis fragment, we can build the full procedure for
-deciding termination of a loop $L$.  This is given as Algorithm~\ref{alg:termination}.
-
-\begin{algorithm}
-{\small
- \caption{Bit-vector termination
- \label{alg:termination}}
- \vspace{-.5em}
- \begin{algorithmic}[1]
- \Statex
-\Function{terminates}{$L$}
-  %\Let{termination-spec}{$\exists W, R. \forall x, x' . \phi(W, R, x, x')$} \Comment{Formula {\bf [CT]}}
-  %\Let{nonterm-spec}{$\exists N, C, x_0, \forall x. \psi(N, C, x_0, x)$} \Comment{Formula {\bf [SNT]}}
-  \Statex  \Comment{Formula {\bf[GT]}:}
-  \Let{spec}{$\exists P_T, P_N. \phi(P_T, x, x') \vee \psi(P_N, y)$}
-  \Let{$(P_T, P_N)$}{{\sc Synthesise}(spec)}
-  \If{decide($\exists x, x' . \lnot \phi(P_T, x, x'))$ = unsat}
-    \State \Return{$L$ terminates}
-  \ElsIf{decide($\exists x . \lnot \psi(P_N, x))$ = unsat}
-    \State \Return{$L$ does not terminate}
-  \EndIf
-\EndFunction
- \end{algorithmic}
- }
-\end{algorithm}
-
-\subsection{Candidate Generation Strategies}
-\noindent{\bf Explicit Proof Search.} The simplest strategy for finding candidates
-is to just exhaustively enumerate them all, starting with the shortest and
-progressively increasing the number of instructions.  This strategy
-is implemented by the {\sc ExplicitSearch} routine.  Since the set of
-$\mathcal{L}$-programs is recursively enumerable, this procedure is complete.
-\\
-
-\noindent{\bf Symbolic Bounded Model Checking.} Another complete method for generating
-candidates is to simply use BMC on the {\sc synth.c} program.  As with explicit
-search, we must progressively increase the length of the $\mathcal{L}$-program we search for
-in order to get a complete search procedure.
-\\
-
-\noindent {\bf Genetic Programming and Incremental Evolution.} \label{sec:gp}
-Our final strategy is genetic programming
-(GP)~\cite{langdon:fogp,brameier2007linear}.  GP provides an adaptive way of
-searching through the space of $\mathcal{L}$-programs for an individual
-that is ``fit'' in some sense.  We measure the
-fitness of an individual by counting the number of tests in {\sc inputs}
-for which it satisfies the specification.
-
-To bootstrap GP in the first iteration of the CEGIS loop, we generate a population
-of random $\mathcal{L}$-programs. We then iteratively evolve this population by
-applying the genetic operators {\sc crossover} and {\sc mutate}.
-{\sc Crossover} combines selected existing programs into new programs,
-whereas {\sc mutate} randomly changes parts of a single program.
-Fitter programs are more likely to be selected.
-
-% The {\sc crossover} operator takes two programs and combines their
-% code in some way to create a new program.  Programs are selected to be
-% {\sc crossover}ed according to their fitness -- fitter programs are more
-% likely to be selected to become parents.
-
-% The {\sc mutate} operator takes a single program and randomly changes parts
-% of its code, again creating a new program.
-
-Rather than generating a random population at the beginning of each subsequent
-iteration of the CEGIS loop, we start with the population we had at the end of the
-previous iteration.  The intuition here is that this population contained
-many individuals that performed well on the $k$ inputs we had before, so
-they will probably continue to perform well on the $k+1$ inputs we have now.
-In the parlance of evolutionary programming, this is known as
-incremental evolution~\cite{Gomez97incrementalevolution}.
-
-% In the setting of program synthesis, genetic programming can be very
-% usefully combined with the concept of incremental evolution.  The idea is
-% that in order to evolve a program that is adequate for some difficult task,
-% it is often helpful to first evolve a population that as an ensemble is good
-% at some simpler variation of the task.  This principle leads us to the
-% observation that each round of the CEGIS loop can be thought of as creating
-% a more difficult problem for the GP synthesiser to solve.  Therefore, if GP
-% finds a candidate program which is correct for the $k$ inputs so far
-% observed, but is not correct for the full specification, that candidate
-% program came from a population that was ``good'' for the specification
-% restrict to those $k$ inputs.  The next time round the CEGIS loop, the
-% problem GP must solve is slightly harder -- any solution to this new problem
-% must be correct for the previous $k$ inputs and also whatever new input was
-% added by the previous iteration of {\sc verif}.  It therefore makes sense to
-% begin the $k+1$st round of evolution with the population at the end of the
-% $k$-th round.
+%% \subsection{The Synthesis Specification}
+
+%% The logic we use to express the synthesis formula, i.e.
+%% the generalised termination specification, is a subset of C that we call
+%% \newC.  The characteristic property of a \newC program is that safety can be
+%% decided for it using a single query to a Bounded Model Checker.  A~\newC
+%% program is just a C program with the following syntactic restrictions:
+%%  all loops in the program must have a constant bound;
+%%  all recursion in the program must be limited to a constant depth;
+%%  all arrays must be statically allocated (i.e.~not using \texttt{malloc}),
+%%  and be of constant size.
+%% Additionally, \newC programs may use nondeterministic values, assumptions
+%% and arbitrary-width types.
+
+%% \subsection{The Proof Language}
+
+%% The proofs generated by the synthesiser take the form of imperative programs
+%% written in the language $\mathcal{L}$.  An $\mathcal{L}$-program is a list
+%% of instructions, each of which matches one of the patterns shown in
+%% Figure~\ref{fig:l-language}.  An instruction has an opcode (such as
+%% \verb|add| for addition) and one or more operands.  An operand is either a
+%% constant, one of the program's inputs or the result of a previous
+%% instruction.  The $\mathcal{L}$ language has various arithmetic and logical
+%% operations, as well as basic branching in the form of the \verb|ite|
+%% (if-then-else) instruction.
+
+%% % The (non-)termination proofs are arbitrary expressions over the grammar given in Figure~\ref{fig:l-language}.  
+%% % %written in a simple RISC-like language $\mathcal{L}$, whose syntax is given in Fig.~\ref{fig:l-language}.  
+%% % \todo{update the figure and make it look more like a grammar.}
+
+%% \begin{figure}
+%%  \begin{center}
+%%  \resizebox{\linewidth}{!}{%
+%%   \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,
+%%  semithick, initial text=]
+
+%%   \matrix[nodes={draw, fill=none, shape=rectangle, minimum height=1cm, minimum width=1.5cm},
+%%           row sep=2cm, column sep=2.5cm, ampersand replacement=\&] {
+%%    \node (synth) {Synthesise};
+%%    \&
+%%    \node (verif) {Verify}; %\\
+%%    %\node[draw=none] {};
+%%    \&
+%%    \node[ellipse] (done) {Done}; \\
+%%   };
+
+%%    \path
+%%     (synth) edge [bend left] node {Candidate program} (verif)
+%%     (verif) edge [bend left] node {Counterexample input} (synth)
+%%     (verif) edge node {Valid} (done);
+%%  \end{tikzpicture}
+%%  }
+%%  \end{center}
+
+%%  \caption{Abstract synthesis refinement loop
+%%  \label{fig:abstract-refinement}}
+%% \end{figure}
+
+%% \begin{figure}
+%% \begin{center}
+%% {\small
+
+%% \setlength{\tabcolsep}{14pt}
+%% Integer arithmetic instructions:
+
+%% \begin{tabular}{llll}
+%%  \verb|add a b| & \verb|sub a b| & \verb|mul a b| & \verb|div a b| \\
+%%  \verb|neg a| &   \verb|mod a b| & \verb|min a b| & \verb|max a b|
+%% \end{tabular}
+
+%% \medskip
+
+%% Bitwise logical and shift instructions:
+
+%% \begin{tabular}{lll}
+%%  \verb|and  a b| & \verb|or   a b| & \verb|xor a b| \\
+%%  \verb|lshr a b| & \verb|ashr a b| & \verb|not a|
+%% \end{tabular}
+
+%% \medskip
+
+%% Unsigned and signed comparison instructions:
+
+%% \begin{tabular}{lll}
+%%  \verb|le  a b| & \verb|lt  a b| & \verb|sle  a b| \\
+%%  \verb|slt a b| & \verb|eq  a b| & \verb|neq  a b| \\
+%% \end{tabular}
+
+%% \medskip
+
+%% Miscellaneous logical instructions:
+
+%% \begin{tabular}{lll}
+%%  \verb|implies a b| & \verb|ite a b c| &  \\
+%% \end{tabular}
+
+%% \medskip
+
+%% \setlength{\tabcolsep}{12pt}
+
+%% Floating-point arithmetic:
+
+%% \begin{tabular}{llll}
+%%  \verb|fadd a b| & \verb|fsub a b| & \verb|fmul a b| & \verb|fdiv a b|
+%% \end{tabular}
+
+%% }
+%% \end{center}
+
+%%  \caption{The language $\mathcal{L}$}
+%%  \label{fig:l-language}
+%% \end{figure}
+
+%% \subsection{The Synthesis Algorithm}
+
+%% We use Counterexample Guided Inductive Synthesis
+%% (CEGIS)~\cite{lezama-thesis,sketch,DBLP:conf/iclp/BrainCVF06} to find an
+%% $\mathcal{L}$-program satisfying our specification.  The core of the CEGIS
+%% algorithm is the refinement loop shown in Figure~\ref{fig:abstract-refinement}
+%% and detailed in Algorithm~\ref{alg:cegis}.
+
+%% %\begin{figure}
+%% \begin{algorithm}
+%% {\small
+%%  \caption{Abstract synthesis algorithm
+%%  \label{alg:cegis}}
+%%  %\vspace{-1.7em}
+%%  %\begin{multicols}{2}
+%%  \begin{algorithmic}[1]
+%%  \Statex
+%% \Function{synth}{{\sc inputs}, $\sigma$}
+%%   \Let{$(i_1, \ldots, i_N)$}{{\sc inputs}}
+%%   \Let{query}{$\exists P, x_0 . \forall i \in \text{\sc inputs} . \sigma(P, x_0, i)$}
+%%   \Let{result}{decide(query)}
+%%   \If{result.satisfiable}
+%%     \State \Return{result.model}
+%%   \Else
+%%     \State \Return{unsatisfiable}
+%%   \EndIf
+%% \EndFunction
+%% \Statex
+%% \Function{verif}{$P, x_0, \sigma$}
+%%   \Let{query}{$\exists x . \lnot \sigma(P, x_0, x)$}
+%%   \Let{result}{decide(query)}
+%%   \If{result.satisfiable}
+%%     \State \Return{result.model}
+%%   \Else
+%%     \State \Return{valid}
+%%   \EndIf
+%% \EndFunction
+%% %\columnbreak
+%% \Statex
+%% \Function{Synthesise}{$\sigma$}
+%%   \Let{{\sc inputs}}{$\emptyset$}
+%%   \Loop
+%%     \Let{candidate}{\Call{synth}{{\sc inputs}}}
+%%     \If{candidate = UNSAT}
+%%       \State \Return{unsatisfiable}
+%%     \EndIf
+%%     \Let{res}{\Call{verif}{candidate}}
+%%     \If{res = valid}
+%%       \State \Return{candidate}
+%%     \Else
+%%       \Let{{\sc inputs}}{{\sc inputs} $\cup$ res}
+%%     \EndIf
+%%   \EndLoop
+%% \EndFunction
+%%  \end{algorithmic}
+%%  %\end{multicols}
+%%  }
+%% \end{algorithm}
+%% %\end{figure}
+
+%% The algorithm is divided into two procedures: {\sc synth} (see Figure~\ref{fig:synth-dfd}) and {\sc
+%% verif}, which interact via a finite set of test vectors {\sc inputs}.
+
+%% The {\sc synth} procedure tries to find existential witnesses $(P, x_0)$
+%% that satisfy the partial specification:
+%% %
+%% \[
+%%  \exists P, x_0 . \forall x \in \text{\sc inputs} . \sigma(P, x_0, x)
+%% \]
+
+%% If {\sc synth} succeeds in finding a witness $(P, x_0)$, this witness is a
+%% candidate solution to the full synthesis formula.  We pass this candidate
+%% solution to {\sc verif} which determines whether it does satisfy
+%% the specification on all inputs by checking satisfiability of the
+%% verification formula:
+%% %
+%% \[
+%%  \exists x . \lnot \sigma(P, x_0, x)
+%% \]
+
+%% If this formula is unsatisfiable, the candidate solution is in fact a
+%% solution to the synthesis formula and so the algorithm terminates. 
+%% Otherwise, the witness $x$ is an input on which the candidate solution fails
+%% to meet the specification.  This witness $x$ is added to the {\sc inputs}
+%% set and the loop iterates again.
+
+%% Concretely, {\sc synth} is implemented as shown in
+%% Figure~\ref{fig:synth-dfd}.  We start by taking the termination
+%% specification and generating a \newC program which takes as input a candidate
+%% $(P, x_0)$ and asserts that the candidate fails to meet the specification on
+%% at least one of the elements of {\sc inputs}.  So this program is unsafe iff
+%% there is some candidate $(P, x_0)$ satisfying the specification for all the
+%% elements of {\sc inputs}.  Finding a new candidate solution is therefore
+%% reduced to model checking this \newC program, which by construction is loop
+%% free.  The model checking is done using a combination of symbolic Bounded Model
+%% Checking (BMC), explicit state enumeration and genetic programming (GP) running in
+%% parallel.
+%% Finally, if any of the model checkers finds a candidate solution, it returns it
+%% and the others are killed.  The form of
+%% this candidate solution is yet another program, this time written in the
+%% proof language $\mathcal{L}$.
+
+%% % As well as efficiency, compiling the original source code of the program
+%% % under analysis has the advantage of guaranteeing that we are verifying the
+%% % exact semantics of the program as understood by a compiler, rather than some
+%% % ad-hoc abstraction of its semantics.
+
+%% Similarly to {\sc synth}, the {\sc verif} procedure is implemented by
+%% creating a \newC program that is unsafe iff some input exists on which the
+%% candidate solution fails to meet the specification.  Again, symbolic and
+%% explicit-state model checking are used to decide the satisfiability of the
+%% verification formula\footnote{GP is not used, since it is inherently incomplete.}.
+
+%% \begin{figure}
+%% \begin{center}
+%% \tikzstyle{file}=[draw, text width=7.0em, text centered,
+%%   minimum height=1.5em]
+%% \tikzstyle{process} = [draw, minimum height=3em, circle]
+%% \tikzstyle{line} = [draw, color=black, -latex']
+
+%% \def\Divide#1#2{%
+%%  \coordinate(a) at ($(#1.east) !.5! (#2.west)$);
+%%  \coordinate(b) at (a |- 0,-3);
+%%  \draw[dotted] (b) -- ++(0, 4.5);
+%% }
+
+%% \resizebox{\linewidth}{!}{
+%% \begin{tikzpicture}[font=\sffamily]
+
+%% \node [file] (pua) {input.c};
+
+%% \path (pua.east)+(2,0) node [file] (spec) {\sc Termination Specification};
+%% \path (spec.south)+(0.0, -1) node [file] (inputs) {\sc Tracked Inputs};
+%% %% \path (synth.south)+(0.0, -0.5) node [file] (tests) {\sc tests.c};
+%% %% \path (tests.south)+(0.0, -0.5) node [file] (interpreter) {\sc interpreter.c};
+%% %% \path (interpreter.south)+(0.0, -0.5) node [file] (spec) {\sc spec.c};
+
+%% %% \path (spec.east)+(2.0, -0.25) node [process] (merged) {merge};
+%% \path (spec.east)+(2.0, -0.75) node [file] (file) {synth.c};
+
+%% \path (file.east)+(2.0, 1.5) node [process] (cbmc) {\sc cbmc};
+%% \path (file.east)+(2.0, 0.0) node [process] (gcc) {\sc Search};
+%% \path (file.east)+(2.0, -1.5) node [process] (gp) {\sc GP};
+
+%% \path (gcc.east)+(2.5, 0.0) node [file] (out) {Candidate Proof};
+
+%% %\path [line] (spec) -- (merged);
+
+%% \path [line] (pua) -- (spec);
+
+%% \path [line] (spec) -- (file);
+%% \path [line] (inputs) -- (file);
+
+%% \path [line] (file) -- (cbmc);
+%% \path [line] (file) -- (gcc);
+%% \path [line] (file) -- (gp);
+
+%% \path [line] (cbmc) -- (out);
+%% \path [line] (gcc) -- (out);
+%% \path [line] (gp) -- (out);
+
+%% \Divide{pua}{spec}{puaspec}
+%% %\Divide{spec}{file}{specfile}
+%% \Divide{file}{gcc}{filegcc}
+%% \Divide{gcc}{out}{gccout}
+
+%% \draw (pua |- 0,-3) node [align=center] {Input program \\ (written in C)};
+
+%% \coordinate (midspec) at ($(spec) !.5! (file)$);
+%% \draw (midspec |- 0,-3) node [align=center] {Automatically generated specification \\ (written in C)};
+%% %\draw (file |- 0,-3) node {Spec};
+%% \draw (gcc |- 0,-3) node [align=center] {Model checker};
+%% \draw (out |- 0,-3) node [align=center] {Proof \\ (written in $\mathcal{L}$)};
+
+%% \end{tikzpicture}
+%% }
+%% \end{center}
+
+%% \caption{Schematic diagram of {\sc synth}}
+%% \label{fig:synth-dfd}
+%% \end{figure}
+
+%% Now that we have a procedure for solving a formula
+%% in the synthesis fragment, we can build the full procedure for
+%% deciding termination of a loop $L$.  This is given as Algorithm~\ref{alg:termination}.
+
+%% \begin{algorithm}
+%% {\small
+%%  \caption{Bit-vector termination
+%%  \label{alg:termination}}
+%%  \vspace{-.5em}
+%%  \begin{algorithmic}[1]
+%%  \Statex
+%% \Function{terminates}{$L$}
+%%   %\Let{termination-spec}{$\exists W, R. \forall x, x' . \phi(W, R, x, x')$} \Comment{Formula {\bf [CT]}}
+%%   %\Let{nonterm-spec}{$\exists N, C, x_0, \forall x. \psi(N, C, x_0, x)$} \Comment{Formula {\bf [SNT]}}
+%%   \Statex  \Comment{Formula {\bf[GT]}:}
+%%   \Let{spec}{$\exists P_T, P_N. \phi(P_T, x, x') \vee \psi(P_N, y)$}
+%%   \Let{$(P_T, P_N)$}{{\sc Synthesise}(spec)}
+%%   \If{decide($\exists x, x' . \lnot \phi(P_T, x, x'))$ = unsat}
+%%     \State \Return{$L$ terminates}
+%%   \ElsIf{decide($\exists x . \lnot \psi(P_N, x))$ = unsat}
+%%     \State \Return{$L$ does not terminate}
+%%   \EndIf
+%% \EndFunction
+%%  \end{algorithmic}
+%%  }
+%% \end{algorithm}
+
+%% \subsection{Candidate Generation Strategies}
+%% \noindent{\bf Explicit Proof Search.} The simplest strategy for finding candidates
+%% is to just exhaustively enumerate them all, starting with the shortest and
+%% progressively increasing the number of instructions.  This strategy
+%% is implemented by the {\sc ExplicitSearch} routine.  Since the set of
+%% $\mathcal{L}$-programs is recursively enumerable, this procedure is complete.
+%% \\
+
+%% \noindent{\bf Symbolic Bounded Model Checking.} Another complete method for generating
+%% candidates is to simply use BMC on the {\sc synth.c} program.  As with explicit
+%% search, we must progressively increase the length of the $\mathcal{L}$-program we search for
+%% in order to get a complete search procedure.
+%% \\
+
+%% \noindent {\bf Genetic Programming and Incremental Evolution.} \label{sec:gp}
+%% Our final strategy is genetic programming
+%% (GP)~\cite{langdon:fogp,brameier2007linear}.  GP provides an adaptive way of
+%% searching through the space of $\mathcal{L}$-programs for an individual
+%% that is ``fit'' in some sense.  We measure the
+%% fitness of an individual by counting the number of tests in {\sc inputs}
+%% for which it satisfies the specification.
+
+%% To bootstrap GP in the first iteration of the CEGIS loop, we generate a population
+%% of random $\mathcal{L}$-programs. We then iteratively evolve this population by
+%% applying the genetic operators {\sc crossover} and {\sc mutate}.
+%% {\sc Crossover} combines selected existing programs into new programs,
+%% whereas {\sc mutate} randomly changes parts of a single program.
+%% Fitter programs are more likely to be selected.
+
+%% % The {\sc crossover} operator takes two programs and combines their
+%% % code in some way to create a new program.  Programs are selected to be
+%% % {\sc crossover}ed according to their fitness -- fitter programs are more
+%% % likely to be selected to become parents.
+
+%% % The {\sc mutate} operator takes a single program and randomly changes parts
+%% % of its code, again creating a new program.
+
+%% Rather than generating a random population at the beginning of each subsequent
+%% iteration of the CEGIS loop, we start with the population we had at the end of the
+%% previous iteration.  The intuition here is that this population contained
+%% many individuals that performed well on the $k$ inputs we had before, so
+%% they will probably continue to perform well on the $k+1$ inputs we have now.
+%% In the parlance of evolutionary programming, this is known as
+%% incremental evolution~\cite{Gomez97incrementalevolution}.
+
+%% % In the setting of program synthesis, genetic programming can be very
+%% % usefully combined with the concept of incremental evolution.  The idea is
+%% % that in order to evolve a program that is adequate for some difficult task,
+%% % it is often helpful to first evolve a population that as an ensemble is good
+%% % at some simpler variation of the task.  This principle leads us to the
+%% % observation that each round of the CEGIS loop can be thought of as creating
+%% % a more difficult problem for the GP synthesiser to solve.  Therefore, if GP
+%% % finds a candidate program which is correct for the $k$ inputs so far
+%% % observed, but is not correct for the full specification, that candidate
+%% % program came from a population that was ``good'' for the specification
+%% % restrict to those $k$ inputs.  The next time round the CEGIS loop, the
+%% % problem GP must solve is slightly harder -- any solution to this new problem
+%% % must be correct for the previous $k$ inputs and also whatever new input was
+%% % added by the previous iteration of {\sc verif}.  It therefore makes sense to
+%% % begin the $k+1$st round of evolution with the population at the end of the
+%% % $k$-th round.
 
@@
 %
-\begin{theorem}\label{thm:synth-sound}
-Algorithm~\ref{alg:cegis} is sound -- if it terminates with witness $(P, x_0)$, then
-$(P, x_0) \models \sigma$.
-\end{theorem}
-
-\begin{proof}
- The procedure {\sc Synthesise} terminates only if {\sc Verif} returns ``valid''.  In that
- case, $\exists x . \lnot \sigma(P, x_0, x)$ is unsatisfiable and so $\forall x . \sigma(P, x_0, x)$ holds.
-\end{proof}
-
-\begin{theorem}
- \label{thm:synth-semicomplete}
- If the existential first-order theory used to express the specification $\sigma$ is decidable
- and the domain of inputs $X$ is finite,
- Algorithm~\ref{alg:cegis} is semi-complete -- if a solution $(P, x_0) \models \sigma$
- exists then Algorithm~\ref{alg:cegis} will terminate.  However, if no program
- satisfies the specification, the algorithm may not terminate.
-\end{theorem}
-
-\begin{proof}
- If the domain $X$ is finite then the loop in procedure {\sc Synthesise} can only
- iterate $\| X \|$ times, since by this time all of the elements of $X$ would have been
- added to the inputs set.  Therefore if the {\sc Synth} procedure always terminates,
- Algorithm~\ref{alg:cegis} does as well.
-
- Since the {\sc ExplicitSearch} routine enumerates all programs (as can be seen by induction on
- the program length $l$), it will eventually enumerate a program that meets the specification
- on whatever set of inputs are currently being tracked, since by assumption such a program
- exists.  Since the first-order theory is
- decidable, the query in {\sc Verif} will succeed for this program, causing the algorithm to terminate.
- The set of correct programs is therefore recursively enumerable and Algorithm~\ref{alg:cegis}
- enumerates this set, so it is semi-complete.
-\end{proof}
-
-\begin{corollary}
-\label{cor:cegis-newc}
- Since safety of \newC programs is decidable, Algorithm~\ref{alg:cegis} is semi-complete
- when instantiated with \newC as a background theory.
-\end{corollary}
-
-% \begin{theorem}
-% \label{thm:sound}
-%  {\sc Headshot} is sound -- if it returns a verdict of termination, the program under
-%  analysis terminates on all inputs.  Similarly if it returns a verdict of non-termination,
-%  the program under analysis has at least one infinite execution.
-% \end{theorem}
-% 
-% \begin{proof}
-%  This is a direct consequence of Theorems~\ref{thm:ut}, \ref{thm:ct}, \ref{thm:ont}, \ref{thm:cnt} and \ref{thm:synth-sound}.
-% \end{proof}
-
-We will now show that $\mathcal{L}$ is expressive enough to encode
-\mbox{(non-)}ter\-mi\-nation proofs for every bit-vector program.
+%% \begin{theorem}\label{thm:synth-sound}
+%% Algorithm~\ref{alg:cegis} is sound -- if it terminates with witness $(P, x_0)$, then
+%% $(P, x_0) \models \sigma$.
+%% \end{theorem}
+
+%% \begin{proof}
+%%  The procedure {\sc Synthesise} terminates only if {\sc Verif} returns ``valid''.  In that
+%%  case, $\exists x . \lnot \sigma(P, x_0, x)$ is unsatisfiable and so $\forall x . \sigma(P, x_0, x)$ holds.
+%% \end{proof}
+
+%% \begin{theorem}
+%%  \label{thm:synth-semicomplete}
+%%  If the existential first-order theory used to express the specification $\sigma$ is decidable
+%%  and the domain of inputs $X$ is finite,
+%%  Algorithm~\ref{alg:cegis} is semi-complete -- if a solution $(P, x_0) \models \sigma$
+%%  exists then Algorithm~\ref{alg:cegis} will terminate.  However, if no program
+%%  satisfies the specification, the algorithm may not terminate.
+%% \end{theorem}
+
+%% \begin{proof}
+%%  If the domain $X$ is finite then the loop in procedure {\sc Synthesise} can only
+%%  iterate $\| X \|$ times, since by this time all of the elements of $X$ would have been
+%%  added to the inputs set.  Therefore if the {\sc Synth} procedure always terminates,
+%%  Algorithm~\ref{alg:cegis} does as well.
+
+%%  Since the {\sc ExplicitSearch} routine enumerates all programs (as can be seen by induction on
+%%  the program length $l$), it will eventually enumerate a program that meets the specification
+%%  on whatever set of inputs are currently being tracked, since by assumption such a program
+%%  exists.  Since the first-order theory is
+%%  decidable, the query in {\sc Verif} will succeed for this program, causing the algorithm to terminate.
+%%  The set of correct programs is therefore recursively enumerable and Algorithm~\ref{alg:cegis}
+%%  enumerates this set, so it is semi-complete.
+%% \end{proof}
+
+%% \begin{corollary}
+%% \label{cor:cegis-newc}
+%%  Since safety of \newC programs is decidable, Algorithm~\ref{alg:cegis} is semi-complete
+%%  when instantiated with \newC as a background theory.
+%% \end{corollary}
+
+%% % \begin{theorem}
+%% % \label{thm:sound}
+%% %  {\sc Headshot} is sound -- if it returns a verdict of termination, the program under
+%% %  analysis terminates on all inputs.  Similarly if it returns a verdict of non-termination,
+%% %  the program under analysis has at least one infinite execution.
+%% % \end{theorem}
+%% % 
+%% % \begin{proof}
+%% %  This is a direct consequence of Theorems~\ref{thm:ut}, \ref{thm:ct}, \ref{thm:ont}, \ref{thm:cnt} and \ref{thm:synth-sound}.
+%% % \end{proof}
+
+%% We will now show that $\mathcal{L}$ is expressive enough to encode
+%% \mbox{(non-)}ter\-mi\-nation proofs for every bit-vector program.
 
@@
 a finite $\mathcal{L}$-program computing $R''$.
-%
-% Since {\sc Kalashnikov} enumerates all programs in order of
-% length, it will eventually check a program computing each permutation of
-% $\mathcal{S}$ and so will eventually check a program computing the ranking
-% function $R''$.  Checking that a single program satisfies the specification
-% is a decidable problem, and the search as a whole terminates.
-%
 \end{proof}
@@
 
-We will now show that the number of iterations of the CEGIS loop is a function
-of the Kolmogorov complexity of the \mbox{(non-)}ter\-mi\-nation proof, and so
-our runtime is asymptotically a function of the Kolmogorov complexity.
-We argue that this gives our procedure various desirable qualities in practical
-applications.
-We first recall the definition of the Kolmogorov complexity
-of a function $f$:
-
-\begin{definition}[Kolmogorov complexity]
- The Kolmogorov complexity $K(f)$ is the length of the shortest program that
- computes~$f$.
-\end{definition}
-
-We can extend this definition slightly to talk about the Kolmogorov complexity of a
-synthesis problem in terms of its specification:
-
-\begin{definition}[Kolmogorov complexity of a synthesis problem]
- The Kolmogorov complexity of a program specification $K(\sigma)$ is the length of the shortest
- program $P$ such that $P \models \sigma$.
-\end{definition}
-
-Let us consider the number of iterations of the CEGIS loop $n$ required for a specification
-$\sigma$.  Since we enumerate candidate programs in order of length, we are always synthesising
-programs with length no greater than $K(\sigma)$ (since when we enumerate the first correct program,
-we will terminate).  So the space of solutions we search over is the space
-of functions computed by $\mathcal{L}$-programs of length no greater than $K(\sigma)$.  Let's
-denote this set $\mathcal{L}(K(\sigma))$.
-Since there are $O(2^{K(\sigma)})$ \emph{programs} of length $K(\sigma)$ and some functions
-will be computed by more than one program, we have $\| \mathcal{L}(K(\sigma)) \| \leq O(2^{K(\sigma)})$.
-
-Each iteration of the CEGIS loop distinguishes at least one incorrect function from the set of correct
-functions, so the loop will iterate no more than $\| \mathcal{L}(K(\sigma)) \|$ times.
-Therefore another bound on our runtime is:
-$$O\left (2^{K(\sigma)} \times NP \left ( K(\sigma) \times \| L \| \right) \right) = O \left(2^{K(\sigma)} \right)$$
-So for even moderately large $K(\sigma)$ our procedure is intractably slow.  Conversely
-if $K(\sigma)$ is small we will be able to find a termination proof, regardless of other parameters
-such as the size of the program under analysis.  To put this another way, under the assumption
-that $K(\sigma)$ is small for a program's termination problem, we will find a termination proof
-quite rapidly.
-
-To give a flavour for the nature of the ``low Kolmogorov complexity'' assumption, 
-we observe that linear functions have low Kolmogorov complexity (since a linear function can
-be computed with at most 1 instruction and 1 multiplication for each program variable),
-so the number of linear functions $f$ with $K(f) \leq \kappa$ grows linearly with $\kappa$.
-Conversely, the number of functions $f$ with $K(f) \leq \kappa$
-grows exponentially with $\kappa$.  This shows that assuming the existence of a linear
-ranking function is much stronger than assuming the existence of a short
-termination proof.  In addition to being a much weaker assumption,
-we argue that low Kolmogorov complexity is a more natural assumption than
-linearity.  Humans tend to write programs they can understand.  It's hard
-to quantify exactly what ``understandable'' means, but we think that
-``having a short proof'' is closer to the mark than ``having a linear proof''.
+%% We will now show that the number of iterations of the CEGIS loop is a function
+%% of the Kolmogorov complexity of the \mbox{(non-)}ter\-mi\-nation proof, and so
+%% our runtime is asymptotically a function of the Kolmogorov complexity.
+%% We argue that this gives our procedure various desirable qualities in practical
+%% applications.
+%% We first recall the definition of the Kolmogorov complexity
+%% of a function $f$:
+
+%% \begin{definition}[Kolmogorov complexity]
+%%  The Kolmogorov complexity $K(f)$ is the length of the shortest program that
+%%  computes~$f$.
+%% \end{definition}
+
+%% We can extend this definition slightly to talk about the Kolmogorov complexity of a
+%% synthesis problem in terms of its specification:
+
+%% \begin{definition}[Kolmogorov complexity of a synthesis problem]
+%%  The Kolmogorov complexity of a program specification $K(\sigma)$ is the length of the shortest
+%%  program $P$ such that $P \models \sigma$.
+%% \end{definition}
+
+%% Let us consider the number of iterations of the CEGIS loop $n$ required for a specification
+%% $\sigma$.  Since we enumerate candidate programs in order of length, we are always synthesising
+%% programs with length no greater than $K(\sigma)$ (since when we enumerate the first correct program,
+%% we will terminate).  So the space of solutions we search over is the space
+%% of functions computed by $\mathcal{L}$-programs of length no greater than $K(\sigma)$.  Let's
+%% denote this set $\mathcal{L}(K(\sigma))$.
+%% Since there are $O(2^{K(\sigma)})$ \emph{programs} of length $K(\sigma)$ and some functions
+%% will be computed by more than one program, we have $\| \mathcal{L}(K(\sigma)) \| \leq O(2^{K(\sigma)})$.
+
+%% Each iteration of the CEGIS loop distinguishes at least one incorrect function from the set of correct
+%% functions, so the loop will iterate no more than $\| \mathcal{L}(K(\sigma)) \|$ times.
+%% Therefore another bound on our runtime is:
+%% $$O\left (2^{K(\sigma)} \times NP \left ( K(\sigma) \times \| L \| \right) \right) = O \left(2^{K(\sigma)} \right)$$
+%% So for even moderately large $K(\sigma)$ our procedure is intractably slow.  Conversely
+%% if $K(\sigma)$ is small we will be able to find a termination proof, regardless of other parameters
+%% such as the size of the program under analysis.  To put this another way, under the assumption
+%% that $K(\sigma)$ is small for a program's termination problem, we will find a termination proof
+%% quite rapidly.
+
+%% To give a flavour for the nature of the ``low Kolmogorov complexity'' assumption, 
+%% we observe that linear functions have low Kolmogorov complexity (since a linear function can
+%% be computed with at most 1 instruction and 1 multiplication for each program variable),
+%% so the number of linear functions $f$ with $K(f) \leq \kappa$ grows linearly with $\kappa$.
+%% Conversely, the number of functions $f$ with $K(f) \leq \kappa$
+%% grows exponentially with $\kappa$.  This shows that assuming the existence of a linear
+%% ranking function is much stronger than assuming the existence of a short
+%% termination proof.  In addition to being a much weaker assumption,
+%% we argue that low Kolmogorov complexity is a more natural assumption than
+%% linearity.  Humans tend to write programs they can understand.  It's hard
+%% to quantify exactly what ``understandable'' means, but we think that
+%% ``having a short proof'' is closer to the mark than ``having a linear proof''.
 
@@
 functions for linear programs over integers (or
-rationals)~\cite{DBLP:conf/pldi/CookPR06, DBLP:conf/cav/LeeWY12,
-DBLP:conf/popl/Ben-AmramG13, DBLP:conf/vmcai/P04,
-DBLP:conf/atva/HeizmannHLP13, DBLP:conf/cav/BradleyMS05,
-DBLP:conf/tacas/CookSZ13, DBLP:conf/cav/KroeningSTW10}.  Among them, Lee et
+rationals)~\cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/KroeningSTW10}.
+Among them, Lee et
 al.~make use of transition predicate abstraction, algorithmic learning, and
@@
 execution proceeds through the loop and require expensive reachability
-analyses~\cite{DBLP:conf/lpe/CodishG03, DBLP:conf/lics/PodelskiR04,
-DBLP:conf/pldi/CookPR06}.  In an attempt to reduce the complexity of
+analyses~\cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}.  In an attempt to reduce the complexity of
 checking the validity of the termination argument, Cook et al.~present an
@@
 
-\bibliographystyle{abbrvnat}
+\bibliographystyle{splncs}
 \bibliography{synth}{}

### 2014-10-17 papers/termination/termination-esop/paper.tex
added back the proof language

@@
 
-%% \subsection{The Proof Language}
-
-%% The proofs generated by the synthesiser take the form of imperative programs
-%% written in the language $\mathcal{L}$.  An $\mathcal{L}$-program is a list
-%% of instructions, each of which matches one of the patterns shown in
-%% Figure~\ref{fig:l-language}.  An instruction has an opcode (such as
-%% \verb|add| for addition) and one or more operands.  An operand is either a
-%% constant, one of the program's inputs or the result of a previous
-%% instruction.  The $\mathcal{L}$ language has various arithmetic and logical
-%% operations, as well as basic branching in the form of the \verb|ite|
-%% (if-then-else) instruction.
-
-%% % The (non-)termination proofs are arbitrary expressions over the grammar given in Figure~\ref{fig:l-language}.  
-%% % %written in a simple RISC-like language $\mathcal{L}$, whose syntax is given in Fig.~\ref{fig:l-language}.  
-%% % \todo{update the figure and make it look more like a grammar.}
-
-%% \begin{figure}
-%%  \begin{center}
-%%  \resizebox{\linewidth}{!}{%
-%%   \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,
-%%  semithick, initial text=]
-
-%%   \matrix[nodes={draw, fill=none, shape=rectangle, minimum height=1cm, minimum width=1.5cm},
-%%           row sep=2cm, column sep=2.5cm, ampersand replacement=\&] {
-%%    \node (synth) {Synthesise};
-%%    \&
-%%    \node (verif) {Verify}; %\\
-%%    %\node[draw=none] {};
-%%    \&
-%%    \node[ellipse] (done) {Done}; \\
-%%   };
-
-%%    \path
-%%     (synth) edge [bend left] node {Candidate program} (verif)
-%%     (verif) edge [bend left] node {Counterexample input} (synth)
-%%     (verif) edge node {Valid} (done);
-%%  \end{tikzpicture}
-%%  }
-%%  \end{center}
-
-%%  \caption{Abstract synthesis refinement loop
-%%  \label{fig:abstract-refinement}}
-%% \end{figure}
-
-%% \begin{figure}
-%% \begin{center}
-%% {\small
-
-%% \setlength{\tabcolsep}{14pt}
-%% Integer arithmetic instructions:
-
-%% \begin{tabular}{llll}
-%%  \verb|add a b| & \verb|sub a b| & \verb|mul a b| & \verb|div a b| \\
-%%  \verb|neg a| &   \verb|mod a b| & \verb|min a b| & \verb|max a b|
-%% \end{tabular}
-
-%% \medskip
-
-%% Bitwise logical and shift instructions:
-
-%% \begin{tabular}{lll}
-%%  \verb|and  a b| & \verb|or   a b| & \verb|xor a b| \\
-%%  \verb|lshr a b| & \verb|ashr a b| & \verb|not a|
-%% \end{tabular}
-
-%% \medskip
-
-%% Unsigned and signed comparison instructions:
-
-%% \begin{tabular}{lll}
-%%  \verb|le  a b| & \verb|lt  a b| & \verb|sle  a b| \\
-%%  \verb|slt a b| & \verb|eq  a b| & \verb|neq  a b| \\
-%% \end{tabular}
-
-%% \medskip
-
-%% Miscellaneous logical instructions:
-
-%% \begin{tabular}{lll}
-%%  \verb|implies a b| & \verb|ite a b c| &  \\
-%% \end{tabular}
-
-%% \medskip
-
-%% \setlength{\tabcolsep}{12pt}
-
-%% Floating-point arithmetic:
-
-%% \begin{tabular}{llll}
-%%  \verb|fadd a b| & \verb|fsub a b| & \verb|fmul a b| & \verb|fdiv a b|
-%% \end{tabular}
-
-%% }
-%% \end{center}
-
-%%  \caption{The language $\mathcal{L}$}
-%%  \label{fig:l-language}
-%% \end{figure}
+\subsection{The Proof Language}
+
+The proofs generated by the synthesiser take the form of imperative programs
+written in the language $\mathcal{L}$.  An $\mathcal{L}$-program is a list
+of instructions, each of which matches one of the patterns shown in
+Figure~\ref{fig:l-language}.  An instruction has an opcode (such as
+\verb|add| for addition) and one or more operands.  An operand is either a
+constant, one of the program's inputs or the result of a previous
+instruction.  The $\mathcal{L}$ language has various arithmetic and logical
+operations, as well as basic branching in the form of the \verb|ite|
+(if-then-else) instruction.
+
+% The (non-)termination proofs are arbitrary expressions over the grammar given in Figure~\ref{fig:l-language}.  
+% %written in a simple RISC-like language $\mathcal{L}$, whose syntax is given in Fig.~\ref{fig:l-language}.  
+% \todo{update the figure and make it look more like a grammar.}
+
+\begin{figure}
+ \begin{center}
+ \resizebox{\linewidth}{!}{%
+  \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,
+ semithick, initial text=]
+
+  \matrix[nodes={draw, fill=none, shape=rectangle, minimum height=1cm, minimum width=1.5cm},
+          row sep=2cm, column sep=2.5cm, ampersand replacement=\&] {
+   \node (synth) {Synthesise};
+   \&
+   \node (verif) {Verify}; %\\
+   %\node[draw=none] {};
+   \&
+   \node[ellipse] (done) {Done}; \\
+  };
+
+   \path
+    (synth) edge [bend left] node {Candidate program} (verif)
+    (verif) edge [bend left] node {Counterexample input} (synth)
+    (verif) edge node {Valid} (done);
+ \end{tikzpicture}
+ }
+ \end{center}
+
+ \caption{Abstract synthesis refinement loop
+ \label{fig:abstract-refinement}}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+{\small
+
+\setlength{\tabcolsep}{14pt}
+Integer arithmetic instructions:
+
+\begin{tabular}{llll}
+ \verb|add a b| & \verb|sub a b| & \verb|mul a b| & \verb|div a b| \\
+ \verb|neg a| &   \verb|mod a b| & \verb|min a b| & \verb|max a b|
+\end{tabular}
+
+\medskip
+
+Bitwise logical and shift instructions:
+
+\begin{tabular}{lll}
+ \verb|and  a b| & \verb|or   a b| & \verb|xor a b| \\
+ \verb|lshr a b| & \verb|ashr a b| & \verb|not a|
+\end{tabular}
+
+\medskip
+
+Unsigned and signed comparison instructions:
+
+\begin{tabular}{lll}
+ \verb|le  a b| & \verb|lt  a b| & \verb|sle  a b| \\
+ \verb|slt a b| & \verb|eq  a b| & \verb|neq  a b| \\
+\end{tabular}
+
+\medskip
+
+Miscellaneous logical instructions:
+
+\begin{tabular}{lll}
+ \verb|implies a b| & \verb|ite a b c| &  \\
+\end{tabular}
+
+\medskip
+
+\setlength{\tabcolsep}{12pt}
+
+Floating-point arithmetic:
+
+\begin{tabular}{llll}
+ \verb|fadd a b| & \verb|fsub a b| & \verb|fmul a b| & \verb|fdiv a b|
+\end{tabular}
+
+}
+\end{center}
+
+ \caption{The language $\mathcal{L}$}
+ \label{fig:l-language}
+\end{figure}
 

### 2014-10-17 papers/termination/termination-esop/paper.tex
removed synthesis refinement loop figure

@@
 
-\begin{figure}
- \begin{center}
- \resizebox{\linewidth}{!}{%
-  \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,
- semithick, initial text=]
-
-  \matrix[nodes={draw, fill=none, shape=rectangle, minimum height=1cm, minimum width=1.5cm},
-          row sep=2cm, column sep=2.5cm, ampersand replacement=\&] {
-   \node (synth) {Synthesise};
-   \&
-   \node (verif) {Verify}; %\\
-   %\node[draw=none] {};
-   \&
-   \node[ellipse] (done) {Done}; \\
-  };
-
-   \path
-    (synth) edge [bend left] node {Candidate program} (verif)
-    (verif) edge [bend left] node {Counterexample input} (synth)
-    (verif) edge node {Valid} (done);
- \end{tikzpicture}
- }
- \end{center}
-
- \caption{Abstract synthesis refinement loop
- \label{fig:abstract-refinement}}
-\end{figure}
+%% \begin{figure}
+%%  \begin{center}
+%%  \resizebox{\linewidth}{!}{%
+%%   \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,
+%%  semithick, initial text=]
+
+%%   \matrix[nodes={draw, fill=none, shape=rectangle, minimum height=1cm, minimum width=1.5cm},
+%%           row sep=2cm, column sep=2.5cm, ampersand replacement=\&] {
+%%    \node (synth) {Synthesise};
+%%    \&
+%%    \node (verif) {Verify}; %\\
+%%    %\node[draw=none] {};
+%%    \&
+%%    \node[ellipse] (done) {Done}; \\
+%%   };
+
+%%    \path
+%%     (synth) edge [bend left] node {Candidate program} (verif)
+%%     (verif) edge [bend left] node {Counterexample input} (synth)
+%%     (verif) edge node {Valid} (done);
+%%  \end{tikzpicture}
+%%  }
+%%  \end{center}
+
+%%  \caption{Abstract synthesis refinement loop
+%%  \label{fig:abstract-refinement}}
+%% \end{figure}
 

### 2014-10-17 papers/termination/termination-esop/paper.tex
fixed soundness and complexity so that they don't refer to the synthesizer

@@
 addition in the Galois field $\mathrm{GF}(2^k)$.}.
+
+\begin{theorem}
+ \label{thm:synth-semicomplete}
+ If the existential first-order theory used to express the specification $\sigma$ is decidable
+ and the domain of inputs $X$ is finite,
+ the second-order SAT decision procedure is semi-complete -- if a solution $(P, x_0) \models \sigma$
+ exists then Algorithm~\ref{alg:cegis} will terminate.  However, if no program
+ satisfies the specification, the algorithm may not terminate.
+\end{theorem}
 
@@
 \section{Soundness, Completeness and Complexity}
-
-Termination for bit-vector programs is known to be
-PSPACE-complete~\cite{DBLP:conf/tacas/CookKRW10}.  We first record the
-expected results for soundness and completeness of our algorithm for
-deciding it.  Our strategy for proving soundness and completeness
-is first to prove that Algorithm~\ref{alg:cegis} (as instantiated
-with \newC as a background logic) is a semi-decision
-procedure.  Next, we will show that $\mathcal{L}$ is expressive enough
+The second-order SAT solver we are using is a semi-decision procedure -- for any
+satisfiable formula, the solver is guaranteed to find a satisfying assignment to
+all the second-order variables~\cite{kalashnikov}. 
+%% Termination for bit-vector programs is known to be
+%% PSPACE-complete~\cite{DBLP:conf/tacas/CookKRW10}.  We first record the
+%% expected results for soundness and completeness of our algorithm for
+%% deciding it.  Our strategy for proving soundness and completeness
+%% is first to prove that Algorithm~\ref{alg:cegis} (as instantiated
+%% with \newC as a background logic) is a semi-decision
+%% procedure.
+In this section, we show that $\mathcal{L}$ is expressive enough
 to compute (non-)termination proofs for every bit-vector program.
-We will then combine these results to show that the procedure as a whole
-terminates with a valid proof for every input program.
+By using this result, %We will then combine this result with the fact that the second-order SAT solver is a semi-decision procedure to 
+we then show that our technique terminates with a valid proof for every input program.
 %
@@
  \label{thm:generalised-sat}
- The generalised termination formula {\bf [GT]} for any loop $L$ is valid
+ The generalised termination formula {\bf [GT]} for any loop $L$ is a tautology
  when $P_N$ and $P_T$ range over \mbox{$\mathcal{L}$-computable} functions.
@@
 \begin{theorem}
- Algorithm~\ref{alg:termination} is sound and complete -- it terminates for all input loops $L$ with
+Our technique %Algorithm~\ref{alg:termination} 
+is sound and complete -- it terminates for all input loops $L$ with
  a correct termination verdict.
@@
 \begin{proof}
- By Theorem~\ref{thm:generalised-sat}, the specification \lstinline!spec! is valid, and hence satisfiable.  Corollary~\ref{cor:cegis-newc}
- shows that Algorithm~\ref{alg:cegis} is semi-complete, and so the call to {\sc Synthesise} terminates with a witness $(P_T, P_N)$.
- If $L$ terminates then $P_T$ is a termination proof (Theorem~\ref{thm:ct}).  Otherwise $P_N$ is a non-termination proof
- (Theorem~\ref{thm:snt}).  Exactly one of these cases holds, and the background theory \newC being decidable means that each
- call to \lstinline|decide()| will terminate, which allows us to determine which of $P_T$ and $P_N$ is a correct proof.
+ By Theorem~\ref{thm:generalised-sat}, the specification \lstinline!spec! is satisfiable.
+ In~\cite{kalashnikov}, we show that the second-order SAT solver is semi-complete, and so is guaranteed to find a satisfying
+ assignment for \lstinline!spec!. %, and so the call to {\sc Synthesise} terminates with a witness $(P_T, P_N)$.
+ If $L$ terminates then $P_T$ is a termination proof (Theorem~\ref{thm:ct}),
+ otherwise $P_N$ is a non-termination proof (Theorem~\ref{thm:snt}).  Exactly one of these purported proofs
+ will be valid, and since we can check each proof with a single call to a SAT solver we simply
+ test both and discard the one that is invalid.
 \end{proof}
 
-Hence, Algorithm~\ref{alg:termination} is sound and complete for bit-vector termination.
-
-Theorem~\ref{thm:synth-semicomplete} provides a time bound -- the CEGIS loop can iterate at most
-$2^{bk}$ times where $k$ is the number of variables used by $L$ and
-$b$ is the number of bits used to store each variable.
-The $i$th iteration of the loop requires solving 2 NP-complete problems of size $i \times l \times \| L\|$
-with $l$ the length of the program to be synthesised, and $\| L \| $ the size of the program-under-analysis
-computing $ L $.  So if
-the CEGIS loop iterates $n$ times, the
-runtime is bounded by
-$$2 \times n \times NP(n \times l \times \| L \|) \leq 2^{bk + 1} NP(2^{bk} \times l \times \| L \|)$$
-Clearly the runtime is affected most strongly by $n$, the total number of iterations of the CEGIS loop.
+%%---- Hence, Algorithm~\ref{alg:termination} is sound and complete for bit-vector termination.
+
+%% Theorem~\ref{thm:synth-semicomplete} provides a time bound -- the CEGIS loop can iterate at most
+%% $2^{bk}$ times where $k$ is the number of variables used by $L$ and
+%% $b$ is the number of bits used to store each variable.
+%% The $i$th iteration of the loop requires solving 2 NP-complete problems of size $i \times l \times \| L\|$
+%% with $l$ the length of the program to be synthesised, and $\| L \| $ the size of the program-under-analysis
+%% computing $ L $.  So if
+%% the CEGIS loop iterates $n$ times, the
+%% runtime is bounded by
+%% $$2 \times n \times NP(n \times l \times \| L \|) \leq 2^{bk + 1} NP(2^{bk} \times l \times \| L \|)$$
+%% Clearly the runtime is affected most strongly by $n$, the total number of iterations of the CEGIS loop.
 

### 2014-10-17 papers/termination/termination-esop/paper.tex
removed aspects related to kalashnikov from the experiments

@@
 
-%% \subsection{Candidate Generation Strategies}
-%% \noindent{\bf Explicit Proof Search.} The simplest strategy for finding candidates
-%% is to just exhaustively enumerate them all, starting with the shortest and
-%% progressively increasing the number of instructions.  This strategy
-%% is implemented by the {\sc ExplicitSearch} routine.  Since the set of
-%% $\mathcal{L}$-programs is recursively enumerable, this procedure is complete.
-%% \\
-
-%% \noindent{\bf Symbolic Bounded Model Checking.} Another complete method for generating
-%% candidates is to simply use BMC on the {\sc synth.c} program.  As with explicit
-%% search, we must progressively increase the length of the $\mathcal{L}$-program we search for
-%% in order to get a complete search procedure.
-%% \\
-
-%% \noindent {\bf Genetic Programming and Incremental Evolution.} \label{sec:gp}
-%% Our final strategy is genetic programming
-%% (GP)~\cite{langdon:fogp,brameier2007linear}.  GP provides an adaptive way of
-%% searching through the space of $\mathcal{L}$-programs for an individual
-%% that is ``fit'' in some sense.  We measure the
-%% fitness of an individual by counting the number of tests in {\sc inputs}
-%% for which it satisfies the specification.
-
-%% To bootstrap GP in the first iteration of the CEGIS loop, we generate a population
-%% of random $\mathcal{L}$-programs. We then iteratively evolve this population by
-%% applying the genetic operators {\sc crossover} and {\sc mutate}.
-%% {\sc Crossover} combines selected existing programs into new programs,
-%% whereas {\sc mutate} randomly changes parts of a single program.
-%% Fitter programs are more likely to be selected.
-
-%% % The {\sc crossover} operator takes two programs and combines their
-%% % code in some way to create a new program.  Programs are selected to be
-%% % {\sc crossover}ed according to their fitness -- fitter programs are more
-%% % likely to be selected to become parents.
-
-%% % The {\sc mutate} operator takes a single program and randomly changes parts
-%% % of its code, again creating a new program.
-
-%% Rather than generating a random population at the beginning of each subsequent
-%% iteration of the CEGIS loop, we start with the population we had at the end of the
-%% previous iteration.  The intuition here is that this population contained
-%% many individuals that performed well on the $k$ inputs we had before, so
-%% they will probably continue to perform well on the $k+1$ inputs we have now.
-%% In the parlance of evolutionary programming, this is known as
-%% incremental evolution~\cite{Gomez97incrementalevolution}.
-
-%% % In the setting of program synthesis, genetic programming can be very
-%% % usefully combined with the concept of incremental evolution.  The idea is
-%% % that in order to evolve a program that is adequate for some difficult task,
-%% % it is often helpful to first evolve a population that as an ensemble is good
-%% % at some simpler variation of the task.  This principle leads us to the
-%% % observation that each round of the CEGIS loop can be thought of as creating
-%% % a more difficult problem for the GP synthesiser to solve.  Therefore, if GP
-%% % finds a candidate program which is correct for the $k$ inputs so far
-%% % observed, but is not correct for the full specification, that candidate
-%% % program came from a population that was ``good'' for the specification
-%% % restrict to those $k$ inputs.  The next time round the CEGIS loop, the
-%% % problem GP must solve is slightly harder -- any solution to this new problem
-%% % must be correct for the previous $k$ inputs and also whatever new input was
-%% % added by the previous iteration of {\sc verif}.  It therefore makes sense to
-%% % begin the $k+1$st round of evolution with the population at the end of the
-%% % $k$-th round.
-
 \section{Soundness, Completeness and Complexity}
@@
 we then show that our technique terminates with a valid proof for every input program.
-%
-%% \begin{theorem}\label{thm:synth-sound}
-%% Algorithm~\ref{alg:cegis} is sound -- if it terminates with witness $(P, x_0)$, then
-%% $(P, x_0) \models \sigma$.
-%% \end{theorem}
-
-%% \begin{proof}
-%%  The procedure {\sc Synthesise} terminates only if {\sc Verif} returns ``valid''.  In that
-%%  case, $\exists x . \lnot \sigma(P, x_0, x)$ is unsatisfiable and so $\forall x . \sigma(P, x_0, x)$ holds.
-%% \end{proof}
-
-%% \begin{theorem}
-%%  \label{thm:synth-semicomplete}
-%%  If the existential first-order theory used to express the specification $\sigma$ is decidable
-%%  and the domain of inputs $X$ is finite,
-%%  Algorithm~\ref{alg:cegis} is semi-complete -- if a solution $(P, x_0) \models \sigma$
-%%  exists then Algorithm~\ref{alg:cegis} will terminate.  However, if no program
-%%  satisfies the specification, the algorithm may not terminate.
-%% \end{theorem}
-
-%% \begin{proof}
-%%  If the domain $X$ is finite then the loop in procedure {\sc Synthesise} can only
-%%  iterate $\| X \|$ times, since by this time all of the elements of $X$ would have been
-%%  added to the inputs set.  Therefore if the {\sc Synth} procedure always terminates,
-%%  Algorithm~\ref{alg:cegis} does as well.
-
-%%  Since the {\sc ExplicitSearch} routine enumerates all programs (as can be seen by induction on
-%%  the program length $l$), it will eventually enumerate a program that meets the specification
-%%  on whatever set of inputs are currently being tracked, since by assumption such a program
-%%  exists.  Since the first-order theory is
-%%  decidable, the query in {\sc Verif} will succeed for this program, causing the algorithm to terminate.
-%%  The set of correct programs is therefore recursively enumerable and Algorithm~\ref{alg:cegis}
-%%  enumerates this set, so it is semi-complete.
-%% \end{proof}
-
-%% \begin{corollary}
-%% \label{cor:cegis-newc}
-%%  Since safety of \newC programs is decidable, Algorithm~\ref{alg:cegis} is semi-complete
-%%  when instantiated with \newC as a background theory.
-%% \end{corollary}
-
-%% % \begin{theorem}
-%% % \label{thm:sound}
-%% %  {\sc Headshot} is sound -- if it returns a verdict of termination, the program under
-%% %  analysis terminates on all inputs.  Similarly if it returns a verdict of non-termination,
-%% %  the program under analysis has at least one infinite execution.
-%% % \end{theorem}
-%% % 
-%% % \begin{proof}
-%% %  This is a direct consequence of Theorems~\ref{thm:ut}, \ref{thm:ct}, \ref{thm:ont}, \ref{thm:cnt} and \ref{thm:synth-sound}.
-%% % \end{proof}
-
-%% We will now show that $\mathcal{L}$ is expressive enough to encode
-%% \mbox{(non-)}ter\-mi\-nation proofs for every bit-vector program.
 
@@
 
-%%---- Hence, Algorithm~\ref{alg:termination} is sound and complete for bit-vector termination.
-
-%% Theorem~\ref{thm:synth-semicomplete} provides a time bound -- the CEGIS loop can iterate at most
-%% $2^{bk}$ times where $k$ is the number of variables used by $L$ and
-%% $b$ is the number of bits used to store each variable.
-%% The $i$th iteration of the loop requires solving 2 NP-complete problems of size $i \times l \times \| L\|$
-%% with $l$ the length of the program to be synthesised, and $\| L \| $ the size of the program-under-analysis
-%% computing $ L $.  So if
-%% the CEGIS loop iterates $n$ times, the
-%% runtime is bounded by
-%% $$2 \times n \times NP(n \times l \times \| L \|) \leq 2^{bk + 1} NP(2^{bk} \times l \times \| L \|)$$
-%% Clearly the runtime is affected most strongly by $n$, the total number of iterations of the CEGIS loop.
-
-%% We will now show that the number of iterations of the CEGIS loop is a function
-%% of the Kolmogorov complexity of the \mbox{(non-)}ter\-mi\-nation proof, and so
-%% our runtime is asymptotically a function of the Kolmogorov complexity.
-%% We argue that this gives our procedure various desirable qualities in practical
-%% applications.
-%% We first recall the definition of the Kolmogorov complexity
-%% of a function $f$:
-
-%% \begin{definition}[Kolmogorov complexity]
-%%  The Kolmogorov complexity $K(f)$ is the length of the shortest program that
-%%  computes~$f$.
-%% \end{definition}
-
-%% We can extend this definition slightly to talk about the Kolmogorov complexity of a
-%% synthesis problem in terms of its specification:
-
-%% \begin{definition}[Kolmogorov complexity of a synthesis problem]
-%%  The Kolmogorov complexity of a program specification $K(\sigma)$ is the length of the shortest
-%%  program $P$ such that $P \models \sigma$.
-%% \end{definition}
-
-%% Let us consider the number of iterations of the CEGIS loop $n$ required for a specification
-%% $\sigma$.  Since we enumerate candidate programs in order of length, we are always synthesising
-%% programs with length no greater than $K(\sigma)$ (since when we enumerate the first correct program,
-%% we will terminate).  So the space of solutions we search over is the space
-%% of functions computed by $\mathcal{L}$-programs of length no greater than $K(\sigma)$.  Let's
-%% denote this set $\mathcal{L}(K(\sigma))$.
-%% Since there are $O(2^{K(\sigma)})$ \emph{programs} of length $K(\sigma)$ and some functions
-%% will be computed by more than one program, we have $\| \mathcal{L}(K(\sigma)) \| \leq O(2^{K(\sigma)})$.
-
-%% Each iteration of the CEGIS loop distinguishes at least one incorrect function from the set of correct
-%% functions, so the loop will iterate no more than $\| \mathcal{L}(K(\sigma)) \|$ times.
-%% Therefore another bound on our runtime is:
-%% $$O\left (2^{K(\sigma)} \times NP \left ( K(\sigma) \times \| L \| \right) \right) = O \left(2^{K(\sigma)} \right)$$
-%% So for even moderately large $K(\sigma)$ our procedure is intractably slow.  Conversely
-%% if $K(\sigma)$ is small we will be able to find a termination proof, regardless of other parameters
-%% such as the size of the program under analysis.  To put this another way, under the assumption
-%% that $K(\sigma)$ is small for a program's termination problem, we will find a termination proof
-%% quite rapidly.
-
-%% To give a flavour for the nature of the ``low Kolmogorov complexity'' assumption, 
-%% we observe that linear functions have low Kolmogorov complexity (since a linear function can
-%% be computed with at most 1 instruction and 1 multiplication for each program variable),
-%% so the number of linear functions $f$ with $K(f) \leq \kappa$ grows linearly with $\kappa$.
-%% Conversely, the number of functions $f$ with $K(f) \leq \kappa$
-%% grows exponentially with $\kappa$.  This shows that assuming the existence of a linear
-%% ranking function is much stronger than assuming the existence of a short
-%% termination proof.  In addition to being a much weaker assumption,
-%% we argue that low Kolmogorov complexity is a more natural assumption than
-%% linearity.  Humans tend to write programs they can understand.  It's hard
-%% to quantify exactly what ``understandable'' means, but we think that
-%% ``having a short proof'' is closer to the mark than ``having a linear proof''.
-
 \section{Experiments}
 
-To evaluate our algorithm, we implemented two tools: one to generate a
-termination specification from a C program, and one to synthesise a proof
-from the specification.  We ran the resulting termination prover,
+To evaluate our algorithm, we implemented a tool that generates a
+termination specification from a C program and calls the second-order SAT solver in \cite{kalashnikov} to obtain a proof.
+We ran the resulting termination prover,
 named {\sc Juggernaut}, on 47
 benchmarks taken from the literature and
-SV-COMP'15~\cite{svcomp15}\footnote{The logs and benchmarks are included in the
-non-anonymous supplementary material submitted with this paper}.%\footnote{The benchmarks, code and log files may be downloaded from the URL \texttt{http://www.cprover.org/juggernaut/}}.
-To
-provide a comparison point, we also ran {\sc ARMC}~\cite{armc-website} on
+SV-COMP'15~\cite{svcomp15}.
+To provide a comparison point, we also ran {\sc ARMC}~\cite{armc-website} on
 the same benchmarks.  Each tool was given a time limit of 180\,s, and was
@@
 
-The final column in Figure~\ref{fig:experiments} gives the number of
-iterations of the CEGIS loop our synthesiser performed.  While the
-theoretical bound for the number of CEGIS iterations is very large
-(exponential in the amount of space used by the program), these results
-demonstrate that in practice the number of iterations required is usually
-small.
+%% The final column in Figure~\ref{fig:experiments} gives the number of
+%% iterations of the CEGIS loop our synthesiser performed.  While the
+%% theoretical bound for the number of CEGIS iterations is very large
+%% (exponential in the amount of space used by the program), these results
+%% demonstrate that in practice the number of iterations required is usually
+%% small.
 
@@
 {\sc Juggernaut} handles the nonlinear cases correctly and rapidly.
-It solves 4 of the 5 nested loops in less than 30\,s\footnote{Although
-2 of these cases timed out with our default parameters, changing one command
-line flag allowed them to go through.  The benchmarks using non-default
-command line flags are marked with a *.}, but times out on the 5th.
-
-Overall, we notice that the GP component was always the fastest to find
-candidate solutions, {\sc ExplicitSearch} was always fastest
-to find counterexample inputs, and {\sc CBMC} was always fastest at
-proving correctness.  Each of these search methods has its strengths,
-and removing any one would have had a significant negative impact on
-overall performance.  Approximately 90\,\% of the total
-runtime was spent in synthesising new candidate solutions, which matches
-our expectations -- the (non-)termination proofs are difficult to
-find, but easy to check.
+It solves 4 of the 5 nested loops in less than 30\,s, but times out on the 5th.
+
+%% Overall, we notice that the GP component was always the fastest to find
+%% candidate solutions, {\sc ExplicitSearch} was always fastest
+%% to find counterexample inputs, and {\sc CBMC} was always fastest at
+%% proving correctness.  Each of these search methods has its strengths,
+%% and removing any one would have had a significant negative impact on
+%% overall performance.  Approximately 90\,\% of the total
+%% runtime was spent in synthesising new candidate solutions, which matches
+%% our expectations -- the (non-)termination proofs are difficult to
+%% find, but easy to check.
 
 In conclusion,
-these experiments confirm our conjecture that synthesis can be used
+these experiments confirm our conjecture that second-order SAT can be used
 effectively to prove termination and non-termination.  In particular,
 for programs with nested loops, nonlinear arithmetic and complex
-termination arguments, the versatility of unrestricted synthesis
+termination arguments, the versatility given by a general purpose solver %unrestricted termination arguments %synthesis
 is very valuable.
@@
 \centering
-\scriptsize
-\begin{tabular}{|l|@{}c@{}||@{}c@{}|r||@{}c@{}|r|r|}
+\small
+\begin{tabular}{|l|@{}c@{}||@{}c@{}|r||@{}c@{}|r|}
 \hline
-          &             & \multicolumn{2}{|c||}{\sc ARMC} & \multicolumn{3}{|c|}{\sc Juggernaut} \\
-Benchmark & \,Expected\, & \,Verdict\, & Time & \,Verdict\, & Time & \#It.\\
+          &             & \multicolumn{2}{|c||}{\sc ARMC} & \multicolumn{2}{|c|}{\sc Juggernaut} \\
+Benchmark & \,Expected\, & \,Verdict\, & Time & \,Verdict\, & Time \\
     \hline
@@
 Key: \tick = terminating, \xmark = non-terminating, ? = unknown (tool terminated with an inconclusive verdict).
-
-Times marked with a * were run with non-default command line flags.
 

### 2014-10-17 papers/termination/termination-esop/paper.tex
explanation on the choice of benchmarks
preliminaries

@@
 
+\subsection{Machine Arithmetic Vs.~Peano Arithmetic} \label{sec:machine.arith}
+
+Physical computers have bounded storage, which means they are unable to
+perform calculations on mathematical integers.  %% For example, if $A$ is the
+%% Ackermann function and $G$ is Graham's number, a physical computer capable
+%% of computing $A(G, G)$ would contain (much!) more matter than is believed to
+%% exist in the universe.  Fortunately, it is rare for a programmer to need
+%% such a large number and so modern computers
+They do their arithmetic over
+fixed-width binary words, otherwise known as bit-vectors.  For the remainder
+of this section, we will say that the bit-vectors we are working with are
+$k$-bits wide, which means that each word can hold one of $2^k$ bit
+patterns.  Typical values for $k$ are 32 and 64.
+
+Machine words can be interpreted as ``signed'' or ``unsigned'' values. 
+Signed values can be negative, while unsigned values cannot.  The encoding
+for signed values is two's complement, where the most significant bit
+$b_{k-1}$ of the word is a ``sign'' bit, whose weight is $-(2^k - 1)$ rather
+than $2^k - 1$.  Two's complement representation has the property that
+$\forall x .  -x = (\mathord{\sim} x) + 1$, where $\mathord{\sim}(\bullet)$
+is bitwise negation.  Two's complement also has the property that addition,
+multiplication and subtraction are defined identically for unsigned and
+signed numbers.
+
+Bit-vector arithmetic is performed modulo $2^k$, which is the source of many
+of the differences between machine arithmetic and Peano
+arithmetic\footnote{ISO C requires that unsigned arithmetic is performed
+modulo $2^k$, whereas the overflow case is undefined for signed arithmetic. 
+In practice, the undefined behaviour is implemented just as if the
+arithmetic had been unsigned.}.  To give an example, $(2^k - 1) + 1 \equiv 0
+\pmod {2^k}$ provides a counterexample to the statement $\forall x. 
+x + 1 > x$, which is a theorem of Peano arithmetic but not of modular
+arithmetic.  When an arithmetic operation has a result greater than $2^k$,
+it is said to ``overflow''.  If an operation does not overflow, its
+machine-arithmetic result is the same as the result of the same operation
+performed on integers.
+
+The final source of disagreement between integer arithmetic and bit-vector
+arithmetic stems from width conversions.  Many programming languages allow
+numeric variables of different types, which can be represented using words
+of different widths.  In C, a \texttt{short} might occupy 16 bits, while an
+\texttt{int} might occupy 32 bits.  When a $k$-bit variable is assigned to a
+$j$-bit variable with $j < k$, the result is truncated $\mathrm{mod}~2^j$.  For
+example, if $x$ is a 32-bit variable and $y$ is a 16-bit variable, $y$ will
+hold the value $0$ after the following code is executed:
+%
+\begin{lstlisting}
+x = 65536;
+y = x;
+\end{lstlisting}
+
+% This gives us a counterexample to the statement $\forall x, y . x = y
+% \Rightarrow (x + 1) = (y + 1)$, which is a theorem of Peano arithmetic.
+
+As well as machine arithmetic differing from Peano arithmetic on the
+operators they have in common, computers have several ``bitwise'' operations
+that are not taken as primitive in the theory of integers.  These operations
+include the Boolean operators \texttt{and, or, not, xor} applied to each
+element of the bit-vector.  Computer programs often make use of these
+operators, which are nonlinear when interpreted in the standard model of
+Peano arithmetic\footnote{Some of these operators can be seen as
+linear in a different algebraic structure, e.g.~\texttt{xor} corresponds to
+addition in the Galois field $\mathrm{GF}(2^k)$.}.
+
 \iffalse
@@
 machine arithmetic and conventional integer arithmetic.
-
-\subsection{Machine Arithmetic Vs.~Peano Arithmetic} \label{sec:machine.arith}
-
-Physical computers have bounded storage, which means they are unable to
-perform calculations on mathematical integers.  For example, if $A$ is the
-Ackermann function and $G$ is Graham's number, a physical computer capable
-of computing $A(G, G)$ would contain (much!) more matter than is believed to
-exist in the universe.  Fortunately, it is rare for a programmer to need
-such a large number and so modern computers do their arithmetic over
-fixed-width binary words, otherwise known as bit-vectors.  For the remainder
-of this section, we will say that the bit-vectors we are working with are
-$k$-bits wide, which means that each word can hold one of $2^k$ bit
-patterns.  Typical values for $k$ are 32 and 64.
-
-Machine words can be interpreted as ``signed'' or ``unsigned'' values. 
-Signed values can be negative, while unsigned values cannot.  The encoding
-for signed values is two's complement, where the most significant bit
-$b_{k-1}$ of the word is a ``sign'' bit, whose weight is $-(2^k - 1)$ rather
-than $2^k - 1$.  Two's complement representation has the property that
-$\forall x .  -x = (\mathord{\sim} x) + 1$, where $\mathord{\sim}(\bullet)$
-is bitwise negation.  Two's complement also has the property that addition,
-multiplication and subtraction are defined identically for unsigned and
-signed numbers.
-
-Bit-vector arithmetic is performed modulo $2^k$, which is the source of many
-of the differences between machine arithmetic and Peano
-arithmetic\footnote{ISO C requires that unsigned arithmetic is performed
-modulo $2^k$, whereas the overflow case is undefined for signed arithmetic. 
-In practice, the undefined behaviour is implemented just as if the
-arithmetic had been unsigned.}.  To give an example, $(2^k - 1) + 1 \equiv 0
-\pmod {2^k}$ provides a counterexample to the statement $\forall x. 
-x + 1 > x$, which is a theorem of Peano arithmetic but not of modular
-arithmetic.  When an arithmetic operation has a result greater than $2^k$,
-it is said to ``overflow''.  If an operation does not overflow, its
-machine-arithmetic result is the same as the result of the same operation
-performed on integers.
-
-The final source of disagreement between integer arithmetic and bit-vector
-arithmetic stems from width conversions.  Many programming languages allow
-numeric variables of different types, which can be represented using words
-of different widths.  In C, a \texttt{short} might occupy 16 bits, while an
-\texttt{int} might occupy 32 bits.  When a $k$-bit variable is assigned to a
-$j$-bit variable with $j < k$, the result is truncated $\mathrm{mod}~2^j$.  For
-example, if $x$ is a 32-bit variable and $y$ is a 16-bit variable, $y$ will
-hold the value $0$ after the following code is executed:
-%
-\begin{lstlisting}
-x = 65536;
-y = x;
-\end{lstlisting}
-
-% This gives us a counterexample to the statement $\forall x, y . x = y
-% \Rightarrow (x + 1) = (y + 1)$, which is a theorem of Peano arithmetic.
-
-As well as machine arithmetic differing from Peano arithmetic on the
-operators they have in common, computers have several ``bitwise'' operations
-that are not taken as primitive in the theory of integers.  These operations
-include the Boolean operators \texttt{and, or, not, xor} applied to each
-element of the bit-vector.  Computer programs often make use of these
-operators, which are nonlinear when interpreted in the standard model of
-Peano arithmetic\footnote{Some of these operators can be seen as
-linear in a different algebraic structure, e.g.~\texttt{xor} corresponds to
-addition in the Galois field $\mathrm{GF}(2^k)$.}.
 
@@
 SV-COMP'15~\cite{svcomp15}.
+We omitted exactly those SVCOMP'15
+benchmarks that made use of arrays or recursion. We do not have arrays in
+our logic and we had not implemented recursion in our frontend (although the
+latter can be syntactically rewritten to our input format).
+
 To provide a comparison point, we also ran {\sc ARMC}~\cite{armc-website} on

### 2014-10-17 papers/termination/termination-esop/paper.tex
added a section on solving the second order SAT formula and cited kalashnikov

@@
 
-\section{Satisfiability of a Synthesis Fragment Formula} \label{sec:synthesis}
-
-In this section we present a solver for the synthesis fragment:
-%
-\[
-  \exists P,~ x . ~\forall~ y . \sigma(P, x, y)
-\]
-%
-where $P$ ranges over functions, and $x$ and $y$ range over ground terms
-(as mentioned, we drop vector notation and write $x$ for $\vec{x}$).
-
-By viewing $\sigma: (X^n \times Y^m \to Z^k) \times X^n \times Y^m  \to
-\mathbb{B}$ as a specification function that returns true iff $P$ computes
-appropriate outputs for the inputs $x$ and $y$, we can regard the
-satisfaction problem for the synthesis fragment as \emph{program synthesis},
-i.e.~the mechanised construction of software that provably satisfies a
-given specification.  A witness $P$ to the satisfiability of some
-formula is then just a program that meets the specification $\sigma$.
-
-Research in the area of program synthesis has been
-particularly fruitful beginning with Alonzo Church's work on the
-\emph{Circuit Synthesis Problem} in the sixties~\cite{church-synth}, and
-continuing with works such as {\sc Brahma}~\cite{brahma} and program
-sketching~\cite{lezama-thesis,sketch,modular-sketch}.
-
-In the context of termination analysis, a witness to the
-satisfiability of a generalised termination specification is a
-proof of \mbox{(non-)termination}.  As mentioned earlier in the paper, our
-analysis must ensure precise treatment of bit-wise operations and arithmetic
-modulo fixed widths by modelling bit-accurate semantics.  The synthesiser
-must therefore be able to compute solutions over machine
-integers and floating-point numbers from the specifications detailed in
-Section~\ref{sec:second.order}.  In the next section, we will begin our treatment of
-bit-accurate synthesis by illustrating some of the differences between
-machine arithmetic and conventional integer arithmetic.
-
-\begin{theorem}
- \label{thm:synth-semicomplete}
- If the existential first-order theory used to express the specification $\sigma$ is decidable
- and the domain of inputs $X$ is finite,
- the second-order SAT decision procedure is semi-complete -- if a solution $(P, x_0) \models \sigma$
- exists then Algorithm~\ref{alg:cegis} will terminate.  However, if no program
- satisfies the specification, the algorithm may not terminate.
-\end{theorem}
-
-%% \subsection{The Synthesis Specification}
-
-%% The logic we use to express the synthesis formula, i.e.
-%% the generalised termination specification, is a subset of C that we call
-%% \newC.  The characteristic property of a \newC program is that safety can be
-%% decided for it using a single query to a Bounded Model Checker.  A~\newC
-%% program is just a C program with the following syntactic restrictions:
-%%  all loops in the program must have a constant bound;
-%%  all recursion in the program must be limited to a constant depth;
-%%  all arrays must be statically allocated (i.e.~not using \texttt{malloc}),
-%%  and be of constant size.
-%% Additionally, \newC programs may use nondeterministic values, assumptions
-%% and arbitrary-width types.
-
-\subsection{The Proof Language}
-
-The proofs generated by the synthesiser take the form of imperative programs
+\subsection{Solving the Second-Order SAT Formula}
+In order to solve the second-order generalised formula {\bf [GT]}, we use the solver described in \cite{kalashnikov}.
+%
+%% \begin{theorem}
+%%  \label{thm:synth-semicomplete}
+%%  If the existential first-order theory used to express the specification $\sigma$ is decidable
+%%  and the domain of inputs $X$ is finite,
+%%  the second-order SAT decision procedure is semi-complete -- if a solution $(P, x_0) \models \sigma$
+%%  exists then Algorithm~\ref{alg:cegis} will terminate.  However, if no program
+%%  satisfies the specification, the algorithm may not terminate.
+%% \end{theorem}
+%
+%
+This second-order SAT solver is a semi-decision procedure -- for any
+satisfiable formula, the solver is guaranteed to find a satisfying assignment to
+all the second-order variables.
+
+In the context of our termination analysis, such a satisfying assignment returned by the solver 
+represents either a proof of termination or non-termination, and
+takes the form of an imperative program
 written in the language $\mathcal{L}$.  An $\mathcal{L}$-program is a list
@@
 
-% The (non-)termination proofs are arbitrary expressions over the grammar given in Figure~\ref{fig:l-language}.  
-% %written in a simple RISC-like language $\mathcal{L}$, whose syntax is given in Fig.~\ref{fig:l-language}.  
-% \todo{update the figure and make it look more like a grammar.}
-
-%% \begin{figure}
-%%  \begin{center}
-%%  \resizebox{\linewidth}{!}{%
-%%   \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,
-%%  semithick, initial text=]
-
-%%   \matrix[nodes={draw, fill=none, shape=rectangle, minimum height=1cm, minimum width=1.5cm},
-%%           row sep=2cm, column sep=2.5cm, ampersand replacement=\&] {
-%%    \node (synth) {Synthesise};
-%%    \&
-%%    \node (verif) {Verify}; %\\
-%%    %\node[draw=none] {};
-%%    \&
-%%    \node[ellipse] (done) {Done}; \\
-%%   };
-
-%%    \path
-%%     (synth) edge [bend left] node {Candidate program} (verif)
-%%     (verif) edge [bend left] node {Counterexample input} (synth)
-%%     (verif) edge node {Valid} (done);
-%%  \end{tikzpicture}
-%%  }
-%%  \end{center}
-
-%%  \caption{Abstract synthesis refinement loop
-%%  \label{fig:abstract-refinement}}
-%% \end{figure}
-
 \begin{figure}
@@
 
-%% \subsection{The Synthesis Algorithm}
-
-%% We use Counterexample Guided Inductive Synthesis
-%% (CEGIS)~\cite{lezama-thesis,sketch,DBLP:conf/iclp/BrainCVF06} to find an
-%% $\mathcal{L}$-program satisfying our specification.  The core of the CEGIS
-%% algorithm is the refinement loop shown in Figure~\ref{fig:abstract-refinement}
-%% and detailed in Algorithm~\ref{alg:cegis}.
-
-%% %\begin{figure}
-%% \begin{algorithm}
-%% {\small
-%%  \caption{Abstract synthesis algorithm
-%%  \label{alg:cegis}}
-%%  %\vspace{-1.7em}
-%%  %\begin{multicols}{2}
-%%  \begin{algorithmic}[1]
-%%  \Statex
-%% \Function{synth}{{\sc inputs}, $\sigma$}
-%%   \Let{$(i_1, \ldots, i_N)$}{{\sc inputs}}
-%%   \Let{query}{$\exists P, x_0 . \forall i \in \text{\sc inputs} . \sigma(P, x_0, i)$}
-%%   \Let{result}{decide(query)}
-%%   \If{result.satisfiable}
-%%     \State \Return{result.model}
-%%   \Else
-%%     \State \Return{unsatisfiable}
-%%   \EndIf
-%% \EndFunction
-%% \Statex
-%% \Function{verif}{$P, x_0, \sigma$}
-%%   \Let{query}{$\exists x . \lnot \sigma(P, x_0, x)$}
-%%   \Let{result}{decide(query)}
-%%   \If{result.satisfiable}
-%%     \State \Return{result.model}
-%%   \Else
-%%     \State \Return{valid}
-%%   \EndIf
-%% \EndFunction
-%% %\columnbreak
-%% \Statex
-%% \Function{Synthesise}{$\sigma$}
-%%   \Let{{\sc inputs}}{$\emptyset$}
-%%   \Loop
-%%     \Let{candidate}{\Call{synth}{{\sc inputs}}}
-%%     \If{candidate = UNSAT}
-%%       \State \Return{unsatisfiable}
-%%     \EndIf
-%%     \Let{res}{\Call{verif}{candidate}}
-%%     \If{res = valid}
-%%       \State \Return{candidate}
-%%     \Else
-%%       \Let{{\sc inputs}}{{\sc inputs} $\cup$ res}
-%%     \EndIf
-%%   \EndLoop
-%% \EndFunction
-%%  \end{algorithmic}
-%%  %\end{multicols}
-%%  }
-%% \end{algorithm}
-%% %\end{figure}
-
-%% The algorithm is divided into two procedures: {\sc synth} (see Figure~\ref{fig:synth-dfd}) and {\sc
-%% verif}, which interact via a finite set of test vectors {\sc inputs}.
-
-%% The {\sc synth} procedure tries to find existential witnesses $(P, x_0)$
-%% that satisfy the partial specification:
-%% %
-%% \[
-%%  \exists P, x_0 . \forall x \in \text{\sc inputs} . \sigma(P, x_0, x)
-%% \]
-
-%% If {\sc synth} succeeds in finding a witness $(P, x_0)$, this witness is a
-%% candidate solution to the full synthesis formula.  We pass this candidate
-%% solution to {\sc verif} which determines whether it does satisfy
-%% the specification on all inputs by checking satisfiability of the
-%% verification formula:
-%% %
-%% \[
-%%  \exists x . \lnot \sigma(P, x_0, x)
-%% \]
-
-%% If this formula is unsatisfiable, the candidate solution is in fact a
-%% solution to the synthesis formula and so the algorithm terminates. 
-%% Otherwise, the witness $x$ is an input on which the candidate solution fails
-%% to meet the specification.  This witness $x$ is added to the {\sc inputs}
-%% set and the loop iterates again.
-
-%% Concretely, {\sc synth} is implemented as shown in
-%% Figure~\ref{fig:synth-dfd}.  We start by taking the termination
-%% specification and generating a \newC program which takes as input a candidate
-%% $(P, x_0)$ and asserts that the candidate fails to meet the specification on
-%% at least one of the elements of {\sc inputs}.  So this program is unsafe iff
-%% there is some candidate $(P, x_0)$ satisfying the specification for all the
-%% elements of {\sc inputs}.  Finding a new candidate solution is therefore
-%% reduced to model checking this \newC program, which by construction is loop
-%% free.  The model checking is done using a combination of symbolic Bounded Model
-%% Checking (BMC), explicit state enumeration and genetic programming (GP) running in
-%% parallel.
-%% Finally, if any of the model checkers finds a candidate solution, it returns it
-%% and the others are killed.  The form of
-%% this candidate solution is yet another program, this time written in the
-%% proof language $\mathcal{L}$.
-
-%% % As well as efficiency, compiling the original source code of the program
-%% % under analysis has the advantage of guaranteeing that we are verifying the
-%% % exact semantics of the program as understood by a compiler, rather than some
-%% % ad-hoc abstraction of its semantics.
-
-%% Similarly to {\sc synth}, the {\sc verif} procedure is implemented by
-%% creating a \newC program that is unsafe iff some input exists on which the
-%% candidate solution fails to meet the specification.  Again, symbolic and
-%% explicit-state model checking are used to decide the satisfiability of the
-%% verification formula\footnote{GP is not used, since it is inherently incomplete.}.
-
-%% \begin{figure}
-%% \begin{center}
-%% \tikzstyle{file}=[draw, text width=7.0em, text centered,
-%%   minimum height=1.5em]
-%% \tikzstyle{process} = [draw, minimum height=3em, circle]
-%% \tikzstyle{line} = [draw, color=black, -latex']
-
-%% \def\Divide#1#2{%
-%%  \coordinate(a) at ($(#1.east) !.5! (#2.west)$);
-%%  \coordinate(b) at (a |- 0,-3);
-%%  \draw[dotted] (b) -- ++(0, 4.5);
-%% }
-
-%% \resizebox{\linewidth}{!}{
-%% \begin{tikzpicture}[font=\sffamily]
-
-%% \node [file] (pua) {input.c};
-
-%% \path (pua.east)+(2,0) node [file] (spec) {\sc Termination Specification};
-%% \path (spec.south)+(0.0, -1) node [file] (inputs) {\sc Tracked Inputs};
-%% %% \path (synth.south)+(0.0, -0.5) node [file] (tests) {\sc tests.c};
-%% %% \path (tests.south)+(0.0, -0.5) node [file] (interpreter) {\sc interpreter.c};
-%% %% \path (interpreter.south)+(0.0, -0.5) node [file] (spec) {\sc spec.c};
-
-%% %% \path (spec.east)+(2.0, -0.25) node [process] (merged) {merge};
-%% \path (spec.east)+(2.0, -0.75) node [file] (file) {synth.c};
-
-%% \path (file.east)+(2.0, 1.5) node [process] (cbmc) {\sc cbmc};
-%% \path (file.east)+(2.0, 0.0) node [process] (gcc) {\sc Search};
-%% \path (file.east)+(2.0, -1.5) node [process] (gp) {\sc GP};
-
-%% \path (gcc.east)+(2.5, 0.0) node [file] (out) {Candidate Proof};
-
-%% %\path [line] (spec) -- (merged);
-
-%% \path [line] (pua) -- (spec);
-
-%% \path [line] (spec) -- (file);
-%% \path [line] (inputs) -- (file);
-
-%% \path [line] (file) -- (cbmc);
-%% \path [line] (file) -- (gcc);
-%% \path [line] (file) -- (gp);
-
-%% \path [line] (cbmc) -- (out);
-%% \path [line] (gcc) -- (out);
-%% \path [line] (gp) -- (out);
-
-%% \Divide{pua}{spec}{puaspec}
-%% %\Divide{spec}{file}{specfile}
-%% \Divide{file}{gcc}{filegcc}
-%% \Divide{gcc}{out}{gccout}
-
-%% \draw (pua |- 0,-3) node [align=center] {Input program \\ (written in C)};
-
-%% \coordinate (midspec) at ($(spec) !.5! (file)$);
-%% \draw (midspec |- 0,-3) node [align=center] {Automatically generated specification \\ (written in C)};
-%% %\draw (file |- 0,-3) node {Spec};
-%% \draw (gcc |- 0,-3) node [align=center] {Model checker};
-%% \draw (out |- 0,-3) node [align=center] {Proof \\ (written in $\mathcal{L}$)};
-
-%% \end{tikzpicture}
-%% }
-%% \end{center}
-
-%% \caption{Schematic diagram of {\sc synth}}
-%% \label{fig:synth-dfd}
-%% \end{figure}
-
-%% Now that we have a procedure for solving a formula
-%% in the synthesis fragment, we can build the full procedure for
-%% deciding termination of a loop $L$.  This is given as Algorithm~\ref{alg:termination}.
-
-%% \begin{algorithm}
-%% {\small
-%%  \caption{Bit-vector termination
-%%  \label{alg:termination}}
-%%  \vspace{-.5em}
-%%  \begin{algorithmic}[1]
-%%  \Statex
-%% \Function{terminates}{$L$}
-%%   %\Let{termination-spec}{$\exists W, R. \forall x, x' . \phi(W, R, x, x')$} \Comment{Formula {\bf [CT]}}
-%%   %\Let{nonterm-spec}{$\exists N, C, x_0, \forall x. \psi(N, C, x_0, x)$} \Comment{Formula {\bf [SNT]}}
-%%   \Statex  \Comment{Formula {\bf[GT]}:}
-%%   \Let{spec}{$\exists P_T, P_N. \phi(P_T, x, x') \vee \psi(P_N, y)$}
-%%   \Let{$(P_T, P_N)$}{{\sc Synthesise}(spec)}
-%%   \If{decide($\exists x, x' . \lnot \phi(P_T, x, x'))$ = unsat}
-%%     \State \Return{$L$ terminates}
-%%   \ElsIf{decide($\exists x . \lnot \psi(P_N, x))$ = unsat}
-%%     \State \Return{$L$ does not terminate}
-%%   \EndIf
-%% \EndFunction
-%%  \end{algorithmic}
-%%  }
-%% \end{algorithm}
-
 \section{Soundness, Completeness and Complexity}
-The second-order SAT solver we are using is a semi-decision procedure -- for any
-satisfiable formula, the solver is guaranteed to find a satisfying assignment to
-all the second-order variables~\cite{kalashnikov}. 
 %% Termination for bit-vector programs is known to be

### 2014-10-17 papers/termination/termination-esop/paper.tex
soundness - minor

@@
 \section{Soundness, Completeness and Complexity}
-%% Termination for bit-vector programs is known to be
-%% PSPACE-complete~\cite{DBLP:conf/tacas/CookKRW10}.  We first record the
-%% expected results for soundness and completeness of our algorithm for
-%% deciding it.  Our strategy for proving soundness and completeness
-%% is first to prove that Algorithm~\ref{alg:cegis} (as instantiated
-%% with \newC as a background logic) is a semi-decision
-%% procedure.
 In this section, we show that $\mathcal{L}$ is expressive enough
@@
 By using this result, %We will then combine this result with the fact that the second-order SAT solver is a semi-decision procedure to 
-we then show that our technique terminates with a valid proof for every input program.
+we then show that our analysis terminates with a valid proof for every input program.
 
@@
 \begin{theorem}
-Our technique %Algorithm~\ref{alg:termination} 
+Our termination analysis %Algorithm~\ref{alg:termination} 
 is sound and complete -- it terminates for all input loops $L$ with

### 2014-10-17 papers/termination/termination-esop/paper.tex
fixed figures

@@
 
-We present a reduction from program \emph{termination} to program
-\emph{synthesis} via a fragment of second-order logic.  This reduction
+We present a reduction from program \emph{termination} to 
+second-order satisfiability. 
+%program
+%\emph{synthesis} via a fragment of second-order logic.  
+We give formulations for termination and non-termination in a fragment of second-order logic 
+with restricted quantification.
+
+existential second-order logic over finite domains is decidable \cite{kalashnikov}.
+
+% decidable for finite-state programs
+
+This reduction
 allows us to generate nonlinear, lexicographic ranking functions and
@@
 \begin{figure*}
+\footnotesize{
 \centering
 \begin{tabular}{ccc}
-
 \begin{subfigure}[b]{0.3\textwidth}
+\small{
 \begin{lstlisting}
@@
  \label{fig:motivation.a}
+}
 \end{subfigure}%
@@
 \begin{subfigure}[b]{0.3\textwidth}
+
 \begin{lstlisting}
@@
 % \end{subfigure}
-
 \end{tabular}
+}
 \caption{Motivational examples, mostly taken from the literature.\label{fig:motivation}}
@@
 \begin{framed}
-\begin{minipage}{0.17\textwidth}
-\begin{lstlisting}[mathescape=true]
+%\resizebox{.25\textwidth}{!}{
+\begin{minipage}{.2\textwidth}
+\begin{lstlisting}[mathescape=true,basicstyle=\tiny]
 $L_1:$
@@
 \end{minipage}
+%}
 \vline
-\begin{minipage}{0.82\textwidth}
+\resizebox{.75\textwidth}{!}{
+\begin{minipage}{.85\textwidth}
  \begin{align*}
@@
 \end{minipage}
+}
 \end{framed}
@@
 \end{figure*}
+
+%% \begin{figure*}
+%% \begin{framed}
+%% \begin{minipage}{0.17\textwidth}
+%% \begin{lstlisting}[mathescape=true]
+%% $L_1:$
+%% while (i<n){
+%%   j = 0;
+
+%% $L_2:$
+%%   while (j$\leq$i){
+%%     j = j + 1;
+%%   }
+
+%%   i = i + 1;
+%% }
+%% \end{lstlisting}
+%% \end{minipage}
+%% \vline
+%% \begin{minipage}{0.82\textwidth}
+%% \begin{align*}
+%%  \exists T_o, R_1, R_2 . \forall i, j, n, i', j', n' . \\
+%%    i < n & \rightarrow T_o(\langle i, j, n \rangle , \langle i, 0, n \rangle) ~ \wedge \\
+%%    j \leq i \wedge T_o(\langle i', j', n' \rangle, \langle i, j, n\rangle ) & \rightarrow R_2(i, j, n) > 0 ~ \wedge \\
+%%      & R_2(i, j, n) > R_2(i, j+1, n) ~ \wedge \\
+%%      & T_o(\langle i', j', n' \rangle, \langle i, j+1, n \rangle) ~ \wedge \\
+%%    i < n \wedge S(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' > i' & \rightarrow R_1(i, j, n) > 0 ~ \wedge \\
+%%                                          & R_1(i, j, n) > R_1(i+1, j, n)
+%%    \end{align*}
+%% \end{minipage}
+%% \end{framed}
+%% \caption{A program with nested loops and its termination formula\label{fig:environment-model}}
+%% \end{figure*}
 
@@
 
+%% \begin{figure*}
+%% \begin{framed}
+%%  \begin{minipage}{0.16\textwidth}
+%% \begin{lstlisting}[mathescape=true]
+%% $L_1$:
+%% while ($G_1$) {
+%%   $P_1$;
+
+%% $L_2$:
+%%   while ($G_2$) {
+%%     $B_2$;
+%%   }
+
+%%   $P_2$;
+%% }
+%% \end{lstlisting}
+%% \end{minipage}
+%% \vline
+%% \begin{minipage}{0.82\textwidth}
+%% \begin{align*}
+%%  \exists N_1, N_2, x_0 . \forall x, x' . \\
+%%   N_1(x_0) & \, \wedge \\
+%%   N_1(x) & \rightarrow G_1(x) \, \wedge \\
+%%   N_1(x) \wedge P_1(x,x') & \rightarrow N_2(x') \, \wedge \\
+%%   G_2(x) \wedge N_2(x) \wedge B_2(x,x') & \rightarrow N_2(x') \, \wedge \\
+%%   \lnot G_2(x) \wedge N_2(x) \wedge P_2(x,x') & \rightarrow N_1(x') 
+%% \end{align*}
+%% \end{minipage}
+%% \end{framed}
+
+%% \caption{Formula encoding non-termination of nested loops \label{fig:nonterm-nested}}
+%% \end{figure*}
+
 \begin{figure*}
 \begin{framed}
- \begin{minipage}{0.16\textwidth}
-\begin{lstlisting}[mathescape=true]
+\begin{minipage}{0.2\textwidth}
+\begin{lstlisting}[mathescape=true,basicstyle=\tiny]
 $L_1$:
@@
 \vline
-\begin{minipage}{0.82\textwidth}
+\begin{minipage}{0.8\textwidth}
 \begin{align*}

### 2014-10-17 papers/termination/termination-esop/paper.tex
fixed the motivational examples figure
change title+keywords

@@
 
-\title{Synthesising Termination and Non-Termination Proofs for Bit-Vector Programs}
+\title{Termination and Non-Termination Proofs for Bit-Vector Programs}
 
@@
 \keywords
-Termination, Non-Termination, Program Synthesis,
+Termination, Non-Termination, 
 Lexicographic Ranking Functions, Bit-vector Ranking Functions, Floating-Point Ranking Functions.
@@
 
-\begin{figure*}
-\footnotesize{
+\begin{figure*}[h]
+\hspace*{-2.5cm}
 \centering
+%\resizebox{.95\textwidth}{!}{
+%\begin{minipage}{\linewidth}
 \begin{tabular}{ccc}
-\begin{subfigure}[b]{0.3\textwidth}
-\small{
+\begin{subfigure}[b]{0.45\textwidth}
 \begin{lstlisting}
@@
  \label{fig:motivation.a}
-}
 \end{subfigure}%
@@
 
-\begin{subfigure}[b]{0.3\textwidth}
+\begin{subfigure}[b]{0.45\textwidth}
 \begin{lstlisting}
@@
 
-\begin{subfigure}[b]{0.3\textwidth}
+\begin{subfigure}[b]{0.45\textwidth}
 \begin{lstlisting}[language=C]
@@
 
-\begin{subfigure}[b]{0.3\textwidth}
-
+\begin{subfigure}[b]{0.45\textwidth}
 \begin{lstlisting}
-while (i < M || j < N) {
+while (i<M || j<N) {
   i = i + 1;
@@
 \end{subfigure} 
-
 &
-
-\begin{subfigure}[b]{0.3\textwidth}
+\begin{subfigure}[b]{0.45\textwidth}
 \begin{lstlisting}
-while (x>0 && y>0 && z>0) {
+while (x>0 && y>0 && z>0){
   if (y > x) {
@@
 \end{subfigure} 
-% 
-% \\
-% 
-% \hline
-% 
-% &
-% 
-% \begin{subfigure}[b]{0.3\textwidth}
-% \begin{lstlisting}
-% while (i < n) {
-%   j = 0;
-%   while (j <= i) {
-%     j = j + 1;
-%   }
-%   i = i + 1;
-% }
-% \end{lstlisting}
-% \caption{Taken from~\cite{DBLP:conf/cav/BrockschmidtCF13}}
-% \end{subfigure}
 \end{tabular}
-}
+%}
 \caption{Motivational examples, mostly taken from the literature.\label{fig:motivation}}

### 2014-10-17 papers/termination/termination-esop/paper.tex
title change
abstract + fixes in introduction + kalashnikov citation

@@
 
-\title{Termination and Non-Termination Proofs for Bit-Vector Programs}
+\title{Unrestricted Termination and Non-Termination Arguments for Bit-Vector Programs}
 
@@
 operating on fixed-width machine integers.
-
-We present a reduction from program \emph{termination} to 
-second-order satisfiability. 
-%program
-%\emph{synthesis} via a fragment of second-order logic.  
-We give formulations for termination and non-termination in a fragment of second-order logic 
-with restricted quantification.
-
-existential second-order logic over finite domains is decidable \cite{kalashnikov}.
-
-% decidable for finite-state programs
-
-This reduction
-allows us to generate nonlinear, lexicographic ranking functions and
+We propose a termination analysis able 
+to generate nonlinear, lexicographic ranking functions and
 nonlinear recurrence sets that are correct for fixed-width machine arithmetic
-and floating-point arithmetic.  We use this reduction to build a sound and
-complete procedure for the termination analysis of finite-state programs with fixed-width
+and floating-point arithmetic
+Our technique is based on a reduction from program \emph{termination} to 
+second-order satisfaction. Thus, we provide 
+formulations for termination and non-termination in a fragment of second-order logic 
+with restricted quantification which was proven decidable over finite domains.
+%existential second-order logic over finite domains is decidable \cite{kalashnikov}. 
+The resulted technique is a sound and complete analysis for the termination 
+of finite-state programs with fixed-width
 integers and IEEE floating-point arithmetic.
@@
 To make solving our constraints tractable, we formulate termination and non-termination inside a fragment of second-order logic with restricted quantification, 
-for which we have built a solver in \cite{}.
+for which we have built a solver in \cite{kalashnikov}.
 Our method is sound and
@@
 \item  We rephrased the termination and non-termination problems as
-second-order $\exists \forall$ satisfaction problems.  This formulation
+second-order %$\exists \forall$ 
+satisfaction problems.  This formulation
 captures the (non-)termination
@@
 handling both bit-vectors and floats. In our implementation we made use of 
-a solver for second-order logic with restricted quantification that we have introduced in \cite{}.
+a solver for a fragment of second-order logic with restricted quantification that is decidable 
+over finite domains \cite{kalashnikov}.
 
@@
 corresponding first-order theory is decidable.
-In \cite{}, we have identified and built a solver for a fragment of second-order logic with restricted quantification, 
+In \cite{kalashnikov}, we have identified and built a solver for a fragment of second-order logic with restricted quantification, 
 which we call second-order SAT (see Definition~\ref{def:2sat}).

### 2014-10-17 papers/termination/termination-esop/paper.tex
spell chacking + abstract fixes

@@
 Our technique is based on a reduction from program \emph{termination} to 
-second-order satisfaction. Thus, we provide 
+second-order \emph{satisfaction}. We provide 
 formulations for termination and non-termination in a fragment of second-order logic 
-with restricted quantification which was proven decidable over finite domains.
+with restricted quantification which is decidable over finite domains \cite{kalashnikov}.
 %existential second-order logic over finite domains is decidable \cite{kalashnikov}. 
@@
 analysis~\cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/KroeningSTW10},
-requiring the support of powerful safety checkers: there is a tradeoff
+requiring the support of powerful safety checkers: there is a trade-off
 between the complexity of the termination arguments and that of checking
@@
 termination arguments.  This means that our ranking functions may involve
-nonlinearity and lexicographic orders: we do not commit to any particular
+non-linearity and lexicographic orders: we do not commit to any particular
 syntactic form, and do not use templates.  Furthermore, our approach allows
@@
 functions and recurrence sets that correctly accounts for the wrap-around
-behavior caused by under- and overflows in bit-vector and floating-point arithmetic.  Our
+behaviour caused by under- and overflows in bit-vector and floating-point arithmetic.  Our
 technique is not restricted to finding linear ranking functions, but can
@@
 feasibility~\cite{DBLP:conf/popl/GuptaHMRX08}.  Chen et al.~prove
-nontermination via reduction to safety
+non-termination via reduction to safety
 proving~\cite{DBLP:conf/tacas/ChenCFNO14}.  Their iterative algorithm uses

### 2014-10-18 papers/termination/termination-esop/paper.tex
Typos

@@
 
-We present a technique that treats uniformly linear and non-linear programs and
+We present a technique that treats linear and nonlinear programs uniformly and
 it is not restricted to finding linear ranking functions, but can
@@
 Each $S_i$ has an associated arity $\mathrm{ar}(S_i)$
- and $S_i \in \mathbb{B}^{\mathrm{ar}(S_i)}$.  Note that 
+ and $S_i \subseteq \mathbb{B}^{\mathrm{ar}(S_i)}$.  Note that 
 $Q_1 x_1 \ldots Q_n x_n . \sigma$
@@
 and termination respectively.  Exactly one of these will be a genuine proof, so we can check
-first one and then the other, which will be exemplified by Algorithm~\ref{alg:termination} once
-we have shown how to solve the formula {\bf [GT]}.
+first one and then the other.
 
@@
 %
-This second-order SAT solver is a semi-decision procedure -- for any
-satisfiable formula, the solver is guaranteed to find a satisfying assignment to
+For any satisfiable formula, the solver is guaranteed to find a satisfying assignment to
 all the second-order variables.
@@
 In this section, we show that $\mathcal{L}$ is expressive enough
-to compute (non-)termination proofs for every bit-vector program.
+to capture  (non-)termination proofs for every bit-vector program.
 By using this result, %We will then combine this result with the fact that the second-order SAT solver is a semi-decision procedure to 
@@
 It should be noted that the comparison here is imperfect, since {\sc ARMC}
-is solving a slightly different problem -- it checks whether the program
+is solving a different problem -- it checks whether the program
 under analysis would terminate if run with unbounded integer variables,

### 2014-11-13 slides/nus-talk.tex
NUS talk slides

--- 2014-06-18 slides/termination.tex
@@
 
+\newcommand{\hpathlen}{\ensuremath{\mathit{pathLength}}\xspace}
+\newcommand{\halloc}{\ensuremath{\mathit{new}}\xspace}
+\newcommand{\hassign}{\ensuremath{\mathit{assign}}\xspace}
+\newcommand{\hlookup}{\ensuremath{\mathit{lookup}}\xspace}
+\newcommand{\hupdate}{\ensuremath{\mathit{update}}\xspace}
+\newcommand{\halias}{\ensuremath{\mathit{alias}}\xspace}
+\newcommand{\hisnull}{\ensuremath{\mathit{isNull}}\xspace}
+\newcommand{\hispath}{\ensuremath{\mathit{isPath}}\xspace}
+\newcommand{\hcircular}{\ensuremath{\mathit{circular}}\xspace}
+%\newcommand{\id}{\ensuremath{\mathrm{id}}\xspace}
+\newcommand{\hsubdivide}{\ensuremath{\mathit{subdivide}}\xspace}
+\newcommand{\hfresh}{\ensuremath{\mathit{fresh}()}\xspace}
+
+\newcommand{\defeq}{\ensuremath{\stackrel{\mathrm{def}}{=}}}
+
+\newcommand{\kalashnikov}{{\sc Kalashnikov}\xspace}
+\newcommand{\shakira}{{\sc Shakira}\xspace}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-%\author{Cristina David \and Daniel Kroening \and Matt Lewis}
-\title{Everything You Ever Wanted To Prove About Termination But Were Afraid To Try}
+\author{Cristina David \and Daniel Kroening \and Matt Lewis}
+\title{Precise, Second-Order Verification}
 %\subtitle{}
-%\institute{Oxford University}
+\institute{Oxford University}
 \date{\today}
@@
 
+\iffalse
 \begin{frame}{Outline}
@@
 \end{frame}
+\fi
 
@@
 
+\begin{frame}[fragile]{Oh, And One More Thing}
+Does this terminate?  If so, what's the ranking function?
+
+\begin{lstlisting}[basicstyle=\normalsize]
+while (x > 0 && y > 0 && z > 0) {
+  if (y > x) {
+    y = z;
+    x = nondet();
+    z = x - 1;
+  } else {
+    z = z - 1;
+    x = nondet();
+    y = x - 1;
+  }
+}
+\end{lstlisting}
+\end{frame}
+
 \section{Advanced Topics}
@@
 
+\begin{frame}[fragile]{Heap}
+ \centering
+\begin{tikzpicture}
+  %\node at (-.75,.25) {$h:$};
+
+  \node (n1) {$\bullet$};
+  \node (n2) [below of=n1] {$\bullet$};
+  \node (n3) [right of=n1] {$\bullet$};
+  \node (n4) [right of=n3] {$\bullet$};
+  \node (n5) [right of=n4,above of=n4] {$\bullet$};
+  \node (n6) [below of=n4] {$\bullet$};
+  \node (x) [left of=n2] {$x$};
+  \node (y) [right of=n5] {$y$};
+  \node (z) [left  of=n1] {$z$};
+  \node (null) [right=2em of n6] {\bf null};
+
+  \path[draw,->] (n1) edge (n3)
+		 (n2) edge (n3)
+		 (n3) edge (n4)
+		 (n4) edge[bend right=45] (n5)
+		 (n5) edge[bend right=45] (n4);
+
+  \path[dashed,->] (x) edge (n2)%
+		   (y) edge (n5)%
+		   (z) edge (n1)%
+		   (null) edge (n6);
+ \end{tikzpicture}
+\end{frame}
+
+\begin{frame}[fragile]{Heap Transformers}
+ \centering
+ \begin{tikzpicture}
+  \node (n1) {$\bullet$};
+  \node [right of=n1] (n2) {$\bullet$};
+  \node (n3) [right of=n2] {$\bullet$};
+
+  \node (y) [below=1em of n1] {$y$};
+  \node (x) [right of=y] {$x$};
+  \node (null) [right of=x] {\bf null};
+
+  \path[draw, ->] (n1) edge (n2) (n2) edge (n3);
+  \path[dashed,->] (x) edge (n3) (y) edge (n1) (null) edge (n3);
+
+  \node (mid) [below right = .5em and 1em of n3] {$\Rightarrow$};
+
+  \node [above right = .5em and 1em of mid] (m1) {$\bullet$};
+  \node (m2) [right of = m1] {$\bullet$};
+  \node (m3) [right of = m2] {$\bullet$};
+
+  \node (y) [below=1em of m1] {$y$};
+  \node (x) [right of=y] {$x$};
+  \node (null) [right of=x] {\bf null};
+
+  \path[draw, ->] (m1) edge (m2) (m2) edge (m3);
+  \path[dashed,->] (x) edge (m2) (y) edge (m1) (null) edge (m3);
+ \end{tikzpicture}
+ }
+
+ \vfill
+ $\hlookup(h, x, y)$ \\
+ $x = y{\rightarrow}next;$
+\end{frame}
+
 \begin{frame}[fragile]{Fin}

### 2014-11-13 slides/nus-talk.tex
NUS slides

@@
 \lstset{language=C,
- basicstyle=\tiny
+ basicstyle=\tiny,
+ mathescape=true
  }
@@
 \definecolor{light-gray}{gray}{0.80}
+
+\usepackage{amsmath}
+\usepackage{xspace}
 
@@
 
-\iffalse
 \begin{frame}{Outline}
@@
 \end{frame}
-\fi
 
-\section{Motivation}
+\section{Program Synthesis and Second-Order Logic}
+
+\section{Termination}
 
@@
 
-\begin{frame}[fragile]{Heap}
- \centering
+\section{Heap}
+
+\begin{frame}[fragile]{Heaps of Singly Linked Lists}
+We would like to use second-order logic to analyse programs using the heap.  To do so, we need a
+heap theory with a decision procedure in NP that we can combine with arithmetic.
+
+\pause
+
+We will restrict ourselves to singly linked lists with lengths, but no data.
+
+\pause
+
+Lists can have (unspecified) cycles and sharing, for example:
+
+\begin{center}
 \begin{tikzpicture}
@@
  \end{tikzpicture}
+\end{center}
 \end{frame}
@@
 \begin{frame}[fragile]{Heap Transformers}
- \centering
+\only<1>{
+x = y$\rightarrow$next
+
+\begin{center}
  \begin{tikzpicture}
@@
 \end{tikzpicture}
+\end{center}
 }
 
+\only<2>{
+x = new()
+
+\begin{center}
+\begin{tikzpicture}[shorten >=1pt,auto]
+  \node (n1) {$\bullet$};
+  \node (n2) [right of=n1] {$\bullet$};
+  \node (n3) [right of=n2] {$\bullet$};
+
+  \node (x) [below of=n1] {$x$};
+  \node (null) [below of=n3] {\bf null};
+
+  \path[draw, ->] (n1) edge (n2)
+                  (n2) edge (n3);
+  \draw[dashed,->] (x) to (n1);
+  \draw[dashed,->] (null) to (n3);
+
+  \node [below right of=n3] (mid) {$\Rightarrow$};
+
+  \node (m1) [above right of=mid]{$\bullet$};
+  \node (m2) [right of=m1] {$\bullet$};
+  \node (m3) [right of=m2] {$\bullet$};
+  \node (m4) [below of=m2] {$\bullet$};
+
+  \node (x2) [below =1em of m4] {$x$};
+  \node (null) [right of=x2] {\bf null};
+each of these 
+  \path[draw, ->] (m1) edge (m2) (m2) edge (m3) (m4) edge (m3);
+  \draw[dashed,->] (x2) to (m4);
+  \draw[dashed,->] (null) to (m3);
+ \end{tikzpicture}
+ \end{center}
+ }
+
+ \only<3>{
+x = y
+
+\begin{center} 
+ \begin{tikzpicture}[shorten >=1pt,auto]
+  \node (n1) {$\bullet$};
+  \node (n2) [right of=n1] {$\bullet$};
+  \node (n3) [right of=n2] {$\bullet$};
+  \node (n4) [below=2em of n2] {$\bullet$};
+
+  \node (y) [below=1em of n4] {$y$};
+  \node (x) [left of=y] {$x$};
+  \node (null) [right of=y] {\bf null};
+
+  \path[draw, ->] (n1) edge (n2)
+                  (n2) edge (n3)
+                  (n4) edge (n3);
+  \path[dashed,->] (x) edge (n1)
+                   (y) edge (n4)
+                   (null) edge (n3);
+
+  \node [below right of=n3] (mid) {$\Rightarrow$};
+
+  \node (m1) [above right of=mid]{$\bullet$};
+  \node (m2) [right of=m1] {$\bullet$};
+  \node (m3) [right of=m2] {$\bullet$};
+  \node (m4) [below=2em of m2] {$\bullet$};
+
+  \node (y2) [below=1em of m4] {$y$};
+  \node (x2) [left of=y2] {$x$};
+  \node (null) [right of=y2] {\bf null};
+
+  \path[draw, ->] (m1) edge (m2)
+                  (m2) edge (m3)
+                  (m4) edge (m3);
+  \path[dashed,->] (x2) edge (m4)
+                   (y2) edge (m4)
+                   (null) edge (m3);
+ \end{tikzpicture}
+ \end{center}
+ }
+
+\only<4>{
+x$\rightarrow$next = y
+
+\begin{center}
+ \begin{tikzpicture}[shorten >=1pt,auto]
+  \node (n1) {$\bullet$};
+  \node (n2) [right of=n1] {$\bullet$};
+  \node (n3) [right of=n2] {$\bullet$};
+  \node (n4) [below=2em of n2] {$\bullet$};
+
+  \node (y) [below=1em of n4] {$y$};
+  \node (x) [left of=y] {$x$};
+  \node (null) [right of=y] {\bf null};
+
+  \path[draw, ->] (n1) edge (n2)
+                  (n2) edge (n3)
+                  (n4) edge (n3);
+  \path[dashed,->] (x) edge (n1)
+                   (y) edge (n4)
+                   (null) edge (n3);
+
+  \node [below right of=n3] (mid) {$\Rightarrow$};
+
+  \node (m1) [above right of=mid]{$\bullet$};
+  \node (m2) [right of=m1] {$\bullet$};
+  \node (m3) [right of=m2] {$\bullet$};
+  \node (m4) [below=2em of m2] {$\bullet$};
+
+  \node (y2) [below=1em of m4] {$y$};
+  \node (x2) [left of=y2] {$x$};
+  \node (null) [right of=y2] {\bf null};
+
+  \path[draw, ->] (m1) edge (m4)
+                  (m2) edge (m3)
+                  (m4) edge (m3);
+  \path[dashed,->] (x2) edge (m1)
+                   (y2) edge (m4)
+                   (null) edge (m3);
+ \end{tikzpicture}
+\end{center}
+}
+
+\end{frame}
+
+\begin{frame}[fragile]{Heap Observations}
+
+\begin{center}
+\begin{minipage}[c]{.4\textwidth}
+\begin{tikzpicture}[>=stealth',shorten >=1pt,auto,scale=.8]
+  %\node at (-.75,.25) {$h:$};
+
+  \node (n1) {$\bullet$};
+  \node (n2) [below of=n1] {$\bullet$};
+  \node (n3) [right of=n1] {$\bullet$};
+  \node (n4) [right of=n3] {$\bullet$};
+  \node (n5) [right of=n4,above of=n4] {$\bullet$};
+  \node (n6) [below of=n4] {$\bullet$};
+  \node (x) [below of=n2] {$x$};
+  \node (y) [right of=n5] {$y$};
+  \node (z) [above of=n1] {$z$};
+  \node (null) [below of=n6] {\bf null};
+
+  \path[draw,->] (n1) edge (n3)
+		 (n2) edge (n3)
+		 (n3) edge (n4)
+		 (n4) edge[bend right=45] (n5)
+		 (n5) edge[bend right=45] (n4);
+
+  \path[dashed,->] (x) edge (n2)%
+		   (y) edge (n5)%
+		   (z) edge (n1)%
+		   (null) edge (n6);
+ \end{tikzpicture}
+ \end{minipage}
+ \begin{minipage}[c]{.4\textwidth}
  \vfill
- $\hlookup(h, x, y)$ \\
- $x = y{\rightarrow}next;$
+ \begin{align*}
+ \hpathlen(x, y) & =  3 \\
+ \hispath(z, y) & = \text{true} \\
+ \hispath(x, z) & = \text{false} \\
+ \halias(x, z) & = \text{false} \\
+ \hisnull(x) & = \text{false} \\
+ \hcircular(y) & = \text{true}
+ \end{align*}
+ \vfill
+ \end{minipage}
+\end{center}
+\end{frame}
+
+\begin{frame}[fragile]{The SLH Theory}
+ SLH is our theory of singly linked lists.  SLH formulae can contain arbitrary boolean combinations of the
+ predicates from the previous slides.
+
+ For example:
+
+ \[
+  h' = \halloc(h, x) \wedge \hisnull(h', x)
+ \]
+
+\end{frame}
+
+\begin{frame}{\shakira}
+ We built a decision procedure for SLH named \shakira, which checks that heaps don't lie.
+
+ \pause
+
+ \shakira works by exploiting a small model property of SLH to build a SAT formula that is equisatisfiabile with the SLH formula.
+\end{frame}
+
+\begin{frame}[fragile]{Edge Subdivision}
+\begin{center}
+ \begin{tikzpicture}
+  \node (h) {$H:$};
+
+  \node (n1) [below right =.1em and .1em of h] {$\bullet$};
+  \node (space) [right of=n1] {};
+  \node (n2) [right of=space] {$\bullet$};
+
+  \node (x) [below of=n1] {$x$};
+  \node (null) [below of=n2] {\bf null};
+
+  \path[draw, ->] (n1) to node[above] {3} (n2);
+  \path[dashed,->] (x) edge (n1)
+                   (null) edge (n2);
+
+  \node (h2) [right=9em of h] {$H':$};
+
+  \node (m1)[below right =.1em and .1em of h2] {$\bullet$};
+  \node (m2) [right of=m1] {$\bullet$};
+  \node (m3) [right of=m2] {$\bullet$};
+
+  \node (x2) [below of=m1] {$x$};
+  \node (null2) [below of=m3] {\bf null};
+
+  \path[draw, ->] (m1) edge node[above] {1} (m2)
+                  (m2) edge node [above] {2} (m3);
+  \path[dashed,->] (x2) edge (m1)
+		   (null2) edge (m3);
+ \end{tikzpicture}
+\end{center}
+
+ \vspace{1em}
+ $H'$ can be generated from $H$ by subdividing an edge.  We say $H'$ is a \emph{subdivision} of $H$.
+\end{frame}
+
+\begin{frame}[fragile]{Heap Homeomorphism}
+
+ \begin{center}
+ \resizebox{\textwidth}{!}{
+ \begin{tikzpicture}
+  \node (h) {$H:$};
+
+  \node (x) [below right =.1em and .1em of h] {$x$};
+  \node (y) [below of=x] {$y$};
+
+  \node (n1) [right of=x] {$\bullet$};
+  \node (n2) [below of=n1] {$\bullet$};
+
+  \path (n1) to node (mid) {} (n2);
+
+  \node (n3) [right of=mid] {$\bullet$};
+  \node (n4) [right of=n3] {$\bullet$};
+
+  \node (n5) [right of=n4] {$\bullet$};
+  \node (space2) [right of=n5] {};
+  \node (n6) [right of=space2] {$\bullet$};
+
+  \path[->,draw] (n1) edge node [above] {1} (n3)
+                 (n2) edge node [below] {2} (n3)
+                 (n3) edge node [above] {4} (n4)
+                 (n4) edge node [above] {2} (n5)
+                 (n5) edge[bend right=70] node [below] {6} (n6)
+                 (n6) edge[bend right=70] node [above] {3} (n5);
+
+  \path[->,dashed] (x) edge (n1)
+                   (y) edge (n2);
+
+  \node (h2) [right=20em of h] {$H':$};
+
+  \node (x2) [below right =.1em and .1em of h2] {$x$};
+  \node (y2) [below of=x2] {$y$};
+
+  \node (m1) [right of=x2] {$\bullet$};
+  \node (m2) [below of=m1] {$\bullet$};
+
+  \path (m1) to node (mid2) {} (m2);
+
+  \node (m3) [right of=mid2] {$\bullet$};
+  \node (m4) [right of=m3] {};
+
+  \node (m5) [right of=m4] {$\bullet$};
+  \node (space2) [right of=m5] {};
+  \node (m6) [right of=space2] {$\bullet$};
+
+  \node (m7) [above of=space2] {$\bullet$};
+
+  \path[->,draw] (m1) edge node [above] {1} (m3)
+                 (m2) edge node [below] {2} (m3)
+                 (m3) edge node [above] {6} (m5)
+                 (m5) edge[bend right=70] node [below] {6} (m6)
+                 (m6) edge[bend right=35] node [above] {1} (m7)
+                 (m7) edge[bend right=35] node [above] {2} (m5);
+
+  \path[->,dashed] (x2) edge (m1)
+                   (y2) edge (m2);
+
+  \node (s) [below right=7em and 8em of h] {$S:$};
+
+  \node (x) [below right =.1em and .1em of s] {$x$};
+  \node (y) [below of=x] {$y$};
+
+  \node (n1) [right of=x] {$\bullet$};
+  \node (n2) [below of=n1] {$\bullet$};
+
+  \path (n1) to node (mid) {} (n2);
+
+  \node (n3) [right of=mid] {$\bullet$};
+  \node (n4) [right of=n3] {$\bullet$};
+
+  \node (n5) [right of=n4] {$\bullet$};
+  \node (space2) [right of=n5] {};
+  \node (n6) [right of=space2] {$\bullet$};
+
+  \node (n7) [above of=space2] {$\bullet$};
+
+  \path[->,draw] (n1) edge node [above] {1} (n3)
+                 (n2) edge node [below] {2} (n3)
+                 (n3) edge node [above] {4} (n4)
+                 (n4) edge node [above] {2} (n5)
+                 (n5) edge[bend right=70] node [below] {6} (n6)
+                 (n6) edge[bend right=35] node [above] {1} (n7)
+                 (n7) edge[bend right=35] node [above] {2} (n5);
+
+  \path[->,dashed] (x) edge (n1)
+                   (y) edge (n2);
+% 
+%   \node (x) [below of=n1] {$x$};
+%   \node (null) [below of=n2] {\bf null};
+% 
+%   \path[draw, ->] (n1) to node[above] {3} (n2);
+%   \path[dashed,->] (x) edge (n1) 
+%                    (null) edge (n2);
+% 
+%   \node (h2) [right=9em of h] {$H':$};
+% 
+%   \node (m1)[below right =.1em and .1em of h2] {$\bullet$};
+%   \node (m2) [right of=m1] {$\bullet$};
+%   \node (m3) [right of=m2] {$\bullet$};
+% 
+%   \node (x2) [below of=m1] {$x$};
+%   \node (null2) [below of=m3] {\bf null};
+% 
+%   \path[draw, ->] (m1) edge node[above] {1} (m2)
+%                   (m2) edge node [above] {2} (m3);
+%   \path[dashed,->] (x2) edge (m1)
+%                    (null2) edge (m3);
+ \end{tikzpicture}
+ }
+ \end{center}
+
+ Both $H$ and $H'$ can be subdivided to produce $S$.  We then say that $H$ and $H'$ are \emph{homeomorphic}.
+\end{frame}
+
+\begin{frame}{SLH Has Small Models}
+ \begin{theorem}[Homeomorphism Preserves Models]
+  If two heaps are homeomorphic, they satisfy exactly the same SLH formulae.
+ \end{theorem}
+
+ \begin{theorem}[Homeomorphisms Can Be Small]
+  Every homeomorphism class has a least element, which has no more than $2P$ nodes ($P$ is the number of program variables).
+ \end{theorem}
+
+ From this it follows that SLH has small models -- we only need to check models up to size $2P$.
 \end{frame}

### 2014-11-13 slides/nus-talk.tex
NUS slides

@@
 
+\begin{frame}{Existential Second-Order Logic}
+ Second-order logic allows quantification over functions as well as ground terms.
+
+ \vspace{1em}
+
+ Existential second-order logic is a restriction where we can only existentially quantify over functions.
+
+\end{frame}
+
+\begin{frame}{Existential Second-Order Logic Example}
+The following is an existential second-order formula:
+
+ \[
+  \exists F . \forall x . F(x) > x
+ \]
+
+ When using Peano arithmetic as a background theory, this formula is satisfiable.  One satisfying assignment would be $F(x) = x + 1$.
+\end{frame}
+
+\begin{frame}{Second-Order SAT}
+ A further restriction is to say that all the first-order variables are boolean propositions.  This
+ gives us the \emph{second-order SAT problem}.
+
+ \vspace{1em}
+
+ Second-order SAT is an extension of propositional SAT, where we are allowed to quantify
+ over functions.
+
+ \vspace{1em}
+
+ Second-order SAT is NEXPTIME-complete.
+\end{frame}
+
+\begin{frame}{Using Second-Order SAT}
+ Second-order SAT is a very general problem and can be used for lots of problems, including:
+
+ \begin{itemize}
+  \item Termination and non-termination
+  \item Safety invariants
+  \item Finding bugs without unwinding loops
+  \item Code refactoring
+  \item Code deobfuscation
+  \item Program synthesis
+  \item Concurrency
+ \end{itemize}
+
+ Also since everything boils down to SAT, we can freely combine with any theory that is decidable in NP.
+\end{frame}
+
+\begin{frame}{Solving Second-Order SAT}
+ To solve second-order SAT instances, we use a program synthesiser.
+
+ \vspace{1em}
+
+ The synthesiser finds programs computing each of the quantified functions.
+\end{frame}
+
+\begin{frame}[fragile]{Example}
+Consider the following second-order SAT problem:
+
+ \[
+  \exists F . \forall x . F(x) < x
+ \]
+
+This formula is satisfiable.  To show this, our program synthesiser might generate the following program:
+
+\begin{center}
+\begin{lstlisting}
+F(unsigned int x) {
+  unsigned int t1 = x / 2;
+  return t1;
+}
+\end{lstlisting}
+\end{center}
+\end{frame}
+
 \section{Termination}
 
-\begin{frame}[fragile]{Motivation}
-Does this program terminate?
-\begin{center}
-\begin{minipage}{0.4\linewidth}
- \begin{lstlisting}[language=C,basicstyle=\normalsize]
-while (x > 0) {
-  x++;
-}
- \end{lstlisting}
-\end{minipage}
-\end{center}
-
-\pause
-
-That depends...
-
-It \emph{does not} terminate if the program operates on integers...
-
-But it \emph{does} terminate if the program operates on fixed-width words
-(-x decreases monotonically and is bounded by -MAXINT).
-
-\end{frame}
-
-\begin{frame}[fragile]{Motivation}
-Which of these guys terminates?
-\begin{center}
-\begin{minipage}{0.45\linewidth}
- \begin{lstlisting}[language=C,basicstyle=\normalsize]
-while (x > 0) {
-  x = (x - 1) & x;
-}
-\end{lstlisting}
-\end{minipage}
-\begin{minipage}{0.45\linewidth}
- \begin{lstlisting}[language=C,basicstyle=\normalsize]
-while (x > 0) {
-  x = (x + 1) & x;
-}
-\end{lstlisting}
-\end{minipage}
-\end{center}
-
-\pause
-
-The left one does (x decreases monotonically), but the right one doesn't ($2 \rightarrow 2 \rightarrow \dots$).
-\end{frame}
-
-\begin{frame}{Motivation}
-Proving program termination has been a big deal for a while.
-
-It'd be nice if we could automatically prove termination for programs that run on computers.
-\end{frame}
-
-\section{Proving Termination}
 \begin{frame}{What Does a Termination Proof Look Like?}
@@
 \end{frame}
-
-\begin{frame}{Finding Ranking Functions}
-Traditionally, people find ranking functions under some assumptions:
-
-\begin{itemize}
- \item Programs are linear
- \item Ranking functions are linear
- \item Program variables are rationals
-\end{itemize}
-
-\vspace{1em}
-
-With these assumptions in hand, one can deploy a very powerful piece of linear algebra: \emph{Farkas's lemma}.
-
-This lemma makes light work of a ton of termination problems.
-
-\vspace{1em}
-
-But it's not for us.
-\end{frame}
+% 
+% \begin{frame}{Finding Ranking Functions}
+% Traditionally, people find ranking functions under some assumptions:
+% 
+% \begin{itemize}
+%  \item Programs are linear
+%  \item Ranking functions are linear
+%  \item Program variables are rationals
+% \end{itemize}
+% 
+% \vspace{1em}
+% 
+% With these assumptions in hand, one can deploy a very powerful piece of linear algebra: \emph{Farkas's lemma}.
+% 
+% This lemma makes light work of a ton of termination problems.
+% 
+% \vspace{1em}
+% 
+% But it's not for us.
+% \end{frame}
 
@@
 
-\section{Program Synthesis}
+% \section{Program Synthesis}
 \iffalse
@@
 
-\section{Advanced Topics}
+% \section{Advanced Topics}
 \begin{frame}{Conditional Termination}

### 2014-12-14 papers/termination/termination-esop/paper.tex
thanks and SV-COMP

@@
 
-\title{Unrestricted Termination and Non-Termination Arguments for Bit-Vector Programs}
+\title{Unrestricted Termination and Non-Termination Arguments for Bit-Vector
+Programs\thanks{Supported by UK EPSRC EP/J012564/1 and ERC project 280053.}}
 
@@
 SV-COMP'15~\cite{svcomp15}.
-We omitted exactly those SVCOMP'15
+We omitted exactly those SV-COMP'15
 benchmarks that made use of arrays or recursion. We do not have arrays in

### 2014-12-14 papers/termination/termination-esop/paper.tex
nudge Fig 1

@@
 \centering
- \begin{tabular}{|ll||c|c|c|c|c|c|c|c|}
+ \begin{tabular}{|l||c|c|c|c|c|c|c|c|}
  \hline
-  & & \multicolumn{8}{c|}{Program} \\
-  & & \multicolumn{2}{c|}{Rationals/Integers} & \multicolumn{2}{c|}{Reals} & \multicolumn{2}{c|}{Bit-vectors} & \multicolumn{2}{c|}{Floats} \\
-  & & L & NL & L & NL & L & NL & L & NL \\
+                   & \multicolumn{8}{c|}{Program} \\
+  Ranking argument & \multicolumn{2}{c|}{Rationals/Integers} & \multicolumn{2}{c|}{Reals} & \multicolumn{2}{c|}{Bit-vectors} & \multicolumn{2}{c|}{Floats} \\
+                   & L & NL & L & NL & L & NL & L & NL \\
   \hline
   \hline
-  \multirow{4}{*}{Ranking} & Linear lexicographic &  \cite{DBLP:conf/popl/Ben-AmramG13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13,DBLP:conf/vmcai/P04} & - & \cite{DBLP:conf/tacas/LeikeH14} & - &\checkmark&\checkmark&\checkmark&\checkmark\\
-   & Linear non-lexicographic & \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10} & \cite{DBLP:conf/vmcai/BradleyMS05} & \cite{DBLP:conf/tacas/LeikeH14} & - & \checkmark~ \cite{DBLP:conf/tacas/CookKRW10} &\checkmark~ \cite{DBLP:conf/tacas/CookKRW10}&\checkmark&\checkmark\\
-   & Nonlinear lexicographic & - & - & - & - &\checkmark&\checkmark&\checkmark&\checkmark\\
-   & Nonlinear non-lexicographic & \cite{DBLP:conf/vmcai/BradleyMS05} &  \cite{DBLP:conf/vmcai/BradleyMS05} & - & - &\checkmark&\checkmark&\checkmark&\checkmark\\
+  Linear lexicographic        & \cite{DBLP:conf/popl/Ben-AmramG13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13,DBLP:conf/vmcai/P04} & -- & \cite{DBLP:conf/tacas/LeikeH14} & -- &\checkmark&\checkmark&\checkmark&\checkmark\\
+  Linear non-lexicographic    & \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10} & \cite{DBLP:conf/vmcai/BradleyMS05} & \cite{DBLP:conf/tacas/LeikeH14} & -- & \checkmark~ \cite{DBLP:conf/tacas/CookKRW10} &\checkmark~ \cite{DBLP:conf/tacas/CookKRW10}&\checkmark&\checkmark\\
+  Nonlinear lexicographic     & -- & -- & -- & -- &\checkmark&\checkmark&\checkmark&\checkmark\\
+  Nonlinear non-lexicographic & \cite{DBLP:conf/vmcai/BradleyMS05} & \cite{DBLP:conf/vmcai/BradleyMS05} & -- & -- &\checkmark&\checkmark&\checkmark&\checkmark\\
   \hline
- \end{tabular}
-
- \caption{Summary of related termination analyses. Legend: \checkmark = we can handle; - = no available works; L = linear; NL = nonlinear.} \label{fig:handletable}
+ \end{tabular}\\
+ Legend: \checkmark = we can handle; --~=~no available works; L = linear; NL = nonlinear
+
+ \caption{Summary of related termination analyses}
+ \label{fig:handletable}
 \end{figure*}

### 2014-12-14 papers/termination/termination-esop/paper.tex
fix Fig 2

@@
 
+\noindent
 {\bf Limitations.} Our algorithm proves termination for transition systems
@@
 their curious properties:
-
 %
@@
 \end{itemize}
+
+\begin{figure*}[p]
+\centering
+\begin{tabular}{ccc}
+\begin{subfigure}[b]{0.45\textwidth}
+\begin{lstlisting}
+while (x > 0) {
+  x = (x - 1) & x;
+}
+\end{lstlisting}
+\caption{Taken from~\cite{DBLP:conf/tacas/CookKRW10}}
+\label{fig:motivation.a}
+\end{subfigure}%
+
+&
+
+\begin{subfigure}[b]{0.45\textwidth}
+\begin{lstlisting}
+while (x != 0) {
+  x = -x / 2;
+}
+\end{lstlisting}
+\caption{}
+ \label{fig:motivation.b}
+\end{subfigure} \\
+\hline
+
+\begin{subfigure}[b]{0.45\textwidth}
+\begin{lstlisting}[language=C]
+while(x > 0) {
+  x++;
+}
+ \end{lstlisting}
+\caption{}
+ \label{fig:motivation.c}
+\end{subfigure}%
+
+&
+
+\begin{subfigure}[b]{0.45\textwidth}
+\begin{lstlisting}
+while (i<M || j<N) {
+  i = i + 1;
+  j = j + 1;
+}
+\end{lstlisting}
+\caption{Taken from~\cite{DBLP:conf/sigsoft/Nori013}}
+ \label{fig:motivation.d}
+\end{subfigure} \\
+\hline
+
+\begin{subfigure}[b]{0.45\textwidth}
+\begin{lstlisting}
+float x;
+
+while (x > 0.0) {
+  x -= 1.0;
+}
+\end{lstlisting}
+\caption{}
+ \label{fig:motivation.e}
+\end{subfigure}%
+
+&
+
+\begin{subfigure}[b]{0.45\textwidth}
+\begin{lstlisting}
+float x;
+
+while (x > 0.0) {
+  x *= 0.5;
+}
+\end{lstlisting}
+\caption{}
+ \label{fig:motivation.f}
+\end{subfigure} \\
+\hline
+
+\begin{subfigure}[b]{0.45\textwidth}
+\begin{lstlisting}
+while (x != 0) {
+  if (x > 0)
+    x--;
+  else
+    x++;
+}
+\end{lstlisting}
+\caption{Taken from \cite{DBLP:conf/tacas/CookSZ13}}
+ \label{fig:motivation.g}
+\end{subfigure}
+
+&
+
+\begin{subfigure}[b]{0.45\textwidth}
+\begin{lstlisting}
+y = 1;
+
+while (x > 0) {
+  x = x - y;
+}
+\end{lstlisting}
+\caption{}
+ \label{fig:motivation.h}
+\end{subfigure} \\
+\hline
+
+\begin{subfigure}[b]{0.45\textwidth}
+\begin{lstlisting}
+while (x>0 && y>0 && z>0){
+  if (y > x) {
+    y = z;
+    x = nondet();
+    z = x - 1;
+  } else {
+    z = z - 1;
+    x = nondet();
+    y = x - 1;
+  }
+}
+\end{lstlisting}
+\caption{Taken from~\cite{BA:mcs}}
+ \label{fig:motivation.i}
+\end{subfigure} 
+\end{tabular}
+\caption{Motivational examples, mostly taken from the literature.\label{fig:motivation}}
+\end{figure*}
 
@@
 functions above were all found completely automatically.
-
-\begin{figure*}[h]
-\hspace*{-2.5cm}
-\centering
-%\resizebox{.95\textwidth}{!}{
-%\begin{minipage}{\linewidth}
-\begin{tabular}{ccc}
-\begin{subfigure}[b]{0.45\textwidth}
-\begin{lstlisting}
-while (x > 0) {
-  x = (x - 1) & x;
-}
-\end{lstlisting}
-\caption{Taken from~\cite{DBLP:conf/tacas/CookKRW10}.}
- \label{fig:motivation.a}
-\end{subfigure}%
-
-&
-
-\begin{subfigure}[b]{0.45\textwidth}
-\begin{lstlisting}
-while (x != 0) {
-  x = -x / 2;
-}
-\end{lstlisting}
-\caption{}
- \label{fig:motivation.b}
-\end{subfigure}%
-
-&
-
-\begin{subfigure}[b]{0.45\textwidth}
-\begin{lstlisting}[language=C]
-while(x > 0) {
-  x++;
-}
- \end{lstlisting}
-\caption{}
- \label{fig:motivation.c}
-\end{subfigure} \\
-
-\hline
-
-\begin{subfigure}[b]{0.45\textwidth}
-\begin{lstlisting}
-while (i<M || j<N) {
-  i = i + 1;
-  j = j + 1;
-}
-\end{lstlisting}
-\caption{Taken from~\cite{DBLP:conf/sigsoft/Nori013}}
- \label{fig:motivation.d}
-\end{subfigure}
-
-&
-
-\begin{subfigure}[b]{0.45\textwidth}
-\begin{lstlisting}
-float x;
-
-while (x > 0.0) {
-  x -= 1.0;
-}
-\end{lstlisting}
-\caption{}
- \label{fig:motivation.e}
-\end{subfigure}
-
-&
-
-\begin{subfigure}[b]{0.45\textwidth}
-\begin{lstlisting}
-float x;
-
-while (x > 0.0) {
-  x *= 0.5;
-}
-\end{lstlisting}
-\caption{}
- \label{fig:motivation.f}
-\end{subfigure} \\
-\hline
-
-\begin{subfigure}[b]{0.45\textwidth}
-\begin{lstlisting}
-while (x != 0) {
-  if (x > 0)
-    x--;
-  else
-    x++;
-}
-\end{lstlisting}
-\caption{Taken from \cite{DBLP:conf/tacas/CookSZ13}}
- \label{fig:motivation.g}
-\end{subfigure}
-
-&
-
-\begin{subfigure}[b]{0.45\textwidth}
-\begin{lstlisting}
-y = 1;
-
-while (x > 0) {
-  x = x - y;
-}
-\end{lstlisting}
-\caption{}
- \label{fig:motivation.h}
-\end{subfigure} 
-&
-\begin{subfigure}[b]{0.45\textwidth}
-\begin{lstlisting}
-while (x>0 && y>0 && z>0){
-  if (y > x) {
-    y = z;
-    x = nondet();
-    z = x - 1;
-  } else {
-    z = z - 1;
-    x = nondet();
-    y = x - 1;
-  }
-}
-\end{lstlisting}
-\caption{Taken from~\cite{BA:mcs}}
- \label{fig:motivation.i}
-\end{subfigure} 
-\end{tabular}
-%}
-\caption{Motivational examples, mostly taken from the literature.\label{fig:motivation}}
-\end{figure*}
 

### 2014-12-14 papers/termination/termination-esop/paper.tex
nudge
make link to ARMC work
nudges

@@
 \usetikzlibrary{arrows, automata, shapes, calc}
+\usepackage{url}
 
@@
 \label{def:2sat}
- \[
-  \exists S_1 \ldots S_m . Q_1 x_1 \ldots Q_n x_n . \sigma
- \]
+\[ \exists S_1 \ldots S_m . Q_1 x_1 \ldots Q_n x_n . \sigma \]
+%
 Where the $S_i$'s range over predicates,
 the $Q_i$'s are either $\exists$ or $\forall$,
-the $x_i$'s range over boolean values,
+the $x_i$'s range over Boolean values,
 and $\sigma$ is a quantifier-free propositional formula
@@
 and $S_i \subseteq \mathbb{B}^{\mathrm{ar}(S_i)}$.  Note that 
-$Q_1 x_1 \ldots Q_n x_n . \sigma$
- is an instance of first-order propositional SAT, i.e. QBF.
+$Q_1 x_1 \ldots Q_n x_n . \sigma$ is the special case of
+a propositional formula with first-order quantification,
+i.e.~QBF.
 \end{definition}
@@
 
-If this formula is satisfiable, $N$ is an open recurrence set for $L$, which proves
+If this formula is satisfiable, $N$~is an open recurrence set for~$L$, which proves
 $L$'s non-termination. The issue with this formula is the additional level of quantifier alternation as compared to second-order SAT
 (it is an $\exists \forall \exists$ formula).  To eliminate the innermost existential quantifier,
-we introduce a Skolem function $C$ that chooses the successor $x'$, which we then existentially quantify over.
+we introduce a Skolem function~$C$ that chooses the successor~$x'$, which we then existentially quantify over.
 This results in formula {\bf [SNT]} (Definition~\ref{def:snt}, Figure~\ref{fig:single_loop}).
@@
 we can avoid if the transition relation $T$ is deterministic.
+%
 \begin{definition}[Determinism]
@@
 \end{definition}
+%
 In order to describe a deterministic \emph{program} in a way that still allows us
 to sensibly talk about termination, we assume the existence of a special sink
-state $s$ with no outgoing transitions and such that $\lnot G(s)$ for any
-of the loop guards $G$.  The program is deterministic if its transition
-relation is deterministic for all states except $s$.
+state~$s$ with no outgoing transitions and such that $\lnot G(s)$ for any
+of the loop guards~$G$.  The program is deterministic if its transition
+relation is deterministic for all states except~$s$.
 
@@
 introduced by Chen et al.~in~\cite{DBLP:conf/tacas/ChenCFNO14}:  for each
-state in the recurrence set $N$, \emph{all} of its successors must be in $N$.
+state in the recurrence set~$N$, \emph{all} of its successors must be in~$N$.
 The existence of a closed recurrence set is equivalent to the satisfiability
@@
 loss of generality, that non-termination of the loop is equivalent
-to the existence of a closed recurrence set.  However if $T$ is
+to the existence of a closed recurrence set.  However, if $T$ is
 non-deterministic, it may be that there is an open recurrence
@@
 It is clear that this loop has many non-terminating executions,
-e.g. the execution where \lstinline!nondet()! always returns 0.
-However each state has a successor
+e.g.~the execution where \lstinline!nondet()! always returns~0.
+However, each state has a successor
 that exits the loop, i.e.~when \lstinline|nondet()| returns
-the value currently stored in \lstinline|x|.  So this loop
+the value currently stored in \lstinline|x|.  Thus, this loop
 has an open recurrence set, but no closed recurrence set
 and hence we cannot give a proof of its non-termination
-with {\bf [CNT]} and instead must use {\bf [SNT]}.
+with {\bf [CNT]} and instead must use~{\bf [SNT]}.
 
@@
 
-Key: \tick = terminating, \xmark = non-terminating, ? = unknown (tool terminated with an inconclusive verdict).
+Key: \tick~=~terminating, \xmark~=~non-terminating, ?~=~unknown (tool terminated with an inconclusive verdict)
 

### 2015-01-16 papers/termination/termination-esop/paper.tex
added amir's commens

@@
   \hline
-  Linear lexicographic        & \cite{DBLP:conf/popl/Ben-AmramG13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13,DBLP:conf/vmcai/P04} & -- & \cite{DBLP:conf/tacas/LeikeH14} & -- &\checkmark&\checkmark&\checkmark&\checkmark\\
-  Linear non-lexicographic    & \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10} & \cite{DBLP:conf/vmcai/BradleyMS05} & \cite{DBLP:conf/tacas/LeikeH14} & -- & \checkmark~ \cite{DBLP:conf/tacas/CookKRW10} &\checkmark~ \cite{DBLP:conf/tacas/CookKRW10}&\checkmark&\checkmark\\
+  Linear lexicographic        & \cite{DBLP:journals/jacm/Ben-AmramG14,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13,DBLP:conf/vmcai/P04} & -- & \cite{DBLP:conf/tacas/LeikeH14} & -- &\checkmark&\checkmark&\checkmark&\checkmark\\
+  Linear non-lexicographic    & \cite{DBLP:conf/popl/Ben-AmramG13,DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10} & \cite{DBLP:conf/vmcai/BradleyMS05} & \cite{DBLP:conf/tacas/LeikeH14} & -- & \checkmark~ \cite{DBLP:conf/tacas/CookKRW10} &\checkmark~ \cite{DBLP:conf/tacas/CookKRW10}&\checkmark&\checkmark\\
   Nonlinear lexicographic     & -- & -- & -- & -- &\checkmark&\checkmark&\checkmark&\checkmark\\
@@
 
-\item In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, program (i)
-admits a \emph{multiphase} ranking function, computed from a multiphase
-ranking template.  Multiphase ranking templates are targeted at programs
-that go through a finite number of phases in their execution.  Each phase is
-ranked with an affine-linear function and the phase is considered to be
-completed once this function becomes non-positive.
-
-In our setting this type of programs does not need special treatment, as we
-can find a nonlinear lexicographic ranking function $R(x, y, z) = (x < y,
+\item %% In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, program (i)
+%% admits a \emph{multiphase} ranking function, computed from a multiphase
+%% ranking template.  Multiphase ranking templates are targeted at programs
+%% that go through a finite number of phases in their execution.  Each phase is
+%% ranked with an affine-linear function and the phase is considered to be
+%% completed once this function becomes non-positive.
+
+%% In our setting this type of programs does not need special treatment, as
+For program (i) we find a nonlinear lexicographic ranking function $R(x, y, z) = (x < y,
 z)$.\footnote{This termination argument is somewhat subtle.  The Boolean
@@
 decrease indefinitely, and so $R$ proves that the loop is well founded.}
+ We are not aware of any linear ranking function for this program.
 

### 2015-01-17 papers/termination/termination-esop/paper.tex
updated citation to the solver
small nudges

@@
 nonlinear recurrence sets that are correct for fixed-width machine arithmetic
-and floating-point arithmetic
+and floating-point arithmetic.
 Our technique is based on a reduction from program \emph{termination} to 
@@
 %existential second-order logic over finite domains is decidable \cite{kalashnikov}. 
-The resulted technique is a sound and complete analysis for the termination 
+The resulting technique is a sound and complete analysis for the termination 
 of finite-state programs with fixed-width
@@
 we will not provide the full algorithm, but rather illustrate it through the example in Figure~\ref{fig:environment-model}.
-Full details of this construction appear in the extended version of this paper.
+%Full details of this construction appear in the extended version of this paper.
 For the current example, the termination formula is given on the right side of 
@@
 Let us assume that we have done some preprocessing of our program which has identified
-loops, straight line code blocks and the control flow between these.  In particular,
+loops, straight-line code blocks and the control flow between these.  In particular,
 the control flow analysis has determined which order these code blocks execute in,
@@
 some ranking function $R''$ that is just a permutation of $\mathcal{S}$.  If
-the number of variables $k > 1$ then in general the ranking function will be
+the number of variables $k > 1$, then in general the ranking function will be
 lexicographic with dimension $\leq k$ and each co-ordinate of the output
@@
  $C : \mathcal{S} \to \mathcal{S}$ is a Skolem function choosing
- a successor for each $x \in N$.  $\mathcal{S}$ is finite, so by Lemma~\ref{lem:l-func} both
+ a successor for each $x \in N$.  The state space $\mathcal{S}$ is finite, so by Lemma~\ref{lem:l-func} both
  $N$ and $C$ are computed by finite $\mathcal{L}$-programs and $x_0$ is just a ground term.

### 2015-04-12 slides/esop-termination.tex
ESOP termination slides

--- 2014-06-18 slides/termination.tex
@@
 
+\lstset{escapeinside={<@}{@>}}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-%\author{Cristina David \and Daniel Kroening \and Matt Lewis}
-\title{Everything You Ever Wanted To Prove About Termination But Were Afraid To Try}
+\author{Cristina David \and Daniel Kroening \and Matt Lewis}
+\title{Unrestricted Termination and Non-Termination Arguments for Bit-Vector Programs}
 %\subtitle{}
-%\institute{Oxford University}
+\institute{Oxford University}
 \date{\today}
@@
 
-\section{Motivation}
-
+\section{What's The Problem?}
 \begin{frame}[fragile]{Motivation}
@@
 
-It'd be nice if we could automatically prove termination for programs that run on computers.
+\vspace{1em}
+
+It's still quite hard to do for the programs that people write, which use bitvectors and all kinds of nonsense.
 \end{frame}
 
-\section{Proving Termination}
-\begin{frame}{What Does a Termination Proof Look Like?}
-We prove that a program terminates by finding a \emph{ranking function}.
+\subsection{Termination Is Hard}
+
+\subsection{Bitvectors Are Hard}
+
+\begin{frame}[fragile]{Bitvector Termination}
+ \begin{columns}[c]
+  \column{.45\textwidth}
+  \emph{If a program terminates for integers, it terminates for bitvectors.}
+
+  \uncover<2->{{\centering \color{red} \Huge \xmark
+
+  }
+
+  \lstinputlisting[language=C,basicstyle=\normalsize,mathescape]{term1.c}
+  }
+
+  \column{.45\textwidth}
+  \emph{If a program terminates for bitvectors, it terminates for integers.}
+
+  \uncover<3->{{\centering \color{red} \Huge \xmark
+
+  }
+
+  \lstinputlisting[language=C,basicstyle=\normalsize]{term2.c}
+  }
+ \end{columns}
+
+ \vspace{1.5em}
+
+ \uncover<4->{\Large \centering Integer and bitvector termination are incomparable!
+
+ }
+
+\end{frame}
+
+\begin{frame}{Bitvector Termination}
+\begin{quote}
+ ``People really don't *think* about termination using bitvector arguments''
+   \hspace*\fill{\small--- Anonymous}
+\end{quote}
+\end{frame}
+
+\subsection{Non-Linearity Is Hard}
+
+\begin{frame}{Example}
+\end{frame}
+
+\section{Our Solution}
+
+\subsection{Existential Second-Order Logic}
+
+\begin{frame}[fragile]{Our Theory}
+ We will try to build termination proofs by solving \emph{second-order constraints}.
 
@@
 
-If we have a transition relation $T$ and a loop guard $g$, a ranking function $R$
-must meet the following criteria:
-\begin{align}
- \forall x, x' \cdot & g(x) \Rightarrow R(x) > 0 ~ \wedge \\
-                     & T(x, x') \Rightarrow R(x) > R(x')
-\end{align}
+ Specifically, we will build constraints of the form:
 
-Or more generally, $R$ must be an order-homomorphism with a well-founded co-domain (aka an ordinal).
+ \begin{center}
+ \begin{columns}
+  \column{.15\textwidth}
+  $\exists {\textcolor{red} F} . \forall \vec{\textcolor{blue} x} . {\textcolor{green} \phi}({\textcolor{red} F}, \vec{\textcolor{blue} x})$
 
-\vspace{1em}
+  \column{.4\textwidth}
+  \begin{minipage}{\textwidth}
+ \begin{itemize}
+  \item[$\textcolor{red} F$] a function.
+  \item[$\textcolor{green} \phi$] a quantifier free formula.
+  \item[$\vec{\textcolor{blue} x}$] the program variables.
+ \end{itemize}
+ \end{minipage}
+ \end{columns}
 
-We are therefore in the business of finding ranking functions.
+ \end{center}
+
+ \vspace{.75em}
+
+ These constraints can be solved using an EPR solver (such as the one in Z3), or by using a program
+ synthesiser to synthesise $\textcolor{red} F$.
 \end{frame}
 
-\begin{frame}{Finding Ranking Functions}
-Traditionally, people find ranking functions under some assumptions:
-
-\begin{itemize}
- \item Programs are linear
- \item Ranking functions are linear
- \item Program variables are rationals
-\end{itemize}
-
-\vspace{1em}
-
-With these assumptions in hand, one can deploy a very powerful piece of linear algebra: \emph{Farkas's lemma}.
-
-This lemma makes light work of a ton of termination problems.
-
-\vspace{1em}
-
-But it's not for us.
-\end{frame}
-
-\begin{frame}{Synthesing Ranking Functions}
-Our approach is to use \emph{program synthesis} to generate ranking functions.
-
-\vspace{1em}
-
-We're going to set up a program synthesis problem that will allow us to solve
-the following 2nd-order satisfaction problem:
-\begin{align*}
- \exists R . \forall x, x' \cdot & g(x) \Rightarrow R(x) > 0 ~ \wedge \\
-                             & T(x, x') \Rightarrow R(x) > R(x')
-\end{align*}
-
-\pause
-
-\vspace{1em}
-
-In other words: ``Is there a ranking function for my program?''
-
-\end{frame}
-
-\begin{frame}{Specifying a Ranking Function}
-Our program synthesiser needs a \emph{specification}, which takes the form of a C function.
-
-\vspace{1em}
-
-The specification function takes two arguments: a candidate program $P$ and an input $x$.
-If $P$ gives the correct output when fed $x$, our specification function returns true,
-otherwise it returns false.
-\end{frame}
-
-\begin{frame}[fragile]{A Specification}
-
-\vspace{-1em}
+\begin{frame}[fragile]{Modelling Loops}
+ For ease of presentation, we will consider loops of the following form:
 
  \begin{center}
-\begin{minipage}{.6\textwidth}
-\begin{lstlisting}[language=C++,basicstyle=\small]
-bool spec(prog_t *p, int x) {
-  int y = exec(p, x);
+ \begin{columns}[c]
+  \column{.15\textwidth}
+  \begin{minipage}{\linewidth}
+ \begin{lstlisting}[language=C,basicstyle=\normalsize]
+<@\textcolor{red}{P}@>;
 
-  if (y >= x)
-    return true;
-  else
-    return false;
+while (<@\textcolor{green}{G}@>) {
+  <@\textcolor{blue}{B}@>;
 }
@@
  \end{minipage}
-\end{center}
 
-\pause
+ \column{.4\textwidth}
 
-Two of the possible programs:
+ \begin{minipage}{\linewidth}
+ \begin{itemize}
+  \item[\textcolor{red}{P}] the prefix to the loop.
+  \item[\textcolor{green}{G}] the loop guard.
+  \item[\textcolor{blue}{B}] the loop body.
+ \end{itemize}
+ \end{minipage}
+ \end{columns}
 
-\begin{itemize}
-  \item[] \texttt{int f(int x) \{ return x; \}} 
-  \item[] \texttt{int f(int x) \{ return MAXINT; \}}
-\end{itemize}
-
-\end{frame}
-
-\begin{frame}[fragile]{Specifying a Ranking Function}
-\vspace{-1em}
-
-\begin{center}
-\begin{minipage}{0.6\textwidth}
-\begin{lstlisting}[language=c++,basicstyle=\small]
-int spec(prog_t *p, int x) {
-  int r1 = exec(p, x);
-
-  if (g(x)) {
-    if (r1 <= 0)
-      return false;
-
-    int y = body(x);
-    int r2 = exec(p, y);
-
-    if (r1 <= r2)
-      return false;
-  }
-
-  return true;
-}
-\end{lstlisting}
-\end{minipage}
-\end{center}
-\end{frame}
-
-\iffalse
-
-\section{Abstract Program Synthesis}
-\begin{frame}{The Synthesis Formula}
-Our high-level goal is to synthesise programs that meet some specification.
-
-\pause
-
-Abstractly, a program $P$ computes some function from an input set $I$ to an output set $O$:
-$$P : I \rightarrow O$$
-
-\pause
-
-A specification is a relation between inputs and outputs:
-$$\sigma \subseteq I \times O$$
-
-\pause
-
-The synthesis problem is to find a program that meets the specification on all inputs:
-$$\exists P . \forall x \in I . \sigma(x, P(x))$$
-\end{frame}
-\fi
-
-\begin{frame}[fragile]{Solving the Synthesis Formula with CEGIS}
- Solving second order formulae is hard.  Instead we'll alternately solve two \emph{first-order} formulae
- in a refinement loop:
-
- \vspace{1cm}
-
-  \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,
- semithick, initial text=]
-
-  \matrix[nodes={draw, fill=none, scale=1, shape=rectangle, minimum height=1cm, minimum width=1.5cm},
-          row sep=2cm, column sep=3cm] {
-   \node (synth) {Synthesise};
-   &
-   \node (verif) {Verify}; %\\
-   %\node[draw=none] {};
-   &
-   \node[ellipse] (done) {Done}; \\
-  };
-
-   \path
-    (synth) edge [bend left] node {Candidate program} (verif)
-    (verif) edge [bend left] node {Counterexample input} (synth)
-    (verif) edge node {Valid} (done);
- \end{tikzpicture}
-\end{frame}
-
-\begin{frame}{First-order Synthesis}
- If we have some small set of test inputs $\{ x_0, \dots, x_N\}$, we can find a program that is correct on just that set:
-
- $$\exists P . \sigma(x_0, P(x_0)) \wedge \dots \wedge \sigma(x_N, P(x_N))$$
-\end{frame}
-
-\begin{frame}{First-order Verification}
- If we have some program $P$ that might be correct, we can check whether it is in fact correct:
-
- $$\exists x . \lnot \sigma(x, P(x))$$
-
- If this formula is \emph{satisfiable}, $P$ is \emph{not} correct and we have found an input on which it is incorrect.
- We can add this input to our set of test inputs and loop back to synthesising a new program.
-\end{frame}
-
-\section{Program Synthesis}
-\iffalse
-\begin{frame}{Concrete Synthesis in C}
- We're using C as an implementation language.  To do that we need:
-
- \begin{itemize}
-  \item A way of encoding programs as \emph{finite structures} in C.
-  \item A way of encoding the first-order synthesis and verification formulae in C.
-  \item A way of checking properties of C programs.
- \end{itemize}
-\end{frame}
-
-\begin{frame}[fragile]{Encoding Programs in C}
- To encode the programs we will synthesise, we just make a very simple language and an interpreter for it.  Programs
- in our language look like this:
-
- \vspace{0.7cm}
-
-\begin{center}
-\begin{minipage}{0.3\linewidth}
-\begin{verbatimtab}
-t1 = neg x
-t2 = add t1 1
-return t2
-\end{verbatimtab}
-\end{minipage}
  \end{center}
@@
 
-\fi
+\subsection{Ranking Functions: Proofs of Termination}
 
-\begin{frame}{Complexity}
-By far the dominant factor in how long synthesis takes is the length of the shortest correct program.
+\subsection{Recurrent Sets: Proofs of Non-Termination}
+\subsection{Using Tautologies}
 
-\vspace{1em}
-
-This is called the \emph{Kolmogorov complexity} of the function being computed.
-
-\vspace{1em}
-
-For termination, this gives us the property that our runtime doesn't depend very heavily
-on the size of the program we are analysing.  It is instead almost entirely determined by the
-size of the shortest program that computes a valid ranking function.
-\end{frame}
-
-\begin{frame}{Genetic Programming}
-The flip side of being so dominated by Kolmogorov complexity is that we
-get \emph{really slow} beyond about 4 instructions.
-
-\vspace{1em}
-
-It turns out that \emph{genetic programming} helps a lot beyond this threshold.
-\end{frame}
-
-\begin{frame}{Genetic Programming}
-Genetic programming is an evolutionary search strategy, which aims to find a program
-that is ``fit'' according to some metric.
-
-\vspace{1em}
-
-It begins by generating a population of random programs.  This population is subjected to
-\emph{evolution} over a period of many generations.
-
-\vspace{1em}
-
-At each generation, each individual's fitness is evaluated.  The fitter an individual is,
-the greater its chance of breeding.
-
-When two programs breed, their code is combined in some way (the crossover operation) and
-the resulting child may be randomly modified (the mutation operator).  This child is copied
-into the next generation.
-\end{frame}
-
-\begin{frame}{Genetic Programming Synthesis}
-To use genetic programming for synthesis, we need to fix a useful fitness metric.
-
-\vspace{1em}
-
-For our purposes, we say that if a program meets the specification on $n$ test vectors, its
-fitness is $n$.
-
-Whenever we add a new test vector, we continue genetic programming with the most recent population
-(the one that produced the most recent candidate program).  This is called \emph{incremental evolution}
-and gives us a big speedup.
-
-\vspace{1em}
-
-Once we've found a program that passes all the tests, we return it and continue round the CEGIS
-loop.
-\end{frame}
-
-\section{Advanced Topics}
-\begin{frame}{Conditional Termination}
-
-\begin{align*}
-\exists I, R \cdot \forall x, x' \cdot & P(x, x') \Rightarrow & I(x') \\
-                                       & I(x) \wedge g(x) \wedge T(x, x') \Rightarrow & R(x) > 0 ~ \wedge \\
-                                       & & R(x) > R(x') ~ \wedge \\
-                                       & & I(x')
-\end{align*}
-\end{frame}
-
-\begin{frame}{Sequential Loops}
-\begin{align*}
-\exists I_1, I_2, R \cdot \forall x, x' \cdot & P(x, x') \Rightarrow & I_1(x') \\
-                                              & I_1(x) \wedge g_1(x) \wedge T_1(x, x') \Rightarrow & I_1(x') \\
-                                              & I_1(x) \wedge \lnot g_1(x) \Rightarrow & I_2(x) \\
-					      & I_2(x) \wedge g_2(x) \wedge T_2(x, x') \Rightarrow & R(x) > 0 ~ \wedge \\
-					      & & R(x) > R(x') ~ \wedge \\
-					      & & I_2(x')
-\end{align*}
-\end{frame}
-
-\begin{frame}{Some Combinatorics}
-A terminating program corresponds to a \emph{labelled partial order} on $2^k$ states.
-
-\vspace{.7em}
-
-Counting partial orders on a finite set turns out to be very hard (it's an open problem!).  However, asymptotics are known:
-\begin{align*}
- P_n & = \left(1 + O \left( \frac{1}{n} \right) \right) \left( \sum_{i=1}^n \sum_{j=1}^{n-1} \binom{n}{i} \binom{n-i}{j} (2^i - 1)^j (2^j - 1)^{n - i - j}\right) \\
-     & \sim O\left( 2^k \right)
-\end{align*}
-
-\end{frame}
-
-\begin{frame}{More Combinatorics}
-Linear functions on bitvectors correspond to \emph{permutations} of the natural ordering on bitvectors.
-
-\vspace{1em}
-
-There are $2^{2k}$ linear functions (although only half of these permute the full space, cf. generators of the group $2^k$).
-
-\vspace{1em}
-
-Each of these functions can show that each of the \emph{unlabelled} partial orders terminates.
-
-The number of unlabelled partial orders over an $n$ element set is related to the number of labelled partial orders as:
-
-$$p_n \sim {P_n \over n!}$$
-\end{frame}
-
-\begin{frame}{The Combinatorial Payoff}
-Since there are $P_n$ terminating functions, and each of the $n^2$ linear functions can prove that
-$p_n \sim {P_n \over n!}$ functions terminate, we can compute the probability that a random terminating
-function can be proved to terminate with a linear argument:
-
-$${2^k \times 2^k \over {2^k}!} \sim {1 \over {2^k}!}$$
-
-\pause
-
-\vspace{1em}
-
-This number is very small, which to some degree justifies our aim of moving beyond linear ranking functions.
-
-\end{frame}
+\section{Experiments}
 

### 2015-04-12 slides/term1.c

@@ -1,0 +1,3 @@
+while (x <= $2^{32} - 1$) {
+  x++;
+}

### 2015-04-12 slides/esop-termination.tex
More esop slides

@@
 \lstset{escapeinside={<@}{@>}}
+
+\newcommand{\red}[1]{\textcolor{red}{#1}}
+\newcommand{\blue}[1]{\textcolor{blue}{#1}}
+\newcommand{\green}[1]{\textcolor{green}{#1}}
 
@@
 
+\vspace{1em}
+
+In our logical encoding, we model \textcolor{green}{G} as a predicate $\textcolor{green}{G}(x)$. \textcolor{red}{P} and \textcolor{blue}{B} are modelled as transition
+relations $\textcolor{red}{P}(x, x')$ and $\textcolor{blue}{B}(x, x')$.
+
+\end{frame}
+
+\begin{frame}[fragile]{Loop Encoding Example}
+
+\begin{columns}[c]
+ \column{.5\textwidth}
+
+ \begin{center}
+ \begin{tabular}{c}
+
+ \begin{lstlisting}[language=C,basicstyle=\normalsize,mathescape]
+<@\textcolor{red}{x = 0}@>;
+
+while (<@\textcolor{green}{x $<$ 100}@>) {
+  <@\textcolor{blue}{x++}@>;
+}
+ \end{lstlisting}
+ \end{tabular}
+ \end{center}
+
+ \column{.5\textwidth}
+\begin{eqnarray*}
+ \textcolor{red}{P} = & \{ (x, x') \mid x' = 0 \} \\
+ \textcolor{green}{G} = & \{ x \mid x < 100 \} \\
+ \textcolor{blue}{B} = & \{ (x, x') \mid x' = x + 1 \}
+\end{eqnarray*}
+\end{columns}
+
 \end{frame}
@@
 \subsection{Ranking Functions: Proofs of Termination}
+
+\begin{frame}[fragile]{Ranking Functions}
+ To prove that a loop $(P, G, B)$ terminates, we find a \emph{ranking function} $R$ which satisfies the following criteria:
+
+ \begin{eqnarray*}
+  \forall x . G(x) \Rightarrow & R(x) > 0 \\
+  \forall x, x' . G(x) \wedge B(x, x') \Rightarrow & R(x) > R(x')
+ \end{eqnarray*}
+
+\pause
+
+\begin{exampleblock}{Example}
+\begin{center}
+\begin{columns}[c]
+ \column{.3\textwidth}
+  \begin{tabular}{c}
+   \begin{lstlisting}[language=c,basicstyle=\normalsize]
+while (x < 100) {
+  x++;
+}
+   \end{lstlisting}
+  \end{tabular}
+ \column{.3\textwidth}
+ $R(x) = 100 - x$
+\end{columns}
+\end{center}
+\end{exampleblock}
+
+\end{frame}
+
+\begin{frame}[fragile]{Conditional Ranking Functions}
+ A loop may not terminate starting in some initial state.  However, the prefix may guarantee
+ that the loop does terminate.
+
+ To prove that such a loop terminates, we must find a \emph{supporting invariant} $I$ along with the
+ ranking function $R$:
+
+ \begin{eqnarray*}
+  \forall x, x' . P(x, x') \Rightarrow & I(x') \\
+  \forall x, x' . I(x) \wedge B(x, x)' \Rightarrow & I(x') \\
+  \forall x . I(x) \wedge G(x) \Rightarrow & R(x) > 0 \\
+  \forall x, x' . I(x) \wedge G(x) \wedge B(x, x') \Rightarrow & R(x) > R(x')
+ \end{eqnarray*}
+\end{frame}
+
+\begin{frame}[fragile]{Conditional Ranking Function Example}
+\begin{exampleblock}{Example}
+\begin{center}
+\begin{columns}[T]
+ \column{.3\textwidth}
+  \begin{tabular}{c}
+   \begin{lstlisting}[language=c,basicstyle=\normalsize]
+y = 1;
+while (x < 100) {
+  x += y;
+}
+   \end{lstlisting}
+  \end{tabular}
+ \column{.3\textwidth}
+\begin{gather*}
+  I =  \{ (x, y) \mid y = 1 \} \\
+  R(x, y) =  100 - x
+\end{gather*}
+\end{columns}
+\end{center}
+\end{exampleblock}
+\end{frame}
+
+\begin{frame}{Second-Order Termination}
+
+\begin{block}{Conditional Termination Formula [CT]}
+ \[
+ \exists \red{I}, \blue{R} .  \forall x, x' . \left ( \begin{array}{rl}
+   P(x, x') \Rightarrow & \red{I}(x') \, \wedge \\
+   \red{I}(x) \wedge G(x) \Rightarrow & \blue{R}(x) > 0 \, \wedge \\
+   \red{I}(x) \wedge G(x) \wedge B(x, x') \Rightarrow & \blue{R}(x) > \blue{R}(x') \wedge \red{I}(x')
+   \end{array} \right )
+ \]
+ \end{block}
+
+ \vspace{1em}
+
+This formula is satisfiable iff there exists a ranking function $\blue{R}$ and supporting invariant $\red{I}$,
+i.e. iff the loop $(P, G, B)$ terminates.
+
+\end{frame}
 

### 2015-04-12 slides/esop-termination.tex
2nd order ranking function in esop slides

@@
  \column{.5\textwidth}
-\begin{eqnarray*}
+\begin{align*}
  \textcolor{red}{P} = & \{ (x, x') \mid x' = 0 \} \\
@@
  \textcolor{blue}{B} = & \{ (x, x') \mid x' = x + 1 \}
-\end{eqnarray*}
+\end{align*}
 \end{columns}
@@
 \begin{frame}[fragile]{Ranking Functions}
- To prove that a loop $(P, G, B)$ terminates, we find a \emph{ranking function} $R$ which satisfies the following criteria:
+ To prove that a loop $(P, G, B)$ terminates, we will try to find a \emph{ranking function} $R$.
 
- \begin{eqnarray*}
+ \vspace{1em}
+
+ A ranking function is some function that decreases on each loop iteration, but which cannot decrease forever.
+
+ \vspace{1em}
+
+ A function $R$ is a ranking function for the look $(P, G, B)$ iff it satisfies the following criteria:
+ \begin{align*}
   \forall x . G(x) \Rightarrow & R(x) > 0 \\
   \forall x, x' . G(x) \wedge B(x, x') \Rightarrow & R(x) > R(x')
- \end{eqnarray*}
+ \end{align*}
+\end{frame}
 
-\pause
-
+\begin{frame}[fragile]{Ranking Function Example}
 \begin{exampleblock}{Example}
@@
  ranking function $R$:
-
- \begin{eqnarray*}
+ \begin{align*}
   \forall x, x' . P(x, x') \Rightarrow & I(x') \\
@@
   \forall x, x' . I(x) \wedge G(x) \wedge B(x, x') \Rightarrow & R(x) > R(x')
- \end{eqnarray*}
+ \end{align*}
 \end{frame}
@@
 
-\subsection{Recurrent Sets: Proofs of Non-Termination}
+\subsection{Recurrence Sets: Proofs of Non-Termination}
+
+\begin{frame}{Recurrence Sets}
+ To prove that a loop terminates, we tried to find a ranking function.
+
+ \vspace{1em}
+
+ In order to prove that a loop does \emph{not} terminate, we will exhibit a \emph{recurrence set}.
+
+ \vspace{1em}
+\end{frame}
+
 \subsection{Using Tautologies}

### 2015-04-12 slides/esop-termination.tex
Non termination

@@
 \begin{frame}{Recurrence Sets}
- To prove that a loop terminates, we tried to find a ranking function.
+ To prove that a loop $(P, G, B)$ terminates, we tried to find a ranking function.
 
@@
 
- In order to prove that a loop does \emph{not} terminate, we will exhibit a \emph{recurrence set}.
+ In order to prove that a loop does \emph{not} terminate, we will exhibit a \emph{recurrence set} --
+ this is a non-empty set of states $S$ that, if entered, will never be left:
+ \begin{align*}
+  \exists x_0 . & S(x_0) \\
+  \forall x . S(x) \Rightarrow & G(x) \\
+  \forall x, x' . S(x) \wedge B(x, x') \Rightarrow & S(x')
+ \end{align*}
+
+\end{frame}
+
+\begin{frame}[fragile]{Recurrence Set Example}
+\begin{exampleblock}{Example}
+\begin{center}
+\begin{columns}[c]
+ \column{.3\textwidth}
+  \begin{tabular}{c}
+   \begin{lstlisting}[language=c,basicstyle=\normalsize]
+while (x > 0) {
+  x = x / 2;
+  x = x + 1;
+}
+   \end{lstlisting}
+  \end{tabular}
+ \column{.3\textwidth}
+ \begin{gather*}
+  x_0 = 2 \\
+  S = \{ 2 \}
+ \end{gather*}
+\end{columns}
+\end{center}
+\end{exampleblock}
+\end{frame}
+
+\begin{frame}[fragile]{Nondeterminism and Non-Termination}
+ If the loop body contains some nondeterminism, we only require that for each $\red{S}$-state, there is
+ \emph{some} successor that is an $\red{S}$-state, rather than requiring that \emph{every} successor is an
+ $\red{S}$-state:
+  \begin{align*}
+  \exists \green{x_0} . & \red{S}(\green{x_0}) \\
+  \forall x . \red{S}(x) \Rightarrow & G(x) \\
+  \forall x \exists x' . \red{S}(x) \wedge B(x, x') \Rightarrow & \red{S}(x')
+ \end{align*}
 
  \vspace{1em}
+
+ \pause
+
+Our syntax does not allow us to existentially quantify over ground terms inside the universal.  To
+get around this, we Skolemise the formula, introducing a successor function $\blue{N}$ which picks
+out the next state for each $\red{S}$-state.
+\end{frame}
+
+\begin{frame}{Second-Order Non-Termination}
+
+\begin{block}{Skolemised Non-Termination Formula [SNT]}
+ \[
+ \exists \green{x_0}, \red{S}, \blue{N} . \forall x . \left ( \begin{array}{rl}
+   \red{S}(\green{x_0}) \, \wedge & \\
+   \red{S}(x) \Rightarrow & G(x) \, \wedge \\
+   \red{S}(x) \Rightarrow & B(x, \blue{N}(x)) \wedge \red{S}(\blue{N}(x)) 
+   \end{array} \right )
+ \]
+ \end{block}
+
+ \vspace{1em}
+
+This formula is satisfiable iff there exists a recurrence set $\blue{S}$ containing
+an initial state $\red{x_0}$, with successor function $\green{N}$, i.e. iff the loop
+does not terminate for some input $\red{x_0}$.
 \end{frame}

### 2015-04-12 slides/esop-termination.tex
Tautologies

@@
 This formula is satisfiable iff there exists a recurrence set $\blue{S}$ containing
-an initial state $\red{x_0}$, with successor function $\green{N}$, i.e. iff the loop
+an initial state $\red{x_0}$, with successor function $\blue{N}$, i.e. iff the loop
 does not terminate for some input $\red{x_0}$.
@@
 \subsection{Using Tautologies}
+
+\begin{frame}{The Power of Tautologies}
+ Our second-order solver works by trying to synthesise programs for each of the existentially quantified functions.
+
+ \vspace{1em}
+
+ This problem is NEXPTIME-hard, but in practice is quite fast if there is some short program that satisfies
+ the constraints.
+
+ \vspace{1em}
+
+ However if the constraints are UNSAT, the solver will take a very long time to discover this.
+\end{frame}
+
+\begin{frame}{The Power of Tautologies}
+ For a given loop, the formula \red{[CT]} is satisfiable iff the loop terminates and the formula \blue{[SNT]} is
+ satisfiable iff it does not terminate.  Therefore, the formula
+ \[
+  \red{\text[CT]} \vee \blue{\text[SNT]}
+ \]
+ is satisfiable for \emph{every} loop.
+
+ \vspace{1em}
+
+ Since our solver produces witnesses to show that a formula is satisfiable, it will terminate with enough
+ information to check whether the \red{[CT]} or \blue{[SNT]} disjunct has been satisfied.
+
+ \vspace{1em}
+
+ This allows us to avoid the expensive case of having to check an UNSAT formula.
+\end{frame}
 

### 2015-04-13 slides/esop-termination.tex
Ranking function as a graph

@@
 \usepackage{pgf}
+\usepackage{dot2texi}
+
+%\usepackage[pdftex]{graphicx}
+\usepackage{graphviz}
 
@@
 \usepackage{tikz}
-\usetikzlibrary{arrows, automata, shapes.multipart, chains, positioning, fit}
+\usetikzlibrary{arrows, automata, shapes.multipart, chains, positioning, fit, calc}
 
@@
 
-\subsection{Termination Is Hard}
-
 \subsection{Bitvectors Are Hard}
@@
  ``People really don't *think* about termination using bitvector arguments''
-   \hspace*\fill{\small--- Anonymous}
+   \hspace*\fill{\small--- Anonymous Reviewer}
 \end{quote}
@@
  \end{align*}
+\end{frame}
+
+\begin{frame}[fragile]{Ranking Functions}
+
+\begin{center}
+\begin{tikzpicture}[>=latex',line join=bevel,scale=0.7,every node/.style={scale=0.7,fill=blue!20}]
+%%
+\node (10) at (101.07bp,-48.672bp) [draw,circle] {{\visible<2->{10}}};
+  \node (1) at (17.434bp,-73.642bp) [draw,circle] {{\visible<2->{1}}};
+  \node (3) at (-66.239bp,-49.026bp) [draw,circle] {{\visible<2->{3}}};
+  \node (2) at (20.808bp,-151.5bp) [draw,circle] {{\visible<2->{2}}};
+  \node (5) at (31.633bp,-3.8169bp) [draw,circle] {{\visible<2->{5}}};
+  \node (4) at (-95.572bp,28.017bp) [draw,circle] {{\visible<2->{4}}};
+  \node (7) at (37.06bp,71.959bp) [draw,circle] {{\visible<2->{7}}};
+  \node (6) at (-42.367bp,91.656bp) [draw,circle] {{\visible<2->{6}}};
+  \node (9) at (110.2bp,36.985bp) [draw,circle] {{\visible<2->{9}}};
+  \node (8) at (-114.03bp,98.041bp) [draw,circle] {{\visible<2->{8}}};
+  \draw [->] (7) ..controls (35.12bp,44.867bp) and (34.356bp,34.203bp)  .. (5);
+  \draw [->] (10) ..controls (104.22bp,-19.15bp) and (105.8bp,-4.2837bp)  .. (9);
+  \draw [->] (7) ..controls (5.6812bp,79.741bp) and (-0.59388bp,81.297bp)  .. (6);
+  \draw [->] (2) ..controls (19.62bp,-124.08bp) and (19.115bp,-112.42bp)  .. (1);
+  \draw [->] (10) ..controls (68.758bp,-58.319bp) and (60.15bp,-60.889bp)  .. (1);
+  \draw [->] (3) ..controls (-33.537bp,-58.646bp) and (-25.058bp,-61.141bp)  .. (1);
+  \draw [->] (6) ..controls (-61.612bp,68.637bp) and (-68.982bp,59.821bp)  .. (4);
+  \draw [->] (9) ..controls (82.091bp,50.426bp) and (75.265bp,53.69bp)  .. (7);
+  \draw [->] (8) ..controls (-107.32bp,72.6bp) and (-105.02bp,63.854bp)  .. (4);
+  \draw [->] (5) ..controls (26.514bp,-28.991bp) and (24.794bp,-37.448bp)  .. (1);
+  \draw [->] (4) ..controls (-85.267bp,0.9491bp) and (-80.662bp,-11.143bp)  .. (3);
+  \draw [->] (8) ..controls (-84.529bp,95.412bp) and (-82.065bp,95.193bp)  .. (6);
+%
+\end{tikzpicture}
+\end{center}
 \end{frame}

### 2015-04-13 slides/esop-termination.tex
nudge
Recurrence sets diagram

@@
 
+\begin{frame}[fragile]{Recurrence Sets}
+
+\begin{center}
+\begin{tikzpicture}[>=latex',line join=bevel,scale=0.7,every node/.style={scale=1,fill=blue!20}]
+%%
+  \node (10) at (107.78bp,-21.147bp) [draw,circle] { };
+  \node (1) at (35.474bp,-65.695bp) [draw,circle] { };
+  \node (3) at (-50.576bp,-70.065bp) [draw,circle] {};
+  \node (2) at (61.283bp,-139.93bp) [draw,circle] {};
+  \node (5) at (-1.8136bp,2.9476bp) [draw,circle] {};
+  \node (7) at (20.058bp,79.448bp) [draw,circle] {};
+  \node (9) at (97.672bp,60.723bp) [draw,circle] {};
+  \visible<2->{\node (4) at (-80.797bp,6.0893bp) [draw,circle,fill=red] {};}
+  \only<1>{\node (4a) at (-80.797bp,6.0893bp) [draw,circle] {};}
+  \node (6) at (-60.664bp,86.061bp) [draw,circle,fill=red] {};
+  \only<1>{\node (6a) at (-60.664bp,86.061bp) [draw,circle] {};}
+  \node (8) at (-128.42bp,61.564bp) [draw,circle,fill=red] {};
+  \only<1>{\node (8a) at (-128.42bp,61.564bp) [draw,circle] {};}
+  \draw [->] (4) ..controls (-98.601bp,26.831bp) and (-103.44bp,32.466bp)  .. (8);
+  \draw [->] (10) ..controls (104.25bp,7.41bp) and (102.62bp,20.665bp)  .. (9);
+  \draw [->] (7) ..controls (-12.21bp,82.092bp) and (-17.962bp,82.563bp)  .. (6);
+  \draw [->] (7) ..controls (12.354bp,52.502bp) and (9.0799bp,41.051bp)  .. (5);
+  \draw [->] (2) ..controls (52.135bp,-113.62bp) and (48.405bp,-102.89bp)  .. (1);
+  \draw [->] (10) ..controls (80.351bp,-38.045bp) and (71.978bp,-43.205bp)  .. (1);
+  \draw [->] (4) ..controls (-48.846bp,4.8184bp) and (-43.811bp,4.6181bp)  .. (5);
+  \draw [->] (3) ..controls (-16.455bp,-68.332bp) and (-8.9323bp,-67.95bp)  .. (1);
+  \draw [->] (6) ..controls (-67.773bp,57.823bp) and (-70.987bp,45.055bp)  .. (4);
+  \draw [->] (9) ..controls (66.898bp,68.148bp) and (61.149bp,69.535bp)  .. (7);
+  \draw [->] (5) ..controls (11.585bp,-21.717bp) and (16.624bp,-30.995bp)  .. (1);
+  \draw [->] (4) ..controls (-70.088bp,-20.897bp) and (-65.468bp,-32.538bp)  .. (3);
+  \draw [->] (8) ..controls (-101.04bp,71.461bp) and (-97.613bp,72.701bp)  .. (6);
+%
+\end{tikzpicture}
+\end{center}
+
+\end{frame}
+
 \begin{frame}[fragile]{Conditional Ranking Functions}

### 2015-04-13 slides/esop-termination.tex
Recurrence sets

@@
 
+\definecolor{tachameleon}{rgb}{0.54118, 0.88627, 0.20392}       % #8ae234
+\definecolor{ta2chameleon}{rgb}{0.45098, 0.82353, 0.086275}     % #73d216
+\definecolor{ta3chameleon}{rgb}{0.30588, 0.60392, 0.023529}     % #4e9a06
+
 \newcommand{\red}[1]{\textcolor{red}{#1}}
 \newcommand{\blue}[1]{\textcolor{blue}{#1}}
-\newcommand{\green}[1]{\textcolor{green}{#1}}
+\newcommand{\green}[1]{\textcolor{ta3chameleon}{#1}}
 
@@
 
+\begin{frame}[fragile]{Motivation}
+Which of these guys terminates?
+\begin{center}
+\begin{minipage}{0.45\linewidth}
+ \begin{lstlisting}[language=C,basicstyle=\normalsize]
+while (x > 0) {
+  x = x * 2;
+}
+\end{lstlisting}
+\end{minipage}
+\begin{minipage}{0.45\linewidth}
+ \begin{lstlisting}[language=C,basicstyle=\normalsize]
+while (x > 0) {
+  x = x * 3;
+}
+\end{lstlisting}
+\end{minipage}
+\end{center}
+
+\pause
+
+The left one does (the position of the leftmost 1 bit increases up to a bound of 32), but the right one doesn't (if $x$ is odd to begin with, it will be odd forever).
+\end{frame}
+
 \begin{frame}{Motivation}
@@
   \column{.15\textwidth}
-  $\exists {\textcolor{red} F} . \forall \vec{\textcolor{blue} x} . {\textcolor{green} \phi}({\textcolor{red} F}, \vec{\textcolor{blue} x})$
+  $\exists \red{F} . \forall \vec{\blue{x}} . \green{\phi}(\red{F}, \vec{\blue{x}})$
 
@@
  \begin{itemize}
-  \item[$\textcolor{red} F$] a function.
-  \item[$\textcolor{green} \phi$] a quantifier free formula.
-  \item[$\vec{\textcolor{blue} x}$] the program variables.
+  \item[$\red{F}$] a function.
+  \item[$\green{\phi}$] a quantifier free formula.
+  \item[$\vec{\blue{x}}$] the program variables.
  \end{itemize}
@@
  These constraints can be solved using an EPR solver (such as the one in Z3), or by using a program
- synthesiser to synthesise $\textcolor{red} F$.
+ synthesiser to synthesise $\red{F}$.
 \end{frame}
@@
  \begin{lstlisting}[language=C,basicstyle=\normalsize]
-<@\textcolor{red}{P}@>;
+<@\red{P}@>;
 
-while (<@\textcolor{green}{G}@>) {
-  <@\textcolor{blue}{B}@>;
+while (<@\green{G}@>) {
+  <@\blue{B}@>;
 }
@@
  \begin{itemize}
-  \item[\textcolor{red}{P}] the prefix to the loop.
-  \item[\textcolor{green}{G}] the loop guard.
-  \item[\textcolor{blue}{B}] the loop body.
+  \item[\red{P}] the prefix to the loop.
+  \item[\green{G}] the loop guard.
+  \item[\blue{B}] the loop body.
  \end{itemize}
@@
 
-In our logical encoding, we model \textcolor{green}{G} as a predicate $\textcolor{green}{G}(x)$. \textcolor{red}{P} and \textcolor{blue}{B} are modelled as transition
-relations $\textcolor{red}{P}(x, x')$ and $\textcolor{blue}{B}(x, x')$.
+In our logical encoding, we model \green{G} as a predicate $\green{G}(x)$. \red{P} and \blue{B} are modelled as transition
+relations $\red{P}(x, x')$ and $\blue{B}(x, x')$.
 
@@
  \begin{lstlisting}[language=C,basicstyle=\normalsize,mathescape]
-<@\textcolor{red}{x = 0}@>;
+<@\red{x = 0}@>;
 
-while (<@\textcolor{green}{x $<$ 100}@>) {
-  <@\textcolor{blue}{x++}@>;
+while (<@\green{x $>$ 0}@>) {
+  <@\blue{x++}@>;
 }
@@
 \begin{align*}
- \textcolor{red}{P} = & \{ (x, x') \mid x' = 0 \} \\
- \textcolor{green}{G} = & \{ x \mid x < 100 \} \\
- \textcolor{blue}{B} = & \{ (x, x') \mid x' = x + 1 \}
+ \red{P} = & \{ (x, x') \mid x' = 0 \} \\
+ \green{G} = & \{ x \mid x > 0 \} \\
+ \blue{B} = & \{ (x, x') \mid x' = x + 1 \}
 \end{align*}
@@
 
- \vspace{1em}
-
- A function $R$ is a ranking function for the look $(P, G, B)$ iff it satisfies the following criteria:
- \begin{align*}
-  \forall x . G(x) \Rightarrow & R(x) > 0 \\
-  \forall x, x' . G(x) \wedge B(x, x') \Rightarrow & R(x) > R(x')
- \end{align*}
+%  \vspace{1em}
+%  
+%  A function $R$ is a ranking function for the loop $(P, G, B)$ iff it satisfies the following criteria:
+%  \begin{align*}
+%   \forall x . G(x) \Rightarrow & R(x) > 0 \\
+%   \forall x, x' . G(x) \wedge B(x, x') \Rightarrow & R(x) > R(x')
+%  \end{align*}
 \end{frame}
@@
 
-\begin{frame}[fragile]{Ranking Function Example}
-\begin{exampleblock}{Example}
+\begin{frame}[fragile]{Ranking Function Examples}
+
 \begin{center}
@@
    \begin{lstlisting}[language=c,basicstyle=\normalsize]
-while (x < 100) {
+while (x > 0) {
   x++;
@@
  \column{.3\textwidth}
- $R(x) = 100 - x$
+ \pause
+ $R(x) = 2^{32} - 1 - x$
 \end{columns}
 \end{center}
-\end{exampleblock}
+
+\pause
+
+\begin{center}
+\begin{columns}[c]
+ \column{.3\textwidth}
+  \begin{tabular}{c}
+   \begin{lstlisting}[language=c,basicstyle=\normalsize]
+while (x > 0) {
+  x = (x - 1) & x;
+}
+   \end{lstlisting}
+  \end{tabular}
+ \column{.3\textwidth}
+ \pause
+ $R(x) = x$
+\end{columns}
+\end{center}
+
+\pause
+
+\begin{center}
+\begin{columns}[c]
+ \column{.3\textwidth}
+  \begin{tabular}{c}
+   \begin{lstlisting}[language=c,basicstyle=\normalsize]
+while (x > 0) {
+  x = x * 2;
+}
+   \end{lstlisting}
+  \end{tabular}
+ \column{.3\textwidth}
+ \pause
+ $R(x) = -x \oplus x$
+\end{columns}
+\end{center}
+
+\end{frame}
+
+% \begin{frame}[fragile]{Conditional Ranking Functions}
+%  A loop may not terminate starting in some initial state.  However, the prefix may guarantee
+%  that the loop does terminate.
+% 
+%  To prove that such a loop terminates, we must find a \emph{supporting invariant} $I$ along with the
+%  ranking function $R$:
+%  \begin{align*}
+%   \forall x, x' . P(x, x') \Rightarrow & I(x') \\
+%   \forall x, x' . I(x) \wedge B(x, x)' \Rightarrow & I(x') \\
+%   \forall x . I(x) \wedge G(x) \Rightarrow & R(x) > 0 \\
+%   \forall x, x' . I(x) \wedge G(x) \wedge B(x, x') \Rightarrow & R(x) > R(x')
+%  \end{align*}
+% \end{frame}
+% 
+% \begin{frame}[fragile]{Conditional Ranking Function Example}
+% \begin{exampleblock}{Example}
+% \begin{center}
+% \begin{columns}[T]
+%  \column{.3\textwidth}
+%   \begin{tabular}{c}
+%    \begin{lstlisting}[language=c,basicstyle=\normalsize]
+% y = 1;
+% while (x < 100) {
+%   x += y;
+% }
+%    \end{lstlisting}
+%   \end{tabular}
+%  \column{.3\textwidth}
+% \begin{gather*}
+%   I =  \{ (x, y) \mid y = 1 \} \\
+%   R(x, y) =  100 - x
+% \end{gather*}
+% \end{columns}
+% \end{center}
+% \end{exampleblock}
+% \end{frame}
+
+\begin{frame}{Second-Order Termination}
+
+\begin{block}{Conditional Termination Formula [CT]}
+ \[
+ \exists \red{I}, \blue{R} .  \forall x, x' . \left ( \begin{array}{rl}
+   P(x, x') \Rightarrow & \red{I}(x') \, \wedge \\
+   \red{I}(x) \wedge G(x) \Rightarrow & \blue{R}(x) > 0 \, \wedge \\
+   \red{I}(x) \wedge G(x) \wedge B(x, x') \Rightarrow & \blue{R}(x) > \blue{R}(x') \wedge \red{I}(x')
+   \end{array} \right )
+ \]
+ \end{block}
+
+ \vspace{1em}
+
+This formula is satisfiable iff there exists a ranking function $\blue{R}$ and supporting invariant $\red{I}$,
+i.e. iff the loop $(P, G, B)$ terminates.
+
+\end{frame}
+
+\subsection{Recurrence Sets: Proofs of Non-Termination}
+
+\begin{frame}{Recurrence Sets}
+ To prove that a loop $(P, G, B)$ terminates, we tried to find a ranking function.
+
+ \vspace{1em}
+
+ In order to prove that a loop does \emph{not} terminate, we will exhibit a \emph{recurrence set}\footnotemark --
+ this is a non-empty set of states $S$ that, if entered, will never be left.
+
+ \footnotetext{Gupta et al. -- \emph{Proving Nontermination}, POPL '08 \vspace{1em} }
 
@@
 
-\begin{frame}[fragile]{Conditional Ranking Functions}
- A loop may not terminate starting in some initial state.  However, the prefix may guarantee
- that the loop does terminate.
-
- To prove that such a loop terminates, we must find a \emph{supporting invariant} $I$ along with the
- ranking function $R$:
- \begin{align*}
-  \forall x, x' . P(x, x') \Rightarrow & I(x') \\
-  \forall x, x' . I(x) \wedge B(x, x)' \Rightarrow & I(x') \\
-  \forall x . I(x) \wedge G(x) \Rightarrow & R(x) > 0 \\
-  \forall x, x' . I(x) \wedge G(x) \wedge B(x, x') \Rightarrow & R(x) > R(x')
- \end{align*}
-\end{frame}
-
-\begin{frame}[fragile]{Conditional Ranking Function Example}
-\begin{exampleblock}{Example}
-\begin{center}
-\begin{columns}[T]
- \column{.3\textwidth}
-  \begin{tabular}{c}
-   \begin{lstlisting}[language=c,basicstyle=\normalsize]
-y = 1;
-while (x < 100) {
-  x += y;
-}
-   \end{lstlisting}
-  \end{tabular}
- \column{.3\textwidth}
-\begin{gather*}
-  I =  \{ (x, y) \mid y = 1 \} \\
-  R(x, y) =  100 - x
-\end{gather*}
-\end{columns}
-\end{center}
-\end{exampleblock}
-\end{frame}
-
-\begin{frame}{Second-Order Termination}
-
-\begin{block}{Conditional Termination Formula [CT]}
- \[
- \exists \red{I}, \blue{R} .  \forall x, x' . \left ( \begin{array}{rl}
-   P(x, x') \Rightarrow & \red{I}(x') \, \wedge \\
-   \red{I}(x) \wedge G(x) \Rightarrow & \blue{R}(x) > 0 \, \wedge \\
-   \red{I}(x) \wedge G(x) \wedge B(x, x') \Rightarrow & \blue{R}(x) > \blue{R}(x') \wedge \red{I}(x')
-   \end{array} \right )
- \]
- \end{block}
-
- \vspace{1em}
-
-This formula is satisfiable iff there exists a ranking function $\blue{R}$ and supporting invariant $\red{I}$,
-i.e. iff the loop $(P, G, B)$ terminates.
-
-\end{frame}
-
-\subsection{Recurrence Sets: Proofs of Non-Termination}
-
-\begin{frame}{Recurrence Sets}
- To prove that a loop $(P, G, B)$ terminates, we tried to find a ranking function.
-
- \vspace{1em}
-
- In order to prove that a loop does \emph{not} terminate, we will exhibit a \emph{recurrence set} --
- this is a non-empty set of states $S$ that, if entered, will never be left:
- \begin{align*}
-  \exists x_0 . & S(x_0) \\
-  \forall x . S(x) \Rightarrow & G(x) \\
-  \forall x, x' . S(x) \wedge B(x, x') \Rightarrow & S(x')
- \end{align*}
-
-\end{frame}
-
 \begin{frame}[fragile]{Recurrence Set Example}
-\begin{exampleblock}{Example}
 \begin{center}
@@
  \column{.3\textwidth}
+ \pause
  \begin{gather*}
@@
 \end{center}
-\end{exampleblock}
-\end{frame}
-
-\begin{frame}[fragile]{Nondeterminism and Non-Termination}
- If the loop body contains some nondeterminism, we only require that for each $\red{S}$-state, there is
- \emph{some} successor that is an $\red{S}$-state, rather than requiring that \emph{every} successor is an
- $\red{S}$-state:
-  \begin{align*}
-  \exists \green{x_0} . & \red{S}(\green{x_0}) \\
-  \forall x . \red{S}(x) \Rightarrow & G(x) \\
-  \forall x \exists x' . \red{S}(x) \wedge B(x, x') \Rightarrow & \red{S}(x')
- \end{align*}
-
- \vspace{1em}
 
@@
 
-Our syntax does not allow us to existentially quantify over ground terms inside the universal.  To
-get around this, we Skolemise the formula, introducing a successor function $\blue{N}$ which picks
-out the next state for each $\red{S}$-state.
+\begin{center}
+\begin{columns}[c]
+ \column{.3\textwidth}
+  \begin{tabular}{c}
+   \begin{lstlisting}[language=c,basicstyle=\normalsize]
+while (x > 0) {
+  x = x * 3;
+}
+   \end{lstlisting}
+  \end{tabular}
+ \column{.3\textwidth}
+ \pause
+ \begin{gather*}
+  x_0 = 1 \\
+  S = \{ x \mid (x \text{ mod } 2) = 1 \}
+ \end{gather*}
+\end{columns}
+\end{center}
+
 \end{frame}
+% 
+% \begin{frame}[fragile]{Nondeterminism and Non-Termination}
+%  If the loop body contains some nondeterminism, we only require that for each $\red{S}$-state, there is
+%  \emph{some} successor that is an $\red{S}$-state, rather than requiring that \emph{every} successor is an
+%  $\red{S}$-state:
+%   \begin{align*}
+%   \exists \green{x_0} . & \red{S}(\green{x_0}) \\
+%   \forall x . \red{S}(x) \Rightarrow & G(x) \\
+%   \forall x \exists x' . \red{S}(x) \wedge B(x, x') \Rightarrow & \red{S}(x')
+%  \end{align*}
+% 
+%  \vspace{1em}
+%  
+%  \pause
+%  
+% Our syntax does not allow us to existentially quantify over ground terms inside the universal.  To
+% get around this, we Skolemise the formula, introducing a successor function $\blue{N}$ which picks
+% out the next state for each $\red{S}$-state.
+% \end{frame}
 

### 2015-04-13 slides/esop-termination.tex
Citation

@@
 %\subtitle{}
-\institute{Oxford University}
+\institute{University of Oxford}
 \date{\today}
@@
 
-\begin{frame}{Example}
-\end{frame}
-
 \section{Our Solution}
@@
 
- These constraints can be solved using an EPR solver (such as the one in Z3), or by using a program
- synthesiser to synthesise $\red{F}$.
+ We solve these constraints by using a program synthesiser to synthesise $\red{F}$\footnotemark.
+
+ \footnotetext{David, Kroening, Lewis -- Second-Order Propositional Satisfiability, arXiv \vspace{1em}}
 \end{frame}
@@
 
-\subsection{Using Tautologies}
-
-\begin{frame}{The Power of Tautologies}
- Our second-order solver works by trying to synthesise programs for each of the existentially quantified functions.
-
- \vspace{1em}
-
- This problem is NEXPTIME-hard, but in practice is quite fast if there is some short program that satisfies
- the constraints.
-
- \vspace{1em}
-
- However if the constraints are UNSAT, the solver will take a very long time to discover this.
-\end{frame}
-
-\begin{frame}{The Power of Tautologies}
- For a given loop, the formula \red{[CT]} is satisfiable iff the loop terminates and the formula \blue{[SNT]} is
- satisfiable iff it does not terminate.  Therefore, the formula
- \[
-  \red{\text[CT]} \vee \blue{\text[SNT]}
- \]
- is satisfiable for \emph{every} loop.
-
- \vspace{1em}
-
- Since our solver produces witnesses to show that a formula is satisfiable, it will terminate with enough
- information to check whether the \red{[CT]} or \blue{[SNT]} disjunct has been satisfied.
-
- \vspace{1em}
-
- This allows us to avoid the expensive case of having to check an UNSAT formula.
-\end{frame}
+% \subsection{Using Tautologies}
+% 
+% \begin{frame}{The Power of Tautologies}
+%  Our second-order solver works by trying to synthesise programs for each of the existentially quantified functions.
+% 
+%  \vspace{1em}
+% 
+%  This problem is NEXPTIME-hard, but in practice is quite fast if there is some short program that satisfies
+%  the constraints.
+% 
+%  \vspace{1em}
+% 
+%  However if the constraints are UNSAT, the solver will take a very long time to discover this.
+% \end{frame}
+% 
+% \begin{frame}{The Power of Tautologies}
+%  For a given loop, the formula \red{[CT]} is satisfiable iff the loop terminates and the formula \blue{[SNT]} is
+%  satisfiable iff it does not terminate.  Therefore, the formula
+%  \[
+%   \red{\text[CT]} \vee \blue{\text[SNT]}
+%  \]
+%  is satisfiable for \emph{every} loop.
+% 
+%  \vspace{1em}
+% 
+%  Since our solver produces witnesses to show that a formula is satisfiable, it will terminate with enough
+%  information to check whether the \red{[CT]} or \blue{[SNT]} disjunct has been satisfied.
+% 
+%  \vspace{1em}
+% 
+%  This allows us to avoid the expensive case of having to check an UNSAT formula.
+% \end{frame}
 
 \section{Experiments}
+
+\begin{frame}{Experiments}
+
+\end{frame}
 

### 2015-04-13 slides/esop-termination.tex
Experiments

@@
 \begin{frame}{Experiments}
+We implemented the {\sc Juggernaut} tool, which is slow but unstoppable.
+
+\vspace{1em}
+
+We compared to {\sc ARMC} on a set of 47 benchmarks taken from SVCOMP and our own
+regression tests.
+
+\begin{center}
+ \begin{tabular}{|l||l|l|}
+  \hline
+  & {\sc ARMC} & {\sc Juggernaut} \\
+  \hline
+  \hline
+  Correct & 28 & 35 \\
+  Incorrect for bitvectors & 3 & 0 \\
+  Unknown & 13 & 0 \\
+  Timeout & 3 & 12 \\
+  \hline
+  Median time (s) & 0.09 & 2.2 \\
+  \hline
+ \end{tabular}
+\end{center}
 

### 2015-04-13 slides/esop-termination.tex
Introduction

@@
 
-\section{What's The Problem?}
+\section{What's Our Problem?}
 \begin{frame}[fragile]{Motivation}
@@
 
-\begin{frame}[fragile]{Motivation}
-Which of these guys terminates?
-\begin{center}
-\begin{minipage}{0.45\linewidth}
- \begin{lstlisting}[language=C,basicstyle=\normalsize]
-while (x > 0) {
-  x = x * 2;
-}
-\end{lstlisting}
-\end{minipage}
-\begin{minipage}{0.45\linewidth}
- \begin{lstlisting}[language=C,basicstyle=\normalsize]
-while (x > 0) {
-  x = x * 3;
-}
-\end{lstlisting}
-\end{minipage}
-\end{center}
-
-\pause
-
-The left one does (the position of the leftmost 1 bit increases up to a bound of 32), but the right one doesn't (if $x$ is odd to begin with, it will be odd forever).
-\end{frame}
+% \begin{frame}[fragile]{Motivation}
+% How about these?
+% \begin{center}
+% \begin{minipage}{0.45\linewidth}
+%  \begin{lstlisting}[language=C,basicstyle=\normalsize]
+% while (x > 0) {
+%   x = x * 2;
+% }
+% \end{lstlisting}
+% \end{minipage}
+% \begin{minipage}{0.45\linewidth}
+%  \begin{lstlisting}[language=C,basicstyle=\normalsize]
+% while (x > 0) {
+%   x = x * 3;
+% }
+% \end{lstlisting}
+% \end{minipage}
+% \end{center}
+% 
+% \pause
+% 
+% The left one does (the position of the leftmost 1 bit increases up to a bound of 32), but the right one doesn't (if $x$ is odd to begin with, it will be odd forever).
+% \end{frame}
 
@@
 
-It's still quite hard to do for the programs that people write, which use bitvectors and all kinds of nonsense.
+It's still quite hard to do for the programs that people write, which use bitvectors and nonlinear operators and all kinds of nonsense.
+
+\vspace{1em}
+
+\begin{block}{Our Goal}
+ We would like to come up with a fully automatic procedure for proving termination or non-termination of
+ arbitrary programs that make use of bitvector and floating point arithmetic.
+\end{block}
+
+% \vspace{1.5em}
+% 
+% \begin{quote}
+%  ``People really don't *think* about termination using bitvector arguments''
+%    \hspace*\fill{\small--- Anonymous Reviewer}
+%    \end{quote}
 \end{frame}
@@
 
-\begin{frame}{Bitvector Termination}
-\begin{quote}
- ``People really don't *think* about termination using bitvector arguments''
-   \hspace*\fill{\small--- Anonymous Reviewer}
-\end{quote}
-\end{frame}
-
-\subsection{Non-Linearity Is Hard}
+% 
+% \begin{frame}{Bitvector Termination}
+% \begin{quote}
+%  ``People really don't *think* about termination using bitvector arguments''
+%    \hspace*\fill{\small--- Anonymous Reviewer}
+% \end{quote}
+% \end{frame}
 
@@
 
-\begin{frame}[fragile]{Our Theory}
+\begin{frame}[fragile]{Our Solution}
  We will try to build termination proofs by solving \emph{second-order constraints}.
@@
 
-\subsection{Ranking Functions: Proofs of Termination}
+\subsection{Proofs of Termination}
 
@@
 
-\subsection{Recurrence Sets: Proofs of Non-Termination}
+\subsection{Proofs of Non-Termination}
 
@@
 
+\begin{frame}[fragile]{Oh, And One More Thing}
+Does this terminate?  If so, what's the ranking function?
+
+\begin{columns}[c]
+\column{.4\textwidth}
+\begin{lstlisting}[basicstyle=\normalsize]
+while (x > 0 && y > 0 && z > 0) {
+  if (y > x) {
+    y = z;
+    x = nondet();
+    z = x - 1;
+  } else {
+    z = z - 1;
+    x = nondet();
+    y = x - 1;
+  }
+}
+\end{lstlisting}
+
+\pause
+\column{.45\textwidth}
+$\red{R(x, y, z) = (x < y, z)}$
+
+\end{columns}
+\end{frame}
+
 \begin{frame}[fragile]{Fin}

### 2015-04-14 slides/esop-termination.tex
Final slides

@@
 
+\usepackage{amssymb}
+
 \pgfdeclareimage[height=6ex]{ou-logo}{images/ou}
@@
 \section{What's Our Problem?}
-\begin{frame}[fragile]{Motivation}
+\subsection{Examples}
+
+\begin{frame}[fragile]{Example}
 Does this program terminate?
@@
 
-\begin{frame}[fragile]{Motivation}
+\begin{frame}[fragile]{Example}
 Which of these guys terminates?
@@
 
-% \begin{frame}[fragile]{Motivation}
+% \begin{frame}[fragile]{Example}
 % How about these?
@@
 
+\begin{frame}{Our Contributions}
+ \begin{itemize}
+  \item We present a sound and complete algorithm for finding \emph{non-linear, lexicographic} ranking functions
+   and \emph{non-linear} recurrence sets for bitvector and floating point programs.
+  \item We implemented the algorithm on top of our second order solver.
+ \end{itemize}
+
+\end{frame}
+
+\subsection{Current Approaches}
+
+\begin{frame}{Current Approaches}
+ Most termination provers make some assumptions, e.g.:
+
+ \vspace{1em}
+
+ \begin{itemize}
+ \item Program variables are \emph{mathematical integers}.
+ \item Programs only use \emph{linear operations}.
+ \item Most terminating programs have \emph{linear ranking functions}.
+ \end{itemize}
+\end{frame}
+
 \subsection{Bitvectors Are Hard}
+
+\begin{frame}{Bitvector Semantics}
+ Programs use fixed width bitvectors for arithmetic.  Arithmetic on a $k-$bit bitvector is performed modulo $2^k$.
+
+ \vspace{1em}
+
+ \begin{tabular}{r|ccc}
+  & Integers & Bitvectors & Example \\
+  \hline
+  $x + 1 > x$ & \green{\checkmark} & \red{\xmark} & $(2^{32} - 1) + 1 = 0$ \\
+  $x \neq 0 \wedge y \neq 0 \Rightarrow xy \neq 0$ & \green{\checkmark} & \red{\xmark} & $2^31 \times 2 = 0$ \\
+ \end{tabular}
+
+\end{frame}
 
@@
 
+ \pause
+
  \vspace{1em}
@@
 
- \footnotetext{David, Kroening, Lewis -- Second-Order Propositional Satisfiability, arXiv \vspace{1em}}
+ \only<2->{\footnotetext{David, Kroening, Lewis -- Second-Order Propositional Satisfiability, arXiv \vspace{1em}}}
 \end{frame}
@@
 
+\begin{frame}{Conclusion}
+\Large
+
+  Bitvector termination is hard and confusing!
+
+  \vspace{2em}
+
+  Specifying it using second-order logic makes it easier.
+
+  \vspace{2em}
+
+  Second-order solvers exist!
+
+\end{frame}
+
 \begin{frame}[fragile]{Fin}
